### 2021-12-18 RefactoringNLS/RefactoringNLS.Rmd
populate repo with Refactoring and TestsDoc

---
title: "Refactoring the `nls()` function in R"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
date: "2021-8-20"
output: 
    pdf_document:
        keep_tex: false
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
link-citations: yes
linkcolor: red
urlcolor: blue
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## require(bookdown) # language engine to display text - does not seem necessary
```

# Abstract

This article reports the particular activities of our Google Summer of Code
project "Improvements to `nls()`" that relate to the R code for that function, which
is intended for the estimation of models written as a formula that has at least
one parameter that is "nonlinear", that is, not estimable via solving a set of 
linear equations. A companion document **Variety in Nonlinear Least Squares Codes** 
presents an overview of methods for the problem which takes a much wider view of the
problem of minimizing a function that can be written as a sum of squared
terms.

Our work has not fully addressed all the issues that we would like to see
resolved, but we believe we have made sufficient progress to demonstrate that
there are worthwhile improvements that can be made to the R function `nls()`.
An important overall consideration in our work has been the maintainability of
the code base that supports the `nls()` functionality, as we believe that the
existing code makes maintenance and improvement very difficult.

# The `nls()` function: strengths and shortcomings

`nls()` is the tool in base R (the distributed software package from [CRAN](https://cran.r-project.org)
for estimating nonlinear statistical models. The function was developed mainly in the 1980s and
1990s by Doug Bates et al., initially for S (see https://en.wikipedia.org/wiki/S_%28programming_language%29) . 
The ideas spring primarily from the book by @bateswatts.

The `nls()` function has a remarkable and quite comprehensive set of capabilities for estimating
nonlinear models that are expressed as formulas. In particular, we note that it

- handles formulas that include R functions
- allows data to be subset
- permits parameters to be indexed over a set of related data
- produces measures of variability (i.e., standard error estimates) for the estimated parameters
- has related profiling capabilities for exploring the likelihood surface as parameters are changed

With such a range of features and a long history, it is not surprising that the code has become untidy 
and overly patched. It is, to
our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
us review some of the issues. We will then propose corrective actions, some of which we have carried out.

## Feature: Convergence and termination tests (FIXED)

A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
(i.e., perfect fit) data was corrected thanks to suggestions by one of us. The release 
notes for R 4.1.0 state:

 - For nls(), an optional constant scaleOffset may be added to the denominator of the relative offset
   convergence test for cases where the fit of a model is expected to be exact, thanks to a proposal by 
   John Nash. nls(*, trace=TRUE) now also shows the convergence criterion.

Within the previous standard documentation (**man** or ".Rd" file) `nls()` warned

>  **The default settings of nls generally fail on artificial “zero-residual” data problems.**

>  The nls function uses a relative-offset convergence criterion that compares the numerical imprecision at
  the current parameter estimates to the residual sum-of-squares. This performs well on data of the form $$y = f(x, \theta) + eps$$

>  (with var(eps) > 0). It fails to indicate convergence on data of the form $$y = f(x, \theta)$$

>  because the criterion amounts to comparing two components of the round-off error. To avoid a zero-divide in
  computing the convergence testing value, a positive constant `scaleOffset` should be added to the denominator
  sum-of-squares; it is set in control; this does not yet apply to algorithm = "port".

It turns out that this issue can be quite easily resolved. The key "convergence test" -- more properly
a "termination test" for the **program** rather than testing for convergence of the underlying **algorithm** --
is the Relative Offset Convergence Criterion (see @BatesWatts81). This works by projecting the proposed
step in the parameter vector on the gradient and estimating how much the sum of squares loss function 
will decrease. To avoid scale issues, we use the current size of the loss function as a measure and divide
by it. When we have "converged", the estimated
decrease is very small, as usually is its ratio to the sum of squares. However, in some cases we have the 
possibility of an exact fit and the sum of squares is (almost) zero and we get the possibility of a 
zero-divide failure.

The issue is easily resolved by adding a small quantity to the loss function. To preserve legacy behaviour,
in 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
with a default value of zero for legacy behaviour. Setting it to a small number -- 1.0 is 
a reasonable choice --
allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
**safeguarded relative offset convergence criterion**.

As noted, we are pleased to report that this improvement is now in the R distributed code as of version 4.1.0.

### Example of a small-residual problem

```
rm(list=ls())
t <- -10:10
y <- 100/(1+.1*exp(-0.51*t))
lform<-y~a/(1+b*exp(-c*t))
ldata<-data.frame(t=t, y=y)
plot(t,y)
lstartbad<-c(a=1, b=1, c=1)
lstart2<-c(a=100, b=10, c=1)
nlsr::nlxb(lform, data=ldata, start=lstart2)
nls(lform, data=ldata, start=lstart2, trace=TRUE)
# Fix with scaleOffset
nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
sessionInfo()
```

Edited output of running this function follows:

```
> rm(list=ls())
> t <- -10:10
> y <- 100/(1+.1*exp(-0.51*t))
> lform<-y~a/(1+b*exp(-c*t))
> ldata<-data.frame(t=t, y=y)
> plot(t,y)
> lstart2<-c(a=100, b=10, c=1)
> nlsr::nlxb(lform, data=ldata, start=lstart2)
nlsr object: x 
residual sumsquares =  1.007e-19  on  21 observations
    after  13    Jacobian and  19 function evaluations
  name            coeff          SE       tstat      pval      gradient    JSingval   
a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6  
b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3  
c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
# Note that this has succeeded. The test in nlsr recognizes small residual problems.
> nls(lform, data=ldata, start=lstart2, trace=TRUE)
40346.    (1.08e+00): par = (100 10 1)
11622.    (2.93e+00): par = (101.47 0.49449 0.71685)
5638.0    (1.08e+01): par = (102.23 0.38062 0.52792)
642.08    (1.04e+01): par = (102.16 0.22422 0.41935)
97.712    (1.79e+01): par = (100.7 0.14774 0.45239)
22.250    (1.78e+02): par = (99.803 0.093868 0.50492)
0.025789  (1.33e+03): par = (100.01 0.10017 0.50916)
6.0571e-08 (7.96e+05): par = (100 0.1 0.51)
4.7017e-19 (1.86e+04): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) : 
  number of iterations exceeded maximum of 50

> nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
40346.      (1.08e+00): par = (100 10 1)
11622.      (2.91e+00): par = (101.47 0.49449 0.71685)
5638.0      (9.23e+00): par = (102.23 0.38062 0.52792)
642.08      (5.17e+00): par = (102.16 0.22422 0.41935)
97.712      (2.31e+00): par = (100.7 0.14774 0.45239)
22.250      (1.11e+00): par = (99.803 0.093868 0.50492)
0.025789    (3.79e-02): par = (100.01 0.10017 0.50916)
6.0571e-08  (5.80e-05): par = (100 0.1 0.51)
4.7017e-19  (1.62e-10): par = (100 0.1 0.51)
Nonlinear regression model
  model: y ~ a/(1 + b * exp(-c * t))
   data: ldata
     a      b      c 
100.00   0.10   0.51 
 residual sum-of-squares: 4.7e-19

Number of iterations to convergence: 8 
Achieved convergence tolerance: 1.62e-10

> sessionInfo()
R version 4.1.0 (2021-05-18)
Platform: x86_64-pc-linux-gnu (64-bit)
Running under: Linux Mint 20.2
```

### More general termination tests

The single convergence criterion of `nls()` leaves out some possibilities that could
be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
the safeguarded relative offset test (**roffset**) as well as a **small sum of 
squares** test (**smallsstest**) that compares the latest evaluated sum of squared
(weighted) residuals to a very small multiple of the initial sum of squares. The
multiple uses a control setting `offset` which defaults to 100.0 and we compute
the 4th power of the machine epsilon times this offset.

```{r meps4, echo=TRUE}
epstol<-100*.Machine$double.eps
e4 <- epstol^4
e4
```

We do note that `nls()` stops after `maxiter` "iterations". However, for almost all
iterative algorithms, the meaning of "iteration" requires careful examination of the
code. Instead, we prefer to record the number of times the residuals or the jacobian have
been computed and put upper limits on these. Our codes exit (terminate) when these limits
are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
but that may simply reflect our history of dealing with more difficult problems as we are
the tool-makers; users consult us when things go wrong.

## Feature: Failure when Jacobian is computationally singular

This is the infamous "singular gradient" termination message of `nls()`. A Google search of

```
R nls "singular gradient"
```

gets over 4000 hits that are spread over the years. 
In some cases this is due
to the failure of the simple finite difference approximation of the Jacobian in the
`numericDeriv()` function that is a part of `nls()`. `nlsr` can use analytic
derivatives, and we can import this functionality to the `nls()` code as an
improvement. See below in the section **Jacobian computation**.

However, the more common source of the issue is that the Jacobian is very close
to singular for some values of the model parameters. In such cases we need to find
an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
@jn77ima). Versions of this have been implemented in packages `minpack.lm` and `nlsr`.
and we have prepared very preliminary versions of an `nls` replacement that can incorporate a
version of the Levenberg-Marquardt stabilization. (There are some issues of 
integration with other code structures and of complexity of the computations that
suggest we should use a simplified LM stabilization.) Unfortunately, making this code
work across all of the features of `nls()` has not proved straightforward.

<!-- ?? JN: I forget exact details, but I recall that we could not get this to work with -->
<!-- some ways in which call could be set up. May need to revisit. this. -->

## Feature: Jacobian computation

`nls()`, with the `numericDeriv()` function, computes the Jacobian as the "gradient"
attribute of the residual vector. This is implemented as a mix of R and C code, but
we have created a rather more compact version entirely in R in this Google Summer of
Code project. See the document **DerivsNLS.pdf**.

Let us consider an example which presents the infamous error message.

```{r code=xfun::read_utf8('SESingGrad.R')}
```

It should be noted that the selfStart models in the `./src/library/stats/R/zzModels.R` file
provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
is within the `nlsModel()` or `nlsModel.pliner()` functions.

## Feature: Subsetting

`nls()` accepts an argument `subset`. Unfortunately, this acts through the mediation of
`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
`/src/library/stats/src/nls.C`.

While the implementation of subset at the level of the call to `nls()` has a certain
attractiveness, it does mean that the programmer of the solver needs to be aware of the
source (and value) of objects such as the data, residuals and Jacobian. By preference, 
we would implement subsetting by means of zero-value weights, with observation counts
(and degrees of freedom) computed via the numbers of non-zero weights.

## Feature: na.action

`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
source code, often being handled behind the scenes after referencing the option `na.action`.
A useful, but possibly dated, description is given in:
https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/

The typical default action, which can be seen by using the command `getOption("na.action")`
is `na.omit`. This option essentially presents computations with data with all observations
containing any missing values (i.e. any row of a data frame with an NA) omitted. 
`na.exclude` does much the same for computations, but keeps the rows with NA elements so 
that predictions are in the correct row position. We recommend that workers actually test 
output to verify behaviour is as wanted.

A succinct description of the issue is given in: https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r

> The only benefit of `na.exclude` over `na.omit` is that the former will retain the 
original number of rows in the data. This may be useful where you need to retain the 
original size of the dataset - for example it is useful when you want to compare 
predicted values to original values. With `na.omit` you will end up with fewer 
rows so you won't as easily be able to compare.

`na.pass` simply passes on data "as is", while `na.fail` will essentially stop if any missing
values are present.

## Feature: model frame

`model` is an argument to the `nls()` function, which is documented

> **model** logical. If true, the model frame is returned as part of the object. Default is FALSE.

Indeed, the argument only gets used when `nls()` is about to return its result object, and the
element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code in the form of an object
`mf`.

## Feature: sources of data

`nls()` can be called without specifying the `data` argument. In this case, it will
search in the available environments (i.e., workspaces) for suitable data objects. 
We do NOT like this approach, but it is "the R way". R allows users to leave many 
objects in the default (.GlobalEnv) workspace. Moreover, users have to actively 
suppress saving this workspace (`.RData`) on exit, and any such file in the path 
when R is launched will be loaded. The overwhelming proportion of R users in our
acquaintance avoid saving the workspace because of the danger of lurking data and
functions which may cause unwanted results.

Nevertheless, to provide compatible behaviour with `nls()`, we will need to ensure
that equivalent behaviour is guaranteed. Furthermore, we need to test that the 
operation is correct.

## Feature: missing start vector and self-starting models

Nonlinear estimation algorithms are almost all iterative and need a set of starting
parameters. `nls()` offers a special class of modeling functions called **selfStart** 
models. There are a number of these in base R (see list below) and others in R packages
such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
`NRAIA`. Unfortunately, the structure of the programming of 
these is such that the methods by which initial parameters are computed is entangled
with the particularities of the `nls()` code. Though there is a `getInitial()` function,
this is not easy to use to simply compute the initial parameter estimates, in part 
because it may call `nls()`. Such circular references are, in our view, dangerous.

In the example below, we show how the `SSlogis` selfStart function can generate a set
of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$

The example shows how these starting parameters can be transformed to those of 
another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$

Let us look at the actual code for `SSlogis()` in `R-devel/src/library/stats/R/zzModels.R`:

```
SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)),
    selfStart(
        function(input, Asym, xmid, scal)
        {
              .expr1 <- xmid - input
              .expr3 <- exp(.e2 <- .expr1/scal)
              .expr4 <- 1 + .expr3
              .value <- Asym/.expr4
              .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")])
              if(all(vapply(.actualArgs, is.name, NA)))
              {
            		  .expr10 <- .expr4^2
                  .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal")))
                  .grad[, "Asym"] <- 1/.expr4
		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10)
		              .grad[, "scal"] <- xm * .e2
                  dimnames(.grad) <- list(NULL, .actualArgs)
                  attr(.value, "gradient") <- .grad
              }
              .value
        },
        initial = function(mCall, data, LHS, ...) {
              xy <- sortedXyData(mCall[["input"]], LHS, data)
              if(nrow(xy) < 4) {
                  stop("too few distinct input values to fit a logistic model")
              }
              z <- xy[["y"]]
              ## transform to proportion, i.e. in (0,1) :
              rng <- range(z); dz <- diff(rng)
              z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz)
              xy[["z"]] <- log(z/(1 - z))		# logit transformation
              aux <- coef(lm(x ~ z, xy))
              pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)),
                               data = xy,
                               start = list(xmid = aux[[1L]], scal = aux[[2L]]),
                               algorithm = "plinear", ...))
              setNames(pars [c(".lin", "xmid", "scal")],
                       mCall[c("Asym", "xmid", "scal")])
        },
        parameters = c("Asym", "xmid", "scal"))
```

We note that the function includes analytic expressions for the Jacobian ("gradient").
These could be possibly be useful to R users, especially if documented. Moreover, we
wonder why the programmers have chosen to save so many quantities in "hidden"
variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
command, making them difficult to access.

In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
to 1. This is, in our view, tolerable, but could possibly be improved by using a set of values
that are slightly different e.g., in the case of a model $$  y \sim a*exp(-b*x) + c*exp(-d*x)$$
it would be useful to have $b$ and $d$ values different so the Jacobian is not singular. Thus,
some sort of sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters might be better and it
can be provided quite simply instead of all 1's.

```{r hobbsSSlogis}
weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
          38.558, 50.156, 62.948, 75.995, 91.972)
tt <- 1:12

NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt))
NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal)
NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal))
NLStestdata <- data.frame(y=weed, tt=tt) 
s0 <- getInitial(NLSformula, NLStestdata)
print(s0)
# We transform the parameters for the NLSformula0 model of original specification.
s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3])
print(as.numeric(s1))
# No actual improvement because nls() has been already used to get the starting values,
# but we do get SEs
hobblog<-nls(NLSformula0, data=NLStestdata, start=s1)
summary(hobblog)
deviance(hobblog)
# nls fails without selfStart -- singular gradient, even on stabilized formula
try(hobblogx<-nls(NLSformulax, data=NLStestdata))
# But Marquardt is able to get a solution easily
library(nlsr)
hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1))
hobblogxx
```

### selfStart models in base R

The following models are provided (in file ./src/library/stats/R/zzModels.R)

```
SSasymp         - asymptotic regression model
SSasympOff      - alternate formulation of asymptotic regression model with offset
SSasympOrig     - exponential curve through the origin to an asymptote
SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input)
SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) -
                          exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe)))
SSfpl           - four parameter logistic model
SSlogis         - three parameter logistic model
SSmicmen        - Michaelis-Menten model for enzyme kinetics
SSgompertz2     - Gompertz model for growth curve data
SSweibull       - Weibull model for growth curve data
```

### Strategic issues in selfStart models

Because the Gauss-Newton algorithm is rather unreliable from many starting sets
of parameters, selfStart models are more than an accessory to `nls()` but a part
of the infrastructure. However, creating such functions is a lot of work, and their
documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
We believe that the focus would better be placed on getting good initial parameters,
possibly with some interactive tools. That is, the emphasis should be on the 
`getInitial()` function, though avoiding the current calls back to `nls()`.

## Issue: documentation of the results of running nls()

The output of `nls()` is an object of class "nls" which has the following structure:

### nls() result output according to the documentation

```
A list of:

m	        an nlsModel object incorporating the model.
data  	    the expression that was passed to nls as the data argument. The actual data values 
              are present in the environment of the m components, e.g., environment(m$conv).
call	    the matched call with several components, notably algorithm.
na.action 	the "na.action" attribute (if any) of the model frame.
dataClasses	the "dataClasses" attribute (if any) of the "terms" attribute of the model frame.
model	    if model = TRUE, the model frame.
weights	    if weights is supplied, the weights.
convInfo	a list with convergence information.
control	    the control list used, see the control argument.

There are also two deprecated items if algorithm = "port" fit only. These are
convergence (a code = 0 for convergence) and message. These are available from convInfo.
```

### Example output

To illustrate, let us run the Croucher example.

```{r nlsoutx}
# Croucher example
xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2)
NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
NLSdata<-data.frame(xdata, ydata)
# Try full output version of nls
library(nlspkg) # use the packaged version of nls()
result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE)
# str(result) -- displays large amount of material - suppressed here
#    as it is too wide for the page
result
ls(result) # to list the elements of the output
ls(result$m) # and in particular the "m" object
```

### Concerns with content of the nls result object

The nls object contains some elements that are awkward to produce by other algorithms.
Moreover, some information that would be useful is not presented obviously.

In the following, we use `result` as the returned object from `nls()`.

The `data` return element is an R symbol. To actually access the data from this 
element, we need to use the syntax

```
eval(parse(text=result$data))
```

However, if the call is made with `model=TRUE`, then there is a returned element
`model` which contains the data, and we can list its contents using

```
ls(result$model)
```

and if there is an element called `xdata`, it can be accessed as `result$model$xdata`.

### Information that is NOT in the nls result object

`nlsr::nlxb()` solves ostensibly the same problem as `nls()` but only claims to return

```
coefficients    A named vector giving the parameter values at the supposed solution.

ssquares        The sum of squared residuals at this set of parameters.

resid           The residual vector at the returned parameters.

jacobian        The jacobian matrix (partial derivatives of residuals w.r.t. the 
                parameters) at the returned parameters.

feval           The number of residual evaluations (sum of squares computations) used.

jeval           The number of Jacobian evaluations used.
```

However, actually looking at the structure of a returned result gives a list of 11
items:
```
 $ resid       : num [1:12] 0.0119 -0.0328 0.092 0.2088 0.3926 ...
  ..- attr(*, "gradient")= num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  .. ..- attr(*, "dimnames")=List of 2
  .. .. ..$ : NULL
  .. .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ jacobian    : num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
  ..- attr(*, "dimnames")=List of 2
  .. ..$ : NULL
  .. ..$ : chr [1:3] "Asym" "xmid" "scal"
 $ feval       : num 31
 $ jeval       : num 23
 $ coefficients: Named num [1:3] 196.19 12.42 3.19
  ..- attr(*, "names")= chr [1:3] "Asym" "xmid" "scal"
 $ ssquares    : num 2.59
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
```
This is still a smaller result object than the one `nls()` returns. Moreover, `nlxb`
explicitly returns the sum of squares as well as the residual vector and Jacobian. The 
counts of evaluations are also returned. Working on this project showed several potential
updates to the `nlsr` documentation.

### Weights in returned functions from nls()

The functions `resid()` (an alias for `residuals()`) and `fitted()` and `lhs()` are UNWEIGHTED.
But if we return `ans` from `nls()` or `minpack.lm::nlsLM` or our new `nlsj` (interim package), 
then `ans$m$resid()` is WEIGHTED.

### Interim output from the "port" algorithm

As the `nls()` **man** page states, when the "port" algorithm is used with the `trace` argument
TRUE, the iterations display the objective function value which is 1/2 the sum of squares (or 
deviance). It is likely that the trace display is embedded in the Fortran of the `nlminb`
routine that is called to execute the "port" algorithm, but the discrepancy is nonetheless
unfortunate for users.

### Failure to return best result achieved

If `nls()` reaches a point where it cannot continue but has not found a point where the
relative offset convergence criterion is met, it may simply exit, especially if a 
"singular gradient" (singular Jacobian) is found. However, this may occur AFTER the 
function has made considerable progress in reducing the sum of squared residuals. 
An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package.
Here is an abbreviated version of that problem and the `nls()` output:

```{r tetrarun}
time=c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
conc = c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
NLSdata <- data.frame(time,conc)
NLSstart <-c(lrc1=-2,lrc2=0.25,A1=150,A2=50) # a starting vector (named!)
NLSformula <-conc ~ A1*exp(-exp(lrc1)*time)+A2*exp(-exp(lrc2)*time)
tryit <- try(nls(NLSformula, data=NLSdata, start=NLSstart, trace=TRUE))
print(tryit)
```

Note that the sum of squares has been reduced from 61216 to 1.6211, but 
unless `trace` is invoked, the user will not get any information about this.
This would be an almost trivial change to the `nls() function and could be 
useful to R users.

## Feature: partially linear models and their specification

Specifying a model to a solver should, ideally, use the same syntax across
solver tools. Unfortunately, R allows multiple approaches.

One obvious case is that nonlinear modeling tools are a superset of linear ones.
Yet the explicit model

```
y ~ a*x + b
```

does not work with the linear modeling function `lm()`, which requires
this model to be specified as

```
y ~ x
```

However, even within `nls()`, we see annoying inconsistencies. Consider the 
following FOUR different calling sequences for the same problem, though the
second is to illustrate how one intuitive choice will not work. In this failed
attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
to try to add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing (end of August 2021) we do
not have a resolution for this issue.

```{r log4ways, echo=TRUE}
DNase1 <- subset(DNase, Run == 1)

## using a selfStart model - do not specify the starting parameters
fm1DNase1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
summary(fm1DNase1)
## the coefficients only:
coef(fm1DNase1)
## including their SE, etc:
coef(summary(fm1DNase1))

## using conditional linearity - leave out the Asym parameter
fm2DNase1 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1,
                 start = list(xmid = 0, scal = 1),
                 algorithm = "plinear")
summary(fm2DNase1)

## using conditional linearity AND Asym does NOT work

fm2aDNase1 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1,
                 start = list(Asym=3, xmid = 0, scal = 1),
                 algorithm = "plinear",
                 trace = TRUE))
summary(fm2aDNase1)

## without conditional linearity
fm3DNase1 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1,
                 start = list(Asym = 3, xmid = 0, scal = 1))
summary(fm3DNase1)

## using Port's nl2sol algorithm
fm4DNase1 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                 data = DNase1,
                 start = list(Asym = 3, xmid = 0, scal = 1),
                 algorithm = "port"))
summary(fm4DNase1)
```

## Issue: code structure

The `nls()` code is structured in a way that inhibits both maintenance and improvement.
In particular, the iterative setup is such that introduction of Marquardt stabilization
is not easily available.

To obtain performance, a lot of the code is in C with consequent calls and returns that
complicate the code. Over time, R has become much more efficient on modern computers, and
the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
could be much reduced by moving code entirely to R.

## Issue: code documentation for maintenance

`setPars()` -- explain weaknesses. Only used by `profile.nls()`

The paucity of documentation is exacerbated by the mixed R/C/Fortran code base.

Following is an email to John Nash from Doug Bates. This is NOT a criticism
of Prof. Bates work, but a reflection on how difficult it is to develop code in this
subject area and to keep it maintainable. We have experienced similar loss of understanding
for some of our own codes.

```
I'm afraid that I don't know the purpose of the recursive call either.  
I know that I wrote the code to use a closure for the response, covariates, etc., 
but I don't recall anything like a recursive call being necessary.

If the R sources were in a git repository I might try to use `git blame` 
to find out when and by whom that was written but they are in an SVN repository, 
I think, and I haven't used it for a long, long time.

I don't think I will be of much help.  My R skills have atrophied to the point 
where I wouldn't even know how to start exploring what is happening in the first 
call as opposed to the recursive call.
```

This was in response to an email to Doug bates on Jun 29, 2021 from John Nash.

```
    Thanks.

    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
    <https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R>

    This has the test problem and the expanded code. Around line 367 is where we are
    scratching our heads. The function code (from nlsModel()) is in the commented lines below
    the call. This is

          # > setPars
          # function(newPars) {
          #   setPars(newPars)
          #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
          #   dev   <<- sum(resid^2) # envir = thisEnv
          #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
          #   QR <<- qr(.swts * gr) # envir = thisEnv
          #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
          # }

    I'm anticipating that we will be able to set up a (possibly inefficient) code
    with documentation that will be easier to follow and test, then gradually figure
    out how to make it more efficient.

    The equivalent from minpack.lm is

    setPars = function(newPars) {
                setPars(newPars)
                assign("resid", .swts * (lhs - assign("rhs", getRHS(),
                    envir = thisEnv)), envir = thisEnv)
                assign("dev", sum(resid^2), envir = thisEnv)
                assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv)
                return(QR$rank < min(dim(QR$qr)))
            }

    In both there is the recursive call, which must have a purpose I don't understand.
```

## Feature: indexed parameters

The **man** file for `nls()` includes the following example of a situation in which
parameters are indexed. It also uses the "plinear" option as an added complication.
Here we use a truncated version of the example to save display space.

```{r nlsindx1}
## The muscle dataset in MASS is from an experiment on muscle
## contraction on 21 animals.  The observed variables are Strip
## (identifier of muscle), Conc (Cacl concentration) and Length
## (resulting length of muscle section).
if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
mm<- MASS::muscle[1:12,] # take only 1st few values of Strip
str(mm)
mm<-droplevels(mm)
str(mm)
nlev <- nlevels(mm)
withAutoprint({
  ## The non linear model considered is
  ##       Length = alpha + beta*exp(-Conc/theta) + error
  ## where theta is constant but alpha and beta may vary with Strip.
  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
  nl <- nlevels(mm$Strip)
  ## We first use the plinear algorithm to fit an overall model,
  ## ignoring that alpha and beta might vary with Strip.
  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
                start = list(th = 1), algorithm = "plinear")
  summary(musc.1)

  ## Then we use nls' indexing feature for parameters in non-linear
  ## models to use the conventional algorithm to fit a model in which
  ## alpha and beta vary with Strip.  The starting values are provided
  ## by the previously fitted model.
  ## Note that with indexed parameters, the starting values must be
  ## given in a list (with names):
  ## ?? but why use b here AND in the new formula??
  b <- coef(musc.1)

  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
  summary(musc.2)
})

```

Note that the answers for the parameters are NOT indexed. e.g. `coef(musc.2)` is a single level
vector of parameters. We do not see `a[1], a[2], a[3]` but `a1, a2, a3`. This is because the model
must integrate all the parameters because `th` is a common parameter across the index `Strip`.

We believe this structure is quite likely to cause confusion and error, and propose an alternative
approach below.

# Goals of our effort

Here are some of the goals we hope to accomplish.

## Code rationalization and documentation

We want

- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
  that matches the version in base R or what is packaged in `nlspkg` as described 
  in the "PkgFromRbase" document.

- try to obtain a cleaner structure for the overall `nls()` infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to maintenance of both
  the existing `nls()` methods and features as well as the new features we would like
  to add.

- try to explain what we do, either in comments or separate maintainer documentation.
  Since we are complaining about the lack of explanatory material for the current
  code, we feel it is incumbent on us to provide such material for our own work, and if
  possible for the existing code.

### Rationalization of formula specifications

Below in "Consistent specifications of partially linear models" we point out that
`nls()` uses a different formula specification from the default for a problem if the 
`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
add `algorithm="plinear"` to the call. Moreover, we believe it makes errors more
likely. We suggest a possible approach to avoiding this issue, but have yet to 
cast this into working code (2021-8-19).

<!-- ``` -->
<!-- y ~ x + z + x*z -->
<!-- y ~ a + b*x +c*z + d * (x*z)  # <-- fully specified -->
<!-- y ~ a*Fn1(b, c, x, z)/Fn2() -->

<!-- ``` -->

### Rationalization of indexed models

Indexed models clearly have a place in some areas of research. However, the current
approach in `nls()` is awkward. The user must use DIFFERENT formulas depending on 
whether the `plinear` algorithm is chosen. This seems to be related to the need for
that approach to use the `lm()` function or its infrastructure, thus employing 
formulas that omit the parameters of models and simply give the variables or functions
thereof, such as interaction terms.

Users, we believe, are much more likely to be comfortable with fully specified formulas.
And such formulas are needed by nonlinear least squares functions such as `minpack.lm::nlsLM`
and `nlsr::nlxb`. Thus we would like the identification of the linear parameters to be,
if possible, automated. At the very least, we should be able to use a structure like

```
algorithm="plinear(parmx, thetaz)"
```

to allow for the full formula to be used with the linear parameters identified.

## Provide tests

We need suitable tests in order:

- to ensure our new `nlsalt` or related packages work properly, in particular, giving
  results comparable to or better than the `nls()` in base R or `nlspkg`;
- to test individual solver functions to ensure they work across the range of calling 
  mechanisms, that is, different ways of supplying inputs to the solver(s);
- to pose "silly" inputs to nonlinear least squares solvers (in R) to see if 
  these bad input exceptions are caught by the programs.

### A test runner program

?? Arkajyoti -- do you want to expand?

When we have a "new" or trial solver function, we would like to know if it gives 
acceptable results on a range of sample problems of different types, starting
parameters, input conditions, constraints, subsets, weights or other settings. 
Ideally we want to be able to get a summary that is easy to read and assess. For
example, one approach would be to list the names of a set of tests with a red, green
or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last
category would be a problem with constraints that the solver is not designed to 
handle.

To accomplish this, we need a suitable "runner" program that can be supplied with
the name of a solver or solvers and a list of test problem cases. Problems generally
have a base setup -- a specification of the function to fit as a formula, some data
and a default starting set of parameters. Other cases can be created by imposing
bounds or mask constraints, subsets of the data, and different starts.

How to set up this "runner" and its supporting infrastructure is non-trivial. While
the pieces are not as complicated as the inter-related parts of the solvers, especially
`nls()`, the categorization of tests, their documentation, and the structuring to make
running them straightforward and easy requires much attention to detail.

Some considerations for our test scrips:

- Is it useful to have a "base" script for each family of test problem, with numbered particular
  cases? That is, if we run the scripts in order, we can avoid some duplication of code and
  data.

- While we have developed some tags to document the test problem families and cases, we believe
  that such tags (essentially summary documentation) will continue to need revision as different
  tools and problems are included in scope of `nlsCompare`.

- Similarly, we expect that there will be ongoing review of the structure of the result files.

# Outputs of the project

The project output is available in the Git repository https://gitlab.com/nashjc/improvenls
?? Arkajyoti -- do you want to change to Github

## Formal reports or documentation

- RefactoringNLS.Rmd: this document which will become the main report of the project.

- TestsDoc.Rmd: a survey of testing tools in R. It has more general possibilities 
  and fits into the subject of regression testing, in which case a 
  more extensive literature review will be needed.  Note that this
  document reflects the work in the the "Problem sets and test infrastructure"
  below.

  <!-- need to incorporate MachID and -->
  <!-- the many tests. Arkajyoti: should we think of working this up into a paper for -->
  <!-- JSS or the R-Journal? My  -->
  <!-- view is that this COULD be a long term side-interest for your academic work, but -->
  <!-- that would depend on your own interests as well as opportunities. -->

## Informal reports

These are documents used to discuss particular aspects of our work. These are part
of the repository https://gitlab.com/nashjc/improvenls.

- DerivsNLS.Rmd: a document to explain different ways in which Jacobian information
  is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is
  a DRAFT of a script to provide examples.

- ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
  project, possibly with wider application to nonlinear least squares in general

- MachineSummary.Rmd: an informal investigation of ways to report the characteristics
  and identity of machines running tests. `MachID.R` offers a concise summary function
  to identify a particular computational system used for tests. 
  Note that there remains an open issue that the get_ram() for Windows 10 in the `benchmarkme`
  package may report the RAM size as "NA". We are collaborating with Colin Gillespie, 
  maintainer of `benchmarkme` to seek a workaround.

- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
  the many choices in their implementation for nonlinear least squares. This is still
  a DRAFT at 2021-8-20.

- PkgFromRbase.Rmd: an explanation of the construction of the `nlspkg` from the
  code in R-base.

  <!-- We could think of putting this in R-bloggers or similar. Or if -->
  <!-- we came up with a more general template, then an R-Journal article. -->

- WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on.
  A project diary.

## Problem sets and test infrastructure

We have several test problems and variants thereof in the `inst/scripts/` directory of
the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare).
We direct the reader to that package for documentation of the test infrastructure, 
in particular the problems and methods files (`problems.csv` and `methods.csv`) and
the various functions invoked by `run.R` to produce an output file in CSV form also.

Towards the end of the project, we have focused our attention on the `nlsCompare`
package, which looks at evaluating and comparing functions for nonlinear least
squares problems. We use many of the same tests as checks that are or will be 
built into our packages for such problems e.g., `nlsalt`. These use the `testthat`
structure and may include verification of outputs that are specific to a package,
such as the upper triangular matrix R of the QR decomposition that has been 
computed for a Jacobian. Since such an object will not be computed by all methods, 
testing it in `nlsCompare` makes no sense, and in that package we concentrate on
the minimum sum of squares and the associated model parameters.

## Code and documentation

A summary of our main results:

- nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch

- nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED. 
  The effort showed that the structure of the programs was difficult to follow, 
  undocumented, and unsuited to adding improvements like the Marquardt stabilization. 
  We were able to get a pure-R version of numericDeriv() and rework most of the functions 
  of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project
  formally ends, but collaboration is likely needed with workers who have a deep knowledge
  of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`.

- nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the 
  fix by JN to the relative offset convergence criterion failure when there are small 
  residuals in problems sent to ``nls()`. Now part of R.

- nlsj: A refactoring of the `nls() functionality. `redesign2107.txt` gives some notes.
  This is an interim package for exploration of ideas and will NOT, as it currently
  is presented, become a distributed package.

- nlsralt: this is intended to be a modified version of Nash and Murdoch package 
  `nlsr` with improvements discovered as a result of this project. At the close of
  this project, this is still a work in progress.

  <!-- For example,  -->
  <!-- `nlsr` does not currently use the `subset` argument correctly. We also need to document -->
  <!-- changes e.g., to numerical derivatives. However, these are tasks for JN rather than AB. -->

## Strategic choices in a nonlinear model estimation

A key difference in approach between `nls()` and `nlsr::nlxb()`, possibly the key
difference, is that `nls()` builds a large infrastructure from which the Gauss-Newton
iteration can be executed and other statistical information such as profiles can be
computed, while `nlxb()` returns quite limited information, and in its execution
computes what is needed on an as and when basis. This follows a path that one of us
(JN) established almost 50 years ago with the software that became @cnm79, using
setup, solver, and post-solution analysis phases to computation. Here, that last 
phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr`
package. `nls()` also has some similar functions, but they are much more tied into
the infrastructure that is created mostly in the `nlsModel()` function.

The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable
entangling of the different functions and capabilities. This is to an extent that even after
both the preparation for and carrying out of the present project, we do not feel confident
to explain the code completely, nor to maintain it. However, we have made some progress toward
such a goal. Part of that has been to render the code in R. In this there remain some tasks
to complete, and they are possibly difficult.

### Opinion for discussion

To advance the stability and maintainability of R, we believe that the functions that carry out
quite advanced calculations like those achieved by `nls()` should be organized to have their 
interactions as orderly as possible. The aspects of `nls()` that have given us the most trouble
are as follows:

- The functions that compute the residuals and jacobians are frequently computed by presuming
  the current data and parameters for those functions are available in an environment. As long
  as the correct environment is used, this provides a surprisingly short syntax to invoke the
  calculations.

- Weights, subsets, and various contextual controls such as that for the `na.action` that tells
  our code what to do with undefined or missing numbers. Again, this makes for a very simple
  invocation of the calculation, but the context is hidden from the user -- and it can be
  difficult for those of us trying to maintain or improve the code to be certain we have the
  context correct.

- The mixing of R and C code makes for high performance, but without developer documentation 
  leaves programmers with a lot of work to do to adjust the code. We believe in keeping code --
  at least a runnable reference version -- in a single programming language. If necessary, by
  measuring ("profiling") the code, we can find bottlenecks and replace just those slower 
  parts of the reference code.

- We believe that a structure that isolates the setup, solve, and post-solution structures 
  for complicated functions reduces the number of objects that must be kept in appropriate 
  alignment at any one stage in a set of calculations.

- We feel that all codes should return the best fit they have found so far, even if there
  are untoward conditions reached, such as a singular Jacobian. This modification could be
  made to `nls()` even in the short term.

### Consistent specifications of partially linear models

Ideally, we believe it would be best to specify all models with a complete set of parameters.
That is, the model should be written down for as it would be used, rather than for estimation. 
The `algorithm="plinear"` setting requires that we leave out the linear parameters. This
means that we cannot simply change the `algorithm` argument without also changing the 
`formula`. Moreover, the partially linear parameters appear in the result object
as `.lin1`, `.lin2`, etc.

A possible syntax to provide consistency would be to specify the partially linear parameters
when the `plinear` option is specified. For example, we could use

```
algorithm="plinear(Asym)"
```

which would allow for the partially linear algorithm to function and the output of different
`algorithm` options to be consistent.

Furthermore, we ask if it is possible for us to devise code that will detect linearity, 
rather than expecting the user to provide the special structure and  setting. Work in this
area appears to tbe quite limited, with @Birkisson2013 being seemingly a sole contribution.
Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
is about a class of models that have almost nothing in common
with the present issue. The algorithm in @Birkisson2013 will need investigating further,
but we suspect that it is complicated enough that it will not easily be adapted to 
an `nls()` style program. Possibly a simpler algorithm may be found.

# Acknowledgement

Hans Werner Borchers was helpful in developing the proposal for this project.

# References

### 2021-12-18 RefactoringNLS/SESingGrad.R

@@ -1,0 +1,29 @@
+#parameters used to generate the data
+# rm(list=ls())
+# https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values
+reala=-3
+realb=5
+realc=0.5
+realr=0.7
+realm=1
+x=1:11 #x values - I have 11 timepoint data
+#linear+exponential function
+y=reala + realb*realr^(x-realm) + realc*x
+#add a bit of noise to avoid zero-residual data
+jitter_y = jitter(y,amount=0.2)
+testdat=data.frame(x,jitter_y)
+strt<-list(a=-3, b=5, c=0.5, r=0.7, m=1)
+#try the regression with similar starting values to the the real parameters
+jform<-jitter_y~a+b*r^(x-m)+c*x
+linexp=try(nls(jform, data=testdat, start=strt, trace=F))
+library(nlsr)
+jmod<-model2rjfun(jform, strt, jacobian=TRUE, gradient=TRUE)
+jmodval<-jmod(strt)
+jmodval
+jmodJ<-attr(jmodval, "gradient")
+cat("Singular values of Jacobian:")
+print(svd(jmodJ)$d)
+# We can also get the singular values of the Jacobian from nlsr::nlxb()
+# BUT note that the values are printed as a column purely to save space
+linexp2<-try(nlxb(jform, data=testdat, start=strt, trace=F))
+linexp2

### 2021-12-18 DerivsNLS/DerivsNLS.Rmd
add derivs and variety papers

---
title: "Jacobian Calculations for nls()"
author:
  - Arkajyoti Bhattacharjee, Indian Institute of Technology, Kanpur
  - John C. Nash, University of Ottawa, Canada
date: "26/05/2021"
output: 
    pdf_document:
        keep_tex: false
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
---

<!-- - Heather Turner, University of Warwick, UK -->

```{r setup, include=FALSE}
rm(list=ls()) # clear the workspace for this document
knitr::opts_chunk$set(echo = TRUE)
## setup for this document
library(nlsr)  # So we have the analytic/symbolic derivatives
library(numDeriv) # numerical derivatives package
library(microbenchmark)  # timing

printsum <- function(xx){ print(summary(xx))} # May be needed
traceval  <-  TRUE  # traceval set TRUE to debug or give full history
#  Set to FALSE when we don't need extensive output
```

# ISSUES

- ExDerivs.R file causes a number of failures in the ORIGINAL numericDeriv.

- Need to verify nlsalt:: version of numericDeriv() matches all cases of nlspkg:: version

- Do we need to get a model frame? How? and How to use it?

## TODOS (mostly from nlsr vignette nlsr-devdoc.Rmd)

- how to insert numerical derivatives when Deriv unable to get result (nlsr)
- approximations for jacfn beyond fwd approximation. How to specify??
- how to force numerical approximations in nlfb() in a 
  manner consistent with that used in `optimx::optimr()`, that is, to
  surround the name of `jacfn` with quotes if it is a numerical approximation,
  or to provide a logical control to `nlxb()` for this purpose.

# Jacobians in nls()

This document source is in file **DerivsNLS.Rmd**.

`nls()` and other nonlinear least squares programs in R need a Jacobian matrix 
calculated at the current set of trial nonlinear model parameters to set up 
the Gauss-Newton equations or their stabilized modifications in methods such as
that of Marquardt (@Marquardt1963). Unfortunately, `nls()` calls
the Jacobian the "gradient", and uses function `numericDeriv()` to compute them.
This document is an attempt to describe different ways to compute the Jacobian
for use in nls() and related software, and to evaluate these approaches from
several perspectives.

In evaluating performance, we need to know the conditions under which the evaluation
was conducted. Thus the computations included in this document, which is built using
`Rmarkdown`, are specific to the computer in which the document is processed. We
will add tables that give the results for different computing environments at the
bottom.

# An example problem

We will use the Hobbs weed infestation problem (@jncnm79, page 120).

```{r hobbsex}
# Data for Hobbs problem
ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
tdat  <-  seq_along(ydat) # for testing

# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
start1  <-  c(b1=1, b2=1, b3=1)
eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt)) # formula -- display structure with str(eunsc)
# Can we convert a string form of this "model" to a formula
ceunsc <- " y ~ b1/(1+b2*exp(-b3*tt))" # This will give character form: str(ceunsc)
# Next line Will be TRUE if we have made the conversion OK
print(as.formula(ceunsc)==eunsc)
weeddata1  <-  data.frame(y=ydat, tt=tdat) ## LOCAL DATA IN DATA FRAMES
weedenv <- list2env(weeddata1) ## Put data in an Environment
# Add the parameter data as "variables" 
weedenv$b1 <- start1[[1]]; weedenv$b2 <- start1[[2]]; weedenv$b3 <- start1[[3]]
# Display content of the Environment with ## ls.str(weedenv)
# We are now set up for computations
```

# Tools for Jacobians

There are a number of ways to get the Jacobian in R.

## numericDeriv() original version from base R

`numericDeriv` is the R function used by `nls()` to evaluate Jacobians for its Gauss-Newton
equations. The R source code is in the file `nls.R`. It calls a C function numeric_deriv
in `nls.c`. These have been extracted in an R package form as `nlspkg` by Duncan Murdoch
as described in our document **PkgFromRbase.Rmd: Making a package from base R files**, and
we will use that version.

In the following we will test and time `numericDeriv()` along with various of its options.

```{r nd1}
rexpr<-call("-",eunsc[[3]], eunsc[[2]]) # Generate the residual "call"
res0<-eval(rexpr, weedenv) # Get the residuals
print(res0) # the base residuals
cat("Sumsquares at 1,1,1 is ",sum(res0^2),"\n")
treseval<-microbenchmark(res0<-eval(rexpr, weedenv))
print(treseval)
rexpr<-call("-",eunsc[[3]], eunsc[[2]]) # This is the "call" that computes the residual
## Try the numericDeriv option
theta<-names(start1)
## suppressMessages(library(nlspkg))
suppressMessages(ndnls<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv))
print(ndnls)
print(sum(ndnls^2))
tndnls<-microbenchmark(ndnls<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv))
print(tndnls)

## numericDeriv also has central difference option, as well as choice of eps parameter
## Central diff
ndnlsc<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE)
print(ndnlsc)
print(sum(ndnlsc^2))
tndnlsc<-microbenchmark(ndnlsc<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE))
print(tndnlsc)

## Forward diff with smaller eps
ndnlsx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, eps=1e-10)
print(ndnlsx)
print(sum(ndnlsx^2))
tndnlsx<-microbenchmark(ndnlsx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, eps=1e-10))
print(tndnlsx)

## Central diff with smaller eps
ndnlscx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=1e-10)
print(ndnlscx)
print(sum(ndnlscx^2))
tndnlscx<-microbenchmark(ndnlscx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=1e-10))
print(tndnlscx)

## Add dir parameter -- the direction of the parameter shift
ndnlsd<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, dir=-1)
# Does dir make a difference? This might be accidental for forward difference.
max(abs(ndnlsd-ndnls))
ndnlscd<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, dir=-1)
# Does dir make a difference? For central diff it should NOT!
max(abs(ndnlscd-ndnlsc))
```

## numericDeriv() alternative pure-R version

This version (see Appendix 2) has C code replaced with R equivalents.

```{r and1}
## Try ExDerivs.R ??
suppressMessages(andnls<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv))
# print(andnls); print(sum(andnls^2))
tandnls<-microbenchmark(andnls<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv))
print(tandnls)
## numericDeriv also has central difference option, as well as choice of eps parameter
## Central diff
andnlsc<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE)
ndnlsc<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE)
# print(andnlsc); print(sum(andnlsc^2))
tandnlsc<-microbenchmark(andnlsc<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE))
print(tandnlsc)

## Forward diff with smaller eps
andnlsx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, eps=1e-10)
ndnlsx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, eps=1e-10)
# print(andnlsx); print(sum(andnlsx^2))
tandnlsx<-microbenchmark(andnlsx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, eps=1e-10))
print(tandnlsx)

## Central diff with smaller eps
andnlscx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=1e-10)
ndnlscx<-nlspkg::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=1e-10)
# print(andnlscx) ; print(sum(andnlscx^2))
tandnlscx<-microbenchmark(andnlscx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=1e-10))
print(tandnlscx)

## Comparisons for Jacobian between nlspkg and nlsalt i.e. R&C vs just R
max(abs(attr(ndnls, "gradient")-attr(andnls,"gradient")))
max(abs(attr(ndnlsc, "gradient")-attr(andnlsc,"gradient")))
max(abs(attr(ndnlsx, "gradient")-attr(andnlsx,"gradient")))
max(abs(attr(ndnlscx, "gradient")-attr(andnlscx,"gradient")))

## Using dir
cat("eps (regular) = ",.Machine$double.eps^(1/2),
    "  eps (central) =",.Machine$double.eps^(1/3),"\n")
andnlsd<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, dir=-1)
max(abs(attr(andnlsd, "gradient")-attr(ndnls,"gradient")))
max(abs(attr(andnlsd, "gradient")-attr(andnls,"gradient")))
andnlscd<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, dir=-1)
max(abs(attr(andnlscd, "gradient")-attr(ndnlsc,"gradient")))

## Try comparisons over different eps sizes
for (ee in 3:10){
andnlsd<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, dir=-1, eps=10^(-ee))
andnlscd<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, dir=-1, eps=10^(-ee))
andnlsx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, eps=10^(-ee))
andnlscx<-nlsalt::numericDeriv(rexpr, theta, rho=weedenv, central=TRUE, eps=10^(-ee))
cat("Regular diff, eps=10^(-",ee,"):",
      max(abs(attr(andnlsd,"gradient")-attr(andnlsx,"gradient"))),"\n")
cat("Central diff, eps=10^(-",ee,"):",
    max(abs(attr(andnlscd,"gradient")-attr(andnlscx,"gradient"))),"\n")
}
```

The `dir` parameter allows us to use a backward difference for the derivative. This 
appears in `nlsModel()` for the case where a parameter is on an upper bound for
the case `algorithm="port"`. It does not check for nearness to the bound, and 
for the lower bound assumes that we are stepping AWAY from the bound in the
default direction (`dir=+1`). None of the code addresses the issue where bounds
are closer together than the step used for the finite difference, so there are
situations where we could crash the code. Nor does the code check if the central
difference is specified when near a bound.

- In the case of lower bounds, a central difference can overstep the bound when
  a parameter is "close" or on the bound.
- In the case of an upper bound, changing the `dir` will not change the central 
  derivative approximation expression and steps in both forward and backward 
  directions of the parameter are taken.

## Symbolic methods from `nlsr`

The package `nlsr` has a function `model2rjfun()` that converts an expression
describing how the residual functions are computed into an R function that
computes the residuals at a particular set of parameters and sets the 
**attribute** "gradient" of the vector of residual values to the Jacobian at
the particular set of parameters. `model2rjfun()` does much the same work
as the `res0<-eval(rexpr, weedenv)` expression evaluation, but adds derivative
expressions to the function.

```{r nlsr1}
# nlsr has function model2rjfun. We can evaluate just the residuals
res0<-model2rjfun(eunsc, start1, data=weeddata1, jacobian=FALSE)
res0(start1)
tres0nlsr<-microbenchmark(res0(start1)) # time it
print(tres0nlsr)
# or the residuals and jacobian
funsc <- model2rjfun(eunsc, start1, data=weeddata1) # from nlsr: creates a function
tmodel2rjfun <- microbenchmark(model2rjfun(eunsc, start1, data=weeddata1))
print(tmodel2rjfun)
## Ways to display information about the residual/jacobian function
# print(funsc); print(funsc(start1)); print(environment(funsc)); print(ls.str(environment(funsc)))
# print(ls(environment(funsc)$data)); eval(eunsc, environment(funsc))
vfunsc<-funsc(start1)
print(vfunsc)
tfunsc<-microbenchmark(funsc(start1))
print(tfunsc)
```

# `numDeriv` package

The package `numDeriv` includes a function `jacobian()` that acts on a user
function `resid()` to produce the Jacobian at a set of parameters by several
choices of approximation.

```{r numDeriv1}
# We use the residual function (without gradient attribute) from nlsr
jnumd<-jacobian(res0, start1) #  uses default "Richardson" method
jnumd
# Timings of the analytic jacobian calculations
tjnumd<-microbenchmark(jnumd<-jacobian(res0, start1))
print(tjnumd)

jnumds<-jacobian(res0, start1, method="simple") #  uses default "Richardson" method
jnumds
# Timings of the analytic jacobian calculations
tjnumds<-microbenchmark(jnumds<-jacobian(res0, start1, method="simple"))
print(tjnumds)

jnumdc<-jacobian(res0, start1, method="complex") #  uses default "Richardson" method
jnumdc
# Timings of the analytic jacobian calculations
tjnumdc<-microbenchmark(jnumdc<-jacobian(res0, start1, method="complex"))
print(tjnumdc)
```

Note that the manual pages for `numDeriv` offer many options for the functions in
the package. We have yet to explore many of these.

# Comparisons

In the following, we are comparing to `vfunsc`, which is the evaluated 
residual vector at `start1=c(1,1,1)` with "gradient" attribute (jacobian)
included, as developed using package `nlsr`. This is taken as the "correct"
result, even though it is possible that the generated order of calculations
may introduce inaccuracies in the supposedly analytic derivatives.

`numericDeriv` computes a similar structure (residuals with "gradient" attribute):
`ndnlsc`: the forward difference result with default `eps` (.Machine$double.eps^(1/2))
`ndnlsc2`: Central difference with default `eps` (.Machine$double.eps^(1/3))
`ndnlscx`: Forward difference with smaller eps=1e-10
`ndnlscx2`: Central difference with smaller eps=1e-10

`jnumd`: numDeriv::jacobian() result with default settings.

```{r compjac1}
## Matrix comparisons -- uncomment code to show these, which use page space
# attr(ndnls, "gradient")-attr(vfunsc,"gradient")
# attr(ndnlsc, "gradient")-attr(vfunsc,"gradient")
# attr(ndnlsx, "gradient")-attr(vfunsc,"gradient")
# attr(ndnlscx, "gradient")-attr(vfunsc,"gradient")
# jnumd - attr(vfunsc,"gradient")
# jnumds - attr(vfunsc,"gradient")
# jnumdc - attr(vfunsc,"gradient")

## Summary comparisons - maximum absolute differences
max(abs(attr(ndnls, "gradient")-attr(vfunsc,"gradient")))
max(abs(attr(ndnlsc, "gradient")-attr(vfunsc,"gradient")))
max(abs(attr(ndnlsx, "gradient")-attr(vfunsc,"gradient")))
max(abs(attr(ndnlscx, "gradient")-attr(vfunsc,"gradient")))
max(abs(jnumd - attr(vfunsc,"gradient")))
max(abs(jnumds - attr(vfunsc,"gradient")))
max(abs(jnumdc - attr(vfunsc,"gradient")))
```

## Observations

Some particular notes:

- the mean time for the default `numericDeriv()` of `nls()` is quite
  fast and its coefficient of variation (sd/mean) is around 0.42. The
  timings are actually very slightly faster than the analytic expressions
  of `nlsr`, but the latter has a COV of 0.36.

- this default method Jacobian unfortunately deviates from the analytical
  computation by a relatively large amount (of the order of 1e-6 for our
  example).

- the central difference version of `numericDeriv()` does better (about 
  three orders of magnitude smaller deviation from the analytic result),
  and the time is comparable with the analytic evaluation.

- making the `eps` parameter smaller degrades the accuracy of the Jacobian
  computed via `numericDeriv()`. This may be counter-intuitive for those 
  unfamiliar with numerical methods. Essentially, a smaller `eps` results in
  subtraction of very close values for the residuals.

- the "simple" option of the `numDeriv` function `jacobian()` gives similarly
  poor accuracy.

- the "Richardson" (default) results of `numDeriv` are of similar accuracy to
  the central difference option of `numericDeriv()` but at a much greater time
  cost -- about 23 times slower.

- on the other hand, the "complex" option gets an essentially analytic result,
  approximately 8 orders of magnitude better than the central difference approximation
  of `numericDeriv`, for a time cost only 3.5 times as great. Unfortunately, not all
  models are amenable to the complex step approximation.

# Cautionary notes on performance results

The results here have been evaluated on a single computer. In fact, while we could
process the Rmarkdown file on any of several machines, the work was mainly carried out
on a machine characterized with the string

```
M21:john-Linux-5.11.0-25-generic|Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz|33482145792bytesRAM 
```

This is a relatively capable tower PC, but otherwise unremarkable.

Nevertheless, we found that running the timings more than once when other tasks were in 
progress did result in variations in the mean and standard deviation of the timings of several
percentage points. We would expect both absolute differences in times and changes in relative
performance with different processors and operating systems, and had thought to carry out some
investigation of such differences. However, such effort seems less valuable than pursuing more
capable nonlinear least squares and derivative code.

# Some notes on derivative computation for nonlinear least squares

In no particular order, we comment on some issues relating to the Jacobian calculations
in nonlinear least squares.

## Nomenclature

R is not in step with many other areas of numerical computation when labeling different
objects in the nonlinear least squares problem. In particular, R uses the term "gradient"
when the object of interest is the Jacobian matrix. In that it is useful in performing 
iterations of the Gauss-Newton or related equations to have the Jacobian associated with
the residuals, and the rows of the Jacobian matrix are "gradients" of the respective
residuals, we can accept the attribute name "gradient" to select the required information.
Moreover, as in package `nlsr` it is very useful to have the Jacobian matrix as an 
attribute of the residual vector, since the main solver function, in this case `nlsr::nlfb()`,
can be called with the same input for the arguments `res` and `jac`. These are the functions
required to compute the residual and the Jacobian, and using the same function for both is
very convenient, but needs some way to return both the residual vector and Jacobian matrix
in a coherent fashion.

## Numerical approximation near constraints

As far as we are aware there is no software that implements a fully safeguarded system to
compute numerical approximation of the Jacobian (or gradients in general optimization) near
constraints. The same statement applies even in the case of the much simpler bounds constraints.
Users have a perverse tendency to devise ways to foil our best efforts. For example, they may
decide that a good way to specify fixed (i.e., masked) parameters that they do not want to vary
during a particular calculation is to specify the lower and upper bound of a parameter at the
same value. Later runs may want the parameter constraints relaxed.

In `nlsr::nlxb()`, users may, in fact, specify masked parameters this way. This is a case of 
"if you can't beat them, join them", but it does provide an easily understood way for users to
fix values.

More tricky is dealing with constraints that are close together. Note that these may arise from,
for example, two linear (planar) constraints that approach at a narrow angle. In the apex where
these constraints intersect, we will have tight bounds on parameters. If the constraint is not
one that is imposed by the nature of the residual or objective function, for example, a log() 
or square root near zero, then we can generally proceed and allow the derivative approximation
to evaluate outside the constraints. Things are decidedly nastier if we do have inadmissible
values of the parameters. This is where analytic Jacobian evaluation is very helpful.

The issue of constraints and the need for a step in parameter values for derivative approximations
was one of the motivations for trying to find analytic derivatives in package `nlsr` and the 
continuing effort to bring them into other R tools.

## A case where a parameter is close to inadmissible

The following example shows that numericDeriv() does a reasonable job of computing
the Jacobian, but the result is still singular.

```{r code=xfun::read_utf8('badJlogmod.R')}
```

# Appendix 1: Base R numericDeriv code

This code is in two files, nls.R and nls.c and is extracted here.

## From nls.R

```{r nlsR, comment=NA, echo=FALSE}
   cat(readLines('nls-numericDeriv.R'), sep = '\n')
```

## From nls.c

```{r nlsc, comment=NA, echo=FALSE}
   cat(readLines('nls-numericDeriv.c'), sep = '\n')
```

# Appendix 2: numericDeriv() from nlsalt package (all in R)

```{r nlsndnew, comment=NA, echo=FALSE}
   cat(readLines('nlsnd.R'), sep = '\n')
```

# References

### 2021-12-18 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
add derivs and variety papers

---
title: "Variety in the Implementation of Nonlinear Least Squares Program Codes"
author: 
   - John C Nash, retired professor, University of Ottawa
date: "16/02/2021"
output: 
    pdf_document:
        keep_tex: false
        toc: true
bibliography: ../BibSupport/ImproveNLS.bib
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## ??require(bookdown) # language engine to display text??
```

# Abstract

There are many ways to structure a Gauss-Newton style nonlinear least squares
program code. In organizing and documenting the nearly half-century of programs
in the Nashlib collection associated with @jncnm79, the author realized that this
variety could be an instructive subject for numerical software designers.

# Underlying algorithms

## Gauss Newton method

An overview of the method is provided by Wikipedia:

\url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}

The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
coelestium in sectionibus conicis solem ambientum*.

In calculus we learn that a stationary point (local maximum, minimum
or saddle point) of a function $f$ occurs where its gradient (or first
derivative) is zero. In multiple dimensions, this is the same as
having the gradient $g$ stationary. The so-called Newton method
uses the second derivatives in the Hessian matrix $H$ defined as

$$  H_{i,j} = \partial^2{f}/ {\partial{x_i} \partial{x_j}}$$

where $f$ is a function of several variables $x_i$, $i=1,...,n$, written
collectively as $x$. The Newton equations are defined as

$$  H \delta = - g $$

where $H$ and $g$ are evaluated at a guess or estimate of $x$. $x$ is then updated
to

$$ x \leftarrow x + \delta $$

and we iterate until there is no change in $x$.

There are some adjustments when our function $f(x)$ can be written as a sum of squares

$$  f(x) = \sum_{i=1}^n {r_i(x)^2} = r^t r $$

The final vector product form is the one I favour because it is the least fussy to write and read.
In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is

$$ H_{j,k} = 2 \sum_{i-1}^n { { [ (\partial r_i/\partial{x_j} ) } ({\partial r_i/\partial{x_k} ) + r_i (\partial^2 r_i/{\partial{x_j} \partial{x_k} ) ]}}} $$

If we define the **Jacobian** matrix

$$ J_{i,j} = \partial r_i/\partial{x_j} $$

then the gradient is

$$ g = 2 J^t r $$

and the first part of the Hessian is

$$  J^t J $$

Arguing that the residuals should be "small", Gauss proposed that the 
Newton equations could be approximated by ignoring
the second term (using elements of the residual times second derivatives 
of the residual).  This gives the **Gauss-Newton**
equations (cancelling the factor 2)

$$ (J^t J)  \delta  =  - J^t r $$

We can use this in an iteration similar to the Newton iteration. Note that these
equations are equivalent to the linear least squares problem that can be written

$$  J \delta  \approx - r $$

## Hartley's method

Conditions for convergence of the Newton iteration and the Gauss-Newton 
iteration are rather a nuisance to verify. In any case we want a solution 
to the original nonlinear least squares problem. 
@Hartley61 proposed that rather than use an iteration

$$ x \leftarrow x + \delta$$

one could do a search along the direction $\delta$. Ideally we would want to **minimize**

$$ f(x + \alpha \delta) $$

with respect to the (scalar) parameter $\alpha$. That one-dimensional minimization can
take quite a lot of computational effort, however, and typically a value of $\alpha$ that
permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the iteration repeated
from

$$ x_{new} = x + \alpha * \delta $$

Clearly there are many tactical choices that give rise to a variety of particular algorithms.
One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
be perpendicular to any downhill direction.

@Hartley61 recommends a quadratic (parabolic) line search where two "new" points
are evaluated, followed by a third try at the suggested minimum of the parabola fitted
to the existing and two new sum of squares function values. Tactics are needed to deal
with the case that the proposed point is not an improvement in the sum of squares.

A different approach is the R function `nls()`, which calls the C routine `nls.c` to compute 
the Gauss Newton direction <!-- (`incr` -- see nls-flowchart.txt) -->
after which a step-halving "line-search" is used.

In `nls.c` we also see that once a satisfactory "new" set of parameters has been 
obtained, we reset the $\alpha$ value to the minimum of $2 * \alpha$ and 1.0. Oddly, there
is no provision for the user to set the tactical choices of the maximum value, the reduction
or the increase in $\alpha$, despite our experience that other choices can be advantageous
for some problems. Moreover, $x_{new}$ is deemed satisfactory if its **deviance** (essentially
the weighted sum of squares of the residuals) is **no greater than** the current best
value. Our experience is that it may be better to insist on a decrease in the value,
or even to use an "acceptable point" approach that enforces a sufficient decrease using
a measure based on the gradient of the deviance projected on $\delta$.

<!-- ?? Should we explain more. Possibly incorporate into code? -->

## Marquardt stabilized Gauss-Newton

@Marquardt1963 is perhaps one of the most important developments in nonlinear least squares
apart from the Gauss-Newton method. There are several ways to view the method.

First, considering that the $J^t J$ may be effectively singular in the computational
environment at hand, the Gauss-Newton method may be unable to compute a search
step that reduces the sum of squared residuals. The right hand side of the 
normal equations, that is, $g = - J^t r$ is the gradient for the sum of squares.
Thus a solution of

<!-- Note mathbf use below -->

$$    \mathbf{1}_n \delta = - g $$
is clearly a downhill version of the gradient. And if we solve

$$  \lambda  \mathbf{1}_n \delta =  - g $$

various values of $\lambda$ will produce steps along the **steepest descents**
direction. Solutions of the Levenberg-Marquardt equations

$$ (J^t J + \lambda  \mathbf{1}_n)  \delta  = A \delta =  - J^t r $$

can be thought of as yielding a step $\delta$ that merges the Gauss-Newton and 
steepest-descents direction. This approach was actually first suggested by 
@Levenberg1944, but it is my opinion that Levenberg 
likely never exercised them in practice, while Marquardt actually had a well-used
program. Before computers were commonly
available, many papers were written with computational
ideas that were given at most a cursory trial. (In 1984, I had the pleasure of showing
Marquardt my own BASIC program running on a Radio Shack TRS80 Model 100 and he  
commented on how the computations were converging to a solution as we watched
the iterations proceed.)

Note that we can define a matrix $K$ which is $J$ augmented below by a suitable diagonal
matrix that is the square root of $\lambda$ times the unit matrix $1_n$. Then the
Levenberg-Marquardt equations are equivalent to

$$  K \delta \approx - r_{augmented} $$

where we have added $n$ zeros to $r$ to make $r_{augmented}$. Similar adjustments
are used for scaled Marquardt methods mentioned below.

Practical Marquardt methods require us to specify an initial $\lambda$ and ways
to adjust its value. For example, if $$SS(x) = r^t r = f(x)$$ is the function 
we wish to minimize to "fit" our nonlinear model, one set of choices is:

- set an initial $\lambda$ of 0.0001

- **A** Compute $J$ and $g$, set up Marquardt equations and find $\delta$

- **B** compute new set of parameters $$x_{new} = x + \delta $$

- if $x_{new} \ne x$, compute $SS(x_{new})$, else quit (**termination**)

- if $SS(x_{new}) < SS(x)$, replace $x$ with $x_{new}$, replace $\lambda$
with $0.4*\lambda$, then retry from **A**

- ELSE increase $\lambda$ to $10*\lambda$ and retry from **B** ($J$ and $g$ do
  NOT need to be recomputed)

In this scheme, apart from the many possibilities that "find $\delta$" allows, 
an algorithm is defined by the initial value, decrease factor 
(e.g., 0.4) and increase factor (e.g., 10) for $\lambda$. However, there are a number of 
further annoying computational 
details concerning underflow or overflow and how we measure equivalence of 
iterates. That is, we need to ensure $\lambda$ is not so "small" that multiplying
by 10 is meaningless, and we need to know what "not equal" means for floating
point vectors in the arithmetic in use.

An example of choices to deal with these details:

- if $\lambda$ is smaller than some pre-set tolerance, set it to the tolerance. 
  (Otherwise do not worry if it is tiny, as long as the sum of squares still gets
   reduced.)

- when comparing $x_{new}$ and $x$, we can use a modest number, e.g, 100.0 which
  we will call $OFFSET$. In the arithmetic of the system at hand, if the floating-point
  number $(x_{new} + OFFSET)$ matches bit for bit $(x + OFFSET)$, then we take the
  unmodified numbers as equal.

Many practitioners dislike using IF statements (comparisons) of floating point
numbers. Tolerances are often suggested to define "small", but this invites users
to make decisions that lead to poor outcomes and which ignore the precision and
other characteristics of the arithmetic at hand.
There are, of course, well-written codes that address the issues of what "small" means,
but such code is generally quite involved.
I have never seen a case where the $OFFSET$ approach caused difficulties, and it
is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
itself and declare equality. When $x$ is very large, $OFFSET$ does not change it and
we compare $x$ with a value that is identical to it.

### Automatic scaling

In @Marquardt1963, it is shown that the use of the diagonal elements of $J^t J$
instead of $\mathbf{1}_n$ in the Marquardt stabilized Gauss-Newton equations is equivalent
to having a similar scale for all the parameters. Call this diagonal matrix $D$.
Generally this is the preferred
approach to Levenberg-Marquardt stabilizations. However, in low-precision 
arithmetic, elements of the diagonal may underflow to zero and cause issues of
singularity in the equations to be solved. A simple workaround (@jn77ima) is to slightly
modify the equations to

$$ (J^t J + \lambda (D + \phi 1_n)  \delta  =  - J^t r $$

where $\phi$ is a modest number such as 1.

<!-- ?? Check the statements about how things are done and redo tests -->
Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
the scaled stabilization, 
which is quite a lot of computational work. The same program, however, 
uses a QR approach to solving the linearized least squares sub-problem.
To avoid the formation of $D$, an experimental package
`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
use ONLY the fixed element (i.e., unit matrix) stabilization. A very
preliminary timing ([2021-6-26]) using the Hobbs unscaled problem shows
a slight improvement for the simplified version in timing with equivalent
solution. ?? check

## Other approaches

There have been many proposed approaches to nonlinear least squares. The following
is a non-exhaustive sample.

### Spiral

@Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart
given but I have never been able to find a code (supposedly Fortran V for the Univac
1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
line-search minimization. Jones mentions matrix inversion, and it is not clear whether
he truly forms the inverse. The main idea in the paper (and presumably the code) is
to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$
is increased. It would be interesting to actually try some of the ideas, which do,
as far as I can determine, have a reasonable basis.

### Approaches to Newton's method for special models

Newton (and his sometime associated worker Raphson) used the algorithm we attribute
to them applied to a very special case and in a manner unrecognizable as that described
above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).

The main attraction of the Gauss-Newton method is that the Jacobian only needs first
derivatives, which are generally much, much easier to compute than the full second
derivatives of the Hessian. However, there have been suggestioned
Newton-like approaches, possibly for special types of models where some short cut
to the Hessian is available. These may or may not make sense in the context of
a given user or group of users. I regard these approaches as part of the
computational infrastructure of the special cases rather than as general nonlinear
least squares methods.

?? MaxLik?? others -- Charles Geyer

https://www.stat.umn.edu/geyer/trust/

CRAN trust package??

?? had something for exponential models??

### Hybrid methods

The central idea of Hartley's method is a line search and that of Marquardt a
parametrized stabilization that can be shown to reduce a sum of squares for large
enough stabilization parameter. There is nothing to prevent use of a line
search additional to the Marquardt stabilization, and there are, of course, 
several reasonable choices for line search. Also we could solve the (possibly
stabilized) Gauss-Newton equations by several quite different numerical linear
algebra techniques, or even approximations. In certain situations, some of these
choices may be important, but here I will not pursue them further.

# Sources of implementation variety

The sources of variety in implementation include:

- structuring of the algorithm, that is, how we set up and sequence the
  parts of the overall computation

- Linking of data to the problem. In particular, R allows **subsetting**
  of data at quite a high level, but this capability is almost totally
  hidden from the user or programmer, so that the feature could be dangerous to
  use. There appears, moreover, to be no example or test with `nls()` 
  to show how to use this facility, even though `subset` is an argument
  to the `nls()` function. This is discussed further below.

- programming language

- possible operating environment features

- solver for the least squares or linear equations sub-problems

- stucture of storage for the solver, that is, compact or full

- sequential or full creation of the Jacobian and residual, since
  it may be done in parts

- how the Jacobian is computed or approximated

- higher level presentation of the problem to the computer, as
  in R's `nls` versus packages `minpack.lm` and `nlsr`, or as
  a general optimization problem.

## Algorithm structure

The R function `nls()` and also the package `minpack.lm` both set up the nonlinear
least squares computation for estimating a model specified as a **formula** (an
R object class) by building an `nlsModel` object, which we will label `m`. This 
object is a set of functions that provide the information needed to compute the 
residuals, jacobian, the search direction, and many other objects required for
the nonlinear least squares solution as well as ancillary information used in
the post-solution analysis and reporting.

In the package `nlsr` (@nlsr-manual), the function `nlxb()` calls `model2rjfun()`
to create functions `res()` and `jac()` that are used in `nlfb()` (the function
`nls.lm()` in `minpack.lm` is very similar) to find a
nonlinear least squares solution. However, post-solution information is built
explicitly AFTER a solution is found.

The essential difference is that `nls()` and `minpack.lm` build a toolbox, `m`,
while `nlsr` creates its required objects as they are needed for the Marquardt
computations.

### Issues with this structure

A major criticism of this approach is that the collection of information needed
at any one time -- the "environment" in which residuals and jacobians are calculated,
as well as the solution methods -- is not at all well-documented. If the software is
to be updated or maintained, the location and purpose of the many elements that
reside in different environments is difficult to determine without much effort.
In my opinion, this renders the structure prone to collapse if there are infrastructure
or other changes. Documentation would go a long way to correcting this problem, but
even better would be clearer structuring. Separation of the problem from the method
would also be useful. At the time of writing of this document, the Gauss-Newton 
METHOD is embedded in the functions created by `nlsModel()` for a given PROBLEM.

## Programming language

We have a versions of the Nashlib Algorithm 23 in BASIC, Fortran, Pascal, and R, 
with Python pending. There may be dialects of these programming languages also,
giving rise to other variations.

## Data linkage to software

Quite reasonably, linking particular data to our problem is mediated by software,
and hence is influenced by the programming language. This is particularly true
for R, since R allows us to **subset** data. Essentially this provides an index
to the rows of data, most easily thought of in terms of a data frame structure.

### Data subsets

The `nls()` function has an argument `subset` and a test included in the 
`Croucher-example1.R` file of this project showed this works as expected. 
However, nowhere have we found a published example of the use of the 
`subset` argument with `nls()`. Worse, it clearly must be applied to
many of the functions used by `nls()` but makes no explicit appearance. Thus it
is VERY difficult to ensure subsetting is correctly applied when we wish to modify
the code, for instance, to provide for newer features especially more reliable
solvers.

Note that the `subset` ARGUMENT to `nls()` is not the same as the `subset()` function
that is usable to restrict the scope of an R object.

OUR HYPOTHESIS: `nls()` uses `subset` to automatically adjust the data to which all
the relevant functions are applied. We note that `minpack.lm::nlsLM` also has a `subset` argument,
but package `nlsr` does not at time of writing (2021-7-12), but will use the **weights** mechanism
of the next subsection of this document at some future release.

### Other approaches to subsetting

In order to provide a sane and well-understood mechanism for subsetting, there
are alternatives for implementation.

- **via weights**: We can weight each "row" that we do NOT wish included as 0.
  This does require us to compute the number of observations as the count of
  non-zero weights e.g, `sum(weights > 0). Incidentally, this underlines the 
  need to ensure weights are non-negative. ??now in nlsj, need to add to nlsr code.??

- **via preprocessing**: We could prepare a function to pre-process the problem and
  create a data collection that is appropriately subset. This may offend against
  the spirit of R/S, but it has the merit that we can easily reassure ourselves
  that we have worked with the correct information, correctly used in our methods.
  ?? Should add this as an option.

PROPOSAL: use the "weights" approach if `nlsj` is called with `subset` active, 
else recommend the preprocessing approach. The preprocessor could also do a lot
of checking, which we might leave out of `nlsj()` in some cases. As of 2021-7-12,
the weights mechanism is working in a trial `nlsj()` and should be transferrable
to package `nlsr` fairly easily.

## Operating environment

Generally, for modern computers, the operating system and its localization do not
have much influence on the way in which we set up nonlinear least squares computations,
though the actual performance of the computations may well be influenced by such factors.
I will comment below about some issues where speed and size of data storage may favour
some approaches over others. However, such issues are less prominent today than in
the past.

One aspect of the computational environment that has proved important to me has
been that of low-precision arithmetic. In the 1970s, it was common to have only
the equivalent of 6 or 7 (decimal) digits of precision. In such environments, 
where, moreover, guard digits and underflow and overflow behaviour were highly
variable features, it was quite common that arguments to functions like exp()
and log() were inadmissible for reasons of computability in the local system
rather than underlying mathematics. In such environments, where there were also
storage limitations, it was important to use very robust algorithms implemented
in ways that were simple but "safe". This led to a number of choices in the
development of the codes in @jncnm79. However, it has turned out that these
choices fare rather well even in more capable computational systems.

## Solver for the least squares or linear equations sub-problems

There are many choices for solver of the Gauss-Newton or related equations
in a nonlinear least squares program. We will consider some of the options.
Note that the choice to set up the normal equations will direct part of the
algorithm selection.

### Solution of the linear normal equations

The Gauss-Newton or Marquardt eqations are a set of linear equations. Moreover, 
the coefficient matrix is non-negative definite and symmetric. Thus it permits of
both general and specialized methods for linear equations. Furthermore, one can also
set up the solution without forming a coefficient matrix $A$ 
by using several matrix decomposition methods. Thus there are many possible procedures.

- Gauss elimination with partial or complete pivoting: The choice of partial or
  complete pivoting changes greatly the program code. However, neither of these
  choices uses the underlying structure and non-negative definite nature of 
  the matrix $A$.

- Variants of Gauss elimination that build matrix decompositions: We can apply
  elimination to produce an LU decomposition of $A$, and this may offer some 
  computational advantages, though in the present case such approaches do not
  seem to be ideal.

- Gauss-Jordan inversion: Here we compute the inverse of $A$ and then work with
  this inverse. That is, we use

  $$\delta = - A^{-1} g$$

  This uses much more computational effort than simply solving for $\delta$, 
  though there is an interesting, and very tiny, method by @BauerReinsch71.
  The sheer ingenuity of this code was magnetic, but it has proved to be of
  limited value.

- Choleski Decomposition and back substitution: There are a number of varieties
  of Choleski decomposition, but they all take advantage of the symmetric and
  non-negative definite structure of $A$. A particularly compact form was used
  in @Nash1987a, though it is debatable whether the saving in storage was not
  heavily outweighed by the complexity of the code and its documentation.

- Eigendecompositions of the sum of squares and cross-products (SSCP) 
  matrix: the eigenvalues of real, symmetric matrices
  such as $A$ can be computed in many ways and the eigendecomposition used to 
  solve the Gauss-Newton equations. The computations are much more costly than
  simple solution of the equations, but the knowledge of the eigenvalues allows
  us to avoid, or at least be aware of, conditioning issues.

### Solution of the least squares sub-problem by matrix decomposition

Here we proposed to solve using the matrix $J$ or an augmented form $K$
directly. There are a number of approaches that can be used, which we
will divide into two classes, QR and SVD methods.

The QR approach aims to decompose the working matrix (call it $K$,
with $m$ rows and $n$ columns)

$$ K = Q R $$

where $Q$ is $m$ by $n$ orthogonal so

$$ Q^t Q = I_n $$

and $R$ is upper triangular and $n$ by $n$. Some variants have $Q$ an $m$ by $m$ orthogonal
matrix and $R$ upper triangular with $m - n$ rows of zeros appended below.

There are many ways of computing such decompostions:

- Householder reflections with or without pivoting options
- Givens plane rotations, similarly
- Gram-Schmidt orthogonalization of the columns of $K$, either directly (one
  column at a time orhogonalized to all previous columns) or in a modified
  fashion that proceeds row-wise. There are a number of detailed choices.

SVD approaches compute a decompostion

$$ K =  U S V^t$$

where U and V have various orthogonality properties and S is a diagonal matrix
with elements called the singular values (the square roots of the eigenvalues
of $K^t K$ actually). There are many variations depending on whether $U$ is 
a full $m$ by $m$ matrix, limited to $m$ by $n$ or even truncated to $m$ by $r$ where 
$r$ is the rank of $K$.

A detail in choosing our approach is the mechanism of inclusion of the
Marquardt (or Marquardt Nash) stabilization. It turns out to be remarkably
easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
has elements $\sqrt {(\lambda)}$ times the diagonal stabilization elements
we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
of the Gauss-Newton.

As we explain below, we do, unfortunately, realize that Marquardt suggested a scaled 
stabiilzation where we want to add the diagonals of $J^t J$ normal equation approach.
This requires that we perform at least a partial computation of the SSCP matrix,
which is relatively costly. For this reason we plan ?? to explore some proxies
for these elements.

### Avoiding re-computation of the Jacobian

The "original" Gauss-Newton iteration offers no guidance in how to deal with a
situation where $x_{new}$ has a poorer fit than $x$. @Hartley61 and @Marquardt1963,
along with most other practical methods, need to recompute the (weighted) residuals
and consequent loss function (sum of squares or deviance) for each value of a step-length
or stabilization parameter ($\lambda$) until either a "better" point $x_{new}$ is found
or we terminate the search.

Thus it is useful to be able to separate the computation of residuals from that of the
Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
historical reasons. This is the setup for both the finite difference approximations of
`nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
as well as the `rjfun()` generated by `nlsr::model2rjfun()`. Nevertheless, it is likley worth 
avoiding duplication if possible, though the computation of Jacobian elements uses a lot of the
same steps as the computation of the residuals.

?? need to explore timings

Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
situations:

- saving residual values when computing derivatives via finite differences
- separation of the Jacobian routine completely from that of the residuals
- other measures?

## Storage stucture

If the choice of approach to Gauss-Newton or Marquardt is to build the normal
equations and hence the sum of squares and cross products (SSCP) matrix, we
know by construnction that this is a symmetric matrix and also non-negative definite.
(Generally, unless we have a strict linear dependence between the columns of
the Jacobian, we can say "positive definite".)
If we use an SSCP form, we can apply algorithms that specifically take advantage of both these
properties, including programs that store the working matrix in special formats. 
Some of these are Algorithms 7, 8 and 9 of Nashlib (@jncnm79). Algorithms 7 and 8 are the
Cholesky decomposition and back-solution using a vector of length `n*(n+1)/2`
to store just the lower triangle of the SSCP matrix. Algorithm 9 inverts this
matrix *in situ*.

The original @jncnm79 Algorithm 23 (Marquardt nonlinear least squares solution) computes
the SSCP matrix $J^t J$  and solves the Marquardt-Nash augmented normal equations
with the Cholesky approach. This was continued in the Second Edition @nlacatvn1060620 and
in @jnmws87. However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
has been to use a QR decomposition of an augmented Jacobian. `nls()` uses a QR decomposition
of the Jacobian without augmentation (referred to as the "gradient", however).

The particular QR calculations in these packages and in R-base are quite well-hidden in internal
code, complicating comparisons of storage, complexity and performance. The present exposition
is an attempt to provide some documentation.

### Storage in Nash/Walker-Smith (1987) BASIC code

In @Nash1987a, each major iteration generates a $A = J^t J$ matrix and the right hand side $-g = J^t r$
at parameters $x$ using functions for the residuals and Jacobian.
The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements (row-wise ordering of $C$). 
The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ is added to each,
where $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in @jn77ima to
overcome potential underflow to zero of the accumulation of the diagonal elements in low-precision
arithmetic. Let us call the augmented SSCP matrix $C$.

Solving $C \delta = -g$, we compute the new sum of squares at $x + \delta$. If this value is greater than
or equal to the current best (i.e., lowest) sum of squares, we increase $\lambda$ and generate a new
$C$ matrix. Note that we do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
adjustment.

The code in @Nash1987a is much more complicated than this description suggests because of

- bounds and masks on the parameters

- tests for "non-computability", where we allow the user to provide an explicit flag that the function cannot
  be computed at the trial parameters. This was included to avoid some of the weaknesses in evaluation of 
  special functions, e.g., `log()` for small arguments or `exp()` for large ones, or for problem specific
  situations where some parameter combinations might be considered inadmissible.

- situations where rounding and truncation errors make the Cholesky decomposition return with an indication
  that $C$ is "singular", at least in the particular computational arithmetic.

### Approach in package `nlsr`

In @nlsr-manual, the approach to the augmented Gauss-Newton equations is nominally equivalent to that in @Nash1987a,
except that we perform the solution using a QR decomposition, in fact a modification of the DQRDC routine from 
@lapack99, which uses Householder transformations.

One of the issues with creating the SSCP matrix $A$ is that we lose information in finite arithmetic 
(See @jncnm79, Chapter 5). A QR approach to the traditional Gauss-Newton codes is to solve

$$  J \delta = -r $$

That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.

To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Let us call this augmented matrix
$K$. We will also need to augment $r$ by an appropriate number of zeros. For the traditional Marquardt method, we
need to use the square roots of the diagonal elements of $J^t J$, that is, an $n$ by $n$ 
diagonal matrix $Z1$ where

$$ Z1_{k,k}  = \sqrt(\lambda) \sqrt(J^t J)_{k,k}  $$

and $r$ is extended by $n$ zeros. For the Nash modification, we add a further $n$ zeros to $r$ and augment
the $K$ matrix with another $n$ by $n$ block $Z2$ which is a diagonal matrix where the diagonal elements are

$$ Z2_{k,k}  = \sqrt(\lambda) \sqrt(\phi)  $$

Unfortunately, there is a need in each major iteration to compute at least the diagonal elements of the
SSCP matrix. It may, in fact, be more efficient to omit the $Z1$ augmentation. Package `nlsralt`, as
mentioned, does this with functions `nlxbx()` (for formula-specified models) and `nlfbx()` (for problems
specified with `res()` and `jac()` functions). This gives the following preliminary example using the 
Hobbs weed problem.

<!-- https://bookdown.org/yihui/rmarkdown-cookbook/source-script.html -->
<!-- ```{r, include=FALSE} -->
<!-- source("your-script.R", local = knitr::knit_global()) -->
<!-- # or sys.source("your-script.R", envir = knitr::knit_global()) -->
<!-- ``` -->

??FIXME??

```{r hobbtime1}
source("HobbsTiming.R", echo=TRUE)
```

### Other storage approaches.

?? do we want to report any other approaches here?
In particular, should we look at minpack.lm to see how the Marquardt stabilization 
is applied? This is a very complicated program, since it calls C to call Fortran,
and the Fortran is layered in a way that each part is well-documented but the
general flow and linkage of the parts is non-obvious.

## Sequential or full Jacobian computation

We could compute a row of the Jacobian plus the corresponding residual element
and process this before computing the next row etc. This means the full Jacobian
does not have to be stored. In Nashlib, Algorithms 3 and 4, we used row-wise data 
entry in linear least squares via Givens' triangularization (QR decompostition),
with the possibility of extending the QR to a singular value decomposition.
Forming the SSCP matrix can also be generated row-wise as well. We do not 
anticipate that any of these approaches will offer advantages that are so
great that we would choose them.

## Analytic or approximate Jacobian

`nls()` and `minpack.lm` use finite difference approximations to compute the Jacobian.
In the case of `minpack.lm` it seems that the standard `numericDeriv()` (from the 
set of routines in file `nls.R`) is used initially for the "gradient", but within
the iteration, the Fortran routine lmdif.f computes its own approximations for
the Jacobian, adding another layer of potential confusion.

The `nlsr` package, however, attempts to compute analytic derivatives using symbolic
and analytic derivative possibilities from other parts of R. A great advantage of
these is that they are for a given point, and do not suffer the issue that a step
must be taken, thereby chancing the violation of a constraint or else an inadmissible
set of arguments for special functions. In my experience, the use of analytic derivatives
is not much faster than the finite difference approximations, and generally finite
difference methods iterate as quickly as the "exact" derivative methods to the 
neighbourhood of a solution. It is in allowing good estimates of convergence 
criteria that the analytic derivatives seem to show their advantage i.e., knowing
we are at a solution rather than the speed in getting there.

## Interfacing to problems

R allows the nonlinear least squares problem to be presented via a formula
for the model. This is generally very attractive to users, since they can
specify their problem in ways that often accord with how they think about them.
Many (most?) researchers do not write program code, so preparing residual and
Jacobian functions is foreign to them.

From the point of view of package or system developers, however, specifications as
formulas pose some particular issues:

- translating a formula to a function for its actual evaluation may be tricky,
  since we must essentially parse the formula. R has some tools for this that
  we use.

- even if we can translate a formula to a function, we need to check that we
  have all the data and parameter inputs to enable the evaluation. Moreover,
  we need to check if the inputs are valid or admissible.

- checking for correctness of inputs and translations can be considered to
  take up most of the code in a reliable and robust package.

# Saving storage

The obvious ways to reduce storage are:

- use a row-wise generation of the Jacobian in either a Givens' QR or SSCP approach.
  This saves space for the Jacobian as well as as well as the working matrices of
  the Gauss-Newton or Marquardt iterations;

- if the number of parameters to estimate is large enough, then a normal equations
  approach using a compact storage of the lower triangle of the SSCP matrix. 
  However, the scale of the saving is really very small in comparison to 
  the size of most programs.

# Measuring performance

Do we plan to use only timing? Or size of code and memory used? ??

# Test problems

```{r settest15}
# set parameters
set.seed(123456)
a <- 1
b <- 2
c <- 3
np <- 15
tt <-1:np
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
plot.new()
plot(tt, yy, type='l')
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
points(tt,y1,type='p', col="green")
y2 <- yy + 5*ev
points(tt,y2,type='p', col="blue")
y3 <- yy + 10*ev
lg3d15 <- data.frame(tt, yy, y1, y2, y3)
points(tt,y3,type='p', col="red")
library(nlsr)
sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
print(sol0)
sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
print(sol1)
sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
print(sol2)
sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
print(sol3)
```

The following is a larger dataset version of this test.

```{r settest150, eval=FALSE}
np <- 150
tt <- (1:np)/10
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
y2 <- yy + 5*ev
y3 <- yy + 10*ev
lg3d150 <- data.frame(tt, yy, y1, y2, y3)
np <- 1500
tt <- (1:np)/100
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
y2 <- yy + 5*ev
y3 <- yy + 10*ev
lg3d1500 <- data.frame(tt, yy, y1, y2, y3)
f0 <- yy ~ a0/(1+b0*exp(-c0*tt))
f1 <- y1 ~ a1/(1+b1*exp(-c1*tt))
f2 <- y2 ~ a2/(2+b2*exp(-c2*tt))
f3 <- y3 ~ a3/(3+b3*exp(-c3*tt))
```

# Implementation comparisons

Here want to explore the ideas. First we will examine the sub-problem of solving
the Gauss-Newton equations or their Marquardt variants.

## Linear least squares and storage considerations

Without going into too many details, we will present the linear least squares 
problem as

$$  A x \tilde = b  $$

In this case $A$ is an $m$ by $n$ matrix with $m \ge n$ and $b$ a vector of lenght $m$.
We write **residuals** as

$$  r  =  A x - b  $$
or as

$$  r_1  = b - A x  $$

Then we wish to minimize the sum of squares $r^t r$. This problem does not necessarily
have a unique solution, but the **minimal length least squares solution** which is 
the $x$ that has the smallest $x' x$ that also minimizes $r' r$ is unique.

### Example setup and run in lm()

Let us set up a simple problem in R:

```{r ls1, echo=TRUE}
# simple linear least squares examples
v <- 1:6
v2 <- v^2
vx <- v+5
one <- rep(1,6)
Ad <- data.frame(one, v, v2)
A <- as.matrix(Ad)
print(A)
Ax <- as.matrix(data.frame(one, v, vx, v2))
print(Ax)
y <- -3 + v + v2
print(y)
set.seed(12345)
ee <- rnorm(6)
ee <- ee - mean(ee)
ye <- y + 0.5*ee
print(ye)
sol1 <- lm.fit(A, y)
print(sol1)
cat("Residual SS=",as.numeric(crossprod(sol1$residuals)),"\n")
sol1e <- lm.fit(A,ye)
print(sol1e)
crossprod(sol1e$residuals)
sol2<-lm.fit(Ax,y)
# Note the NA in the coefficients -- Ax is effectively singular
print(sol2)
S2 <- sol2$coefficients
J <- which(is.na(S2))
S2[J]<-0
crossprod(S2)
```

The above uses the intermediate code in function `lm.fit()`. This uses a QR solver, but
it is written as a wrapper in C calling a Fortran routine (in the Fortran 77 dialect).

?? Should we put in the structure and where the code is located?

I believe that the structure of lm() predates the availability of the family
of `qr.xxx()` functions for R. These allow us to access the QR approach 
directly.

```{r ls2, echo=TRUE}
x <- qr.solve(A,y)
print(x)
xe<-qr.solve(A, ye)
print(xe)
# But then we get an error when we try to solve the singular system
# This was NOT caught above.
xx <- try(qr.solve(Ax,y))
print(xx)
xxe<-try(qr.solve(Ax, ye))
print(xxe)
```

## Traditional Normal equations approach

The historically traditional method for solving the **linear** least squares problem was
to apply calculus to set the partial derivatives of the sum of squares with respect to
each parameter to zero. This forms the **normal equations**

$$ A^t A x  = A^t b $$

This was attractive to early computational workers, since while $A$ is $m$ by $n$, $A^t A$
is only $n$ by $n$. Unfortunately, this **sum of squares and cross-products** (SSCP) matrix
can make the solution less reliable, and this is discussed with examples in @jncnm79 and
@nlacatvn1060620.

```{r lm5, echo=TRUE}
AtA <- t(A)%*%A
print(AtA)
Aty <- t(A)%*%y
print(t(Aty))
x<-solve(AtA,Aty)
print(t(x))
```

Let us try this with the extended matrix `Ax`

```{r lm6, echo=TRUE}
AxA <- t(Ax)%*%Ax
print(AxA)
Axy <- t(Ax)%*%y
print(t(Axy))
xx<-try(solve(AxA,Axy))
print(t(xx))
```

### Dealing with singularity

The `lm.fit()` function has a parameter `singular.ok` which defaults to TRUE.
By setting this to FALSE, we get.

```{r lm3, echo=TRUE}
sol2b<-try(lm.fit(Ax,y, singular.ok=FALSE))
print(sol2b)
```

We've already seen above that the solution `sol2` has a size of

```{r num1}
eval(as.numeric(crossprod(S2)))
```

The solution `sol2` is, it turns out, not unique, since the variables `v` and
`vx` and `one` are related, namely

```{r num2, echo=TRUE}
print(vx - (v+5*one))
```
Thus we can find a "new" solution as follows and show (essentially) the same 
residuals

```{r num3, echo=TRUE}
res2<-Ax%*%S2-y
print(t(res2))
print(S2)
S2b<-S2+c(10,2,-2,0)
res2b<-Ax%*%S2b-y
print(t(res2b))
cat("Sum of squares of S2b=", as.numeric(crossprod(S2b)),"\n")
```

### Approximate solution -- Ridge regression

An approximate solution via the normal equations can be found by adding a small 
diagonal matrix to the sum of squares and
cross products `AxA`.

```{r lm7, echo=TRUE}
AxA <- AxA + diag(rep(1e-10,4))
print(AxA)
xxx<-try(solve(AxA,Axy))
print(t(xxx))
print(t(Ax %*% xxx - y))
```

We note that this solution is rather different from `S2` or `S2b`.
There is a large (and often misguided and confusing) literature 
about this sort of approach under the title **ridge regression**.
Note that the Marquardt stabilization of the Gauss-Newton equations
uses the same general idea.

### Dealing with singularity in QR solutions

We already saw that we got errors when trying to use `qr.solve()`, but there
are ways to use the QR decomposition that overcome the singularity. Here
are some illustrations. Note that the size of the solution as measured by
the sum of squares of the coefficients (ignoring the NA usng `na.rm=TRUE`)
are different. Moreover, they are different from the minimum length solution
in the next subsection.

```{r qrsing1, echo=TRUE}
## This fails
xx <- try(qr.solve(Ax,y))
print(xx)
## So does this
xxe<-try(qr.solve(Ax, ye))
print(xxe)
## Let us compute a QR decomposition, using LINPACK then LAPACK
qrd1<-qr(Ax, LAPACK=FALSE)
qrd2<-qr(Ax, LAPACK=TRUE)
# and get the solutions
xx1 <- try(qr.coef(qrd1,y))
print(xx1)
xx2 <- try(qr.coef(qrd2,y))
print(xx2)
# and computer the sum of squares of the coefficients (size of solution)
try(sum(xx1^2))
try(sum(xx1^2, na.rm=TRUE))
try(sum(xx2^2))
```
### Minimum length least squares solution

The minimum length least squares solution, which is unique, 
is found using the Moore-Penrose inverse of `Ax`
(the article https://en.wikipedia.org/wiki/Moore%E2%80%93Penrose_inverse gives a quite good
overview) which can be computed from the Singular Value Decomposition. R has a function `svd()`
which is adequate to our needs here.

```{r lm4, echo=TRUE}
Z<-svd(Ax) # get the svd
D1 <- 1/Z$d # invert S, the singular values diagonal matrix
print(D1) 
D1[4]<-0 # to remove linear dependency (small singvals are set to zero)
# D1 is diagonal of S+ now
# minimum length LS solution A+ = V S+ t(U)
minsol2 <- Z$v %*% (diag(D1) %*% (t(Z$u) %*% y))
print(minsol2)
cat("SS of minsol2=",as.numeric(crossprod(minsol2)),"\n")
resminsol<-Ax%*%minsol2-y
print(t(resminsol))
```

### QR decomposition variants

There are a number of ways to compute a QR decomposition.

- Householder transformations have good vector-computation performance and properties.
  They use matrix transformations that can be called **reflections** to be used to
  build the Q matrix as a product of these transformation matrices as the initial
  rectangular matrix $A$ is rendered upper triangular, or $R$

- Givens transformations (@Bindel2002Givens) are **plane rotations** that zero one element at a time of
  the current working matrix that starts out as $A$. Once again, $Q$ is a product of
  the transformations and $R$ the result of zeroing out sub-diagonal elements.

- The Gram-Schmidt method, which has several forms, builds the Q matrix by 
  subtracting a multiple of the first column of $A$ from the second and then 
  normalizing those columns so their sum of squares is 1. The process is then 
  applied to the third column by subtracting multiples of the first two. The $R$
  matrix is formed from the norms and subtraction multipliers in the process. 
  There are row- and column-wise approaches. 
  See https://en.wikipedia.org/wiki/Gram%E2%80%93Schmidt_process

### Using a QR approach

Let us try to form a QR decomposition of $A$, for example with Givens rotations.

$$ A =  Q R$$

where $Q$ is orthogonal (by construction for plane rotations) and $R$ is upper triangular.
We can rewrite our original form of the least squares problem as

$$  Q^t A = Q^t Q R = R \tilde= Q^t b$$

$R$ is an upper triangular matrix $R_n$ stacked on an $m-n$ by $n$ matrix of zeros. But 
$z = Q^t b$ can be thought of as $n$-vector $z_1$ stacked on ($m-n$)-vector $z_2$. 
It can easily be shown (we won't do so here) that a least squares solution is the
rather easily found (by back-substitution) solution of

$$  R_n x = z_1$$

and the minimal sum of squares turns out to be the cross-product $z_2^t z_2$. Sometimes
the elements of $z_2$ are called **uncorrelated residuals**. The solution for $x$ can
actually be formed in the space used to store $z_1$ as a further storage saving,
since back-substitution forms the elements of $x$ in reverse order.

All this is very nice, but how can we use the ideas to both avoid forming the SSCP
matrix and keep our storage requirements low?

Let us think of the row-wise application of the Givens transformations, and use a 
working array that is $n+1$ by $n+1$. (We can actually add more columns if we have
more than one $b$ vector.)

Suppose we put the first $n+1$ rows of a merged $A | b$ working matrix into this
storage and apply the row-wise Givens transformations until we have an $n$ by $n$
upper triangular matrix in the first $n$ rows and columns of our working array.
We further want row $n+1$ to have $n$ zeros (which is possible by simple transformations)
and a single number in the $n+1$, $n+1$ position. This is the first element of $z_2$.
We can write it out to external storage if was want to have it available, or else
we can begin to accumulate the sum of squares.

We then put row $n+2$ of [$A | b$] into the bottom row of our working storage and
eliminate the first $n$ columns of this row with Givens transformations. This
gives us another element of $z_2$. Repeat until all the data has been processed.

We can at this point solve for $x$. Algorithm 4 of @jncnm79, however, applies the one-sided
Jacobi method to get a singular value decomposition of $A$ allowing of a minimal
length least squares solution as well as some useful diagnostic information about
the condition of our problem. This was also published as @LefkovitchNash1976.

### QR approach to the Marquardt stabilization

We can arrange to solve the normal equations approximately as under Ridge Regression
above by adding a unit matrix scaled by e.g., 1e-5, and padding the RHS with zeros.
The solution is "almost but not quite" the same because the formation of the sum of squares and
cross products matrix explicitly in the normal equations can introduce errors that
are partially avoidable in the QR approach below.

```{r lm7a, echo=TRUE}
Dx <- diag(rep(1e-5,4))
print(Dx)
Ap<-rbind(Ax, Dx)
print(Ap)
## we need to pad y with zeros
yp<-c(y, rep(0,4))
xxp<-qr.solve(Ap,yp)
print(xxp)
## Previous solution via normal equations
print(as.vector(xxx))
```

### Choices when solving the normal equations

Whether the normal equations arise from the original Gauss-Newton or Marquardt
approaches to nonlinear least squares, we have several choices of how to solve
them.

Traditional solutions to the solution of linear equations used elimination 
techniques under names such as **Gaussian Elimination**, **Gauss-Doolittle**,
**Crout's Method**, **Gauss-Jordan** and many others. All these can be considered
as LU decompositions. That is, they can be expressed as a transformation of the
matrix $A = J^t J$ or its Marquardt stabilized variant into a product of a lower
triangular and an upper triangular matrix, possibly with permutations of rows and/or
columns. Row permutations of A give

$$ P A = L U$$

The origin of our matrix $A$ means that it is at least non-negative definite. 
This means that we can form a special LU decomposition that does NOT need
permutations and has

$$ A = L L^t$$

which is called the Choleski decomposition. In practice, we need $A$ computationally
non-singular (not just theoretically so), but the Marquardt stabilization can be used
to ensure this.

If we are very worried about storage space, we can note that $A$ is symmetric, so
that we only need to store the lower (or upper) triangle, including diagonals. 
And some bright programmers realized we could do our storage in a single array,
so that a matrix of order $n$ could be stored in a vector of length $n (n+1)/2$.
@Nash1987a uses this Cholesky decomposition structure in BASIC code.

Similarly, a Gauss Jordan variant could be programmed using such a format for
$A$ and with only a temporary storage of size $n$, matrix $A$ could be overwritten
with its inverse. This is the code of @BauerReinsch71 which is Algorithm 9 of
@jncnm79.

Both these "single vector" storage programs are challenging to check and maintain.
While I embraced them in the age when 16K of storage was the norm for personal
computers, they are simply too finicky to follow easily, and highly prone to
coding and maintenance errors.

# Implementation choices for an R package

??2021 update to nlsr and ideas for the "Improvements to nls()" project.

# References

### 2021-12-18 DerivsNLS/badJlogmod.R

@@ -1,0 +1,51 @@
+# File: badJ2.R
+# A problem illustrating poor numeric Jacobian ?? 20211217 does not fail??
+form<-y ~ 10*a*(8+b*log(1-0.049*c*x)) # the model formula
+# This model uses log near a small argument, which skirts the dangerous
+# value of 0. The parameters a, b, c could all be 1 "safely" as a start.
+x<-3*(1:10) # define x
+np<-length(x)
+a<-1.01
+b<-.9
+eps<-1e-6
+c<-1/(max(x)*.049)-eps
+c<-0.68
+cat("c =",c,"\n")
+yexact<-10*a*(8+b*log(1-0.049*c*x))
+y <- yexact+0.2*runif(np) # compute a y
+# y <- yexact+0.1*runif(np) # compute a y
+df<-data.frame(x=x, y=y)
+plot(x,y) # for information
+st<-c(a=1, b=1,c=c) # set the "default" starting vector
+n0<-try(nls(form, start=st, data=df)) # and watch the fun as this fails. ??DOES NOT FAIL!
+summary(n0)
+library(nlsr) # but this will work
+n1<-nlxb(form, start=st, data=df)
+n1
+coef(n1)-coef(n0)
+jmod<-model2rjfun(form, pvec=st,data=data.frame(x=x, y=y)) # extract the model
+Jatst<-jmod(st) # compute this at the start from package nlsr
+Jatst<-attr(Jatst,"gradient") # and extract the Jacobian
+#
+# Now try to compute Jacobian produced by nls()
+env<-environment(form) # We need the environment of the formula
+eform<-eval(form, envir=env) # and the evaluated expression
+localdata<-list2env(as.list(st), parent=env)
+jnlsatst<-numericDeriv(form[[3L]], theta=names(st), rho=localdata)
+Jnls<-attr(jnlsatst,"gradient")
+Jnls # from nls()
+Jatst # from nlsr -- analytic derivative
+max(abs(Jnls-Jatst))
+svd(Jnls)$d
+svd(Jatst)$d
+# Even start at the solution?
+n0c<-try(nls(form, start=coef(n1), data=data.frame(x=x, y=y), trace=TRUE))
+summary(n0c)
+## attempts with nlsj 
+library(nlsj)
+n0jn<-try(nlsj(form, start=st, data=df, trace=TRUE,control=nlsj.control(derivmeth="numericDeriv")))
+summary(n0jn)
[…]

### 2021-12-18 DerivsNLS/nlsnd.R

@@ -1,0 +1,79 @@
+#  File src/library/stats/R/nlsnd.R
+#  Part of the modified R package, https://www.R-project.org
+#
+#  Copyright (C) 2000-2020 The R Core Team
+#  Copyright (C) 1999-1999 Saikat DebRoy, Douglas M. Bates, Jose C. Pinheiro
+#  J C Nash 2021
+#
+#  This program is free software; you can redistribute it and/or modify
+#  it under the terms of the GNU General Public License as published by
+#  the Free Software Foundation; either version 2 of the License, or
+#  (at your option) any later version.
+#
+#  This program is distributed in the hope that it will be useful,
+#  but WITHOUT ANY WARRANTY; without even the implied warranty of
+#  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+#  GNU General Public License for more details.
+#
+#  A copy of the GNU General Public License is available at
+#  https://www.R-project.org/Licenses/
+
+###
+###            numeric Jacobian for Nonlinear least squares for R
+###
+
+numericDeriv <- function(expr, theta, rho = parent.frame(), dir = 1,
+                 eps = .Machine$double.eps ^ (1/if(central) 3 else 2), central = FALSE)
+## Note: this expr must be set up as a call to work properly according to JN??
+## ?? we set eps conditional on central. But central set AFTER eps. Is this OK.
+{   ndtrace<-FALSE
+#    if(ndtrace) cat("numericDeriv-Alt\n")
+    dir <- rep_len(dir, length(theta))
+    stopifnot(is.finite(eps), eps > 0)
+    rho1 <- new.env(FALSE, rho, 0)
+    if (!is.character(theta) ) {stop("'theta' should be of type character")}
+    if (is.null(rho)) {
+            stop("use of NULL environment is defunct")
+            #        rho <- R_BaseEnv;
+    } else {
+          if(! is.environment(rho)) {stop("'rho' should be an environment")}
+          #    int nprot = 3;
+    }
+    if( ! ((length(dir) == length(theta) ) & (is.numeric(dir) ) ) )
+              {stop("'dir' is not a numeric vector of the correct length") }
+    if(is.na(central)) { stop("'central' is NA, but must be TRUE or FALSE") }
+    res0 <- eval(expr, rho) # the base residuals. ?? C has a check for REAL ANS=res0
[…]

### 2021-12-18 MachineSummary/MachineSummary.Rmd
Add machine summary and pkgfromRbase

---
title: "A Machine Profile Summary"
author: 
  - John C. Nash, University of Ottawa, Canada
date: "14/06/2021"
bibliography: ../BibSupport/ImproveNLS.bib
output: 
  pdf_document:
    keep_tex: false
---

<!-- date: "`r format(Sys.time(), '%d %B, %Y')`" -->

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Providing a characterization of a particular computing environment for R calculations

I have looked at this.

If we want to compare results computed on different machines, we need a way to provide 
measures and identifiers of the particular computing environment at hand. This document
attempts to summarize some possibilities available in the R statistical software and language.

# Some possible information desired

There are multiple ways to get information of the type useful for characterizing the
computing infrastructure used for a particular test or timing. We will need to
ensure that identifiers of particular pieces of information are unique, and that 
common identifiers really do specify identical information. Otherwise, we need
to develop new names for our set of information.

## Temporal information

Most computatins should have a time/date stamp that is unlikely to be confused with any
other, at least in combination with a machine name and/or other tags.

```{r tdstamp}
#get a timestamp
tsstr<- format(Sys.time(), "%Y%m%d%H%M") # tsstr == time stamp string in form YYYYmmddHHMM
cat("Date and Time stamp:",tsstr,"\n")
```
## Identifier for the computing environment

Some of the important information elements concerning the computing environment that 
we need for reporting tests are

- machine name. While most systems (and Linux in particular) offer to let the user
  provide a machine name, there are generally defaults that many people accept, and
  these are often uninformative and may be non-unique. We note that VirtualBox with
  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
  generated we are not sure. Settings / System allows "rename this PC".

- operating system and version

- compiler or interpreter version. For our needs, it is obvious that the R version
  will be important. However, if any code is compiled or linked to libraries, we
  would like to know that. In particular, versions of compilers (e.g., gfortran, 
  gcc) or BLAS or LAPACK libraries used will affect performance. Linux generally
  displays the BLAS and LAPACK **filenames** in `sessionInfo()`, but to get the
  version information, one needs to dig deeper, for example, using operating
  system commands. For the present, we will omit such extra information unless
  we can extract it easily within R.

- specific variations, if any, that should be noted. Here we may want to note if
  a machine is running background tasks, or if some special steps have been 
  taken to speed up or slow down the operation e.g., overclocking.

## Hardware information

Some of the factors relating to hardware that could be important are:

- cpu
- cores
- operating cycle speed (While the cpu is generally specified )
- RAM total
- RAM available, possibly indicating swap space and usage
- RAM speed (could be variable)

## Software information

?? what to put in.

## R tools for machine information

### sessionInfo()

```{r sessionInfox}
si <- sessionInfo()
si <- as.vector(si)
si
```

### benchmarkme

- uses `proc.time()` for timings
- has single core and parallel timings
- ?? does it do GPUs
- provides far too much for most needs i.e., a succinct summary of
  how capable a computing system should be

```{r benchmarkmex}
library(benchmarkmeData)
library(benchmarkme)
ls(package:benchmarkmeData)
ls(package:benchmarkme)
lsf.str("package:benchmarkme")
lsf.str("package:benchmarkmeData")
## This next line takes a lot of time to run, so is commented out here
# benchmark_std()
get_byte_compiler()
gla<-get_linear_algebra()
gsd<-get_sys_details()
```

?? get_ram may not work in Windows 10.

### calceps.R

This is a port of Mike Malcolm's ENVRON to R. ??ref. It computes the 
machine precision, radix and number of radix digits.

?? get file into chunk

```
calceps.R

- calceps.R example

### Sys.info()

```{r sysinfox}
Sys.info()
```

## Issues relateing to RNG

## Issues relating to compilation of R

- LAPACK
- BLAS
- others?

# Tools for accessing and clearing environments and dataframes

?? should this be here -- probably to list things in the workspace

```
sys.frame()
sys.frames()
sys.status()
sys.on.exit()
sys.parents()
sys.calls()
```

??JN exp(sin(cos())) benchmark

# Choices

For use in recording tests of R functions and packages for optimization and nonlinear 
least squares, it seems that the `benchmarkme` function `get_sys_details()`
provides more than sufficient information for our needs.

From the above discussion, the following offers a possible compact solution.

```{r code=xfun::read_utf8('MachID.R')}
```

Note that we are reading from the file `MachID.R` so that this file is usable elsewhere
and the current document does not become out of sync with the working MachID.

### 2021-12-18 MachineSummary/MachineSummary.Rmd
tidy authorship

@@
 author: 
-  - John C. Nash, University of Ottawa, Canada
+   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
+   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
 date: "14/06/2021"

### 2021-12-18 PkgFromRbase/PkgFromRbase.Rmd
tidy authorship

---
title: "Making a package from base R files"
author: 
   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
   - based on contribution from Duncan Murdoch
date: "11/06/2021"
output:
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Background

This article tries to explain an approach to developing alternative versions of
functions which are in the distributed base of R. Our interest was in developing
improvements to the nls() function and related features in R as part of a Google
Summer of Code project for which Arkajyoti Bhattacharjee is the funded student.
However, nls() has many tentacles involving a number of files and functions that
may or may not be called as `nls()` is executed.

Part of the difficulty in carrying out such development of alternative versions
is that one needs to be able to execute the new variants in parallel with the
existing ones. A heavy-effort approach would be to have separate full sets of
R code and build each system and run them separately. That is, we could have
two versions of R.

Attempts to provide R functions to switch between the sets of working functions
-- e.g., `ReplaceNLS.R` and `RestoreNLS.R` --
were not promising, and not pursued.

Duncan Murdoch suggested, and generously provided an almost-complete prototype of,
a package of interlocked functions that provide the set of `nls()` capabilities.
This package `nlspkg` has been cloned to `nlsalt` which is being modified
to provide the new capabilities or new expression of the code. Moreover, these
capabilities, such as a function `Anyfn()` can be tested side by side by use
of the double colon syntax, namely,

```
nlspkg::Anyfn() # existing functionality or code
# and
nlsalt::Anyfn() # New functionality of code
```

# Packaging from base R files

Duncan Murdoch described his process of developing the prototype of 
`nlspkg`:

```
Identify the main R functions I wanted in the package, and copy that
code into ./R.

Identify the C/Fortran functions that are referenced, and copy the
source for that into ./src.

Try to build, and get lots of errors about missing functions. Find
those and copy as above.

There was one case (the `%||%` operator) where I thought the code used
an unexported object from utils; the base-internals.R function just used
::: to get it, but a better solution would be to copy the source (which
is pretty trivial), and it turns out, is what stats did (the duplicate
source is in AIC.R).

Some files contained functions or methods that weren't relevant to us
(e.g. confint.lm).  I commented those out so I didn't need to worry
about the functions they used, and imported the generic (i.e. confint)
from stats.

In the C code, we were lucky because the nls code only made use of one C
function that's not in the API:  R_NewEnv. (That was called from
numeric_deriv.) There's an R function to do that, so I called it and
passed the result in.

Finally, once it would build, I tried running R CMD check.  This
identified all the missing help pages, which I copied over.  They didn't
need any modifications.

Because nls isn't really base functionality, all of this was a lot
easier than if I had tried to move something that works more with the
internals.  That's probably true for most functions in stats, but would
not be true for other base packages (except datasets).
```

After I downloaded the ZIP file of Duncan's package from his Github repository, 
I built and checked the package. There were two WARNING messages:

```
* checking DESCRIPTION meta-information ... WARNING
Non-standard license specification:
  GPL-2, GPL-3
Standardizable: FALSE
* checking for missing documentation entries ... WARNING
Undocumented code objects:
  ‘C_bvalus’ ‘C_lowesp’ ‘C_lowesw’ ‘C_nls_iter’ ‘C_numeric_deriv’
  ‘C_port_ivset’ ‘C_port_nlminb’ ‘C_port_nlsb’ ‘C_pppred’ ‘C_rbart’
  ‘C_setppr’ ‘C_setsmu’ ‘C_smart’ ‘C_supsmu’ ‘anova.nls’
  ‘anovalist.nls’ ‘coef.nls’ ‘confint.nls’ ‘deviance.nls’
  ‘df.residual.nls’ ‘fitted.nls’ ‘formula.nls’ ‘logLik.nls’ ‘nlsModel’
  ‘nlsModel.plinear’ ‘nls_port_fit’ ‘nobs.nls’ ‘plot.profile.nls’
  ‘port_cpos’ ‘port_get_named_v’ ‘port_msg’ ‘port_v_nms’ ‘predict.nls’
  ‘print.nls’ ‘print.summary.nls’ ‘profiler’ ‘profiler.nls’
  ‘residuals.nls’ ‘summary.nls’ ‘vcov.nls’ ‘weights.nls’
All user-level objects in a package should have documentation entries.
See chapter ‘Writing R documentation files’ in the ‘Writing R
Extensions’ manual.
```

The first WARNING can be overcome by changing the License line in the package
DESCRIPTION file from 
```
License: GPL-2, GPL-3
```
to
```
License: GPL-2 | GPL-3
```
as noted in \url{https://www.r-project.org/Licenses/}.

The warnings about documentation come from two sources:

- many of the `nls()` features parallel generic functions such as `residuals`,
`predict`, `coef`, and so on, so are generically documented in R.

- the rest of the "undocumented" functions are C functions (identifiable as
their names have underscore characters in the names). It is these functions
that our "Improvements to nls()" project aims to replace where sensible to
do so with R functions. They may then be documented as needed.

### Name issues -- IMPORTANT

It is worth mentioning that the NAMESPACE file provided by Duncan is an
important part of the packaging. It could be worthwhile to document the
elements in that file needed for a package such as `nlspkg`. In particular,
however, we must note that in renaming the package from `nlspkg` to something
else, the change of name must be mirrored in the line

```
useDynLib(nlspkg, .registration = TRUE, .fixes = "C_")
```

where `nlspkg` in this line must be changed appropriately.

We also noted that the `S3method()` entries are needed to link the generic
functions to similar modeling functions in R. Thus `anova`, `coef`, etc.
are declared as S3 methods.

As a convenience to get the package running, Duncan used the shortcut

`exportPattern("^[[:alpha:]]+")`

to ensure functions were made available to the user. This makes visible all 
the internal C functions, named `C_(something)`. We have suppressed this blanket
declaration and explicitly exported the `nls` and `numericDeric` functions. 
One or two other routines may need to be exported to allow full functioning
of the package.

We note that name change is ALSO required in `src/init.c`
in the line

```
void attribute_visible R_init_nlspkg(DllInfo *dll)
```

For version control, the file `.gitignore` has two entries: one is `*.0` and
the other, initially, was `nlspkg.so`. I changed the latter to `*.so` and this
appears to be satisfactory for general use.

# Developing an alternative set of functions and alternative package

The alternative package is being developed by replacing calls to C functions
with R work-alike functions. A preliminary example that already is set up and
has passed preliminary checks is the function `numericDeriv()` from the 
file `R/nls.R` with called elements from the file `src/nls.c`. Note that some
code needs to be commented out or deleted from the C file(s) and headers
to avoid errors when building and checking the alternative.

To prepare the `nlsalt` package a directory of that name was created and
the contents of the `nlspkg` directory copied over. The DESCRIPTION file
was edited to provide the `nlsalt` name and the result checked with 
build and CHECK.

At this stage, we can begin to substitute for calls to C code and remove that
code once replacement R material is present. Frequent build and CHECK cycles
are helpful.

For Rstudio users, note that each package directory needs its own (name).Rproj
file, created using Rstudio's "File/New Project" etc. dialog. However, the package
directories with their Rproj files can be within the umbrella version control 
project `improvenls` for the Git version control. (We are using a Gitlab
repository.)

# Windows issue

While all the above worked well in a Linux environment, when we tried running
the build and CHECK in a Windows environment, we found we got errors. (I normally
run in Linux Mint 20.1, but Arkajyoti Bhattacharjee uses Windows 10. Each of us
has a VirtualBox VM for the alternative OS to allow for checking of our work.)
There is a load error for functions from the BLAS (Basic Linear Algebra Subroutines)
collection in a 32 bit architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
feature. Using the `CMD.exe` terminal window showed that `R CMD build` worked, but
that `R CMD INSTALL` gave the load errors.

After some exchange of emails with others and with the
R-package-devel mailing list, a Google search using terms suggested in the emails
found the posting

\url{https://stackoverflow.com/questions/42118561/error-in-r-cmd-shlib-compiling-c-code}

This suggested that a file `Makevars.win` is needed in the `src/` directory of the package(s),
and that this file should contain just one line

```
PKG_LIBS = $(LAPACK_LIBS) $(BLAS_LIBS) $(FLIBS)
```

I did, however, note that it is not clearly indicated that the file needs to be in 
`src/`, nor (until revealed by `R CMD check`) that the file must be `Makevars.win` and
not `makevars.win`. There is a lot of material on `Makevars` in the Writing R
Extensions manual, but these points do not seem to be presented.

# Downstream benefit

The approach to replacing functionality in base R suggested here has a follow-on
benefit should some or all of the improvements we hope to be able to make are 
eventually included in R's distribution files. That benefit is that the initial
package can be renamed to, for example, `nlslegacy` so that R users can, should
they need, use legacy capabilities. Indeed, from preliminary timings of some 
functions, it appears that pure R code runs 2-3 times slower than the mixed 
R and C versions. On the other hand, the pure R code in our early cases is much
shorter and, we suspect, much easier to maintain. Furthermore, it simplifies
the packaging of base R for alternative interpreter or compiler versions of the R
language.

# Acknowledgements

It is a pleasure to note the help and encouragement of Duncan Murdoch, Arkajyoti
Bhattacharjee, Dirk Eddelbuettel and Brett Klamer.

### 2022-01-18 MachineSummary/MachineSummary.Rmd
update MachineSummary

--- 2021-12-18 MachineSummary/MachineSummary.Rmd
@@
 author: 
-  - John C. Nash, University of Ottawa, Canada
-date: "14/06/2021"
+   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
+   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
+date: "14/06/2021 revised 18/01/2022"
 bibliography: ../BibSupport/ImproveNLS.bib
@@
 
-I have looked at this.
-
 If we want to compare results computed on different machines, we need a way to provide 
-measures and identifiers of the particular computing environment at hand. This document
+measures and identifiers of the particular computing environment at hand. This write-up
 attempts to summarize some possibilities available in the R statistical software and language.
@@
 There are multiple ways to get information of the type useful for characterizing the
-computing infrastructure used for a particular test or timing. We will need to
+computing infrastructure used for a particular test or timing. However, we need to
 ensure that identifiers of particular pieces of information are unique, and that 
-common identifiers really do specify identical information. Otherwise, we need
-to develop new names for our set of information.
+common identifiers really do specify identical, or at least equivalent, information. 
+Otherwise, new names for some pieces of our set of capability information are needed.
 
@@
 
-Most computatins should have a time/date stamp that is unlikely to be confused with any
-other, at least in combination with a machine name and/or other tags.
+When we wish to analyze performance results, it is extremely helpful to have 
+information on when the calculations were performed. 
+Most computations should have a time/date stamp that is unlikely to be confused with any
+other, at least in combination with a machine name and/or other tags. Here is one
+suggestion based on the R `Sys.time()` function.
 
@@
 ```
-## Identifier for the computing environment
+
+## Identifier for the computing software environment
 
@@
 
-- machine name. While most systems (and Linux in particular) offer to let the user
+- a machine name. While most systems (and Linux in particular) offer to let the user
   provide a machine name, there are generally defaults that many people accept, and
@@
 
-- operating system and version
+- operating system and version. For Linux systems, we note that besides the distribution
+  variant, e.g., Linux Mint 20.3 MATE 64 bit, it is necessary to add the Linux kernel
+  identifier, e.g.,  5.13.0-25-generic, from the `uname` command.
 
@@
 
-- cpu
-- cores
-- operating cycle speed (While the cpu is generally specified )
-- RAM total
-- RAM available, possibly indicating swap space and usage
-- RAM speed (could be variable)
+- cpu (i.e., processor)
+- number of cores
+- operating cycle speed (this is sometimes specific to the cpu specified )
+- RAM size total
+- RAM size available, possibly indicating swap space and usage
+- RAM speed (which could be variable)
+- GPU (i.e, additional processors) information.
 
@@
 
-?? what to put in.
+Clearly it is important to identify which software packages are active, and the
+R `sessionInfo()` command is useful for this. However, there may be additional 
+features that should be mentioned if the user has modified the software in any
+way.
 
-## R tools for machine information
+The use of tools to carry out computations in parallel, using multiple cores or
+GPUs, is clearly an issue. In our opinion, such possibilities are not well supported
+by common R functions, or indeed in other computing languages, though we would very
+much like to hear of tools to provide such information.
 
-### sessionInfo()
+# R tools for machine information
+
+## sessionInfo()
 
@@
 
-### benchmarkme
+## benchmarkme
+
+?? Need to comment on
 
@@
 
-### calceps.R
+## calceps.R
 
+Until the IEEE 754 standard (??ref) became availaible in the 1980s, 
+computers used multiple forms of floating point arithmetic. With
+some ingenuity, it is possible to detect many aspects of the 
+floating point properties.
 This is a port of Mike Malcolm's ENVRON to R. ??ref. It computes the 
@@
 
-?? get file into chunk
+?? mention Kahan's Paranoia
 
+```{r envrontest}
+envron<-function(){
+D1<-as.numeric(1) # use variable to avoid constants when double precision invoked
+E5 <- 10 #arbitrary scaling for additive equality tests
+B9 <- 1E+35 # big number, not necessarily biggest possible
+E6 <- 1 # initial value for radix 
+E9 <- 1 # initial value for machine precision
+D0 <- E6 
+repeat{
+  E9 <- E9/2 #start of loop to decrease estimated machine precision
+  D0 <- E6+E9 #force storage of sum into a floating-point scalar
+  if (D0 <= E6) break
+} # repeat reduction while (1+E9) > 1
+E9 <- E9*2 # restore smallest E9 which gives (1+E9) > 1
+repeat {
+   E6 <- E6+1 # try different radix values
+   D0 <- E6+E9
+   if (D0 <= E6) break
+} # until a shift is observed
+J1 <- 1 # initial count of radix digits in mantissa
+E9 <- 1 # use radix for exact machine precision
+repeat {  
+  E9 <- E9/E6 #loop while dividing by radix
+  J1 <- J1+1 #increment counter
+  D0 <- D1+E9 #add tp 1
+  if (D0<=D1) break #test and repeat until equality
+}
+E9 <- E9*E6 #recover last value of machine precision
+J1 <- J1-1 #and adjust the number of digits
+mpvals<-list(eps=E9, radix=E6, ndigits=J1)
+}
+mp<-envron()
+E9<-mp$eps
+E6<-mp$radix
+J1<-mp$ndigits
+cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
+cat( "E9*1E+16=",E9*1E+16,"\n")
+cat( "E6 = ",E6,"  -- the radix of arithmetic","\n")
+cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers","\n")
+cat( "E6^(-J1+1) * 1E+16=",1E+16*(E6^(-J1+1)),"\n")
+cat(".Machine$double.eps=",.Machine$double.eps,"\n")
 ```
-calceps.R
 
@@
 
-### Sys.info()
+??JN exp(sin(cos())) benchmark
+
+## Sys.info()
 
@@
 
-## Issues relateing to RNG
+## Issues relating to random number generation
 
@@
 ```
-
-??JN exp(sin(cos())) benchmark
-
 # Choices

### 2022-01-18 MachineSummary/MachineSummary.Rmd
updated

@@
 attempts to summarize some possibilities available in the R statistical software and language.
+Note that one of our interests is to have tools that work in R **across different platforms**.
+That is, we want to be able to call an R function to get machine and computing environment
+characterizations regardless of the particular platform which we are currently using.
+Moreover, we want to have a succinct output of such information.
 
@@
 
+An important tool for acquiring machine information is the CRAN package
+`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
+function of this package for some Windows 10 versions which is now 
+corrected (??check it is uploaded). ??Do we wish to comment on the
+"GB" issue?
+
 ?? Need to comment on
@@
 
-?? get_ram may not work in Windows 10.
-
-## calceps.R
+## Calculating the machine precision
 
@@
 cat(".Machine$double.eps=",.Machine$double.eps,"\n")
+cat("From R  .Machine:\n")
+cat("double.eps=", .Machine$double.eps,"\n", 
+    "double.base=", .Machine$double.base,"\n", 
+    "double.digits=", .Machine$double.digits,"\n" )
 ```
 
-- calceps.R example
+## Simple performance examples
 
-??JN exp(sin(cos())) benchmark
+There are many performance examples used for testing the speed of computers. A 
+well-known example is the @linbench03 LINPACK test used for ranking supercomputers.
+These generally employ linear algebra calculations. The example below, by
+contrast, focuses on the special function computations. Variations on this
+simple performance test have been used by one of the authors for nearly half
+a century. Note the use of the `microbenchmark` timing package (@microbench13).
+
+```{r loopesc}
+loopesc <- function(nn){
+  ss<-0
+  for (i in 1:nn) {
+    xx<-exp(sin(cos(1.0*i)))
+    ss<-ss+xx
+  }
+  xx
+}
+require("microbenchmark")
+nvals<-c(1000, 10000, 100000, 1000000)
+tvals<-rep(NA,length(nvals))
+i<-0
+for (nn in nvals) {
+  i<-i+1
+  cat(nn,"\n")
+  tt<-microbenchmark(loopesc(nn), unit='us')
+  print(tt)
+  tvals[i]<-mean(tt$time)
+}
+plot(log10(nvals),log(tvals))
+
+```
 
@@
 
-From the above discussion, the following offers a possible compact solution.
+From the above discussion, the following offers a possible compact solution. Users may
+wish to modify this to their own particular needs.
 
@@
 and the current document does not become out of sync with the working MachID.
+
+## References

### 2022-01-18 MachineSummary/MachineSummary.Rmd
minor edit

@@
 function of this package for some Windows 10 versions which is now 
-corrected (??check it is uploaded). ??Do we wish to comment on the
+corrected on https://github.com/csgillespie/benchmarkme, but at the 
+time of writing is not yet on the CRAN repository. ??Do we wish to comment on the
 "GB" issue?

### 2022-01-19 PkgFromRbase/PkgFromRbase.Rmd
Edits to articles and blog posts

@@
    - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
-   - based on contribution from Duncan Murdoch
-date: "11/06/2021"
+   - based on communications from Duncan Murdoch
+date: "11/06/2021, revised 19/01/2022"
+bibliography: ../BibSupport/ImproveNLS.bib
 output:
@@
 improvements to the nls() function and related features in R as part of a Google
-Summer of Code project for which Arkajyoti Bhattacharjee is the funded student.
+Summer of Code project for which Arkajyoti Bhattacharjee was the funded student.
 However, nls() has many tentacles involving a number of files and functions that
@@
 existing ones. A heavy-effort approach would be to have separate full sets of
-R code and build each system and run them separately. That is, we could have
-two versions of R.
+R code and build each system and run them separately. That is, we want to have
+two or more versions of R in the same computing system.
 
-Attempts to provide R functions to switch between the sets of working functions
--- e.g., `ReplaceNLS.R` and `RestoreNLS.R` --
-were not promising, and not pursued.
+There are several ways to do this:
+
+- build or install the two versions of R in separate directory trees and
+  invoke them with separate launch scripts or programs, such as described
+  in https://support.rstudio.com/hc/en-us/articles/215488098-Installing-multiple-versions-of-R-on-Linux ;
+
+- use virtual machines with different R versions;
+
+- run the different R versions in containers, e.g., @Rocker2017;
+
+- move the required functionality to a package that can be loaded as needed
+  to mask the base functionality and replace it with the functions we wish
+  to test or use. This last approach is discussed here.
+
+We did attempt to provide R functions to switch between alternative sets of working functions
+-- e.g., `ReplaceNLS.R` and `RestoreNLS.R` -- but this approach 
+was not promising, and not pursued.
 
@@
 a package of interlocked functions that provide the set of `nls()` capabilities.
-This package `nlspkg` has been cloned to `nlsalt` which is being modified
-to provide the new capabilities or new expression of the code. Moreover, these
-capabilities, such as a function `Anyfn()` can be tested side by side by use
-of the double colon syntax, namely,
+This package `nlspkg` was copied to `nlsalt` then modified
+to provide the new capabilities or new expression of the code. The new 
+capabilities or code, such as a function `Anyfn()` can be tested side by side with 
+a "standard" version (if one exists) by use of the double colon syntax, namely,
 
@@
 
-After I downloaded the ZIP file of Duncan's package from his Github repository, 
-I built and checked the package. There were two WARNING messages:
+After we downloaded the ZIP file of Duncan's package from his Github repository, 
+we built and checked the package. There were two WARNING messages:
 
@@
 
-- the rest of the "undocumented" functions are C functions (identifiable as
-their names have underscore characters in the names). It is these functions
-that our "Improvements to nls()" project aims to replace where sensible to
+- the rest of the "undocumented" functions are C functions (identifiable easily
+as their names contain underscore characters). It is these functions
+that our "Improvements to nls()" project aims to replace where it is sensible to
 do so with R functions. They may then be documented as needed.
@@
 It is worth mentioning that the NAMESPACE file provided by Duncan is an
-important part of the packaging. It could be worthwhile to document the
-elements in that file needed for a package such as `nlspkg`. In particular,
-however, we must note that in renaming the package from `nlspkg` to something
+important part of the packaging. It could be worthwhile to document each
+elements in that file needed for a package such as `nlspkg`, but at the
+time of writing, this is yet to be carried out. However, we must note 
+that in renaming the package from `nlspkg` to something
 else, the change of name must be mirrored in the line
@@
 
-We also noted that the `S3method()` entries are needed to link the generic
+We also noted that the `S3method()` entries of the NAMESPACE files 
+are needed to link the generic
 functions to similar modeling functions in R. Thus `anova`, `coef`, etc.
@@
 the internal C functions, named `C_(something)`. We have suppressed this blanket
-declaration and explicitly exported the `nls` and `numericDeric` functions. 
+declaration and explicitly exported the `nls` and `numericDeriv` functions. 
 One or two other routines may need to be exported to allow full functioning
@@
 
-We note that name change is ALSO required in `src/init.c`
-in the line
+We note that name change is ALSO required in `src/init.c` in the line
 
@@
 For version control, the file `.gitignore` has two entries: one is `*.0` and
-the other, initially, was `nlspkg.so`. I changed the latter to `*.so` and this
+the other, initially, was `nlspkg.so`. We changed the latter to `*.so` and this
 appears to be satisfactory for general use.
 
-# Developing an alternative set of functions and alternative package
+# Developing an alternative set of functions and an alternative package
 
-The alternative package is being developed by replacing calls to C functions
+We wished to build our alternative package by replacing calls to C functions
 with R work-alike functions. A preliminary example that already is set up and
@@
 code needs to be commented out or deleted from the C file(s) and headers
-to avoid errors when building and checking the alternative.
+to avoid errors when building and checking the alternative package.
 
@@
 directories with their Rproj files can be within the umbrella version control 
-project `improvenls` for the Git version control. (We are using a Gitlab
+project `improvenls` for the Git version control. (We used a Gitlab
 repository.)
@@
 While all the above worked well in a Linux environment, when we tried running
-the build and CHECK in a Windows environment, we found we got errors. (I normally
-run in Linux Mint 20.1, but Arkajyoti Bhattacharjee uses Windows 10. Each of us
+the build and CHECK in a Windows environment, we found errors. (JN normally
+runs in Linux Mint 20.x, but AB uses Windows 10. Each of us
 has a VirtualBox VM for the alternative OS to allow for checking of our work.)
 There is a load error for functions from the BLAS (Basic Linear Algebra Subroutines)
-collection in a 32 bit architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
+collection in a 32 bit Windows architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
 feature. Using the `CMD.exe` terminal window showed that `R CMD build` worked, but
@@
 ```
-
-I did, however, note that it is not clearly indicated that the file needs to be in 
+We did, however, note that it is not clearly indicated that this file needs to be in 
 `src/`, nor (until revealed by `R CMD check`) that the file must be `Makevars.win` and
 not `makevars.win`. There is a lot of material on `Makevars` in the Writing R
-Extensions manual, but these points do not seem to be presented.
+Extensions manual (@R-extns), but these points do not seem to be presented, or at least are
+not obvious.
+
+Our packaged version of nls() from July 2021 is at 
+https://gitlab.com/nashjc/improvenls/-/tree/master/nlspkg 
+Note that this is drawn from the version of R of that date (R 4.1.0), so will not necessarily reflect
+the distributed version of R when this document is read.
 
@@
 
-It is a pleasure to note the help and encouragement of Duncan Murdoch, Arkajyoti
-Bhattacharjee, Dirk Eddelbuettel and Brett Klamer.
+It is a pleasure to note the help and encouragement of Duncan Murdoch, 
+Dirk Eddelbuettel and Brett Klamer.
+
+# References

### 2022-01-19 RefactoringNLS/RefactoringNLS.Rmd
Edits to articles and blog posts

@@
 
-The project output is available in the Git repository https://gitlab.com/nashjc/improvenls
-?? Arkajyoti -- do you want to change to Github
+The project output is available in the Git repository https://github.com/nashjc/RNonlinearLS .
 
@@
 These are documents used to discuss particular aspects of our work. These are part
-of the repository https://gitlab.com/nashjc/improvenls.
+of the repository https://github.com/nashjc/RNonlinearLS.
 
@@
 - ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
-  project, possibly with wider application to nonlinear least squares in general
-
-- MachineSummary.Rmd: an informal investigation of ways to report the characteristics
-  and identity of machines running tests. `MachID.R` offers a concise summary function
+  project, possibly with wider application to nonlinear least squares in general.
+
+- MachineSummary.Rmd: an investigation of ways to report the characteristics
+  and identity of machines running tests. `MachID.R` offers a suggested concise summary function
   to identify a particular computational system used for tests.
-  Note that there remains an open issue that the get_ram() for Windows 10 in the `benchmarkme`
-  package may report the RAM size as "NA". We are collaborating with Colin Gillespie, 
-  maintainer of `benchmarkme` to seek a workaround.
 
@@
 
-Hans Werner Borchers was helpful in developing the proposal for this project.
+Hans Werner Borchers has been helpful in developing the proposal for this project
+and in comments on this and related work.
 

### 2022-01-19 MachineSummary/MachID.R

@@ -2,6 +2,7 @@ sy<-Sys.info()
 tsstr<- format(Sys.time(), "%Y%m%d%H%M")
 cpu<-benchmarkme::get_cpu()
 ram<-benchmarkme::get_ram()
-machid<-paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
+# machid<-paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
               "|",cpu$model_name,"|",ram,"bytesRAM", sep='')
 cat(machid,"\n")
+

### 2022-01-19 PkgFromRbase/PkgFromRbase.Rmd
edits from AB

@@
 functions which are in the distributed base of R. Our interest was in developing
-improvements to the nls() function and related features in R as part of a Google
+improvements to the `nls()` function and related features in R as part of a Google
 Summer of Code project for which Arkajyoti Bhattacharjee was the funded student.
-However, nls() has many tentacles involving a number of files and functions that
+However, `nls()` has many tentacles involving a number of files and functions that
 may or may not be called as `nls()` is executed.
@@
   invoke them with separate launch scripts or programs, such as described
-  in https://support.rstudio.com/hc/en-us/articles/215488098-Installing-multiple-versions-of-R-on-Linux ;
+  in \url{https://support.rstudio.com/hc/en-us/articles/215488098-Installing-multiple-versions-of-R-on-Linux};
 
@@
 # and
-nlsalt::Anyfn() # New functionality of code
+nlsalt::Anyfn() # New functionality of code.
 ```
@@
 
-After we downloaded the ZIP file of Duncan's package from his Github repository, 
+After we downloaded the ZIP file of Duncan's package from his GitHub repository, 
 we built and checked the package. There were two WARNING messages:
@@
 
-For Rstudio users, note that each package directory needs its own (name).Rproj
-file, created using Rstudio's "File/New Project" etc. dialog. However, the package
+For RStudio users, note that each package directory needs its own (name).Rproj
+file, created using RStudio's "File/New Project" etc. dialog. However, the package
 directories with their Rproj files can be within the umbrella version control 
-project `improvenls` for the Git version control. (We used a Gitlab
+project `improvenls` for the Git version control. (We used a GitLab
 repository.)
@@
 runs in Linux Mint 20.x, but AB uses Windows 10. Each of us
+<!-- Decided abbreviations are obvious, so not explained. -->
 has a VirtualBox VM for the alternative OS to allow for checking of our work.)
 There is a load error for functions from the BLAS (Basic Linear Algebra Subroutines)
-collection in a 32 bit Windows architecture. This is revealed by the Rstudio "Build / Clean and rebuild"
+collection in a 32 bit Windows architecture. This is revealed by the RStudio "Build / Clean and Rebuild"
 feature. Using the `CMD.exe` terminal window showed that `R CMD build` worked, but
@@
 
-\url{https://stackoverflow.com/questions/42118561/error-in-r-cmd-shlib-compiling-c-code}
+\url{https://stackoverflow.com/questions/42118561/error-in-r-cmd-shlib-compiling-c-code}.
 
 This suggested that a file `Makevars.win` is needed in the `src/` directory of the package(s),
-and that this file should contain just one line
+and that this file should contain just the following line.
+<!-- Changed sentence to avoid period in line below which might cause errors if left in code. -->
 
@@
 
-Our packaged version of nls() from July 2021 is at 
-https://gitlab.com/nashjc/improvenls/-/tree/master/nlspkg 
+Our packaged version of `nls()` from July 2021 is at
+
+\url{https://gitlab.com/nashjc/improvenls/-/tree/master/nlspkg}.
+
 Note that this is drawn from the version of R of that date (R 4.1.0), so will not necessarily reflect

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
adjust "Variety"

@@
    - John C Nash, retired professor, University of Ottawa
-date: "16/02/2021"
+date: "16/02/2021 revised 20/01/2022"
 output: 
@@
 knitr::opts_chunk$set(echo = TRUE)
-## ??require(bookdown) # language engine to display text??
+## require(bookdown) # language engine to display text??
 ```
@@
 variety could be an instructive subject for numerical software designers.
+
+# The Problem to be solved
+
+Given a function
+
+$$  f(x) = \sum_{i=1}^n {r_i(x)^2} = r^t r $$
+
+where $x$ is a set of real valued parameters, we wish to find the minimum of $f(x)$
+with respect to x. There may be additional conditions that add to the complexity of
+the problem such as
+
+- bounds (or box) constraints of the form  $$ l_i <= x_i <= u_i $$ alternatively
+  written in vector form $$ l <= x <=u $$
+
+- masks (or fixed parameters), for example, $$ x_i = value_i $$. Such constraints
+  reduce the dimensionality of the nonlinear least squares problem, and we could
+  specify the problem without the masked parameters. However, it is sometimes very
+  convenient to be able to write our problem with such paramters that may later be
+  altered in value or allowed to vary in an optimization of the sum of squares.
+
+- inequality constraints, which in general may be written as 
+  $$ C_k(x) >=0 $$, which can be written as a vector form $$ C(x) >= 0 $$
+  A special and important case is that of **linear inequalities** $$ A x >= b $$,
+  that is, $$ C(x) = A x - b $$.
 

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
more changes to Variety

@@
 
-Given a function
+Given a function that can be written as a sum of squared residuals
 
@@
 - inequality constraints, which in general may be written as 
-  $$ C_k(x) >=0 $$, which can be written as a vector form $$ C(x) >= 0 $$
-  A special and important case is that of **linear inequalities** $$ A x >= b $$,
-  that is, $$ C(x) = A x - b $$.
+  $$ C_k(x) >=0 $$ which can be written as a vector form $$ C(x) >= 0 $$
+  A special and important case is that of **linear inequalities** $$ A x >= b $$
+  that is, $$ C(x) = A x - b $$
 
@@
 
-An overview of the method is provided by Wikipedia:
-
+An overview of this method is provided by Wikipedia: 
 \url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}
-
 The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
@@
 derivative) is zero. In multiple dimensions, this is the same as
-having the gradient $g$ stationary. The so-called Newton method
+having the gradient $g$ stationary. The so-called **Newton method**
 uses the second derivatives in the Hessian matrix $H$ defined as
@@
 
-and we iterate until there is no change in $x$.
-
-There are some adjustments when our function $f(x)$ can be written as a sum of squares
-
-$$  f(x) = \sum_{i=1}^n {r_i(x)^2} = r^t r $$
-
-The final vector product form is the one I favour because it is the least fussy to write and read.
-In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is
+and we iterate until there is no change in $x$, assuming that the method does indeed
+converge for the problem at hand. Newton (and his sometimes associated worker Raphson) 
+used the algorithm we attribute
+to them in a very special case and in a manner unrecognizable as that described
+above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).
+
+For our present situation, the $f(x)$ can be written as a sum of squares as above. It is worth
+repeating the vector form of this
+
+$$  f(x) = r^t r $$
+
+because it is less fussy to write and read. This is clear when we expand the evaluation
+of the Hessian of this sum of squares function.
 
@@
 
-then the gradient is
+then the gradient, in vector form, is
 
@@
 Newton equations could be approximated by ignoring
-the second term (using elements of the residual times second derivatives 
+the second term (which have elements of the residual vector times second derivatives 
 of the residual).  This gives the **Gauss-Newton**
-equations (cancelling the factor 2)
+equations (cancelling a factor 2)
 
@@
 $$  J \delta  \approx - r $$
+
+Like most other treatments, we will skate over the thin ice that $J^t J$ is a 
+good approximation of the Hessian and that the resulting method converges.
 
@@
 
-with respect to the (scalar) parameter $\alpha$. That one-dimensional minimization can
+with respect to the (scalar) parameter $\alpha$. This one-dimensional minimization can
 take quite a lot of computational effort, however, and typically a value of $\alpha$ that
@@
 $$ x_{new} = x + \alpha * \delta $$
-
-Clearly there are many tactical choices that give rise to a variety of particular algorithms.
-One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
-there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
-be perpendicular to any downhill direction.
 
@@
 <!-- ?? Should we explain more. Possibly incorporate into code? -->
+
+Clearly there are many tactical choices that give rise to a variety of particular algorithms.
+One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
+there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
+be perpendicular to any downhill direction.
 
@@
 
-- **A** Compute $J$ and $g$, set up Marquardt equations and find $\delta$
-
-- **B** compute new set of parameters $$x_{new} = x + \delta $$
+- **A** Compute $J$ and $g$,
+
+- **B** set up Marquardt equations for current $\lambda$ and find $\delta$, 
+   compute new set of parameters $$x_{new} = x + \delta $$
 
@@
 There are, of course, well-written codes that address the issues of what "small" means,
-but such code is generally quite involved.
+but such code is generally quite involved. The R function `all.equal()` presents
+a well-considered approach. However,
 I have never seen a case where the $OFFSET$ approach caused difficulties, and it
@@
 `nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
-use ONLY the fixed element (i.e., unit matrix) stabilization. A very
-preliminary timing ([2021-6-26]) using the Hobbs unscaled problem shows
-a slight improvement for the simplified version in timing with equivalent
-solution. ?? check
+use ONLY the fixed element (i.e., unit matrix) stabilization. A
+preliminary timing (using @microbench13) with the Hobbs unscaled problem 
+(@cnm79, Example 12.2) shows
+a slight advantage for `nlsralt` in execution timing with an equivalent solution.
 
@@
 
-There have been many proposed approaches to nonlinear least squares. The following
-is a non-exhaustive sample.
+While the above ideas are the dominant theme in nonlinear least squares, there have been many 
+other proposed approaches. The following is a non-exhaustive sample.
 
@@
 1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
-line-search minimization. Jones mentions matrix inversion, and it is not clear whether
+line-search minimization. Jones mentions matrix inversion, but it is not clear whether
 he truly forms the inverse. The main idea in the paper (and presumably the code) is
@@
 is increased. It would be interesting to actually try some of the ideas, which do,
-as far as I can determine, have a reasonable basis.
+as far as I can determine, have a reasonable basis and could possibly offer some
+efficiencies.
 
 ### Approaches to Newton's method for special models
-
-Newton (and his sometime associated worker Raphson) used the algorithm we attribute
-to them applied to a very special case and in a manner unrecognizable as that described
-above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).
 
@@
 derivatives, which are generally much, much easier to compute than the full second
-derivatives of the Hessian. However, there have been suggestioned
+derivatives of the Hessian. However, there have been suggested
 Newton-like approaches, possibly for special types of models where some short cut
-to the Hessian is available. These may or may not make sense in the context of
-a given user or group of users. I regard these approaches as part of the
-computational infrastructure of the special cases rather than as general nonlinear
-least squares methods.
-
-?? MaxLik?? others -- Charles Geyer
-
-https://www.stat.umn.edu/geyer/trust/
-
-CRAN trust package??
-
-?? had something for exponential models??
+to the Hessian is available. I regard these approaches as part of the
+computational infrastructure of the special cases, as it is usually 
+difficult to apply them outside of their context.
+
+Two such examples are the R packages `maxLik` (@maxLik2011) and `trust` (@p-trust).
 
@@
 The central idea of Hartley's method is a line search and that of Marquardt a
-parametrized stabilization that can be shown to reduce a sum of squares for large
+parametrized stabilization that can be shown to reduce a sum of squares for a large
 enough stabilization parameter. There is nothing to prevent use of a line
@@
 algebra techniques, or even approximations. In certain situations, some of these
-choices may be important, but here I will not pursue them further.
+choices may be important. Unfortunately, the effort to benchmark the possibilities
+seems higher than any benefit in replacing quite good existing methods.
 
@@
 
-The sources of variety in implementation include:
+The sources of variety in implementation of nonlinear least-squares methods include:
 
@@
 - sequential or full creation of the Jacobian and residual, since
-  it may be done in parts
+  this may be done in parts
 

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
incremental edits -- still much to do

@@
 
+<!-- expand ??, include some BHHH stuff -->
+
 ### Hybrid methods
@@
   in R's `nls` versus packages `minpack.lm` and `nlsr`, or as
-  a general optimization problem.
+  a general optimization problem
+
+- incorporation of constraints on the parameters, namely bounds,
+  masks or inequalities.
 
@@
 
-### Issues with this structure
-
-A major criticism of this approach is that the collection of information needed
+### Issues with the `nls()` structure
+
+A major criticism of the `nls()` approach is that the collection of information needed
 at any one time -- the "environment" in which residuals and jacobians are calculated,
-as well as the solution methods -- is not at all well-documented. If the software is
+as well as the solution methods -- is not sufficiently well-documented. If the software is
 to be updated or maintained, the location and purpose of the many elements that
-reside in different environments is difficult to determine without much effort.
+reside in different environments is difficult to determine without effort.
 In my opinion, this renders the structure prone to collapse if there are infrastructure
@@
 easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
-has elements $\sqrt {(\lambda)}$ times the diagonal stabilization elements
+has elements $\sqrt {(\lambda)}$ times the square roots of the 
+diagonal stabilization elements
 we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
@@
 or stabilization parameter ($\lambda$) until either a "better" point $x_{new}$ is found
-or we terminate the search.
-
-Thus it is useful to be able to separate the computation of residuals from that of the
+or we terminate the search. However, as we have seen, Hartley does not have to adjust
+the search direction.
+
+Thus it could be helpful to separate the computation of residuals from that of the
 Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
@@
 `nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
-as well as the `rjfun()` generated by `nlsr::model2rjfun()`. Nevertheless, it is likley worth 
-avoiding duplication if possible, though the computation of Jacobian elements uses a lot of the
-same steps as the computation of the residuals.
-
-?? need to explore timings
-
+as well as the `rjfun()` generated by `nlsr::model2rjfun()`. On the other hand, 
+the computation of Jacobian elements can share many calculations with the computation of the residuals. 
 Evidence is needed as to which de-duplication efforts give worthwhile savings, and in what
-situations:
-
-- saving residual values when computing derivatives via finite differences
-- separation of the Jacobian routine completely from that of the residuals
-- other measures?
+situations, but is likely of importance only to users with extremely time consuming problems.
 
@@
 equations and hence the sum of squares and cross products (SSCP) matrix, we
-know by construnction that this is a symmetric matrix and also non-negative definite.
+know by construction that matrix is symmetric and also non-negative definite.
 (Generally, unless we have a strict linear dependence between the columns of
@@
 
+?? sort out $A$ and $C$
+
 ### Storage in Nash/Walker-Smith (1987) BASIC code
@@
 at parameters $x$ using functions for the residuals and Jacobian.
-The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements (row-wise ordering of $C$). 
+The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements.
 The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ is added to each,
-where $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in @jn77ima to
+then saved in matrix $C$. $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in
+@jn77ima to
 overcome potential underflow to zero of the accumulation of the diagonal elements in low-precision
-arithmetic. Let us call the augmented SSCP matrix $C$.
+arithmetic.
 
@@
 or equal to the current best (i.e., lowest) sum of squares, we increase $\lambda$ and generate a new
-$C$ matrix. Note that we do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
-adjustment.
-
-The code in @Nash1987a is much more complicated than this description suggests because of
+$C$ matrix. We do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
+adjustment to $A$ to get a new $C$.
+
+The code in @Nash1987a is much more complicated than this description, since we should address
 
@@
   special functions, e.g., `log()` for small arguments or `exp()` for large ones, or for problem specific
-  situations where some parameter combinations might be considered inadmissible.
+  situations where some parameter combinations might be considered inadmissible. (Readers may note that
+  the codes were written before @IEEE754-1985 was applied to most BASICs. The author was a member of
+  that standardization committee.)
 
@@
 
-In @nlsr-manual, the approach to the augmented Gauss-Newton equations is nominally equivalent to that in @Nash1987a,
+In @nlsr-manual, the approach to the augmented Gauss-Newton equations is nominally equivalent to that in
+@Nash1987a,
 except that we perform the solution using a QR decomposition, in fact a modification of the DQRDC routine from 
@@
 
-That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.
+That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newtx
 
 To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Let us call this augmented matrix
-$K$. We will also need to augment $r$ by an appropriate number of zeros. For the traditional Marquardt method, we
+$K$. We will also need to augment $r$ by an appropriate number of zeros. 
+For the traditional Marquardt method, we
 need to use the square roots of the diagonal elements of $J^t J$, that is, an $n$ by $n$ 

### 2022-01-20 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
more incremental

@@
 
-$$ (J^t J + \lambda  \mathbf{1}_n)  \delta  = A \delta =  - J^t r $$
+$$ (J^t J + \lambda  \mathbf{1}_n)  \delta  = C \delta =  - J^t r $$
 
@@
 the coefficient matrix is non-negative definite and symmetric. Thus it permits of
-both general and specialized methods for linear equations. Furthermore, one can also
-set up the solution without forming a coefficient matrix $A$ 
+both general and specialized solution methods for linear equations. Furthermore, one can 
+often set up the solution to a linear least squares problem without forming a coefficient matrix $C$ 
 by using several matrix decomposition methods. Thus there are many possible procedures.
@@
   choices uses the underlying structure and non-negative definite nature of 
-  the matrix $A$.
+  the matrix $A = J^t J$ or stabilized versions $C$ thereof.
 
 - Variants of Gauss elimination that build matrix decompositions: We can apply
-  elimination to produce an LU decomposition of $A$, and this may offer some 
+  elimination to produce an LU decomposition of $C$, and this may offer some 
   computational advantages, though in the present case such approaches do not
@@
 
-- Gauss-Jordan inversion: Here we compute the inverse of $A$ and then work with
+- Gauss-Jordan inversion: Here we compute the inverse of $C$ and then work with
   this inverse. That is, we use
 
-  $$\delta = - A^{-1} g$$
+  $$\delta = - C^{-1} g$$
 
@@
 
-- Choleski Decomposition and back substitution: There are a number of varieties
-  of Choleski decomposition, but they all take advantage of the symmetric and
-  non-negative definite structure of $A$. A particularly compact form was used
+- Cholesky Decomposition and back substitution: There are a number of varieties
+  of Cholesky decomposition, but they all take advantage of the symmetric and
+  non-negative definite structure of $C$. A particularly compact form was used
   in @Nash1987a, though it is debatable whether the saving in storage was not
@@
   matrix: the eigenvalues of real, symmetric matrices
-  such as $A$ can be computed in many ways and the eigendecomposition used to 
+  such as $C$ can be computed in many ways and the eigendecomposition used to 
   solve the Gauss-Newton equations. The computations are much more costly than
@@
 
-Here we proposed to solve using the matrix $J$ or an augmented form $K$
+Here we propose to solve the linear least squares sub-problem 
+using the matrix $J$ or an augmented form $K$
 directly. There are a number of approaches that can be used, which we
-will divide into two classes, QR and SVD methods.
+will divide into two classes, QR and singular value decomposition (SVD) methods.
 
@@
 
-There are many ways of computing such decompostions:
-
-- Householder reflections with or without pivoting options
-- Givens plane rotations, similarly
+There are many ways of computing such decompositions:
+
+- Householder reflections with or without pivoting options;
+- Givens plane rotations, again with pivoting variants;
 - Gram-Schmidt orthogonalization of the columns of $K$, either directly (one
@@
 A detail in choosing our approach is the mechanism of inclusion of the
-Marquardt (or Marquardt Nash) stabilization. It turns out to be remarkably
+Marquardt (or Marquardt-Nash) stabilization. It turns out to be remarkably
 easy, if a little tedious. We simply augment $J$ by a diagonal matrix that
-has elements $\sqrt {(\lambda)}$ times the square roots of the 
-diagonal stabilization elements
-we wish to use. Then $K^t K$ is the appropriate $A$ for the Marquardt version
-of the Gauss-Newton.
-
-As we explain below, we do, unfortunately, realize that Marquardt suggested a scaled 
-stabiilzation where we want to add the diagonals of $J^t J$ normal equation approach.
-This requires that we perform at least a partial computation of the SSCP matrix,
-which is relatively costly. For this reason we plan ?? to explore some proxies
-for these elements.
+has elements that the square roots of  $\lambda (D + \phi 1_n)$. 
+Then $K^t K$ is the appropriate $C$ for the Marquardt version
+of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $\lambda D$
+and $\lambda \phi 1_n$.
+
+Unfortunately, if we decide to use the stabilization involving $D$, then 
+we must perform at least a partial computation of the SSCP matrix,
+which is relatively costly. For this reason, it may be preferred to use
+the unscaled stabilization using $1_n$ only. It may then be important to
+provide an explicit (and fixed) parameter scaling.
+
+?? manual scaling -- is it part of nlsr??
 
@@
 Thus it could be helpful to separate the computation of residuals from that of the
-Jacobian. However, the structure of R is such that it is much easier to have the Jacobian
+Jacobian. However, the structure of R is such that it is often much easier to have the Jacobian
 as an **attribute** of the residuals object. Indeed, it is the "gradient" attribute for 
-historical reasons. This is the setup for both the finite difference approximations of
-`nlspkg::numericDeriv()` (the Jacobian routine in base R for nls() in file `./src/library/stats/R/nls.R`)
-as well as the `rjfun()` generated by `nlsr::model2rjfun()`. On the other hand, 
+historical reasons. This is the setup for both the finite difference approximations to the 
+Jacobian in 
+`nlspkg::numericDeriv()` which is the Jacobian approximation routine in base R for nls() in file
+`./src/library/stats/R/nls.R`. It is also the structure of `rjfun()` generated by `nlsr::model2rjfun()`. 
+On the other hand, 
 the computation of Jacobian elements can share many calculations with the computation of the residuals. 
@@
 
-## Storage stucture
+## Storage of algorithm objects
 
@@
 with the Cholesky approach. This was continued in the Second Edition @nlacatvn1060620 and
-in @jnmws87. However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
-has been to use a QR decomposition of an augmented Jacobian. `nls()` uses a QR decomposition
-of the Jacobian without augmentation (referred to as the "gradient", however).
-
-The particular QR calculations in these packages and in R-base are quite well-hidden in internal
-code, complicating comparisons of storage, complexity and performance. The present exposition
-is an attempt to provide some documentation.
-
-?? sort out $A$ and $C$
+in @jnmws87. Even storing the full $A$ matrix uses only $n^2$ numbers.
+
+However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
+has been to use a QR decomposition of an augmented Jacobian. Similarly, `nls()` uses a QR decomposition
+of the Jacobian (referred to as the "gradient") without augmentation. Clearly the $K$ matrix is
+at least $m$ by $n$, which for most nonlinear modelling problems is much larger than the $C$ matrix.
+The particular QR calculations in `nls()` and its packaged version are quite well-hidden in internal
+code, complicating comparisons of storage, complexity and performance.
 
@@
 
-which is called the Choleski decomposition. In practice, we need $A$ computationally
+which is called the Cholesky decomposition. In practice, we need $A$ computationally
 non-singular (not just theoretically so), but the Marquardt stabilization can be used

### 2022-01-21 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
clean up hobbs timingx

@@
 program code. In organizing and documenting the nearly half-century of programs
-in the Nashlib collection associated with @jncnm79, the author realized that this
-variety could be an instructive subject for numerical software designers.
+in the Nashlib collection associated with @jncnm79 and with similar packages
+for the R Project for Statistical Computing, the author realized that an overview
+of this variety could be instructive for numerical software designers.
 
@@
 preliminary timing (using @microbench13) with the Hobbs unscaled problem 
-(@cnm79, Example 12.2) shows
-a slight advantage for `nlsralt` in execution timing with an equivalent solution.
+(@cnm79, Example 12.2), presented later, shows
+a slight advantage for `nlsralt` in execution timing, but we caution
+that this is but a single example.
 
@@
 Then $K^t K$ is the appropriate $C$ for the Marquardt version
-of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $\lambda D$
-and $\lambda \phi 1_n$.
+of the Gauss-Newton. Clearly it is equivalent to augment $J$ by **both** $Z1 = \lambda D$
+and $Z2 = \lambda \phi 1_n$.
 
@@
 the unscaled stabilization using $1_n$ only. It may then be important to
-provide an explicit (and fixed) parameter scaling.
-
-?? manual scaling -- is it part of nlsr??
+provide an explicit (and fixed) parameter scaling. The function minimization
+package `optimx` (@p-optimx) includes such a possibility, but appears to be
+missing from nonlinear least squares possibilities in R.
 
@@
 One of the issues with creating the SSCP matrix $A$ is that we lose information in finite arithmetic 
-(See @jncnm79, Chapter 5). A QR approach to the traditional Gauss-Newton codes is to solve
+(See @jncnm79, Chapter 5). As noted above, a QR approach to the traditional Gauss-Newton codes is to solve
 
@@
 
-That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newtx
-
-To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Let us call this augmented matrix
-$K$. We will also need to augment $r$ by an appropriate number of zeros. 
-For the traditional Marquardt method, we
-need to use the square roots of the diagonal elements of $J^t J$, that is, an $n$ by $n$ 
-diagonal matrix $Z1$ where
+That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.
+
+To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Also as described above,
+augmented matrix $K$ is a row-wise appending $Z1$ and, for the Nash modification $Z2$, to $J$.
+We will also need to augment $r$ by an appropriate number of zeros. 
+As a reminder, $Z1$ and $Z2$ are given by
 
 $$ Z1_{k,k}  = \sqrt(\lambda) \sqrt(J^t J)_{k,k}  $$
-
-and $r$ is extended by $n$ zeros. For the Nash modification, we add a further $n$ zeros to $r$ and augment
-the $K$ matrix with another $n$ by $n$ block $Z2$ which is a diagonal matrix where the diagonal elements are
+and
 
@@
 
-Unfortunately, there is a need in each major iteration to compute at least the diagonal elements of the
-SSCP matrix. It may, in fact, be more efficient to omit the $Z1$ augmentation. Package `nlsralt`, as
+We have noted that, at each major iteration, the construction of $Z1$ requires the diagonal elements of the
+SSCP matrix. Thus it could be more efficient to omit the $Z1$ augmentation and use an explicit scaling
+in the original setup. Package `nlsralt`, as
 mentioned, does this with functions `nlxbx()` (for formula-specified models) and `nlfbx()` (for problems
@@
 <!-- ``` -->
-
-??FIXME??
 

### 2022-01-21 VarietyInNLS/HobbsTiming.R

@@ -15,7 +15,6 @@ weeddata1  <-  data.frame(y=ydat, tt=tdat)
 
 library(microbenchmark)
 library(nlsralt)
-library(nlsr)
 tnlsrh1<-microbenchmark(nlsrh1<-nlxb(eunsc, data=weeddata1, start=start1))
 tnlsrxh1<-microbenchmark(nlsrxh1<-nlxbx(eunsc, data=weeddata1, start=start1))
 nlsrh1

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
add sstimer

@@
 
-### Other storage approaches.
-
-?? do we want to report any other approaches here?
-In particular, should we look at minpack.lm to see how the Marquardt stabilization 
-is applied? This is a very complicated program, since it calls C to call Fortran,
-and the Fortran is layered in a way that each part is well-documented but the
-general flow and linkage of the parts is non-obvious.
-
-## Sequential or full Jacobian computation
-
-We could compute a row of the Jacobian plus the corresponding residual element
-and process this before computing the next row etc. This means the full Jacobian
-does not have to be stored. In Nashlib, Algorithms 3 and 4, we used row-wise data 
-entry in linear least squares via Givens' triangularization (QR decompostition),
+<!-- ### Other storage approaches. -->
+
+<!-- ?? do we want to report any other approaches here? -->
+<!-- In particular, should we look at minpack.lm to see how the Marquardt stabilization  -->
+<!-- is applied? This is a very complicated program, since it calls C to call Fortran, -->
+<!-- and the Fortran is layered in a way that each part is well-documented but the -->
+<!-- general flow and linkage of the parts is non-obvious. -->
+
+### Sequential or full Jacobian computation
+
+It is possible to compute a row of the Jacobian plus the corresponding residual element
+and process these, for example, into the SSCP matrix $J^t J$ and gradient $- J^t r$  before computing 
+the next row etc. This means the full Jacobian does not have to be stored. 
+In Nashlib, Algorithms 3 and 4, we used row-wise data 
+entry in **linear** least squares via Givens' triangularization (QR decompostition),
 with the possibility of extending the QR to a singular value decomposition.
-Forming the SSCP matrix can also be generated row-wise as well. We do not 
-anticipate that any of these approaches will offer advantages that are so
-great that we would choose them.
+While such approaches were forced upon us in the early 1970s when they were devised,
+we do not anticipate that any of these approaches will now offer enough advantage that
+we would use them as a general choice. They may, however, be appropriate for some 
+problems where the number of observations is extremely large.
 
@@
 set of arguments for special functions. In my experience, the use of analytic derivatives
-is not much faster than the finite difference approximations, and generally finite
-difference methods iterate as quickly as the "exact" derivative methods to the 
-neighbourhood of a solution. It is in allowing good estimates of convergence 
-criteria that the analytic derivatives seem to show their advantage i.e., knowing
-we are at a solution rather than the speed in getting there.
+does **not** give a much faster solution than the finite difference approximations.
+Indeed, finite difference methods iterate as quickly as the "exact" derivative methods to the 
+neighbourhood of a solution. This is much along the same lines as secant versus Newton
+methods for solving nonlinear equations. Analytic derivatives do, however, show their
+advantage in allowing us to more reliably compute convergence 
+criteria. That is, we know we are at a solution and do not attempt further iterations.
 
@@
 
-R allows the nonlinear least squares problem to be presented via a formula
+R allows the nonlinear least squares problem to be presented via a **formula**
 for the model. This is generally very attractive to users, since they can
 specify their problem in ways that often accord with how they think about them.
-Many (most?) researchers do not write program code, so preparing residual and
+Many, possibly most, researchers do not write program code, so preparing residual and
 Jacobian functions is foreign to them.
@@
   since we must essentially parse the formula. R has some tools for this that
-  we use.
+  we use. Nevertheless, the automated conversion may risk setting up calculations
+  in ways that could incur ill-conditioning.
 
@@
 
-- checking for correctness of inputs and translations can be considered to
+- checking for correctness of inputs and translations may
   take up most of the code in a reliable and robust package.
@@
 
-- use a row-wise generation of the Jacobian in either a Givens' QR or SSCP approach.
-  This saves space for the Jacobian as well as as well as the working matrices of
+- use a row-wise generation of the Jacobian in either a Givens' QR or SSCP approach, 
+  as mentioned above. 
+  This saves space for the Jacobian as well as as well as some of the working storage of
   the Gauss-Newton or Marquardt iterations;
@@
 - if the number of parameters to estimate is large enough, then a normal equations
-  approach using a compact storage of the lower triangle of the SSCP matrix. 
+  approach using a compact storage of the lower triangle of the SSCP matrix may help. 
   However, the scale of the saving is really very small in comparison to 
@@
 
-Do we plan to use only timing? Or size of code and memory used? ??
+Most workers report execution timings as the principal measure of performance.
+This is generally acceptable for comparing programs, and we often then attribute
+the results to the underlying methods, though clearly programming choices can have
+a quite large effect.
+
+For example, we can form sums of squares in R using `for` loops, a vectorized
+`sum()` or the `crossprod()` function. These have different performance profiles.
+Let us form the sums of squares of different length random vectors.
+
+```
+ n        t(forloop) : ratio          t(sum) : ratio       t(crossprod)    all.equal 
+100           31701  :  2.1996        15541  :  1.0784           14412       TRUE 
+1000          35604  :  13.004       2900.1  :  1.0592            2737.9     TRUE 
+10000        323969  :  27.826        21051  :  1.8081           11643       TRUE 
+1e+05       3120277  :  30.604       291234  :  2.8564          101957       TRUE 
+1e+06      31140093  :  29.231      4701221  :  4.4131         1065296       TRUE 
+1e+07     317191471  :  21.776     40540123  :  2.7832        14565799       TRUE 
+```
+
+These are averages of 100 executions of each sum, and we would expect them to be
+stable, but there is some variation, likely due to other processes running in an
+active computer, as seen by rerunning the script `sstimer.R`.
+
+```
+ n        t(forloop) : ratio          t(sum) : ratio        t(crossprod)   all.equal 
+100           35643  :  2.4688        13258  :  0.91831           14437      TRUE 
+1000          33172  :  12.142         3689  :  1.3503             2731.9    TRUE 
+10000        324269  :  27.958        28028  :  2.4166            11598      TRUE 
+1e+05       3132152  :  30.765       304086  :  2.9868           101810      TRUE 
+1e+06      31262348  :  28.846      3759337  :  3.4688          1083772      TRUE 
+1e+07     313522217  :  21.538     38746311  :  2.6618         14556401      TRUE 
+```
+
+We also note the very strange result for `n=1000`. ?? No explanation.
+Moreover, results in a Windows 10 virtual machine (under VirtualBox)
+give the results
+
+```
+ n      t(forloop) : ratio            t(sum) : ratio        t(crossprod)   all.equal 
+100         43781  :  2.592126        14442  :  0.8550622         16890      TRUE 
+1000        34241  :  12.59323         2452  :  0.9018021          2719      TRUE 
+10000      348003  :  19.6058         15038  :  0.8472113         17750      TRUE 
+1e+05     3404639  :  20.32475       293930  :  1.75468           67512      TRUE 
+1e+06    34379874  :  19.36951      3571310  :  2.012065        1774948      TRUE 
+1e+07   376569416  :  17.45253     36871902  :  1.70887        21576778      TRUE 
+```
+
+It is also possible to measure the computational effort in terms of Jacobian and residual
+evaluations, as well as the number of "iterations" used to achieve a solution. The
+Hobbs timing above comparing `nlxb()` and `nlxbx()` (i.e., scaled and unscaled Marquardt
+approaches) show different numbers of Jacobian computations, which for these methods are
+the same as the number of major iterations.
+
+In situations where storage is critical, or for gauging the complexity of code, it may
+be useful to report program size or number of lines of code. These measures are all
+influenced by comments, program structure, and computing environment issues.
 
 # Test problems
+
+It is worth including some sample test problems in this article to illustrate
+the discussion.
+
+?? here or later
 

### 2022-01-22 VarietyInNLS/sstimer.R

@@ -1,0 +1,29 @@
+# crossprod timer
+library(microbenchmark)
+suml<-function(vv) {
+    ss<-0.0
+    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
+    ss  
+}
+sums<-function(vv) {
+  ss<-sum(vv^2)
+  ss  
+}
+sumc<-function(vv) {
+  ss<-as.numeric(crossprod(vv))
+  ss  
+}
+ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
+set.seed(1234)
+cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
+for (nn in ll ){
+   set.seed(1234)
+   vv <- runif(nn)
+   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
+   tsums<-microbenchmark(ss<-sums(vv), unit="us")
+   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
+   ml<-mean(tsuml$time)
+   ms<-mean(tsums$time)
+   mc<-mean(tsumc$time)
+   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
+}

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
sstimer stuff

@@
 `sum()` or the `crossprod()` function. These have different performance profiles.
-Let us form the sums of squares of different length random vectors.
+We can time the sums of squares calculation of different length random vectors
+using these three approaches and get the ratio to the time using `crossprod()`.
+This reveals that the time for summing 100 elements is much longer than that
+of summing 1000, which seems very strange. If we first "compile" the functions
+using `compiler::cmpfun()` we get the second line of timings, which are ordered
+by the size of the vector. Thus we are suspicious that R is trying to compile
+the functions "on the fly" and this may be taking time when the program
+is initiated. Timings in microseconds were run on a machine with the profile (@RMachID2022)
+
+```
+M21:john-Linux-5.13.0-27-generic|Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz|33474072576bytesRAM 
+```
 
@@
  n         t(forloop) : ratio         t(sum) : ratio        t(crossprod)  all.equal 
-100           31701  :  2.1996        15541  :  1.0784           14412       TRUE 
-1000          35604  :  13.004       2900.1  :  1.0592            2737.9     TRUE 
-10000        323969  :  27.826        21051  :  1.8081           11643       TRUE 
-1e+05       3120277  :  30.604       291234  :  2.8564          101957       TRUE 
-1e+06      31140093  :  29.231      4701221  :  4.4131         1065296       TRUE 
-1e+07     317191471  :  21.776     40540123  :  2.7832        14565799       TRUE 
-```
-
-These are averages of 100 executions of each sum, and we would expect them to be
-stable, but there is some variation, likely due to other processes running in an
-active computer, as seen by rerunning the script `sstimer.R`.
-
-```
- n        t(forloop) : ratio          t(sum) : ratio        t(crossprod)   all.equal 
-100           35643  :  2.4688        13258  :  0.91831           14437      TRUE 
-1000          33172  :  12.142         3689  :  1.3503             2731.9    TRUE 
-10000        324269  :  27.958        28028  :  2.4166            11598      TRUE 
-1e+05       3132152  :  30.765       304086  :  2.9868           101810      TRUE 
-1e+06      31262348  :  28.846      3759337  :  3.4688          1083772      TRUE 
-1e+07     313522217  :  21.538     38746311  :  2.6618         14556401      TRUE 
-```
-
-We also note the very strange result for `n=1000`. ?? No explanation.
-Moreover, results in a Windows 10 virtual machine (under VirtualBox)
-give the results
-
-```
- n      t(forloop) : ratio            t(sum) : ratio        t(crossprod)   all.equal 
-100         43781  :  2.592126        14442  :  0.8550622         16890      TRUE 
-1000        34241  :  12.59323         2452  :  0.9018021          2719      TRUE 
-10000      348003  :  19.6058         15038  :  0.8472113         17750      TRUE 
-1e+05     3404639  :  20.32475       293930  :  1.75468           67512      TRUE 
-1e+06    34379874  :  19.36951      3571310  :  2.012065        1774948      TRUE 
-1e+07   376569416  :  17.45253     36871902  :  1.70887        21576778      TRUE 
+100       26686.85 :  1.883607     12721.49  :  0.8979062   14167.95      TRUE 
+           3968.35 :  2.074802       953.36  :  0.4984524    1912.64 
+1000      33413.23 :  12.08929      2372.34  :  0.8583399    2763.87      TRUE 
+          32894.62 :  11.96442      2319.56  :  0.8436696    2749.37 
+10000    325475.7  :  28.14976     15694.36  :  1.357375    11562.29      TRUE 
+         331670.1  :  28.24715     16511.91  :  1.40626     11741.72 
+1e+05   3234528    :  31.81066    346601.8   :  3.408729   101680.6       TRUE 
+        3178637    :  30.75105    346969     :  3.356677   103366.8 
+1e+06  31728237    :  29.08711   4057012     :  3.719297  1090801         TRUE 
+       31837624    :  29.11575   4051945     :  3.705535  1093485 
 ```

### 2022-01-22 MachineSummary/MachID.R

@@ -2,7 +2,7 @@ sy<-Sys.info()
 tsstr<- format(Sys.time(), "%Y%m%d%H%M")
 cpu<-benchmarkme::get_cpu()
 ram<-benchmarkme::get_ram()
-# machid<-paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
+machid<-paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
               "|",cpu$model_name,"|",ram,"bytesRAM", sep='')
 cat(machid,"\n")
 

### 2022-01-22 VarietyInNLS/sstimer2.R

@@ -13,15 +13,21 @@ sumc<-function(vv) {
   ss<-as.numeric(crossprod(vv))
   ss  
 }
-## ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
+
+csuml<-compiler::cmpfun(suml)
+csumc<-compiler::cmpfun(sumc)
+csums<-compiler::cmpfun(sums)
+
+# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
 ll <- c(100, 1000, 10000, 100000, 1000000)
+
 cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
 for (nn in ll ){
    set.seed(1234)
    vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
-   tsums<-microbenchmark(ss<-sums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
+   tsuml<-microbenchmark(sl<-csuml(vv), unit="us")
+   tsums<-microbenchmark(ss<-csums(vv), unit="us")
+   tsumc<-microbenchmark(sc<-csumc(vv), unit="us")
    ml<-mean(tsuml$time)
    ms<-mean(tsums$time)
    mc<-mean(tsumc$time)

### 2022-01-22 VarietyInNLS/sstimer3.R

@@ -13,15 +13,18 @@ sumc<-function(vv) {
   ss<-as.numeric(crossprod(vv))
   ss  
 }
+
+
 ## ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
 ll <- c(100, 1000, 10000, 100000, 1000000)
+
 cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
 for (nn in ll ){
    set.seed(1234)
    vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
-   tsums<-microbenchmark(ss<-sums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
+   tsuml<-microbenchmark(sl<-suml(vv), unit="us", setup=suml(vv))
+   tsums<-microbenchmark(ss<-sums(vv), unit="us", setup=sums(vv))
+   tsumc<-microbenchmark(sc<-sumc(vv), unit="us", setup=sumc(vv))
    ml<-mean(tsuml$time)
    ms<-mean(tsums$time)
    mc<-mean(tsumc$time)

### 2022-01-22 VarietyInNLS/sstimer4.R

@@ -13,8 +13,14 @@ sumc<-function(vv) {
   ss<-as.numeric(crossprod(vv))
   ss  
 }
-## ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
+
+csuml<-compiler::cmpfun(suml)
+csumc<-compiler::cmpfun(sumc)
+csums<-compiler::cmpfun(sums)
+
+# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
 ll <- c(100, 1000, 10000, 100000, 1000000)
+
 cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
 for (nn in ll ){
    set.seed(1234)
@@ -22,8 +28,15 @@ for (nn in ll ){
    tsuml<-microbenchmark(sl<-suml(vv), unit="us")
    tsums<-microbenchmark(ss<-sums(vv), unit="us")
    tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
+   tcsuml<-microbenchmark(sl<-csuml(vv), unit="us")
+   tcsums<-microbenchmark(ss<-csums(vv), unit="us")
+   tcsumc<-microbenchmark(sc<-csumc(vv), unit="us")
    ml<-mean(tsuml$time)
    ms<-mean(tsums$time)
    mc<-mean(tsumc$time)
+   mcl<-mean(tcsuml$time)
+   mcs<-mean(tcsums$time)
+   mcc<-mean(tcsumc$time)
    cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
+   cat("\t",mcl," : ",mcl/mcc,"\t",mcs," : ",mcs/mcc,"\t",mcc,"\n")   
 }

### 2022-01-22 VarietyInNLS/sstimer5.R
add direct statements

@@ -1,0 +1,49 @@
+# crossprod timer
+library(microbenchmark)
+suml<-function(vv) {
+    ss<-0.0
+    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
+    ss  
+}
+sums<-function(vv) {
+  ss<-sum(vv^2)
+  ss  
+}
+sumc<-function(vv) {
+  ss<-as.numeric(crossprod(vv))
+  ss  
+}
+
+csuml<-compiler::cmpfun(suml)
+csumc<-compiler::cmpfun(sumc)
+csums<-compiler::cmpfun(sums)
+
+# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
+ll <- c(100, 1000, 10000, 100000, 1000000)
+
+cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
+for (nn in ll ){
+   set.seed(1234)
+   vv <- runif(nn)
+   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
+   tsums<-microbenchmark(ss<-sums(vv), unit="us")
+   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
+   tcsuml<-microbenchmark(sl<-csuml(vv), unit="us")
+   tcsums<-microbenchmark(ss<-csums(vv), unit="us")
+   tcsumc<-microbenchmark(sc<-csumc(vv), unit="us")
+   tsumld<-microbenchmark({sl<-0;for(ii in 1:nn){sl<-sl+vv[ii]^2}}, unit="us")
+   tsumsd<-microbenchmark(ss<-sum(vv^2), unit="us")
+   tsumcd<-microbenchmark(sc<-as.numeric(crossprod(vv)), unit="us")
+   ml<-mean(tsuml$time)
+   ms<-mean(tsums$time)
+   mc<-mean(tsumc$time)
+   mcl<-mean(tcsuml$time)
+   mcs<-mean(tcsums$time)
+   mcc<-mean(tcsumc$time)
+   mld<-mean(tsumld$time)
+   msd<-mean(tsumsd$time)
+   mcd<-mean(tsumcd$time)
+   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
[…]

### 2022-01-22 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Tidying article

@@
 
-- programming language
+- programming language (which might include called functions that have been
+  constructed outside the current computing environment)
 
@@
 
-# Saving storage
+## Saving storage
 
@@
 using these three approaches and get the ratio to the time using `crossprod()`.
-This reveals that the time for summing 100 elements is much longer than that
+This reveals that the time for summing 100 elements may be much longer than that
 of summing 1000, which seems very strange. If we first "compile" the functions
 using `compiler::cmpfun()` we get the second line of timings, which are ordered
-by the size of the vector. Thus we are suspicious that R is trying to compile
-the functions "on the fly" and this may be taking time when the program
-is initiated. Timings in microseconds were run on a machine with the profile (@RMachID2022)
+by the size of the vector. The third line times just the direct statements, and 
+we see that explicit loops executed directly are quite inefficient unless compiled
+first. Note that `crossprod()` calls code from whatever BLAS is linked to R (@BLAS79 
+and subsequent evolution of these ideas).
+
+Because of the differences and orderings of the timings, we are suspicious that R 
+is trying to compile functions "on the fly" and this may be taking time when the program
+is initiated. R is constantly being reworked, with new releases several times per
+year, with possible changes in how program code is parsed and executed. 
+Timings in microseconds were run on a machine with the profile (see @RMachID2022)
 
@@
            3968.35 :  2.074802       953.36  :  0.4984524    1912.64 
+        1767952    :  1172.25       1120.33  :  0.7428407 	 1508.17 
 1000      33413.23 :  12.08929      2372.34  :  0.8583399    2763.87      TRUE 
           32894.62 :  11.96442      2319.56  :  0.8436696    2749.37 
+        1827674    :  747.1422 	    2645.11  :  1.081305 	   2446.22 
 10000    325475.7  :  28.14976     15694.36  :  1.357375    11562.29      TRUE 
          331670.1  :  28.24715     16511.91  :  1.40626     11741.72 
+        2090288    :  180.5873     15792.13  :  1.364338    11574.94 
 1e+05   3234528    :  31.81066    346601.8   :  3.408729   101680.6       TRUE 
         3178637    :  30.75105    346969     :  3.356677   103366.8 
+        5104998    :  48.30487    361242.5   :  3.418175 	 105682.9 
 1e+06  31728237    :  29.08711   4057012     :  3.719297  1090801         TRUE 
        31837624    :  29.11575   4051945     :  3.705535  1093485 
+       33884253    :  27.14376 	 4560980     :  3.653678  1248326 
 ```
@@
 influenced by comments, program structure, and computing environment issues.
-
-# Test problems
-
-It is worth including some sample test problems in this article to illustrate
-the discussion.
-
-?? here or later
-
-```{r settest15}
-# set parameters
-set.seed(123456)
-a <- 1
-b <- 2
-c <- 3
-np <- 15
-tt <-1:np
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-plot.new()
-plot(tt, yy, type='l')
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-points(tt,y1,type='p', col="green")
-y2 <- yy + 5*ev
-points(tt,y2,type='p', col="blue")
-y3 <- yy + 10*ev
-lg3d15 <- data.frame(tt, yy, y1, y2, y3)
-points(tt,y3,type='p', col="red")
-library(nlsr)
-sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
-print(sol0)
-sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
-print(sol1)
-sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
-print(sol2)
-sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
-print(sol3)
-```
-
-The following is a larger dataset version of this test.
-
-```{r settest150, eval=FALSE}
-np <- 150
-tt <- (1:np)/10
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-y2 <- yy + 5*ev
-y3 <- yy + 10*ev
-lg3d150 <- data.frame(tt, yy, y1, y2, y3)
-np <- 1500
-tt <- (1:np)/100
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-y2 <- yy + 5*ev
-y3 <- yy + 10*ev
-lg3d1500 <- data.frame(tt, yy, y1, y2, y3)
-f0 <- yy ~ a0/(1+b0*exp(-c0*tt))
-f1 <- y1 ~ a1/(1+b1*exp(-c1*tt))
-f2 <- y2 ~ a2/(2+b2*exp(-c2*tt))
-f3 <- y3 ~ a3/(3+b3*exp(-c3*tt))
-```
 
@@
 
-# Implementation choices for an R package
-
-??2021 update to nlsr and ideas for the "Improvements to nls()" project.
+<!-- # Implementation choices for an R package -->
+
+<!-- ??2021 update to nlsr and ideas for the "Improvements to nls()" project. -->
+
+# Test problems
+
+It is worth including some sample test problems in this article to illustrate
+the discussion.
+
+?? here or later
+
+```{r settest15}
+# set parameters
+set.seed(123456)
+a <- 1
+b <- 2
+c <- 3
+np <- 15
+tt <-1:np
+yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
+plot.new()
+plot(tt, yy, type='l')
+set.seed(123456)
+ev <- runif(np)
+ev <- ev - mean(ev)
+y1 <- yy + ev
+points(tt,y1,type='p', col="green")
+y2 <- yy + 5*ev
+points(tt,y2,type='p', col="blue")
+y3 <- yy + 10*ev
+lg3d15 <- data.frame(tt, yy, y1, y2, y3)
+points(tt,y3,type='p', col="red")
+library(nlsr)
+sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
+print(sol0)
+sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
+print(sol1)
+sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
+print(sol2)
+sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
+print(sol3)
+```
+
+The following is a larger dataset version of this test.
+
+```{r settest150, eval=FALSE}
+np <- 150
+tt <- (1:np)/10
+yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
+set.seed(123456)
+ev <- runif(np)
+ev <- ev - mean(ev)
+y1 <- yy + ev
+y2 <- yy + 5*ev
+y3 <- yy + 10*ev
+lg3d150 <- data.frame(tt, yy, y1, y2, y3)
+np <- 1500
+tt <- (1:np)/100
+yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
+set.seed(123456)
+ev <- runif(np)
+ev <- ev - mean(ev)
+y1 <- yy + ev
+y2 <- yy + 5*ev
+y3 <- yy + 10*ev
+lg3d1500 <- data.frame(tt, yy, y1, y2, y3)
+f0 <- yy ~ a0/(1+b0*exp(-c0*tt))
+f1 <- y1 ~ a1/(1+b1*exp(-c1*tt))
+f2 <- y2 ~ a2/(2+b2*exp(-c2*tt))
+f3 <- y3 ~ a3/(3+b3*exp(-c3*tt))
+```
 

### 2022-01-27 PkgFromRbase/PkgFromRbaseBlog.Rmd
blog version

--- 2022-01-19 PkgFromRbase/PkgFromRbase.Rmd
@@
 author: 
-   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
-   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
+   - John C. Nash, retired professor, Telfer School of Management, University of Ottawa
+   - Arkajyoti Bhattacharjee, Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur
    - based on communications from Duncan Murdoch
@@
 output:
-  pdf_document: default
+  html_document: default
 ---
@@
 important part of the packaging. It could be worthwhile to document each
-elements in that file needed for a package such as `nlspkg`, but at the
+element in that file needed for a package such as `nlspkg`, but at the
 time of writing, this is yet to be carried out. However, we must note 

### 2022-05-27 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Start edit again May 2022

@@
 
-```{r settest15}
+```{r settest15, fig.width=5}
 # set parameters
@@
 yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-plot.new()
-plot(tt, yy, type='l')
+# plot.new()
+plot(tt, yy)
+#plot(tt, yy, type='l')
 set.seed(123456)

### 2022-05-27 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Minor edits to get clean output

@@
 author: 
-   - John C Nash, retired professor, University of Ottawa
-date: "16/02/2021 revised 20/01/2022"
+   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
+   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
+date: "16/02/2021 revised 27/05/2022"
 output: 
@@
 in the Nashlib collection associated with @jncnm79 and with similar packages
-for the R Project for Statistical Computing, the author realized that an overview
+for the R Project for Statistical Computing, the authors realized that an overview
 of this variety could be instructive for numerical software designers.
@@
 
-Given a function that can be written as a sum of squared residuals
-
-$$  f(x) = \sum_{i=1}^n {r_i(x)^2} = r^t r $$
-
-where $x$ is a set of real valued parameters, we wish to find the minimum of $f(x)$
-with respect to x. There may be additional conditions that add to the complexity of
+Numerical analysts will generally want to solve a **nonlinear least squares** problem,
+that is, minimize an objective function with respect to a set of $n$ parameters, $x$, where the 
+objective  $f()$ can be written as a sum of $m$ squared elements.
+
+$$  f(x, Y) = \sum_{i=1}^m {r_i(x,Y)^2} $$
+
+where $Y$ represents the ensemble of **exogenous** data that is not changed by alteration
+of the parameters $x$. The functions $r_i()$ need not have the same form. An example is
+the well know Rosenbrock banana shaped valley, which has $n = n = 2$ and
+
+$$ r_1(x) = 10*(x_2 - x_1 ^ 2) $$
+
+$$ r_2(x) = (1 - x_1) $$
+
+There is no exogenous data here.
+
+There may be additional conditions that add to the complexity of
 the problem such as
@@
 
+Many other researchers will want to solve a **nonlinear regression** problem, where
+the $r_i()$ are all of the same form. The (unscaled) Hobbs weed infestation problem (@jncnm79, page 120) uses
+
+$$ r_i(x) = x_1/(1+x_2 * exp(-x_3 * i)) - y_i$$
+
+where the exogenous data is a single vector of numbers $y$. In nonlinear regression, we are mostly
+interested in "fitting a model to data", and the "minimizing a sum of squares" is the approach.
+
+While the term "nonlinear" devolves from functions where parameters appear to degree 1, a more
+useful classification is according to whether a fixed number of computations can estimate the
+parameters. That is, problems can be considered "nonlinear" if we cannot compute the best parameter values
+in a pre-defined fixed number of computations and must use an iterative approach.
+
+In what follows, we will primarily take the numerical analyst's viewpoint. That is, we will be looking
+mostly at the calculations for the minimization of the sum of squares. However, the purpose of these
+calculations can dictate much of the structure of the software in which such computations appear.
+
 # Underlying algorithms
@@
 
-An overview of this method is provided by Wikipedia: 
-\url{https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm}
+An overview of the method is provided by Wikipedia: *https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm*.
+
 The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
@@
 derivative) is zero. In multiple dimensions, this is the same as
-having the gradient $g$ stationary. The so-called **Newton method**
+having the gradient $g$ stationary. The so-called Newton method
 uses the second derivatives in the Hessian matrix $H$ defined as
@@
 to them in a very special case and in a manner unrecognizable as that described
-above.  (See \url{https://en.wikipedia.org/wiki/Newton%27s_method}).
-
-For our present situation, the $f(x)$ can be written as a sum of squares as above. It is worth
-repeating the vector form of this
-
-$$  f(x) = r^t r $$
-
-because it is less fussy to write and read. This is clear when we expand the evaluation
-of the Hessian of this sum of squares function.
+above.
+
+There are some adjustments when our function $f(x)$ can be written as a sum of squares
+
+$$  f(x) = \sum_{i=1}^n {r_i(x)^2} = r^t r $$
+
+The final vector product form is the one I favour because it is the least fussy to write and read.
+In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is
 
@@
 Newton equations could be approximated by ignoring
-the second term (which have elements of the residual vector times second derivatives 
+the second term (which has elements of the residual vector times second derivatives 
 of the residual).  This gives the **Gauss-Newton**
@@
 $$ x_{new} = x + \alpha * \delta $$
+
+Clearly there are many tactical choices that give rise to a variety of particular algorithms.
+One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
+there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
+be perpendicular to any downhill direction.
 
@@
 
-<!-- ?? Should we explain more. Possibly incorporate into code? -->
-
-Clearly there are many tactical choices that give rise to a variety of particular algorithms.
-One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
-there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
-be perpendicular to any downhill direction.
-
 ## Marquardt stabilized Gauss-Newton
@@
 available, many papers were written with computational
-ideas that were given at most a cursory trial. (In 1984, I had the pleasure of showing
-Marquardt my own BASIC program running on a Radio Shack TRS80 Model 100 and he  
-commented on how the computations were converging to a solution as we watched
+ideas that were given at most a cursory trial. (In 1984, JN had the pleasure of showing
+Marquardt his own BASIC version of this program running on a Radio Shack TRS80 Model 100 and 
+Marquardt commented on how the computations were converging to a solution as we watched
 the iterations proceed.)
@@
 matrix that is the square root of $\lambda$ times the unit matrix $1_n$. Then the
-Levenberg-Marquardt equations are equivalent to
+Levenberg-Marquardt equations are equivalent to solving the linear least squares problem
 
@@
 a well-considered approach. However,
-I have never seen a case where the $OFFSET$ approach caused difficulties, and it
+We have never seen a case where the $OFFSET$ approach caused difficulties, and it
 is trivial to code. When $x$ is tiny, we compare $OFFSET$ to 
@@
 
-<!-- ?? Check the statements about how things are done and redo tests -->
 Note that we form $J^t J$ in `nlsr::nlfb()` to provide the $D$ matrix for 
@@
 To avoid the formation of $D$, an experimental package
-`nlsralt` was been built to add functions `nlxbx()` and `nlfbx()` that
-use ONLY the fixed element (i.e., unit matrix) stabilization. A
-preliminary timing (using @microbench13) with the Hobbs unscaled problem 
-(@cnm79, Example 12.2), presented later, shows
-a slight advantage for `nlsralt` in execution timing, but we caution
-that this is but a single example.
+`nlsralt` has been built to add functions `nlxbx()` and `nlfbx()` that
+use ONLY the fixed element (i.e., unit matrix) stabilization. A very
+preliminary timing ([2021-6-26]) using the Hobbs unscaled problem shows
+a slight improvement for the simplified version in timing with equivalent
+solution. More extensive testing is, we feel, warranted to show if this
+approach offers a useful gain in performance while preserving robustness
+in finding the parameter estimates.
 
@@
 
+Newton (and his sometime associated worker Raphson) used the algorithm we attribute
+to them applied to a very special case and in a manner unrecognizable as that described
+above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*).
+
 The main attraction of the Gauss-Newton method is that the Jacobian only needs first
 derivatives, which are generally much, much easier to compute than the full second
-derivatives of the Hessian. However, there have been suggested
+derivatives of the Hessian. However, there have been suggestioned
 Newton-like approaches, possibly for special types of models where some short cut
-to the Hessian is available. I regard these approaches as part of the
-computational infrastructure of the special cases, as it is usually 
-difficult to apply them outside of their context.
-
-Two such examples are the R packages `maxLik` (@maxLik2011) and `trust` (@p-trust).
-
-<!-- expand ??, include some BHHH stuff -->
+to the Hessian is available. These may or may not make sense in the context of
+a given user or group of users. I regard these approaches as part of the
+computational infrastructure of the special cases rather than as general nonlinear
+least squares methods. The `maxLik` package (@maxLik2011) is a prime example.
 
@@
 
-We have a versions of the Nashlib Algorithm 23 in BASIC, Fortran, Pascal, and R, 
-with Python pending. There may be dialects of these programming languages also,
-giving rise to other variations.
+We have a versions of the Nashlib Algorithm 23 in BASIC, Fortran, Pascal, Matlab/Octave and R, 
+with Python under consideration. There may be dialects of these programming languages,
+giving rise to other variations. Features of these computing environments can greatly
+ease or hinder the building of well-structured codes. See
+*git@github.com:pcolsen/Nash-Compact-Numerical-Methods.git*.
 
@@
 
-The `nls()` function has an argument `subset` and a test included in the 
-`Croucher-example1.R` file of this project showed this works as expected. 
+The `nls()` function has an argument `subset`.  Here is an example of its use.
+
+```{r croucher}
+# Croucher-example1.R -- https://walkingrandomly.com/?p=5254
+
+xdata = c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
+ydata = c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
+Cform <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) # formula
+Cstart<-list(p1=1,p2=0.2) # starting parameters
+Cdata<-data.frame(xdata, ydata) # data frame of the data
+Csubset<-3:8 # index of rows to use from data frame
+
+# fit using all the data
+fitfull<-nls(Cform, data=Cdata, start=Cstart)
+fitfull
+# fit using subset data
+fitsub<-nls(Cform, data=Cdata, subset=Csubset, start=Cstart)
+fitsub
+```
+
+A more extensive version of this with explicit extraction of the data 
+showed this works as expected. 
 However, nowhere have we found a published example of the use of the 
-`subset` argument with `nls()`. Worse, it clearly must be applied to
+`subset` argument with `nls()`. Worse, it appears to be applied to
 many of the functions used by `nls()` but makes no explicit appearance. Thus it
-is VERY difficult to ensure subsetting is correctly applied when we wish to modify
-the code, for instance, to provide for newer features especially more reliable
-solvers.
+is VERY difficult to ensure subseting is correctly applied when we wish to modify
+the code, for instance, to provide for more reliable solvers.
 
 Note that the `subset` ARGUMENT to `nls()` is not the same as the `subset()` function
-that is usable to restrict the scope of an R object.
-
-OUR HYPOTHESIS: `nls()` uses `subset` to automatically adjust the data to which all
-the relevant functions are applied. We note that `minpack.lm::nlsLM` also has a `subset` argument,
-but package `nlsr` does not at time of writing (2021-7-12), but will use the **weights** mechanism
-of the next subsection of this document at some future release.
+that is usable to restrict the scope of an R object. The potential confusion that a
+common name implies is an additional concern.
 
@@
 
-In order to provide a sane and well-understood mechanism for subsetting, there
+In order to provide a sane and well-understood mechanism for subseting, there
 are alternatives for implementation.
@@
   non-zero weights e.g, `sum(weights > 0). Incidentally, this underlines the 
-  need to ensure weights are non-negative. ??now in nlsj, need to add to nlsr code.??
+  need to ensure weights are non-negative.
 
 - **via preprocessing**: We could prepare a function to pre-process the problem and
-  create a data collection that is appropriately subset. This may offend against
+  create a dataframe that is appropriately subset. This may offend against
   the spirit of R/S, but it has the merit that we can easily reassure ourselves
-  that we have worked with the correct information, correctly used in our methods.
-  ?? Should add this as an option.
-
-PROPOSAL: use the "weights" approach if `nlsj` is called with `subset` active, 
-else recommend the preprocessing approach. The preprocessor could also do a lot
-of checking, which we might leave out of `nlsj()` in some cases. As of 2021-7-12,
-the weights mechanism is working in a trial `nlsj()` and should be transferrable
-to package `nlsr` fairly easily.
+  that we have worked with the correct information.
+
+  <!-- ?? Should add this as an option. -->
+
+<!-- PROPOSAL: use the "weights" approach if `nlsj` is called with `subset` active,  -->
+<!-- else recommend the preprocessing approach. The preprocessor could also do a lot -->
+<!-- of checking, which we might leave out of `nlsj()` in some cases. As of 2021-7-12, -->
+<!-- the weights mechanism is working in a trial `nlsj()` and should be transferrable -->
+<!-- to package `nlsr` fairly easily. -->
 

### 2022-05-27 nlsj/R/nlsj.R
Add nlsj package (not complete)

@@ -1,0 +1,507 @@
+nlsj <- function (formula, data = parent.frame(), start=NULL, 
+            control = nlsj.control(),
+            algorithm = "default", weights=NULL, subset=NULL, trace = FALSE,
+            na.action, model=FALSE, lower = -Inf, upper = Inf, ...) {
+# ?? left out -- FIXME??    na.action, model = FALSE, (masked from nlxb)
+# ?? at this stage ONLY treat "default", but will add bounds
+# ?? data in .GlobalEnv -- should be OK
+##?? Should a lot of this material be in nlsjModel() to build tools for problem??
+# DOES NOT CALL nlsjModel, but does everything here
+
+##?? ... args may NOT be well-defined for this function. CAUTION  
+  
+# Controls
+   if (is.null(control$derivmeth)) control$derivmeth="default" # for safety
+   epstol <- (.Machine$double.eps * control$offset) 
+   epsh <- sqrt(epstol)
+   epstol4 <- epstol^4 # used for smallsstest
+   ##?? may want these in nlsj.control
+   if (control$derivmeth == "numericDeriv") warning("Forcing numericDeriv")
+#   cat("control:"); print(control)
+#   tmp <- readline("cont.")
+#   control$watch<-TRUE
+
+# Algorithm
+   if (is.null(algorithm)) algorithm<-"default"
+   algchoice <- c("default", "port", "plinear", "marquardt")
+   alg <- which(algorithm %in% algchoice)
+   switch(alg,
+      "default" = if (trace) cat("nlsj: Using default algorithm\n"),
+      "marquardt" = if (trace) cat("nlsj: Using marquardt algorithm\n"),
+      # all other choices
+      { msg <- paste("Algorithm choice '",alg,
+                     "' not yet implemented or does not exist")
+        stop(msg) }
+   ) # end switch choices
+
+getlen <- function(lnames) {
+      nn<-length(lnames)
+      ll<-rep(NA,nn)
+      for (i in 1:nn) ll[i]=length(get(lnames[i]))
+      ll
+}
+   
+# Data
+   stopifnot(inherits(formula, "formula"))
+   vnames <- all.vars(formula) # all names in the formula
+   if (missing(data)) {  ## rather than   #  if (is.null(data)) {
+      warning("Data is not declared explicitly. Caution!")
+      data <- environment(formula) # this will handle variables in the parent frame??
+      dnames <- vnames[which(vnames %in% ls(data))]
+      ldata<-getlen(vnames) # lengths of data
[…]

### 2022-05-27 nlsj/inst/XCrouchBd.R

@@ -1,0 +1,71 @@
+# XCrouchBd.R
+
+rm(list=ls())
+xdata = c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
+ydata = c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
+Cform <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
+Cstart<-list(p1=1,p2=0.2)
+Cdata<-data.frame(xdata, ydata)
+library(nlsj)
+library(nlsr)
+library(minpack.lm)
+
+
+Clower<-c(0,0)
+Cupper<-c(1.5, 1.5)
+
+fjb<-nlsj(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper)
+fjb
+fjbm<-nlsj(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper, algorithm="marquardt")
+fjbm
+fnb<-nls(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper, algorithm="port")
+fnb
+fxb<-nlxb(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper)
+fxb
+cbrj<-model2rjfun(Cform, data=Cdata, pvec=Cstart)
+fmb<-nlsLM(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper)
+fmb
+crossprod(cbrj(coef(fxb)))
+deviance(fnb)
+coef(fnb)
+deviance(fjbm)
+coef(fjbm)
+crossprod(cbrj(coef(fjbm)))
+deviance(fmb)
+coef(fmb)
+crossprod(cbrj(coef(fxb)))
+coef(fxb)
+# ----------------------------------------------------------------
+# Now change the bounds so p2 is smaller than allowed, p1 allowed bigger
+Cstart<-list(p1=1,p2=0.2)
+
+Clower<-c(0.4, 0.2)
+Cupper<-c(2.5, 0.5)
+
+# Note that "default" does not get good bounds, marquardt better
+fjb<-nlsj(Cform, data=Cdata, start=Cstart, trace=TRUE, lower=Clower, upper=Cupper)
+fjb
[…]

### 2022-05-27 nlsj/tests/nlsj.R

@@ -1,0 +1,397 @@
+#  File src/library/stats/tests/nlsj.R
+#  Part of the R package, https://www.R-project.org
+#
+#  This program is free software; you can redistribute it and/or modify
+#  it under the terms of the GNU General Public License as published by
+#  the Free Software Foundation; either version 2 of the License, or
+#  (at your option) any later version.
+#
+#  This program is distributed in the hope that it will be useful,
+#  but WITHOUT ANY WARRANTY; without even the implied warranty of
+#  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
+#  GNU General Public License for more details.
+#
+#  A copy of the GNU General Public License is available at
+#  https://www.R-project.org/Licenses/
+
+## tests of nlsj, especially of weighted fits
+library(nlsj) # ??added 210716 JN
+
+library(stats)
+options(digits = 5) # to avoid trivial printed differences
+options(useFancyQuotes = FALSE) # avoid fancy quotes in o/p
+options(show.nlsj.convergence = FALSE) # avoid non-diffable output
+options(warn = 1)
+
+have_MASS <- requireNamespace('MASS', quietly = TRUE)
+
+pdf("nlsj-test.pdf")
+
+## utility for comparing nlsj() results:  [TODO: use more often below]
+.n <- function(r) r[names(r) != "call"]
+
+## selfStart.default() w/ no parameters:
+logist <- deriv( ~Asym/(1+exp(-(x-xmid)/scal)), c("Asym", "xmid", "scal"),
+		function(x, Asym, xmid, scal){} )
+logistInit <- function(mCall, LHS, data) {
+    xy <- sortedXyData(mCall[["x"]], LHS, data)
+    if(nrow(xy) < 3) stop("Too few distinct input values to fit a logistic")
+    Asym <- max(abs(xy[,"y"]))
+    if (Asym != max(xy[,"y"])) Asym <- -Asym  # negative asymptote
+    xmid <- NLSstClosestX(xy, 0.5 * Asym)
[…]

### 2022-05-27 nlsj/R/nlsj.R
Attempts to fix nlsj -- not working yet as the identification of variables and parameters is incorrect for linear and quadratic models. Why??

@@ -1,5 +1,5 @@
 nlsj <- function (formula, data = parent.frame(), start=NULL, 
-            control = nlsj.control(),
+            control = list(),
             algorithm = "default", weights=NULL, subset=NULL, trace = FALSE,
             na.action, model=FALSE, lower = -Inf, upper = Inf, ...) {
 # ?? left out -- FIXME??    na.action, model = FALSE, (masked from nlxb)
@@ -10,6 +10,18 @@ nlsj <- function (formula, data = parent.frame(), start=NULL,
 
 ##?? ... args may NOT be well-defined for this function. CAUTION  
   
+ctrl <- nlsj.control()
+ncontrol <- names(control)
+nctrl <- names(ctrl)
+for (onename in ncontrol) {
+    if (onename %in% nctrl) {
+       if (! is.null(control[onename]) || ! is.na(control[onename]) )
+       ctrl[onename]<-control[onename]
+    }
+}
+control<-ctrl
+ctrl<-NULL
+
 # Controls
    if (is.null(control$derivmeth)) control$derivmeth="default" # for safety
    epstol <- (.Machine$double.eps * control$offset) 
@@ -48,9 +60,8 @@ getlen <- function(lnames) {
       warning("Data is not declared explicitly. Caution!")
       data <- environment(formula) # this will handle variables in the parent frame??
       dnames <- vnames[which(vnames %in% ls(data))]
-      ldata<-getlen(vnames) # lengths of data
-      dnames <- vnames[which(ldata > 1)]
-      pnames <- vnames[which(ldata == 1)] # could 
+      ldata <- getlen(dnames)
+      pnames <- dnames[ which(ldata == 1)] 
    }
    else if (is.list(data)){
           data <- list2env(data, parent = environment(formula))
@@ -349,6 +360,8 @@ getlen <- function(lnames) {
       # ?? Don't continue with Gauss-Newton; delta can't be computed
       # ?? marquardt is using J (augmented Jacobian) in convergence test. This may
       # ?? not be exactly what we want. Leave it there for now.
+      cat("control$smallsstest =",control$smallsstest,"\n") # ??diagnostic
+      cat("ssmin and epstol4:",ssmin," ",epstol4,"\n")      # ??diagnostic
       convInfo <- convCrit() # This is the convergence test
       if (control$watch) print(convInfo)
       if (marqalg && trace) cat("slam =",slam," ")

### 2022-05-27 nlsj/R/nlsjextra.R

@@ -48,7 +48,7 @@ print.nlsj <- function(x, digits = max(3L, getOption("digits") - 3L), ...)
 	"residual sum-of-squares: ", format(x$m$deviance(), digits = digits),
 	"\n", sep = "")
 #    .p.nlsj.convInfo(x, digits = digits)
-    print(x$convInfo)
+#    print(x$convInfo)
     invisible(x)
 }
 
@@ -84,7 +84,7 @@ print.summary.nlsj <-
     }
 
 #    .p.nlsj.convInfo(x, digits = digits)
-    print(x$convInfo)
+#    print(x$convInfo)
 
     if(nzchar(mess <- naprint(x$na.action))) cat("  (", mess, ")\n", sep = "")
     cat("\n")

### 2022-05-27 nlsj/tests/nlsj.R

@@ -55,7 +55,7 @@ plot(x,y)
 curve(a+b*x+c*x^2, add = TRUE)
 ## IGNORE_RDIFF_BEGIN ## nlsj does not call "port" (yet??)
 ## nlsj(y ~ a+b*x+c*I(x^2), start = c(a=1, b=1, c=0.1), algorithm = "port")
-nlsj(y ~ a+b*x+c*I(x^2), start = c(a=1, b=1, c=0.1))
+nlsj(y ~ a+b*x+c*x^2, start = c(a=1, b=1, c=0.1))
 # ?? (fm <- nlsj(y ~ a+b*x+c*I(x^2), start = c(a=1, b=1, c=0.1),
 # ??           algorithm = "port", lower = c(0, 0, 0)))
 ## IGNORE_RDIFF_END

### 2022-05-27 nlsj/R/nlsj.R
Seem to have fixed extraction of data and parameter names from formula and data, but probably will find other examples that fail

@@ -46,22 +46,41 @@ ctrl<-NULL
         stop(msg) }
    ) # end switch choices
 
-getlen <- function(lnames) {
-      nn<-length(lnames)
-      ll<-rep(NA,nn)
-      for (i in 1:nn) ll[i]=length(get(lnames[i]))
-      ll
+getlen <- function(ln) {
+  nn<-length(ln)
+  ll<-rep(NA,nn)
+  for (i in 1:nn) ll[i]=length(get(ln[i]))
+  ll
+}
+
+datpar <- function(frm){ # return dn and pn for data and parameter names
+  # check formulas
+  vn <- all.vars(frm)
+  cat("vn:"); print(vn)
+  data<-environment(frm)
+  lstdat <- ls(data)
+  cat("lstdat:"); print(lstdat)
+  dn <- vn[which(vn %in% lstdat)]
+  cat("dn:"); print(dn)
+  pn <- vn[- which(vn %in% dn)]
+  cat("pn:"); print(pn)
+  ll <- getlen(dn)
+  cat("ll:"); print(ll)
+  if (any(ll < 2)) {
+     pn<-dn[which(ll < 2)]
+     dn <- dn[ -which(dn %in% pn) ]
+  }
+  res<-list(dn=dn, pn=pn)
+  res
 }
    
 # Data
    stopifnot(inherits(formula, "formula"))
    vnames <- all.vars(formula) # all names in the formula
    if (missing(data)) {  ## rather than   #  if (is.null(data)) {
-      warning("Data is not declared explicitly. Caution!")
-      data <- environment(formula) # this will handle variables in the parent frame??
-      dnames <- vnames[which(vnames %in% ls(data))]
-      ldata <- getlen(dnames)
-      pnames <- dnames[ which(ldata == 1)] 
+      dp <- datpar(formula)
+      dnames <- dp$dn
+      pnames <- dp$pn
    }
    else if (is.list(data)){
           data <- list2env(data, parent = environment(formula))

### 2022-05-28 RefactoringNLS/RefactoringNLS.Rmd
Edits and cleanup 20220528

@@
     pdf_document:
-        keep_tex: false
+        keep_tex: true
         toc: true
@@
 
-This article reports the particular activities of our Google Summer of Code
-project "Improvements to `nls()`" that relate to the R code for that function, which
-is intended for the estimation of models written as a formula that has at least
-one parameter that is "nonlinear", that is, not estimable via solving a set of 
+This article considers the features and limitations of the R function `nls()`. It
+arose from a Google Summer of Code project "Improvements to `nls()`", a tool 
+intended for the estimation of models written as a formula that has at least
+one parameter that is "nonlinear". That is, the model is not estimable via solving a 
+finite set of 
 linear equations. A companion document **Variety in Nonlinear Least Squares Codes** 
@@
 problem of minimizing a function that can be written as a sum of squared
-terms.
-
-Our work has not fully addressed all the issues that we would like to see
-resolved, but we believe we have made sufficient progress to demonstrate that
-there are worthwhile improvements that can be made to the R function `nls()`.
+terms, and of the even more general case of nonlinear regression.
+
 An important overall consideration in our work has been the maintainability of
 the code base that supports the `nls()` functionality, as we believe that the
-existing code makes maintenance and improvement very difficult.
+existing code makes maintenance and improvement very difficult. Moreover, legacy
+applications and examples mean some improvments are blocked. Thus we have not
+addressed all the issues raised, but we believe that the discussion and trials
+so far are relevant to improving both `nls()` and R generally.
 
@@
 our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
-us review some of the issues. We will then propose corrective actions, some of which we have carried out.
+us review some of the issues. We will then propose corrective actions, some of which we have carried out
+either in demonstrations or in the distributed coed.
 
@@
 A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
-(i.e., perfect fit) data was corrected thanks to suggestions by one of us. The release 
-notes for R 4.1.0 state:
-
- - For nls(), an optional constant scaleOffset may be added to the denominator of the relative offset
-   convergence test for cases where the fit of a model is expected to be exact, thanks to a proposal by 
-   John Nash. nls(*, trace=TRUE) now also shows the convergence criterion.
-
-Within the previous standard documentation (**man** or ".Rd" file) `nls()` warned
-
->  **The default settings of nls generally fail on artificial “zero-residual” data problems.**
-
->  The nls function uses a relative-offset convergence criterion that compares the numerical imprecision at
-  the current parameter estimates to the residual sum-of-squares. This performs well on data of the form $$y = f(x, \theta) + eps$$
-
->  (with var(eps) > 0). It fails to indicate convergence on data of the form $$y = f(x, \theta)$$
-
->  because the criterion amounts to comparing two components of the round-off error. To avoid a zero-divide in
-  computing the convergence testing value, a positive constant `scaleOffset` should be added to the denominator
-  sum-of-squares; it is set in control; this does not yet apply to algorithm = "port".
-
-It turns out that this issue can be quite easily resolved. The key "convergence test" -- more properly
-a "termination test" for the **program** rather than testing for convergence of the underlying **algorithm** --
+(i.e., perfect fit) data was corrected thanks to suggestions by one of us.
+
+In the manual page for `nls()` in R 4.0.0 there is the warning
+
+  **Do not use `nls` on artificial "zero-residual" data.**
+
+  The `nls` function uses a relative-offset convergence criterion
+  that compares the numerical imprecision at the current parameter
+  estimates to the residual sum-of-squares.  This performs well on data of
+  the form
+
+  $$  y = f(x, \theta) + eps $$
+
+  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form
+
+  $$  y = f(x, \theta)  $$
+
+  because the criterion amounts to comparing two components of the round-off 
+  error.
+
+  If you wish to test `nls` on artificial data please add a noise component, 
+  as shown in the example below.
+
+This amounts to admitting R cannot solve perfectly well-posed problems. The 
+suggestion that one needs to pollute data with errors is one that should be offensive
+to good science.
+
+It turns out that this issue can be easily resolved. The key "convergence test" -- more properly
+a "termination test" for the **program** rather than for convergence of the underlying **algorithm** --
 is the Relative Offset Convergence Criterion (see @BatesWatts81). This works by projecting the proposed
 step in the parameter vector on the gradient and estimating how much the sum of squares loss function 
-will decrease. To avoid scale issues, we use the current size of the loss function as a measure and divide
-by it. When we have "converged", the estimated
-decrease is very small, as usually is its ratio to the sum of squares. However, in some cases we have the 
-possibility of an exact fit and the sum of squares is (almost) zero and we get the possibility of a 
+will decrease. This is divide by the current size of the loss function to avoid scale issues.
+When we have "converged", the estimated
+decrease is very small, as usually is its ratio to the sum of squares. However, with small residuals,
+the sum of squares loss function is (almost) zero and we get the possibility of a 
 zero-divide failure.
 
-The issue is easily resolved by adding a small quantity to the loss function. To preserve legacy behaviour,
-in 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
-with a default value of zero for legacy behaviour. Setting it to a small number -- 1.0 is 
+Adding a small quantity to the loss function before dividing avoids trouble. 
+In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
+with a default value of zero. Setting it to a small number -- 1.0 is 
 a reasonable choice --
 allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
-**safeguarded relative offset convergence criterion**.
-
-As noted, we are pleased to report that this improvement is now in the R distributed code as of version 4.1.0.
+**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
+We are pleased to note that this improvement is now in the R distributed code as of version 4.1.0.
 
@@
 ```
-rm(list=ls())
 t <- -10:10
@@
 ```
-> rm(list=ls())
 > t <- -10:10
@@
 > plot(t,y)
+> lstartbad<-c(a=1, b=1, c=1)
 > lstart2<-c(a=100, b=10, c=1)
@@
 c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
-# Note that this has succeeded. The test in nlsr recognizes small residual problems.
 > nls(lform, data=ldata, start=lstart2, trace=TRUE)
@@
 1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
+... (approx 40 lines omitted)
 1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
@@
   number of iterations exceeded maximum of 50
-
+> # Fix with scaleOffset
 > nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
@@
  residual sum-of-squares: 4.7e-19
-
 Number of iterations to convergence: 8 
@@
 
-> sessionInfo()
-R version 4.1.0 (2021-05-18)
-Platform: x86_64-pc-linux-gnu (64-bit)
-Running under: Linux Mint 20.2
+Running under: Linux Mint 20.3
 ```
@@
 squares** test (**smallsstest**) that compares the latest evaluated sum of squared
-(weighted) residuals to a very small multiple of the initial sum of squares. The
-multiple uses a control setting `offset` which defaults to 100.0 and we compute
-the 4th power of the machine epsilon times this offset.
+(weighted) residuals to `e4` times the initial sum of squares. The value of 
+`e4` is computed as
 
@@
 
-We do note that `nls()` stops after `maxiter` "iterations". However, for almost all
+We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
+Unfortunately, for almost all
 iterative algorithms, the meaning of "iteration" requires careful examination of the
-code. Instead, we prefer to record the number of times the residuals or the jacobian have
+code. we prefer to express such tests using the number of times the residuals or the jacobian have
 been computed and put upper limits on these. Our codes exit (terminate) when these limits
@@
 `numericDeriv()` function that is a part of `nls()`. `nlsr` can use analytic
-derivatives, and we can import this functionality to the `nls()` code as an
+derivatives, and we could import this functionality to the `nls()` code as an
 improvement. See below in the section **Jacobian computation**.
 
-However, the more common source of the issue is that the Jacobian is very close
+A common source of the issue is that the Jacobian is very close
 to singular for some values of the model parameters. In such cases we need to find
@@
 @jn77ima). Versions of this have been implemented in packages `minpack.lm` and `nlsr`.
-and we have prepared very preliminary versions of an `nls` replacement that can incorporate a
-version of the Levenberg-Marquardt stabilization. (There are some issues of 
-integration with other code structures and of complexity of the computations that
-suggest we should use a simplified LM stabilization.) Unfortunately, making this code
-work across all of the features of `nls()` has not proved straightforward.
-
-<!-- ?? JN: I forget exact details, but I recall that we could not get this to work with -->
-<!-- some ways in which call could be set up. May need to revisit. this. -->
+and we have prepared experimental versions of `nls` replacements that can incorporate
+stabilizations. Integration of such ideas with other code structures so that all the
+features of `nls()` work properly has unfortunately proved difficult.
+
+Let us consider an example which presents the infamous 'singular gradient' error message.
+This was posted on StackExchange. 
+*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*
+
+```{r SESingGrad}
+reala=-3
+realb=5
+realc=0.5
+realr=0.7
+realm=1
+x=1:11 #x values 11 timepoint data
+#linear+exponential function
+y=reala + realb*realr^(x-realm) + realc*x
+testdat=data.frame(x,y)
+strt<-list(a=-3, b=5, c=0.5, r=0.7, m=1)
+jform<-y~a+b*r^(x-m)+c*x # Formula
+linexp=try(nls(jform, data=testdat, start=strt, trace=F))
+library(nlsr)
+# Note singular values of Jacobian in rightmost column
+linexp2<-try(nlxb(jform, data=testdat, start=strt, trace=F))
+linexp2
+```
+
+In the above, we see that the Jacobian is essentially singular as shown by its singular
+values. Note that we DISPLAY the singular values in a single column for convenience in
+the output. The values do NOT correspond to individual parameters, but are a property
+of the whole problem.
 
@@
 attribute of the residual vector. This is implemented as a mix of R and C code, but
-we have created a rather more compact version entirely in R in this Google Summer of
-Code project. See the document **DerivsNLS.pdf**.
-
-Let us consider an example which presents the infamous error message.
-
-```{r code=xfun::read_utf8('SESingGrad.R')}
-```
+we have created a rather more compact version entirely in R.
+Code project. See the companion document **Jacobian Calculations for nls()**
+(?? need citation -- currently **DerivsNLS.pdf**).
+
+The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
+of `nls()` will use a numerical approximation. In the most recent example above,
+the singularity is not due to the numerical approximation.
 
@@
 defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
-is within the `nlsModel()` or `nlsModel.pliner()` functions.
+is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
+almost certainly contributes to the performance of `nls()` on selfStart models.
 
@@
 we would implement subsetting by means of zero-value weights, with observation counts
-(and degrees of freedom) computed via the numbers of non-zero weights.
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
 
@@
 
-A succinct description of the issue is given in: https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r
+A succinct description of the issue is given in: *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*
 

### 2022-05-28 RefactoringNLS/testplot.Rmd
Edits and cleanup 20220528

---
title: "testplot"
author: "John C. Nash"
date: '2022-05-21'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown

This is an R Markdown document. Markdown is a simple formatting syntax for authoring HTML, PDF, and MS Word documents. For more details on using R Markdown see <http://rmarkdown.rstudio.com>.

When you click the **Knit** button a document will be generated that includes both content as well as the output of any embedded R code chunks within the document. You can embed an R code chunk like this:

## Including Plots

You can also embed plots, for example:

```{r settest15, fig.width=6}
# set parameters
set.seed(123456)
a <- 1
b <- 2
c <- 3
np <- 15
tt <-1:np
yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
# plot.new()
# plot(tt, yy)
plot(tt, yy, type='l')
set.seed(123456)
ev <- runif(np)
ev <- ev - mean(ev)
y1 <- yy + ev
points(tt,y1,type='p', col="green")
y2 <- yy + 5*ev
points(tt,y2,type='p', col="blue")
y3 <- yy + 10*ev
lg3d15 <- data.frame(tt, yy, y1, y2, y3)
 points(tt,y3,type='p', col="red")
library(nlsr)
sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
print(sol0)
sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
print(sol1)
sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
print(sol2)
sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
print(sol3)
```

### 2022-05-29 RefactoringNLS/RefactoringNLS.Rmd
add refs to bibliography inc. R-bloggers JN and RB PkgFromRbase22. Edits and improvements to Refactoring

@@
 The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
-of `nls()` will use a numerical approximation. In the most recent example above,
-the singularity is not due to the numerical approximation.
-
-It should be noted that the selfStart models in the `./src/library/stats/R/zzModels.R` file
+of `nls()` (and, by extension, `minpack.lm::nlsLM`) will use a numerical approximation. 
+This can lead to different behaviour:
+
+- A pasture regrowth problem (@Huet2004, page 1, based on @Ratkowsky1983) has a poorly
+  conditioned Jacobian and `nls()` fails with "singular gradient". 
+  Worse, numerical approximation to the Jacobian give the error
+  "singular gradient matrix at initial parameter estimates" for `minpack.lm::nlsLM`
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of `nlsr::nlxb` give a solution.
+
+- Karl Schilling (private communication) provided an example where a model specified
+  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
+  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
+  this is undefined. In such cases, we observed that `nls()` and `minpack.lm::nlsLM`
+  found a solution, though it can be debated whether such lucky accidents can be
+  taken as an advantage.
+
+Note that the selfStart models in the `./src/library/stats/R/zzModels.R` file
 provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
@@
 element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
-be confusing.) However, the model frame is used within the function code in the form of an object
+be confusing.) However, the model frame is used within the function code in the form of the object
 `mf`.
@@
 
-Nevertheless, to provide compatible behaviour with `nls()`, we will need to ensure
-that equivalent behaviour is guaranteed. Furthermore, we need to test that the 
+Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
+need to ensure equivalent behaviour, but users should test that the 
 operation is correct.
@@
 such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
-`NRAIA`. Unfortunately, the structure of the programming of 
-these is such that the methods by which initial parameters are computed is entangled
-with the particularities of the `nls()` code. Though there is a `getInitial()` function,
+`NRAIA`. Unfortunately, the structure of **selfStart** codes
+is such that the methods by which initial parameters are computed is entangled
+with the particulars of the `nls()` code. Though there is a `getInitial()` function,
 this is not easy to use to simply compute the initial parameter estimates, in part 
@@
 of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
-is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$
-
-The example shows how these starting parameters can be transformed to those of 
-another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$
+is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$, but we show how the starting parameters
+for this model can be transformed to those of 
+another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$.
 
@@
 
-We note that the function includes analytic expressions for the Jacobian ("gradient").
-These could be possibly be useful to R users, especially if documented. Moreover, we
+This R function includes analytic expressions for the Jacobian ("gradient").
+These could be be useful to R users, especially if documented. Moreover, we
 wonder why the programmers have chosen to save so many quantities in "hidden"
@@
 
-Because the Gauss-Newton algorithm is rather unreliable from many starting sets
+Because the Gauss-Newton algorithm is unreliable from many starting sets
 of parameters, selfStart models are more than an accessory to `nls()` but a part
@@
 
-The nls object contains some elements that are awkward to produce by other algorithms.
-Moreover, some information that would be useful is not presented obviously.
+The nls object contains some elements that are awkward to produce by other algorithms,
+but some information that would be useful is not presented obviously.
 
@@
 counts of evaluations are also returned. Working on this project showed several potential
-updates to the `nlsr` documentation.
+updates to the `nlsr` documentation. Note that the singular values of the Jacobian are
+actually computed in the `print` and `summary` methods for the result.
 
@@
 Specifying a model to a solver should, ideally, use the same syntax across
-solver tools. Unfortunately, R allows multiple approaches.
+solver tools. Unfortunately, R allows multiple approaches within different
+modelling tools, and within `nls()` itself.
 
@@
 
-However, even within `nls()`, we see annoying inconsistencies. Consider the 
+Within `nls()`, consider the 
 following FOUR different calling sequences for the same problem, though the
-second is to illustrate how one intuitive choice will not work. In this failed
+fm2a -- an intuitive choice -- will not work. In this failed
 attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
 to try to add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing (end of August 2021) we do
-not have a resolution for this issue.
+like to see a consistent syntax. At the time of writing we do
+not foresee a resolution for this issue.
 
@@
 ## the coefficients only:
-coef(fm1DNase1)
+# coef(fm1DNase1)
 ## including their SE, etc:
-coef(summary(fm1DNase1))
+# coef(summary(fm1DNase1))
 
@@
 
-## Issue: code structure
+## Issue: code structure and documentation for maintenance
 
@@
 In particular, the iterative setup is such that introduction of Marquardt stabilization
-is not easily available.
-
-To obtain performance, a lot of the code is in C with consequent calls and returns that
+is not easily available. To obtain performance a lot of the code was in C
+with consequent calls and returns that
 complicate the code. Over time, R has become much more efficient on modern computers, and
 the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
-could be much reduced by moving code entirely to R.
-
-## Issue: code documentation for maintenance
-
-`setPars()` -- explain weaknesses. Only used by `profile.nls()`
-
-The paucity of documentation is exacerbated by the mixed R/C/Fortran code base.
-
+could be reduced by moving code entirely to R.
+
+While R and its packages are generally very well documented for usage, that documentation
+rarely extends as far as it might in directions helpful for code maintenance. 
+The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
+where seemingly trivial differences in things like representation of numbers or
+base index for arrays lead to traps for unwary programmers.
+
+For `nls()` the concern is demonstrated by a short email correspondence.
 Following is an email to John Nash from Doug Bates. This is NOT a criticism
@@
 
-If the R sources were in a git repository I might try to use `git blame` 
-to find out when and by whom that was written but they are in an SVN repository, 
-I think, and I haven't used it for a long, long time.
+...
 
@@
 
-# Goals of our effort
-
-Here are some of the goals we hope to accomplish.
+# Goals of our effort to improve `nls()`
 
@@
   that matches the version in base R or what is packaged in `nlspkg` as described 
-  in the "PkgFromRbase" document.
+  in @PkgFromRbase22.
 

### 2022-05-29 RefactoringNLS/RefactoringNLS.Rmd
More edits -- now at a review point

@@
    - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
-date: "2021-8-20"
+date: "2022-5-29"
 output: 
@@
 result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE)
+# We can display the result with several commands
 # str(result) -- displays large amount of material - suppressed here
 #    as it is too wide for the page
-result
-ls(result) # to list the elements of the output
-ls(result$m) # and in particular the "m" object
+# result # displays the object
+# ls(result) # to list the elements of the output
+# ls(result$m) # and in particular the "m" object
 ```
@@
 like to see a consistent syntax. At the time of writing we do
-not foresee a resolution for this issue.
-
-```{r log4ways, echo=TRUE}
+not foresee a resolution for this issue. In the example we have NOT evaluated
+the commands to save space.
+
+```{r log4ways, echo=TRUE, eval=FALSE}
 DNase1 <- subset(DNase, Run == 1)
@@
 
-For `nls()` the concern is demonstrated by a short email correspondence.
-Following is an email to John Nash from Doug Bates. This is NOT a criticism
-of Prof. Bates work, but a reflection on how difficult it is to develop code in this
-subject area and to keep it maintainable. We have experienced similar loss of understanding
-for some of our own codes.
-
-```
-I'm afraid that I don't know the purpose of the recursive call either.  
-I know that I wrote the code to use a closure for the response, covariates, etc., 
-but I don't recall anything like a recursive call being necessary.
-
+For `nls()` the concern is demonstrated by a short email query from John Nash
+to Doug Bates and his reply.
+This is NOT a criticism of Prof. Bates (or any other) work, but a reflection on 
+how difficult it is to develop code in this subject area and to keep it maintainable. 
+We have experienced similar loss of understanding for some of our own codes.
+
+```
     ...
-
-I don't think I will be of much help.  My R skills have atrophied to the point 
-where I wouldn't even know how to start exploring what is happening in the first 
-call as opposed to the recursive call.
-```
-
-This was in response to an email to Doug bates on Jun 29, 2021 from John Nash.
-
-```
-    Thanks.
-
     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
-    <https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R>
-
-    This has the test problem and the expanded code. Around line 367 is where we are
-    scratching our heads. The function code (from nlsModel()) is in the commented lines below
-    the call. This is
+    ... has the test problem and the expanded code. Around line 367 is where we are
+    scratching our heads. The function code (from nlsModel()) is in the commented 
+    lines below the call. This is
 
@@
     In both there is the recursive call, which must have a purpose I don't understand.
+```
+The reply was
+
+```
+    I'm afraid that I don't know the purpose of the recursive call either.  
+    I know that I wrote the code to use a closure for the response, covariates, etc., 
+    but I don't recall anything like a recursive call being necessary.
+    ...
+
+    I don't think I will be of much help.  My R skills have atrophied to the point 
+    where I wouldn't even know how to start exploring what is happening in the first 
+    call as opposed to the recursive call.
 ```
@@
 
-- try to obtain a cleaner structure for the overall `nls()` infrastructure. By this we
+- to streamline the overall `nls()` infrastructure. By this we
   mean a re-factoring of the routines so they are better suited to maintenance of both
-  the existing `nls()` methods and features as well as the new features we would like
-  to add.
-
-- try to explain what we do, either in comments or separate maintainer documentation.
+  the existing `nls()` methods and features as well as new features we or others
+  would like to add.
+
+- to explain what we do, either in comments or separate maintainer documentation.
   Since we are complaining about the lack of explanatory material for the current
@@
 
-Below in "Consistent specifications of partially linear models" we point out that
-`nls()` uses a different formula specification from the default for a problem if the 
+We have seen that
+`nls()` uses a different formula specification from the default if the 
 `plinear` algorithm is used. This is unfortunate, since the user cannot then simply
-add `algorithm="plinear"` to the call. Moreover, we believe it makes errors more
-likely. We suggest a possible approach to avoiding this issue, but have yet to 
-cast this into working code (2021-8-19).
+add `algorithm="plinear"` to the call. This makes errors more
+likely.
+
+While not consistent with the `lm()` function and specification of linear models, 
+we believe a complete set of parameters should be specified for nonlinear models.
+That is, the model formula should be written down as it would be computed, 
+rather than for estimation. Thus `y ~ a*x + b` rather than `y ~ x`.
+
+Moreover, we want to avoid the partially linear parameters appearing in the result object
+as `.lin1`, `.lin2`, etc., where there actual position in the model is not explicit.
+
+A possible workaround to allow consistency would be to specify the partially linear parameters
+when the `plinear` option is specified. For example, we could use `algorithm="plinear(Asym)"`
+while keeping the general model formula from the default specification. This would allow 
+for the output of different `algorithm` options to be consistent. However, we have not
+yet tried to code such a change.
+
+A further complication of model specification in R is that some formulas require
+the user to surround a term with `I(  )` to inhibit interpretation of arithmetic 
+operators. We would prefer that formulas be usable without this complication, which
+has limited documentation and examples.
+
+It is also of interest to consider code that will detect partial linearity, 
+rather than expecting the user to provide the special structure and settings. Work in this
+area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution
+we can find.
+Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
+is about a class of models that have almost nothing in common
+with the present issue. The algorithm in @Birkisson2013 will need investigating further,
+but we suspect that it is complicated enough that it will not easily be adapted to 
+a program like `nls()`. Possibly a simpler algorithm may be found.
 
@@
 
-Indexed models clearly have a place in some areas of research. However, the current
-approach in `nls()` is awkward. The user must use DIFFERENT formulas depending on 
-whether the `plinear` algorithm is chosen. This seems to be related to the need for
-that approach to use the `lm()` function or its infrastructure, thus employing 
-formulas that omit the parameters of models and simply give the variables or functions
-thereof, such as interaction terms.
-
-Users, we believe, are much more likely to be comfortable with fully specified formulas.
-And such formulas are needed by nonlinear least squares functions such as `minpack.lm::nlsLM`
-and `nlsr::nlxb`. Thus we would like the identification of the linear parameters to be,
-if possible, automated. At the very least, we should be able to use a structure like
-
-```
-algorithm="plinear(parmx, thetaz)"
-```
-
-to allow for the full formula to be used with the linear parameters identified.
-
-## Provide tests
-
-We need suitable tests in order:
+Indexed models clearly have a place in some areas of research. 
+We need to be able to process models, as the example above, where the
+model formula is 
+`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` and `Strip` is the index. Data
+`Length` and `Conc` are available for all observations. Parameter `th` applies
+to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
+could be an experimental run, or some other categorization.
+
+Given that `[ ]` are the R method to index arrays, their presence in a model
+formula signals indexing. The question is how to perform the correct calculations.
+
+While tedious (??maybe dummies or fastdummies packages make it less so?), 
+we could set up the model as a sum of similar models with dummy
+variables to turn "on" or "off" the particular case, transforming the model
+parameters to simple, non-indexed ones (as does `nls()`).
+
+Wrapper functions to prepare the estimation calculations and present the
+results in a form that matches the modelling formula would avoid adding
+complexity to `nls()` and other tools.
+
+## Provide tests and use-case examples
+
+We need suitable tests and use-case examples in order:
 
@@
 
-### A test runner program
+### A test runner program ?? NLSCompare
 
@@
 
-Some considerations for our test scrips:
+Some considerations for our test scripts:
 
@@
 
-# Outputs of the project
-
-The project output is available in the Git repository https://github.com/nashjc/RNonlinearLS .
-
-## Formal reports or documentation
-
-- RefactoringNLS.Rmd: this document which will become the main report of the project.
-
-- TestsDoc.Rmd: a survey of testing tools in R. It has more general possibilities 
+# Progress so far
+
+Much of our work is reported in https://github.com/nashjc/RNonlinearLS .
+
+## Reports and articles
+
+- The present article is the central report of the Google Summer or Code 2021 project
+  "Improvements to nls()".
+
+- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
+  the many choices in their implementation for nonlinear least squares. This is still
+  a DRAFT at 2021-8-20. ?? fix
+
+- TestsDoc.Rmd (??citation) is a survey of testing tools in R. It has more 
+  general possibilities 
   and fits into the subject of regression testing, in which case a 
@@
 
-## Informal reports
-
-These are documents used to discuss particular aspects of our work. These are part
-of the repository https://github.com/nashjc/RNonlinearLS.
-
-- DerivsNLS.Rmd: a document to explain different ways in which Jacobian information
+- DerivsNLS.Rmd ??citation: a document to explain different ways in which Jacobian information
   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is
@@
 
-- ImproveNLS.bib: a consolidated BibTex bibliography for all documents in this
-  project, possibly with wider application to nonlinear least squares in general.
-
-- MachineSummary.Rmd: an investigation of ways to report the characteristics
+- The BibTex bibliography ImproveNLS.bib (??citation or reference) was set up 
+  for use with all documents in this project, but has wider application to nonlinear 
+  least squares projects in general.
+
+- MachineSummary.Rmd (??citation -- blog post??) is an investigation of ways to report 
+  the characteristics
   and identity of machines running tests. `MachID.R` offers a suggested concise summary function
@@
 
-- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
-  the many choices in their implementation for nonlinear least squares. This is still
-  a DRAFT at 2021-8-20.
-
-- PkgFromRbase.Rmd: an explanation of the construction of the `nlspkg` from the
+- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
   code in R-base.
 
-  <!-- We could think of putting this in R-bloggers or similar. Or if -->
-  <!-- we came up with a more general template, then an R-Journal article. -->
-
-- WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on.
-  A project diary.
+- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
+  a review of optimization in R. Ideas from the present work have been instrumental
+  in the creation of @NashWires22.
+
+<!-- ?? probably don't need to include -->
+<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
+<!--   A project diary. -->
 
 ## Problem sets and test infrastructure
+
+?? need to clean up
 
@@
   undocumented, and unsuited to adding improvements like the Marquardt stabilization. 
-  We were able to get a pure-R version of numericDeriv() and rework most of the functions 
+  We were able to get a pure-R version of `numericDeriv()` and rework most of the functions 
   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project
@@
 
-- nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the 
-  fix by JN to the relative offset convergence criterion failure when there are small 
-  residuals in problems sent to ``nls()`. Now part of R.
-
-- nlsj: A refactoring of the `nls() functionality. `redesign2107.txt` gives some notes.
+<!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  -->
+<!--   fix by JN to the relative offset convergence criterion failure when there are small  -->
+<!--   residuals in problems sent to ``nls()`. Now part of R. -->
+
+- the safeguarded relative offset convergence criterion that allows small-residual problems
+  to ba handled with `nls()` is now in the R base code.
+
+- `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features??
   This is an interim package for exploration of ideas and will NOT, as it currently
@@
 - nlsralt: this is intended to be a modified version of Nash and Murdoch package 
-  `nlsr` with improvements discovered as a result of this project. At the close of
-  this project, this is still a work in progress.
+  `nlsr` to examine possible improvements discovered as a result of this project. 
+  This is EXPERIMENTAL.
 
@@
 
-A key difference in approach between `nls()` and `nlsr::nlxb()`, possibly the key
-difference, is that `nls()` builds a large infrastructure from which the Gauss-Newton
+The key difference in approach between `nls()` and `nlsr::nlxb()` is that `nls()` 
+builds a large infrastructure from which the Gauss-Newton
 iteration can be executed and other statistical information such as profiles can be
 computed, while `nlxb()` returns quite limited information, and in its execution
-computes what is needed on an as and when basis. This follows a path that one of us
+computes what is needed on an "as and when" basis. This follows a path that one of us
 (JN) established almost 50 years ago with the software that became @cnm79, using
@@
 package. `nls()` also has some similar functions, but they are much more tied into
-the infrastructure that is created mostly in the `nlsModel()` function.
+the infrastructure that is created, mostly in the `nlsModel()` function.
 
 The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable
-entangling of the different functions and capabilities. This is to an extent that even after
+entangling of the different functions and capabilities. Even after
 both the preparation for and carrying out of the present project, we do not feel confident
-to explain the code completely, nor to maintain it. However, we have made some progress toward
-such a goal. Part of that has been to render the code in R. In this there remain some tasks
-to complete, and they are possibly difficult.
+to explain the code completely, nor to maintain it. However, we have made some 
+forays into render parts of the code in R, and to understand where difficulties
+lie.
 
@@
 
-To advance the stability and maintainability of R, we believe that the functions that carry out
-quite advanced calculations like those achieved by `nls()` should be organized to have their 
-interactions as orderly as possible. The aspects of `nls()` that have given us the most trouble
-are as follows:
+To advance the stability and maintainability of R, we believe the program objects (R functions)
+that are created by tools such as `nls()` should have minimal interactions and side-effects.
+The aspects of `nls()` that have given us the most trouble are as follows:
 
@@
   as the correct environment is used, this provides a surprisingly short syntax to invoke the
-  calculations.
-
-- Weights, subsets, and various contextual controls such as that for the `na.action` that tells
+  calculations. The danger is that the wrong data is accessed if the internal search finds a
+  valid name that is not the object we want.
+
+- Weights, subsets, and various contextual controls such as that for the `na.action` that tell
   our code what to do with undefined or missing numbers. Again, this makes for a very simple
-  invocation of the calculation, but the context is hidden from the user -- and it can be
+  invocation of the calculation, but the context is hidden from the user. It can be
   difficult for those of us trying to maintain or improve the code to be certain we have the
@@
 
-- The mixing of R and C code makes for high performance, but without developer documentation 
-  leaves programmers with a lot of work to do to adjust the code. We believe in keeping code --
+- The mixing of R and C code made for high performance in the past, but without 
+  developer documentation, programmers face a lot of work to fix code. We believe in keeping code --
   at least a runnable reference version -- in a single programming language. If necessary, by
@@
 
-### Consistent specifications of partially linear models
-
-Ideally, we believe it would be best to specify all models with a complete set of parameters.
-That is, the model should be written down for as it would be used, rather than for estimation. 
-The `algorithm="plinear"` setting requires that we leave out the linear parameters. This
-means that we cannot simply change the `algorithm` argument without also changing the 
-`formula`. Moreover, the partially linear parameters appear in the result object
-as `.lin1`, `.lin2`, etc.
-
-A possible syntax to provide consistency would be to specify the partially linear parameters
-when the `plinear` option is specified. For example, we could use
-
-```
-algorithm="plinear(Asym)"
-```
-
-which would allow for the partially linear algorithm to function and the output of different
-`algorithm` options to be consistent.
-
-Furthermore, we ask if it is possible for us to devise code that will detect linearity, 
-rather than expecting the user to provide the special structure and  setting. Work in this
-area appears to tbe quite limited, with @Birkisson2013 being seemingly a sole contribution.
-Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
-is about a class of models that have almost nothing in common
-with the present issue. The algorithm in @Birkisson2013 will need investigating further,
-but we suspect that it is complicated enough that it will not easily be adapted to 
-an `nls()` style program. Possibly a simpler algorithm may be found.
-
 # Acknowledgement

### 2022-05-31 RefactoringNLS/RefactoringNLS.Rmd
add note in acknowledgement -- needs action

@@
 existing code makes maintenance and improvement very difficult. Moreover, legacy
-applications and examples mean some improvments are blocked. Thus we have not
+applications and examples mean some improvements are blocked. Thus we have not
 addressed all the issues raised, but we believe that the discussion and trials
@@
 us review some of the issues. We will then propose corrective actions, some of which we have carried out
-either in demonstrations or in the distributed coed.
+either in demonstrations or in the distributed code.
 
@@
 
-  **Do not use `nls` on artificial "zero-residual" data.**
-
-  The `nls` function uses a relative-offset convergence criterion
+>  **Do not use `nls` on artificial "zero-residual" data.**
+
+>  The `nls` function uses a relative-offset convergence criterion
   that compares the numerical imprecision at the current parameter
@@
 
-  $$  y = f(x, \theta) + eps $$
-
-  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form
-
-  $$  y = f(x, \theta)  $$
-
-  because the criterion amounts to comparing two components of the round-off 
+>  $$  y = f(x, \theta) + eps $$
+
+>  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form
+
+>  $$  y = f(x, \theta)  $$
+
+>  because the criterion amounts to comparing two components of the round-off 
   error.
 
-  If you wish to test `nls` on artificial data please add a noise component, 
+>  If you wish to test `nls` on artificial data please add a noise component, 
   as shown in the example below.
@@
 t <- -10:10
-y <- 100/(1+.1*exp(-0.51*t))
-lform<-y~a/(1+b*exp(-c*t))
-ldata<-data.frame(t=t, y=y)
-plot(t,y)
-lstartbad<-c(a=1, b=1, c=1)
-lstart2<-c(a=100, b=10, c=1)
-nlsr::nlxb(lform, data=ldata, start=lstart2)
-nls(lform, data=ldata, start=lstart2, trace=TRUE)
+y <- 100/(1 + .1 * exp(-0.51 * t))
+lform <- y ~ a/(1 + b * exp(-c * t))
+ldata <- data.frame(t = t, y = y)
+plot(t, y)
+lstartbad <- c(a = 1, b = 1, c = 1)
+lstart2 <- c(a = 100, b = 10, c = 1)
+nlsr::nlxb(lform, data = ldata, start = lstart2)
+nls(lform, data = ldata, start = lstart2, trace = TRUE)
 # Fix with scaleOffset
-nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
+nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
 sessionInfo()
@@
 > t <- -10:10
-> y <- 100/(1+.1*exp(-0.51*t))
-> lform<-y~a/(1+b*exp(-c*t))
-> ldata<-data.frame(t=t, y=y)
+> y <- 100/(1 + .1 * exp(-0.51 * t))
+> lform <- y ~ a/(1 + b * exp(-c * t))
+> ldata <- data.frame(t=t, y=y)
 > plot(t,y)
-> lstartbad<-c(a=1, b=1, c=1)
-> lstart2<-c(a=100, b=10, c=1)
-> nlsr::nlxb(lform, data=ldata, start=lstart2)
+> lstartbad <- c(a = 1, b = 1, c = 1)
+> lstart2 <- c(a = 100, b = 10, c = 1)
+> nlsr::nlxb(lform, data = ldata, start = lstart2)
 nlsr object: x 
@@
 c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
-> nls(lform, data=ldata, start=lstart2, trace=TRUE)
+> nls(lform, data = ldata, start = lstart2, trace = TRUE)
 40346.    (1.08e+00): par = (100 10 1)
@@
 > # Fix with scaleOffset
-> nls(lform, data=ldata, start=lstart2, trace=TRUE, control=list(scaleOffset=1.0))
+> nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
 40346.      (1.08e+00): par = (100 10 1)
@@
 ```{r SESingGrad}
-reala=-3
-realb=5
-realc=0.5
-realr=0.7
-realm=1
-x=1:11 #x values 11 timepoint data
+reala = -3
+realb = 5
+realc = 0.5
+realr = 0.7
+realm = 1
+x = 1:11 #x values; 11 timepoint data
 #linear+exponential function
-y=reala + realb*realr^(x-realm) + realc*x
-testdat=data.frame(x,y)
-strt<-list(a=-3, b=5, c=0.5, r=0.7, m=1)
-jform<-y~a+b*r^(x-m)+c*x # Formula
-linexp=try(nls(jform, data=testdat, start=strt, trace=F))
+y=reala + realb * realr^(x - realm) + realc * x
+testdat = data.frame(x, y)
+strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1)
+jform <- y ~ a + b * r^(x - m) + c * x # Formula
+linexp=try(nls(jform, data = testdat, start = strt, trace = F))
 library(nlsr)
 # Note singular values of Jacobian in rightmost column
-linexp2<-try(nlxb(jform, data=testdat, start=strt, trace=F))
+linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
 linexp2
@@
 A useful, but possibly dated, description is given in:
-https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/
+https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/.
 
@@
 
+??Heather Turner ... (co-mentor ?? how to word?)
+
 # References

### 2022-06-02 RefactoringNLS/RefactoringNLS.Rmd
Add section on bounds

@@
 
+## Feature: bounds on parameters
+
+There are many situations where the context of a problem constrains the values of parameters.
+For example, one of us was asked many years ago to estimate a model. One paramater returned
+a negative value. The client complained "But that is supposed to be the number of grain elevators
+in Saskatchewan". That is, the number should have been a positive integer, and likely less
+than a few thousand.
+
+`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
+the manual states
+
+> The algorithm = "port" code appears unfinished, and does not even check that the starting value is within the bounds. Use with caution, especially where bounds are supplied.
+
+This is clearly unsatisfactory for general use. Inclusion of bounds on parameters in nonlinear
+least squares computations is relatively straightforward, and it is part of the default method
+for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
+script, which we do not run here for reasons of space, shows that the solutions found are not
+optimal for the Hobbs test problem, even in its scaled form.
+
+```{r chkbound, eval=FALSE}
+#### bounds with formula specification of problem
+## Use the Hobbs Weed problem
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weedframe <- data.frame(y=weed, tt=tt)
+st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
+## Unscaled model
+wmodu <- y ~ b1/(1+b2*exp(-b3*tt))
+## Scaled model
+wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt))
+## We can provide the residual and Jacobian as functions
+# Scaled Hobbs problem
+shobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("shobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  100.0*x[1]/(1+x[2]*10.*exp(-0.1*x[3]*tt)) - y
+}
+
+shobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-0.1*x[3]*tt)
+  zz  <-  100.0/(1+10.*x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -0.1*x[1]*zz*zz*yy
+  jj[tt,3]   <-   0.01*x[1]*zz*zz*yy*x[2]*tt
+  attr(jj, "gradient") <- jj
+  jj
+}
+
+##----- Hobbsbounded unique code starts here ------
+require(minpack.lm)
+traceval<-FALSE
+cat("Infeasible start test\n")
+start1inf <- c(b1=4, b2=4, b3=4) # b3 OUT OF BOUNDS for next few tries
+## Infeasible start! No warning message!
+anlM2i <- try(nlsLM(wmodu, start=start1inf, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
+                    trace=traceval))
+
+# feasible start i.e. on or within bounds
+start1<-st
+anlM1 <- try(nlsLM(wmodu, start=start1, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
+                    trace=traceval))
+print(anlM1)
+
+# Hobbs scaled problem with bounds, formula specification
+
+require(nlsr)
+anlxb1 <- nlxb(wmods, start=start1, data=weedframe, lower=c(0,0,0),
+                  upper=c(2,6,3), trace=traceval)
+print(anlxb1)
+
+## nlsLM seems NOT to work properly with bounds
+anlsLM1b <- nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
+                 upper=c(2,6,3), trace=traceval)
+print(anlsLM1b)
+newst<-coef(anlsLM1b)
+print(newst)
+anlsLM1bx <- nlsLM(wmods, start=newst, data=weedframe, lower=c(0,0,0),
+                   upper=c(2,6,3))
+print(anlsLM1bx)
+
+cat("Single number bounds:\n")
+anlsLM1b1 <- try(nlsLM(wmods, start=start1, data=weedframe, lower=0,
+                   upper=3))
+print(anlsLM1b1)
+anlsLM1b1a <- try(nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
+                   upper=c(3,3,3)))
+print(anlsLM1b1a)
+
+anlxb1b1<- try(nlxb(wmods, start=start1, data=weedframe, lower=0,
+                     upper=3))
+print(anlxb1b1)
+
+anlm1b <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
+                  upper=c(2,6,3))
+print(anlm1b)
+
+# functional presentation of problem  -- seems to work 
+anlm1bx <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
+                   upper=c(2,6,3))
+print(anlm1bx)
+
+anlfb1bx <- nlfb(start=start1, resfn=shobbs.res, jacfn=shobbs.jac, lower=c(0,0,0),
+                  upper=c(2,6,3))
+print(anlfb1bx)
+```
+
+### Philosophical considerations
+
+Bounds on parameters raise some interesting and difficult questions about how uncertainty
+in parameter estimates should be computed or reported. That is, the traditional "standard
+errors" are generally taken to imply symmetric intervals about the point estimate in which
+the parameter may be expected to be found with some probability under certain assumptions.
+Bounds change those assumptions drastically. Hence `nlsr::nlxb()` does not compute standard
+errors nor their derived statistics when bounds are active.
+
 # Goals of our effort to improve `nls()`

### 2022-06-12 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
cleanup of Variety. Removed test problems as too long

@@
 ease or hinder the building of well-structured codes. See
-*git@github.com:pcolsen/Nash-Compact-Numerical-Methods.git*.
+*git@github.com:pcolsen/Nash-Compact-Numerical-Methods.git*. The principal merit
+of having these multiple implementations is that they illustrate the interaction between
+algorithm and programming language. Unfortunately, creating them is a lot of effort for
+no appreciable new capability. That is, they are an interesting intellectual exercise
+that may incidentally afford minor benefits to a few individual users.
 
@@
 The above uses the intermediate code in function `lm.fit()`. This uses a QR solver, but
-it is written as a wrapper in C calling a Fortran routine (in the Fortran 77 dialect).
-
-?? Should we put in the structure and where the code is located?
+it is written as an R wrapper calling a C routine, which in turn calls a local Fortran 77
+driver for Linpack (@Dongarra79) routines
+"dqrdc" and "dqrsl". The R wrapper is part of the R source distribution
+(here we refer to R-4.0.0, but I believe that the location of the code has not changed over a
+fairly long sequence of releases). It is found in src/library/stats/R/lm.R. This calls
+the C routine "Cdqrls" in src/library/stats/src/lm.c. The Fortran (strictly Fortran 77)
+code "dqrls.f" is found in src/appl/.
 
 I believe that the structure of lm() predates the availability of the family
-of `qr.xxx()` functions for R. These allow us to access the QR approach 
+of `qr.xxx()` functions for R, which allow us to access the QR approach 
 directly.
@@
 ```
+
+This rather long-winded pursuit of the convoluted code structure underlines how 
+much effort must be expended, and how much expertise over several programming
+languages is needed to properly alter or maintain the code for `nls()` and other
+functions within R that do so much computation.
 

### 2022-06-13 VarietyInNLS/HobbsTiming.R
HobbsTiming is in improvenls/Tests/Examples

@@ -1,24 +1,0 @@
-## HobbsTiming.R -- computes jacobian of Hobbs at 1,1,1 using numericDeriv()
-# rm(list=ls()) # clear the workspace (does not completely remove things, need to restart R sometimes)
-
-traceval  <-  FALSE  # traceval set TRUE to debug or give full history
-
-# Data for Hobbs problem
-ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
-            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
-tdat  <-  seq_along(ydat) # for testing
-
-# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
-start1  <-  c(b1=1, b2=1, b3=1)
-eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt))
-weeddata1  <-  data.frame(y=ydat, tt=tdat)
-
-library(microbenchmark)
-library(nlsralt)
-tnlsrh1<-microbenchmark(nlsrh1<-nlxb(eunsc, data=weeddata1, start=start1))
-tnlsrxh1<-microbenchmark(nlsrxh1<-nlxbx(eunsc, data=weeddata1, start=start1))
-nlsrh1
-nlsrxh1
-tnlsrh1
-tnlsrxh1
-

### 2022-06-13 RefactoringNLS/SESingGrad.R
In improvenls/Tests/Examples

@@ -1,29 +1,0 @@
-#parameters used to generate the data
-# rm(list=ls())
-# https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values
-reala=-3
-realb=5
-realc=0.5
-realr=0.7
-realm=1
-x=1:11 #x values - I have 11 timepoint data
-#linear+exponential function
-y=reala + realb*realr^(x-realm) + realc*x
-#add a bit of noise to avoid zero-residual data
-jitter_y = jitter(y,amount=0.2)
-testdat=data.frame(x,jitter_y)
-strt<-list(a=-3, b=5, c=0.5, r=0.7, m=1)
-#try the regression with similar starting values to the the real parameters
-jform<-jitter_y~a+b*r^(x-m)+c*x
-linexp=try(nls(jform, data=testdat, start=strt, trace=F))
-library(nlsr)
-jmod<-model2rjfun(jform, strt, jacobian=TRUE, gradient=TRUE)
-jmodval<-jmod(strt)
-jmodval
-jmodJ<-attr(jmodval, "gradient")
-cat("Singular values of Jacobian:")
-print(svd(jmodJ)$d)
-# We can also get the singular values of the Jacobian from nlsr::nlxb()
-# BUT note that the values are printed as a column purely to save space
-linexp2<-try(nlxb(jform, data=testdat, start=strt, trace=F))
-linexp2

### 2022-06-13 RefactoringNLS/RefactoringNLS.Rmd
Updates to two articles

@@
 
+??fix next bits
+<!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
+<!--   would, however, be a painful and mainly political activity. -->
+
+<!-- - The tools for analytic derivatives that exist in parts of `nls()` for self-starting models and -->
+<!--   in `nlsr` for problems specified as models that use expressions for which derivatives exist  -->
+<!--   should be more generally available, but the task is one of tedium and limited reward to the -->
+<!--   persons who do it, while benefiting the R community substantially. -->
+
+<!-- - Similarly, default methods need stabilized solvers e.g., Marquardt methods, though to date -->
+<!--   a mechanism to incorporate such modifications in `nls()` has eluded us. -->
+
+<!-- - Bounds constraints are sufficiently important that they should be part of the default -->
+<!--   methods -->
+<!--   (rather than algorithm = "port" for `nls()`). `minpack.lm` should handle them correctly. -->
+
 # Acknowledgement

### 2022-06-13 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Updates to two articles

@@
 
-# Test problems
-
-It is worth including some sample test problems in this article to illustrate
-the discussion.
-
-?? here or later
-
-```{r settest15, fig.width=5}
-# set parameters
-set.seed(123456)
-a <- 1
-b <- 2
-c <- 3
-np <- 15
-tt <-1:np
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-# plot.new()
-plot(tt, yy)
-#plot(tt, yy, type='l')
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-points(tt,y1,type='p', col="green")
-y2 <- yy + 5*ev
-points(tt,y2,type='p', col="blue")
-y3 <- yy + 10*ev
-lg3d15 <- data.frame(tt, yy, y1, y2, y3)
- points(tt,y3,type='p', col="red")
-library(nlsr)
-sol0 <- nlxb(yy ~ a0/(1+b0*exp(-c0*tt)), data=lg3d15, start=list(a0=1, b0=1, c0=1))
-print(sol0)
-sol1 <- nlxb(y1 ~ a1/(1+b1*exp(-c1*tt)), data=lg3d15, start=list(a1=1, b1=1, c1=1))
-print(sol1)
-sol2 <- nlxb(y2 ~ a2/(2+b2*exp(-c2*tt)), data=lg3d15, start=list(a2=1, b2=1, c2=1))
-print(sol2)
-sol3 <- nlxb(y3 ~ a3/(3+b3*exp(-c3*tt)), data=lg3d15, start=list(a3=1, b3=1, c3=1))
-print(sol3)
-```
-
-The following is a larger dataset version of this test.
-
-```{r settest150, eval=FALSE}
-np <- 150
-tt <- (1:np)/10
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-y2 <- yy + 5*ev
-y3 <- yy + 10*ev
-lg3d150 <- data.frame(tt, yy, y1, y2, y3)
-np <- 1500
-tt <- (1:np)/100
-yy <- 100*a/(1+10*b*exp(-0.1*c*tt))
-set.seed(123456)
-ev <- runif(np)
-ev <- ev - mean(ev)
-y1 <- yy + ev
-y2 <- yy + 5*ev
-y3 <- yy + 10*ev
-lg3d1500 <- data.frame(tt, yy, y1, y2, y3)
-f0 <- yy ~ a0/(1+b0*exp(-c0*tt))
-f1 <- y1 ~ a1/(1+b1*exp(-c1*tt))
-f2 <- y2 ~ a2/(2+b2*exp(-c2*tt))
-f3 <- y3 ~ a3/(3+b3*exp(-c3*tt))
-```
+# Summary discussion
+
+This article has attempted to provide an overview of how software for nonlinear least
+squares computations may be structured and implemented, taking note that users may wish
+to solve modeling problems rather than minimize a sum of squared functions. We have seen
+that different implementation goals and structural decisions greatly influence the
+nature, capability, performance and features of different codes. Moreover, while the
+treatment here has taken a panoramic view, we have been interested primarily in what
+is availabel to R users.
+
+Having reviewed the tools available, we note that there are some serious concerns. Of these,
+we will list those we consider to be the most important:
+
+- tools such as the `nls()` function in R has so many legacy uses that it is likely
+  impossible to deprecate it. However, it has also (as we detail in ??RefactoringNLS??)
+  non-maintainable, but it seems unlikely that any
+  further development, nor even repair of current deficiencies, will be forthcoming.
+  This underlines the importance of designing software for its long-term maintenance,
+  and this is often inimical to using currently novel programming tools.
+
+- The rich variety of ways of structuring and solving the Gauss-Newton equations, stabilized
+  or not, allows for codes to be implemented to take advantage of special situations. 
+  However, for most use-cases, an approach using reliable matrix decompositions such
+  as modern QR techniques should suffice. These allow Marquardt stabilizations to be
+  applied relatively easily in most, but not all, situations.
+
+- The needs of different user leads to difficult choices and implementation methods for
+  handling how problems are interfaced to the computational infrastructure and results
+  are returned and reported. These need to be well-documented, paying attention to 
+  the different ways in which users understand their problems.
 

### 2022-06-13 VarietyInNLS/HobbsTiming.R
Restoring some files that somehow disappeared from repo

@@ -1,0 +1,24 @@
+## HobbsTiming.R -- computes jacobian of Hobbs at 1,1,1 using numericDeriv()
+# rm(list=ls()) # clear the workspace (does not completely remove things, need to restart R sometimes)
+
+traceval  <-  FALSE  # traceval set TRUE to debug or give full history
+
+# Data for Hobbs problem
+ydat  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+            38.558, 50.156, 62.948, 75.995, 91.972) # for testing
+tdat  <-  seq_along(ydat) # for testing
+
+# A simple starting vector -- must have named parameters for nlxb, nls, wrapnlsr.
+start1  <-  c(b1=1, b2=1, b3=1)
+eunsc  <-   y ~ b1/(1+b2*exp(-b3*tt))
+weeddata1  <-  data.frame(y=ydat, tt=tdat)
+
+library(microbenchmark)
+library(nlsralt)
+tnlsrh1<-microbenchmark(nlsrh1<-nlxb(eunsc, data=weeddata1, start=start1))
+tnlsrxh1<-microbenchmark(nlsrxh1<-nlxbx(eunsc, data=weeddata1, start=start1))
+nlsrh1
+nlsrxh1
+tnlsrh1
+tnlsrxh1
+

### 2022-06-13 MachineSummary/MachineSummary.Rmd
Edits to articles and blog posts
changed data time format:aesthetics?

@@
 #get a timestamp
-tsstr<- format(Sys.time(), "%Y%m%d%H%M") # tsstr == time stamp string in form YYYYmmddHHMM
+tsstr<- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
 cat("Date and Time stamp:",tsstr,"\n")

### 2022-06-13 MachineSummary/MachID.R
added spaces; modified ram output to make it succinct

@@ -1,8 +1,8 @@
-sy<-Sys.info()
-tsstr<- format(Sys.time(), "%Y%m%d%H%M")
-cpu<-benchmarkme::get_cpu()
-ram<-benchmarkme::get_ram()
-machid<-paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
-              "|",cpu$model_name,"|",ram,"bytesRAM", sep='')
+sy <- Sys.info()
+tsstr <- format(Sys.time(), "%Y%m%d%H%M")
+cpu <- benchmarkme::get_cpu()
+ram <- benchmarkme::get_ram()
+machid <- paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],
+              "|",cpu$model_name,"|",round(ram/1000^3,2)," GB RAM", sep='')
 cat(machid,"\n")
 

### 2022-06-13 MachineSummary/MachID.R
changed tsstr format as in the main doc

@@ -1,5 +1,5 @@
 sy <- Sys.info()
-tsstr <- format(Sys.time(), "%Y%m%d%H%M")
+tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M")
 cpu <- benchmarkme::get_cpu()
 ram <- benchmarkme::get_ram()
 machid <- paste(sy["nodename"],":",sy["user"],"-",sy["sysname"],"-",sy["release"],

### 2022-06-13 MachineSummary/MachineSummary.Rmd
updated machine summary

--- 2022-01-18 MachineSummary/MachineSummary.Rmd
@@
     keep_tex: false
+link-citations: yes
+linkcolor: blue
 ---
@@
 #get a timestamp
-tsstr<- format(Sys.time(), "%Y%m%d%H%M") # tsstr == time stamp string in form YYYYmmddHHMM
-cat("Date and Time stamp:",tsstr,"\n")
+tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
+cat("Date and Time stamp:", tsstr, "\n")
 ```
@@
 
-- operating system and version. For Linux systems, we note that besides the distribution
-  variant, e.g., Linux Mint 20.3 MATE 64 bit, it is necessary to add the Linux kernel
-  identifier, e.g.,  5.13.0-25-generic, from the `uname` command.
+- operating system and version.
+For Linux systems, we note that besides the distribution variant, e.g., Linux Mint 20.3 MATE 64 bit, it is necessary to add the Linux kernel identifier, e.g.,  5.13.0-25-generic, from the `uname` command.
 
@@
 `benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
-function of this package for some Windows 10 versions which is now 
-corrected (??check it is uploaded). ??Do we wish to comment on the
-"GB" issue?
+function of this package for some Windows 10 versions. It has been 
+corrected in the latest version 1.0.8, with contributions from one of the authors (AB).
+
+<!-- on https://github.com/csgillespie/benchmarkme, but at the -->
+<!-- time of writing is not yet on the CRAN repository. ??Do we wish to comment on the -->
+<!-- "GB" issue? -->
 
@@
 get_byte_compiler()
-gla<-get_linear_algebra()
-gsd<-get_sys_details()
+gla <- get_linear_algebra()
+gsd <- get_sys_details()
 ```
@@
 
-Until the IEEE 754 standard (??ref) became availaible in the 1980s, 
+Until the IEEE 754 standard (?? @institute1985ieee) became available in the 1980s, 
 computers used multiple forms of floating point arithmetic. With
@@
 floating point properties.
-This is a port of Mike Malcolm's ENVRON to R. ??ref. It computes the 
+This is a port(??part) of Mike Malcolm's ENVRON to R. ??ref. It computes the 
 machine precision, radix and number of radix digits.
@@
 ```{r envrontest}
-envron<-function(){
-D1<-as.numeric(1) # use variable to avoid constants when double precision invoked
-E5 <- 10 #arbitrary scaling for additive equality tests
+envron <- function(){
+D1 <- as.numeric(1) # use variable to avoid constants when double precision invoked
+E5 <- 10 # arbitrary scaling for additive equality tests
 B9 <- 1E+35 # big number, not necessarily biggest possible
@@
 repeat{
-  E9 <- E9/2 #start of loop to decrease estimated machine precision
-  D0 <- E6+E9 #force storage of sum into a floating-point scalar
+  E9 <- E9 / 2 # start of loop to decrease estimated machine precision
+  D0 <- E6 + E9 # force storage of sum into a floating-point scalar
   if (D0 <= E6) break
 } # repeat reduction while (1+E9) > 1
-E9 <- E9*2 # restore smallest E9 which gives (1+E9) > 1
+E9 <- E9 * 2 # restore smallest E9 which gives (1 + E9) > 1
 repeat {
-   E6 <- E6+1 # try different radix values
-   D0 <- E6+E9
+   E6 <- E6 + 1 # try different radix values
+   D0 <- E6 + E9
    if (D0 <= E6) break
@@
 repeat {  
-  E9 <- E9/E6 #loop while dividing by radix
-  J1 <- J1+1 #increment counter
-  D0 <- D1+E9 #add tp 1
-  if (D0<=D1) break #test and repeat until equality
+  E9 <- E9 / E6 # loop while dividing by radix
+  J1 <- J1 + 1 # increment counter
+  D0 <- D1 + E9 # add tp 1
+  if (D0 <= D1) break # test and repeat until equality
 }
-E9 <- E9*E6 #recover last value of machine precision
-J1 <- J1-1 #and adjust the number of digits
-mpvals<-list(eps=E9, radix=E6, ndigits=J1)
+E9 <- E9 * E6 # recover last value of machine precision
+J1 <- J1 - 1 # and adjust the number of digits
+mpvals<-list(eps = E9, radix = E6, ndigits = J1)
 }
-mp<-envron()
-E9<-mp$eps
-E6<-mp$radix
-J1<-mp$ndigits
+mp <- envron()
+E9 <- mp$eps
+E6 <- mp$radix
+J1 <- mp$ndigits
 cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
-cat( "E9*1E+16=",E9*1E+16,"\n")
-cat( "E6 = ",E6,"  -- the radix of arithmetic","\n")
-cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers","\n")
-cat( "E6^(-J1+1) * 1E+16=",1E+16*(E6^(-J1+1)),"\n")
-cat(".Machine$double.eps=",.Machine$double.eps,"\n")
+cat( "E9*1E+16=", E9 * 1E+16, "\n")
+cat( "E6 = ",E6,"  -- the radix of arithmetic", "\n")
+cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
+cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")
+cat(".Machine$double.eps=", .Machine$double.eps, "\n")
 cat("From R  .Machine:\n")
-cat("double.eps=", .Machine$double.eps,"\n", 
-    "double.base=", .Machine$double.base,"\n", 
-    "double.digits=", .Machine$double.digits,"\n" )
+cat("double.eps =", .Machine$double.eps, "\n", 
+    "double.base =", .Machine$double.base, "\n", 
+    "double.digits =", .Machine$double.digits, "\n" )
 ```
@@
 loopesc <- function(nn){
-  ss<-0
+  ss <- 0
   for (i in 1:nn) {
-    xx<-exp(sin(cos(1.0*i)))
-    ss<-ss+xx
+    xx <- exp(sin(cos(1.0*i)))
+    ss <- ss + xx
   }
@@
 nvals<-c(1000, 10000, 100000, 1000000)
-tvals<-rep(NA,length(nvals))
+tvals<-rep(NA, length(nvals))
 i<-0
 for (nn in nvals) {
-  i<-i+1
-  cat(nn,"\n")
-  tt<-microbenchmark(loopesc(nn), unit='us')
+  i <- i + 1
+  cat(nn, "\n")
+  tt <- microbenchmark(loopesc(nn), unit = 'us')
   print(tt)
-  tvals[i]<-mean(tt$time)
+  tvals[i] <- mean(tt$time)
 }
-plot(log10(nvals),log(tvals))
+plot(log10(nvals), log(tvals))
 

### 2022-06-13 MachineSummary/MachineSummary.Rmd
JN first major edit of MachineSummary

@@
 
-# Providing a characterization of a particular computing environment for R calculations
+# Abstract
 
-If we want to compare results computed on different machines, we need a way to provide 
-measures and identifiers of the particular computing environment at hand. This write-up
-attempts to summarize some possibilities available in the R statistical software and language.
-Note that one of our interests is to have tools that work in R **across different platforms**.
-That is, we want to be able to call an R function to get machine and computing environment
-characterizations regardless of the particular platform which we are currently using.
-Moreover, we want to have a succinct output of such information.
+Many statistical calculations draw on system resources of memory, storage, arithmetic,
+instructional processing and communications. A given computational environment will
+have a particular collection of such resources. With workers now commonly working 
+separately and sharing their work virtually, we should consider how to characterize
+the systems under which calculations are carried out in order that we can diagnose
+differences in results or performance of the computing methods. Our objective here
+is to consider how a succinct labeling may provide a **machine profile summary**
+to assist such analysis and comparisons. To focus our work, we limit our attention
+to R as the software environment, but the ideas are more generally applicable.
 
-# Some possible information desired
+# Characterizing a particular computing environment for R
 
-There are multiple ways to get information of the type useful for characterizing the
-computing infrastructure used for a particular test or timing. However, we need to
-ensure that identifiers of particular pieces of information are unique, and that 
-common identifiers really do specify identical, or at least equivalent, information. 
-Otherwise, new names for some pieces of our set of capability information are needed.
+In 2021, we undertook an investigation of tools in R for nonlinear least squares
+calculations under the Google Summer of Code initiative. In the process, we recognized 
+that we would be comparing calculations
+performed on **different computing environments** at our disposal. Some of these were
+based on different hardware, while others were different combinations of given hardware
+with alternative operating software. In one case a machine failed and had to be replaced,
+underlining the need to be able to compare results from the defunct system with its
+replacements.
+
+A useful context for our ideas is provided by the Wikipedia article on Computer Performance
+(https://en.wikipedia.org/wiki/Computer_performance). We seek a way to provide succinct
+measures and identifiers of the particular computing environment at hand, leading to a
+**profile** for a given computing environment for which we will provide a 
+short but meaningful label unique within our project context. This article
+considers some possibilities available in the R statistical software and language
+that assist this effort, but the ideas are extensible to other programming systems.
+
+# Desired information
+
+There may be multiple terms used to describe some of the information we want, as
+well as different interpretations of some of the units of measure. For example
+"Gigabytes", or GB, may be $1000^3$ or $1024^3$ bytes. ?? AB? Is that correct?
 
@@
 
-When we wish to analyze performance results, it is extremely helpful to have 
+For tracking work, it is important to have 
 information on when the calculations were performed. 
@@
 other, at least in combination with a machine name and/or other tags. Here is one
-suggestion based on the R `Sys.time()` function.
+suggestion based on the R `Sys.time()` function. Moreover, as we were working on opposite
+sides of the globe, it could be important to either specify time zone or to work with a
+single "clock" such as UTC (ref?? or expansion?)
 
@@
 cat("Date and Time stamp:", tsstr, "\n")
+# and here is a way to get the UTC time
+tutc<-as.POSIXlt(Sys.time(), tz = "UTC")
+print(tutc)
+utctstmp<-format(tutc, "%Y/%m/%d|%H:%M")
+print(utctstmp)
 ```
 
-## Identifier for the computing software environment
+## Computing software environment
 
@@
   a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
-  generated we are not sure. Settings / System allows "rename this PC".
+  generated we are not sure. Settings / System allows "rename this PC". Clearly 
+  we may want to extend the name if particular software is involved, as follows.
 
-- operating system and version.
-For Linux systems, we note that besides the distribution variant, e.g., Linux Mint 20.3 MATE 64 bit, it is necessary to add the Linux kernel identifier, e.g.,  5.13.0-25-generic, from the `uname` command.
+- operating system and version. For Linux systems, we note that besides the 
+  distribution variant, e.g., Linux Mint 20.3 MATE 64 bit, it is necessary to add 
+  the Linux kernel identifier, e.g.,  5.13.0-25-generic, from the `uname` command.
 
@@
   version information, one needs to dig deeper, for example, using operating
-  system commands. For the present, we will omit such extra information unless
-  we can extract it easily within R.
+  system commands. 
+  <!-- ??For the present, we will omit such extra information unless -->
+  <!-- we can extract it easily within R. -->
 
@@
 features that should be mentioned if the user has modified the software in any
-way.
+way, or is using customized packages.
 
@@
 
-## benchmarkme
+## Machine precision and .Machine
 
-An important tool for acquiring machine information is the CRAN package
-`benchmarkme` (@benchmarkme). We found a minor bug in the `get_ram()`
-function of this package for some Windows 10 versions. It has been 
-corrected in the latest version 1.0.8, with contributions from one of the authors (AB).
-
-<!-- on https://github.com/csgillespie/benchmarkme, but at the -->
-<!-- time of writing is not yet on the CRAN repository. ??Do we wish to comment on the -->
-<!-- "GB" issue? -->
-
-?? Need to comment on
-
-- uses `proc.time()` for timings
-- has single core and parallel timings
-- ?? does it do GPUs
-- provides far too much for most needs i.e., a succinct summary of
-  how capable a computing system should be
-
-```{r benchmarkmex}
-library(benchmarkmeData)
-library(benchmarkme)
-ls(package:benchmarkmeData)
-ls(package:benchmarkme)
-lsf.str("package:benchmarkme")
-lsf.str("package:benchmarkmeData")
-## This next line takes a lot of time to run, so is commented out here
-# benchmark_std()
-get_byte_compiler()
-gla <- get_linear_algebra()
-gsd <- get_sys_details()
-```
-
-## Calculating the machine precision
-
-Until the IEEE 754 standard (?? @institute1985ieee) became available in the 1980s, 
+Until the IEEE 754 standard (@institute1985ieee) became available in the 1980s, 
 computers used multiple forms of floating point arithmetic. With
 some ingenuity, it is possible to detect many aspects of the 
-floating point properties.
-This is a port(??part) of Mike Malcolm's ENVRON to R. ??ref. It computes the 
-machine precision, radix and number of radix digits.
+floating point properties. In particular, we may be interested in the
+**machine precision**, that is, the smallest number `eps` such that 
+`1 + eps` is greater than `1`. R performs computations in IEEE "double"
+precision, and provides the R variable `.Machine` that has " information on 
+the numerical characteristics of the machine R is running on", including
+the machine precision as `.Machine$double.eps`.
 
-?? mention Kahan's Paranoia
+The information in `.Machine` is documented as based on @Cody1988Alg665, which is 
+an extension of Mike Malcolm's ENVRON (@Malcolm72). A port of this to R is
+presented below. It computes the 
+machine precision, radix and number of radix digits. A more detailed 
+tool was W. Kahan's PARANOIA (@Karpinski85). The rise of IEEE arithmetic has
+set aside the routine use of such tools, but recent innovations in so-called
+"half-precision", of which there is more than one variety, may result in 
+renewed interest. See https://en.wikipedia.org/wiki/Half-precision_floating-point_format.
 
@@
 
-## Simple performance examples
+## Sys.info()
+
+This R function returns information on the operating system and the generic type of the 
+processor.
+
+```{r sysinfox}
+Sys.info()
+```
+
+## benchmarkme
+
+The CRAN package `benchmarkme` (@benchmarkme) provides a wide range of
+information on system in which it is running, indeed a superset of 
+previous tools mentioned. It is primarily designed to 
+benchmark the capabilities of a particular computing system, so highly appropriate
+to our current effort, though its output is not succinct. Note that we found a 
+minor bug in the `get_ram()`
+function of this package for some Windows 10 versions. Our thanks for 
+Colin Gillespie, the author, for making  
+corrections in version 1.0.8, based on communications with one of us (AB).
+
+?? Need to comment on
+
+?? AB: I found the plots of the benchmarks difficult to read. Which end of the
+plots are "better"?
+
+<!-- - uses `proc.time()` for timings -->
+<!-- - has single core and parallel timings -->
+<!-- - ?? does it do GPUs -->
+<!-- - provides far too much for most needs i.e., a succinct summary of -->
+<!--   how capable a computing system should be -->
+
+<!--  ```{r benchmarkmex}  -->
+<!-- # library(benchmarkmeData) -->
+<!-- # library(benchmarkme) -->
+<!-- # ls(package:benchmarkmeData) -->
+<!-- # ls(package:benchmarkme) -->
+<!-- # lsf.str("package:benchmarkme") # list functions in -->
+<!-- # lsf.str("package:benchmarkmeData") -->
+<!-- ## This next line takes a lot of time to run, so is commented out here -->
+<!-- # benchmark_std() -->
+<!-- # get_byte_compiler() -->
+<!-- # gla <- get_linear_algebra() -->
+<!-- # gsd <- get_sys_details() -->
+<!--  ``` -->
+
+## Some other performance examples
 
@@
 simple performance test have been used by one of the authors for nearly half
-a century. Note the use of the `microbenchmark` timing package (@microbench13).
+a century. We used the `microbenchmark` timing package (@microbench13),
+which allows us to see the variation in timings, which can be the result of
+other activities and processes in our computing systems.
+
+Note that it appears that the first cycle of the loop is very slow. ?? is this
+byte-compiling?
 
@@
 }
+
 require("microbenchmark")
-nvals<-c(1000, 10000, 100000, 1000000)
-tvals<-rep(NA, length(nvals))
-i<-0
-for (nn in nvals) {
-  i <- i + 1
-  cat(nn, "\n")
+# nvals<-c(1000, 10000, 100000, 1000000)
+# tvals<-rep(NA, length(nvals))
+# i<-0
+# for (nn in nvals) {
+#   i <- i + 1
+#   cat(nn, "\n")
+  nn <- 10000
+  library(compiler)
+  le <- cmpfun(loopesc)
+  te <- microbenchmark(le(nn), unit='us')
+  print(te)
+  boxplot(te)
   tt <- microbenchmark(loopesc(nn), unit = 'us')
   print(tt)
-  tvals[i] <- mean(tt$time)
-}
-plot(log10(nvals), log(tvals))
+  boxplot(tt)
+#  ht<-hist(tt$time)
+#  plot(ht, main="Histogram of loopesc times")
+  tt2 <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))
+  print(tt2)
+  boxplot(tt2)
 
+#  ht2<-hist(tt2$time)
+#  plot(ht2, main="Histogram of loopesc times after 2 warmup cycles")
+# }
+# plot(log10(nvals), log(tvals))
 ```
 
-## Sys.info()
+## Missing tools
 
-```{r sysinfox}
-Sys.info()
-```
+While we cannot be certain that they do not exist, there do not appear to be tools
+readily available to allow us to characterize the following features of R computations
+that might be relevant to understanding machine performance.
 
-## Issues relating to random number generation
+- Some users will want to know key information about (pseudo-)random number generation
+  used in their calculations.
+  We suspect that users will need to custom-code ways to get such information
+  about their computations. For example, it is useful to have the specification
+  of the generator and the seed(s) used. For parallel computations, it is important
+  in some applications to use the same set of numbers for each thread, while in
+  others we want to use separate segments of the stream of pseudo-random numbers. 
+  From the specification of the generator, it should be possible to know the 
+  period (how many numbers are generated before the sequence repeats) and 
+  properties concerning correlations between generated numbers. However, in our
+  experience, such information requires a lot of research to uncover.
 
-## Issues relating to compilation of R
+- R allows for byte compilation of code, and there are several ways in which this
+  can be done. `benchmarkme` returns some information about the general setting used,
+  but details would need user intervention if locally written code is used.
 
-- LAPACK
-- BLAS
-- others?
+- Compilation of code written in other programming languages and how the resultant
+  routines are called will similarly require user attention. Fortunately, some 
+  information is available for heavily used software such as the BLAS and LAPACK,
+  for which `sessionInfo()` reports the libraries used. There are, however, several
+  alternatives, some of which claim high-performance.
 
-# Tools for accessing and clearing environments and dataframes
+<!-- # Tools for accessing and clearing environments and dataframes -->
 
-?? should this be here -- probably to list things in the workspace
+<!-- ?? should this be here -- probably to list things in the workspace? -->
 
-```
-sys.frame()
-sys.frames()
-sys.status()
-sys.on.exit()
-sys.parents()
-sys.calls()
-```
-# Choices
+<!-- ``` -->
+<!-- sys.frame() -->
+<!-- sys.frames() -->
+<!-- sys.status() -->
+<!-- sys.on.exit() -->
+<!-- sys.parents() -->
+<!-- sys.calls() -->
+<!-- ``` -->
 
-For use in recording tests of R functions and packages for optimization and nonlinear 
-least squares, it seems that the `benchmarkme` function `get_sys_details()`
-provides more than sufficient information for our needs.
+# A modest profile
 
-From the above discussion, the following offers a possible compact solution. Users may
+For most uses in recording tests of R functions and packages for optimization and nonlinear 
+least squares, which has been our focus, it seems that the `benchmarkme` function 
+`get_sys_details()` provides more than sufficient information for our needs. 
+The following script offers a possible compact solution. Users may
 wish to modify this to their own particular needs.
@@
 
-Note that we are reading from the file `MachID.R` so that this file is usable elsewhere
-and the current document does not become out of sync with the working MachID.
+<!-- Note that we are reading from the file `MachID.R` so that this file is usable elsewhere -->
+<!-- and the current document does not become out of sync with the working MachID. -->
 

### 2022-06-14 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Minor edits

@@
 - tools such as the `nls()` function in R has so many legacy uses that it is likely
-  impossible to deprecate it. However, it has also (as we detail in ??RefactoringNLS??)
-  non-maintainable, but it seems unlikely that any
+  impossible to deprecate it. However, as we detail in a related article, "Refactoring 
+  the `nls()` function in R" 
+  <!-- ??RefactoringNLS??,  -->
+  it is practically non-maintainable. It seems unlikely that any
   further development, nor even repair of current deficiencies, will be forthcoming.

### 2022-06-14 MachineSummary/MachineSummary.Rmd
added gb vs gib

@@
 well as different interpretations of some of the units of measure. For example
-"Gigabytes", or GB, may be $1000^3$ or $1024^3$ bytes. ?? AB? Is that correct?
+"Gigabytes" may be $1000^3$ or $1024^3$ bytes. To settle the ambiguity, the
+international standard *IEC 80000-13:2008 Quantities and units - Part 13: Information science and technology*(??ref https://en.wikipedia.org/wiki/Gigabyte, https://en.wikipedia.org/wiki/ISO/IEC_80000#Part_13:_Information_science_and_technology) defines 1 Gigabyte (1 GB) as 1000^3 bytes, whereas 1 Gibibyte (1 GiB) as 1024^3 bytes.
+
+<!-- ?? AB? Is that correct? -->
 

### 2022-06-14 MachineSummary/MachineSummary.Rmd

@@
 # and here is a way to get the UTC time
-tutc<-as.POSIXlt(Sys.time(), tz = "UTC")
+tutc <- as.POSIXlt(Sys.time(), tz = "UTC")
 print(tutc)
-utctstmp<-format(tutc, "%Y/%m/%d|%H:%M")
+utctstmp <- format(tutc, "%Y/%m/%d|%H:%M")
 print(utctstmp)
@@
 `1 + eps` is greater than `1`. R performs computations in IEEE "double"
-precision, and provides the R variable `.Machine` that has " information on 
+precision, and provides the R variable `.Machine` that has "information on 
 the numerical characteristics of the machine R is running on", including
@@
 cat( "E9*1E+16=", E9 * 1E+16, "\n")
-cat( "E6 = ",E6,"  -- the radix of arithmetic", "\n")
-cat( "J1 = ",J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
+cat( "E6 = ", E6,"  -- the radix of arithmetic", "\n")
+cat( "J1 = ", J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
 cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")
@@
     "double.base =", .Machine$double.base, "\n", 
-    "double.digits =", .Machine$double.digits, "\n" )
+    "double.digits =", .Machine$double.digits, "\n")
 ```
@@
 function of this package for some Windows 10 versions. Our thanks for 
-Colin Gillespie, the author, for making  
-corrections in version 1.0.8, based on communications with one of us (AB).
+Colin Gillespie, the author, for making  corrections in version 1.0.8, based on communications with one of us (AB).
 

### 2022-06-14 MachineSummary/loopesctimes.R
clean up background, make copies unversioned

@@ -1,0 +1,29 @@
+loopesc <- function(nn){
+  ss <- 0
+  for (i in 1:nn) {
+    xx <- exp(sin(cos(1.0*i)))
+    ss <- ss + xx
+  }
+  xx
+}
+
+require("microbenchmark")
+# nvals<-c(1000, 10000, 100000, 1000000)
+# tvals<-rep(NA, length(nvals))
+# i<-0
+# for (nn in nvals) {
+#   i <- i + 1
+#   cat(nn, "\n")
+nn <- 10000
+tt <- microbenchmark(loopesc(nn), unit = 'us')
+print(tt)
+ht<-hist(tt$time)
+plot(ht, main="Histogram of loopesc times")
+tmp<-readlines("next?")
+graphics.off()
+tt2 <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))
+print(tt2)
+ht2<-hist(tt2$time)
+plot(ht2, main="Histogram of loopesc times after 2 warmup cycles")
+# }
+# plot(log10(nvals), log(tvals))

### 2022-06-14 MachineSummary/MachineSummary.Rmd
Clean up sections except boxplots

@@
 There may be multiple terms used to describe some of the information we want, as
-well as different interpretations of some of the units of measure. For example
-"Gigabytes" may be $1000^3$ or $1024^3$ bytes. To settle the ambiguity, the
-international standard *IEC 80000-13:2008 Quantities and units - Part 13: Information science and technology*(??ref https://en.wikipedia.org/wiki/Gigabyte, https://en.wikipedia.org/wiki/ISO/IEC_80000#Part_13:_Information_science_and_technology) defines 1 Gigabyte (1 GB) as 1000^3 bytes, whereas 1 Gibibyte (1 GiB) as 1024^3 bytes.
-
-<!-- ?? AB? Is that correct? -->
+well as different interpretations of some of the units of measure. For example,
+storage for disks or memory has decimal and binary schemes of units. Thus 
+(see https://en.wikipedia.org/wiki/Gigabyte) there is a movement to use, for example, Gigabyte or GB to mean usits of 1000^3 bytes, while the Gigibyte or Gib is 1024^3.
 
@@
 sides of the globe, it could be important to either specify time zone or to work with a
-single "clock" such as UTC (ref?? or expansion?)
+single "clock" such as UTC. R provides tools to allow this, as in the following example.
 
@@
 - operating cycle speed (this is sometimes specific to the cpu specified )
+
+?? how do we reconcile?
+
+<!-- From get_sys_details() -->
+<!-- $cpu$model_name -->
+<!-- [1] "Intel(R) Core(TM) i5-10400 CPU @ 2.90GHz" -->
+
+<!-- From inxi -C -->
+<!-- john@M21:~$ inxi -C -->
+<!-- CPU:       Topology: 6-Core model: Intel Core i5-10400 bits: 64 type: MT MCP L2 cache: 12.0 MiB  -->
+<!--            Speed: 800 MHz min/max: 800/4300 MHz Core speeds (MHz): 1: 800 2: 800 3: 800 4: 800 5: 800 6: 800 7: 800 8: 800  -->
+<!--            9: 800 10: 800 11: 800 12: 800  -->
+
+?? also noted that get_sys_details shows 16 cores on my laptop, but inxi shows 8??
+
 - RAM size total
@@
 - RAM speed (which could be variable) 
+  ?? this is, as I understand the output, in inxi (for Linux). Not as far as I know in benchmarkme.
+
 - GPU (i.e, additional processors) information. 
+  ?? is there a way to get this? I don't have a GPU on my desktop machine.
 
@@
 Clearly it is important to identify which software packages are active, and the
-R `sessionInfo()` command is useful for this. However, there may be additional 
+R `sessionInfo()` command provides this. However, there may be additional 
 features that should be mentioned if the user has modified the software in any
@@
 least squares, which has been our focus, it seems that the `benchmarkme` function 
-`get_sys_details()` provides more than sufficient information for our needs. 
-The following script offers a possible compact solution. Users may
-wish to modify this to their own particular needs.
+`get_sys_details()` provides more than sufficient information for most of our needs. 
+A workable profile for our project is captured in the following script,
+which allows a 1-line compact sketch of the system in use.
 
@@
 
+For other applications, it is clear that users will make other choices to meet their particular needs.
+
+Our efforts noted in this article underline the following principles.
+
+- Relatively comprehensive tools such as package `benchmarkme` are needed to 
+  make possible machine profiles to allow sensible comparison of computational 
+  performance across different platforms.
+
+- If a project uses a common software environment such as R, then the tools for
+  capturing profile data should be available from that environment, and run 
+  under different operating systems and hardware. In that respect, `benchmarkme`
+  has been very helpful to us.
+
+- No tool covers all the possibilities, and users may have to do extra work to 
+  capture information they need. Moreover, the challenges of cross-platform 
+  computing ensure that some of the results of different tools will be inaccurate.
+
+- Summarizing capability information for computing environments to provide a 
+  short-form profile is a challenge that requires care and attention. Indeed,
+  it is much easier to gather a large body of information than to devise a 
+  compact but useful precis of it.
+
+We welcome conversation on how machine capability may be summarized and reported
+to allow sensible comparison of computational performance of algorithms and 
+systems.
+
 ## References

### 2022-06-14 MachineSummary/MachineSummary.Rmd
Fix boxplots

@@
 
-Note that it appears that the first cycle of the loop is very slow. ?? is this
-byte-compiling?
+Note that it appears that the first cycle of the loop is very slow. Fortunately 
+`microbenchmark` seems aware of this, and we can specify a number of "warmup"
+cycles. In the present case, we tried 2 cycles. We also tried byte compiling
+the loopesc function with the `compiler` package. Note that rerunning the 
+code several times gave slightly different boxplots, but the general characteristics
+were preserved.
 
-```{r loopesc}
+```{r loopesc, echo=TRUE}
 loopesc <- function(nn){
@@
 }
+```
 
-require("microbenchmark")
-# nvals<-c(1000, 10000, 100000, 1000000)
-# tvals<-rep(NA, length(nvals))
-# i<-0
-# for (nn in nvals) {
-#   i <- i + 1
-#   cat(nn, "\n")
-  nn <- 10000
-  library(compiler)
-  le <- cmpfun(loopesc)
-  te <- microbenchmark(le(nn), unit='us')
-  print(te)
-  boxplot(te)
-  tt <- microbenchmark(loopesc(nn), unit = 'us')
-  print(tt)
-  boxplot(tt)
-#  ht<-hist(tt$time)
-#  plot(ht, main="Histogram of loopesc times")
-  tt2 <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))
-  print(tt2)
-  boxplot(tt2)
-
-#  ht2<-hist(tt2$time)
-#  plot(ht2, main="Histogram of loopesc times after 2 warmup cycles")
-# }
-# plot(log10(nvals), log(tvals))
+```{r looptime, code=readLines("loopbox.R"), echo=FALSE}
 ```

### 2022-06-14 MachineSummary/loopbox.R

@@ -1,0 +1,20 @@
+# loopbox.R
+loopesc <- function(nn){
+  ss <- 0
+  for (i in 1:nn) {
+    xx <- exp(sin(cos(1.0*i)))
+    ss <- ss + xx
+  }
+  xx
+}
+
+require("microbenchmark")
+nn <- 10000
+library(compiler)
+le <- cmpfun(loopesc)
+tcompiled <- microbenchmark(le(nn), unit='us')$time
+tcoldstart <- microbenchmark(loopesc(nn), unit = 'us')$time
+twarmstart <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=10))$time
+tdf <- data.frame(tcoldstart, twarmstart, tcompiled)
+boxplot(x=as.list(log(tdf)), main="log times for 10000 loops in microsecs")
+

### 2022-06-15 VarietyInNLS/VarietyInNonlinearLeastSquaresCodes.Rmd
Begin simplifying and shortening Variety

@@
 Numerical analysts will generally want to solve a **nonlinear least squares** problem,
-that is, minimize an objective function with respect to a set of $n$ parameters, $x$, where the 
+that is, minimize an objective function with respect to a set of $n$ parameters, 
+$x$, where the 
 objective  $f()$ can be written as a sum of $m$ squared elements.
@@
 
-There is no exogenous data here.
-
-There may be additional conditions that add to the complexity of
-the problem such as
+There is no exogenous data here. However, there may be additional 
+conditions that add to the complexity of the problem such as
 
@@
 
-where the exogenous data is a single vector of numbers $y$. In nonlinear regression, we are mostly
-interested in "fitting a model to data", and the "minimizing a sum of squares" is the approach.
+where the exogenous data is a single vector of numbers $y$. 
+In nonlinear regression, we are mostly interested in "fitting a model to data", 
+and the "minimizing a sum of squares" is the approach.
 
@@
 useful classification is according to whether a fixed number of computations can estimate the
-parameters. That is, problems can be considered "nonlinear" if we cannot compute the best parameter values
+parameters. That is, problems can be considered "nonlinear" if we cannot 
+compute the best parameter values
 in a pre-defined fixed number of computations and must use an iterative approach.
 
-In what follows, we will primarily take the numerical analyst's viewpoint. That is, we will be looking
-mostly at the calculations for the minimization of the sum of squares. However, the purpose of these
-calculations can dictate much of the structure of the software in which such computations appear.
+In what follows, we will primarily take the numerical analyst's viewpoint. 
+That is, we will be looking mostly at the calculations for the minimization of 
+the sum of squares. However, the purpose of these calculations can dictate much 
+of the structure of the software in which such computations appear.
 
@@
 An overview of the method is provided by Wikipedia: *https://en.wikipedia.org/wiki/Gauss%E2%80%93Newton_algorithm*.
-
-The method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
+Indeed, the method was described by C.F.Gauss in 1809 in *Theoria motus corporum 
 coelestium in sectionibus conicis solem ambientum*.
@@
 $$ x \leftarrow x + \delta $$
-
 and we iterate until there is no change in $x$, assuming that the method does indeed
@@
 
-The final vector product form is the one I favour because it is the least fussy to write and read.
-In particular, when we try to evaluate the Hessian of this sum of squares function, we see that it is
+The final vector product form is the one I favour because it is the 
+least fussy to write and read. In particular, when we try to evaluate the Hessian 
+of this sum of squares function, we see that it is
 
@@
 $$  J^t J $$
-
 Arguing that the residuals should be "small", Gauss proposed that the 
@@
 $$ (J^t J)  \delta  =  - J^t r $$
-
 We can use this in an iteration similar to the Newton iteration. Note that these
@@
 $$  J \delta  \approx - r $$
-
 Like most other treatments, we will skate over the thin ice that $J^t J$ is a 
@@
 $$ x \leftarrow x + \delta$$
-
 one could do a search along the direction $\delta$. Ideally we would want to **minimize**
@@
 $$ f(x + \alpha \delta) $$
-
 with respect to the (scalar) parameter $\alpha$. This one-dimensional minimization can
 take quite a lot of computational effort, however, and typically a value of $\alpha$ that
-permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the iteration repeated
-from
+permits a suitable reduction in the sum of squares $f(x)$ is accepted, and the 
+iteration repeated from
 
 $$ x_{new} = x + \alpha * \delta $$
-
 Clearly there are many tactical choices that give rise to a variety of particular algorithms.
-One concern is that the direction $\delta$ may give no lower value of the sum of squares, since
-there is an approximation involved in using $J^t J$ rather than $H$. For example, $\delta$ could
-be perpendicular to any downhill direction.
+One concern is that the direction $\delta$ may give no lower value of the sum of squares, 
+since there is an approximation involved in using $J^t J$ rather than $H$. 
+For example, $\delta$ could be perpendicular to any downhill direction.
 
@@
 
-In `nls.c` we also see that once a satisfactory "new" set of parameters has been 
+In R's `nls.c` we also see that once a satisfactory "new" set of parameters has been 
 obtained, we reset the $\alpha$ value to the minimum of $2 * \alpha$ and 1.0. Oddly, there
@@
 $$  K \delta \approx - r_{augmented} $$
-
 where we have added $n$ zeros to $r$ to make $r_{augmented}$. Similar adjustments
@@
 
+<!-- ### Variants -->
+
+<!-- The central idea of Hartley's method is a line search and that of Marquardt a -->
+<!-- parametrized stabilization that can be shown to reduce a sum of squares for a large -->
+<!-- enough stabilization parameter. There is nothing to prevent use of a line -->
+<!-- search additional to the Marquardt stabilization, and there are, of course,  -->
+<!-- several reasonable choices for line search. Also we could solve the (possibly -->
+<!-- stabilized) Gauss-Newton equations by several quite different numerical linear -->
+<!-- algebra techniques, or even approximations. In certain situations, some of these -->
+<!-- choices may be important. Unfortunately, the effort to benchmark the possibilities -->
+<!-- seems higher than any benefit in replacing quite good existing methods. -->
+
 ## Other approaches
 
-While the above ideas are the dominant theme in nonlinear least squares, there have been many 
-other proposed approaches. The following is a non-exhaustive sample.
-
-### Spiral
-
-@Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart
-given but I have never been able to find a code (supposedly Fortran V for the Univac
-1108.) The motivations are to avoid both matrix equation solutions and one-dimensional
-line-search minimization. Jones mentions matrix inversion, but it is not clear whether
-he truly forms the inverse. The main idea in the paper (and presumably the code) is
-to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$
-is increased. It would be interesting to actually try some of the ideas, which do,
-as far as I can determine, have a reasonable basis and could possibly offer some
-efficiencies.
-
-### Approaches to Newton's method for special models
+While the above ideas are the dominant themes in nonlinear least squares, 
+there have been many other approaches. 
+<!-- The following is a non-exhaustive sample. -->
+
+<!-- ### Spiral  -->
+
+<!-- @Jones70Spiral seems only to be mentioned in a single paper. There is a flowchart -->
+<!-- given but I have never been able to find a code (supposedly Fortran V for the Univac -->
+<!-- 1108.) The motivations are to avoid both matrix equation solutions and one-dimensional -->
+<!-- line-search minimization. Jones mentions matrix inversion, but it is not clear whether -->
+<!-- he truly forms the inverse. The main idea in the paper (and presumably the code) is -->
+<!-- to use a geometric spiral that mimics the Marquadt search trajectory as $\lambda$ -->
+<!-- is increased. It would be interesting to actually try some of the ideas, which do, -->
+<!-- as far as I can determine, have a reasonable basis and could possibly offer some -->
+<!-- efficiencies. -->
+
+### Using Newton's method for special models
 
@@
 to them applied to a very special case and in a manner unrecognizable as that described
-above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*).
-
-The main attraction of the Gauss-Newton method is that the Jacobian only needs first
-derivatives, which are generally much, much easier to compute than the full second
-derivatives of the Hessian. However, there have been suggestioned
-Newton-like approaches, possibly for special types of models where some short cut
-to the Hessian is available. These may or may not make sense in the context of
-a given user or group of users. I regard these approaches as part of the
-computational infrastructure of the special cases rather than as general nonlinear
-least squares methods. The `maxLik` package (@maxLik2011) is a prime example.
-
-### Hybrid methods
-
-The central idea of Hartley's method is a line search and that of Marquardt a
-parametrized stabilization that can be shown to reduce a sum of squares for a large
-enough stabilization parameter. There is nothing to prevent use of a line
-search additional to the Marquardt stabilization, and there are, of course, 
-several reasonable choices for line search. Also we could solve the (possibly
-stabilized) Gauss-Newton equations by several quite different numerical linear
-algebra techniques, or even approximations. In certain situations, some of these
-choices may be important. Unfortunately, the effort to benchmark the possibilities
-seems higher than any benefit in replacing quite good existing methods.
+above.  (See *https://en.wikipedia.org/wiki/Newton%27s_method*). Since Newton's
+method requires the full Hessian and its second derivatives, we do not see it applied
+except when the Hessian is easily and efficiently calculated.
+The Gauss-Newton method needs only first derivatives for the Jacobian. Generally, 
+these are much easier to compute. However, for special types of models where some short cut
+to the Hessian is available, Newton may be worthwhile. The `maxLik` package (@maxLik2011) 
+for R is an example.
+
+### Using general optimization methods
+
+There is no prohibition on applying algorithms for general function minimization, 
+even with constraints, to problems where the objective is a sum of squared functions.
+General methods also contend with issues of approximating the Hessian
+or ensuring iterates are satisfactory, but there may be additional advantages
+with a sum of squared function.
+
+?? nlminb (from Gay etc. nlr)
+?? nl2sol (??)
+?? DUD ??
+
+# Differences in user interface
 
@@
 However, in the now defunct @jnnlmrt12 and successor @nlsr-manual, the choice
-has been to use a QR decomposition of an augmented Jacobian. Similarly, `nls()` uses a QR decomposition
-of the Jacobian (referred to as the "gradient") without augmentation. Clearly the $K$ matrix is
-at least $m$ by $n$, which for most nonlinear modelling problems is much larger than the $C$ matrix.
-The particular QR calculations in `nls()` and its packaged version are quite well-hidden in internal
+has been to use a QR decomposition of an augmented Jacobian. Similarly, `nls()` uses 
+a QR decomposition of the Jacobian (referred to as the "gradient") without augmentation.
+Clearly the $K$ matrix is at least $m$ by $n$, which for most nonlinear modelling problems 
+is much larger than the $C$ matrix.
+The particular QR calculations in `nls()` and its packaged version are 
+quite well-hidden in internal
 code, complicating comparisons of storage, complexity and performance.
@@
 
-In @Nash1987a, each major iteration generates a $A = J^t J$ matrix and the right hand side $-g = J^t r$
+In @Nash1987a, each major iteration generates a $A = J^t J$ matrix and the right 
+hand side $-g = J^t r$
 at parameters $x$ using functions for the residuals and Jacobian.
 The lower triangle of $A$ is stored as a vector of $n (n+1)/2$ elements.
-The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ is added to each,
-then saved in matrix $C$. $\lambda$ is our Marquardt parameter and $\phi$ is the adjustment suggested in
-@jn77ima to
-overcome potential underflow to zero of the accumulation of the diagonal elements in low-precision
-arithmetic.
-
-Solving $C \delta = -g$, we compute the new sum of squares at $x + \delta$. If this value is greater than
-or equal to the current best (i.e., lowest) sum of squares, we increase $\lambda$ and generate a new
-$C$ matrix. We do not, however, need to build a new SSCP matrix or gradient $g$, just make the diagonal
-adjustment to $A$ to get a new $C$.
+The diagonal elements are then multiplied by $(1 + \lambda)$ and $\lambda * \phi$ 
+is added to each,
+then saved in matrix $C$. $\lambda$ is our Marquardt parameter and $\phi$ 
+is the adjustment suggested in @jn77ima to
+overcome potential underflow to zero of the accumulation of the 
+diagonal elements in low-precision arithmetic.
+
+Solving $C \delta = -g$, we compute the new sum of squares at $x + \delta$. 
+If this value is greater than or equal to the current best (i.e., lowest) sum of 
+squares, we increase $\lambda$ and generate a new
+$C$ matrix. We do not, however, need to build a new SSCP matrix or gradient $g$, 
+just make the diagonal adjustment to $A$ to get a new $C$.
 
@@
 
-- tests for "non-computability", where we allow the user to provide an explicit flag that the function cannot
-  be computed at the trial parameters. This was included to avoid some of the weaknesses in evaluation of 
-  special functions, e.g., `log()` for small arguments or `exp()` for large ones, or for problem specific
-  situations where some parameter combinations might be considered inadmissible. (Readers may note that
-  the codes were written before @IEEE754-1985 was applied to most BASICs. The author was a member of
-  that standardization committee.)
-
-- situations where rounding and truncation errors make the Cholesky decomposition return with an indication
-  that $C$ is "singular", at least in the particular computational arithmetic.
+- tests for "non-computability", where we allow the user to provide an explicit 
+  flag that the function cannot be computed at the trial parameters. This was included 
+  to avoid some of the weaknesses in evaluation of special functions, e.g., `log()` 
+  for small arguments or `exp()` for large ones, or for problem specific
+  situations where some parameter combinations might be considered inadmissible. 
+  (Readers may note that the codes were written before @IEEE754-1985 was applied to 
+  most BASICs. The author was a member of that standardization committee.)
+
+- situations where rounding and truncation errors make the Cholesky decomposition 
+  return with an indication that $C$ is "singular", at least in the particular 
+  computational arithmetic.
 
@@
 
-In @nlsr-manual, the approach to the augmented Gauss-Newton equations is nominally equivalent to that in
-@Nash1987a,
-except that we perform the solution using a QR decomposition, in fact a modification of the DQRDC routine from 
+In @nlsr-manual, the approach to the augmented Gauss-Newton equations is nominally 
+equivalent to that in @Nash1987a, except that we perform the solution using a QR
+decomposition, in fact a modification of the DQRDC routine from 
 @lapack99, which uses Householder transformations.
 
-One of the issues with creating the SSCP matrix $A$ is that we lose information in finite arithmetic 
-(See @jncnm79, Chapter 5). As noted above, a QR approach to the traditional Gauss-Newton codes is to solve
+One of the issues with creating the SSCP matrix $A$ is that we lose information 
+in finite arithmetic 
+(See @jncnm79, Chapter 5). As noted above, a QR approach to the traditional 
+Gauss-Newton codes is to solve
 
@@
 
-That is, we assume a generalized inverse of $J$ can be applied to both sides of the Gauss-Newton equations.
-
-To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. Also as described above,
+That is, we assume a generalized inverse of $J$ can be applied to both sides 
+of the Gauss-Newton equations.
+
+To move to the Marquardt variant of the Gauss-Newton, we need to augment $J$. 
+Also as described above,
 augmented matrix $K$ is a row-wise appending $Z1$ and, for the Nash modification $Z2$, to $J$.
@@
 
-We have noted that, at each major iteration, the construction of $Z1$ requires the diagonal elements of the
-SSCP matrix. Thus it could be more efficient to omit the $Z1$ augmentation and use an explicit scaling
+We have noted that, at each major iteration, the construction of $Z1$ requires 
+the diagonal elements of the
+SSCP matrix. Thus it could be more efficient to omit the $Z1$ augmentation 
+and use an explicit scaling
 in the original setup. Package `nlsralt`, as
@@
 
-# Implementation comparisons
+# Implementation considerations
 
@@
 above by adding a unit matrix scaled by e.g., 1e-5, and padding the RHS with zeros.
-The solution is "almost but not quite" the same because the formation of the sum of squares and
+The solution is "almost but not quite" the same because the formation of the 
+sum of squares and
 cross products matrix explicitly in the normal equations can introduce errors that
 are partially avoidable in the QR approach below.
+
+?? More 77 item
 

### 2022-06-15 RefactoringNLS/RefactoringNLS.Rmd
Minor edits
Continued work on Refactoring

@@
 
-This article considers the features and limitations of the R function `nls()`. It
-arose from a Google Summer of Code project "Improvements to `nls()`", a tool 
-intended for the estimation of models written as a formula that has at least
-one parameter that is "nonlinear". That is, the model is not estimable via solving a 
-finite set of 
-linear equations. A companion document **Variety in Nonlinear Least Squares Codes** 
-presents an overview of methods for the problem which takes a much wider view of the
-problem of minimizing a function that can be written as a sum of squared
-terms, and of the even more general case of nonlinear regression.
-
-An important overall consideration in our work has been the maintainability of
-the code base that supports the `nls()` functionality, as we believe that the
-existing code makes maintenance and improvement very difficult. Moreover, legacy
-applications and examples mean some improvements are blocked. Thus we have not
-addressed all the issues raised, but we believe that the discussion and trials
-so far are relevant to improving both `nls()` and R generally.
+Based on work for a Google Summer of Code project "Improvements to `nls()`",
+we consider the features and limitations of the R function `nls()`
+in the context of improving and rationalizing R tools for nonlinear regression.
+
+An important consideration is the  maintainability of
+the code base that provides the functionality `nls()` claims to offer.
+Our work suggests that the
+existing code makes maintenance and improvement very difficult,
+with legacy applications and examples blocking some important updates.
+Discussion of these matters is relevant to improving R generally, as
+well as its nonlinear estimation tools.
 
@@
 `nls()` is the tool in base R (the distributed software package from [CRAN](https://cran.r-project.org)
-for estimating nonlinear statistical models. The function was developed mainly in the 1980s and
-1990s by Doug Bates et al., initially for S (see https://en.wikipedia.org/wiki/S_%28programming_language%29) . 
-The ideas spring primarily from the book by @bateswatts.
-
-The `nls()` function has a remarkable and quite comprehensive set of capabilities for estimating
-nonlinear models that are expressed as formulas. In particular, we note that it
+for estimating nonlinear statistical models. The function dates to the 1980s and the work related
+to @bateswatts in S (see https://en.wikipedia.org/wiki/S_%28programming_language%29).
+
+The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
+nonlinear models that are expressed as **formulas**. In particular, we note that it
 
@@
 - has related profiling capabilities for exploring the likelihood surface as parameters are changed
-
-With such a range of features and a long history, it is not surprising that the code has become untidy 
+- links to a number of pre-coded ("selfStart") models
+
+With such a range of features and long history, the code has become untidy 
 and overly patched. It is, to
-our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved. Let
-us review some of the issues. We will then propose corrective actions, some of which we have carried out
-either in demonstrations or in the distributed code.
+our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
+as we suggest below.
 
@@
 
-This amounts to admitting R cannot solve perfectly well-posed problems. The 
-suggestion that one needs to pollute data with errors is one that should be offensive
-to good science.
-
-It turns out that this issue can be easily resolved. The key "convergence test" -- more properly
-a "termination test" for the **program** rather than for convergence of the underlying **algorithm** --
-is the Relative Offset Convergence Criterion (see @BatesWatts81). This works by projecting the proposed
-step in the parameter vector on the gradient and estimating how much the sum of squares loss function 
-will decrease. This is divide by the current size of the loss function to avoid scale issues.
+This amounts to admitting R cannot solve well-posed problems unless data is polluted
+with errors.
+
+This issue can be easily resolved. The "termination test" for the **program** rather than for 
+"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
+(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
+and estimates how much the sum of squares loss function should decrease. This estimate is divided 
+by the current size of the loss function to avoid scale issues.
 When we have "converged", the estimated
-decrease is very small, as usually is its ratio to the sum of squares. However, with small residuals,
+decrease is very small. However, with small residuals,
 the sum of squares loss function is (almost) zero and we get the possibility of a 
@@
 
-### Example of a small-residual problem
-
-```
-t <- -10:10
-y <- 100/(1 + .1 * exp(-0.51 * t))
-lform <- y ~ a/(1 + b * exp(-c * t))
-ldata <- data.frame(t = t, y = y)
-plot(t, y)
-lstartbad <- c(a = 1, b = 1, c = 1)
-lstart2 <- c(a = 100, b = 10, c = 1)
-nlsr::nlxb(lform, data = ldata, start = lstart2)
-nls(lform, data = ldata, start = lstart2, trace = TRUE)
-# Fix with scaleOffset
-nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
-sessionInfo()
-```
-
-Edited output of running this function follows:
-
-```
-> t <- -10:10
-> y <- 100/(1 + .1 * exp(-0.51 * t))
-> lform <- y ~ a/(1 + b * exp(-c * t))
-> ldata <- data.frame(t=t, y=y)
-> plot(t,y)
-> lstartbad <- c(a = 1, b = 1, c = 1)
-> lstart2 <- c(a = 100, b = 10, c = 1)
-> nlsr::nlxb(lform, data = ldata, start = lstart2)
-nlsr object: x 
-residual sumsquares =  1.007e-19  on  21 observations
-    after  13    Jacobian and  19 function evaluations
-  name            coeff          SE       tstat      pval      gradient    JSingval   
-a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6  
-b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3  
-c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791  
-> nls(lform, data = ldata, start = lstart2, trace = TRUE)
-40346.    (1.08e+00): par = (100 10 1)
-11622.    (2.93e+00): par = (101.47 0.49449 0.71685)
-5638.0    (1.08e+01): par = (102.23 0.38062 0.52792)
-642.08    (1.04e+01): par = (102.16 0.22422 0.41935)
-97.712    (1.79e+01): par = (100.7 0.14774 0.45239)
-22.250    (1.78e+02): par = (99.803 0.093868 0.50492)
-0.025789  (1.33e+03): par = (100.01 0.10017 0.50916)
-6.0571e-08 (7.96e+05): par = (100 0.1 0.51)
-4.7017e-19 (1.86e+04): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-... (approx 40 lines omitted)
-1.2440e-27 (5.71e-01): par = (100 0.1 0.51)
-Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) : 
-  number of iterations exceeded maximum of 50
-> # Fix with scaleOffset
-> nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0))
-40346.      (1.08e+00): par = (100 10 1)
-11622.      (2.91e+00): par = (101.47 0.49449 0.71685)
-5638.0      (9.23e+00): par = (102.23 0.38062 0.52792)
-642.08      (5.17e+00): par = (102.16 0.22422 0.41935)
-97.712      (2.31e+00): par = (100.7 0.14774 0.45239)
-22.250      (1.11e+00): par = (99.803 0.093868 0.50492)
-0.025789    (3.79e-02): par = (100.01 0.10017 0.50916)
-6.0571e-08  (5.80e-05): par = (100 0.1 0.51)
-4.7017e-19  (1.62e-10): par = (100 0.1 0.51)
-Nonlinear regression model
-  model: y ~ a/(1 + b * exp(-c * t))
-   data: ldata
-     a      b      c 
-100.00   0.10   0.51 
- residual sum-of-squares: 4.7e-19
-Number of iterations to convergence: 8 
-Achieved convergence tolerance: 1.62e-10
-
-Running under: Linux Mint 20.3
-```
+<!-- ### Example of a small-residual problem -->
+
+<!-- ``` -->
+<!-- t <- -10:10 -->
+<!-- y <- 100/(1 + .1 * exp(-0.51 * t)) -->
+<!-- lform <- y ~ a/(1 + b * exp(-c * t)) -->
+<!-- ldata <- data.frame(t = t, y = y) -->
+<!-- plot(t, y) -->
+<!-- lstartbad <- c(a = 1, b = 1, c = 1) -->
+<!-- lstart2 <- c(a = 100, b = 10, c = 1) -->
+<!-- nlsr::nlxb(lform, data = ldata, start = lstart2) -->
+<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
+<!-- # Fix with scaleOffset -->
+<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
+<!-- sessionInfo() -->
+<!-- ``` -->
+
+<!-- Edited output of as small example follows: -->
+
+<!-- ``` -->
+<!-- > t <- -10:10 -->
+<!-- > y <- 100/(1 + .1 * exp(-0.51 * t)) -->
+<!-- > lform <- y ~ a/(1 + b * exp(-c * t)) -->
+<!-- > ldata <- data.frame(t=t, y=y) -->
+<!-- > plot(t,y) -->
+<!-- > lstartbad <- c(a = 1, b = 1, c = 1) -->
+<!-- > lstart2 <- c(a = 100, b = 10, c = 1) -->
+<!-- > nlsr::nlxb(lform, data = ldata, start = lstart2) -->
+<!-- nlsr object: x  -->
+<!-- residual sumsquares =  1.007e-19  on  21 observations -->
+<!--     after  13    Jacobian and  19 function evaluations -->
+<!--   name            coeff          SE       tstat      pval      gradient    JSingval    -->
+<!-- a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6   -->
+<!-- b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3   -->
+<!-- c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791   -->
+<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
+<!-- 40346.    (1.08e+00): par = (100 10 1) -->
+<!-- 11622.    (2.93e+00): par = (101.47 0.49449 0.71685) -->
+<!-- 5638.0    (1.08e+01): par = (102.23 0.38062 0.52792) -->
+<!-- 642.08    (1.04e+01): par = (102.16 0.22422 0.41935) -->
+<!-- 97.712    (1.79e+01): par = (100.7 0.14774 0.45239) -->
+<!-- 22.250    (1.78e+02): par = (99.803 0.093868 0.50492) -->
+<!-- 0.025789  (1.33e+03): par = (100.01 0.10017 0.50916) -->
+<!-- 6.0571e-08 (7.96e+05): par = (100 0.1 0.51) -->
+<!-- 4.7017e-19 (1.86e+04): par = (100 0.1 0.51) -->
+<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
+<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
+<!-- ... (approx 40 lines omitted) -->
+<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
+<!-- Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) :  -->
+<!--   number of iterations exceeded maximum of 50 -->
+<!-- > # Fix with scaleOffset -->
+<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
+<!-- 40346.      (1.08e+00): par = (100 10 1) -->
+<!-- 11622.      (2.91e+00): par = (101.47 0.49449 0.71685) -->
+<!-- 5638.0      (9.23e+00): par = (102.23 0.38062 0.52792) -->
+<!-- 642.08      (5.17e+00): par = (102.16 0.22422 0.41935) -->
+<!-- 97.712      (2.31e+00): par = (100.7 0.14774 0.45239) -->
+<!-- 22.250      (1.11e+00): par = (99.803 0.093868 0.50492) -->
+<!-- 0.025789    (3.79e-02): par = (100.01 0.10017 0.50916) -->
+<!-- 6.0571e-08  (5.80e-05): par = (100 0.1 0.51) -->
+<!-- 4.7017e-19  (1.62e-10): par = (100 0.1 0.51) -->
+<!-- Nonlinear regression model -->
+<!--   model: y ~ a/(1 + b * exp(-c * t)) -->
+<!--    data: ldata -->
+<!--      a      b      c  -->
+<!-- 100.00   0.10   0.51  -->
+<!--  residual sum-of-squares: 4.7e-19 -->
+<!-- Number of iterations to convergence: 8  -->
+<!-- Achieved convergence tolerance: 1.62e-10 -->
+<!-- ``` -->
 
@@
 squares** test (**smallsstest**) that compares the latest evaluated sum of squared
-(weighted) residuals to `e4` times the initial sum of squares. The value of 
-`e4` is computed as
-
-```{r meps4, echo=TRUE}
-epstol<-100*.Machine$double.eps
-e4 <- epstol^4
-e4
-```
+(weighted) residuals to `e4` times the initial sum of squares, where
+`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
 
 We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
-Unfortunately, for almost all
-iterative algorithms, the meaning of "iteration" requires careful examination of the
-code. we prefer to express such tests using the number of times the residuals or the jacobian have
+Unfortunately, the meaning of "iteration" varies among programs and requires 
+careful examination of the
+code. We prefer to use the number of times the residuals or the jacobian have
 been computed and put upper limits on these. Our codes exit (terminate) when these limits
 are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
-but that may simply reflect our history of dealing with more difficult problems as we are
-the tool-makers; users consult us when things go wrong.
+but that may reflect the more difficult problems we have encountered because 
+users consult us when standard tools have given unsatisfactory results.
 
@@
 
-This is the infamous "singular gradient" termination message of `nls()`. A Google search of
+This refers to the infamous "singular gradient" termination message of `nls()`. 
+A Google search of
 
@@
 
-gets over 4000 hits that are spread over the years. 
-In some cases this is due
-to the failure of the simple finite difference approximation of the Jacobian in the
-`numericDeriv()` function that is a part of `nls()`. `nlsr` can use analytic
-derivatives, and we could import this functionality to the `nls()` code as an
-improvement. See below in the section **Jacobian computation**.
-
-A common source of the issue is that the Jacobian is very close
+gets over 4000 hits that are spread over some years. This could be because the 
+Jacobian is poorly approximated (see **Jacobian computation** below). However,
+it is common in nonlinear least squares computations that the Jacobian is very close
 to singular for some values of the model parameters. In such cases we need to find
@@
 work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
-@jn77ima). Versions of this have been implemented in packages `minpack.lm` and `nlsr`.
-and we have prepared experimental versions of `nls` replacements that can incorporate
-stabilizations. Integration of such ideas with other code structures so that all the
-features of `nls()` work properly has unfortunately proved difficult.
-
-Let us consider an example which presents the infamous 'singular gradient' error message.
-This was posted on StackExchange. 
-*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*
+@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
+In our work we prepared experimental prototypes of `nls` that incorporate
+stabilizations, but integration with all the features of `nls()` has proved difficult.
+
+Let us consider an 'singular gradient' example 
+*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
+`nlsr::nlxb()` displays the singular values of the Jacobian, so the `nls()` error message is correct,
+but unhelpful in obtaining a solution.
 
 ```{r SESingGrad}
-reala = -3
-realb = 5
-realc = 0.5
-realr = 0.7
-realm = 1
-x = 1:11 #x values; 11 timepoint data
-#linear+exponential function
-y=reala + realb * realr^(x - realm) + realc * x
-testdat = data.frame(x, y)
-strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1)
+reala = -3; realb = 5; realc = 0.5; realr = 0.7; realm = 1 # underlying parameters
+x = 1:11 # x values; 11 data points
+y=reala + realb * realr^(x - realm) + realc * x # linear+exponential function
+testdat = data.frame(x, y) # save in a data frame
+strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
 jform <- y ~ a + b * r^(x - m) + c * x # Formula
+library(nlsr)
+linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
+linexp2 # Note singular values of Jacobian in rightmost column
+# We get an error with nls()
 linexp=try(nls(jform, data = testdat, start = strt, trace = F))
-library(nlsr)
-# Note singular values of Jacobian in rightmost column
-linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
-linexp2
-```
-
-In the above, we see that the Jacobian is essentially singular as shown by its singular
-values. Note that we DISPLAY the singular values in a single column for convenience in
-the output. The values do NOT correspond to individual parameters, but are a property
-of the whole problem.
+```
+
+The Jacobian is essentially singular as shown by its singular
+values. Note that these are **displayed** in a single column in
+the output to provide a compact layout, but the values do **NOT** correspond to 
+the individual parameters in 
+whose row they appear; they are a property of the whole problem.
 
@@
 we have created a rather more compact version entirely in R.
-Code project. See the companion document **Jacobian Calculations for nls()**
-(?? need citation -- currently **DerivsNLS.pdf**).
+<!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
+<!-- (?? need citation -- currently **DerivsNLS.pdf**). -->
 
 The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
-of `nls()` (and, by extension, `minpack.lm::nlsLM`) will use a numerical approximation. 
+of `nls()` (and, by extension, `minpack.lm::nlsLM`) mostly use a numerical approximation. 
 This can lead to different behaviour:
@@
 
-Note that the selfStart models in the `./src/library/stats/R/zzModels.R` file
+Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file
 provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
@@
 is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
-almost certainly contributes to the performance of `nls()` on selfStart models.
+almost certainly contributes to the good performance of `nls()` on selfStart models.
 
@@
 
-`nls()` accepts an argument `subset`. Unfortunately, this acts through the mediation of
+`nls()` accepts an argument `subset`. This acts through the mediation of
 `model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
@@
 The typical default action, which can be seen by using the command `getOption("na.action")`
-is `na.omit`. This option essentially presents computations with data with all observations
-containing any missing values (i.e. any row of a data frame with an NA) omitted. 
+is `na.omit`. This option essentially omits from computations any observations
+containing missing values (i.e. any row of a data frame containing an NA). 
 `na.exclude` does much the same for computations, but keeps the rows with NA elements so 
@@
 
+Our concern with `na.action` is that users may be unaware of the effects of an option
+setting they may not even be aware has been set. Is `na.fail` a safer default?
+
 ## Feature: model frame
 
-`model` is an argument to the `nls()` function, which is documented
+`model` is an argument to the `nls()` function, which is documented as
 
@@
 be confusing.) However, the model frame is used within the function code in the form of the object
-`mf`.
+`mf`. We feel that users could benefit from more extensive documentation and examples of its use.
 
@@
 need to ensure equivalent behaviour, but users should test that the 
-operation is correct.
+operation is as they intend.
 
@@
 parameters. `nls()` offers a special class of modeling functions called **selfStart** 
-models. There are a number of these in base R (see list below) and others in R packages
-such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as the now-archived package
-`NRAIA`. Unfortunately, the structure of **selfStart** codes
-is such that the methods by which initial parameters are computed is entangled
+models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
+and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
+the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
+entangle the calculation of starting values for parameters
 with the particulars of the `nls()` code. Though there is a `getInitial()` function,
@@
 
-In the example below, we show how the `SSlogis` selfStart function can generate a set
+Other concerns are revealed by the example below. Here the `SSlogis` selfStart function 
+is used to generate a set
 of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
-is $$ y \sim Asym/(1+exp((xmid-tt)/scal)) $$, but we show how the starting parameters
+is $y \sim Asym/(1+exp((xmid-tt)/scal))$, but we show how the starting parameters
 for this model can be transformed to those of 
-another form of the model, namely, $$ y \sim b1/(1 + b2*exp(-b3*t)) $$.
-
-Let us look at the actual code for `SSlogis()` in `R-devel/src/library/stats/R/zzModels.R`:
-
-```
-SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)),
-    selfStart(
-        function(input, Asym, xmid, scal)
-        {
-              .expr1 <- xmid - input
-              .expr3 <- exp(.e2 <- .expr1/scal)
-              .expr4 <- 1 + .expr3
-              .value <- Asym/.expr4
-              .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")])
-              if(all(vapply(.actualArgs, is.name, NA)))
-              {
-            		  .expr10 <- .expr4^2
-                  .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal")))
-                  .grad[, "Asym"] <- 1/.expr4
-		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10)
-		              .grad[, "scal"] <- xm * .e2
-                  dimnames(.grad) <- list(NULL, .actualArgs)
-                  attr(.value, "gradient") <- .grad
-              }
-              .value
-        },
-        initial = function(mCall, data, LHS, ...) {
-              xy <- sortedXyData(mCall[["input"]], LHS, data)
-              if(nrow(xy) < 4) {
-                  stop("too few distinct input values to fit a logistic model")
-              }
-              z <- xy[["y"]]
-              ## transform to proportion, i.e. in (0,1) :
-              rng <- range(z); dz <- diff(rng)
-              z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz)
-              xy[["z"]] <- log(z/(1 - z))		# logit transformation
-              aux <- coef(lm(x ~ z, xy))
-              pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)),
-                               data = xy,
-                               start = list(xmid = aux[[1L]], scal = aux[[2L]]),
-                               algorithm = "plinear", ...))
-              setNames(pars [c(".lin", "xmid", "scal")],
-                       mCall[c("Asym", "xmid", "scal")])
-        },
-        parameters = c("Asym", "xmid", "scal"))
-```
-
+another form of the model, namely, $y \sim b1/(1 + b2*exp(-b3*t))$.
+
+The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
+<!-- ``` -->
+<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
+<!--     selfStart( -->
+<!--         function(input, Asym, xmid, scal) -->
+<!--         { -->
+<!--               .expr1 <- xmid - input -->
+<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
+<!--               .expr4 <- 1 + .expr3 -->
+<!--               .value <- Asym/.expr4 -->
+<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
+<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
+<!--               { -->
+<!--             		  .expr10 <- .expr4^2 -->
+<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
+<!--                   .grad[, "Asym"] <- 1/.expr4 -->
+<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
+<!-- 		              .grad[, "scal"] <- xm * .e2 -->
+<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
+<!--                   attr(.value, "gradient") <- .grad -->
+<!--               } -->
+<!--               .value -->
+<!--         }, -->
+<!--         initial = function(mCall, data, LHS, ...) { -->
+<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
+<!--               if(nrow(xy) < 4) { -->
+<!--                   stop("too few distinct input values to fit a logistic model") -->
+<!--               } -->
+<!--               z <- xy[["y"]] -->
+<!--               ## transform to proportion, i.e. in (0,1) : -->
+<!--               rng <- range(z); dz <- diff(rng) -->
+<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
+<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
+<!--               aux <- coef(lm(x ~ z, xy)) -->
+<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
+<!--                                data = xy, -->
+<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
+<!--                                algorithm = "plinear", ...)) -->
+<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
+<!--                        mCall[c("Asym", "xmid", "scal")]) -->
+<!--         }, -->
+<!--         parameters = c("Asym", "xmid", "scal")) -->
+<!-- ``` -->
 This R function includes analytic expressions for the Jacobian ("gradient").
@@
 In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
-to 1. This is, in our view, tolerable, but could possibly be improved by using a set of values
-that are slightly different e.g., in the case of a model $$  y \sim a*exp(-b*x) + c*exp(-d*x)$$
-it would be useful to have $b$ and $d$ values different so the Jacobian is not singular. Thus,
-some sort of sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters might be better and it
-can be provided quite simply instead of all 1's.
-
-```{r hobbsSSlogis}
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-
-NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt))
-NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal)
-NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal))
-NLStestdata <- data.frame(y=weed, tt=tt) 
-s0 <- getInitial(NLSformula, NLStestdata)
-print(s0)
-# We transform the parameters for the NLSformula0 model of original specification.
-s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3])
-print(as.numeric(s1))
-# No actual improvement because nls() has been already used to get the starting values,
-# but we do get SEs
-hobblog<-nls(NLSformula0, data=NLStestdata, start=s1)
-summary(hobblog)
-deviance(hobblog)
-# nls fails without selfStart -- singular gradient, even on stabilized formula
-try(hobblogx<-nls(NLSformulax, data=NLStestdata))
-# But Marquardt is able to get a solution easily
-library(nlsr)
-hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1))
-hobblogxx
-```
-
-### selfStart models in base R
-
-The following models are provided (in file ./src/library/stats/R/zzModels.R)
-
-```
-SSasymp         - asymptotic regression model
-SSasympOff      - alternate formulation of asymptotic regression model with offset
-SSasympOrig     - exponential curve through the origin to an asymptote
-SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input)
-SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) -
-                          exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe)))
-SSfpl           - four parameter logistic model
-SSlogis         - three parameter logistic model
-SSmicmen        - Michaelis-Menten model for enzyme kinetics
-SSgompertz2     - Gompertz model for growth curve data
-SSweibull       - Weibull model for growth curve data
-```
+to 1. This is, in our view, tolerable, but could be improved by using a set of values
+that are all slightly different, which, in the case of the example model $y \sim a*exp(-b*x) + c*exp(-d*x)$
+would avoid a singular Jacobiam because $b$ and $d$ were equal in value. 
+A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
+could be provided quite simply instead of all 1's.
+
+<!-- ```{r hobbsSSlogis} -->
+<!-- weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, -->
+<!--           38.558, 50.156, 62.948, 75.995, 91.972) -->
+<!-- tt <- 1:12 -->
+
+<!-- NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt)) -->
+<!-- NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal) -->
+<!-- NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal)) -->
+<!-- NLStestdata <- data.frame(y=weed, tt=tt)  -->
+<!-- s0 <- getInitial(NLSformula, NLStestdata) -->
+<!-- print(s0) -->
+<!-- # We transform the parameters for the NLSformula0 model of original specification. -->
+<!-- s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3]) -->
+<!-- print(as.numeric(s1)) -->
+<!-- # No actual improvement because nls() has been already used to get the starting values, -->
+<!-- # but we do get SEs -->
+<!-- hobblog<-nls(NLSformula0, data=NLStestdata, start=s1) -->
+<!-- summary(hobblog) -->
+<!-- deviance(hobblog) -->
+<!-- # nls fails without selfStart -- singular gradient, even on stabilized formula -->
+<!-- try(hobblogx<-nls(NLSformulax, data=NLStestdata)) -->
+<!-- # But Marquardt is able to get a solution easily -->
+<!-- library(nlsr) -->
+<!-- hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1)) -->
+<!-- hobblogxx -->
+<!-- ``` -->
+
+<!-- ### selfStart models in base R -->
+
+<!-- The following models are provided (in file ./src/library/stats/R/zzModels.R) -->
+
+<!-- ``` -->
+<!-- SSasymp         - asymptotic regression model -->
+<!-- SSasympOff      - alternate formulation of asymptotic regression model with offset -->
+<!-- SSasympOrig     - exponential curve through the origin to an asymptote -->
+<!-- SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input) -->
+<!-- SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) - -->
+<!--                           exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe))) -->
+<!-- SSfpl           - four parameter logistic model -->
+<!-- SSlogis         - three parameter logistic model -->
+<!-- SSmicmen        - Michaelis-Menten model for enzyme kinetics -->
+<!-- SSgompertz2     - Gompertz model for growth curve data -->
+<!-- SSweibull       - Weibull model for growth curve data -->
+<!-- ``` -->
 
@@
 
-Because the Gauss-Newton algorithm is unreliable from many starting sets
-of parameters, selfStart models are more than an accessory to `nls()` but a part
-of the infrastructure. However, creating such functions is a lot of work, and their
+Because the Gauss-Newton algorithm is unreliable from many sets of starting
+of parameters, selfStart models are a part of the `nls()` infrastructure rather
+than an accessory. However, creating such functions is a lot of work, and their
 documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
-We believe that the focus would better be placed on getting good initial parameters,
-possibly with some interactive tools. That is, the emphasis should be on the 
-`getInitial()` function, though avoiding the current calls back to `nls()`.
-
-## Issue: documentation of the results of running nls()
+We believe that the focus should be placed on getting good initial parameters,
+that is `getInitial()` function, though avoiding the current calls back to `nls()`.
+Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
+worth considering.
+
+## Issue: returned output of nls() and its documentation
 
 The output of `nls()` is an object of class "nls" which has the following structure:
-
-### nls() result output according to the documentation
 
@@
 
-### Example output
-
-To illustrate, let us run the Croucher example.
-
-```{r nlsoutx}
-# Croucher example
-xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9)
-ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001)
-p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2)
-NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata)
-NLSdata<-data.frame(xdata, ydata)
-# Try full output version of nls
-library(nlspkg) # use the packaged version of nls()
-result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE)
-# We can display the result with several commands
-# str(result) -- displays large amount of material - suppressed here
-#    as it is too wide for the page
-# result # displays the object
-# ls(result) # to list the elements of the output
-# ls(result$m) # and in particular the "m" object
-```
-
-### Concerns with content of the nls result object
-
-The nls object contains some elements that are awkward to produce by other algorithms,
-but some information that would be useful is not presented obviously.
+<!-- ### Example output -->
+
+<!-- To illustrate, let us run the Croucher example. -->
+
+<!-- ```{r nlsoutx} -->
+<!-- # Croucher example -->
+<!-- xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9) -->
+<!-- ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001) -->
+<!-- p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2) -->
+<!-- NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) -->
+<!-- NLSdata<-data.frame(xdata, ydata) -->
+<!-- # Try full output version of nls -->
+<!-- library(nlspkg) # use the packaged version of nls() -->
+<!-- result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE) -->
+<!-- # We can display the result with several commands -->
+<!-- # str(result) -- displays large amount of material - suppressed here -->
+<!-- #    as it is too wide for the page -->
+<!-- # result # displays the object -->
+<!-- # ls(result) # to list the elements of the output -->
+<!-- # ls(result$m) # and in particular the "m" object -->
+<!-- ``` -->
+
+<!-- ### Concerns with content of the nls result object -->
+
+The `nls` object contains some elements that are awkward to produce by other algorithms,
+but some information that would be useful is not presented obviously. Moreover, the
+complexity of the object is a challenge to users.
 
@@
 
-### Information that is NOT in the nls result object
-
-`nlsr::nlxb()` solves ostensibly the same problem as `nls()` but only claims to return
+Let us compare the `nls()` result with that from 
+`nlsr::nlxb()`, with ostensibly solves the same problem:
 
@@
 coefficients    A named vector giving the parameter values at the supposed solution.
-
 ssquares        The sum of squared residuals at this set of parameters.
-
 resid           The residual vector at the returned parameters.
-
 jacobian        The jacobian matrix (partial derivatives of residuals w.r.t. the 
                 parameters) at the returned parameters.
-
 feval           The number of residual evaluations (sum of squares computations) used.
-
 jeval           The number of Jacobian evaluations used.
@@
 However, actually looking at the structure of a returned result gives a list of 11
-items:
-```
- $ resid       : num [1:12] 0.0119 -0.0328 0.092 0.2088 0.3926 ...
-  ..- attr(*, "gradient")= num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
-  .. ..- attr(*, "dimnames")=List of 2
-  .. .. ..$ : NULL
-  .. .. ..$ : chr [1:3] "Asym" "xmid" "scal"
- $ jacobian    : num [1:12, 1:3] 0.0271 0.0367 0.0496 0.0666 0.089 ...
-  ..- attr(*, "dimnames")=List of 2
-  .. ..$ : NULL
-  .. ..$ : chr [1:3] "Asym" "xmid" "scal"
- $ feval       : num 31
- $ jeval       : num 23
- $ coefficients: Named num [1:3] 196.19 12.42 3.19
-  ..- attr(*, "names")= chr [1:3] "Asym" "xmid" "scal"
- $ ssquares    : num 2.59
+items. The extra 5 are:
+
+```
  $ lower       : num [1:3] -Inf -Inf -Inf
@@
 ```
-This is still a smaller result object than the one `nls()` returns. Moreover, `nlxb`
-explicitly returns the sum of squares as well as the residual vector and Jacobian. The 
-counts of evaluations are also returned. Working on this project showed several potential
-updates to the `nlsr` documentation. Note that the singular values of the Jacobian are
-actually computed in the `print` and `summary` methods for the result.
+The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
+Moreover, `nlxb`
+explicitly returns the sum of squares as well as the residual vector and Jacobian. 
+The counts of evaluations are also returned.
+(Note that the singular values of the Jacobian are
+actually computed in the `print` and `summary` methods for the result.) 
+We noted several potential updates to the `nlsr` documentation as well as that
+for `nls()` as we made our comparisons.
 
@@
 But if we return `ans` from `nls()` or `minpack.lm::nlsLM` or our new `nlsj` (interim package), 
-then `ans$m$resid()` is WEIGHTED.
+then `ans$m$resid()` is WEIGHTED. This needs clear documentation so users can get the answers
+intended.
 
@@
 function has made considerable progress in reducing the sum of squared residuals. 
-An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package.
-Here is an abbreviated version of that problem and the `nls()` output:
+<!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
+<!-- ?? AB: Are we far enough along to include reference? -->
+Here is an abbreviated example:
 
@@
 unless `trace` is invoked, the user will not get any information about this.
-This would be an almost trivial change to the `nls() function and could be 
+This almost trivial change to the `nls()` function could be 
 useful to R users.
@@
 One obvious case is that nonlinear modeling tools are a superset of linear ones.
-Yet the explicit model
-
-```
-y ~ a*x + b
-```
-
-does not work with the linear modeling function `lm()`, which requires
-this model to be specified as
-
-```
-y ~ x
-```
+Yet the explicit model `y ~ a*x + b` does not work with the linear modeling 
+function `lm()`, which requires this model to be specified as `y ~ x`.
 
 Within `nls()`, consider the 
-following FOUR different calling sequences for the same problem, though the
-fm2a -- an intuitive choice -- will not work. In this failed
+following FOUR different calling sequences for the same problem, though an intuitive choice,
+labelled fm2a, will not work. In this failed
 attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
@@
 ```{r log4ways, echo=TRUE, eval=FALSE}
-DNase1 <- subset(DNase, Run == 1)
-
+DNase1 <- subset(DNase, Run == 1) # select the data
 ## using a selfStart model - do not specify the starting parameters
-fm1DNase1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
-summary(fm1DNase1)
-## the coefficients only:
-# coef(fm1DNase1)
-## including their SE, etc:
-# coef(summary(fm1DNase1))
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
 
 ## using conditional linearity - leave out the Asym parameter
-fm2DNase1 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
-                 data = DNase1,
-                 start = list(xmid = 0, scal = 1),
+fm2 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
+                 data = DNase1, start = list(xmid = 0, scal = 1),
                  algorithm = "plinear")
-summary(fm2DNase1)
-
-## using conditional linearity AND Asym does NOT work
-
-fm2aDNase1 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
-                 data = DNase1,
-                 start = list(Asym=3, xmid = 0, scal = 1),
-                 algorithm = "plinear",
-                 trace = TRUE))
-summary(fm2aDNase1)
+summary(fm2)
 
 ## without conditional linearity
-fm3DNase1 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+fm3 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
                  data = DNase1,
                  start = list(Asym = 3, xmid = 0, scal = 1))
-summary(fm3DNase1)
+summary(fm3)
 
 ## using Port's nl2sol algorithm
-fm4DNase1 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
-                 data = DNase1,
-                 start = list(Asym = 3, xmid = 0, scal = 1),
+fm4 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
                  algorithm = "port"))
-summary(fm4DNase1)
+summary(fm4)
+
+## using conditional linearity AND Asym does NOT work
+fm2a <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
+                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                 algorithm = "plinear", trace = TRUE))
+summary(fm2a)
 ```
@@
     ...
-
-    I don't think I will be of much help.  My R skills have atrophied to the point 
-    where I wouldn't even know how to start exploring what is happening in the first 
-    call as opposed to the recursive call.
-```
+```
+<!--     I don't think I will be of much help.  My R skills have atrophied to the point  -->
+<!--     where I wouldn't even know how to start exploring what is happening in the first  -->
+<!--     call as opposed to the recursive call. -->
+<!-- ``` -->
 
@@
 parameters are indexed. It also uses the "plinear" option as an added complication.
-Here we use a truncated version of the example to save display space.
-
-```{r nlsindx1}
-## The muscle dataset in MASS is from an experiment on muscle
-## contraction on 21 animals.  The observed variables are Strip
-## (identifier of muscle), Conc (Cacl concentration) and Length
-## (resulting length of muscle section).
-if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
-mm<- MASS::muscle[1:12,] # take only 1st few values of Strip
-str(mm)
-mm<-droplevels(mm)
-str(mm)
-nlev <- nlevels(mm)
-withAutoprint({
-  ## The non linear model considered is
-  ##       Length = alpha + beta*exp(-Conc/theta) + error
-  ## where theta is constant but alpha and beta may vary with Strip.
-  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
-  nl <- nlevels(mm$Strip)
-  ## We first use the plinear algorithm to fit an overall model,
-  ## ignoring that alpha and beta might vary with Strip.
-  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
-                start = list(th = 1), algorithm = "plinear")
-  summary(musc.1)
-
-  ## Then we use nls' indexing feature for parameters in non-linear
-  ## models to use the conventional algorithm to fit a model in which
-  ## alpha and beta vary with Strip.  The starting values are provided
-  ## by the previously fitted model.
-  ## Note that with indexed parameters, the starting values must be
-  ## given in a list (with names):
-  ## ?? but why use b here AND in the new formula??
-  b <- coef(musc.1)
-
-  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
-                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
-  summary(musc.2)
-})
-
-```
-
-Note that the answers for the parameters are NOT indexed. e.g. `coef(musc.2)` is a single level
-vector of parameters. We do not see `a[1], a[2], a[3]` but `a1, a2, a3`. This is because the model
-must integrate all the parameters because `th` is a common parameter across the index `Strip`.
-
-We believe this structure is quite likely to cause confusion and error, and propose an alternative
-approach below.
+
+<!-- Here we use a truncated version of the example to save display space. -->
+
+<!-- ```{r nlsindx1} -->
+<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
+<!-- ## contraction on 21 animals.  The observed variables are Strip -->
+<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
+<!-- ## (resulting length of muscle section). -->
+<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
+<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip -->
+<!-- str(mm) -->
+<!-- mm<-droplevels(mm) -->
+<!-- str(mm) -->
+<!-- nlev <- nlevels(mm) -->
+<!-- withAutoprint({ -->
+<!--   ## The non linear model considered is -->
+<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
+<!--   ## where theta is constant but alpha and beta may vary with Strip. -->
+<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
+<!--   nl <- nlevels(mm$Strip) -->
+<!--   ## We first use the plinear algorithm to fit an overall model, -->
+<!--   ## ignoring that alpha and beta might vary with Strip. -->
+<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
+<!--                 start = list(th = 1), algorithm = "plinear") -->
+<!--   summary(musc.1) -->
+
+<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
+<!--   ## models to use the conventional algorithm to fit a model in which -->
+<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
+<!--   ## by the previously fitted model. -->
+<!--   ## Note that with indexed parameters, the starting values must be -->
+<!--   ## given in a list (with names): -->
+<!--   ## ?? but why use b here AND in the new formula?? -->
+<!--   b <- coef(musc.1) -->
+
+<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
+<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
+<!--   summary(musc.2) -->
+<!-- }) -->
+
+<!-- ``` -->
+
+Running the example reveals that the answers for the parameters are NOT indexed. 
+<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
+We do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
+<!-- This is because the model -->
+<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
+This is no doubt because programming for indexed parameters is extremely challenging.
+However, we believe the current structure is could cause confusion and error, and 
+propose an alternative approach below.
 
@@
 There are many situations where the context of a problem constrains the values of parameters.
-For example, one of us was asked many years ago to estimate a model. One paramater returned
-a negative value. The client complained "But that is supposed to be the number of grain elevators
-in Saskatchewan". That is, the number should have been a positive integer, and likely less
+For example, one of us was asked many years ago to estimate a model where one parameter estimate
+was negative. The client complained "But that is supposed to be the number of grain elevators
+in Saskatchewan". The number should have been a positive integer, and likely less
 than a few thousand.
@@
 
-> The algorithm = "port" code appears unfinished, and does not even check that the starting value is within the bounds. Use with caution, especially where bounds are supplied.
+> The algorithm = "port" code appears unfinished, and does not even check that the starting 
+  value is within the bounds. Use with caution, especially where bounds are supplied.
 
@@
 Hans Werner Borchers has been helpful in developing the proposal for this project
-and in comments on this and related work.
-
-??Heather Turner ... (co-mentor ?? how to word?)
+and in comments on this and related work. Heather Turner co-mentored the project 
+and helped guide the progress of the work.??
 

### 2022-06-15 RefactoringNLS/bhobbsX.R

@@ -1,0 +1,92 @@
+# bhobbsX.R
+#### bounds with formula specification of problem
+## Use the Hobbs Weed problem
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weedframe <- data.frame(y=weed, tt=tt)
+st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
+## Unscaled model
+wmodu <- y ~ b1/(1+b2*exp(-b3*tt))
+## Scaled model
+wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt))
+## We can provide the residual and Jacobian as functions
+# Scaled Hobbs problem
+shobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("shobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  100.0*x[1]/(1+x[2]*10.*exp(-0.1*x[3]*tt)) - y
+}
+
+shobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-0.1*x[3]*tt)
+  zz  <-  100.0/(1+10.*x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -0.1*x[1]*zz*zz*yy
+  jj[tt,3]   <-   0.01*x[1]*zz*zz*yy*x[2]*tt
+  attr(jj, "gradient") <- jj
+  jj
+}
+
+##----- Hobbsbounded unique code starts here ------
+require(minpack.lm)
+traceval<-FALSE
+cat("Infeasible start test\n")
+start1inf <- c(b1=4, b2=4, b3=4) # b3 OUT OF BOUNDS for next few tries
+## Infeasible start! No warning message!
+anlM2i <- try(nlsLM(wmodu, start=start1inf, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
+                    trace=traceval))
+
+# feasible start i.e. on or within bounds
+start1<-st
[…]

### 2022-06-15 RefactoringNLS/RefactoringNLS.Rmd
Pruned!

@@
    - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
-date: "2022-5-29"
+date: "2022-6-15"
 output: 
@@
 we have created a rather more compact version entirely in R. 
+See https://github.com/nashjc/RNonlinearLS/DerivsNLS .
 <!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
@@
 
-This is clearly unsatisfactory for general use. Inclusion of bounds on parameters in nonlinear
-least squares computations is relatively straightforward, and it is part of the default method
+This is clearly unsatisfactory for general use, but does find bounded solutions. 
+Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
+straightforward, and it is part of the default method
 for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
-script, which we do not run here for reasons of space, shows that the solutions found are not
-optimal for the Hobbs test problem, even in its scaled form.
-
-```{r chkbound, eval=FALSE}
-#### bounds with formula specification of problem
-## Use the Hobbs Weed problem
+script, which we do not run here for reasons of space, shows that it does not find a
+good solution for the scaled Hobbs test problem.
+
+```{r chkbound}
+# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
 weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
@@
 tt <- 1:12
-weedframe <- data.frame(y=weed, tt=tt)
+wf <- data.frame(y=weed, tt=tt)
 st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
-## Unscaled model
-wmodu <- y ~ b1/(1+b2*exp(-b3*tt))
-## Scaled model
-wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt))
-## We can provide the residual and Jacobian as functions
-# Scaled Hobbs problem
-shobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
-  # This variant uses looping
-  if(length(x) != 3) stop("shobbs.res -- parameter vector n!=3")
-  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
-           38.558, 50.156, 62.948, 75.995, 91.972)
-  tt  <-  1:12
-  res  <-  100.0*x[1]/(1+x[2]*10.*exp(-0.1*x[3]*tt)) - y
-}
-
-shobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
-  jj  <-  matrix(0.0, 12, 3)
-  tt  <-  1:12
-  yy  <-  exp(-0.1*x[3]*tt)
-  zz  <-  100.0/(1+10.*x[2]*yy)
-  jj[tt,1]   <-   zz
-  jj[tt,2]   <-   -0.1*x[1]*zz*zz*yy
-  jj[tt,3]   <-   0.01*x[1]*zz*zz*yy*x[2]*tt
-  attr(jj, "gradient") <- jj
-  jj
-}
-
-##----- Hobbsbounded unique code starts here ------
+wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt)) ## Scaled model
 require(minpack.lm)
-traceval<-FALSE
-cat("Infeasible start test\n")
-start1inf <- c(b1=4, b2=4, b3=4) # b3 OUT OF BOUNDS for next few tries
-## Infeasible start! No warning message!
-anlM2i <- try(nlsLM(wmodu, start=start1inf, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
-                    trace=traceval))
-
-# feasible start i.e. on or within bounds
-start1<-st
-anlM1 <- try(nlsLM(wmodu, start=start1, data=weedframe, lower=c(0,0,0), upper=c(2, 6, 3), 
-                    trace=traceval))
-print(anlM1)
-
+require(nlsr)
 # Hobbs scaled problem with bounds, formula specification
-
-require(nlsr)
-anlxb1 <- nlxb(wmods, start=start1, data=weedframe, lower=c(0,0,0),
-                  upper=c(2,6,3), trace=traceval)
+anlxb1 <- nlxb(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+cat("Using nlsr::nlxb():")
 print(anlxb1)
-
 ## nlsLM seems NOT to work properly with bounds
-anlsLM1b <- nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
-                 upper=c(2,6,3), trace=traceval)
+anlsLM1b <- nlsLM(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+cat("\n"); cat("using minpack.lm::nlsLM():")
 print(anlsLM1b)
-newst<-coef(anlsLM1b)
-print(newst)
-anlsLM1bx <- nlsLM(wmods, start=newst, data=weedframe, lower=c(0,0,0),
-                   upper=c(2,6,3))
-print(anlsLM1bx)
-
-cat("Single number bounds:\n")
-anlsLM1b1 <- try(nlsLM(wmods, start=start1, data=weedframe, lower=0,
-                   upper=3))
-print(anlsLM1b1)
-anlsLM1b1a <- try(nlsLM(wmods, start=start1, data=weedframe, lower=c(0,0,0),
-                   upper=c(3,3,3)))
-print(anlsLM1b1a)
-
-anlxb1b1<- try(nlxb(wmods, start=start1, data=weedframe, lower=0,
-                     upper=3))
-print(anlxb1b1)
-
-anlm1b <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
-                  upper=c(2,6,3))
-print(anlm1b)
-
-# functional presentation of problem  -- seems to work 
-anlm1bx <- nls.lm(par=start1, fn=shobbs.res, jac=shobbs.jac, lower=c(0,0,0),
-                   upper=c(2,6,3))
-print(anlm1bx)
-
-anlfb1bx <- nlfb(start=start1, resfn=shobbs.res, jacfn=shobbs.jac, lower=c(0,0,0),
-                  upper=c(2,6,3))
-print(anlfb1bx)
+anlsb<-nls(wmods, start=st, data=wf, algorithm="port", lower=c(0,0,0), upper=c(2,6,3))
+cat("Using nls with 'port':")
+print(anlsb)
 ```
@@
 the parameter may be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions drastically. Hence `nlsr::nlxb()` does not compute standard
+Bounds change those assumptions. Hence `nlsr::nlxb()` does not compute standard
 errors nor their derived statistics when bounds are active.
 
-# Goals of our effort to improve `nls()`
+# Goals and progress of our effort to improve `nls()`
 
@@
 - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
-  that matches the version in base R or what is packaged in `nlspkg` as described 
+  that matches the features in base R or what is packaged in `nlspkg` as described 
   in @PkgFromRbase22.
@@
 - to streamline the overall `nls()` infrastructure. By this we
-  mean a re-factoring of the routines so they are better suited to maintenance of both
+  mean a re-factoring the routines so they are better suited to maintenance of both
   the existing `nls()` methods and features as well as new features we or others
@@
 
+These goals echo themes in the recent discussion @Vidoni21 and accompanying 
+commentary.
+
 ### Rationalization of formula specifications
@@
 `plinear` algorithm is used. This is unfortunate, since the user cannot then simply
-add `algorithm="plinear"` to the call. This makes errors more
-likely.
-
-While not consistent with the `lm()` function and specification of linear models, 
-we believe a complete set of parameters should be specified for nonlinear models.
-That is, the model formula should be written down as it would be computed, 
-rather than for estimation. Thus `y ~ a*x + b` rather than `y ~ x`.
-
-Moreover, we want to avoid the partially linear parameters appearing in the result object
-as `.lin1`, `.lin2`, etc., where there actual position in the model is not explicit.
-
-A possible workaround to allow consistency would be to specify the partially linear parameters
-when the `plinear` option is specified. For example, we could use `algorithm="plinear(Asym)"`
-while keeping the general model formula from the default specification. This would allow 
+add `algorithm="plinear"` to the call, making errors more likely.
+
+While not consistent with the `lm()` specification of linear models, 
+nonlinear models need all parameters specified. 
+That is, the model formula should be written down as it would be computed.
+Thus a linear model (which is valid input for nonlinear estimatin) should
+be specified `y ~ a*x + b` rather than `y ~ x`. 
+Moreover, we want to avoid the partially linear parameters appearing 
+in the result object as `.lin1`, `.lin2`, etc., where the actual position 
+of parameters in the model is not explicit.
+
+A possible workaround to allow consistency would be to specify the 
+partially linear parameters when the `plinear` option is specified. For example, 
+we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
+from the default specification. This would allow 
 for the output of different `algorithm` options to be consistent. However, we have not
-yet tried to code such a change.
+yet tried to code such a change, and welcome discussion from those working with
+other packages using model formulas.
 
@@
 the user to surround a term with `I(  )` to inhibit interpretation of arithmetic 
-operators. We would prefer that formulas be usable without this complication, which
-has limited documentation and examples.
-
-It is also of interest to consider code that will detect partial linearity, 
-rather than expecting the user to provide the special structure and settings. Work in this
-area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution
-we can find.
-Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin 
-is about a class of models that have almost nothing in common
-with the present issue. The algorithm in @Birkisson2013 will need investigating further,
-but we suspect that it is complicated enough that it will not easily be adapted to 
-a program like `nls()`. Possibly a simpler algorithm may be found.
+operators. We would prefer that formulas be usable without this complication.
+
+<!-- It is also of interest to consider code that will detect partial linearity,  -->
+<!-- rather than expecting the user to provide the special structure and settings. Work in this -->
+<!-- area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution -->
+<!-- we can find. -->
+<!-- Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin  -->
+<!-- is about a class of models that have almost nothing in common -->
+<!-- with the present issue. The algorithm in @Birkisson2013 will need investigating further, -->
+<!-- but we suspect that it is complicated enough that it will not easily be adapted to  -->
+<!-- a program like `nls()`. Possibly a simpler algorithm may be found. -->
 
@@
 Indexed models clearly have a place in some areas of research. 
-We need to be able to process models, as the example above, where the
-model formula is 
-`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` and `Strip` is the index. Data
-`Length` and `Conc` are available for all observations. Parameter `th` applies
+<!-- We need to be able to process models, as the example above, where the -->
+<!-- model formula is  -->
+We need to be able to process models such as 
+`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` where `Strip` is the index. Data
+`Length` and `Conc` are available for all observations and parameter `th` applies
 to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
-could be an experimental run, or some other categorization.
+could be an experimental run number, or some other categorization.
 
@@
 
-While tedious (??maybe dummies or fastdummies packages make it less so?), 
-we could set up the model as a sum of similar models with dummy
-variables to turn "on" or "off" the particular case, transforming the model
-parameters to simple, non-indexed ones (as does `nls()`).
-
-Wrapper functions to prepare the estimation calculations and present the
-results in a form that matches the modelling formula would avoid adding
-complexity to `nls()` and other tools.
-
-## Provide tests and use-case examples
-
-We need suitable tests and use-case examples in order:
-
-- to ensure our new `nlsalt` or related packages work properly, in particular, giving
-  results comparable to or better than the `nls()` in base R or `nlspkg`;
+<!-- While tedious (??maybe dummies or fastdummies packages make it less so?),  -->
+<!-- we could set up the model as a sum of similar models with dummy -->
+<!-- variables to turn "on" or "off" the particular case, transforming the model -->
+<!-- parameters to simple, non-indexed ones (as does `nls()`). -->
+
+Our view is that inclusion of the indexing **within** `nls()` introduces 
+complexity that could be avoided by a suitable wrapper function. The modeling
+formula must be expanded to encompass all the index cases, likely with indexed
+parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
+we believe the wrapper should be able to process output to present the indexed
+parameters correctly.
+
+### Streamlining code
+
+We are pessimistic that the overall structure of `nls()` can be streamlined due
+to the entanglement of so many features with the complicated mix of R, C and
+Fortran. Indeed, despite digging into the code over some months, we do not feel
+confident that we sufficiently understand it nor that we could maintain it. On the other
+hand, we do believe equivalent functionality can be built, but with at least
+some differences in how the features will be accessed.
+
+## Tests and use-case examples
+
+Maintainers of packages need suitable tests and use-case examples in order:
+
+- to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace;
 - to test individual solver functions to ensure they work across the range of calling 
   mechanisms, that is, different ways of supplying inputs to the solver(s);
-- to pose "silly" inputs to nonlinear least squares solvers (in R) to see if 
-  these bad input exceptions are caught by the programs.
-
-### A test runner program ?? NLSCompare
-
-?? Arkajyoti -- do you want to expand?
-
-When we have a "new" or trial solver function, we would like to know if it gives 
-acceptable results on a range of sample problems of different types, starting
-parameters, input conditions, constraints, subsets, weights or other settings. 
-Ideally we want to be able to get a summary that is easy to read and assess. For
-example, one approach would be to list the names of a set of tests with a red, green
-or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last
-category would be a problem with constraints that the solver is not designed to 
-handle.
-
-To accomplish this, we need a suitable "runner" program that can be supplied with
-the name of a solver or solvers and a list of test problem cases. Problems generally
-have a base setup -- a specification of the function to fit as a formula, some data
-and a default starting set of parameters. Other cases can be created by imposing
-bounds or mask constraints, subsets of the data, and different starts.
-
-How to set up this "runner" and its supporting infrastructure is non-trivial. While
-the pieces are not as complicated as the inter-related parts of the solvers, especially
-`nls()`, the categorization of tests, their documentation, and the structuring to make
-running them straightforward and easy requires much attention to detail.
-
-Some considerations for our test scripts:
-
-- Is it useful to have a "base" script for each family of test problem, with numbered particular
-  cases? That is, if we run the scripts in order, we can avoid some duplication of code and
-  data.
-
-- While we have developed some tags to document the test problem families and cases, we believe
-  that such tags (essentially summary documentation) will continue to need revision as different
-  tools and problems are included in scope of `nlsCompare`.
-
-- Similarly, we expect that there will be ongoing review of the structure of the result files.
-
-# Progress so far
-
-Much of our work is reported in https://github.com/nashjc/RNonlinearLS .
-
-## Reports and articles
-
-- The present article is the central report of the Google Summer or Code 2021 project
-  "Improvements to nls()".
-
-- VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and
-  the many choices in their implementation for nonlinear least squares. This is still
-  a DRAFT at 2021-8-20. ?? fix
-
-- TestsDoc.Rmd (??citation) is a survey of testing tools in R. It has more 
-  general possibilities 
-  and fits into the subject of regression testing, in which case a 
-  more extensive literature review will be needed.  Note that this
-  document reflects the work in the the "Problem sets and test infrastructure"
-  below.
+- to pose "silly" inputs to see if 
+  these bad inputs are caught by the programs.
+
+Such goals align with the aims of **unit testing** 
+(e.g., https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211), as well
+as with R package checking tools ?? reference??. ??Other refs.
+
+<!-- ??AB: perhaps we should put in the refernces?? -->
+<!-- I've emailed author to ask if there is preferred citation. JN -->
+
+### A testing package: NLSCompare
+
+One of us has developed a working prototype package at https://github.com/ArkaB-DS/nlsCompare .
+A primary design objective of this is to allow the summarisation of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
+
+<!-- ?? Arkajyoti -- do you want to expand? -->
+<!-- Trying to trim a lot of words, so am makign reference to package. Is it ready -->
+<!-- for others to use? -->
+
+<!-- When we have a "new" or trial solver function, we would like to know if it gives  -->
+<!-- acceptable results on a range of sample problems of different types, starting -->
+<!-- parameters, input conditions, constraints, subsets, weights or other settings.  -->
+<!-- Ideally we want to be able to get a summary that is easy to read and assess. For -->
+<!-- example, one approach would be to list the names of a set of tests with a red, green -->
+<!-- or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last -->
+<!-- category would be a problem with constraints that the solver is not designed to  -->
+<!-- handle.  -->
+
+<!-- To accomplish this, we need a suitable "runner" program that can be supplied with -->
+<!-- the name of a solver or solvers and a list of test problem cases. Problems generally -->
+<!-- have a base setup -- a specification of the function to fit as a formula, some data -->
+<!-- and a default starting set of parameters. Other cases can be created by imposing -->
+<!-- bounds or mask constraints, subsets of the data, and different starts.  -->
+
+<!-- How to set up this "runner" and its supporting infrastructure is non-trivial. While -->
+<!-- the pieces are not as complicated as the inter-related parts of the solvers, especially -->
+<!-- `nls()`, the categorization of tests, their documentation, and the structuring to make -->
+<!-- running them straightforward and easy requires much attention to detail. -->
+
+<!-- Some considerations for our test scripts: -->
+
+<!-- - Is it useful to have a "base" script for each family of test problem, with numbered particular -->
+<!--   cases? That is, if we run the scripts in order, we can avoid some duplication of code and -->
+<!--   data. -->
+
+<!-- - While we have developed some tags to document the test problem families and cases, we believe -->
+<!--   that such tags (essentially summary documentation) will continue to need revision as different -->
+<!--   tools and problems are included in scope of `nlsCompare`. -->
+
+<!-- - Similarly, we expect that there will be ongoing review of the structure of the result files. -->
+
+<!-- # Progress so far -->
+
+<!-- Much of our work is reported in https://github.com/nashjc/RNonlinearLS . -->
+
+<!-- ## Reports and articles -->
+
+<!-- - The present article is the central report of the Google Summer or Code 2021 project -->
+<!--   "Improvements to nls()". -->
+
+<!-- - VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and -->
+<!--   the many choices in their implementation for nonlinear least squares. This is still -->
+<!--   a DRAFT at 2021-8-20. ?? fix -->
+
+<!-- ?? Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
+<!-- useful. Perhaps incorporate in the nlsCompare repo. -->
+<!-- - TestsDoc.Rmd (??citation) is a survey of testing tools in R. It has more  -->
+<!--   general possibilities  -->
+<!--   and fits into the subject of regression testing, in which case a  -->
+<!--   more extensive literature review will be needed.  Note that this -->
+<!--   document reflects the work in the the "Problem sets and test infrastructure" -->
+<!--   below. -->
 
@@
 
-- DerivsNLS.Rmd ??citation: a document to explain different ways in which Jacobian information
-  is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is
-  a DRAFT of a script to provide examples.
-
-- The BibTex bibliography ImproveNLS.bib (??citation or reference) was set up 
-  for use with all documents in this project, but has wider application to nonlinear 
-  least squares projects in general.
-
-- MachineSummary.Rmd (??citation -- blog post??) is an investigation of ways to report 
-  the characteristics
-  and identity of machines running tests. `MachID.R` offers a suggested concise summary function
-  to identify a particular computational system used for tests.
+<!-- ?? needs more focus of where it may appear. -->
+<!-- - DerivsNLS.Rmd ??citation: a document to explain different ways in which Jacobian information -->
+<!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->
+<!--   a DRAFT of a script to provide examples. -->
+
+## Documentation and resources
+
+In our investigation, we built several resources, which are now part of 
+the repository https://github.com/nashjc/RNonlinearLS/. Particular items
+are:
+
+- A BibTex bibliography for use with all documents in this project, but which 
+  has wider application to nonlinear least squares projects in general
+  (https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib)
+
+- `MachID.R` offers a suggested concise summary function
+  to identify a particular computational system used for tests. A discussion 
+  of how it was built and the resources needed across platforms is given in
+  at https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary
 
 - @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
-  code in R-base.
+  `nls()` code in R-base.
 
@@
 
-## Problem sets and test infrastructure
-
-?? need to clean up
-
-We have several test problems and variants thereof in the `inst/scripts/` directory of
-the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare).
-We direct the reader to that package for documentation of the test infrastructure, 
-in particular the problems and methods files (`problems.csv` and `methods.csv`) and
-the various functions invoked by `run.R` to produce an output file in CSV form also.
-
-Towards the end of the project, we have focused our attention on the `nlsCompare`
-package, which looks at evaluating and comparing functions for nonlinear least
-squares problems. We use many of the same tests as checks that are or will be 
-built into our packages for such problems e.g., `nlsalt`. These use the `testthat`
-structure and may include verification of outputs that are specific to a package,
-such as the upper triangular matrix R of the QR decomposition that has been 
-computed for a Jacobian. Since such an object will not be computed by all methods, 
-testing it in `nlsCompare` makes no sense, and in that package we concentrate on
-the minimum sum of squares and the associated model parameters.
-
-## Code and documentation
-
-A summary of our main results:
-
-- nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch
-
-- nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED. 
-  The effort showed that the structure of the programs was difficult to follow, 
-  undocumented, and unsuited to adding improvements like the Marquardt stabilization. 
-  We were able to get a pure-R version of `numericDeriv()` and rework most of the functions 
-  of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project
-  formally ends, but collaboration is likely needed with workers who have a deep knowledge
-  of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`.
-
-<!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  -->
-<!--   fix by JN to the relative offset convergence criterion failure when there are small  -->
-<!--   residuals in problems sent to ``nls()`. Now part of R. -->
-
-- the safeguarded relative offset convergence criterion that allows small-residual problems
-  to ba handled with `nls()` is now in the R base code.
-
-- `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features??
-  This is an interim package for exploration of ideas and will NOT, as it currently
-  is presented, become a distributed package.
-
-- nlsralt: this is intended to be a modified version of Nash and Murdoch package 
-  `nlsr` to examine possible improvements discovered as a result of this project. 
-  This is EXPERIMENTAL.
+<!-- ?? does not seem to add enough? -->
+<!-- ## Problem sets and test infrastructure -->
+
+<!-- ?? need to clean up -->
+
+<!-- We have several test problems and variants thereof in the `inst/scripts/` directory of -->
+<!-- the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare). -->
+<!-- We direct the reader to that package for documentation of the test infrastructure,  -->
+<!-- in particular the problems and methods files (`problems.csv` and `methods.csv`) and -->
+<!-- the various functions invoked by `run.R` to produce an output file in CSV form also. -->
+
+<!-- Towards the end of the project, we have focused our attention on the `nlsCompare` -->
+<!-- package, which looks at evaluating and comparing functions for nonlinear least -->
+<!-- squares problems. We use many of the same tests as checks that are or will be  -->
+<!-- built into our packages for such problems e.g., `nlsalt`. These use the `testthat` -->
+<!-- structure and may include verification of outputs that are specific to a package, -->
+<!-- such as the upper triangular matrix R of the QR decomposition that has been  -->
+<!-- computed for a Jacobian. Since such an object will not be computed by all methods,  -->
+<!-- testing it in `nlsCompare` makes no sense, and in that package we concentrate on -->
+<!-- the minimum sum of squares and the associated model parameters. -->
+
+<!-- ## Code development -->
+
+<!-- A summary of our main results: -->
+
+<!-- - nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch -->
+
+<!-- - nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED.  -->
+<!--   The effort showed that the structure of the programs was difficult to follow,  -->
+<!--   undocumented, and unsuited to adding improvements like the Marquardt stabilization.  -->
+<!--   We were able to get a pure-R version of `numericDeriv()` and rework most of the functions  -->
+<!--   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project -->
+<!--   formally ends, but collaboration is likely needed with workers who have a deep knowledge -->
+<!--   of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`. -->
+
+<!-- <!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  --> 
+<!-- <!--   fix by JN to the relative offset convergence criterion failure when there are small  --> 
+<!-- <!--   residuals in problems sent to ``nls()`. Now part of R. -->
+
+<!-- - the safeguarded relative offset convergence criterion that allows small-residual problems -->
+<!--   to ba handled with `nls()` is now in the R base code. -->
+
+<!-- - `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features?? -->
+<!--   This is an interim package for exploration of ideas and will NOT, as it currently -->
+<!--   is presented, become a distributed package. -->
+
+<!-- - nlsralt: this is intended to be a modified version of Nash and Murdoch package  -->
+<!--   `nlsr` to examine possible improvements discovered as a result of this project.  -->
+<!--   This is EXPERIMENTAL. -->
 
@@
 
-## Strategic choices in a nonlinear model estimation
-
-The key difference in approach between `nls()` and `nlsr::nlxb()` is that `nls()` 
-builds a large infrastructure from which the Gauss-Newton
+# Strategic choices in nonlinear model estimation
+
+It is possible that `nls()` will remain more or less as it has been for the past
+two decades. Given its importance to R, a focus of discussion should be what needs 
+fixing and how that should be approached, and we welcome an opportunity to participate in 
+such conversations.
+
+From a design point of view, a key difference in approach between `nls()` and 
+`nlsr::nlxb()` is that `nls()` 
+builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
 iteration can be executed and other statistical information such as profiles can be
-computed, while `nlxb()` returns quite limited information, and in its execution
+computed, while `nlxb()` returns quite limited information, and 
 computes what is needed on an "as and when" basis. This follows a path that one of us
-(JN) established almost 50 years ago with the software that became @cnm79, using
-setup, solver, and post-solution analysis phases to computation. Here, that last 
-phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr`
-package. `nls()` also has some similar functions, but they are much more tied into
-the infrastructure that is created, mostly in the `nlsModel()` function.
-
-The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable
-entangling of the different functions and capabilities. Even after
-both the preparation for and carrying out of the present project, we do not feel confident
-to explain the code completely, nor to maintain it. However, we have made some 
-forays into render parts of the code in R, and to understand where difficulties
-lie.
-
-### Opinion for discussion
+(JN) established almost 50 years ago with the software that became @cnm79, separating
+setup, solver, and post-solution analysis phases of computations. 
+<!-- Here, that last  -->
+<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
+<!-- package. -->
+
+<!-- The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable -->
+<!-- entangling of the different functions and capabilities. Even after -->
+<!-- both the preparation for and carrying out of the present project, we do not feel confident -->
+<!-- to explain the code completely, nor to maintain it. However, we have made some  -->
+<!-- forays into render parts of the code in R, and to understand where difficulties -->
+<!-- lie. -->
+
+## Opinions
 
@@
 that are created by tools such as `nls()` should have minimal interactions and side-effects.
-The aspects of `nls()` that have given us the most trouble are as follows:
-
-- The functions that compute the residuals and jacobians are frequently computed by presuming
-  the current data and parameters for those functions are available in an environment. As long
-  as the correct environment is used, this provides a surprisingly short syntax to invoke the
+The aspects of `nls()` that have given us the most trouble:
+
+- The functions that compute the residuals and jacobians often presume
+  the data and parameters needed are available in a particular environment. As long
+  as the correct environment is used, this provides a short syntax to invoke the
   calculations. The danger is that the wrong data is accessed if the internal search finds a
@@
 
-- Weights, subsets, and various contextual controls such as that for the `na.action` that tell
-  our code what to do with undefined or missing numbers. Again, this makes for a very simple
-  invocation of the calculation, but the context is hidden from the user. It can be
+- Weights, subsets, and various contextual controls such as that for the `na.action` option
+  are similarly taken from the first available source. 
+  Again, this makes for a very simple
+  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
   difficult for those of us trying to maintain or improve the code to be certain we have the
@@
 
-- The mixing of R and C code made for high performance in the past, but without 
-  developer documentation, programmers face a lot of work to fix code. We believe in keeping code --
-  at least a runnable reference version -- in a single programming language. If necessary, by
-  measuring ("profiling") the code, we can find bottlenecks and replace just those slower 
+- The mixing of R, C and Fortran code was necessary for computational performance in the past.
+  Lacking decent  
+  developer documentation, programmers now face a lot of work to fix or improve code. 
+  We believe in having at least a working reference version of code in a single programming language. 
+  If necessary, by
+  measuring ("profiling") the code, we can find bottlenecks and, if necessary, substitute just those slower 
   parts of the reference code.
 
-- We believe that a structure that isolates the setup, solve, and post-solution structures 
-  for complicated functions reduces the number of objects that must be kept in appropriate 
-  alignment at any one stage in a set of calculations.
-
-- We feel that all codes should return the best fit they have found so far, even if there
+- A design that isolates the setup, solution, and post-solution parts of
+  for complicated calculations reduces the number of objects that must be kept in 
+  alignment.
+
+- All iterative codes should return the best solution they have found so far, even if there
   are untoward conditions reached, such as a singular Jacobian. This modification could be
-  made to `nls()` even in the short term.
-
-??fix next bits
+  made to `nls()` in the short term.
+
+- Given the existence of examples of good practices such as analytic derivatives, stabilized 
+  solution of Gauss-Newton equations and bounds constrained parameters, base R tools should 
+  be moving to incorporate them.
+
+<!-- ??fix next bits -->
 <!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
@@
 and in comments on this and related work. Heather Turner co-mentored the project 
-and helped guide the progress of the work.??
+and helped guide the progress of the work.
 

### 2022-06-16 DNase1Models.R
minor fixups

@@ -1,0 +1,28 @@
+DNase1 <- subset(DNase, Run == 1) # select the data
+## using a selfStart model - do not specify the starting parameters
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
+
+## using conditional linearity - leave out the Asym parameter
+fm2 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
+           data = DNase1, start = list(xmid = 0, scal = 1),
+           algorithm = "plinear")
+summary(fm2)
+## without conditional linearity
+fm3 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+           data = DNase1,
+           start = list(Asym = 3, xmid = 0, scal = 1))
+summary(fm3)
+
+## using Port's nl2sol algorithm
+fm4 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+               data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+               algorithm = "port"))
+summary(fm4)
+
+## using conditional linearity AND Asym does NOT work
+fm2a <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
+                data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                algorithm = "plinear", trace = TRUE))
+summary(fm2a)
+

### 2022-06-16 nlsj/R/nlsj.R

@@ -22,16 +22,13 @@ for (onename in ncontrol) {
 control<-ctrl
 ctrl<-NULL
 
-# Controls
+# Controls we need to have. ?? use ctrldefault() idea?
    if (is.null(control$derivmeth)) control$derivmeth="default" # for safety
    epstol <- (.Machine$double.eps * control$offset) 
    epsh <- sqrt(epstol)
    epstol4 <- epstol^4 # used for smallsstest
    ##?? may want these in nlsj.control
    if (control$derivmeth == "numericDeriv") warning("Forcing numericDeriv")
-#   cat("control:"); print(control)
-#   tmp <- readline("cont.")
-#   control$watch<-TRUE
 
 # Algorithm
    if (is.null(algorithm)) algorithm<-"default"

### 2022-06-20 MachineSummary/MachineSummary.Rmd
minor changes

@@
 In 2021, we undertook an investigation of tools in R for nonlinear least squares
-calculations under the Google Summer of Code initiative. In the process, we recognized 
+calculations under the Google Summer of Code program. In the process, we recognized 
 that we would be comparing calculations
@@
 storage for disks or memory has decimal and binary schemes of units. Thus 
-(see https://en.wikipedia.org/wiki/Gigabyte) there is a movement to use, for example, Gigabyte or GB to mean usits of 1000^3 bytes, while the Gigibyte or Gib is 1024^3.
+(see https://en.wikipedia.org/wiki/Gigabyte) there is a movement to use, for example, Gigabyte or GB to mean units of $1000^3$ bytes, while the Gibibyte or GiB is $1024^3$` bytes.
 
@@
 suggestion based on the R `Sys.time()` function. Moreover, as we were working on opposite
-sides of the globe, it could be important to either specify time zone or to work with a
+sides of the globe, it deemed important to either specify time zone or to work with a
 single "clock" such as UTC. R provides tools to allow this, as in the following example.
@@
 ```{r tdstamp}
-#get a timestamp
-tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYYmmddHHMM
+# get a time stamp
+tsstr <- format(Sys.time(), "%Y/%m/%d|%H:%M") # tsstr == time stamp string in form YYYY/mm/dd|HH:MM
 cat("Date and Time stamp:", tsstr, "\n")
@@
   these are often uninformative and may be non-unique. We note that VirtualBox with
-  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. Where this name was 
-  generated we are not sure. Settings / System allows "rename this PC". Clearly 
+  a Windows 10 guest machine gave the name DESKTOP-HF4CKVA. We are not sure where this name was generated. Settings / System allows "rename this PC". Clearly 
   we may want to extend the name if particular software is involved, as follows.
@@
 - number of cores
-- operating cycle speed (this is sometimes specific to the cpu specified )
+- operating cycle speed (this is sometimes specific to the cpu specified)
 
@@
 J1 <- mp$ndigits
-cat("E9 = ",E9,"  -- the machine precision,  E9=MIN (X  1+X>1)\n")
+cat("E9 = ", E9, "  -- the machine precision,  E9=MIN (X  1+X>1)\n")
 cat( "E9*1E+16=", E9 * 1E+16, "\n")
-cat( "E6 = ", E6,"  -- the radix of arithmetic", "\n")
-cat( "J1 = ", J1,"  -- the number of radix digits in mantissa of FP numbers", "\n")
+cat( "E6 = ", E6, "  -- the radix of arithmetic", "\n")
+cat( "J1 = ", J1, "  -- the number of radix digits in mantissa of FP numbers", "\n")
 cat( "E6^(-J1+1) * 1E+16=", 1E+16*(E6^(-J1+1)), "\n")

### 2022-06-20 RefactoringNLS/RefactoringNLS.Rmd
a few changes

@@
 
-In the manual page for `nls()` in R 4.0.0 there is the warning
+In the manual page for `nls()` in R 4.0.0 there is the warning:
 
@@
 been computed and put upper limits on these. Our codes exit (terminate) when these limits
-are reached. Generally we prefer larger limits than the default `maxiter=50` of `nls()`,
+are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
 but that may reflect the more difficult problems we have encountered because 
@@
 it is common in nonlinear least squares computations that the Jacobian is very close
-to singular for some values of the model parameters. In such cases we need to find
+to singular for some values of the model parameters. In such cases, we need to find
 an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
@@
 @jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
-In our work we prepared experimental prototypes of `nls` that incorporate
-stabilizations, but integration with all the features of `nls()` has proved difficult.
+In our work, we prepared experimental prototypes of `nls` that incorporate
+stabilization, but integration with all the features of `nls()` has proved difficult.
 
@@
 ```{r SESingGrad}
-reala = -3; realb = 5; realc = 0.5; realr = 0.7; realm = 1 # underlying parameters
-x = 1:11 # x values; 11 data points
-y=reala + realb * realr^(x - realm) + realc * x # linear+exponential function
-testdat = data.frame(x, y) # save in a data frame
+reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
+x <- 1:11 # x values; 11 data points
+y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
+testdat <- data.frame(x, y) # save in a data frame
 strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
@@
 # We get an error with nls()
-linexp=try(nls(jform, data = testdat, start = strt, trace = F))
+linexp <- try(nls(jform, data = testdat, start = strt, trace = F))
 ```
@@
 containing missing values (i.e. any row of a data frame containing an NA). 
-`na.exclude` does much the same for computations, but keeps the rows with NA elements so 
+`na.exclude` does much of the same for computations, but keeps the rows with NA elements so 
 that predictions are in the correct row position. We recommend that workers actually test 
-output to verify behaviour is as wanted.
+output to verify the behaviour is as wanted.
 
@@
 
-`model` is an argument to the `nls()` function, which is documented as
+`model` is an argument to the `nls()` function, which is documented as:
 
@@
 objects in the default (.GlobalEnv) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit, and any such file in the path 
-when R is launched will be loaded. The overwhelming proportion of R users in our
+suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
+when R is launched, will be loaded. The overwhelming proportion of R users in our
 acquaintance avoid saving the workspace because of the danger of lurking data and
@@
 
-Other concerns are revealed by the example below. Here the `SSlogis` selfStart function 
+Other concerns are revealed by the example below. Here, the `SSlogis` selfStart function 
 is used to generate a set
 of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
-is $y \sim Asym/(1+exp((xmid-tt)/scal))$, but we show how the starting parameters
+is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
 for this model can be transformed to those of 
-another form of the model, namely, $y \sim b1/(1 + b2*exp(-b3*t))$.
+another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.
 
@@
 This R function includes analytic expressions for the Jacobian ("gradient").
-These could be be useful to R users, especially if documented. Moreover, we
+These could be useful to R users, especially if documented. Moreover, we
 wonder why the programmers have chosen to save so many quantities in "hidden"
@@
 to 1. This is, in our view, tolerable, but could be improved by using a set of values
-that are all slightly different, which, in the case of the example model $y \sim a*exp(-b*x) + c*exp(-d*x)$
-would avoid a singular Jacobiam because $b$ and $d$ were equal in value. 
+that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
+would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
 A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
@@
 
-```
-A list of:
-
-m	        an nlsModel object incorporating the model.
-data  	    the expression that was passed to nls as the data argument. The actual data values 
-              are present in the environment of the m components, e.g., environment(m$conv).
-call	    the matched call with several components, notably algorithm.
-na.action 	the "na.action" attribute (if any) of the model frame.
-dataClasses	the "dataClasses" attribute (if any) of the "terms" attribute of the model frame.
-model	    if model = TRUE, the model frame.
-weights	    if weights is supplied, the weights.
-convInfo	a list with convergence information.
-control	    the control list used, see the control argument.
-
-There are also two deprecated items if algorithm = "port" fit only. These are
-convergence (a code = 0 for convergence) and message. These are available from convInfo.
-```
+>A list of
+
+>`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.
+
+>`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.
+
+>`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.
+
+>`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.
+
+>`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.
+
+>`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.
+
+>`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.
+
+>`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.
+
+>`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.
+
+>`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.
+
+> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.
 
@@
 The `data` return element is an R symbol. To actually access the data from this 
-element, we need to use the syntax
+element, we need to use the syntax:
 
@@
 However, if the call is made with `model=TRUE`, then there is a returned element
-`model` which contains the data, and we can list its contents using
+`model` which contains the data, and we can list its contents using:
 
@@
 
-```
-coefficients    A named vector giving the parameter values at the supposed solution.
-ssquares        The sum of squared residuals at this set of parameters.
-resid           The residual vector at the returned parameters.
-jacobian        The jacobian matrix (partial derivatives of residuals w.r.t. the 
-                parameters) at the returned parameters.
-feval           The number of residual evaluations (sum of squares computations) used.
-jeval           The number of Jacobian evaluations used.
-```
+>`coefficients`    A named vector giving the parameter values at the supposed solution.
+
+>`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.
+
+>`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.
+
+>`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
+                parameters) at the \newline
+$~~~~~~~~~~~~~~~~~~~~$ returned parameters.
+
+>`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.
+
+>`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.
 
@@
 ```{r tetrarun}
-time=c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
-conc = c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
+time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
+conc <- c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
 NLSdata <- data.frame(time,conc)
-NLSstart <-c(lrc1=-2,lrc2=0.25,A1=150,A2=50) # a starting vector (named!)
-NLSformula <-conc ~ A1*exp(-exp(lrc1)*time)+A2*exp(-exp(lrc2)*time)
-tryit <- try(nls(NLSformula, data=NLSdata, start=NLSstart, trace=TRUE))
+NLSstart <- c(lrc1 = -2, lrc2 = 0.25, A1 = 150, A2 = 50) # a starting vector (named!)
+NLSformula <- conc ~ A1 * exp(-exp(lrc1) * time) + A2 * exp(-exp(lrc2) * time)
+tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
 print(tryit)
@@
 One obvious case is that nonlinear modeling tools are a superset of linear ones.
-Yet the explicit model `y ~ a*x + b` does not work with the linear modeling 
+Yet the explicit model `y ~ a * x + b` does not work with the linear modeling 
 function `lm()`, which requires this model to be specified as `y ~ x`.
@@
 to try to add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing we do
-not foresee a resolution for this issue. In the example we have NOT evaluated
+like to see a consistent syntax. At the time of writing, we do
+not foresee a resolution for this issue. In the example, we have NOT evaluated
 the commands to save space.
@@
 ## using conditional linearity - leave out the Asym parameter
-fm2 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
+fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
                  data = DNase1, start = list(xmid = 0, scal = 1),
@@
 ## without conditional linearity
-fm3 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
                  data = DNase1,
@@
 ## using Port's nl2sol algorithm
-fm4 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
+fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
                  data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
@@
 ## using conditional linearity AND Asym does NOT work
-fm2a <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
+fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
                  data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
@@
 to Doug Bates and his reply.
-This is NOT a criticism of Prof. Bates (or any other) work, but a reflection on 
+This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on 
 how difficult it is to develop code in this subject area and to keep it maintainable. 
@@
 
-> The algorithm = "port" code appears unfinished, and does not even check that the starting 
+> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
   value is within the bounds. Use with caution, especially where bounds are supplied.
@@
 tt <- 1:12
-wf <- data.frame(y=weed, tt=tt)
-st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
-wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt)) ## Scaled model
+wf <- data.frame(y = weed, tt = tt)
+st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
+wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
 require(minpack.lm)
@@
 # Hobbs scaled problem with bounds, formula specification
-anlxb1 <- nlxb(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
 cat("Using nlsr::nlxb():")
@@
 ## nlsLM seems NOT to work properly with bounds
-anlsLM1b <- nlsLM(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
 cat("\n"); cat("using minpack.lm::nlsLM():")
 print(anlsLM1b)
-anlsb<-nls(wmods, start=st, data=wf, algorithm="port", lower=c(0,0,0), upper=c(2,6,3))
+anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", lower = c(0, 0, 0), upper = c(2, 6, 3))
 cat("Using nls with 'port':")
@@
 the parameter may be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions. Hence `nlsr::nlxb()` does not compute standard
+Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
 errors nor their derived statistics when bounds are active.
@@
 
-We want
+We want:
 
@@
 - to streamline the overall `nls()` infrastructure. By this we
-  mean a re-factoring the routines so they are better suited to maintenance of both
+  mean a re-factoring of the routines so they are better suited to maintenance of both
   the existing `nls()` methods and features as well as new features we or others
@@
 Thus a linear model (which is valid input for nonlinear estimatin) should
-be specified `y ~ a*x + b` rather than `y ~ x`. 
+be specified `y ~ a * x + b` rather than `y ~ x`. 
 Moreover, we want to avoid the partially linear parameters appearing 
@@
 A further complication of model specification in R is that some formulas require
-the user to surround a term with `I(  )` to inhibit interpretation of arithmetic 
+the user to surround a term with `I()` to inhibit interpretation of arithmetic 
 operators. We would prefer that formulas be usable without this complication.
@@
 We need to be able to process models such as 
-`Length ~ a[Strip] + b[Strip] * exp(-Conc/th)` where `Strip` is the index. Data
+`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)` where `Strip` is the index. Data
 `Length` and `Conc` are available for all observations and parameter `th` applies
@@
 - The mixing of R, C and Fortran code was necessary for computational performance in the past.
-  Lacking decent  
-  developer documentation, programmers now face a lot of work to fix or improve code. 
+  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
   We believe in having at least a working reference version of code in a single programming language. 

### 2022-06-20 RefactoringNLS/RefactoringNLS.tex

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Refactoring the nls() function in R},
  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
  colorlinks=true,
  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi

\title{Refactoring the \texttt{nls()} function in R}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2022-6-15}

\begin{document}
\maketitle

{
\hypersetup{linkcolor=}
\setcounter{tocdepth}{2}
\tableofcontents
}
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Based on work for a Google Summer of Code project ``Improvements to
\texttt{nls()}'', we consider the features and limitations of the R
function \texttt{nls()} in the context of improving and rationalizing R
tools for nonlinear regression.

An important consideration is the maintainability of the code base that
provides the functionality \texttt{nls()} claims to offer. Our work
suggests that the existing code makes maintenance and improvement very
difficult, with legacy applications and examples blocking some important
updates. Discussion of these matters is relevant to improving R
generally, as well as its nonlinear estimation tools.

\hypertarget{the-nls-function-strengths-and-shortcomings}{%
\section{\texorpdfstring{The \texttt{nls()} function: strengths and
shortcomings}{The nls() function: strengths and shortcomings}}\label{the-nls-function-strengths-and-shortcomings}}

\texttt{nls()} is the tool in base R (the distributed software package
from \href{https://cran.r-project.org}{CRAN} for estimating nonlinear
statistical models. The function dates to the 1980s and the work related
to D. M. Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S
(see
\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).

The \texttt{nls()} function has a remarkable and comprehensive set of
capabilities for estimating nonlinear models that are expressed as
\textbf{formulas}. In particular, we note that it

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions
\item
  allows data to be subset
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to a number of pre-coded (``selfStart'') models
\end{itemize}

With such a range of features and long history, the code has become
untidy and overly patched. It is, to our mind, essentially
unmaintainable. Moreover, its underlying methods can and should be
improved, as we suggest below.

\hypertarget{feature-convergence-and-termination-tests-fixed}{%
\subsection{Feature: Convergence and termination tests
(FIXED)}\label{feature-convergence-and-termination-tests-fixed}}

A previous issue with \texttt{nls()} that prevented it from providing
parameter estimates for zero-residual (i.e., perfect fit) data was
corrected thanks to suggestions by one of us.

In the manual page for \texttt{nls()} in R 4.0.0 there is the warning:

\begin{quote}
\textbf{Do not use \texttt{nls} on artificial ``zero-residual'' data.}
\end{quote}

\begin{quote}
The \texttt{nls} function uses a relative-offset convergence criterion
that compares the numerical imprecision at the current parameter
estimates to the residual sum-of-squares. This performs well on data of
the form
\end{quote}

\begin{quote}
\[  y = f(x, \theta) + eps \]
\end{quote}

\begin{quote}
(with \texttt{var(eps)\ \textgreater{}\ 0}). It fails to indicate
convergence on data of the form
\end{quote}

\begin{quote}
\[  y = f(x, \theta)  \]
\end{quote}

\begin{quote}
because the criterion amounts to comparing two components of the
round-off error.
\end{quote}

\begin{quote}
If you wish to test \texttt{nls} on artificial data please add a noise
component, as shown in the example below.
\end{quote}

This amounts to admitting R cannot solve well-posed problems unless data
is polluted with errors.

This issue can be easily resolved. The ``termination test'' for the
\textbf{program} rather than for ``convergence'' of the underlying
\textbf{algorithm} is the Relative Offset Convergence Criterion (see
Douglas M. Bates and Watts
(\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
proposed step in the parameter vector on the gradient and estimates how
much the sum of squares loss function should decrease. This estimate is
divided by the current size of the loss function to avoid scale issues.
When we have ``converged'', the estimated decrease is very small.
However, with small residuals, the sum of squares loss function is
(almost) zero and we get the possibility of a zero-divide failure.

Adding a small quantity to the loss function before dividing avoids
trouble. In 2021, one of us (J. Nash) proposed that
\texttt{nls.control()} have an additional parameter \texttt{scaleOffset}
with a default value of zero. Setting it to a small number -- 1.0 is a
reasonable choice -- allows small-residual problems (i.e., near-exact
fits) to be dealt with easily. We call this the \textbf{safeguarded
relative offset convergence criterion}. The default value gives the
legacy behaviour. We are pleased to note that this improvement is now in
the R distributed code as of version 4.1.0.

\hypertarget{more-general-termination-tests}{%
\subsubsection{More general termination
tests}\label{more-general-termination-tests}}

The single convergence criterion of \texttt{nls()} leaves out some
possibilities that could be useful for some problems. The package
\texttt{nlsr} (John C. Nash and Murdoch
(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
safeguarded relative offset test (\textbf{roffset}) as well as a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

We note that \texttt{nls()} uses a termination test to stop after
\texttt{maxiter} ``iterations''. Unfortunately, the meaning of
``iteration'' varies among programs and requires careful examination of
the code. We prefer to use the number of times the residuals or the
jacobian have been computed and put upper limits on these. Our codes
exit (terminate) when these limits are reached. Generally we prefer
larger limits than the default \texttt{maxiter\ =\ 50} of
\texttt{nls()}, but that may reflect the more difficult problems we have
encountered because users consult us when standard tools have given
unsatisfactory results.

\hypertarget{feature-failure-when-jacobian-is-computationally-singular}{%
\subsection{Feature: Failure when Jacobian is computationally
singular}\label{feature-failure-when-jacobian-is-computationally-singular}}

This refers to the infamous ``singular gradient'' termination message of
\texttt{nls()}. A Google search of

\begin{verbatim}
R nls "singular gradient"
\end{verbatim}

gets over 4000 hits that are spread over some years. This could be
because the Jacobian is poorly approximated (see \textbf{Jacobian
computation} below). However, it is common in nonlinear least squares
computations that the Jacobian is very close to singular for some values
of the model parameters. In such cases, we need to find an alternative
algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most
common work-around is the Levenberg-Marquardt stabilization (see
Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
(\protect\hyperlink{ref-Levenberg1944}{1944}), John C. Nash
(\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
implemented in packages \texttt{minpack.lm} and \texttt{nlsr}. In our
work, we prepared experimental prototypes of \texttt{nls} that
incorporate stabilization, but integration with all the features of
\texttt{nls()} has proved difficult.

Let us consider an `singular gradient' example
\emph{\url{https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values}}.
\texttt{nlsr::nlxb()} displays the singular values of the Jacobian, so
the \texttt{nls()} error message is correct, but unhelpful in obtaining
a solution.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{reala }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{; realb }\OtherTok{\textless{}{-}} \DecValTok{5}\NormalTok{; realc }\OtherTok{\textless{}{-}} \FloatTok{0.5}\NormalTok{; realr }\OtherTok{\textless{}{-}} \FloatTok{0.7}\NormalTok{; realm }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# underlying parameters}
\NormalTok{x }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{11} \CommentTok{\# x values; 11 data points}
\NormalTok{y }\OtherTok{\textless{}{-}}\NormalTok{ reala }\SpecialCharTok{+}\NormalTok{ realb }\SpecialCharTok{*}\NormalTok{ realr}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ realm) }\SpecialCharTok{+}\NormalTok{ realc }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# linear + exponential function}
\NormalTok{testdat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(x, y) }\CommentTok{\# save in a data frame}
\NormalTok{strt }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{, }\AttributeTok{b =} \DecValTok{5}\NormalTok{, }\AttributeTok{c =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{r =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{m =} \DecValTok{1}\NormalTok{) }\CommentTok{\# give programs a good start}
\NormalTok{jform }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b }\SpecialCharTok{*}\NormalTok{ r}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ m) }\SpecialCharTok{+}\NormalTok{ c }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# Formula}
\FunctionTok{library}\NormalTok{(nlsr)}
\NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =}\NormalTok{ F))}
\NormalTok{linexp2 }\CommentTok{\# Note singular values of Jacobian in rightmost column}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  0  on  11 observations
##     after  1    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## a                     -3            NA         NA         NA           0       26.49  
## b                      5            NA         NA         NA           0       9.615  
## c                    0.5            NA         NA         NA           0       2.466  
## r                    0.7            NA         NA         NA           0      0.1098  
## m                      1            NA         NA         NA           0   1.118e-16
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# We get an error with nls()}
\NormalTok{linexp }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =}\NormalTok{ F))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nlsModel(formula, mf, start, wts, scaleOffset = scOff, nDcentral = nDcntr) : 
##   singular gradient matrix at initial parameter estimates
\end{verbatim}

The Jacobian is essentially singular as shown by its singular values.
Note that these are \textbf{displayed} in a single column in the output
to provide a compact layout, but the values do \textbf{NOT} correspond
to the individual parameters in whose row they appear; they are a
property of the whole problem.

\hypertarget{feature-jacobian-computation}{%
\subsection{Feature: Jacobian
computation}\label{feature-jacobian-computation}}

\texttt{nls()}, with the \texttt{numericDeriv()} function, computes the
Jacobian as the ``gradient'' attribute of the residual vector. This is
implemented as a mix of R and C code, but we have created a rather more
compact version entirely in R. See
\url{https://github.com/nashjc/RNonlinearLS/DerivsNLS} .

The Jacobian used by \texttt{nlsr::nlxb()} is computed from analytic
expressions, while that of \texttt{nls()} (and, by extension,
\texttt{minpack.lm::nlsLM}) mostly use a numerical approximation. This
can lead to different behaviour:

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al.
  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
  conditioned Jacobian and \texttt{nls()} fails with ``singular
  gradient''. Worse, numerical approximation to the Jacobian give the
  error ``singular gradient matrix at initial parameter estimates'' for
  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
  unable to take effect, while the analytic derivatives of
  \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a
  model specified with the formula
  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
  which \texttt{x\ =\ 0}, this is undefined. In such cases, we observed
  that \texttt{nls()} and \texttt{minpack.lm::nlsLM} found a solution,
  though it can be debated whether such lucky accidents can be taken as
  an advantage.
\end{itemize}

Note that the selfStart models in the base R
\texttt{./src/library/stats/R/zzModels.R} file provide the Jacobian in
the ``gradient'' attribute of the ``one-sided'' formula that defines
each model, and these Jacobians may be the analytic forms. The
\texttt{nls()} function, after computing the ``right hand side'' or
\texttt{rhs} of the residual, then checks to see if the ``gradient''
attribute is defined, and, if not, uses \texttt{numericDeriv} to compute
a Jacobian into that attribute. This code is within the
\texttt{nlsModel()} or \texttt{nlsModel.pliner()} functions. The use of
analytic Jacobians almost certainly contributes to the good performance
of \texttt{nls()} on selfStart models.

\hypertarget{feature-subsetting}{%
\subsection{Feature: Subsetting}\label{feature-subsetting}}

\texttt{nls()} accepts an argument \texttt{subset}. This acts through
the mediation of \texttt{model.frame} and is not clearly obvious in the
source code files \texttt{/src/library/stats/R/nls.R} and
\texttt{/src/library/stats/src/nls.C}.

While the implementation of subset at the level of the call to
\texttt{nls()} has a certain attractiveness, it does mean that the
programmer of the solver needs to be aware of the source (and value) of
objects such as the data, residuals and Jacobian. By preference, we
would implement subsetting by means of zero-value weights, with
observation counts (and degrees of freedom) computed via the numbers of
non-zero weights. Alternatively, we would extract a working dataframe
from the relevant elements in the original.

\hypertarget{feature-na.action}{%
\subsection{Feature: na.action}\label{feature-na.action}}

\texttt{na.action} is an argument to the \texttt{nls()} function, but it
does not appear obviously in the source code, often being handled behind
the scenes after referencing the option \texttt{na.action}. A useful,
but possibly dated, description is given in:
\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}.

The typical default action, which can be seen by using the command
\texttt{getOption("na.action")} is \texttt{na.omit}. This option
essentially omits from computations any observations containing missing
values (i.e.~any row of a data frame containing an NA).
\texttt{na.exclude} does much of the same for computations, but keeps
the rows with NA elements so that predictions are in the correct row
position. We recommend that workers actually test output to verify the
behaviour is as wanted.

A succinct description of the issue is given in:
\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}

\begin{quote}
The only benefit of \texttt{na.exclude} over \texttt{na.omit} is that
the former will retain the original number of rows in the data. This may
be useful where you need to retain the original size of the dataset -
for example it is useful when you want to compare predicted values to
original values. With \texttt{na.omit} you will end up with fewer rows
so you won't as easily be able to compare.
\end{quote}

\texttt{na.pass} simply passes on data ``as is'', while \texttt{na.fail}
will essentially stop if any missing values are present.

Our concern with \texttt{na.action} is that users may be unaware of the
effects of an option setting they may not even be aware has been set. Is
\texttt{na.fail} a safer default?

\hypertarget{feature-model-frame}{%
\subsection{Feature: model frame}\label{feature-model-frame}}

\texttt{model} is an argument to the \texttt{nls()} function, which is
documented as:

\begin{quote}
\textbf{model} logical. If true, the model frame is returned as part of
the object. Default is FALSE.
\end{quote}

Indeed, the argument only gets used when \texttt{nls()} is about to
return its result object, and the element \texttt{model} is NULL unless
the calling argument \texttt{model} is TRUE. (Using the same name could
be confusing.) However, the model frame is used within the function code
in the form of the object \texttt{mf}. We feel that users could benefit
from more extensive documentation and examples of its use.

\hypertarget{feature-sources-of-data}{%
\subsection{Feature: sources of data}\label{feature-sources-of-data}}

\texttt{nls()} can be called without specifying the \texttt{data}
argument. In this case, it will search in the available environments
(i.e., workspaces) for suitable data objects. We do NOT like this
approach, but it is ``the R way''. R allows users to leave many objects
in the default (.GlobalEnv) workspace. Moreover, users have to actively
suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any
such file in the path, when R is launched, will be loaded. The
overwhelming proportion of R users in our acquaintance avoid saving the
workspace because of the danger of lurking data and functions which may
cause unwanted results.

Nevertheless, to provide compatible behaviour with \texttt{nls()},
competing programs need to ensure equivalent behaviour, but users should
test that the operation is as they intend.

\hypertarget{feature-missing-start-vector-and-self-starting-models}{%
\subsection{Feature: missing start vector and self-starting
models}\label{feature-missing-start-vector-and-self-starting-models}}

Nonlinear estimation algorithms are almost all iterative and need a set
of starting parameters. \texttt{nls()} offers a special class of
modeling functions called \textbf{selfStart} models. There are a number
of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
others in R packages such as CRAN package \texttt{nlraa} (Miguez
(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
now-archived package \texttt{NRAIA}. Unfortunately, some
\textbf{selfStart} codes entangle the calculation of starting values for
parameters with the particulars of the \texttt{nls()} code. Though there
is a \texttt{getInitial()} function, this is not easy to use to simply
compute the initial parameter estimates, in part because it may call
\texttt{nls()}. Such circular references are, in our view, dangerous.

Other concerns are revealed by the example below. Here, the
\texttt{SSlogis} selfStart function is used to generate a set of initial
parameters for a 3-parameter logistic curve. The form used by
\texttt{SSlogis} is
\(y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))\), but we
show how the starting parameters for this model can be transformed to
those of another form of the model, namely,
\(y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))\).

The code for \texttt{SSlogis()} is in
\texttt{./src/library/stats/R/zzModels.R}. This R function includes
analytic expressions for the Jacobian (``gradient''). These could be
useful to R users, especially if documented. Moreover, we wonder why the
programmers have chosen to save so many quantities in ``hidden''
variables, i.e., with names preceded by ``.''. These are then not
displayed by the \texttt{ls()} command, making them difficult to access.

In the event that a selfStart model is not available, \texttt{nls()}
sets all the starting parameters to 1. This is, in our view, tolerable,
but could be improved by using a set of values that are all slightly
different, which, in the case of the example model
\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
avoid a singular Jacobian because \(b\) and \(d\) were equal in value. A
sequence like 1.0, 1.1, 1.2, 1.3 for the four parameters could be
provided quite simply instead of all 1's.

\hypertarget{strategic-issues-in-selfstart-models}{%
\subsubsection{Strategic issues in selfStart
models}\label{strategic-issues-in-selfstart-models}}

Because the Gauss-Newton algorithm is unreliable from many sets of
starting of parameters, selfStart models are a part of the
\texttt{nls()} infrastructure rather than an accessory. However,
creating such functions is a lot of work, and their documentation (file
\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
believe that the focus should be placed on getting good initial
parameters, that is \texttt{getInitial()} function, though avoiding the
current calls back to \texttt{nls()}. Interactive tools, such as
``visual fitting'' (John C. Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering.

\hypertarget{issue-returned-output-of-nls-and-its-documentation}{%
\subsection{Issue: returned output of nls() and its
documentation}\label{issue-returned-output-of-nls-and-its-documentation}}

The output of \texttt{nls()} is an object of class ``nls'' which has the
following structure:

\begin{quote}
A list of
\end{quote}

\begin{quote}
\texttt{m} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ an \texttt{nlsModel} object
incorporating the model.
\end{quote}

\begin{quote}
\texttt{data} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~the expression that was
passed to \texttt{nls} as the data argument. The actual \newline
\(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) data values are present in the
environment of the \texttt{m} components, e.g.,
\newline \(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\)
\texttt{environment(m\$conv)}.
\end{quote}

\begin{quote}
\texttt{call} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the matched call with several
components, notably \texttt{algorithm}.
\end{quote}

\begin{quote}
\texttt{na.action} ~ ~ ~ ~ ~ ~ ~ ~ ~ the \texttt{"na.action"} attribute
(if any) of the model frame.
\end{quote}

\begin{quote}
\texttt{dataClasses} ~ ~ ~ ~ ~ ~ ~ \(~\)the \texttt{"dataClasses"}
attribute (if any) of the ``\texttt{terms}'' attribute of the \newline
\(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) model frame.
\end{quote}

\begin{quote}
\texttt{model} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~if \texttt{model\ =\ TRUE}, the
model frame.
\end{quote}

\begin{quote}
\texttt{weights} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ if \texttt{weights} is supplied,
the weights.
\end{quote}

\begin{quote}
\texttt{convInfo} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ a list with convergence
information.
\end{quote}

\begin{quote}
\texttt{control} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the control \texttt{list} used,
see the \texttt{control} argument.
\end{quote}

\begin{quote}
\texttt{convergence,\ message} ~for an \texttt{algorithm\ =\ "port"} fit
only, a convergence code (\texttt{0} for \newline
\(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\)convergence) and message.
\end{quote}

\begin{quote}
~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ To use these is \emph{deprecated}, as
they are available from \texttt{convInfo} now.
\end{quote}

The \texttt{nls} object contains some elements that are awkward to
produce by other algorithms, but some information that would be useful
is not presented obviously. Moreover, the complexity of the object is a
challenge to users.

In the following, we use \texttt{result} as the returned object from
\texttt{nls()}.

The \texttt{data} return element is an R symbol. To actually access the
data from this element, we need to use the syntax:

\begin{verbatim}
eval(parse(text=result$data))
\end{verbatim}

However, if the call is made with \texttt{model=TRUE}, then there is a
returned element \texttt{model} which contains the data, and we can list
its contents using:

\begin{verbatim}
ls(result$model)
\end{verbatim}

and if there is an element called \texttt{xdata}, it can be accessed as
\texttt{result\$model\$xdata}.

Let us compare the \texttt{nls()} result with that from
\texttt{nlsr::nlxb()}, with ostensibly solves the same problem:

\begin{quote}
\texttt{coefficients} A named vector giving the parameter values at the
supposed solution.
\end{quote}

\begin{quote}
\texttt{ssquares} \(~~~~~\) The sum of squared residuals at this set of
parameters.
\end{quote}

\begin{quote}
\texttt{resid} \(~~~~~~~~~~\) The residual vector at the returned
parameters.
\end{quote}

\begin{quote}
\texttt{jacobian} \(~~~~~\) The jacobian matrix (partial derivatives of
residuals w.r.t. the parameters) at the \newline
\(~~~~~~~~~~~~~~~~~~~~\) returned parameters.
\end{quote}

\begin{quote}
\texttt{feval} \(~~~~~~~~~~\) The number of residual evaluations (sum of
squares computations) used.
\end{quote}

\begin{quote}
\texttt{jeval} \(~~~~~~~~~~\) The number of Jacobian evaluations used.
\end{quote}

However, actually looking at the structure of a returned result gives a
list of 11 items. The extra 5 are:

\begin{verbatim}
 $ lower       : num [1:3] -Inf -Inf -Inf
 $ upper       : num [1:3] Inf Inf Inf
 $ maskidx     : int(0) 
 $ weights     : NULL
 $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
 - attr(*, "class")= chr "nlsr"
\end{verbatim}

The result object from \texttt{nlsr::nlxb()} is still much smaller than
the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
returns the sum of squares as well as the residual vector and Jacobian.
The counts of evaluations are also returned. (Note that the singular
values of the Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result.) We noted several potential
updates to the \texttt{nlsr} documentation as well as that for
\texttt{nls()} as we made our comparisons.

\hypertarget{weights-in-returned-functions-from-nls}{%
\subsubsection{Weights in returned functions from
nls()}\label{weights-in-returned-functions-from-nls}}

The functions \texttt{resid()} (an alias for \texttt{residuals()}) and
\texttt{fitted()} and \texttt{lhs()} are UNWEIGHTED. But if we return
\texttt{ans} from \texttt{nls()} or \texttt{minpack.lm::nlsLM} or our
new \texttt{nlsj} (interim package), then \texttt{ans\$m\$resid()} is
WEIGHTED. This needs clear documentation so users can get the answers
intended.

\hypertarget{interim-output-from-the-port-algorithm}{%
\subsubsection{Interim output from the ``port''
algorithm}\label{interim-output-from-the-port-algorithm}}

As the \texttt{nls()} \textbf{man} page states, when the ``port''
algorithm is used with the \texttt{trace} argument TRUE, the iterations
display the objective function value which is 1/2 the sum of squares (or
deviance). It is likely that the trace display is embedded in the
Fortran of the \texttt{nlminb} routine that is called to execute the
``port'' algorithm, but the discrepancy is nonetheless unfortunate for
users.

\hypertarget{failure-to-return-best-result-achieved}{%
\subsubsection{Failure to return best result
achieved}\label{failure-to-return-best-result-achieved}}

If \texttt{nls()} reaches a point where it cannot continue but has not
found a point where the relative offset convergence criterion is met, it
may simply exit, especially if a ``singular gradient'' (singular
Jacobian) is found. However, this may occur AFTER the function has made
considerable progress in reducing the sum of squared residuals. Here is
an abbreviated example:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
\NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
\NormalTok{NLSstart }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{lrc1 =} \SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\AttributeTok{lrc2 =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{A1 =} \DecValTok{150}\NormalTok{, }\AttributeTok{A2 =} \DecValTok{50}\NormalTok{) }\CommentTok{\# a starting vector (named!)}
\NormalTok{NLSformula }\OtherTok{\textless{}{-}}\NormalTok{ conc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ A1 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc1) }\SpecialCharTok{*}\NormalTok{ time) }\SpecialCharTok{+}\NormalTok{ A2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc2) }\SpecialCharTok{*}\NormalTok{ time)}
\NormalTok{tryit }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(NLSformula, }\AttributeTok{data =}\NormalTok{ NLSdata, }\AttributeTok{start =}\NormalTok{ NLSstart, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      61216. (3.56e+03): par = (-2 0.25 150 50)
## 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
## 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
## Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(tryit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
## attr(,"class")
## [1] "try-error"
## attr(,"condition")
## <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
\end{verbatim}

Note that the sum of squares has been reduced from 61216 to 1.6211, but
unless \texttt{trace} is invoked, the user will not get any information
about this. This almost trivial change to the \texttt{nls()} function
could be useful to R users.

\hypertarget{feature-partially-linear-models-and-their-specification}{%
\subsection{Feature: partially linear models and their
specification}\label{feature-partially-linear-models-and-their-specification}}

Specifying a model to a solver should, ideally, use the same syntax
across solver tools. Unfortunately, R allows multiple approaches within
different modelling tools, and within \texttt{nls()} itself.

One obvious case is that nonlinear modeling tools are a superset of
linear ones. Yet the explicit model
\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
linear modeling function \texttt{lm()}, which requires this model to be
specified as \texttt{y\ \textasciitilde{}\ x}.

Within \texttt{nls()}, consider the following FOUR different calling
sequences for the same problem, though an intuitive choice, labelled
fm2a, will not work. In this failed attempt, putting the \texttt{Asym}
parameter in the model causes the \texttt{plinear} algorithm to try to
add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do not
foresee a resolution for this issue. In the example, we have NOT
evaluated the commands to save space.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DNase1 }\OtherTok{\textless{}{-}} \FunctionTok{subset}\NormalTok{(DNase, Run }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\CommentTok{\# select the data}
\DocumentationTok{\#\# using a selfStart model {-} do not specify the starting parameters}
\NormalTok{fm1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(}\FunctionTok{log}\NormalTok{(conc), Asym, xmid, scal), DNase1)}
\FunctionTok{summary}\NormalTok{(fm1)}

\DocumentationTok{\#\# using conditional linearity {-} leave out the Asym parameter}
\NormalTok{fm2 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fm2)}

\DocumentationTok{\#\# without conditional linearity}
\NormalTok{fm3 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1,}
                 \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm3)}

\DocumentationTok{\#\# using Port\textquotesingle{}s nl2sol algorithm}
\NormalTok{fm4 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm4)}

\DocumentationTok{\#\# using conditional linearity AND Asym does NOT work}
\NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm2a)}
\end{Highlighting}
\end{Shaded}

\hypertarget{issue-code-structure-and-documentation-for-maintenance}{%
\subsection{Issue: code structure and documentation for
maintenance}\label{issue-code-structure-and-documentation-for-maintenance}}

The \texttt{nls()} code is structured in a way that inhibits both
maintenance and improvement. In particular, the iterative setup is such
that introduction of Marquardt stabilization is not easily available. To
obtain performance a lot of the code was in C with consequent calls and
returns that complicate the code. Over time, R has become much more
efficient on modern computers, and the need to use compiled C and
Fortran is less critical. Moreover, the burden for maintenance could be
reduced by moving code entirely to R.

While R and its packages are generally very well documented for usage,
that documentation rarely extends as far as it might in directions
helpful for code maintenance. The paucity of such documentation is
exacerbated by the mixed R/C/Fortran code base, where seemingly trivial
differences in things like representation of numbers or base index for
arrays lead to traps for unwary programmers.

For \texttt{nls()} the concern is demonstrated by a short email query
from John Nash to Doug Bates and his reply. This is NOT a criticism of
Prof.~Bates' (or any other) work, but a reflection on how difficult it
is to develop code in this subject area and to keep it maintainable. We
have experienced similar loss of understanding for some of our own
codes.

\begin{verbatim}
    ...
    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
    ... has the test problem and the expanded code. Around line 367 is where we are
    scratching our heads. The function code (from nlsModel()) is in the commented 
    lines below the call. This is

          # > setPars
          # function(newPars) {
          #   setPars(newPars)
          #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
          #   dev   <<- sum(resid^2) # envir = thisEnv
          #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
          #   QR <<- qr(.swts * gr) # envir = thisEnv
          #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
          # }

    I'm anticipating that we will be able to set up a (possibly inefficient) code
    with documentation that will be easier to follow and test, then gradually figure
    out how to make it more efficient.

    The equivalent from minpack.lm is

    setPars = function(newPars) {
                setPars(newPars)
                assign("resid", .swts * (lhs - assign("rhs", getRHS(),
                    envir = thisEnv)), envir = thisEnv)
                assign("dev", sum(resid^2), envir = thisEnv)
                assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv)
                return(QR$rank < min(dim(QR$qr)))
            }

    In both there is the recursive call, which must have a purpose I don't understand.
\end{verbatim}

The reply was

\begin{verbatim}
    I'm afraid that I don't know the purpose of the recursive call either.  
    I know that I wrote the code to use a closure for the response, covariates, etc., 
    but I don't recall anything like a recursive call being necessary.
    ...
\end{verbatim}

\hypertarget{feature-indexed-parameters}{%
\subsection{Feature: indexed
parameters}\label{feature-indexed-parameters}}

The \textbf{man} file for \texttt{nls()} includes the following example
of a situation in which parameters are indexed. It also uses the
``plinear'' option as an added complication.

Running the example reveals that the answers for the parameters are NOT
indexed. We do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but
\texttt{a1,\ a2,\ a3}. This is no doubt because programming for indexed
parameters is extremely challenging. However, we believe the current
structure is could cause confusion and error, and propose an alternative
approach below.

\hypertarget{feature-bounds-on-parameters}{%
\subsection{Feature: bounds on
parameters}\label{feature-bounds-on-parameters}}

There are many situations where the context of a problem constrains the
values of parameters. For example, one of us was asked many years ago to
estimate a model where one parameter estimate was negative. The client
complained ``But that is supposed to be the number of grain elevators in
Saskatchewan''. The number should have been a positive integer, and
likely less than a few thousand.

\texttt{nls()} can impose bounds on parameters, but only if the
\texttt{port} algorithm is used. Unfortunately, the manual states

\begin{quote}
The \texttt{algorithm\ =\ "port"} code appears unfinished, and does not
even check that the starting value is within the bounds. Use with
caution, especially where bounds are supplied.
\end{quote}

This is clearly unsatisfactory for general use, but does find bounded
solutions. Inclusion of bounds on parameters in nonlinear least squares
computations is relatively straightforward, and it is part of the
default method for package \texttt{nlsr}. Package \texttt{minpack.lm}
also includes provision for bounds, but the following script, which we
do not run here for reasons of space, shows that it does not find a good
solution for the scaled Hobbs test problem.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# bhobbsX.R \#\# bounded formula specification of problem using Hobbs Weed problem}
\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{wf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{y =}\NormalTok{ weed, }\AttributeTok{tt =}\NormalTok{ tt)}
\NormalTok{st }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
\NormalTok{wmods }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}} \DecValTok{100} \SpecialCharTok{*}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ b3 }\SpecialCharTok{*}\NormalTok{ tt)) }\DocumentationTok{\#\# Scaled model}
\FunctionTok{require}\NormalTok{(minpack.lm)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Loading required package: minpack.lm
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{require}\NormalTok{(nlsr)}
\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
\NormalTok{anlxb1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nlsr::nlxb():"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Using nlsr::nlxb():
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(anlxb1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## nlsr object: x 
## residual sumsquares =  9.4726  on  12 observations
##     after  16    Jacobian and  17 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1                     2U           NA         NA         NA           0       23.35  
## b2               4.43325            NA         NA         NA   1.832e-10           0  
## b3                     3U           NA         NA         NA           0           0
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\DocumentationTok{\#\# nlsLM seems NOT to work properly with bounds}
\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\FunctionTok{cat}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{); }\FunctionTok{cat}\NormalTok{(}\StringTok{"using minpack.lm::nlsLM():"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## using minpack.lm::nlsLM():
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(anlsLM1b)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Nonlinear regression model
##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
##    data: wf
## b1 b2 b3 
##  2  6  3 
##  residual sum-of-squares: 881
## 
## Number of iterations to convergence: 2 
## Achieved convergence tolerance: 1.49e-08
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Using nls with 'port':
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(anlsb)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Nonlinear regression model
##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
##    data: wf
##   b1   b2   b3 
## 2.00 4.43 3.00 
##  residual sum-of-squares: 9.47
## 
## Algorithm "port", convergence message: both X-convergence and relative convergence (5)
\end{verbatim}

\hypertarget{philosophical-considerations}{%
\subsubsection{Philosophical
considerations}\label{philosophical-considerations}}

Bounds on parameters raise some interesting and difficult questions
about how uncertainty in parameter estimates should be computed or
reported. That is, the traditional ``standard errors'' are generally
taken to imply symmetric intervals about the point estimate in which the
parameter may be expected to be found with some probability under
certain assumptions. Bounds change those assumptions. Hence,
\texttt{nlsr::nlxb()} does not compute standard errors nor their derived
statistics when bounds are active.

\hypertarget{goals-and-progress-of-our-effort-to-improve-nls}{%
\section{\texorpdfstring{Goals and progress of our effort to improve
\texttt{nls()}}{Goals and progress of our effort to improve nls()}}\label{goals-and-progress-of-our-effort-to-improve-nls}}

\hypertarget{code-rationalization-and-documentation}{%
\subsection{Code rationalization and
documentation}\label{code-rationalization-and-documentation}}

We want:

\begin{itemize}
\item
  to provide a packaged version of \texttt{nls()} (call it
  \texttt{nlsalt}) coded entirely in R that matches the features in base
  R or what is packaged in \texttt{nlspkg} as described in John C. Nash
  and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
\item
  to streamline the overall \texttt{nls()} infrastructure. By this we
  mean a re-factoring of the routines so they are better suited to
  maintenance of both the existing \texttt{nls()} methods and features
  as well as new features we or others would like to add.
\item
  to explain what we do, either in comments or separate maintainer
  documentation. Since we are complaining about the lack of explanatory
  material for the current code, we feel it is incumbent on us to
  provide such material for our own work, and if possible for the
  existing code.
\end{itemize}

These goals echo themes in the recent discussion Vidoni
(\protect\hyperlink{ref-Vidoni21}{2021}) and accompanying commentary.

\hypertarget{rationalization-of-formula-specifications}{%
\subsubsection{Rationalization of formula
specifications}\label{rationalization-of-formula-specifications}}

We have seen that \texttt{nls()} uses a different formula specification
from the default if the \texttt{plinear} algorithm is used. This is
unfortunate, since the user cannot then simply add
\texttt{algorithm="plinear"} to the call, making errors more likely.

While not consistent with the \texttt{lm()} specification of linear
models, nonlinear models need all parameters specified. That is, the
model formula should be written down as it would be computed. Thus a
linear model (which is valid input for nonlinear estimatin) should be
specified \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
\texttt{y\ \textasciitilde{}\ x}. Moreover, we want to avoid the
partially linear parameters appearing in the result object as
\texttt{.lin1}, \texttt{.lin2}, etc., where the actual position of
parameters in the model is not explicit.

A possible workaround to allow consistency would be to specify the
partially linear parameters when the \texttt{plinear} option is
specified. For example, we could use \texttt{algorithm="plinear(Asym)"}
while keeping the general model formula from the default specification.
This would allow for the output of different \texttt{algorithm} options
to be consistent. However, we have not yet tried to code such a change,
and welcome discussion from those working with other packages using
model formulas.

A further complication of model specification in R is that some formulas
require the user to surround a term with \texttt{I()} to inhibit
interpretation of arithmetic operators. We would prefer that formulas be
usable without this complication.

\hypertarget{rationalization-of-indexed-models}{%
\subsubsection{Rationalization of indexed
models}\label{rationalization-of-indexed-models}}

Indexed models clearly have a place in some areas of research. We need
to be able to process models such as
\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)}
where \texttt{Strip} is the index. Data \texttt{Length} and
\texttt{Conc} are available for all observations and parameter
\texttt{th} applies to every fitted point, but \texttt{a{[}Strip{]}} and
\texttt{b{[}Strip{]}} depend on the index, which could be an
experimental run number, or some other categorization.

Given that \texttt{{[}\ {]}} are the R method to index arrays, their
presence in a model formula signals indexing. The question is how to
perform the correct calculations.

Our view is that inclusion of the indexing \textbf{within}
\texttt{nls()} introduces complexity that could be avoided by a suitable
wrapper function. The modeling formula must be expanded to encompass all
the index cases, likely with indexed parameters renamed e.g.,
\texttt{a{[}2{]}} becomes \texttt{a2} as at present with \texttt{nls()}.
However, we believe the wrapper should be able to process output to
present the indexed parameters correctly.

\hypertarget{streamlining-code}{%
\subsubsection{Streamlining code}\label{streamlining-code}}

We are pessimistic that the overall structure of \texttt{nls()} can be
streamlined due to the entanglement of so many features with the
complicated mix of R, C and Fortran. Indeed, despite digging into the
code over some months, we do not feel confident that we sufficiently
understand it nor that we could maintain it. On the other hand, we do
believe equivalent functionality can be built, but with at least some
differences in how the features will be accessed.

\hypertarget{tests-and-use-case-examples}{%
\subsection{Tests and use-case
examples}\label{tests-and-use-case-examples}}

Maintainers of packages need suitable tests and use-case examples in
order:

\begin{itemize}
\tightlist
\item
  to ensure packages work properly, in particular, giving results
  comparable to or better than the functions they are to replace;
\item
  to test individual solver functions to ensure they work across the
  range of calling mechanisms, that is, different ways of supplying
  inputs to the solver(s);
\item
  to pose ``silly'' inputs to see if these bad inputs are caught by the
  programs.
\end{itemize}

Such goals align with the aims of \textbf{unit testing} (e.g.,
\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}), as
well as with R package checking tools ?? reference??. ??Other refs.

\hypertarget{a-testing-package-nlscompare}{%
\subsubsection{A testing package:
NLSCompare}\label{a-testing-package-nlscompare}}

One of us has developed a working prototype package at
\url{https://github.com/ArkaB-DS/nlsCompare} . A primary design
objective of this is to allow the summarisation of multiple tests in a
compact output. The prototype has a vignette to illustrate its use.

\hypertarget{documentation-and-resources}{%
\subsection{Documentation and
resources}\label{documentation-and-resources}}

In our investigation, we built several resources, which are now part of
the repository \url{https://github.com/nashjc/RNonlinearLS/}. Particular
items are:

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib})
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}
\item
  John C. Nash and Bhattacharjee
  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
  in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of John C. Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

\hypertarget{strategic-choices-in-nonlinear-model-estimation}{%
\section{Strategic choices in nonlinear model
estimation}\label{strategic-choices-in-nonlinear-model-estimation}}

It is possible that \texttt{nls()} will remain more or less as it has
been for the past two decades. Given its importance to R, a focus of
discussion should be what needs fixing and how that should be
approached, and we welcome an opportunity to participate in such
conversations.

From a design point of view, a key difference in approach between
\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()} builds a
large infrastructure, especially the \texttt{nlsModel()} function, from
which the Gauss-Newton iteration can be executed and other statistical
information such as profiles can be computed, while \texttt{nlxb()}
returns quite limited information, and computes what is needed on an
``as and when'' basis. This follows a path that one of us (JN)
established almost 50 years ago with the software that became John C.
Nash (\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver,
and post-solution analysis phases of computations.

\hypertarget{opinions}{%
\subsection{Opinions}\label{opinions}}

To advance the stability and maintainability of R, we believe the
program objects (R functions) that are created by tools such as
\texttt{nls()} should have minimal interactions and side-effects. The
aspects of \texttt{nls()} that have given us the most trouble:

\begin{itemize}
\item
  The functions that compute the residuals and jacobians often presume
  the data and parameters needed are available in a particular
  environment. As long as the correct environment is used, this provides
  a short syntax to invoke the calculations. The danger is that the
  wrong data is accessed if the internal search finds a valid name that
  is not the object we want.
\item
  Weights, subsets, and various contextual controls such as that for the
  \texttt{na.action} option are similarly taken from the first available
  source. Again, this makes for a very simple invocation of
  calculations, but the context is hidden from the user. Furthermore, it
  can be difficult for those of us trying to maintain or improve the
  code to be certain we have the context correct.
\item
  The mixing of R, C and Fortran code was necessary for computational
  performance in the past. Lacking decent developer documentation,
  programmers now face a lot of work to fix or improve code. We believe
  in having at least a working reference version of code in a single
  programming language. If necessary, by measuring (``profiling'') the
  code, we can find bottlenecks and, if necessary, substitute just those
  slower parts of the reference code.
\item
  A design that isolates the setup, solution, and post-solution parts of
  for complicated calculations reduces the number of objects that must
  be kept in alignment.
\item
  All iterative codes should return the best solution they have found so
  far, even if there are untoward conditions reached, such as a singular
  Jacobian. This modification could be made to \texttt{nls()} in the
  short term.
\item
  Given the existence of examples of good practices such as analytic
  derivatives, stabilized solution of Gauss-Newton equations and bounds
  constrained parameters, base R tools should be moving to incorporate
  them.
\end{itemize}

\hypertarget{acknowledgement}{%
\section{Acknowledgement}\label{acknowledgement}}

Hans Werner Borchers has been helpful in developing the proposal for
this project and in comments on this and related work. Heather Turner
co-mentored the project and helped guide the progress of the work.

\hypertarget{references}{%
\section*{References}\label{references}}
\addcontentsline{toc}{section}{References}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis
and Its Applications}. Wiley.

\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
Bates, Douglas M., and Donald G. Watts. 1981. {``A Relative Offset
Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
\emph{Technometrics} 23 (2): 179--83.

\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004.
\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with
{S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain
Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied
Mathematics} 2: 164-\/-168.

\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation
of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics}
11 (2): 431--41.

\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for
Agricultural Applications}}.
\url{https://CRAN.R-project.org/package=nlraa}.

\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on
a Small Computer.''} \emph{Journal of the Institute for Mathematics and
Its Applications} 19: 231--37.

\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear
Algebra and Function Minimisation}. Bristol: Adam Hilger.

\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
---------. 2022. {``Function Minimization and Nonlinear Least Squares in
r.''} \emph{WIREs Computational Statistics}, no. e1580.
https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package
from Base {R} Files.''} \emph{R-Bloggers}, January.
\url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
Nash, John C, and Duncan Murdoch. 2019. \emph{Nlsr: Functions for
Nonlinear Least Squares Solutions}.

\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation
Combining Visual Fitting with Optimization Methods.''} In
\emph{Proceedings of the Section on Physical and Engineering Sciences of
the American Statistical Association}, 256--61. American Statistical
Association.

\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified
Practical Approach}. New York; Basel: Marcel Dekker Inc.

\leavevmode\vadjust pre{\hypertarget{ref-Vidoni21}{}}%
Vidoni, Melina. 2021. {``{Software Engineering and R Programming: A Call
for Research}.''} \emph{{The R Journal}} 13 (2): 6--14.
\url{https://doi.org/10.32614/RJ-2021-108}.

\end{CSLReferences}

\end{document}

### 2022-06-21 RefactoringNLS/RefactoringNLS.Rmd
bracket mismatch

@@
 
-`nls()` is the tool in base R (the distributed software package from [CRAN](https://cran.r-project.org)
-for estimating nonlinear statistical models. The function dates to the 1980s and the work related
-to @bateswatts in S (see https://en.wikipedia.org/wiki/S_%28programming_language%29).
+`nls()` is the tool in base R, the primary software software distribution
+from the Comprehensive R Archive Network (https://cran.r-project.org)
+for estimating nonlinear statistical models. The function dates to the 1980s and the work
+related to @bateswatts in S (see https://en.wikipedia.org/wiki/S_%28programming_language%29).
 

### 2022-06-21 RefactoringNLS/RefactoringNLS.tex
bracket mismatch

@@
 
-\texttt{nls()} is the tool in base R (the distributed software package
-from \href{https://cran.r-project.org}{CRAN} for estimating nonlinear
-statistical models. The function dates to the 1980s and the work related
-to D. M. Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S
-(see
+\texttt{nls()} is the tool in base R, the primary software software
+distribution from the Comprehensive R Archive Network
+(\url{https://cran.r-project.org}) for estimating nonlinear statistical
+models. The function dates to the 1980s and the work related to D. M.
+Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
 \url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).
@@
 ##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## a                     -3            NA         NA         NA           0       26.49  
-## b                      5            NA         NA         NA           0       9.615  
-## c                    0.5            NA         NA         NA           0       2.466  
-## r                    0.7            NA         NA         NA           0      0.1098  
-## m                      1            NA         NA         NA           0   1.118e-16
+## a                     -3             0       -Inf          0           0       26.49  
+## b                      5             0        Inf          0           0       9.615  
+## c                    0.5             0        Inf          0           0       2.466  
+## r                    0.7             0        Inf          0           0      0.1098  
+## m                      1             0        Inf          0           0   1.949e-09
 \end{verbatim}
@@
 ## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   1.832e-10           0  
+## b2               4.43325            NA         NA         NA   1.015e-10           0  
 ## b3                     3U           NA         NA         NA           0           0

### 2022-06-21 RefactoringNLS/bhobbsXsmall.R
remove temporary files from versioning, add test example to preserve

@@ -1,0 +1,19 @@
+# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+wf <- data.frame(y=weed, tt=tt)
+st <- c(b1=1, b2=1, b3=1) # a default starting vector (named!)
+wmods <- y ~ 100*b1/(1+10*b2*exp(-0.1*b3*tt)) ## Scaled model
+require(minpack.lm)
+require(nlsr)
+# Hobbs scaled problem with bounds, formula specification
+anlxb1 <- nlxb(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+cat("Using nlsr::nlxb():")
+print(anlxb1)
+## nlsLM seems NOT to work properly with bounds
+anlsLM1b <- nlsLM(wmods, start=st, data=wf, lower=c(0,0,0), upper=c(2,6,3))
+cat("\n"); cat("using minpack.lm::nlsLM():")
+print(anlsLM1b)
+anlsb<-nls(wmods, start=st, data=wf, algorithm="port", lower=c(0,0,0), upper=c(2,6,3))
+print(anlsb)

### 2022-06-21 RefactoringNLS/RefactoringNLS.Rmd
minor edits, acknowledge Gillespie

@@
 Such goals align with the aims of **unit testing** 
-(e.g., https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211), as well
-as with R package checking tools ?? reference??. ??Other refs.
+(e.g., https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211, @HWtestthat11, @HWdevtools21
+and the conventional R package testing tools).
 
@@
 
-<!-- ?? Arkajyoti -- do you want to expand? -->
-<!-- Trying to trim a lot of words, so am makign reference to package. Is it ready -->
+<!-- Arkajyoti -- do you want to expand? -->
+<!-- No. I think it would be tangential to the goal of the paper. We can have a -->
+<!-- separate article for that, maybe? -->
+<!-- Trying to trim a lot of words, so am making reference to package. Is it ready -->
 <!-- for others to use? -->
@@
 <!--   the many choices in their implementation for nonlinear least squares. This is still -->
-<!--   a DRAFT at 2021-8-20. ?? fix -->
-
-<!-- ?? Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
+<!--   a DRAFT at 2022-6-20. -->
+
+<!-- Response to query about TestsDoc -->
+<!-- 4. Testing infrastructure, I think, is a general topic and not -->
+<!-- particularly relevant to nls(); i.e., it is important for all packages. -->
+<!-- So, I don't think we need to include that in RefactoringNLS. Could be a -->
+<!-- separate blog maybe?   -->
+
+<!-- Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
 <!-- useful. Perhaps incorporate in the nlsCompare repo. -->
-<!-- - TestsDoc.Rmd (??citation) is a survey of testing tools in R. It has more  -->
+<!-- - TestsDoc.Rmd (?citation) is a survey of testing tools in R. It has more  -->
 <!--   general possibilities  -->
@@
 
-<!-- ?? needs more focus of where it may appear. -->
-<!-- - DerivsNLS.Rmd ??citation: a document to explain different ways in which Jacobian information -->
+<!-- This needs more focus of where it may appear. I'm thinking of a blog post -- JN-->
+<!-- - DerivsNLS.Rmd ?citation: a document to explain different ways in which Jacobian information -->
 <!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->

### 2022-06-21 RefactoringNLS/RefactoringNLS.tex
minor edits, acknowledge Gillespie

@@
 ## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   1.015e-10           0  
+## b2               4.43325            NA         NA         NA   1.583e-10           0  
 ## b3                     3U           NA         NA         NA           0           0
@@
 Such goals align with the aims of \textbf{unit testing} (e.g.,
-\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}), as
-well as with R package checking tools ?? reference??. ??Other refs.
+\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211},
+Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
+(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
+package testing tools).
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
+Wickham, Hadley. 2011. {``Testthat: Get Started with Testing.''}
+\emph{The R Journal} 3: 5--10.
+\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
+Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
+\emph{Devtools: Tools to Make Developing r Packages Easier}.
+\url{https://CRAN.R-project.org/package=devtools}.
+
 \end{CSLReferences}

### 2022-07-22 RefactoringNLS/RefactoringNLS.Rmd
minor edit to termination criteria
edits to recognize developments July 2022

@@
 
-An important consideration is the  maintainability of
+Important considerations are the usability and  maintainability of
 the code base that provides the functionality `nls()` claims to offer.
@@
 
-- handles formulas that include R functions
-- allows data to be subset
+- handles formulas that include R functions, even ones which call calculations in
+  other programming languages
+- allows data to be weighted or subset
+- can estimate bound constrained parameters
+- provides a mechanism for handling partially linear models
 - permits parameters to be indexed over a set of related data
@@
 
-This amounts to admitting R cannot solve well-posed problems unless data is polluted
+This amounted to admitting R cannot solve well-posed problems unless data is polluted
 with errors.
@@
 
-The single convergence criterion of `nls()` leaves out some possibilities that could
+The convergence criterion above, the principal termination control of `nls()`, leaves out 
+some possibilities that could
 be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
@@
 
-Let us consider an 'singular gradient' example 
+Let us consider a 'singular gradient' example 
 *https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
-`nlsr::nlxb()` displays the singular values of the Jacobian, so the `nls()` error message is correct,
-but unhelpful in obtaining a solution.
+`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that the `nls()` error message is correct,
+but this is not helpful in obtaining a solution.
 
@@
 The Jacobian is essentially singular as shown by its singular
-values. Note that these are **displayed** in a single column in
+values. Note that these are **displayed** by package `nlsr` in a single column in
 the output to provide a compact layout, but the values do **NOT** correspond to 
@@
 
-The Jacobian used by `nlsr::nlxb()` is computed from analytic expressions, while that
-of `nls()` (and, by extension, `minpack.lm::nlsLM`) mostly use a numerical approximation. 
-This can lead to different behaviour:
+As far as we can understand the logic in `nls()`, the Jacobian computation during
+parameter estimation is carried out entirely within the called C-language program,
+and the R code `function numericDeriv()`, part of `./src/library/stats/R/nls.R`
+in the R distribution source code, is used to provide Jacobian information in 
+the `nlsModel()` and `nlsModel.plinear()` functions.
+
+The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, while that
+of `nls()` mostly uses a numerical approximation. Some selfStart models do provide
+"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
+version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from the 
+original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
+can lead to different behaviour, though in our experience, one has to look for the
+differences, as they are generally minor.
 
@@
 
-Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file
+Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file may
 provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
@@
 defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
-is within the `nlsModel()` or `nlsModel.pliner()` functions. The use of analytic Jacobians
+is within the `nlsModel()` or `nlsModel.plinear()` functions. The use of analytic Jacobians
 almost certainly contributes to the good performance of `nls()` on selfStart models.
+
+## Feature: Weights on observations
+
+?? does resid return weighted or unweighted residuals
+
+?? do we want unweighted (raw) residuals? or are predictions / fits just as good?
 
@@
 
-## Feature: sources of data
+## Feature: Sources of data
 
@@
 with the particulars of the `nls()` code. Though there is a `getInitial()` function,
-this is not easy to use to simply compute the initial parameter estimates, in part 
-because it may call `nls()`. Such circular references are, in our view, dangerous.
+this is not easy to use to simply compute the initial parameter estimates outside of
+`nls()`, in part 
+because it may call that function. Such circular references are, in our view, dangerous.
+Moreover, we believe that it would be helpful to have selfStart models that allow
+users to explicitly provide the starting parameters.
 
@@
 
+We also note that the introduction of `scaleOffset` in R 4.1.1 to deal with the 
+convergence test for small residual problems now requires that the `getInitial()`
+function have dot-arguments (`...`) in its argument list. This illustrates the
+entanglement of many features in `nls()`.
+
 ## Issue: returned output of nls() and its documentation

### 2022-07-22 RefactoringNLS/RefactoringNLS.tex
edits to recognize developments July 2022

@@
 
-An important consideration is the maintainability of the code base that
-provides the functionality \texttt{nls()} claims to offer. Our work
-suggests that the existing code makes maintenance and improvement very
-difficult, with legacy applications and examples blocking some important
-updates. Discussion of these matters is relevant to improving R
-generally, as well as its nonlinear estimation tools.
+Important considerations are the usability and maintainability of the
+code base that provides the functionality \texttt{nls()} claims to
+offer. Our work suggests that the existing code makes maintenance and
+improvement very difficult, with legacy applications and examples
+blocking some important updates. Discussion of these matters is relevant
+to improving R generally, as well as its nonlinear estimation tools.
 
@@
 \item
-  handles formulas that include R functions
-\item
-  allows data to be subset
+  handles formulas that include R functions, even ones which call
+  calculations in other programming languages
+\item
+  allows data to be weighted or subset
+\item
+  can estimate bound constrained parameters
+\item
+  provides a mechanism for handling partially linear models
 \item
@@
 
-This amounts to admitting R cannot solve well-posed problems unless data
-is polluted with errors.
+This amounted to admitting R cannot solve well-posed problems unless
+data is polluted with errors.
 
@@
 
-The single convergence criterion of \texttt{nls()} leaves out some
-possibilities that could be useful for some problems. The package
-\texttt{nlsr} (John C. Nash and Murdoch
+The convergence criterion above, the principal termination control of
+\texttt{nls()}, leaves out some possibilities that could be useful for
+some problems. The package \texttt{nlsr} (John C. Nash and Murdoch
 (\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
@@
 
-Let us consider an `singular gradient' example
+Let us consider a `singular gradient' example
 \emph{\url{https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values}}.
-\texttt{nlsr::nlxb()} displays the singular values of the Jacobian, so
-the \texttt{nls()} error message is correct, but unhelpful in obtaining
-a solution.
+\texttt{nlsr::nlxb()} displays the singular values of the Jacobian,
+which confirm that the \texttt{nls()} error message is correct, but this
+is not helpful in obtaining a solution.
 
@@
 ##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## a                     -3             0       -Inf          0           0       26.49  
-## b                      5             0        Inf          0           0       9.615  
-## c                    0.5             0        Inf          0           0       2.466  
-## r                    0.7             0        Inf          0           0      0.1098  
-## m                      1             0        Inf          0           0   1.949e-09
+## a                     -3            NA         NA         NA           0       26.49  
+## b                      5            NA         NA         NA           0       9.615  
+## c                    0.5            NA         NA         NA           0       2.466  
+## r                    0.7            NA         NA         NA           0      0.1098  
+## m                      1            NA         NA         NA           0   1.398e-16
 \end{verbatim}
@@
 The Jacobian is essentially singular as shown by its singular values.
-Note that these are \textbf{displayed} in a single column in the output
-to provide a compact layout, but the values do \textbf{NOT} correspond
-to the individual parameters in whose row they appear; they are a
-property of the whole problem.
+Note that these are \textbf{displayed} by package \texttt{nlsr} in a
+single column in the output to provide a compact layout, but the values
+do \textbf{NOT} correspond to the individual parameters in whose row
+they appear; they are a property of the whole problem.
 
@@
 
-The Jacobian used by \texttt{nlsr::nlxb()} is computed from analytic
-expressions, while that of \texttt{nls()} (and, by extension,
-\texttt{minpack.lm::nlsLM}) mostly use a numerical approximation. This
-can lead to different behaviour:
+As far as we can understand the logic in \texttt{nls()}, the Jacobian
+computation during parameter estimation is carried out entirely within
+the called C-language program, and the R code
+\texttt{function\ numericDeriv()}, part of
+\texttt{./src/library/stats/R/nls.R} in the R distribution source code,
+is used to provide Jacobian information in the \texttt{nlsModel()} and
+\texttt{nlsModel.plinear()} functions.
+
+The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from
+analytic expressions, while that of \texttt{nls()} mostly uses a
+numerical approximation. Some selfStart models do provide ``gradient''
+information. \texttt{minpack.lm::nlsLM()} invokes
+\texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
+it appears to use an internal approximate Jacobian code from the
+original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such
+differences in approach can lead to different behaviour, though in our
+experience, one has to look for the differences, as they are generally
+minor.
 
@@
 Note that the selfStart models in the base R
-\texttt{./src/library/stats/R/zzModels.R} file provide the Jacobian in
-the ``gradient'' attribute of the ``one-sided'' formula that defines
+\texttt{./src/library/stats/R/zzModels.R} file may provide the Jacobian
+in the ``gradient'' attribute of the ``one-sided'' formula that defines
 each model, and these Jacobians may be the analytic forms. The
@@
 a Jacobian into that attribute. This code is within the
-\texttt{nlsModel()} or \texttt{nlsModel.pliner()} functions. The use of
+\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
 analytic Jacobians almost certainly contributes to the good performance
 of \texttt{nls()} on selfStart models.
+
+\hypertarget{feature-weights-on-observations}{%
+\subsection{Feature: Weights on
+observations}\label{feature-weights-on-observations}}
+
+?? does resid return weighted or unweighted residuals
+
+?? do we want unweighted (raw) residuals? or are predictions / fits just
+as good?
 
@@
 \hypertarget{feature-sources-of-data}{%
-\subsection{Feature: sources of data}\label{feature-sources-of-data}}
+\subsection{Feature: Sources of data}\label{feature-sources-of-data}}
 
@@
 is a \texttt{getInitial()} function, this is not easy to use to simply
-compute the initial parameter estimates, in part because it may call
-\texttt{nls()}. Such circular references are, in our view, dangerous.
+compute the initial parameter estimates outside of \texttt{nls()}, in
+part because it may call that function. Such circular references are, in
+our view, dangerous. Moreover, we believe that it would be helpful to
+have selfStart models that allow users to explicitly provide the
+starting parameters.
 
@@
 considering.
+
+We also note that the introduction of \texttt{scaleOffset} in R 4.1.1 to
+deal with the convergence test for small residual problems now requires
+that the \texttt{getInitial()} function have dot-arguments
+(\texttt{...}) in its argument list. This illustrates the entanglement
+of many features in \texttt{nls()}.
 
@@
 ## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   1.583e-10           0  
+## b2               4.43325            NA         NA         NA   1.832e-10           0  
 ## b3                     3U           NA         NA         NA           0           0

### 2022-07-24 RefactoringNLS/RefactoringNLS.Rmd
Working on weights -- seems to be a bug in nlxb/nlfb for weights

@@
 
+                  -----------------------------------------------------------
+
 <!-- ### Example output -->
@@
 The `nls` object contains some elements that are awkward to produce by other algorithms,
-but some information that would be useful is not presented obviously. Moreover, the
+but some information that would be useful is not presented in a clear manner. Moreover, the
 complexity of the object is a challenge to users.
@@
 
+<!-- nlsr2 has some extra info ?? -->
+
 ### Weights in returned functions from nls()
@@
 
+SHORT EXAMPLE??
+
+```{r nlswtx, echo=TRUE}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal))
+nowt
+nowt$m$resid()
+inpar<-getInitial(weed ~ SSlogis(tt, Asym, xmid, scal), data=weeddf)
+print(inpar)
+library(nlsr2)
+nowtx<-nlxb(frmlogis, start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
+nowtx
+nowtxi<-nlxb(frmlogis, start=inpar, trace=TRUE)
+nowtxi
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+usewt$m$resid()
+nmod <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+usewtx<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts,
+              control=list(japprox="jafwd"), trace=TRUE)
+usewtx
+usewtxi<-nlxb(frmlogis, start=inpar, weights=wts,
+              weights=wts, control=list(japprox="jafwd"), trace=TRUE)
+usewtxi
+# usewtx$resid
+```
+
+?? NOTE -- nlxb NOT handling weights correctly here
 ### Interim output from the "port" algorithm

### 2022-07-24 RefactoringNLS/RefactoringNLS.tex
Working on weights -- seems to be a bug in nlxb/nlfb for weights

@@
 
+\begin{verbatim}
+              -----------------------------------------------------------
+\end{verbatim}
+
 The \texttt{nls} object contains some elements that are awkward to
 produce by other algorithms, but some information that would be useful
-is not presented obviously. Moreover, the complexity of the object is a
-challenge to users.
+is not presented in a clear manner. Moreover, the complexity of the
+object is a challenge to users.
 
@@
 intended.
+
+SHORT EXAMPLE??
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
+          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
+\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
+\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
+\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
+\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
+\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal))}
+\NormalTok{nowt}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Nonlinear regression model
+##   model: weed ~ SSlogis(tt, Asym, xmid, scal)
+##    data: parent.frame()
+##   Asym   xmid   scal 
+## 196.19  12.42   3.19 
+##  residual sum-of-squares: 2.59
+## 
+## Number of iterations to convergence: 0 
+## Achieved convergence tolerance: 1.13e-06
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
+##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
+## attr(,"gradient")
+##           Asym     xmid    scal
+##  [1,] 0.027117  -1.6229  5.8103
+##  [2,] 0.036737  -2.1769  7.1111
+##  [3,] 0.049596  -2.8997  8.5628
+##  [4,] 0.066645  -3.8266 10.1000
+##  [5,] 0.089005  -4.9881 11.6015
+##  [6,] 0.117921  -6.3988 12.8762
+##  [7,] 0.154635  -8.0418 13.6607
+##  [8,] 0.200186  -9.8498 13.6432
+##  [9,] 0.255106 -11.6901 12.5267
+## [10,] 0.319083 -13.3660 10.1313
+## [11,] 0.390688 -14.6444  6.5083
+## [12,] 0.467334 -15.3139  2.0038
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{inpar}\OtherTok{\textless{}{-}}\FunctionTok{getInitial}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{data=}\NormalTok{weeddf)}
+\FunctionTok{print}\NormalTok{(inpar)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##     Asym     xmid     scal 
+## 196.1862  12.4173   3.1891
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{library}\NormalTok{(nlsr2)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 
+## Attaching package: 'nlsr2'
+\end{verbatim}
+
+\begin{verbatim}
+## The following objects are masked from 'package:nlsr':
+## 
+##     codeDeriv, findSubexprs, fnDeriv, isCALL, isMINUSONE, isONE,
+##     isZERO, model2rjfun, model2ssgrfun, modelexpr, newDeriv,
+##     newSimplification, nlfb, nlsDeriv, nlsSimplify, nlxb, resgr, resss,
+##     rjfundoc, sysDerivs, sysSimplifications, wrapnlsr
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nowtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{control=}\FunctionTok{list}\NormalTok{(}\AttributeTok{japprox=}\StringTok{"jafwd"}\NormalTok{), }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Using approximation  jafwd   is.character(ctrl$japprox)= TRUE 
+## <<lamda: 1e-04  SS= 23527 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
+## <<lamda: 4e-05  SS= 10887 ( 0.8264 ) at  Asym = 52.337  xmid = 48.234  scal = 180.33 f/j  2 / 1
+## <<lamda: 0.016  SS= 8948.4 ( 0.41672 ) at  Asym = 80.893  xmid = 48.394  scal = 169.52 f/j  6 / 2
+## <<lamda: 0.0064  SS= 8853.5 ( 0.068832 ) at  Asym = 89.895  xmid = 66.542  scal = 118.27 f/j  7 / 3
+## <<lamda: 0.0256  SS= 8673.7 ( 0.080434 ) at  Asym = 103.26  xmid = 69.707  scal = 91.474 f/j  9 / 4
+## <<lamda: 0.1024  SS= 8563.7 ( 0.059227 ) at  Asym = 110.79  xmid = 67.869  scal = 81.345 f/j  11 / 5
+## <<lamda: 0.04096  SS= 8392.1 ( 0.12852 ) at  Asym = 129.64  xmid = 62.354  scal = 48.414 f/j  12 / 6
+## <<lamda: 0.16384  SS= 7777.5 ( 0.17269 ) at  Asym = 139.58  xmid = 48.216  scal = 39.581 f/j  14 / 7
+## <<lamda: 0.65536  SS= 7547.4 ( 0.11307 ) at  Asym = 142.31  xmid = 43.841  scal = 33.749 f/j  16 / 8
+## <<lamda: 0.26214  SS= 6213 ( 0.22183 ) at  Asym = 149.97  xmid = 28.124  scal = 17.452 f/j  17 / 9
+## <<lamda: 1.0486  SS= 2919.7 ( 0.33273 ) at  Asym = 152.81  xmid = 15.369  scal = 7.5443 f/j  19 / 10
+## <<lamda: 4.1943  SS= 1903.2 ( 0.74156 ) at  Asym = 153.41  xmid = 9.8394  scal = 1.8083 f/j  21 / 11
+## <<lamda: 1.6777  SS= 113.84 ( 0.73921 ) at  Asym = 153.46  xmid = 10.631  scal = 2.5864 f/j  22 / 12
+## <<lamda: 0.67109  SS= 11.222 ( 0.66287 ) at  Asym = 153.64  xmid = 10.972  scal = 2.8505 f/j  23 / 13
+## <<lamda: 0.26844  SS= 10.289 ( 0.19346 ) at  Asym = 154.03  xmid = 11.02  scal = 2.8798 f/j  24 / 14
+## <<lamda: 0.10737  SS= 9.8337 ( 0.11332 ) at  Asym = 154.95  xmid = 11.055  scal = 2.889 f/j  25 / 15
+## <<lamda: 0.04295  SS= 8.8717 ( 0.16811 ) at  Asym = 157.04  xmid = 11.133  scal = 2.9084 f/j  26 / 16
+## <<lamda: 0.01718  SS= 7.2069 ( 0.21397 ) at  Asym = 161.34  xmid = 11.29  scal = 2.9469 f/j  27 / 17
+## <<lamda: 0.0068719  SS= 5.1659 ( 0.30108 ) at  Asym = 168.61  xmid = 11.548  scal = 3.0079 f/j  28 / 18
+## <<lamda: 0.0027488  SS= 3.5835 ( 0.33934 ) at  Asym = 178.1  xmid = 11.867  scal = 3.079 f/j  29 / 19
+## <<lamda: 0.0010995  SS= 2.8237 ( 0.28331 ) at  Asym = 187.23  xmid = 12.155  scal = 3.1385 f/j  30 / 20
+## <<lamda: 0.0004398  SS= 2.6123 ( 0.16163 ) at  Asym = 193.29  xmid = 12.334  scal = 3.1734 f/j  31 / 21
+## <<lamda: 0.00017592  SS= 2.588 ( 0.056549 ) at  Asym = 195.68  xmid = 12.403  scal = 3.1863 f/j  32 / 22
+## <<lamda: 7.0369e-05  SS= 2.5873 ( 0.010514 ) at  Asym = 196.14  xmid = 12.416  scal = 3.1889 f/j  33 / 23
+## <<lamda: 2.8147e-05  SS= 2.5873 ( 0.0009007 ) at  Asym = 196.18  xmid = 12.417  scal = 3.1891 f/j  34 / 24
+## <<lamda: 1.1259e-05  SS= 2.5873 ( 3.4226e-05 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  35 / 25
+## <<lamda: 4.5036e-06  SS= 2.5873 ( 6.5424e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  36 / 26
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nowtx}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## nlsr2  object: x 
+## residual sumsquares =  2.5873  on  12 observations
+##     after  27    Jacobian and  36 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## Asym             196.186         11.31      17.35  3.167e-08  -5.714e-10       44.93  
+## xmid             12.4173        0.3346      37.11  3.716e-11  -6.654e-08        15.6  
+## scal             3.18908        0.0698      45.69  5.768e-12   1.567e-07      0.0474
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
+\NormalTok{usewt}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Nonlinear regression model
+##   model: weed ~ SSlogis(tt, Asym, xmid, scal)
+##    data: parent.frame()
+##   Asym   xmid   scal 
+## 199.86  12.50   3.19 
+##  weighted residual sum-of-squares: 0.0196
+## 
+## Number of iterations to convergence: 2 
+## Achieved convergence tolerance: 8.29e-06
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+## attr(,"gradient")
+##           Asym     xmid    scal
+##  [1,] 0.026498  -1.6157  5.8228
+##  [2,] 0.035901  -2.1679  7.1333
+##  [3,] 0.048474  -2.8890  8.6006
+##  [4,] 0.065153  -3.8149 10.1617
+##  [5,] 0.087046  -4.9775 11.6983
+##  [6,] 0.115387  -6.3933 13.0222
+##  [7,] 0.151425  -8.0483 13.8709
+##  [8,] 0.196222  -9.8787 13.9297
+##  [9,] 0.250362 -11.7553 12.8918
+## [10,] 0.313611 -13.4827 10.5608
+## [11,] 0.384641 -14.8251  6.9663
+## [12,] 0.460954 -15.5631  2.4357
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# usewtx\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym, xmid, scal), start=inpar, weights=wts,}
+\CommentTok{\#              control=list(japprox="jafwd"), trace=TRUE)}
+\CommentTok{\# usewtx}
+\CommentTok{\# usewtx$resid}
+\end{Highlighting}
+\end{Shaded}
 
@@
 \begin{verbatim}
-## nlsr object: x 
+## nlsr2  object: anlxb1 
 ## residual sumsquares =  9.4726  on  12 observations
-##     after  16    Jacobian and  17 function evaluations
+##     after  12    Jacobian and  12 function evaluations
 ##   name            coeff          SE       tstat      pval      gradient    JSingval   
 ## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   1.832e-10           0  
+## b2               4.43325            NA         NA         NA   -2.16e-07           0  
 ## b3                     3U           NA         NA         NA           0           0

### 2022-07-24 RefactoringNLS/RefactoringNLS.Rmd
More corrections

--- 2022-07-22 RefactoringNLS/RefactoringNLS.Rmd
@@
 ```
+
+<!-- ?? nlme::gnls() -- should we note it? -->
 
@@
 intended.
+
+SHORT EXAMPLE??
+
+```{r nlswtx, echo=TRUE}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal))
+nowt
+nowt$m$resid()
+inpar<-getInitial(weed ~ SSlogis(tt, Asym, xmid, scal), data=weeddf)
+print(inpar)
+library(nlsr2)
+nowtx<-nlxb(frmlogis, start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
+nowtx
+# nowtSSx0<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
+# nowtSsx0
+# nowtSSx<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=100, xmid=2, scal=1), trace=TRUE, 
+#               control=list(japprox="jand"))
+# nowtSSx
+nowtxi<-nlxb(frmlogis, start=inpar, trace=TRUE)
+nowtxi
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+usewt$m$resid()
+nmod <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+usewtx<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts, trace=TRUE)
+usewtx
+# NOTE jafwd and jaback fail!
+usewtxn<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts, trace=TRUE, control=list(japprox="jacentral"))
+usewtxn
+usewtxi<-nlxb(frmlogis, start=inpar, weights=wts, trace=TRUE)
+usewtxi
+# usewtx$resid
+```
 
@@
 
-One obvious case is that nonlinear modeling tools are a superset of linear ones.
-Yet the explicit model `y ~ a * x + b` does not work with the linear modeling 
+The nonlinear modeling model specifications are, of course, a development of linear ones.
+Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
 function `lm()`, which requires this model to be specified as `y ~ x`.
 
-Within `nls()`, consider the 
-following FOUR different calling sequences for the same problem, though an intuitive choice,
-labelled fm2a, will not work. In this failed
+Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
+an intuitive choice, labelled fm2a, that does not work. In this failed
 attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm

### 2022-07-24 RefactoringNLS/RefactoringNLS.tex
More corrections

@@
 \NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
+\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
 \NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
@@
 \begin{Highlighting}[]
-\NormalTok{nowtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{control=}\FunctionTok{list}\NormalTok{(}\AttributeTok{japprox=}\StringTok{"jafwd"}\NormalTok{), }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using approximation  jafwd   is.character(ctrl$japprox)= TRUE 
+\NormalTok{nowtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
 ## <<lamda: 1e-04  SS= 23527 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
@@
 ## <<lamda: 0.1024  SS= 8563.7 ( 0.059227 ) at  Asym = 110.79  xmid = 67.869  scal = 81.345 f/j  11 / 5
-## <<lamda: 0.04096  SS= 8392.1 ( 0.12852 ) at  Asym = 129.64  xmid = 62.354  scal = 48.414 f/j  12 / 6
+## <<lamda: 0.04096  SS= 8392.1 ( 0.12852 ) at  Asym = 129.64  xmid = 62.354  scal = 48.413 f/j  12 / 6
 ## <<lamda: 0.16384  SS= 7777.5 ( 0.17269 ) at  Asym = 139.58  xmid = 48.216  scal = 39.581 f/j  14 / 7
-## <<lamda: 0.65536  SS= 7547.4 ( 0.11307 ) at  Asym = 142.31  xmid = 43.841  scal = 33.749 f/j  16 / 8
-## <<lamda: 0.26214  SS= 6213 ( 0.22183 ) at  Asym = 149.97  xmid = 28.124  scal = 17.452 f/j  17 / 9
-## <<lamda: 1.0486  SS= 2919.7 ( 0.33273 ) at  Asym = 152.81  xmid = 15.369  scal = 7.5443 f/j  19 / 10
-## <<lamda: 4.1943  SS= 1903.2 ( 0.74156 ) at  Asym = 153.41  xmid = 9.8394  scal = 1.8083 f/j  21 / 11
-## <<lamda: 1.6777  SS= 113.84 ( 0.73921 ) at  Asym = 153.46  xmid = 10.631  scal = 2.5864 f/j  22 / 12
-## <<lamda: 0.67109  SS= 11.222 ( 0.66287 ) at  Asym = 153.64  xmid = 10.972  scal = 2.8505 f/j  23 / 13
-## <<lamda: 0.26844  SS= 10.289 ( 0.19346 ) at  Asym = 154.03  xmid = 11.02  scal = 2.8798 f/j  24 / 14
-## <<lamda: 0.10737  SS= 9.8337 ( 0.11332 ) at  Asym = 154.95  xmid = 11.055  scal = 2.889 f/j  25 / 15
+## <<lamda: 0.65536  SS= 7547.4 ( 0.11307 ) at  Asym = 142.31  xmid = 43.841  scal = 33.748 f/j  16 / 8
+## <<lamda: 0.26214  SS= 6212.9 ( 0.22183 ) at  Asym = 149.97  xmid = 28.123  scal = 17.451 f/j  17 / 9
+## <<lamda: 1.0486  SS= 2919.3 ( 0.33274 ) at  Asym = 152.81  xmid = 15.368  scal = 7.5437 f/j  19 / 10
+## <<lamda: 4.1943  SS= 1904 ( 0.74162 ) at  Asym = 153.41  xmid = 9.8392  scal = 1.8081 f/j  21 / 11
+## <<lamda: 1.6777  SS= 113.91 ( 0.73923 ) at  Asym = 153.46  xmid = 10.631  scal = 2.5863 f/j  22 / 12
+## <<lamda: 0.67109  SS= 11.223 ( 0.6629 ) at  Asym = 153.64  xmid = 10.972  scal = 2.8504 f/j  23 / 13
+## <<lamda: 0.26844  SS= 10.289 ( 0.19356 ) at  Asym = 154.03  xmid = 11.02  scal = 2.8798 f/j  24 / 14
+## <<lamda: 0.10737  SS= 9.8336 ( 0.11332 ) at  Asym = 154.95  xmid = 11.055  scal = 2.889 f/j  25 / 15
 ## <<lamda: 0.04295  SS= 8.8717 ( 0.16811 ) at  Asym = 157.04  xmid = 11.133  scal = 2.9084 f/j  26 / 16
@@
 ## <<lamda: 7.0369e-05  SS= 2.5873 ( 0.010514 ) at  Asym = 196.14  xmid = 12.416  scal = 3.1889 f/j  33 / 23
-## <<lamda: 2.8147e-05  SS= 2.5873 ( 0.0009007 ) at  Asym = 196.18  xmid = 12.417  scal = 3.1891 f/j  34 / 24
-## <<lamda: 1.1259e-05  SS= 2.5873 ( 3.4226e-05 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  35 / 25
-## <<lamda: 4.5036e-06  SS= 2.5873 ( 6.5424e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  36 / 26
+## <<lamda: 2.8147e-05  SS= 2.5873 ( 0.00090069 ) at  Asym = 196.18  xmid = 12.417  scal = 3.1891 f/j  34 / 24
+## <<lamda: 1.1259e-05  SS= 2.5873 ( 3.4237e-05 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  35 / 25
+## <<lamda: 4.5036e-06  SS= 2.5873 ( 6.5557e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  36 / 26
 \end{verbatim}
@@
 ##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             196.186         11.31      17.35  3.167e-08  -5.714e-10       44.93  
-## xmid             12.4173        0.3346      37.11  3.716e-11  -6.654e-08        15.6  
-## scal             3.18908        0.0698      45.69  5.768e-12   1.567e-07      0.0474
+## Asym             196.186         11.31      17.35  3.167e-08  -3.432e-10       44.93  
+## xmid             12.4173        0.3346      37.11  3.716e-11  -5.273e-08        15.6  
+## scal             3.18908        0.0698      45.69  5.768e-12   1.752e-07      0.0474
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# nowtSSx0\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=1, xmid=1, scal=1), trace=TRUE)}
+\CommentTok{\# nowtSsx0}
+\CommentTok{\# nowtSSx\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=100, xmid=2, scal=1), trace=TRUE, }
+\CommentTok{\#               control=list(japprox="jand"))}
+\CommentTok{\# nowtSSx}
+\NormalTok{nowtxi}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\NormalTok{inpar, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## <<lamda: 1e-04  SS= 2.5873 ( NA ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  1 / 0
+## <<lamda: 4e-05  SS= 2.5873 ( 9.4859e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  2 / 1
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nowtxi}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## nlsr2  object: x 
+## residual sumsquares =  2.5873  on  12 observations
+##     after  2    Jacobian and  2 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## Asym             196.186         11.31      17.35  3.167e-08  -1.769e-09       44.93  
+## xmid             12.4173        0.3346      37.11  3.716e-11  -2.902e-08        15.6  
+## scal             3.18908        0.0698      45.69  5.768e-12   3.896e-08      0.0474
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\CommentTok{\# usewtx\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym, xmid, scal), start=inpar, weights=wts,}
-\CommentTok{\#              control=list(japprox="jafwd"), trace=TRUE)}
-\CommentTok{\# usewtx}
+\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
+\NormalTok{usewtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis,  }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
+## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
+## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
+## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
+## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
+## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
+## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
+## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
+## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
+## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
+## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
+## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
+## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
+## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
+## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
+## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
+## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
+## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
+## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
+## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
+## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
+## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
+## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
+## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
+## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.2336e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewtx}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## nlsr2  object: x 
+## residual sumsquares =  0.019617  on  12 observations
+##     after  25    Jacobian and  31 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## Asym             199.859         17.46      11.45  1.149e-06  -2.201e-11       7.858  
+## xmid             12.4994        0.4131      30.26    2.3e-10    2.22e-09        0.79  
+## scal             3.19088       0.04323      73.81  7.781e-14  -1.511e-08    0.002673
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# }\AlertTok{NOTE}\CommentTok{ jafwd and jaback fail!}
+\NormalTok{usewtxn}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis,  }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{, }\AttributeTok{control=}\FunctionTok{list}\NormalTok{(}\AttributeTok{japprox=}\StringTok{"jacentral"}\NormalTok{))}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Using approximation  jacentral   is.character(ctrl$japprox)= TRUE 
+## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
+## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
+## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
+## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
+## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
+## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
+## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
+## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
+## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
+## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
+## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
+## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
+## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
+## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
+## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
+## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
+## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
+## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
+## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
+## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
+## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
+## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
+## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
+## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
+## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.24e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewtxn}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## nlsr2  object: x 
+## residual sumsquares =  0.019617  on  12 observations
+##     after  25    Jacobian and  31 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## Asym             199.859         17.46      11.45  1.149e-06  -2.105e-11       7.858  
+## xmid             12.4994        0.4131      30.26    2.3e-10   2.212e-09        0.79  
+## scal             3.19088       0.04323      73.81  7.781e-14  -1.514e-08    0.002673
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewtxi}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\NormalTok{inpar, }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## <<lamda: 1e-04  SS= 0.020914 ( NA ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  1 / 0
+## <<lamda: 4e-05  SS= 0.019704 ( 0.028209 ) at  Asym = 196.45  xmid = 12.419  scal = 3.1833 f/j  2 / 1
+## <<lamda: 1.6e-05  SS= 0.019678 ( 0.0029702 ) at  Asym = 196.99  xmid = 12.432  scal = 3.1845 f/j  3 / 2
+## <<lamda: 6.4e-06  SS= 0.019645 ( 0.0031987 ) at  Asym = 197.91  xmid = 12.453  scal = 3.1866 f/j  4 / 3
+## <<lamda: 2.56e-06  SS= 0.019623 ( 0.0031067 ) at  Asym = 198.96  xmid = 12.478  scal = 3.1889 f/j  5 / 4
+## <<lamda: 1.024e-06  SS= 0.019617 ( 0.0017597 ) at  Asym = 199.63  xmid = 12.494  scal = 3.1904 f/j  6 / 5
+## <<lamda: 4.096e-07  SS= 0.019617 ( 0.00050079 ) at  Asym = 199.83  xmid = 12.499  scal = 3.1908 f/j  7 / 6
+## <<lamda: 1.6384e-07  SS= 0.019617 ( 6.2448e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  8 / 7
+## <<lamda: 6.5536e-08  SS= 0.019617 ( 2.9857e-06 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  9 / 8
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{usewtxi}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## nlsr2  object: x 
+## residual sumsquares =  0.019617  on  12 observations
+##     after  9    Jacobian and  9 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## Asym             199.859         17.46      11.45  1.149e-06  -2.056e-10       7.858  
+## xmid             12.4994        0.4131      30.26    2.3e-10   9.074e-09        0.79  
+## scal             3.19088       0.04323      73.81  7.781e-14  -5.703e-08    0.002673
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{\# usewtx$resid}

### 2022-07-24 RefactoringNLS/RefactoringNLS.Rmd
more fixes

@@
 In particular, the iterative setup is such that introduction of Marquardt stabilization
-is not easily available. To obtain performance a lot of the code was in C
-with consequent calls and returns that
-complicate the code. Over time, R has become much more efficient on modern computers, and
+is not easily available. To obtain performance when the code was written, a large portion
+was written in C with consequent calls and returns that
+severely complicate the code. Over time, R has become much more efficient on modern computers, and
 the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
@@
 While R and its packages are generally very well documented for usage, that documentation
-rarely extends as far as it might in directions helpful for code maintenance. 
+rarely extends as far as it might for code maintenance. 
 The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
@@
 
-For `nls()` the concern is demonstrated by a short email query from John Nash
-to Doug Bates and his reply.
-This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on 
-how difficult it is to develop code in this subject area and to keep it maintainable. 
-We have experienced similar loss of understanding for some of our own codes.
-
-```
-    ...
-    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
-    ... has the test problem and the expanded code. Around line 367 is where we are
-    scratching our heads. The function code (from nlsModel()) is in the commented 
-    lines below the call. This is
-
-          # > setPars
-          # function(newPars) {
-          #   setPars(newPars)
-          #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
-          #   dev   <<- sum(resid^2) # envir = thisEnv
-          #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
-          #   QR <<- qr(.swts * gr) # envir = thisEnv
-          #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
-          # }
-
-    I'm anticipating that we will be able to set up a (possibly inefficient) code
-    with documentation that will be easier to follow and test, then gradually figure
-    out how to make it more efficient.
-
-    The equivalent from minpack.lm is
-
-    setPars = function(newPars) {
-                setPars(newPars)
-                assign("resid", .swts * (lhs - assign("rhs", getRHS(),
-                    envir = thisEnv)), envir = thisEnv)
-                assign("dev", sum(resid^2), envir = thisEnv)
-                assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv)
-                return(QR$rank < min(dim(QR$qr)))
-            }
-
-    In both there is the recursive call, which must have a purpose I don't understand.
-```
-The reply was
-
-```
-    I'm afraid that I don't know the purpose of the recursive call either.  
-    I know that I wrote the code to use a closure for the response, covariates, etc., 
-    but I don't recall anything like a recursive call being necessary.
-    ...
-```
-<!--     I don't think I will be of much help.  My R skills have atrophied to the point  -->
-<!--     where I wouldn't even know how to start exploring what is happening in the first  -->
-<!--     call as opposed to the recursive call. -->
-<!-- ``` -->
+<!-- For `nls()` the concern is demonstrated by a short email query from John Nash -->
+<!-- to Doug Bates and his reply. -->
+<!-- This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on  -->
+<!-- how difficult it is to develop code in this subject area and to keep it maintainable.  -->
+<!-- We have experienced similar loss of understanding for some of our own codes. -->
+
+<!-- ``` -->
+<!--     ... -->
+<!--     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R -->
+<!--     ... has the test problem and the expanded code. Around line 367 is where we are -->
+<!--     scratching our heads. The function code (from nlsModel()) is in the commented  -->
+<!--     lines below the call. This is -->
+
+<!--           # > setPars -->
+<!--           # function(newPars) { -->
+<!--           #   setPars(newPars) -->
+<!--           #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x} -->
+<!--           #   dev   <<- sum(resid^2) # envir = thisEnv -->
+<!--           #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr) -->
+<!--           #   QR <<- qr(.swts * gr) # envir = thisEnv -->
+<!--           #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix -->
+<!--           # } -->
+
+<!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
+<!--     with documentation that will be easier to follow and test, then gradually figure -->
+<!--     out how to make it more efficient. -->
+
+<!--     The equivalent from minpack.lm is -->
+
+<!--     setPars = function(newPars) { -->
+<!--                 setPars(newPars) -->
+<!--                 assign("resid", .swts * (lhs - assign("rhs", getRHS(), -->
+<!--                     envir = thisEnv)), envir = thisEnv) -->
+<!--                 assign("dev", sum(resid^2), envir = thisEnv) -->
+<!--                 assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv) -->
+<!--                 return(QR$rank < min(dim(QR$qr))) -->
+<!--             } -->
+
+<!--     In both there is the recursive call, which must have a purpose I don't understand. -->
+<!-- ``` -->
+<!-- The reply was -->
+
+<!-- ``` -->
+<!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
+<!--     I know that I wrote the code to use a closure for the response, covariates, etc.,  -->
+<!--     but I don't recall anything like a recursive call being necessary. -->
+<!--     ... -->
+<!-- ``` -->
+<!-- <!--     I don't think I will be of much help.  My R skills have atrophied to the point  --> -->
+<!-- <!--     where I wouldn't even know how to start exploring what is happening in the first  --> -->
+<!-- <!--     call as opposed to the recursive call. --> -->
+<!-- <!-- ``` --> -->
 

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Some cleanup of bib file

@@
 \usepackage{xcolor}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\hypersetup{
-  pdftitle={Refactoring the nls() function in R},
-  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
-  colorlinks=true,
-  linkcolor={red},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
-  pdfcreator={LaTeX via pandoc}}
-\urlstyle{same} % disable monospaced font for URLs
 \usepackage[margin=1in]{geometry}
@@
 \fi
+\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
+\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
+\urlstyle{same} % disable monospaced font for URLs
+\hypersetup{
+  pdftitle={Refactoring the nls() function in R},
+  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
+  colorlinks=true,
+  linkcolor={red},
+  filecolor={Maroon},
+  citecolor={Blue},
+  urlcolor={blue},
+  pdfcreator={LaTeX via pandoc}}
 
@@
 (\url{https://cran.r-project.org}) for estimating nonlinear statistical
-models. The function dates to the 1980s and the work related to D. M.
-Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
+models. The function dates to the 1980s and the work related to Bates
+and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
 \url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).
@@
 \textbf{algorithm} is the Relative Offset Convergence Criterion (see
-Douglas M. Bates and Watts
+Bates, Douglas M. and Watts, Donald G.
 (\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
@@
 \texttt{nls()}, leaves out some possibilities that could be useful for
-some problems. The package \texttt{nlsr} (John C. Nash and Murdoch
+some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
 (\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
@@
 Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
-(\protect\hyperlink{ref-Levenberg1944}{1944}), John C. Nash
+(\protect\hyperlink{ref-Levenberg1944}{1944}), Nash
 (\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
@@
 current calls back to \texttt{nls()}. Interactive tools, such as
-``visual fitting'' (John C. Nash and Velleman
+``visual fitting'' (Nash and Velleman
 (\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
@@
 \begin{verbatim}
-## Using approximation  jacentral   is.character(ctrl$japprox)= TRUE 
+## Using approximation  jacentral  
+##       is.character(ctrl$japprox)= TRUE 
 ## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
@@
 
-One obvious case is that nonlinear modeling tools are a superset of
-linear ones. Yet the explicit model
+The nonlinear modeling model specifications are, of course, a
+development of linear ones. Unfortunately, the explicit model
 \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
@@
 
-Within \texttt{nls()}, consider the following FOUR different calling
-sequences for the same problem, though an intuitive choice, labelled
-fm2a, will not work. In this failed attempt, putting the \texttt{Asym}
-parameter in the model causes the \texttt{plinear} algorithm to try to
-add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing, we do not
-foresee a resolution for this issue. In the example, we have NOT
-evaluated the commands to save space.
+Within \texttt{nls()}, consider the following FOUR different
+specifications for the same problem, plus an intuitive choice, labelled
+fm2a, that does not work. In this failed attempt, putting the
+\texttt{Asym} parameter in the model causes the \texttt{plinear}
+algorithm to try to add another term to the model. We believe this is
+unfortunate, and would like to see a consistent syntax. At the time of
+writing, we do not foresee a resolution for this issue. In the example,
+we have NOT evaluated the commands to save space.
 
@@
 \end{Shaded}
-
-\hypertarget{issue-code-structure-and-documentation-for-maintenance}{%
-\subsection{Issue: code structure and documentation for
-maintenance}\label{issue-code-structure-and-documentation-for-maintenance}}
-
-The \texttt{nls()} code is structured in a way that inhibits both
-maintenance and improvement. In particular, the iterative setup is such
-that introduction of Marquardt stabilization is not easily available. To
-obtain performance a lot of the code was in C with consequent calls and
-returns that complicate the code. Over time, R has become much more
-efficient on modern computers, and the need to use compiled C and
-Fortran is less critical. Moreover, the burden for maintenance could be
-reduced by moving code entirely to R.
-
-While R and its packages are generally very well documented for usage,
-that documentation rarely extends as far as it might in directions
-helpful for code maintenance. The paucity of such documentation is
-exacerbated by the mixed R/C/Fortran code base, where seemingly trivial
-differences in things like representation of numbers or base index for
-arrays lead to traps for unwary programmers.
-
-For \texttt{nls()} the concern is demonstrated by a short email query
-from John Nash to Doug Bates and his reply. This is NOT a criticism of
-Prof.~Bates' (or any other) work, but a reflection on how difficult it
-is to develop code in this subject area and to keep it maintainable. We
-have experienced similar loss of understanding for some of our own
-codes.
-
-\begin{verbatim}
-    ...
-    https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R
-    ... has the test problem and the expanded code. Around line 367 is where we are
-    scratching our heads. The function code (from nlsModel()) is in the commented 
-    lines below the call. This is
-
-          # > setPars
-          # function(newPars) {
-          #   setPars(newPars)
-          #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
-          #   dev   <<- sum(resid^2) # envir = thisEnv
-          #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
-          #   QR <<- qr(.swts * gr) # envir = thisEnv
-          #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
-          # }
-
-    I'm anticipating that we will be able to set up a (possibly inefficient) code
-    with documentation that will be easier to follow and test, then gradually figure
-    out how to make it more efficient.
-
-    The equivalent from minpack.lm is
-
-    setPars = function(newPars) {
-                setPars(newPars)
-                assign("resid", .swts * (lhs - assign("rhs", getRHS(),
-                    envir = thisEnv)), envir = thisEnv)
-                assign("dev", sum(resid^2), envir = thisEnv)
-                assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv)
-                return(QR$rank < min(dim(QR$qr)))
-            }
-
-    In both there is the recursive call, which must have a purpose I don't understand.
-\end{verbatim}
-
-The reply was
-
-\begin{verbatim}
-    I'm afraid that I don't know the purpose of the recursive call either.  
-    I know that I wrote the code to use a closure for the response, covariates, etc., 
-    but I don't recall anything like a recursive call being necessary.
-    ...
-\end{verbatim}
 
@@
 approach below.
+
+\hypertarget{issue-code-structure-and-documentation-for-maintenance}{%
+\subsection{Issue: code structure and documentation for
+maintenance}\label{issue-code-structure-and-documentation-for-maintenance}}
+
+The \texttt{nls()} code is structured in a way that inhibits both
+maintenance and improvement. In particular, the iterative setup is such
+that introduction of Marquardt stabilization is not easily available. To
+obtain performance when the code was written, a large portion was
+written in C with consequent calls and returns that severely complicate
+the code. Over time, R has become much more efficient on modern
+computers, and the need to use compiled C and Fortran is less critical.
+Moreover, the burden for maintenance could be reduced by moving code
+entirely to R.
+
+While R and its packages are generally very well documented for usage,
+that documentation rarely extends as far as it might for code
+maintenance. The paucity of such documentation is exacerbated by the
+mixed R/C/Fortran code base, where seemingly trivial differences in
+things like representation of numbers or base index for arrays lead to
+traps for unwary programmers.
+
+--\textgreater{} --\textgreater{} --\textgreater{} --\textgreater{}
 
@@
   \texttt{nlsalt}) coded entirely in R that matches the features in base
-  R or what is packaged in \texttt{nlspkg} as described in John C. Nash
-  and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
+  R or what is packaged in \texttt{nlspkg} as described in Nash and
+  Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
 \item
@@
 \item
-  John C. Nash and Bhattacharjee
-  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
-  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
-  in R-base.
+  Nash and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022})
+  is an explanation of the construction of the \texttt{nlspkg} from the
+  \texttt{nls()} code in R-base.
 \item
@@
   invited to prepare a review of optimization in R. Ideas from the
-  present work have been instrumental in the creation of John C. Nash
+  present work have been instrumental in the creation of Nash
   (\protect\hyperlink{ref-NashWires22}{2022}).
@@
 ``as and when'' basis. This follows a path that one of us (JN)
-established almost 50 years ago with the software that became John C.
-Nash (\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver,
-and post-solution analysis phases of computations.
+established almost 50 years ago with the software that became Nash
+(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
+post-solution analysis phases of computations.
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
-Bates, Douglas M., and Donald G. Watts. 1981. {``A Relative Offset
+Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
 Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
+John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
+Nonlinear Least Squares Solutions}}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
@@
 \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
-Nash, John C, and Duncan Murdoch. 2019. \emph{Nlsr: Functions for
-Nonlinear Least Squares Solutions}.
 

### 2022-07-27 RefactoringNLS/RefactoringNLS.Rmd
Some cleanup of bib file
Edits to formula section

@@
 
-Specifying a model to a solver should, ideally, use the same syntax across
+The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
+effective then general approaches in finding good solutions to nonlinear least squares 
+problems when some of the parameters appear linearly. However, setting up the 
+calculations, that is, identifying which parameters are linear, is not a trivial
+task.
+
+Moreover, specifying a model to a solver should, ideally, use the same syntax across
 solver tools. Unfortunately, R allows multiple approaches within different
-modelling tools, and within `nls()` itself.
+modelling tools, and within `nls()` itself. This is highlighted by the case of
+partially linear models.
 
@@
 Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
-an intuitive choice, labelled fm2a, that does not work. In this failed
+an intuitive choice, labelled `fm2a`, that does NOT work. In this failed
 attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
@@
 ```
+
+## Feature: indexed parameters
+
+Indexed parameters -- those whose names have subscripts on some index variable -- 
+are a useful idea to simplify model specification in some cases. However, they
+are also a focus of trouble.
+
+The **man** file for `nls()` includes the following example of a situation in which
+parameters are indexed. It also uses the "plinear" option as an added complication.
+
+Here we use a truncated version of the example to save display space.
+
+```{r nlsindx1}
+## The muscle dataset in MASS is from an experiment on muscle
+## contraction on 21 animals.  The observed variables are Strip
+## (identifier of muscle), Conc (Cacl concentration) and Length
+## (resulting length of muscle section).
+if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
+mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE)
+mm<-droplevels(mm) # remove unused levels after truncation
+nlev <- nlevels(mm)
+withAutoprint({
+  ## The non linear model considered is
+  ##       Length = alpha + beta*exp(-Conc/theta) + error
+  ## where theta is constant. For now alpha and beta do NOT vary with Strip.
+  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
+  nl <- nlevels(mm$Strip)
+  ## We first use the plinear algorithm to fit an overall model,
+  ## ignoring that alpha and beta might vary with Strip.
+  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
+                start = list(th = 1), algorithm = "plinear")
+  summary(musc.1)
+
+  ## Then we use nls' indexing feature for parameters in non-linear
+  ## models to use the conventional algorithm to fit a model in which
+  ## alpha and beta vary with Strip.  The starting values are provided
+  ## by the previously fitted model.
+  ## Note that with indexed parameters, the starting values must be
+  ## given in a list (with names):
+  ## ?? but why use b here AND in the new formula??
+  b <- coef(musc.1)
+  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
+                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
+  summary(musc.2)
+})
+```
+
+Running the example reveals that the answers for the parameters are NOT indexed. 
+<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
+That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
+<!-- This is because the model -->
+<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
+This is no doubt because programming for indexed parameters is extremely challenging.
+<!-- However, we believe the current structure could cause confusion and error, and  -->
+<!-- propose an alternative approach below. -->
 
@@
 <!-- <!-- ``` --> -->
-
-## Feature: indexed parameters
-
-The **man** file for `nls()` includes the following example of a situation in which
-parameters are indexed. It also uses the "plinear" option as an added complication.
-
-<!-- Here we use a truncated version of the example to save display space. -->
-
-<!-- ```{r nlsindx1} -->
-<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
-<!-- ## contraction on 21 animals.  The observed variables are Strip -->
-<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
-<!-- ## (resulting length of muscle section). -->
-<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
-<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip -->
-<!-- str(mm) -->
-<!-- mm<-droplevels(mm) -->
-<!-- str(mm) -->
-<!-- nlev <- nlevels(mm) -->
-<!-- withAutoprint({ -->
-<!--   ## The non linear model considered is -->
-<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
-<!--   ## where theta is constant but alpha and beta may vary with Strip. -->
-<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
-<!--   nl <- nlevels(mm$Strip) -->
-<!--   ## We first use the plinear algorithm to fit an overall model, -->
-<!--   ## ignoring that alpha and beta might vary with Strip. -->
-<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
-<!--                 start = list(th = 1), algorithm = "plinear") -->
-<!--   summary(musc.1) -->
-
-<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
-<!--   ## models to use the conventional algorithm to fit a model in which -->
-<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
-<!--   ## by the previously fitted model. -->
-<!--   ## Note that with indexed parameters, the starting values must be -->
-<!--   ## given in a list (with names): -->
-<!--   ## ?? but why use b here AND in the new formula?? -->
-<!--   b <- coef(musc.1) -->
-
-<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
-<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
-<!--   summary(musc.2) -->
-<!-- }) -->
-
-<!-- ``` -->
-
-Running the example reveals that the answers for the parameters are NOT indexed. 
-<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
-We do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
-<!-- This is because the model -->
-<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
-This is no doubt because programming for indexed parameters is extremely challenging.
-However, we believe the current structure is could cause confusion and error, and 
-propose an alternative approach below.
 
@@
   that matches the features in base R or what is packaged in `nlspkg` as described 
-  in @PkgFromRbase22.
+  in @PkgFromRbase22. The R version can either serve as the operational code or
+  as a reference version to facilitate the maintenance of C, Fortran, or other 
+  versions of the software.
 
@@
 
-- to explain what we do, either in comments or separate maintainer documentation.
+- to explain what we do, either in comments, examples, or separate maintainer documentation.
   Since we are complaining about the lack of explanatory material for the current
@@
 
-While not consistent with the `lm()` specification of linear models, 
-nonlinear models need all parameters specified. 
+Nonlinear models, apart from those using the "plinear" algorithm, need all parameters specified.
 That is, the model formula should be written down as it would be computed.
-Thus a linear model (which is valid input for nonlinear estimatin) should
+Thus a linear model (which is valid input for nonlinear estimation) should
 be specified `y ~ a * x + b` rather than `y ~ x`. 

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Edits to formula section

@@
 \texttt{nls()}, leaves out some possibilities that could be useful for
-some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
-(\protect\hyperlink{ref-nlsr-manual}{2019})) already offers both the
-safeguarded relative offset test (\textbf{roffset}) as well as a
-\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
-the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
-times the initial sum of squares, where
+some problems. The package \texttt{nlsr}
+((\protect\hyperlink{ref-nlsr-manual}{\textbf{nlsr-manual?}})) already
+offers both the safeguarded relative offset test (\textbf{roffset}) as
+well as a \textbf{small sum of squares} test (\textbf{smallsstest}) that
+compares the latest evaluated sum of squared (weighted) residuals to
+\texttt{e4} times the initial sum of squares, where
 \texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
@@
 
-Specifying a model to a solver should, ideally, use the same syntax
-across solver tools. Unfortunately, R allows multiple approaches within
-different modelling tools, and within \texttt{nls()} itself.
+The variable projection method (Golub and Pereyra
+(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
+(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
+effective then general approaches in finding good solutions to nonlinear
+least squares problems when some of the parameters appear linearly.
+However, setting up the calculations, that is, identifying which
+parameters are linear, is not a trivial task.
+
+Moreover, specifying a model to a solver should, ideally, use the same
+syntax across solver tools. Unfortunately, R allows multiple approaches
+within different modelling tools, and within \texttt{nls()} itself. This
+is highlighted by the case of partially linear models.
 
@@
 specifications for the same problem, plus an intuitive choice, labelled
-fm2a, that does not work. In this failed attempt, putting the
+\texttt{fm2a}, that does NOT work. In this failed attempt, putting the
 \texttt{Asym} parameter in the model causes the \texttt{plinear}
@@
 
+Indexed parameters -- those whose names have subscripts on some index
+variable -- are a useful idea to simplify model specification in some
+cases. However, they are also a focus of trouble.
+
 The \textbf{man} file for \texttt{nls()} includes the following example
@@
 ``plinear'' option as an added complication.
+
+Here we use a truncated version of the example to save display space.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# The muscle dataset in MASS is from an experiment on muscle}
+\DocumentationTok{\#\# contraction on 21 animals.  The observed variables are Strip}
+\DocumentationTok{\#\# (identifier of muscle), Conc (Cacl concentration) and Length}
+\DocumentationTok{\#\# (resulting length of muscle section).}
+\ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!} \FunctionTok{requireNamespace}\NormalTok{(}\StringTok{"MASS"}\NormalTok{, }\AttributeTok{quietly =} \ConstantTok{TRUE}\NormalTok{)) }\FunctionTok{stop}\NormalTok{(}\StringTok{"Need MASS pkg"}\NormalTok{)}
+\NormalTok{mm}\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\NormalTok{muscle[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{12}\NormalTok{,] }\CommentTok{\# take only 1st few values of Strip}
+\FunctionTok{str}\NormalTok{(mm)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 'data.frame':    12 obs. of  3 variables:
+##  $ Strip : Factor w/ 21 levels "S01","S02","S03",..: 1 1 1 1 2 2 2 2 3 3 ...
+##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
+##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{mm}\OtherTok{\textless{}{-}}\FunctionTok{droplevels}\NormalTok{(mm)}
+\FunctionTok{str}\NormalTok{(mm)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## 'data.frame':    12 obs. of  3 variables:
+##  $ Strip : Factor w/ 3 levels "S01","S02","S03": 1 1 1 1 2 2 2 2 3 3 ...
+##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
+##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nlev }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm)}
+\FunctionTok{withAutoprint}\NormalTok{(\{}
+  \DocumentationTok{\#\# The non linear model considered is}
+  \DocumentationTok{\#\#       Length = alpha + beta*exp({-}Conc/theta) + error}
+  \DocumentationTok{\#\# where theta is constant but alpha and beta may vary with Strip.}
+  \FunctionTok{with}\NormalTok{(mm, }\FunctionTok{table}\NormalTok{(Strip)) }\CommentTok{\# 2, 3 or 4 obs per strip}
+\NormalTok{  nl }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm}\SpecialCharTok{$}\NormalTok{Strip)}
+  \DocumentationTok{\#\# We first use the plinear algorithm to fit an overall model,}
+  \DocumentationTok{\#\# ignoring that alpha and beta might vary with Strip.}
+\NormalTok{  musc}\FloatTok{.1} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}} \FunctionTok{cbind}\NormalTok{(}\DecValTok{1}\NormalTok{, }\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th)), mm,}
+                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{th =} \DecValTok{1}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
+  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
+
+  \DocumentationTok{\#\# Then we use nls\textquotesingle{} indexing feature for parameters in non{-}linear}
+  \DocumentationTok{\#\# models to use the conventional algorithm to fit a model in which}
+  \DocumentationTok{\#\# alpha and beta vary with Strip.  The starting values are provided}
+  \DocumentationTok{\#\# by the previously fitted model.}
+  \DocumentationTok{\#\# Note that with indexed parameters, the starting values must be}
+  \DocumentationTok{\#\# given in a list (with names):}
+  \DocumentationTok{\#\# ?? but why use b here AND in the new formula??}
+\NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
+
+\NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
+                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{2}\NormalTok{], nl), }\AttributeTok{b =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{3}\NormalTok{], nl), }\AttributeTok{th =}\NormalTok{ b[}\DecValTok{1}\NormalTok{]))}
+  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.2}\NormalTok{)}
+\NormalTok{\})}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## > with(mm, table(Strip))
+## Strip
+## S01 S02 S03 
+##   4   4   4 
+## > nl <- nlevels(mm$Strip)
+## > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
+## +     algorithm = "plinear")
+## > summary(musc.1)
+## 
+## Formula: Length ~ cbind(1, exp(-Conc/th))
+## 
+## Parameters:
+##       Estimate Std. Error t value Pr(>|t|)    
+## th       0.624      0.222    2.81   0.0203 *  
+## .lin1   25.684      1.441   17.82  2.5e-08 ***
+## .lin2  -26.631      6.147   -4.33   0.0019 ** 
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Residual standard error: 2.96 on 9 degrees of freedom
+## 
+## Number of iterations to convergence: 5 
+## Achieved convergence tolerance: 4.71e-07
+## 
+## > b <- coef(musc.1)
+## > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
+## +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
+## > summary(musc.2)
+## 
+## Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
+## 
+## Parameters:
+##    Estimate Std. Error t value Pr(>|t|)    
+## a1  22.9277     0.5154   44.49  1.1e-07 ***
+## a2  27.8606     0.5150   54.09  4.1e-08 ***
+## a3  28.3426     1.0771   26.31  1.5e-06 ***
+## b1 -44.1558    12.3900   -3.56    0.016 *  
+## b2 -43.8084    12.3148   -3.56    0.016 *  
+## b3 -32.8421     2.1553  -15.24  2.2e-05 ***
+## th   0.5548     0.0808    6.86    0.001 ** 
+## ---
+## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+## 
+## Residual standard error: 0.798 on 5 degrees of freedom
+## 
+## Number of iterations to convergence: 6 
+## Achieved convergence tolerance: 1.94e-06
+\end{verbatim}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
+Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of
+Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables
+Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsr-manual}{}}%
-John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
-Nonlinear Least Squares Solutions}}.
-
 \leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
----------. 2022. {``Function Minimization and Nonlinear Least Squares in
-r.''} \emph{WIREs Computational Statistics}, no. e1580.
+---------. 2022. {``{Function minimization and nonlinear least squares
+in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
 https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
+O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
+Nonlinear Least Squares Problems.''} \emph{Computational Optimization
+and Applications} 54 (3): 579--93.
+
 \leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
@@
 \leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
-Wickham, Hadley. 2011. {``Testthat: Get Started with Testing.''}
+Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
 \emph{The R Journal} 3: 5--10.
@@
 Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
-\emph{Devtools: Tools to Make Developing r Packages Easier}.
+\emph{{devtools: Tools to Make Developing R Packages Easier}}.
 \url{https://CRAN.R-project.org/package=devtools}.

### 2022-07-27 RefactoringNLS/RefactoringNLS.Rmd
Cleanup edits

@@
 some possibilities that could
-be useful for some problems. The package `nlsr` (@nlsr-manual) already offers both 
+be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
 the safeguarded relative offset test (**roffset**) as well as a **small sum of 
@@
 `e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+(NOTE: For the purposes of this article, we use a developmental version of package
+`nlsr` labelled `nlsr2`.)
 
@@
 jform <- y ~ a + b * r^(x - m) + c * x # Formula
-library(nlsr)
-linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = F))
+library(nlsr2)
+linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
 linexp2 # Note singular values of Jacobian in rightmost column
 # We get an error with nls()
-linexp <- try(nls(jform, data = testdat, start = strt, trace = F))
+linexp <- try(nls(jform, data = testdat, start = strt, trace = FALSE))
 ```
@@
 
-The functions `resid()` (an alias for `residuals()`) and `fitted()` and `lhs()` are UNWEIGHTED.
-But if we return `ans` from `nls()` or `minpack.lm::nlsLM` or our new `nlsj` (interim package), 
-then `ans$m$resid()` is WEIGHTED. This needs clear documentation so users can get the answers
-intended.
-
-SHORT EXAMPLE??
+The function `resid()` (an alias for `residuals()`) `nlsModel()` is WEIGHTED, as are those in the
+structured object `m` returned by `nls()` or `minpack.lm::nlsLM`,
+e.g., `ans$m$resid()`.
 
@@
 Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal))
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
 nowt
-nowt$m$resid()
-inpar<-getInitial(weed ~ SSlogis(tt, Asym, xmid, scal), data=weeddf)
-print(inpar)
-library(nlsr2)
-nowtx<-nlxb(frmlogis, start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
-nowtx
-# nowtSSx0<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
-# nowtSsx0
-# nowtSSx<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=100, xmid=2, scal=1), trace=TRUE, 
-#               control=list(japprox="jand"))
-# nowtSSx
-nowtxi<-nlxb(frmlogis, start=inpar, trace=TRUE)
-nowtxi
+nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
 usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
 usewt
-usewt$m$resid()
-nmod <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
-usewtx<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts, trace=TRUE)
-usewtx
-# NOTE jafwd and jaback fail!
-usewtxn<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts, trace=TRUE, control=list(japprox="jacentral"))
-usewtxn
-usewtxi<-nlxb(frmlogis, start=inpar, weights=wts, trace=TRUE)
-usewtxi
-# usewtx$resid
+usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
+nmod$resid()
 ```
@@
 in the result object as `.lin1`, `.lin2`, etc., where the actual position 
-of parameters in the model is not explicit.
+of parameters in the model is not explicit. We are unclear why these parameters
+begin with a dot, that is, that they have the "hidden" indicator.
 
@@
 `Length` and `Conc` are available for all observations and parameter `th` applies
-to every fitted point, but `a[Strip]` and `b[Strip]` depend on the index, which
+to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
 could be an experimental run number, or some other categorization.
@@
 
-Our view is that inclusion of the indexing **within** `nls()` introduces 
+Our view is that inclusion of the indexing **within** `nls()` or similar tools introduces 
 complexity that could be avoided by a suitable wrapper function. The modeling
@@
 parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
-we believe the wrapper should be able to process output to present the indexed
-parameters correctly.
+a wrapper could process output to reconstitute the indexing of parameters.
 
@@
   mechanisms, that is, different ways of supplying inputs to the solver(s);
-- to pose "silly" inputs to see if 
-  these bad inputs are caught by the programs.
+- to pose "silly" inputs to see if these bad inputs are caught by the programs.
 
@@
   the data and parameters needed are available in a particular environment. As long
-  as the correct environment is used, this provides a short syntax to invoke the
+  as the correct environment is used, this provides a compact syntax to invoke the
   calculations. The danger is that the wrong data is accessed if the internal search finds a
@@
 
-- Weights, subsets, and various contextual controls such as that for the `na.action` option
+- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
   are similarly taken from the first available source. 
@@
   If necessary, by
-  measuring ("profiling") the code, we can find bottlenecks and, if necessary, substitute just those slower 
+  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
   parts of the reference code.
@@
   for complicated calculations reduces the number of objects that must be kept in 
-  alignment.
+  alignment. On the other hand, it may mean users have to add more commands to complete
+  their computations.
 
@@
   are untoward conditions reached, such as a singular Jacobian. This modification could be
-  made to `nls()` in the short term.
+  made to `nls()` in the short term, correcting the issue identified above under "Failure to
+  return best result achieved".
 
 - Given the existence of examples of good practices such as analytic derivatives, stabilized 
-  solution of Gauss-Newton equations and bounds constrained parameters, base R tools should 
+  solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
   be moving to incorporate them.

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Cleanup edits

@@
 \texttt{nls()}, leaves out some possibilities that could be useful for
-some problems. The package \texttt{nlsr}
-((\protect\hyperlink{ref-nlsr-manual}{\textbf{nlsr-manual?}})) already
-offers both the safeguarded relative offset test (\textbf{roffset}) as
-well as a \textbf{small sum of squares} test (\textbf{smallsstest}) that
-compares the latest evaluated sum of squared (weighted) residuals to
-\texttt{e4} times the initial sum of squares, where
+some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
+(\protect\hyperlink{ref-nlsr2019manual}{2019})) already offers both the
+safeguarded relative offset test (\textbf{roffset}) as well as a
+\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
+the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
+times the initial sum of squares, where
 \texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
-approximately 2.43e-55.
+approximately 2.43e-55. (NOTE: For the purposes of this article, we use
+a developmental version of package \texttt{nlsr} labelled
+\texttt{nlsr2}.)
 
@@
 \NormalTok{jform }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b }\SpecialCharTok{*}\NormalTok{ r}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ m) }\SpecialCharTok{+}\NormalTok{ c }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# Formula}
-\FunctionTok{library}\NormalTok{(nlsr)}
-\NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =}\NormalTok{ F))}
+\FunctionTok{library}\NormalTok{(nlsr2)}
+\NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
 \NormalTok{linexp2 }\CommentTok{\# Note singular values of Jacobian in rightmost column}
@@
 \begin{verbatim}
-## nlsr object: x 
+## nlsr2  object: x 
 ## residual sumsquares =  0  on  11 observations
@@
 \CommentTok{\# We get an error with nls()}
-\NormalTok{linexp }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =}\NormalTok{ F))}
+\NormalTok{linexp }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
 \end{Highlighting}
@@
 
-The functions \texttt{resid()} (an alias for \texttt{residuals()}) and
-\texttt{fitted()} and \texttt{lhs()} are UNWEIGHTED. But if we return
-\texttt{ans} from \texttt{nls()} or \texttt{minpack.lm::nlsLM} or our
-new \texttt{nlsj} (interim package), then \texttt{ans\$m\$resid()} is
-WEIGHTED. This needs clear documentation so users can get the answers
-intended.
-
-SHORT EXAMPLE??
+The function \texttt{resid()} (an alias for \texttt{residuals()})
+\texttt{nlsModel()} is WEIGHTED, as are those in the structured object
+\texttt{m} returned by \texttt{nls()} or \texttt{minpack.lm::nlsLM},
+e.g., \texttt{ans\$m\$resid()}.
 
@@
 \NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
-\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal))}
+\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
 \NormalTok{nowt}
@@
 \begin{Highlighting}[]
-\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{inpar}\OtherTok{\textless{}{-}}\FunctionTok{getInitial}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{data=}\NormalTok{weeddf)}
-\FunctionTok{print}\NormalTok{(inpar)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##     Asym     xmid     scal 
-## 196.1862  12.4173   3.1891
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(nlsr2)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 
-## Attaching package: 'nlsr2'
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:nlsr':
-## 
-##     codeDeriv, findSubexprs, fnDeriv, isCALL, isMINUSONE, isONE,
-##     isZERO, model2rjfun, model2ssgrfun, modelexpr, newDeriv,
-##     newSimplification, nlfb, nlsDeriv, nlsSimplify, nlxb, resgr, resss,
-##     rjfundoc, sysDerivs, sysSimplifications, wrapnlsr
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nowtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## <<lamda: 1e-04  SS= 23527 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
-## <<lamda: 4e-05  SS= 10887 ( 0.8264 ) at  Asym = 52.337  xmid = 48.234  scal = 180.33 f/j  2 / 1
-## <<lamda: 0.016  SS= 8948.4 ( 0.41672 ) at  Asym = 80.893  xmid = 48.394  scal = 169.52 f/j  6 / 2
-## <<lamda: 0.0064  SS= 8853.5 ( 0.068832 ) at  Asym = 89.895  xmid = 66.542  scal = 118.27 f/j  7 / 3
-## <<lamda: 0.0256  SS= 8673.7 ( 0.080434 ) at  Asym = 103.26  xmid = 69.707  scal = 91.474 f/j  9 / 4
-## <<lamda: 0.1024  SS= 8563.7 ( 0.059227 ) at  Asym = 110.79  xmid = 67.869  scal = 81.345 f/j  11 / 5
-## <<lamda: 0.04096  SS= 8392.1 ( 0.12852 ) at  Asym = 129.64  xmid = 62.354  scal = 48.413 f/j  12 / 6
-## <<lamda: 0.16384  SS= 7777.5 ( 0.17269 ) at  Asym = 139.58  xmid = 48.216  scal = 39.581 f/j  14 / 7
-## <<lamda: 0.65536  SS= 7547.4 ( 0.11307 ) at  Asym = 142.31  xmid = 43.841  scal = 33.748 f/j  16 / 8
-## <<lamda: 0.26214  SS= 6212.9 ( 0.22183 ) at  Asym = 149.97  xmid = 28.123  scal = 17.451 f/j  17 / 9
-## <<lamda: 1.0486  SS= 2919.3 ( 0.33274 ) at  Asym = 152.81  xmid = 15.368  scal = 7.5437 f/j  19 / 10
-## <<lamda: 4.1943  SS= 1904 ( 0.74162 ) at  Asym = 153.41  xmid = 9.8392  scal = 1.8081 f/j  21 / 11
-## <<lamda: 1.6777  SS= 113.91 ( 0.73923 ) at  Asym = 153.46  xmid = 10.631  scal = 2.5863 f/j  22 / 12
-## <<lamda: 0.67109  SS= 11.223 ( 0.6629 ) at  Asym = 153.64  xmid = 10.972  scal = 2.8504 f/j  23 / 13
-## <<lamda: 0.26844  SS= 10.289 ( 0.19356 ) at  Asym = 154.03  xmid = 11.02  scal = 2.8798 f/j  24 / 14
-## <<lamda: 0.10737  SS= 9.8336 ( 0.11332 ) at  Asym = 154.95  xmid = 11.055  scal = 2.889 f/j  25 / 15
-## <<lamda: 0.04295  SS= 8.8717 ( 0.16811 ) at  Asym = 157.04  xmid = 11.133  scal = 2.9084 f/j  26 / 16
-## <<lamda: 0.01718  SS= 7.2069 ( 0.21397 ) at  Asym = 161.34  xmid = 11.29  scal = 2.9469 f/j  27 / 17
-## <<lamda: 0.0068719  SS= 5.1659 ( 0.30108 ) at  Asym = 168.61  xmid = 11.548  scal = 3.0079 f/j  28 / 18
-## <<lamda: 0.0027488  SS= 3.5835 ( 0.33934 ) at  Asym = 178.1  xmid = 11.867  scal = 3.079 f/j  29 / 19
-## <<lamda: 0.0010995  SS= 2.8237 ( 0.28331 ) at  Asym = 187.23  xmid = 12.155  scal = 3.1385 f/j  30 / 20
-## <<lamda: 0.0004398  SS= 2.6123 ( 0.16163 ) at  Asym = 193.29  xmid = 12.334  scal = 3.1734 f/j  31 / 21
-## <<lamda: 0.00017592  SS= 2.588 ( 0.056549 ) at  Asym = 195.68  xmid = 12.403  scal = 3.1863 f/j  32 / 22
-## <<lamda: 7.0369e-05  SS= 2.5873 ( 0.010514 ) at  Asym = 196.14  xmid = 12.416  scal = 3.1889 f/j  33 / 23
-## <<lamda: 2.8147e-05  SS= 2.5873 ( 0.00090069 ) at  Asym = 196.18  xmid = 12.417  scal = 3.1891 f/j  34 / 24
-## <<lamda: 1.1259e-05  SS= 2.5873 ( 3.4237e-05 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  35 / 25
-## <<lamda: 4.5036e-06  SS= 2.5873 ( 6.5557e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  36 / 26
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nowtx}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## nlsr2  object: x 
-## residual sumsquares =  2.5873  on  12 observations
-##     after  27    Jacobian and  36 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             196.186         11.31      17.35  3.167e-08  -3.432e-10       44.93  
-## xmid             12.4173        0.3346      37.11  3.716e-11  -5.273e-08        15.6  
-## scal             3.18908        0.0698      45.69  5.768e-12   1.752e-07      0.0474
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# nowtSSx0\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=1, xmid=1, scal=1), trace=TRUE)}
-\CommentTok{\# nowtSsx0}
-\CommentTok{\# nowtSSx\textless{}{-}nlxb(weed \textasciitilde{} SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=100, xmid=2, scal=1), trace=TRUE, }
-\CommentTok{\#               control=list(japprox="jand"))}
-\CommentTok{\# nowtSSx}
-\NormalTok{nowtxi}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\NormalTok{inpar, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## <<lamda: 1e-04  SS= 2.5873 ( NA ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  1 / 0
-## <<lamda: 4e-05  SS= 2.5873 ( 9.4859e-07 ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  2 / 1
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nowtxi}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## nlsr2  object: x 
-## residual sumsquares =  2.5873  on  12 observations
-##     after  2    Jacobian and  2 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             196.186         11.31      17.35  3.167e-08  -1.769e-09       44.93  
-## xmid             12.4173        0.3346      37.11  3.716e-11  -2.902e-08        15.6  
-## scal             3.18908        0.0698      45.69  5.768e-12   3.896e-08      0.0474
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
@@
 \begin{Highlighting}[]
-\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# WEIGHTED. Does NOT take coefficients.}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{usewtx}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis,  }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
-## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
-## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
-## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
-## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
-## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
-## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
-## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
-## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
-## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
-## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
-## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
-## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
-## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
-## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
-## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
-## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
-## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
-## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
-## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
-## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
-## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
-## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
-## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
-## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.2336e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewtx}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## nlsr2  object: x 
-## residual sumsquares =  0.019617  on  12 observations
-##     after  25    Jacobian and  31 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             199.859         17.46      11.45  1.149e-06  -2.201e-11       7.858  
-## xmid             12.4994        0.4131      30.26    2.3e-10    2.22e-09        0.79  
-## scal             3.19088       0.04323      73.81  7.781e-14  -1.511e-08    0.002673
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# }\AlertTok{NOTE}\CommentTok{ jafwd and jaback fail!}
-\NormalTok{usewtxn}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis,  }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{, }\AttributeTok{control=}\FunctionTok{list}\NormalTok{(}\AttributeTok{japprox=}\StringTok{"jacentral"}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using approximation  jacentral  
-##       is.character(ctrl$japprox)= TRUE 
-## <<lamda: 1e-04  SS= 82.379 ( NA ) at  Asym = 1  xmid = 1  scal = 1 f/j  1 / 0
-## <<lamda: 4e-05  SS= 27.961 ( 0.87591 ) at  Asym = 23.072  xmid = 24.605  scal = 33.424 f/j  2 / 1
-## <<lamda: 0.0016  SS= 27.179 ( 0.10632 ) at  Asym = 27.088  xmid = 25.624  scal = 24.768 f/j  5 / 2
-## <<lamda: 0.0064  SS= 26.326 ( 0.11129 ) at  Asym = 30.015  xmid = 24.489  scal = 21.971 f/j  7 / 3
-## <<lamda: 0.00256  SS= 26.011 ( 0.18194 ) at  Asym = 36.922  xmid = 21.283  scal = 12.464 f/j  8 / 4
-## <<lamda: 0.01024  SS= 21.446 ( 0.27496 ) at  Asym = 39.825  xmid = 16.241  scal = 10.675 f/j  10 / 5
-## <<lamda: 0.04096  SS= 19.865 ( 0.18158 ) at  Asym = 40.684  xmid = 14.706  scal = 8.9148 f/j  12 / 6
-## <<lamda: 0.016384  SS= 11.81 ( 0.32805 ) at  Asym = 42.981  xmid = 9.344  scal = 4.7286 f/j  13 / 7
-## <<lamda: 0.065536  SS= 3.8548 ( 0.46972 ) at  Asym = 43.701  xmid = 6.3095  scal = 2.7564 f/j  15 / 8
-## <<lamda: 0.026214  SS= 2.8386 ( 0.45514 ) at  Asym = 44.458  xmid = 5.503  scal = 2.0895 f/j  16 / 9
-## <<lamda: 0.010486  SS= 2.4496 ( 0.2365 ) at  Asym = 46.866  xmid = 5.9479  scal = 2.3 f/j  17 / 10
-## <<lamda: 0.0041943  SS= 2.0123 ( 0.21463 ) at  Asym = 51.378  xmid = 6.3516  scal = 2.3762 f/j  18 / 11
-## <<lamda: 0.0016777  SS= 1.4462 ( 0.2703 ) at  Asym = 59.51  xmid = 7.03  scal = 2.5092 f/j  19 / 12
-## <<lamda: 0.00067109  SS= 0.91272 ( 0.306 ) at  Asym = 71.787  xmid = 7.8952  scal = 2.6532 f/j  20 / 13
-## <<lamda: 0.00026844  SS= 0.51816 ( 0.29993 ) at  Asym = 88.001  xmid = 8.8228  scal = 2.7857 f/j  21 / 14
-## <<lamda: 0.00010737  SS= 0.26988 ( 0.25879 ) at  Asym = 107.49  xmid = 9.727  scal = 2.901 f/j  22 / 15
-## <<lamda: 4.295e-05  SS= 0.13173 ( 0.20437 ) at  Asym = 129.22  xmid = 10.558  scal = 2.9978 f/j  23 / 16
-## <<lamda: 1.718e-05  SS= 0.062773 ( 0.16799 ) at  Asym = 151.62  xmid = 11.279  scal = 3.0748 f/j  24 / 17
-## <<lamda: 6.8719e-06  SS= 0.032464 ( 0.12324 ) at  Asym = 172.29  xmid = 11.849  scal = 3.1314 f/j  25 / 18
-## <<lamda: 2.7488e-06  SS= 0.021923 ( 0.076781 ) at  Asym = 188.07  xmid = 12.235  scal = 3.1673 f/j  26 / 19
-## <<lamda: 1.0995e-06  SS= 0.01977 ( 0.035429 ) at  Asym = 196.72  xmid = 12.43  scal = 3.1847 f/j  27 / 20
-## <<lamda: 4.398e-07  SS= 0.019619 ( 0.0092109 ) at  Asym = 199.45  xmid = 12.49  scal = 3.19 f/j  28 / 21
-## <<lamda: 1.7592e-07  SS= 0.019617 ( 0.00087595 ) at  Asym = 199.84  xmid = 12.499  scal = 3.1908 f/j  29 / 22
-## <<lamda: 7.0369e-08  SS= 0.019617 ( 4.7993e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  30 / 23
-## <<lamda: 2.8147e-08  SS= 0.019617 ( 8.24e-07 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  31 / 24
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewtxn}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## nlsr2  object: x 
-## residual sumsquares =  0.019617  on  12 observations
-##     after  25    Jacobian and  31 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             199.859         17.46      11.45  1.149e-06  -2.105e-11       7.858  
-## xmid             12.4994        0.4131      30.26    2.3e-10   2.212e-09        0.79  
-## scal             3.19088       0.04323      73.81  7.781e-14  -1.514e-08    0.002673
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewtxi}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(frmlogis, }\AttributeTok{start=}\NormalTok{inpar, }\AttributeTok{weights=}\NormalTok{wts, }\AttributeTok{trace=}\ConstantTok{TRUE}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## <<lamda: 1e-04  SS= 0.020914 ( NA ) at  Asym = 196.19  xmid = 12.417  scal = 3.1891 f/j  1 / 0
-## <<lamda: 4e-05  SS= 0.019704 ( 0.028209 ) at  Asym = 196.45  xmid = 12.419  scal = 3.1833 f/j  2 / 1
-## <<lamda: 1.6e-05  SS= 0.019678 ( 0.0029702 ) at  Asym = 196.99  xmid = 12.432  scal = 3.1845 f/j  3 / 2
-## <<lamda: 6.4e-06  SS= 0.019645 ( 0.0031987 ) at  Asym = 197.91  xmid = 12.453  scal = 3.1866 f/j  4 / 3
-## <<lamda: 2.56e-06  SS= 0.019623 ( 0.0031067 ) at  Asym = 198.96  xmid = 12.478  scal = 3.1889 f/j  5 / 4
-## <<lamda: 1.024e-06  SS= 0.019617 ( 0.0017597 ) at  Asym = 199.63  xmid = 12.494  scal = 3.1904 f/j  6 / 5
-## <<lamda: 4.096e-07  SS= 0.019617 ( 0.00050079 ) at  Asym = 199.83  xmid = 12.499  scal = 3.1908 f/j  7 / 6
-## <<lamda: 1.6384e-07  SS= 0.019617 ( 6.2448e-05 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  8 / 7
-## <<lamda: 6.5536e-08  SS= 0.019617 ( 2.9857e-06 ) at  Asym = 199.86  xmid = 12.499  scal = 3.1909 f/j  9 / 8
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewtxi}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## nlsr2  object: x 
-## residual sumsquares =  0.019617  on  12 observations
-##     after  9    Jacobian and  9 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             199.859         17.46      11.45  1.149e-06  -2.056e-10       7.858  
-## xmid             12.4994        0.4131      30.26    2.3e-10   9.074e-09        0.79  
-## scal             3.19088       0.04323      73.81  7.781e-14  -5.703e-08    0.002673
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# usewtx$resid}
-\end{Highlighting}
-\end{Shaded}
+\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
+\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
+## [11] 1.6572 1.4214
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts=}\NormalTok{wts)}
+\NormalTok{nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\end{verbatim}
 
@@
 \ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!} \FunctionTok{requireNamespace}\NormalTok{(}\StringTok{"MASS"}\NormalTok{, }\AttributeTok{quietly =} \ConstantTok{TRUE}\NormalTok{)) }\FunctionTok{stop}\NormalTok{(}\StringTok{"Need MASS pkg"}\NormalTok{)}
-\NormalTok{mm}\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\NormalTok{muscle[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{12}\NormalTok{,] }\CommentTok{\# take only 1st few values of Strip}
-\FunctionTok{str}\NormalTok{(mm)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 'data.frame':    12 obs. of  3 variables:
-##  $ Strip : Factor w/ 21 levels "S01","S02","S03",..: 1 1 1 1 2 2 2 2 3 3 ...
-##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
-##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{mm}\OtherTok{\textless{}{-}}\FunctionTok{droplevels}\NormalTok{(mm)}
-\FunctionTok{str}\NormalTok{(mm)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 'data.frame':    12 obs. of  3 variables:
-##  $ Strip : Factor w/ 3 levels "S01","S02","S03": 1 1 1 1 2 2 2 2 3 3 ...
-##  $ Conc  : num  1 2 3 4 1 2 3 4 0.25 0.5 ...
-##  $ Length: num  15.8 20.8 22.6 23.8 20.6 26.8 28.4 27 7.2 15.4 ...
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
+\NormalTok{mm}\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\NormalTok{muscle[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{12}\NormalTok{,] }\CommentTok{\# take only 1st few values of Strip (TRUNCATION OF EXAMPLE)}
+\NormalTok{mm}\OtherTok{\textless{}{-}}\FunctionTok{droplevels}\NormalTok{(mm) }\CommentTok{\# remove unused levels after truncation}
 \NormalTok{nlev }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm)}
@@
   \DocumentationTok{\#\#       Length = alpha + beta*exp({-}Conc/theta) + error}
-  \DocumentationTok{\#\# where theta is constant but alpha and beta may vary with Strip.}
+  \DocumentationTok{\#\# where theta is constant. For now alpha and beta do NOT vary with Strip.}
   \FunctionTok{with}\NormalTok{(mm, }\FunctionTok{table}\NormalTok{(Strip)) }\CommentTok{\# 2, 3 or 4 obs per strip}
@@
 \NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
-
 \NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
@@
 Running the example reveals that the answers for the parameters are NOT
-indexed. We do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but
-\texttt{a1,\ a2,\ a3}. This is no doubt because programming for indexed
-parameters is extremely challenging. However, we believe the current
-structure is could cause confusion and error, and propose an alternative
-approach below.
+indexed. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
+but \texttt{a1,\ a2,\ a3}. This is no doubt because programming for
+indexed parameters is extremely challenging.
 
@@
 \FunctionTok{require}\NormalTok{(nlsr)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Loading required package: nlsr
+\end{verbatim}
+
+\begin{verbatim}
+## 
+## Attaching package: 'nlsr'
+\end{verbatim}
+
+\begin{verbatim}
+## The following objects are masked from 'package:nlsr2':
+## 
+##     codeDeriv, findSubexprs, fnDeriv, isCALL, isMINUSONE, isONE,
+##     isZERO, model2rjfun, model2ssgrfun, modelexpr, newDeriv,
+##     newSimplification, nlfb, nlsDeriv, nlsSimplify, nlxb, resgr, resss,
+##     rjfundoc, sysDerivs, sysSimplifications, wrapnlsr
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
 \CommentTok{\# Hobbs scaled problem with bounds, formula specification}
@@
 \begin{verbatim}
-## nlsr2  object: anlxb1 
+## nlsr object: x 
 ## residual sumsquares =  9.4726  on  12 observations
-##     after  12    Jacobian and  12 function evaluations
+##     after  16    Jacobian and  17 function evaluations
 ##   name            coeff          SE       tstat      pval      gradient    JSingval   
 ## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   -2.16e-07           0  
+## b2               4.43325            NA         NA         NA   1.832e-10           0  
 ## b3                     3U           NA         NA         NA           0           0
@@
   R or what is packaged in \texttt{nlspkg} as described in Nash and
-  Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}).
+  Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}). The R
+  version can either serve as the operational code or as a reference
+  version to facilitate the maintenance of C, Fortran, or other versions
+  of the software.
 \item
@@
 \item
-  to explain what we do, either in comments or separate maintainer
-  documentation. Since we are complaining about the lack of explanatory
-  material for the current code, we feel it is incumbent on us to
-  provide such material for our own work, and if possible for the
+  to explain what we do, either in comments, examples, or separate
+  maintainer documentation. Since we are complaining about the lack of
+  explanatory material for the current code, we feel it is incumbent on
+  us to provide such material for our own work, and if possible for the
   existing code.
@@
 
-While not consistent with the \texttt{lm()} specification of linear
-models, nonlinear models need all parameters specified. That is, the
-model formula should be written down as it would be computed. Thus a
-linear model (which is valid input for nonlinear estimatin) should be
-specified \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
+Nonlinear models, apart from those using the ``plinear'' algorithm, need
+all parameters specified. That is, the model formula should be written
+down as it would be computed. Thus a linear model (which is valid input
+for nonlinear estimation) should be specified
+\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
 \texttt{y\ \textasciitilde{}\ x}. Moreover, we want to avoid the
@@
 \texttt{.lin1}, \texttt{.lin2}, etc., where the actual position of
-parameters in the model is not explicit.
+parameters in the model is not explicit. We are unclear why these
+parameters begin with a dot, that is, that they have the ``hidden''
+indicator.
 
@@
 \texttt{th} applies to every fitted point, but \texttt{a{[}Strip{]}} and
-\texttt{b{[}Strip{]}} depend on the index, which could be an
+\texttt{b{[}Strip{]}} apply only to data indexed, which could be an
 experimental run number, or some other categorization.
@@
 Our view is that inclusion of the indexing \textbf{within}
-\texttt{nls()} introduces complexity that could be avoided by a suitable
-wrapper function. The modeling formula must be expanded to encompass all
-the index cases, likely with indexed parameters renamed e.g.,
-\texttt{a{[}2{]}} becomes \texttt{a2} as at present with \texttt{nls()}.
-However, we believe the wrapper should be able to process output to
-present the indexed parameters correctly.
+\texttt{nls()} or similar tools introduces complexity that could be
+avoided by a suitable wrapper function. The modeling formula must be
+expanded to encompass all the index cases, likely with indexed
+parameters renamed e.g., \texttt{a{[}2{]}} becomes \texttt{a2} as at
+present with \texttt{nls()}. However, a wrapper could process output to
+reconstitute the indexing of parameters.
 
@@
   environment. As long as the correct environment is used, this provides
-  a short syntax to invoke the calculations. The danger is that the
+  a compact syntax to invoke the calculations. The danger is that the
   wrong data is accessed if the internal search finds a valid name that
@@
 \item
-  Weights, subsets, and various contextual controls such as that for the
-  \texttt{na.action} option are similarly taken from the first available
-  source. Again, this makes for a very simple invocation of
+  Weights, subsets, and various contextual controls (such as that for
+  the \texttt{na.action} option) are similarly taken from the first
+  available source. Again, this makes for a very simple invocation of
   calculations, but the context is hidden from the user. Furthermore, it
@@
   programming language. If necessary, by measuring (``profiling'') the
-  code, we can find bottlenecks and, if necessary, substitute just those
-  slower parts of the reference code.
+  code, we can find bottlenecks and substitute for those slower parts of
+  the reference code.
 \item
@@
   for complicated calculations reduces the number of objects that must
-  be kept in alignment.
+  be kept in alignment. On the other hand, it may mean users have to add
+  more commands to complete their computations.
 \item
@@
   Jacobian. This modification could be made to \texttt{nls()} in the
-  short term.
+  short term, correcting the issue identified above under ``Failure to
+  return best result achieved''.
 \item
   Given the existence of examples of good practices such as analytic
-  derivatives, stabilized solution of Gauss-Newton equations and bounds
-  constrained parameters, base R tools should be moving to incorporate
-  them.
+  derivatives, stabilized solution of Gauss-Newton equations and
+  bounds-constrained parameters, base R tools should be moving to
+  incorporate them.
 \end{itemize}
@@
 {S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlsr2019manual}{}}%
+John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
+Nonlinear Least Squares Solutions}}.
 

### 2022-07-27 RefactoringNLS/nlswtx-full.R

@@ -1,0 +1,40 @@
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+strt <- c(Asym=Asym, xmid=xmid, scal=scal)
+modlogis<-model2rjfun(frmlogis, strt)
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+nowt
+nlsmod0<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian
+nlsmod0
+nlsrmod<-modlogis(coef(nowt)) # generate model using nlsr function
+nlsrmod
+inpar<-getInitial(weed ~ SSlogis(tt, Asym, xmid, scal), data=weeddf)
+print(inpar)
+nowtx<-nlxb(frmlogis, start=c(Asym=1, xmid=1, scal=1), trace=FALSE)
+nowtx
+# nowtSSx0<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=1, xmid=1, scal=1), trace=TRUE)
+# nowtSsx0
+# nowtSSx<-nlxb(weed ~ SSlogis(tt, Asym=1, xmid=1, scal=1), start=c(Asym=100, xmid=2, scal=1), trace=TRUE, 
+#               control=list(japprox="jand"))
+# nowtSSx
+nowtxi<-nlxb(frmlogis, start=inpar, trace=FALSE)
+nowtxi
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+usewt$m$resid() 
+nmod <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod$resid()
+usewtx<-nlxb(frmlogis,  start=c(Asym=1, xmid=1, scal=1), weights=wts, trace=TRUE)
+usewtx
+# NOTE jafwd and jaback fail!
[…]

### 2022-07-27 RefactoringNLS/RefactoringNLS.Rmd
Minor cleanup

@@
 <!-- We have experienced similar loss of understanding for some of our own codes. -->
-
-<!-- ``` -->
 <!--     ... -->
@@
 <!--     lines below the call. This is -->
-
 <!--           # > setPars -->
@@
 <!--           # } -->
-
 <!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
@@
 <!--     out how to make it more efficient. -->
-
 <!--     The equivalent from minpack.lm is -->
-
 <!--     setPars = function(newPars) { -->
@@
 <!--             } -->
-
 <!--     In both there is the recursive call, which must have a purpose I don't understand. -->
-<!-- ``` -->
 <!-- The reply was -->
-
-<!-- ``` -->
 <!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
@@
 <!--     but I don't recall anything like a recursive call being necessary. -->
-<!--     ... -->
+<!--      I don't think I will be of much help.  My R skills have atrophied to the point  --> 
+<!--      where I wouldn't even know how to start exploring what is happening in the first  -->
+<!--      call as opposed to the recursive call. --> 
 <!--  ``` -->
-<!-- <!--     I don't think I will be of much help.  My R skills have atrophied to the point  --> -->
-<!-- <!--     where I wouldn't even know how to start exploring what is happening in the first  --> -->
-<!-- <!--     call as opposed to the recursive call. --> -->
-<!-- <!-- ``` --> -->
 
@@
 require(minpack.lm)
-require(nlsr)
 # Hobbs scaled problem with bounds, formula specification

### 2022-07-27 RefactoringNLS/RefactoringNLS.tex
Minor cleanup

@@
 
---\textgreater{} --\textgreater{} --\textgreater{} --\textgreater{}
-
 \hypertarget{feature-bounds-on-parameters}{%
@@
 \begin{Highlighting}[]
-\FunctionTok{require}\NormalTok{(nlsr)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: nlsr
-\end{verbatim}
-
-\begin{verbatim}
-## 
-## Attaching package: 'nlsr'
-\end{verbatim}
-
-\begin{verbatim}
-## The following objects are masked from 'package:nlsr2':
-## 
-##     codeDeriv, findSubexprs, fnDeriv, isCALL, isMINUSONE, isONE,
-##     isZERO, model2rjfun, model2ssgrfun, modelexpr, newDeriv,
-##     newSimplification, nlfb, nlsDeriv, nlsSimplify, nlxb, resgr, resss,
-##     rjfundoc, sysDerivs, sysSimplifications, wrapnlsr
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
 \CommentTok{\# Hobbs scaled problem with bounds, formula specification}

### 2023-01-09 RefactoringNLS/RefactoringNLS.Rmd
updating paper

@@
 `e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
-(NOTE: For the purposes of this article, we use a developmental version of package
-`nlsr` labelled `nlsr2`.)
 
@@
 jform <- y ~ a + b * r^(x - m) + c * x # Formula
-library(nlsr2)
+library(nlsr)
 linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
@@
 parameter estimation is carried out entirely within the called C-language program,
-and the R code `function numericDeriv()`, part of `./src/library/stats/R/nls.R`
-in the R distribution source code, is used to provide Jacobian information in 
-the `nlsModel()` and `nlsModel.plinear()` functions.
-
-The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, while that
+and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
+in the R distribution source code. This is used to provide Jacobian information in 
+the `nlsModel()` and `nlsModel.plinear()` functions, which are not exported for 
+general use.
+
+The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
+while that
 of `nls()` mostly uses a numerical approximation. Some selfStart models do provide
 "gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
-version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from the 
-original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, though in our experience, one has to look for the
-differences, as they are generally minor.
+version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
+the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
+can lead to different behaviour, though in our experience differences are usually minor.
 
@@
 Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file may
-provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines each
-model, and these Jacobians may be the analytic forms. The `nls()` function, after computing the
-"right hand side" or `rhs` of the residual, then checks to see if the "gradient" attribute is
-defined, and, if not, uses `numericDeriv` to compute a Jacobian into that attribute. This code
-is within the `nlsModel()` or `nlsModel.plinear()` functions. The use of analytic Jacobians
+provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
+each model, and these Jacobians may be the analytic forms. The `nls()` function, after
+computing the "right hand side" or `rhs` of the residual, then checks to see if the
+"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
+into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
+functions. The use of analytic Jacobians
 almost certainly contributes to the good performance of `nls()` on selfStart models.
@@
 
-<!-- nlsr2 has some extra info ?? -->
-
 ### Weights in returned functions from nls()
@@
 usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+source("nlsModel.R")
 nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)

### 2023-01-09 RefactoringNLS/RefactoringNLS.tex
updating paper

@@
 \texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
-approximately 2.43e-55. (NOTE: For the purposes of this article, we use
-a developmental version of package \texttt{nlsr} labelled
-\texttt{nlsr2}.)
+approximately 2.43e-55.
 
@@
 \NormalTok{jform }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b }\SpecialCharTok{*}\NormalTok{ r}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ m) }\SpecialCharTok{+}\NormalTok{ c }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# Formula}
-\FunctionTok{library}\NormalTok{(nlsr2)}
+\FunctionTok{library}\NormalTok{(nlsr)}
 \NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
@@
 \begin{verbatim}
-## nlsr2  object: x 
 ## residual sumsquares =  0  on  11 observations
@@
 ## r                    0.7            NA         NA         NA           0      0.1098  
-## m                      1            NA         NA         NA           0   1.398e-16
+## m                      1            NA         NA         NA           0   9.798e-17
 \end{verbatim}
@@
 computation during parameter estimation is carried out entirely within
-the called C-language program, and the R code
-\texttt{function\ numericDeriv()}, part of
-\texttt{./src/library/stats/R/nls.R} in the R distribution source code,
-is used to provide Jacobian information in the \texttt{nlsModel()} and
-\texttt{nlsModel.plinear()} functions.
+the called C-language program, and the R code function
+\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
+the R distribution source code. This is used to provide Jacobian
+information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
+functions, which are not exported for general use.
 
@@
 differences in approach can lead to different behaviour, though in our
-experience, one has to look for the differences, as they are generally
-minor.
+experience differences are usually minor.
 
@@
 ## Number of iterations to convergence: 0 
-## Achieved convergence tolerance: 1.13e-06
+## Achieved convergence tolerance: 1.12e-06
 \end{verbatim}
@@
 \begin{Highlighting}[]
+\FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
 \NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
@@
 ## Number of iterations to convergence: 6 
-## Achieved convergence tolerance: 1.94e-06
+## Achieved convergence tolerance: 2.03e-06
 \end{verbatim}
@@
 \begin{verbatim}
-## nlsr2  object: anlxb1 
 ## residual sumsquares =  9.4726  on  12 observations

### 2023-01-09 RefactoringNLS/RefactoringNLS.Rmd
reedits Jan 2023

@@
 `nls()` is the tool in base R, the primary software software distribution
-from the Comprehensive R Archive Network (https://cran.r-project.org)
+from the Comprehensive R Archive Network (*https://cran.r-project.org*),
 for estimating nonlinear statistical models. The function dates to the 1980s and the work
-related to @bateswatts in S (see https://en.wikipedia.org/wiki/S_%28programming_language%29).
+related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
 
@@
 we have created a rather more compact version entirely in R. 
-See https://github.com/nashjc/RNonlinearLS/DerivsNLS .
+See *https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS*.
 <!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
@@
 in the R distribution source code. This is used to provide Jacobian information in 
-the `nlsModel()` and `nlsModel.plinear()` functions, which are not exported for 
+the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
 general use.
@@
 
-Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file may
+Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file 
+and in package `nlraa` (@MiguezNLRAA2021) may
 provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
@@
 
-?? does resid return weighted or unweighted residuals
-
-?? do we want unweighted (raw) residuals? or are predictions / fits just as good?
+`nls()`, `nlsr::nlxb()` and `minpack.lm::nlsLM()` all have a `weights` argument that
+specifies a vector of weights the same length as the number of residuals. Each residual
+is multiplied by the square root of the corresponding weight. The values returned by the
+`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
+used to compute raw residuals.
 
@@
 attractiveness, it does mean that the programmer of the solver needs to be aware of the
-source (and value) of objects such as the data, residuals and Jacobian. By preference, 
+source (and value) of objects such as the data, residuals and Jacobian. This is overly
+complicated. By preference, 
 we would implement subsetting by means of zero-value weights, with observation counts
@@
 A useful, but possibly dated, description is given in:
-https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/.
+*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
 
@@
 A succinct description of the issue is given in: *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*
+where the "Answer" states
 
@@
 We do NOT like this approach, but it is "the R way". R allows users to leave many 
-objects in the default (.GlobalEnv) workspace. Moreover, users have to actively 
+objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
 suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
@@
 Moreover, we believe that it would be helpful to have selfStart models that allow
-users to explicitly provide the starting parameters.
+users to explicitly provide values other than those suggested for the starting parameters.
 
@@
 
-Because the Gauss-Newton algorithm is unreliable from many sets of starting
-of parameters, selfStart models are a part of the `nls()` infrastructure rather
-than an accessory. However, creating such functions is a lot of work, and their
+By providing starting values that are likely to be reasonable and by including 
+Jacobian code in the selfStart model code, some of the deficiencies of the 
+Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
+of work, and their
 documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
@@
 actually computed in the `print` and `summary` methods for the result.) 
-We noted several potential updates to the `nlsr` documentation as well as that
-for `nls()` as we made our comparisons.
+As we made our comparisons, We noted several potential updates to the `nlsr` documentation
+as well as that for `nls()`.
 
@@
 
-The function `resid()` (an alias for `residuals()`) `nlsModel()` is WEIGHTED, as are those in the
+As already noted, the function `resid()` (an alias for `residuals()`) is WEIGHTED, as are those in the
 structured object `m` returned by `nls()` or `minpack.lm::nlsLM`,
-e.g., `ans$m$resid()`.
+e.g., `ans$m$resid()`. The function `nlsModel()`, which we have had to extract from the 
+base R code and explicitly `source()` because it is not exported to the working namespace,
+allows us to compute residuals for particular coefficient sets.
 
@@
 deviance). It is likely that the trace display is embedded in the Fortran of the `nlminb`
-routine that is called to execute the "port" algorithm, but the discrepancy is nonetheless
+routine that is called to execute the "port" algorithm, but the factor of 2 discrepancy is nonetheless
 unfortunate for users.
@@
 solver tools. Unfortunately, R allows multiple approaches within different
-modelling tools, and within `nls()` itself. This is highlighted by the case of
+modeling tools, and within `nls()` itself. This is highlighted by the case of
 partially linear models.
@@
 
-Indexed parameters -- those whose names have subscripts on some index variable -- 
+Indexed parameters -- those whose names have subscripts on some element that
+is part of the coefficient list -- 
 are a useful idea to simplify model specification in some cases. However, they
@@
 <!-- propose an alternative approach below. -->
+
+## Feature: bounds on parameters
+
+There are many situations where the context of a problem constrains the values of parameters.
+For example, one of us was asked many years ago to estimate a model where one parameter estimate
+was negative. The client complained "But that is supposed to be the number of grain elevators
+in Saskatchewan". The number should have been a positive integer, and likely less
+than a few thousand.
+
+`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
+the manual states
+
+> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
+  value is within the bounds. Use with caution, especially where bounds are supplied.
+
+This is clearly unsatisfactory for general use, but does find bounded solutions. 
+Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
+straightforward, and it is part of the default method
+for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
+script, which we do not run here for reasons of space, shows that it does not find a
+good solution for the scaled Hobbs test problem.
+
+```{r chkbound}
+# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+wf <- data.frame(y = weed, tt = tt)
+st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
+wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
+require(minpack.lm)
+# Hobbs scaled problem with bounds, formula specification
+anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("Using nlsr::nlxb():")
+print(anlxb1)
+## nlsLM seems NOT to work properly with bounds
+anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("\n"); cat("using minpack.lm::nlsLM():")
+print(anlsLM1b)
+anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("Using nls with 'port':")
+print(anlsb)
+```
+
+### Philosophical considerations
+
+Bounds on parameters raise some interesting and difficult questions about how uncertainty
+in parameter estimates should be computed or reported. That is, the traditional "standard
+errors" are generally taken to imply symmetric intervals about the point estimate in which
+the parameter may be expected to be found with some probability under certain assumptions.
+Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
+errors nor their derived statistics when bounds are active.
 
@@
 
-## Feature: bounds on parameters
-
-There are many situations where the context of a problem constrains the values of parameters.
-For example, one of us was asked many years ago to estimate a model where one parameter estimate
-was negative. The client complained "But that is supposed to be the number of grain elevators
-in Saskatchewan". The number should have been a positive integer, and likely less
-than a few thousand.
-
-`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
-the manual states
-
-> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
-  value is within the bounds. Use with caution, especially where bounds are supplied.
-
-This is clearly unsatisfactory for general use, but does find bounded solutions. 
-Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
-straightforward, and it is part of the default method
-for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
-script, which we do not run here for reasons of space, shows that it does not find a
-good solution for the scaled Hobbs test problem.
-
-```{r chkbound}
-# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-wf <- data.frame(y = weed, tt = tt)
-st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
-wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
-require(minpack.lm)
-# Hobbs scaled problem with bounds, formula specification
-anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("Using nlsr::nlxb():")
-print(anlxb1)
-## nlsLM seems NOT to work properly with bounds
-anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("\n"); cat("using minpack.lm::nlsLM():")
-print(anlsLM1b)
-anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("Using nls with 'port':")
-print(anlsb)
-```
-
-### Philosophical considerations
-
-Bounds on parameters raise some interesting and difficult questions about how uncertainty
-in parameter estimates should be computed or reported. That is, the traditional "standard
-errors" are generally taken to imply symmetric intervals about the point estimate in which
-the parameter may be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
-errors nor their derived statistics when bounds are active.
-
 # Goals and progress of our effort to improve `nls()`
@@
 
-We want:
+When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
 
@@
 
-These goals echo themes in the recent discussion @Vidoni21 and accompanying 
+These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
 commentary.
 
-### Rationalization of formula specifications
+## Rationalization of formula specifications
 
@@
 `plinear` algorithm is used. This is unfortunate, since the user cannot then simply
-add `algorithm="plinear"` to the call, making errors more likely.
-
-Nonlinear models, apart from those using the "plinear" algorithm, need all parameters specified.
+add `algorithm="plinear"` to the standard call, making errors more likely.
+
+Nonlinear models, apart from those using the "plinear" algorithm, need all the parameters specified.
 That is, the model formula should be written down as it would be computed.
@@
 
-### Rationalization of indexed models
+## Rationalization of indexed models
 
@@
 
-### Streamlining code
+## Streamlining code
 
@@
 Such goals align with the aims of **unit testing** 
-(e.g., https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211, @HWtestthat11, @HWdevtools21
+(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
 and the conventional R package testing tools).
 
-<!-- ??AB: perhaps we should put in the refernces?? -->
-<!-- I've emailed author to ask if there is preferred citation. JN -->
-
-### A testing package: NLSCompare
-
-One of us has developed a working prototype package at https://github.com/ArkaB-DS/nlsCompare .
+## A testing package: NLSCompare
+
+One of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
 A primary design objective of this is to allow the summarisation of multiple tests in
@@
 In our investigation, we built several resources, which are now part of 
-the repository https://github.com/nashjc/RNonlinearLS/. Particular items
+the repository *https://github.com/nashjc/RNonlinearLS/*. Particular items
 are:
@@
   has wider application to nonlinear least squares projects in general
-  (https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib)
+  (*https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib*).
 
@@
   of how it was built and the resources needed across platforms is given in
-  at https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary
+  at *https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary*.
 
@@
 It is possible that `nls()` will remain more or less as it has been for the past
-two decades. Given its importance to R, a focus of discussion should be what needs 
-fixing and how that should be approached, and we welcome an opportunity to participate in 
-such conversations.
-
+two decades. Given its importance to R, the focus of discussion should be the measures
+needed to secure its continued operation for legacy purposes and how that may be approached.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program objects (R functions)
+that are created by tools such as `nls()` should have minimal cross-linkages and side-effects.
+The aspects of `nls()` that have given us the most trouble:
+
+- The functions that compute the residuals and jacobians often presume
+  the data and parameters needed are available in a particular environment. As long
+  as the correct environment is used, this provides a compact syntax to invoke the
+  calculations. The danger is that the wrong data is accessed if the internal search finds a
+  valid name that is not the object we want.
+
+- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
+  are similarly taken from the first available source, making for a very simple
+  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
+  difficult for those of us trying to maintain or improve the code to be certain we have the
+  context correct.
+
+- The mixing of R, C and Fortran code was necessary for computational performance in the past.
+  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
+  We believe in having at least a working reference version of code in a single programming language. 
+  If necessary, by
+  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
+  parts of the reference code.
+
+- A design that isolates the setup, solution, and post-solution parts of
+  for complicated calculations reduces the number of objects that must be kept in 
+  alignment. On the other hand, it may mean users have to add more commands to complete
+  their computations.
+
+- All iterative codes should return the best solution they have found so far, even if there
+  are untoward conditions reached, such as a singular Jacobian. This modification could be
+  made to `nls()` in the short term, correcting the issue identified above under "Failure to
+  return best result achieved".
+
+- Given the existence of examples of good practices such as analytic derivatives, stabilized 
+  solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
+  be moving to incorporate them.
+
+We anticipate that tools like `nlsr` will become more attractive as users discover
+the increased robustness and general capabilities, along with the more recent documentation.
 From a design point of view, a key difference in approach between `nls()` and 
@@
 
-## Opinions
-
-To advance the stability and maintainability of R, we believe the program objects (R functions)
-that are created by tools such as `nls()` should have minimal interactions and side-effects.
-The aspects of `nls()` that have given us the most trouble:
-
-- The functions that compute the residuals and jacobians often presume
-  the data and parameters needed are available in a particular environment. As long
-  as the correct environment is used, this provides a compact syntax to invoke the
-  calculations. The danger is that the wrong data is accessed if the internal search finds a
-  valid name that is not the object we want.
-
-- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
-  are similarly taken from the first available source. 
-  Again, this makes for a very simple
-  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
-  difficult for those of us trying to maintain or improve the code to be certain we have the
-  context correct.
-
-- The mixing of R, C and Fortran code was necessary for computational performance in the past.
-  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
-  We believe in having at least a working reference version of code in a single programming language. 
-  If necessary, by
-  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
-  parts of the reference code.
-
-- A design that isolates the setup, solution, and post-solution parts of
-  for complicated calculations reduces the number of objects that must be kept in 
-  alignment. On the other hand, it may mean users have to add more commands to complete
-  their computations.
-
-- All iterative codes should return the best solution they have found so far, even if there
-  are untoward conditions reached, such as a singular Jacobian. This modification could be
-  made to `nls()` in the short term, correcting the issue identified above under "Failure to
-  return best result achieved".
-
-- Given the existence of examples of good practices such as analytic derivatives, stabilized 
-  solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
-  be moving to incorporate them.
-
 <!-- ??fix next bits -->
@@
 and in comments on this and related work. Heather Turner co-mentored the project 
-and helped guide the progress of the work.
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of selfStart models and improve package `nlsr` in that respect.
 

### 2023-01-09 RefactoringNLS/RefactoringNLS.tex
reedits Jan 2023

@@
 distribution from the Comprehensive R Archive Network
-(\url{https://cran.r-project.org}) for estimating nonlinear statistical
-models. The function dates to the 1980s and the work related to Bates
-and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
-\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}).
+(\emph{\url{https://cran.r-project.org}}), for estimating nonlinear
+statistical models. The function dates to the 1980s and the work related
+to Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
+\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
 
@@
 ## r                    0.7            NA         NA         NA           0      0.1098  
-## m                      1            NA         NA         NA           0   9.798e-17
+## m                      1            NA         NA         NA           0   1.311e-16
 \end{verbatim}
@@
 compact version entirely in R. See
-\url{https://github.com/nashjc/RNonlinearLS/DerivsNLS} .
+\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
 
@@
 information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
-functions, which are not exported for general use.
+functions, which are \textbf{not} exported for general use.
 
@@
 Note that the selfStart models in the base R
-\texttt{./src/library/stats/R/zzModels.R} file may provide the Jacobian
-in the ``gradient'' attribute of the ``one-sided'' formula that defines
-each model, and these Jacobians may be the analytic forms. The
-\texttt{nls()} function, after computing the ``right hand side'' or
-\texttt{rhs} of the residual, then checks to see if the ``gradient''
-attribute is defined, and, if not, uses \texttt{numericDeriv} to compute
-a Jacobian into that attribute. This code is within the
-\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
-analytic Jacobians almost certainly contributes to the good performance
-of \texttt{nls()} on selfStart models.
+\texttt{./src/library/stats/R/zzModels.R} file and in package
+\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
+may provide the Jacobian in the ``gradient'' attribute of the
+``one-sided'' formula that defines each model, and these Jacobians may
+be the analytic forms. The \texttt{nls()} function, after computing the
+``right hand side'' or \texttt{rhs} of the residual, then checks to see
+if the ``gradient'' attribute is defined, and, if not, uses
+\texttt{numericDeriv} to compute a Jacobian into that attribute. This
+code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
+functions. The use of analytic Jacobians almost certainly contributes to
+the good performance of \texttt{nls()} on selfStart models.
 
@@
 
-?? does resid return weighted or unweighted residuals
-
-?? do we want unweighted (raw) residuals? or are predictions / fits just
-as good?
+\texttt{nls()}, \texttt{nlsr::nlxb()} and \texttt{minpack.lm::nlsLM()}
+all have a \texttt{weights} argument that specifies a vector of weights
+the same length as the number of residuals. Each residual is multiplied
+by the square root of the corresponding weight. The values returned by
+the \texttt{residuals()} function are weighted, and the
+\texttt{fitted()} or \texttt{predict()} function are used to compute raw
+residuals.
 
@@
 programmer of the solver needs to be aware of the source (and value) of
-objects such as the data, residuals and Jacobian. By preference, we
-would implement subsetting by means of zero-value weights, with
-observation counts (and degrees of freedom) computed via the numbers of
-non-zero weights. Alternatively, we would extract a working dataframe
-from the relevant elements in the original.
+objects such as the data, residuals and Jacobian. This is overly
+complicated. By preference, we would implement subsetting by means of
+zero-value weights, with observation counts (and degrees of freedom)
+computed via the numbers of non-zero weights. Alternatively, we would
+extract a working dataframe from the relevant elements in the original.
 
@@
 but possibly dated, description is given in:
-\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}.
+\emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.
 
@@
 \emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}
+where the ``Answer'' states
 
@@
 approach, but it is ``the R way''. R allows users to leave many objects
-in the default (.GlobalEnv) workspace. Moreover, users have to actively
-suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any
-such file in the path, when R is launched, will be loaded. The
-overwhelming proportion of R users in our acquaintance avoid saving the
-workspace because of the danger of lurking data and functions which may
-cause unwanted results.
+in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
+actively suppress saving this workspace (\texttt{.RData}) on exit;
+otherwise, any such file in the path, when R is launched, will be
+loaded. The overwhelming proportion of R users in our acquaintance avoid
+saving the workspace because of the danger of lurking data and functions
+which may cause unwanted results.
 
@@
 our view, dangerous. Moreover, we believe that it would be helpful to
-have selfStart models that allow users to explicitly provide the
-starting parameters.
+have selfStart models that allow users to explicitly provide values
+other than those suggested for the starting parameters.
 
@@
 
-Because the Gauss-Newton algorithm is unreliable from many sets of
-starting of parameters, selfStart models are a part of the
-\texttt{nls()} infrastructure rather than an accessory. However,
+By providing starting values that are likely to be reasonable and by
+including Jacobian code in the selfStart model code, some of the
+deficiencies of the Gauss-Newton algorithm are mitigated. However,
 creating such functions is a lot of work, and their documentation (file
@@
 values of the Jacobian are actually computed in the \texttt{print} and
-\texttt{summary} methods for the result.) We noted several potential
-updates to the \texttt{nlsr} documentation as well as that for
-\texttt{nls()} as we made our comparisons.
+\texttt{summary} methods for the result.) As we made our comparisons, We
+noted several potential updates to the \texttt{nlsr} documentation as
+well as that for \texttt{nls()}.
 
@@
 
-The function \texttt{resid()} (an alias for \texttt{residuals()})
-\texttt{nlsModel()} is WEIGHTED, as are those in the structured object
+As already noted, the function \texttt{resid()} (an alias for
+\texttt{residuals()}) is WEIGHTED, as are those in the structured object
 \texttt{m} returned by \texttt{nls()} or \texttt{minpack.lm::nlsLM},
-e.g., \texttt{ans\$m\$resid()}.
+e.g., \texttt{ans\$m\$resid()}. The function \texttt{nlsModel()}, which
+we have had to extract from the base R code and explicitly
+\texttt{source()} because it is not exported to the working namespace,
+allows us to compute residuals for particular coefficient sets.
 
@@
 Fortran of the \texttt{nlminb} routine that is called to execute the
-``port'' algorithm, but the discrepancy is nonetheless unfortunate for
-users.
+``port'' algorithm, but the factor of 2 discrepancy is nonetheless
+unfortunate for users.
 
@@
 syntax across solver tools. Unfortunately, R allows multiple approaches
-within different modelling tools, and within \texttt{nls()} itself. This
+within different modeling tools, and within \texttt{nls()} itself. This
 is highlighted by the case of partially linear models.
@@
 
-Indexed parameters -- those whose names have subscripts on some index
-variable -- are a useful idea to simplify model specification in some
-cases. However, they are also a focus of trouble.
+Indexed parameters -- those whose names have subscripts on some element
+that is part of the coefficient list -- are a useful idea to simplify
+model specification in some cases. However, they are also a focus of
+trouble.
 
@@
 ## Number of iterations to convergence: 6 
-## Achieved convergence tolerance: 2.03e-06
+## Achieved convergence tolerance: 1.95e-06
 \end{verbatim}
@@
 indexed parameters is extremely challenging.
+
+\hypertarget{feature-bounds-on-parameters}{%
+\subsection{Feature: bounds on
+parameters}\label{feature-bounds-on-parameters}}
+
+There are many situations where the context of a problem constrains the
+values of parameters. For example, one of us was asked many years ago to
+estimate a model where one parameter estimate was negative. The client
+complained ``But that is supposed to be the number of grain elevators in
+Saskatchewan''. The number should have been a positive integer, and
+likely less than a few thousand.
+
+\texttt{nls()} can impose bounds on parameters, but only if the
+\texttt{port} algorithm is used. Unfortunately, the manual states
+
+\begin{quote}
+The \texttt{algorithm\ =\ "port"} code appears unfinished, and does not
+even check that the starting value is within the bounds. Use with
+caution, especially where bounds are supplied.
+\end{quote}
+
+This is clearly unsatisfactory for general use, but does find bounded
+solutions. Inclusion of bounds on parameters in nonlinear least squares
+computations is relatively straightforward, and it is part of the
+default method for package \texttt{nlsr}. Package \texttt{minpack.lm}
+also includes provision for bounds, but the following script, which we
+do not run here for reasons of space, shows that it does not find a good
+solution for the scaled Hobbs test problem.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# bhobbsX.R \#\# bounded formula specification of problem using Hobbs Weed problem}
+\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
+          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
+\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
+\NormalTok{wf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{y =}\NormalTok{ weed, }\AttributeTok{tt =}\NormalTok{ tt)}
+\NormalTok{st }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
+\NormalTok{wmods }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}} \DecValTok{100} \SpecialCharTok{*}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ b3 }\SpecialCharTok{*}\NormalTok{ tt)) }\DocumentationTok{\#\# Scaled model}
+\FunctionTok{require}\NormalTok{(minpack.lm)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Loading required package: minpack.lm
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
+\NormalTok{anlxb1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
+\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nlsr::nlxb():"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Using nlsr::nlxb():
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{print}\NormalTok{(anlxb1)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## residual sumsquares =  9.4726  on  12 observations
+##     after  12    Jacobian and  12 function evaluations
+##   name            coeff          SE       tstat      pval      gradient    JSingval   
+## b1                     2U           NA         NA         NA           0       23.35  
+## b2               4.43325            NA         NA         NA   -2.16e-07           0  
+## b3                     3U           NA         NA         NA           0           0
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# nlsLM seems NOT to work properly with bounds}
+\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
+\FunctionTok{cat}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{); }\FunctionTok{cat}\NormalTok{(}\StringTok{"using minpack.lm::nlsLM():"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## using minpack.lm::nlsLM():
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{print}\NormalTok{(anlsLM1b)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Nonlinear regression model
+##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
+##    data: wf
+## b1 b2 b3 
+##  2  6  3 
+##  residual sum-of-squares: 881
+## 
+## Number of iterations to convergence: 2 
+## Achieved convergence tolerance: 1.49e-08
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
+\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Using nls with 'port':
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\FunctionTok{print}\NormalTok{(anlsb)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## Nonlinear regression model
+##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
+##    data: wf
+##   b1   b2   b3 
+## 2.00 4.43 3.00 
+##  residual sum-of-squares: 9.47
+## 
+## Algorithm "port", convergence message: both X-convergence and relative convergence (5)
+\end{verbatim}
+
+\hypertarget{philosophical-considerations}{%
+\subsubsection{Philosophical
+considerations}\label{philosophical-considerations}}
+
+Bounds on parameters raise some interesting and difficult questions
+about how uncertainty in parameter estimates should be computed or
+reported. That is, the traditional ``standard errors'' are generally
+taken to imply symmetric intervals about the point estimate in which the
+parameter may be expected to be found with some probability under
+certain assumptions. Bounds change those assumptions. Hence,
+\texttt{nlsr::nlxb()} does not compute standard errors nor their derived
+statistics when bounds are active.
 
@@
 
-\hypertarget{feature-bounds-on-parameters}{%
-\subsection{Feature: bounds on
-parameters}\label{feature-bounds-on-parameters}}
-
-There are many situations where the context of a problem constrains the
-values of parameters. For example, one of us was asked many years ago to
-estimate a model where one parameter estimate was negative. The client
-complained ``But that is supposed to be the number of grain elevators in
-Saskatchewan''. The number should have been a positive integer, and
-likely less than a few thousand.
-
-\texttt{nls()} can impose bounds on parameters, but only if the
-\texttt{port} algorithm is used. Unfortunately, the manual states
-
-\begin{quote}
-The \texttt{algorithm\ =\ "port"} code appears unfinished, and does not
-even check that the starting value is within the bounds. Use with
-caution, especially where bounds are supplied.
-\end{quote}
-
-This is clearly unsatisfactory for general use, but does find bounded
-solutions. Inclusion of bounds on parameters in nonlinear least squares
-computations is relatively straightforward, and it is part of the
-default method for package \texttt{nlsr}. Package \texttt{minpack.lm}
-also includes provision for bounds, but the following script, which we
-do not run here for reasons of space, shows that it does not find a good
-solution for the scaled Hobbs test problem.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# bhobbsX.R \#\# bounded formula specification of problem using Hobbs Weed problem}
-\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
-          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{wf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{y =}\NormalTok{ weed, }\AttributeTok{tt =}\NormalTok{ tt)}
-\NormalTok{st }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
-\NormalTok{wmods }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}} \DecValTok{100} \SpecialCharTok{*}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ b3 }\SpecialCharTok{*}\NormalTok{ tt)) }\DocumentationTok{\#\# Scaled model}
-\FunctionTok{require}\NormalTok{(minpack.lm)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: minpack.lm
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
-\NormalTok{anlxb1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nlsr::nlxb():"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using nlsr::nlxb():
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlxb1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  9.4726  on  12 observations
-##     after  12    Jacobian and  12 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   -2.16e-07           0  
-## b3                     3U           NA         NA         NA           0           0
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\# nlsLM seems NOT to work properly with bounds}
-\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{); }\FunctionTok{cat}\NormalTok{(}\StringTok{"using minpack.lm::nlsLM():"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## using minpack.lm::nlsLM():
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlsLM1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
-##    data: wf
-## b1 b2 b3 
-##  2  6  3 
-##  residual sum-of-squares: 881
-## 
-## Number of iterations to convergence: 2 
-## Achieved convergence tolerance: 1.49e-08
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using nls with 'port':
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlsb)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
-##    data: wf
-##   b1   b2   b3 
-## 2.00 4.43 3.00 
-##  residual sum-of-squares: 9.47
-## 
-## Algorithm "port", convergence message: both X-convergence and relative convergence (5)
-\end{verbatim}
-
-\hypertarget{philosophical-considerations}{%
-\subsubsection{Philosophical
-considerations}\label{philosophical-considerations}}
-
-Bounds on parameters raise some interesting and difficult questions
-about how uncertainty in parameter estimates should be computed or
-reported. That is, the traditional ``standard errors'' are generally
-taken to imply symmetric intervals about the point estimate in which the
-parameter may be expected to be found with some probability under
-certain assumptions. Bounds change those assumptions. Hence,
-\texttt{nlsr::nlxb()} does not compute standard errors nor their derived
-statistics when bounds are active.
-
 \hypertarget{goals-and-progress-of-our-effort-to-improve-nls}{%
@@
 
-We want:
+When we started \emph{Improvements to nls()} for the 2021 Google Summer
+of Code we wanted:
 
@@
 
-These goals echo themes in the recent discussion Vidoni
+These goals echo themes in the recent discussion by Vidoni
 (\protect\hyperlink{ref-Vidoni21}{2021}) and accompanying commentary.
@@
 \hypertarget{rationalization-of-formula-specifications}{%
-\subsubsection{Rationalization of formula
+\subsection{Rationalization of formula
 specifications}\label{rationalization-of-formula-specifications}}
@@
 unfortunate, since the user cannot then simply add
-\texttt{algorithm="plinear"} to the call, making errors more likely.
+\texttt{algorithm="plinear"} to the standard call, making errors more
+likely.
 
 Nonlinear models, apart from those using the ``plinear'' algorithm, need
-all parameters specified. That is, the model formula should be written
-down as it would be computed. Thus a linear model (which is valid input
-for nonlinear estimation) should be specified
+all the parameters specified. That is, the model formula should be
+written down as it would be computed. Thus a linear model (which is
+valid input for nonlinear estimation) should be specified
 \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
@@
 \hypertarget{rationalization-of-indexed-models}{%
-\subsubsection{Rationalization of indexed
+\subsection{Rationalization of indexed
 models}\label{rationalization-of-indexed-models}}
@@
 \hypertarget{streamlining-code}{%
-\subsubsection{Streamlining code}\label{streamlining-code}}
+\subsection{Streamlining code}\label{streamlining-code}}
 
@@
 Such goals align with the aims of \textbf{unit testing} (e.g.,
-\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211},
+\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
 Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
@@
 \hypertarget{a-testing-package-nlscompare}{%
-\subsubsection{A testing package:
+\subsection{A testing package:
 NLSCompare}\label{a-testing-package-nlscompare}}
@@
 One of us has developed a working prototype package at
-\url{https://github.com/ArkaB-DS/nlsCompare} . A primary design
+\emph{\url{https://github.com/ArkaB-DS/nlsCompare}} . A primary design
 objective of this is to allow the summarisation of multiple tests in a
@@
 In our investigation, we built several resources, which are now part of
-the repository \url{https://github.com/nashjc/RNonlinearLS/}. Particular
-items are:
+the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
+Particular items are:
 
@@
   general
-  (\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib})
+  (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
 \item
@@
   platforms is given in at
-  \url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}
+  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
 \item
@@
 It is possible that \texttt{nls()} will remain more or less as it has
-been for the past two decades. Given its importance to R, a focus of
-discussion should be what needs fixing and how that should be
-approached, and we welcome an opportunity to participate in such
-conversations.
-
-From a design point of view, a key difference in approach between
-\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()} builds a
-large infrastructure, especially the \texttt{nlsModel()} function, from
-which the Gauss-Newton iteration can be executed and other statistical
-information such as profiles can be computed, while \texttt{nlxb()}
-returns quite limited information, and computes what is needed on an
-``as and when'' basis. This follows a path that one of us (JN)
-established almost 50 years ago with the software that became Nash
-(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
-post-solution analysis phases of computations.
-
-\hypertarget{opinions}{%
-\subsection{Opinions}\label{opinions}}
+been for the past two decades. Given its importance to R, the focus of
+discussion should be the measures needed to secure its continued
+operation for legacy purposes and how that may be approached. We welcome
+an opportunity to participate in such conversations.
 
@@
 program objects (R functions) that are created by tools such as
-\texttt{nls()} should have minimal interactions and side-effects. The
+\texttt{nls()} should have minimal cross-linkages and side-effects. The
 aspects of \texttt{nls()} that have given us the most trouble:
@@
   the \texttt{na.action} option) are similarly taken from the first
-  available source. Again, this makes for a very simple invocation of
-  calculations, but the context is hidden from the user. Furthermore, it
-  can be difficult for those of us trying to maintain or improve the
-  code to be certain we have the context correct.
+  available source, making for a very simple invocation of calculations,
+  but the context is hidden from the user. Furthermore, it can be
+  difficult for those of us trying to maintain or improve the code to be
+  certain we have the context correct.
 \item
@@
 
+We anticipate that tools like \texttt{nlsr} will become more attractive
+as users discover the increased robustness and general capabilities,
+along with the more recent documentation. From a design point of view, a
+key difference in approach between \texttt{nls()} and
+\texttt{nlsr::nlxb()} is that \texttt{nls()} builds a large
+infrastructure, especially the \texttt{nlsModel()} function, from which
+the Gauss-Newton iteration can be executed and other statistical
+information such as profiles can be computed, while \texttt{nlxb()}
+returns quite limited information, and computes what is needed on an
+``as and when'' basis. This follows a path that one of us (JN)
+established almost 50 years ago with the software that became Nash
+(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
+post-solution analysis phases of computations.
+
 \hypertarget{acknowledgement}{%
@@
 co-mentored the project and helped guide the progress of the work.
+Exchanges with Fernando Miguez helped to clarify aspects of selfStart
+models and improve package \texttt{nlsr} in that respect.
 

### 2023-01-28 RefactoringNLS/RefactoringNLS.Rmd
note on profiling
Incorporating AB edits

@@
 ---
-title: "Refactoring the `nls()` function in R"
+title: "Refactoring `nls()`: a constructive critique"
 author: 
@@
 with legacy applications and examples blocking some important updates.
-Discussion of these matters is relevant to improving R generally, as
+Discussion of these matters is relevant to improving R generally as
 well as its nonlinear estimation tools.
@@
 - has related profiling capabilities for exploring the likelihood surface as parameters are changed
-- links to a number of pre-coded ("selfStart") models
+- links to a number of pre-coded (`selfStart`) models
 
@@
 **safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
-We are pleased to note that this improvement is now in the R distributed code as of version 4.1.0.
+We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.
 
@@
 while that
-of `nls()` mostly uses a numerical approximation. Some selfStart models do provide
+of `nls()` mostly uses a numerical approximation. Some `selfStart` models do provide
 "gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
@@
 
-Note that the selfStart models in the base R `./src/library/stats/R/zzModels.R` file 
+Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` file 
 and in package `nlraa` (@MiguezNLRAA2021) may
@@
 functions. The use of analytic Jacobians
-almost certainly contributes to the good performance of `nls()` on selfStart models.
+almost certainly contributes to the good performance of `nls()` on `selfStart` models.
 
@@
 Our concern with `na.action` is that users may be unaware of the effects of an option
-setting they may not even be aware has been set. Is `na.fail` a safer default?
+setting they may not even be aware has been set. Should `na.fail` be the default?
 
@@
 Nonlinear estimation algorithms are almost all iterative and need a set of starting
-parameters. `nls()` offers a special class of modeling functions called **selfStart** 
+parameters. `nls()` offers a special class of modeling functions called **`selfStart`** 
 models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
 and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
-the now-archived package `NRAIA`. Unfortunately, some **selfStart** codes
+the now-archived package `NRAIA`. Unfortunately, some **`selfStart`** codes
 entangle the calculation of starting values for parameters
@@
 because it may call that function. Such circular references are, in our view, dangerous.
-Moreover, we believe that it would be helpful to have selfStart models that allow
+Moreover, we believe that it would be helpful to have `selfStart` models that allow
 users to explicitly provide values other than those suggested for the starting parameters.
 
-Other concerns are revealed by the example below. Here, the `SSlogis` selfStart function 
+Other concerns are revealed by the example below. Here, the `SSlogis` `selfStart` function 
 is used to generate a set
@@
 
-In the event that a selfStart model is not available, `nls()` sets all the starting parameters 
+In the event that a `selfStart` model is not available, `nls()` sets all the starting parameters 
 to 1. This is, in our view, tolerable, but could be improved by using a set of values
@@
 
-### Strategic issues in selfStart models
+### Strategic issues in `selfStart` models
 
 By providing starting values that are likely to be reasonable and by including 
-Jacobian code in the selfStart model code, some of the deficiencies of the 
+Jacobian code in the `selfStart` model code, some of the deficiencies of the 
 Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
@@
 Let us compare the `nls()` result with that from 
-`nlsr::nlxb()`, with ostensibly solves the same problem:
+`nlsr::nlxb()`, which ostensibly solves the same problem:
 
@@
 actually computed in the `print` and `summary` methods for the result.) 
-As we made our comparisons, We noted several potential updates to the `nlsr` documentation
+As we made our comparisons, we noted several potential updates to the `nlsr` documentation
 as well as that for `nls()`.
@@
 The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
-effective then general approaches in finding good solutions to nonlinear least squares 
+effective than general approaches in finding good solutions to nonlinear least squares 
 problems when some of the parameters appear linearly. However, setting up the 
@@
 
-The nonlinear modeling model specifications are, of course, a development of linear ones.
+The nonlinear modeling specifications are, of course, a development of linear ones.
 Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
@@
   ## given in a list (with names):
-  ## ?? but why use b here AND in the new formula??
   b <- coef(musc.1)
@@
 print(anlsLM1b)
-anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", lower = c(0, 0, 0), upper = c(2, 6, 3))
+anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", 
+           lower = c(0, 0, 0), upper = c(2, 6, 3))
 cat("Using nls with 'port':")
@@
 We need to be able to process models such as 
-`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)` where `Strip` is the index. Data
+`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
 `Length` and `Conc` are available for all observations and parameter `th` applies
@@
 
-Given that `[ ]` are the R method to index arrays, their presence in a model
+Given that `[ ]` provides the R method to index arrays, the presence of square
+brackets in a model
 formula signals indexing. The question is how to perform the correct calculations.
@@
 (JN) established almost 50 years ago with the software that became @cnm79, separating
-setup, solver, and post-solution analysis phases of computations. 
+setup, solver, and post-solution analysis phases of computations. There is, however,
+the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver
+of `nlsr` then calls `nls()` to build the object of class `nls` for which profile log-likelihood
+information can be computed.
 <!-- Here, that last  -->
@@
 and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
-clarify aspects of selfStart models and improve package `nlsr` in that respect.
+clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
+Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
+to succinctly summarize computing environments.
 

### 2023-01-31 RefactoringNLS/RefactoringNLS.tex
note on profiling

@@
 \hypersetup{
-  pdftitle={Refactoring the nls() function in R},
+  pdftitle={Refactoring nls(): a constructive critique},
   pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
@@
 
-\title{Refactoring the \texttt{nls()} function in R}
+\title{Refactoring \texttt{nls()}: a constructive critique}
 \author{John C. Nash
@@
 blocking some important updates. Discussion of these matters is relevant
-to improving R generally, as well as its nonlinear estimation tools.
+to improving R generally as well as its nonlinear estimation tools.
 
@@
 \item
-  links to a number of pre-coded (``selfStart'') models
+  links to a number of pre-coded (\texttt{selfStart}) models
 \end{itemize}
@@
 legacy behaviour. We are pleased to note that this improvement is now in
-the R distributed code as of version 4.1.0.
+the R distributed code since version 4.1.0.
 
@@
 analytic expressions, while that of \texttt{nls()} mostly uses a
-numerical approximation. Some selfStart models do provide ``gradient''
-information. \texttt{minpack.lm::nlsLM()} invokes
+numerical approximation. Some \texttt{selfStart} models do provide
+``gradient'' information. \texttt{minpack.lm::nlsLM()} invokes
 \texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
@@
 
-Note that the selfStart models in the base R
+Note that the \texttt{selfStart} models in the base R
 \texttt{./src/library/stats/R/zzModels.R} file and in package
@@
 functions. The use of analytic Jacobians almost certainly contributes to
-the good performance of \texttt{nls()} on selfStart models.
+the good performance of \texttt{nls()} on \texttt{selfStart} models.
 
@@
 Our concern with \texttt{na.action} is that users may be unaware of the
-effects of an option setting they may not even be aware has been set. Is
-\texttt{na.fail} a safer default?
+effects of an option setting they may not even be aware has been set.
+Should \texttt{na.fail} be the default?
 
@@
 of starting parameters. \texttt{nls()} offers a special class of
-modeling functions called \textbf{selfStart} models. There are a number
-of these in base R (\texttt{./src/library/stats/R/zzModels.R}) and
-others in R packages such as CRAN package \texttt{nlraa} (Miguez
+modeling functions called \textbf{\texttt{selfStart}} models. There are
+a number of these in base R (\texttt{./src/library/stats/R/zzModels.R})
+and others in R packages such as CRAN package \texttt{nlraa} (Miguez
 (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
 now-archived package \texttt{NRAIA}. Unfortunately, some
-\textbf{selfStart} codes entangle the calculation of starting values for
-parameters with the particulars of the \texttt{nls()} code. Though there
-is a \texttt{getInitial()} function, this is not easy to use to simply
-compute the initial parameter estimates outside of \texttt{nls()}, in
-part because it may call that function. Such circular references are, in
-our view, dangerous. Moreover, we believe that it would be helpful to
-have selfStart models that allow users to explicitly provide values
-other than those suggested for the starting parameters.
+\textbf{\texttt{selfStart}} codes entangle the calculation of starting
+values for parameters with the particulars of the \texttt{nls()} code.
+Though there is a \texttt{getInitial()} function, this is not easy to
+use to simply compute the initial parameter estimates outside of
+\texttt{nls()}, in part because it may call that function. Such circular
+references are, in our view, dangerous. Moreover, we believe that it
+would be helpful to have \texttt{selfStart} models that allow users to
+explicitly provide values other than those suggested for the starting
+parameters.
 
 Other concerns are revealed by the example below. Here, the
-\texttt{SSlogis} selfStart function is used to generate a set of initial
-parameters for a 3-parameter logistic curve. The form used by
+\texttt{SSlogis} \texttt{selfStart} function is used to generate a set
+of initial parameters for a 3-parameter logistic curve. The form used by
 \texttt{SSlogis} is
@@
 
-In the event that a selfStart model is not available, \texttt{nls()}
-sets all the starting parameters to 1. This is, in our view, tolerable,
-but could be improved by using a set of values that are all slightly
-different, which, in the case of the example model
+In the event that a \texttt{selfStart} model is not available,
+\texttt{nls()} sets all the starting parameters to 1. This is, in our
+view, tolerable, but could be improved by using a set of values that are
+all slightly different, which, in the case of the example model
 \(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
@@
 \hypertarget{strategic-issues-in-selfstart-models}{%
-\subsubsection{Strategic issues in selfStart
-models}\label{strategic-issues-in-selfstart-models}}
+\subsubsection{\texorpdfstring{Strategic issues in \texttt{selfStart}
+models}{Strategic issues in selfStart models}}\label{strategic-issues-in-selfstart-models}}
 
 By providing starting values that are likely to be reasonable and by
-including Jacobian code in the selfStart model code, some of the
-deficiencies of the Gauss-Newton algorithm are mitigated. However,
+including Jacobian code in the \texttt{selfStart} model code, some of
+the deficiencies of the Gauss-Newton algorithm are mitigated. However,
 creating such functions is a lot of work, and their documentation (file
@@
 Let us compare the \texttt{nls()} result with that from
-\texttt{nlsr::nlxb()}, with ostensibly solves the same problem:
+\texttt{nlsr::nlxb()}, which ostensibly solves the same problem:
 
@@
 values of the Jacobian are actually computed in the \texttt{print} and
-\texttt{summary} methods for the result.) As we made our comparisons, We
+\texttt{summary} methods for the result.) As we made our comparisons, we
 noted several potential updates to the \texttt{nlsr} documentation as
@@
 (\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
-effective then general approaches in finding good solutions to nonlinear
+effective than general approaches in finding good solutions to nonlinear
 least squares problems when some of the parameters appear linearly.
@@
 
-The nonlinear modeling model specifications are, of course, a
-development of linear ones. Unfortunately, the explicit model
+The nonlinear modeling specifications are, of course, a development of
+linear ones. Unfortunately, the explicit model
 \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
@@
   \DocumentationTok{\#\# given in a list (with names):}
-  \DocumentationTok{\#\# ?? but why use b here AND in the new formula??}
 \NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
@@
 \begin{Highlighting}[]
-\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
+\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }
+           \AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
 \FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
@@
 to be able to process models such as
-\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)}
+\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)},
 where \texttt{Strip} is the index. Data \texttt{Length} and
@@
 
-Given that \texttt{{[}\ {]}} are the R method to index arrays, their
-presence in a model formula signals indexing. The question is how to
-perform the correct calculations.
+Given that \texttt{{[}\ {]}} provides the R method to index arrays, the
+presence of square brackets in a model formula signals indexing. The
+question is how to perform the correct calculations.
 
@@
 (\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
-post-solution analysis phases of computations.
+post-solution analysis phases of computations. There is, however, the
+wrapper \texttt{wrapnlsr()} that solves the nonlinear estimation problem
+with the solver of \texttt{nlsr} then calls \texttt{nls()} to build the
+object of class \texttt{nls} for which profile log-likelihood
+information can be computed.
 
@@
 co-mentored the project and helped guide the progress of the work.
-Exchanges with Fernando Miguez helped to clarify aspects of selfStart
-models and improve package \texttt{nlsr} in that respect.
+Exchanges with Fernando Miguez helped to clarify aspects of
+\texttt{selfStart} models and improve package \texttt{nlsr} in that
+respect.
 

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
convert paper to rjtools

---
title: "Refactoring `nls()`: a constructive critique"
date: "2023-02-03"
abstract: >
   Based on work for a Google Summer of Code project "Improvements to `nls()`",
   we consider the features and limitations of the R function `nls()`
   in the context of improving and rationalizing R tools for nonlinear regression.
   Important considerations are the usability and  maintainability of
   the code base that provides the functionality `nls()` claims to offer.
   Our work suggests that the
   existing code makes maintenance and improvement very difficult,
   with legacy applications and examples blocking some important updates.
   Discussion of these matters is relevant to improving R generally as
   well as its nonlinear estimation tools.

draft: true
author:  
  # see ?rjournal_article for more information
    - name: John C. Nash
      affiliation: retired professor, University of Ottawa
      address:
      - Telfer School of Management
      - Ottawa ON Canada K1N 6N5
      orcid: 0000-0002-2762-8039
      email: profjcnash@gmail.com
    - name: Arkajyoti Bhattacharjee
      affiliation: Indian Institute of Technology
      address:
      - Department of Mathematics and Statistics
      - Kanpur
      email: arkastat98@gmail.com
type: package
output: 
  rjtools::rjournal_web_article:
    self_contained: yes
    toc: no
bibliography: ../BibSupport/ImproveNLS.bib

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

?? to sort out
- note that AB now at Ohio State
- ORCID for AB?

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
convert paper to rjtools

@@
 
+# The `nls()` function: strengths and shortcomings
+
+`nls()` is the tool in base R, the primary software software distribution
+from the Comprehensive R Archive Network (*https://cran.r-project.org*),
+for estimating nonlinear statistical models. The function dates to the 1980s and the work
+related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+
+The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
+nonlinear models that are expressed as **formulas**. In particular, we note that it
+
+- handles formulas that include R functions, even ones which call calculations in
+  other programming languages
+- allows data to be weighted or subset
+- can estimate bound constrained parameters
+- provides a mechanism for handling partially linear models
+- permits parameters to be indexed over a set of related data
+- produces measures of variability (i.e., standard error estimates) for the estimated parameters
+- has related profiling capabilities for exploring the likelihood surface as parameters are changed
+- links to a number of pre-coded (`selfStart`) models
+
+With such a range of features and long history, the code has become untidy 
+and overly patched. It is, to
+our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
+as we suggest below.
+
+## Feature: Convergence and termination tests (FIXED)
+
+A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
+(i.e., perfect fit) data was corrected thanks to suggestions by one of us.
+
+In the manual page for `nls()` in R 4.0.0 there is the warning:
+
+>  **Do not use `nls` on artificial "zero-residual" data.**
+
+>  The `nls` function uses a relative-offset convergence criterion
+  that compares the numerical imprecision at the current parameter
+  estimates to the residual sum-of-squares.  This performs well on data of
+  the form
+
+>  $$  y = f(x, \theta) + eps $$
+
+>  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form
+
+>  $$  y = f(x, \theta)  $$
+
+>  because the criterion amounts to comparing two components of the round-off 
+  error.
+
+>  If you wish to test `nls` on artificial data please add a noise component, 
+  as shown in the example below.
+
+This amounted to admitting R cannot solve well-posed problems unless data is polluted
+with errors.
+
+This issue can be easily resolved. The "termination test" for the **program** rather than for 
+"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
+(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
+and estimates how much the sum of squares loss function should decrease. This estimate is divided 
+by the current size of the loss function to avoid scale issues.
+When we have "converged", the estimated
+decrease is very small. However, with small residuals,
+the sum of squares loss function is (almost) zero and we get the possibility of a 
+zero-divide failure.
+
+Adding a small quantity to the loss function before dividing avoids trouble. 
+In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
+with a default value of zero. Setting it to a small number -- 1.0 is 
+a reasonable choice --
+allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
+**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
+We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.
+
+# References
+
 ?? to sort out

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
Add another section Jacobian singular

@@
 
+### More general termination tests
+
+The convergence criterion above, the principal termination control of `nls()`, leaves out 
+some possibilities that could
+be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
+the safeguarded relative offset test (**roffset**) as well as a **small sum of 
+squares** test (**smallsstest**) that compares the latest evaluated sum of squared
+(weighted) residuals to `e4` times the initial sum of squares, where
+`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+
+We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
+Unfortunately, the meaning of "iteration" varies among programs and requires 
+careful examination of the
+code. We prefer to use the number of times the residuals or the jacobian have
+been computed and put upper limits on these. Our codes exit (terminate) when these limits
+are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
+but that may reflect the more difficult problems we have encountered because 
+users consult us when standard tools have given unsatisfactory results.
+
+## Feature: Failure when Jacobian is computationally singular
+
+This refers to the infamous "singular gradient" termination message of `nls()`. 
+A Google search of
+
+```
+R nls "singular gradient"
+```
+
+gets over 4000 hits that are spread over some years. This could be because the 
+Jacobian is poorly approximated (see **Jacobian computation** below). However,
+it is common in nonlinear least squares computations that the Jacobian is very close
+to singular for some values of the model parameters. In such cases, we need to find
+an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
+work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
+@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
+In our work, we prepared experimental prototypes of `nls` that incorporate
+stabilization, but integration with all the features of `nls()` has proved difficult.
+
+Let us consider a 'singular gradient' example 
+*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
+`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that the `nls()` error message is correct,
+but this is not helpful in obtaining a solution.
+
+```{r SESingGrad}
+reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
+x <- 1:11 # x values; 11 data points
+y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
+testdat <- data.frame(x, y) # save in a data frame
+strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
+jform <- y ~ a + b * r^(x - m) + c * x # Formula
+library(nlsr)
+linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
+linexp2 # Note singular values of Jacobian in rightmost column
+# We get an error with nls()
+linexp <- try(nls(jform, data = testdat, start = strt, trace = FALSE))
+```
+
+The Jacobian is essentially singular as shown by its singular
+values. Note that these are **displayed** by package `nlsr` in a single column in
+the output to provide a compact layout, but the values do **NOT** correspond to 
+the individual parameters in 
+whose row they appear; they are a property of the whole problem.
+
+## Feature: Jacobian computation
+
+`nls()`, with the `numericDeriv()` function, computes the Jacobian as the "gradient"
+attribute of the residual vector. This is implemented as a mix of R and C code, but
+we have created a rather more compact version entirely in R. 
+See *https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS*.
+<!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
+<!-- (?? need citation -- currently **DerivsNLS.pdf**). -->
+
+As far as we can understand the logic in `nls()`, the Jacobian computation during
+parameter estimation is carried out entirely within the called C-language program,
+and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
+in the R distribution source code. This is used to provide Jacobian information in 
+the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
+general use.
+
+The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
+while that
+of `nls()` mostly uses a numerical approximation. Some `selfStart` models do provide
+"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
+version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
+the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
+can lead to different behaviour, though in our experience differences are usually minor.
+
+- A pasture regrowth problem (@Huet2004, page 1, based on @Ratkowsky1983) has a poorly
+  conditioned Jacobian and `nls()` fails with "singular gradient". 
+  Worse, numerical approximation to the Jacobian give the error
+  "singular gradient matrix at initial parameter estimates" for `minpack.lm::nlsLM`
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of `nlsr::nlxb` give a solution.
+
+- Karl Schilling (private communication) provided an example where a model specified
+  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
+  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
+  this is undefined. In such cases, we observed that `nls()` and `minpack.lm::nlsLM`
+  found a solution, though it can be debated whether such lucky accidents can be
+  taken as an advantage.
+
+Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` file 
+and in package `nlraa` (@MiguezNLRAA2021) may
+provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
+each model, and these Jacobians may be the analytic forms. The `nls()` function, after
+computing the "right hand side" or `rhs` of the residual, then checks to see if the
+"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
+into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
+functions. The use of analytic Jacobians
+almost certainly contributes to the good performance of `nls()` on `selfStart` models.
+
 # References
@@
 ?? to sort out
-- note that AB now at Ohio State
+- note that AB now at Ohio State?
 - ORCID for AB?

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
Add another section

--- 2023-02-03 RefactoringNLS/JNAB230203.Rmd
@@
 
+### More general termination tests
+
+The convergence criterion above, the principal termination control of `nls()`, leaves out 
+some possibilities that could
+be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
+the safeguarded relative offset test (**roffset**) as well as a **small sum of 
+squares** test (**smallsstest**) that compares the latest evaluated sum of squared
+(weighted) residuals to `e4` times the initial sum of squares, where
+`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+
+We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
+Unfortunately, the meaning of "iteration" varies among programs and requires 
+careful examination of the
+code. We prefer to use the number of times the residuals or the jacobian have
+been computed and put upper limits on these. Our codes exit (terminate) when these limits
+are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
+but that may reflect the more difficult problems we have encountered because 
+users consult us when standard tools have given unsatisfactory results.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
Add weights section

--- 2023-02-03 RefactoringNLS/JNAB230203.Rmd
@@
 bibliography: ../BibSupport/ImproveNLS.bib
-
 ---
@@
 
+## Feature: Weights on observations
+
+`nls()`, `nlsr::nlxb()` and `minpack.lm::nlsLM()` all have a `weights` argument that
+specifies a vector of weights the same length as the number of residuals. Each residual
+is multiplied by the square root of the corresponding weight. The values returned by the
+`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
+used to compute raw residuals.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add subset section

@@
 
+## Feature: Subsetting
+
+`nls()` accepts an argument `subset`. This acts through the mediation of
+`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
+`/src/library/stats/src/nls.C`.
+
+While the implementation of subset at the level of the call to `nls()` has a certain
+attractiveness, it does mean that the programmer of the solver needs to be aware of the
+source (and value) of objects such as the data, residuals and Jacobian. This is overly
+complicated. By preference, 
+we would implement subsetting by means of zero-value weights, with observation counts
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add na.action section

@@
 
+## Feature: na.action
+
+`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
+source code, often being handled behind the scenes after referencing the option `na.action`.
+A useful, but possibly dated, description is given in:
+*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
+
+The typical default action, which can be seen by using the command `getOption("na.action")`
+is `na.omit`. This option essentially omits from computations any observations
+containing missing values (i.e. any row of a data frame containing an NA). 
+`na.exclude` does much of the same for computations, but keeps the rows with NA elements so 
+that predictions are in the correct row position. We recommend that workers actually test 
+output to verify the behaviour is as wanted.
+
+A succinct description of the issue is given in: *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*
+where the "Answer" states
+
+> The only benefit of `na.exclude` over `na.omit` is that the former will retain the 
+original number of rows in the data. This may be useful where you need to retain the 
+original size of the dataset - for example it is useful when you want to compare 
+predicted values to original values. With `na.omit` you will end up with fewer 
+rows so you won't as easily be able to compare.
+
+`na.pass` simply passes on data "as is", while `na.fail` will essentially stop if any missing
+values are present.
+
+Our concern with `na.action` is that users may be unaware of the effects of an option
+setting they may not even be aware has been set. Should `na.fail` be the default?
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add model frame section
add sources of data section

@@
 
+## Feature: model frame
+
+`model` is an argument to the `nls()` function, which is documented as:
+
+> **model** logical. If true, the model frame is returned as part of the object. Default is FALSE.
+
+Indeed, the argument only gets used when `nls()` is about to return its result object, and the
+element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
+be confusing.) However, the model frame is used within the function code in the form of the object
+`mf`. We feel that users could benefit from more extensive documentation and examples of its use.
+
+## Feature: Sources of data
+
+`nls()` can be called without specifying the `data` argument. In this case, it will
+search in the available environments (i.e., workspaces) for suitable data objects. 
+We do NOT like this approach, but it is "the R way". R allows users to leave many 
+objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
+suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
+when R is launched, will be loaded. The overwhelming proportion of R users in our
+acquaintance avoid saving the workspace because of the danger of lurking data and
+functions which may cause unwanted results.
+
+Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
+need to ensure equivalent behaviour, but users should test that the 
+operation is as they intend.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
First part of selfStart section

@@
 
+## Feature: missing start vector and self-starting models
+
+Nonlinear estimation algorithms are almost all iterative and need a set of starting
+parameters. `nls()` offers a special class of modeling functions called **`selfStart`** 
+models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
+and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
+the now-archived package `NRAIA`. Unfortunately, some **`selfStart`** codes
+entangle the calculation of starting values for parameters
+with the particulars of the `nls()` code. Though there is a `getInitial()` function,
+this is not easy to use to simply compute the initial parameter estimates outside of
+`nls()`, in part 
+because it may call that function. Such circular references are, in our view, dangerous.
+Moreover, we believe that it would be helpful to have `selfStart` models that allow
+users to explicitly provide values other than those suggested for the starting parameters.
+
+Other concerns are revealed by the example below. Here, the `SSlogis` `selfStart` function 
+is used to generate a set
+of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
+is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
+for this model can be transformed to those of 
+another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.
+
+The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
+<!-- ``` -->
+<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
+<!--     selfStart( -->
+<!--         function(input, Asym, xmid, scal) -->
+<!--         { -->
+<!--               .expr1 <- xmid - input -->
+<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
+<!--               .expr4 <- 1 + .expr3 -->
+<!--               .value <- Asym/.expr4 -->
+<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
+<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
+<!--               { -->
+<!--             		  .expr10 <- .expr4^2 -->
+<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
+<!--                   .grad[, "Asym"] <- 1/.expr4 -->
+<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
+<!-- 		              .grad[, "scal"] <- xm * .e2 -->
+<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
+<!--                   attr(.value, "gradient") <- .grad -->
+<!--               } -->
+<!--               .value -->
+<!--         }, -->
+<!--         initial = function(mCall, data, LHS, ...) { -->
+<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
+<!--               if(nrow(xy) < 4) { -->
+<!--                   stop("too few distinct input values to fit a logistic model") -->
+<!--               } -->
+<!--               z <- xy[["y"]] -->
+<!--               ## transform to proportion, i.e. in (0,1) : -->
+<!--               rng <- range(z); dz <- diff(rng) -->
+<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
+<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
+<!--               aux <- coef(lm(x ~ z, xy)) -->
+<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
+<!--                                data = xy, -->
+<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
+<!--                                algorithm = "plinear", ...)) -->
+<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
+<!--                        mCall[c("Asym", "xmid", "scal")]) -->
+<!--         }, -->
+<!--         parameters = c("Asym", "xmid", "scal")) -->
+<!-- ``` -->
+This R function includes analytic expressions for the Jacobian ("gradient").
+These could be useful to R users, especially if documented. Moreover, we
+wonder why the programmers have chosen to save so many quantities in "hidden"
+variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
+command, making them difficult to access.
+
+In the event that a `selfStart` model is not available, `nls()` sets all the starting parameters 
+to 1. This is, in our view, tolerable, but could be improved by using a set of values
+that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
+would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
+A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
+could be provided quite simply instead of all 1's.
+
+<!-- ```{r hobbsSSlogis} -->
+<!-- weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, -->
+<!--           38.558, 50.156, 62.948, 75.995, 91.972) -->
+<!-- tt <- 1:12 -->
+
+<!-- NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt)) -->
+<!-- NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal) -->
+<!-- NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal)) -->
+<!-- NLStestdata <- data.frame(y=weed, tt=tt)  -->
+<!-- s0 <- getInitial(NLSformula, NLStestdata) -->
+<!-- print(s0) -->
+<!-- # We transform the parameters for the NLSformula0 model of original specification. -->
+<!-- s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3]) -->
+<!-- print(as.numeric(s1)) -->
+<!-- # No actual improvement because nls() has been already used to get the starting values, -->
+<!-- # but we do get SEs -->
+<!-- hobblog<-nls(NLSformula0, data=NLStestdata, start=s1) -->
+<!-- summary(hobblog) -->
+<!-- deviance(hobblog) -->
+<!-- # nls fails without selfStart -- singular gradient, even on stabilized formula -->
+<!-- try(hobblogx<-nls(NLSformulax, data=NLStestdata)) -->
+<!-- # But Marquardt is able to get a solution easily -->
+<!-- library(nlsr) -->
+<!-- hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1)) -->
+<!-- hobblogxx -->
+<!-- ``` -->
+
+<!-- ### selfStart models in base R -->
+
+<!-- The following models are provided (in file ./src/library/stats/R/zzModels.R) -->
+
+<!-- ``` -->
+<!-- SSasymp         - asymptotic regression model -->
+<!-- SSasympOff      - alternate formulation of asymptotic regression model with offset -->
+<!-- SSasympOrig     - exponential curve through the origin to an asymptote -->
+<!-- SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input) -->
+<!-- SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) - -->
+<!--                           exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe))) -->
+<!-- SSfpl           - four parameter logistic model -->
+<!-- SSlogis         - three parameter logistic model -->
+<!-- SSmicmen        - Michaelis-Menten model for enzyme kinetics -->
+<!-- SSgompertz2     - Gompertz model for growth curve data -->
+<!-- SSweibull       - Weibull model for growth curve data -->
+<!-- ``` -->
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
Strategic issues in selfStart

@@
 
+### Strategic issues in `selfStart` models
+
+By providing starting values that are likely to be reasonable and by including 
+Jacobian code in the `selfStart` model code, some of the deficiencies of the 
+Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
+of work, and their
+documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
+We believe that the focus should be placed on getting good initial parameters,
+that is `getInitial()` function, though avoiding the current calls back to `nls()`.
+Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
+worth considering.
+
+We also note that the introduction of `scaleOffset` in R 4.1.1 to deal with the 
+convergence test for small residual problems now requires that the `getInitial()`
+function have dot-arguments (`...`) in its argument list. This illustrates the
+entanglement of many features in `nls()`.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
returned output of nls()

@@
 
+## Issue: returned output of nls() and its documentation
+
+The output of `nls()` is an object of class "nls" which has the following structure:
+
+>A list of
+
+>`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.
+
+>`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.
+
+>`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.
+
+>`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.
+
+>`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.
+
+>`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.
+
+>`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.
+
+>`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.
+
+>`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.
+
+>`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.
+
+> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.
+
+                  -----------------------------------------------------------
+
+<!-- ### Example output -->
+
+<!-- To illustrate, let us run the Croucher example. -->
+
+<!-- ```{r nlsoutx} -->
+<!-- # Croucher example -->
+<!-- xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9) -->
+<!-- ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001) -->
+<!-- p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2) -->
+<!-- NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) -->
+<!-- NLSdata<-data.frame(xdata, ydata) -->
+<!-- # Try full output version of nls -->
+<!-- library(nlspkg) # use the packaged version of nls() -->
+<!-- result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE) -->
+<!-- # We can display the result with several commands -->
+<!-- # str(result) -- displays large amount of material - suppressed here -->
+<!-- #    as it is too wide for the page -->
+<!-- # result # displays the object -->
+<!-- # ls(result) # to list the elements of the output -->
+<!-- # ls(result$m) # and in particular the "m" object -->
+<!-- ``` -->
+
+<!-- ### Concerns with content of the nls result object -->
+
+The `nls` object contains some elements that are awkward to produce by other algorithms,
+but some information that would be useful is not presented in a clear manner. Moreover, the
+complexity of the object is a challenge to users.
+
+In the following, we use `result` as the returned object from `nls()`.
+
+The `data` return element is an R symbol. To actually access the data from this 
+element, we need to use the syntax:
+
+```
+eval(parse(text=result$data))
+```
+
+However, if the call is made with `model=TRUE`, then there is a returned element
+`model` which contains the data, and we can list its contents using:
+
+```
+ls(result$model)
+```
+
+and if there is an element called `xdata`, it can be accessed as `result$model$xdata`.
+
+Let us compare the `nls()` result with that from 
+`nlsr::nlxb()`, which ostensibly solves the same problem:
+
+>`coefficients`    A named vector giving the parameter values at the supposed solution.
+
+>`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.
+
+>`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.
+
+>`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
+                parameters) at the \newline
+$~~~~~~~~~~~~~~~~~~~~$ returned parameters.
+
+>`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.
+
+>`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.
+
+However, actually looking at the structure of a returned result gives a list of 11
+items. The extra 5 are:
+
+```
+ $ lower       : num [1:3] -Inf -Inf -Inf
+ $ upper       : num [1:3] Inf Inf Inf
+ $ maskidx     : int(0) 
+ $ weights     : NULL
+ $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
+  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
+ - attr(*, "class")= chr "nlsr"
+```
+The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
+Moreover, `nlxb`
+explicitly returns the sum of squares as well as the residual vector and Jacobian. 
+The counts of evaluations are also returned.
+(Note that the singular values of the Jacobian are
+actually computed in the `print` and `summary` methods for the result.) 
+As we made our comparisons, we noted several potential updates to the `nlsr` documentation
+as well as that for `nls()`.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
returned weights

@@
 
+### Weights in returned functions from nls()
+
+As already noted, the function `resid()` (an alias for `residuals()`) is WEIGHTED, as are those in the
+structured object `m` returned by `nls()` or `minpack.lm::nlsLM`,
+e.g., `ans$m$resid()`. The function `nlsModel()`, which we have had to extract from the 
+base R code and explicitly `source()` because it is not exported to the working namespace,
+allows us to compute residuals for particular coefficient sets.
+
+```{r nlswtx, echo=TRUE}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+nowt
+nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+source("nlsModel.R")
+nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
+nmod$resid()
+```
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
interim output of port etc.

@@
 
+### Interim output from the "port" algorithm
+
+As the `nls()` **man** page states, when the "port" algorithm is used with the `trace` argument
+TRUE, the iterations display the objective function value which is 1/2 the sum of squares (or 
+deviance). It is likely that the trace display is embedded in the Fortran of the `nlminb`
+routine that is called to execute the "port" algorithm, but the factor of 2 discrepancy is nonetheless
+unfortunate for users.
+
+### Failure to return best result achieved
+
+If `nls()` reaches a point where it cannot continue but has not found a point where the
+relative offset convergence criterion is met, it may simply exit, especially if a 
+"singular gradient" (singular Jacobian) is found. However, this may occur AFTER the 
+function has made considerable progress in reducing the sum of squared residuals. 
+<!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
+<!-- ?? AB: Are we far enough along to include reference? -->
+Here is an abbreviated example:
+
+```{r tetrarun}
+time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
+conc <- c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
+NLSdata <- data.frame(time,conc)
+NLSstart <- c(lrc1 = -2, lrc2 = 0.25, A1 = 150, A2 = 50) # a starting vector (named!)
+NLSformula <- conc ~ A1 * exp(-exp(lrc1) * time) + A2 * exp(-exp(lrc2) * time)
+tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
+print(tryit)
+```
+
+Note that the sum of squares has been reduced from 61216 to 1.6211, but 
+unless `trace` is invoked, the user will not get any information about this.
+This almost trivial change to the `nls()` function could be 
+useful to R users.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
Partially linear parameters

@@
 
+## Feature: partially linear models and their specification
+
+The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
+effective than general approaches in finding good solutions to nonlinear least squares 
+problems when some of the parameters appear linearly. However, setting up the 
+calculations, that is, identifying which parameters are linear, is not a trivial
+task.
+
+Moreover, specifying a model to a solver should, ideally, use the same syntax across
+solver tools. Unfortunately, R allows multiple approaches within different
+modeling tools, and within `nls()` itself. This is highlighted by the case of
+partially linear models.
+
+The nonlinear modeling specifications are, of course, a development of linear ones.
+Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
+function `lm()`, which requires this model to be specified as `y ~ x`.
+
+Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
+an intuitive choice, labelled `fm2a`, that does NOT work. In this failed
+attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
+to try to add another term to the model. We believe this is unfortunate, and would
+like to see a consistent syntax. At the time of writing, we do
+not foresee a resolution for this issue. In the example, we have NOT evaluated
+the commands to save space.
+
+```{r log4ways, echo=TRUE, eval=FALSE}
+DNase1 <- subset(DNase, Run == 1) # select the data
+## using a selfStart model - do not specify the starting parameters
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
+
+## using conditional linearity - leave out the Asym parameter
+fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(xmid = 0, scal = 1),
+                 algorithm = "plinear")
+summary(fm2)
+
+## without conditional linearity
+fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1,
+                 start = list(Asym = 3, xmid = 0, scal = 1))
+summary(fm3)
+
+## using Port's nl2sol algorithm
+fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+                 algorithm = "port"))
+summary(fm4)
+
+## using conditional linearity AND Asym does NOT work
+fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
+                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                 algorithm = "plinear", trace = TRUE))
+summary(fm2a)
+```
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add indexed parameters section

@@
 
+## Feature: indexed parameters
+
+Indexed parameters -- those whose names have subscripts on some element that
+is part of the coefficient list -- 
+are a useful idea to simplify model specification in some cases. However, they
+are also a focus of trouble.
+
+The **man** file for `nls()` includes the following example of a situation in which
+parameters are indexed. It also uses the "plinear" option as an added complication.
+
+Here we use a truncated version of the example to save display space.
+
+```{r nlsindx1}
+## The muscle dataset in MASS is from an experiment on muscle
+## contraction on 21 animals.  The observed variables are Strip
+## (identifier of muscle), Conc (Cacl concentration) and Length
+## (resulting length of muscle section).
+if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
+mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE)
+mm<-droplevels(mm) # remove unused levels after truncation
+nlev <- nlevels(mm)
+withAutoprint({
+  ## The non linear model considered is
+  ##       Length = alpha + beta*exp(-Conc/theta) + error
+  ## where theta is constant. For now alpha and beta do NOT vary with Strip.
+  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
+  nl <- nlevels(mm$Strip)
+  ## We first use the plinear algorithm to fit an overall model,
+  ## ignoring that alpha and beta might vary with Strip.
+  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
+                start = list(th = 1), algorithm = "plinear")
+  summary(musc.1)
+
+  ## Then we use nls' indexing feature for parameters in non-linear
+  ## models to use the conventional algorithm to fit a model in which
+  ## alpha and beta vary with Strip.  The starting values are provided
+  ## by the previously fitted model.
+  ## Note that with indexed parameters, the starting values must be
+  ## given in a list (with names):
+  b <- coef(musc.1)
+  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
+                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
+  summary(musc.2)
+})
+```
+
+Running the example reveals that the answers for the parameters are NOT indexed. 
+<!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
+That is, we do not see `a[1], a[2], a[3]` but `a1, a2, a3`. 
+<!-- This is because the model -->
+<!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
+This is no doubt because programming for indexed parameters is extremely challenging.
+<!-- However, we believe the current structure could cause confusion and error, and  -->
+<!-- propose an alternative approach below. -->
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
bounds on parameters

@@
 
+## Feature: bounds on parameters
+
+There are many situations where the context of a problem constrains the values of parameters.
+For example, one of us was asked many years ago to estimate a model where one parameter estimate
+was negative. The client complained "But that is supposed to be the number of grain elevators
+in Saskatchewan". The number should have been a positive integer, and likely less
+than a few thousand.
+
+`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
+the manual states
+
+> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
+  value is within the bounds. Use with caution, especially where bounds are supplied.
+
+This is clearly unsatisfactory for general use, but does find bounded solutions. 
+Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
+straightforward, and it is part of the default method
+for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
+script, which we do not run here for reasons of space, shows that it does not find a
+good solution for the scaled Hobbs test problem.
+
+```{r chkbound}
+# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+wf <- data.frame(y = weed, tt = tt)
+st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
+wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
+require(minpack.lm)
+# Hobbs scaled problem with bounds, formula specification
+anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("Using nlsr::nlxb():")
+print(anlxb1)
+## nlsLM seems NOT to work properly with bounds
+anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("\n"); cat("using minpack.lm::nlsLM():")
+print(anlsLM1b)
+anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", 
+           lower = c(0, 0, 0), upper = c(2, 6, 3))
+cat("Using nls with 'port':")
+print(anlsb)
+```
+
+### Philosophical considerations
+
+Bounds on parameters raise some interesting and difficult questions about how uncertainty
+in parameter estimates should be computed or reported. That is, the traditional "standard
+errors" are generally taken to imply symmetric intervals about the point estimate in which
+the parameter may be expected to be found with some probability under certain assumptions.
+Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
+errors nor their derived statistics when bounds are active.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add code structure

@@
 
+## Issue: code structure and documentation for maintenance
+
+The `nls()` code is structured in a way that inhibits both maintenance and improvement.
+In particular, the iterative setup is such that introduction of Marquardt stabilization
+is not easily available. To obtain performance when the code was written, a large portion
+was written in C with consequent calls and returns that
+severely complicate the code. Over time, R has become much more efficient on modern computers, and
+the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
+could be reduced by moving code entirely to R.
+
+While R and its packages are generally very well documented for usage, that documentation
+rarely extends as far as it might for code maintenance. 
+The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
+where seemingly trivial differences in things like representation of numbers or
+base index for arrays lead to traps for unwary programmers.
+
+<!-- For `nls()` the concern is demonstrated by a short email query from John Nash -->
+<!-- to Doug Bates and his reply. -->
+<!-- This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on  -->
+<!-- how difficult it is to develop code in this subject area and to keep it maintainable.  -->
+<!-- We have experienced similar loss of understanding for some of our own codes. -->
+<!--     ... -->
+<!--     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R -->
+<!--     ... has the test problem and the expanded code. Around line 367 is where we are -->
+<!--     scratching our heads. The function code (from nlsModel()) is in the commented  -->
+<!--     lines below the call. This is -->
+<!--           # > setPars -->
+<!--           # function(newPars) { -->
+<!--           #   setPars(newPars) -->
+<!--           #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x} -->
+<!--           #   dev   <<- sum(resid^2) # envir = thisEnv -->
+<!--           #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr) -->
+<!--           #   QR <<- qr(.swts * gr) # envir = thisEnv -->
+<!--           #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix -->
+<!--           # } -->
+<!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
+<!--     with documentation that will be easier to follow and test, then gradually figure -->
+<!--     out how to make it more efficient. -->
+<!--     The equivalent from minpack.lm is -->
+<!--     setPars = function(newPars) { -->
+<!--                 setPars(newPars) -->
+<!--                 assign("resid", .swts * (lhs - assign("rhs", getRHS(), -->
+<!--                     envir = thisEnv)), envir = thisEnv) -->
+<!--                 assign("dev", sum(resid^2), envir = thisEnv) -->
+<!--                 assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv) -->
+<!--                 return(QR$rank < min(dim(QR$qr))) -->
+<!--             } -->
+<!--     In both there is the recursive call, which must have a purpose I don't understand. -->
+<!-- The reply was -->
+<!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
+<!--     I know that I wrote the code to use a closure for the response, covariates, etc.,  -->
+<!--     but I don't recall anything like a recursive call being necessary. -->
+<!--      I don't think I will be of much help.  My R skills have atrophied to the point  --> 
+<!--      where I wouldn't even know how to start exploring what is happening in the first  -->
+<!--      call as opposed to the recursive call. --> 
+<!--  ``` -->
+
+# Goals and progress of our effort to improve `nls()`
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
add rest

@@
 
+## Code rationalization and documentation
+
+When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
+
+- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
+  that matches the features in base R or what is packaged in `nlspkg` as described 
+  in @PkgFromRbase22. The R version can either serve as the operational code or
+  as a reference version to facilitate the maintenance of C, Fortran, or other 
+  versions of the software.
+
+- to streamline the overall `nls()` infrastructure. By this we
+  mean a re-factoring of the routines so they are better suited to maintenance of both
+  the existing `nls()` methods and features as well as new features we or others
+  would like to add.
+
+- to explain what we do, either in comments, examples, or separate maintainer documentation.
+  Since we are complaining about the lack of explanatory material for the current
+  code, we feel it is incumbent on us to provide such material for our own work, and if
+  possible for the existing code.
+
+These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
+commentary.
+
+## Rationalization of formula specifications
+
+We have seen that
+`nls()` uses a different formula specification from the default if the 
+`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
+add `algorithm="plinear"` to the standard call, making errors more likely.
+
+Nonlinear models, apart from those using the "plinear" algorithm, need all the parameters specified.
+That is, the model formula should be written down as it would be computed.
+Thus a linear model (which is valid input for nonlinear estimation) should
+be specified `y ~ a * x + b` rather than `y ~ x`. 
+Moreover, we want to avoid the partially linear parameters appearing 
+in the result object as `.lin1`, `.lin2`, etc., where the actual position 
+of parameters in the model is not explicit. We are unclear why these parameters
+begin with a dot, that is, that they have the "hidden" indicator.
+
+A possible workaround to allow consistency would be to specify the 
+partially linear parameters when the `plinear` option is specified. For example, 
+we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
+from the default specification. This would allow 
+for the output of different `algorithm` options to be consistent. However, we have not
+yet tried to code such a change, and welcome discussion from those working with
+other packages using model formulas.
+
+A further complication of model specification in R is that some formulas require
+the user to surround a term with `I()` to inhibit interpretation of arithmetic 
+operators. We would prefer that formulas be usable without this complication.
+
+<!-- It is also of interest to consider code that will detect partial linearity,  -->
+<!-- rather than expecting the user to provide the special structure and settings. Work in this -->
+<!-- area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution -->
+<!-- we can find. -->
+<!-- Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin  -->
+<!-- is about a class of models that have almost nothing in common -->
+<!-- with the present issue. The algorithm in @Birkisson2013 will need investigating further, -->
+<!-- but we suspect that it is complicated enough that it will not easily be adapted to  -->
+<!-- a program like `nls()`. Possibly a simpler algorithm may be found. -->
+
+<!-- ``` -->
+<!-- y ~ x + z + x*z -->
+<!-- y ~ a + b*x +c*z + d * (x*z)  # <-- fully specified -->
+<!-- y ~ a*Fn1(b, c, x, z)/Fn2() -->
+
+<!-- ``` -->
+
+## Rationalization of indexed models
+
+Indexed models clearly have a place in some areas of research. 
+<!-- We need to be able to process models, as the example above, where the -->
+<!-- model formula is  -->
+We need to be able to process models such as 
+`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
+`Length` and `Conc` are available for all observations and parameter `th` applies
+to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
+could be an experimental run number, or some other categorization.
+
+Given that `[ ]` provides the R method to index arrays, the presence of square
+brackets in a model
+formula signals indexing. The question is how to perform the correct calculations.
+
+<!-- While tedious (??maybe dummies or fastdummies packages make it less so?),  -->
+<!-- we could set up the model as a sum of similar models with dummy -->
+<!-- variables to turn "on" or "off" the particular case, transforming the model -->
+<!-- parameters to simple, non-indexed ones (as does `nls()`). -->
+
+Our view is that inclusion of the indexing **within** `nls()` or similar tools introduces 
+complexity that could be avoided by a suitable wrapper function. The modeling
+formula must be expanded to encompass all the index cases, likely with indexed
+parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
+a wrapper could process output to reconstitute the indexing of parameters.
+
+## Streamlining code
+
+We are pessimistic that the overall structure of `nls()` can be streamlined due
+to the entanglement of so many features with the complicated mix of R, C and
+Fortran. Indeed, despite digging into the code over some months, we do not feel
+confident that we sufficiently understand it nor that we could maintain it. On the other
+hand, we do believe equivalent functionality can be built, but with at least
+some differences in how the features will be accessed.
+
+## Tests and use-case examples
+
+Maintainers of packages need suitable tests and use-case examples in order:
+
+- to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace;
+- to test individual solver functions to ensure they work across the range of calling 
+  mechanisms, that is, different ways of supplying inputs to the solver(s);
+- to pose "silly" inputs to see if these bad inputs are caught by the programs.
+
+Such goals align with the aims of **unit testing** 
+(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
+and the conventional R package testing tools).
+
+## A testing package: NLSCompare
+
+One of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
+A primary design objective of this is to allow the summarisation of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
+
+<!-- Arkajyoti -- do you want to expand? -->
+<!-- No. I think it would be tangential to the goal of the paper. We can have a -->
+<!-- separate article for that, maybe? -->
+<!-- Trying to trim a lot of words, so am making reference to package. Is it ready -->
+<!-- for others to use? -->
+
+<!-- When we have a "new" or trial solver function, we would like to know if it gives  -->
+<!-- acceptable results on a range of sample problems of different types, starting -->
+<!-- parameters, input conditions, constraints, subsets, weights or other settings.  -->
+<!-- Ideally we want to be able to get a summary that is easy to read and assess. For -->
+<!-- example, one approach would be to list the names of a set of tests with a red, green -->
+<!-- or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last -->
+<!-- category would be a problem with constraints that the solver is not designed to  -->
+<!-- handle.  -->
+
+<!-- To accomplish this, we need a suitable "runner" program that can be supplied with -->
+<!-- the name of a solver or solvers and a list of test problem cases. Problems generally -->
+<!-- have a base setup -- a specification of the function to fit as a formula, some data -->
+<!-- and a default starting set of parameters. Other cases can be created by imposing -->
+<!-- bounds or mask constraints, subsets of the data, and different starts.  -->
+
+<!-- How to set up this "runner" and its supporting infrastructure is non-trivial. While -->
+<!-- the pieces are not as complicated as the inter-related parts of the solvers, especially -->
+<!-- `nls()`, the categorization of tests, their documentation, and the structuring to make -->
+<!-- running them straightforward and easy requires much attention to detail. -->
+
+<!-- Some considerations for our test scripts: -->
+
+<!-- - Is it useful to have a "base" script for each family of test problem, with numbered particular -->
+<!--   cases? That is, if we run the scripts in order, we can avoid some duplication of code and -->
+<!--   data. -->
+
+<!-- - While we have developed some tags to document the test problem families and cases, we believe -->
+<!--   that such tags (essentially summary documentation) will continue to need revision as different -->
+<!--   tools and problems are included in scope of `nlsCompare`. -->
+
+<!-- - Similarly, we expect that there will be ongoing review of the structure of the result files. -->
+
+<!-- # Progress so far -->
+
+<!-- Much of our work is reported in https://github.com/nashjc/RNonlinearLS . -->
+
+<!-- ## Reports and articles -->
+
+<!-- - The present article is the central report of the Google Summer or Code 2021 project -->
+<!--   "Improvements to nls()". -->
+
+<!-- - VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and -->
+<!--   the many choices in their implementation for nonlinear least squares. This is still -->
+<!--   a DRAFT at 2022-6-20. -->
+
+<!-- Response to query about TestsDoc -->
+<!-- 4. Testing infrastructure, I think, is a general topic and not -->
+<!-- particularly relevant to nls(); i.e., it is important for all packages. -->
+<!-- So, I don't think we need to include that in RefactoringNLS. Could be a -->
+<!-- separate blog maybe?   -->
+
+<!-- Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
+<!-- useful. Perhaps incorporate in the nlsCompare repo. -->
+<!-- - TestsDoc.Rmd (?citation) is a survey of testing tools in R. It has more  -->
+<!--   general possibilities  -->
+<!--   and fits into the subject of regression testing, in which case a  -->
+<!--   more extensive literature review will be needed.  Note that this -->
+<!--   document reflects the work in the the "Problem sets and test infrastructure" -->
+<!--   below. -->
+
+  <!-- need to incorporate MachID and -->
+  <!-- the many tests. Arkajyoti: should we think of working this up into a paper for -->
+  <!-- JSS or the R-Journal? My  -->
+  <!-- view is that this COULD be a long term side-interest for your academic work, but -->
+  <!-- that would depend on your own interests as well as opportunities. -->
+
+<!-- This needs more focus of where it may appear. I'm thinking of a blog post -- JN-->
+<!-- - DerivsNLS.Rmd ?citation: a document to explain different ways in which Jacobian information -->
+<!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->
+<!--   a DRAFT of a script to provide examples. -->
+
+## Documentation and resources
+
+In our investigation, we built several resources, which are now part of 
+the repository *https://github.com/nashjc/RNonlinearLS/*. Particular items
+are:
+
+- A BibTex bibliography for use with all documents in this project, but which 
+  has wider application to nonlinear least squares projects in general
+  (*https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib*).
+
+- `MachID.R` offers a suggested concise summary function
+  to identify a particular computational system used for tests. A discussion 
+  of how it was built and the resources needed across platforms is given in
+  at *https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary*.
+
+- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
+  `nls()` code in R-base.
+
+- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
+  a review of optimization in R. Ideas from the present work have been instrumental
+  in the creation of @NashWires22.
+
+<!-- ?? probably don't need to include -->
+<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
+<!--   A project diary. -->
+
+<!-- ?? does not seem to add enough? -->
+<!-- ## Problem sets and test infrastructure -->
+
+<!-- ?? need to clean up -->
+
+<!-- We have several test problems and variants thereof in the `inst/scripts/` directory of -->
+<!-- the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare). -->
+<!-- We direct the reader to that package for documentation of the test infrastructure,  -->
+<!-- in particular the problems and methods files (`problems.csv` and `methods.csv`) and -->
+<!-- the various functions invoked by `run.R` to produce an output file in CSV form also. -->
+
+<!-- Towards the end of the project, we have focused our attention on the `nlsCompare` -->
+<!-- package, which looks at evaluating and comparing functions for nonlinear least -->
+<!-- squares problems. We use many of the same tests as checks that are or will be  -->
+<!-- built into our packages for such problems e.g., `nlsalt`. These use the `testthat` -->
+<!-- structure and may include verification of outputs that are specific to a package, -->
+<!-- such as the upper triangular matrix R of the QR decomposition that has been  -->
+<!-- computed for a Jacobian. Since such an object will not be computed by all methods,  -->
+<!-- testing it in `nlsCompare` makes no sense, and in that package we concentrate on -->
+<!-- the minimum sum of squares and the associated model parameters. -->
+
+<!-- ## Code development -->
+
+<!-- A summary of our main results: -->
+
+<!-- - nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch -->
+
+<!-- - nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED.  -->
+<!--   The effort showed that the structure of the programs was difficult to follow,  -->
+<!--   undocumented, and unsuited to adding improvements like the Marquardt stabilization.  -->
+<!--   We were able to get a pure-R version of `numericDeriv()` and rework most of the functions  -->
+<!--   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project -->
+<!--   formally ends, but collaboration is likely needed with workers who have a deep knowledge -->
+<!--   of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`. -->
+
+<!-- <!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  --> 
+<!-- <!--   fix by JN to the relative offset convergence criterion failure when there are small  --> 
+<!-- <!--   residuals in problems sent to ``nls()`. Now part of R. -->
+
+<!-- - the safeguarded relative offset convergence criterion that allows small-residual problems -->
+<!--   to ba handled with `nls()` is now in the R base code. -->
+
+<!-- - `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features?? -->
+<!--   This is an interim package for exploration of ideas and will NOT, as it currently -->
+<!--   is presented, become a distributed package. -->
+
+<!-- - nlsralt: this is intended to be a modified version of Nash and Murdoch package  -->
+<!--   `nlsr` to examine possible improvements discovered as a result of this project.  -->
+<!--   This is EXPERIMENTAL. -->
+
+<!-- For example,  -->
+<!-- `nlsr` does not currently use the `subset` argument correctly. We also need to document -->
+<!-- changes e.g., to numerical derivatives. However, these are tasks for JN rather than AB. -->
+
+# Strategic choices in nonlinear model estimation
+
+It is possible that `nls()` will remain more or less as it has been for the past
+two decades. Given its importance to R, the focus of discussion should be the measures
+needed to secure its continued operation for legacy purposes and how that may be approached.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program objects (R functions)
+that are created by tools such as `nls()` should have minimal cross-linkages and side-effects.
+The aspects of `nls()` that have given us the most trouble:
+
+- The functions that compute the residuals and jacobians often presume
+  the data and parameters needed are available in a particular environment. As long
+  as the correct environment is used, this provides a compact syntax to invoke the
+  calculations. The danger is that the wrong data is accessed if the internal search finds a
+  valid name that is not the object we want.
+
+- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
+  are similarly taken from the first available source, making for a very simple
+  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
+  difficult for those of us trying to maintain or improve the code to be certain we have the
+  context correct.
+
+- The mixing of R, C and Fortran code was necessary for computational performance in the past.
+  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
+  We believe in having at least a working reference version of code in a single programming language. 
+  If necessary, by
+  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
+  parts of the reference code.
+
+- A design that isolates the setup, solution, and post-solution parts of
+  for complicated calculations reduces the number of objects that must be kept in 
+  alignment. On the other hand, it may mean users have to add more commands to complete
+  their computations.
+
+- All iterative codes should return the best solution they have found so far, even if there
+  are untoward conditions reached, such as a singular Jacobian. This modification could be
+  made to `nls()` in the short term, correcting the issue identified above under "Failure to
+  return best result achieved".
+
+- Given the existence of examples of good practices such as analytic derivatives, stabilized 
+  solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
+  be moving to incorporate them.
+
+We anticipate that tools like `nlsr` will become more attractive as users discover
+the increased robustness and general capabilities, along with the more recent documentation.
+From a design point of view, a key difference in approach between `nls()` and 
+`nlsr::nlxb()` is that `nls()` 
+builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
+iteration can be executed and other statistical information such as profiles can be
+computed, while `nlxb()` returns quite limited information, and 
+computes what is needed on an "as and when" basis. This follows a path that one of us
+(JN) established almost 50 years ago with the software that became @cnm79, separating
+setup, solver, and post-solution analysis phases of computations. There is, however,
+the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver
+of `nlsr` then calls `nls()` to build the object of class `nls` for which profile log-likelihood
+information can be computed.
+<!-- Here, that last  -->
+<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
+<!-- package. -->
+
+<!-- The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable -->
+<!-- entangling of the different functions and capabilities. Even after -->
+<!-- both the preparation for and carrying out of the present project, we do not feel confident -->
+<!-- to explain the code completely, nor to maintain it. However, we have made some  -->
+<!-- forays into render parts of the code in R, and to understand where difficulties -->
+<!-- lie. -->
+
+<!-- ??fix next bits -->
+<!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
+<!--   would, however, be a painful and mainly political activity. -->
+
+<!-- - The tools for analytic derivatives that exist in parts of `nls()` for self-starting models and -->
+<!--   in `nlsr` for problems specified as models that use expressions for which derivatives exist  -->
+<!--   should be more generally available, but the task is one of tedium and limited reward to the -->
+<!--   persons who do it, while benefiting the R community substantially. -->
+
+<!-- - Similarly, default methods need stabilized solvers e.g., Marquardt methods, though to date -->
+<!--   a mechanism to incorporate such modifications in `nls()` has eluded us. -->
+
+<!-- - Bounds constraints are sufficiently important that they should be part of the default -->
+<!--   methods -->
+<!--   (rather than algorithm = "port" for `nls()`). `minpack.lm` should handle them correctly. -->
+
+# Acknowledgement
+
+Hans Werner Borchers has been helpful in developing the proposal for this project
+and in comments on this and related work. Heather Turner co-mentored the project 
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
+Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
+to succinctly summarize computing environments.
+
 # References

### 2023-02-03 RefactoringNLS/JNAB230203.Rmd
cleanup line length

--- 2023-02-03 RefactoringNLS/JNAB230203.Rmd
@@
 from the Comprehensive R Archive Network (*https://cran.r-project.org*),
-for estimating nonlinear statistical models. The function dates to the 1980s and the work
-related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+for estimating nonlinear statistical models. The function dates to the 1980s and 
+the work related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
 
-The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
-nonlinear models that are expressed as **formulas**. In particular, we note that it
+The `nls()` function has a remarkable and comprehensive set of capabilities for
+estimating nonlinear models that are expressed as **formulas**. In particular, we note that it
 
@@
 - permits parameters to be indexed over a set of related data
-- produces measures of variability (i.e., standard error estimates) for the estimated parameters
-- has related profiling capabilities for exploring the likelihood surface as parameters are changed
+- produces measures of variability (i.e., standard error estimates) for the 
+  estimated parameters
+- has related profiling capabilities for exploring the likelihood surface as
+  parameters are changed
 - links to a number of pre-coded (`selfStart`) models
@@
 and overly patched. It is, to
-our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
-as we suggest below.
+our mind, essentially unmaintainable. Moreover, its underlying methods can and 
+should be improved, as we suggest below.
 
@@
 
-A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
-(i.e., perfect fit) data was corrected thanks to suggestions by one of us.
+A previous issue with `nls()` that prevented it from providing parameter 
+estimates for zero-residual
+(i.e., perfect fit) data was corrected using a suggestion by one of us.
 
@@
 This issue can be easily resolved. The "termination test" for the **program** rather than for 
-"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
-(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
-and estimates how much the sum of squares loss function should decrease. This estimate is divided 
+"convergence" of the underlying **algorithm** is the Relative Offset 
+Convergence Criterion (see @BatesWatts81). This projects the proposed step 
+in the parameter vector on the gradient 
+and estimates how much the sum of squares loss function should decrease. 
+This estimate is divided 
 by the current size of the loss function to avoid scale issues.
@@
 Adding a small quantity to the loss function before dividing avoids trouble. 
-In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
+In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional
+parameter `scaleOffset` 
 with a default value of zero. Setting it to a small number -- 1.0 is 
-a reasonable choice --
-allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
-**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
-We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.
+a reasonable choice -- allows small-residual problems (i.e., near-exact fits) 
+to be dealt with easily. We call this the 
+**safeguarded relative offset convergence criterion**. The default value 
+gives the legacy behaviour. We are pleased to note that this improvement is 
+now in the R distributed code since version 4.1.0.
 
@@
 
-The convergence criterion above, the principal termination control of `nls()`, leaves out 
-some possibilities that could
-be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
+The convergence criterion above, the principal termination control of `nls()`, 
+leaves out some possibilities that could be useful for some problems. The 
+package `nlsr` (@nlsr2019manual) already offers both 
 the safeguarded relative offset test (**roffset**) as well as a **small sum of 
@@
 code. We prefer to use the number of times the residuals or the jacobian have
-been computed and put upper limits on these. Our codes exit (terminate) when these limits
-are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
+been computed and put upper limits on these. Our codes exit (terminate) when 
+these limits are reached. Generally we prefer larger limits than the default 
+`maxiter = 50` of `nls()`,
 but that may reflect the more difficult problems we have encountered because 
@@
 an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
-work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
-@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
+work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963,
+@Levenberg1944, @jn77ima), and versions of it have been implemented in 
+packages `minpack.lm` and `nlsr`.
 In our work, we prepared experimental prototypes of `nls` that incorporate
@@
 *https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
-`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that the `nls()` error message is correct,
-but this is not helpful in obtaining a solution.
+`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that 
+the `nls()` error message is correct, but this is not helpful in obtaining a
+solution.
 
@@
 See *https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS*.
-<!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
-<!-- (?? need citation -- currently **DerivsNLS.pdf**). -->
 
@@
 
-The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
-while that
-of `nls()` mostly uses a numerical approximation. Some `selfStart` models do provide
-"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
-version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
-the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, though in our experience differences are usually minor.
+The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic
+expressions, while that of `nls()` mostly uses a numerical approximation. Some
+`selfStart` models do provide "gradient" information. `minpack.lm::nlsLM()` 
+invokes `numericDeriv()` in its local version of `nlsModel()`, but it appears 
+to use an internal approximate Jacobian code from the original Fortran 
+`minpack` code, namely, `lmdif.f`. Such differences in approach can lead to 
+different behaviour, though in our experience differences are usually minor.
 
@@
 
-Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` file 
-and in package `nlraa` (@MiguezNLRAA2021) may
-provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
-each model, and these Jacobians may be the analytic forms. The `nls()` function, after
-computing the "right hand side" or `rhs` of the residual, then checks to see if the
-"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
+Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` 
+file and in package `nlraa` (@MiguezNLRAA2021) may
+provide the Jacobian in the "gradient" attribute of the "one-sided" formula that 
+defines each model, and these Jacobians may be the analytic forms. The `nls()` 
+function, after computing the "right hand side" or `rhs` of the residual, then 
+checks to see if the "gradient" attribute is defined, and, if not, uses
+`numericDeriv` to compute a Jacobian
 into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
-functions. The use of analytic Jacobians
-almost certainly contributes to the good performance of `nls()` on `selfStart` models.
+functions. The use of analytic Jacobians almost certainly contributes to the good
+performance of `nls()` on `selfStart` models.
 
@@
 `nls()`, `nlsr::nlxb()` and `minpack.lm::nlsLM()` all have a `weights` argument that
-specifies a vector of weights the same length as the number of residuals. Each residual
-is multiplied by the square root of the corresponding weight. The values returned by the
-`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
-used to compute raw residuals.
+specifies a vector of weights the same length as the number of residuals. Each 
+residual is multiplied by the square root of the corresponding weight. The values
+returned by the `residuals()` function are weighted, and the `fitted()` or
+`predict()` function are used to compute raw residuals.
 
@@
 `nls()` accepts an argument `subset`. This acts through the mediation of
-`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
-`/src/library/stats/src/nls.C`.
+`model.frame` and is not clearly obvious in the source code files
+`/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`.
 
 While the implementation of subset at the level of the call to `nls()` has a certain
-attractiveness, it does mean that the programmer of the solver needs to be aware of the
-source (and value) of objects such as the data, residuals and Jacobian. This is overly
-complicated. By preference, 
+attractiveness, it does mean that the programmer of the solver needs to be aware of 
+the source (and value) of objects such as the data, residuals and Jacobian. This 
+is overly complicated. By preference, 
 we would implement subsetting by means of zero-value weights, with observation counts
@@
 
-`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
-source code, often being handled behind the scenes after referencing the option `na.action`.
-A useful, but possibly dated, description is given in:
+`na.action` is an argument to the `nls()` function, but it does not appear obviously 
+in the source code, often being handled behind the scenes after referencing the
+option `na.action`. A useful, but possibly dated, description is given in:
 *https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
 
-The typical default action, which can be seen by using the command `getOption("na.action")`
+The typical default action, which can be seen by using the command 
+`getOption("na.action")`
 is `na.omit`. This option essentially omits from computations any observations
 containing missing values (i.e. any row of a data frame containing an NA). 
-`na.exclude` does much of the same for computations, but keeps the rows with NA elements so 
-that predictions are in the correct row position. We recommend that workers actually test 
-output to verify the behaviour is as wanted.
+`na.exclude` does much of the same for computations, but keeps the rows with NA 
+elements so that predictions are in the correct row position. We recommend that
+workers actually test output to verify the behaviour is as wanted.
 
@@
 
-`na.pass` simply passes on data "as is", while `na.fail` will essentially stop if any missing
-values are present.
+`na.pass` simply passes on data "as is", while `na.fail` will essentially stop 
+if any missing values are present.
 
@@
 
-Indeed, the argument only gets used when `nls()` is about to return its result object, and the
-element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
-be confusing.) However, the model frame is used within the function code in the form of the object
-`mf`. We feel that users could benefit from more extensive documentation and examples of its use.
+Indeed, the argument only gets used when `nls()` is about to return its result 
+object, and the element `model` is NULL unless the calling argument `model` is 
+TRUE. (Using the same name could be confusing.) However, the model frame is 
+used within the function code in the form of the object `mf`. We feel that 
+users could benefit from more extensive documentation and examples of its use.
 
@@
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
-when R is launched, will be loaded. The overwhelming proportion of R users in our
-acquaintance avoid saving the workspace because of the danger of lurking data and
-functions which may cause unwanted results.
+suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the 
+path, when R is launched, will be loaded. The overwhelming proportion of R users 
+in our acquaintance avoid saving the workspace because of the danger of lurking 
+data and functions which may cause unwanted results.
 
@@
 Nonlinear estimation algorithms are almost all iterative and need a set of starting
-parameters. `nls()` offers a special class of modeling functions called **`selfStart`** 
-models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
+parameters. `nls()` offers a special class of modeling functions called
+**`selfStart`** models. There are a number of these in base R
+(`./src/library/stats/R/zzModels.R`)
 and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
@@
 `nls()`, in part 
-because it may call that function. Such circular references are, in our view, dangerous.
-Moreover, we believe that it would be helpful to have `selfStart` models that allow
-users to explicitly provide values other than those suggested for the starting parameters.
+because it may call that function. Such circular references are, in our view, 
+dangerous. Moreover, we believe that it would be helpful to have `selfStart` 
+models that allow users to explicitly provide values other than those suggested 
+for the starting parameters.
 
-Other concerns are revealed by the example below. Here, the `SSlogis` `selfStart` function 
-is used to generate a set
+Other concerns are revealed by the example below. Here, the `SSlogis` 
+`selfStart` function is used to generate a set
 of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
-is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
-for this model can be transformed to those of 
-another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.
+is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show 
+how the starting parameters for this model can be transformed to those of 
+another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\,
+b2\,*\,exp(-b3\,*\,t))$.
 
 The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
-<!-- ``` -->
-<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
-<!--     selfStart( -->
-<!--         function(input, Asym, xmid, scal) -->
-<!--         { -->
-<!--               .expr1 <- xmid - input -->
-<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
-<!--               .expr4 <- 1 + .expr3 -->
-<!--               .value <- Asym/.expr4 -->
-<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
-<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
-<!--               { -->
-<!--             		  .expr10 <- .expr4^2 -->
-<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
-<!--                   .grad[, "Asym"] <- 1/.expr4 -->
-<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
-<!-- 		              .grad[, "scal"] <- xm * .e2 -->
-<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
-<!--                   attr(.value, "gradient") <- .grad -->
-<!--               } -->
-<!--               .value -->
-<!--         }, -->
-<!--         initial = function(mCall, data, LHS, ...) { -->
-<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
-<!--               if(nrow(xy) < 4) { -->
-<!--                   stop("too few distinct input values to fit a logistic model") -->
-<!--               } -->
-<!--               z <- xy[["y"]] -->
-<!--               ## transform to proportion, i.e. in (0,1) : -->
-<!--               rng <- range(z); dz <- diff(rng) -->
-<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
-<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
-<!--               aux <- coef(lm(x ~ z, xy)) -->
-<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
-<!--                                data = xy, -->
-<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
-<!--                                algorithm = "plinear", ...)) -->
-<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
-<!--                        mCall[c("Asym", "xmid", "scal")]) -->
-<!--         }, -->
-<!--         parameters = c("Asym", "xmid", "scal")) -->
-<!-- ``` -->
 This R function includes analytic expressions for the Jacobian ("gradient").
@@
 wonder why the programmers have chosen to save so many quantities in "hidden"
-variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
-command, making them difficult to access.
+variables, i.e., with names preceded by ".". These are then not displayed by the 
+`ls()` command, making them difficult to access.
 
-In the event that a `selfStart` model is not available, `nls()` sets all the starting parameters 
-to 1. This is, in our view, tolerable, but could be improved by using a set of values
-that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
+In the event that a `selfStart` model is not available, `nls()` sets all the 
+starting parameters to 1. This is, in our view, tolerable, but could be improved 
+by using a set of values that are all slightly different, which, in the case of 
+the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
 would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
@@
 could be provided quite simply instead of all 1's.
-
-<!-- ```{r hobbsSSlogis} -->
-<!-- weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, -->
-<!--           38.558, 50.156, 62.948, 75.995, 91.972) -->
-<!-- tt <- 1:12 -->
-
-<!-- NLSformula0 <- y ~ b1/(1+b2*exp(-b3*tt)) -->
-<!-- NLSformula <- y ~ SSlogis(tt, Asym, xmid, scal) -->
-<!-- NLSformulax <- y ~ Asym/(1+exp((xmid-tt)/scal)) -->
-<!-- NLStestdata <- data.frame(y=weed, tt=tt)  -->
-<!-- s0 <- getInitial(NLSformula, NLStestdata) -->
-<!-- print(s0) -->
-<!-- # We transform the parameters for the NLSformula0 model of original specification. -->
-<!-- s1<-list(b1=s0[1], b2=exp(s0[2]/s0[3]), b3=1/s0[3]) -->
-<!-- print(as.numeric(s1)) -->
-<!-- # No actual improvement because nls() has been already used to get the starting values, -->
-<!-- # but we do get SEs -->
-<!-- hobblog<-nls(NLSformula0, data=NLStestdata, start=s1) -->
-<!-- summary(hobblog) -->
-<!-- deviance(hobblog) -->
-<!-- # nls fails without selfStart -- singular gradient, even on stabilized formula -->
-<!-- try(hobblogx<-nls(NLSformulax, data=NLStestdata)) -->
-<!-- # But Marquardt is able to get a solution easily -->
-<!-- library(nlsr) -->
-<!-- hobblogxx<-nlxb(NLSformulax, data=NLStestdata, start=c(Asym=1, xmid=1, scal=1)) -->
-<!-- hobblogxx -->
-<!-- ``` -->
-
-<!-- ### selfStart models in base R -->
-
-<!-- The following models are provided (in file ./src/library/stats/R/zzModels.R) -->
-
-<!-- ``` -->
-<!-- SSasymp         - asymptotic regression model -->
-<!-- SSasympOff      - alternate formulation of asymptotic regression model with offset -->
-<!-- SSasympOrig     - exponential curve through the origin to an asymptote -->
-<!-- SSbiexp         - y ~ ~ A1 * exp(-exp(lrc1)*input) + A2 * exp(-exp(lrc2) * input) -->
-<!-- SSfol           - y ~ Dose * (exp(lKe + lKa - lCl) * (exp(-exp(lKe) * input) - -->
-<!--                           exp(-exp(lKa) * input))/(exp(lKa) - exp(lKe))) -->
-<!-- SSfpl           - four parameter logistic model -->
-<!-- SSlogis         - three parameter logistic model -->
-<!-- SSmicmen        - Michaelis-Menten model for enzyme kinetics -->
-<!-- SSgompertz2     - Gompertz model for growth curve data -->
-<!-- SSweibull       - Weibull model for growth curve data -->
-<!-- ``` -->
 
@@
 
-                  -----------------------------------------------------------
-
-<!-- ### Example output -->
-
-<!-- To illustrate, let us run the Croucher example. -->
-
-<!-- ```{r nlsoutx} -->
-<!-- # Croucher example -->
-<!-- xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9) -->
-<!-- ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001) -->
-<!-- p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2) -->
-<!-- NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) -->
-<!-- NLSdata<-data.frame(xdata, ydata) -->
-<!-- # Try full output version of nls -->
-<!-- library(nlspkg) # use the packaged version of nls() -->
-<!-- result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE) -->
-<!-- # We can display the result with several commands -->
-<!-- # str(result) -- displays large amount of material - suppressed here -->
-<!-- #    as it is too wide for the page -->
-<!-- # result # displays the object -->
-<!-- # ls(result) # to list the elements of the output -->
-<!-- # ls(result$m) # and in particular the "m" object -->
-<!-- ``` -->
-
-<!-- ### Concerns with content of the nls result object -->
-
-The `nls` object contains some elements that are awkward to produce by other algorithms,
-but some information that would be useful is not presented in a clear manner. Moreover, the
-complexity of the object is a challenge to users.
+The `nls` object contains some elements that are awkward to produce by other 
+algorithms, but some information that would be useful is not presented in a 
+clear manner. Moreover, the complexity of the object is a challenge to users.
 
@@
 ```
-The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
-Moreover, `nlxb`
+The result object from `nlsr::nlxb()` is still much smaller than the one `nls()`
+returns. Moreover, `nlxb`
 explicitly returns the sum of squares as well as the residual vector and Jacobian. 
@@
 actually computed in the `print` and `summary` methods for the result.) 
-As we made our comparisons, we noted several potential updates to the `nlsr` documentation
-as well as that for `nls()`.
+As we made our comparisons, we noted several potential updates to the `nlsr`
+documentation as well as that for `nls()`.
 
@@
 
-As already noted, the function `resid()` (an alias for `residuals()`) is WEIGHTED, as are those in the
-structured object `m` returned by `nls()` or `minpack.lm::nlsLM`,
-e.g., `ans$m$resid()`. The function `nlsModel()`, which we have had to extract from the 
-base R code and explicitly `source()` because it is not exported to the working namespace,
-allows us to compute residuals for particular coefficient sets.
+As already noted, the function `resid()` (an alias for `residuals()`) is 
+WEIGHTED, as are those in the structured object `m` returned by `nls()` or
+`minpack.lm::nlsLM`, e.g., `ans$m$resid()`. The function `nlsModel()`, which 
+we have had to extract from the base R code and explicitly `source()` because 
+it is not exported to the working namespace, allows us to compute residuals 
+for particular coefficient sets.
 
@@
 
-As the `nls()` **man** page states, when the "port" algorithm is used with the `trace` argument
-TRUE, the iterations display the objective function value which is 1/2 the sum of squares (or 
-deviance). It is likely that the trace display is embedded in the Fortran of the `nlminb`
-routine that is called to execute the "port" algorithm, but the factor of 2 discrepancy is nonetheless
+As the `nls()` **man** page states, when the "port" algorithm is used with the 
+`trace` argument TRUE, the iterations display the objective function value which 
+is 1/2 the sum of squares (or deviance). It is likely that the trace display is
+embedded in the Fortran of the `nlminb` routine that is called to execute the 
+"port" algorithm, but the factor of 2 discrepancy is nonetheless
 unfortunate for users.
@@
 
-If `nls()` reaches a point where it cannot continue but has not found a point where the
-relative offset convergence criterion is met, it may simply exit, especially if a 
-"singular gradient" (singular Jacobian) is found. However, this may occur AFTER the 
-function has made considerable progress in reducing the sum of squared residuals. 
-<!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
-<!-- ?? AB: Are we far enough along to include reference? -->
+If `nls()` reaches a point where it cannot continue but has not found a point 
+where the relative offset convergence criterion is met, it may simply exit,
+especially if a "singular gradient" (singular Jacobian) is found. However, this 
+may occur AFTER the function has made considerable progress in reducing the sum 
+of squared residuals.
+
 Here is an abbreviated example:
@@
 The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
-effective than general approaches in finding good solutions to nonlinear least squares 
-problems when some of the parameters appear linearly. However, setting up the 
-calculations, that is, identifying which parameters are linear, is not a trivial
-task.
+effective than general approaches in finding good solutions to nonlinear least 
+squares problems when some of the parameters appear linearly. However, setting 
+up the calculations, that is, identifying which parameters are linear, is not a
+trivial task.
 
@@
 The nonlinear modeling specifications are, of course, a development of linear ones.
-Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
-function `lm()`, which requires this model to be specified as `y ~ x`.
+Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear 
+modeling function `lm()`, which requires this model to be specified as `y ~ x`.
 
-Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
-an intuitive choice, labelled `fm2a`, that does NOT work. In this failed
-attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
-to try to add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing, we do
-not foresee a resolution for this issue. In the example, we have NOT evaluated
-the commands to save space.
+Within `nls()`, consider the following FOUR different specifications for the same 
+problem, plus an intuitive choice, labelled `fm2a`, that does NOT work. In this
+failed attempt, putting the `Asym` parameter in the model causes the `plinear`
+algorithm to try to add another term to the model. We believe this is 
+unfortunate, and would like to see a consistent syntax. At the time of writing, 
+we do not foresee a resolution for this issue. In the example, we have NOT 
+evaluated the commands to save space.
 
@@
 
-There are many situations where the context of a problem constrains the values of parameters.
-For example, one of us was asked many years ago to estimate a model where one parameter estimate
-was negative. The client complained "But that is supposed to be the number of grain elevators
-in Saskatchewan". The number should have been a positive integer, and likely less
-than a few thousand.
+There are many situations where the context of a problem constrains the values of 
+parameters. For example, one of us was asked many years ago to estimate a model 
+where one parameter estimate was negative. The client complained "But that is
+supposed to be the number of grain elevators in Saskatchewan". The number 
+should have been a positive integer, and likely less than a few thousand.
 
-`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
-the manual states
+`nls()` can impose bounds on parameters, but only if the `port` algorithm is used.
+Unfortunately, the manual states
 
@@
 This is clearly unsatisfactory for general use, but does find bounded solutions. 
-Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
-straightforward, and it is part of the default method
-for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
-script, which we do not run here for reasons of space, shows that it does not find a
-good solution for the scaled Hobbs test problem.
+Inclusion of bounds on parameters in nonlinear least squares computations is
+relatively straightforward, and it is part of the default method
+for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but 
+the following script
+<!-- , which we do not run here for reasons of space,  -->
+shows that it does not find a good solution for the scaled Hobbs test problem.
 
@@
 
-Bounds on parameters raise some interesting and difficult questions about how uncertainty
-in parameter estimates should be computed or reported. That is, the traditional "standard
-errors" are generally taken to imply symmetric intervals about the point estimate in which
-the parameter may be expected to be found with some probability under certain assumptions.
+Bounds on parameters raise some interesting and difficult questions about how 
+uncertainty in parameter estimates should be computed or reported. That is, the
+traditional "standard errors" are generally taken to imply symmetric intervals 
+about the point estimate in which the parameter may be expected to be found with 
+some probability under certain assumptions.
 Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
 errors nor their derived statistics when bounds are active.
+
+## Issue: code structure and documentation for maintenance
+
+The `nls()` code is structured in a way that inhibits both maintenance and 
+improvement. In particular, the iterative setup is such that introduction 
+of Marquardt stabilization is not easily available. To obtain performance 
+when the code was written, a large portion was written in C with consequent 
+calls and returns that severely complicate the code. Over time, R has become 
+much more efficient on modern computers, and the need to use compiled C and 
+Fortran is less critical. Moreover, the burden for maintenance
+could be reduced by moving code entirely to R.
+
+While R and its packages are generally very well documented for usage, that 
+documentation rarely extends as far as it might for code maintenance. 
+The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
+where seemingly trivial differences in things like representation of numbers or
+base index for arrays lead to traps for unwary programmers.
+
+# Goals and progress of our effort to improve `nls()`
+
+## Code rationalization and documentation
+
+When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
+
+- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
+  that matches the features in base R or what is packaged in `nlspkg` as described 
+  in @PkgFromRbase22. The R version can either serve as the operational code or
+  as a reference version to facilitate the maintenance of C, Fortran, or other 
+  versions of the software.
+
+- to streamline the overall `nls()` infrastructure. By this we
+  mean a re-factoring of the routines so they are better suited to maintenance of both
+  the existing `nls()` methods and features as well as new features we or others
+  would like to add.
+
+- to explain what we do, either in comments, examples, or separate maintainer
+  documentation. Since we are complaining about the lack of explanatory material 
+  for the current code, we feel it is incumbent on us to provide such material 
+  for our own work, and if possible for the existing code.
+
+These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
+commentary.
+
+## Rationalization of formula specifications
+
+We have seen that
+`nls()` uses a different formula specification from the default if the 
+`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
+add `algorithm="plinear"` to the standard call, making errors more likely.
+
+Nonlinear models, apart from those using the "plinear" algorithm, need all the
+parameters specified.
+That is, the model formula should be written down as it would be computed.
+Thus a linear model (which is valid input for nonlinear estimation) should
+be specified `y ~ a * x + b` rather than `y ~ x`. 
+Moreover, we want to avoid the partially linear parameters appearing 
+in the result object as `.lin1`, `.lin2`, etc., where the actual position 
+of parameters in the model is not explicit. We are unclear why these parameters
+begin with a dot, that is, that they have the "hidden" indicator.
+
+A possible workaround to allow consistency would be to specify the 
+partially linear parameters when the `plinear` option is specified. For example, 
+we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
+from the default specification. This would allow 
+for the output of different `algorithm` options to be consistent. However, we have 
+not yet tried to code such a change, and welcome discussion from those working with
+other packages using model formulas.
+
+A further complication of model specification in R is that some formulas require
+the user to surround a term with `I()` to inhibit interpretation of arithmetic 
+operators. We would prefer that formulas be usable without this complication.
+
+## Rationalization of indexed models
+
+Indexed models clearly have a place in some areas of research. 
+<!-- We need to be able to process models, as the example above, where the -->
+<!-- model formula is  -->
+We need to be able to process models such as 
+`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
+`Length` and `Conc` are available for all observations and parameter `th` applies
+to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
+could be an experimental run number, or some other categorization.
+
+Given that `[ ]` provides the R method to index arrays, the presence of square
+brackets in a model
+formula signals indexing. The question is how to perform the correct calculations.
+
+Our view is that inclusion of the indexing **within** `nls()` or similar tools 
+introduces complexity that could be avoided by a suitable wrapper function. The
+modeling formula must be expanded to encompass all the index cases, likely with
+indexed parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`.
+However, a wrapper could process output to reconstitute the indexing of parameters.
+
+## Streamlining code
+
+We are pessimistic that the overall structure of `nls()` can be streamlined due
+to the entanglement of so many features with the complicated mix of R, C and
+Fortran. Indeed, despite digging into the code over some months, we do not feel
+confident that we sufficiently understand it nor that we could maintain it. On 
+the other hand, we do believe equivalent functionality can be built, but with 
+at least some differences in how the features will be accessed.
+
+## Tests and use-case examples
+
+Maintainers of packages need suitable tests and use-case examples in order:
+
+- to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace;
+- to test individual solver functions to ensure they work across the range of 
+  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
+- to pose "silly" inputs to see if these bad inputs are caught by the programs.
+
+Such goals align with the aims of **unit testing** 
+(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
+and the conventional R package testing tools).
+
+## A testing package: NLSCompare
+
+One of us has developed a working prototype package at
+*https://github.com/ArkaB-DS/nlsCompare* .
+A primary design objective of this is to allow the summarisation of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
+
+## Documentation and resources
+
+In our investigation, we built several resources, which are now part of 
+the repository *https://github.com/nashjc/RNonlinearLS/*. Particular items
+are:
+
+- A BibTex bibliography for use with all documents in this project, but which 
+  has wider application to nonlinear least squares projects in general
+  (*https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib*).
+
+- `MachID.R` offers a suggested concise summary function
+  to identify a particular computational system used for tests. A discussion 
+  of how it was built and the resources needed across platforms is given in
+  at *https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary*.
+
+- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
+  `nls()` code in R-base.
+
+- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
+  a review of optimization in R. Ideas from the present work have been instrumental
+  in the creation of @NashWires22.
+
+# Strategic choices in nonlinear model estimation
+
+It is possible that `nls()` will remain more or less as it has been for the past
+two decades. Given its importance to R, the focus of discussion should be the 
+measures needed to secure its continued operation for legacy purposes and how 
+that may be approached.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program objects (R 
+functions) that are created by tools such as `nls()` should have minimal
+cross-linkages and side-effects.
+
+The aspects of `nls()` that have given us the most trouble:
+
+- The functions that compute the residuals and jacobians often presume
+  the data and parameters needed are available in a particular environment. As long
+  as the correct environment is used, this provides a compact syntax to invoke the
+  calculations. The danger is that the wrong data is accessed if the internal 
+  search finds a valid name that is not the object we want.
+
+- Weights, subsets, and various contextual controls (such as that for the 
+  `na.action` option) are similarly taken from the first available source, 
+  making for a very simple invocation of calculations, but the context is 
+  hidden from the user. Furthermore, it can be difficult for those of us 
+  trying to maintain or improve the code to be certain we have the
+  context correct.
+
+- The mixing of R, C and Fortran code was necessary for computational performance 
+  in the past. Lacking decent developer documentation, programmers now face a 
+  lot of work to fix or improve code. We believe in having at least a working
+  reference version of code in a single programming language. If necessary, by
+  measuring ("profiling") the code, we can find bottlenecks and substitute for 
+  those slower parts of the reference code.
+
+- A design that isolates the setup, solution, and post-solution parts of
+  for complicated calculations reduces the number of objects that must be kept in 
+  alignment. On the other hand, it may mean users have to add more commands to
+  complete their computations.
+
+- All iterative codes should return the best solution they have found so far, 
+  even if there are untoward conditions reached, such as a singular Jacobian. 
+  This modification could be made to `nls()` in the short term, correcting the 
+  issue identified above under "Failure to return best result achieved".
+
+- Given the existence of examples of good practices such as analytic derivatives,
+  stabilized solution of Gauss-Newton equations and bounds-constrained parameters,
+  base R tools should be moving to incorporate them.
+
+We anticipate that tools like `nlsr` will become more attractive as users discover
+the increased robustness and general capabilities, along with the more recent
+documentation. From a design point of view, a key difference in approach between
+`nls()` and `nlsr::nlxb()` is that `nls()` 
+builds a large infrastructure, especially the `nlsModel()` function,  from 
+which the Gauss-Newton
+iteration can be executed and other statistical information such as profiles can be
+computed, while `nlxb()` returns quite limited information, and 
+computes what is needed on an "as and when" basis. This follows a path that one 
+of us (JN) established almost 50 years ago with the software that became @cnm79,
+separating setup, solver, and post-solution analysis phases of computations. 
+There is, however, the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver of `nlsr` then calls `nls()` to build the object of 
+class `nls` for which profile log-likelihood information can be computed.
+
+# Acknowledgement
+
+Hans Werner Borchers has been helpful in developing the proposal for this project
+and in comments on this and related work. Heather Turner co-mentored the project 
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
+Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
+to succinctly summarize computing environments.
 

### 2023-03-02 RefactoringNLS/JNAB230203.tex
adding files needed to develop first R-Journal submisison

--- 2023-01-31 RefactoringNLS/RefactoringNLS.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={Refactoring nls(): a constructive critique},
-  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
-  colorlinks=true,
-  linkcolor={red},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
-  pdfcreator={LaTeX via pandoc}}
-
+% !TeX root = RJwrapper.tex
 \title{Refactoring \texttt{nls()}: a constructive critique}
-\author{John C. Nash
-\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
-Bhattacharjee
-\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
-\date{2022-6-15}
-
-\begin{document}
+\author{by John C. Nash and Arkajyoti Bhattacharjee}
+
 \maketitle
 
-{
-\hypersetup{linkcolor=}
-\setcounter{tocdepth}{2}
-\tableofcontents
+\abstract{%
+Based on work for a Google Summer of Code project ``Improvements to \texttt{nls()}'', we consider the features and limitations of the R function \texttt{nls()} in the context of improving and rationalizing R tools for nonlinear regression.
+Important considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer.
+Our work suggests that the existing code makes maintenance and improvement very difficult, with legacy applications and examples blocking some important updates.
+Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
 }
-\hypertarget{abstract}{%
-\section{Abstract}\label{abstract}}
-
-Based on work for a Google Summer of Code project ``Improvements to
-\texttt{nls()}'', we consider the features and limitations of the R
-function \texttt{nls()} in the context of improving and rationalizing R
-tools for nonlinear regression.
-
-Important considerations are the usability and maintainability of the
-code base that provides the functionality \texttt{nls()} claims to
-offer. Our work suggests that the existing code makes maintenance and
-improvement very difficult, with legacy applications and examples
-blocking some important updates. Discussion of these matters is relevant
-to improving R generally as well as its nonlinear estimation tools.
 
 \hypertarget{the-nls-function-strengths-and-shortcomings}{%
-\section{\texorpdfstring{The \texttt{nls()} function: strengths and
-shortcomings}{The nls() function: strengths and shortcomings}}\label{the-nls-function-strengths-and-shortcomings}}
-
-\texttt{nls()} is the tool in base R, the primary software software
-distribution from the Comprehensive R Archive Network
-(\emph{\url{https://cran.r-project.org}}), for estimating nonlinear
-statistical models. The function dates to the 1980s and the work related
-to Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) in S (see
-\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
-
-The \texttt{nls()} function has a remarkable and comprehensive set of
-capabilities for estimating nonlinear models that are expressed as
-\textbf{formulas}. In particular, we note that it
+\section{\texorpdfstring{The \texttt{nls()} function: strengths and shortcomings}{The nls() function: strengths and shortcomings}}\label{the-nls-function-strengths-and-shortcomings}}
+
+\texttt{nls()} is the tool in base R, the primary software software distribution
+from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}),
+for estimating nonlinear statistical models. The function dates to the 1980s and
+the work related to Bates and Watts (1988) in S (see \emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
+
+The \texttt{nls()} function has a remarkable and comprehensive set of capabilities for
+estimating nonlinear models that are expressed as \textbf{formulas}. In particular, we note that it
 
@@
 \item
-  handles formulas that include R functions, even ones which call
-  calculations in other programming languages
+  handles formulas that include R functions, even ones which call calculations in
+  other programming languages
 \item
@@
 \item
-  produces measures of variability (i.e., standard error estimates) for
-  the estimated parameters
-\item
-  has related profiling capabilities for exploring the likelihood
-  surface as parameters are changed
+  produces measures of variability (i.e., standard error estimates) for the
+  estimated parameters
+\item
+  has related profiling capabilities for exploring the likelihood surface as
+  parameters are changed
 \item
@@
 
-With such a range of features and long history, the code has become
-untidy and overly patched. It is, to our mind, essentially
-unmaintainable. Moreover, its underlying methods can and should be
-improved, as we suggest below.
+With such a range of features and long history, the code has become untidy
+and overly patched. It is, to
+our mind, essentially unmaintainable. Moreover, its underlying methods can and
+should be improved, as we suggest below.
 
 \hypertarget{feature-convergence-and-termination-tests-fixed}{%
-\subsection{Feature: Convergence and termination tests
-(FIXED)}\label{feature-convergence-and-termination-tests-fixed}}
-
-A previous issue with \texttt{nls()} that prevented it from providing
-parameter estimates for zero-residual (i.e., perfect fit) data was
-corrected thanks to suggestions by one of us.
+\subsection{Feature: Convergence and termination tests (FIXED)}\label{feature-convergence-and-termination-tests-fixed}}
+
+A previous issue with \texttt{nls()} that prevented it from providing parameter
+estimates for zero-residual
+(i.e., perfect fit) data was corrected using a suggestion by one of us.
 
@@
 \begin{quote}
-(with \texttt{var(eps)\ \textgreater{}\ 0}). It fails to indicate
-convergence on data of the form
+(with \texttt{var(eps)\ \textgreater{}\ 0}). It fails to indicate convergence on data of the form
 \end{quote}
@@
 \begin{quote}
-because the criterion amounts to comparing two components of the
-round-off error.
-\end{quote}
-
-\begin{quote}
-If you wish to test \texttt{nls} on artificial data please add a noise
-component, as shown in the example below.
-\end{quote}
-
-This amounted to admitting R cannot solve well-posed problems unless
-data is polluted with errors.
-
-This issue can be easily resolved. The ``termination test'' for the
-\textbf{program} rather than for ``convergence'' of the underlying
-\textbf{algorithm} is the Relative Offset Convergence Criterion (see
-Bates, Douglas M. and Watts, Donald G.
-(\protect\hyperlink{ref-BatesWatts81}{1981})). This projects the
-proposed step in the parameter vector on the gradient and estimates how
-much the sum of squares loss function should decrease. This estimate is
-divided by the current size of the loss function to avoid scale issues.
-When we have ``converged'', the estimated decrease is very small.
-However, with small residuals, the sum of squares loss function is
-(almost) zero and we get the possibility of a zero-divide failure.
-
-Adding a small quantity to the loss function before dividing avoids
-trouble. In 2021, one of us (J. Nash) proposed that
-\texttt{nls.control()} have an additional parameter \texttt{scaleOffset}
-with a default value of zero. Setting it to a small number -- 1.0 is a
-reasonable choice -- allows small-residual problems (i.e., near-exact
-fits) to be dealt with easily. We call this the \textbf{safeguarded
-relative offset convergence criterion}. The default value gives the
-legacy behaviour. We are pleased to note that this improvement is now in
-the R distributed code since version 4.1.0.
+because the criterion amounts to comparing two components of the round-off
+error.
+\end{quote}
+
+\begin{quote}
+If you wish to test \texttt{nls} on artificial data please add a noise component,
+as shown in the example below.
+\end{quote}
+
+This amounted to admitting R cannot solve well-posed problems unless data is polluted
+with errors.
+
+This issue can be easily resolved. The ``termination test'' for the \textbf{program} rather than for
+``convergence'' of the underlying \textbf{algorithm} is the Relative Offset
+Convergence Criterion (see Bates, Douglas M. and Watts, Donald G. (1981)). This projects the proposed step
+in the parameter vector on the gradient
+and estimates how much the sum of squares loss function should decrease.
+This estimate is divided
+by the current size of the loss function to avoid scale issues.
+When we have ``converged'', the estimated
+decrease is very small. However, with small residuals,
+the sum of squares loss function is (almost) zero and we get the possibility of a
+zero-divide failure.
+
+Adding a small quantity to the loss function before dividing avoids trouble.
+In 2021, one of us (J. Nash) proposed that \texttt{nls.control()} have an additional
+parameter \texttt{scaleOffset}
+with a default value of zero. Setting it to a small number -- 1.0 is
+a reasonable choice -- allows small-residual problems (i.e., near-exact fits)
+to be dealt with easily. We call this the
+\textbf{safeguarded relative offset convergence criterion}. The default value
+gives the legacy behaviour. We are pleased to note that this improvement is
+now in the R distributed code since version 4.1.0.
 
 \hypertarget{more-general-termination-tests}{%
-\subsubsection{More general termination
-tests}\label{more-general-termination-tests}}
-
-The convergence criterion above, the principal termination control of
-\texttt{nls()}, leaves out some possibilities that could be useful for
-some problems. The package \texttt{nlsr} (John C Nash and Duncan Murdoch
-(\protect\hyperlink{ref-nlsr2019manual}{2019})) already offers both the
-safeguarded relative offset test (\textbf{roffset}) as well as a
-\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
-the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
-times the initial sum of squares, where
-\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
-approximately 2.43e-55.
-
-We note that \texttt{nls()} uses a termination test to stop after
-\texttt{maxiter} ``iterations''. Unfortunately, the meaning of
-``iteration'' varies among programs and requires careful examination of
-the code. We prefer to use the number of times the residuals or the
-jacobian have been computed and put upper limits on these. Our codes
-exit (terminate) when these limits are reached. Generally we prefer
-larger limits than the default \texttt{maxiter\ =\ 50} of
-\texttt{nls()}, but that may reflect the more difficult problems we have
-encountered because users consult us when standard tools have given
-unsatisfactory results.
+\subsubsection{More general termination tests}\label{more-general-termination-tests}}
+
+The convergence criterion above, the principal termination control of \texttt{nls()},
+leaves out some possibilities that could be useful for some problems. The
+package \texttt{nlsr} (John C Nash and Duncan Murdoch (2019)) already offers both
+the safeguarded relative offset test (\textbf{roffset}) as well as a \textbf{small sum of
+squares} test (\textbf{smallsstest}) that compares the latest evaluated sum of squared
+(weighted) residuals to \texttt{e4} times the initial sum of squares, where
+\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.
+
+We note that \texttt{nls()} uses a termination test to stop after \texttt{maxiter} ``iterations''.
+Unfortunately, the meaning of ``iteration'' varies among programs and requires
+careful examination of the
+code. We prefer to use the number of times the residuals or the jacobian have
+been computed and put upper limits on these. Our codes exit (terminate) when
+these limits are reached. Generally we prefer larger limits than the default
+\texttt{maxiter\ =\ 50} of \texttt{nls()},
+but that may reflect the more difficult problems we have encountered because
+users consult us when standard tools have given unsatisfactory results.
 
 \hypertarget{feature-failure-when-jacobian-is-computationally-singular}{%
-\subsection{Feature: Failure when Jacobian is computationally
-singular}\label{feature-failure-when-jacobian-is-computationally-singular}}
-
-This refers to the infamous ``singular gradient'' termination message of
-\texttt{nls()}. A Google search of
+\subsection{Feature: Failure when Jacobian is computationally singular}\label{feature-failure-when-jacobian-is-computationally-singular}}
+
+This refers to the infamous ``singular gradient'' termination message of \texttt{nls()}.
+A Google search of
 
@@
 
-gets over 4000 hits that are spread over some years. This could be
-because the Jacobian is poorly approximated (see \textbf{Jacobian
-computation} below). However, it is common in nonlinear least squares
-computations that the Jacobian is very close to singular for some values
-of the model parameters. In such cases, we need to find an alternative
-algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most
-common work-around is the Levenberg-Marquardt stabilization (see
-Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}), Levenberg
-(\protect\hyperlink{ref-Levenberg1944}{1944}), Nash
-(\protect\hyperlink{ref-jn77ima}{1977})), and versions of it have been
-implemented in packages \texttt{minpack.lm} and \texttt{nlsr}. In our
-work, we prepared experimental prototypes of \texttt{nls} that
-incorporate stabilization, but integration with all the features of
-\texttt{nls()} has proved difficult.
+\noindent
+gets over 4000 hits that are spread over some years. This could be because the
+Jacobian is poorly approximated (see \textbf{Jacobian computation} below). However,
+it is common in nonlinear least squares computations that the Jacobian is very close
+to singular for some values of the model parameters. In such cases, we need to find
+an alternative algorithm to the Gauss-Newton iteration of \texttt{nls()}. The most common
+work-around is the Levenberg-Marquardt stabilization (see Marquardt (1963),
+Levenberg (1944), Nash (1977)), and versions of it have been implemented in
+packages \texttt{minpack.lm} and \texttt{nlsr}.
+In our work, we prepared experimental prototypes of \texttt{nls} that incorporate
+stabilization, but integration with all the features of \texttt{nls()} has proved difficult.
 
@@
 \emph{\url{https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values}}.
-\texttt{nlsr::nlxb()} displays the singular values of the Jacobian,
-which confirm that the \texttt{nls()} error message is correct, but this
-is not helpful in obtaining a solution.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{reala }\OtherTok{\textless{}{-}} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{; realb }\OtherTok{\textless{}{-}} \DecValTok{5}\NormalTok{; realc }\OtherTok{\textless{}{-}} \FloatTok{0.5}\NormalTok{; realr }\OtherTok{\textless{}{-}} \FloatTok{0.7}\NormalTok{; realm }\OtherTok{\textless{}{-}} \DecValTok{1} \CommentTok{\# underlying parameters}
-\NormalTok{x }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{11} \CommentTok{\# x values; 11 data points}
-\NormalTok{y }\OtherTok{\textless{}{-}}\NormalTok{ reala }\SpecialCharTok{+}\NormalTok{ realb }\SpecialCharTok{*}\NormalTok{ realr}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ realm) }\SpecialCharTok{+}\NormalTok{ realc }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# linear + exponential function}
-\NormalTok{testdat }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(x, y) }\CommentTok{\# save in a data frame}
-\NormalTok{strt }\OtherTok{\textless{}{-}} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \SpecialCharTok{{-}}\DecValTok{3}\NormalTok{, }\AttributeTok{b =} \DecValTok{5}\NormalTok{, }\AttributeTok{c =} \FloatTok{0.5}\NormalTok{, }\AttributeTok{r =} \FloatTok{0.7}\NormalTok{, }\AttributeTok{m =} \DecValTok{1}\NormalTok{) }\CommentTok{\# give programs a good start}
-\NormalTok{jform }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a }\SpecialCharTok{+}\NormalTok{ b }\SpecialCharTok{*}\NormalTok{ r}\SpecialCharTok{\^{}}\NormalTok{(x }\SpecialCharTok{{-}}\NormalTok{ m) }\SpecialCharTok{+}\NormalTok{ c }\SpecialCharTok{*}\NormalTok{ x }\CommentTok{\# Formula}
-\FunctionTok{library}\NormalTok{(nlsr)}
-\NormalTok{linexp2 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
-\NormalTok{linexp2 }\CommentTok{\# Note singular values of Jacobian in rightmost column}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  0  on  11 observations
-##     after  1    Jacobian and  1 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## a                     -3            NA         NA         NA           0       26.49  
-## b                      5            NA         NA         NA           0       9.615  
-## c                    0.5            NA         NA         NA           0       2.466  
-## r                    0.7            NA         NA         NA           0      0.1098  
-## m                      1            NA         NA         NA           0   1.311e-16
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# We get an error with nls()}
-\NormalTok{linexp }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(jform, }\AttributeTok{data =}\NormalTok{ testdat, }\AttributeTok{start =}\NormalTok{ strt, }\AttributeTok{trace =} \ConstantTok{FALSE}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nlsModel(formula, mf, start, wts, scaleOffset = scOff, nDcentral = nDcntr) : 
-##   singular gradient matrix at initial parameter estimates
-\end{verbatim}
-
-The Jacobian is essentially singular as shown by its singular values.
-Note that these are \textbf{displayed} by package \texttt{nlsr} in a
-single column in the output to provide a compact layout, but the values
-do \textbf{NOT} correspond to the individual parameters in whose row
-they appear; they are a property of the whole problem.
+\texttt{nlsr::nlxb()} displays the singular values of the Jacobian, which confirm that
+the \texttt{nls()} error message is correct, but this is not helpful in obtaining a
+solution.
+
+\begin{verbatim}
+#> residual sumsquares =  0  on  11 observations
+#>     after  1    Jacobian and  1 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> a                     -3            NA         NA         NA           0       26.49  
+#> b                      5            NA         NA         NA           0       9.615  
+#> c                    0.5            NA         NA         NA           0       2.466  
+#> r                    0.7            NA         NA         NA           0      0.1098  
+#> m                      1            NA         NA         NA           0   9.798e-17
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nlsModel(formula, mf, start, wts, scaleOffset = scOff, nDcentral = nDcntr) : 
+#>   singular gradient matrix at initial parameter estimates
+\end{verbatim}
+
+The Jacobian is essentially singular as shown by its singular
+values. Note that these are \textbf{displayed} by package \texttt{nlsr} in a single column in
+the output to provide a compact layout, but the values do \textbf{NOT} correspond to
+the individual parameters in
+whose row they appear; they are a property of the whole problem.
 
 \hypertarget{feature-jacobian-computation}{%
-\subsection{Feature: Jacobian
-computation}\label{feature-jacobian-computation}}
-
-\texttt{nls()}, with the \texttt{numericDeriv()} function, computes the
-Jacobian as the ``gradient'' attribute of the residual vector. This is
-implemented as a mix of R and C code, but we have created a rather more
-compact version entirely in R. See
-\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
-
-As far as we can understand the logic in \texttt{nls()}, the Jacobian
-computation during parameter estimation is carried out entirely within
-the called C-language program, and the R code function
-\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
-the R distribution source code. This is used to provide Jacobian
-information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
-functions, which are \textbf{not} exported for general use.
-
-The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from
-analytic expressions, while that of \texttt{nls()} mostly uses a
-numerical approximation. Some \texttt{selfStart} models do provide
-``gradient'' information. \texttt{minpack.lm::nlsLM()} invokes
-\texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but
-it appears to use an internal approximate Jacobian code from the
-original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such
-differences in approach can lead to different behaviour, though in our
-experience differences are usually minor.
+\subsection{Feature: Jacobian computation}\label{feature-jacobian-computation}}
+
+\texttt{nls()}, with the \texttt{numericDeriv()} function, computes the Jacobian as the ``gradient''
+attribute of the residual vector. This is implemented as a mix of R and C code, but
+we have created a rather more compact version entirely in R.
+See \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
+
+As far as we can understand the logic in \texttt{nls()}, the Jacobian computation during
+parameter estimation is carried out entirely within the called C-language program,
+and the R code function \texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R}
+in the R distribution source code. This is used to provide Jacobian information in
+the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are \textbf{not} exported for
+general use.
+
+The Jacobian used by \texttt{nlsr::nlxb()} is, by default, computed from analytic
+expressions, while that of \texttt{nls()} mostly uses a numerical approximation. Some
+\texttt{selfStart} models do provide ``gradient'' information. \texttt{minpack.lm::nlsLM()}
+invokes \texttt{numericDeriv()} in its local version of \texttt{nlsModel()}, but it appears
+to use an internal approximate Jacobian code from the original Fortran
+\texttt{minpack} code, namely, \texttt{lmdif.f}. Such differences in approach can lead to
+different behaviour, though in our experience differences are usually minor.
 
@@
 \item
-  A pasture regrowth problem (Huet et al.
-  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
-  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
-  conditioned Jacobian and \texttt{nls()} fails with ``singular
-  gradient''. Worse, numerical approximation to the Jacobian give the
-  error ``singular gradient matrix at initial parameter estimates'' for
-  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
-  unable to take effect, while the analytic derivatives of
-  \texttt{nlsr::nlxb} give a solution.
-\item
-  Karl Schilling (private communication) provided an example where a
-  model specified with the formula
-  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
-  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
-  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
-  which \texttt{x\ =\ 0}, this is undefined. In such cases, we observed
-  that \texttt{nls()} and \texttt{minpack.lm::nlsLM} found a solution,
-  though it can be debated whether such lucky accidents can be taken as
-  an advantage.
+  A pasture regrowth problem (Huet et al. (2004), page 1, based on Ratkowsky (1983)) has a poorly
+  conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
+  Worse, numerical approximation to the Jacobian give the error
+  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of \texttt{nlsr::nlxb} give a solution.
+\item
+  Karl Schilling (private communication) provided an example where a model specified
+  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
+  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
+  this is undefined. In such cases, we observed that \texttt{nls()} and \texttt{minpack.lm::nlsLM}
+  found a solution, though it can be debated whether such lucky accidents can be
+  taken as an advantage.
 \end{itemize}
 
-Note that the \texttt{selfStart} models in the base R
-\texttt{./src/library/stats/R/zzModels.R} file and in package
-\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
-may provide the Jacobian in the ``gradient'' attribute of the
-``one-sided'' formula that defines each model, and these Jacobians may
-be the analytic forms. The \texttt{nls()} function, after computing the
-``right hand side'' or \texttt{rhs} of the residual, then checks to see
-if the ``gradient'' attribute is defined, and, if not, uses
-\texttt{numericDeriv} to compute a Jacobian into that attribute. This
-code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
-functions. The use of analytic Jacobians almost certainly contributes to
-the good performance of \texttt{nls()} on \texttt{selfStart} models.
+Note that the \texttt{selfStart} models in the base R \texttt{./src/library/stats/R/zzModels.R}
+file and in package \texttt{nlraa} (Miguez (2021)) may
+provide the Jacobian in the ``gradient'' attribute of the ``one-sided'' formula that
+defines each model, and these Jacobians may be the analytic forms. The \texttt{nls()}
+function, after computing the ``right hand side'' or \texttt{rhs} of the residual, then
+checks to see if the ``gradient'' attribute is defined, and, if not, uses
+\texttt{numericDeriv} to compute a Jacobian
+into that attribute. This code is within the \texttt{nlsModel()} or \texttt{nlsModel.plinear()}
+functions. The use of analytic Jacobians almost certainly contributes to the good
+performance of \texttt{nls()} on \texttt{selfStart} models.
 
 \hypertarget{feature-weights-on-observations}{%
-\subsection{Feature: Weights on
-observations}\label{feature-weights-on-observations}}
-
-\texttt{nls()}, \texttt{nlsr::nlxb()} and \texttt{minpack.lm::nlsLM()}
-all have a \texttt{weights} argument that specifies a vector of weights
-the same length as the number of residuals. Each residual is multiplied
-by the square root of the corresponding weight. The values returned by
-the \texttt{residuals()} function are weighted, and the
-\texttt{fitted()} or \texttt{predict()} function are used to compute raw
-residuals.
+\subsection{Feature: Weights on observations}\label{feature-weights-on-observations}}
+
+\texttt{nls()}, \texttt{nlsr::nlxb()} and \texttt{minpack.lm::nlsLM()} all have a \texttt{weights} argument that
+specifies a vector of weights the same length as the number of residuals. Each
+residual is multiplied by the square root of the corresponding weight. The values
+returned by the \texttt{residuals()} function are weighted, and the \texttt{fitted()} or
+\texttt{predict()} function are used to compute raw residuals.
 
@@
 
-\texttt{nls()} accepts an argument \texttt{subset}. This acts through
-the mediation of \texttt{model.frame} and is not clearly obvious in the
-source code files \texttt{/src/library/stats/R/nls.R} and
-\texttt{/src/library/stats/src/nls.C}.
-
-While the implementation of subset at the level of the call to
-\texttt{nls()} has a certain attractiveness, it does mean that the
-programmer of the solver needs to be aware of the source (and value) of
-objects such as the data, residuals and Jacobian. This is overly
-complicated. By preference, we would implement subsetting by means of
-zero-value weights, with observation counts (and degrees of freedom)
-computed via the numbers of non-zero weights. Alternatively, we would
-extract a working dataframe from the relevant elements in the original.
+\texttt{nls()} accepts an argument \texttt{subset}. This acts through the mediation of
+\texttt{model.frame} and is not clearly obvious in the source code files
+\texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
+
+While the implementation of subset at the level of the call to \texttt{nls()} has a certain
+attractiveness, it does mean that the programmer of the solver needs to be aware of
+the source (and value) of objects such as the data, residuals and Jacobian. This
+is overly complicated. By preference,
+we would implement subsetting by means of zero-value weights, with observation counts
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
 
@@
 
-\texttt{na.action} is an argument to the \texttt{nls()} function, but it
-does not appear obviously in the source code, often being handled behind
-the scenes after referencing the option \texttt{na.action}. A useful,
-but possibly dated, description is given in:
+\texttt{na.action} is an argument to the \texttt{nls()} function, but it does not appear obviously
+in the source code, often being handled behind the scenes after referencing the
+option \texttt{na.action}. A useful, but possibly dated, description is given in:
 \emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.
@@
 The typical default action, which can be seen by using the command
-\texttt{getOption("na.action")} is \texttt{na.omit}. This option
-essentially omits from computations any observations containing missing
-values (i.e.~any row of a data frame containing an NA).
-\texttt{na.exclude} does much of the same for computations, but keeps
-the rows with NA elements so that predictions are in the correct row
-position. We recommend that workers actually test output to verify the
-behaviour is as wanted.
-
-A succinct description of the issue is given in:
-\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}
+\texttt{getOption("na.action")}
+is \texttt{na.omit}. This option essentially omits from computations any observations
+containing missing values (i.e.~any row of a data frame containing an NA).
+\texttt{na.exclude} does much of the same for computations, but keeps the rows with NA
+elements so that predictions are in the correct row position. We recommend that
+workers actually test output to verify the behaviour is as wanted.
+
+A succinct description of the issue is given in: \emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}
 where the ``Answer'' states
@@
 \begin{quote}
-The only benefit of \texttt{na.exclude} over \texttt{na.omit} is that
-the former will retain the original number of rows in the data. This may
-be useful where you need to retain the original size of the dataset -
-for example it is useful when you want to compare predicted values to
-original values. With \texttt{na.omit} you will end up with fewer rows
-so you won't as easily be able to compare.
-\end{quote}
-
-\texttt{na.pass} simply passes on data ``as is'', while \texttt{na.fail}
-will essentially stop if any missing values are present.
-
-Our concern with \texttt{na.action} is that users may be unaware of the
-effects of an option setting they may not even be aware has been set.
-Should \texttt{na.fail} be the default?
+The only benefit of \texttt{na.exclude} over \texttt{na.omit} is that the former will retain the
+original number of rows in the data. This may be useful where you need to retain the
+original size of the dataset - for example it is useful when you want to compare
+predicted values to original values. With \texttt{na.omit} you will end up with fewer
+rows so you won't as easily be able to compare.
+\end{quote}
+
+\texttt{na.pass} simply passes on data ``as is'', while \texttt{na.fail} will essentially stop
+if any missing values are present.
+
+Our concern with \texttt{na.action} is that users may be unaware of the effects of an option
+setting they may not even be aware has been set. Should \texttt{na.fail} be the default?
 
@@
 
-\texttt{model} is an argument to the \texttt{nls()} function, which is
-documented as:
-
-\begin{quote}
-\textbf{model} logical. If true, the model frame is returned as part of
-the object. Default is FALSE.
-\end{quote}
-
-Indeed, the argument only gets used when \texttt{nls()} is about to
-return its result object, and the element \texttt{model} is NULL unless
-the calling argument \texttt{model} is TRUE. (Using the same name could
-be confusing.) However, the model frame is used within the function code
-in the form of the object \texttt{mf}. We feel that users could benefit
-from more extensive documentation and examples of its use.
+\texttt{model} is an argument to the \texttt{nls()} function, which is documented as:
+
+\begin{quote}
+\textbf{model} logical. If true, the model frame is returned as part of the object. Default is FALSE.
+\end{quote}
+
+Indeed, the argument only gets used when \texttt{nls()} is about to return its result
+object, and the element \texttt{model} is NULL unless the calling argument \texttt{model} is
+TRUE. (Using the same name could be confusing.) However, the model frame is
+used within the function code in the form of the object \texttt{mf}. We feel that
+users could benefit from more extensive documentation and examples of its use.
 
@@
 
-\texttt{nls()} can be called without specifying the \texttt{data}
-argument. In this case, it will search in the available environments
-(i.e., workspaces) for suitable data objects. We do NOT like this
-approach, but it is ``the R way''. R allows users to leave many objects
-in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
-actively suppress saving this workspace (\texttt{.RData}) on exit;
-otherwise, any such file in the path, when R is launched, will be
-loaded. The overwhelming proportion of R users in our acquaintance avoid
-saving the workspace because of the danger of lurking data and functions
-which may cause unwanted results.
-
-Nevertheless, to provide compatible behaviour with \texttt{nls()},
-competing programs need to ensure equivalent behaviour, but users should
-test that the operation is as they intend.
+\texttt{nls()} can be called without specifying the \texttt{data} argument. In this case, it will
+search in the available environments (i.e., workspaces) for suitable data objects.
+We do NOT like this approach, but it is ``the R way''. R allows users to leave many
+objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
+suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any such file in the
+path, when R is launched, will be loaded. The overwhelming proportion of R users
+in our acquaintance avoid saving the workspace because of the danger of lurking
+data and functions which may cause unwanted results.
+
+Nevertheless, to provide compatible behaviour with \texttt{nls()}, competing programs
+need to ensure equivalent behaviour, but users should test that the
+operation is as they intend.
 
 \hypertarget{feature-missing-start-vector-and-self-starting-models}{%
-\subsection{Feature: missing start vector and self-starting
-models}\label{feature-missing-start-vector-and-self-starting-models}}
-
-Nonlinear estimation algorithms are almost all iterative and need a set
-of starting parameters. \texttt{nls()} offers a special class of
-modeling functions called \textbf{\texttt{selfStart}} models. There are
-a number of these in base R (\texttt{./src/library/stats/R/zzModels.R})
-and others in R packages such as CRAN package \texttt{nlraa} (Miguez
-(\protect\hyperlink{ref-MiguezNLRAA2021}{2021})), as well as the
-now-archived package \texttt{NRAIA}. Unfortunately, some
-\textbf{\texttt{selfStart}} codes entangle the calculation of starting
-values for parameters with the particulars of the \texttt{nls()} code.
-Though there is a \texttt{getInitial()} function, this is not easy to
-use to simply compute the initial parameter estimates outside of
-\texttt{nls()}, in part because it may call that function. Such circular
-references are, in our view, dangerous. Moreover, we believe that it
-would be helpful to have \texttt{selfStart} models that allow users to
-explicitly provide values other than those suggested for the starting
-parameters.
-
-Other concerns are revealed by the example below. Here, the
-\texttt{SSlogis} \texttt{selfStart} function is used to generate a set
-of initial parameters for a 3-parameter logistic curve. The form used by
-\texttt{SSlogis} is
-\(y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))\), but we
-show how the starting parameters for this model can be transformed to
-those of another form of the model, namely,
-\(y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))\).
-
-The code for \texttt{SSlogis()} is in
-\texttt{./src/library/stats/R/zzModels.R}. This R function includes
-analytic expressions for the Jacobian (``gradient''). These could be
-useful to R users, especially if documented. Moreover, we wonder why the
-programmers have chosen to save so many quantities in ``hidden''
-variables, i.e., with names preceded by ``.''. These are then not
-displayed by the \texttt{ls()} command, making them difficult to access.
-
-In the event that a \texttt{selfStart} model is not available,
-\texttt{nls()} sets all the starting parameters to 1. This is, in our
-view, tolerable, but could be improved by using a set of values that are
-all slightly different, which, in the case of the example model
-\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
-avoid a singular Jacobian because \(b\) and \(d\) were equal in value. A
-sequence like 1.0, 1.1, 1.2, 1.3 for the four parameters could be
-provided quite simply instead of all 1's.
+\subsection{Feature: missing start vector and self-starting models}\label{feature-missing-start-vector-and-self-starting-models}}
+
+Nonlinear estimation algorithms are almost all iterative and need a set of starting
+parameters. \texttt{nls()} offers a special class of modeling functions called
+\textbf{\texttt{selfStart}} models. There are a number of these in base R
+(\texttt{./src/library/stats/R/zzModels.R})
+and others in R packages such as CRAN package \texttt{nlraa} (Miguez (2021)), as well as
+the now-archived package \texttt{NRAIA}. Unfortunately, some \textbf{\texttt{selfStart}} codes
+entangle the calculation of starting values for parameters
+with the particulars of the \texttt{nls()} code. Though there is a \texttt{getInitial()} function,
+this is not easy to use to simply compute the initial parameter estimates outside of
+\texttt{nls()}, in part
+because it may call that function. Such circular references are, in our view,
+dangerous. Moreover, we believe that it would be helpful to have \texttt{selfStart}
+models that allow users to explicitly provide values other than those suggested
+for the starting parameters.
+
+Other concerns are revealed by the example below. Here, the \texttt{SSlogis}
+\texttt{selfStart} function is used to generate a set
+of initial parameters for a 3-parameter logistic curve. The form used by \texttt{SSlogis}
+is \(y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))\), but we show
+how the starting parameters for this model can be transformed to those of
+another form of the model, namely, \(y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))\).
+
+The code for \texttt{SSlogis()} is in \texttt{./src/library/stats/R/zzModels.R}.
+This R function includes analytic expressions for the Jacobian (``gradient'').
+These could be useful to R users, especially if documented. Moreover, we
+wonder why the programmers have chosen to save so many quantities in ``hidden''
+variables, i.e., with names preceded by ``.''. These are then not displayed by the
+\texttt{ls()} command, making them difficult to access.
+
+In the event that a \texttt{selfStart} model is not available, \texttt{nls()} sets all the
+starting parameters to 1. This is, in our view, tolerable, but could be improved
+by using a set of values that are all slightly different, which, in the case of
+the example model \(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\)
+would avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
+A sequence like 1.0, 1.1, 1.2, 1.3 for the four parameters
+could be provided quite simply instead of all 1's.
 
 \hypertarget{strategic-issues-in-selfstart-models}{%
-\subsubsection{\texorpdfstring{Strategic issues in \texttt{selfStart}
-models}{Strategic issues in selfStart models}}\label{strategic-issues-in-selfstart-models}}
-
-By providing starting values that are likely to be reasonable and by
-including Jacobian code in the \texttt{selfStart} model code, some of
-the deficiencies of the Gauss-Newton algorithm are mitigated. However,
-creating such functions is a lot of work, and their documentation (file
-\texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated. We
-believe that the focus should be placed on getting good initial
-parameters, that is \texttt{getInitial()} function, though avoiding the
-current calls back to \texttt{nls()}. Interactive tools, such as
-``visual fitting'' (Nash and Velleman
-(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
-considering.
-
-We also note that the introduction of \texttt{scaleOffset} in R 4.1.1 to
-deal with the convergence test for small residual problems now requires
-that the \texttt{getInitial()} function have dot-arguments
-(\texttt{...}) in its argument list. This illustrates the entanglement
-of many features in \texttt{nls()}.
+\subsubsection{\texorpdfstring{Strategic issues in \texttt{selfStart} models}{Strategic issues in selfStart models}}\label{strategic-issues-in-selfstart-models}}
+
+By providing starting values that are likely to be reasonable and by including
+Jacobian code in the \texttt{selfStart} model code, some of the deficiencies of the
+Gauss-Newton algorithm are mitigated. However, creating such functions is a lot
+of work, and their
+documentation (file \texttt{./src/library/stats/man/selfStart.Rd}) is quite complicated.
+We believe that the focus should be placed on getting good initial parameters,
+that is \texttt{getInitial()} function, though avoiding the current calls back to \texttt{nls()}.
+Interactive tools, such as ``visual fitting'' (Nash and Velleman (1996)) might be
+worth considering.
+
+We also note that the introduction of \texttt{scaleOffset} in R 4.1.1 to deal with the
+convergence test for small residual problems now requires that the \texttt{getInitial()}
+function have dot-arguments (\texttt{...}) in its argument list. This illustrates the
+entanglement of many features in \texttt{nls()}.
 
 \hypertarget{issue-returned-output-of-nls-and-its-documentation}{%
-\subsection{Issue: returned output of nls() and its
-documentation}\label{issue-returned-output-of-nls-and-its-documentation}}
-
-The output of \texttt{nls()} is an object of class ``nls'' which has the
-following structure:
+\subsection{Issue: returned output of nls() and its documentation}\label{issue-returned-output-of-nls-and-its-documentation}}
+
+The output of \texttt{nls()} is an object of class ``nls'' which has the following structure:
 
@@
 \begin{quote}
-\texttt{m} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ an \texttt{nlsModel} object
-incorporating the model.
-\end{quote}
-
-\begin{quote}
-\texttt{data} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~the expression that was
-passed to \texttt{nls} as the data argument. The actual \newline
-\(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) data values are present in the
-environment of the \texttt{m} components, e.g.,
-\newline \(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\)
-\texttt{environment(m\$conv)}.
-\end{quote}
-
-\begin{quote}
-\texttt{call} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the matched call with several
-components, notably \texttt{algorithm}.
-\end{quote}
-
-\begin{quote}
-\texttt{na.action} ~ ~ ~ ~ ~ ~ ~ ~ ~ the \texttt{"na.action"} attribute
-(if any) of the model frame.
-\end{quote}
-
-\begin{quote}
-\texttt{dataClasses} ~ ~ ~ ~ ~ ~ ~ \(~\)the \texttt{"dataClasses"}
-attribute (if any) of the ``\texttt{terms}'' attribute of the \newline
+\texttt{m} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ an \texttt{nlsModel} object incorporating the model.
+\end{quote}
+
+\begin{quote}
+\texttt{data} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~the expression that was passed to \texttt{nls} as the data argument. The actual \newline
+\(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) data values are present in the environment of the \texttt{m} components, e.g., \newline \(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) \texttt{environment(m\$conv)}.
+\end{quote}
+
+\begin{quote}
+\texttt{call} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the matched call with several components, notably \texttt{algorithm}.
+\end{quote}
+
+\begin{quote}
+\texttt{na.action} ~ ~ ~ ~ ~ ~ ~ ~ ~ the \texttt{"na.action"} attribute (if any) of the model frame.
+\end{quote}
+
+\begin{quote}
+\texttt{dataClasses} ~ ~ ~ ~ ~ ~ ~ \(~\)the \texttt{"dataClasses"} attribute (if any) of the ``\texttt{terms}'' attribute of the \newline
 \(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\) model frame.
@@
 \begin{quote}
-\texttt{model} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~if \texttt{model\ =\ TRUE}, the
-model frame.
-\end{quote}
-
-\begin{quote}
-\texttt{weights} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ if \texttt{weights} is supplied,
-the weights.
-\end{quote}
-
-\begin{quote}
-\texttt{convInfo} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ a list with convergence
-information.
-\end{quote}
-
-\begin{quote}
-\texttt{control} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the control \texttt{list} used,
-see the \texttt{control} argument.
-\end{quote}
-
-\begin{quote}
-\texttt{convergence,\ message} ~for an \texttt{algorithm\ =\ "port"} fit
-only, a convergence code (\texttt{0} for \newline
+\texttt{model} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~if \texttt{model\ =\ TRUE}, the model frame.
+\end{quote}
+
+\begin{quote}
+\texttt{weights} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ if \texttt{weights} is supplied, the weights.
+\end{quote}
+
+\begin{quote}
+\texttt{convInfo} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~~ a list with convergence information.
+\end{quote}
+
+\begin{quote}
+\texttt{control} ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ the control \texttt{list} used, see the \texttt{control} argument.
+\end{quote}
+
+\begin{quote}
+\texttt{convergence,\ message} ~for an \texttt{algorithm\ =\ "port"} fit only, a convergence code (\texttt{0} for \newline
 \(~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~\)convergence) and message.
@@
 \begin{quote}
-~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ To use these is \emph{deprecated}, as
-they are available from \texttt{convInfo} now.
-\end{quote}
-
-\begin{verbatim}
-              -----------------------------------------------------------
-\end{verbatim}
-
-The \texttt{nls} object contains some elements that are awkward to
-produce by other algorithms, but some information that would be useful
-is not presented in a clear manner. Moreover, the complexity of the
-object is a challenge to users.
-
-In the following, we use \texttt{result} as the returned object from
-\texttt{nls()}.
-
-The \texttt{data} return element is an R symbol. To actually access the
-data from this element, we need to use the syntax:
+~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ ~ To use these is \emph{deprecated}, as they are available from \texttt{convInfo} now.
+\end{quote}
+
+The \texttt{nls} object contains some elements that are awkward to produce by other
+algorithms, but some information that would be useful is not presented in a
+clear manner. Moreover, the complexity of the object is a challenge to users.
+
+In the following, we use \texttt{result} as the returned object from \texttt{nls()}.
+
+The \texttt{data} return element is an R symbol. To actually access the data from this
+element, we need to use the syntax:
 
@@
 
-However, if the call is made with \texttt{model=TRUE}, then there is a
-returned element \texttt{model} which contains the data, and we can list
-its contents using:
+However, if the call is made with \texttt{model=TRUE}, then there is a returned element
+\texttt{model} which contains the data, and we can list its contents using:
 
@@
 
-and if there is an element called \texttt{xdata}, it can be accessed as
-\texttt{result\$model\$xdata}.
+and if there is an element called \texttt{xdata}, it can be accessed as \texttt{result\$model\$xdata}.
 
@@
 \begin{quote}
-\texttt{coefficients} A named vector giving the parameter values at the
-supposed solution.
-\end{quote}
-
-\begin{quote}
-\texttt{ssquares} \(~~~~~\) The sum of squared residuals at this set of
-parameters.
-\end{quote}
-
-\begin{quote}
-\texttt{resid} \(~~~~~~~~~~\) The residual vector at the returned
-parameters.
-\end{quote}
-
-\begin{quote}
-\texttt{jacobian} \(~~~~~\) The jacobian matrix (partial derivatives of
-residuals w.r.t. the parameters) at the \newline
+\texttt{coefficients} A named vector giving the parameter values at the supposed solution.
+\end{quote}
+
+\begin{quote}
+\texttt{ssquares} \(~~~~~\) The sum of squared residuals at this set of parameters.
+\end{quote}
+
+\begin{quote}
+\texttt{resid} \(~~~~~~~~~~\) The residual vector at the returned parameters.
+\end{quote}
+
+\begin{quote}
+\texttt{jacobian} \(~~~~~\) The jacobian matrix (partial derivatives of residuals w.r.t. the
+parameters) at the \newline
 \(~~~~~~~~~~~~~~~~~~~~\) returned parameters.
@@
 \begin{quote}
-\texttt{feval} \(~~~~~~~~~~\) The number of residual evaluations (sum of
-squares computations) used.
+\texttt{feval} \(~~~~~~~~~~\) The number of residual evaluations (sum of squares computations) used.
 \end{quote}
@@
 
-However, actually looking at the structure of a returned result gives a
-list of 11 items. The extra 5 are:
+However, actually looking at the structure of a returned result gives a list of 11
+items. The extra 5 are:
 
@@
 
-The result object from \texttt{nlsr::nlxb()} is still much smaller than
-the one \texttt{nls()} returns. Moreover, \texttt{nlxb} explicitly
-returns the sum of squares as well as the residual vector and Jacobian.
-The counts of evaluations are also returned. (Note that the singular
-values of the Jacobian are actually computed in the \texttt{print} and
-\texttt{summary} methods for the result.) As we made our comparisons, we
-noted several potential updates to the \texttt{nlsr} documentation as
-well as that for \texttt{nls()}.
+The result object from \texttt{nlsr::nlxb()} is still much smaller than the one \texttt{nls()}
+returns. Moreover, \texttt{nlxb}
+explicitly returns the sum of squares as well as the residual vector and Jacobian.
+The counts of evaluations are also returned.
+(Note that the singular values of the Jacobian are
+actually computed in the \texttt{print} and \texttt{summary} methods for the result.)
+As we made our comparisons, we noted several potential updates to the \texttt{nlsr}
+documentation as well as that for \texttt{nls()}.
 
 \hypertarget{weights-in-returned-functions-from-nls}{%
-\subsubsection{Weights in returned functions from
-nls()}\label{weights-in-returned-functions-from-nls}}
-
-As already noted, the function \texttt{resid()} (an alias for
-\texttt{residuals()}) is WEIGHTED, as are those in the structured object
-\texttt{m} returned by \texttt{nls()} or \texttt{minpack.lm::nlsLM},
-e.g., \texttt{ans\$m\$resid()}. The function \texttt{nlsModel()}, which
-we have had to extract from the base R code and explicitly
-\texttt{source()} because it is not exported to the working namespace,
-allows us to compute residuals for particular coefficient sets.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
-          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
-\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
-\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
-\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
-\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
-\NormalTok{nowt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: weed ~ SSlogis(tt, Asym, xmid, scal)
-##    data: parent.frame()
-##   Asym   xmid   scal 
-## 196.19  12.42   3.19 
-##  residual sum-of-squares: 2.59
-## 
-## Number of iterations to convergence: 0 
-## Achieved convergence tolerance: 1.12e-06
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
-##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
-## attr(,"gradient")
-##           Asym     xmid    scal
-##  [1,] 0.027117  -1.6229  5.8103
-##  [2,] 0.036737  -2.1769  7.1111
-##  [3,] 0.049596  -2.8997  8.5628
-##  [4,] 0.066645  -3.8266 10.1000
-##  [5,] 0.089005  -4.9881 11.6015
-##  [6,] 0.117921  -6.3988 12.8762
-##  [7,] 0.154635  -8.0418 13.6607
-##  [8,] 0.200186  -9.8498 13.6432
-##  [9,] 0.255106 -11.6901 12.5267
-## [10,] 0.319083 -13.3660 10.1313
-## [11,] 0.390688 -14.6444  6.5083
-## [12,] 0.467334 -15.3139  2.0038
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
-\NormalTok{usewt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: weed ~ SSlogis(tt, Asym, xmid, scal)
-##    data: parent.frame()
-##   Asym   xmid   scal 
-## 199.86  12.50   3.19 
-##  weighted residual sum-of-squares: 0.0196
-## 
-## Number of iterations to convergence: 2 
-## Achieved convergence tolerance: 8.29e-06
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# WEIGHTED. Does NOT take coefficients.}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
-## attr(,"gradient")
-##           Asym     xmid    scal
-##  [1,] 0.026498  -1.6157  5.8228
-##  [2,] 0.035901  -2.1679  7.1333
-##  [3,] 0.048474  -2.8890  8.6006
-##  [4,] 0.065153  -3.8149 10.1617
-##  [5,] 0.087046  -4.9775 11.6983
-##  [6,] 0.115387  -6.3933 13.0222
-##  [7,] 0.151425  -8.0483 13.8709
-##  [8,] 0.196222  -9.8787 13.9297
-##  [9,] 0.250362 -11.7553 12.8918
-## [10,] 0.313611 -13.4827 10.5608
-## [11,] 0.384641 -14.8251  6.9663
-## [12,] 0.460954 -15.5631  2.4357
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
-\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
-## [11] 1.6572 1.4214
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}
+
+As already noted, the function \texttt{resid()} (an alias for \texttt{residuals()}) is
+WEIGHTED, as are those in the structured object \texttt{m} returned by \texttt{nls()} or
+\texttt{minpack.lm::nlsLM}, e.g., \texttt{ans\$m\$resid()}. The function \texttt{nlsModel()}, which
+we have had to extract from the base R code and explicitly \texttt{source()} because
+it is not exported to the working namespace, allows us to compute residuals
+for particular coefficient sets.
+
+\begin{verbatim}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+nowt
+\end{verbatim}
+
+\begin{verbatim}
+#> Nonlinear regression model
+#>   model: weed ~ SSlogis(tt, Asym, xmid, scal)
+#>    data: parent.frame()
+#>   Asym   xmid   scal 
+#> 196.19  12.42   3.19 
+#>  residual sum-of-squares: 2.59
+#> 
+#> Number of iterations to convergence: 0 
+#> Achieved convergence tolerance: 1.12e-06
+\end{verbatim}
+
+\begin{verbatim}
+nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
+#>  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
+#> attr(,"gradient")
+#>           Asym     xmid    scal
+#>  [1,] 0.027117  -1.6229  5.8103
+#>  [2,] 0.036737  -2.1769  7.1111
+#>  [3,] 0.049596  -2.8997  8.5628
+#>  [4,] 0.066645  -3.8266 10.1000
+#>  [5,] 0.089005  -4.9881 11.6015
+#>  [6,] 0.117921  -6.3988 12.8762
+#>  [7,] 0.154635  -8.0418 13.6607
+#>  [8,] 0.200186  -9.8498 13.6432
+#>  [9,] 0.255106 -11.6901 12.5267
+#> [10,] 0.319083 -13.3660 10.1313
+#> [11,] 0.390688 -14.6444  6.5083
+#> [12,] 0.467334 -15.3139  2.0038
+\end{verbatim}
+
+\begin{verbatim}
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+\end{verbatim}
+
+\begin{verbatim}
+#> Nonlinear regression model
+#>   model: weed ~ SSlogis(tt, Asym, xmid, scal)
+#>    data: parent.frame()
+#>   Asym   xmid   scal 
+#> 199.86  12.50   3.19 
+#>  weighted residual sum-of-squares: 0.0196
+#> 
+#> Number of iterations to convergence: 2 
+#> Achieved convergence tolerance: 8.29e-06
+\end{verbatim}
+
+\begin{verbatim}
+usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+#> attr(,"gradient")
+#>           Asym     xmid    scal
+#>  [1,] 0.026498  -1.6157  5.8228
+#>  [2,] 0.035901  -2.1679  7.1333
+#>  [3,] 0.048474  -2.8890  8.6006
+#>  [4,] 0.065153  -3.8149 10.1617
+#>  [5,] 0.087046  -4.9775 11.6983
+#>  [6,] 0.115387  -6.3933 13.0222
+#>  [7,] 0.151425  -8.0483 13.8709
+#>  [8,] 0.196222  -9.8787 13.9297
+#>  [9,] 0.250362 -11.7553 12.8918
+#> [10,] 0.313611 -13.4827 10.5608
+#> [11,] 0.384641 -14.8251  6.9663
+#> [12,] 0.460954 -15.5631  2.4357
+\end{verbatim}
+
+\begin{verbatim}
+source("nlsModel.R")
+nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
+#> [11] 1.6572 1.4214
+\end{verbatim}
+
+\begin{verbatim}
+nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
+nmod$resid()
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
 \end{verbatim}
@@
 \hypertarget{interim-output-from-the-port-algorithm}{%
-\subsubsection{Interim output from the ``port''
-algorithm}\label{interim-output-from-the-port-algorithm}}
-
-As the \texttt{nls()} \textbf{man} page states, when the ``port''
-algorithm is used with the \texttt{trace} argument TRUE, the iterations
-display the objective function value which is 1/2 the sum of squares (or
-deviance). It is likely that the trace display is embedded in the
-Fortran of the \texttt{nlminb} routine that is called to execute the
+\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}
+
+As the \texttt{nls()} \textbf{man} page states, when the ``port'' algorithm is used with the
+\texttt{trace} argument TRUE, the iterations display the objective function value which
+is 1/2 the sum of squares (or deviance). It is likely that the trace display is
+embedded in the Fortran of the \texttt{nlminb} routine that is called to execute the
 ``port'' algorithm, but the factor of 2 discrepancy is nonetheless
@@
 \hypertarget{failure-to-return-best-result-achieved}{%
-\subsubsection{Failure to return best result
-achieved}\label{failure-to-return-best-result-achieved}}
-
-If \texttt{nls()} reaches a point where it cannot continue but has not
-found a point where the relative offset convergence criterion is met, it
-may simply exit, especially if a ``singular gradient'' (singular
-Jacobian) is found. However, this may occur AFTER the function has made
-considerable progress in reducing the sum of squared residuals. Here is
-an abbreviated example:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
-\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
-\NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
-\NormalTok{NLSstart }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{lrc1 =} \SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\AttributeTok{lrc2 =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{A1 =} \DecValTok{150}\NormalTok{, }\AttributeTok{A2 =} \DecValTok{50}\NormalTok{) }\CommentTok{\# a starting vector (named!)}
-\NormalTok{NLSformula }\OtherTok{\textless{}{-}}\NormalTok{ conc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ A1 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc1) }\SpecialCharTok{*}\NormalTok{ time) }\SpecialCharTok{+}\NormalTok{ A2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc2) }\SpecialCharTok{*}\NormalTok{ time)}
-\NormalTok{tryit }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(NLSformula, }\AttributeTok{data =}\NormalTok{ NLSdata, }\AttributeTok{start =}\NormalTok{ NLSstart, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 61216.    (3.56e+03): par = (-2 0.25 150 50)
-## 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
-## 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
-## Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(tryit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
-## attr(,"class")
-## [1] "try-error"
-## attr(,"condition")
-## <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
+\subsubsection{Failure to return best result achieved}\label{failure-to-return-best-result-achieved}}
+
+If \texttt{nls()} reaches a point where it cannot continue but has not found a point
+where the relative offset convergence criterion is met, it may simply exit,
+especially if a ``singular gradient'' (singular Jacobian) is found. However, this
+may occur AFTER the function has made considerable progress in reducing the sum
+of squared residuals.
+
+Here is an abbreviated example:
+
+\begin{verbatim}
+#> 61216.    (3.56e+03): par = (-2 0.25 150 50)
+#> 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
+#> 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
+#> Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
+#> attr(,"class")
+#> [1] "try-error"
+#> attr(,"condition")
+#> <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
 \end{verbatim}
@@
 Note that the sum of squares has been reduced from 61216 to 1.6211, but
-unless \texttt{trace} is invoked, the user will not get any information
-about this. This almost trivial change to the \texttt{nls()} function
-could be useful to R users.
+unless \texttt{trace} is invoked, the user will not get any information about this.
+This almost trivial change to the \texttt{nls()} function could be
+useful to R users.
 
 \hypertarget{feature-partially-linear-models-and-their-specification}{%
-\subsection{Feature: partially linear models and their
-specification}\label{feature-partially-linear-models-and-their-specification}}
-
-The variable projection method (Golub and Pereyra
-(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
-(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
-effective than general approaches in finding good solutions to nonlinear
-least squares problems when some of the parameters appear linearly.
-However, setting up the calculations, that is, identifying which
-parameters are linear, is not a trivial task.
-
-Moreover, specifying a model to a solver should, ideally, use the same
-syntax across solver tools. Unfortunately, R allows multiple approaches
-within different modeling tools, and within \texttt{nls()} itself. This
-is highlighted by the case of partially linear models.
-
-The nonlinear modeling specifications are, of course, a development of
-linear ones. Unfortunately, the explicit model
-\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the
-linear modeling function \texttt{lm()}, which requires this model to be
-specified as \texttt{y\ \textasciitilde{}\ x}.
-
-Within \texttt{nls()}, consider the following FOUR different
-specifications for the same problem, plus an intuitive choice, labelled
-\texttt{fm2a}, that does NOT work. In this failed attempt, putting the
-\texttt{Asym} parameter in the model causes the \texttt{plinear}
+\subsection{Feature: partially linear models and their specification}\label{feature-partially-linear-models-and-their-specification}}
+
+The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is generally much more
+effective than general approaches in finding good solutions to nonlinear least
+squares problems when some of the parameters appear linearly. However, setting
+up the calculations, that is, identifying which parameters are linear, is not a
+trivial task.
+
+Moreover, specifying a model to a solver should, ideally, use the same syntax across
+solver tools. Unfortunately, R allows multiple approaches within different
+modeling tools, and within \texttt{nls()} itself. This is highlighted by the case of
+partially linear models.
+
+The nonlinear modeling specifications are, of course, a development of linear ones.
+Unfortunately, the explicit model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the linear
+modeling function \texttt{lm()}, which requires this model to be specified as \texttt{y\ \textasciitilde{}\ x}.
+
+Within \texttt{nls()}, consider the following FOUR different specifications for the same
+problem, plus an intuitive choice, labelled \texttt{fm2a}, that does NOT work. In this
+failed attempt, putting the \texttt{Asym} parameter in the model causes the \texttt{plinear}
 algorithm to try to add another term to the model. We believe this is
-unfortunate, and would like to see a consistent syntax. At the time of
-writing, we do not foresee a resolution for this issue. In the example,
-we have NOT evaluated the commands to save space.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{DNase1 }\OtherTok{\textless{}{-}} \FunctionTok{subset}\NormalTok{(DNase, Run }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\CommentTok{\# select the data}
-\DocumentationTok{\#\# using a selfStart model {-} do not specify the starting parameters}
-\NormalTok{fm1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(}\FunctionTok{log}\NormalTok{(conc), Asym, xmid, scal), DNase1)}
-\FunctionTok{summary}\NormalTok{(fm1)}
-
-\DocumentationTok{\#\# using conditional linearity {-} leave out the Asym parameter}
-\NormalTok{fm2 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
-\FunctionTok{summary}\NormalTok{(fm2)}
-
-\DocumentationTok{\#\# without conditional linearity}
-\NormalTok{fm3 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1,}
-                 \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm3)}
-
-\DocumentationTok{\#\# using Port\textquotesingle{}s nl2sol algorithm}
-\NormalTok{fm4 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm4)}
-
-\DocumentationTok{\#\# using conditional linearity AND Asym does NOT work}
-\NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm2a)}
-\end{Highlighting}
-\end{Shaded}
+unfortunate, and would like to see a consistent syntax. At the time of writing,
+we do not foresee a resolution for this issue. In the example, we have NOT
+evaluated the commands to save space.
+
+\begin{verbatim}
+DNase1 <- subset(DNase, Run == 1) # select the data
+## using a selfStart model - do not specify the starting parameters
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
+
+## using conditional linearity - leave out the Asym parameter
+fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(xmid = 0, scal = 1),
+                 algorithm = "plinear")
+summary(fm2)
+
+## without conditional linearity
+fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1,
+                 start = list(Asym = 3, xmid = 0, scal = 1))
+summary(fm3)
+
+## using Port's nl2sol algorithm
+fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+                 algorithm = "port"))
+summary(fm4)
+
+## using conditional linearity AND Asym does NOT work
+fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
+                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                 algorithm = "plinear", trace = TRUE))
+summary(fm2a)
+\end{verbatim}
 
 \hypertarget{feature-indexed-parameters}{%
-\subsection{Feature: indexed
-parameters}\label{feature-indexed-parameters}}
-
-Indexed parameters -- those whose names have subscripts on some element
-that is part of the coefficient list -- are a useful idea to simplify
-model specification in some cases. However, they are also a focus of
-trouble.
-
-The \textbf{man} file for \texttt{nls()} includes the following example
-of a situation in which parameters are indexed. It also uses the
-``plinear'' option as an added complication.
+\subsection{Feature: indexed parameters}\label{feature-indexed-parameters}}
+
+Indexed parameters -- those whose names have subscripts on some element that
+is part of the coefficient list --
+are a useful idea to simplify model specification in some cases. However, they
+are also a focus of trouble.
+
+The \textbf{man} file for \texttt{nls()} includes the following example of a situation in which
+parameters are indexed. It also uses the ``plinear'' option as an added complication.
 
@@
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\# The muscle dataset in MASS is from an experiment on muscle}
-\DocumentationTok{\#\# contraction on 21 animals.  The observed variables are Strip}
-\DocumentationTok{\#\# (identifier of muscle), Conc (Cacl concentration) and Length}
-\DocumentationTok{\#\# (resulting length of muscle section).}
-\ControlFlowTok{if}\NormalTok{(}\SpecialCharTok{!} \FunctionTok{requireNamespace}\NormalTok{(}\StringTok{"MASS"}\NormalTok{, }\AttributeTok{quietly =} \ConstantTok{TRUE}\NormalTok{)) }\FunctionTok{stop}\NormalTok{(}\StringTok{"Need MASS pkg"}\NormalTok{)}
-\NormalTok{mm}\OtherTok{\textless{}{-}}\NormalTok{ MASS}\SpecialCharTok{::}\NormalTok{muscle[}\DecValTok{1}\SpecialCharTok{:}\DecValTok{12}\NormalTok{,] }\CommentTok{\# take only 1st few values of Strip (TRUNCATION OF EXAMPLE)}
-\NormalTok{mm}\OtherTok{\textless{}{-}}\FunctionTok{droplevels}\NormalTok{(mm) }\CommentTok{\# remove unused levels after truncation}
-\NormalTok{nlev }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm)}
-\FunctionTok{withAutoprint}\NormalTok{(\{}
-  \DocumentationTok{\#\# The non linear model considered is}
-  \DocumentationTok{\#\#       Length = alpha + beta*exp({-}Conc/theta) + error}
-  \DocumentationTok{\#\# where theta is constant. For now alpha and beta do NOT vary with Strip.}
-  \FunctionTok{with}\NormalTok{(mm, }\FunctionTok{table}\NormalTok{(Strip)) }\CommentTok{\# 2, 3 or 4 obs per strip}
-\NormalTok{  nl }\OtherTok{\textless{}{-}} \FunctionTok{nlevels}\NormalTok{(mm}\SpecialCharTok{$}\NormalTok{Strip)}
-  \DocumentationTok{\#\# We first use the plinear algorithm to fit an overall model,}
-  \DocumentationTok{\#\# ignoring that alpha and beta might vary with Strip.}
-\NormalTok{  musc}\FloatTok{.1} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}} \FunctionTok{cbind}\NormalTok{(}\DecValTok{1}\NormalTok{, }\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th)), mm,}
-                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{th =} \DecValTok{1}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
-  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
-
-  \DocumentationTok{\#\# Then we use nls\textquotesingle{} indexing feature for parameters in non{-}linear}
-  \DocumentationTok{\#\# models to use the conventional algorithm to fit a model in which}
-  \DocumentationTok{\#\# alpha and beta vary with Strip.  The starting values are provided}
-  \DocumentationTok{\#\# by the previously fitted model.}
-  \DocumentationTok{\#\# Note that with indexed parameters, the starting values must be}
-  \DocumentationTok{\#\# given in a list (with names):}
-\NormalTok{  b }\OtherTok{\textless{}{-}} \FunctionTok{coef}\NormalTok{(musc}\FloatTok{.1}\NormalTok{)}
-\NormalTok{  musc}\FloatTok{.2} \OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(Length }\SpecialCharTok{\textasciitilde{}}\NormalTok{ a[Strip] }\SpecialCharTok{+}\NormalTok{ b[Strip]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{Conc}\SpecialCharTok{/}\NormalTok{th), }\AttributeTok{data=}\NormalTok{mm,}
-                \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{a =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{2}\NormalTok{], nl), }\AttributeTok{b =} \FunctionTok{rep}\NormalTok{(b[}\DecValTok{3}\NormalTok{], nl), }\AttributeTok{th =}\NormalTok{ b[}\DecValTok{1}\NormalTok{]))}
-  \FunctionTok{summary}\NormalTok{(musc}\FloatTok{.2}\NormalTok{)}
-\NormalTok{\})}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## > with(mm, table(Strip))
-## Strip
-## S01 S02 S03 
-##   4   4   4 
-## > nl <- nlevels(mm$Strip)
-## > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
-## +     algorithm = "plinear")
-## > summary(musc.1)
-## 
-## Formula: Length ~ cbind(1, exp(-Conc/th))
-## 
-## Parameters:
-##       Estimate Std. Error t value Pr(>|t|)    
-## th       0.624      0.222    2.81   0.0203 *  
-## .lin1   25.684      1.441   17.82  2.5e-08 ***
-## .lin2  -26.631      6.147   -4.33   0.0019 ** 
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Residual standard error: 2.96 on 9 degrees of freedom
-## 
-## Number of iterations to convergence: 5 
-## Achieved convergence tolerance: 4.71e-07
-## 
-## > b <- coef(musc.1)
-## > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
-## +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
-## > summary(musc.2)
-## 
-## Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
-## 
-## Parameters:
-##    Estimate Std. Error t value Pr(>|t|)    
-## a1  22.9277     0.5154   44.49  1.1e-07 ***
-## a2  27.8606     0.5150   54.09  4.1e-08 ***
-## a3  28.3426     1.0771   26.31  1.5e-06 ***
-## b1 -44.1558    12.3900   -3.56    0.016 *  
-## b2 -43.8084    12.3148   -3.56    0.016 *  
-## b3 -32.8421     2.1553  -15.24  2.2e-05 ***
-## th   0.5548     0.0808    6.86    0.001 ** 
-## ---
-## Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
-## 
-## Residual standard error: 0.798 on 5 degrees of freedom
-## 
-## Number of iterations to convergence: 6 
-## Achieved convergence tolerance: 1.95e-06
-\end{verbatim}
-
-Running the example reveals that the answers for the parameters are NOT
-indexed. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
-but \texttt{a1,\ a2,\ a3}. This is no doubt because programming for
-indexed parameters is extremely challenging.
+\begin{verbatim}
+#> > with(mm, table(Strip))
+#> Strip
+#> S01 S02 S03 
+#>   4   4   4 
+#> > nl <- nlevels(mm$Strip)
+#> > musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, start = list(th = 1), 
+#> +     algorithm = "plinear")
+#> > summary(musc.1)
+#> 
+#> Formula: Length ~ cbind(1, exp(-Conc/th))
+#> 
+#> Parameters:
+#>       Estimate Std. Error t value Pr(>|t|)    
+#> th       0.624      0.222    2.81   0.0203 *  
+#> .lin1   25.684      1.441   17.82  2.5e-08 ***
+#> .lin2  -26.631      6.147   -4.33   0.0019 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> Residual standard error: 2.96 on 9 degrees of freedom
+#> 
+#> Number of iterations to convergence: 5 
+#> Achieved convergence tolerance: 4.71e-07
+#> 
+#> > b <- coef(musc.1)
+#> > musc.2 <- nls(Length ~ a[Strip] + b[Strip] * exp(-Conc/th), data = mm, 
+#> +     start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
+#> > summary(musc.2)
+#> 
+#> Formula: Length ~ a[Strip] + b[Strip] * exp(-Conc/th)
+#> 
+#> Parameters:
+#>    Estimate Std. Error t value Pr(>|t|)    
+#> a1  22.9277     0.5154   44.49  1.1e-07 ***
+#> a2  27.8606     0.5150   54.09  4.1e-08 ***
+#> a3  28.3426     1.0771   26.31  1.5e-06 ***
+#> b1 -44.1558    12.3900   -3.56    0.016 *  
+#> b2 -43.8084    12.3148   -3.56    0.016 *  
+#> b3 -32.8421     2.1553  -15.24  2.2e-05 ***
+#> th   0.5548     0.0808    6.86    0.001 ** 
+#> ---
+#> Signif. codes:  0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
+#> 
+#> Residual standard error: 0.798 on 5 degrees of freedom
+#> 
+#> Number of iterations to convergence: 6 
+#> Achieved convergence tolerance: 2.03e-06
+\end{verbatim}
+
+Running the example reveals that the answers for the parameters are NOT indexed.
+That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
+This is no doubt because programming for indexed parameters is extremely challenging.
 
 \hypertarget{feature-bounds-on-parameters}{%
-\subsection{Feature: bounds on
-parameters}\label{feature-bounds-on-parameters}}
-
-There are many situations where the context of a problem constrains the
-values of parameters. For example, one of us was asked many years ago to
-estimate a model where one parameter estimate was negative. The client
-complained ``But that is supposed to be the number of grain elevators in
-Saskatchewan''. The number should have been a positive integer, and
-likely less than a few thousand.
-
-\texttt{nls()} can impose bounds on parameters, but only if the
-\texttt{port} algorithm is used. Unfortunately, the manual states
-
-\begin{quote}
-The \texttt{algorithm\ =\ "port"} code appears unfinished, and does not
-even check that the starting value is within the bounds. Use with
-caution, especially where bounds are supplied.
-\end{quote}
-
-This is clearly unsatisfactory for general use, but does find bounded
-solutions. Inclusion of bounds on parameters in nonlinear least squares
-computations is relatively straightforward, and it is part of the
-default method for package \texttt{nlsr}. Package \texttt{minpack.lm}
-also includes provision for bounds, but the following script, which we
-do not run here for reasons of space, shows that it does not find a good
-solution for the scaled Hobbs test problem.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# bhobbsX.R \#\# bounded formula specification of problem using Hobbs Weed problem}
-\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
-          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{wf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(}\AttributeTok{y =}\NormalTok{ weed, }\AttributeTok{tt =}\NormalTok{ tt)}
-\NormalTok{st }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
-\NormalTok{wmods }\OtherTok{\textless{}{-}}\NormalTok{ y }\SpecialCharTok{\textasciitilde{}} \DecValTok{100} \SpecialCharTok{*}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{ b3 }\SpecialCharTok{*}\NormalTok{ tt)) }\DocumentationTok{\#\# Scaled model}
-\FunctionTok{require}\NormalTok{(minpack.lm)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Loading required package: minpack.lm
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
-\NormalTok{anlxb1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nlsr::nlxb():"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using nlsr::nlxb():
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlxb1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  9.4726  on  12 observations
-##     after  12    Jacobian and  12 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1                     2U           NA         NA         NA           0       23.35  
-## b2               4.43325            NA         NA         NA   -2.16e-07           0  
-## b3                     3U           NA         NA         NA           0           0
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\# nlsLM seems NOT to work properly with bounds}
-\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"}\SpecialCharTok{\textbackslash{}n}\StringTok{"}\NormalTok{); }\FunctionTok{cat}\NormalTok{(}\StringTok{"using minpack.lm::nlsLM():"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## using minpack.lm::nlsLM():
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlsLM1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
-##    data: wf
-## b1 b2 b3 
-##  2  6  3 
-##  residual sum-of-squares: 881
-## 
-## Number of iterations to convergence: 2 
-## Achieved convergence tolerance: 1.49e-08
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsb}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(wmods, }\AttributeTok{start =}\NormalTok{ st, }\AttributeTok{data =}\NormalTok{ wf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{, }
-           \AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{, }\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{, }\DecValTok{6}\NormalTok{, }\DecValTok{3}\NormalTok{))}
-\FunctionTok{cat}\NormalTok{(}\StringTok{"Using nls with \textquotesingle{}port\textquotesingle{}:"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Using nls with 'port':
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlsb)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Nonlinear regression model
-##   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
-##    data: wf
-##   b1   b2   b3 
-## 2.00 4.43 3.00 
-##  residual sum-of-squares: 9.47
-## 
-## Algorithm "port", convergence message: both X-convergence and relative convergence (5)
+\subsection{Feature: bounds on parameters}\label{feature-bounds-on-parameters}}
+
+There are many situations where the context of a problem constrains the values of
+parameters. For example, one of us was asked many years ago to estimate a model
+where one parameter estimate was negative. The client complained ``But that is
+supposed to be the number of grain elevators in Saskatchewan''. The number
+should have been a positive integer, and likely less than a few thousand.
+
+\texttt{nls()} can impose bounds on parameters, but only if the \texttt{port} algorithm is used.
+Unfortunately, the manual states
+
+\begin{quote}
+The \texttt{algorithm\ =\ "port"} code appears unfinished, and does not even check that the starting
+value is within the bounds. Use with caution, especially where bounds are supplied.
+\end{quote}
+
+This is clearly unsatisfactory for general use, but does find bounded solutions.
+Inclusion of bounds on parameters in nonlinear least squares computations is
+relatively straightforward, and it is part of the default method
+for package \texttt{nlsr}. Package \texttt{minpack.lm} also includes provision for bounds, but
+the following script
+shows that it does not find a good solution for the scaled Hobbs test problem.
+
+\begin{verbatim}
+#> Using nlsr::nlxb():
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  9.4726  on  12 observations
+#>     after  12    Jacobian and  12 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> b1                     2U           NA         NA         NA           0       23.35  
+#> b2               4.43325            NA         NA         NA   -2.16e-07           0  
+#> b3                     3U           NA         NA         NA           0           0
+\end{verbatim}
+
+\begin{verbatim}
+#> using minpack.lm::nlsLM():
+\end{verbatim}
+
+\begin{verbatim}
+#> Nonlinear regression model
+#>   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
+#>    data: wf
+#> b1 b2 b3 
+#>  2  6  3 
+#>  residual sum-of-squares: 881
+#> 
+#> Number of iterations to convergence: 2 
+#> Achieved convergence tolerance: 1.49e-08
+\end{verbatim}
+
+\begin{verbatim}
+#> Using nls with 'port':
+\end{verbatim}
+
+\begin{verbatim}
+#> Nonlinear regression model
+#>   model: y ~ 100 * b1/(1 + 10 * b2 * exp(-0.1 * b3 * tt))
+#>    data: wf
+#>   b1   b2   b3 
+#> 2.00 4.43 3.00 
+#>  residual sum-of-squares: 9.47
+#> 
+#> Algorithm "port", convergence message: both X-convergence and relative convergence (5)
 \end{verbatim}
@@
 \hypertarget{philosophical-considerations}{%
-\subsubsection{Philosophical
-considerations}\label{philosophical-considerations}}
-
-Bounds on parameters raise some interesting and difficult questions
-about how uncertainty in parameter estimates should be computed or
-reported. That is, the traditional ``standard errors'' are generally
-taken to imply symmetric intervals about the point estimate in which the
-parameter may be expected to be found with some probability under
-certain assumptions. Bounds change those assumptions. Hence,
-\texttt{nlsr::nlxb()} does not compute standard errors nor their derived
-statistics when bounds are active.
+\subsubsection{Philosophical considerations}\label{philosophical-considerations}}
+
+Bounds on parameters raise some interesting and difficult questions about how
+uncertainty in parameter estimates should be computed or reported. That is, the
+traditional ``standard errors'' are generally taken to imply symmetric intervals
+about the point estimate in which the parameter may be expected to be found with
+some probability under certain assumptions.
+Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not compute standard
+errors nor their derived statistics when bounds are active.
 
 \hypertarget{issue-code-structure-and-documentation-for-maintenance}{%
-\subsection{Issue: code structure and documentation for
-maintenance}\label{issue-code-structure-and-documentation-for-maintenance}}
-
-The \texttt{nls()} code is structured in a way that inhibits both
-maintenance and improvement. In particular, the iterative setup is such
-that introduction of Marquardt stabilization is not easily available. To
-obtain performance when the code was written, a large portion was
-written in C with consequent calls and returns that severely complicate
-the code. Over time, R has become much more efficient on modern
-computers, and the need to use compiled C and Fortran is less critical.
-Moreover, the burden for maintenance could be reduced by moving code
-entirely to R.
-
-While R and its packages are generally very well documented for usage,
-that documentation rarely extends as far as it might for code
-maintenance. The paucity of such documentation is exacerbated by the
-mixed R/C/Fortran code base, where seemingly trivial differences in
-things like representation of numbers or base index for arrays lead to
-traps for unwary programmers.
+\subsection{Issue: code structure and documentation for maintenance}\label{issue-code-structure-and-documentation-for-maintenance}}
+
+The \texttt{nls()} code is structured in a way that inhibits both maintenance and
+improvement. In particular, the iterative setup is such that introduction
+of Marquardt stabilization is not easily available. To obtain performance
+when the code was written, a large portion was written in C with consequent
+calls and returns that severely complicate the code. Over time, R has become
+much more efficient on modern computers, and the need to use compiled C and
+Fortran is less critical. Moreover, the burden for maintenance
+could be reduced by moving code entirely to R.
+
+While R and its packages are generally very well documented for usage, that
+documentation rarely extends as far as it might for code maintenance.
+The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
+where seemingly trivial differences in things like representation of numbers or
+base index for arrays lead to traps for unwary programmers.
 
 \hypertarget{goals-and-progress-of-our-effort-to-improve-nls}{%
-\section{\texorpdfstring{Goals and progress of our effort to improve
-\texttt{nls()}}{Goals and progress of our effort to improve nls()}}\label{goals-and-progress-of-our-effort-to-improve-nls}}
+\section{\texorpdfstring{Goals and progress of our effort to improve \texttt{nls()}}{Goals and progress of our effort to improve nls()}}\label{goals-and-progress-of-our-effort-to-improve-nls}}
 
 \hypertarget{code-rationalization-and-documentation}{%
-\subsection{Code rationalization and
-documentation}\label{code-rationalization-and-documentation}}
-
-When we started \emph{Improvements to nls()} for the 2021 Google Summer
-of Code we wanted:
+\subsection{Code rationalization and documentation}\label{code-rationalization-and-documentation}}
+
+When we started \emph{Improvements to nls()} for the 2021 Google Summer of Code we wanted:
 
@@
 \item
-  to provide a packaged version of \texttt{nls()} (call it
-  \texttt{nlsalt}) coded entirely in R that matches the features in base
-  R or what is packaged in \texttt{nlspkg} as described in Nash and
-  Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022}). The R
-  version can either serve as the operational code or as a reference
-  version to facilitate the maintenance of C, Fortran, or other versions
-  of the software.
+  to provide a packaged version of \texttt{nls()} (call it \texttt{nlsalt}) coded entirely in R
+  that matches the features in base R or what is packaged in \texttt{nlspkg} as described
+  in Nash and Bhattacharjee (2022). The R version can either serve as the operational code or
+  as a reference version to facilitate the maintenance of C, Fortran, or other
+  versions of the software.
 \item
   to streamline the overall \texttt{nls()} infrastructure. By this we
-  mean a re-factoring of the routines so they are better suited to
-  maintenance of both the existing \texttt{nls()} methods and features
-  as well as new features we or others would like to add.
-\item
-  to explain what we do, either in comments, examples, or separate
-  maintainer documentation. Since we are complaining about the lack of
-  explanatory material for the current code, we feel it is incumbent on
-  us to provide such material for our own work, and if possible for the
-  existing code.
+  mean a re-factoring of the routines so they are better suited to maintenance of both
+  the existing \texttt{nls()} methods and features as well as new features we or others
+  would like to add.
+\item
+  to explain what we do, either in comments, examples, or separate maintainer
+  documentation. Since we are complaining about the lack of explanatory material
+  for the current code, we feel it is incumbent on us to provide such material
+  for our own work, and if possible for the existing code.
 \end{itemize}
 
-These goals echo themes in the recent discussion by Vidoni
-(\protect\hyperlink{ref-Vidoni21}{2021}) and accompanying commentary.
+These goals echo themes in the recent discussion by Vidoni (2021) and accompanying
+commentary.
 
 \hypertarget{rationalization-of-formula-specifications}{%
-\subsection{Rationalization of formula
-specifications}\label{rationalization-of-formula-specifications}}
-
-We have seen that \texttt{nls()} uses a different formula specification
-from the default if the \texttt{plinear} algorithm is used. This is
-unfortunate, since the user cannot then simply add
-\texttt{algorithm="plinear"} to the standard call, making errors more
-likely.
-
-Nonlinear models, apart from those using the ``plinear'' algorithm, need
-all the parameters specified. That is, the model formula should be
-written down as it would be computed. Thus a linear model (which is
-valid input for nonlinear estimation) should be specified
-\texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than
-\texttt{y\ \textasciitilde{}\ x}. Moreover, we want to avoid the
-partially linear parameters appearing in the result object as
-\texttt{.lin1}, \texttt{.lin2}, etc., where the actual position of
-parameters in the model is not explicit. We are unclear why these
-parameters begin with a dot, that is, that they have the ``hidden''
-indicator.
+\subsection{Rationalization of formula specifications}\label{rationalization-of-formula-specifications}}
+
+We have seen that
+\texttt{nls()} uses a different formula specification from the default if the
+\texttt{plinear} algorithm is used. This is unfortunate, since the user cannot then simply
+add \texttt{algorithm="plinear"} to the standard call, making errors more likely.
+
+Nonlinear models, apart from those using the ``plinear'' algorithm, need all the
+parameters specified.
+That is, the model formula should be written down as it would be computed.
+Thus a linear model (which is valid input for nonlinear estimation) should
+be specified \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} rather than \texttt{y\ \textasciitilde{}\ x}.
+Moreover, we want to avoid the partially linear parameters appearing
+in the result object as \texttt{.lin1}, \texttt{.lin2}, etc., where the actual position
+of parameters in the model is not explicit. We are unclear why these parameters
+begin with a dot, that is, that they have the ``hidden'' indicator.
 
 A possible workaround to allow consistency would be to specify the
-partially linear parameters when the \texttt{plinear} option is
-specified. For example, we could use \texttt{algorithm="plinear(Asym)"}
-while keeping the general model formula from the default specification.
-This would allow for the output of different \texttt{algorithm} options
-to be consistent. However, we have not yet tried to code such a change,
-and welcome discussion from those working with other packages using
-model formulas.
-
-A further complication of model specification in R is that some formulas
-require the user to surround a term with \texttt{I()} to inhibit
-interpretation of arithmetic operators. We would prefer that formulas be
-usable without this complication.
+partially linear parameters when the \texttt{plinear} option is specified. For example,
+we could use \texttt{algorithm="plinear(Asym)"} while keeping the general model formula
+from the default specification. This would allow
+for the output of different \texttt{algorithm} options to be consistent. However, we have
+not yet tried to code such a change, and welcome discussion from those working with
+other packages using model formulas.
+
+A further complication of model specification in R is that some formulas require
+the user to surround a term with \texttt{I()} to inhibit interpretation of arithmetic
+operators. We would prefer that formulas be usable without this complication.
 
 \hypertarget{rationalization-of-indexed-models}{%
-\subsection{Rationalization of indexed
-models}\label{rationalization-of-indexed-models}}
-
-Indexed models clearly have a place in some areas of research. We need
-to be able to process models such as
-\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)},
-where \texttt{Strip} is the index. Data \texttt{Length} and
-\texttt{Conc} are available for all observations and parameter
-\texttt{th} applies to every fitted point, but \texttt{a{[}Strip{]}} and
-\texttt{b{[}Strip{]}} apply only to data indexed, which could be an
-experimental run number, or some other categorization.
-
-Given that \texttt{{[}\ {]}} provides the R method to index arrays, the
-presence of square brackets in a model formula signals indexing. The
-question is how to perform the correct calculations.
-
-Our view is that inclusion of the indexing \textbf{within}
-\texttt{nls()} or similar tools introduces complexity that could be
-avoided by a suitable wrapper function. The modeling formula must be
-expanded to encompass all the index cases, likely with indexed
-parameters renamed e.g., \texttt{a{[}2{]}} becomes \texttt{a2} as at
-present with \texttt{nls()}. However, a wrapper could process output to
-reconstitute the indexing of parameters.
+\subsection{Rationalization of indexed models}\label{rationalization-of-indexed-models}}
+
+Indexed models clearly have a place in some areas of research.
+We need to be able to process models such as
+\texttt{Length\ \textasciitilde{}\ a{[}Strip{]}\ +\ b{[}Strip{]}\ *\ exp(-Conc\ /\ th)}, where \texttt{Strip} is the index. Data
+\texttt{Length} and \texttt{Conc} are available for all observations and parameter \texttt{th} applies
+to every fitted point, but \texttt{a{[}Strip{]}} and \texttt{b{[}Strip{]}} apply only to data indexed, which
+could be an experimental run number, or some other categorization.
+
+Given that \texttt{{[}\ {]}} provides the R method to index arrays, the presence of square
+brackets in a model
+formula signals indexing. The question is how to perform the correct calculations.
+
+Our view is that inclusion of the indexing \textbf{within} \texttt{nls()} or similar tools
+introduces complexity that could be avoided by a suitable wrapper function. The
+modeling formula must be expanded to encompass all the index cases, likely with
+indexed parameters renamed e.g., \texttt{a{[}2{]}} becomes \texttt{a2} as at present with \texttt{nls()}.
+However, a wrapper could process output to reconstitute the indexing of parameters.
 
@@
 
-We are pessimistic that the overall structure of \texttt{nls()} can be
-streamlined due to the entanglement of so many features with the
-complicated mix of R, C and Fortran. Indeed, despite digging into the
-code over some months, we do not feel confident that we sufficiently
-understand it nor that we could maintain it. On the other hand, we do
-believe equivalent functionality can be built, but with at least some
-differences in how the features will be accessed.
+We are pessimistic that the overall structure of \texttt{nls()} can be streamlined due
+to the entanglement of so many features with the complicated mix of R, C and
+Fortran. Indeed, despite digging into the code over some months, we do not feel
+confident that we sufficiently understand it nor that we could maintain it. On
+the other hand, we do believe equivalent functionality can be built, but with
+at least some differences in how the features will be accessed.
 
 \hypertarget{tests-and-use-case-examples}{%
-\subsection{Tests and use-case
-examples}\label{tests-and-use-case-examples}}
-
-Maintainers of packages need suitable tests and use-case examples in
-order:
+\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}}
+
+Maintainers of packages need suitable tests and use-case examples in order:
 
@@
 \item
-  to ensure packages work properly, in particular, giving results
-  comparable to or better than the functions they are to replace;
-\item
-  to test individual solver functions to ensure they work across the
-  range of calling mechanisms, that is, different ways of supplying
-  inputs to the solver(s);
-\item
-  to pose ``silly'' inputs to see if these bad inputs are caught by the
-  programs.
+  to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace;
+\item
+  to test individual solver functions to ensure they work across the range of
+  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
+\item
+  to pose ``silly'' inputs to see if these bad inputs are caught by the programs.
 \end{itemize}
 
-Such goals align with the aims of \textbf{unit testing} (e.g.,
-\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
-Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
-(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
-package testing tools).
+Such goals align with the aims of \textbf{unit testing}
+(e.g., \emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}}, Wickham (2011), Wickham et al. (2021)
+and the conventional R package testing tools).
 
 \hypertarget{a-testing-package-nlscompare}{%
-\subsection{A testing package:
-NLSCompare}\label{a-testing-package-nlscompare}}
+\subsection{A testing package: NLSCompare}\label{a-testing-package-nlscompare}}
 
 One of us has developed a working prototype package at
-\emph{\url{https://github.com/ArkaB-DS/nlsCompare}} . A primary design
-objective of this is to allow the summarisation of multiple tests in a
-compact output. The prototype has a vignette to illustrate its use.
+\emph{\url{https://github.com/ArkaB-DS/nlsCompare}} .
+A primary design objective of this is to allow the summarisation of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
 
 \hypertarget{documentation-and-resources}{%
-\subsection{Documentation and
-resources}\label{documentation-and-resources}}
+\subsection{Documentation and resources}\label{documentation-and-resources}}
 
 In our investigation, we built several resources, which are now part of
-the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
-Particular items are:
+the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}. Particular items
+are:
 
@@
 \item
-  A BibTex bibliography for use with all documents in this project, but
-  which has wider application to nonlinear least squares projects in
-  general
+  A BibTex bibliography for use with all documents in this project, but which
+  has wider application to nonlinear least squares projects in general
   (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
 \item
-  \texttt{MachID.R} offers a suggested concise summary function to
-  identify a particular computational system used for tests. A
-  discussion of how it was built and the resources needed across
-  platforms is given in at
-  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
-\item
-  Nash and Bhattacharjee (\protect\hyperlink{ref-PkgFromRbase22}{2022})
-  is an explanation of the construction of the \texttt{nlspkg} from the
+  \texttt{MachID.R} offers a suggested concise summary function
+  to identify a particular computational system used for tests. A discussion
+  of how it was built and the resources needed across platforms is given in
+  at \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
+\item
+  Nash and Bhattacharjee (2022) is an explanation of the construction of the \texttt{nlspkg} from the
   \texttt{nls()} code in R-base.
 \item
-  As the 2021 Summer of Code period was ending, one of us (JN) was
-  invited to prepare a review of optimization in R. Ideas from the
-  present work have been instrumental in the creation of Nash
-  (\protect\hyperlink{ref-NashWires22}{2022}).
+  As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
+  a review of optimization in R. Ideas from the present work have been instrumental
+  in the creation of Nash (2022).
 \end{itemize}
@@
 \hypertarget{strategic-choices-in-nonlinear-model-estimation}{%
-\section{Strategic choices in nonlinear model
-estimation}\label{strategic-choices-in-nonlinear-model-estimation}}
-
-It is possible that \texttt{nls()} will remain more or less as it has
-been for the past two decades. Given its importance to R, the focus of
-discussion should be the measures needed to secure its continued
-operation for legacy purposes and how that may be approached. We welcome
-an opportunity to participate in such conversations.
-
-To advance the stability and maintainability of R, we believe the
-program objects (R functions) that are created by tools such as
-\texttt{nls()} should have minimal cross-linkages and side-effects. The
-aspects of \texttt{nls()} that have given us the most trouble:
+\section{Strategic choices in nonlinear model estimation}\label{strategic-choices-in-nonlinear-model-estimation}}
+
+It is possible that \texttt{nls()} will remain more or less as it has been for the past
+two decades. Given its importance to R, the focus of discussion should be the
+measures needed to secure its continued operation for legacy purposes and how
+that may be approached.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program objects (R
+functions) that are created by tools such as \texttt{nls()} should have minimal
+cross-linkages and side-effects.
+
+The aspects of \texttt{nls()} that have given us the most trouble:
 
@@
   The functions that compute the residuals and jacobians often presume
-  the data and parameters needed are available in a particular
-  environment. As long as the correct environment is used, this provides
-  a compact syntax to invoke the calculations. The danger is that the
-  wrong data is accessed if the internal search finds a valid name that
-  is not the object we want.
-\item
-  Weights, subsets, and various contextual controls (such as that for
-  the \texttt{na.action} option) are similarly taken from the first
-  available source, making for a very simple invocation of calculations,
-  but the context is hidden from the user. Furthermore, it can be
-  difficult for those of us trying to maintain or improve the code to be
-  certain we have the context correct.
-\item
-  The mixing of R, C and Fortran code was necessary for computational
-  performance in the past. Lacking decent developer documentation,
-  programmers now face a lot of work to fix or improve code. We believe
-  in having at least a working reference version of code in a single
-  programming language. If necessary, by measuring (``profiling'') the
-  code, we can find bottlenecks and substitute for those slower parts of
-  the reference code.
+  the data and parameters needed are available in a particular environment. As long
+  as the correct environment is used, this provides a compact syntax to invoke the
+  calculations. The danger is that the wrong data is accessed if the internal
+  search finds a valid name that is not the object we want.
+\item
+  Weights, subsets, and various contextual controls (such as that for the
+  \texttt{na.action} option) are similarly taken from the first available source,
+  making for a very simple invocation of calculations, but the context is
+  hidden from the user. Furthermore, it can be difficult for those of us
+  trying to maintain or improve the code to be certain we have the
+  context correct.
+\item
+  The mixing of R, C and Fortran code was necessary for computational performance
+  in the past. Lacking decent developer documentation, programmers now face a
+  lot of work to fix or improve code. We believe in having at least a working
+  reference version of code in a single programming language. If necessary, by
+  measuring (``profiling'') the code, we can find bottlenecks and substitute for
+  those slower parts of the reference code.
 \item
   A design that isolates the setup, solution, and post-solution parts of
-  for complicated calculations reduces the number of objects that must
-  be kept in alignment. On the other hand, it may mean users have to add
-  more commands to complete their computations.
-\item
-  All iterative codes should return the best solution they have found so
-  far, even if there are untoward conditions reached, such as a singular
-  Jacobian. This modification could be made to \texttt{nls()} in the
-  short term, correcting the issue identified above under ``Failure to
-  return best result achieved''.
-\item
-  Given the existence of examples of good practices such as analytic
-  derivatives, stabilized solution of Gauss-Newton equations and
-  bounds-constrained parameters, base R tools should be moving to
-  incorporate them.
+  for complicated calculations reduces the number of objects that must be kept in
+  alignment. On the other hand, it may mean users have to add more commands to
+  complete their computations.
+\item
+  All iterative codes should return the best solution they have found so far,
+  even if there are untoward conditions reached, such as a singular Jacobian.
+  This modification could be made to \texttt{nls()} in the short term, correcting the
+  issue identified above under ``Failure to return best result achieved''.
+\item
+  Given the existence of examples of good practices such as analytic derivatives,
+  stabilized solution of Gauss-Newton equations and bounds-constrained parameters,
+  base R tools should be moving to incorporate them.
 \end{itemize}
 
-We anticipate that tools like \texttt{nlsr} will become more attractive
-as users discover the increased robustness and general capabilities,
-along with the more recent documentation. From a design point of view, a
-key difference in approach between \texttt{nls()} and
-\texttt{nlsr::nlxb()} is that \texttt{nls()} builds a large
-infrastructure, especially the \texttt{nlsModel()} function, from which
-the Gauss-Newton iteration can be executed and other statistical
-information such as profiles can be computed, while \texttt{nlxb()}
-returns quite limited information, and computes what is needed on an
-``as and when'' basis. This follows a path that one of us (JN)
-established almost 50 years ago with the software that became Nash
-(\protect\hyperlink{ref-cnm79}{1979}), separating setup, solver, and
-post-solution analysis phases of computations. There is, however, the
-wrapper \texttt{wrapnlsr()} that solves the nonlinear estimation problem
-with the solver of \texttt{nlsr} then calls \texttt{nls()} to build the
-object of class \texttt{nls} for which profile log-likelihood
-information can be computed.
+We anticipate that tools like \texttt{nlsr} will become more attractive as users discover
+the increased robustness and general capabilities, along with the more recent
+documentation. From a design point of view, a key difference in approach between
+\texttt{nls()} and \texttt{nlsr::nlxb()} is that \texttt{nls()}
+builds a large infrastructure, especially the \texttt{nlsModel()} function, from
+which the Gauss-Newton
+iteration can be executed and other statistical information such as profiles can be
+computed, while \texttt{nlxb()} returns quite limited information, and
+computes what is needed on an ``as and when'' basis. This follows a path that one
+of us (JN) established almost 50 years ago with the software that became Nash (1979),
+separating setup, solver, and post-solution analysis phases of computations.
+There is, however, the wrapper \texttt{wrapnlsr()} that solves the nonlinear estimation problem with the solver of \texttt{nlsr} then calls \texttt{nls()} to build the object of
+class \texttt{nls} for which profile log-likelihood information can be computed.
 
@@
 
-Hans Werner Borchers has been helpful in developing the proposal for
-this project and in comments on this and related work. Heather Turner
-co-mentored the project and helped guide the progress of the work.
-Exchanges with Fernando Miguez helped to clarify aspects of
-\texttt{selfStart} models and improve package \texttt{nlsr} in that
-respect.
+Hans Werner Borchers has been helpful in developing the proposal for this project
+and in comments on this and related work. Heather Turner co-mentored the project
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of \texttt{selfStart} models and improve package \texttt{nlsr} in that respect.
+Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our attempts
+to succinctly summarize computing environments.
 
 \hypertarget{references}{%
-\section*{References}\label{references}}
+\section{References}\label{references}}
+
+\hypertarget{references-1}{%
+\section*{References}\label{references-1}}
 \addcontentsline{toc}{section}{References}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
-Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis
-and Its Applications}. Wiley.
+Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
 
 \leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
-Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
-Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
-\emph{Technometrics} 23 (2): 179--83.
+Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
-Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of
-Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables
-Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
+Golub, G.
+H., and V.
+Pereyra.
+1973.
+{``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
-Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004.
-\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with
-{S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.
+Huet, S., A.
+Bouvier, M.-A.
+Poursat, and E.
+Jolivet.
+2004.
+\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with {S-PLUS} Examples, 2nd Edition}.
+Berlin \& New York: Springer-Verlag.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsr2019manual}{}}%
-John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for
-Nonlinear Least Squares Solutions}}.
+John C Nash, and Duncan Murdoch. 2019. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
-Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain
-Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied
-Mathematics} 2: 164-\/-168.
+Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
-Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation
-of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics}
-11 (2): 431--41.
+Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
 
 \leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
-Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for
-Agricultural Applications}}.
-\url{https://CRAN.R-project.org/package=nlraa}.
+Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
-Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on
-a Small Computer.''} \emph{Journal of the Institute for Mathematics and
-Its Applications} 19: 231--37.
+Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
 
 \leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
----------. 1979. \emph{Compact Numerical Methods for Computers: Linear
-Algebra and Function Minimisation}. Bristol: Adam Hilger.
+---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
 \leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
----------. 2022. {``{Function minimization and nonlinear least squares
-in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
-https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
+---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
-Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package
-from Base {R} Files.''} \emph{R-Bloggers}, January.
-\url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
+Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
-Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation
-Combining Visual Fitting with Optimization Methods.''} In
-\emph{Proceedings of the Section on Physical and Engineering Sciences of
-the American Statistical Association}, 256--61. American Statistical
-Association.
+Nash, John C., and Paul Velleman.
+1996.
+{``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61.
+American Statistical Association.
 
 \leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
-O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
-Nonlinear Least Squares Problems.''} \emph{Computational Optimization
-and Applications} 54 (3): 579--93.
+O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
-Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified
-Practical Approach}. New York; Basel: Marcel Dekker Inc.
+Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Vidoni21}{}}%
-Vidoni, Melina. 2021. {``{Software Engineering and R Programming: A Call
-for Research}.''} \emph{{The R Journal}} 13 (2): 6--14.
-\url{https://doi.org/10.32614/RJ-2021-108}.
+Vidoni, Melina. 2021. {``{Software Engineering and R Programming: A Call for Research}.''} \emph{{The R Journal}} 13 (2): 6--14. \url{https://doi.org/10.32614/RJ-2021-108}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
-Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
-\emph{The R Journal} 3: 5--10.
-\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
+Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
-Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
-\emph{{devtools: Tools to Make Developing R Packages Easier}}.
-\url{https://CRAN.R-project.org/package=devtools}.
+Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.
 
@@
 
-\end{document}
+\bibliography{../BibSupport/ImproveNLS.bib}
+
+\address{%
+John C. Nash\\
+retired professor, University of Ottawa\\%
+Telfer School of Management\\ Ottawa ON Canada K1N 6N5\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0002-2762-8039}{0000-0002-2762-8039}}\\%
+\href{mailto:profjcnash@gmail.com}{\nolinkurl{profjcnash@gmail.com}}%
+}
+
+\address{%
+Arkajyoti Bhattacharjee\\
+Indian Institute of Technology\\%
+Department of Mathematics and Statistics\\ Kanpur\\
+%
+%
+%
+\href{mailto:arkastat98@gmail.com}{\nolinkurl{arkastat98@gmail.com}}%
+}

### 2023-03-02 JNABsub/JNAB230203.Rmd
Submission 2 start

--- 2023-02-03 RefactoringNLS/JNAB230203.Rmd
@@
 ---
-title: "Refactoring `nls()`: a constructive critique"
+title: "Refactoring 'nls()': a Constructive Critique"
 date: "2023-02-03"
@@
 
+\noindent
 gets over 4000 hits that are spread over some years. This could be because the 
@@
 # References
-
-?? to sort out
-- note that AB now at Ohio State?
-- ORCID for AB?

### 2023-03-02 RefactoringNLS/ComparingNLStools.Rmd
Submission 2 start

--- 2023-01-28 RefactoringNLS/RefactoringNLS.Rmd
@@
 ---
-title: "Refactoring `nls()`: a constructive critique"
+title: "A comparison on R tools for nonlinear least squares modeling"
 author: 
@@
    - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
-date: "2022-6-15"
+date: "2023-3-2"
 output: 
@@
         toc: true
-bibliography: ../BibSupport/ImproveNLS.bib
+bibliography: ./ImproveNLS.bib
 link-citations: yes
@@
 
-Based on work for a Google Summer of Code project "Improvements to `nls()`",
-we consider the features and limitations of the R function `nls()`
+In a Google Summer of Code project "Improvements to `nls()`",
+we considered the features and limitations of the R function `nls()`
 in the context of improving and rationalizing R tools for nonlinear regression.
-
-Important considerations are the usability and  maintainability of
+The rich features of this function are weakened by a number of deficiencies and
+inconsistencies, in particular a lack of stabilization of the Gauss-Newton solver.
+Further considerations are the usability and  maintainability of
 the code base that provides the functionality `nls()` claims to offer.
-Our work suggests that the
-existing code makes maintenance and improvement very difficult,
-with legacy applications and examples blocking some important updates.
-Discussion of these matters is relevant to improving R generally as
-well as its nonlinear estimation tools.
+Various packages, including our `nlsr`, provide alternative capabilities.
+This article considers the differences in goals, approach and features of
+different tools for nonlinear least squares modeling in R. Discussion of 
+these matters is relevant to improving R generally as well as its 
+nonlinear estimation tools.
 
@@
 from the Comprehensive R Archive Network (*https://cran.r-project.org*),
-for estimating nonlinear statistical models. The function dates to the 1980s and the work
-related to @bateswatts in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
-
-The `nls()` function has a remarkable and comprehensive set of capabilities for estimating
-nonlinear models that are expressed as **formulas**. In particular, we note that it
+for estimating nonlinear statistical models. The function dates to the 1980s and the 
+work related to @bateswatts in S (see 
+*https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+
+The `nls()` function has a remarkable and comprehensive set of capabilities for 
+estimating nonlinear models that are expressed as **formulas**. In particular, we
+note that it
 
@@
 With such a range of features and long history, the code has become untidy 
-and overly patched. It is, to
-our mind, essentially unmaintainable. Moreover, its underlying methods can and should be improved, 
-as we suggest below.
+and overly patched, difficult to maintain, and its underlying methods could
+be improved. Various workers have developed packages to overcome these 
+concerns, and we will address some of these here.
 

### 2023-03-02 JNABsub/JNAB230203.R

@@ -1,0 +1,139 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit JNAB230203.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----SESingGrad---------------------------------------------------------------
+reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
+x <- 1:11 # x values; 11 data points
+y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
+testdat <- data.frame(x, y) # save in a data frame
+strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
+jform <- y ~ a + b * r^(x - m) + c * x # Formula
+library(nlsr)
+linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
+linexp2 # Note singular values of Jacobian in rightmost column
+# We get an error with nls()
+linexp <- try(nls(jform, data = testdat, start = strt, trace = FALSE))
+
+
+## ----nlswtx, echo=TRUE--------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+nowt
+nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+usewt
+usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+source("nlsModel.R")
+nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
[…]

### 2023-03-02 JNABsub/nlsModel.R

@@ -1,0 +1,170 @@
+#  From file src/library/stats/R/nls.R
+#  Part of the R package, https://www.R-project.org
+#
+#  Copyright (C) 2000-2020 The R Core Team
+#  Copyright (C) 1999-1999 Saikat DebRoy, Douglas M. Bates, Jose C. Pinheiro
+#
+#  The nlsModel() function extracted here.
+
+nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
+{
+    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
+    env <- new.env(hash = TRUE, parent = environment(form))
+    for(i in names(data)) env[[i]] <- data[[i]]
+    ind <- as.list(start)
+    parLength <- 0L
+    for(i in names(ind) ) {
+        temp <- start[[i]]
+        storage.mode(temp) <- "double"
+        env[[i]] <- temp
+        ind[[i]] <- parLength + seq_along(temp)
+        parLength <- parLength + length(temp)
+    }
+    getPars.noVarying <- function() unlist(mget(names(ind), env))
+    getPars <- getPars.noVarying
+    internalPars <- getPars()
+
+    if(!is.null(upper)) upper <- rep_len(upper, parLength)
+    useParams <- rep_len(TRUE, parLength)
+    lhs <- eval(form[[2L]], envir = env)
+    rhs <- eval(form[[3L]], envir = env)
+    .swts <- if(!missing(wts) && length(wts))
+        sqrt(wts) else rep_len(1, length(rhs))
+    env$.swts <- .swts
+    resid <- .swts * (lhs - rhs)
+    dev <- sum(resid^2)
+    if(is.null(attr(rhs, "gradient"))) {
+        getRHS.noVarying <- function() {
+            if(is.null(upper)) # always for "default"
+                numericDeriv(form[[3L]], names(ind), env, central = nDcentral)
+            else # possibly with "port"
+                numericDeriv(form[[3L]], names(ind), env,
+                             dir = ## ifelse(internalPars < upper, 1, -1)
+                                 -1 + 2*(internalPars < upper), central = nDcentral)
+        ## 210701JN: Above line should check for nDcentral=FALSE when approaching
+        ## upper bound, as central diff is same whether dir is +1 or -1. ???
+        }
[…]

### 2023-03-02 hobbSSlogistry.R

@@ -1,0 +1,512 @@
+?microbenchmark
+source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopesctimes.R")
+library(benchmarkme)
+get_sys_details
+sessionInfo
+sessionInfo()
+?COMPILE
+library(compiler)
+?compile
+cmpfun(loopesc)
+le<-cmpfun(loopesc)
+microbenchmark(le)
+microbenchmark(le, unit=us)
+microbenchmark(le, unit="us")
+microbenchmark(loopesc, unit="us")
+nn
+microbenchmark(le(nn), unit="us")
+microbenchmark(loopesc(nn), unit="us")
+microbenchmark()
+?microbenchmark
+boxplot.microbenchmark(t2)
+boxplot.microbenchmark(tt2)
+library(microbenchmark)
+?boxplot.microbenchmark
+# loopbox.R
+loopesc <- function(nn){
+ss <- 0
+for (i in 1:nn) {
+xx <- exp(sin(cos(1.0*i)))
+ss <- ss + xx
+}
+xx
+}
+require("microbenchmark")
+nn <- 10000
+library(compiler)
+le <- cmpfun(loopesc)
+tcompiled <- microbenchmark(le(nn), unit='us')$time
+#  print(te)
+#  boxplot(te)
+tcoldstart <- microbenchmark(loopesc(nn), unit = 'us')$time
+#  print(tt)
+#  boxplot(tt)
+#  ht<-hist(tt$time)
+#  plot(ht, main="Histogram of loopesc times")
+twarmstart <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))$time
+tdf <- data.frame(tcompiled, tcoldstart, twarmstart)
+str(tdf)
+names(tdf)
+?data.frame
+boxplot(x=as.list(tdf))
+boxplot(x=as.list(log(tdf))
+)
+?boxplot
+tdf <- data.frame(tcompiled, tcoldstart, twarmstart)
+boxplot(x=as.list(log(tdf)), main="log times for 10000 loops in microsecs")
+source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
+source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
+source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
[…]

### 2023-03-02 RefactoringNLS/ComparingNLStools.Rmd
new submission

@@
 nonlinear estimation tools.
+
+## Scope of our comparison
+
+The tools we will compare are the base-R `nls()` function 
+and a number of packages that are available on the CRAN repository.
+
+Important:
+
+nlsr
+minpack.lm
+gslnls
+nls2
+
+Worth commentary:
+
+crsnls
+NISTnls
+nlshelper
+nlsic
+nlsMicrobio
+nlsmsn
+nls.multstart
+nlstac
+nlstools
+onls
+
+Mention only:
+
+easynls
+
+## Comparison dimensions
+
+**nlsr** has functions to find a set of parameters that minimizes the sum of 
+squares (or deviance) 
+of differences between a model specified either as a formula using function 
+`nlxb()` or as both residual
+and Jacobian R functions using function `nlfb()`. As described below, the 
+Jacobian may be approximated
+by particular control settings.
+
+Existing R tools that are similar to `nlxb()` are `nls()` and `minpack.lm::nlsLM()`;  
+`minpack.lm::nls.lm()` is similar to `nlfb()`.
+
+There are several important differences in approach and results, some of which 
+are detailed in this document.
+
+### Output object
+
+The output object of `nlxb()` is smaller than that of `nls()` and is focused on 
+the solution and its properties
+rather than the **modeling** task. That is, package `nlsr` emphasizes the solution 
+of the nonlinear least squares problem rather than the estimation of a nonlinear 
+model that fits or explains the data. `nls()` and `nlsLM` return an object of 
+class `nls`, which allows for
+a number of specialized modeling and diagnostic extensions. For compatibility, 
+the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
+This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
+`nls` object. Unless particular modeling features are needed, the use of 
+`wrapnlsr()` is unnecessary and wasteful of resources.
+
+### Jacobian calculation
+
+Internally, `nlxb()` uses symbolic and automatic differentiation tools to create 
+the residual and Jacobian functions needed for `nlfb()` then uses it to solve 
+the relevant nonlinear least squares minimization problem.
+
+As with other nonlinear least squares packages, we provide the Jacobian code as 
+the "gradient" attribute of the Jacobian function. This lets us embed the code 
+for the Jacobian as this attribute of the **residual** function so that the call 
+to `nlfb()` can be made with the same name used for 
+both residual and Jacobian function arguments. This programming trick saves a 
+lot of trouble for the package developer, but it can be a nuisance for users 
+trying to understand the code.
+
+Generally `nls()` and `nlsLM()` use a fairly simple forward difference 
+approximation of derivatives for the Jacobian, though a central approximation 
+can be specified in control parameters. Package `nlsr` provides four approximation 
+options, with a further control `ndstep` for the size of the
+step used in the approximation.
+
+### Stabilization of Gauss-Newton computations
+
+The iteration in all the major tools mentioned is the solution of the Gauss-Newton 
+equations. `nls()` uses a variant of an approach suggested by @Hartley61, while 
+`nlsr` and `minpack.lm` use variants of the method published by @Marquardt1963, 
+though it was suggested earlier by @Levenberg1944. The details of the method in 
+`nlsr` are discussed in  @JNABRefactoringNLS22.
+Though it is unlikely to be useful in general, control settings for `nlxb()` or `nlfb()`
+allow for those routines to perform a variety of Hartley and Marquardt algorithms. This
+has been useful for learning about the algorithms, but tangential to simply finding
+parameters of nonlinear models.
+
+In general, the Levenberg-Marquardt stabilization is important in obtaining solutions.
+The default method of `nls()` terminates with `singular gradient` unnecessarily in 
+too many instances.
+
+### Programming language
+
+An important choice made in developing `nlsr` was to code entirely within the R 
+programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
+Generally, I believe that keeping to a single programming language allows for 
+easier maintenance and upgrades. On the other hand, R is usually somewhat slower 
+in performing computations because it keeps track of names and structures and 
+because it is usually interpreted rather than compiled. In recent years the 
+performance penalty for using code entirely in R has been much reduced
+with the just-in-time compiler and other improvements, so that using an all-R 
+package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
+less of a performance cost than the attempt to ensure the solution has been found, 
+which forces more iterations to be used.
+
+---OLD VERSION BELOW ------------------------
 

### 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd
new submission
update for dimensions of comparison +1

@@
 and a number of packages that are available on the CRAN repository.
-
-Important:
-
-nlsr
-minpack.lm
-gslnls
-nls2
-
-Worth commentary:
-
+Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
+**minpack.lm** (@minpacklm12), 
+gslnls (@gslnls23), and nls2 (@nls2-22).
+
+Other CRAN packages worth Commentary: ??
+
+onls
 crsnls
@@
 nlstools
-onls
-
-Mention only:
-
 easynls
+nlraa
+optimx ??
+
+It is likely that there are a number of other tools that could be relevant to
+this comparison in the Bioconductor (??ref) collection and on repositories like
+Github (??ref) and Gitlab (??ref), but we will not pursue those possibilities 
+here.
 
 ## Comparison dimensions
+
+The tools for nonlinear estimation we consider have been built with different
+goals and contexts. We will make our comparison in a number of directions.
+We will make a number of specific notes, but our main commentary will consider
+the following categories.
+
+### Modeling versus minimizing a sum of squares
+
+The goal of many statisticians and other workers is to estimate a model. The
+minimization of a sum of squared residuals -- the nonlinear least squares problem --
+is the mechanism to estimate the model. Some of the tools we review focus on the
+modeling, so attempt to provide measures of uncertainty and other information. 
+The interface to modeling tools will generally try to streamline that task. By
+contrast, nonlinear least squares tools are trying to solve the more limited
+numerical task.
+
+### Programming language(s)
+
+R tools are our interest, but the tools we consider often use other programming
+languages. There are advantages in using a well-established program that has
+been much used and tested. Early versions of R were relatively sluggish in 
+computational speed. The downside is that we must interface between two or more
+programming languages. Changes to one or the other language or its library support
+force maintenance issues. Moreover, the pool of expertise with languages like C
+and Fortran is diminishing.
+
+### Internal solver stabilization
+
+Early nonlinear least squares programs generally used the Gauss-Newton method 
+(?? ref). This is often very effective and efficient, but for a quite large 
+proportion of problems can fail to find the solution. Stabilization of the 
+computations is relatively straightforward in principle, and overcomes most
+of the failures, though sometimes at the cost of more calculations.
+
+### Jacobian calculation
+
+The necessarily iterative solution of a nonlinear least squares problem 
+to carry out nonlinear modeling requires that we can find smaller values
+of a sum of squared residuals. This implies there is a "downhill" direction.
+The gradient of the residual sum of squares is computed as the product
+of the Jacobian matrix: the partial derivatives of the residual vector 
+with respect to the modeling parameters.
+(?? do we need to present this, or give a ref)
+
+Most researchers find the effort to interface the differential calculus 
+required to specify the Jacobian to a nonlinear estimation tool very 
+challenging. The work-around is generally to use a numerical approximation,
+but there is the possibility of using symbolic or automatic differentiation,
+and there are particular models where workers have provided code for the
+Jacobian.
+
+### Output
+
+?? DO WE WANT A COMPARISON TABLE
 
@@
 and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
-clarify aspects of `selfStart` models and improve package `nlsr` in that respect.
-Colin Gillespie (package `benchmarkme`) has been helpful in guiding our attempts
-to succinctly summarize computing environments.
+clarify aspects of `selfStart` models and instigated the "Introduction to nlsr" 
+vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
+attempts to succinctly summarize computing environments.
 

### 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd
update for dimensions of comparison

@@
 (?? do we need to present this, or give a ref)
-
-Most researchers find the effort to interface the differential calculus 
-required to specify the Jacobian to a nonlinear estimation tool very 
-challenging. The work-around is generally to use a numerical approximation,
-but there is the possibility of using symbolic or automatic differentiation,
-and there are particular models where workers have provided code for the
-Jacobian.
 

### 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd

--- 2023-03-03 RefactoringNLS/ComparingNLStools.Rmd
@@
 
-onls
-crsnls
-NISTnls
-nlshelper
-nlsic
-nlsMicrobio
-nlsmsn
-nls.multstart
-nlstac
-nlstools
-easynls
-nlraa
-optimx ??
+??put in proper refs to each
+
+onls -- when there are errors in variables, this minimizes the orthogonal 
+residuals
+
+crsnls -- Controlled random search. No vignette!
+
+NISTnls -- NIST problems for testing nls methods
+
+nlshelper -- A few utilities for summarizing, testing, and plotting non-linear
+    regression models fit with nls(), nlsList() or nlme()
+
+nlsic -- We solve non linear least squares problems with optional
+    equality and/or inequality constraints.
+
+nlsMicrobio -- Data sets and nonlinear regression models dedicated to predictive microbiology.
+
+nlsmsn -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) regression, details in Garay, Lachos and Abanto-Valle (2011)
+
+nls.multstart -- Non-linear least squares regression with the Levenberg-Marquardt algorithm using multiple starting values for increasing the chance that the minimum found is the global minimum.
+
+nlstac -- Set of functions implementing the algorithm described in Fernandez 
+    Torvisco et al. (2018) for fitting separable nonlinear regression curves.
+    ?? check ref and add to bib.??
+
+nlstools -- Several tools for assessing the quality of fit of a gaussian nonlinear model are provided.
+
+easynls -- Fit and plot some nonlinear models. ?? no vignette!
+
+nlraa -- a set of nonlinear *selfStart* models. Most include analytic Jacobian
+code. (@MiguezNLRAA2021)
+
+optimx -- not really a nonlinear least squares method, but has general optimizers
+that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
+squares.
 
@@
 this comparison in the Bioconductor (??ref) collection and on repositories like
-Github (??ref) and Gitlab (??ref), but we will not pursue those possibilities 
-here.
+Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
+not pursue those possibilities here.
 
@@
 We will make a number of specific notes, but our main commentary will consider
-the following categories.
+the following categories. These overlap, for example in that the emphasis on modeling
+versus solving the nonlinear least squares problem will result in different
+output.f
 
@@
 
+### Constraints on parameters
+
+Reality often imposes conditions on parameters. While general constraints can
+be difficult to include, bounds on parameters can be accommodated by many 
+solvers. For some reason, fixed parameters (masks) are not often considered,
+but are very convenient when a parameter has a commonly accepted value that we
+later wish to allow to be estimated from the data.
+
 ### Output
 
+While the aim of nonlinear estimation is to return a set of parameters for a
+particular model, there is a variety of additional information which may be
+provided. From the perspective of the nonlinear least squares problem, there
+are measures of work and information on the quality of the solution, while
+for statistical purposes measures of uncertainty in the parameters are helpful.
+
+Some tools allow constraints to be imposed on the solution, and it is helpful
+to know if these constraints are active at the proposed solution.
+
 ?? DO WE WANT A COMPARISON TABLE
+
+## An illustrative example
+
+The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
+task which is seemingly straightforward but turns out to be quite nasty. Because
+this problem has a very succinct statement, it often provides the "short 
+reproducible example" much beseeched of contributors to R mailing lists.
+
+The data and a graph of it is given below.
+
+```{r ex01, echo=TRUE}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+```
+
+Three suggested models for this data are (with names to allow for easy reference)
+
+**Logistic3U**:
+$$  y \approx  b_1 / (1 + b_2 * exp(- b_3 * t)) $$
+
+**Logistic3S**:
+$$ y \approx  100 * c_1 / (1 + 10 * c_2 * exp(- 0.1 * c_3 * t)) $$
+
+**Logistic3T**:
+$$ y \approx Asym / (1 + exp((xmid - t)/scal)) $$
+
+The functions above are equivalent, but the first is generally more awkward
+to solve numerically due to its poor scaling. The parameters of the three forms
+are related as follows:
+
+$$   Asym =  b_1 = 100 * c_1 $$
+$$ exp(xmid/scal)  =  b_2 = 10 * c_2 $$
+$$ 1/scal  =  b_3 $$
+To allow for simpler discussion, let us say that the parameters for a vector
+$p$ and the model function is called $model(p)$. The residuals can be written
+either as
+
+$$ r_1 = y - model(p)$$
+
+or
+
+$$ r = model(p) - y $$
+
+since when squared the loss function has an identical value. In both cases, the
+residual vector is a function of $p$. The second form
+has a minor advantage in avoiding a potential sign error 
+if we need to differentiate to get gradients etc.
+
+### Loss function
+
+The general approach of nonlinear optimizers, of which nonlinear least squares
+solvers are usually a subset, is to try to reduce the loss function (or objective
+function) defined by the sum of squares of the residuals ($y - model()$) by 
+altering the model parameters from some starting guess. In all general 
+methods of which we are aware, this is an iterative process.
+
+We note several useful quantities
+
+Loss function:
+$$ S(p) = r' r$$
+Gradient of $S(p)$:
+$$ 2 * J' r$$
+where the Jacobian $J$ is given by elements
+$$ J_{i,j} = \partial r_i / \partial p_j $$
+and the Hessian is defined by elements
+$$ \partial ^2 S(p) / {\partial p_i \partial p_j} $$
+?? how to succinctly write the Hessian in terms of J' J
+
+`nlsr` tries to find solutions from very poor initial parameter values. While
+this is not always possible, `nslr` has generally been able to succeed in finding
+solutions, even when all parameters are started at 1. Let us compare its results
+with those of `nls()` and `nlsLM()`.
+
+### Problem setup
+
+```{r ex02set, echo=TRUE}
+# model formulas
+frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
+frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
+frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+#
+# Starting parameter sets
+stu1<-c(b1=1, b2=1, b3=1)
+sts1<-c(c1=1, c2=1, c3=1)
+stt1<-c(Asym=1, xmid=1, scal=1)
+```
+
+-------------------------------------------------
 

### 2023-03-04 RefactoringNLS/ComparingNLStools.Rmd
More Comparing -- try nls

@@
 
+### Algorithm and termination choices
+
+Iterative algorithms need to be stopped at some point, and how termination
+is forced may alter results. Furthermore, even supposedly similar algorithms
+may differ greatly in detail.
+
 ### Constraints on parameters
@@
 later wish to allow to be estimated from the data.
+
+### Special capabilities
+
+?? weighting, indexed parameters, partially linear ...
 
@@
 
-### Loss function
+### The loss function and its minimization
 
@@
 
+Let us consider there are $n$ parameters and $m$ residuals. 
 We note several useful quantities
@@
 Loss function:
-$$ S(p) = r' r$$
+$$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
 Gradient of $S(p)$:
@@
 and the Hessian is defined by elements
-$$ \partial ^2 S(p) / {\partial p_i \partial p_j} $$
-?? how to succinctly write the Hessian in terms of J' J
-
-`nlsr` tries to find solutions from very poor initial parameter values. While
-this is not always possible, `nslr` has generally been able to succeed in finding
-solutions, even when all parameters are started at 1. Let us compare its results
-with those of `nls()` and `nlsLM()`.
+$$ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} $$
+If we expand the Hessian for nonlinear least squares problems, we find
+
+$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
+Let us use $D_{i,j}$ for the elements of the second term of this expression.
+What is generally called **Newton's method** for function minimization tries
+to set the gradient to zero (to find a stationary point of the function $S(p)$).
+This leads to the **Newton's equation**
+
+$$ H \delta = -g $$
+given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
+to $p + \delta$ and iterate, hopefully to converge on a solution.
+
+Applying this to a sum of squares problem gives
+
+$$ H \delta = (J'J + D) \delta = - J' r$$
+In this expression, only the elements of $D$ have second partial derivatives. 
+Gauss, attempting to model planetary orbits, had small residuals, and noted
+that these multiplied the second partial derivatives of $r$, so he approximated
+
+$$ H \approx J' J$$
+by assuming $D \approx 0$. This results in the Gauss-Newton method where we
+solve
+
+$$ J' J \delta = - J' r$$
+though we can avoid some loss of accuracy by NOT forming the inner product matrix
+$J' J$ by solving the linear least squares matrix problem
+
+$$ J \delta \approx -r $$
+which may be accomplished by a variety of matrix decomposition methods.
+
+In reality, there are many problems were $D$ should not be ignored, but the work
+to compute it precisely is considerable. A number of work-arounds have been 
+proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
+the most commonly used. Henceforth we will simply say "Marquardt" as we believe
+he was the first to actually incorporate the ideas into a practical computer
+program.
+
+The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
+a multiple of the diagonal part of $J' J$. For machines with exceptionally low
+precision arithmetic found on some minicomputers in the 1970s, one of us suggested
+a linear combination of these ideas, as with low precision, some elements of 
+$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
+$D$, especially the scaling multiple, as well as possible line search along the
+direction $\delta$ rather than a unit step, give rise to a panoply of variant
+algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
+work well.
 
 ### Problem setup
+
+Let us specify in R the three model formulae and set some starting
+values for parameters. These starting points are NOT equivalent, and
+are deliberately crude choices. Workers performing many calculations 
+of a similar nature should try to provide good starting points to reduce
+computation time and the chance a false solution is found.
 
@@
 
+It is also possible to provide R functions for the residual and Jacobian. 
+This is usually more work for the user if the formula setup is possible.
+To illustrate, we show the functions for the unscaled 3 parameter logistic.
+
+```{r ex02fn, echo=TRUE}
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3]*tt)
+  zz  <-  1.0/(1+x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -x[1]*zz*zz*yy
+  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
+  attr(jj, "gradient") <- jj
+  jj
+}
+```
+
+## Estimation of models specified as formulas
+
+Using a formula specification was one of the principal advances made
+when `nls()` became available in S sometime in the 1980s. It uses a
+Gauss-Newton (i.e., unstabilized) iteration with step reduction line
+search. This works very efficiently as long as $J$ is not ill-conditioned.
+In a one-parameter problem, think of $J$ as a very small number so that
+we need to watch for a zero-divide or 0/0 situation in computing $\delta$.
+
+```{r ex02nls}
+unls1<-try(nls(formula=frmu, start=stu1, data=weeddf))
+summary(unls1)
+snls1<-try(nls(formula=frms, start=sts1, data=weeddf))
+summary(snls1)
+tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
+summary(tnls1)
+cat("\n")
+```
+
 -------------------------------------------------
+
+`nlsr` tries to find solutions from very poor initial parameter values. While
+this is not always possible, `nslr` has generally been able to succeed in finding
+solutions, even when all parameters are started at 1. Let us compare its results
+with those of `nls()` and `nlsLM()`.
 

### 2023-03-04 RefactoringNLS/ComparingNLStools.Rmd
more added

@@
 reproducible example" much beseeched of contributors to R mailing lists. 
+Nevertheless, this is not an artificially created test problem, but one that
+came from a field researcher.
 
@@
 search. This works very efficiently as long as $J$ is not ill-conditioned.
-In a one-parameter problem, think of $J$ as a very small number so that
-we need to watch for a zero-divide or 0/0 situation in computing $\delta$.
+In a one-parameter problem with ill-conditioning, think of $J$ as a 
+very small number so that we need to watch for a zero-divide or 0/0 
+situation in computing $\delta$.
+The following show `nls()` does poorly on this example problem.
 
@@
 
+Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
+and `gslnls::gsl_nls()`. We use default methods and controls in these examples.
+
+### Solution attempts with nlsr
+
+```{r ex02nlsr}
+library(nlsr)
+unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
+print(unlx1) 
+pshort(unlx1) # A short form output
+snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
+# pshort(snlx1)
+print(snlx1)
+tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
+# pshort(tnlx1)
+print(tnlx1)
+```
+
+### Solution attempts with minpack.lm
+
+```{r ex02minpack}
+library(minpack.lm)
+unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
+summary(unlm1) # Use summary() to get display
+unlm1 # Note the difference. Use this form to get sum of squares
+# pnlm0(unlm1)  # Short form of output
+snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
+# pnlm0(snlm1)
+summary(snlm1)
+## Following structure needed to deal with M1mac issue
+tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
+if (inherits(tnlm1, "try-error")) {
+   cat("Failure to compute solution -- likely singular Jacobian\n")
+} else {  
+   pnlm0(tnlm1) # short form to give sum of squares
+   summary(tnlm1)
+}   
+```
+
+### Solution attempts with gslnls
+
+```{r ex02gslnls}
+library(gslnls)
+ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
+summary(ugslnls1) # Use summary() to get display
+sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
+summary(sgslnls1) # Use summary() to get display
+tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
+summary(tgslnls1) # Use summary() to get display
+```
+
+### Comparison notes - formula setup
+
+`nlsr::nlxb()` is set up to use `print()` to output standard errors and 
+singular values of the Jacobian (for diagnostic purposes). By contrast, 
+`minpack.lm::nlsLM()` and `nls()`
+use `summary()`, which does NOT display the sum of squares, while `print()` 
+gives the sum of squares, but not the standard error of the residuals.
+
+The singular values displayed by `print.nlsr()` (the internal name for the 
+adaptation of the generic `print()`) are displayed in a column to the right 
+of the coefficient  and standard error display, but are NOT specific to the 
+parameters. Their position is purely for efficient use of page space. The 
+most common use of the singular values is to gauge how "nearly singular" the 
+Jacobian is at the solution, and the ratio of the largest to smallest of the 
+singular values is a simple but effective measure. In the above example, we note 
+that the scaled logistic has the smallest ratio.
+
+The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
+large sum of squares, which may suggest that these program have failed. 
+Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
+offer the singular values, we need to extract the Jacobian and compute its
+singular values. The following script shows how to do this, using what is called
+the `gradient` element in the returned solution for these solvers.
+
+```{r ex04singval, echo=TRUE, eval=TRUE}  
+# for nlsLM
+if (inherits(tnlm1, "try-error"))  {
+   print("Cannot compute solution -- likely singular Jacobian")
+ } else {  
+   JtnlsLM <- tnlm1$m$gradient() # actually the Jacobian
+   svd(JtnlsLM)$d # Singular values
+}   
+# for gsl_nls
+if (inherits(tgslnls1, "try-error")) {
+   cat("Cannot compute solution -- likely singular Jacobian")
+} else {  
+   JtnlsLM <- tgslnls1$m$gradient()
+   svd(JtnlsLM)$d # Singular values
+}   
+```
+
+We see that there are differences in detail, but the more important result is that
+two out of three singular values are essentially 0. Our Jacobian is singular, and no 
+method of the Gauss-Newton type should be able to continue. Indeed, from this set of
+parameters, `nlxb` also stops, even though the code in `nlsr` tries to find solutions
+from very poor initial parameter values.
+
+```{r ex05, echo=TRUE}  
+stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
+print(badstart)
+```
+
+## Functional specification of problems
+
+?? what do we want here??
+
+## Design goals, termination tests and output objects
+
+The output object of `nlxb()` is smaller than the class `nls` object returned
+by `nls()`, `nlsLM()` and `gsl_nls()`. Package `nlsr` emphasizes the solution 
+of the nonlinear least squares problem rather than the estimation of a nonlinear 
+model that fits or explains the data. `nls()` and `nlsLM` return an object of 
+class `nls`, which allows for a number of specialized modeling and diagnostic extensions. For compatibility, 
+the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
+This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
+`nls` object. Unless particular modeling features are needed, the use of 
+`wrapnlsr()` is unnecessary and wasteful of resources.
+
+The design goals of the different tools are also evident in the so-called
+"convergence tests" for the iterative solvers. In the manual page for `nls()` 
+in R 4.0.0 there was the warning:
+
+>  **Do not use `nls` on artificial "zero-residual" data.**
+
+and suggested adding small perturbations to the data. This amounted to 
+admitting `nls()` cannot solve well-posed problems unless data is polluted
+with errors. We dispute that zero-residual problems are always artificial,
+since problems in function approximation and nonlinear equations can be 
+approached with nonlinear least squares. Moreover, the matter can
+can be easily resolved. The "termination test" for the **program** rather than for 
+"convergence" of the underlying **algorithm** is the Relative Offset Convergence
+Criterion (see @BatesWatts81). This projects the proposed step in the parameter
+vector on the gradient and estimates how much the sum of squares loss function 
+should decrease. This estimate is divided 
+by the current size of the loss function to avoid scale issues.
+When we have "converged", the estimated
+decrease is very small. However, with small residuals,
+the sum of squares loss function is (almost) zero and we get the possibility of a 
+zero-divide failure.
+
+Adding a small quantity to the loss function before dividing avoids trouble. 
+In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
+number -- 1.0 is a reasonable choice -- allows small-residual problems 
+(i.e., near-exact fits) to be dealt with easily. We call this the
+**safeguarded relative offset convergence criterion**. The default value 
+gives the legacy behaviour.
+We are pleased to note that this improvement is now in the R distributed code 
+since version 4.1.0, and it has been a part of `nlsr` since it was introduced.
+
+More general termination tests can be used. `nlsr` a **small sum of 
+squares** test (**smallsstest**) that compares the latest evaluated sum of squared
+(weighted) residuals to `e4` times the initial sum of squares, where
+`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+
+Termination after what may be considered excessive computation is also important.
+`nls()` stops after `maxiter` "iterations".  
+Unfortunately, the meaning of "iteration" varies among programs and requires 
+careful examination of the
+code. We prefer to use the number of times the residuals or the jacobian have
+been computed and put upper limits on these. Our codes exit (terminate) 
+when these limits are reached. Generally we prefer larger limits than the 
+default `maxiter = 50` of `nls()`,
+but that may reflect the more difficult problems we have encountered because 
+users consult us when standard tools have given unsatisfactory results.
+
+## Jacobian calculation
+
+Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration. 
+By default, `nlsr::nlxb()` will try to evaluate this using analytic 
+expressions using symbolic and automatic differentiation tools. 
+When using formula specification of the model, `nls()`, `minpack.lm::nlsLM()`
+and `gslnls::gsl_nls()` use a finite difference approximation to compute the
+Jacobian, though `gsl_nls()` does have an option to attempt symbolic expressions.
+
+When The Jacobian code as 
+the "gradient" attribute of the output of the Jacobian function. This 
+lets us embed the code 
+for the Jacobian as this attribute of the **residual** function so that the call 
+to `nlfb()` can be made with the same name used for 
+both residual and Jacobian function arguments. This programming trick saves a 
+lot of trouble for the package developer, but it can be a nuisance for users 
+trying to understand the code.
+
+Generally `nls()`, `nlsLM()` and `gsl_nls()` use a fairly 
+simple forward difference approximation of derivatives for the Jacobian
+from the base-R `stats` package called `numericDeriv()`. A central 
+approximation can be specified in control parameters. 
+Package `nlsr` provides, via appropriate calling syntax, four approximation 
+options, with a further control `ndstep` for the size of the
+step used in the approximation.
+
+Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
+and most `selfStart` models that automatically provide starting parameters
+also include such code. There is documentation in R
+of `selfStart`, but the construction of them is non-trivial. Their use with
+`nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
+generally can use very crude starting values, we have rarely needed them.
+
 -------------------------------------------------
-
-`nlsr` tries to find solutions from very poor initial parameter values. While
-this is not always possible, `nslr` has generally been able to succeed in finding
-solutions, even when all parameters are started at 1. Let us compare its results
-with those of `nls()` and `nlsLM()`.
 
@@
 are detailed in this document.
-
-### Output object
-
-The output object of `nlxb()` is smaller than that of `nls()` and is focused on 
-the solution and its properties
-rather than the **modeling** task. That is, package `nlsr` emphasizes the solution 
-of the nonlinear least squares problem rather than the estimation of a nonlinear 
-model that fits or explains the data. `nls()` and `nlsLM` return an object of 
-class `nls`, which allows for
-a number of specialized modeling and diagnostic extensions. For compatibility, 
-the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
-This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
-`nls` object. Unless particular modeling features are needed, the use of 
-`wrapnlsr()` is unnecessary and wasteful of resources.
-
-### Jacobian calculation
-
-Internally, `nlxb()` uses symbolic and automatic differentiation tools to create 
-the residual and Jacobian functions needed for `nlfb()` then uses it to solve 
-the relevant nonlinear least squares minimization problem.
-
-As with other nonlinear least squares packages, we provide the Jacobian code as 
-the "gradient" attribute of the Jacobian function. This lets us embed the code 
-for the Jacobian as this attribute of the **residual** function so that the call 
-to `nlfb()` can be made with the same name used for 
-both residual and Jacobian function arguments. This programming trick saves a 
-lot of trouble for the package developer, but it can be a nuisance for users 
-trying to understand the code.
-
-Generally `nls()` and `nlsLM()` use a fairly simple forward difference 
-approximation of derivatives for the Jacobian, though a central approximation 
-can be specified in control parameters. Package `nlsr` provides four approximation 
-options, with a further control `ndstep` for the size of the
-step used in the approximation.
 
@@
 concerns, and we will address some of these here.
-
-## Feature: Convergence and termination tests (FIXED)
-
-A previous issue with `nls()` that prevented it from providing parameter estimates for zero-residual
-(i.e., perfect fit) data was corrected thanks to suggestions by one of us.
-
-In the manual page for `nls()` in R 4.0.0 there is the warning:
-
->  **Do not use `nls` on artificial "zero-residual" data.**
-
->  The `nls` function uses a relative-offset convergence criterion
-  that compares the numerical imprecision at the current parameter
-  estimates to the residual sum-of-squares.  This performs well on data of
-  the form
-
->  $$  y = f(x, \theta) + eps $$
-
->  (with `var(eps) > 0`).  It fails to indicate convergence on data of the form
-
->  $$  y = f(x, \theta)  $$
-
->  because the criterion amounts to comparing two components of the round-off 
-  error.
-
->  If you wish to test `nls` on artificial data please add a noise component, 
-  as shown in the example below.
-
-This amounted to admitting R cannot solve well-posed problems unless data is polluted
-with errors.
-
-This issue can be easily resolved. The "termination test" for the **program** rather than for 
-"convergence" of the underlying **algorithm** is the Relative Offset Convergence Criterion 
-(see @BatesWatts81). This projects the proposed step in the parameter vector on the gradient 
-and estimates how much the sum of squares loss function should decrease. This estimate is divided 
-by the current size of the loss function to avoid scale issues.
-When we have "converged", the estimated
-decrease is very small. However, with small residuals,
-the sum of squares loss function is (almost) zero and we get the possibility of a 
-zero-divide failure.
-
-Adding a small quantity to the loss function before dividing avoids trouble. 
-In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` 
-with a default value of zero. Setting it to a small number -- 1.0 is 
-a reasonable choice --
-allows small-residual problems (i.e., near-exact fits) to be dealt with easily. We call this the
-**safeguarded relative offset convergence criterion**. The default value gives the legacy behaviour.
-We are pleased to note that this improvement is now in the R distributed code since version 4.1.0.
-
-<!-- ### Example of a small-residual problem -->
-
-<!-- ``` -->
-<!-- t <- -10:10 -->
-<!-- y <- 100/(1 + .1 * exp(-0.51 * t)) -->
-<!-- lform <- y ~ a/(1 + b * exp(-c * t)) -->
-<!-- ldata <- data.frame(t = t, y = y) -->
-<!-- plot(t, y) -->
-<!-- lstartbad <- c(a = 1, b = 1, c = 1) -->
-<!-- lstart2 <- c(a = 100, b = 10, c = 1) -->
-<!-- nlsr::nlxb(lform, data = ldata, start = lstart2) -->
-<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
-<!-- # Fix with scaleOffset -->
-<!-- nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
-<!-- sessionInfo() -->
-<!-- ``` -->
-
-<!-- Edited output of as small example follows: -->
-
-<!-- ``` -->
-<!-- > t <- -10:10 -->
-<!-- > y <- 100/(1 + .1 * exp(-0.51 * t)) -->
-<!-- > lform <- y ~ a/(1 + b * exp(-c * t)) -->
-<!-- > ldata <- data.frame(t=t, y=y) -->
-<!-- > plot(t,y) -->
-<!-- > lstartbad <- c(a = 1, b = 1, c = 1) -->
-<!-- > lstart2 <- c(a = 100, b = 10, c = 1) -->
-<!-- > nlsr::nlxb(lform, data = ldata, start = lstart2) -->
-<!-- nlsr object: x  -->
-<!-- residual sumsquares =  1.007e-19  on  21 observations -->
-<!--     after  13    Jacobian and  19 function evaluations -->
-<!--   name            coeff          SE       tstat      pval      gradient    JSingval    -->
-<!-- a                    100     2.679e-11  3.732e+12  1.863e-216  -6.425e-11       626.6   -->
-<!-- b                    0.1      3.78e-13  2.646e+11  9.125e-196  -3.393e-08       112.3   -->
-<!-- c                   0.51       6.9e-13  7.391e+11  8.494e-204   1.503e-08       2.791   -->
-<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE) -->
-<!-- 40346.    (1.08e+00): par = (100 10 1) -->
-<!-- 11622.    (2.93e+00): par = (101.47 0.49449 0.71685) -->
-<!-- 5638.0    (1.08e+01): par = (102.23 0.38062 0.52792) -->
-<!-- 642.08    (1.04e+01): par = (102.16 0.22422 0.41935) -->
-<!-- 97.712    (1.79e+01): par = (100.7 0.14774 0.45239) -->
-<!-- 22.250    (1.78e+02): par = (99.803 0.093868 0.50492) -->
-<!-- 0.025789  (1.33e+03): par = (100.01 0.10017 0.50916) -->
-<!-- 6.0571e-08 (7.96e+05): par = (100 0.1 0.51) -->
-<!-- 4.7017e-19 (1.86e+04): par = (100 0.1 0.51) -->
-<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
-<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
-<!-- ... (approx 40 lines omitted) -->
-<!-- 1.2440e-27 (5.71e-01): par = (100 0.1 0.51) -->
-<!-- Error in nls(lform, data = ldata, start = lstart2, trace = TRUE) :  -->
-<!--   number of iterations exceeded maximum of 50 -->
-<!-- > # Fix with scaleOffset -->
-<!-- > nls(lform, data = ldata, start = lstart2, trace = TRUE, control = list(scaleOffset = 1.0)) -->
-<!-- 40346.      (1.08e+00): par = (100 10 1) -->
-<!-- 11622.      (2.91e+00): par = (101.47 0.49449 0.71685) -->
-<!-- 5638.0      (9.23e+00): par = (102.23 0.38062 0.52792) -->
-<!-- 642.08      (5.17e+00): par = (102.16 0.22422 0.41935) -->
-<!-- 97.712      (2.31e+00): par = (100.7 0.14774 0.45239) -->
-<!-- 22.250      (1.11e+00): par = (99.803 0.093868 0.50492) -->
-<!-- 0.025789    (3.79e-02): par = (100.01 0.10017 0.50916) -->
-<!-- 6.0571e-08  (5.80e-05): par = (100 0.1 0.51) -->
-<!-- 4.7017e-19  (1.62e-10): par = (100 0.1 0.51) -->
-<!-- Nonlinear regression model -->
-<!--   model: y ~ a/(1 + b * exp(-c * t)) -->
-<!--    data: ldata -->
-<!--      a      b      c  -->
-<!-- 100.00   0.10   0.51  -->
-<!--  residual sum-of-squares: 4.7e-19 -->
-<!-- Number of iterations to convergence: 8  -->
-<!-- Achieved convergence tolerance: 1.62e-10 -->
-<!-- ``` -->
-
-### More general termination tests
-
-The convergence criterion above, the principal termination control of `nls()`, leaves out 
-some possibilities that could
-be useful for some problems. The package `nlsr` (@nlsr2019manual) already offers both 
-the safeguarded relative offset test (**roffset**) as well as a **small sum of 
-squares** test (**smallsstest**) that compares the latest evaluated sum of squared
-(weighted) residuals to `e4` times the initial sum of squares, where
-`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
-
-We note that `nls()` uses a termination test to stop after `maxiter` "iterations". 
-Unfortunately, the meaning of "iteration" varies among programs and requires 
-careful examination of the
-code. We prefer to use the number of times the residuals or the jacobian have
-been computed and put upper limits on these. Our codes exit (terminate) when these limits
-are reached. Generally we prefer larger limits than the default `maxiter = 50` of `nls()`,
-but that may reflect the more difficult problems we have encountered because 
-users consult us when standard tools have given unsatisfactory results.
 

### 2023-03-05 RefactoringNLS/ComparingNLStools.Rmd
continue work

@@
 It is likely that there are a number of other tools that could be relevant to
-this comparison in the Bioconductor (??ref) collection and on repositories like
+this comparison in the Bioconductor (@Bioconductor) collection 
+and on repositories like
 Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
@@
 
-R tools are our interest, but the tools we consider often use other programming
+R tools are our interest, but they often call programs in other programming
 languages. There are advantages in using a well-established program that has
@@
 computational speed. The downside is that we must interface between two or more
-programming languages. Changes to one or the other language or its library support
+programming languages, adding to the risk of errors. Changes to one or the 
+other language or its library support may
 force maintenance issues. Moreover, the pool of expertise with languages like C
@@
 Early nonlinear least squares programs generally used the Gauss-Newton method 
-(?? ref). This is often very effective and efficient, but for a quite large 
+(@Hartley1961). This is often very effective and efficient, but for a quite large 
 proportion of problems can fail to find the solution. Stabilization of the 
@@
 ?? DO WE WANT A COMPARISON TABLE
+
+### Documentation
+
+?? presence of vignettes, demos, (good?) examples
 
@@
 $D$, especially the scaling multiple, as well as possible line search along the
-direction $\delta$ rather than a unit step, give rise to a panoply of variant
+direction $\delta$ rather than a unit step (@Hartley1961), 
+give rise to a panoply of variant
 algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
@@
 situation in computing $\delta$.
-The following show `nls()` does poorly on this example problem.
+The following show `nls()` does poorly on this example problem. Note that `nlsr`
+offers function `pnlm0()` to give a 1-line display of class nls method results,
+and `pshort()` for `nlsr`.
+
+```{r shortforms, echo=FALSE}
+library(nlsr)
+```
 
@@
 summary(tnls1)
-cat("\n")
 ```
@@
 Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
-and `gslnls::gsl_nls()`. We use default methods and controls in these examples.
+and `gslnls::gsl_nls()`, which we will call the **class nls** methods.
+We use default methods and controls in these examples.
 
@@
 snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
-# pshort(snlx1)
-print(snlx1)
+pshort(snlx1) # or print(snlx1)
 tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
-# pshort(tnlx1)
-print(tnlx1)
+pshort(tnlx1) # alternatively print(tnlx1)
 ```
@@
 summary(unlm1) # Use summary() to get display
-unlm1 # Note the difference. Use this form to get sum of squares
+# unlm1 # Note the difference. Use this form to get sum of squares
 # pnlm0(unlm1)  # Short form of output
 snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
-# pnlm0(snlm1)
-summary(snlm1)
+pnlm0(snlm1)
+# summary(snlm1)
 ## Following structure needed to deal with M1mac issue
@@
 } else {  
-   pnlm0(tnlm1) # short form to give sum of squares
-   summary(tnlm1)
+   pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
 }   
@@
 summary(ugslnls1) # Use summary() to get display
+pnlm0(ugslnls1) # to get sum of squares
 sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
-summary(sgslnls1) # Use summary() to get display
+pnlm0(sgslnls1) # Use summary() to get display
 tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
-summary(tgslnls1) # Use summary() to get display
+pnlm0(tgslnls1) 
 ```
@@
 
-?? what do we want here??
+We illustrate how to solve nonlinear least squares problems using
+a function to define the residual. Note that `gsl_nls()` requires
+a vector `y` that is the expected value of what we have called the
+residual, but in this case is actually the model. `gsl_nls` uses 
+a numerical approximation for the Jacobian if the argument `jac` 
+is missing.
+
+```{r exhobbsfn, eval=TRUE}
+require(minpack.lm)
+require(nlsr)
+require(gslnls)
+hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
+print(hobnlfb)
+hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
+summary(hobnlm)  
+hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
+summary(hobgsln)
+hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12),jac=hobbs.jac)
+summary(hobgsl)
+```
 
@@
 also include such code. There is documentation in R
-of `selfStart`, but the construction of them is non-trivial. Their use with
+of `selfStart` models, but the construction of them is non-trivial. Their use with
 `nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
@@
 
--------------------------------------------------
-
-**nlsr** has functions to find a set of parameters that minimizes the sum of 
-squares (or deviance) 
-of differences between a model specified either as a formula using function 
-`nlxb()` or as both residual
-and Jacobian R functions using function `nlfb()`. As described below, the 
-Jacobian may be approximated
-by particular control settings.
-
-Existing R tools that are similar to `nlxb()` are `nls()` and `minpack.lm::nlsLM()`;  
-`minpack.lm::nls.lm()` is similar to `nlfb()`.
-
-There are several important differences in approach and results, some of which 
-are detailed in this document.
-
-### Stabilization of Gauss-Newton computations
-
-The iteration in all the major tools mentioned is the solution of the Gauss-Newton 
-equations. `nls()` uses a variant of an approach suggested by @Hartley61, while 
-`nlsr` and `minpack.lm` use variants of the method published by @Marquardt1963, 
-though it was suggested earlier by @Levenberg1944. The details of the method in 
-`nlsr` are discussed in  @JNABRefactoringNLS22.
-Though it is unlikely to be useful in general, control settings for `nlxb()` or `nlfb()`
-allow for those routines to perform a variety of Hartley and Marquardt algorithms. This
+## Bounds constraints on parameters
+
+In some cases, we know that parameters cannot be bigger or smaller than some
+externally known limits. Cash reserves cannot be negative. Animals have a 
+minimum need for water. Airplanes cannot carry more than a known or 
+legislated cargo weight. Such limits can be built into models, but there 
+are some important details for using the tools in R.
+
+- `nls()` can only impose bounds if the `algorithm="port"` argument is 
+used in the call. Unfortunately, the documentation warns us:
+
+  *The algorithm = "port" code appears unfinished, and does not even 
+   check that the starting value is within the bounds. Use with 
+   caution, especially where bounds are supplied.*
+
+- `gsl_nls()` does not offer bounds.
+
+- `nlsLM()` includes bounds in the standard call, but we have observed cases
+   where it fails to get the correct answer. From examination of the code,
+   we believe the authors have not taken into account all possibilities, though
+   it should be noted that all programs have some weakness in
+   regard to constrained optimization. Software creators have to work with
+   assumptions on the extremity of scale that they are willing to countenance,
+   and sometimes problems will be outside the scope envisaged.
+
+```{r ex10, echo=TRUE} 
+# Start MUST be feasible i.e. on or within bounds
+anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
+             upper=c(2,6,3), algorithm='port')
+pnls0(anlshob1b) #  check the answer (short form)
+# nlsLM seems NOT to work with bounds in this example
+anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pnls0(anlsLM1b)
+# also no warning if starting out of bounds, but gets good answer!!
+st4<-c(c1=4, c2=4, c3=4)
+anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pnls0(anlsLMob)
+# Try nlsr::nlxb()
+anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pshort(anlx1b)
+```
+
+Let us try to fix (mask) the first parameter in the first two example problems.
+
+```{r ex10m, echo=TRUE}
+# Hobbsmaskx.R -- masks with formula specification of problem
+require(nlsr)
+require(minpack.lm)
+stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
+sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
+traceval<-FALSE
+# fix first parameter
+anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+			upper=c(200, 60, .3), trace=traceval))
+print(anxbmsk1)
+anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+			upper=c(200, 60, .3), trace=traceval))
+summary(anlM1)
+# nls gives warnings
+anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+		upper=c(200, 60, .3),  algorithm="port", trace=traceval))
+summary(anlsmsk1)
+# Hobbs scaled problem with bounds, formula specification
+anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                  upper=c(2,6,3))
+print(anlxmsks1)
+anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
+             upper=c(2,6,3), algorithm='port')
+summary(anlshmsk1)
+cat("More precisely...crossprod(resid(anlshmsk1))=",crossprod(resid(anlshmsk1)),"\n")
+# nlsLM does not always work with bounds
+anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                 upper=c(2,6,3))
+summary(anlsLMmsks1)
+
+# Test with all parameters masked
+anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+print(anlxmskall)
+```
+
+`nlsr` has an output format that indicates the constraint status of the parameter
+estimates. For `nlsr` we have **chosen** to suppress calculation of 
+approximate standard 
+errors in the parameters when constraints are active, though we believe this 
+worthy of discussion and further investigation.
+
+## Stabilization of Gauss-Newton computations
+
+All four major mentioned must solve some variant of the Gauss-Newton 
+equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
+while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
+offers an accelerated Marquardt method and three alternative methods. We have 
+not explored all five methods with various control settings.
+Though it is unlikely to be useful in general, 
+control settings for `nlxb()` or `nlfb()` allow for those routines to perform 
+a variety of Hartley and Marquardt algorithms. This
 has been useful for learning about the algorithms, but tangential to simply finding
@@
 
-In general, the Levenberg-Marquardt stabilization is important in obtaining solutions.
-The default method of `nls()` terminates with `singular gradient` unnecessarily in 
-too many instances.
+In general, the Levenberg-Marquardt stabilization is important in obtaining 
+solutions in methods of the Gauss-Newton family, as `nls()` terminates 
+too frequently and unnecessarily with `singular gradient` errors.
 
@@
 programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
+As far as we are aware, gslnls
 Generally, I believe that keeping to a single programming language allows for 
@@
 The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
-<!-- ``` -->
-<!-- SSlogis <- selfStart(~ Asym/(1 + exp((xmid - input)/scal)), -->
-<!--     selfStart( -->
-<!--         function(input, Asym, xmid, scal) -->
-<!--         { -->
-<!--               .expr1 <- xmid - input -->
-<!--               .expr3 <- exp(.e2 <- .expr1/scal) -->
-<!--               .expr4 <- 1 + .expr3 -->
-<!--               .value <- Asym/.expr4 -->
-<!--               .actualArgs <- as.list(match.call()[c("Asym", "xmid", "scal")]) -->
-<!--               if(all(vapply(.actualArgs, is.name, NA))) -->
-<!--               { -->
-<!--             		  .expr10 <- .expr4^2 -->
-<!--                   .grad <- array(0, c(length(.value), 3L), list(NULL, c("Asym", "xmid", "scal"))) -->
-<!--                   .grad[, "Asym"] <- 1/.expr4 -->
-<!-- 		              .grad[, "xmid"] <- - (xm <- Asym * .expr3/scal/.expr10) -->
-<!-- 		              .grad[, "scal"] <- xm * .e2 -->
-<!--                   dimnames(.grad) <- list(NULL, .actualArgs) -->
-<!--                   attr(.value, "gradient") <- .grad -->
-<!--               } -->
-<!--               .value -->
-<!--         }, -->
-<!--         initial = function(mCall, data, LHS, ...) { -->
-<!--               xy <- sortedXyData(mCall[["input"]], LHS, data) -->
-<!--               if(nrow(xy) < 4) { -->
-<!--                   stop("too few distinct input values to fit a logistic model") -->
-<!--               } -->
-<!--               z <- xy[["y"]] -->
-<!--               ## transform to proportion, i.e. in (0,1) : -->
-<!--               rng <- range(z); dz <- diff(rng) -->
-<!--               z <- (z - rng[1L] + 0.05 * dz)/(1.1 * dz) -->
-<!--               xy[["z"]] <- log(z/(1 - z))		# logit transformation -->
-<!--               aux <- coef(lm(x ~ z, xy)) -->
-<!--               pars <- coef(nls(y ~ 1/(1 + exp((xmid - x)/scal)), -->
-<!--                                data = xy, -->
-<!--                                start = list(xmid = aux[[1L]], scal = aux[[2L]]), -->
-<!--                                algorithm = "plinear", ...)) -->
-<!--               setNames(pars [c(".lin", "xmid", "scal")], -->
-<!--                        mCall[c("Asym", "xmid", "scal")]) -->
-<!--         }, -->
-<!--         parameters = c("Asym", "xmid", "scal")) -->
-<!-- ``` -->
+
 This R function includes analytic expressions for the Jacobian ("gradient").
@@
                   -----------------------------------------------------------
-
-<!-- ### Example output -->
-
-<!-- To illustrate, let us run the Croucher example. -->
-
-<!-- ```{r nlsoutx} -->
-<!-- # Croucher example -->
-<!-- xdata <- c(-2,-1.64,-1.33,-0.7,0,0.45,1.2,1.64,2.32,2.9) -->
-<!-- ydata <- c(0.699369,0.700462,0.695354,1.03905,1.97389,2.41143,1.91091,0.919576,-0.730975,-1.42001) -->
-<!-- p1<- 1; p2<-0.2; NLSstart<-list(p1=p1,p2=p2) -->
-<!-- NLSformula <- ydata ~ p1*cos(p2*xdata) + p2*sin(p1*xdata) -->
-<!-- NLSdata<-data.frame(xdata, ydata) -->
-<!-- # Try full output version of nls -->
-<!-- library(nlspkg) # use the packaged version of nls() -->
-<!-- result<-nls(NLSformula, data=NLSdata, start=NLSstart, model=TRUE) -->
-<!-- # We can display the result with several commands -->
-<!-- # str(result) -- displays large amount of material - suppressed here -->
-<!-- #    as it is too wide for the page -->
-<!-- # result # displays the object -->
-<!-- # ls(result) # to list the elements of the output -->
-<!-- # ls(result$m) # and in particular the "m" object -->
-<!-- ``` -->
-
-<!-- ### Concerns with content of the nls result object -->
 
@@
 
-```{r chkbound}
-# bhobbsX.R ## bounded formula specification of problem using Hobbs Weed problem
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-wf <- data.frame(y = weed, tt = tt)
-st <- c(b1 = 1, b2 = 1, b3 = 1) # a default starting vector (named!)
-wmods <- y ~ 100 * b1 / (1 + 10 * b2 * exp(-0.1 * b3 * tt)) ## Scaled model
-require(minpack.lm)
-# Hobbs scaled problem with bounds, formula specification
-anlxb1 <- nlxb(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("Using nlsr::nlxb():")
-print(anlxb1)
-## nlsLM seems NOT to work properly with bounds
-anlsLM1b <- nlsLM(wmods, start = st, data = wf, lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("\n"); cat("using minpack.lm::nlsLM():")
-print(anlsLM1b)
-anlsb<-nls(wmods, start = st, data = wf, algorithm = "port", 
-           lower = c(0, 0, 0), upper = c(2, 6, 3))
-cat("Using nls with 'port':")
-print(anlsb)
-```
-
 ### Philosophical considerations
@@
 base index for arrays lead to traps for unwary programmers.
-
-<!-- For `nls()` the concern is demonstrated by a short email query from John Nash -->
-<!-- to Doug Bates and his reply. -->
-<!-- This is NOT a criticism of Prof. Bates' (or any other) work, but a reflection on  -->
-<!-- how difficult it is to develop code in this subject area and to keep it maintainable.  -->
-<!-- We have experienced similar loss of understanding for some of our own codes. -->
-<!--     ... -->
-<!--     https://gitlab.com/nashjc/improvenls/-/blob/master/Croucher-expandednlsnoc.R -->
-<!--     ... has the test problem and the expanded code. Around line 367 is where we are -->
-<!--     scratching our heads. The function code (from nlsModel()) is in the commented  -->
-<!--     lines below the call. This is -->
-<!--           # > setPars -->
-<!--           # function(newPars) { -->
-<!--           #   setPars(newPars) -->
-<!--           #   resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x} -->
-<!--           #   dev   <<- sum(resid^2) # envir = thisEnv -->
-<!--           #   if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr) -->
-<!--           #   QR <<- qr(.swts * gr) # envir = thisEnv -->
-<!--           #   (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix -->
-<!--           # } -->
-<!--     I'm anticipating that we will be able to set up a (possibly inefficient) code -->
-<!--     with documentation that will be easier to follow and test, then gradually figure -->
-<!--     out how to make it more efficient. -->
-<!--     The equivalent from minpack.lm is -->
-<!--     setPars = function(newPars) { -->
-<!--                 setPars(newPars) -->
-<!--                 assign("resid", .swts * (lhs - assign("rhs", getRHS(), -->
-<!--                     envir = thisEnv)), envir = thisEnv) -->
-<!--                 assign("dev", sum(resid^2), envir = thisEnv) -->
-<!--                 assign("QR", qr(.swts * attr(rhs, "gradient")), envir = thisEnv) -->
-<!--                 return(QR$rank < min(dim(QR$qr))) -->
-<!--             } -->
-<!--     In both there is the recursive call, which must have a purpose I don't understand. -->
-<!-- The reply was -->
-<!--     I'm afraid that I don't know the purpose of the recursive call either.   -->
-<!--     I know that I wrote the code to use a closure for the response, covariates, etc.,  -->
-<!--     but I don't recall anything like a recursive call being necessary. -->
-<!--      I don't think I will be of much help.  My R skills have atrophied to the point  --> 
-<!--      where I wouldn't even know how to start exploring what is happening in the first  -->
-<!--      call as opposed to the recursive call. --> 
-<!--  ``` -->
 
@@
 
-<!-- Arkajyoti -- do you want to expand? -->
-<!-- No. I think it would be tangential to the goal of the paper. We can have a -->
-<!-- separate article for that, maybe? -->
-<!-- Trying to trim a lot of words, so am making reference to package. Is it ready -->
-<!-- for others to use? -->
-
-<!-- When we have a "new" or trial solver function, we would like to know if it gives  -->
-<!-- acceptable results on a range of sample problems of different types, starting -->
-<!-- parameters, input conditions, constraints, subsets, weights or other settings.  -->
-<!-- Ideally we want to be able to get a summary that is easy to read and assess. For -->
-<!-- example, one approach would be to list the names of a set of tests with a red, green -->
-<!-- or yellow dot beside the name for FAILURE, SUCCESS, or "NOT APPLICABLE". In the last -->
-<!-- category would be a problem with constraints that the solver is not designed to  -->
-<!-- handle.  -->
-
-<!-- To accomplish this, we need a suitable "runner" program that can be supplied with -->
-<!-- the name of a solver or solvers and a list of test problem cases. Problems generally -->
-<!-- have a base setup -- a specification of the function to fit as a formula, some data -->
-<!-- and a default starting set of parameters. Other cases can be created by imposing -->
-<!-- bounds or mask constraints, subsets of the data, and different starts.  -->
-
-<!-- How to set up this "runner" and its supporting infrastructure is non-trivial. While -->
-<!-- the pieces are not as complicated as the inter-related parts of the solvers, especially -->
-<!-- `nls()`, the categorization of tests, their documentation, and the structuring to make -->
-<!-- running them straightforward and easy requires much attention to detail. -->
-
-<!-- Some considerations for our test scripts: -->
-
-<!-- - Is it useful to have a "base" script for each family of test problem, with numbered particular -->
-<!--   cases? That is, if we run the scripts in order, we can avoid some duplication of code and -->
-<!--   data. -->
-
-<!-- - While we have developed some tags to document the test problem families and cases, we believe -->
-<!--   that such tags (essentially summary documentation) will continue to need revision as different -->
-<!--   tools and problems are included in scope of `nlsCompare`. -->
-
-<!-- - Similarly, we expect that there will be ongoing review of the structure of the result files. -->
-
-<!-- # Progress so far -->
-
-<!-- Much of our work is reported in https://github.com/nashjc/RNonlinearLS . -->
-
-<!-- ## Reports and articles -->
-
-<!-- - The present article is the central report of the Google Summer or Code 2021 project -->
-<!--   "Improvements to nls()". -->
-
-<!-- - VarietyInNonlinearLeastSquaresCodes.Rmd: a review of the different algorithms and -->
-<!--   the many choices in their implementation for nonlinear least squares. This is still -->
-<!--   a DRAFT at 2022-6-20. -->
-
-<!-- Response to query about TestsDoc -->
-<!-- 4. Testing infrastructure, I think, is a general topic and not -->
-<!-- particularly relevant to nls(); i.e., it is important for all packages. -->
-<!-- So, I don't think we need to include that in RefactoringNLS. Could be a -->
-<!-- separate blog maybe?   -->
-
-<!-- Arkajyoti: I think TestsDoc is too preliminary to mention, though the links might be -->
-<!-- useful. Perhaps incorporate in the nlsCompare repo. -->
-<!-- - TestsDoc.Rmd (?citation) is a survey of testing tools in R. It has more  -->
-<!--   general possibilities  -->
-<!--   and fits into the subject of regression testing, in which case a  -->
-<!--   more extensive literature review will be needed.  Note that this -->
-<!--   document reflects the work in the the "Problem sets and test infrastructure" -->
-<!--   below. -->
-
-  <!-- need to incorporate MachID and -->
-  <!-- the many tests. Arkajyoti: should we think of working this up into a paper for -->
-  <!-- JSS or the R-Journal? My  -->
-  <!-- view is that this COULD be a long term side-interest for your academic work, but -->
-  <!-- that would depend on your own interests as well as opportunities. -->
-
-<!-- This needs more focus of where it may appear. I'm thinking of a blog post -- JN-->
-<!-- - DerivsNLS.Rmd ?citation: a document to explain different ways in which Jacobian information -->
-<!--   is supplied to nonlinear least squares computation in R. File `ExDerivs.R` is -->
-<!--   a DRAFT of a script to provide examples. -->
-
 ## Documentation and resources
@@
   in the creation of @NashWires22.
-
-<!-- ?? probably don't need to include -->
-<!-- - WorkingDocument4ImproveNLS.Rmd: essentially a way to record what we have worked on. -->
-<!--   A project diary. -->
-
-<!-- ?? does not seem to add enough? -->
-<!-- ## Problem sets and test infrastructure -->
-
-<!-- ?? need to clean up -->
-
-<!-- We have several test problems and variants thereof in the `inst/scripts/` directory of -->
-<!-- the `nlsCompare` package available on Github (https://github.com/ArkaB-DS/nlsCompare). -->
-<!-- We direct the reader to that package for documentation of the test infrastructure,  -->
-<!-- in particular the problems and methods files (`problems.csv` and `methods.csv`) and -->
-<!-- the various functions invoked by `run.R` to produce an output file in CSV form also. -->
-
-<!-- Towards the end of the project, we have focused our attention on the `nlsCompare` -->
-<!-- package, which looks at evaluating and comparing functions for nonlinear least -->
-<!-- squares problems. We use many of the same tests as checks that are or will be  -->
-<!-- built into our packages for such problems e.g., `nlsalt`. These use the `testthat` -->
-<!-- structure and may include verification of outputs that are specific to a package, -->
-<!-- such as the upper triangular matrix R of the QR decomposition that has been  -->
-<!-- computed for a Jacobian. Since such an object will not be computed by all methods,  -->
-<!-- testing it in `nlsCompare` makes no sense, and in that package we concentrate on -->
-<!-- the minimum sum of squares and the associated model parameters. -->
-
-<!-- ## Code development -->
-
-<!-- A summary of our main results: -->
-
-<!-- - nlspkg: a packaged version of the `nls() code from R-base. Thanks to Duncan Murdoch -->
-
-<!-- - nlsalt: attempt to mirror `nls() behaviour entirely in R. This is UNFINISHED.  -->
-<!--   The effort showed that the structure of the programs was difficult to follow,  -->
-<!--   undocumented, and unsuited to adding improvements like the Marquardt stabilization.  -->
-<!--   We were able to get a pure-R version of `numericDeriv()` and rework most of the functions  -->
-<!--   of `nlsModel` (but not `nlsModel.plinear`). This work may continue after the project -->
-<!--   formally ends, but collaboration is likely needed with workers who have a deep knowledge -->
-<!--   of R internals. `nls-flowchart.txt` was a start at documentation of the structure of `nls()`. -->
-
-<!-- <!-- - nls-changes-for-small-residuals-in-nls-R-4.0.2.zip: collected material for the  --> 
-<!-- <!--   fix by JN to the relative offset convergence criterion failure when there are small  --> 
-<!-- <!--   residuals in problems sent to ``nls()`. Now part of R. -->
-
-<!-- - the safeguarded relative offset convergence criterion that allows small-residual problems -->
-<!--   to ba handled with `nls()` is now in the R base code. -->
-
-<!-- - `nlsj` is a refactoring of some `nls()` functionality. ?? What are key features?? -->
-<!--   This is an interim package for exploration of ideas and will NOT, as it currently -->
-<!--   is presented, become a distributed package. -->
-
-<!-- - nlsralt: this is intended to be a modified version of Nash and Murdoch package  -->
-<!--   `nlsr` to examine possible improvements discovered as a result of this project.  -->
-<!--   This is EXPERIMENTAL. -->
-
-<!-- For example,  -->
-<!-- `nlsr` does not currently use the `subset` argument correctly. We also need to document -->
-<!-- changes e.g., to numerical derivatives. However, these are tasks for JN rather than AB. -->
 

### 2023-03-06 RefactoringNLS/ComparingNLStools.Rmd
More edits of new version

@@
 
+## The `nls()` function
+
+`nls()` is the tool in base R, the primary software software distribution
+from the Comprehensive R Archive Network (*https://cran.r-project.org*),
+for estimating nonlinear statistical models. The function dates to the 1980s and the 
+work related to @bateswatts in S (see 
+*https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+
+The `nls()` function has a remarkable and comprehensive set of capabilities for 
+estimating nonlinear models that are expressed as **formulas**. In particular, we
+note that it
+
+- handles formulas that include R functions, even ones which call calculations in
+  other programming languages
+- allows data to be weighted or subset
+- can estimate bound constrained parameters
+- provides a mechanism for handling partially linear models
+- permits parameters to be indexed over a set of related data
+- produces measures of variability (i.e., standard error estimates) for the 
+estimated parameters
+- has related profiling capabilities for exploring the likelihood surface as 
+parameters are changed
+- links to a number of pre-coded (`selfStart`) models.
+
+With such a range of features and long history, the code has become untidy 
+and overly patched, difficult to maintain, and its underlying methods could
+be improved. Various workers have developed packages to overcome these 
+concerns, and we will address some of these here.
+
 ## Scope of our comparison
@@
 ```
+
+One of the useful features of `nls()` is the possibility of a `selfStart` model,
+where starting parameter values are not required. We discuss aspects of this
+feature at several points below. However, in the event that a `selfStart` model 
+is not available, `nls()` sets all the starting parameters 
+to 1. This is, in our view, tolerable, but could be improved by using a set of values
+that are all slightly different, which, in the case of the example 
+model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
+would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
+Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
+parameters are fairly obvious. However, users may want to avoid equal parameter
+starts in situations like this when setting starting values.
 
@@
 
+Here we see the infamous "singular gradient" termination message of `nls()`. 
+Questions appear quite frequently on the R-help mailing list where users
+have encountered this termination result.
+
 Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
-and `gslnls::gsl_nls()`, which we will call the **class nls** methods.
+and `gslnls::gsl_nls()`.
 We use default methods and controls in these examples.
@@
 ```
+
+Though we have found solutions, the Jacobian is essentially singular as 
+shown by its singular
+values. Note that these are **displayed** by package `nlsr` in a single column in
+the output to provide a compact layout, but the values do **NOT** correspond to 
+the individual parameters in 
+whose row they appear; they are a property of the whole problem.
 
@@
 
+### When to compute ancillary information
+
+Tools that produce a class `nls` output object create a rich set of functions
+and structures that are then used in a variety of modeling tasks. By contrast,
+`nlsr` computes quantities as they are requested or needed, with additional
+features in separate functions. These two approaches
+lead to different consequences for performance and how features are
+provided. `nlsr` has antecendants in the methods of @cnm79, where storage for
+data and programs was at a ridiculous premium in the small computers of the era.
+Thus the code in `nlsr` is likely of value for workers to copy and modify it
+for customized tools.
+
 ## Jacobian calculation
@@
 
+### Jacobian code in selfStart models
+
 Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
@@
 also include such code. There is documentation in R
-of `selfStart` models, but the construction of them is non-trivial. Their use with
+of `selfStart` models, but the construction of them is non-trivial. A number 
+of such models are included with base R in `./src/library/stats/R/zzModels.R` file 
+with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
+These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
+that defines
+each model, and these Jacobians may be the analytic forms.
+
+The `nls()` function, after
+computing the "right hand side" or `rhs` of the residual, then checks to see if the
+"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a 
+Jacobian
+into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
+functions. The use of analytic Jacobians
+almost certainly contributes to the good performance of `nls()` on `selfStart` 
+models.
+
+The use of `selfStart` models with
 `nlsr` is described in the "Introduction to nlsr" vignette. However, since `nlsr`
-generally can use very crude starting values, we have rarely needed them.
+generally can use very crude starting values, we have rarely needed them, though
+it should be pointed out that our work is primarily diagnostic. If we were carrying
+out a large number of similar estimations, then good initial parameters are 
+critical to efficiency.
+
+In considering `selfStart` models, we noted that the base-R function `SSlogis`
+is intended to solve our example problem **Logistic3T**, but found initial 
+parameters for `nls()` by using a completely separate solver, then rerunning the
+solution with the default algorithm from the derived -- and final -- parameters.
+To provide simpler estimates of starting parameters, the function `SSlogisJN`
+is now part of package `nlsr`.
 
@@
 ```
+
+### Fixed parameters (masks)
 
@@
 errors in the parameters when constraints are active, though we believe this 
-worthy of discussion and further investigation.
+policy worthy of discussion and further investigation.
 
@@
 programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
-As far as we are aware, gslnls
-Generally, I believe that keeping to a single programming language allows for 
-easier maintenance and upgrades. On the other hand, R is usually somewhat slower 
-in performing computations because it keeps track of names and structures and 
+`gslnls` is an R wrapper to various C-language routines in the Gnu Scientific 
+Library (@GSL-manual).
+Generally, we believe that keeping to a single programming language can allow for 
+easier maintenance and upgrades. It also avoids some work when there are changes 
+or upgrades to libraries for the non-R languages. 
+On the other hand, R is usually somewhat slower 
+than some other computing systems because it keeps track of names and structures and 
 because it is usually interpreted rather than compiled. In recent years the 
@@
 package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
-less of a performance cost than the attempt to ensure the solution has been found, 
-which forces more iterations to be used.
+less of a performance cost than a policy of aggressively seeking a solution,
+which can cause more iterations to be used.
+
+## Data sources to problems
+
+`nls()` can be called without specifying the `data` argument. In this case, it will
+search in the available environments (i.e., workspaces) for suitable data objects. 
+We do NOT like this approach, but it is "the R way". R allows users to leave many 
+objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
+suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
+the path, 
+when R is launched, will be loaded. The overwhelming proportion of R users in our
+acquaintance avoid saving the workspace because of the danger of lurking data and
+functions which may cause unwanted results.
+
+Nevertheless, to provide compatible behaviour with `nls()` and R generally, 
+programs need to ensure equivalent behaviour, but users should test that the 
+operation is as they intend.
+
+### Feature: Subsetting
+
+`nls()` and other class `nls` tools accept an argument `subset`.
+This acts through the mediation of `model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`.
+
+Having `subset` at the level of the call to a function like
+`nls()` saves effort, but it does mean that the programmer of the 
+solver needs to be aware of the
+source (and value) of objects such as the data, residuals and Jacobian. 
+This is overly complicated. By preference, 
+we would implement subsetting by means of zero-value weights, with observation counts
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
+
+### Feature: na.action
+
+`na.action` is an argument to the `nls()` function, but it does not appear obviously in the source code, often being handled behind the scenes after referencing the option `na.action`.
+Yet this feature also changes the data supplied to our nonlinear
+least squares solver.
+
+A useful, but possibly dated, description is given in:
+*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
+The typical default action, which can be seen by using the command
+`getOption("na.action")`
+is `na.omit`. This option essentially omits from computations any observations
+containing missing values (i.e. any row of a data frame containing an NA). 
+`na.exclude` does much of the same for computations, but keeps the rows with NA 
+elements so that predictions are in the correct row position. We recommend that 
+workers actually test output to verify the behaviour is as wanted.
+See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
+
+As with `subset`, our concern with `na.action` is that users may be unaware of the
+effects of an option they may not even be aware has been set. Should `na.fail` be the default?
+
+## Feature: model frame
+
+`model` is an argument to the `nls()` and related functions, which is documented as:
+
+> **model** logical. If true, the model frame is returned as part of the object. Default is FALSE.
+
+Indeed, the argument only gets used when `nls()` is about to return its result
+object, and the
+element `model` is NULL unless the calling argument `model` is TRUE. (Using the same 
+name could
+be confusing.) However, the model frame is used within the function code in the form 
+of the object
+`mf`. We feel that users could benefit from more extensive documentation and 
+examples of its use since it is used to implement features like `subset`.
+
+## Weights on observations
+
+All four main tools we consider here allow a `weights` argument that
+specifies a vector of fixed weights the same length as the number of residuals. 
+Each residual is multiplied by the square root of the corresponding weight. 
+Where available, the values returned by the
+`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
+used to compute raw residuals.
+
+While fixed weights may be useful, there are a number of problems for which
+we want weights that are determined at least partially from the model 
+parameters, for
+example, a measure of the standard deviation of observations.
+
+Such dynamic weighting situations are discussed in the vignette 
+"Introduction to nlsr" of package `nlsr` in section 
+*Weights that are functions of the model parameters*. Care is advised
+in applying such ideas.
 
 ---OLD VERSION BELOW ------------------------
-
-# The `nls()` function: strengths and shortcomings
-
-`nls()` is the tool in base R, the primary software software distribution
-from the Comprehensive R Archive Network (*https://cran.r-project.org*),
-for estimating nonlinear statistical models. The function dates to the 1980s and the 
-work related to @bateswatts in S (see 
-*https://en.wikipedia.org/wiki/S_%28programming_language%29*).
-
-The `nls()` function has a remarkable and comprehensive set of capabilities for 
-estimating nonlinear models that are expressed as **formulas**. In particular, we
-note that it
-
-- handles formulas that include R functions, even ones which call calculations in
-  other programming languages
-- allows data to be weighted or subset
-- can estimate bound constrained parameters
-- provides a mechanism for handling partially linear models
-- permits parameters to be indexed over a set of related data
-- produces measures of variability (i.e., standard error estimates) for the estimated parameters
-- has related profiling capabilities for exploring the likelihood surface as parameters are changed
-- links to a number of pre-coded (`selfStart`) models
-
-With such a range of features and long history, the code has become untidy 
-and overly patched, difficult to maintain, and its underlying methods could
-be improved. Various workers have developed packages to overcome these 
-concerns, and we will address some of these here.
-
-## Feature: Failure when Jacobian is computationally singular
-
-This refers to the infamous "singular gradient" termination message of `nls()`. 
-A Google search of
-
-```
-R nls "singular gradient"
-```
-
-gets over 4000 hits that are spread over some years. This could be because the 
-Jacobian is poorly approximated (see **Jacobian computation** below). However,
-it is common in nonlinear least squares computations that the Jacobian is very close
-to singular for some values of the model parameters. In such cases, we need to find
-an alternative algorithm to the Gauss-Newton iteration of `nls()`. The most common
-work-around is the Levenberg-Marquardt stabilization (see @Marquardt1963, @Levenberg1944, 
-@jn77ima), and versions of it have been implemented in packages `minpack.lm` and `nlsr`.
-In our work, we prepared experimental prototypes of `nls` that incorporate
-stabilization, but integration with all the features of `nls()` has proved difficult.
-
-Let us consider a 'singular gradient' example 
-*https://stats.stackexchange.com/questions/13053/singular-gradient-error-in-nls-with-correct-starting-values*.
-`nlsr::nlxb()` displays the singular values of the Jacobian, which confirm that the `nls()` error message is correct,
-but this is not helpful in obtaining a solution.
-
-```{r SESingGrad}
-reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
-x <- 1:11 # x values; 11 data points
-y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
-testdat <- data.frame(x, y) # save in a data frame
-strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
-jform <- y ~ a + b * r^(x - m) + c * x # Formula
-library(nlsr)
-linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
-linexp2 # Note singular values of Jacobian in rightmost column
-# We get an error with nls()
-linexp <- try(nls(jform, data = testdat, start = strt, trace = FALSE))
-```
-
-The Jacobian is essentially singular as shown by its singular
-values. Note that these are **displayed** by package `nlsr` in a single column in
-the output to provide a compact layout, but the values do **NOT** correspond to 
-the individual parameters in 
-whose row they appear; they are a property of the whole problem.
 
@@
   taken as an advantage.
-
-Note that the `selfStart` models in the base R `./src/library/stats/R/zzModels.R` file 
-and in package `nlraa` (@MiguezNLRAA2021) may
-provide the Jacobian in the "gradient" attribute of the "one-sided" formula that defines
-each model, and these Jacobians may be the analytic forms. The `nls()` function, after
-computing the "right hand side" or `rhs` of the residual, then checks to see if the
-"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a Jacobian
-into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
-functions. The use of analytic Jacobians
-almost certainly contributes to the good performance of `nls()` on `selfStart` models.
-
-## Feature: Weights on observations
-
-`nls()`, `nlsr::nlxb()` and `minpack.lm::nlsLM()` all have a `weights` argument that
-specifies a vector of weights the same length as the number of residuals. Each residual
-is multiplied by the square root of the corresponding weight. The values returned by the
-`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
-used to compute raw residuals.
-
-## Feature: Subsetting
-
-`nls()` accepts an argument `subset`. This acts through the mediation of
-`model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and 
-`/src/library/stats/src/nls.C`.
-
-While the implementation of subset at the level of the call to `nls()` has a certain
-attractiveness, it does mean that the programmer of the solver needs to be aware of the
-source (and value) of objects such as the data, residuals and Jacobian. This is overly
-complicated. By preference, 
-we would implement subsetting by means of zero-value weights, with observation counts
-(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
-we would extract a working dataframe from the relevant elements in the original.
-
-## Feature: na.action
-
-`na.action` is an argument to the `nls()` function, but it does not appear obviously in the
-source code, often being handled behind the scenes after referencing the option `na.action`.
-A useful, but possibly dated, description is given in:
-*https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/*.
-
-The typical default action, which can be seen by using the command `getOption("na.action")`
-is `na.omit`. This option essentially omits from computations any observations
-containing missing values (i.e. any row of a data frame containing an NA). 
-`na.exclude` does much of the same for computations, but keeps the rows with NA elements so 
-that predictions are in the correct row position. We recommend that workers actually test 
-output to verify the behaviour is as wanted.
-
-A succinct description of the issue is given in: *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*
-where the "Answer" states
-
-> The only benefit of `na.exclude` over `na.omit` is that the former will retain the 
-original number of rows in the data. This may be useful where you need to retain the 
-original size of the dataset - for example it is useful when you want to compare 
-predicted values to original values. With `na.omit` you will end up with fewer 
-rows so you won't as easily be able to compare.
-
-`na.pass` simply passes on data "as is", while `na.fail` will essentially stop if any missing
-values are present.
-
-Our concern with `na.action` is that users may be unaware of the effects of an option
-setting they may not even be aware has been set. Should `na.fail` be the default?
-
-## Feature: model frame
-
-`model` is an argument to the `nls()` function, which is documented as:
-
-> **model** logical. If true, the model frame is returned as part of the object. Default is FALSE.
-
-Indeed, the argument only gets used when `nls()` is about to return its result object, and the
-element `model` is NULL unless the calling argument `model` is TRUE. (Using the same name could
-be confusing.) However, the model frame is used within the function code in the form of the object
-`mf`. We feel that users could benefit from more extensive documentation and examples of its use.
-
-## Feature: Sources of data
-
-`nls()` can be called without specifying the `data` argument. In this case, it will
-search in the available environments (i.e., workspaces) for suitable data objects. 
-We do NOT like this approach, but it is "the R way". R allows users to leave many 
-objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit; otherwise, any such file in the path, 
-when R is launched, will be loaded. The overwhelming proportion of R users in our
-acquaintance avoid saving the workspace because of the danger of lurking data and
-functions which may cause unwanted results.
-
-Nevertheless, to provide compatible behaviour with `nls()`, competing programs 
-need to ensure equivalent behaviour, but users should test that the 
-operation is as they intend.
 
@@
 command, making them difficult to access.
-
-In the event that a `selfStart` model is not available, `nls()` sets all the starting parameters 
-to 1. This is, in our view, tolerable, but could be improved by using a set of values
-that are all slightly different, which, in the case of the example model $y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)$
-would avoid a singular Jacobian because $b$ and $d$ were equal in value. 
-A sequence like  1.0, 1.1, 1.2, 1.3 for the four parameters 
-could be provided quite simply instead of all 1's.
 

### 2023-03-06 RefactoringNLS/ComparingNLStools.Rmd
more edits

@@
 parameters are changed
-- links to a number of pre-coded (`selfStart`) models.
+- links to a number of pre-coded (`selfStart`) models that do not require
+  initial parameter values.
 
@@
 
+### Returned results of nls() and other tools
+
+The output of `nls()` is an object of class "nls" which has the following structure:
+
+>A list of
+
+>`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.
+
+>`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.
+
+>`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.
+
+>`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.
+
+>`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.
+
+>`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.
+
+>`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.
+
+>`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.
+
+>`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.
+
+>`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
+$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.
+
+> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.
+
+                  -----------------------------------------------------------
+
+The complexity of this object is a challenge to users. Let us use 
+`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
+`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
+actually access the data from this element, we need to use the syntax:
+
+```
+eval(parse(text=result$data))
+```
+
+However, if the call is made with `model=TRUE`, then there is a returned element
+`model` which contains the data, and we can list its contents using:
+
+```
+ls(result$model)
+```
+
+and if there is an element called `xdata` that can be accessed as
+`result$model$xdata`.
+
+By contrast `nlsr::nlxb()` returns
+
+>`coefficients`    A named vector giving the parameter values at the supposed solution.
+
+>`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.
+
+>`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.
+
+>`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
+                parameters) at the \newline
+$~~~~~~~~~~~~~~~~~~~~$ returned parameters.
+
+>`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.
+
+>`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.
+
+However, actually looking at the structure of a returned result gives a list of 11
+items. The extra 5 are:
+
+```
+ $ lower       : num [1:3] -Inf -Inf -Inf
+ $ upper       : num [1:3] Inf Inf Inf
+ $ maskidx     : int(0) 
+ $ weights     : NULL
+ $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
+  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
+ - attr(*, "class")= chr "nlsr"
+```
+The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` 
+returns. Moreover, `nlxb`
+explicitly returns the sum of squares as well as the residual vector and Jacobian. 
+The counts of evaluations are also returned.
+
 ### When to compute ancillary information
@@
 `nlsr` computes quantities as they are requested or needed, with additional
-features in separate functions. These two approaches
+features in separate functions. For example, the singular values of the Jacobian 
+are actually computed in the `print` and `summary` methods for the result. 
+These two approaches
 lead to different consequences for performance and how features are
@@
 expressions using symbolic and automatic differentiation tools. 
-When using formula specification of the model, `nls()`, `minpack.lm::nlsLM()`
+When using a formula specification of the model, `nls()`, `minpack.lm::nlsLM()`
 and `gslnls::gsl_nls()` use a finite difference approximation to compute the
 Jacobian, though `gsl_nls()` does have an option to attempt symbolic expressions.
-
-When The Jacobian code as 
-the "gradient" attribute of the output of the Jacobian function. This 
-lets us embed the code 
-for the Jacobian as this attribute of the **residual** function so that the call 
-to `nlfb()` can be made with the same name used for 
+Package `nlsr` provides, via appropriate calling syntax, four numeric approximation 
+options for the Jacobian, with a further control `ndstep` for the size of the
+step used in the approximation.
+
+Using the "gradient" attribute of the output of the Jacobian function to hold
+this matrix lets us embed this attribute of the **residual** function as well,
+so that the call to `nlfb()` can be made with the same name used for 
 both residual and Jacobian function arguments. This programming trick saves a 
@@
 
-Generally `nls()`, `nlsLM()` and `gsl_nls()` use a fairly 
-simple forward difference approximation of derivatives for the Jacobian
-from the base-R `stats` package called `numericDeriv()`. A central 
-approximation can be specified in control parameters. 
-Package `nlsr` provides, via appropriate calling syntax, four approximation 
-options, with a further control `ndstep` for the size of the
-step used in the approximation.
+As far as we can understand the logic in `nls()`, the Jacobian computation during
+parameter estimation is carried out entirely within the called C-language program,
+and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
+in the R distribution source code. This is used to provide Jacobian information in 
+the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
+general use. `gsl_nls()` also appears to use `numericDeriv()`.
+
+`numericDeriv()` uses a 
+simple forward difference approximation of derivatives, though a central 
+difference approximation can be specified in control parameters. 
+We are unclear why `numericDeriv()` in base R calls `C_numeric_deriv`,
+as we were easily able to create a more compact version entirely in R. 
+See *https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS*.
+
+`minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
+version of `nlsModel()`, but it appears to use an internal approximate Jacobian 
+code from 
+the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
+can lead to different behaviour, though in our experience the differences are usually
+very minor.
+
+Minor numerical differences can give annoying results with ill-conditioned
+problems.
+
+- A pasture regrowth problem (@Huet2004, page 1, based on @Ratkowsky1983) has a
+  poorly conditioned Jacobian and `nls()` fails with "singular gradient". 
+  Worse, numerical approximation to the Jacobian gives the error
+  "singular gradient matrix at initial parameter estimates" for `minpack.lm::nlsLM`
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of `nlsr::nlxb` give a solution.
+
+- Karl Schilling (private communication) provided an example where a model specified
+  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
+  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
+  the derivative is undefined, while the model can be computed. In such cases, 
+  we observed that `nls()` and 
+  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.
 
@@
 with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
+There are also some in the now-archived package `NRAIA`.
 These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
@@
 In considering `selfStart` models, we noted that the base-R function `SSlogis`
-is intended to solve our example problem **Logistic3T**, but found initial 
-parameters for `nls()` by using a completely separate solver, then rerunning the
+is intended to solve our example problem **Logistic3T**, but that it finds initial 
+parameters for `nls()` by using a completely separate solver, then reruns the
 solution with the default algorithm from the derived -- and final -- parameters.
@@
 
+The `SSLogis` case is one where the `getInitial()` function (used to find
+starting values) actually calls `nls()` with a non-default algorithm. 
+Such circular references are, in our view, prone to creating errors
+in code maintenance.
+Users may also want to use these models with explicit starting values 
+other than those suggested by `getInitial()`.
+We are also surprised that the 
+analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
+saves many quantities in "hidden"
+variables, i.e., with names preceded by ".". 
+These are then not displayed by the `ls()`
+command, making them difficult to access by users who may wish to create
+their own selfStart model by a copy and edit.
+Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
+worth considering as a way to find starting parameters, but we know of no
+R capability of this type.
+
+As a side note, the introduction of `scaleOffset` in R 4.1.1 to deal with the 
+convergence test for small residual problems now requires that the `getInitial()`
+function have dot-arguments (`...`) in its argument list. This illustrates the
+entanglement of many features in `nls()` that complicate its maintenance and
+improvement.
+
 ## Bounds constraints on parameters
+
+?? Arkajyoti -- we can probably trim these examples
 
@@
 - `gsl_nls()` does not offer bounds.
+
+- bounds are part of the default method for package `nlsr`.
 
@@
 
+### Philosophical considerations
+
+Bounds on parameters raise some interesting and difficult questions about how 
+uncertainty in parameter estimates should be computed or reported. That is, the
+traditional "standard errors" are generally taken to imply symmetric intervals about
+the point estimate in which the parameter may be expected to be found with some
+probability under certain assumptions.
+Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
+errors nor their derived statistics when bounds are active.
+
 ### Fixed parameters (masks)
@@
 anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, .3), trace=traceval))
+			upper=c(200, 60, 3), trace=traceval))
 print(anxbmsk1)
 anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, .3), trace=traceval))
+			upper=c(200, 60, 3), trace=traceval))
 summary(anlM1)
@@
 anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-		upper=c(200, 60, .3),  algorithm="port", trace=traceval))
+		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
 summary(anlsmsk1)
@@
 anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                  upper=c(2,6,3))
+                  upper=c(2,6,30))
 print(anlxmsks1)
@@
 approximate standard 
-errors in the parameters when constraints are active, though we believe this 
+errors in the parameters when constraints are active because their meaning
+under constraints is unclear, though we believe this 
 policy worthy of discussion and further investigation.
@@
 
-All four major mentioned must solve some variant of the Gauss-Newton 
+All four major tools illustrated must solve some variant of the Gauss-Newton 
 equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
@@
 example, a measure of the standard deviation of observations.
-
 Such dynamic weighting situations are discussed in the vignette 
@@
 
----OLD VERSION BELOW ------------------------
-
-## Feature: Jacobian computation
-
-`nls()`, with the `numericDeriv()` function, computes the Jacobian as the "gradient"
-attribute of the residual vector. This is implemented as a mix of R and C code, but
-we have created a rather more compact version entirely in R. 
-See *https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS*.
-<!-- Code project. See the companion document **Jacobian Calculations for nls()** -->
-<!-- (?? need citation -- currently **DerivsNLS.pdf**). -->
-
-As far as we can understand the logic in `nls()`, the Jacobian computation during
-parameter estimation is carried out entirely within the called C-language program,
-and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
-in the R distribution source code. This is used to provide Jacobian information in 
-the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
-general use.
-
-The Jacobian used by `nlsr::nlxb()` is, by default, computed from analytic expressions, 
-while that
-of `nls()` mostly uses a numerical approximation. Some `selfStart` models do provide
-"gradient" information. `minpack.lm::nlsLM()` invokes `numericDeriv()` in its local 
-version of `nlsModel()`, but it appears to use an internal approximate Jacobian code from 
-the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, though in our experience differences are usually minor.
-
-- A pasture regrowth problem (@Huet2004, page 1, based on @Ratkowsky1983) has a poorly
-  conditioned Jacobian and `nls()` fails with "singular gradient". 
-  Worse, numerical approximation to the Jacobian give the error
-  "singular gradient matrix at initial parameter estimates" for `minpack.lm::nlsLM`
-  so that the Marquardt stabilization is unable to take effect, while the analytic
-  derivatives of `nlsr::nlxb` give a solution.
-
-- Karl Schilling (private communication) provided an example where a model specified
-  with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
-  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
-  this is undefined. In such cases, we observed that `nls()` and `minpack.lm::nlsLM`
-  found a solution, though it can be debated whether such lucky accidents can be
-  taken as an advantage.
-
-## Feature: missing start vector and self-starting models
-
-Nonlinear estimation algorithms are almost all iterative and need a set of starting
-parameters. `nls()` offers a special class of modeling functions called **`selfStart`** 
-models. There are a number of these in base R (`./src/library/stats/R/zzModels.R`)
-and others in R packages such as CRAN package `nlraa` (@MiguezNLRAA2021), as well as 
-the now-archived package `NRAIA`. Unfortunately, some **`selfStart`** codes
-entangle the calculation of starting values for parameters
-with the particulars of the `nls()` code. Though there is a `getInitial()` function,
-this is not easy to use to simply compute the initial parameter estimates outside of
-`nls()`, in part 
-because it may call that function. Such circular references are, in our view, dangerous.
-Moreover, we believe that it would be helpful to have `selfStart` models that allow
-users to explicitly provide values other than those suggested for the starting parameters.
-
-Other concerns are revealed by the example below. Here, the `SSlogis` `selfStart` function 
-is used to generate a set
-of initial parameters for a 3-parameter logistic curve. The form used by `SSlogis`
-is $y \,\sim\, Asym\,/\, (1 \,+\, exp((xmid\,-\,tt)\,/\,scal))$, but we show how the starting parameters
-for this model can be transformed to those of 
-another form of the model, namely, $y \,\sim\, b1\,/\,(1 \,+\, b2\,*\,exp(-b3\,*\,t))$.
-
-The code for `SSlogis()` is in `./src/library/stats/R/zzModels.R`.
-
-This R function includes analytic expressions for the Jacobian ("gradient").
-These could be useful to R users, especially if documented. Moreover, we
-wonder why the programmers have chosen to save so many quantities in "hidden"
-variables, i.e., with names preceded by ".". These are then not displayed by the `ls()`
-command, making them difficult to access.
-
-### Strategic issues in `selfStart` models
-
-By providing starting values that are likely to be reasonable and by including 
-Jacobian code in the `selfStart` model code, some of the deficiencies of the 
-Gauss-Newton algorithm are mitigated. However, creating such functions is a lot 
-of work, and their
-documentation (file `./src/library/stats/man/selfStart.Rd`) is quite complicated.
-We believe that the focus should be placed on getting good initial parameters,
-that is `getInitial()` function, though avoiding the current calls back to `nls()`.
-Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
-worth considering.
-
-We also note that the introduction of `scaleOffset` in R 4.1.1 to deal with the 
-convergence test for small residual problems now requires that the `getInitial()`
-function have dot-arguments (`...`) in its argument list. This illustrates the
-entanglement of many features in `nls()`.
-
-## Issue: returned output of nls() and its documentation
-
-The output of `nls()` is an object of class "nls" which has the following structure:
-
->A list of
-
->`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.
-
->`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.
-
->`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.
-
->`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.
-
->`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.
-
->`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.
-
->`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.
-
->`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.
-
->`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.
-
->`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.
-
-> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.
-
-                  -----------------------------------------------------------
-
-The `nls` object contains some elements that are awkward to produce by other algorithms,
-but some information that would be useful is not presented in a clear manner. Moreover, the
-complexity of the object is a challenge to users.
-
-In the following, we use `result` as the returned object from `nls()`.
-
-The `data` return element is an R symbol. To actually access the data from this 
-element, we need to use the syntax:
-
-```
-eval(parse(text=result$data))
-```
-
-However, if the call is made with `model=TRUE`, then there is a returned element
-`model` which contains the data, and we can list its contents using:
-
-```
-ls(result$model)
-```
-
-and if there is an element called `xdata`, it can be accessed as `result$model$xdata`.
-
-Let us compare the `nls()` result with that from 
-`nlsr::nlxb()`, which ostensibly solves the same problem:
-
->`coefficients`    A named vector giving the parameter values at the supposed solution.
-
->`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.
-
->`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.
-
->`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
-                parameters) at the \newline
-$~~~~~~~~~~~~~~~~~~~~$ returned parameters.
-
->`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.
-
->`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.
-
-However, actually looking at the structure of a returned result gives a list of 11
-items. The extra 5 are:
-
-```
- $ lower       : num [1:3] -Inf -Inf -Inf
- $ upper       : num [1:3] Inf Inf Inf
- $ maskidx     : int(0) 
- $ weights     : NULL
- $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
-  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
- - attr(*, "class")= chr "nlsr"
-```
-The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` returns. 
-Moreover, `nlxb`
-explicitly returns the sum of squares as well as the residual vector and Jacobian. 
-The counts of evaluations are also returned.
-(Note that the singular values of the Jacobian are
-actually computed in the `print` and `summary` methods for the result.) 
-As we made our comparisons, we noted several potential updates to the `nlsr` documentation
-as well as that for `nls()`.
-
 ### Weights in returned functions from nls()
 
-As already noted, the function `resid()` (an alias for `residuals()`) is WEIGHTED, as are those in the
-structured object `m` returned by `nls()` or `minpack.lm::nlsLM`,
-e.g., `ans$m$resid()`. The function `nlsModel()`, which we have had to extract from the 
-base R code and explicitly `source()` because it is not exported to the working namespace,
+The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
+as does, for example, `result$m$resid()`. The function `nlsModel()`, which we 
+have had to extract from the base R code and explicitly `source()` because it 
+is not exported to the working namespace,
 allows us to compute residuals for particular coefficient sets.
@@
 
+## Minor complaints
+
 ### Interim output from the "port" algorithm
 
-As the `nls()` **man** page states, when the "port" algorithm is used with the `trace` argument
-TRUE, the iterations display the objective function value which is 1/2 the sum of squares (or 
-deviance). It is likely that the trace display is embedded in the Fortran of the `nlminb`
-routine that is called to execute the "port" algorithm, but the factor of 2 discrepancy is nonetheless
+As the `nls()` **man** page states, when the "port" algorithm is used with the
+`trace` argument TRUE, the iterations display the objective function value which 
+is 1/2 the sum of squares (or deviance). It is likely that the trace display is
+embedded in the Fortran of the `nlminb` routine that is called to execute 
+the "port" algorithm, but the factor of 2 discrepancy is nonetheless
 unfortunate for users.
@@
 
-If `nls()` reaches a point where it cannot continue but has not found a point where the
-relative offset convergence criterion is met, it may simply exit, especially if a 
-"singular gradient" (singular Jacobian) is found. However, this may occur AFTER the 
-function has made considerable progress in reducing the sum of squared residuals. 
+If `nls()` reaches a point where it cannot continue but has not found a point 
+where the relative offset convergence criterion is met, it may simply exit,
+especially if a "singular gradient" (singular Jacobian) is found. However, 
+this may occur AFTER the function has made considerable progress in reducing 
+the sum of squared residuals. 
 <!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
@@
 
-## Feature: partially linear models and their specification
-
-The variable projection method (@Golub1973, @OlearyRust13) is generally much more 
-effective than general approaches in finding good solutions to nonlinear least squares 
+## Estimating models that are partially linear
+
+Many nonlinear models can be expressed so that some of the parameters 
+appear linearly. In our logistic examples, the asymptote parameters are
+an illustration. In such cases, specialized methods can give spectactularly
+better performance. The variable projection method (@Golub1973, @OlearyRust13) is 
+generally much more effective than general approaches in finding good solutions 
+to nonlinear least squares 
 problems when some of the parameters appear linearly. However, setting up the 
@@
 
-Moreover, specifying a model to a solver should, ideally, use the same syntax across
-solver tools. Unfortunately, R allows multiple approaches within different
-modeling tools, and within `nls()` itself. This is highlighted by the case of
-partially linear models.
-
-The nonlinear modeling specifications are, of course, a development of linear ones.
-Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear modeling 
-function `lm()`, which requires this model to be specified as `y ~ x`.
-
-Within `nls()`, consider the following FOUR different specifications for the same problem, plus 
-an intuitive choice, labelled `fm2a`, that does NOT work. In this failed
-attempt, putting the `Asym` parameter in the model causes the `plinear` algorithm
+`nls()` has an option `algorithm="plinear"` that allows some partially linear
+models to be solved. The other tools, as far as we are aware, do not offer any
+such capability, with no external package providing it. 
+The feature in `nls()` has, however, some awkwardness.
+In particular, we would like to specify a model to a solver consistently
+across solver methods. Unfortunately, within `nls()` itself we must use
+different specifications with different `algorithm` options.
+
+The nonlinear model specifications are, of course, a development of linear ones.
+Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear
+modeling function `lm()`, which requires this model to be specified as `y ~ x`.
+Within `nls()`, consider the following FOUR different specifications for the same 
+problem, plus an intuitive choice, labelled `fm2a`, that does NOT work. 
+In this failed attempt, putting the `Asym` parameter in the model causes the
+`plinear` algorithm
 to try to add another term to the model. We believe this is unfortunate, and would
@@
 
-## Feature: indexed parameters
+## Models with indexed parameters
 
@@
 
-## Feature: bounds on parameters
-
-There are many situations where the context of a problem constrains the values of parameters.
-For example, one of us was asked many years ago to estimate a model where one parameter estimate
-was negative. The client complained "But that is supposed to be the number of grain elevators
-in Saskatchewan". The number should have been a positive integer, and likely less
-than a few thousand.
-
-`nls()` can impose bounds on parameters, but only if the `port` algorithm is used. Unfortunately,
-the manual states
-
-> The `algorithm = "port"` code appears unfinished, and does not even check that the starting 
-  value is within the bounds. Use with caution, especially where bounds are supplied.
-
-This is clearly unsatisfactory for general use, but does find bounded solutions. 
-Inclusion of bounds on parameters in nonlinear least squares computations is relatively 
-straightforward, and it is part of the default method
-for package `nlsr`. Package `minpack.lm` also includes provision for bounds, but the following
-script, which we do not run here for reasons of space, shows that it does not find a
-good solution for the scaled Hobbs test problem.
-
-### Philosophical considerations
-
-Bounds on parameters raise some interesting and difficult questions about how uncertainty
-in parameter estimates should be computed or reported. That is, the traditional "standard
-errors" are generally taken to imply symmetric intervals about the point estimate in which
-the parameter may be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
-errors nor their derived statistics when bounds are active.
+If such models are of enough importance, we believe it would be relatively
+straightforward to prepare an external package to provide for their estimation,
+including an appropriately formatted output object so the indexing of the 
+parameters can be used in other calculations.
+
+---OLD VERSION BELOW ------------------------
 

### 2023-03-07 RefactoringNLS/ComparingNLStools.Rmd
minor change
more about refactoring

@@
 are a useful idea to simplify model specification in some cases. However, they
-are also a focus of trouble.
+are also a focus of trouble. Typically we have some common parameters and 
+others that are tied to particular cases.
 
@@
 
----OLD VERSION BELOW ------------------------
-
-## Issue: code structure and documentation for maintenance
-
-The `nls()` code is structured in a way that inhibits both maintenance and improvement.
-In particular, the iterative setup is such that introduction of Marquardt stabilization
-is not easily available. To obtain performance when the code was written, a large portion
-was written in C with consequent calls and returns that
-severely complicate the code. Over time, R has become much more efficient on modern computers, and
-the need to use compiled C and Fortran is less critical. Moreover, the burden for maintenance
-could be reduced by moving code entirely to R.
-
-While R and its packages are generally very well documented for usage, that documentation
-rarely extends as far as it might for code maintenance. 
-The paucity of such documentation is exacerbated by the mixed R/C/Fortran code base,
-where seemingly trivial differences in things like representation of numbers or
-base index for arrays lead to traps for unwary programmers.
-
-# Goals and progress of our effort to improve `nls()`
-
-## Code rationalization and documentation
-
-When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
-
-- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
-  that matches the features in base R or what is packaged in `nlspkg` as described 
-  in @PkgFromRbase22. The R version can either serve as the operational code or
-  as a reference version to facilitate the maintenance of C, Fortran, or other 
-  versions of the software.
-
-- to streamline the overall `nls()` infrastructure. By this we
-  mean a re-factoring of the routines so they are better suited to maintenance of both
-  the existing `nls()` methods and features as well as new features we or others
-  would like to add.
-
-- to explain what we do, either in comments, examples, or separate maintainer documentation.
-  Since we are complaining about the lack of explanatory material for the current
-  code, we feel it is incumbent on us to provide such material for our own work, and if
-  possible for the existing code.
-
-These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
-commentary.
-
-## Rationalization of formula specifications
-
-We have seen that
-`nls()` uses a different formula specification from the default if the 
-`plinear` algorithm is used. This is unfortunate, since the user cannot then simply
-add `algorithm="plinear"` to the standard call, making errors more likely.
-
-Nonlinear models, apart from those using the "plinear" algorithm, need all the parameters specified.
-That is, the model formula should be written down as it would be computed.
-Thus a linear model (which is valid input for nonlinear estimation) should
-be specified `y ~ a * x + b` rather than `y ~ x`. 
-Moreover, we want to avoid the partially linear parameters appearing 
-in the result object as `.lin1`, `.lin2`, etc., where the actual position 
-of parameters in the model is not explicit. We are unclear why these parameters
-begin with a dot, that is, that they have the "hidden" indicator.
-
-A possible workaround to allow consistency would be to specify the 
-partially linear parameters when the `plinear` option is specified. For example, 
-we could use `algorithm="plinear(Asym)"` while keeping the general model formula 
-from the default specification. This would allow 
-for the output of different `algorithm` options to be consistent. However, we have not
-yet tried to code such a change, and welcome discussion from those working with
-other packages using model formulas.
-
-A further complication of model specification in R is that some formulas require
-the user to surround a term with `I()` to inhibit interpretation of arithmetic 
-operators. We would prefer that formulas be usable without this complication.
-
-<!-- It is also of interest to consider code that will detect partial linearity,  -->
-<!-- rather than expecting the user to provide the special structure and settings. Work in this -->
-<!-- area appears to be quite limited, with @Birkisson2013 seemingly a sole contribution -->
-<!-- we can find. -->
-<!-- Unfortunately, the mention of a "Linear and Nonlinear Discoverer" in @Zhang2011LinNonlin  -->
-<!-- is about a class of models that have almost nothing in common -->
-<!-- with the present issue. The algorithm in @Birkisson2013 will need investigating further, -->
-<!-- but we suspect that it is complicated enough that it will not easily be adapted to  -->
-<!-- a program like `nls()`. Possibly a simpler algorithm may be found. -->
-
-<!-- ``` -->
-<!-- y ~ x + z + x*z -->
-<!-- y ~ a + b*x +c*z + d * (x*z)  # <-- fully specified -->
-<!-- y ~ a*Fn1(b, c, x, z)/Fn2() -->
-
-<!-- ``` -->
-
-## Rationalization of indexed models
-
-Indexed models clearly have a place in some areas of research. 
-<!-- We need to be able to process models, as the example above, where the -->
-<!-- model formula is  -->
-We need to be able to process models such as 
-`Length ~ a[Strip] + b[Strip] * exp(-Conc / th)`, where `Strip` is the index. Data
-`Length` and `Conc` are available for all observations and parameter `th` applies
-to every fitted point, but `a[Strip]` and `b[Strip]` apply only to data indexed, which
-could be an experimental run number, or some other categorization.
-
-Given that `[ ]` provides the R method to index arrays, the presence of square
-brackets in a model
-formula signals indexing. The question is how to perform the correct calculations.
-
-<!-- While tedious (??maybe dummies or fastdummies packages make it less so?),  -->
-<!-- we could set up the model as a sum of similar models with dummy -->
-<!-- variables to turn "on" or "off" the particular case, transforming the model -->
-<!-- parameters to simple, non-indexed ones (as does `nls()`). -->
-
-Our view is that inclusion of the indexing **within** `nls()` or similar tools introduces 
-complexity that could be avoided by a suitable wrapper function. The modeling
-formula must be expanded to encompass all the index cases, likely with indexed
-parameters renamed e.g., `a[2]` becomes `a2` as at present with `nls()`. However,
-a wrapper could process output to reconstitute the indexing of parameters.
-
-## Streamlining code
-
-We are pessimistic that the overall structure of `nls()` can be streamlined due
-to the entanglement of so many features with the complicated mix of R, C and
-Fortran. Indeed, despite digging into the code over some months, we do not feel
-confident that we sufficiently understand it nor that we could maintain it. On the other
-hand, we do believe equivalent functionality can be built, but with at least
-some differences in how the features will be accessed.
-
 ## Tests and use-case examples
@@
   results comparable to or better than the functions they are to replace;
-- to test individual solver functions to ensure they work across the range of calling 
-  mechanisms, that is, different ways of supplying inputs to the solver(s);
+- to test individual solver functions to ensure they work across the range of 
+  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
 - to pose "silly" inputs to see if these bad inputs are caught by the programs.
@@
 
-## A testing package: NLSCompare
-
-One of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
+In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
 A primary design objective of this is to allow the summarisation of multiple tests in
@@
 
-# Strategic choices in nonlinear model estimation
+## Should nls() be refactored?
+
+When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
+
+- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
+  that matches the features in base R or what is packaged in `nlspkg` as described 
+  in @PkgFromRbase22. Our intent in providing a purely R version of the code was
+  either to replace `nls()` or to serve as a reference version aiding maintenance 
+  of C, Fortran, or other components of the software.
+
+- to streamline the overall `nls()` infrastructure. By this we mean a 
+  refactoring of the routines so they are better suited to maintenance of both
+  the existing `nls()` methods and features as well as new features we or others
+  would like to add.
+
+- to explain what we do, either in comments, examples, or separate maintainer 
+  documentation. While R and its packages are generally very well documented for
+  usage, that documentation
+  rarely extends as far as it might for code maintenance. The paucity of such
+  documentation is exacerbated by the mixed R/C/Fortran code base,
+  where seemingly trivial differences in things like representation of numbers or
+  base index for arrays lead to traps for unwary programmers.
+
+These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
+commentary.
+
+A serious obstacle to changing `nls()` is that there are many legacy applications
+and examples. Furthermore, with `nls()` as part of the base R distribution, users
+are drawn to using it even when it is less than ideal. A possible way forward
+is to
+
+- use a packaged version of `nls()` to provide legacy features, with notification
+  to users that there are other tools available
+
+- if maintainers are willing, to merge and rationalize the capabilities of 
+  `minpack.lm`, `nlsr` and `gslnls`
+
+- to encourage the (separate) development of capabilities such as indexed
+  parameter models if these are in demand.
+
+This direction does not, of course, resolve the inconsistencies in model 
+specification in formulas for nonlinear models, which undoubtedly is an issue
+for other modeling tools.
+
+A possible workaround to allow consistency for the `plinear` algorithm 
+would be to specify the linear parameters when the `plinear` option is specified. 
+For example, we could use `algorithm="plinear(Asym)"` while keeping the general 
+model formula from the default specification. This would allow 
+for the output of different `algorithm` options to be consistent. However, we 
+have not yet tried to code such a change, and welcome discussion from those working
+with other packages using model formulas.
+
+A further complication of model specification in R is that some formulas require
+the user to surround a term with `I()` to inhibit interpretation of arithmetic 
+operators. We would prefer that formulas be usable without this complication.
+
+---OLD VERSION BELOW ------------------------
+
+## Future of nonlinear model estimation in R
 
 It is possible that `nls()` will remain more or less as it has been for the past
-two decades. Given its importance to R, the focus of discussion should be the measures
-needed to secure its continued operation for legacy purposes and how that may be approached.
-We welcome an opportunity to participate in such conversations.
+two decades. Given its importance to R, the focus of discussion should be the
+measures needed to secure its continued operation for legacy purposes and how that
+may be approached. We welcome an opportunity to participate in such conversations.
 

### 2023-03-13 RefactoringNLS/ComparingNLStools.Rmd
update bib and smaller pkgs

@@
 
-The tools we will compare are the base-R `nls()` function 
+The tools we will consider are the base-R `nls()` function 
 and a number of packages that are available on the CRAN repository.
@@
 
-onls -- when there are errors in variables, this minimizes the orthogonal 
+onls (@R-onls-manual) -- when there are errors in variables, this minimizes the orthogonal 
 residuals

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
update with citations and descriptions

@@
 Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
-**minpack.lm** (@minpacklm12), 
-gslnls (@gslnls23), and nls2 (@nls2-22).
-
-Other CRAN packages worth Commentary: ??
-
-??put in proper refs to each
-
-onls (@R-onls-manual) -- when there are errors in variables, this minimizes the orthogonal 
-residuals
-
-crsnls -- Controlled random search. No vignette!
-
-NISTnls -- NIST problems for testing nls methods
-
-nlshelper -- A few utilities for summarizing, testing, and plotting non-linear
-    regression models fit with nls(), nlsList() or nlme()
-
-nlsic -- We solve non linear least squares problems with optional
-    equality and/or inequality constraints.
-
-nlsMicrobio -- Data sets and nonlinear regression models dedicated to predictive microbiology.
+**minpack.lm** (@minpacklm12), gslnls (@gslnls23), and nls2 (@nls2-22).
+
+There are a number of other CRAN packages worth commentary, and we will give
+some capsule comments later. Packages outside of CRAN, even if they might be
+worthy of review, will not be included here.
+
+??move down
+## Other CRAN packages
+
+`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
+is the vertical difference between the dependent variable and the functional
+model. `onls` minimizes the sum of orthogonal residuals. The objective is
+therefore different, and involves nontrivial extra calculation. A vignette with
+the package and the blog article https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
+give some description with illustrative graphs. `onls` appears to be limited to 
+problems with one independent and one dependent variable. The Wikipedia article
+https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
+ideas. Our view is that the approach needs wider discussion and tutorial examples
+to allow its merits to be judged.
+
+`crsnls` (@R-crsnls-manual) -- This package allows nonlinear estimation by 
+controlled random search via two methods. There is unfortunately no vignette.
+A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
+a small fraction of the time required by either of the methods in `crsnls`.
+
+NISTnls (@NISTnls-manual) -- This package provides R code and data for a set of
+nonlinear least squares problems from the U.S. National Institute for Standards
+and Technology for testing solvers. We caution that the problems are generally 
+poorly conditioned and may not represent real-world situations.
+
+nlshelper (@nlshelper-manual) -- This package, which unfortunately lacks a
+vignette, provides a few utilities for summarizing, testing, and plotting 
+non-linear regression models estimated with nls(), nlsList() or nlme() that
+are linked or grouped in some way.
+
+nlsic (@nlsic-manual) -- This solves non linear least squares problems with optional
+equality and/or inequality constraints. It is clearly not about modeling, and
+the input and output are quite different from class `nls` methods.
+
+`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models dedicated to predictive microbiology, including a vignette, by authors of the
+`nlstools` package who have a strong record in the subject.
+
+`nlstools` (@p-nlstools) -- Several tools for assessing the quality of fit of a gaussian nonlinear model are provided.
 
@@
 
-nls.multstart -- Non-linear least squares regression with the Levenberg-Marquardt algorithm using multiple starting values for increasing the chance that the minimum found is the global minimum.
-
-nlstac -- Set of functions implementing the algorithm described in Fernandez 
+`nls.multstart` (@nlsmultstart) -- Non-linear least squares regression using
+AIC scores with the Levenberg-Marquardt algorithm using multiple starting values for increasing the chance that the minimum found is the global minimum.
+
+`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
+motivations, but a longer history and using `nls()` within multiple trials.
+
+`nlstac` (@m-nlstac) -- Set of functions implementing the algorithm described in Fernandez 
     Torvisco et al. (2018) for fitting separable nonlinear regression curves.
+
     ?? check ref and add to bib.??
-
-nlstools -- Several tools for assessing the quality of fit of a gaussian nonlinear model are provided.
 

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
cleanup

@@
 
-# Abstract
+## Abstract
 
@@
 There are a number of other CRAN packages worth commentary, and we will give
-some capsule comments later. Packages outside of CRAN, even if they might be
-worthy of review, will not be included here.
-
-??move down
+some capsule comments later. It is likely that there are a number of other 
+tools that could be relevant to
+this comparison in the Bioconductor (@Bioconductor) collection 
+and on repositories like
+Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
+not pursue those possibilities here.
+
+## Comparison dimensions
+
+The tools for nonlinear estimation we consider have been built with different
+goals and contexts. We will make our comparison in a number of directions.
+We will make a number of specific notes, but our main commentary will consider
+the following categories. These overlap, for example in that the emphasis on modeling
+versus solving the nonlinear least squares problem will result in different
+output.f
+
+### Modeling versus minimizing a sum of squares
+
+The goal of many statisticians and other workers is to estimate a model. The
+minimization of a sum of squared residuals -- the nonlinear least squares problem --
+is the mechanism to estimate the model. Some of the tools we review focus on the
+modeling, so attempt to provide measures of uncertainty and other information. 
+The interface to modeling tools will generally try to streamline that task. By
+contrast, nonlinear least squares tools are trying to solve the more limited
+numerical task.
+
+### Programming language(s)
+
+R tools are our interest, but they often call programs in other programming
+languages. There are advantages in using a well-established program that has
+been much used and tested. Early versions of R were relatively sluggish in 
+computational speed. The downside is that we must interface between two or more
+programming languages, adding to the risk of errors. Changes to one or the 
+other language or its library support may
+force maintenance issues. Moreover, the pool of expertise with languages like C
+and Fortran is diminishing.
+
+### Internal solver stabilization
+
+Early nonlinear least squares programs generally used the Gauss-Newton method 
+(@Hartley1961). This is often very effective and efficient, but for a quite large 
+proportion of problems can fail to find the solution. Stabilization of the 
+computations is relatively straightforward in principle, and overcomes most
+of the failures, though sometimes at the cost of more calculations.
+
+### Jacobian calculation
+
+The necessarily iterative solution of a nonlinear least squares problem 
+to carry out nonlinear modeling requires that we can find smaller values
+of a sum of squared residuals. This implies there is a "downhill" direction.
+The gradient of the residual sum of squares is computed as the product
+of the Jacobian matrix: the partial derivatives of the residual vector 
+with respect to the modeling parameters.
+(?? do we need to present this, or give a ref)
+
+Most researchers find the effort to interface the differential calculus 
+required to specify the Jacobian to a nonlinear estimation tool very 
+challenging. The work-around is generally to use a numerical approximation,
+but there is the possibility of using symbolic or automatic differentiation,
+and there are particular models where workers have provided code for the
+Jacobian.
+
+### Algorithm and termination choices
+
+Iterative algorithms need to be stopped at some point, and how termination
+is forced may alter results. Furthermore, even supposedly similar algorithms
+may differ greatly in detail.
+
+### Constraints on parameters
+
+Reality often imposes conditions on parameters. While general constraints can
+be difficult to include, bounds on parameters can be accommodated by many 
+solvers. For some reason, fixed parameters (masks) are not often considered,
+but are very convenient when a parameter has a commonly accepted value that we
+later wish to allow to be estimated from the data.
+
+### Special capabilities
+
+Users often have particular problems that introduce other requirements. It is
+common to weight observations where their reliability is thought to differ.
+Some situations present a number of related situations where a combined model
+with some common and some local parameters are present, and the local parameters
+could be indexed. The common exponential decay and related **partially linear**
+models (sometimes called **separable** models) is a difficult one that can 
+benefit from special treatment.
+
+### Output
+
+While the aim of nonlinear estimation is to return a set of parameters for a
+particular model, there is a variety of additional information which may be
+provided. From the perspective of the nonlinear least squares problem, there
+are measures of work and information on the quality of the solution, while
+for statistical purposes measures of uncertainty in the parameters are helpful.
+
+Some tools allow constraints to be imposed on the solution, and it is helpful
+to know if these constraints are active at the proposed solution.
+
+?? DO WE WANT A COMPARISON TABLE
+
+### Documentation
+
+We believe that the availability of good documentation with examples that users
+can modify to deal with their own problems is essential. The presence of vignettes, 
+demos, and tests indicates a package is well-supported.
+
 ## Other CRAN packages
@@
 
-`nlstools` (@p-nlstools) -- Several tools for assessing the quality of fit of a gaussian nonlinear model are provided.
-
-nlsmsn -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) regression, details in Garay, Lachos and Abanto-Valle (2011)
+`nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation of nonlinear models, particularly
+using `nls()`. The vignette is actually a journal article, and the authors have considerable experience in the subject.
+
+`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) regression, with details in Garay, Lachos and Abanto-Valle (2011).
+The problem here is to minimize an objective which is modified from the traditional sum of squared residuals.
 
@@
 `nlstac` (@m-nlstac) -- Set of functions implementing the algorithm described in Fernandez 
-    Torvisco et al. (2018) for fitting separable nonlinear regression curves.
-
-    ?? check ref and add to bib.??
-
-easynls -- Fit and plot some nonlinear models. ?? no vignette!
-
-nlraa -- a set of nonlinear *selfStart* models. Most include analytic Jacobian
-code. (@MiguezNLRAA2021)
-
-optimx -- not really a nonlinear least squares method, but has general optimizers
+    Torvisco et al. (2018) for fitting separable nonlinear regression curves. The special 
+    class of problem for which this package is intended is an important and difficult one.
+    No vignette is provided, unfortunately.
+
+`easynls`  -- Fit and plot some nonlinear models. Thirteen models are treated, but there is minimal 
+documentation and no vignette. Package `nlraa` is to be preferred.
+
+`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models. Most include analytic Jacobian
+code.
+
+`optimx` (@p-optimx) -- not really a nonlinear least squares method, but has general optimizers
 that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
-squares.
-
-It is likely that there are a number of other tools that could be relevant to
-this comparison in the Bioconductor (@Bioconductor) collection 
-and on repositories like
-Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
-not pursue those possibilities here.
-
-## Comparison dimensions
-
-The tools for nonlinear estimation we consider have been built with different
-goals and contexts. We will make our comparison in a number of directions.
-We will make a number of specific notes, but our main commentary will consider
-the following categories. These overlap, for example in that the emphasis on modeling
-versus solving the nonlinear least squares problem will result in different
-output.f
-
-### Modeling versus minimizing a sum of squares
-
-The goal of many statisticians and other workers is to estimate a model. The
-minimization of a sum of squared residuals -- the nonlinear least squares problem --
-is the mechanism to estimate the model. Some of the tools we review focus on the
-modeling, so attempt to provide measures of uncertainty and other information. 
-The interface to modeling tools will generally try to streamline that task. By
-contrast, nonlinear least squares tools are trying to solve the more limited
-numerical task.
-
-### Programming language(s)
-
-R tools are our interest, but they often call programs in other programming
-languages. There are advantages in using a well-established program that has
-been much used and tested. Early versions of R were relatively sluggish in 
-computational speed. The downside is that we must interface between two or more
-programming languages, adding to the risk of errors. Changes to one or the 
-other language or its library support may
-force maintenance issues. Moreover, the pool of expertise with languages like C
-and Fortran is diminishing.
-
-### Internal solver stabilization
-
-Early nonlinear least squares programs generally used the Gauss-Newton method 
-(@Hartley1961). This is often very effective and efficient, but for a quite large 
-proportion of problems can fail to find the solution. Stabilization of the 
-computations is relatively straightforward in principle, and overcomes most
-of the failures, though sometimes at the cost of more calculations.
-
-### Jacobian calculation
-
-The necessarily iterative solution of a nonlinear least squares problem 
-to carry out nonlinear modeling requires that we can find smaller values
-of a sum of squared residuals. This implies there is a "downhill" direction.
-The gradient of the residual sum of squares is computed as the product
-of the Jacobian matrix: the partial derivatives of the residual vector 
-with respect to the modeling parameters.
-(?? do we need to present this, or give a ref)
-
-Most researchers find the effort to interface the differential calculus 
-required to specify the Jacobian to a nonlinear estimation tool very 
-challenging. The work-around is generally to use a numerical approximation,
-but there is the possibility of using symbolic or automatic differentiation,
-and there are particular models where workers have provided code for the
-Jacobian.
-
-### Algorithm and termination choices
-
-Iterative algorithms need to be stopped at some point, and how termination
-is forced may alter results. Furthermore, even supposedly similar algorithms
-may differ greatly in detail.
-
-### Constraints on parameters
-
-Reality often imposes conditions on parameters. While general constraints can
-be difficult to include, bounds on parameters can be accommodated by many 
-solvers. For some reason, fixed parameters (masks) are not often considered,
-but are very convenient when a parameter has a commonly accepted value that we
-later wish to allow to be estimated from the data.
-
-### Special capabilities
-
-?? weighting, indexed parameters, partially linear ...
-
-### Output
-
-While the aim of nonlinear estimation is to return a set of parameters for a
-particular model, there is a variety of additional information which may be
-provided. From the perspective of the nonlinear least squares problem, there
-are measures of work and information on the quality of the solution, while
-for statistical purposes measures of uncertainty in the parameters are helpful.
-
-Some tools allow constraints to be imposed on the solution, and it is helpful
-to know if these constraints are active at the proposed solution.
-
-?? DO WE WANT A COMPARISON TABLE
-
-### Documentation
-
-?? presence of vignettes, demos, (good?) examples
+squares, as well as other objectives. This is not generally recommended if nonlinear least 
+squares programs can be easily used, but provides a way to approach problems when others are
+unavailable.
 

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
fixup some refs

@@
 therefore different, and involves nontrivial extra calculation. A vignette with
-the package and the blog article https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
+the package and the blog article 
+https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
 give some description with illustrative graphs. `onls` appears to be limited to 
@@
 https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
-ideas. Our view is that the approach needs wider discussion and tutorial examples
-to allow its merits to be judged.
+ideas, with references to the literature. Our view is that the approach needs wider 
+discussion and tutorial examples to allow its merits to be judged.
 
@@
 A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
-a small fraction of the time required by either of the methods in `crsnls`.
-
-NISTnls (@NISTnls-manual) -- This package provides R code and data for a set of
+a small fraction of the time required by either of the methods in `crsnls`. The
+method discussed in @tvrdik2007 claims better reliability in finding solutions
+than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
+on the extreme NIST examples (@NISTnls-manual).
+
+`NISTnls` (@NISTnls-manual) -- This package provides R code and data for a set of
 nonlinear least squares problems from the U.S. National Institute for Standards
@@
 
-nlshelper (@nlshelper-manual) -- This package, which unfortunately lacks a
+`nlshelper` (@nlshelper-manual) -- This package, which unfortunately lacks a
 vignette, provides a few utilities for summarizing, testing, and plotting 
@@
 
-nlsic (@nlsic-manual) -- This solves non linear least squares problems with optional
-equality and/or inequality constraints. It is clearly not about modeling, and
+`nlsic` (@nlsic-manual) -- This solves non linear least squares problems with optional
+equality and/or inequality constraints. It is clearly **not** about modeling, and
 the input and output are quite different from class `nls` methods.

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
Edits

@@
 
-`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models dedicated to predictive microbiology, including a vignette, by authors of the
+`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models dedicated 
+to predictive microbiology, including a vignette, by authors of the
 `nlstools` package who have a strong record in the subject.
 
-`nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation of nonlinear models, particularly
-using `nls()`. The vignette is actually a journal article, and the authors have considerable experience in the subject.
-
-`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) regression, with details in Garay, Lachos and Abanto-Valle (2011).
-The problem here is to minimize an objective which is modified from the traditional sum of squared residuals.
+`nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation 
+of nonlinear models, particularly using `nls()`. The vignette is actually a journal article, 
+and the authors have considerable experience in the subject.
+
+`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
+regression, with details in Garay, Lachos and Abanto-Valle (2011). The problem 
+here is to minimize an objective which is modified from the traditional sum of squared residuals.
 
 `nls.multstart` (@nlsmultstart) -- Non-linear least squares regression using
-AIC scores with the Levenberg-Marquardt algorithm using multiple starting values for increasing the chance that the minimum found is the global minimum.
+AIC scores with the Levenberg-Marquardt algorithm using multiple starting values 
+for increasing the chance that the minimum found is the global minimum.
 
 `nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
-motivations, but a longer history and using `nls()` within multiple trials.
-
-`nlstac` (@m-nlstac) -- Set of functions implementing the algorithm described in Fernandez 
+motivations, but uses `nls()` within multiple trials. The package and its
+author have a long history in the R community.
+
+`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in Fernandez 
     Torvisco et al. (2018) for fitting separable nonlinear regression curves. The special 
@@
 $$ 1/scal  =  b_3 $$
-To allow for simpler discussion, let us say that the parameters for a vector
+To allow for simpler discussion, let us say that the parameters form a (named) vector
 $p$ and the model function is called $model(p)$. The residuals can be written
@@
 
-since when squared the loss function has an identical value. In both cases, the
+since the sumsquares loss function has an identical value. In both cases, the
 residual vector is a function of $p$. The second form
@@
 solvers are usually a subset, is to try to reduce the loss function (or objective
-function) defined by the sum of squares of the residuals ($y - model()$) by 
+function) defined by the sum of squares of the residuals by 
 altering the model parameters from some starting guess. In all general 
@@
 though we can avoid some loss of accuracy by NOT forming the inner product matrix
-$J' J$ by solving the linear least squares matrix problem
+$J' J$ and solving the linear least squares matrix problem
 
@@
 precision arithmetic found on some minicomputers in the 1970s, one of us suggested
-a linear combination of these ideas, as with low precision, some elements of 
+a linear combination of these ideas. In low precision, some elements of 
 $J' J$ could underflow to zero (@jn77ima). We note that the various choices for
@@
 Program modifications to give a sequence like  1.0, 1.1, 1.2, 1.3 for the four 
-parameters are fairly obvious. However, users may want to avoid equal parameter
-starts in situations like this when setting starting values.
+parameters are fairly obvious.
 
@@
 very small number so that we need to watch for a zero-divide or 0/0 
-situation in computing $\delta$.
-The following show `nls()` does poorly on this example problem. Note that `nlsr`
-offers function `pnlm0()` to give a 1-line display of class nls method results,
+situation in computing $\delta$. Below we see `nls()` does poorly on this 
+example problem. Note that `nlsr`
+offers function `pnlm0()` to give a 1-line display of results of class `nls`,
 and `pshort()` for `nlsr`.
@@
 
-The singular values displayed by `print.nlsr()` (the internal name for the 
-adaptation of the generic `print()`) are displayed in a column to the right 
-of the coefficient  and standard error display, but are NOT specific to the 
-parameters. Their position is purely for efficient use of page space. The 
-most common use of the singular values is to gauge how "nearly singular" the 
-Jacobian is at the solution, and the ratio of the largest to smallest of the 
+The singular values allow us to gauge how "nearly singular" the 
+Jacobian is at the solution, and the ratio of the smallest to largest of the 
 singular values is a simple but effective measure. In the above example, we note 
-that the scaled logistic has the smallest ratio.
+that the scaled logistic has the largest ("least singular") ratio.
 
 The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
-large sum of squares, which may suggest that these program have failed. 
+large sum of squares, which may suggest that these programs have failed. 
 Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
@@
 
-More general termination tests can be used. `nlsr` a **small sum of 
+More general termination tests can be used. `nlsr` has a **small sum of 
 squares** test (**smallsstest**) that compares the latest evaluated sum of squared
@@
 Using the "gradient" attribute of the output of the Jacobian function to hold
-this matrix lets us embed this attribute of the **residual** function as well,
+this matrix lets us embed this attribute in the **residual** function as well,
 so that the call to `nlfb()` can be made with the same name used for 
@@
 As far as we can understand the logic in `nls()`, the Jacobian computation during
-parameter estimation is carried out entirely within the called C-language program,
-and the R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
+parameter estimation is carried out within the called C-language program
+and its wrapper R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
 in the R distribution source code. This is used to provide Jacobian information in 
@@
 the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, though in our experience the differences are usually
+can lead to different behaviour, though in our experience these differences are usually
 very minor.
 
-Minor numerical differences can give annoying results with ill-conditioned
+However, minor numerical differences can give annoying results with ill-conditioned
 problems.
@@
 models to be solved. The other tools, as far as we are aware, do not offer any
-such capability, with no external package providing it. 
-The feature in `nls()` has, however, some awkwardness.
-In particular, we would like to specify a model to a solver consistently
-across solver methods. Unfortunately, within `nls()` itself we must use
+such capability. The `nlstac` package uses a different algorithm.
+
+Within `nls()` itself we must, unfortunately, use
 different specifications with different `algorithm` options.
-
 The nonlinear model specifications are, of course, a development of linear ones.
-Unfortunately, the explicit model `y ~ a * x + b` does not work with the linear
+The explicit model `y ~ a * x + b` does not work with the linear
 modeling function `lm()`, which requires this model to be specified as `y ~ x`.

### 2023-03-14 RefactoringNLS/ComparingNLStools.Rmd
yet more edits

@@
 
----OLD VERSION BELOW ------------------------
-
 ## Future of nonlinear model estimation in R
@@
 that are created by tools such as `nls()` should have minimal cross-linkages and side-effects.
-The aspects of `nls()` that have given us the most trouble:
-
-- The functions that compute the residuals and jacobians often presume
-  the data and parameters needed are available in a particular environment. As long
-  as the correct environment is used, this provides a compact syntax to invoke the
+The aspects of `nls()` that have given us the most trouble follow:
+
+- R tools often presume data and parameters needed are available in an accessible environment. 
+  As long as the correct environment is used, this provides a compact syntax to invoke the
   calculations. The danger is that the wrong data is accessed if the internal search finds a
-  valid name that is not the object we want.
-
-- Weights, subsets, and various contextual controls (such as that for the `na.action` option)
-  are similarly taken from the first available source, making for a very simple
-  invocation of calculations, but the context is hidden from the user. Furthermore, it can be
-  difficult for those of us trying to maintain or improve the code to be certain we have the
-  context correct.
-
-- The mixing of R, C and Fortran code was necessary for computational performance in the past.
-  Lacking decent developer documentation, programmers now face a lot of work to fix or improve code. 
-  We believe in having at least a working reference version of code in a single programming language. 
-  If necessary, by
-  measuring ("profiling") the code, we can find bottlenecks and substitute for those slower 
-  parts of the reference code.
-
-- A design that isolates the setup, solution, and post-solution parts of
-  for complicated calculations reduces the number of objects that must be kept in 
-  alignment. On the other hand, it may mean users have to add more commands to complete
-  their computations.
-
-- All iterative codes should return the best solution they have found so far, even if there
-  are untoward conditions reached, such as a singular Jacobian. This modification could be
-  made to `nls()` in the short term, correcting the issue identified above under "Failure to
-  return best result achieved".
-
-- Given the existence of examples of good practices such as analytic derivatives, stabilized 
+  valid name that is not the object we want, or if `subset` or `na.action` settings modify
+  the selection, or `weights` are applied.
+
+- Mixing of R, C and Fortran code adds to the burden of following the program logic.
+
+- The class `nls` structure simplifies calls with its rich set of functions, but
+  also adds to the task of following what has been done. A design that isolates the 
+  setup, solution, and post-solution parts of complicated calculations reduces the 
+  number of objects that must be kept in alignment.
+
+Given the existence of examples of good practices such as analytic derivatives, stabilized 
 solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
-  be moving to incorporate them.
-
-We anticipate that tools like `nlsr` will become more attractive as users discover
-the increased robustness and general capabilities, along with the more recent documentation.
-From a design point of view, a key difference in approach between `nls()` and 
-`nlsr::nlxb()` is that `nls()` 
-builds a large infrastructure, especially the `nlsModel()` function,  from which the Gauss-Newton
-iteration can be executed and other statistical information such as profiles can be
-computed, while `nlxb()` returns quite limited information, and 
-computes what is needed on an "as and when" basis. This follows a path that one of us
-(JN) established almost 50 years ago with the software that became @cnm79, separating
-setup, solver, and post-solution analysis phases of computations. There is, however,
-the wrapper `wrapnlsr()` that solves the nonlinear estimation problem with the solver
-of `nlsr` then calls `nls()` to build the object of class `nls` for which profile log-likelihood
-information can be computed.
-<!-- Here, that last  -->
-<!-- phase is not part of `nlxb()`, but is the domain of other functions in the `nlsr` -->
-<!-- package. -->
-
-<!-- The `nls()` approach, as implemented in the base R code and `nlspkg` leads to considerable -->
-<!-- entangling of the different functions and capabilities. Even after -->
-<!-- both the preparation for and carrying out of the present project, we do not feel confident -->
-<!-- to explain the code completely, nor to maintain it. However, we have made some  -->
-<!-- forays into render parts of the code in R, and to understand where difficulties -->
-<!-- lie. -->
-
-<!-- ??fix next bits -->
-<!-- - The different dialects of modeling expressions as detailed above should be rationalized. This -->
-<!--   would, however, be a painful and mainly political activity. -->
-
-<!-- - The tools for analytic derivatives that exist in parts of `nls()` for self-starting models and -->
-<!--   in `nlsr` for problems specified as models that use expressions for which derivatives exist  -->
-<!--   should be more generally available, but the task is one of tedium and limited reward to the -->
-<!--   persons who do it, while benefiting the R community substantially. -->
-
-<!-- - Similarly, default methods need stabilized solvers e.g., Marquardt methods, though to date -->
-<!--   a mechanism to incorporate such modifications in `nls()` has eluded us. -->
-
-<!-- - Bounds constraints are sufficiently important that they should be part of the default -->
-<!--   methods -->
-<!--   (rather than algorithm = "port" for `nls()`). `minpack.lm` should handle them correctly. -->
+be moving to incorporate them. These capabilities are available now by using several tools,
+but it would be helpful if they were unified.
 
@@
 
-Hans Werner Borchers has been helpful in developing the proposal for this project
+Hans Werner Borchers was helpful in developing GSoC project motivating this article
 and in comments on this and related work. Heather Turner co-mentored the project 

### 2023-03-15 RefactoringNLS/ComparingNLStools.Rmd
more edits

@@
 with respect to the modeling parameters.
-(?? do we need to present this, or give a ref)
 
@@
 to know if these constraints are active at the proposed solution.
-
-?? DO WE WANT A COMPARISON TABLE
 
@@
 ```{r exhobbsfn, eval=TRUE}
-require(minpack.lm)
-require(nlsr)
-require(gslnls)
+require(minpack.lm); require(nlsr); require(gslnls)
 hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
@@
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-summary(hobnlm)  
+pnlm0(hobnlm)  
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
-summary(hobgsln)
-hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12),jac=hobbs.jac)
-summary(hobgsl)
+pnlm0(hobgsln)
+hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
+pnlm0(hobgsl) # using analytic Jacobian
 ```
@@
 
->  **Do not use `nls` on artificial "zero-residual" data.**
-
-and suggested adding small perturbations to the data. This amounted to 
+>  *Do not use `nls` on artificial "zero-residual" data.*
+
+with suggested addition of small perturbations to the data. This amounted to 
 admitting `nls()` cannot solve well-posed problems unless data is polluted
@@
 since problems in function approximation and nonlinear equations can be 
-approached with nonlinear least squares. Moreover, the matter can
+approached with nonlinear least squares. Moreover, the matter 
 can be easily resolved. The "termination test" for the **program** rather than for 
@@
 Criterion (see @BatesWatts81). This projects the proposed step in the parameter
-vector on the gradient and estimates how much the sum of squares loss function 
+vector onto the gradient and estimates how much the sum of squares loss function 
 should decrease. This estimate is divided 
 by the current size of the loss function to avoid scale issues.
-When we have "converged", the estimated
+When we have "converged", the estimated relative
 decrease is very small. However, with small residuals,
@@
 Adding a small quantity to the loss function before dividing avoids trouble. 
-In 2021, one of us (J. Nash) proposed that `nls.control()` have an additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
+In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
+additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
 number -- 1.0 is a reasonable choice -- allows small-residual problems 
@@
 We are pleased to note that this improvement is now in the R distributed code 
-since version 4.1.0, and it has been a part of `nlsr` since it was introduced.
+since version 4.1.0; it has been a part of `nlsr` since it was introduced.
 
@@
 Termination after what may be considered excessive computation is also important.
-`nls()` stops after `maxiter` "iterations". 
-Unfortunately, the meaning of "iteration" varies among programs and requires 
-careful examination of the
-code. We prefer to use the number of times the residuals or the jacobian have
-been computed and put upper limits on these. Our codes exit (terminate) 
+`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
+examination of the code. We prefer to use the number of times the residuals or 
+the jacobian have been computed and put upper limits on these. Our codes exit (terminate) 
 when these limits are reached. Generally we prefer larger limits than the 
-default `maxiter = 50` of `nls()`,
-but that may reflect the more difficult problems we have encountered because 
-users consult us when standard tools have given unsatisfactory results.
+default `maxiter = 50` of `nls()`, avoiding the risk of stopping early.
 
@@
 Tools that produce a class `nls` output object create a rich set of functions
-and structures that are then used in a variety of modeling tasks. By contrast,
+and structures that are then used in a variety of modeling tasks, including the
+least squares solution. By contrast,
 `nlsr` computes quantities as they are requested or needed, with additional
@@
 of `selfStart` models, but the construction of them is non-trivial. A number 
-of such models are included with base R in `./src/library/stats/R/zzModels.R` file 
+of such models are included with base R in `./src/library/stats/R/zzModels.R`,  
 with package `nlraa` (@MiguezNLRAA2021) providing a richer set. 
@@
 computing the "right hand side" or `rhs` of the residual, then checks to see if the
-"gradient" attribute is defined, and, if not, uses `numericDeriv` to compute a 
+"gradient" attribute is defined, and, if not, uses `numericDeriv()` to compute a 
 Jacobian
@@
 In considering `selfStart` models, we noted that the base-R function `SSlogis`
-is intended to solve our example problem **Logistic3T**, but that it finds initial 
-parameters for `nls()` by using a completely separate solver, then reruns the
-solution with the default algorithm from the derived -- and final -- parameters.
-To provide simpler estimates of starting parameters, the function `SSlogisJN`
+is intended to solve our example problem **Logistic3T**.  This example is one 
+where the `getInitial()` function (used to find
+starting values) actually calls `nls()` recursively with a non-default algorithm
+to get "starting" parameters. Then it returns to running the default algorithm
+unnecessarily. Such circular references are, in our view, prone to creating errors
+in code maintenance. 
+To provide simpler starting parameters, the function `SSlogisJN`
 is now part of package `nlsr`.
 
-The `SSLogis` case is one where the `getInitial()` function (used to find
-starting values) actually calls `nls()` with a non-default algorithm. 
-Such circular references are, in our view, prone to creating errors
-in code maintenance.
-Users may also want to use these models with explicit starting values 
+Users may also want to use `selfStart` models with explicit starting values 
 other than those suggested by `getInitial()`. 
@@
 analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
-saves many quantities in "hidden"
+and others save many quantities in "hidden"
 variables, i.e., with names preceded by ".". 
@@
 ## Bounds constraints on parameters
-
-?? Arkajyoti -- we can probably trim these examples
 
@@
 # Hobbsmaskx.R -- masks with formula specification of problem
-require(nlsr)
-require(minpack.lm)
+require(nlsr); require(minpack.lm); traceval<-FALSE
 stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
 sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
-traceval<-FALSE
 # fix first parameter

### 2023-03-15 RefactoringNLS/ComparingNLStools.Rmd
finish copy edit

@@
 			upper=c(200, 60, 3), trace=traceval))
-summary(anlM1)
-# nls gives warnings
+pnlm0(anlM1)
 anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
 		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
-summary(anlsmsk1)
+pnls0(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
@@
 anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
-             upper=c(2,6,3), algorithm='port')
-summary(anlshmsk1)
-cat("More precisely...crossprod(resid(anlshmsk1))=",crossprod(resid(anlshmsk1)),"\n")
-# nlsLM does not always work with bounds
+             upper=c(2,6,30), algorithm='port')
+pnls0(anlshmsk1)
 anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                 upper=c(2,6,3))
-summary(anlsLMmsks1)
+                 upper=c(2,6,30))
+pnlm0(anlsLMmsks1)
 
@@
 
-All four major tools illustrated must solve some variant of the Gauss-Newton 
+All four major tools illustrated solve some variant of the Gauss-Newton 
 equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
@@
 not explored all five methods with various control settings.
-Though it is unlikely to be useful in general, 
-control settings for `nlxb()` or `nlfb()` allow for those routines to perform 
-a variety of Hartley and Marquardt algorithms. This
-has been useful for learning about the algorithms, but tangential to simply finding
-parameters of nonlinear models.
+Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
+algorithm variants.
 
@@
 Library (@GSL-manual).
-Generally, we believe that keeping to a single programming language can allow for 
+Generally, keeping to a single programming language can allow for 
 easier maintenance and upgrades. It also avoids some work when there are changes 
 or upgrades to libraries for the non-R languages. 
-On the other hand, R is usually somewhat slower 
-than some other computing systems because it keeps track of names and structures and 
-because it is usually interpreted rather than compiled. In recent years the 
+R is usually considered slower 
+than most computing environments because it keeps track of objects and 
+because it is usually interpreted. In recent years the 
 performance penalty for using code entirely in R has been much reduced
-with the just-in-time compiler and other improvements, so that using an all-R 
-package offers acceptable performance. Indeed, in `nlsr` the use of R may be 
-less of a performance cost than a policy of aggressively seeking a solution,
+with the just-in-time compiler and other improvements. All-R computation 
+may now offer acceptable performance. In `nlsr` the use of R may be a
+smaller performance cost than the aggressive approach to a solution,
 which can cause more iterations to be used.
@@
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
-the path, 
-when R is launched, will be loaded. The overwhelming proportion of R users in our
-acquaintance avoid saving the workspace because of the danger of lurking data and
+suppress saving this workspace (`.RData`) on exit, otherwise any such file in 
+the path will be loaded on startup. R users in our
+acquaintance avoid saving the workspace because lurking data and
 functions which may cause unwanted results.
-
-Nevertheless, to provide compatible behaviour with `nls()` and R generally, 
-programs need to ensure equivalent behaviour, but users should test that the 
-operation is as they intend.
+Program alternatives to `nls()` still need to ensure they work with data in 
+the available environments.
 
@@
 
-`nls()` and other class `nls` tools accept an argument `subset`.
-This acts through the mediation of `model.frame` and is not clearly obvious in the source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`.
-
+`nls()` and other class `nls` tools accept an argument `subset`. This acts through the 
+mediation of `model.frame`, which not obvious in the 
+source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`. 
 Having `subset` at the level of the call to a function like
@@
 solver needs to be aware of the
-source (and value) of objects such as the data, residuals and Jacobian. 
-This is overly complicated. By preference, 
-we would implement subsetting by means of zero-value weights, with observation counts
+origin (and value) of objects such as the data, residuals and Jacobian. 
+By preference, 
+we would implement subsetting by zero-value weights, with observation counts
 (and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
@@
 
-`na.action` is an argument to the `nls()` function, but it does not appear obviously in the source code, often being handled behind the scenes after referencing the option `na.action`.
-Yet this feature also changes the data supplied to our nonlinear
-least squares solver.
+`na.action` is an argument to the `nls()` function, but it does not appear 
+obviously in the source code, often being handled behind the scenes after 
+referencing the option `na.action`. Yet this feature also changes the data 
+supplied to our nonlinear least squares solver.
 
@@
 `getOption("na.action")`
-is `na.omit`. This option essentially omits from computations any observations
+is `na.omit`. This option removes from computations any observations
 containing missing values (i.e. any row of a data frame containing an NA). 
-`na.exclude` does much of the same for computations, but keeps the rows with NA 
+`na.exclude` does much of the same for solver computations, but keeps the rows with NA 
 elements so that predictions are in the correct row position. We recommend that 
@@
 name could
-be confusing.) However, the model frame is used within the function code in the form 
+be confusing.) Despite this, the model frame is used within the function code in the form 
 of the object
@@
 "Introduction to nlsr" of package `nlsr` in section 
-*Weights that are functions of the model parameters*. Care is advised
+*Weights that are functions of the model parameters* and 
+`minpack.lm` offers a function `wfct()` to 
+facilitate such weighting. Care is advised
 in applying such ideas.
@@
 
+?? Arkajyoti: I think this example may get too long or be too unclear.
+
 ```{r nlswtx, echo=TRUE}
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-weeddf <- data.frame(tt, weed)
 wts <- 0.5^tt # simple weights
@@
 nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-nowt
-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+attr(rnowt, "gradient")<-NULL; rnowt
 usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-usewt
-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+attr(rusewt,"gradient")<-NULL; rusewt
 source("nlsModel.R")
 nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+attr(rn0,"gradient")<-NULL; rn0
 nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
-nmod$resid()
+rn<-nmod$resid()
+attr(rn,"gradient")<-NULL; rn
 ```
@@
 
-Many nonlinear models can be expressed so that some of the parameters 
-appear linearly. In our logistic examples, the asymptote parameters are
-an illustration. In such cases, specialized methods can give spectactularly
-better performance. The variable projection method (@Golub1973, @OlearyRust13) is 
+The variable projection method (@Golub1973, @OlearyRust13) is 
 generally much more effective than general approaches in finding good solutions 
 to nonlinear least squares 
-problems when some of the parameters appear linearly. However, setting up the 
+problems when some of the parameters appear linearly. In our logistic examples, 
+the asymptote parameters are an illustration.However, setting up the 
 calculations, that is, identifying which parameters are linear, is not a trivial
@@
 models to be solved. The other tools, as far as we are aware, do not offer any
-such capability. The `nlstac` package uses a different algorithm.
+such capability. The `nlstac` package uses a different algorithm for similar goals.
 
@@
 different specifications with different `algorithm` options.
-The nonlinear model specifications are, of course, a development of linear ones.
-The explicit model `y ~ a * x + b` does not work with the linear
+For example, the explicit model `y ~ a * x + b` does not work with the linear
 modeling function `lm()`, which requires this model to be specified as `y ~ x`.
@@
 parameters are indexed. It also uses the "plinear" option as an added complication.
-
 Here we use a truncated version of the example to save display space.
@@
 
-A serious obstacle to changing `nls()` is that there are many legacy applications
-and examples. Furthermore, with `nls()` as part of the base R distribution, users
-are drawn to using it even when it is less than ideal. A possible way forward
+We found the legacy code and examples prevented us achieving a revised `nls()` 
+with all its current capabilities. Unfortunately, `nls()` is part of the base R distribution, 
+so users are drawn to using it even when it is less than ideal. A possible way forward
 is to

### 2023-05-02 RefactoringNLS/ComparingNLStools.Rmd
trying to streamline

@@
 Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
-**minpack.lm** (@minpacklm12), gslnls (@gslnls23), and nls2 (@nls2-22).
+**minpack.lm** (@minpacklm12), **gslnls** (@gslnls23), and **nls2** (@nls2-22).
 
@@
 The tools for nonlinear estimation we consider have been built with different
-goals and contexts. We will make our comparison in a number of directions.
-We will make a number of specific notes, but our main commentary will consider
-the following categories. These overlap, for example in that the emphasis on modeling
-versus solving the nonlinear least squares problem will result in different
-output.f
+goals and contexts. We will make our comparison considering the overlapping
+headings summarized below.
 
@@
 minimization of a sum of squared residuals -- the nonlinear least squares problem --
-is the mechanism to estimate the model. Some of the tools we review focus on the
-modeling, so attempt to provide measures of uncertainty and other information. 
+is the mechanism to estimate the model. Some tools focus on modeling, 
+so attempt to provide measures of uncertainty and other information. 
 The interface to modeling tools will generally try to streamline that task. By
@@
 of a sum of squared residuals. This implies there is a "downhill" direction.
-The gradient of the residual sum of squares is computed as the product
-of the Jacobian matrix: the partial derivatives of the residual vector 
+The gradient of the residual sum of squares is computed as its product
+with the Jacobian matrix: the partial derivatives of the residual vector 
 with respect to the modeling parameters.
 
-Most researchers find the effort to interface the differential calculus 
-required to specify the Jacobian to a nonlinear estimation tool very 
-challenging. The work-around is generally to use a numerical approximation,
-but there is the possibility of using symbolic or automatic differentiation,
-and there are particular models where workers have provided code for the
+Most researchers are challenged to program the differential calculus 
+required to specify the Jacobian to a nonlinear estimation tool. 
+The work-around is generally to use a numerical approximation,
+but there is the possibility of using symbolic or automatic differentiation.
+There are also particular models where workers have provided code for the
 Jacobian.
@@
 
-Reality often imposes conditions on parameters. While general constraints can
+Reality imposes conditions on parameters. While general constraints can
 be difficult to include, bounds on parameters can be accommodated by many 
 solvers. For some reason, fixed parameters (masks) are not often considered,
-but are very convenient when a parameter has a commonly accepted value that we
+but are convenient when a parameter has a commonly accepted value that we
 later wish to allow to be estimated from the data.
@@
 
-## Other CRAN packages
+## Other CRAN packages considered
 
@@
 is the vertical difference between the dependent variable and the functional
-model. `onls` minimizes the sum of orthogonal residuals. The objective is
+model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
 therefore different, and involves nontrivial extra calculation. A vignette with
@@
 
-`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models dedicated 
-to predictive microbiology, including a vignette, by authors of the
+`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models 
+dedicated  to predictive microbiology, including a vignette, by authors of the
 `nlstools` package who have a strong record in the subject.
@@
 `nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation 
-of nonlinear models, particularly using `nls()`. The vignette is actually a journal article, 
-and the authors have considerable experience in the subject.
+of nonlinear models, particularly using `nls()`. The vignette is actually a journal
+article,  and the authors have considerable experience in the subject.
 
@@
 regression, with details in Garay, Lachos and Abanto-Valle (2011). The problem 
-here is to minimize an objective which is modified from the traditional sum of squared residuals.
+here is to minimize an objective which is modified from the traditional sum of squared
+residuals.
 
@@
 
-`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in Fernandez 
-    Torvisco et al. (2018) for fitting separable nonlinear regression curves. The special 
-    class of problem for which this package is intended is an important and difficult one.
-    No vignette is provided, unfortunately.
-
-`easynls`  -- Fit and plot some nonlinear models. Thirteen models are treated, but there is minimal 
-documentation and no vignette. Package `nlraa` is to be preferred.
-
-`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models. Most include analytic Jacobian
-code.
-
-`optimx` (@p-optimx) -- not really a nonlinear least squares method, but has general optimizers
-that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
-squares, as well as other objectives. This is not generally recommended if nonlinear least 
-squares programs can be easily used, but provides a way to approach problems when others are
-unavailable.
+`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
+Fernandez Torvisco et al. (2018) for fitting separable nonlinear regression curves. The 
+special class of problem for which this package is intended is an important and 
+difficult one. No vignette is provided, unfortunately.
+
+`easynls`  -- Fit and plot some nonlinear models. Thirteen models are treated, but 
+there is minimal documentation and no vignette. Package `nlraa` is to be preferred.
+
+`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models. Most include
+analytic Jacobian code.
+
+`optimx` (@p-optimx) -- not really a nonlinear least squares method, but has general
+optimizers that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
+squares, as well as other objectives. This is not generally recommended if nonlinear 
+least squares programs can be easily used, but provides a way to approach problems 
+when others are unavailable.
 

### 2023-05-02 RefactoringNLS/ComparingNLStools.Rmd

@@
 in the context of improving and rationalizing R tools for nonlinear regression.
-The rich features of this function are weakened by a number of deficiencies and
+The rich features of this function are weakened by several deficiencies and
 inconsistencies, in particular a lack of stabilization of the Gauss-Newton solver.
@@
 Various packages, including our `nlsr`, provide alternative capabilities.
-This article considers the differences in goals, approach and features of
+This article considers the differences in goals, approaches, and features of
 different tools for nonlinear least squares modeling in R. Discussion of 
@@
 
-`nls()` is the tool in base R, the primary software software distribution
+`nls()` is the tool in base R, the primary software distribution
 from the Comprehensive R Archive Network (*https://cran.r-project.org*),
@@
 
-- handles formulas that include R functions, even ones which call calculations in
+- handles formulas that include R functions, even ones that call calculations in
   other programming languages
 - allows data to be weighted or subset
-- can estimate bound constrained parameters
+- can estimate bound-constrained parameters
 - provides a mechanism for handling partially linear models
@@
 parameters are changed
-- links to a number of pre-coded (`selfStart`) models that do not require
+- links to many pre-coded (`selfStart`) models that do not require
   initial parameter values.
@@
 The tools we will consider are the base-R `nls()` function 
-and a number of packages that are available on the CRAN repository.
+and some packages that are available on the CRAN repository.
 Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
@@
 
-There are a number of other CRAN packages worth commentary, and we will give
-some capsule comments later. It is likely that there are a number of other 
+There are some other CRAN packages worth commentary, and we will give
+some capsule comments later. There are likely several other 
 tools that could be relevant to
@@
 and on repositories like
-Github (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
+GitHub (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
 not pursue those possibilities here.
@@
 languages. There are advantages in using a well-established program that has
-been much used and tested. Early versions of R were relatively sluggish in 
+been used and tested. Early versions of R were relatively sluggish in 
 computational speed. The downside is that we must interface between two or more
@@
 required to specify the Jacobian to a nonlinear estimation tool. 
-The work-around is generally to use a numerical approximation,
+The workaround is generally to use a numerical approximation,
 but there is the possibility of using symbolic or automatic differentiation.
@@
 be difficult to include, bounds on parameters can be accommodated by many 
-solvers. For some reason, fixed parameters (masks) are not often considered,
+solvers. For some reason, fixed parameters (masks) are not often considered
 but are convenient when a parameter has a commonly accepted value that we
@@
 
-While the aim of nonlinear estimation is to return a set of parameters for a
+While nonlinear estimation aims to return a set of parameters for a
 particular model, there is a variety of additional information which may be
@@
 Some tools allow constraints to be imposed on the solution, and it is helpful
-to know if these constraints are active at the proposed solution.
+to know if these constraints are active in the proposed solution.
 
@@
 We believe that the availability of good documentation with examples that users
-can modify to deal with their own problems is essential. The presence of vignettes, 
+can modify to deal with their problems is essential. The presence of vignettes, 
 demos, and tests indicates a package is well-supported.
@@
 model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
-therefore different, and involves nontrivial extra calculation. A vignette with
+therefore different and involves nontrivial extra calculation. A vignette with
 the package and the blog article 
@@
 
-`nlsic` (@nlsic-manual) -- This solves non linear least squares problems with optional
+`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
 equality and/or inequality constraints. It is clearly **not** about modeling, and
@@
 `nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
-regression, with details in Garay, Lachos and Abanto-Valle (2011). The problem 
-here is to minimize an objective which is modified from the traditional sum of squared
+regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem 
+here is to minimize an objective that is modified from the traditional sum of squared
 residuals.
@@
 `nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
-motivations, but uses `nls()` within multiple trials. The package and its
+motivation, but uses `nls()` within multiple trials. The package and its
 author have a long history in the R community.
@@
 The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
-task which is seemingly straightforward but turns out to be quite nasty. Because
+task that is seemingly straightforward but turns out to be quite nasty. Because
 this problem has a very succinct statement, it often provides the "short 
@@
 
-The data and a graph of it is given below.
+The data and a graph of it are given below.
 
@@
 to set the gradient to zero (to find a stationary point of the function $S(p)$).
-This leads to the **Newton's equation**
+This leads to **Newton's equation**
 
@@
 In reality, there are many problems were $D$ should not be ignored, but the work
-to compute it precisely is considerable. A number of work-arounds have been 
+to compute it precisely is considerable. Many workarounds have been 
 proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
-the most commonly used. Henceforth we will simply say "Marquardt" as we believe
-he was the first to actually incorporate the ideas into a practical computer
+the most commonly used. Henceforth, we will simply say "Marquardt", as we believe,
+was the first to actually incorporate the ideas into a practical computer
 program.
@@
 Let us specify in R the three model formulae and set some starting
-values for parameters. These starting points are NOT equivalent, and
+values for parameters. These starting points are NOT equivalent and
 are deliberately crude choices. Workers performing many calculations 
@@
 where starting parameter values are not required. We discuss aspects of this
-feature at several points below. However, in the event that a `selfStart` model 
+feature at several points below. However, if a `selfStart` model 
 is not available, `nls()` sets all the starting parameters 
@@
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-summary(unlm1) # Use summary() to get display
+faraway::sumary(unlm1) # Use summary() to get display
 # unlm1 # Note the difference. Use this form to get sum of squares
@@
    we believe the authors have not taken into account all possibilities, though
-   it should be noted that all programs have some weakness in
-   regard to constrained optimization. Software creators have to work with
+   it should be noted that all programs have some weakness regarding constrained
+   optimization. Software creators have to work with
    assumptions on the extremity of scale that they are willing to countenance,
@@
 
-While fixed weights may be useful, there are a number of problems for which
+While fixed weights may be useful, there are many problems for which
 we want weights that are determined at least partially from the model 
@@
 `trace` argument TRUE, the iterations display the objective function value which 
-is 1/2 the sum of squares (or deviance). It is likely that the trace display is
+is 1/2 the sum of squares (or deviance). The trace display is likely
 embedded in the Fortran of the `nlminb` routine that is called to execute 
@@
 
-### Failure to return best result achieved
+### Failure to return the best result achieved
 

### 2023-05-03 RefactoringNLS/ComparingNLStools.Rmd
update with some simplifications

@@
 to compute it precisely is considerable. Many workarounds have been 
-proposed, of which the Levenberg-Marquardt (@Levenberg1944, @Marquardt1963) is
-the most commonly used. Henceforth, we will simply say "Marquardt", as we believe,
-was the first to actually incorporate the ideas into a practical computer
-program.
+proposed, of which the Levenberg-Marquardt stabilization 
+(@Levenberg1944, @Marquardt1963) is
+the most commonly used. We will use "Marquardt", as we believe he
+first incorporated the ideas into a practical computer program.
 
 The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
-a multiple of the diagonal part of $J' J$. For machines with exceptionally low
-precision arithmetic found on some minicomputers in the 1970s, one of us suggested
-a linear combination of these ideas. In low precision, some elements of 
-$J' J$ could underflow to zero (@jn77ima). We note that the various choices for
-$D$, especially the scaling multiple, as well as possible line search along the
+a multiple of the diagonal part of $J' J$. In low precision, some elements of 
+$J' J$ could underflow to zero (@jn77ima), and a linear combination of both
+choices was suggested as a compromise. Various choices for
+$D$, as well as a possible line search along the
 direction $\delta$ rather than a unit step (@Hartley1961), 
@@
 
-Let us specify in R the three model formulae and set some starting
+Let us specify in R the three model formulas and set some starting
 values for parameters. These starting points are NOT equivalent and
@@
 is not available, `nls()` sets all the starting parameters 
-to 1. This is, in our view, tolerable, but could be improved by using a set of values
+to 1. This is tolerable, but could be improved by using a set of values
 that are all slightly different, which, in the case of the example 
@@
 library(minpack.lm)
+# ??not needed -- library(nlsr) # for pnlm0
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-faraway::sumary(unlm1) # Use summary() to get display
+# summary not exported by faraway??
+# faraway::summary(unlm1) # Use summary() to get display
 # unlm1 # Note the difference. Use this form to get sum of squares
-# pnlm0(unlm1)  # Short form of output
+pnlm0(unlm1)  # Short form of output
 snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
@@
 of the nonlinear least squares problem rather than the estimation of a nonlinear 
-model that fits or explains the data. `nls()` and `nlsLM` return an object of 
-class `nls`, which allows for a number of specialized modeling and diagnostic extensions. For compatibility, 
+model that fits or explains the data. The object of class `nls` allows for 
+a number of specialized modeling and diagnostic extensions. For compatibility, 
 the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
@@
 when these limits are reached. Generally we prefer larger limits than the 
-default `maxiter = 50` of `nls()`, avoiding the risk of stopping early.
+default `maxiter = 50` of `nls()` to avoid stopping early.
 
 ### Returned results of nls() and other tools
+
+?? Can we simplify or suggest use of `str()` ??
 
@@
 To provide simpler starting parameters, the function `SSlogisJN`
-is now part of package `nlsr`.
+is now part of package `nlsr`, but is most useful for `nls()`.
 
@@
 
-### Feature: na.action
+### Feature: na.action (missing value treatment)
 
@@
 
-?? Arkajyoti: I think this example may get too long or be too unclear.
+<!-- ?? Arkajyoti: I think this example may get too long or be too unclear. -->
 
@@
 
+<!-- Ben Bolker mentioned a possibility of using a nonlinear model in  -->
+<!-- ??I think?? nlme with some sort of linear model form of the parameters. -->
+<!-- May need to mention. -->
+
 ## Tests and use-case examples
@@
 We found the legacy code and examples prevented us achieving a revised `nls()` 
-with all its current capabilities. Unfortunately, `nls()` is part of the base R distribution, 
+with all its current capabilities. Unfortunately, `nls()` is part of the base R
+distribution, 
 so users are drawn to using it even when it is less than ideal. A possible way forward
@@
 
-To advance the stability and maintainability of R, we believe the program objects (R functions)
-that are created by tools such as `nls()` should have minimal cross-linkages and side-effects.
+To advance the stability and maintainability of R, we believe the program 
+objects (R functions)
+that are created by tools such as `nls()` should have minimal 
+cross-linkages and side-effects.
 The aspects of `nls()` that have given us the most trouble follow:
 
-- R tools often presume data and parameters needed are available in an accessible environment. 
-  As long as the correct environment is used, this provides a compact syntax to invoke the
-  calculations. The danger is that the wrong data is accessed if the internal search finds a
+- R tools often presume data and parameters needed are available in an 
+  accessible environment. As long as the correct environment is used, this provides 
+  a compact syntax to invoke the
+  calculations. The danger is that the wrong data is accessed if the 
+  internal search finds a
   valid name that is not the object we want, or if `subset` or `na.action` settings modify
@@
 
-Given the existence of examples of good practices such as analytic derivatives, stabilized 
+Given the existence of examples of good practices such as analytic derivatives, 
+stabilized 
 solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
-be moving to incorporate them. These capabilities are available now by using several tools,
-but it would be helpful if they were unified.
+be moving to incorporate them. These capabilities are available now by 
+using several tools, but it would be helpful if they were unified.
 

### 2023-05-09 RefactoringNLS/ComparingNLStools.Rmd
more edit

@@
 
-<!-- ?? nlme::gnls() -- should we note it? -->
-
 ## Abstract
@@
 Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
-**minpack.lm** (@minpacklm12), **gslnls** (@gslnls23), and **nls2** (@nls2-22).
-
-There are some other CRAN packages worth commentary, and we will give
-some capsule comments later. There are likely several other 
-tools that could be relevant to
-this comparison in the Bioconductor (@Bioconductor) collection 
-and on repositories like
-GitHub (https://github.com) and Gitlab (https://about.gitlab.com), but we will 
-not pursue those possibilities here.
-
-## Comparison dimensions
-
-The tools for nonlinear estimation we consider have been built with different
-goals and contexts. We will make our comparison considering the overlapping
-headings summarized below.
-
-### Modeling versus minimizing a sum of squares
-
-The goal of many statisticians and other workers is to estimate a model. The
-minimization of a sum of squared residuals -- the nonlinear least squares problem --
-is the mechanism to estimate the model. Some tools focus on modeling, 
-so attempt to provide measures of uncertainty and other information. 
-The interface to modeling tools will generally try to streamline that task. By
-contrast, nonlinear least squares tools are trying to solve the more limited
-numerical task.
-
-### Programming language(s)
-
-R tools are our interest, but they often call programs in other programming
-languages. There are advantages in using a well-established program that has
-been used and tested. Early versions of R were relatively sluggish in 
-computational speed. The downside is that we must interface between two or more
-programming languages, adding to the risk of errors. Changes to one or the 
-other language or its library support may
-force maintenance issues. Moreover, the pool of expertise with languages like C
-and Fortran is diminishing.
-
-### Internal solver stabilization
-
-Early nonlinear least squares programs generally used the Gauss-Newton method 
-(@Hartley1961). This is often very effective and efficient, but for a quite large 
-proportion of problems can fail to find the solution. Stabilization of the 
-computations is relatively straightforward in principle, and overcomes most
-of the failures, though sometimes at the cost of more calculations.
-
-### Jacobian calculation
-
-The necessarily iterative solution of a nonlinear least squares problem 
-to carry out nonlinear modeling requires that we can find smaller values
-of a sum of squared residuals. This implies there is a "downhill" direction.
-The gradient of the residual sum of squares is computed as its product
-with the Jacobian matrix: the partial derivatives of the residual vector 
-with respect to the modeling parameters.
-
-Most researchers are challenged to program the differential calculus 
-required to specify the Jacobian to a nonlinear estimation tool. 
-The workaround is generally to use a numerical approximation,
-but there is the possibility of using symbolic or automatic differentiation.
-There are also particular models where workers have provided code for the
-Jacobian.
-
-### Algorithm and termination choices
-
-Iterative algorithms need to be stopped at some point, and how termination
-is forced may alter results. Furthermore, even supposedly similar algorithms
-may differ greatly in detail.
-
-### Constraints on parameters
-
-Reality imposes conditions on parameters. While general constraints can
-be difficult to include, bounds on parameters can be accommodated by many 
-solvers. For some reason, fixed parameters (masks) are not often considered
-but are convenient when a parameter has a commonly accepted value that we
-later wish to allow to be estimated from the data.
-
-### Special capabilities
-
-Users often have particular problems that introduce other requirements. It is
-common to weight observations where their reliability is thought to differ.
-Some situations present a number of related situations where a combined model
-with some common and some local parameters are present, and the local parameters
-could be indexed. The common exponential decay and related **partially linear**
-models (sometimes called **separable** models) is a difficult one that can 
-benefit from special treatment.
-
-### Output
-
-While nonlinear estimation aims to return a set of parameters for a
-particular model, there is a variety of additional information which may be
-provided. From the perspective of the nonlinear least squares problem, there
-are measures of work and information on the quality of the solution, while
-for statistical purposes measures of uncertainty in the parameters are helpful.
-
-Some tools allow constraints to be imposed on the solution, and it is helpful
-to know if these constraints are active in the proposed solution.
-
-### Documentation
-
-We believe that the availability of good documentation with examples that users
-can modify to deal with their problems is essential. The presence of vignettes, 
-demos, and tests indicates a package is well-supported.
-
-## Other CRAN packages considered
+**minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
+general nonlinear least-squares solvers.
+
+While we will provide capsule comments for some other CRAN packages, 
+we will not pursue those in the Bioconductor (@Bioconductor) 
+collection and on repositories such as 
+GitHub (https://github.com) and Gitlab (https://about.gitlab.com).
+
+## Some other CRAN packages for nonlinear modeling
 
@@
 `NISTnls` (@NISTnls-manual) -- This package provides R code and data for a set of
-nonlinear least squares problems from the U.S. National Institute for Standards
-and Technology for testing solvers. We caution that the problems are generally 
-poorly conditioned and may not represent real-world situations.
+(numerically ill-conditioned) nonlinear least squares problems from the U.S. National 
+Institute for Standards and Technology. These may not represent real-world situations.
 
@@
 
-`nls2` (@nls2-22) -- Nonlinear least squares by brute force has some similar
-motivation, but uses `nls()` within multiple trials. The package and its
-author have a long history in the R community.
+`nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
+motivations, but uses `nls()` within multiple trials.
 
@@
 
-`optimx` (@p-optimx) -- not really a nonlinear least squares method, but has general
-optimizers that can be applied to minimize a nonlinear function e.g., a nonlinear sum of 
-squares, as well as other objectives. This is not generally recommended if nonlinear 
-least squares programs can be easily used, but provides a way to approach problems 
-when others are unavailable.
+`optimx` (@p-optimx) -- This provides 
+optimizers that can be applied to minimize a nonlinear function which could be
+a nonlinear sum of 
+squares. NOT generally recommended if nonlinear 
+least squares programs can be easily used, but provides a check and alternative
+solvers.
 
@@
 The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
-task that is seemingly straightforward but turns out to be quite nasty. Because
-this problem has a very succinct statement, it often provides the "short 
-reproducible example" much beseeched of contributors to R mailing lists. 
-Nevertheless, this is not an artificially created test problem, but one that
-came from a field researcher.
-
-The data and a graph of it are given below.
+task that seems straightforward but is quite nasty. Its very succinct statement 
+provides the "short reproducible example" much requested on R mailing lists. 
+This is a real problem from a field researcher. The data and graph follow.
 
@@
 $p$ and the model function is called $model(p)$. The residuals can be written
-either as
-
-$$ r_1 = y - model(p)$$
-
-or
-
-$$ r = model(p) - y $$
-
-since the sumsquares loss function has an identical value. In both cases, the
-residual vector is a function of $p$. The second form
-has a minor advantage in avoiding a potential sign error 
-if we need to differentiate to get gradients etc.
-
-### The loss function and its minimization
-
-The general approach of nonlinear optimizers, of which nonlinear least squares
-solvers are usually a subset, is to try to reduce the loss function (or objective
-function) defined by the sum of squares of the residuals by 
-altering the model parameters from some starting guess. In all general 
-methods of which we are aware, this is an iterative process.
+either as $r_1 = y - model(p)$ or $r = model(p) - y$
+since their sum of squares has an identical value. The second form
+avoids a potential sign error if we need to evaluate derivatives.
+
+We wish to minimize the sum of squared residuals, which is our loss (or objective) function. 
+Starting with some guess for the parameters, we aim to alter these parameters to obtain a smaller
+loss function. We then iterate until we can make no further progress.
 
 Let us consider there are $n$ parameters and $m$ residuals. 
-We note several useful quantities
-
-Loss function:
+We note several useful quantities. The loss function is 
 $$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
-Gradient of $S(p)$:
+The gradient of $S(p)$ is
 $$ 2 * J' r$$
@@
 If we expand the Hessian for nonlinear least squares problems, we find
-
 $$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
@@
 This leads to **Newton's equation**
-
 $$ H \delta = -g $$
-given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
+Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
 to $p + \delta$ and iterate, hopefully to converge on a solution. 
-
 Applying this to a sum of squares problem gives
-
 $$ H \delta = (J'J + D) \delta = - J' r$$
@@
 that these multiplied the second partial derivatives of $r$, so he approximated
-
 $$ H \approx J' J$$
@@
 solve 
-
 $$ J' J \delta = - J' r$$
@@
 $J' J$ and solving the linear least squares matrix problem 
-
 $$ J \delta \approx -r $$
-which may be accomplished by a variety of matrix decomposition methods.
+by one of several matrix decomposition methods.
 
 In reality, there are many problems were $D$ should not be ignored, but the work
-to compute it precisely is considerable. Many workarounds have been 
+to compute it precisely is considerable. Many work-arounds have been 
 proposed, of which the Levenberg-Marquardt stabilization 
 (@Levenberg1944, @Marquardt1963) is
-the most commonly used. We will use "Marquardt", as we believe he
+the most commonly used. For convenience, we will use "Marquardt", as we believe he
 first incorporated the ideas into a practical computer program.
 
-The suggestion is that $D$ be replaced by a multiple of the unit matrix or else
+The usual suggestion is that $D$ be replaced by a multiple of the unit matrix or else
 a multiple of the diagonal part of $J' J$. In low precision, some elements of 
-$J' J$ could underflow to zero (@jn77ima), and a linear combination of both
-choices was suggested as a compromise. Various choices for
+$J' J$ could underflow to zero (@jn77ima), and one of us suggested a linear combination of both
+choices as a compromise. Various choices for
 $D$, as well as a possible line search along the
 direction $\delta$ rather than a unit step (@Hartley1961), 
-give rise to a panoply of variant
+give rise to several variant
 algorithms. "Marquardt's method" is a family of methods. Fortunately, most choices
@@
 One of the useful features of `nls()` is the possibility of a `selfStart` model,
-where starting parameter values are not required. We discuss aspects of this
-feature at several points below. However, if a `selfStart` model 
+where starting parameter values are not required. However, if a `selfStart` model 
 is not available, `nls()` sets all the starting parameters 
@@
 
-Using a formula specification was one of the principal advances made
-when `nls()` became available in S sometime in the 1980s. It uses a
-Gauss-Newton (i.e., unstabilized) iteration with step reduction line
+Using a formula specification was a principal advantage made
+with `nls()` when it became available in S sometime in the 1980s. It uses a
+Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
 search. This works very efficiently as long as $J$ is not ill-conditioned.
-In a one-parameter problem with ill-conditioning, think of $J$ as a 
-very small number so that we need to watch for a zero-divide or 0/0 
-situation in computing $\delta$. Below we see `nls()` does poorly on this 
-example problem. Note that `nlsr`
-offers function `pnlm0()` to give a 1-line display of results of class `nls`,
-and `pshort()` for `nlsr`.
+Below we see `nls()` does poorly on the example problem. To save page space,
+we use 1-line result display functions from package `nlsr`, nameley `pnlm0()` 
+and `pshort()`.
 
@@
 library(nlsr)
+library(minpack.lm)
 ```
@@
 unls1<-try(nls(formula=frmu, start=stu1, data=weeddf))
-summary(unls1)
 snls1<-try(nls(formula=frms, start=sts1, data=weeddf))
-summary(snls1)
 tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
-summary(tnls1)
 ```
@@
 Here we see the infamous "singular gradient" termination message of `nls()`.
-Questions appear quite frequently on the R-help mailing list where users
-have encountered this termination result.
-
-Let us consider three alternatives to `nls()`: `nlsr::nlxb()`, `minpack.lm::nlsLM()`
-and `gslnls::gsl_nls()`.
-We use default methods and controls in these examples.
 
@@
 ```{r ex02nlsr}
-library(nlsr)
 unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
 print(unlx1) 
-pshort(unlx1) # A short form output
 snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
-pshort(snlx1) # or print(snlx1)
+pshort(snlx1) # a short-form output
 tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
@@
 ```{r ex02minpack}
-library(minpack.lm)
-# ??not needed -- library(nlsr) # for pnlm0
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-# summary not exported by faraway??
-# faraway::summary(unlm1) # Use summary() to get display
-# unlm1 # Note the difference. Use this form to get sum of squares
 pnlm0(unlm1)  # Short form of output
@@
 pnlm0(snlm1)
-# summary(snlm1)
-## Following structure needed to deal with M1mac issue
 tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
-if (inherits(tnlm1, "try-error")) {
-   cat("Failure to compute solution -- likely singular Jacobian\n")
-} else {  
 pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
-}   
 ```
@@
 ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
-summary(ugslnls1) # Use summary() to get display
 pnlm0(ugslnls1) # to get sum of squares
@@
 
-### Comparison notes - formula setup
-
-`nlsr::nlxb()` is set up to use `print()` to output standard errors and 
+### Comparison notes for formula setup solutions
+
+`nlsr::nlxb()` uses `print()` to output standard errors and 
 singular values of the Jacobian (for diagnostic purposes). By contrast, 
@@
 Jacobian is at the solution, and the ratio of the smallest to largest of the 
-singular values is a simple but effective measure. In the above example, we note 
-that the scaled logistic has the largest ("least singular") ratio.
-
-The results from `nlsLM` and `gsl_nls` for the transformed model have a very 
+singular values is a simple but effective measure. The ratios are
+4.6641e-05 for Logistic3U, 0.021022 for Logistic3S and 0.001055 for
+Logistic3T, so Logistic3S is the "least singular".
+
+The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a very 
 large sum of squares, which may suggest that these programs have failed. 
@@
 offer the singular values, we need to extract the Jacobian and compute its
-singular values. The following script shows how to do this, using what is called
+singular values. The following script shows how to do this, using as Jacobian what is called
 the `gradient` element in the returned solution for these solvers.
@@
 two out of three singular values are essentially 0. Our Jacobian is singular, and no 
-method of the Gauss-Newton type should be able to continue. Indeed, from this set of
-parameters, `nlxb` also stops, even though the code in `nlsr` tries to find solutions
-from very poor initial parameter values.
+method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
+reported at this saddle point, `nlsr::nlxb()` cannot proceed.
 
@@
 ```{r exhobbsfn, eval=TRUE}
-require(minpack.lm); require(nlsr); require(gslnls)
 hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
-print(hobnlfb)
+pshort(hobnlfb) # use print(hobnlfb) for more detail
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
@@
 # References
+
+------------------------------
+
+## Comparison dimensions
+
+The tools for nonlinear estimation we consider have been built with different
+goals and contexts. We will make our comparison considering the overlapping
+headings summarized below.
+
+### Modeling versus minimizing a sum of squares
+
+The goal of many statisticians and other workers is to estimate a model. The
+minimization of a sum of squared residuals -- the nonlinear least squares problem --
+is the mechanism to estimate the model. Some tools focus on modeling, 
+so attempt to provide measures of uncertainty and other information. 
+The interface to modeling tools will generally try to streamline that task. By
+contrast, nonlinear least squares tools are trying to solve the more limited
+numerical task.
+
+### Programming language(s)
+
+R tools are our interest, but they often call programs in other programming
+languages. There are advantages in using a well-established program that has
+been used and tested. Early versions of R were relatively sluggish in 
+computational speed. The downside is that we must interface between two or more
+programming languages, adding to the risk of errors. Changes to one or the 
+other language or its library support may
+force maintenance issues. Moreover, the pool of expertise with languages like C
+and Fortran is diminishing.
+
+### Internal solver stabilization
+
+Early nonlinear least squares programs generally used the Gauss-Newton method 
+(@Hartley1961). This is often very effective and efficient, but for a quite large 
+proportion of problems can fail to find the solution. Stabilization of the 
+computations is relatively straightforward in principle, and overcomes most
+of the failures, though sometimes at the cost of more calculations.
+
+### Jacobian calculation
+
+The necessarily iterative solution of a nonlinear least squares problem 
+to carry out nonlinear modeling requires that we can find smaller values
+of a sum of squared residuals. This implies there is a "downhill" direction.
+The gradient of the residual sum of squares is computed as its product
+with the Jacobian matrix: the partial derivatives of the residual vector 
+with respect to the modeling parameters.
+
+Most researchers are challenged to program the differential calculus 
+required to specify the Jacobian to a nonlinear estimation tool. 
+The workaround is generally to use a numerical approximation,
+but there is the possibility of using symbolic or automatic differentiation.
+There are also particular models where workers have provided code for the
+Jacobian.
+
+### Algorithm and termination choices
+
+Iterative algorithms need to be stopped at some point, and how termination
+is forced may alter results. Furthermore, even supposedly similar algorithms
+may differ greatly in detail.
+
+### Constraints on parameters
+
+Reality imposes conditions on parameters. While general constraints can
+be difficult to include, bounds on parameters can be accommodated by many 
+solvers. For some reason, fixed parameters (masks) are not often considered
+but are convenient when a parameter has a commonly accepted value that we
+later wish to allow to be estimated from the data.
+
+### Special capabilities
+
+Users often have particular problems that introduce other requirements. It is
+common to weight observations where their reliability is thought to differ.
+Some situations present a number of related situations where a combined model
+with some common and some local parameters are present, and the local parameters
+could be indexed. The common exponential decay and related **partially linear**
+models (sometimes called **separable** models) is a difficult one that can 
+benefit from special treatment.
+
+### Output
+
+While nonlinear estimation aims to return a set of parameters for a
+particular model, there is a variety of additional information which may be
+provided. From the perspective of the nonlinear least squares problem, there
+are measures of work and information on the quality of the solution, while
+for statistical purposes measures of uncertainty in the parameters are helpful.
+
+Some tools allow constraints to be imposed on the solution, and it is helpful
+to know if these constraints are active in the proposed solution.
+
+### Documentation
+
+We believe that the availability of good documentation with examples that users
+can modify to deal with their problems is essential. The presence of vignettes, 
+demos, and tests indicates a package is well-supported.

### 2023-05-09 RefactoringNLS/ComparingNLStools.Rmd
finish first pass edit to shorten

@@
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnlm0(hobnlm)  
+pnlm1(hobnlm)  
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
@@
 
-The design goals of the different tools are also evident in the so-called
+The design goals of the different tools may also be revealed in the so-called
 "convergence tests" for the iterative solvers. In the manual page for `nls()` 
@@
 
-with suggested addition of small perturbations to the data. This amounted to 
-admitting `nls()` cannot solve well-posed problems unless data is polluted
-with errors. We dispute that zero-residual problems are always artificial,
+with suggested addition of small perturbations to the data. This admits 
+`nls()` could not solve well-posed problems unless data is polluted
+with errors. Zero-residual problems are not always artificial,
 since problems in function approximation and nonlinear equations can be 
-approached with nonlinear least squares. Moreover, the matter 
-can be easily resolved. The "termination test" for the **program** rather than for 
-"convergence" of the underlying **algorithm** is the Relative Offset Convergence
-Criterion (see @BatesWatts81). This projects the proposed step in the parameter
-vector onto the gradient and estimates how much the sum of squares loss function 
-should decrease. This estimate is divided 
-by the current size of the loss function to avoid scale issues.
-When we have "converged", the estimated relative
-decrease is very small. However, with small residuals,
-the sum of squares loss function is (almost) zero and we get the possibility of a 
-zero-divide failure.
-
-Adding a small quantity to the loss function before dividing avoids trouble. 
+approached with nonlinear least squares. Fortunately, a small adjustment
+to the "termination test" for the **program**, rather than for 
+"convergence" of the underlying **algorithm**, fixes the defect. The test
+is the Relative Offset Convergence
+Criterion (see @BatesWatts81). This scales an estimated reduction in the loss
+function by its current value. If the loss function is very small, we are close
+to a zero-divide. Adding a small quantity to the divisor avoids trouble. 
 In 2021, one of us (J. Nash) proposed that `nls.control()` have an 
-additional parameter `scaleOffset` with a default value of zero. Setting it to a small 
-number -- 1.0 is a reasonable choice -- allows small-residual problems 
+additional parameter `scaleOffset` with a default value of zero. Setting it to a 
+small number -- 1.0 is a reasonable choice -- allows small-residual problems 
 (i.e., near-exact fits) to be dealt with easily. We call this the
-**safeguarded relative offset convergence criterion**. The default value 
-gives the legacy behaviour.
-We are pleased to note that this improvement is now in the R distributed code 
-since version 4.1.0; it has been a part of `nlsr` since it was introduced.
-
-More general termination tests can be used. `nlsr` has a **small sum of 
+**safeguarded relative offset convergence criterion**, and it has been in `nlsr`
+since it was introduced. The default value 
+gives the legacy behaviour. This improvement has been in the R distributed code 
+since version 4.1.0.
+
+Additional termination tests can be used. `nlsr` has a **small sum of 
 squares** test (**smallsstest**) that compares the latest evaluated sum of squared
@@
 `nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
-examination of the code. We prefer to use the number of times the residuals or 
-the jacobian have been computed and put upper limits on these. Our codes exit (terminate) 
-when these limits are reached. Generally we prefer larger limits than the 
+examination of the code. We terminate execution when the number of residual or 
+jacobian evaluations exceed set limits. Generally we prefer larger limits than the 
 default `maxiter = 50` of `nls()` to avoid stopping early.
@@
 
-?? Can we simplify or suggest use of `str()` ??
-
-The output of `nls()` is an object of class "nls" which has the following structure:
-
->A list of
-
->`m` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp;  an `nlsModel` object incorporating the model.
-
->`data`	&nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;the expression that was passed to `nls` as the data argument. The actual \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$  data values are present in the environment of the `m` components, e.g., \newline $~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ `environment(m$conv)`.
-
->`call` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the matched call with several components, notably `algorithm`.
-
->`na.action` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the `"na.action"` attribute (if any) of the model frame.
-
->`dataClasses` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; $~$the `"dataClasses"` attribute (if any) of the "`terms`" attribute of the \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$ model frame.
-
->`model` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp;if `model = TRUE`, the model frame.
-
->`weights` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; if `weights` is supplied, the weights.
-
->`convInfo` &nbsp; &nbsp;  &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp;&nbsp; a list with convergence information.
-
->`control` &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; the control `list` used, see the `control` argument.
-
->`convergence, message` &nbsp;for an `algorithm = "port"` fit only, a convergence code (`0` for  \newline
-$~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~$convergence) and message.
-
-> &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; &nbsp; To use these is *deprecated*, as they are available from `convInfo` now.
-
-                  -----------------------------------------------------------
+The output of `nls()` is an object of class "nls" which has quite rich structure
+described in the manual file or revealed by applying the `str()` function to the
+result of `nls()`.
 
@@
 actually access the data from this element, we need to use the syntax:
-
 ```
@@
 ```
-
 However, if the call is made with `model=TRUE`, then there is a returned element
 `model` which contains the data, and we can list its contents using:
-
 ```
@@
 ```
-
 and if there is an element called `xdata` that can be accessed as
@@
 
-By contrast `nlsr::nlxb()` returns
-
->`coefficients`    A named vector giving the parameter values at the supposed solution.
-
->`ssquares` $~~~~~$ The sum of squared residuals at this set of parameters.
-
->`resid` $~~~~~~~~~~$ The residual vector at the returned parameters.
-
->`jacobian` $~~~~~$ The jacobian matrix (partial derivatives of residuals w.r.t. the 
-                parameters) at the \newline
-$~~~~~~~~~~~~~~~~~~~~$ returned parameters.
-
->`feval` $~~~~~~~~~~$ The number of residual evaluations (sum of squares computations) used.
-
->`jeval` $~~~~~~~~~~$ The number of Jacobian evaluations used.
-
-However, actually looking at the structure of a returned result gives a list of 11
-items. The extra 5 are:
-
-```
- $ lower       : num [1:3] -Inf -Inf -Inf
- $ upper       : num [1:3] Inf Inf Inf
- $ maskidx     : int(0) 
- $ weights     : NULL
- $ formula     :Class 'formula'  language y ~ Asym/(1 + exp((xmid - tt)/scal))
-  .. ..- attr(*, ".Environment")=<environment: R_GlobalEnv> 
- - attr(*, "class")= chr "nlsr"
-```
-The result object from `nlsr::nlxb()` is still much smaller than the one `nls()` 
-returns. Moreover, `nlxb`
-explicitly returns the sum of squares as well as the residual vector and Jacobian. 
-The counts of evaluations are also returned.
+By contrast `nlsr::nlxb()` returns a much simpler structure of 11 items
+in one-level. Moreover, `nlxb` explicitly returns the sum of squares, the residual vector, Jacobian and counts of evaluations.
 
@@
 Using the "gradient" attribute of the output of the Jacobian function to hold
-this matrix lets us embed this attribute in the **residual** function as well,
+the Jacobian matrix lets us embed this in the **residual** function as well,
 so that the call to `nlfb()` can be made with the same name used for 
@@
 parameter estimation is carried out within the called C-language program
-and its wrapper R code function `numericDeriv()`, part of `./src/library/stats/R/nls.R`
+and its wrapper R code function `numericDeriv()`, part of
+`./src/library/stats/R/nls.R`
 in the R distribution source code. This is used to provide Jacobian information in 
@@
 the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, though in our experience these differences are usually
-very minor.
-
-However, minor numerical differences can give annoying results with ill-conditioned
-problems.
+can lead to different behaviour, usually minor, but sometimes annoying with
+ill-conditioned problems.
 
@@
   derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
-  the derivative is undefined, while the model can be computed. In such cases, 
+  the derivative is undefined, but the model can be computed. In such cases, 
   we observed that `nls()` and 
@@
 
-Analytic Jacobian code can be provided to `nls()`, `nlsLM()` and `gsl_nls()`,
-and most `selfStart` models that automatically provide starting parameters
+Analytic Jacobian code can be provided to all the solvers discussed.
+Most `selfStart` models that automatically provide starting parameters
 also include such code. There is documentation in R
-of `selfStart` models, but the construction of them is non-trivial. A number 
+of `selfStart` models, but their construction is non-trivial. A number 
 of such models are included with base R in `./src/library/stats/R/zzModels.R`,  
@@
 These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
-that defines
-each model, and these Jacobians may be the analytic forms.
+that defines each model, and these Jacobians are often the analytic forms.
 
 The `nls()` function, after
-computing the "right hand side" or `rhs` of the residual, then checks to see if the
-"gradient" attribute is defined, and, if not, uses `numericDeriv()` to compute a 
-Jacobian
-into that attribute. This code is within the `nlsModel()` or `nlsModel.plinear()`
-functions. The use of analytic Jacobians
+computing the "right hand side" or `rhs` of the residual, checks to see if the
+"gradient" attribute is defined, and, otherwise using `numericDeriv()` to compute a 
+Jacobian into that attribute. This code is within the `nlsModel()` or
+`nlsModel.plinear()` functions. The use of analytic Jacobians
 almost certainly contributes to the good performance of `nls()` on `selfStart` 
@@
 it should be pointed out that our work is primarily diagnostic. If we were carrying
-out a large number of similar estimations, then good initial parameters are 
+out a large number of similar estimations, such initial parameters are 
 critical to efficiency.
@@
 In considering `selfStart` models, we noted that the base-R function `SSlogis`
-is intended to solve our example problem **Logistic3T**.  This example is one 
-where the `getInitial()` function (used to find
-starting values) actually calls `nls()` recursively with a non-default algorithm
-to get "starting" parameters. Then it returns to running the default algorithm
-unnecessarily. Such circular references are, in our view, prone to creating errors
-in code maintenance. 
+is intended to solve problem **Logistic3T** above.  When this function is used
+via `getInitial()` to  find
+starting values, it actually calls `nls()` with the 'plinear' algorithm 
+and finds a solution. It then passes the solution coefficients to the default 
+algorithm unnecessarily. Moreover, the double call is, in our view, prone to creating 
+errors in code maintenance. 
 To provide simpler starting parameters, the function `SSlogisJN`
@@
 
-Users may also want to use `selfStart` models with explicit starting values 
-other than those suggested by `getInitial()`. 
+Users may also want to profit from the Jacobian code of `selfStart` models 
+but supply explicit starting values other than those suggested by `getInitial()`. 
+This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
+starting parameters, and uses `getInitial()' to find them from the selfStart
+model. The same model is used in the formula for the `nlxb()` call.
+
+<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym, xmid, scale), start=stt1, data=weeddf)) -->
+<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym=1, xmid=1, scale=1),  data=weeddf)) -->
+<!-- tssx <-try(nls(weed ~ SSlogis(tt, Asym, xmid, scale),  data=weeddf))summary(tssx) -->
+
 We are also surprised that the 
 analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
-and others save many quantities in "hidden"
+and others save quantities in "hidden"
 variables, i.e., with names preceded by ".". 
@@
 Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
-worth considering as a way to find starting parameters, but we know of no
+worth considering as another way to find starting parameters, but we know of no
 R capability of this type.
@@
 In some cases, we know that parameters cannot be bigger or smaller than some
-externally known limits. Cash reserves cannot be negative. Animals have a 
-minimum need for water. Airplanes cannot carry more than a known or 
-legislated cargo weight. Such limits can be built into models, but there 
+externally known limits. Such limits should be built into models, but there 
 are some important details for using the tools in R.
@@
    we believe the authors have not taken into account all possibilities, though
-   it should be noted that all programs have some weakness regarding constrained
-   optimization. Software creators have to work with
-   assumptions on the extremity of scale that they are willing to countenance,
-   and sometimes problems will be outside the scope envisaged.
+   all programs have some weakness regarding constrained optimization in that
+   programmers have to work with assumptions on the scale of numbers, 
+   and some problems will be outside the scope envisaged.
 
@@
 
-Bounds on parameters raise some interesting and difficult questions about how 
+Bounds on parameters raise some interesting questions about how 
 uncertainty in parameter estimates should be computed or reported. That is, the
@@
 estimates. For `nlsr` we have **chosen** to suppress calculation of 
-approximate standard 
-errors in the parameters when constraints are active because their meaning
-under constraints is unclear, though we believe this 
+approximate standard errors in the parameters when constraints are active because
+their meaning under constraints is unclear, though we believe this 
 policy worthy of discussion and further investigation.
@@
 while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
-offers an accelerated Marquardt method and three alternative methods. We have 
-not explored all five methods with various control settings.
+offers an accelerated Marquardt method and three alternative methods which we
+have not investigated.
 Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
-algorithm variants.
-
-In general, the Levenberg-Marquardt stabilization is important in obtaining 
+algorithm variants. In general, the Levenberg-Marquardt stabilization 
+is important in obtaining 
 solutions in methods of the Gauss-Newton family, as `nls()` terminates 
@@
 
-`nls()` and other class `nls` tools accept an argument `subset`. This acts through the 
-mediation of `model.frame`, which not obvious in the 
+`nls()` and other class `nls` tools accept an argument `subset`. This acts through 
+the mediation of `model.frame`, which not obvious in the 
 source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`. 
@@
 obviously in the source code, often being handled behind the scenes after 
-referencing the option `na.action`. Yet this feature also changes the data 
+referencing the option `na.action`, yet this feature also changes the data 
 supplied to our nonlinear least squares solver. 
-
 A useful, but possibly dated, description is given in:
@@
 containing missing values (i.e. any row of a data frame containing an NA). 
-`na.exclude` does much of the same for solver computations, but keeps the rows with NA 
-elements so that predictions are in the correct row position. We recommend that 
+`na.exclude` does much of the same for solver computations, but keeps the rows with 
+NA elements so that predictions are in the correct row position. We recommend that 
 workers actually test output to verify the behaviour is as wanted.
 See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
-
 As with `subset`, our concern with `na.action` is that users may be unaware of the
@@
 name could
-be confusing.) Despite this, the model frame is used within the function code in the form 
-of the object
+be confusing.) Despite this, the model frame is used within the function code in 
+the form of the object
 `mf`. We feel that users could benefit from more extensive documentation and 
@@
 allows us to compute residuals for particular coefficient sets.
-
-<!-- ?? Arkajyoti: I think this example may get too long or be too unclear. -->
 
@@
 
-## Minor complaints
+## Minor issues with nonlinear least-squares tools
 
@@
 problems when some of the parameters appear linearly. In our logistic examples, 
-the asymptote parameters are an illustration.However, setting up the 
-calculations, that is, identifying which parameters are linear, is not a trivial
-task.
+the asymptote parameters are an illustration. However, identifying which parameters
+are linear and communicating this information to estimating functions 
+is not a trivial task.
 
@@
 
-Indexed parameters -- those whose names have subscripts on some element that
-is part of the coefficient list -- 
-are a useful idea to simplify model specification in some cases. However, they
-are also a focus of trouble. Typically we have some common parameters and 
+Some models have several common parameters and 
 others that are tied to particular cases. 
-
-The **man** file for `nls()` includes the following example of a situation in which
-parameters are indexed. It also uses the "plinear" option as an added complication.
-Here we use a truncated version of the example to save display space.
-
-```{r nlsindx1}
-## The muscle dataset in MASS is from an experiment on muscle
-## contraction on 21 animals.  The observed variables are Strip
-## (identifier of muscle), Conc (Cacl concentration) and Length
-## (resulting length of muscle section).
-if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg")
-mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE)
-mm<-droplevels(mm) # remove unused levels after truncation
-nlev <- nlevels(mm)
-withAutoprint({
-  ## The non linear model considered is
-  ##       Length = alpha + beta*exp(-Conc/theta) + error
-  ## where theta is constant. For now alpha and beta do NOT vary with Strip.
-  with(mm, table(Strip)) # 2, 3 or 4 obs per strip
-  nl <- nlevels(mm$Strip)
-  ## We first use the plinear algorithm to fit an overall model,
-  ## ignoring that alpha and beta might vary with Strip.
-  musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm,
-                start = list(th = 1), algorithm = "plinear")
-  summary(musc.1)
-
-  ## Then we use nls' indexing feature for parameters in non-linear
-  ## models to use the conventional algorithm to fit a model in which
-  ## alpha and beta vary with Strip.  The starting values are provided
-  ## by the previously fitted model.
-  ## Note that with indexed parameters, the starting values must be
-  ## given in a list (with names):
-  b <- coef(musc.1)
-  musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm,
-                start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1]))
-  summary(musc.2)
-})
-```
-
-Running the example reveals that the answers for the parameters are NOT indexed. 
+The **man** file for `nls()` includes an example of a situation in which
+parameters are indexed but which uses the "plinear" option as an added complication.
+Running this example reveals that the answers for the parameters are NOT indexed
+as in a vector. 
 <!-- e.g. `coef(musc.2)` is a single level vector of parameters.  -->
@@
 <!-- must integrate all the parameters because `th` is a common parameter across the index `Strip`. -->
-This is no doubt because programming for indexed parameters is extremely challenging.
+This is no doubt because programming for indexed parameters is challenging.
 <!-- However, we believe the current structure could cause confusion and error, and  -->
 <!-- propose an alternative approach below. -->
-
-If such models are of enough importance, we believe it would be relatively
-straightforward to prepare an external package to provide for their estimation,
-including an appropriately formatted output object so the indexing of the 
-parameters can be used in other calculations.
+We note that there are capabilities in packages for mixed-effects modelling
+such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (p-lme4) for 
+estimating models of such type.
 
@@
 
+<!-- ===```{r nlsindx1} -->
+<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
+<!-- ## contraction on 21 animals.  The observed variables are Strip -->
+<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
+<!-- ## (resulting length of muscle section). -->
+<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
+<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE) -->
+<!-- mm<-droplevels(mm) # remove unused levels after truncation -->
+<!-- nlev <- nlevels(mm) -->
+<!-- withAutoprint({ -->
+<!--   ## The non linear model considered is -->
+<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
+<!--   ## where theta is constant. For now alpha and beta do NOT vary with Strip. -->
+<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
+<!--   nl <- nlevels(mm$Strip) -->
+<!--   ## We first use the plinear algorithm to fit an overall model, -->
+<!--   ## ignoring that alpha and beta might vary with Strip. -->
+<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
+<!--                 start = list(th = 1), algorithm = "plinear") -->
+<!--   summary(musc.1) -->
+
+<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
+<!--   ## models to use the conventional algorithm to fit a model in which -->
+<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
+<!--   ## by the previously fitted model. -->
+<!--   ## Note that with indexed parameters, the starting values must be -->
+<!--   ## given in a list (with names): -->
+<!--   b <- coef(musc.1) -->
+<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
+<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
+<!--   summary(musc.2) -->
+<!-- }) -->
+<!-- === ``` -->
+
 ## Tests and use-case examples
 
-Maintainers of packages need suitable tests and use-case examples in order:
-
+Maintainers of packages need suitable tests and use-case examples in order
 - to ensure packages work properly, in particular, giving
-  results comparable to or better than the functions they are to replace;
+results comparable to or better than the functions they are to replace. 
 - to test individual solver functions to ensure they work across the range of 
@@
 and the conventional R package testing tools).
-
 In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
@@
 
-## Should nls() be refactored?
-
-When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted:
-
-- to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R 
-  that matches the features in base R or what is packaged in `nlspkg` as described 
-  in @PkgFromRbase22. Our intent in providing a purely R version of the code was
-  either to replace `nls()` or to serve as a reference version aiding maintenance 
-  of C, Fortran, or other components of the software.
-
-- to streamline the overall `nls()` infrastructure. By this we mean a 
-  refactoring of the routines so they are better suited to maintenance of both
-  the existing `nls()` methods and features as well as new features we or others
-  would like to add.
-
-- to explain what we do, either in comments, examples, or separate maintainer 
-  documentation. While R and its packages are generally very well documented for
-  usage, that documentation
-  rarely extends as far as it might for code maintenance. The paucity of such
-  documentation is exacerbated by the mixed R/C/Fortran code base,
-  where seemingly trivial differences in things like representation of numbers or
-  base index for arrays lead to traps for unwary programmers.
-
-These goals echo themes in the recent discussion by @Vidoni21 and accompanying 
-commentary.
-
-We found the legacy code and examples prevented us achieving a revised `nls()` 
-with all its current capabilities. Unfortunately, `nls()` is part of the base R
-distribution, 
-so users are drawn to using it even when it is less than ideal. A possible way forward
-is to
-
-- use a packaged version of `nls()` to provide legacy features, with notification
-  to users that there are other tools available
-
-- if maintainers are willing, to merge and rationalize the capabilities of 
-  `minpack.lm`, `nlsr` and `gslnls`
-
-- to encourage the (separate) development of capabilities such as indexed
-  parameter models if these are in demand.
-
-This direction does not, of course, resolve the inconsistencies in model 
-specification in formulas for nonlinear models, which undoubtedly is an issue
-for other modeling tools.
-
-A possible workaround to allow consistency for the `plinear` algorithm 
-would be to specify the linear parameters when the `plinear` option is specified. 
-For example, we could use `algorithm="plinear(Asym)"` while keeping the general 
-model formula from the default specification. This would allow 
-for the output of different `algorithm` options to be consistent. However, we 
-have not yet tried to code such a change, and welcome discussion from those working
-with other packages using model formulas.
-
-A further complication of model specification in R is that some formulas require
-the user to surround a term with `I()` to inhibit interpretation of arithmetic 
-operators. We would prefer that formulas be usable without this complication.
+<!-- ## Should nls() be refactored? -->
+
+<!-- When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted: -->
+
+<!-- - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R  -->
+<!--   that matches the features in base R or what is packaged in `nlspkg` as described  -->
+<!--   in @PkgFromRbase22. Our intent in providing a purely R version of the code was -->
+<!--   either to replace `nls()` or to serve as a reference version aiding maintenance  -->
+<!--   of C, Fortran, or other components of the software. -->
+
+<!-- - to streamline the overall `nls()` infrastructure. By this we mean a  -->
+<!--   refactoring of the routines so they are better suited to maintenance of both -->
+<!--   the existing `nls()` methods and features as well as new features we or others -->
+<!--   would like to add. -->
+
+<!-- - to explain what we do, either in comments, examples, or separate maintainer  -->
+<!--   documentation. While R and its packages are generally very well documented for -->
+<!--   usage, that documentation -->
+<!--   rarely extends as far as it might for code maintenance. The paucity of such -->
+<!--   documentation is exacerbated by the mixed R/C/Fortran code base, -->
+<!--   where seemingly trivial differences in things like representation of numbers or -->
+<!--   base index for arrays lead to traps for unwary programmers. -->
+
+<!-- These goals echo themes in the recent discussion by @Vidoni21 and accompanying  -->
+<!-- commentary. -->
+
+<!-- We found the legacy code and examples prevented us achieving a revised `nls()`  -->
+<!-- with all its current capabilities. Unfortunately, `nls()` is part of the base R -->
+<!-- distribution,  -->
+<!-- so users are drawn to using it even when it is less than ideal. A possible way forward -->
+<!-- is to  -->
+
+<!-- - use a packaged version of `nls()` to provide legacy features, with notification -->
+<!--   to users that there are other tools available -->
+
+<!-- - if maintainers are willing, to merge and rationalize the capabilities of  -->
+<!--   `minpack.lm`, `nlsr` and `gslnls` -->
+
+<!-- - to encourage the (separate) development of capabilities such as indexed -->
+<!--   parameter models if these are in demand. -->
+
+<!-- This direction does not, of course, resolve the inconsistencies in model  -->
+<!-- specification in formulas for nonlinear models, which undoubtedly is an issue -->
+<!-- for other modeling tools.   -->
+
+<!-- A possible workaround to allow consistency for the `plinear` algorithm  -->
+<!-- would be to specify the linear parameters when the `plinear` option is specified.  -->
+<!-- For example, we could use `algorithm="plinear(Asym)"` while keeping the general  -->
+<!-- model formula from the default specification. This would allow  -->
+<!-- for the output of different `algorithm` options to be consistent. However, we  -->
+<!-- have not yet tried to code such a change, and welcome discussion from those working -->
+<!-- with other packages using model formulas. -->
+
+<!-- A further complication of model specification in R is that some formulas require -->
+<!-- the user to surround a term with `I()` to inhibit interpretation of arithmetic  -->
+<!-- operators. We would prefer that formulas be usable without this complication. -->
 
@@
 
-It is possible that `nls()` will remain more or less as it has been for the past
-two decades. Given its importance to R, the focus of discussion should be the
-measures needed to secure its continued operation for legacy purposes and how that
-may be approached. We welcome an opportunity to participate in such conversations.
+Given its importance to R, it is possible that `nls()` will remain more or less 
+as it has been for the past several decades. If so, the focus of discussion 
+should be the measures needed to secure its continued operation for legacy 
+purposes and how that may be accomplished. 
+We welcome an opportunity to participate in such conversations.
 
@@
 cross-linkages and side-effects.
-The aspects of `nls()` that have given us the most trouble follow:
-
-- R tools often presume data and parameters needed are available in an 
-  accessible environment. As long as the correct environment is used, this provides 
-  a compact syntax to invoke the
-  calculations. The danger is that the wrong data is accessed if the 
-  internal search finds a
-  valid name that is not the object we want, or if `subset` or `na.action` settings modify
-  the selection, or `weights` are applied.
-
-- Mixing of R, C and Fortran code adds to the burden of following the program logic.
+The aspects of `nls()` that concern us follow.
+
+- R tools presume data and parameters needed are available in an 
+  accessible environment. This provides a compact syntax to invoke the
+  calculations, but the wrong data can be used if the 
+  internal search finds a valid name that is not the object we want, 
+  or if `subset` or `na.action` settings modify the selection, or 
+  `weights` are applied.
+
+- Mixing of R, C and Fortran code adds to the burden of following the 
+  program logic.
 
 - The class `nls` structure simplifies calls with its rich set of functions, but
-  also adds to the task of following what has been done. A design that isolates the 
-  setup, solution, and post-solution parts of complicated calculations reduces the 
-  number of objects that must be kept in alignment.
+  also adds to the task of understanding what has been done. A design that isolates
+  the setup, solution, and post-solution parts of complicated calculations reduces 
+  the number of objects that must be kept in alignment.
 
 Given the existence of examples of good practices such as analytic derivatives, 
-stabilized 
-solution of Gauss-Newton equations and bounds-constrained parameters, base R tools should 
+stabilized solution of Gauss-Newton equations and bounds-constrained parameters, 
+base R tools should 
 be moving to incorporate them. These capabilities are available now by 
@@
 
-------------------------------
-
-## Comparison dimensions
-
-The tools for nonlinear estimation we consider have been built with different
-goals and contexts. We will make our comparison considering the overlapping
-headings summarized below.
-
-### Modeling versus minimizing a sum of squares
-
-The goal of many statisticians and other workers is to estimate a model. The
-minimization of a sum of squared residuals -- the nonlinear least squares problem --
-is the mechanism to estimate the model. Some tools focus on modeling, 
-so attempt to provide measures of uncertainty and other information. 
-The interface to modeling tools will generally try to streamline that task. By
-contrast, nonlinear least squares tools are trying to solve the more limited
-numerical task.
-
-### Programming language(s)
-
-R tools are our interest, but they often call programs in other programming
-languages. There are advantages in using a well-established program that has
-been used and tested. Early versions of R were relatively sluggish in 
-computational speed. The downside is that we must interface between two or more
-programming languages, adding to the risk of errors. Changes to one or the 
-other language or its library support may
-force maintenance issues. Moreover, the pool of expertise with languages like C
-and Fortran is diminishing.
-
-### Internal solver stabilization
-
-Early nonlinear least squares programs generally used the Gauss-Newton method 
-(@Hartley1961). This is often very effective and efficient, but for a quite large 
-proportion of problems can fail to find the solution. Stabilization of the 
-computations is relatively straightforward in principle, and overcomes most
-of the failures, though sometimes at the cost of more calculations.
-
-### Jacobian calculation
-
-The necessarily iterative solution of a nonlinear least squares problem 
-to carry out nonlinear modeling requires that we can find smaller values
-of a sum of squared residuals. This implies there is a "downhill" direction.
-The gradient of the residual sum of squares is computed as its product
-with the Jacobian matrix: the partial derivatives of the residual vector 
-with respect to the modeling parameters.
-
-Most researchers are challenged to program the differential calculus 
-required to specify the Jacobian to a nonlinear estimation tool. 
-The workaround is generally to use a numerical approximation,
-but there is the possibility of using symbolic or automatic differentiation.
-There are also particular models where workers have provided code for the
-Jacobian.
-
-### Algorithm and termination choices
-
-Iterative algorithms need to be stopped at some point, and how termination
-is forced may alter results. Furthermore, even supposedly similar algorithms
-may differ greatly in detail.
-
-### Constraints on parameters
-
-Reality imposes conditions on parameters. While general constraints can
-be difficult to include, bounds on parameters can be accommodated by many 
-solvers. For some reason, fixed parameters (masks) are not often considered
-but are convenient when a parameter has a commonly accepted value that we
-later wish to allow to be estimated from the data.
-
-### Special capabilities
-
-Users often have particular problems that introduce other requirements. It is
-common to weight observations where their reliability is thought to differ.
-Some situations present a number of related situations where a combined model
-with some common and some local parameters are present, and the local parameters
-could be indexed. The common exponential decay and related **partially linear**
-models (sometimes called **separable** models) is a difficult one that can 
-benefit from special treatment.
-
-### Output
-
-While nonlinear estimation aims to return a set of parameters for a
-particular model, there is a variety of additional information which may be
-provided. From the perspective of the nonlinear least squares problem, there
-are measures of work and information on the quality of the solution, while
-for statistical purposes measures of uncertainty in the parameters are helpful.
-
-Some tools allow constraints to be imposed on the solution, and it is helpful
-to know if these constraints are active in the proposed solution.
-
-### Documentation
-
-We believe that the availability of good documentation with examples that users
-can modify to deal with their problems is essential. The presence of vignettes, 
-demos, and tests indicates a package is well-supported.
+<!-- ------------------------------ -->
+
+<!-- ## Comparison dimensions -->
+
+<!-- The tools for nonlinear estimation we consider have been built with different -->
+<!-- goals and contexts. We will make our comparison considering the overlapping -->
+<!-- headings summarized below. -->
+
+<!-- ### Modeling versus minimizing a sum of squares -->
+
+<!-- The goal of many statisticians and other workers is to estimate a model. The -->
+<!-- minimization of a sum of squared residuals -- the nonlinear least squares problem -- -->
+<!-- is the mechanism to estimate the model. Some tools focus on modeling,  -->
+<!-- so attempt to provide measures of uncertainty and other information.  -->
+<!-- The interface to modeling tools will generally try to streamline that task. By -->
+<!-- contrast, nonlinear least squares tools are trying to solve the more limited -->
+<!-- numerical task. -->
+
+<!-- ### Programming language(s) -->
+
+<!-- R tools are our interest, but they often call programs in other programming -->
+<!-- languages. There are advantages in using a well-established program that has -->
+<!-- been used and tested. Early versions of R were relatively sluggish in  -->
+<!-- computational speed. The downside is that we must interface between two or more -->
+<!-- programming languages, adding to the risk of errors. Changes to one or the  -->
+<!-- other language or its library support may -->
+<!-- force maintenance issues. Moreover, the pool of expertise with languages like C -->
+<!-- and Fortran is diminishing. -->
+
+<!-- ### Internal solver stabilization -->
+
+<!-- Early nonlinear least squares programs generally used the Gauss-Newton method  -->
+<!-- (@Hartley1961). This is often very effective and efficient, but for a quite large  -->
+<!-- proportion of problems can fail to find the solution. Stabilization of the  -->
+<!-- computations is relatively straightforward in principle, and overcomes most -->
+<!-- of the failures, though sometimes at the cost of more calculations. -->
+
+<!-- ### Jacobian calculation -->
+
+<!-- The necessarily iterative solution of a nonlinear least squares problem  -->
+<!-- to carry out nonlinear modeling requires that we can find smaller values -->
+<!-- of a sum of squared residuals. This implies there is a "downhill" direction. -->
+<!-- The gradient of the residual sum of squares is computed as its product -->
+<!-- with the Jacobian matrix: the partial derivatives of the residual vector  -->
+<!-- with respect to the modeling parameters. -->
+
+<!-- Most researchers are challenged to program the differential calculus  -->
+<!-- required to specify the Jacobian to a nonlinear estimation tool.  -->
+<!-- The workaround is generally to use a numerical approximation, -->
+<!-- but there is the possibility of using symbolic or automatic differentiation. -->
+<!-- There are also particular models where workers have provided code for the -->
+<!-- Jacobian. -->
+
+<!-- ### Algorithm and termination choices -->
+
+<!-- Iterative algorithms need to be stopped at some point, and how termination -->
+<!-- is forced may alter results. Furthermore, even supposedly similar algorithms -->
+<!-- may differ greatly in detail.  -->
+
+<!-- ### Constraints on parameters -->
+
+<!-- Reality imposes conditions on parameters. While general constraints can -->
+<!-- be difficult to include, bounds on parameters can be accommodated by many  -->
+<!-- solvers. For some reason, fixed parameters (masks) are not often considered -->
+<!-- but are convenient when a parameter has a commonly accepted value that we -->
+<!-- later wish to allow to be estimated from the data. -->
+
+<!-- ### Special capabilities -->
+
+<!-- Users often have particular problems that introduce other requirements. It is -->
+<!-- common to weight observations where their reliability is thought to differ. -->
+<!-- Some situations present a number of related situations where a combined model -->
+<!-- with some common and some local parameters are present, and the local parameters -->
+<!-- could be indexed. The common exponential decay and related **partially linear** -->
+<!-- models (sometimes called **separable** models) is a difficult one that can  -->
+<!-- benefit from special treatment.  -->
+
+<!-- ### Output  -->
+
+<!-- While nonlinear estimation aims to return a set of parameters for a -->
+<!-- particular model, there is a variety of additional information which may be -->
+<!-- provided. From the perspective of the nonlinear least squares problem, there -->
+<!-- are measures of work and information on the quality of the solution, while -->
+<!-- for statistical purposes measures of uncertainty in the parameters are helpful. -->
+
+<!-- Some tools allow constraints to be imposed on the solution, and it is helpful -->
+<!-- to know if these constraints are active in the proposed solution. -->
+
+<!-- ### Documentation -->
+
+<!-- We believe that the availability of good documentation with examples that users -->
+<!-- can modify to deal with their problems is essential. The presence of vignettes,  -->
+<!-- demos, and tests indicates a package is well-supported. -->

### 2023-05-11 RefactoringNLS/ComparingNLStools.Rmd
minor update in editing

@@
 ---
-title: "A comparison on R tools for nonlinear least squares modeling"
+title: "A comparison of R tools for nonlinear least squares modeling"
 author: 
@@
         keep_tex: true
-        toc: true
+        toc: false
 bibliography: ./ImproveNLS.bib
@@
 
-In a Google Summer of Code project "Improvements to `nls()`",
-we considered the features and limitations of the R function `nls()`
-in the context of improving and rationalizing R tools for nonlinear regression.
-The rich features of this function are weakened by several deficiencies and
-inconsistencies, in particular a lack of stabilization of the Gauss-Newton solver.
+Our Google Summer of Code project "Improvements to `nls()`" considered the features and limitations of the R function `nls()` with the aim 
+of improving and rationalizing R tools for nonlinear regression.
+The rich features of `nls()` are weakened by several deficiencies and
+inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
 Further considerations are the usability and  maintainability of
@@
 Various packages, including our `nlsr`, provide alternative capabilities.
-This article considers the differences in goals, approaches, and features of
+We consider the differences in goals, approaches, and features of
 different tools for nonlinear least squares modeling in R. Discussion of 
@@
 
-`nls()` is the tool in base R, the primary software distribution
-from the Comprehensive R Archive Network (*https://cran.r-project.org*),
-for estimating nonlinear statistical models. The function dates to the 1980s and the 
-work related to @bateswatts in S (see 
+`nls()` is the tool for estimating nonlinear statistical models in base R, the 
+primary software distribution
+from the Comprehensive R Archive Network (*https://cran.r-project.org*).
+The function dates to the 1980s and the work related to @bateswatts in S (see 
 *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
@@
 we will not pursue those in the Bioconductor (@Bioconductor) 
-collection and on repositories such as 
+collection, nor those on repositories such as 
 GitHub (https://github.com) and Gitlab (https://about.gitlab.com).

### 2023-05-11 RefactoringNLS/ComparingNLStools.Rmd
JN final edited and reduced

@@
 https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
-ideas, with references to the literature. Our view is that the approach needs wider 
-discussion and tutorial examples to allow its merits to be judged.
+ideas, with references to the literature. The approach needs a wider 
+discussion and tutorial examples to allow its merits to be judged than can be
+included here.
 
@@
 than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
-on the extreme NIST examples (@NISTnls-manual).
+on the extreme NIST examples mentioned next.
 
@@
 equality and/or inequality constraints. It is clearly **not** about modeling, and
-the input and output are quite different from class `nls` methods.
+the input and output are quite different from class `nls` methods. However, there do
+not appear to be other R packages with these capabilities.
 
@@
 dedicated  to predictive microbiology, including a vignette, by authors of the
-`nlstools` package who have a strong record in the subject.
+`nlstools` package.
 
@@
 `nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
-motivations, but uses `nls()` within multiple trials.
+motivations to `nls.multstart`, but uses `nls()` within multiple trials. The author
+has extensive expertise in R.
 
@@
 
-`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models. Most include
-analytic Jacobian code.
+`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models, primarily from
+agriculture. Most include analytic Jacobian code.
 
@@
 
+where we will use `weed` for $y$ and `tt` for $t$. 
 The functions above are equivalent, but the first is generally more awkward
@@
 Let us consider there are $n$ parameters and $m$ residuals. 
-We note several useful quantities. The loss function is 
+The loss function is 
 $$ S(p) = r' r = \sum_{i=1}^m { r_i^2 }$$
@@
 a multiple of the diagonal part of $J' J$. In low precision, some elements of 
-$J' J$ could underflow to zero (@jn77ima), and one of us suggested a linear combination of both
-choices as a compromise. Various choices for
+$J' J$ could underflow to zero (@jn77ima), and a linear combination of both
+choices is an effective compromise. Various choices for
 $D$, as well as a possible line search along the
@@
 of a similar nature should try to provide good starting points to reduce
-computation time and the chance a false solution is found.
+computation time and avoid finding a false solution.
 
@@
 Below we see `nls()` does poorly on the example problem. To save page space,
-we use 1-line result display functions from package `nlsr`, nameley `pnlm0()` 
+we use 1-line result display functions from package `nlsr`, namely `pnls()` 
 and `pshort()`.
@@
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-pnlm0(unlm1)  # Short form of output
+pnls(unlm1)  # Short form of output
 snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
-pnlm0(snlm1)
+pnls(snlm1)
 tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
-pnlm0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
+pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
 ```
@@
 ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
-pnlm0(ugslnls1) # to get sum of squares
+pnls(ugslnls1) # to get sum of squares
 sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
-pnlm0(sgslnls1) # Use summary() to get display
+pnls(sgslnls1) # Use summary() to get display
 tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
-pnlm0(tgslnls1) 
-```
-
-### Comparison notes for formula setup solutions
+pnls(tgslnls1) 
+```
+
+### Comparison notes for formula-setup solutions
 
@@
 a numerical approximation for the Jacobian if the argument `jac` 
-is missing.
+is missing. Note function `nlsr::pnlslm()` for a 1-line display of
+the results of `minpack.lm::nls.lm()`.
 
@@
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnlm1(hobnlm)  
+pnlslm(hobnlm)  
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
-pnlm0(hobgsln)
+pnls(hobgsln)
 hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
-pnlm0(hobgsl) # using analytic Jacobian
+pnls(hobgsl) # using analytic Jacobian
 ```
@@
 
-The output of `nls()` is an object of class "nls" which has quite rich structure
+As mentioned, the output of `nls()` is an object of class "nls" which has a quite rich structure
 described in the manual file or revealed by applying the `str()` function to the
@@
 ```
-and if there is an element called `xdata` that can be accessed as
+and if there is an element called `xdata`, then it can be accessed as
 `result$model$xdata`.
@@
 the Jacobian matrix lets us embed this in the **residual** function as well,
-so that the call to `nlfb()` can be made with the same name used for 
+so that the call to `nlsr::nlfb()` can be made with the same name used for 
 both residual and Jacobian function arguments. This programming trick saves a 
@@
 computing the "right hand side" or `rhs` of the residual, checks to see if the
-"gradient" attribute is defined, and, otherwise using `numericDeriv()` to compute a 
+"gradient" attribute is defined, otherwise using `numericDeriv()` to compute a 
 Jacobian into that attribute. This code is within the `nlsModel()` or
@@
 starting values, it actually calls `nls()` with the 'plinear' algorithm 
-and finds a solution. It then passes the solution coefficients to the default 
-algorithm unnecessarily. Moreover, the double call is, in our view, prone to creating 
+and finds a (full) solution. It then passes the solution coefficients to the default 
+algorithm unnecessarily. Moreover, the implicit double call is, in our view, prone to creating 
 errors in code maintenance. 
@@
 This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
-starting parameters, and uses `getInitial()' to find them from the selfStart
-model. The same model is used in the formula for the `nlxb()` call.
+starting parameters, and can either use `getInitial()` to find them from the `selfStart`
+model or provide explicit values, but the `selfStart` model is used in the formula 
+for the `nlxb()` call.
 
@@
 We are also surprised that the 
-analytic expressions for the Jacobian ("gradient") in the `SSLogis` R function 
+analytic expressions for the Jacobian ("gradient") in the `SSLogis` function 
 and others save quantities in "hidden"
@@
 command, making them difficult to access by users who may wish to create
-their own selfStart model by a copy and edit.
+their own selfStart model via copy and edit.
 Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
@@
 
-In some cases, we know that parameters cannot be bigger or smaller than some
+For many problems we know that parameters cannot be bigger or smaller than some
 externally known limits. Such limits should be built into models, but there 
@@
              upper=c(2,6,3), algorithm='port')
-pnls0(anlshob1b) #  check the answer (short form)
+pnls(anlshob1b) #  check the answer (short form)
 # nlsLM seems NOT to work with bounds in this example
 anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls0(anlsLM1b)
+pnls(anlsLM1b)
 # also no warning if starting out of bounds, but gets good answer!!
@@
 anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls0(anlsLMob)
+pnls(anlsLMob)
 # Try nlsr::nlxb()
@@
 			upper=c(200, 60, 3), trace=traceval))
-pnlm0(anlM1)
+pnls(anlM1)
 anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
 		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
-pnls0(anlsmsk1)
+pnls(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
@@
              upper=c(2,6,30), algorithm='port')
-pnls0(anlshmsk1)
+pnls(anlshmsk1)
 anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
-pnlm0(anlsLMmsks1)
+pnls(anlsLMmsks1)
 
@@
 the path will be loaded on startup. R users in our
-acquaintance avoid saving the workspace because lurking data and
-functions which may cause unwanted results.
-Program alternatives to `nls()` still need to ensure they work with data in 
-the available environments.
+acquaintance avoid saving the workspace because of lurking data and
+functions that may cause unwanted results.
 
@@
 obviously in the source code, often being handled behind the scenes after 
-referencing the option `na.action`, yet this feature also changes the data 
+referencing the option `na.action`. This feature also changes the data 
 supplied to our nonlinear least squares solver. 
@@
 "Introduction to nlsr" of package `nlsr` in section 
-*Weights that are functions of the model parameters* and 
+*Weights that are functions of the model parameters*.
 `minpack.lm` offers a function `wfct()` to 
@@
 are linear and communicating this information to estimating functions 
-is not a trivial task.
-
-`nls()` has an option `algorithm="plinear"` that allows some partially linear
+is not a trivial task. `nls()` has an option `algorithm="plinear"` that allows some partially linear
 models to be solved. The other tools, as far as we are aware, do not offer any
@@
 Maintainers of packages need suitable tests and use-case examples in order
+
 - to ensure packages work properly, in particular, giving

### 2023-05-14 RefactoringNLS/ComparingNLStools.tex
minor updates

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
%
\documentclass[
]{article}
\usepackage{amsmath,amssymb}
\usepackage{lmodern}
\usepackage{iftex}
\ifPDFTeX
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\usepackage[margin=1in]{geometry}
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{graphicx}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
\newlength{\cslhangindent}
\setlength{\cslhangindent}{1.5em}
\newlength{\csllabelwidth}
\setlength{\csllabelwidth}{3em}
\newlength{\cslentryspacingunit} % times entry-spacing
\setlength{\cslentryspacingunit}{\parskip}
\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
 {% don't indent paragraphs
  \setlength{\parindent}{0pt}
  % turn on hanging indent if param 1 is 1
  \ifodd #1
  \let\oldpar\par
  \def\par{\hangindent=\cslhangindent\oldpar}
  \fi
  % set entry spacing
  \setlength{\parskip}{#2\cslentryspacingunit}
 }%
 {}
\usepackage{calc}
\newcommand{\CSLBlock}[1]{#1\hfill\break}
\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={A comparison of R tools for nonlinear least squares modeling},
  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
  colorlinks=true,
  linkcolor={red},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={blue},
  pdfcreator={LaTeX via pandoc}}

\title{A comparison of R tools for nonlinear least squares modeling}
\author{John C. Nash
\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
Bhattacharjee
\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
\date{2023-3-2}

\begin{document}
\maketitle

\hypertarget{abstract}{%
\subsection{Abstract}\label{abstract}}

Our Google Summer of Code project ``Improvements to \texttt{nls()}''
considered the features and limitations of the R function \texttt{nls()}
with the aim of improving and rationalizing R tools for nonlinear
regression. The rich features of \texttt{nls()} are weakened by several
deficiencies and inconsistencies such as a lack of stabilization of the
Gauss-Newton solver. Further considerations are the usability and
maintainability of the code base that provides the functionality
\texttt{nls()} claims to offer. Various packages, including our
\texttt{nlsr}, provide alternative capabilities. We consider the
differences in goals, approaches, and features of different tools for
nonlinear least squares modeling in R. Discussion of these matters is
relevant to improving R generally as well as its nonlinear estimation
tools.

\hypertarget{the-nls-function}{%
\subsection{\texorpdfstring{The \texttt{nls()}
function}{The nls() function}}\label{the-nls-function}}

\texttt{nls()} is the tool for estimating nonlinear statistical models
in base R, the primary software distribution from the Comprehensive R
Archive Network (\emph{\url{https://cran.r-project.org}}). The function
dates to the 1980s and the work related to Bates and Watts
(\protect\hyperlink{ref-bateswatts}{1988}) in S (see
\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).

The \texttt{nls()} function has a remarkable and comprehensive set of
capabilities for estimating nonlinear models that are expressed as
\textbf{formulas}. In particular, we note that it

\begin{itemize}
\tightlist
\item
  handles formulas that include R functions, even ones that call
  calculations in other programming languages
\item
  allows data to be weighted or subset
\item
  can estimate bound-constrained parameters
\item
  provides a mechanism for handling partially linear models
\item
  permits parameters to be indexed over a set of related data
\item
  produces measures of variability (i.e., standard error estimates) for
  the estimated parameters
\item
  has related profiling capabilities for exploring the likelihood
  surface as parameters are changed
\item
  links to many pre-coded (\texttt{selfStart}) models that do not
  require initial parameter values.
\end{itemize}

With such a range of features and long history, the code has become
untidy and overly patched, difficult to maintain, and its underlying
methods could be improved. Various workers have developed packages to
overcome these concerns, and we will address some of these here.

\hypertarget{scope-of-our-comparison}{%
\subsection{Scope of our comparison}\label{scope-of-our-comparison}}

The tools we will consider are the base-R \texttt{nls()} function and
some packages that are available on the CRAN repository. Of these, we
will pay particular attention to \textbf{nlsr} (John C Nash and Duncan
Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
\textbf{minpack.lm} (Elzhov et al.
(\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau
(\protect\hyperlink{ref-gslnls23}{2023})) which are general nonlinear
least-squares solvers.

While we will provide capsule comments for some other CRAN packages, we
will not pursue those in the Bioconductor (Gentleman et al.
(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
repositories such as GitHub (\url{https://github.com}) and Gitlab
(\url{https://about.gitlab.com}).

\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
\subsection{Some other CRAN packages for nonlinear
modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}

\texttt{onls} (Andrej-Nikolai Spiess
(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
in estimating nonlinear models is the vertical difference between the
dependent variable and the functional model. \texttt{onls} minimizes the
sum of \textbf{orthogonal} residuals. The objective is therefore
different and involves nontrivial extra calculation. A vignette with the
package and the blog article
\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
give some description with illustrative graphs. \texttt{onls} appears to
be limited to problems with one independent and one dependent variable.
The Wikipedia article
\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
overview of some ideas, with references to the literature. The approach
needs a wider discussion and tutorial examples to allow its merits to be
judged than can be included here.

\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
-- This package allows nonlinear estimation by controlled random search
via two methods. There is unfortunately no vignette. A modest trial we
carried out showed \texttt{nlsr::nlxb()} gave the same results in a
small fraction of the time required by either of the methods in
\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
reliability in finding solutions than a Levenberg-Marquardt code
(actually from Matlab), but the tests were conducted on the extreme NIST
examples mentioned next.

\texttt{NISTnls} (National Institutes for Standards and R port by
Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
package provides R code and data for a set of (numerically
ill-conditioned) nonlinear least squares problems from the U.S. National
Institute for Standards and Technology. These may not represent
real-world situations.

\texttt{nlshelper} (Duursma
(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
unfortunately lacks a vignette, provides a few utilities for
summarizing, testing, and plotting non-linear regression models
estimated with nls(), nlsList() or nlme() that are linked or grouped in
some way.

\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
This solves nonlinear least squares problems with optional equality
and/or inequality constraints. It is clearly \textbf{not} about
modeling, and the input and output are quite different from class
\texttt{nls} methods. However, there do not appear to be other R
packages with these capabilities.

\texttt{nlsMicrobio} (Baty and Delignette-Muller
(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
nonlinear regression models dedicated to predictive microbiology,
including a vignette, by authors of the \texttt{nlstools} package.

\texttt{nlstools} (Baty and Delignette-Muller
(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
several tools for aiding the estimation of nonlinear models,
particularly using \texttt{nls()}. The vignette is actually a journal
article, and the authors have considerable experience in the subject.

\texttt{nlsmsn} (Prates, Lachos, and Garay
(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
Lachos, and Abanto-Valle (2011). The problem here is to minimize an
objective that is modified from the traditional sum of squared
residuals.

\texttt{nls.multstart} (Padfield and Matheson
(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
squares regression using AIC scores with the Levenberg-Marquardt
algorithm using multiple starting values for increasing the chance that
the minimum found is the global minimum.

\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
Nonlinear least squares by brute force has similar motivations to
\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
The author has extensive expertise in R.

\texttt{nlstac} (Rodriguez-Arias et al.
(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
implementing the algorithm described in Fernandez Torvisco et al.~(2018)
for fitting separable nonlinear regression curves. The special class of
problem for which this package is intended is an important and difficult
one. No vignette is provided, unfortunately.

\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models
are treated, but there is minimal documentation and no vignette. Package
\texttt{nlraa} is to be preferred.

\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
-- a set of nonlinear \emph{selfStart} models, primarily from
agriculture. Most include analytic Jacobian code.

\texttt{optimx} (John C. Nash and Varadhan
(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
that can be applied to minimize a nonlinear function which could be a
nonlinear sum of squares. NOT generally recommended if nonlinear least
squares programs can be easily used, but provides a check and
alternative solvers.

\hypertarget{an-illustrative-example}{%
\subsection{An illustrative example}\label{an-illustrative-example}}

The Hobbs weed infestation problem (John C. Nash
(\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
task that seems straightforward but is quite nasty. Its very succinct
statement provides the ``short reproducible example'' much requested on
R mailing lists. This is a real problem from a field researcher. The
data and graph follow.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
\FunctionTok{plot}\NormalTok{(weeddf, }\AttributeTok{main=}\StringTok{"Hobbs weed infestation data"}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

\includegraphics{ComparingNLStools_files/figure-latex/ex01-1.pdf}

Three suggested models for this data are (with names to allow for easy
reference)

\textbf{Logistic3U}: \[  y \approx  b_1 / (1 + b_2 * exp(- b_3 * t)) \]

\textbf{Logistic3S}:
\[ y \approx  100 * c_1 / (1 + 10 * c_2 * exp(- 0.1 * c_3 * t)) \]

\textbf{Logistic3T}: \[ y \approx Asym / (1 + exp((xmid - t)/scal)) \]

where we will use \texttt{weed} for \(y\) and \texttt{tt} for \(t\). The
functions above are equivalent, but the first is generally more awkward
to solve numerically due to its poor scaling. The parameters of the
three forms are related as follows:

\[   Asym =  b_1 = 100 * c_1 \] \[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
\[ 1/scal  =  b_3 \] To allow for simpler discussion, let us say that
the parameters form a (named) vector \(p\) and the model function is
called \(model(p)\). The residuals can be written either as
\(r_1 = y - model(p)\) or \(r = model(p) - y\) since their sum of
squares has an identical value. The second form avoids a potential sign
error if we need to evaluate derivatives.

We wish to minimize the sum of squared residuals, which is our loss (or
objective) function. Starting with some guess for the parameters, we aim
to alter these parameters to obtain a smaller loss function. We then
iterate until we can make no further progress.

Let us consider there are \(n\) parameters and \(m\) residuals. The loss
function is \[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\] The gradient of
\(S(p)\) is \[ 2 * J' r\] where the Jacobian \(J\) is given by elements
\[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is defined
by elements
\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \] If we
expand the Hessian for nonlinear least squares problems, we find
\[  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
Let us use \(D_{i,j}\) for the elements of the second term of this
expression. What is generally called \textbf{Newton's method} for
function minimization tries to set the gradient to zero (to find a
stationary point of the function \(S(p)\)). This leads to
\textbf{Newton's equation} \[ H \delta = -g \] Given a set of parameters
\(p\), we solve this equation for \(\delta\), adjust \(p\) to
\(p + \delta\) and iterate, hopefully to converge on a solution.
Applying this to a sum of squares problem gives
\[ H \delta = (J'J + D) \delta = - J' r\] In this expression, only the
elements of \(D\) have second partial derivatives. Gauss, attempting to
model planetary orbits, had small residuals, and noted that these
multiplied the second partial derivatives of \(r\), so he approximated
\[ H \approx J' J\] by assuming \(D \approx 0\). This results in the
Gauss-Newton method where we solve \[ J' J \delta = - J' r\] though we
can avoid some loss of accuracy by NOT forming the inner product matrix
\(J' J\) and solving the linear least squares matrix problem
\[ J \delta \approx -r \] by one of several matrix decomposition
methods.

In reality, there are many problems were \(D\) should not be ignored,
but the work to compute it precisely is considerable. Many work-arounds
have been proposed, of which the Levenberg-Marquardt stabilization
(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt
(\protect\hyperlink{ref-Marquardt1963}{1963})) is the most commonly
used. For convenience, we will use ``Marquardt'', as we believe he first
incorporated the ideas into a practical computer program.

The usual suggestion is that \(D\) be replaced by a multiple of the unit
matrix or else a multiple of the diagonal part of \(J' J\). In low
precision, some elements of \(J' J\) could underflow to zero (John C.
Nash (\protect\hyperlink{ref-jn77ima}{1977})), and a linear combination
of both choices is an effective compromise. Various choices for \(D\),
as well as a possible line search along the direction \(\delta\) rather
than a unit step (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})),
give rise to several variant algorithms. ``Marquardt's method'' is a
family of methods. Fortunately, most choices work well.

\hypertarget{problem-setup}{%
\subsubsection{Problem setup}\label{problem-setup}}

Let us specify in R the three model formulas and set some starting
values for parameters. These starting points are NOT equivalent and are
deliberately crude choices. Workers performing many calculations of a
similar nature should try to provide good starting points to reduce
computation time and avoid finding a false solution.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# model formulas}
\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{b2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3}\SpecialCharTok{*}\NormalTok{tt))}
\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{c1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\DecValTok{10}\SpecialCharTok{*}\NormalTok{c2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1}\SpecialCharTok{*}\NormalTok{c3}\SpecialCharTok{*}\NormalTok{tt))}
\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid}\SpecialCharTok{{-}}\NormalTok{tt)}\SpecialCharTok{/}\NormalTok{scal))}
\CommentTok{\#}
\CommentTok{\# Starting parameter sets}
\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{1}\NormalTok{, }\AttributeTok{b2=}\DecValTok{1}\NormalTok{, }\AttributeTok{b3=}\DecValTok{1}\NormalTok{)}
\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{1}\NormalTok{, }\AttributeTok{c2=}\DecValTok{1}\NormalTok{, }\AttributeTok{c3=}\DecValTok{1}\NormalTok{)}
\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{)}
\end{Highlighting}
\end{Shaded}

One of the useful features of \texttt{nls()} is the possibility of a
\texttt{selfStart} model, where starting parameter values are not
required. However, if a \texttt{selfStart} model is not available,
\texttt{nls()} sets all the starting parameters to 1. This is tolerable,
but could be improved by using a set of values that are all slightly
different, which, in the case of the example model
\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
Program modifications to give a sequence like 1.0, 1.1, 1.2, 1.3 for the
four parameters are fairly obvious.

It is also possible to provide R functions for the residual and
Jacobian. This is usually more work for the user if the formula setup is
possible. To illustrate, we show the functions for the unscaled 3
parameter logistic.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Logistic3U}
\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
  \CommentTok{\# This variant uses looping}
  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
\NormalTok{\}}

\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)}
\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{yy)}
\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}
\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}\SpecialCharTok{*}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt}
  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
\NormalTok{  jj}
\NormalTok{\}}
\end{Highlighting}
\end{Shaded}

\hypertarget{estimation-of-models-specified-as-formulas}{%
\subsection{Estimation of models specified as
formulas}\label{estimation-of-models-specified-as-formulas}}

Using a formula specification was a principal advantage made with
\texttt{nls()} when it became available in S sometime in the 1980s. It
uses a Gauss-Newton (i.e., unstabilized) iteration with a step reduction
line search. This works very efficiently as long as \(J\) is not
ill-conditioned. Below we see \texttt{nls()} does poorly on the example
problem. To save page space, we use 1-line result display functions from
package \texttt{nlsr}, namely \texttt{pnls()} and \texttt{pshort()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frmu, start = stu1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frms, start = sts1, data = weeddf) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Error in nls(formula = frmt, start = stt1, data = weeddf) : 
##   singular gradient
\end{verbatim}

Here we see the infamous ``singular gradient'' termination message of
\texttt{nls()}.

\hypertarget{solution-attempts-with-nlsr}{%
\subsubsection{Solution attempts with
nlsr}\label{solution-attempts-with-nlsr}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{print}\NormalTok{(unlx1) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  2.5873  on  12 observations
##     after  19    Jacobian and  25 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
## b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
## b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
\end{verbatim}

Though we have found solutions, the Jacobian is essentially singular as
shown by its singular values. Note that these are \textbf{displayed} by
package \texttt{nlsr} in a single column in the output to provide a
compact layout, but the values do \textbf{NOT} correspond to the
individual parameters in whose row they appear; they are a property of
the whole problem.

\hypertarget{solution-attempts-with-minpack.lm}{%
\subsubsection{Solution attempts with
minpack.lm}\label{solution-attempts-with-minpack.lm}}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{unlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{snlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{pnls}\NormalTok{(snlm1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tnlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
\FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
\end{verbatim}

\hypertarget{solution-attempts-with-gslnls}{%
\subsubsection{Solution attempts with
gslnls}\label{solution-attempts-with-gslnls}}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{library}\NormalTok{(gslnls)}
\NormalTok{ugslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
\FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{sgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
\FunctionTok{pnls}\NormalTok{(sgslnls1) }\CommentTok{\# Use summary() to get display}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{tgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
\FunctionTok{pnls}\NormalTok{(tgslnls1) }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20847  scal = -1744.7; 47  itns
\end{verbatim}

\hypertarget{comparison-notes-for-formula-setup-solutions}{%
\subsubsection{Comparison notes for formula-setup
solutions}\label{comparison-notes-for-formula-setup-solutions}}

\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors
and singular values of the Jacobian (for diagnostic purposes). By
contrast, \texttt{minpack.lm::nlsLM()} and \texttt{nls()} use
\texttt{summary()}, which does NOT display the sum of squares, while
\texttt{print()} gives the sum of squares, but not the standard error of
the residuals.

The singular values allow us to gauge how ``nearly singular'' the
Jacobian is at the solution, and the ratio of the smallest to largest of
the singular values is a simple but effective measure. The ratios are
4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
Logistic3T, so Logistic3S is the ``least singular''.

The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the
transformed model Logistic3T have a very large sum of squares, which may
suggest that these programs have failed. Since \texttt{nls()},
\texttt{nlsLM()}, and \texttt{gsl\_nls()} do not offer singular values,
we need to extract the Jacobian and compute its singular values. The
following script shows how to do this, using as Jacobian what is called
the \texttt{gradient} element in the returned solution for these
solvers.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# for nlsLM}
\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tnlm1, }\StringTok{"try{-}error"}\NormalTok{))  \{}
   \FunctionTok{print}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
\NormalTok{ \} }\ControlFlowTok{else}\NormalTok{ \{  }
\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tnlm1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{() }\CommentTok{\# actually the Jacobian}
   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
\NormalTok{\}   }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 3.4641e+00 3.2530e-10 8.9643e-12
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# for gsl\_nls}
\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tgslnls1, }\StringTok{"try{-}error"}\NormalTok{)) \{}
   \FunctionTok{cat}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
\NormalTok{\} }\ControlFlowTok{else}\NormalTok{ \{  }
\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tgslnls1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{()}
   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
\NormalTok{\}   }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] 3.4641e+00 1.0111e-08 4.7443e-11
\end{verbatim}

We see that there are differences in detail, but the more important
result is that two out of three singular values are essentially 0. Our
Jacobian is singular, and no method of the Gauss-Newton type should be
able to continue. Indeed, from the parameters reported at this saddle
point, \texttt{nlsr::nlxb()} cannot proceed.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{stspecial}\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
\NormalTok{badstart}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stspecial, }\AttributeTok{data=}\NormalTok{weeddf)}
\FunctionTok{print}\NormalTok{(badstart)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  9205.4  on  12 observations
##     after  2    Jacobian and  2 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## Asym             35.5321            NA         NA         NA  -9.694e-09       3.464  
## xmid               43376            NA         NA         NA  -1.742e-09    2.61e-10  
## scal             -2935.4            NA         NA         NA    -2.4e-08    7.12e-16
\end{verbatim}

\hypertarget{functional-specification-of-problems}{%
\subsection{Functional specification of
problems}\label{functional-specification-of-problems}}

We illustrate how to solve nonlinear least squares problems using a
function to define the residual. Note that \texttt{gsl\_nls()} requires
a vector \texttt{y} that is the expected value of what we have called
the residual, but in this case, is actually the model. \texttt{gsl\_nls}
uses a numerical approximation for the Jacobian if the argument
\texttt{jac} is missing. Note function \texttt{nlsr::pnlslm()} for a
1-line display of the results of \texttt{minpack.lm::nls.lm()}.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobnlfb}\OtherTok{\textless{}{-}}\FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{resfn=}\NormalTok{hobbs.res, }\AttributeTok{jacfn=}\NormalTok{hobbs.jac)}
\FunctionTok{pshort}\NormalTok{(hobnlfb) }\CommentTok{\# use print(hobnlfb) for more detail}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobnlm}\OtherTok{\textless{}{-}}\FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
\FunctionTok{pnlslm}\NormalTok{(hobnlm)  }
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobgsln}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(hobgsln)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{hobgsl}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{), }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
\FunctionTok{pnls}\NormalTok{(hobgsl) }\CommentTok{\# using analytic Jacobian}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
\end{verbatim}

\hypertarget{design-goals-termination-tests-and-output-objects}{%
\subsection{Design goals, termination tests, and output
objects}\label{design-goals-termination-tests-and-output-objects}}

The output object of \texttt{nlxb()} is smaller than the class
\texttt{nls} object returned by \texttt{nls()}, \texttt{nlsLM()}, and
\texttt{gsl\_nls()}. Package \texttt{nlsr} emphasizes the solution of
the nonlinear least squares problem rather than the estimation of a
nonlinear model that fits or explains the data. The object of class
\texttt{nls} allows for a number of specialized modeling and diagnostic
extensions. For compatibility, the \texttt{nlsr} package has the
function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias.
This uses \texttt{nlxb()} to find good parameters, then calls
\texttt{nls()} to return the class \texttt{nls} object. Unless
particular modeling features are needed, the use of \texttt{wrapnlsr()}
is unnecessary and wasteful of resources.

The design goals of the different tools may also be revealed in the
so-called ``convergence tests'' for the iterative solvers. In the manual
page for \texttt{nls()} in R 4.0.0 there was the warning:

\begin{quote}
\emph{Do not use \texttt{nls} on artificial ``zero-residual'' data.}
\end{quote}

with suggested addition of small perturbations to the data. This admits
\texttt{nls()} could not solve well-posed problems unless data is
polluted with errors. Zero-residual problems are not always artificial,
since problems in function approximation and nonlinear equations can be
approached with nonlinear least squares. Fortunately, a small adjustment
to the ``termination test'' for the \textbf{program}, rather than for
the ``convergence'' of the underlying \textbf{algorithm}, fixes the
defect. The test is the Relative Offset Convergence Criterion (see
Bates, Douglas M. and Watts, Donald G.
(\protect\hyperlink{ref-BatesWatts81}{1981})). This scales an estimated
reduction in the loss function by its current value. If the loss
function is very small, we are close to a zero-divide. Adding a small
quantity to the divisor avoids trouble. In 2021, one of us (J. Nash)
proposed that \texttt{nls.control()} have an additional parameter
\texttt{scaleOffset} with a default value of zero. Setting it to a small
number -- 1.0 is a reasonable choice -- allows small-residual problems
(i.e., near-exact fits) to be dealt with easily. We call this the
\textbf{safeguarded relative offset convergence criterion}, and it has
been in \texttt{nlsr} since it was introduced. The default value gives
the legacy behaviour. This improvement has been in the R distributed
code since version 4.1.0.

Additional termination tests can be used. \texttt{nlsr} has a
\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
times the initial sum of squares, where
\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
approximately 2.43e-55.

Termination after what may be considered excessive computation is also
important. \texttt{nls()} stops after \texttt{maxiter} ``iterations''.
The meaning of ``iteration'' may require an examination of the code. We
terminate execution when the number of residual or Jacobian evaluations
exceed set limits. Generally, we prefer larger limits than the default
\texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.

\hypertarget{returned-results-of-nls-and-other-tools}{%
\subsubsection{Returned results of nls() and other
tools}\label{returned-results-of-nls-and-other-tools}}

As mentioned, the output of \texttt{nls()} is an object of class ``nls''
which has a quite rich structure described in the manual file or
revealed by applying the \texttt{str()} function to the result of
\texttt{nls()}.

The complexity of this object is a challenge to users. Let us use the
\texttt{result} as the returned object from \texttt{nls()},
\texttt{minpack.lm::nlsLM()} or, \texttt{gslnls::gsl\_nls()}. For
example, the \texttt{data} return element is an R symbol. To actually
access the data from this element, we need to use the syntax:

\begin{verbatim}
eval(parse(text=result$data))
\end{verbatim}

However, if the call is made with \texttt{model=TRUE}, then there is a
returned element \texttt{model} which contains the data, and we can list
its contents using:

\begin{verbatim}
ls(result$model)
\end{verbatim}

and if there is an element called \texttt{xdata}, then it can be
accessed as \texttt{result\$model\$xdata}.

By contrast, \texttt{nlsr::nlxb()} returns a much simpler structure of
11 items in one level. Moreover, \texttt{nlxb} explicitly returns the
sum of squares, the residual vector, Jacobian, and counts of
evaluations.

\hypertarget{when-to-compute-ancillary-information}{%
\subsubsection{When to compute ancillary
information}\label{when-to-compute-ancillary-information}}

Tools that produce a class \texttt{nls} output object create a rich set
of functions and structures that are then used in a variety of modeling
tasks, including the least squares solution. By contrast, \texttt{nlsr}
computes quantities as they are requested or needed, with additional
features in separate functions. For example, the singular values of the
Jacobian are actually computed in the \texttt{print} and
\texttt{summary} methods for the result. These two approaches lead to
different consequences for performance and how features are provided.
\texttt{nlsr} has antecedents in the methods of John C. Nash
(\protect\hyperlink{ref-cnm79}{1979}), where storage for data and
programs was at a ridiculous premium in the small computers of the era.
Thus the code in \texttt{nlsr} is likely of value for workers to copy
and modify for customized tools.

\hypertarget{jacobian-calculation}{%
\subsection{Jacobian calculation}\label{jacobian-calculation}}

Gauss-Newton/Marquardt methods all need a Jacobian matrix at each
iteration. By default, \texttt{nlsr::nlxb()} will try to evaluate this
using analytic expressions using symbolic and automatic differentiation
tools. When using a formula specification of the model, \texttt{nls()},
\texttt{minpack.lm::nlsLM()} and \texttt{gslnls::gsl\_nls()} use a
finite difference approximation to compute the Jacobian, though
\texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
Package \texttt{nlsr} provides, via appropriate calling syntax, four
numeric approximation options for the Jacobian, with a further control
\texttt{ndstep} for the size of the step used in the approximation.

Using the ``gradient'' attribute of the output of the Jacobian function
to hold the Jacobian matrix lets us embed this in the \textbf{residual}
function as well, so that the call to \texttt{nlsr::nlfb()} can be made
with the same name used for both residual and Jacobian function
arguments. This programming trick saves a lot of trouble for the package
developer, but it can be a nuisance for users trying to understand the
code.

As far as we can understand the logic in \texttt{nls()}, the Jacobian
computation during parameter estimation is carried out within the called
C-language program and its wrapper R code function
\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
the R distribution source code. This is used to provide Jacobian
information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
functions, which are \textbf{not} exported for general use.
\texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.

\texttt{numericDeriv()} uses a simple forward difference approximation
of derivatives, though a central difference approximation can be
specified in control parameters. We are unclear why
\texttt{numericDeriv()} in base R calls \texttt{C\_numeric\_deriv}, as
we were easily able to create a more compact version entirely in R. See
\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.

\texttt{minpack.lm::nlsLM()} invokes \texttt{numericDeriv()} in its
local version of \texttt{nlsModel()}, but it appears to use an internal
approximate Jacobian code from the original Fortran \texttt{minpack}
code, namely, \texttt{lmdif.f}. Such differences in approach can lead to
different behaviour, usually minor, but sometimes annoying with
ill-conditioned problems.

\begin{itemize}
\item
  A pasture regrowth problem (Huet et al.
  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
  conditioned Jacobian and \texttt{nls()} fails with ``singular
  gradient''. Worse, numerical approximation to the Jacobian gives the
  error ``singular gradient matrix at initial parameter estimates'' for
  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
  unable to take effect, while the analytic derivatives of
  \texttt{nlsr::nlxb} give a solution.
\item
  Karl Schilling (private communication) provided an example where a
  model specified with the formula
  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
  which \texttt{x\ =\ 0}, the derivative is undefined, but the model can
  be computed. In such cases, we observed that \texttt{nls()} and
  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a
  lucky accident.
\end{itemize}

\hypertarget{jacobian-code-in-selfstart-models}{%
\subsubsection{Jacobian code in selfStart
models}\label{jacobian-code-in-selfstart-models}}

Analytic Jacobian code can be provided to all the solvers discussed.
Most \texttt{selfStart} models that automatically provide starting
parameters also include such code. There is documentation in R of
\texttt{selfStart} models, but their construction is non-trivial. A
number of such models are included with base R in
\texttt{./src/library/stats/R/zzModels.R}, with package \texttt{nlraa}
(Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})) providing a
richer set. There are also some in the now-archived package
\texttt{NRAIA}. These provide the Jacobian in the ``gradient'' attribute
of the ``one-sided'' formula that defines each model, and these
Jacobians are often the analytic forms.

The \texttt{nls()} function, after computing the ``right-hand side'' or
\texttt{rhs} of the residual, checks to see if the ``gradient''
attribute is defined, otherwise using \texttt{numericDeriv()} to compute
a Jacobian into that attribute. This code is within the
\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
analytic Jacobians almost certainly contributes to the good performance
of \texttt{nls()} on \texttt{selfStart} models.

The use of \texttt{selfStart} models with \texttt{nlsr} is described in
the ``Introduction to nlsr'' vignette. However, since \texttt{nlsr}
generally can use very crude starting values, we have rarely needed
them, though it should be pointed out that our work is primarily
diagnostic. If we were carrying out a large number of similar
estimations, such initial parameters are critical to efficiency.

In considering \texttt{selfStart} models, we noted that the base-R
function \texttt{SSlogis} is intended to solve problem
\textbf{Logistic3T} above. When this function is used via
\texttt{getInitial()} to find starting values, it actually calls
\texttt{nls()} with the `plinear' algorithm and finds a (full) solution.
It then passes the solution coefficients to the default algorithm
unnecessarily. Moreover, the implicit double call is, in our view, prone
to creating errors in code maintenance. To provide simpler starting
parameters, the function \texttt{SSlogisJN} is now part of the package
\texttt{nlsr}, but is most useful for \texttt{nls()}.

Users may also want to profit from the Jacobian code of
\texttt{selfStart} models but supply explicit starting values other than
those suggested by \texttt{getInitial()}. This does not appear to be
possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
starting parameters, and can either use \texttt{getInitial()} to find
them from the \texttt{selfStart} model or provide explicit values, but
the \texttt{selfStart} model is used in the formula for the
\texttt{nlxb()} call.

We are also surprised that the analytic expressions for the Jacobian
(``gradient'') in the \texttt{SSLogis} function and others save
quantities in ``hidden'' variables, i.e., with names preceded by ``.''.
These are then not displayed by the \texttt{ls()} command, making them
difficult to access by users who may wish to create their own selfStart
model via copy and edit. Interactive tools, such as ``visual fitting''
(John C. Nash and Velleman
(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
considering as another way to find starting parameters, but we know of
no R capability of this type.

As a side note, the introduction of \texttt{scaleOffset} in R 4.1.1 to
deal with the convergence test for small residual problems now requires
that the \texttt{getInitial()} function have dot-arguments
(\texttt{...}) in its argument list. This illustrates the entanglement
of many features in \texttt{nls()} that complicate its maintenance and
improvement.

\hypertarget{bounds-constraints-on-parameters}{%
\subsection{Bounds constraints on
parameters}\label{bounds-constraints-on-parameters}}

For many problems, we know that parameters cannot be bigger or smaller
than some externally known limits. Such limits should be built into
models, but there are some important details for using the tools in R.

\begin{itemize}
\item
  \texttt{nls()} can only impose bounds if the \texttt{algorithm="port"}
  argument is used in the call. Unfortunately, the documentation warns
  us:

  \emph{The algorithm = ``port'' code appears unfinished, and does not
  even check that the starting value is within the bounds. Use with
  caution, especially where bounds are supplied.}
\item
  \texttt{gsl\_nls()} does not offer bounds.
\item
  bounds are part of the default method for package \texttt{nlsr}.
\item
  \texttt{nlsLM()} includes bounds in the standard call, but we have
  observed cases where it fails to get the correct answer. From an
  examination of the code, we believe the authors have not taken into
  account all possibilities, though all programs have some weakness
  regarding constrained optimization in that programmers have to work
  with assumptions on the scale of numbers, and some problems will be
  outside the scope envisaged.
\end{itemize}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Start MUST be feasible i.e. on or within bounds}
\NormalTok{anlshob1b }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
\FunctionTok{pnls}\NormalTok{(anlshob1b) }\CommentTok{\#  check the answer (short form)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlshob1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 10  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# nlsLM seems NOT to work with bounds in this example}
\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(anlsLM1b)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlsLM1b  -- ss= 881.02 :  c1 = 2  c2 = 6  c3 = 3; 2  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# also no warning if starting out of bounds, but gets a good answer!!}
\NormalTok{st4}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{4}\NormalTok{, }\AttributeTok{c2=}\DecValTok{4}\NormalTok{, }\AttributeTok{c3=}\DecValTok{4}\NormalTok{)}
\NormalTok{anlsLMob }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{st4, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(anlsLMob)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlsLMob  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 4  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Try nlsr::nlxb()}
\NormalTok{anlx1b }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
\FunctionTok{pshort}\NormalTok{(anlx1b)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlx1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 12 res/ 12 jac
\end{verbatim}

\hypertarget{philosophical-considerations}{%
\subsubsection{Philosophical
considerations}\label{philosophical-considerations}}

Bounds on parameters raise some interesting questions about how
uncertainty in parameter estimates should be computed or reported. That
is, the traditional ``standard errors'' are generally taken to imply
symmetric intervals about the point estimate in which the parameter may
be expected to be found with some probability under certain assumptions.
Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not
compute standard errors nor their derived statistics when bounds are
active.

\hypertarget{fixed-parameters-masks}{%
\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}}

Let us try to fix (mask) the first parameter in the first two example
problems.

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Hobbsmaskx.R {-}{-} masks with formula specification of the problem}
\FunctionTok{require}\NormalTok{(nlsr); }\FunctionTok{require}\NormalTok{(minpack.lm); traceval}\OtherTok{\textless{}{-}}\ConstantTok{FALSE}
\NormalTok{stu }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{200}\NormalTok{, }\AttributeTok{b2=}\DecValTok{50}\NormalTok{, }\AttributeTok{b3=}\FloatTok{0.3}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
\NormalTok{sts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{2}\NormalTok{, }\AttributeTok{c2=}\DecValTok{5}\NormalTok{, }\AttributeTok{c3=}\DecValTok{3}\NormalTok{) }\CommentTok{\# a default scaled starting vector (named!)}
\CommentTok{\# fix first parameter}
\NormalTok{anxbmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
\FunctionTok{print}\NormalTok{(anxbmsk1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  2.6182  on  12 observations
##     after  4    Jacobian and  4 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## b1                   200U M         NA         NA         NA           0          NA  
## b2               49.5108          1.12      44.21  8.421e-13  -2.887e-07        1022  
## b3              0.311461      0.002278      136.8  1.073e-17   0.0001635      0.4569
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlM1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
\FunctionTok{pnls}\NormalTok{(anlM1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlM1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 4  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlsmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
        \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{),  }\AttributeTok{algorithm=}\StringTok{"port"}\NormalTok{, }\AttributeTok{trace=}\NormalTok{traceval))}
\FunctionTok{pnls}\NormalTok{(anlsmsk1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlsmsk1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 5  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
\NormalTok{anlxmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
                  \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
\FunctionTok{print}\NormalTok{(anlxmsks1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## residual sumsquares =  2.6182  on  12 observations
##     after  4    Jacobian and  4 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## c1                     2U M         NA         NA         NA           0          NA  
## c2               4.95108         0.112      44.21  8.421e-13  -2.981e-06       104.2  
## c3               3.11461       0.02278      136.8  1.073e-17   1.583e-05       4.482
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlshmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{trace=}\NormalTok{traceval, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
\FunctionTok{pnls}\NormalTok{(anlshmsk1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlshmsk1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 5  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{anlsLMmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
                 \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
\FunctionTok{pnls}\NormalTok{(anlsLMmsks1)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## anlsLMmsks1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 4  itns
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{\# Test with all parameters masked}
\NormalTok{anlxmskall}\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\NormalTok{sts, }\AttributeTok{upper=}\NormalTok{sts))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in nlfb(start = pnum, resfn = trjfn, jacfn = trjfn, trace = trace, :
## All parameters are masked
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(anlxmskall)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## Warning in summary.nlsr(x): No unmasked parameters
\end{verbatim}

\begin{verbatim}
## residual sumsquares =  158.23  on  12 observations
##     after  0    Jacobian and  1 function evaluations
##   name            coeff          SE       tstat      pval      gradient    JSingval   
## c1                     2U M         NA         NA         NA          NA          NA  
## c2                     5U M         NA         NA         NA          NA          NA  
## c3                     3U M         NA         NA         NA          NA          NA
\end{verbatim}

\texttt{nlsr} has an output format that indicates the constraint status
of the parameter estimates. For \texttt{nlsr}, we have \textbf{chosen}
to suppress the calculation of approximate standard errors in the
parameters when constraints are active because their meaning under
constraints is unclear, though we believe this policy worthy of
discussion and further investigation.

\hypertarget{stabilization-of-gauss-newton-computations}{%
\subsection{Stabilization of Gauss-Newton
computations}\label{stabilization-of-gauss-newton-computations}}

All four major tools illustrated solve some variant of the Gauss-Newton
equations. \texttt{nls()} uses a modification of an approach suggested
by Hartley (\protect\hyperlink{ref-Hartley1961}{1961}), while
\texttt{nlsr}, \texttt{gslnls}, and \texttt{minpack.lm} use flavours of
Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}). \texttt{gslnls}
offers an accelerated Marquardt method and three alternative methods
which we have not investigated. Control settings for \texttt{nlxb()} or
\texttt{nlfb()} allow exploration of Hartley and Marquardt algorithm
variants. In general, the Levenberg-Marquardt stabilization is important
in obtaining solutions in methods of the Gauss-Newton family, as
\texttt{nls()} terminates too frequently and unnecessarily with
\texttt{singular\ gradient} errors.

\hypertarget{programming-language}{%
\subsubsection{Programming language}\label{programming-language}}

An important choice made in developing \texttt{nlsr} was to code
entirely within the R programming language. \texttt{nls()} uses a mix of
R, C, and Fortran, as does \texttt{minpack.lm}. \texttt{gslnls} is an R
wrapper to various C-language routines in the Gnu Scientific Library
(Galassi et al. (\protect\hyperlink{ref-GSL-manual}{2009})). Generally,
keeping to a single programming language can allow for easier
maintenance and upgrades. It also avoids some work when there are
changes or upgrades to libraries for the non-R languages. R is usually
considered slower than most computing environments because it keeps
track of objects and because it is usually interpreted. In recent years,
the performance penalty for using code entirely in R has been much
reduced with the just-in-time compiler and other improvements. All-R
computation may now offer acceptable performance. In \texttt{nlsr}, the
use of R may be a smaller performance cost than the aggressive approach
to a solution, which can cause more iterations to be used.

\hypertarget{data-sources-for-problems}{%
\subsection{Data sources for problems}\label{data-sources-for-problems}}

\texttt{nls()} can be called without specifying the \texttt{data}
argument. In this case, it will search in the available environments
(i.e., workspaces) for suitable data objects. We do NOT like this
approach, but it is ``the R way''. R allows users to leave many objects
in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
actively suppress saving this workspace (\texttt{.RData}) on exit;
otherwise, any such file in the path will be loaded on startup. R users
in our acquaintance avoid saving the workspace because of lurking data
and functions that may cause unwanted results.

\hypertarget{feature-subsetting}{%
\subsubsection{Feature: Subsetting}\label{feature-subsetting}}

\texttt{nls()} and other class \texttt{nls} tools accept an argument
\texttt{subset}. This acts through the mediation of
\texttt{model.frame}, which not obvious in the source code files
\texttt{/src/library/stats/R/nls.R} and
\texttt{/src/library/stats/src/nls.C}. Having \texttt{subset} at the
level of the call to a function like \texttt{nls()} saves effort, but it
does mean that the programmer of the solver needs to be aware of the
origin (and value) of objects such as the data, residuals and Jacobian.
By preference, we would implement subsetting by zero-value weights, with
observation counts (and degrees of freedom) computed via the numbers of
non-zero weights. Alternatively, we would extract a working dataframe
from the relevant elements in the original.

\hypertarget{feature-na.action-missing-value-treatment}{%
\subsubsection{Feature: na.action (missing value
treatment)}\label{feature-na.action-missing-value-treatment}}

\texttt{na.action} is an argument to the \texttt{nls()} function, but it
does not appear obviously in the source code, often being handled behind
the scenes after referencing the option \texttt{na.action}. This feature
also changes the data supplied to our nonlinear least squares solver. A
useful, but possibly dated, description is given in:
\emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.
The typical default action, which can be seen by using the command
\texttt{getOption("na.action")} is \texttt{na.omit}. This option removes
from computations any observations containing missing values (i.e.~any
row of a data frame containing an NA). \texttt{na.exclude} does much of
the same for solver computations, but keeps the rows with NA elements so
that predictions are in the correct row position. We recommend that
workers actually test output to verify the behaviour is as wanted. See
\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}.
As with \texttt{subset}, our concern with \texttt{na.action} is that
users may be unaware of the effects of an option they may not even be
aware has been set. Should \texttt{na.fail} be the default?

\hypertarget{feature-model-frame}{%
\subsection{Feature: model frame}\label{feature-model-frame}}

\texttt{model} is an argument to the \texttt{nls()} and related
functions, which is documented as:

\begin{quote}
\textbf{model} logical. If true, the model frame is returned as part of
the object. Default is FALSE.
\end{quote}

Indeed, the argument only gets used when \texttt{nls()} is about to
return its result object, and the element \texttt{model} is NULL unless
the calling argument \texttt{model} is TRUE. (Using the same name could
be confusing.) Despite this, the model frame is used within the function
code in the form of the object \texttt{mf}. We feel that users could
benefit from more extensive documentation and examples of its use since
it is used to implement features like \texttt{subset}.

\hypertarget{weights-on-observations}{%
\subsection{Weights on observations}\label{weights-on-observations}}

All four main tools we consider here allow a \texttt{weights} argument
that specifies a vector of fixed weights the same length as the number
of residuals. Each residual is multiplied by the square root of the
corresponding weight. Where available, the values returned by the
\texttt{residuals()} function are weighted, and the \texttt{fitted()} or
\texttt{predict()} function are used to compute raw residuals.

While fixed weights may be useful, there are many problems for which we
want weights that are determined at least partially from the model
parameters, for example, a measure of the standard deviation of
observations. Such dynamic weighting situations are discussed in the
vignette ``Introduction to nlsr'' of package \texttt{nlsr} in section
\emph{Weights that are functions of the model parameters}.
\texttt{minpack.lm} offers a function \texttt{wfct()} to facilitate such
weighting. Care is advised in applying such ideas.

\hypertarget{weights-in-returned-functions-from-nls}{%
\subsubsection{Weights in returned functions from
nls()}\label{weights-in-returned-functions-from-nls}}

The function \texttt{resid()} (an alias for \texttt{residuals()}) gives
WEIGHTED residuals, as does, for example, \texttt{result\$m\$resid()}.
The function \texttt{nlsModel()}, which we have had to extract from the
base R code and explicitly \texttt{source()} because it is not exported
to the working namespace, allows us to compute residuals for particular
coefficient sets.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
\NormalTok{rnowt}\OtherTok{\textless{}{-}}\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.}
\FunctionTok{attr}\NormalTok{(rnowt, }\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rnowt}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
\NormalTok{rusewt}\OtherTok{\textless{}{-}}\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# WEIGHTED. Does NOT take coefficients.}
\FunctionTok{attr}\NormalTok{(rusewt,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rusewt}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
\NormalTok{rn0}\OtherTok{\textless{}{-}}\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
\FunctionTok{attr}\NormalTok{(rn0,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn0}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
## [11] 1.6572 1.4214
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts=}\NormalTok{wts)}
\NormalTok{rn}\OtherTok{\textless{}{-}}\NormalTok{nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
\FunctionTok{attr}\NormalTok{(rn,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
\end{verbatim}

\hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
\subsection{Minor issues with nonlinear least-squares
tools}\label{minor-issues-with-nonlinear-least-squares-tools}}

\hypertarget{interim-output-from-the-port-algorithm}{%
\subsubsection{Interim output from the ``port''
algorithm}\label{interim-output-from-the-port-algorithm}}

As the \texttt{nls()} \textbf{man} page states, when the ``port''
algorithm is used with the \texttt{trace} argument TRUE, the iterations
display the objective function value which is 1/2 the sum of squares (or
deviance). The trace display is likely embedded in the Fortran of the
\texttt{nlminb} routine that is called to execute the ``port''
algorithm, but the factor of 2 discrepancy is nonetheless unfortunate
for users.

\hypertarget{failure-to-return-the-best-result-achieved}{%
\subsubsection{Failure to return the best result
achieved}\label{failure-to-return-the-best-result-achieved}}

If \texttt{nls()} reaches a point where it cannot continue but has not
found a point where the relative offset convergence criterion is met, it
may simply exit, especially if a ``singular gradient'' (singular
Jacobian) is found. However, this may occur AFTER the function has made
considerable progress in reducing the sum of squared residuals. Here is
an abbreviated example:

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
\NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
\NormalTok{NLSstart }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{lrc1 =} \SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\AttributeTok{lrc2 =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{A1 =} \DecValTok{150}\NormalTok{, }\AttributeTok{A2 =} \DecValTok{50}\NormalTok{) }\CommentTok{\# a starting vector (named!)}
\NormalTok{NLSformula }\OtherTok{\textless{}{-}}\NormalTok{ conc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ A1 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc1) }\SpecialCharTok{*}\NormalTok{ time) }\SpecialCharTok{+}\NormalTok{ A2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc2) }\SpecialCharTok{*}\NormalTok{ time)}
\NormalTok{tryit }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(NLSformula, }\AttributeTok{data =}\NormalTok{ NLSdata, }\AttributeTok{start =}\NormalTok{ NLSstart, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
##      61216. (3.56e+03): par = (-2 0.25 150 50)
## 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
## 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
## Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
##   singular gradient
\end{verbatim}

\begin{Shaded}
\begin{Highlighting}[]
\FunctionTok{print}\NormalTok{(tryit)}
\end{Highlighting}
\end{Shaded}

\begin{verbatim}
## [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
## attr(,"class")
## [1] "try-error"
## attr(,"condition")
## <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
\end{verbatim}

Note that the sum of squares has been reduced from 61216 to 1.6211, but
unless \texttt{trace} is invoked, the user will not get any information
about this. This almost trivial change to the \texttt{nls()} function
could be useful to R users.

\hypertarget{estimating-models-that-are-partially-linear}{%
\subsection{Estimating models that are partially
linear}\label{estimating-models-that-are-partially-linear}}

The variable projection method (Golub and Pereyra
(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
effective than general approaches in finding good solutions to nonlinear
least squares problems when some of the parameters appear linearly. In
our logistic examples, the asymptote parameters are an illustration.
However, identifying which parameters are linear and communicating this
information to estimating functions is not a trivial task.
\texttt{nls()} has an option \texttt{algorithm="plinear"} that allows
some partially linear models to be solved. The other tools, as far as we
are aware, do not offer any such capability. The \texttt{nlstac} package
uses a different algorithm for similar goals.

Within \texttt{nls()} itself we must, unfortunately, use different
specifications with different \texttt{algorithm} options. For example,
the explicit model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not
work with the linear modeling function \texttt{lm()}, which requires
this model to be specified as \texttt{y\ \textasciitilde{}\ x}. Within
\texttt{nls()}, consider the following FOUR different specifications for
the same problem, plus an intuitive choice, labeled \texttt{fm2a}, that
does NOT work. In this failed attempt, putting the \texttt{Asym}
parameter in the model causes the \texttt{plinear} algorithm to try to
add another term to the model. We believe this is unfortunate, and would
like to see a consistent syntax. At the time of writing, we do not
foresee a resolution for this issue. In the example, we have NOT
evaluated the commands to save space.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{DNase1 }\OtherTok{\textless{}{-}} \FunctionTok{subset}\NormalTok{(DNase, Run }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\CommentTok{\# select the data}
\DocumentationTok{\#\# using a selfStart model {-} do not specify the starting parameters}
\NormalTok{fm1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(}\FunctionTok{log}\NormalTok{(conc), Asym, xmid, scal), DNase1)}
\FunctionTok{summary}\NormalTok{(fm1)}

\DocumentationTok{\#\# using conditional linearity {-} leave out the Asym parameter}
\NormalTok{fm2 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
\FunctionTok{summary}\NormalTok{(fm2)}

\DocumentationTok{\#\# without conditional linearity}
\NormalTok{fm3 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1,}
                 \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm3)}

\DocumentationTok{\#\# using Port\textquotesingle{}s nl2sol algorithm}
\NormalTok{fm4 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm4)}

\DocumentationTok{\#\# using conditional linearity AND Asym does NOT work}
\NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
\FunctionTok{summary}\NormalTok{(fm2a)}
\end{Highlighting}
\end{Shaded}

\hypertarget{models-with-indexed-parameters}{%
\subsection{Models with indexed
parameters}\label{models-with-indexed-parameters}}

Some models have several common parameters and others that are tied to
particular cases. The \textbf{man} file for \texttt{nls()} includes an
example of a situation in which parameters are indexed but which uses
the ``plinear'' option as an added complication. Running this example
reveals that the answers for the parameters are NOT indexed as in a
vector. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
but \texttt{a1,\ a2,\ a3}. This is no doubt because programming for
indexed parameters is challenging. We note that there are capabilities
in packages for mixed-effects modeling such as \texttt{nlme} (Pinheiro
et al. (\protect\hyperlink{ref-p-nlme}{2013})), \texttt{bbmle} (Bolker
and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4}
(p-lme4) for estimating models of such type.

\hypertarget{tests-and-use-case-examples}{%
\subsection{Tests and use-case
examples}\label{tests-and-use-case-examples}}

Maintainers of packages need suitable tests and use-case examples in
order

\begin{itemize}
\tightlist
\item
  to ensure packages work properly, in particular, giving results
  comparable to or better than the functions they are to replace.
\item
  to test individual solver functions to ensure they work across the
  range of calling mechanisms, that is, different ways of supplying
  inputs to the solver(s);
\item
  to pose ``silly'' inputs to see if these bad inputs are caught by the
  programs.
\end{itemize}

Such goals align with the aims of \textbf{unit testing} (e.g.,
\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
package testing tools). In our work, one of us has developed a working
prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}}
. A primary design objective of this is to allow the summarisation of
multiple tests in a compact output. The prototype has a vignette to
illustrate its use.

\hypertarget{documentation-and-resources}{%
\subsection{Documentation and
resources}\label{documentation-and-resources}}

In our investigation, we built several resources, which are now part of
the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
Particular items are:

\begin{itemize}
\item
  A BibTex bibliography for use with all documents in this project, but
  which has wider application to nonlinear least squares projects in
  general
  (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
\item
  \texttt{MachID.R} offers a suggested concise summary function to
  identify a particular computational system used for tests. A
  discussion of how it was built and the resources needed across
  platforms is given in at
  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
\item
  John C. Nash and Bhattacharjee
  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
  in R-base.
\item
  As the 2021 Summer of Code period was ending, one of us (JN) was
  invited to prepare a review of optimization in R. Ideas from the
  present work have been instrumental in the creation of John C. Nash
  (\protect\hyperlink{ref-NashWires22}{2022}).
\end{itemize}

\hypertarget{future-of-nonlinear-model-estimation-in-r}{%
\subsection{Future of nonlinear model estimation in
R}\label{future-of-nonlinear-model-estimation-in-r}}

Given its importance to R, it is possible that \texttt{nls()} will
remain more or less as it has been for the past several decades. If so,
the focus of discussion should be the measures needed to secure its
continued operation for legacy purposes and how that may be
accomplished. We welcome an opportunity to participate in such
conversations.

To advance the stability and maintainability of R, we believe the
program objects (R functions) that are created by tools such as
\texttt{nls()} should have minimal cross-linkages and side-effects. The
aspects of \texttt{nls()} that concern us follow.

\begin{itemize}
\item
  R tools presume data and parameters needed are available in an
  accessible environment. This provides a compact syntax to invoke the
  calculations, but the wrong data can be used if the internal search
  finds a valid name that is not the object we want, or if
  \texttt{subset} or \texttt{na.action} settings modify the selection,
  or \texttt{weights} are applied.
\item
  Mixing of R, C and Fortran code adds to the burden of following the
  program logic.
\item
  The class \texttt{nls} structure simplifies calls with its rich set of
  functions, but also adds to the task of understanding what has been
  done. A design that isolates the setup, solution, and post-solution
  parts of complicated calculations reduces the number of objects that
  must be kept in alignment.
\end{itemize}

Given the existence of examples of good practices such as analytic
derivatives, stabilized solution of Gauss-Newton equations, and
bounds-constrained parameters, base R tools should be moving to
incorporate them. These capabilities are available now by using several
tools, but it would be helpful if they were unified.

\hypertarget{acknowledgments}{%
\section{Acknowledgments}\label{acknowledgments}}

Hans Werner Borchers was helpful in developing the GSoC project
motivating this article and in comments on this and related work.
Heather Turner co-mentored the project and helped guide the progress of
the work. Exchanges with Fernando Miguez helped to clarify aspects of
\texttt{selfStart} models and instigated the ``Introduction to nlsr''
vignette. Colin Gillespie (package \texttt{benchmarkme}) has been
helpful in guiding our attempts to succinctly summarize computing
environments.

\hypertarget{references}{%
\section{References}\label{references}}

\hypertarget{refs}{}
\begin{CSLReferences}{1}{0}
\leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear
Least-Squares Regression}.
\url{https://CRAN.R-project.org/package=onls}.

\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis
and Its Applications}. Wiley.

\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
\emph{Technometrics} 23 (2): 179--83.

\leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools:
Tools for Nonlinear Regression Diagnostics}.

\leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression
Models Dedicated to Predictive Microbiology}.

\leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for
General Maximum Likelihood Estimation}.
\url{http://CRAN.R-project.org/package=bbmle}.

\leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}.
\url{https://CRAN.R-project.org/package=gslnls}.

\leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for
Non-Linear Regression}.
\url{https://CRAN.R-project.org/package=nlshelper}.

\leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben
Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt
Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for
Bounds}. R Project for Statistical Computing.
\url{http://CRAN.R-project.org/package=minpack.lm}.

\leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard
Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third
Edition, for GSL Version 1.12 (3. Ed.).}

\leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad,
Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004.
{``Bioconductor: Open Software Development for Computational Biology and
Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
\url{https://doi.org/10.1186/gb-2004-5-10-r80}.

\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of
Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables
Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.

\leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute
Force}. \url{https://CRAN.R-project.org/package=nls2}.

\leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting
of Non-Linear Regression Functions by Least Squares.''}
\emph{Technometrics} 3: 269-\/-280.

\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004.
\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with
{S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.

\leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for
Nonlinear Least Squares Solutions}}.

\leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. {``Adaptive
Population-Based Search: Application to Estimation of Nonlinear
Regression Parameters.''} \emph{{Computational Statistics \& Data
Analysis}} 52 (2): 713--24.
\url{https://doi.org/10.1016/j.csda.2006.10.014}.

\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain
Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied
Mathematics} 2: 164-\/-168.

\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation
of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics}
11 (2): 431--41.

\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for
Agricultural Applications}}.
\url{https://CRAN.R-project.org/package=nlraa}.

\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on
a Small Computer.''} \emph{Journal of the Institute for Mathematics and
Its Applications} 19: 231--37.

\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
---------. 1979. \emph{Compact Numerical Methods for Computers: Linear
Algebra and Function Minimisation}. Bristol: Adam Hilger.

\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
---------. 2022. {``{Function minimization and nonlinear least squares
in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
https://doi.org/\url{https://doi.org/10.1002/wics.1580}.

\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package
from Base {R} Files.''} \emph{R-Bloggers}, January.
\url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.

\leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A
Replacement and Extension of the {optim()} Function}. Nash Information
Services Inc.; Johns Hopkins University.

\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation
Combining Visual Fitting with Optimization Methods.''} In
\emph{Proceedings of the Section on Physical and Engineering Sciences of
the American Statistical Association}, 256--61. American Statistical
Association.

\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
National Institutes for Standards, original from, and Technology
http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by
Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples
from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.

\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
Nonlinear Least Squares Problems.''} \emph{Computational Optimization
and Applications} 54 (3): 579--93.

\leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart:
Robust Non-Linear Regression Using AIC Scores}.
\url{https://CRAN.R-project.org/package=nls.multstart}.

\leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R
Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.

\leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn:
Fitting Nonlinear Models with Scale Mixture of Skew-Normal
Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.

\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified
Practical Approach}. New York; Basel: Marcel Dekker Inc.

\leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and
Rafael Benitez. 2020. \emph{Nlstac: An r Package for Fitting Separable
Nonlinear Models}. \url{https://CRAN.R-project.org/package=nlstac}.

\leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%
Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with
Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.

\leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters
Estimation by 'CRS4HC' and 'CRS4HCe'}.
\url{https://CRAN.R-project.org/package=crsnls}.

\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
\emph{The R Journal} 3: 5--10.
\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.

\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
\emph{{devtools: Tools to Make Developing R Packages Easier}}.
\url{https://CRAN.R-project.org/package=devtools}.

\end{CSLReferences}

\end{document}

### 2023-05-14 RefactoringNLS/ComparingNLStools.Rmd

@@
 ---
-title: "A comparison of R tools for nonlinear least squares modeling"
+title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
 author:
-   - John C. Nash \thanks{ retired professor, Telfer School of Management, University of Ottawa}
-   - Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}
+- John C. Nash \thanks{ retired professor, Telfer School of Management, University
+  of Ottawa}
+- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
+  Institute of Technology, Kanpur}
 date: "2023-3-2"
@@
   pdf_document:
-        keep_tex: true
-        toc: false
+    keep_tex: yes
+    toc: no
+  word_document:
+    toc: no
 bibliography: ./ImproveNLS.bib
@@
 singular values is a simple but effective measure. The ratios are
-4.6641e-05 for Logistic3U, 0.021022 for Logistic3S and 0.001055 for
+4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
 Logistic3T, so Logistic3S is the "least singular".
@@
 Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
-offer the singular values, we need to extract the Jacobian and compute its
+offer singular values, we need to extract the Jacobian and compute its
 singular values. The following script shows how to do this, using as Jacobian what is called
@@
 a vector `y` that is the expected value of what we have called the
-residual, but in this case is actually the model. `gsl_nls` uses 
+residual, but in this case, is actually the model. `gsl_nls` uses 
 a numerical approximation for the Jacobian if the argument `jac` 
@@
 
-## Design goals, termination tests and output objects
+## Design goals, termination tests, and output objects
 
 The output object of `nlxb()` is smaller than the class `nls` object returned
-by `nls()`, `nlsLM()` and `gsl_nls()`. Package `nlsr` emphasizes the solution 
+by `nls()`, `nlsLM()`, and `gsl_nls()`. Package `nlsr` emphasizes the solution 
 of the nonlinear least squares problem rather than the estimation of a nonlinear 
@@
 a number of specialized modeling and diagnostic extensions. For compatibility, 
-the `nlsr` package has function `wrapnlsr()`, for which `nlsr()` is an alias. 
+the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias. 
 This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
@@
 approached with nonlinear least squares. Fortunately, a small adjustment
-to the "termination test" for the **program**, rather than for 
+to the "termination test" for the **program**, rather than for the
 "convergence" of the underlying **algorithm**, fixes the defect. The test
@@
 Termination after what may be considered excessive computation is also important.
-`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require
+`nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
 examination of the code. We terminate execution when the number of residual or 
-jacobian evaluations exceed set limits. Generally we prefer larger limits than the 
+Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the 
 default `maxiter = 50` of `nls()` to avoid stopping early.
@@
 
-The complexity of this object is a challenge to users. Let us use 
-`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or
+The complexity of this object is a challenge to users. Let us use the
+`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
 `gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
@@
 
-By contrast `nlsr::nlxb()` returns a much simpler structure of 11 items
-in one-level. Moreover, `nlxb` explicitly returns the sum of squares, the residual vector, Jacobian and counts of evaluations.
+By contrast, `nlsr::nlxb()` returns a much simpler structure of 11 items
+in one level. Moreover, `nlxb` explicitly returns the sum of squares, the residual vector, Jacobian, and counts of evaluations.
 
@@
 lead to different consequences for performance and how features are
-provided. `nlsr` has antecendants in the methods of @cnm79, where storage for
+provided. `nlsr` has antecedents in the methods of @cnm79, where storage for
 data and programs was at a ridiculous premium in the small computers of the era.
-Thus the code in `nlsr` is likely of value for workers to copy and modify it
+Thus the code in `nlsr` is likely of value for workers to copy and modify
 for customized tools.
@@
 The `nls()` function, after
-computing the "right hand side" or `rhs` of the residual, checks to see if the
+computing the "right-hand side" or `rhs` of the residual, checks to see if the
 "gradient" attribute is defined, otherwise using `numericDeriv()` to compute a 
@@
 To provide simpler starting parameters, the function `SSlogisJN`
-is now part of package `nlsr`, but is most useful for `nls()`.
+is now part of the package `nlsr`, but is most useful for `nls()`.
 
@@
 
-For many problems we know that parameters cannot be bigger or smaller than some
+For many problems, we know that parameters cannot be bigger or smaller than some
 externally known limits. Such limits should be built into models, but there 
@@
 - `nlsLM()` includes bounds in the standard call, but we have observed cases
-   where it fails to get the correct answer. From examination of the code,
+   where it fails to get the correct answer. From an examination of the code,
    we believe the authors have not taken into account all possibilities, though
@@
 pnls(anlsLM1b)
-# also no warning if starting out of bounds, but gets good answer!!
+# also no warning if starting out of bounds, but gets a good answer!!
 st4<-c(c1=4, c2=4, c3=4)
@@
 ```{r ex10m, echo=TRUE}
-# Hobbsmaskx.R -- masks with formula specification of problem
+# Hobbsmaskx.R -- masks with formula specification of the problem
 require(nlsr); require(minpack.lm); traceval<-FALSE
@@
 `nlsr` has an output format that indicates the constraint status of the parameter
-estimates. For `nlsr` we have **chosen** to suppress calculation of 
+estimates. For `nlsr`, we have **chosen** to suppress the calculation of 
 approximate standard errors in the parameters when constraints are active because
@@
 equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
-while `nlsr`, `gslnls` and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
+while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
 offers an accelerated Marquardt method and three alternative methods which we
@@
 An important choice made in developing `nlsr` was to code entirely within the R 
-programming language. `nls()` uses a mix of R, C and Fortran, as does `minpack.lm`. 
+programming language. `nls()` uses a mix of R, C, and Fortran, as does `minpack.lm`. 
 `gslnls` is an R wrapper to various C-language routines in the Gnu Scientific 
@@
 than most computing environments because it keeps track of objects and 
-because it is usually interpreted. In recent years the 
+because it is usually interpreted. In recent years, the 
 performance penalty for using code entirely in R has been much reduced
 with the just-in-time compiler and other improvements. All-R computation 
-may now offer acceptable performance. In `nlsr` the use of R may be a
+may now offer acceptable performance. In `nlsr`, the use of R may be a
 smaller performance cost than the aggressive approach to a solution,
@@
 
-## Data sources to problems
+## Data sources for problems
 
@@
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit, otherwise any such file in 
+suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
 the path will be loaded on startup. R users in our
@@
 Within `nls()`, consider the following FOUR different specifications for the same 
-problem, plus an intuitive choice, labelled `fm2a`, that does NOT work. 
+problem, plus an intuitive choice, labeled `fm2a`, that does NOT work. 
 In this failed attempt, putting the `Asym` parameter in the model causes the
@@
 <!-- propose an alternative approach below. -->
-We note that there are capabilities in packages for mixed-effects modelling
+We note that there are capabilities in packages for mixed-effects modeling
 such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (p-lme4) for 
@@
 Given the existence of examples of good practices such as analytic derivatives, 
-stabilized solution of Gauss-Newton equations and bounds-constrained parameters, 
+stabilized solution of Gauss-Newton equations, and bounds-constrained parameters, 
 base R tools should 
@@
 
-# Acknowledgement
-
-Hans Werner Borchers was helpful in developing GSoC project motivating this article
+# Acknowledgments
+
+Hans Werner Borchers was helpful in developing the GSoC project motivating this article
 and in comments on this and related work. Heather Turner co-mentored the project 

### 2023-05-17 JNABsub20230515/ComparingNLStools.Rmd
add submission may 23

--- 2023-05-14 RefactoringNLS/ComparingNLStools.Rmd
@@
 title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
-author:
-- John C. Nash \thanks{ retired professor, Telfer School of Management, University
-  of Ottawa}
-- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
-  Institute of Technology, Kanpur}
-date: "2023-3-2"
-output:
-  pdf_document:
-    keep_tex: yes
-    toc: no
-  word_document:
-    toc: no
-bibliography: ./ImproveNLS.bib
-link-citations: yes
-linkcolor: red
-urlcolor: blue
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-## require(bookdown) # language engine to display text - does not seem necessary
-```
-
-## Abstract
-
-Our Google Summer of Code project "Improvements to `nls()`" considered the features and limitations of the R function `nls()` with the aim 
+date: "2023-5-16"
+abstract: >
+  Our Google Summer of Code project "Improvements to `nls()`" considered the 
+  features and limitations of the R function `nls()` with the aim 
   of improving and rationalizing R tools for nonlinear regression.
@@
   nonlinear estimation tools.
+
+draft: true
+author:
+  # see ?rjournal_article for more information
+    - name: John C. Nash
+      affiliation: retired professor, University of Ottawa
+      address:
+      - Telfer School of Management
+      - Ottawa ON Canada K1N 6N5
+      orcid: 0000-0002-2762-8039
+      email: profjcnash@gmail.com
+    - name: Arkajyoti Bhattacharjee
+      affiliation: Indian Institute of Technology
+      address:
+      - Department of Mathematics and Statistics
+      - Kanpur
+      email: arkastat98@gmail.com
+type: package
+output:
+  rjtools::rjournal_web_article:
+    self_contained: yes
+    toc: no
+bibliography: ./ComparingNLStools.bib
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+```
 
@@
 `nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
-Fernandez Torvisco et al. (2018) for fitting separable nonlinear regression curves. The 
+@Torvisco2018 for fitting separable nonlinear regression curves. The 
 special class of problem for which this package is intended is an important and 
@@
 since it was introduced. The default value 
-gives the legacy behaviour. This improvement has been in the R distributed code 
+gives the legacy behavior. This improvement has been in the R distributed code 
 since version 4.1.0.
@@
 `e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+%% ^ is picked up in rjtools check
 
@@
 the original Fortran `minpack` code, namely, `lmdif.f`. Such differences in approach
-can lead to different behaviour, usually minor, but sometimes annoying with
+can lead to different behavior, usually minor, but sometimes annoying with
 ill-conditioned problems.
@@
 equations. `nls()` uses a modification of an approach suggested by @Hartley1961,
-while `nlsr`, `gslnls`, and `minpack.lm` use flavours of @Marquardt1963. `gslnls`
+while `nlsr`, `gslnls`, and `minpack.lm` use flavors of @Marquardt1963. `gslnls`
 offers an accelerated Marquardt method and three alternative methods which we
@@
 NA elements so that predictions are in the correct row position. We recommend that 
-workers actually test output to verify the behaviour is as wanted.
+workers actually test output to verify the behavior is as wanted.
 See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
@@
 The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
-as does, for example, `result$m$resid()`. The function `nlsModel()`, which we 
-have had to extract from the base R code and explicitly `source()` because it 
-is not exported to the working namespace,
+as does, for example, `result$m$resid()`. The function `nlsModel()` 
 allows us to compute residuals for particular coefficient sets.
+We have had to extract from the base R code and include it via a code chunk 
+(not echoed here for space) because it is not exported to the working namespace,
+We could also explicitly `source()` this code.
+
+```{r nlsmodelsource, echo=FALSE}
+#  From file src/library/stats/R/nls.R
+#  Part of the R package, https://www.R-project.org
+#
+#  Copyright (C) 2000-2020 The R Core Team
+#  Copyright (C) 1999-1999 Saikat DebRoy, Douglas M. Bates, Jose C. Pinheiro
+#
+#  The nlsModel() function extracted here.
+
+nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
+{
+    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
+    env <- new.env(hash = TRUE, parent = environment(form))
+    for(i in names(data)) env[[i]] <- data[[i]]
+    ind <- as.list(start)
+    parLength <- 0L
+    for(i in names(ind) ) {
+        temp <- start[[i]]
+        storage.mode(temp) <- "double"
+        env[[i]] <- temp
+        ind[[i]] <- parLength + seq_along(temp)
+        parLength <- parLength + length(temp)
+    }
+    getPars.noVarying <- function() unlist(mget(names(ind), env))
+    getPars <- getPars.noVarying
+    internalPars <- getPars()
+
+    if(!is.null(upper)) upper <- rep_len(upper, parLength)
+    useParams <- rep_len(TRUE, parLength)
+    lhs <- eval(form[[2L]], envir = env)
+    rhs <- eval(form[[3L]], envir = env)
+    .swts <- if(!missing(wts) && length(wts))
+        sqrt(wts) else rep_len(1, length(rhs))
+    env$.swts <- .swts
+    resid <- .swts * (lhs - rhs)
+    dev <- sum(resid^2)
+    if(is.null(attr(rhs, "gradient"))) {
+        getRHS.noVarying <- function() {
+            if(is.null(upper)) # always for "default"
+                numericDeriv(form[[3L]], names(ind), env, central = nDcentral)
+            else # possibly with "port"
+                numericDeriv(form[[3L]], names(ind), env,
+                             dir = ## ifelse(internalPars < upper, 1, -1)
+                                 -1 + 2*(internalPars < upper), central = nDcentral)
+        ## 210701JN: Above line should check for nDcentral=FALSE when approaching
+        ## upper bound, as central diff is same whether dir is +1 or -1. ???
+        }
+        getRHS <- getRHS.noVarying
+        rhs <- getRHS()
+    } else {
+        getRHS.noVarying <- function() eval(form[[3L]], envir = env)
+        getRHS <- getRHS.noVarying
+    }
+    dimGrad <- dim(attr(rhs, "gradient"))
+    marg <- length(dimGrad)
+    if(marg > 0L) {
+        gradSetArgs <- vector("list", marg + 1L)
+        for(i in 2L:marg)
+            gradSetArgs[[i]] <- rep_len(TRUE, dimGrad[i-1L])
+        useParams <- rep_len(TRUE, dimGrad[marg])
+    } else {
+        gradSetArgs <- vector("list", 2L)
+        useParams <- rep_len(TRUE, length(attr(rhs, "gradient")))
+    }
+    npar <- length(useParams)
+    gradSetArgs[[1L]] <- (~attr(ans, "gradient"))[[2L]]
+    gradCall <-
+        switch(length(gradSetArgs) - 1L,
+               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], drop = FALSE),
+               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
+                    drop = FALSE),
+               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
+                    gradSetArgs[[3L]], drop = FALSE),
+               call("[", gradSetArgs[[1L]], gradSetArgs[[2L]], gradSetArgs[[2L]],
+                    gradSetArgs[[3L]], gradSetArgs[[4L]], drop = FALSE))
+    getRHS.varying <- function()
+    {
+        ans <- getRHS.noVarying()
+        attr(ans, "gradient") <- eval(gradCall)
+        ans
+    }
+    if(length(gr <- attr(rhs, "gradient")) == 1L)
+		    attr(rhs, "gradient") <- gr <- as.vector(gr)
+    QR <- qr(.swts * gr)
+    qrDim <- min(dim(QR$qr))
+    if(QR$rank < qrDim)
+        stop("singular gradient matrix at initial parameter estimates")
+
+    getPars.varying <- function() unlist(mget(names(ind), env))[useParams]
+    setPars.noVarying <- function(newPars)
+    {
+        internalPars <<- newPars # envir = thisEnv
+        for(i in names(ind))
+            env[[i]] <- unname(newPars[ ind[[i]] ])
+    }
+    setPars.varying <- function(newPars)
+    {
+        internalPars[useParams] <<- newPars
+        for(i in names(ind))
+            env[[i]] <- unname(internalPars[ ind[[i]] ])
+    }
+    setPars <- setPars.noVarying
+
+    if(scaleOffset) scaleOffset <- (length(resid)-npar) * scaleOffset^2
+    convCrit <- function() {
+        if(npar == 0) return(0)
+        rr <- qr.qty(QR, c(resid)) # rotated residual vector
+        sqrt( sum(rr[1L:npar]^2) / (scaleOffset + sum(rr[-(1L:npar)]^2)))
+    }
+
+    on.exit(remove(i, data, parLength, start, temp, m, gr,
+                   marg, dimGrad, qrDim, gradSetArgs))
+    ## must use weighted resid for use with "port" algorithm.
+    m <-
+	list(resid = function() resid,
+	     fitted = function() rhs,
+	     formula = function() form,
+	     deviance = function() dev, ##?? PROBLEM -- does not get updated
+	     lhs = function() lhs,
+	     gradient = function() .swts * attr(rhs, "gradient"),
+	     conv = function() convCrit(),
+	     incr = function() qr.coef(QR, resid),
+	     setVarying = function(vary = rep_len(TRUE, np)) {
+                 np <- length(useParams)
+		 useParams <<- useP <-
+                     if(is.character(vary)) {
+                         temp <- logical(np)
+                         temp[unlist(ind[vary])] <- TRUE
+                         temp
+                     } else if(is.logical(vary) && length(vary) != np)
+                         stop("setVarying : 'vary' length must match length of parameters")
+                     else
+                         vary # envir = thisEnv
+		 gradCall[[length(gradCall) - 1L]] <<- useP
+		 if(all(useP)) {
+		     setPars <<- setPars.noVarying
+		     getPars <<- getPars.noVarying
+		     getRHS  <<-  getRHS.noVarying
+		     npar    <<- length(useP)
+		 } else {
+		     setPars <<- setPars.varying
+		     getPars <<- getPars.varying
+		     getRHS  <<-  getRHS.varying
+		     npar    <<- sum(useP)
+		 }
+	     },
+	     setPars = function(newPars) {
+		 setPars(newPars)
+		 resid <<- .swts * (lhs - (rhs <<- getRHS())) # envir = thisEnv {2 x}
+		 dev   <<- sum(resid^2) # envir = thisEnv
+		 if(length(gr <- attr(rhs, "gradient")) == 1L) gr <- c(gr)
+		 QR <<- qr(.swts * gr) # envir = thisEnv
+		 (QR$rank < min(dim(QR$qr))) # to catch the singular gradient matrix
+	     },
+	     getPars = function() getPars(),
+	     getAllPars = function() getPars(),
+	     getEnv = function() env,
+	     trace = function() {
+		 d <- getOption("digits")
+		 cat(sprintf("%-*s (%.2e): par = (%s)\n", d+4L+2L*(scaleOffset > 0),
+			     formatC(dev, digits=d, flag="#"),
+			     convCrit(),
+			     paste(vapply(getPars(), format, ""), collapse=" ")))
+	     },
+	     Rmat = function() qr.R(QR),
+	     predict = function(newdata = list(), qr = FALSE)
+                 eval(form[[3L]], as.list(newdata), env)
+	     )
+    class(m) <- "nlsModel"
+    m
+}
+```
 
@@
 attr(rusewt,"gradient")<-NULL; rusewt
-source("nlsModel.R")
+## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
 nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
@@
 In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
-A primary design objective of this is to allow the summarisation of multiple tests in
+A primary design objective of this is to allow the summarization of multiple tests in
 a compact output. The prototype has a vignette to illustrate its use.

### 2023-05-17 JNABsub20230515/ComparingNLStools.tex
add submission may 23

--- 2023-05-14 RefactoringNLS/ComparingNLStools.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{lmodern}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math}
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
-\hypersetup{
-  pdftitle={A comparison of R tools for nonlinear least squares modeling},
-  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
-  colorlinks=true,
-  linkcolor={red},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\title{A comparison of R tools for nonlinear least squares modeling}
-\author{John C. Nash
-\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
-Bhattacharjee
-\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
-\date{2023-3-2}
-
-\begin{document}
+% !TeX root = RJwrapper.tex
+\title{A Comparison of R Tools for Nonlinear Least Squares Modeling}
+\author{by John C. Nash and Arkajyoti Bhattacharjee}
+
 \maketitle
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
-
-Our Google Summer of Code project ``Improvements to \texttt{nls()}''
-considered the features and limitations of the R function \texttt{nls()}
-with the aim of improving and rationalizing R tools for nonlinear
-regression. The rich features of \texttt{nls()} are weakened by several
-deficiencies and inconsistencies such as a lack of stabilization of the
-Gauss-Newton solver. Further considerations are the usability and
-maintainability of the code base that provides the functionality
-\texttt{nls()} claims to offer. Various packages, including our
-\texttt{nlsr}, provide alternative capabilities. We consider the
-differences in goals, approaches, and features of different tools for
-nonlinear least squares modeling in R. Discussion of these matters is
-relevant to improving R generally as well as its nonlinear estimation
-tools.
+\abstract{%
+Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression.
+The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
+Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} claims to offer.
+Various packages, including our \texttt{nlsr}, provide alternative capabilities.
+We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R.
+Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
+}
 
 \hypertarget{the-nls-function}{%
-\subsection{\texorpdfstring{The \texttt{nls()}
-function}{The nls() function}}\label{the-nls-function}}
-
-\texttt{nls()} is the tool for estimating nonlinear statistical models
-in base R, the primary software distribution from the Comprehensive R
-Archive Network (\emph{\url{https://cran.r-project.org}}). The function
-dates to the 1980s and the work related to Bates and Watts
-(\protect\hyperlink{ref-bateswatts}{1988}) in S (see
+\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}
+
+\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
+primary software distribution
+from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
+The function dates to the 1980s and the work related to Bates and Watts (1988) in S (see
 \emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
 
-The \texttt{nls()} function has a remarkable and comprehensive set of
-capabilities for estimating nonlinear models that are expressed as
-\textbf{formulas}. In particular, we note that it
+The \texttt{nls()} function has a remarkable and comprehensive set of capabilities for
+estimating nonlinear models that are expressed as \textbf{formulas}. In particular, we
+note that it
 
@@
 \item
-  handles formulas that include R functions, even ones that call
-  calculations in other programming languages
+  handles formulas that include R functions, even ones that call calculations in
+  other programming languages
 \item
@@
 \item
-  produces measures of variability (i.e., standard error estimates) for
-  the estimated parameters
-\item
-  has related profiling capabilities for exploring the likelihood
-  surface as parameters are changed
-\item
-  links to many pre-coded (\texttt{selfStart}) models that do not
-  require initial parameter values.
+  produces measures of variability (i.e., standard error estimates) for the
+  estimated parameters
+\item
+  has related profiling capabilities for exploring the likelihood surface as
+  parameters are changed
+\item
+  links to many pre-coded (\texttt{selfStart}) models that do not require
+  initial parameter values.
 \end{itemize}
 
-With such a range of features and long history, the code has become
-untidy and overly patched, difficult to maintain, and its underlying
-methods could be improved. Various workers have developed packages to
-overcome these concerns, and we will address some of these here.
+With such a range of features and long history, the code has become untidy
+and overly patched, difficult to maintain, and its underlying methods could
+be improved. Various workers have developed packages to overcome these
+concerns, and we will address some of these here.
 
@@
 
-The tools we will consider are the base-R \texttt{nls()} function and
-some packages that are available on the CRAN repository. Of these, we
-will pay particular attention to \textbf{nlsr} (John C Nash and Duncan
-Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
-\textbf{minpack.lm} (Elzhov et al.
-(\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau
-(\protect\hyperlink{ref-gslnls23}{2023})) which are general nonlinear
-least-squares solvers.
-
-While we will provide capsule comments for some other CRAN packages, we
-will not pursue those in the Bioconductor (Gentleman et al.
-(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
-repositories such as GitHub (\url{https://github.com}) and Gitlab
-(\url{https://about.gitlab.com}).
+The tools we will consider are the base-R \texttt{nls()} function
+and some packages that are available on the CRAN repository.
+Of these, we will pay particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch (2023)),
+\textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
+general nonlinear least-squares solvers.
+
+While we will provide capsule comments for some other CRAN packages,
+we will not pursue those in the Bioconductor (Gentleman et al. (2004))
+collection, nor those on repositories such as
+GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}).
 
 \hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
-\subsection{Some other CRAN packages for nonlinear
-modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
-
-\texttt{onls} (Andrej-Nikolai Spiess
-(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
-in estimating nonlinear models is the vertical difference between the
-dependent variable and the functional model. \texttt{onls} minimizes the
-sum of \textbf{orthogonal} residuals. The objective is therefore
-different and involves nontrivial extra calculation. A vignette with the
-package and the blog article
+\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
+
+\texttt{onls} (Andrej-Nikolai Spiess (2022)) -- The usual optimization in estimating nonlinear models
+is the vertical difference between the dependent variable and the functional
+model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
+therefore different and involves nontrivial extra calculation. A vignette with
+the package and the blog article
 \url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
-give some description with illustrative graphs. \texttt{onls} appears to
-be limited to problems with one independent and one dependent variable.
-The Wikipedia article
-\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
-overview of some ideas, with references to the literature. The approach
-needs a wider discussion and tutorial examples to allow its merits to be
-judged than can be included here.
-
-\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
--- This package allows nonlinear estimation by controlled random search
-via two methods. There is unfortunately no vignette. A modest trial we
-carried out showed \texttt{nlsr::nlxb()} gave the same results in a
-small fraction of the time required by either of the methods in
-\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
-Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
-reliability in finding solutions than a Levenberg-Marquardt code
-(actually from Matlab), but the tests were conducted on the extreme NIST
-examples mentioned next.
-
-\texttt{NISTnls} (National Institutes for Standards and R port by
-Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
-package provides R code and data for a set of (numerically
-ill-conditioned) nonlinear least squares problems from the U.S. National
-Institute for Standards and Technology. These may not represent
-real-world situations.
-
-\texttt{nlshelper} (Duursma
-(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
-unfortunately lacks a vignette, provides a few utilities for
-summarizing, testing, and plotting non-linear regression models
-estimated with nls(), nlsList() or nlme() that are linked or grouped in
-some way.
-
-\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
-This solves nonlinear least squares problems with optional equality
-and/or inequality constraints. It is clearly \textbf{not} about
-modeling, and the input and output are quite different from class
-\texttt{nls} methods. However, there do not appear to be other R
-packages with these capabilities.
-
-\texttt{nlsMicrobio} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
-nonlinear regression models dedicated to predictive microbiology,
-including a vignette, by authors of the \texttt{nlstools} package.
-
-\texttt{nlstools} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
-several tools for aiding the estimation of nonlinear models,
-particularly using \texttt{nls()}. The vignette is actually a journal
+give some description with illustrative graphs. \texttt{onls} appears to be limited to
+problems with one independent and one dependent variable. The Wikipedia article
+\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
+ideas, with references to the literature. The approach needs a wider
+discussion and tutorial examples to allow its merits to be judged than can be
+included here.
+
+\texttt{crsnls} (Tvrdík (2016)) -- This package allows nonlinear estimation by
+controlled random search via two methods. There is unfortunately no vignette.
+A modest trial we carried out showed \texttt{nlsr::nlxb()} gave the same results in
+a small fraction of the time required by either of the methods in \texttt{crsnls}. The
+method discussed in Josef Tvrdík and Ivan Křivý and Ladislav Mišík (2007) claims better reliability in finding solutions
+than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
+on the extreme NIST examples mentioned next.
+
+\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
+(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
+Institute for Standards and Technology. These may not represent real-world situations.
+
+\texttt{nlshelper} (Duursma (2017)) -- This package, which unfortunately lacks a
+vignette, provides a few utilities for summarizing, testing, and plotting
+non-linear regression models estimated with nls(), nlsList() or nlme() that
+are linked or grouped in some way.
+
+\texttt{nlsic} (Sokol (2022)) -- This solves nonlinear least squares problems with optional
+equality and/or inequality constraints. It is clearly \textbf{not} about modeling, and
+the input and output are quite different from class \texttt{nls} methods. However, there do
+not appear to be other R packages with these capabilities.
+
+\texttt{nlsMicrobio} (Baty and Delignette-Muller (2014)) -- Data sets and nonlinear regression models
+dedicated to predictive microbiology, including a vignette, by authors of the
+\texttt{nlstools} package.
+
+\texttt{nlstools} (Baty and Delignette-Muller (2013)) -- This package provides several tools for aiding the estimation
+of nonlinear models, particularly using \texttt{nls()}. The vignette is actually a journal
 article, and the authors have considerable experience in the subject.
 
-\texttt{nlsmsn} (Prates, Lachos, and Garay
-(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
-scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
-Lachos, and Abanto-Valle (2011). The problem here is to minimize an
-objective that is modified from the traditional sum of squared
+\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN)
+regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
+here is to minimize an objective that is modified from the traditional sum of squared
 residuals.
 
-\texttt{nls.multstart} (Padfield and Matheson
-(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
-squares regression using AIC scores with the Levenberg-Marquardt
-algorithm using multiple starting values for increasing the chance that
-the minimum found is the global minimum.
-
-\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
-Nonlinear least squares by brute force has similar motivations to
-\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
-The author has extensive expertise in R.
-
-\texttt{nlstac} (Rodriguez-Arias et al.
-(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
-implementing the algorithm described in Fernandez Torvisco et al.~(2018)
-for fitting separable nonlinear regression curves. The special class of
-problem for which this package is intended is an important and difficult
-one. No vignette is provided, unfortunately.
-
-\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models
-are treated, but there is minimal documentation and no vignette. Package
-\texttt{nlraa} is to be preferred.
-
-\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
--- a set of nonlinear \emph{selfStart} models, primarily from
+\texttt{nls.multstart} (Padfield and Matheson (2020)) -- Non-linear least squares regression using
+AIC scores with the Levenberg-Marquardt algorithm using multiple starting values
+for increasing the chance that the minimum found is the global minimum.
+
+\texttt{nls2} (Grothendieck (2022)) -- Nonlinear least squares by brute force has similar
+motivations to \texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials. The author
+has extensive expertise in R.
+
+\texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
+Torvisco, Fernandez, and Sanchez (2018) for fitting separable nonlinear regression curves. The
+special class of problem for which this package is intended is an important and
+difficult one. No vignette is provided, unfortunately.
+
+\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models are treated, but
+there is minimal documentation and no vignette. Package \texttt{nlraa} is to be preferred.
+
+\texttt{nlraa} (Miguez (2021)) -- a set of nonlinear \emph{selfStart} models, primarily from
 agriculture. Most include analytic Jacobian code.
 
-\texttt{optimx} (John C. Nash and Varadhan
-(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
-that can be applied to minimize a nonlinear function which could be a
-nonlinear sum of squares. NOT generally recommended if nonlinear least
-squares programs can be easily used, but provides a check and
-alternative solvers.
+\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
+optimizers that can be applied to minimize a nonlinear function which could be
+a nonlinear sum of
+squares. NOT generally recommended if nonlinear
+least squares programs can be easily used, but provides a check and alternative
+solvers.
 
@@
 
-The Hobbs weed infestation problem (John C. Nash
-(\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
-task that seems straightforward but is quite nasty. Its very succinct
-statement provides the ``short reproducible example'' much requested on
-R mailing lists. This is a real problem from a field researcher. The
-data and graph follow.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
-          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
-\FunctionTok{plot}\NormalTok{(weeddf, }\AttributeTok{main=}\StringTok{"Hobbs weed infestation data"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
+task that seems straightforward but is quite nasty. Its very succinct statement
+provides the ``short reproducible example'' much requested on R mailing lists.
+This is a real problem from a field researcher. The data and graph follow.
+
+\begin{verbatim}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+\end{verbatim}
 
@@
 
-Three suggested models for this data are (with names to allow for easy
-reference)
-
-\textbf{Logistic3U}: \[  y \approx  b_1 / (1 + b_2 * exp(- b_3 * t)) \]
+Three suggested models for this data are (with names to allow for easy reference)
+
+\textbf{Logistic3U}:
+\[  y \approx  b_1 / (1 + b_2 * exp(- b_3 * t)) \]
 
@@
 
-\textbf{Logistic3T}: \[ y \approx Asym / (1 + exp((xmid - t)/scal)) \]
-
-where we will use \texttt{weed} for \(y\) and \texttt{tt} for \(t\). The
-functions above are equivalent, but the first is generally more awkward
-to solve numerically due to its poor scaling. The parameters of the
-three forms are related as follows:
-
-\[   Asym =  b_1 = 100 * c_1 \] \[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
-\[ 1/scal  =  b_3 \] To allow for simpler discussion, let us say that
-the parameters form a (named) vector \(p\) and the model function is
-called \(model(p)\). The residuals can be written either as
-\(r_1 = y - model(p)\) or \(r = model(p) - y\) since their sum of
-squares has an identical value. The second form avoids a potential sign
-error if we need to evaluate derivatives.
-
-We wish to minimize the sum of squared residuals, which is our loss (or
-objective) function. Starting with some guess for the parameters, we aim
-to alter these parameters to obtain a smaller loss function. We then
-iterate until we can make no further progress.
-
-Let us consider there are \(n\) parameters and \(m\) residuals. The loss
-function is \[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\] The gradient of
-\(S(p)\) is \[ 2 * J' r\] where the Jacobian \(J\) is given by elements
-\[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is defined
-by elements
-\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \] If we
-expand the Hessian for nonlinear least squares problems, we find
+\textbf{Logistic3T}:
+\[ y \approx Asym / (1 + exp((xmid - t)/scal)) \]
+
+where we will use \texttt{weed} for \(y\) and \texttt{tt} for \(t\).
+The functions above are equivalent, but the first is generally more awkward
+to solve numerically due to its poor scaling. The parameters of the three forms
+are related as follows:
+
+\[   Asym =  b_1 = 100 * c_1 \]
+\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
+\[ 1/scal  =  b_3 \]
+To allow for simpler discussion, let us say that the parameters form a (named) vector
+\(p\) and the model function is called \(model(p)\). The residuals can be written
+either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
+since their sum of squares has an identical value. The second form
+avoids a potential sign error if we need to evaluate derivatives.
+
+We wish to minimize the sum of squared residuals, which is our loss (or objective) function.
+Starting with some guess for the parameters, we aim to alter these parameters to obtain a smaller
+loss function. We then iterate until we can make no further progress.
+
+Let us consider there are \(n\) parameters and \(m\) residuals.
+The loss function is
+\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
+The gradient of \(S(p)\) is
+\[ 2 * J' r\]
+where the Jacobian \(J\) is given by elements
+\[ J_{i,j} = \partial r_i / \partial p_j \]
+and the Hessian is defined by elements
+\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
+If we expand the Hessian for nonlinear least squares problems, we find
 \[  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
-Let us use \(D_{i,j}\) for the elements of the second term of this
-expression. What is generally called \textbf{Newton's method} for
-function minimization tries to set the gradient to zero (to find a
-stationary point of the function \(S(p)\)). This leads to
-\textbf{Newton's equation} \[ H \delta = -g \] Given a set of parameters
-\(p\), we solve this equation for \(\delta\), adjust \(p\) to
-\(p + \delta\) and iterate, hopefully to converge on a solution.
+Let us use \(D_{i,j}\) for the elements of the second term of this expression.
+What is generally called \textbf{Newton's method} for function minimization tries
+to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
+This leads to \textbf{Newton's equation}
+\[ H \delta = -g \]
+Given a set of parameters \(p\), we solve this equation for \(\delta\), adjust \(p\)
+to \(p + \delta\) and iterate, hopefully to converge on a solution.
 Applying this to a sum of squares problem gives
-\[ H \delta = (J'J + D) \delta = - J' r\] In this expression, only the
-elements of \(D\) have second partial derivatives. Gauss, attempting to
-model planetary orbits, had small residuals, and noted that these
-multiplied the second partial derivatives of \(r\), so he approximated
-\[ H \approx J' J\] by assuming \(D \approx 0\). This results in the
-Gauss-Newton method where we solve \[ J' J \delta = - J' r\] though we
-can avoid some loss of accuracy by NOT forming the inner product matrix
+\[ H \delta = (J'J + D) \delta = - J' r\]
+In this expression, only the elements of \(D\) have second partial derivatives.
+Gauss, attempting to model planetary orbits, had small residuals, and noted
+that these multiplied the second partial derivatives of \(r\), so he approximated
+\[ H \approx J' J\]
+by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
+solve
+\[ J' J \delta = - J' r\]
+though we can avoid some loss of accuracy by NOT forming the inner product matrix
 \(J' J\) and solving the linear least squares matrix problem
-\[ J \delta \approx -r \] by one of several matrix decomposition
-methods.
-
-In reality, there are many problems were \(D\) should not be ignored,
-but the work to compute it precisely is considerable. Many work-arounds
-have been proposed, of which the Levenberg-Marquardt stabilization
-(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt
-(\protect\hyperlink{ref-Marquardt1963}{1963})) is the most commonly
-used. For convenience, we will use ``Marquardt'', as we believe he first
-incorporated the ideas into a practical computer program.
-
-The usual suggestion is that \(D\) be replaced by a multiple of the unit
-matrix or else a multiple of the diagonal part of \(J' J\). In low
-precision, some elements of \(J' J\) could underflow to zero (John C.
-Nash (\protect\hyperlink{ref-jn77ima}{1977})), and a linear combination
-of both choices is an effective compromise. Various choices for \(D\),
-as well as a possible line search along the direction \(\delta\) rather
-than a unit step (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})),
-give rise to several variant algorithms. ``Marquardt's method'' is a
-family of methods. Fortunately, most choices work well.
+\[ J \delta \approx -r \]
+by one of several matrix decomposition methods.
+
+In reality, there are many problems were \(D\) should not be ignored, but the work
+to compute it precisely is considerable. Many work-arounds have been
+proposed, of which the Levenberg-Marquardt stabilization
+(Levenberg (1944), Marquardt (1963)) is
+the most commonly used. For convenience, we will use ``Marquardt'', as we believe he
+first incorporated the ideas into a practical computer program.
+
+The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
+a multiple of the diagonal part of \(J' J\). In low precision, some elements of
+\(J' J\) could underflow to zero (John C. Nash (1977)), and a linear combination of both
+choices is an effective compromise. Various choices for
+\(D\), as well as a possible line search along the
+direction \(\delta\) rather than a unit step (Hartley (1961)),
+give rise to several variant
+algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
+work well.
 
@@
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are NOT equivalent and are
-deliberately crude choices. Workers performing many calculations of a
-similar nature should try to provide good starting points to reduce
+values for parameters. These starting points are NOT equivalent and
+are deliberately crude choices. Workers performing many calculations
+of a similar nature should try to provide good starting points to reduce
 computation time and avoid finding a false solution.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# model formulas}
-\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{b2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3}\SpecialCharTok{*}\NormalTok{tt))}
-\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{c1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\DecValTok{10}\SpecialCharTok{*}\NormalTok{c2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1}\SpecialCharTok{*}\NormalTok{c3}\SpecialCharTok{*}\NormalTok{tt))}
-\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid}\SpecialCharTok{{-}}\NormalTok{tt)}\SpecialCharTok{/}\NormalTok{scal))}
-\CommentTok{\#}
-\CommentTok{\# Starting parameter sets}
-\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{1}\NormalTok{, }\AttributeTok{b2=}\DecValTok{1}\NormalTok{, }\AttributeTok{b3=}\DecValTok{1}\NormalTok{)}
-\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{1}\NormalTok{, }\AttributeTok{c2=}\DecValTok{1}\NormalTok{, }\AttributeTok{c3=}\DecValTok{1}\NormalTok{)}
-\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-One of the useful features of \texttt{nls()} is the possibility of a
-\texttt{selfStart} model, where starting parameter values are not
-required. However, if a \texttt{selfStart} model is not available,
-\texttt{nls()} sets all the starting parameters to 1. This is tolerable,
-but could be improved by using a set of values that are all slightly
-different, which, in the case of the example model
-\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
-avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
-Program modifications to give a sequence like 1.0, 1.1, 1.2, 1.3 for the
-four parameters are fairly obvious.
-
-It is also possible to provide R functions for the residual and
-Jacobian. This is usually more work for the user if the formula setup is
-possible. To illustrate, we show the functions for the unscaled 3
-parameter logistic.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Logistic3U}
-\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
-  \CommentTok{\# This variant uses looping}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
-\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
-           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
-\NormalTok{\}}
-
-\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
-\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
-\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)}
-\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{yy)}
-\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
-\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}
-\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}\SpecialCharTok{*}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt}
-  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
-\NormalTok{  jj}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+# model formulas
+frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
+frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
+frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+#
+# Starting parameter sets
+stu1<-c(b1=1, b2=1, b3=1)
+sts1<-c(c1=1, c2=1, c3=1)
+stt1<-c(Asym=1, xmid=1, scal=1)
+\end{verbatim}
+
+One of the useful features of \texttt{nls()} is the possibility of a \texttt{selfStart} model,
+where starting parameter values are not required. However, if a \texttt{selfStart} model
+is not available, \texttt{nls()} sets all the starting parameters
+to 1. This is tolerable, but could be improved by using a set of values
+that are all slightly different, which, in the case of the example
+model \(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\)
+would avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
+Program modifications to give a sequence like 1.0, 1.1, 1.2, 1.3 for the four
+parameters are fairly obvious.
+
+It is also possible to provide R functions for the residual and Jacobian.
+This is usually more work for the user if the formula setup is possible.
+To illustrate, we show the functions for the unscaled 3 parameter logistic.
+
+\begin{verbatim}
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3]*tt)
+  zz  <-  1.0/(1+x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -x[1]*zz*zz*yy
+  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
+  attr(jj, "gradient") <- jj
+  jj
+}
+\end{verbatim}
 
 \hypertarget{estimation-of-models-specified-as-formulas}{%
-\subsection{Estimation of models specified as
-formulas}\label{estimation-of-models-specified-as-formulas}}
-
-Using a formula specification was a principal advantage made with
-\texttt{nls()} when it became available in S sometime in the 1980s. It
-uses a Gauss-Newton (i.e., unstabilized) iteration with a step reduction
-line search. This works very efficiently as long as \(J\) is not
-ill-conditioned. Below we see \texttt{nls()} does poorly on the example
-problem. To save page space, we use 1-line result display functions from
-package \texttt{nlsr}, namely \texttt{pnls()} and \texttt{pshort()}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frmu, start = stu1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frms, start = sts1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frmt, start = stt1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-Here we see the infamous ``singular gradient'' termination message of
-\texttt{nls()}.
+\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}
+
+Using a formula specification was a principal advantage made
+with \texttt{nls()} when it became available in S sometime in the 1980s. It uses a
+Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
+search. This works very efficiently as long as \(J\) is not ill-conditioned.
+Below we see \texttt{nls()} does poorly on the example problem. To save page space,
+we use 1-line result display functions from package \texttt{nlsr}, namely \texttt{pnls()}
+and \texttt{pshort()}.
+
+\begin{verbatim}
+#> Error in nls(formula = frmu, start = stu1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nls(formula = frms, start = sts1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nls(formula = frmt, start = stt1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+Here we see the infamous ``singular gradient'' termination message of \texttt{nls()}.
 
 \hypertarget{solution-attempts-with-nlsr}{%
-\subsubsection{Solution attempts with
-nlsr}\label{solution-attempts-with-nlsr}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{print}\NormalTok{(unlx1) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.5873  on  12 observations
-##     after  19    Jacobian and  25 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
-## b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
-## b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
+\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+
+\begin{verbatim}
+#> residual sumsquares =  2.5873  on  12 observations
+#>     after  19    Jacobian and  25 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
+#> b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
+#> b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
+\end{verbatim}
+
+\begin{verbatim}
+#> snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
+\end{verbatim}
+
+\begin{verbatim}
+#> tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
 \end{verbatim}
@@
 Though we have found solutions, the Jacobian is essentially singular as
-shown by its singular values. Note that these are \textbf{displayed} by
-package \texttt{nlsr} in a single column in the output to provide a
-compact layout, but the values do \textbf{NOT} correspond to the
-individual parameters in whose row they appear; they are a property of
-the whole problem.
+shown by its singular
+values. Note that these are \textbf{displayed} by package \texttt{nlsr} in a single column in
+the output to provide a compact layout, but the values do \textbf{NOT} correspond to
+the individual parameters in
+whose row they appear; they are a property of the whole problem.
 
 \hypertarget{solution-attempts-with-minpack.lm}{%
-\subsubsection{Solution attempts with
-minpack.lm}\label{solution-attempts-with-minpack.lm}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{pnls}\NormalTok{(snlm1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
-\FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
+\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}
+
+\begin{verbatim}
+#> unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
 \end{verbatim}
@@
 \hypertarget{solution-attempts-with-gslnls}{%
-\subsubsection{Solution attempts with
-gslnls}\label{solution-attempts-with-gslnls}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(gslnls)}
-\NormalTok{ugslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
-\FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
-\FunctionTok{pnls}\NormalTok{(sgslnls1) }\CommentTok{\# Use summary() to get display}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
-\FunctionTok{pnls}\NormalTok{(tgslnls1) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20847  scal = -1744.7; 47  itns
+\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}
+
+\begin{verbatim}
+#> ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
 \end{verbatim}
@@
 \hypertarget{comparison-notes-for-formula-setup-solutions}{%
-\subsubsection{Comparison notes for formula-setup
-solutions}\label{comparison-notes-for-formula-setup-solutions}}
-
-\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors
-and singular values of the Jacobian (for diagnostic purposes). By
-contrast, \texttt{minpack.lm::nlsLM()} and \texttt{nls()} use
-\texttt{summary()}, which does NOT display the sum of squares, while
-\texttt{print()} gives the sum of squares, but not the standard error of
-the residuals.
+\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}
+
+\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors and
+singular values of the Jacobian (for diagnostic purposes). By contrast,
+\texttt{minpack.lm::nlsLM()} and \texttt{nls()}
+use \texttt{summary()}, which does NOT display the sum of squares, while \texttt{print()}
+gives the sum of squares, but not the standard error of the residuals.
 
 The singular values allow us to gauge how ``nearly singular'' the
-Jacobian is at the solution, and the ratio of the smallest to largest of
-the singular values is a simple but effective measure. The ratios are
+Jacobian is at the solution, and the ratio of the smallest to largest of the
+singular values is a simple but effective measure. The ratios are
 4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
@@
 
-The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the
-transformed model Logistic3T have a very large sum of squares, which may
-suggest that these programs have failed. Since \texttt{nls()},
-\texttt{nlsLM()}, and \texttt{gsl\_nls()} do not offer singular values,
-we need to extract the Jacobian and compute its singular values. The
-following script shows how to do this, using as Jacobian what is called
-the \texttt{gradient} element in the returned solution for these
-solvers.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# for nlsLM}
-\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tnlm1, }\StringTok{"try{-}error"}\NormalTok{))  \{}
-   \FunctionTok{print}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
-\NormalTok{ \} }\ControlFlowTok{else}\NormalTok{ \{  }
-\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tnlm1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{() }\CommentTok{\# actually the Jacobian}
-   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
-\NormalTok{\}   }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 3.4641e+00 3.2530e-10 8.9643e-12
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# for gsl\_nls}
-\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tgslnls1, }\StringTok{"try{-}error"}\NormalTok{)) \{}
-   \FunctionTok{cat}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
-\NormalTok{\} }\ControlFlowTok{else}\NormalTok{ \{  }
-\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tgslnls1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{()}
-   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
-\NormalTok{\}   }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 3.4641e+00 1.0111e-08 4.7443e-11
-\end{verbatim}
-
-We see that there are differences in detail, but the more important
-result is that two out of three singular values are essentially 0. Our
-Jacobian is singular, and no method of the Gauss-Newton type should be
-able to continue. Indeed, from the parameters reported at this saddle
-point, \texttt{nlsr::nlxb()} cannot proceed.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{stspecial}\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
-\NormalTok{badstart}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stspecial, }\AttributeTok{data=}\NormalTok{weeddf)}
-\FunctionTok{print}\NormalTok{(badstart)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  9205.4  on  12 observations
-##     after  2    Jacobian and  2 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             35.5321            NA         NA         NA  -9.694e-09       3.464  
-## xmid               43376            NA         NA         NA  -1.742e-09    2.61e-10  
-## scal             -2935.4            NA         NA         NA    -2.4e-08    7.12e-16
+The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the transformed model Logistic3T have a very
+large sum of squares, which may suggest that these programs have failed.
+Since \texttt{nls()}, \texttt{nlsLM()}, and \texttt{gsl\_nls()} do not
+offer singular values, we need to extract the Jacobian and compute its
+singular values. The following script shows how to do this, using as Jacobian what is called
+the \texttt{gradient} element in the returned solution for these solvers.
+
+\begin{verbatim}
+# for nlsLM
+if (inherits(tnlm1, "try-error"))  {
+   print("Cannot compute solution -- likely singular Jacobian")
+ } else {  
+   JtnlsLM <- tnlm1$m$gradient() # actually the Jacobian
+   svd(JtnlsLM)$d # Singular values
+}   
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] 3.4641e+00 3.2530e-10 8.9643e-12
+\end{verbatim}
+
+\begin{verbatim}
+# for gsl_nls
+if (inherits(tgslnls1, "try-error")) {
+   cat("Cannot compute solution -- likely singular Jacobian")
+} else {  
+   JtnlsLM <- tgslnls1$m$gradient()
+   svd(JtnlsLM)$d # Singular values
+}   
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] 3.4641e+00 9.4495e-09 4.1324e-11
+\end{verbatim}
+
+We see that there are differences in detail, but the more important result is that
+two out of three singular values are essentially 0. Our Jacobian is singular, and no
+method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
+reported at this saddle point, \texttt{nlsr::nlxb()} cannot proceed.
+
+\begin{verbatim}
+stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
+print(badstart)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  9205.4  on  12 observations
+#>     after  2    Jacobian and  2 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> Asym             35.5321            NA         NA         NA  -9.694e-09       3.464  
+#> xmid               43376            NA         NA         NA  -1.742e-09    2.61e-10  
+#> scal             -2935.4            NA         NA         NA    -2.4e-08    7.12e-16
 \end{verbatim}
@@
 \hypertarget{functional-specification-of-problems}{%
-\subsection{Functional specification of
-problems}\label{functional-specification-of-problems}}
-
-We illustrate how to solve nonlinear least squares problems using a
-function to define the residual. Note that \texttt{gsl\_nls()} requires
-a vector \texttt{y} that is the expected value of what we have called
-the residual, but in this case, is actually the model. \texttt{gsl\_nls}
-uses a numerical approximation for the Jacobian if the argument
-\texttt{jac} is missing. Note function \texttt{nlsr::pnlslm()} for a
-1-line display of the results of \texttt{minpack.lm::nls.lm()}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobnlfb}\OtherTok{\textless{}{-}}\FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{resfn=}\NormalTok{hobbs.res, }\AttributeTok{jacfn=}\NormalTok{hobbs.jac)}
-\FunctionTok{pshort}\NormalTok{(hobnlfb) }\CommentTok{\# use print(hobnlfb) for more detail}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobnlm}\OtherTok{\textless{}{-}}\FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
-\FunctionTok{pnlslm}\NormalTok{(hobnlm)  }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobgsln}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(hobgsln)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobgsl}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{), }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
-\FunctionTok{pnls}\NormalTok{(hobgsl) }\CommentTok{\# using analytic Jacobian}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+\subsection{Functional specification of problems}\label{functional-specification-of-problems}}
+
+We illustrate how to solve nonlinear least squares problems using
+a function to define the residual. Note that \texttt{gsl\_nls()} requires
+a vector \texttt{y} that is the expected value of what we have called the
+residual, but in this case, is actually the model. \texttt{gsl\_nls} uses
+a numerical approximation for the Jacobian if the argument \texttt{jac}
+is missing. Note function \texttt{nlsr::pnlslm()} for a 1-line display of
+the results of \texttt{minpack.lm::nls.lm()}.
+
+\begin{verbatim}
+#> hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
+\end{verbatim}
+
+\begin{verbatim}
+#> hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+\end{verbatim}
+
+\begin{verbatim}
+#> hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
 \end{verbatim}
@@
 \hypertarget{design-goals-termination-tests-and-output-objects}{%
-\subsection{Design goals, termination tests, and output
-objects}\label{design-goals-termination-tests-and-output-objects}}
-
-The output object of \texttt{nlxb()} is smaller than the class
-\texttt{nls} object returned by \texttt{nls()}, \texttt{nlsLM()}, and
-\texttt{gsl\_nls()}. Package \texttt{nlsr} emphasizes the solution of
-the nonlinear least squares problem rather than the estimation of a
-nonlinear model that fits or explains the data. The object of class
-\texttt{nls} allows for a number of specialized modeling and diagnostic
-extensions. For compatibility, the \texttt{nlsr} package has the
-function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias.
-This uses \texttt{nlxb()} to find good parameters, then calls
-\texttt{nls()} to return the class \texttt{nls} object. Unless
-particular modeling features are needed, the use of \texttt{wrapnlsr()}
-is unnecessary and wasteful of resources.
-
-The design goals of the different tools may also be revealed in the
-so-called ``convergence tests'' for the iterative solvers. In the manual
-page for \texttt{nls()} in R 4.0.0 there was the warning:
+\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}
+
+The output object of \texttt{nlxb()} is smaller than the class \texttt{nls} object returned
+by \texttt{nls()}, \texttt{nlsLM()}, and \texttt{gsl\_nls()}. Package \texttt{nlsr} emphasizes the solution
+of the nonlinear least squares problem rather than the estimation of a nonlinear
+model that fits or explains the data. The object of class \texttt{nls} allows for
+a number of specialized modeling and diagnostic extensions. For compatibility,
+the \texttt{nlsr} package has the function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias.
+This uses \texttt{nlxb()} to find good parameters, then calls \texttt{nls()} to return the class
+\texttt{nls} object. Unless particular modeling features are needed, the use of
+\texttt{wrapnlsr()} is unnecessary and wasteful of resources.
+
+The design goals of the different tools may also be revealed in the so-called
+``convergence tests'' for the iterative solvers. In the manual page for \texttt{nls()}
+in R 4.0.0 there was the warning:
 
@@
 with suggested addition of small perturbations to the data. This admits
-\texttt{nls()} could not solve well-posed problems unless data is
-polluted with errors. Zero-residual problems are not always artificial,
+\texttt{nls()} could not solve well-posed problems unless data is polluted
+with errors. Zero-residual problems are not always artificial,
 since problems in function approximation and nonlinear equations can be
 approached with nonlinear least squares. Fortunately, a small adjustment
-to the ``termination test'' for the \textbf{program}, rather than for
-the ``convergence'' of the underlying \textbf{algorithm}, fixes the
-defect. The test is the Relative Offset Convergence Criterion (see
-Bates, Douglas M. and Watts, Donald G.
-(\protect\hyperlink{ref-BatesWatts81}{1981})). This scales an estimated
-reduction in the loss function by its current value. If the loss
-function is very small, we are close to a zero-divide. Adding a small
-quantity to the divisor avoids trouble. In 2021, one of us (J. Nash)
-proposed that \texttt{nls.control()} have an additional parameter
-\texttt{scaleOffset} with a default value of zero. Setting it to a small
-number -- 1.0 is a reasonable choice -- allows small-residual problems
+to the ``termination test'' for the \textbf{program}, rather than for the
+``convergence'' of the underlying \textbf{algorithm}, fixes the defect. The test
+is the Relative Offset Convergence
+Criterion (see Bates, Douglas M. and Watts, Donald G. (1981)). This scales an estimated reduction in the loss
+function by its current value. If the loss function is very small, we are close
+to a zero-divide. Adding a small quantity to the divisor avoids trouble.
+In 2021, one of us (J. Nash) proposed that \texttt{nls.control()} have an
+additional parameter \texttt{scaleOffset} with a default value of zero. Setting it to a
+small number -- 1.0 is a reasonable choice -- allows small-residual problems
 (i.e., near-exact fits) to be dealt with easily. We call this the
-\textbf{safeguarded relative offset convergence criterion}, and it has
-been in \texttt{nlsr} since it was introduced. The default value gives
-the legacy behaviour. This improvement has been in the R distributed
-code since version 4.1.0.
-
-Additional termination tests can be used. \texttt{nlsr} has a
-\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
-the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
-times the initial sum of squares, where
-\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
-approximately 2.43e-55.
-
-Termination after what may be considered excessive computation is also
-important. \texttt{nls()} stops after \texttt{maxiter} ``iterations''.
-The meaning of ``iteration'' may require an examination of the code. We
-terminate execution when the number of residual or Jacobian evaluations
-exceed set limits. Generally, we prefer larger limits than the default
-\texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
+\textbf{safeguarded relative offset convergence criterion}, and it has been in \texttt{nlsr}
+since it was introduced. The default value
+gives the legacy behavior. This improvement has been in the R distributed code
+since version 4.1.0.
+
+Additional termination tests can be used. \texttt{nlsr} has a \textbf{small sum of
+squares} test (\textbf{smallsstest}) that compares the latest evaluated sum of squared
+(weighted) residuals to \texttt{e4} times the initial sum of squares, where
+\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.
+\%\% \^{} is picked up in rjtools check
+
+Termination after what may be considered excessive computation is also important.
+\texttt{nls()} stops after \texttt{maxiter} ``iterations''. The meaning of ``iteration'' may require an
+examination of the code. We terminate execution when the number of residual or
+Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the
+default \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
 
 \hypertarget{returned-results-of-nls-and-other-tools}{%
-\subsubsection{Returned results of nls() and other
-tools}\label{returned-results-of-nls-and-other-tools}}
-
-As mentioned, the output of \texttt{nls()} is an object of class ``nls''
-which has a quite rich structure described in the manual file or
-revealed by applying the \texttt{str()} function to the result of
-\texttt{nls()}.
+\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}
+
+As mentioned, the output of \texttt{nls()} is an object of class ``nls'' which has a quite rich structure
+described in the manual file or revealed by applying the \texttt{str()} function to the
+result of \texttt{nls()}.
 
 The complexity of this object is a challenge to users. Let us use the
-\texttt{result} as the returned object from \texttt{nls()},
-\texttt{minpack.lm::nlsLM()} or, \texttt{gslnls::gsl\_nls()}. For
-example, the \texttt{data} return element is an R symbol. To actually
-access the data from this element, we need to use the syntax:
+\texttt{result} as the returned object from \texttt{nls()}, \texttt{minpack.lm::nlsLM()} or,
+\texttt{gslnls::gsl\_nls()}. For example, the \texttt{data} return element is an R symbol. To
+actually access the data from this element, we need to use the syntax:
 
@@
 
-and if there is an element called \texttt{xdata}, then it can be
-accessed as \texttt{result\$model\$xdata}.
-
-By contrast, \texttt{nlsr::nlxb()} returns a much simpler structure of
-11 items in one level. Moreover, \texttt{nlxb} explicitly returns the
-sum of squares, the residual vector, Jacobian, and counts of
-evaluations.
+and if there is an element called \texttt{xdata}, then it can be accessed as
+\texttt{result\$model\$xdata}.
+
+By contrast, \texttt{nlsr::nlxb()} returns a much simpler structure of 11 items
+in one level. Moreover, \texttt{nlxb} explicitly returns the sum of squares, the residual vector, Jacobian, and counts of evaluations.
 
 \hypertarget{when-to-compute-ancillary-information}{%
-\subsubsection{When to compute ancillary
-information}\label{when-to-compute-ancillary-information}}
-
-Tools that produce a class \texttt{nls} output object create a rich set
-of functions and structures that are then used in a variety of modeling
-tasks, including the least squares solution. By contrast, \texttt{nlsr}
-computes quantities as they are requested or needed, with additional
-features in separate functions. For example, the singular values of the
-Jacobian are actually computed in the \texttt{print} and
-\texttt{summary} methods for the result. These two approaches lead to
-different consequences for performance and how features are provided.
-\texttt{nlsr} has antecedents in the methods of John C. Nash
-(\protect\hyperlink{ref-cnm79}{1979}), where storage for data and
-programs was at a ridiculous premium in the small computers of the era.
-Thus the code in \texttt{nlsr} is likely of value for workers to copy
-and modify for customized tools.
+\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}
+
+Tools that produce a class \texttt{nls} output object create a rich set of functions
+and structures that are then used in a variety of modeling tasks, including the
+least squares solution. By contrast,
+\texttt{nlsr} computes quantities as they are requested or needed, with additional
+features in separate functions. For example, the singular values of the Jacobian
+are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
+These two approaches
+lead to different consequences for performance and how features are
+provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
+data and programs was at a ridiculous premium in the small computers of the era.
+Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
+for customized tools.
 
@@
 
-Gauss-Newton/Marquardt methods all need a Jacobian matrix at each
-iteration. By default, \texttt{nlsr::nlxb()} will try to evaluate this
-using analytic expressions using symbolic and automatic differentiation
-tools. When using a formula specification of the model, \texttt{nls()},
-\texttt{minpack.lm::nlsLM()} and \texttt{gslnls::gsl\_nls()} use a
-finite difference approximation to compute the Jacobian, though
-\texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
-Package \texttt{nlsr} provides, via appropriate calling syntax, four
-numeric approximation options for the Jacobian, with a further control
-\texttt{ndstep} for the size of the step used in the approximation.
-
-Using the ``gradient'' attribute of the output of the Jacobian function
-to hold the Jacobian matrix lets us embed this in the \textbf{residual}
-function as well, so that the call to \texttt{nlsr::nlfb()} can be made
-with the same name used for both residual and Jacobian function
-arguments. This programming trick saves a lot of trouble for the package
-developer, but it can be a nuisance for users trying to understand the
-code.
-
-As far as we can understand the logic in \texttt{nls()}, the Jacobian
-computation during parameter estimation is carried out within the called
-C-language program and its wrapper R code function
-\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
-the R distribution source code. This is used to provide Jacobian
-information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
-functions, which are \textbf{not} exported for general use.
-\texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.
-
-\texttt{numericDeriv()} uses a simple forward difference approximation
-of derivatives, though a central difference approximation can be
-specified in control parameters. We are unclear why
-\texttt{numericDeriv()} in base R calls \texttt{C\_numeric\_deriv}, as
-we were easily able to create a more compact version entirely in R. See
-\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
-
-\texttt{minpack.lm::nlsLM()} invokes \texttt{numericDeriv()} in its
-local version of \texttt{nlsModel()}, but it appears to use an internal
-approximate Jacobian code from the original Fortran \texttt{minpack}
-code, namely, \texttt{lmdif.f}. Such differences in approach can lead to
-different behaviour, usually minor, but sometimes annoying with
+Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration.
+By default, \texttt{nlsr::nlxb()} will try to evaluate this using analytic
+expressions using symbolic and automatic differentiation tools.
+When using a formula specification of the model, \texttt{nls()}, \texttt{minpack.lm::nlsLM()}
+and \texttt{gslnls::gsl\_nls()} use a finite difference approximation to compute the
+Jacobian, though \texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
+Package \texttt{nlsr} provides, via appropriate calling syntax, four numeric approximation
+options for the Jacobian, with a further control \texttt{ndstep} for the size of the
+step used in the approximation.
+
+Using the ``gradient'' attribute of the output of the Jacobian function to hold
+the Jacobian matrix lets us embed this in the \textbf{residual} function as well,
+so that the call to \texttt{nlsr::nlfb()} can be made with the same name used for
+both residual and Jacobian function arguments. This programming trick saves a
+lot of trouble for the package developer, but it can be a nuisance for users
+trying to understand the code.
+
+As far as we can understand the logic in \texttt{nls()}, the Jacobian computation during
+parameter estimation is carried out within the called C-language program
+and its wrapper R code function \texttt{numericDeriv()}, part of
+\texttt{./src/library/stats/R/nls.R}
+in the R distribution source code. This is used to provide Jacobian information in
+the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are \textbf{not} exported for
+general use. \texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.
+
+\texttt{numericDeriv()} uses a
+simple forward difference approximation of derivatives, though a central
+difference approximation can be specified in control parameters.
+We are unclear why \texttt{numericDeriv()} in base R calls \texttt{C\_numeric\_deriv},
+as we were easily able to create a more compact version entirely in R.
+See \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
+
+\texttt{minpack.lm::nlsLM()} invokes \texttt{numericDeriv()} in its local
+version of \texttt{nlsModel()}, but it appears to use an internal approximate Jacobian
+code from
+the original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such differences in approach
+can lead to different behavior, usually minor, but sometimes annoying with
 ill-conditioned problems.
@@
 \item
-  A pasture regrowth problem (Huet et al.
-  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
-  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
-  conditioned Jacobian and \texttt{nls()} fails with ``singular
-  gradient''. Worse, numerical approximation to the Jacobian gives the
-  error ``singular gradient matrix at initial parameter estimates'' for
-  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
-  unable to take effect, while the analytic derivatives of
-  \texttt{nlsr::nlxb} give a solution.
-\item
-  Karl Schilling (private communication) provided an example where a
-  model specified with the formula
-  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
-  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
-  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
-  which \texttt{x\ =\ 0}, the derivative is undefined, but the model can
-  be computed. In such cases, we observed that \texttt{nls()} and
-  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a
-  lucky accident.
+  A pasture regrowth problem (Huet et al. (2004), page 1, based on Ratkowsky (1983)) has a
+  poorly conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
+  Worse, numerical approximation to the Jacobian gives the error
+  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of \texttt{nlsr::nlxb} give a solution.
+\item
+  Karl Schilling (private communication) provided an example where a model specified
+  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
+  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
+  the derivative is undefined, but the model can be computed. In such cases,
+  we observed that \texttt{nls()} and
+  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a lucky accident.
 \end{itemize}
@@
 \hypertarget{jacobian-code-in-selfstart-models}{%
-\subsubsection{Jacobian code in selfStart
-models}\label{jacobian-code-in-selfstart-models}}
+\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}
 
 Analytic Jacobian code can be provided to all the solvers discussed.
-Most \texttt{selfStart} models that automatically provide starting
-parameters also include such code. There is documentation in R of
-\texttt{selfStart} models, but their construction is non-trivial. A
-number of such models are included with base R in
-\texttt{./src/library/stats/R/zzModels.R}, with package \texttt{nlraa}
-(Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})) providing a
-richer set. There are also some in the now-archived package
-\texttt{NRAIA}. These provide the Jacobian in the ``gradient'' attribute
-of the ``one-sided'' formula that defines each model, and these
-Jacobians are often the analytic forms.
-
-The \texttt{nls()} function, after computing the ``right-hand side'' or
-\texttt{rhs} of the residual, checks to see if the ``gradient''
-attribute is defined, otherwise using \texttt{numericDeriv()} to compute
-a Jacobian into that attribute. This code is within the
-\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
-analytic Jacobians almost certainly contributes to the good performance
-of \texttt{nls()} on \texttt{selfStart} models.
-
-The use of \texttt{selfStart} models with \texttt{nlsr} is described in
-the ``Introduction to nlsr'' vignette. However, since \texttt{nlsr}
-generally can use very crude starting values, we have rarely needed
-them, though it should be pointed out that our work is primarily
-diagnostic. If we were carrying out a large number of similar
-estimations, such initial parameters are critical to efficiency.
-
-In considering \texttt{selfStart} models, we noted that the base-R
-function \texttt{SSlogis} is intended to solve problem
-\textbf{Logistic3T} above. When this function is used via
-\texttt{getInitial()} to find starting values, it actually calls
-\texttt{nls()} with the `plinear' algorithm and finds a (full) solution.
-It then passes the solution coefficients to the default algorithm
-unnecessarily. Moreover, the implicit double call is, in our view, prone
-to creating errors in code maintenance. To provide simpler starting
-parameters, the function \texttt{SSlogisJN} is now part of the package
-\texttt{nlsr}, but is most useful for \texttt{nls()}.
-
-Users may also want to profit from the Jacobian code of
-\texttt{selfStart} models but supply explicit starting values other than
-those suggested by \texttt{getInitial()}. This does not appear to be
-possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
-starting parameters, and can either use \texttt{getInitial()} to find
-them from the \texttt{selfStart} model or provide explicit values, but
-the \texttt{selfStart} model is used in the formula for the
-\texttt{nlxb()} call.
-
-We are also surprised that the analytic expressions for the Jacobian
-(``gradient'') in the \texttt{SSLogis} function and others save
-quantities in ``hidden'' variables, i.e., with names preceded by ``.''.
-These are then not displayed by the \texttt{ls()} command, making them
-difficult to access by users who may wish to create their own selfStart
-model via copy and edit. Interactive tools, such as ``visual fitting''
-(John C. Nash and Velleman
-(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
-considering as another way to find starting parameters, but we know of
-no R capability of this type.
-
-As a side note, the introduction of \texttt{scaleOffset} in R 4.1.1 to
-deal with the convergence test for small residual problems now requires
-that the \texttt{getInitial()} function have dot-arguments
-(\texttt{...}) in its argument list. This illustrates the entanglement
-of many features in \texttt{nls()} that complicate its maintenance and
+Most \texttt{selfStart} models that automatically provide starting parameters
+also include such code. There is documentation in R
+of \texttt{selfStart} models, but their construction is non-trivial. A number
+of such models are included with base R in \texttt{./src/library/stats/R/zzModels.R},
+with package \texttt{nlraa} (Miguez (2021)) providing a richer set.
+There are also some in the now-archived package \texttt{NRAIA}.
+These provide the Jacobian in the ``gradient'' attribute of the ``one-sided'' formula
+that defines each model, and these Jacobians are often the analytic forms.
+
+The \texttt{nls()} function, after
+computing the ``right-hand side'' or \texttt{rhs} of the residual, checks to see if the
+``gradient'' attribute is defined, otherwise using \texttt{numericDeriv()} to compute a
+Jacobian into that attribute. This code is within the \texttt{nlsModel()} or
+\texttt{nlsModel.plinear()} functions. The use of analytic Jacobians
+almost certainly contributes to the good performance of \texttt{nls()} on \texttt{selfStart}
+models.
+
+The use of \texttt{selfStart} models with
+\texttt{nlsr} is described in the ``Introduction to nlsr'' vignette. However, since \texttt{nlsr}
+generally can use very crude starting values, we have rarely needed them, though
+it should be pointed out that our work is primarily diagnostic. If we were carrying
+out a large number of similar estimations, such initial parameters are
+critical to efficiency.
+
+In considering \texttt{selfStart} models, we noted that the base-R function \texttt{SSlogis}
+is intended to solve problem \textbf{Logistic3T} above. When this function is used
+via \texttt{getInitial()} to find
+starting values, it actually calls \texttt{nls()} with the `plinear' algorithm
+and finds a (full) solution. It then passes the solution coefficients to the default
+algorithm unnecessarily. Moreover, the implicit double call is, in our view, prone to creating
+errors in code maintenance.
+To provide simpler starting parameters, the function \texttt{SSlogisJN}
+is now part of the package \texttt{nlsr}, but is most useful for \texttt{nls()}.
+
+Users may also want to profit from the Jacobian code of \texttt{selfStart} models
+but supply explicit starting values other than those suggested by \texttt{getInitial()}.
+This does not appear to be possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
+starting parameters, and can either use \texttt{getInitial()} to find them from the \texttt{selfStart}
+model or provide explicit values, but the \texttt{selfStart} model is used in the formula
+for the \texttt{nlxb()} call.
+
+We are also surprised that the
+analytic expressions for the Jacobian (``gradient'') in the \texttt{SSLogis} function
+and others save quantities in ``hidden''
+variables, i.e., with names preceded by ``.''.
+These are then not displayed by the \texttt{ls()}
+command, making them difficult to access by users who may wish to create
+their own selfStart model via copy and edit.
+Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (1996)) might be
+worth considering as another way to find starting parameters, but we know of no
+R capability of this type.
+
+As a side note, the introduction of \texttt{scaleOffset} in R 4.1.1 to deal with the
+convergence test for small residual problems now requires that the \texttt{getInitial()}
+function have dot-arguments (\texttt{...}) in its argument list. This illustrates the
+entanglement of many features in \texttt{nls()} that complicate its maintenance and
 improvement.
@@
 \hypertarget{bounds-constraints-on-parameters}{%
-\subsection{Bounds constraints on
-parameters}\label{bounds-constraints-on-parameters}}
-
-For many problems, we know that parameters cannot be bigger or smaller
-than some externally known limits. Such limits should be built into
-models, but there are some important details for using the tools in R.
+\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}}
+
+For many problems, we know that parameters cannot be bigger or smaller than some
+externally known limits. Such limits should be built into models, but there
+are some important details for using the tools in R.
 
@@
 \item
-  \texttt{nls()} can only impose bounds if the \texttt{algorithm="port"}
-  argument is used in the call. Unfortunately, the documentation warns
-  us:
-
-  \emph{The algorithm = ``port'' code appears unfinished, and does not
-  even check that the starting value is within the bounds. Use with
+  \texttt{nls()} can only impose bounds if the \texttt{algorithm="port"} argument is
+  used in the call. Unfortunately, the documentation warns us:
+
+  \emph{The algorithm = ``port'' code appears unfinished, and does not even
+  check that the starting value is within the bounds. Use with
   caution, especially where bounds are supplied.}
@@
 \item
-  \texttt{nlsLM()} includes bounds in the standard call, but we have
-  observed cases where it fails to get the correct answer. From an
-  examination of the code, we believe the authors have not taken into
-  account all possibilities, though all programs have some weakness
-  regarding constrained optimization in that programmers have to work
-  with assumptions on the scale of numbers, and some problems will be
-  outside the scope envisaged.
+  \texttt{nlsLM()} includes bounds in the standard call, but we have observed cases
+  where it fails to get the correct answer. From an examination of the code,
+  we believe the authors have not taken into account all possibilities, though
+  all programs have some weakness regarding constrained optimization in that
+  programmers have to work with assumptions on the scale of numbers,
+  and some problems will be outside the scope envisaged.
 \end{itemize}
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Start MUST be feasible i.e. on or within bounds}
-\NormalTok{anlshob1b }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
-\FunctionTok{pnls}\NormalTok{(anlshob1b) }\CommentTok{\#  check the answer (short form)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlshob1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 10  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# nlsLM seems NOT to work with bounds in this example}
-\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLM1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLM1b  -- ss= 881.02 :  c1 = 2  c2 = 6  c3 = 3; 2  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# also no warning if starting out of bounds, but gets a good answer!!}
-\NormalTok{st4}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{4}\NormalTok{, }\AttributeTok{c2=}\DecValTok{4}\NormalTok{, }\AttributeTok{c3=}\DecValTok{4}\NormalTok{)}
-\NormalTok{anlsLMob }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{st4, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLMob)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLMob  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Try nlsr::nlxb()}
-\NormalTok{anlx1b }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pshort}\NormalTok{(anlx1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlx1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 12 res/ 12 jac
+\begin{verbatim}
+# Start MUST be feasible i.e. on or within bounds
+anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
+             upper=c(2,6,3), algorithm='port')
+pnls(anlshob1b) #  check the answer (short form)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlshob1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 10  itns
+\end{verbatim}
+
+\begin{verbatim}
+# nlsLM seems NOT to work with bounds in this example
+anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pnls(anlsLM1b)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLM1b  -- ss= 881.02 :  c1 = 2  c2 = 6  c3 = 3; 2  itns
+\end{verbatim}
+
+\begin{verbatim}
+# also no warning if starting out of bounds, but gets a good answer!!
+st4<-c(c1=4, c2=4, c3=4)
+anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pnls(anlsLMob)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLMob  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Try nlsr::nlxb()
+anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+pshort(anlx1b)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlx1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 12 res/ 12 jac
 \end{verbatim}
@@
 \hypertarget{philosophical-considerations}{%
-\subsubsection{Philosophical
-considerations}\label{philosophical-considerations}}
+\subsubsection{Philosophical considerations}\label{philosophical-considerations}}
 
 Bounds on parameters raise some interesting questions about how
-uncertainty in parameter estimates should be computed or reported. That
-is, the traditional ``standard errors'' are generally taken to imply
-symmetric intervals about the point estimate in which the parameter may
-be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not
-compute standard errors nor their derived statistics when bounds are
-active.
+uncertainty in parameter estimates should be computed or reported. That is, the
+traditional ``standard errors'' are generally taken to imply symmetric intervals about
+the point estimate in which the parameter may be expected to be found with some
+probability under certain assumptions.
+Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not compute standard
+errors nor their derived statistics when bounds are active.
 
@@
 
-Let us try to fix (mask) the first parameter in the first two example
-problems.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbsmaskx.R {-}{-} masks with formula specification of the problem}
-\FunctionTok{require}\NormalTok{(nlsr); }\FunctionTok{require}\NormalTok{(minpack.lm); traceval}\OtherTok{\textless{}{-}}\ConstantTok{FALSE}
-\NormalTok{stu }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{200}\NormalTok{, }\AttributeTok{b2=}\DecValTok{50}\NormalTok{, }\AttributeTok{b3=}\FloatTok{0.3}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
-\NormalTok{sts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{2}\NormalTok{, }\AttributeTok{c2=}\DecValTok{5}\NormalTok{, }\AttributeTok{c3=}\DecValTok{3}\NormalTok{) }\CommentTok{\# a default scaled starting vector (named!)}
-\CommentTok{\# fix first parameter}
-\NormalTok{anxbmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
-\FunctionTok{print}\NormalTok{(anxbmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.6182  on  12 observations
-##     after  4    Jacobian and  4 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1                   200U M         NA         NA         NA           0          NA  
-## b2               49.5108          1.12      44.21  8.421e-13  -2.887e-07        1022  
-## b3              0.311461      0.002278      136.8  1.073e-17   0.0001635      0.4569
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlM1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
-\FunctionTok{pnls}\NormalTok{(anlM1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlM1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-        \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{),  }\AttributeTok{algorithm=}\StringTok{"port"}\NormalTok{, }\AttributeTok{trace=}\NormalTok{traceval))}
-\FunctionTok{pnls}\NormalTok{(anlsmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsmsk1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 5  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
-\NormalTok{anlxmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                  \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
-\FunctionTok{print}\NormalTok{(anlxmsks1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.6182  on  12 observations
-##     after  4    Jacobian and  4 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## c1                     2U M         NA         NA         NA           0          NA  
-## c2               4.95108         0.112      44.21  8.421e-13  -2.981e-06       104.2  
-## c3               3.11461       0.02278      136.8  1.073e-17   1.583e-05       4.482
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlshmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{trace=}\NormalTok{traceval, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
-\FunctionTok{pnls}\NormalTok{(anlshmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlshmsk1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 5  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsLMmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                 \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLMmsks1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLMmsks1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Test with all parameters masked}
-\NormalTok{anlxmskall}\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\NormalTok{sts, }\AttributeTok{upper=}\NormalTok{sts))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in nlfb(start = pnum, resfn = trjfn, jacfn = trjfn, trace = trace, :
-## All parameters are masked
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlxmskall)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in summary.nlsr(x): No unmasked parameters
-\end{verbatim}
-
-\begin{verbatim}
-## residual sumsquares =  158.23  on  12 observations
-##     after  0    Jacobian and  1 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## c1                     2U M         NA         NA         NA          NA          NA  
-## c2                     5U M         NA         NA         NA          NA          NA  
-## c3                     3U M         NA         NA         NA          NA          NA
-\end{verbatim}
-
-\texttt{nlsr} has an output format that indicates the constraint status
-of the parameter estimates. For \texttt{nlsr}, we have \textbf{chosen}
-to suppress the calculation of approximate standard errors in the
-parameters when constraints are active because their meaning under
-constraints is unclear, though we believe this policy worthy of
-discussion and further investigation.
+Let us try to fix (mask) the first parameter in the first two example problems.
+
+\begin{verbatim}
+# Hobbsmaskx.R -- masks with formula specification of the problem
+require(nlsr); require(minpack.lm); traceval<-FALSE
+stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
+sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
+# fix first parameter
+anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+            upper=c(200, 60, 3), trace=traceval))
+print(anxbmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  2.6182  on  12 observations
+#>     after  4    Jacobian and  4 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> b1                   200U M         NA         NA         NA           0          NA  
+#> b2               49.5108          1.12      44.21  8.421e-13  -2.887e-07        1022  
+#> b3              0.311461      0.002278      136.8  1.073e-17   0.0001635      0.4569
+\end{verbatim}
+
+\begin{verbatim}
+anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+            upper=c(200, 60, 3), trace=traceval))
+pnls(anlM1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlM1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+        upper=c(200, 60, 3),  algorithm="port", trace=traceval))
+pnls(anlsmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsmsk1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 5  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Hobbs scaled problem with bounds, formula specification
+anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                  upper=c(2,6,30))
+print(anlxmsks1)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  2.6182  on  12 observations
+#>     after  4    Jacobian and  4 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> c1                     2U M         NA         NA         NA           0          NA  
+#> c2               4.95108         0.112      44.21  8.421e-13  -2.981e-06       104.2  
+#> c3               3.11461       0.02278      136.8  1.073e-17   1.583e-05       4.482
+\end{verbatim}
+
+\begin{verbatim}
+anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
+             upper=c(2,6,30), algorithm='port')
+pnls(anlshmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlshmsk1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 5  itns
+\end{verbatim}
+
+\begin{verbatim}
+anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                 upper=c(2,6,30))
+pnls(anlsLMmsks1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLMmsks1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Test with all parameters masked
+anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+print(anlxmskall)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  158.23  on  12 observations
+#>     after  0    Jacobian and  1 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> c1                     2U M         NA         NA         NA          NA          NA  
+#> c2                     5U M         NA         NA         NA          NA          NA  
+#> c3                     3U M         NA         NA         NA          NA          NA
+\end{verbatim}
+
+\texttt{nlsr} has an output format that indicates the constraint status of the parameter
+estimates. For \texttt{nlsr}, we have \textbf{chosen} to suppress the calculation of
+approximate standard errors in the parameters when constraints are active because
+their meaning under constraints is unclear, though we believe this
+policy worthy of discussion and further investigation.
 
 \hypertarget{stabilization-of-gauss-newton-computations}{%
-\subsection{Stabilization of Gauss-Newton
-computations}\label{stabilization-of-gauss-newton-computations}}
+\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}}
 
 All four major tools illustrated solve some variant of the Gauss-Newton
-equations. \texttt{nls()} uses a modification of an approach suggested
-by Hartley (\protect\hyperlink{ref-Hartley1961}{1961}), while
-\texttt{nlsr}, \texttt{gslnls}, and \texttt{minpack.lm} use flavours of
-Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}). \texttt{gslnls}
-offers an accelerated Marquardt method and three alternative methods
-which we have not investigated. Control settings for \texttt{nlxb()} or
-\texttt{nlfb()} allow exploration of Hartley and Marquardt algorithm
-variants. In general, the Levenberg-Marquardt stabilization is important
-in obtaining solutions in methods of the Gauss-Newton family, as
-\texttt{nls()} terminates too frequently and unnecessarily with
-\texttt{singular\ gradient} errors.
+equations. \texttt{nls()} uses a modification of an approach suggested by Hartley (1961),
+while \texttt{nlsr}, \texttt{gslnls}, and \texttt{minpack.lm} use flavors of Marquardt (1963). \texttt{gslnls}
+offers an accelerated Marquardt method and three alternative methods which we
+have not investigated.
+Control settings for \texttt{nlxb()} or \texttt{nlfb()} allow exploration of Hartley and Marquardt
+algorithm variants. In general, the Levenberg-Marquardt stabilization
+is important in obtaining
+solutions in methods of the Gauss-Newton family, as \texttt{nls()} terminates
+too frequently and unnecessarily with \texttt{singular\ gradient} errors.
 
@@
 
-An important choice made in developing \texttt{nlsr} was to code
-entirely within the R programming language. \texttt{nls()} uses a mix of
-R, C, and Fortran, as does \texttt{minpack.lm}. \texttt{gslnls} is an R
-wrapper to various C-language routines in the Gnu Scientific Library
-(Galassi et al. (\protect\hyperlink{ref-GSL-manual}{2009})). Generally,
-keeping to a single programming language can allow for easier
-maintenance and upgrades. It also avoids some work when there are
-changes or upgrades to libraries for the non-R languages. R is usually
-considered slower than most computing environments because it keeps
-track of objects and because it is usually interpreted. In recent years,
-the performance penalty for using code entirely in R has been much
-reduced with the just-in-time compiler and other improvements. All-R
-computation may now offer acceptable performance. In \texttt{nlsr}, the
-use of R may be a smaller performance cost than the aggressive approach
-to a solution, which can cause more iterations to be used.
+An important choice made in developing \texttt{nlsr} was to code entirely within the R
+programming language. \texttt{nls()} uses a mix of R, C, and Fortran, as does \texttt{minpack.lm}.
+\texttt{gslnls} is an R wrapper to various C-language routines in the Gnu Scientific
+Library (Galassi et al. (2009)).
+Generally, keeping to a single programming language can allow for
+easier maintenance and upgrades. It also avoids some work when there are changes
+or upgrades to libraries for the non-R languages.
+R is usually considered slower
+than most computing environments because it keeps track of objects and
+because it is usually interpreted. In recent years, the
+performance penalty for using code entirely in R has been much reduced
+with the just-in-time compiler and other improvements. All-R computation
+may now offer acceptable performance. In \texttt{nlsr}, the use of R may be a
+smaller performance cost than the aggressive approach to a solution,
+which can cause more iterations to be used.
 
@@
 
-\texttt{nls()} can be called without specifying the \texttt{data}
-argument. In this case, it will search in the available environments
-(i.e., workspaces) for suitable data objects. We do NOT like this
-approach, but it is ``the R way''. R allows users to leave many objects
-in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
-actively suppress saving this workspace (\texttt{.RData}) on exit;
-otherwise, any such file in the path will be loaded on startup. R users
-in our acquaintance avoid saving the workspace because of lurking data
-and functions that may cause unwanted results.
+\texttt{nls()} can be called without specifying the \texttt{data} argument. In this case, it will
+search in the available environments (i.e., workspaces) for suitable data objects.
+We do NOT like this approach, but it is ``the R way''. R allows users to leave many
+objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
+suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any such file in
+the path will be loaded on startup. R users in our
+acquaintance avoid saving the workspace because of lurking data and
+functions that may cause unwanted results.
 
@@
 
-\texttt{nls()} and other class \texttt{nls} tools accept an argument
-\texttt{subset}. This acts through the mediation of
-\texttt{model.frame}, which not obvious in the source code files
-\texttt{/src/library/stats/R/nls.R} and
-\texttt{/src/library/stats/src/nls.C}. Having \texttt{subset} at the
-level of the call to a function like \texttt{nls()} saves effort, but it
-does mean that the programmer of the solver needs to be aware of the
+\texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
+the mediation of \texttt{model.frame}, which not obvious in the
+source code files \texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
+Having \texttt{subset} at the level of the call to a function like
+\texttt{nls()} saves effort, but it does mean that the programmer of the
+solver needs to be aware of the
 origin (and value) of objects such as the data, residuals and Jacobian.
-By preference, we would implement subsetting by zero-value weights, with
-observation counts (and degrees of freedom) computed via the numbers of
-non-zero weights. Alternatively, we would extract a working dataframe
-from the relevant elements in the original.
+By preference,
+we would implement subsetting by zero-value weights, with observation counts
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
 
 \hypertarget{feature-na.action-missing-value-treatment}{%
-\subsubsection{Feature: na.action (missing value
-treatment)}\label{feature-na.action-missing-value-treatment}}
-
-\texttt{na.action} is an argument to the \texttt{nls()} function, but it
-does not appear obviously in the source code, often being handled behind
-the scenes after referencing the option \texttt{na.action}. This feature
-also changes the data supplied to our nonlinear least squares solver. A
-useful, but possibly dated, description is given in:
+\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}}
+
+\texttt{na.action} is an argument to the \texttt{nls()} function, but it does not appear
+obviously in the source code, often being handled behind the scenes after
+referencing the option \texttt{na.action}. This feature also changes the data
+supplied to our nonlinear least squares solver.
+A useful, but possibly dated, description is given in:
 \emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.
 The typical default action, which can be seen by using the command
-\texttt{getOption("na.action")} is \texttt{na.omit}. This option removes
-from computations any observations containing missing values (i.e.~any
-row of a data frame containing an NA). \texttt{na.exclude} does much of
-the same for solver computations, but keeps the rows with NA elements so
-that predictions are in the correct row position. We recommend that
-workers actually test output to verify the behaviour is as wanted. See
-\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}.
-As with \texttt{subset}, our concern with \texttt{na.action} is that
-users may be unaware of the effects of an option they may not even be
-aware has been set. Should \texttt{na.fail} be the default?
+\texttt{getOption("na.action")}
+is \texttt{na.omit}. This option removes from computations any observations
+containing missing values (i.e.~any row of a data frame containing an NA).
+\texttt{na.exclude} does much of the same for solver computations, but keeps the rows with
+NA elements so that predictions are in the correct row position. We recommend that
+workers actually test output to verify the behavior is as wanted.
+See \emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}.
+As with \texttt{subset}, our concern with \texttt{na.action} is that users may be unaware of the
+effects of an option they may not even be aware has been set. Should \texttt{na.fail} be the default?
 
@@
 
-\texttt{model} is an argument to the \texttt{nls()} and related
-functions, which is documented as:
+\texttt{model} is an argument to the \texttt{nls()} and related functions, which is documented as:
 
 \begin{quote}
-\textbf{model} logical. If true, the model frame is returned as part of
-the object. Default is FALSE.
+\textbf{model} logical. If true, the model frame is returned as part of the object. Default is FALSE.
 \end{quote}
 
-Indeed, the argument only gets used when \texttt{nls()} is about to
-return its result object, and the element \texttt{model} is NULL unless
-the calling argument \texttt{model} is TRUE. (Using the same name could
-be confusing.) Despite this, the model frame is used within the function
-code in the form of the object \texttt{mf}. We feel that users could
-benefit from more extensive documentation and examples of its use since
-it is used to implement features like \texttt{subset}.
+Indeed, the argument only gets used when \texttt{nls()} is about to return its result
+object, and the
+element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. (Using the same
+name could
+be confusing.) Despite this, the model frame is used within the function code in
+the form of the object
+\texttt{mf}. We feel that users could benefit from more extensive documentation and
+examples of its use since it is used to implement features like \texttt{subset}.
 
@@
 
-All four main tools we consider here allow a \texttt{weights} argument
-that specifies a vector of fixed weights the same length as the number
-of residuals. Each residual is multiplied by the square root of the
-corresponding weight. Where available, the values returned by the
-\texttt{residuals()} function are weighted, and the \texttt{fitted()} or
-\texttt{predict()} function are used to compute raw residuals.
-
-While fixed weights may be useful, there are many problems for which we
-want weights that are determined at least partially from the model
-parameters, for example, a measure of the standard deviation of
-observations. Such dynamic weighting situations are discussed in the
-vignette ``Introduction to nlsr'' of package \texttt{nlsr} in section
+All four main tools we consider here allow a \texttt{weights} argument that
+specifies a vector of fixed weights the same length as the number of residuals.
+Each residual is multiplied by the square root of the corresponding weight.
+Where available, the values returned by the
+\texttt{residuals()} function are weighted, and the \texttt{fitted()} or \texttt{predict()} function are
+used to compute raw residuals.
+
+While fixed weights may be useful, there are many problems for which
+we want weights that are determined at least partially from the model
+parameters, for
+example, a measure of the standard deviation of observations.
+Such dynamic weighting situations are discussed in the vignette
+``Introduction to nlsr'' of package \texttt{nlsr} in section
 \emph{Weights that are functions of the model parameters}.
-\texttt{minpack.lm} offers a function \texttt{wfct()} to facilitate such
-weighting. Care is advised in applying such ideas.
+\texttt{minpack.lm} offers a function \texttt{wfct()} to
+facilitate such weighting. Care is advised
+in applying such ideas.
 
 \hypertarget{weights-in-returned-functions-from-nls}{%
-\subsubsection{Weights in returned functions from
-nls()}\label{weights-in-returned-functions-from-nls}}
-
-The function \texttt{resid()} (an alias for \texttt{residuals()}) gives
-WEIGHTED residuals, as does, for example, \texttt{result\$m\$resid()}.
-The function \texttt{nlsModel()}, which we have had to extract from the
-base R code and explicitly \texttt{source()} because it is not exported
-to the working namespace, allows us to compute residuals for particular
-coefficient sets.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
-\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
-\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
-\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
-\NormalTok{rnowt}\OtherTok{\textless{}{-}}\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian.
-Does NOT take coefficients.}
-\FunctionTok{attr}\NormalTok{(rnowt, }\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rnowt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
-##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
-\NormalTok{rusewt}\OtherTok{\textless{}{-}}\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# WEIGHTED. Does NOT take coefficients.}
-\FunctionTok{attr}\NormalTok{(rusewt,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rusewt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
-\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{rn0}\OtherTok{\textless{}{-}}\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
-\FunctionTok{attr}\NormalTok{(rn0,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn0}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
-## [11] 1.6572 1.4214
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{rn}\OtherTok{\textless{}{-}}\NormalTok{nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
-\FunctionTok{attr}\NormalTok{(rn,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}
+
+The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
+as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}, which we
+have had to extract from the base R code and explicitly \texttt{source()} because it
+is not exported to the working namespace,
+allows us to compute residuals for particular coefficient sets.
+
+\begin{verbatim}
+wts <- 0.5^tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
+Asym<-1; xmid<-1; scal<-1
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+attr(rnowt, "gradient")<-NULL; rnowt
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
+#>  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
+\end{verbatim}
+
+\begin{verbatim}
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+attr(rusewt,"gradient")<-NULL; rusewt
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\end{verbatim}
+
+\begin{verbatim}
+source("nlsModel.R")
+nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+attr(rn0,"gradient")<-NULL; rn0
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
+#> [11] 1.6572 1.4214
+\end{verbatim}
+
+\begin{verbatim}
+nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
+rn<-nmod$resid()
+attr(rn,"gradient")<-NULL; rn
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
 \end{verbatim}
@@
 \hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
-\subsection{Minor issues with nonlinear least-squares
-tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
+\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
 
 \hypertarget{interim-output-from-the-port-algorithm}{%
-\subsubsection{Interim output from the ``port''
-algorithm}\label{interim-output-from-the-port-algorithm}}
-
-As the \texttt{nls()} \textbf{man} page states, when the ``port''
-algorithm is used with the \texttt{trace} argument TRUE, the iterations
-display the objective function value which is 1/2 the sum of squares (or
-deviance). The trace display is likely embedded in the Fortran of the
-\texttt{nlminb} routine that is called to execute the ``port''
-algorithm, but the factor of 2 discrepancy is nonetheless unfortunate
-for users.
+\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}
+
+As the \texttt{nls()} \textbf{man} page states, when the ``port'' algorithm is used with the
+\texttt{trace} argument TRUE, the iterations display the objective function value which
+is 1/2 the sum of squares (or deviance). The trace display is likely
+embedded in the Fortran of the \texttt{nlminb} routine that is called to execute
+the ``port'' algorithm, but the factor of 2 discrepancy is nonetheless
+unfortunate for users.
 
 \hypertarget{failure-to-return-the-best-result-achieved}{%
-\subsubsection{Failure to return the best result
-achieved}\label{failure-to-return-the-best-result-achieved}}
-
-If \texttt{nls()} reaches a point where it cannot continue but has not
-found a point where the relative offset convergence criterion is met, it
-may simply exit, especially if a ``singular gradient'' (singular
-Jacobian) is found. However, this may occur AFTER the function has made
-considerable progress in reducing the sum of squared residuals. Here is
-an abbreviated example:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
-\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
-\NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
-\NormalTok{NLSstart }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{lrc1 =} \SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\AttributeTok{lrc2 =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{A1 =} \DecValTok{150}\NormalTok{, }\AttributeTok{A2 =} \DecValTok{50}\NormalTok{) }\CommentTok{\# a starting vector (named!)}
-\NormalTok{NLSformula }\OtherTok{\textless{}{-}}\NormalTok{ conc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ A1 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc1) }\SpecialCharTok{*}\NormalTok{ time) }\SpecialCharTok{+}\NormalTok{ A2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc2) }\SpecialCharTok{*}\NormalTok{ time)}
-\NormalTok{tryit }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(NLSformula, }\AttributeTok{data =}\NormalTok{ NLSdata, }\AttributeTok{start =}\NormalTok{ NLSstart, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##      61216. (3.56e+03): par = (-2 0.25 150 50)
-## 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
-## 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
-## Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(tryit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
-## attr(,"class")
-## [1] "try-error"
-## attr(,"condition")
-## <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
+\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}}
+
+If \texttt{nls()} reaches a point where it cannot continue but has not found a point
+where the relative offset convergence criterion is met, it may simply exit,
+especially if a ``singular gradient'' (singular Jacobian) is found. However,
+this may occur AFTER the function has made considerable progress in reducing
+the sum of squared residuals.
+Here is an abbreviated example:
+
+\begin{verbatim}
+#> 61216.    (3.56e+03): par = (-2 0.25 150 50)
+#> 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
+#> 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
+#> Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
+#> attr(,"class")
+#> [1] "try-error"
+#> attr(,"condition")
+#> <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
 \end{verbatim}
@@
 Note that the sum of squares has been reduced from 61216 to 1.6211, but
-unless \texttt{trace} is invoked, the user will not get any information
-about this. This almost trivial change to the \texttt{nls()} function
-could be useful to R users.
+unless \texttt{trace} is invoked, the user will not get any information about this.
+This almost trivial change to the \texttt{nls()} function could be
+useful to R users.
 
 \hypertarget{estimating-models-that-are-partially-linear}{%
-\subsection{Estimating models that are partially
-linear}\label{estimating-models-that-are-partially-linear}}
-
-The variable projection method (Golub and Pereyra
-(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
-(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
-effective than general approaches in finding good solutions to nonlinear
-least squares problems when some of the parameters appear linearly. In
-our logistic examples, the asymptote parameters are an illustration.
-However, identifying which parameters are linear and communicating this
-information to estimating functions is not a trivial task.
-\texttt{nls()} has an option \texttt{algorithm="plinear"} that allows
-some partially linear models to be solved. The other tools, as far as we
-are aware, do not offer any such capability. The \texttt{nlstac} package
-uses a different algorithm for similar goals.
-
-Within \texttt{nls()} itself we must, unfortunately, use different
-specifications with different \texttt{algorithm} options. For example,
-the explicit model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not
-work with the linear modeling function \texttt{lm()}, which requires
-this model to be specified as \texttt{y\ \textasciitilde{}\ x}. Within
-\texttt{nls()}, consider the following FOUR different specifications for
-the same problem, plus an intuitive choice, labeled \texttt{fm2a}, that
-does NOT work. In this failed attempt, putting the \texttt{Asym}
-parameter in the model causes the \texttt{plinear} algorithm to try to
-add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing, we do not
-foresee a resolution for this issue. In the example, we have NOT
-evaluated the commands to save space.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{DNase1 }\OtherTok{\textless{}{-}} \FunctionTok{subset}\NormalTok{(DNase, Run }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\CommentTok{\# select the data}
-\DocumentationTok{\#\# using a selfStart model {-} do not specify the starting parameters}
-\NormalTok{fm1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(}\FunctionTok{log}\NormalTok{(conc), Asym, xmid, scal), DNase1)}
-\FunctionTok{summary}\NormalTok{(fm1)}
-
-\DocumentationTok{\#\# using conditional linearity {-} leave out the Asym parameter}
-\NormalTok{fm2 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
-\FunctionTok{summary}\NormalTok{(fm2)}
-
-\DocumentationTok{\#\# without conditional linearity}
-\NormalTok{fm3 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1,}
-                 \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm3)}
-
-\DocumentationTok{\#\# using Port\textquotesingle{}s nl2sol algorithm}
-\NormalTok{fm4 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm4)}
-
-\DocumentationTok{\#\# using conditional linearity AND Asym does NOT work}
-\NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm2a)}
-\end{Highlighting}
-\end{Shaded}
+\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}}
+
+The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is
+generally much more effective than general approaches in finding good solutions
+to nonlinear least squares
+problems when some of the parameters appear linearly. In our logistic examples,
+the asymptote parameters are an illustration. However, identifying which parameters
+are linear and communicating this information to estimating functions
+is not a trivial task. \texttt{nls()} has an option \texttt{algorithm="plinear"} that allows some partially linear
+models to be solved. The other tools, as far as we are aware, do not offer any
+such capability. The \texttt{nlstac} package uses a different algorithm for similar goals.
+
+Within \texttt{nls()} itself we must, unfortunately, use
+different specifications with different \texttt{algorithm} options.
+For example, the explicit model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work with the linear
+modeling function \texttt{lm()}, which requires this model to be specified as \texttt{y\ \textasciitilde{}\ x}.
+Within \texttt{nls()}, consider the following FOUR different specifications for the same
+problem, plus an intuitive choice, labeled \texttt{fm2a}, that does NOT work.
+In this failed attempt, putting the \texttt{Asym} parameter in the model causes the
+\texttt{plinear} algorithm
+to try to add another term to the model. We believe this is unfortunate, and would
+like to see a consistent syntax. At the time of writing, we do
+not foresee a resolution for this issue. In the example, we have NOT evaluated
+the commands to save space.
+
+\begin{verbatim}
+DNase1 <- subset(DNase, Run == 1) # select the data
+## using a selfStart model - do not specify the starting parameters
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
+
+## using conditional linearity - leave out the Asym parameter
+fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(xmid = 0, scal = 1),
+                 algorithm = "plinear")
+summary(fm2)
+
+## without conditional linearity
+fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1,
+                 start = list(Asym = 3, xmid = 0, scal = 1))
+summary(fm3)
+
+## using Port's nl2sol algorithm
+fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+                 algorithm = "port"))
+summary(fm4)
+
+## using conditional linearity AND Asym does NOT work
+fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
+                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                 algorithm = "plinear", trace = TRUE))
+summary(fm2a)
+\end{verbatim}
 
 \hypertarget{models-with-indexed-parameters}{%
-\subsection{Models with indexed
-parameters}\label{models-with-indexed-parameters}}
-
-Some models have several common parameters and others that are tied to
-particular cases. The \textbf{man} file for \texttt{nls()} includes an
-example of a situation in which parameters are indexed but which uses
-the ``plinear'' option as an added complication. Running this example
-reveals that the answers for the parameters are NOT indexed as in a
-vector. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
-but \texttt{a1,\ a2,\ a3}. This is no doubt because programming for
-indexed parameters is challenging. We note that there are capabilities
-in packages for mixed-effects modeling such as \texttt{nlme} (Pinheiro
-et al. (\protect\hyperlink{ref-p-nlme}{2013})), \texttt{bbmle} (Bolker
-and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4}
-(p-lme4) for estimating models of such type.
+\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}}
+
+Some models have several common parameters and
+others that are tied to particular cases.
+The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
+parameters are indexed but which uses the ``plinear'' option as an added complication.
+Running this example reveals that the answers for the parameters are NOT indexed
+as in a vector.
+That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
+This is no doubt because programming for indexed parameters is challenging.
+We note that there are capabilities in packages for mixed-effects modeling
+such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (p-lme4) for
+estimating models of such type.
 
 \hypertarget{tests-and-use-case-examples}{%
-\subsection{Tests and use-case
-examples}\label{tests-and-use-case-examples}}
-
-Maintainers of packages need suitable tests and use-case examples in
-order
+\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}}
+
+Maintainers of packages need suitable tests and use-case examples in order
 
@@
 \item
-  to ensure packages work properly, in particular, giving results
-  comparable to or better than the functions they are to replace.
-\item
-  to test individual solver functions to ensure they work across the
-  range of calling mechanisms, that is, different ways of supplying
-  inputs to the solver(s);
-\item
-  to pose ``silly'' inputs to see if these bad inputs are caught by the
-  programs.
+  to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace.
+\item
+  to test individual solver functions to ensure they work across the range of
+  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
+\item
+  to pose ``silly'' inputs to see if these bad inputs are caught by the programs.
 \end{itemize}
 
-Such goals align with the aims of \textbf{unit testing} (e.g.,
-\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
-Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
-(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
-package testing tools). In our work, one of us has developed a working
-prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}}
-. A primary design objective of this is to allow the summarisation of
-multiple tests in a compact output. The prototype has a vignette to
-illustrate its use.
+Such goals align with the aims of \textbf{unit testing}
+(e.g., \emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}}, Wickham (2011), Wickham et al. (2021)
+and the conventional R package testing tools).
+In our work, one of us has developed a working prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}} .
+A primary design objective of this is to allow the summarization of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
 
 \hypertarget{documentation-and-resources}{%
-\subsection{Documentation and
-resources}\label{documentation-and-resources}}
+\subsection{Documentation and resources}\label{documentation-and-resources}}
 
 In our investigation, we built several resources, which are now part of
-the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
-Particular items are:
+the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}. Particular items
+are:
 
@@
 \item
-  A BibTex bibliography for use with all documents in this project, but
-  which has wider application to nonlinear least squares projects in
-  general
+  A BibTex bibliography for use with all documents in this project, but which
+  has wider application to nonlinear least squares projects in general
   (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
 \item
-  \texttt{MachID.R} offers a suggested concise summary function to
-  identify a particular computational system used for tests. A
-  discussion of how it was built and the resources needed across
-  platforms is given in at
-  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
-\item
-  John C. Nash and Bhattacharjee
-  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
-  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
-  in R-base.
-\item
-  As the 2021 Summer of Code period was ending, one of us (JN) was
-  invited to prepare a review of optimization in R. Ideas from the
-  present work have been instrumental in the creation of John C. Nash
-  (\protect\hyperlink{ref-NashWires22}{2022}).
+  \texttt{MachID.R} offers a suggested concise summary function
+  to identify a particular computational system used for tests. A discussion
+  of how it was built and the resources needed across platforms is given in
+  at \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
+\item
+  John C. Nash and Bhattacharjee (2022) is an explanation of the construction of the \texttt{nlspkg} from the
+  \texttt{nls()} code in R-base.
+\item
+  As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
+  a review of optimization in R. Ideas from the present work have been instrumental
+  in the creation of John C. Nash (2022).
 \end{itemize}
@@
 \hypertarget{future-of-nonlinear-model-estimation-in-r}{%
-\subsection{Future of nonlinear model estimation in
-R}\label{future-of-nonlinear-model-estimation-in-r}}
-
-Given its importance to R, it is possible that \texttt{nls()} will
-remain more or less as it has been for the past several decades. If so,
-the focus of discussion should be the measures needed to secure its
-continued operation for legacy purposes and how that may be
-accomplished. We welcome an opportunity to participate in such
-conversations.
-
-To advance the stability and maintainability of R, we believe the
-program objects (R functions) that are created by tools such as
-\texttt{nls()} should have minimal cross-linkages and side-effects. The
-aspects of \texttt{nls()} that concern us follow.
+\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}}
+
+Given its importance to R, it is possible that \texttt{nls()} will remain more or less
+as it has been for the past several decades. If so, the focus of discussion
+should be the measures needed to secure its continued operation for legacy
+purposes and how that may be accomplished.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program
+objects (R functions)
+that are created by tools such as \texttt{nls()} should have minimal
+cross-linkages and side-effects.
+The aspects of \texttt{nls()} that concern us follow.
 
@@
   accessible environment. This provides a compact syntax to invoke the
-  calculations, but the wrong data can be used if the internal search
-  finds a valid name that is not the object we want, or if
-  \texttt{subset} or \texttt{na.action} settings modify the selection,
-  or \texttt{weights} are applied.
+  calculations, but the wrong data can be used if the
+  internal search finds a valid name that is not the object we want,
+  or if \texttt{subset} or \texttt{na.action} settings modify the selection, or
+  \texttt{weights} are applied.
 \item
@@
 \item
-  The class \texttt{nls} structure simplifies calls with its rich set of
-  functions, but also adds to the task of understanding what has been
-  done. A design that isolates the setup, solution, and post-solution
-  parts of complicated calculations reduces the number of objects that
-  must be kept in alignment.
+  The class \texttt{nls} structure simplifies calls with its rich set of functions, but
+  also adds to the task of understanding what has been done. A design that isolates
+  the setup, solution, and post-solution parts of complicated calculations reduces
+  the number of objects that must be kept in alignment.
 \end{itemize}
 
-Given the existence of examples of good practices such as analytic
-derivatives, stabilized solution of Gauss-Newton equations, and
-bounds-constrained parameters, base R tools should be moving to
-incorporate them. These capabilities are available now by using several
-tools, but it would be helpful if they were unified.
+Given the existence of examples of good practices such as analytic derivatives,
+stabilized solution of Gauss-Newton equations, and bounds-constrained parameters,
+base R tools should
+be moving to incorporate them. These capabilities are available now by
+using several tools, but it would be helpful if they were unified.
 
@@
 
-Hans Werner Borchers was helpful in developing the GSoC project
-motivating this article and in comments on this and related work.
-Heather Turner co-mentored the project and helped guide the progress of
-the work. Exchanges with Fernando Miguez helped to clarify aspects of
-\texttt{selfStart} models and instigated the ``Introduction to nlsr''
-vignette. Colin Gillespie (package \texttt{benchmarkme}) has been
-helpful in guiding our attempts to succinctly summarize computing
-environments.
+Hans Werner Borchers was helpful in developing the GSoC project motivating this article
+and in comments on this and related work. Heather Turner co-mentored the project
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of \texttt{selfStart} models and instigated the ``Introduction to nlsr''
+vignette. Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our
+attempts to succinctly summarize computing environments.
 
@@
 \section{References}\label{references}}
+
+\hypertarget{references-1}{%
+\section*{References}\label{references-1}}
+\addcontentsline{toc}{section}{References}
 
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
-Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear
-Least-Squares Regression}.
-\url{https://CRAN.R-project.org/package=onls}.
+Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
-Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis
-and Its Applications}. Wiley.
+Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
 
 \leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
-Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
-Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
-\emph{Technometrics} 23 (2): 179--83.
+Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
-Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools:
-Tools for Nonlinear Regression Diagnostics}.
+Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
----------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression
-Models Dedicated to Predictive Microbiology}.
+---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
-Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for
-General Maximum Likelihood Estimation}.
-\url{http://CRAN.R-project.org/package=bbmle}.
+Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
-Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}.
-\url{https://CRAN.R-project.org/package=gslnls}.
+Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
-Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for
-Non-Linear Regression}.
-\url{https://CRAN.R-project.org/package=nlshelper}.
+Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
-Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben
-Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt
-Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for
-Bounds}. R Project for Statistical Computing.
+Elzhov, Timur V., Katharine M.
+Mullen, Andrej-Nikolai Spiess, and Ben Bolker.
+2012.
+\emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}.
+R Project for Statistical Computing.
 \url{http://CRAN.R-project.org/package=minpack.lm}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
-Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard
-Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third
-Edition, for GSL Version 1.12 (3. Ed.).}
+Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
-Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad,
-Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004.
-{``Bioconductor: Open Software Development for Computational Biology and
-Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
+Gentleman, Robert C., Vincent J.
+Carey, Douglas M.
+Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al.
+2004.
+{``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
 \url{https://doi.org/10.1186/gb-2004-5-10-r80}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
-Golub, G. H., and V. Pereyra. 1973. {``The Differentiation of
-Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables
-Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
+Golub, G.
+H., and V.
+Pereyra.
+1973.
+{``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
-Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute
-Force}. \url{https://CRAN.R-project.org/package=nls2}.
+Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
-Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting
-of Non-Linear Regression Functions by Least Squares.''}
-\emph{Technometrics} 3: 269-\/-280.
+Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
-Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004.
-\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with
-{S-PLUS} Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.
+Huet, S., A.
+Bouvier, M.-A.
+Poursat, and E.
+Jolivet.
+2004.
+\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with {S-PLUS} Examples, 2nd Edition}.
+Berlin \& New York: Springer-Verlag.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
-John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for
-Nonlinear Least Squares Solutions}}.
+John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
-Josef Tvrdík and Ivan Křivý and Ladislav Mišík. 2007. {``Adaptive
-Population-Based Search: Application to Estimation of Nonlinear
-Regression Parameters.''} \emph{{Computational Statistics \& Data
-Analysis}} 52 (2): 713--24.
+Josef Tvrdík and Ivan Křivý and Ladislav Mišík.
+2007.
+{``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.''} \emph{{Computational Statistics \& Data Analysis}} 52 (2): 713--24.
 \url{https://doi.org/10.1016/j.csda.2006.10.014}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
-Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain
-Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied
-Mathematics} 2: 164-\/-168.
+Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
-Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation
-of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics}
-11 (2): 431--41.
+Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
 
 \leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
-Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for
-Agricultural Applications}}.
-\url{https://CRAN.R-project.org/package=nlraa}.
+Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
-Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on
-a Small Computer.''} \emph{Journal of the Institute for Mathematics and
-Its Applications} 19: 231--37.
+Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
 
 \leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
----------. 1979. \emph{Compact Numerical Methods for Computers: Linear
-Algebra and Function Minimisation}. Bristol: Adam Hilger.
+---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
 \leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
----------. 2022. {``{Function minimization and nonlinear least squares
-in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
-https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
+---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
-Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package
-from Base {R} Files.''} \emph{R-Bloggers}, January.
-\url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
+Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
-Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A
-Replacement and Extension of the {optim()} Function}. Nash Information
-Services Inc.; Johns Hopkins University.
+Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
-Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation
-Combining Visual Fitting with Optimization Methods.''} In
-\emph{Proceedings of the Section on Physical and Engineering Sciences of
-the American Statistical Association}, 256--61. American Statistical
-Association.
+Nash, John C., and Paul Velleman.
+1996.
+{``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61.
+American Statistical Association.
 
 \leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
-National Institutes for Standards, original from, and Technology
-http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by
-Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples
-from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.
+National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.
+2012.
+\emph{NISTnls: Nonlinear Least Squares Examples from NIST}.
+\url{https://CRAN.R-project.org/package=NISTnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
-O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
-Nonlinear Least Squares Problems.''} \emph{Computational Optimization
-and Applications} 54 (3): 579--93.
+O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
-Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart:
-Robust Non-Linear Regression Using AIC Scores}.
-\url{https://CRAN.R-project.org/package=nls.multstart}.
+Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
-Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R
-Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
+Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
-Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn:
-Fitting Nonlinear Models with Scale Mixture of Skew-Normal
-Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
+Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
-Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified
-Practical Approach}. New York; Basel: Marcel Dekker Inc.
+Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
 \leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
-Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and
-Rafael Benitez. 2020. \emph{Nlstac: An r Package for Fitting Separable
-Nonlinear Models}. \url{https://CRAN.R-project.org/package=nlstac}.
+Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez.
+2020.
+\emph{Nlstac: An r Package for Fitting Separable Nonlinear Models}.
+\url{https://CRAN.R-project.org/package=nlstac}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%
-Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with
-Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
+Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Torvisco2018}{}}%
+Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez.
+2018.
+{``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Niˇs, Serbia}.
+\url{https://doi.org/doi.org/10.2298/FIL1812233T}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
-Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters
-Estimation by 'CRS4HC' and 'CRS4HCe'}.
-\url{https://CRAN.R-project.org/package=crsnls}.
+Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
-Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
-\emph{The R Journal} 3: 5--10.
-\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
+Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
-Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
-\emph{{devtools: Tools to Make Developing R Packages Easier}}.
-\url{https://CRAN.R-project.org/package=devtools}.
+Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.
 
@@
 
-\end{document}
+\bibliography{./ImproveNLS.bib}
+
+\address{%
+John C. Nash\\
+retired professor, University of Ottawa\\%
+Telfer School of Management\\ Ottawa ON Canada K1N 6N5\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0002-2762-8039}{0000-0002-2762-8039}}\\%
+\href{mailto:profjcnash@gmail.com}{\nolinkurl{profjcnash@gmail.com}}%
+}
+
+\address{%
+Arkajyoti Bhattacharjee\\
+Indian Institute of Technology\\%
+Department of Mathematics and Statistics\\ Kanpur\\
+%
+%
+%
+\href{mailto:arkastat98@gmail.com}{\nolinkurl{arkastat98@gmail.com}}%
+}

### 2023-05-17 DNase1Models.R

@@ -1,28 +1,0 @@
-DNase1 <- subset(DNase, Run == 1) # select the data
-## using a selfStart model - do not specify the starting parameters
-fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
-summary(fm1)
-
-## using conditional linearity - leave out the Asym parameter
-fm2 <- nls(density ~ 1/(1 + exp((xmid - log(conc))/scal)),
-           data = DNase1, start = list(xmid = 0, scal = 1),
-           algorithm = "plinear")
-summary(fm2)
-## without conditional linearity
-fm3 <- nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
-           data = DNase1,
-           start = list(Asym = 3, xmid = 0, scal = 1))
-summary(fm3)
-
-## using Port's nl2sol algorithm
-fm4 <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)),
-               data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
-               algorithm = "port"))
-summary(fm4)
-
-## using conditional linearity AND Asym does NOT work
-fm2a <- try(nls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
-                data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
-                algorithm = "plinear", trace = TRUE))
-summary(fm2a)
-

### 2023-05-17 JNABsub20230207/JNAB230203.R

@@ -1,139 +1,0 @@
-# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
-# Please edit JNAB230203.Rmd to modify this file
-
-## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-
-
-## ----SESingGrad---------------------------------------------------------------
-reala <- -3; realb <- 5; realc <- 0.5; realr <- 0.7; realm <- 1 # underlying parameters
-x <- 1:11 # x values; 11 data points
-y <- reala + realb * realr^(x - realm) + realc * x # linear + exponential function
-testdat <- data.frame(x, y) # save in a data frame
-strt <- list(a = -3, b = 5, c = 0.5, r = 0.7, m = 1) # give programs a good start
-jform <- y ~ a + b * r^(x - m) + c * x # Formula
-library(nlsr)
-linexp2 <- try(nlxb(jform, data = testdat, start = strt, trace = FALSE))
-linexp2 # Note singular values of Jacobian in rightmost column
-# We get an error with nls()
-linexp <- try(nls(jform, data = testdat, start = strt, trace = FALSE))
-
-
-## ----nlswtx, echo=TRUE--------------------------------------------------------
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-weeddf <- data.frame(tt, weed)
-wts <- 0.5^tt # simple weights
-frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
-Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-nowt
-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
-usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-usewt
-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
-source("nlsModel.R")
-nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
[…]

### 2023-05-17 JNABsub20230207/nlsModel.R

@@ -1,170 +1,0 @@
-#  From file src/library/stats/R/nls.R
-#  Part of the R package, https://www.R-project.org
-#
-#  Copyright (C) 2000-2020 The R Core Team
-#  Copyright (C) 1999-1999 Saikat DebRoy, Douglas M. Bates, Jose C. Pinheiro
-#
-#  The nlsModel() function extracted here.
-
-nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
-{
-    ## thisEnv <- environment() # shared by all functions in the 'm' list; variable no longer needed
-    env <- new.env(hash = TRUE, parent = environment(form))
-    for(i in names(data)) env[[i]] <- data[[i]]
-    ind <- as.list(start)
-    parLength <- 0L
-    for(i in names(ind) ) {
-        temp <- start[[i]]
-        storage.mode(temp) <- "double"
-        env[[i]] <- temp
-        ind[[i]] <- parLength + seq_along(temp)
-        parLength <- parLength + length(temp)
-    }
-    getPars.noVarying <- function() unlist(mget(names(ind), env))
-    getPars <- getPars.noVarying
-    internalPars <- getPars()
-
-    if(!is.null(upper)) upper <- rep_len(upper, parLength)
-    useParams <- rep_len(TRUE, parLength)
-    lhs <- eval(form[[2L]], envir = env)
-    rhs <- eval(form[[3L]], envir = env)
-    .swts <- if(!missing(wts) && length(wts))
-        sqrt(wts) else rep_len(1, length(rhs))
-    env$.swts <- .swts
-    resid <- .swts * (lhs - rhs)
-    dev <- sum(resid^2)
-    if(is.null(attr(rhs, "gradient"))) {
-        getRHS.noVarying <- function() {
-            if(is.null(upper)) # always for "default"
-                numericDeriv(form[[3L]], names(ind), env, central = nDcentral)
-            else # possibly with "port"
-                numericDeriv(form[[3L]], names(ind), env,
-                             dir = ## ifelse(internalPars < upper, 1, -1)
-                                 -1 + 2*(internalPars < upper), central = nDcentral)
-        ## 210701JN: Above line should check for nDcentral=FALSE when approaching
-        ## upper bound, as central diff is same whether dir is +1 or -1. ???
-        }
[…]

### 2023-05-17 JNABsub20230515/ComparingNLStools.R

@@ -1,0 +1,231 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit ComparingNLStools.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----ex01, echo=TRUE----------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE-------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
+frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
+frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+#
+# Starting parameter sets
+stu1<-c(b1=1, b2=1, b3=1)
+sts1<-c(c1=1, c2=1, c3=1)
+stt1<-c(Asym=1, xmid=1, scal=1)
+
+
+## ----ex02fn, echo=TRUE--------------------------------------------------------
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3]*tt)
+  zz  <-  1.0/(1+x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -x[1]*zz*zz*yy
[…]

### 2023-05-17 hobbSSlogistry.R

@@ -1,512 +1,0 @@
-?microbenchmark
-source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopesctimes.R")
-library(benchmarkme)
-get_sys_details
-sessionInfo
-sessionInfo()
-?COMPILE
-library(compiler)
-?compile
-cmpfun(loopesc)
-le<-cmpfun(loopesc)
-microbenchmark(le)
-microbenchmark(le, unit=us)
-microbenchmark(le, unit="us")
-microbenchmark(loopesc, unit="us")
-nn
-microbenchmark(le(nn), unit="us")
-microbenchmark(loopesc(nn), unit="us")
-microbenchmark()
-?microbenchmark
-boxplot.microbenchmark(t2)
-boxplot.microbenchmark(tt2)
-library(microbenchmark)
-?boxplot.microbenchmark
-# loopbox.R
-loopesc <- function(nn){
-ss <- 0
-for (i in 1:nn) {
-xx <- exp(sin(cos(1.0*i)))
-ss <- ss + xx
-}
-xx
-}
-require("microbenchmark")
-nn <- 10000
-library(compiler)
-le <- cmpfun(loopesc)
-tcompiled <- microbenchmark(le(nn), unit='us')$time
-#  print(te)
-#  boxplot(te)
-tcoldstart <- microbenchmark(loopesc(nn), unit = 'us')$time
-#  print(tt)
-#  boxplot(tt)
-#  ht<-hist(tt$time)
-#  plot(ht, main="Histogram of loopesc times")
-twarmstart <- microbenchmark(loopesc(nn), unit = 'us', control=list(warmup=2))$time
-tdf <- data.frame(tcompiled, tcoldstart, twarmstart)
-str(tdf)
-names(tdf)
-?data.frame
-boxplot(x=as.list(tdf))
-boxplot(x=as.list(log(tdf))
-)
-?boxplot
-tdf <- data.frame(tcompiled, tcoldstart, twarmstart)
-boxplot(x=as.list(log(tdf)), main="log times for 10000 loops in microsecs")
-source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
-source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
-source("~/current/GSoC2021/RNonlinearLS/MachineSummary/loopbox.R")
[…]

### 2023-05-17 VarietyInNLS/sstimer.R
remove Variety

@@ -1,29 +1,0 @@
-# crossprod timer
-library(microbenchmark)
-suml<-function(vv) {
-    ss<-0.0
-    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
-    ss  
-}
-sums<-function(vv) {
-  ss<-sum(vv^2)
-  ss  
-}
-sumc<-function(vv) {
-  ss<-as.numeric(crossprod(vv))
-  ss  
-}
-## ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
-ll <- c(100, 1000, 10000, 100000, 1000000)
-cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
-for (nn in ll ){
-   set.seed(1234)
-   vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
-   tsums<-microbenchmark(ss<-sums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
-   ml<-mean(tsuml$time)
-   ms<-mean(tsums$time)
-   mc<-mean(tsumc$time)
-   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
-}

### 2023-05-17 VarietyInNLS/sstimer2.R

@@ -1,35 +1,0 @@
-# crossprod timer
-library(microbenchmark)
-suml<-function(vv) {
-    ss<-0.0
-    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
-    ss  
-}
-sums<-function(vv) {
-  ss<-sum(vv^2)
-  ss  
-}
-sumc<-function(vv) {
-  ss<-as.numeric(crossprod(vv))
-  ss  
-}
-
-csuml<-compiler::cmpfun(suml)
-csumc<-compiler::cmpfun(sumc)
-csums<-compiler::cmpfun(sums)
-
-# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
-ll <- c(100, 1000, 10000, 100000, 1000000)
-
-cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
-for (nn in ll ){
-   set.seed(1234)
-   vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-csuml(vv), unit="us")
-   tsums<-microbenchmark(ss<-csums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-csumc(vv), unit="us")
-   ml<-mean(tsuml$time)
-   ms<-mean(tsums$time)
-   mc<-mean(tsumc$time)
-   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
-}

### 2023-05-17 VarietyInNLS/sstimer3.R

@@ -1,32 +1,0 @@
-# crossprod timer
-library(microbenchmark)
-suml<-function(vv) {
-    ss<-0.0
-    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
-    ss  
-}
-sums<-function(vv) {
-  ss<-sum(vv^2)
-  ss  
-}
-sumc<-function(vv) {
-  ss<-as.numeric(crossprod(vv))
-  ss  
-}
-
-
-## ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
-ll <- c(100, 1000, 10000, 100000, 1000000)
-
-cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
-for (nn in ll ){
-   set.seed(1234)
-   vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us", setup=suml(vv))
-   tsums<-microbenchmark(ss<-sums(vv), unit="us", setup=sums(vv))
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us", setup=sumc(vv))
-   ml<-mean(tsuml$time)
-   ms<-mean(tsums$time)
-   mc<-mean(tsumc$time)
-   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
-}

### 2023-05-17 VarietyInNLS/sstimer4.R

@@ -1,42 +1,0 @@
-# crossprod timer
-library(microbenchmark)
-suml<-function(vv) {
-    ss<-0.0
-    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
-    ss  
-}
-sums<-function(vv) {
-  ss<-sum(vv^2)
-  ss  
-}
-sumc<-function(vv) {
-  ss<-as.numeric(crossprod(vv))
-  ss  
-}
-
-csuml<-compiler::cmpfun(suml)
-csumc<-compiler::cmpfun(sumc)
-csums<-compiler::cmpfun(sums)
-
-# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
-ll <- c(100, 1000, 10000, 100000, 1000000)
-
-cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
-for (nn in ll ){
-   set.seed(1234)
-   vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
-   tsums<-microbenchmark(ss<-sums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
-   tcsuml<-microbenchmark(sl<-csuml(vv), unit="us")
-   tcsums<-microbenchmark(ss<-csums(vv), unit="us")
-   tcsumc<-microbenchmark(sc<-csumc(vv), unit="us")
-   ml<-mean(tsuml$time)
-   ms<-mean(tsums$time)
-   mc<-mean(tsumc$time)
-   mcl<-mean(tcsuml$time)
-   mcs<-mean(tcsums$time)
-   mcc<-mean(tcsumc$time)
-   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
-   cat("\t",mcl," : ",mcl/mcc,"\t",mcs," : ",mcs/mcc,"\t",mcc,"\n")   
-}

### 2023-05-17 VarietyInNLS/sstimer5.R

@@ -1,49 +1,0 @@
-# crossprod timer
-library(microbenchmark)
-suml<-function(vv) {
-    ss<-0.0
-    for (i in 1:length(vv)) {ss<-ss+vv[i]^2}
-    ss  
-}
-sums<-function(vv) {
-  ss<-sum(vv^2)
-  ss  
-}
-sumc<-function(vv) {
-  ss<-as.numeric(crossprod(vv))
-  ss  
-}
-
-csuml<-compiler::cmpfun(suml)
-csumc<-compiler::cmpfun(sumc)
-csums<-compiler::cmpfun(sums)
-
-# ll <- c(100, 1000, 10000, 100000, 1000000, 10000000)
-ll <- c(100, 1000, 10000, 100000, 1000000)
-
-cat(" n  \t  t(forloop) : ratio \t  t(sum) : ratio \t t(crossprod) \t all.equal \n")
-for (nn in ll ){
-   set.seed(1234)
-   vv <- runif(nn)
-   tsuml<-microbenchmark(sl<-suml(vv), unit="us")
-   tsums<-microbenchmark(ss<-sums(vv), unit="us")
-   tsumc<-microbenchmark(sc<-sumc(vv), unit="us")
-   tcsuml<-microbenchmark(sl<-csuml(vv), unit="us")
-   tcsums<-microbenchmark(ss<-csums(vv), unit="us")
-   tcsumc<-microbenchmark(sc<-csumc(vv), unit="us")
-   tsumld<-microbenchmark({sl<-0;for(ii in 1:nn){sl<-sl+vv[ii]^2}}, unit="us")
-   tsumsd<-microbenchmark(ss<-sum(vv^2), unit="us")
-   tsumcd<-microbenchmark(sc<-as.numeric(crossprod(vv)), unit="us")
-   ml<-mean(tsuml$time)
-   ms<-mean(tsums$time)
-   mc<-mean(tsumc$time)
-   mcl<-mean(tcsuml$time)
-   mcs<-mean(tcsums$time)
-   mcc<-mean(tcsumc$time)
-   mld<-mean(tsumld$time)
-   msd<-mean(tsumsd$time)
-   mcd<-mean(tsumcd$time)
-   cat(nn,"\t",ml," : ",ml/mc,"\t",ms," : ",ms/mc,"\t",mc,"\t",all.equal(sl, ss, sc),"\n")   
[…]

### 2023-05-17 nlsj/R/nlsj.R
remove nlsj

@@ -1,536 +1,0 @@
-nlsj <- function (formula, data = parent.frame(), start=NULL, 
-            control = list(),
-            algorithm = "default", weights=NULL, subset=NULL, trace = FALSE,
-            na.action, model=FALSE, lower = -Inf, upper = Inf, ...) {
-# ?? left out -- FIXME??    na.action, model = FALSE, (masked from nlxb)
-# ?? at this stage ONLY treat "default", but will add bounds
-# ?? data in .GlobalEnv -- should be OK
-##?? Should a lot of this material be in nlsjModel() to build tools for problem??
-# DOES NOT CALL nlsjModel, but does everything here
-
-##?? ... args may NOT be well-defined for this function. CAUTION  
-  
-ctrl <- nlsj.control()
-ncontrol <- names(control)
-nctrl <- names(ctrl)
-for (onename in ncontrol) {
-    if (onename %in% nctrl) {
-       if (! is.null(control[onename]) || ! is.na(control[onename]) )
-       ctrl[onename]<-control[onename]
-    }
-}
-control<-ctrl
-ctrl<-NULL
-
-# Controls we need to have. ?? use ctrldefault() idea?
-   if (is.null(control$derivmeth)) control$derivmeth="default" # for safety
-   epstol <- (.Machine$double.eps * control$offset) 
-   epsh <- sqrt(epstol)
-   epstol4 <- epstol^4 # used for smallsstest
-   ##?? may want these in nlsj.control
-   if (control$derivmeth == "numericDeriv") warning("Forcing numericDeriv")
-
-# Algorithm
-   if (is.null(algorithm)) algorithm<-"default"
-   algchoice <- c("default", "port", "plinear", "marquardt")
-   alg <- which(algorithm %in% algchoice)
-   switch(alg,
-      "default" = if (trace) cat("nlsj: Using default algorithm\n"),
-      "marquardt" = if (trace) cat("nlsj: Using marquardt algorithm\n"),
-      # all other choices
-      { msg <- paste("Algorithm choice '",alg,
-                     "' not yet implemented or does not exist")
-        stop(msg) }
-   ) # end switch choices
-
-getlen <- function(ln) {
-  nn<-length(ln)
-  ll<-rep(NA,nn)
-  for (i in 1:nn) ll[i]=length(get(ln[i]))
-  ll
-}
-
-datpar <- function(frm){ # return dn and pn for data and parameter names
-  # check formulas
[…]

### 2023-05-17 nlsj/R/nlsjextra.R

@@ -1,272 +1,0 @@
-coef.nlsj <- function(object, ...) object$m$getPars() # not getAllPars since no plinear yet
-
-summary.nlsj <-
-    function (object, correlation = FALSE, symbolic.cor = FALSE, ...)
-{
-    r <- as.vector(object$m$resid()) # These are weighted residuals.
-    w <- object$weights
-    n <- if (!is.null(w)) sum(w > 0) else length(r)
-    param <- coef(object)
-    pnames <- names(param)
-    p <- length(param)
-    rdf <- n - p
-    resvar <- if(rdf <= 0) NaN else deviance(object)/rdf
-    XtXinv <- chol2inv(object$m$Rmat())
-    dimnames(XtXinv) <- list(pnames, pnames)
-    se <- sqrt(diag(XtXinv) * resvar)
-    tval <- param/se
-    param <- cbind(param, se, tval, 2 * pt(abs(tval), rdf, lower.tail = FALSE))
-    dimnames(param) <-
-        list(pnames, c("Estimate", "Std. Error", "t value", "Pr(>|t|)"))
-    ans <- list(formula = formula(object), residuals = r, sigma = sqrt(resvar),
-                df = c(p, rdf), cov.unscaled = XtXinv,
-                call = object$call,
-                convInfo = object$convInfo,
-                control = object$control,
-                na.action = object$na.action,
-                coefficients = param,
-                parameters = param)# never documented, for back-compatibility
-    if(correlation && rdf > 0) {
-        ans$correlation <- (XtXinv * resvar)/outer(se, se)
-        ans$symbolic.cor <- symbolic.cor
-    }
-    ## if(identical(object$call$algorithm, "port"))
-    ##     ans$message <- object$message
-    class(ans) <- "summary.nlsj"
-    ans
-}
-
-print.nlsj <- function(x, digits = max(3L, getOption("digits") - 3L), ...)
-{
-    cat("Nonlinear regression model\n")
-    cat("  model: ", deparse(formula(x)), "\n", sep = "")
-    cat("   data: ", deparse(x$data), "\n", sep = "")
-#    print(x$m$getAllPars(), digits = digits, ...)
-#?? getAllPars includes plinear ones??
-     print(x$m$getPars(), digits = digits, ...)
-    cat(" ", if(!is.null(x$weights) && diff(range(x$weights))) "weighted ",
-	"residual sum-of-squares: ", format(x$m$deviance(), digits = digits),
-	"\n", sep = "")
[…]

### 2023-05-17 nlsj/tests/nlsj.R

@@ -1,397 +1,0 @@
-#  File src/library/stats/tests/nlsj.R
-#  Part of the R package, https://www.R-project.org
-#
-#  This program is free software; you can redistribute it and/or modify
-#  it under the terms of the GNU General Public License as published by
-#  the Free Software Foundation; either version 2 of the License, or
-#  (at your option) any later version.
-#
-#  This program is distributed in the hope that it will be useful,
-#  but WITHOUT ANY WARRANTY; without even the implied warranty of
-#  MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE.  See the
-#  GNU General Public License for more details.
-#
-#  A copy of the GNU General Public License is available at
-#  https://www.R-project.org/Licenses/
-
-## tests of nlsj, especially of weighted fits
-library(nlsj) # ??added 210716 JN
-
-library(stats)
-options(digits = 5) # to avoid trivial printed differences
-options(useFancyQuotes = FALSE) # avoid fancy quotes in o/p
-options(show.nlsj.convergence = FALSE) # avoid non-diffable output
-options(warn = 1)
-
-have_MASS <- requireNamespace('MASS', quietly = TRUE)
-
-pdf("nlsj-test.pdf")
-
-## utility for comparing nlsj() results:  [TODO: use more often below]
-.n <- function(r) r[names(r) != "call"]
-
-## selfStart.default() w/ no parameters:
-logist <- deriv( ~Asym/(1+exp(-(x-xmid)/scal)), c("Asym", "xmid", "scal"),
-		function(x, Asym, xmid, scal){} )
-logistInit <- function(mCall, LHS, data) {
-    xy <- sortedXyData(mCall[["x"]], LHS, data)
-    if(nrow(xy) < 3) stop("Too few distinct input values to fit a logistic")
-    Asym <- max(abs(xy[,"y"]))
-    if (Asym != max(xy[,"y"])) Asym <- -Asym  # negative asymptote
-    xmid <- NLSstClosestX(xy, 0.5 * Asym)
[…]

### 2023-05-17 RefactoringNLS/Hobbsmaskx.R
add odds and ends

@@ -1,0 +1,38 @@
+# Hobbsmaskx.R
+## masks in Hobbs with formula specification of problem
+require(nlsr)
+require(minpack.lm)
+stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
+sts <- c(b1=2, b2=5, b3=3) # a default scaled starting vector (named!)
+traceval<-FALSE
+
+# fix first parameter
+anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+			upper=c(200, 60, .3), trace=traceval))
+print(anxbmsk1)
+anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+			upper=c(200, 60, .3), trace=traceval))
+print(anlM1)
+# nls gives warnings
+anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
+		upper=c(200, 60, .3),  algorithm="port", trace=traceval))
+print(anlsmsk1)
+
+
+# Hobbs scaled problem with bounds, formula specification
+anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
+             upper=c(2,6,3), algorithm='port')
+print(anlshmsk1)
+cat("More precisely...crossprod(resid(anlshmsk1))=",crossprod(resid(anlshmsk1)),"\n")
+# nlsLM does not always work with bounds
+anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                 upper=c(2,6,3))
+print(anlsLMmsks1)
+
+anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
+                  upper=c(2,6,3))
+print(anlxmsks1)
+
+# Test with all parameters masked
+anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+print(anlxmskall)

### 2023-05-17 RefactoringNLS/crs4hcEx.R
add tests of other packages

@@ -1,0 +1,31 @@
+# crs4hcEx.R
+library(microbenchmark)
+library(crsnls)
+x <- c(1,2,3,5,7,10)
+y <- c(109,149,149,191,213,224)
+df <- data.frame(x=x, y=y)
+lowerBounds <- c(1, 0.1)
+upperBounds <- c(1000, 2)
+pnam <- c("b1", "b2")
+names(lowerBounds)<-pnam
+names(upperBounds)<-pnam
+tmod<-microbenchmark(mod <- crs4hc(y ~ b1 * (1-exp(-b2*x)), df, lowerBounds, upperBounds))
+mod
+cat("evaluations =", mod$numOfEvals,"\n")
+# str(mod)
+
+tmde<-microbenchmark(mde <- crs4hce(y ~ b1 * (1-exp(-b2*x)), df, lowerBounds, upperBounds))
+mde
+cat("evaluations =", mde$numOfEvals,"\n")
+# str(mde)
+library(nlsr)
+tmodn <- microbenchmark(modn <- nlxb(formula = y ~ b1 * (1-exp(-b2*x)), data=df, lower=lowerBounds, 
+             upper=upperBounds, start=0.5*(lowerBounds+upperBounds)))
+tmodn
+modn
+summary(tmod)
+summary(tmde)
+summary(tmodn)
+mod$model$estimates
+mde$model$estimates
+coef(modn)

### 2023-05-17 RefactoringNLS/onlsEx.R

@@ -1,0 +1,87 @@
+library(onls)
+## 1A. The DNase data from 'nls',
+## use all generic functions.
+DNase1 <- subset(DNase, Run == 1)
+DNase1$density <- sapply(DNase1$density, function(x) rnorm(1, x, 0.1 * x))
+mod1 <- onls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
+             data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1))
+print(mod1)
+plot(mod1)
+summary(mod1)
+predict(mod1, newdata = data.frame(conc = 6))
+logLik(mod1)
+deviance(mod1)
+formula(mod1)
+weights(mod1)
+df.residual(mod1)
+fitted(mod1)
+residuals(mod1)
+vcov(mod1)
+coef(mod1)
+
+DNase2 <- DNase1
+DNase2$conc <- DNase2$conc * 2
+mod2 <- update(mod1, data = DNase2)
+print(mod2)
+
+
+## 1B. Same model as above, but using the restricted
+## predictor range which results in non-orthogonality
+## of some points.
+onls(density ~ Asym/(1 + exp((xmid - log(conc))/scal)), 
+     data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+     extend = c(0, 0))
+
+## 2. Example from odrpack_guide.pdf, 2.C.i, pages 39ff.
+x <- c(0, 0, 5, 7, 7.5, 10, 16, 26, 30, 34, 34.5, 100)
+y <- c(1265, 1263.6, 1258, 1254, 1253, 1249.8, 1237, 1218, 1220.6, 
+       1213.8, 1215.5, 1212)
+DAT <- data.frame(x, y)
+mod3 <- onls(y ~ b1 + b2 * (exp(b3 * x) -1)^2, data = DAT, 
+             start = list(b1 = 1500, b2 = -50, b3 = -0.1))
+deviance_o(mod3) # 21.445 as in page 47
+coef(mod3) # 1264.65481/-54.01838/-0.08785 as in page 48
+
+## 3. Example from Algorithm 676: ODRPACK, page 355 + 356.
+x <- c(0, 10, 20, 30, 40, 50, 60, 70, 80, 85, 90, 95, 100, 105)
+y <- c(4.14, 8.52, 16.31, 32.18, 64.62, 98.76, 151.13, 224.74, 341.35, 
[…]

### 2023-05-18 JNABsub20230518/ComparingNLStools.Rmd
May 18 fixes

--- 2023-05-17 JNABsub20230515/ComparingNLStools.Rmd
@@
 title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
-date: "2023-5-16"
+date: "2023-5-18"
 abstract: >
@@
 author:
-  # see ?rjournal_article for more information
     - name: John C. Nash
@@
 bibliography: ./ComparingNLStools.bib
+---
 
@@
 
+### Principal messages
+
+Our work was aimed at unifying nonlinear modeling functionality in R,
+ideally in a refactored `nls()` function.
+
+For R **users**, we would advise adapting working scripts, preferably those
+with documentation and using recent tools. If there is a suspicion that there
+may be ill-conditioning, package `nlsr` or the example we give below of how to
+find singular values of the Jacobian allow these diagnostics to be calculated.
+
+For R **developers** we invite and encourage discussion of the design choices, since these
+have downstream implications for ease of use, adaptation to new features and 
+efficiency of ongoing maintenance.
+
 ## Some other CRAN packages for nonlinear modeling
@@
 The gradient of $S(p)$ is
-$$ 2 * J' r$$
+$$ g = 2 * J' r$$
 where the Jacobian $J$ is given by elements
@@
 If we expand the Hessian for nonlinear least squares problems, we find
-$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
+$$  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
 Let us use $D_{i,j}$ for the elements of the second term of this expression.
@@
 Applying this to a sum of squares problem gives
-$$ H \delta = (J'J + D) \delta = - J' r$$
+$$ 0.5 * H \delta = (J'J + D) \delta = - J' r$$
 In this expression, only the elements of $D$ have second partial derivatives. 
@@
 that these multiplied the second partial derivatives of $r$, so he approximated
-$$ H \approx J' J$$
+$$ 0.5 * H \approx J' J$$
 by assuming $D \approx 0$. This results in the Gauss-Newton method where we
@@
 
-As mentioned, the output of `nls()` is an object of class "nls" which has a quite rich structure
-described in the manual file or revealed by applying the `str()` function to the
-result of `nls()`.
-
-The complexity of this object is a challenge to users. Let us use the
-`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
-`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
-actually access the data from this element, we need to use the syntax:
+As mentioned, the output of `nls()`, `minpack.lm::nlsLM()`, or
+`gslnls::gsl_nls()` is an object of class "nls" which has a quite rich structure
+described in the manual files or revealed by applying the `str()` function to the
+result of `nls()`.  The complexity of this object is a challenge to users. Let us 
+use for example `result <- snlm1` as the returned object from `nlsLM()` for the 
+Logistic3S problem. . The `data` return element 
+is an R symbol. To actually access the data from this element, we need to use 
+the syntax:
 ```
@@
 probability under certain assumptions.
-Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
-errors nor their derived statistics when bounds are active.
+Bounds change those assumptions and hence the interpretation of estimates of
+uncertainty, whether by traditional approximations from the $J' J$ matrix or from methods
+such as profile likelihood or bootstrap. At the time of writing, 
+`nlsr::nlxb()` does not compute standard errors nor their derived statistics 
+when bounds are active to avoid providing misleading information.
 
@@
 
-<!-- Ben Bolker mentioned a possibility of using a nonlinear model in  -->
-<!-- ??I think?? nlme with some sort of linear model form of the parameters. -->
-<!-- May need to mention. -->
-
-<!-- ===```{r nlsindx1} -->
-<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
-<!-- ## contraction on 21 animals.  The observed variables are Strip -->
-<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
-<!-- ## (resulting length of muscle section). -->
-<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
-<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE) -->
-<!-- mm<-droplevels(mm) # remove unused levels after truncation -->
-<!-- nlev <- nlevels(mm) -->
-<!-- withAutoprint({ -->
-<!--   ## The non linear model considered is -->
-<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
-<!--   ## where theta is constant. For now alpha and beta do NOT vary with Strip. -->
-<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
-<!--   nl <- nlevels(mm$Strip) -->
-<!--   ## We first use the plinear algorithm to fit an overall model, -->
-<!--   ## ignoring that alpha and beta might vary with Strip. -->
-<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
-<!--                 start = list(th = 1), algorithm = "plinear") -->
-<!--   summary(musc.1) -->
-
-<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
-<!--   ## models to use the conventional algorithm to fit a model in which -->
-<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
-<!--   ## by the previously fitted model. -->
-<!--   ## Note that with indexed parameters, the starting values must be -->
-<!--   ## given in a list (with names): -->
-<!--   b <- coef(musc.1) -->
-<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
-<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
-<!--   summary(musc.2) -->
-<!-- }) -->
-<!-- === ``` -->
-
 ## Tests and use-case examples
@@
 
-<!-- ## Should nls() be refactored? -->
-
-<!-- When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted: -->
-
-<!-- - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R  -->
-<!--   that matches the features in base R or what is packaged in `nlspkg` as described  -->
-<!--   in @PkgFromRbase22. Our intent in providing a purely R version of the code was -->
-<!--   either to replace `nls()` or to serve as a reference version aiding maintenance  -->
-<!--   of C, Fortran, or other components of the software. -->
-
-<!-- - to streamline the overall `nls()` infrastructure. By this we mean a  -->
-<!--   refactoring of the routines so they are better suited to maintenance of both -->
-<!--   the existing `nls()` methods and features as well as new features we or others -->
-<!--   would like to add. -->
-
-<!-- - to explain what we do, either in comments, examples, or separate maintainer  -->
-<!--   documentation. While R and its packages are generally very well documented for -->
-<!--   usage, that documentation -->
-<!--   rarely extends as far as it might for code maintenance. The paucity of such -->
-<!--   documentation is exacerbated by the mixed R/C/Fortran code base, -->
-<!--   where seemingly trivial differences in things like representation of numbers or -->
-<!--   base index for arrays lead to traps for unwary programmers. -->
-
-<!-- These goals echo themes in the recent discussion by @Vidoni21 and accompanying  -->
-<!-- commentary. -->
-
-<!-- We found the legacy code and examples prevented us achieving a revised `nls()`  -->
-<!-- with all its current capabilities. Unfortunately, `nls()` is part of the base R -->
-<!-- distribution,  -->
-<!-- so users are drawn to using it even when it is less than ideal. A possible way forward -->
-<!-- is to  -->
-
-<!-- - use a packaged version of `nls()` to provide legacy features, with notification -->
-<!--   to users that there are other tools available -->
-
-<!-- - if maintainers are willing, to merge and rationalize the capabilities of  -->
-<!--   `minpack.lm`, `nlsr` and `gslnls` -->
-
-<!-- - to encourage the (separate) development of capabilities such as indexed -->
-<!--   parameter models if these are in demand. -->
-
-<!-- This direction does not, of course, resolve the inconsistencies in model  -->
-<!-- specification in formulas for nonlinear models, which undoubtedly is an issue -->
-<!-- for other modeling tools.   -->
-
-<!-- A possible workaround to allow consistency for the `plinear` algorithm  -->
-<!-- would be to specify the linear parameters when the `plinear` option is specified.  -->
-<!-- For example, we could use `algorithm="plinear(Asym)"` while keeping the general  -->
-<!-- model formula from the default specification. This would allow  -->
-<!-- for the output of different `algorithm` options to be consistent. However, we  -->
-<!-- have not yet tried to code such a change, and welcome discussion from those working -->
-<!-- with other packages using model formulas. -->
-
-<!-- A further complication of model specification in R is that some formulas require -->
-<!-- the user to surround a term with `I()` to inhibit interpretation of arithmetic  -->
-<!-- operators. We would prefer that formulas be usable without this complication. -->
-
 ## Future of nonlinear model estimation in R
@@
 # References
-
-<!-- ------------------------------ -->
-
-<!-- ## Comparison dimensions -->
-
-<!-- The tools for nonlinear estimation we consider have been built with different -->
-<!-- goals and contexts. We will make our comparison considering the overlapping -->
-<!-- headings summarized below. -->
-
-<!-- ### Modeling versus minimizing a sum of squares -->
-
-<!-- The goal of many statisticians and other workers is to estimate a model. The -->
-<!-- minimization of a sum of squared residuals -- the nonlinear least squares problem -- -->
-<!-- is the mechanism to estimate the model. Some tools focus on modeling,  -->
-<!-- so attempt to provide measures of uncertainty and other information.  -->
-<!-- The interface to modeling tools will generally try to streamline that task. By -->
-<!-- contrast, nonlinear least squares tools are trying to solve the more limited -->
-<!-- numerical task. -->
-
-<!-- ### Programming language(s) -->
-
-<!-- R tools are our interest, but they often call programs in other programming -->
-<!-- languages. There are advantages in using a well-established program that has -->
-<!-- been used and tested. Early versions of R were relatively sluggish in  -->
-<!-- computational speed. The downside is that we must interface between two or more -->
-<!-- programming languages, adding to the risk of errors. Changes to one or the  -->
-<!-- other language or its library support may -->
-<!-- force maintenance issues. Moreover, the pool of expertise with languages like C -->
-<!-- and Fortran is diminishing. -->
-
-<!-- ### Internal solver stabilization -->
-
-<!-- Early nonlinear least squares programs generally used the Gauss-Newton method  -->
-<!-- (@Hartley1961). This is often very effective and efficient, but for a quite large  -->
-<!-- proportion of problems can fail to find the solution. Stabilization of the  -->
-<!-- computations is relatively straightforward in principle, and overcomes most -->
-<!-- of the failures, though sometimes at the cost of more calculations. -->
-
-<!-- ### Jacobian calculation -->
-
-<!-- The necessarily iterative solution of a nonlinear least squares problem  -->
-<!-- to carry out nonlinear modeling requires that we can find smaller values -->
-<!-- of a sum of squared residuals. This implies there is a "downhill" direction. -->
-<!-- The gradient of the residual sum of squares is computed as its product -->
-<!-- with the Jacobian matrix: the partial derivatives of the residual vector  -->
-<!-- with respect to the modeling parameters. -->
-
-<!-- Most researchers are challenged to program the differential calculus  -->
-<!-- required to specify the Jacobian to a nonlinear estimation tool.  -->
-<!-- The workaround is generally to use a numerical approximation, -->
-<!-- but there is the possibility of using symbolic or automatic differentiation. -->
-<!-- There are also particular models where workers have provided code for the -->
-<!-- Jacobian. -->
-
-<!-- ### Algorithm and termination choices -->
-
-<!-- Iterative algorithms need to be stopped at some point, and how termination -->
-<!-- is forced may alter results. Furthermore, even supposedly similar algorithms -->
-<!-- may differ greatly in detail.  -->
-
-<!-- ### Constraints on parameters -->
-
-<!-- Reality imposes conditions on parameters. While general constraints can -->
-<!-- be difficult to include, bounds on parameters can be accommodated by many  -->
-<!-- solvers. For some reason, fixed parameters (masks) are not often considered -->
-<!-- but are convenient when a parameter has a commonly accepted value that we -->
-<!-- later wish to allow to be estimated from the data. -->
-
-<!-- ### Special capabilities -->
-
-<!-- Users often have particular problems that introduce other requirements. It is -->
-<!-- common to weight observations where their reliability is thought to differ. -->
-<!-- Some situations present a number of related situations where a combined model -->
-<!-- with some common and some local parameters are present, and the local parameters -->
-<!-- could be indexed. The common exponential decay and related **partially linear** -->
-<!-- models (sometimes called **separable** models) is a difficult one that can  -->
-<!-- benefit from special treatment.  -->
-
-<!-- ### Output  -->
-
-<!-- While nonlinear estimation aims to return a set of parameters for a -->
-<!-- particular model, there is a variety of additional information which may be -->
-<!-- provided. From the perspective of the nonlinear least squares problem, there -->
-<!-- are measures of work and information on the quality of the solution, while -->
-<!-- for statistical purposes measures of uncertainty in the parameters are helpful. -->
-
-<!-- Some tools allow constraints to be imposed on the solution, and it is helpful -->
-<!-- to know if these constraints are active in the proposed solution. -->
-
-<!-- ### Documentation -->
-
-<!-- We believe that the availability of good documentation with examples that users -->
-<!-- can modify to deal with their problems is essential. The presence of vignettes,  -->
-<!-- demos, and tests indicates a package is well-supported. -->

### 2023-05-18 JNABsub20230518/ComparingNLStools.tex
May 18 fixes

--- 2023-05-17 JNABsub20230515/ComparingNLStools.tex
@@
 GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}).
+
+\hypertarget{principal-messages}{%
+\subsubsection{Principal messages}\label{principal-messages}}
+
+Our work was aimed at unifying nonlinear modeling functionality in R,
+ideally in a refactored \texttt{nls()} function.
+
+For R \textbf{users}, we would advise adapting working scripts, preferably those
+with documentation and using recent tools. If there is a suspicion that there
+may be ill-conditioning, package \texttt{nlsr} or the example we give below of how to
+find singular values of the Jacobian allow these diagnostics to be calculated.
+
+For R \textbf{developers} we invite and encourage discussion of the design choices, since these
+have downstream implications for ease of use, adaptation to new features and
+efficiency of ongoing maintenance.
 
@@
 The gradient of \(S(p)\) is
-\[ 2 * J' r\]
+\[ g = 2 * J' r\]
 where the Jacobian \(J\) is given by elements
@@
 If we expand the Hessian for nonlinear least squares problems, we find
-\[  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
+\[  0.5 * H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
 Let us use \(D_{i,j}\) for the elements of the second term of this expression.
@@
 Applying this to a sum of squares problem gives
-\[ H \delta = (J'J + D) \delta = - J' r\]
+\[ 0.5 * H \delta = (J'J + D) \delta = - J' r\]
 In this expression, only the elements of \(D\) have second partial derivatives.
@@
 that these multiplied the second partial derivatives of \(r\), so he approximated
-\[ H \approx J' J\]
+\[ 0.5 * H \approx J' J\]
 by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
@@
 
-As mentioned, the output of \texttt{nls()} is an object of class ``nls'' which has a quite rich structure
-described in the manual file or revealed by applying the \texttt{str()} function to the
-result of \texttt{nls()}.
-
-The complexity of this object is a challenge to users. Let us use the
-\texttt{result} as the returned object from \texttt{nls()}, \texttt{minpack.lm::nlsLM()} or,
-\texttt{gslnls::gsl\_nls()}. For example, the \texttt{data} return element is an R symbol. To
-actually access the data from this element, we need to use the syntax:
+As mentioned, the output of \texttt{nls()}, \texttt{minpack.lm::nlsLM()}, or
+\texttt{gslnls::gsl\_nls()} is an object of class ``nls'' which has a quite rich structure
+described in the manual files or revealed by applying the \texttt{str()} function to the
+result of \texttt{nls()}. The complexity of this object is a challenge to users. Let us
+use for example \texttt{result\ \textless{}-\ snlm1} as the returned object from \texttt{nlsLM()} for the
+Logistic3S problem. . The \texttt{data} return element
+is an R symbol. To actually access the data from this element, we need to use
+the syntax:
 
@@
 probability under certain assumptions.
-Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not compute standard
-errors nor their derived statistics when bounds are active.
+Bounds change those assumptions and hence the interpretation of estimates of
+uncertainty, whether by traditional approximations from the \(J' J\) matrix or from methods
+such as profile likelihood or bootstrap. At the time of writing,
+\texttt{nlsr::nlxb()} does not compute standard errors nor their derived statistics
+when bounds are active to avoid providing misleading information.
 
@@
 The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
-as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}, which we
-have had to extract from the base R code and explicitly \texttt{source()} because it
-is not exported to the working namespace,
+as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
 allows us to compute residuals for particular coefficient sets.
+We have had to extract from the base R code and include it via a code chunk
+(not echoed here for space) because it is not exported to the working namespace,
+We could also explicitly \texttt{source()} this code.
 
@@
 \begin{verbatim}
-source("nlsModel.R")
+## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
 nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
@@
 
-\bibliography{./ImproveNLS.bib}
+\bibliography{./ComparingNLStools.bib}
 

### 2023-05-18 JNABsub20230518/ComparingNLStools.R

@@ -1,0 +1,404 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit ComparingNLStools.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----ex01, echo=TRUE----------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE-------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
+frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
+frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+#
+# Starting parameter sets
+stu1<-c(b1=1, b2=1, b3=1)
+sts1<-c(c1=1, c2=1, c3=1)
+stt1<-c(Asym=1, xmid=1, scal=1)
+
+
+## ----ex02fn, echo=TRUE--------------------------------------------------------
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3]*tt)
+  zz  <-  1.0/(1+x[2]*yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -x[1]*zz*zz*yy
+  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
[…]

### 2023-05-22 JNABsub20230518/ComparingNLStools.Rmd
updates to bib

@@
 $$ exp(xmid/scal)  =  b_2 = 10 * c_2 $$
-$$ 1/scal  =  b_3 $$
+$$ 1/scal  =  b_3 = 0.1*c3 $$
 To allow for simpler discussion, let us say that the parameters form a (named) vector
@@
 unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
+cat("unlx1:")
 print(unlx1) 
@@
 `e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
-%% ^ is picked up in rjtools check
 
@@
 options for the Jacobian, with a further control `ndstep` for the size of the
-step used in the approximation.
+step used in the approximation. These options allow programming choices to be
+examined. Users can largely ignore them.
 
@@
 starting parameters, and can either use `getInitial()` to find them from the `selfStart`
-model or provide explicit values, but the `selfStart` model is used in the formula 
-for the `nlxb()` call.
+model or provide explicit values, but the formula used in the call to `nlxb()`
+still involves the `selfStart` function.
 
@@
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit; otherwise, any such file in 
+suppress saving this workspace (`.RData`) on exit; otherwise any such file in 
 the path will be loaded on startup. R users in our
@@
 `nls()` and other class `nls` tools accept an argument `subset`. This acts through 
-the mediation of `model.frame`, which not obvious in the 
-source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`. 
+the mediation of `model.frame`, which is not obvious in the 
+source code files 
+`/src/library/stats/R/nls.R` and 
+`/src/library/stats/src/nls.C`. 
 Having `subset` at the level of the call to a function like
@@
 allows us to compute residuals for particular coefficient sets.
-We have had to extract from the base R code and include it via a code chunk 
+We have had to extract `nlsModel()` from the base R code and include it via a code chunk 
 (not echoed here for space) because it is not exported to the working namespace,
@@
 rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0
+attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
 nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
 rn<-nmod$resid()
-attr(rn,"gradient")<-NULL; rn
+attr(rn,"gradient")<-NULL; rn # same as rusewt
 ```
@@
 unless `trace` is invoked, the user will not get any information about this.
-This almost trivial change to the `nls()` function could be 
-useful to R users.
+Changing this in the `nls()` function could be useful to R users, and would
+be almost trivial.
 
@@
 The variable projection method (@Golub1973, @OlearyRust13) is 
-generally much more effective than general approaches in finding good solutions 
+usually much more effective than general approaches in finding good solutions 
 to nonlinear least squares 
@@
 We note that there are capabilities in packages for mixed-effects modeling
-such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (p-lme4) for 
+such as `nlme` (@p-nlme), `bbmle` (@p-bbmle), and `lme4` (@p-lme4) for 
 estimating models of such type.
@@
 attempts to succinctly summarize computing environments.
-
-# References

### 2023-05-22 JNABsub20230518/ComparingNLStools.tex
updates to bib

@@
 
-\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
+\texttt{NISTnls} ((\emph{NISTnls: Nonlinear Least Squares Examples from Nist} 2012)) -- This package provides R code and data for a set of
 (numerically ill-conditioned) nonlinear least squares problems from the U.S. National
@@
 
-\texttt{optimx} (John C. Nash and Varadhan (2011)) -- This provides
+\texttt{optimx} (Nash and Varadhan (2011)) -- This provides
 optimizers that can be applied to minimize a nonlinear function which could be
@@
 
-The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
+The Hobbs weed infestation problem (Nash (1979, 120)) is a growth curve modeling
 task that seems straightforward but is quite nasty. Its very succinct statement
@@
 \[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
-\[ 1/scal  =  b_3 \]
+\[ 1/scal  =  b_3 = 0.1*c3 \]
 To allow for simpler discussion, let us say that the parameters form a (named) vector
@@
 a multiple of the diagonal part of \(J' J\). In low precision, some elements of
-\(J' J\) could underflow to zero (John C. Nash (1977)), and a linear combination of both
+\(J' J\) could underflow to zero (Nash (1977)), and a linear combination of both
 choices is an effective compromise. Various choices for
@@
 \subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+
+\begin{verbatim}
+#> unlx1:
+\end{verbatim}
 
@@
 \texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is approximately 2.43e-55.
-\%\% \^{} is picked up in rjtools check
 
@@
 lead to different consequences for performance and how features are
-provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (1979), where storage for
+provided. \texttt{nlsr} has antecedents in the methods of Nash (1979), where storage for
 data and programs was at a ridiculous premium in the small computers of the era.
@@
 options for the Jacobian, with a further control \texttt{ndstep} for the size of the
-step used in the approximation.
+step used in the approximation. These options allow programming choices to be
+examined. Users can largely ignore them.
 
@@
 starting parameters, and can either use \texttt{getInitial()} to find them from the \texttt{selfStart}
-model or provide explicit values, but the \texttt{selfStart} model is used in the formula
-for the \texttt{nlxb()} call.
+model or provide explicit values, but the formula used in the call to \texttt{nlxb()}
+still involves the \texttt{selfStart} function.
 
@@
 their own selfStart model via copy and edit.
-Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (1996)) might be
+Interactive tools, such as ``visual fitting'' (Nash and Velleman (1996)) might be
 worth considering as another way to find starting parameters, but we know of no
@@
 objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
-suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any such file in
+suppress saving this workspace (\texttt{.RData}) on exit; otherwise any such file in
 the path will be loaded on startup. R users in our
@@
 \texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
-the mediation of \texttt{model.frame}, which not obvious in the
-source code files \texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
+the mediation of \texttt{model.frame}, which is not obvious in the
+source code files
+\texttt{/src/library/stats/R/nls.R} and
+\texttt{/src/library/stats/src/nls.C}.
 Having \texttt{subset} at the level of the call to a function like
@@
 allows us to compute residuals for particular coefficient sets.
-We have had to extract from the base R code and include it via a code chunk
+We have had to extract \texttt{nlsModel()} from the base R code and include it via a code chunk
 (not echoed here for space) because it is not exported to the working namespace,
@@
 rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0
+attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
 \end{verbatim}
@@
 rn<-nmod$resid()
-attr(rn,"gradient")<-NULL; rn
+attr(rn,"gradient")<-NULL; rn # same as rusewt
 \end{verbatim}
@@
 unless \texttt{trace} is invoked, the user will not get any information about this.
-This almost trivial change to the \texttt{nls()} function could be
-useful to R users.
+Changing this in the \texttt{nls()} function could be useful to R users, and would
+be almost trivial.
 
@@
 The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is
-generally much more effective than general approaches in finding good solutions
+usually much more effective than general approaches in finding good solutions
 to nonlinear least squares
@@
 We note that there are capabilities in packages for mixed-effects modeling
-such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (p-lme4) for
+such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (Bates et al. (2015)) for
 estimating models of such type.
@@
 \hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{references-1}{%
-\section*{References}\label{references-1}}
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}
@@
 \hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
+\begin{cslreferences}
+\leavevmode\hypertarget{ref-R-onls-manual}{}%
 Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
+\leavevmode\hypertarget{ref-bateswatts}{}%
 Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
 
-\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
-Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
-
-\leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
+\leavevmode\hypertarget{ref-p-lme4}{}%
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker.
+2015.
+``Fitting Linear Mixed-Effects Models Using lme4.'' \emph{Journal of Statistical Software} 67 (1): 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
+
+\leavevmode\hypertarget{ref-BatesWatts81}{}%
+Bates, Douglas M., and Watts, Donald G. 1981. ``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.'' \emph{Technometrics} 23 (2): 179--83.
+
+\leavevmode\hypertarget{ref-p-nlstools}{}%
 Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
----------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
+\leavevmode\hypertarget{ref-nlsMicrobio-manual}{}%
+---------. 2014. \emph{NlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
+
+\leavevmode\hypertarget{ref-p-bbmle}{}%
 Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
+\leavevmode\hypertarget{ref-gslnls23}{}%
 Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
+\leavevmode\hypertarget{ref-nlshelper-manual}{}%
 Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
+\leavevmode\hypertarget{ref-minpacklm12}{}%
 Elzhov, Timur V., Katharine M.
@@
 2012.
-\emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}.
+\emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in Minpack, Plus Support for Bounds}.
 R Project for Statistical Computing.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
-Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
-
-\leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
+\leavevmode\hypertarget{ref-GSL-manual}{}%
+Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for Gsl Version 1.12 (3. Ed.).}
+
+\leavevmode\hypertarget{ref-Bioconductor}{}%
 Gentleman, Robert C., Vincent J.
@@
 2004.
-{``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
+``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.'' \emph{Genome Biology} 5 (R80).
 \url{https://doi.org/10.1186/gb-2004-5-10-r80}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
-Golub, G.
-H., and V.
-Pereyra.
-1973.
-{``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.''} \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
+\leavevmode\hypertarget{ref-Golub1973}{}%
+Golub, G. H., and V. Pereyra. 1973. ``The Differentiation of Pseudo-Inverses and Nonlinear Least Squares Problems Whose Variables Separate.'' \emph{SIAM Journal of Numerical Analysis} 10 (2): 413--32.
+
+\leavevmode\hypertarget{ref-nls2-22}{}%
 Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
-Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
-Huet, S., A.
-Bouvier, M.-A.
-Poursat, and E.
-Jolivet.
-2004.
-\emph{Statistical Tools for Nonlinear Regression: A Practical Guide with {S-PLUS} Examples, 2nd Edition}.
-Berlin \& New York: Springer-Verlag.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
-John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
+\leavevmode\hypertarget{ref-Hartley1961}{}%
+Hartley, H. O. 1961. ``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.'' \emph{Technometrics} 3: 269--80.
+
+\leavevmode\hypertarget{ref-Huet2004}{}%
+Huet, S., A. Bouvier, M.-A. Poursat, and E. Jolivet. 2004. \emph{Statistical Tools for Nonlinear Regression: A Practical Guide with S-PLUS Examples, 2nd Edition}. Berlin \& New York: Springer-Verlag.
+
+\leavevmode\hypertarget{ref-nlsr2023manual}{}%
+John C Nash, and Duncan Murdoch. 2023. \emph{nlsr: Functions for Nonlinear Least Squares Solutions}.
+
+\leavevmode\hypertarget{ref-tvrdik2007}{}%
 Josef Tvrdík and Ivan Křivý and Ladislav Mišík.
 2007.
-{``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.''} \emph{{Computational Statistics \& Data Analysis}} 52 (2): 713--24.
+``Adaptive Population-Based Search: Application to Estimation of Nonlinear Regression Parameters.'' \emph{Computational Statistics \& Data Analysis} 52 (2): 713--24.
 \url{https://doi.org/10.1016/j.csda.2006.10.014}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
-Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
-
-\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
-Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
-
-\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
-Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
-Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
-
-\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
+\leavevmode\hypertarget{ref-Levenberg1944}{}%
+Levenberg, Kenneth. 1944. ``A Method for the Solution of Certain Non-Linear Problems in Least Squares.'' \emph{Quarterly of Applied Mathematics} 2: 164--68.
+
+\leavevmode\hypertarget{ref-Marquardt1963}{}%
+Marquardt, Donald W. 1963. ``An Algorithm for Least-Squares Estimation of Nonlinear Parameters.'' \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
+
+\leavevmode\hypertarget{ref-MiguezNLRAA2021}{}%
+Miguez, Fernando. 2021. \emph{nlraa: Nonlinear Regression for Agricultural Applications}. \url{https://CRAN.R-project.org/package=nlraa}.
+
+\leavevmode\hypertarget{ref-jn77ima}{}%
+Nash, John C. 1977. ``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.'' \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
+
+\leavevmode\hypertarget{ref-cnm79}{}%
 ---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
-\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
----------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
-Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
-Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
+\leavevmode\hypertarget{ref-NashWires22}{}%
+---------. 2022. ``Function minimization and nonlinear least squares in R.'' \emph{WIREs Computational Statistics}, no. e1580. \url{https://doi.org/https://doi.org/10.1002/wics.1580}.
+
+\leavevmode\hypertarget{ref-PkgFromRbase22}{}%
+Nash, John C., and Arkajyoti Bhattacharjee. 2022. ``Making a Package from Base R Files.'' \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
+
+\leavevmode\hypertarget{ref-p-optimx}{}%
+Nash, John C, and Ravi Varadhan. 2011. \emph{\textup{Optimx}: A Replacement and Extension of the optim() Function}. Nash Information Services Inc.; Johns Hopkins University.
+
+\leavevmode\hypertarget{ref-nash1996nonlinear}{}%
 Nash, John C., and Paul Velleman.
 1996.
-{``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61.
+``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.'' In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61.
 American Statistical Association.
 
-\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
-National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.
-2012.
-\emph{NISTnls: Nonlinear Least Squares Examples from NIST}.
-\url{https://CRAN.R-project.org/package=NISTnls}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
-O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
-
-\leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
-Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
+\leavevmode\hypertarget{ref-NISTnls-manual}{}%
+\emph{NISTnls: Nonlinear Least Squares Examples from Nist}. 2012. \url{https://CRAN.R-project.org/package=NISTnls}.
+
+\leavevmode\hypertarget{ref-OlearyRust13}{}%
+O'Leary, Dianne P., and Bert W. Rust. 2013. ``Variable Projection for Nonlinear Least Squares Problems.'' \emph{Computational Optimization and Applications} 54 (3): 579--93.
+
+\leavevmode\hypertarget{ref-nlsmultstart}{}%
+Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using Aic Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
+
+\leavevmode\hypertarget{ref-p-nlme}{}%
 Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
+\leavevmode\hypertarget{ref-m-nlsmsn}{}%
 Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
+\leavevmode\hypertarget{ref-Ratkowsky1983}{}%
 Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
-\leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
+\leavevmode\hypertarget{ref-m-nlstac}{}%
 Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez.
 2020.
-\emph{Nlstac: An r Package for Fitting Separable Nonlinear Models}.
+\emph{Nlstac: An R Package for Fitting Separable Nonlinear Models}.
 \url{https://CRAN.R-project.org/package=nlstac}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%
+\leavevmode\hypertarget{ref-nlsic-manual}{}%
 Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Torvisco2018}{}}%
+\leavevmode\hypertarget{ref-Torvisco2018}{}%
 Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez.
 2018.
-{``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Niˇs, Serbia}.
+``A New Algorithm to Fit Exponential Decays Without Initial Guess.'' Faculty of Sciences and Mathematics, University of Niˇs, Serbia.
 \url{https://doi.org/doi.org/10.2298/FIL1812233T}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
-Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
-Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
-Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.
-
-\end{CSLReferences}
+\leavevmode\hypertarget{ref-R-crsnls-manual}{}%
+Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'Crs4hc' and 'Crs4hce'}. \url{https://CRAN.R-project.org/package=crsnls}.
+
+\leavevmode\hypertarget{ref-HWtestthat11}{}%
+Wickham, Hadley. 2011. ``testthat: Get Started with Testing.'' \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
+
+\leavevmode\hypertarget{ref-HWdevtools21}{}%
+Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{devtools: Tools to Make Developing R Packages Easier}. \url{https://CRAN.R-project.org/package=devtools}.
+\end{cslreferences}
 

### 2023-05-22 JNABsub20230518/ComparingNLStools.R

@@ -62,6 +62,7 @@ tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
 
 ## ----ex02nlsr-----------------------------------------------------------------
 unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
+cat("unlx1:")
 print(unlx1) 
 snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
 pshort(snlx1) # a short-form output
@@ -356,10 +357,10 @@ attr(rusewt,"gradient")<-NULL; rusewt
 ## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
 nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
 rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0
+attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
 nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
 rn<-nmod$resid()
-attr(rn,"gradient")<-NULL; rn
+attr(rn,"gradient")<-NULL; rn # same as rusewt
 
 
 ## ----tetrarun-----------------------------------------------------------------

### 2023-07-01 RefactoringNLS/revision202307/ReviewResponses.txt
start revisions

Review 1

Referee report of “A Comparison of R Tools for Nonlinear Least
Squares Modeling”.
ID: 2023-15
General
This is a long paper that covers many topics related to nonlinear least squares fitting with R. This
mixture of many topics can be confusing and I think the paper would benefit from a clearer definition
of its purpose and focus. Once a clearer focus is established it will be easier to decide which topics are
relevant to that purpose.
The paper’s title indicates that the purpose is to compare and contrast facilities in base R, meaning
the nls function and related facilities in the stats package, and several external packages for nonlinear
least-squares modeling that are available on CRAN. There is a listing of packages related to nonlinear
least squares on page 2 but the majority of the discussion is focussed on nls, in the stats package that
is part of base R, and the nlsr package (Nash and Murdoch 2023), written by this paper’s authors and
others (on CRAN this package is said to be the work of four authors but only two of them are cited in
the paper).
There are also indications that the paper is about replacing nls in base R with functions from nlsr. The
abstract begins by citing the Google Summer of Code project “Improvements to nls()” and the first
sentence in the “Principal messages”” section indicates that the purpose is to create a refactored nls()
based on the work described here. Toward the end of the paper, the section on “Future of nonlinear
model estimation in R” also indicates that the goal is to enhance or replace nls() and related parts
of base R.
So is the purpose of the paper:
• To introduce the nlsr package and describe its capabilities?
• To compare and contrast several available packages for nonlinear least-squares?
• To advocate that the authors’ code replace nls in base R?
I would recommend concentrating on the first objective — introducing the nlsr package and demonstrating
the advantages that it provides relative to nls — and I will address my comments to that purpose. I
hope this approach will be of use to the authors and the editor.
It is certainly appropriate when describing the nlsr package to contrast it with facilities like nls() in
base R. To help with this I will provide some background on the development of nls() and the stats
package in general.
Detailed comments
nls is S and in R
Although nls() has been part of R for over 25 years, it originated much earlier in the “Statistical Models
in S” project, documented in Chambers and Hastie (1992), and based on work starting in the mid-1980’s.
The purpose was to create a unified set of tools for fitting statistical models, including nonlinear models
with a scalar objective, such as a log-likelihood, or as a nonlinear regression model. Many capabilities for
R, including S3 classes and methods, the formula language for linear predictors, and model frames and
model matrices, were based directly on the results of this project in S. (As described by Ross Ihaka, one
of the two original developers of R, their goal was to produce a language that was “not unlike S” so the
resemblance was not accidental.)
1What was called QPE (quantitative programming environment), then “The New S Language”, then “S3”
represented a huge step forward for nonlinear modeling because it could represent model functions as
expressions (i.e. abstract syntax trees) that could be manipulated in the system itself. This allowed for
symbolic differentiation.
The original implementation of the deriv function for symbolic differentiation was written in S. When
porting deriv to R in the mid-1990’s Ross Ihaka moved quite a bit of the functionality to C code because
it was comparatively easy to do so and, on the hardware of the day, the C implementation provided a
considerable speed advantage. The speed of the differentiation itself isn’t usually a problem - it’s the
common-sub-expression elimination that can be the bottleneck.
This legacy is also the reason that the Jacobian matrix is the "gradient" attribute of the returned value
of the model function. For a nonlinear least squares problem, where the model function returns a vector,
this deriv attribute should be called "Jacobian" but the version for the scalar objective, for which the
vector of partial derivatives is the gradient, was written first. Because deriv only has access to the
expression and not to the model frame it is not possible to determine if the value will be a scalar or a
vector and the name of the attribute was left as "gradient". A further complication is that numerical
values in S and R are always stored as vectors, with the understanding that a scalar is represented as
a vector of length 1. Even if it were desirable to distinguish between scalars and vectors, internally it
is impossible to determine if a vector of length 1 is intended to be a scalar or to be a vector that just
happens to have length 1.
The legacy is also the reason that initial data manipulation, such as evaluating a subset or na.action, is
performed by non-standard evaluation of a model.frame. The non-standard evaluation means that, within
the nls function evaluation, the code accesses its own call (with match.call) and re-writes the function
name to be stats::model.frame then evaluates the rewritten call (around line 580 of nls.R). Handling
all the special cases is complicated but the purpose is so that arguments like subset or na.action and
conversion of formulas to model frames to model matrices are handled consistently across the modeling
functions in the stats package and without code duplication.
Thomas Lumley has called this “replace the function name in my own call” approach the “standard,
non-standard evaluation idiom” in R.
Another aspect of this legacy is the expected behavior of print, summary, plot, etc. applied to an object
representing a fitted model. Peculiarities like summary producing longer descriptions than print or the
choice of printing the sum of squared residuals or the estimated standard deviation of the residuals are
for consistency with the other model output.
It is also why the extractor for the parameter values is coef, even though the parameters in a nonlin-
ear regression model are not necessarily coefficients, and why they are shown in the summary using
print.coefmat.
These choices, which allow users to transfer experience from one modelling situation to another, are
neither arbitrary nor trivial.
Function closure with a shared environment
One of the major differences between the S and R implementations of nls is the use of function closures
in R to allow several functions in the m field of an nls model (see below for an example) shared access to
parameter values, covariate values, the response vector and various pieces of auxillary information.
In R if a function is defined during the evaluation of another function its “closure” includes the evaluation
environment of its parent. If several functions are defined and returned, say in a list, then they have
access to this common environment. This shared environment is where parameter updates and subsequent
evaluation of the model function, the Jacobian and various decompositions occur.
Self-starting models in S and R
Self-starting models were developed for nlme package for S, Pinheiro and Bates (2000), to automate
the fitting of nonlinear regression models to repeated measures data collected on several observational
2units, e.g. several different subjects. Rather than requiring the analyst to provide starting values for each
subject’s data, the steps of formulating and refining such starting values were encapsulated in functions.
These functions also used symbolic derivatives to provide Jacobian evaluations and frequently fit a reduced
form of the model using the plinear algorithm to refine the starting values, resulting in the “starting
values” being the actual parameter estimates. These are fed into the default algorithm to produce a
summary but with the understanding that the default algorithm will converge at the first evaluation
(i.e. 0 iterative steps) if the convergence criterion is the relative offset. The point here is that the relative
offset criterion depends only on the current parameter values and not on changes in the parameter values
so there is only one additional evaluation required after the plinear algorithm converges.
For the illustrative example starting at the bottom of page 2 of this paper the SSlogis self-starting
model can be used.
weeddf <- data.frame(
tt = 1:12,
weed = c(
5.308, 7.24, 9.638, 12.866, 17.069, 23.192,
31.443, 38.558, 50.156, 62.948, 75.995, 91.972
)
)
summary(hobbsm1 <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf))
Formula: weed ~ SSlogis(tt, Asym, xmid, scal)
Parameters:
Estimate Std. Error t value Pr(>|t|)
Asym 196.1862
11.3069
17.35 3.17e-08 ***
xmid 12.4173
0.3346
37.11 3.72e-11 ***
scal
3.1891
0.0698
45.69 5.77e-12 ***
---
Signif. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Residual standard error: 0.5362 on 9 degrees of freedom
Number of iterations to convergence: 0
Achieved convergence tolerance: 1.118e-06
str(hobbsm1)
List of 6
$ m
:List of 16
..$ resid
:function ()
..$ fitted
:function ()
..$ formula
:function ()
..$ deviance :function ()
..$ lhs
:function ()
..$ gradient :function ()
..$ conv
:function ()
..$ incr
:function ()
..$ setVarying:function (vary = rep_len(TRUE, np))
..$ setPars
:function (newPars)
..$ getPars
:function ()
..$ getAllPars:function ()
..$ getEnv
:function ()
3..$ trace
:function ()
..$ Rmat
:function ()
..$ predict
:function (newdata = list(), qr = FALSE)
..- attr(*, "class")= chr "nlsModel"
$ convInfo
:List of 5
..$ isConv
: logi TRUE
..$ finIter
: int 0
..$ finTol
: num 1.12e-06
..$ stopCode
: int 0
..$ stopMessage: chr "converged"
$ data
: symbol weeddf
$ call
: language nls(formula = weed ~ SSlogis(tt, Asym, xmid, scal), data = weeddf, algorithm
$ dataClasses: Named chr "numeric"
..- attr(*, "names")= chr "tt"
$ control
:List of 7
..$ maxiter
: num 50
..$ tol
: num 1e-05
..$ minFactor : num 0.000977
..$ printEval : logi FALSE
..$ warnOnly
: logi FALSE
..$ scaleOffset: num 0
..$ nDcentral : logi FALSE
- attr(*, "class")= chr "nls"
This only takes a few milliseconds on a laptop computer
microbenchmark::microbenchmark(nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf), times=100L)
Unit: milliseconds
expr
min
lq
mean
nls(weed ~ SSlogis(tt, Asym, xmid, scal), weeddf) 3.483984 3.624459 4.599596
median
uq
max neval
3.92275 5.035838 13.76178
100
although it turns out that the SSlogisJN formulation is slightly faster in this case
summary(hobbsm2 <- nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf))
Formula: weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal)
Parameters:
Estimate Std. Error t value Pr(>|t|)
Asym 196.1862
11.3069
17.35 3.17e-08 ***
xmid 12.4173
0.3346
37.11 3.72e-11 ***
scal
3.1891
0.0698
45.69 5.77e-12 ***
---
Signif. codes: 0 '***' 0.001 '**' 0.01 '*' 0.05 '.' 0.1 ' ' 1
Residual standard error: 0.5362 on 9 degrees of freedom
Number of iterations to convergence: 3
Achieved convergence tolerance: 2.419e-06
microbenchmark::microbenchmark(nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf), times=10
4Unit: milliseconds
expr
min
lq
nls(weed ~ nlsr::SSlogisJN(tt, Asym, xmid, scal), weeddf) 1.960326 2.069342
mean
median
uq
max neval
2.492946 2.184441 2.554391 11.43154
100
Some of the discussion in the “Jacobian code in selfStart models” section (p. 9) of this paper claims
that the plinear fit followed by a fit using the default approach would be difficult to maintain and could
be very slow but that has not been the case. It is known (and, in fact, stated in the documentation for
SSlogis) that the call to the default nls algorithm will always declare convergence immediately.
Also in that section it is claimed that deriv creates “hidden variables” because the default for the tag
argument to deriv is .expr. This tag is only used in bindings within the evaluation environment of the
model function, which is never visible to the user unless they are inspecting the function evaluation with
a symbolic debugger. The only reason for the default tag starting with "." is to avoid conflicts with
parameter or covariate names in the model expression. But it is only the default.
Recommendations
As stated in the “General” section, I think the purpose of the paper should be clarified. If the purpose
is to introduce the nlsr package and contrast it with the existing implementation of nls in the stats
package, then it might be helpful to understand why nls was implemented the way it was.
I hope that this unconventional referee’s report is of use to the authors and the editor.
References
Chambers, John M., and Trevor J. Hastie, eds. 1992. Statistical Models in S. Routledge. https:
//doi.org/10.1201/9780203738535.
Nash, John C, and Duncan Murdoch. 2023. Nlsr: Functions for Nonlinear Least Squares Solutions -
Updated 2022. https://CRAN.R-project.org/package=nlsr.
Pinheiro, José, and Douglas Bates. 2000. Mixed-Effects Models in S and S-PLUS. Springer New York.
5
\

Review 2

Referee report: A Comparison of R Tools for Nonlinear Least
Squares Modeling
Id: 2023-15
2023-06-28
General

This article provides a vignette-style overview of the nlsr-package and highlights its design choices and
behavior in comparison to base R’s nls() function as well as the nonlinear least squares solvers provided by
the R-packages minpack.lm and gslnls. The article’s main contribution is to make R users (more) aware of
the historic shortcomings and pitfalls in using R’s nls() function and how the nlsr-package is able to serve
as a more reliable alternative. Overall, the article is well-written and structured into coherent subthemes
and sections. Given that nls() is the de facto standard for solving nonlinear least squares problems in R, I
believe this article provides valuable insights and is well-suited for publication in the R Journal after some
revisions and minor clarifications.

My main advice is to revise the title and introduction of the article. As it currently stands, it is suggested
that the article provides an (objective) comparison of nonlinear least squares solvers available in R. The body
of the article reads more like a vignette for the nlsr-package, explaining its design choices and comparing its
behavior to nls() and other NLS solvers based on a few selected examples. As such, a more appropriate title
could be, for instance, The current state of R tools for nonlinear least squares modeling or similar.

2:01

We appreciate the positive comments and agree with the suggested title change.
Both referees have requested revisions to the introduction, and we hope that we have addressed the
comments.

Detailed comments
1. The Hobbs weed example as an illustration of the performance of different NLS solvers is lacking a
thorough comparison. I understand that the aim of the authors is to provide a data example that
demonstrates the reliability of nlsr in favor of nls(), but nls() converges to the same solution as
nlsr for all model setups when using the Port algorithm. This goes unnoticed in the text. Also,
the model setup (Logistic3T) for which gsl_nls() fails can easily be made to converge correctly by
changing either the algorithm or the scaling method, see the Illustrating code section. My advice
is to provide a more comprehensive comparison of the available NLS solvers, or to pick a different
data example that better highlights the benefits of nlsr. For instance, trying to solve the BoxBOD
regression problem (https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), nls() fails with
starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton and Port algorithm, whereas nlsr::nlxb()
converges correctly.
2. The section Returned results of nls() and other tools discusses the complexity of S3-objects of class
nls, illustrated by the retrieval of the input data from an nls-object. It seems ill-advised to run
eval(parse(text=result$data)), since it evaluates the name of the data object in the user’s envi-
ronment, which may not exist. Instead, one should use model = TRUE to store the input data in the
returned nls-object. This behavior is consistent with lm(), so perhaps less confusing than currently
presented in the manuscript. The authors mention that nlxb() returns a much simpler structure of 11
items in one level, but I did not manage to find the input data object in the nlsr-object returned by
nlxb().
3. The section Functional specification of problems mentions the use of nlsr::nlfb() to solve nonlinear
least squares problems using a function to define the residuals. Since nlsr::nlfb() only has a notion
of the residuals, how is it able to disentangle model and response, for the purpose of e.g. generating
1model predictions? The authors mention that the function wrapnlsr() can be used to return an object
of class nls, but this does not seem to work for problems defined through a residual function as in
nlsr::nlfb()? If indeed the case, how does one evaluate e.g. model predictions, parameter confidence
intervals, etc. when using nlsr::nlfb()? This remains unclear to me.
4. The section Philosophical considerations hints at the challenge of performing inference when parameter
bound-constraints are involved in the optimization. In case of physical or model constraints on the
parameters, a common approach is to reparametrize the model to be unconstrained in the parameters.
Perhaps the authors can highlight an example scenario where imposing parameter constraints is
preferable to model reparametrization (the included data examples with parameter constraints appear
somewhat artificial), or elaborate further on the possible downsides of reparametrizing the model, if any.
5. The section Programming language makes the argument that keeping to a single programming language
can allow for easier maintenance and upgrades, and that the performance penalty for using code entirely
in R has been much reduced in recent years. I agree that there is not much of a performance penalty
when solving small problems, such as the data examples included in the article. However, when solving
problems using a large amount of data or a large number of parameters, I have experienced that the
run-time (as well as memory usage) can become an actual bottleneck. As an example, consider the
Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) included in the Illustrating code
section.
Minor comments
• Avoid emphasizing words by writing in all capitals. Examples: NOT (pg. 14, 15), WEIGHTED (pg.
13) and more.
• The link https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211 (as well as the links to github)
may be subject to change or become unavailable in the future.
Illustrating code
## Hobbs weed example
weeddf <- data.frame(
tt = 1:12,
weed = c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192,
31.443, 38.558, 50.156, 62.948, 75.995, 91.972)
)
### nls-port all setups
nls(
formula = weed ~ b1 / (1 + b2 * exp(-b3 * tt)),
data = weeddf,
start = c(b1 = 1, b2 = 1, b3 = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ b1/(1 + b2 * exp(-b3 * tt))
#>
data: weeddf
#>
b1
b2
b3
#> 196.1863 49.0916
0.3136
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
nls(
formula = weed ~ 100 * c1 / (1 + 10 * c2 * exp(-0.1 * c3 * tt)),
2data = weeddf,
start = c(c1 = 1, c2 = 1, c3 = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ 100 * c1/(1 + 10 * c2 * exp(-0.1 * c3 * tt))
#>
data: weeddf
#>
c1
c2
c3
#> 1.962 4.909 3.136
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
nls(
formula = weed ~ Asym /(1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "port"
)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm "port", convergence message: relative convergence (4)
### gsl_nls Logistic3T setup
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "lmaccel"
)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm: multifit/levenberg-marquardt+accel, (scaling: more, solver: qr)
#>
#> Number of iterations to convergence: 22
#> Achieved convergence tolerance: 1.18e-11
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "dogleg"
)
3#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm: multifit/dogleg, (scaling: more, solver: qr)
#>
#> Number of iterations to convergence: 16
#> Achieved convergence tolerance: 1.599e-13
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "ddogleg"
)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm: multifit/double-dogleg, (scaling: more, solver: qr)
#>
#> Number of iterations to convergence: 16
#> Achieved convergence tolerance: 1.14e-11
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
algorithm = "subspace2D"
)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm: multifit/2D-subspace, (scaling: more, solver: qr)
#>
#> Number of iterations to convergence: 19
#> Achieved convergence tolerance: 9.783e-13
gslnls::gsl_nls(
fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
data = weeddf,
start = c(Asym = 1, xmid = 1, scal = 1),
control = list(scale = "levenberg")
4)
#> Nonlinear regression model
#>
model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
#>
data: weeddf
#>
Asym
xmid
scal
#> 196.186 12.417
3.189
#> residual sum-of-squares: 2.587
#>
#> Algorithm: multifit/levenberg-marquardt, (scaling: levenberg, solver: qr)
#>
#> Number of iterations to convergence: 24
#> Achieved convergence tolerance: 7.065e-13
## NOT RUN
## Penalty function I [More, Garbow, Hillstrom 1981]
alpha <- 1e-5
n <- 1000
### residuals
fres <- function(x) {
c(sqrt(alpha) * (x - 1), sum(xˆ2) - 0.25)
}
### jacobian
fjac <- function(x) {
jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
attr(jj, "gradient") <- jj
jj
}
bench::mark(
"large problem" = nlsr::nlfb(start = 1:n, resfn = fres, jacfn = fjac),
iterations = 1
)
# # A tibble: 1 × 13
#
expression
min
median `itr/sec` mem_alloc `gc/sec`
#
<bch:expr>
<bch:tm> <bch:tm>
<dbl> <bch:byt>
<dbl>
# 1 large problem
3.24m
3.24m
0.00515
9.23GB
0.226
References
Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. “Testing Unconstrained Optimization
Software.” ACM Transactions on Mathematical Software 7: 17–41.
5

### 2023-07-03 RefactoringNLS/revision202307/ReviewResponses.txt
Responses -- first draft

@@
 General
-This is a long paper that covers many topics related to nonlinear least squares fitting with R. This
-mixture of many topics can be confusing and I think the paper would benefit from a clearer definition
-of its purpose and focus. Once a clearer focus is established it will be easier to decide which topics are
-relevant to that purpose.
-The paper’s title indicates that the purpose is to compare and contrast facilities in base R, meaning
-the nls function and related facilities in the stats package, and several external packages for nonlinear
-least-squares modeling that are available on CRAN. There is a listing of packages related to nonlinear
-least squares on page 2 but the majority of the discussion is focussed on nls, in the stats package that
-is part of base R, and the nlsr package (Nash and Murdoch 2023), written by this paper’s authors and
-others (on CRAN this package is said to be the work of four authors but only two of them are cited in
-the paper).
-There are also indications that the paper is about replacing nls in base R with functions from nlsr. The
-abstract begins by citing the Google Summer of Code project “Improvements to nls()” and the first
-sentence in the “Principal messages”” section indicates that the purpose is to create a refactored nls()
-based on the work described here. Toward the end of the paper, the section on “Future of nonlinear
-model estimation in R” also indicates that the goal is to enhance or replace nls() and related parts
+This is a long paper that covers many topics related to nonlinear least 
+squares fitting with R. This mixture of many topics can be confusing and 
+I think the paper would benefit from a clearer definition of its purpose 
+and focus. Once a clearer focus is established it will be easier to decide 
+which topics are relevant to that purpose.
+
+The paper’s title indicates that the purpose is to compare and contrast 
+facilities in base R, meaning the nls function and related facilities 
+in the stats package, and several external packages for nonlinear
+least-squares modeling that are available on CRAN. There is a listing 
+of packages related to nonlinear least squares on page 2 but the majority 
+of the discussion is focussed on nls, in the stats package that is part 
+of base R, and the nlsr package (Nash and Murdoch 2023), written by this 
+paper’s authors and others (on CRAN this package is said to be the work 
+of four authors but only two of them are cited in the paper).
+
+Response 1:01
+
+Fernando Miguez and Arkajyoti Bhattacharjee, who are listed as contributors,
+provided comments that were very helpful in revising package nlsr, but Nash 
+and Murdoch provided all the code to our knowledge.
+
+There are 
+also indications that the paper is about replacing nls in base R with 
+functions from nlsr. The abstract begins by citing the Google Summer 
+of Code project “Improvements to nls()” and the first sentence in the 
+“Principal messages”” section indicates that the purpose is to create 
+a refactored nls() based on the work described here. Toward the end of 
+the paper, the section on “Future of nonlinear model estimation in R” 
+also indicates that the goal is to enhance or replace nls() and related parts
 of base R.
+
 So is the purpose of the paper:
@@
 • To advocate that the authors’ code replace nls in base R?
-I would recommend concentrating on the first objective — introducing the nlsr package and demonstrating
-the advantages that it provides relative to nls — and I will address my comments to that purpose. I
+
+I would recommend concentrating on the first objective — introducing 
+the nlsr package and demonstrating the advantages that it provides 
+relative to nls — and I will address my comments to that purpose. I
 hope this approach will be of use to the authors and the editor.
-It is certainly appropriate when describing the nlsr package to contrast it with facilities like nls() in
-base R. To help with this I will provide some background on the development of nls() and the stats
+It is certainly appropriate when describing the nlsr package to 
+contrast it with facilities like nls() in base R. To help with this 
+I will provide some background on the development of nls() and the stats
 package in general.
+
+Response 1:02
+     This overlaps with comments by the second referee. See Response 2:01.
+
 Detailed comments
+
 nls is S and in R
-Although nls() has been part of R for over 25 years, it originated much earlier in the “Statistical Models
-in S” project, documented in Chambers and Hastie (1992), and based on work starting in the mid-1980’s.
-The purpose was to create a unified set of tools for fitting statistical models, including nonlinear models
-with a scalar objective, such as a log-likelihood, or as a nonlinear regression model. Many capabilities for
-R, including S3 classes and methods, the formula language for linear predictors, and model frames and
-model matrices, were based directly on the results of this project in S. (As described by Ross Ihaka, one
-of the two original developers of R, their goal was to produce a language that was “not unlike S” so the
-resemblance was not accidental.)
-1What was called QPE (quantitative programming environment), then “The New S Language”, then “S3”
-represented a huge step forward for nonlinear modeling because it could represent model functions as
-expressions (i.e. abstract syntax trees) that could be manipulated in the system itself. This allowed for
-symbolic differentiation.
-The original implementation of the deriv function for symbolic differentiation was written in S. When
-porting deriv to R in the mid-1990’s Ross Ihaka moved quite a bit of the functionality to C code because
-it was comparatively easy to do so and, on the hardware of the day, the C implementation provided a
-considerable speed advantage. The speed of the differentiation itself isn’t usually a problem - it’s the
+
+Although nls() has been part of R for over 25 years, it originated 
+much earlier in the “Statistical Models in S” project, documented 
+in Chambers and Hastie (1992), and based on work starting in the mid-1980’s.
+The purpose was to create a unified set of tools for fitting statistical 
+models, including nonlinear models with a scalar objective, such as a 
+log-likelihood, or as a nonlinear regression model. Many capabilities for
+R, including S3 classes and methods, the formula language for linear 
+predictors, and model frames and model matrices, were based directly on 
+the results of this project in S. (As described by Ross Ihaka, one of the 
+two original developers of R, their goal was to produce a language that 
+was “not unlike S” so the resemblance was not accidental.)
+
+What was called QPE (quantitative programming environment), then “The New 
+S Language”, then “S3” represented a huge step forward for nonlinear 
+modeling because it could represent model functions as expressions 
+(i.e. abstract syntax trees) that could be manipulated in the system 
+itself. This allowed for symbolic differentiation.
+
+The original implementation of the deriv function for symbolic 
+differentiation was written in S. When porting deriv to R in the 
+mid-1990’s Ross Ihaka moved quite a bit of the functionality to C code 
+because it was comparatively easy to do so and, on the hardware of the 
+day, the C implementation provided a considerable speed advantage. The 
+speed of the differentiation itself isn’t usually a problem - it’s the
 common-sub-expression elimination that can be the bottleneck.
-This legacy is also the reason that the Jacobian matrix is the "gradient" attribute of the returned value
-of the model function. For a nonlinear least squares problem, where the model function returns a vector,
-this deriv attribute should be called "Jacobian" but the version for the scalar objective, for which the
-vector of partial derivatives is the gradient, was written first. Because deriv only has access to the
-expression and not to the model frame it is not possible to determine if the value will be a scalar or a
-vector and the name of the attribute was left as "gradient". A further complication is that numerical
-values in S and R are always stored as vectors, with the understanding that a scalar is represented as
-a vector of length 1. Even if it were desirable to distinguish between scalars and vectors, internally it
-is impossible to determine if a vector of length 1 is intended to be a scalar or to be a vector that just
-happens to have length 1.
-The legacy is also the reason that initial data manipulation, such as evaluating a subset or na.action, is
-performed by non-standard evaluation of a model.frame. The non-standard evaluation means that, within
-the nls function evaluation, the code accesses its own call (with match.call) and re-writes the function
-name to be stats::model.frame then evaluates the rewritten call (around line 580 of nls.R). Handling
-all the special cases is complicated but the purpose is so that arguments like subset or na.action and
-conversion of formulas to model frames to model matrices are handled consistently across the modeling
-functions in the stats package and without code duplication.
-Thomas Lumley has called this “replace the function name in my own call” approach the “standard,
-non-standard evaluation idiom” in R.
-Another aspect of this legacy is the expected behavior of print, summary, plot, etc. applied to an object
-representing a fitted model. Peculiarities like summary producing longer descriptions than print or the
-choice of printing the sum of squared residuals or the estimated standard deviation of the residuals are
-for consistency with the other model output.
-It is also why the extractor for the parameter values is coef, even though the parameters in a nonlin-
-ear regression model are not necessarily coefficients, and why they are shown in the summary using
-print.coefmat.
-These choices, which allow users to transfer experience from one modelling situation to another, are
-neither arbitrary nor trivial.
+
+This legacy is also the reason that the Jacobian matrix is the "gradient" 
+attribute of the returned value of the model function. For a nonlinear 
+least squares problem, where the model function returns a vector, this 
+deriv attribute should be called "Jacobian" but the version for the 
+scalar objective, for which the vector of partial derivatives is the 
+gradient, was written first. Because deriv only has access to the 
+expression and not to the model frame it is not possible to determine 
+if the value will be a scalar or a vector and the name of the attribute 
+was left as "gradient". A further complication is that numerical values 
+in S and R are always stored as vectors, with the understanding that a 
+scalar is represented as a vector of length 1. Even if it were desirable 
+to distinguish between scalars and vectors, internally it is impossible 
+to determine if a vector of length 1 is intended to be a scalar or to 
+be a vector that just happens to have length 1.
+
+The legacy is also the reason that initial data manipulation, such as 
+evaluating a subset or na.action, is performed by non-standard evaluation 
+of a model.frame. The non-standard evaluation means that, within the 
+nls function evaluation, the code accesses its own call (with match.call) 
+and re-writes the function name to be stats::model.frame then evaluates 
+the rewritten call (around line 580 of nls.R). Handling all the special 
+cases is complicated but the purpose is so that arguments like subset 
+or na.action and conversion of formulas to model frames to model matrices 
+are handled consistently across the modeling functions in the stats 
+package and without code duplication.
+
+Thomas Lumley has called this “replace the function name in my own call” 
+approach the “standard, non-standard evaluation idiom” in R.
+
+Another aspect of this legacy is the expected behavior of print, summary, 
+plot, etc. applied to an object representing a fitted model. Peculiarities 
+like summary producing longer descriptions than print or the choice of 
+printing the sum of squared residuals or the estimated standard deviation 
+of the residuals are for consistency with the other model output.
+
+Response 1:06 We now mention the background to summary/print.
+
+It is also why the extractor for the parameter values is coef, even though 
+the parameters in a nonlinear regression model are not necessarily 
+coefficients, and why they are shown in the summary using print.coefmat.
+
+These choices, which allow users to transfer experience from one modelling 
+situation to another, are neither arbitrary nor trivial.
+
 Function closure with a shared environment
-One of the major differences between the S and R implementations of nls is the use of function closures
-in R to allow several functions in the m field of an nls model (see below for an example) shared access to
-parameter values, covariate values, the response vector and various pieces of auxillary information.
-In R if a function is defined during the evaluation of another function its “closure” includes the evaluation
-environment of its parent. If several functions are defined and returned, say in a list, then they have
-access to this common environment. This shared environment is where parameter updates and subsequent
-evaluation of the model function, the Jacobian and various decompositions occur.
+
+One of the major differences between the S and R implementations of nls is 
+the use of function closures in R to allow several functions in the m field 
+of an nls model (see below for an example) shared access to parameter values, 
+covariate values, the response vector and various pieces of auxillary 
+information.
+
+In R if a function is defined during the evaluation of another function its 
+“closure” includes the evaluation environment of its parent. If several 
+functions are defined and returned, say in a list, then they have access 
+to this common environment. This shared environment is where parameter 
+updates and subsequent evaluation of the model function, the Jacobian and 
+various decompositions occur.
+
 Self-starting models in S and R
-Self-starting models were developed for nlme package for S, Pinheiro and Bates (2000), to automate
-the fitting of nonlinear regression models to repeated measures data collected on several observational
-2units, e.g. several different subjects. Rather than requiring the analyst to provide starting values for each
-subject’s data, the steps of formulating and refining such starting values were encapsulated in functions.
-These functions also used symbolic derivatives to provide Jacobian evaluations and frequently fit a reduced
-form of the model using the plinear algorithm to refine the starting values, resulting in the “starting
-values” being the actual parameter estimates. These are fed into the default algorithm to produce a
-summary but with the understanding that the default algorithm will converge at the first evaluation
-(i.e. 0 iterative steps) if the convergence criterion is the relative offset. The point here is that the relative
-offset criterion depends only on the current parameter values and not on changes in the parameter values
-so there is only one additional evaluation required after the plinear algorithm converges.
-For the illustrative example starting at the bottom of page 2 of this paper the SSlogis self-starting
-model can be used.
+
+Self-starting models were developed for nlme package for S, Pinheiro and 
+Bates (2000), to automate the fitting of nonlinear regression models to 
+repeated measures data collected on several observational units, e.g. 
+several different subjects. Rather than requiring the analyst to provide 
+starting values for each subject’s data, the steps of formulating and 
+refining such starting values were encapsulated in functions. These 
+functions also used symbolic derivatives to provide Jacobian evaluations 
+and frequently fit a reduced form of the model using the plinear algorithm 
+to refine the starting values, resulting in the “starting values” being 
+the actual parameter estimates. These are fed into the default algorithm 
+to produce a summary but with the understanding that the default algorithm 
+will converge at the first evaluation (i.e. 0 iterative steps) if the 
+convergence criterion is the relative offset. The point here is that the 
+relative offset criterion depends only on the current parameter values 
+and not on changes in the parameter values so there is only one additional 
+evaluation required after the plinear algorithm converges.
+
+For the illustrative example starting at the bottom of page 2 of this 
+paper the SSlogis self-starting model can be used.
+
 weeddf <- data.frame(
@@
 100
-Some of the discussion in the “Jacobian code in selfStart models” section (p. 9) of this paper claims
-that the plinear fit followed by a fit using the default approach would be difficult to maintain and could
-be very slow but that has not been the case. It is known (and, in fact, stated in the documentation for
-SSlogis) that the call to the default nls algorithm will always declare convergence immediately.
-Also in that section it is claimed that deriv creates “hidden variables” because the default for the tag
-argument to deriv is .expr. This tag is only used in bindings within the evaluation environment of the
-model function, which is never visible to the user unless they are inspecting the function evaluation with
-a symbolic debugger. The only reason for the default tag starting with "." is to avoid conflicts with
-parameter or covariate names in the model expression. But it is only the default.
+
+Some of the discussion in the “Jacobian code in selfStart models” section 
+(p. 9) of this paper claims that the plinear fit followed by a fit using 
+the default approach would be difficult to maintain and could be very slow
+but that has not been the case. It is known (and, in fact, stated in the 
+documentation for SSlogis) that the call to the default nls algorithm will 
+always declare convergence immediately.
+
+Response 1:03 Can we have a pointer to the documentation of the immediate 
+      convergence? We have checked the documentation ?SSlogis, as well as 
+      selfStart and getInitial, but do not see a mention in R 4.3.1.
+      ??Interim response --
+
+Also in that section it is claimed that deriv creates “hidden variables” 
+because the default for the tag argument to deriv is .expr. This tag is 
+only used in bindings within the evaluation environment of the model 
+function, which is never visible to the user unless they are inspecting 
+the function evaluation with a symbolic debugger. The only reason for 
+the default tag starting with "." is to avoid conflicts with parameter 
+or covariate names in the model expression. But it is only the default.
+
+Response 1:04 Thanks. Reworded to explain the "why".
+
 Recommendations
-As stated in the “General” section, I think the purpose of the paper should be clarified. If the purpose
-is to introduce the nlsr package and contrast it with the existing implementation of nls in the stats
-package, then it might be helpful to understand why nls was implemented the way it was.
-I hope that this unconventional referee’s report is of use to the authors and the editor.
+
+As stated in the “General” section, I think the purpose of the paper 
+should be clarified. If the purpose is to introduce the nlsr package 
+and contrast it with the existing implementation of nls in the stats
+package, then it might be helpful to understand why nls was implemented 
+the way it was.
+
+Response 1:05 The historical context could be very useful to the R community.
+    Would the editors consider asking that Referee 1 consider recasting
+    the material as a companion note?
+
+    See also 2:01 for clarification of purpose.
+
+I hope that this unconventional referee’s report is of use to the authors 
+and the editor.
+
 References
-Chambers, John M., and Trevor J. Hastie, eds. 1992. Statistical Models in S. Routledge. https:
-//doi.org/10.1201/9780203738535.
-Nash, John C, and Duncan Murdoch. 2023. Nlsr: Functions for Nonlinear Least Squares Solutions -
-Updated 2022. https://CRAN.R-project.org/package=nlsr.
-Pinheiro, José, and Douglas Bates. 2000. Mixed-Effects Models in S and S-PLUS. Springer New York.
-5
-\
+
+Chambers, John M., and Trevor J. Hastie, eds. 1992. Statistical Models 
+in S. Routledge. https://doi.org/10.1201/9780203738535.
+
+Nash, John C, and Duncan Murdoch. 2023. Nlsr: Functions for Nonlinear 
+Least Squares Solutions - Updated 2022. https://CRAN.R-project.org/package=nlsr.
+
+Pinheiro, José, and Douglas Bates. 2000. Mixed-Effects Models in S and 
+S-PLUS. Springer New York.
 
@@
 2023-06-28
+
 General
 
-This article provides a vignette-style overview of the nlsr-package and highlights its design choices and
-behavior in comparison to base R’s nls() function as well as the nonlinear least squares solvers provided by
-the R-packages minpack.lm and gslnls. The article’s main contribution is to make R users (more) aware of
-the historic shortcomings and pitfalls in using R’s nls() function and how the nlsr-package is able to serve
-as a more reliable alternative. Overall, the article is well-written and structured into coherent subthemes
-and sections. Given that nls() is the de facto standard for solving nonlinear least squares problems in R, I
-believe this article provides valuable insights and is well-suited for publication in the R Journal after some
-revisions and minor clarifications.
+This article provides a vignette-style overview of the nlsr-package and 
+highlights its design choices and behavior in comparison to base R’s nls() 
+function as well as the nonlinear least squares solvers provided by the 
+R-packages minpack.lm and gslnls. The article’s main contribution is to 
+make R users (more) aware of the historic shortcomings and pitfalls in 
+using R’s nls() function and how the nlsr-package is able to serve as a 
+more reliable alternative. Overall, the article is well-written and 
+structured into coherent subthemes and sections. Given that nls() is 
+the de facto standard for solving nonlinear least squares problems in R, I
+believe this article provides valuable insights and is well-suited for 
+publication in the R Journal after some revisions and minor clarifications.
 
-My main advice is to revise the title and introduction of the article. As it currently stands, it is suggested
-that the article provides an (objective) comparison of nonlinear least squares solvers available in R. The body
-of the article reads more like a vignette for the nlsr-package, explaining its design choices and comparing its
-behavior to nls() and other NLS solvers based on a few selected examples. As such, a more appropriate title
-could be, for instance, The current state of R tools for nonlinear least squares modeling or similar.
+My main advice is to revise the title and introduction of the article. As 
+it currently stands, it is suggested that the article provides an 
+(objective) comparison of nonlinear least squares solvers available in R. 
+The body of the article reads more like a vignette for the nlsr-package, 
+explaining its design choices and comparing its behavior to nls() and other 
+NLS solvers based on a few selected examples. As such, a more appropriate title
+could be, for instance, The current state of R tools for nonlinear least 
+squares modeling or similar.
 
-2:01
+Response 2:01
 
 We appreciate the positive comments and agree with the suggested title change.
-Both referees have requested revisions to the introduction, and we hope that we have addressed the
-comments.
+
+Response 2:02
+
+Both referees have requested revisions to the introduction, and we hope that 
+we have addressed the comments, essentially by clarifying the goals and how
+they arose.
 
 Detailed comments
-1. The Hobbs weed example as an illustration of the performance of different NLS solvers is lacking a
-thorough comparison. I understand that the aim of the authors is to provide a data example that
-demonstrates the reliability of nlsr in favor of nls(), but nls() converges to the same solution as
-nlsr for all model setups when using the Port algorithm. This goes unnoticed in the text. Also,
-the model setup (Logistic3T) for which gsl_nls() fails can easily be made to converge correctly by
-changing either the algorithm or the scaling method, see the Illustrating code section. My advice
-is to provide a more comprehensive comparison of the available NLS solvers, or to pick a different
-data example that better highlights the benefits of nlsr. For instance, trying to solve the BoxBOD
-regression problem (https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), nls() fails with
-starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton and Port algorithm, whereas nlsr::nlxb()
-converges correctly.
-2. The section Returned results of nls() and other tools discusses the complexity of S3-objects of class
-nls, illustrated by the retrieval of the input data from an nls-object. It seems ill-advised to run
-eval(parse(text=result$data)), since it evaluates the name of the data object in the user’s envi-
-ronment, which may not exist. Instead, one should use model = TRUE to store the input data in the
-returned nls-object. This behavior is consistent with lm(), so perhaps less confusing than currently
-presented in the manuscript. The authors mention that nlxb() returns a much simpler structure of 11
-items in one level, but I did not manage to find the input data object in the nlsr-object returned by
-nlxb().
-3. The section Functional specification of problems mentions the use of nlsr::nlfb() to solve nonlinear
-least squares problems using a function to define the residuals. Since nlsr::nlfb() only has a notion
-of the residuals, how is it able to disentangle model and response, for the purpose of e.g. generating
-1model predictions? The authors mention that the function wrapnlsr() can be used to return an object
-of class nls, but this does not seem to work for problems defined through a residual function as in
-nlsr::nlfb()? If indeed the case, how does one evaluate e.g. model predictions, parameter confidence
-intervals, etc. when using nlsr::nlfb()? This remains unclear to me.
-4. The section Philosophical considerations hints at the challenge of performing inference when parameter
-bound-constraints are involved in the optimization. In case of physical or model constraints on the
-parameters, a common approach is to reparametrize the model to be unconstrained in the parameters.
-Perhaps the authors can highlight an example scenario where imposing parameter constraints is
-preferable to model reparametrization (the included data examples with parameter constraints appear
-somewhat artificial), or elaborate further on the possible downsides of reparametrizing the model, if any.
-5. The section Programming language makes the argument that keeping to a single programming language
-can allow for easier maintenance and upgrades, and that the performance penalty for using code entirely
-in R has been much reduced in recent years. I agree that there is not much of a performance penalty
-when solving small problems, such as the data examples included in the article. However, when solving
-problems using a large amount of data or a large number of parameters, I have experienced that the
-run-time (as well as memory usage) can become an actual bottleneck. As an example, consider the
-Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) included in the Illustrating code
-section.
+
+1. The Hobbs weed example as an illustration of the performance of different 
+NLS solvers is lacking a thorough comparison. I understand that the aim of 
+the authors is to provide a data example that demonstrates the reliability 
+of nlsr in favor of nls(), but nls() converges to the same solution as nlsr 
+for all model setups when using the Port algorithm. This goes unnoticed in 
+the text.
+
+Response 2:03 "port" cases added. The "appears unfinished" comment in the 
+       documentation led us to avoid its use. This is the sort of issue we 
+       would very much like to see resolved, since "port" is the only tool
+       for bounds constraints.
+
+Also, the model setup (Logistic3T) for which gsl_nls() fails can easily be 
+made to converge correctly by changing either the algorithm or the scaling 
+method, see the Illustrating code section. My advice is to provide a more 
+comprehensive comparison of the available NLS solvers, or to pick a different
+data example that better highlights the benefits of nlsr. For instance, trying 
+to solve the BoxBOD regression problem 
+(https://www.itl.nist.gov/div898/strd/nls/data/boxbod.shtml), 
+nls() fails with starting values θ1 = 1, θ2 = 1 using both the Gauss-Newton 
+and Port algorithm, whereas nlsr::nlxb() converges correctly.
+
+Response 2:04 We have added a paragraph
+
+	We consider the code inside package `nlsr` to be competently coded, 
+	with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
+	in common with `minpack.lm` and `glsnls`. The success rate is high,
+        but is not perfect. However, our principal message is that we obtain 
+        diagnostic information in the form of the singular values of the Jacobian 
+	that allow us to try alternative starts, algorithms or programs in cases 
+	where there are indications of early termination.
+
+   We appreciate the pointers on other examples. However, a substantive comparison
+   would make for a very long and likely boring paper. The BoxBOD example is, in
+   many ways, similar to the Hobbs example in its nasty characteristics. 
+   Some years ago, one of us created a package NISTopt to use the NIST examples 
+   on function minimization tools (as per package NISTnls). See, for example, 
+   https://rdrr.io/rforge/NISTopt/ 
+   Experiences with these problems have led to an opinion (i.e., open to debate)
+   as to the usefulness of such extreme cases. They underline a point of view that
+   good diagnostics are more helpful than a "usually succeeds" level of reliability.
+
+The section Returned results of nls() and other tools discusses the complexity 
+of S3-objects of class nls, illustrated by the retrieval of the input data 
+from an nls-object. It seems ill-advised to run eval(parse(text=result$data)), 
+since it evaluates the name of the data object in the user’s environment, 
+which may not exist. Instead, one should use model = TRUE to store the input 
+data in the returned nls-object. This behavior is consistent with lm(), so 
+perhaps less confusing than currently presented in the manuscript. The 
+authors mention that nlxb() returns a much simpler structure of 11
+items in one level, but I did not manage to find the input data object in 
+the nlsr-object returned by nlxb().
+
+Response 2:05 ??
+
+The section Functional specification of problems mentions the use of 
+nlsr::nlfb() to solve nonlinear least squares problems using a function 
+to define the residuals. Since nlsr::nlfb() only has a notion of the 
+residuals, how is it able to disentangle model and response, for the 
+purpose of e.g. generating model predictions? The authors mention that 
+the function wrapnlsr() can be used to return an object of class nls, 
+but this does not seem to work for problems defined through a residual 
+function as in nlsr::nlfb()? If indeed the case, how does one evaluate 
+e.g. model predictions, parameter confidence intervals, etc. when using 
+nlsr::nlfb()? This remains unclear to me.
+
+Response 2:06 The referee's comments point out that we have not explained
+    clearly enough the important differences that arise between modeling
+    and minimizing a nonlinear sum of squares. We have added a short 
+    exposition using the Rosenbrock problem. The difference in context
+    is important and we thank the referee, as other readers would surely
+    have had similar concerns.
+
+4. The section Philosophical considerations hints at the challenge of 
+performing inference when parameter bound-constraints are involved in 
+the optimization. In case of physical or model constraints on the
+parameters, a common approach is to reparametrize the model to be 
+unconstrained in the parameters. Perhaps the authors can highlight an 
+example scenario where imposing parameter constraints is preferable to 
+model reparametrization (the included data examples with parameter 
+constraints appear somewhat artificial), or elaborate further on 
+the possible downsides of reparametrizing the model, if any.
+
+Response 2:07 Reparametrization is likely the appropriate approach if
+    traditional measures of dispersion are wanted.
+
+    Opinion: One of us (JN) has been interested for > 25 years in 
+    comparing the approaches of numerical analysts and statisticians.
+    The former are insterested in the nonlinear LS problem, for which
+    constraints are a natural characteristic of some problems, while the
+    latter want nonlinear models, where bounds constraints can be a
+    nuisance to the dispersion ideas. I'll not pretend my knowledge is
+    sufficiently broad to be able to capture all the nuances, but I
+    believe a dialog on these matters could be very fruitful in long
+    term better understanding. As far as I am aware, the literature on
+    this area from a statistical viewpoint is quite thin.
+
+    For the paper: We agree the examples are essentially artificial. They
+    are intended to show how to use the features. A discussion of an upper 
+    limit on the Asym parameter (i.e., weed density in some units) or the 
+    scal parameter (measuring how fast this is reached) might be possible, 
+    but the people who provided the problem are no longer available.
+
+    We have added two small comments on recommended usage of bounds. One
+    is to control for wild excursions in the optimization trajectory, as 
+    we believe this is sensible when bounds constraints are available, 
+    as in nlsr::nlxb. We also note that in cases where a simple and 
+    obvious transformation is available, such as squaring a parameter to 
+    ensure positivity, this is likely a better approach than imposing bounds.
+
+5. The section Programming language makes the argument that keeping to a 
+single programming language can allow for easier maintenance and upgrades, 
+and that the performance penalty for using code entirely in R has been 
+much reduced in recent years. I agree that there is not much of a 
+performance penalty when solving small problems, such as the data examples 
+included in the article. However, when solving problems using a large amount 
+of data or a large number of parameters, I have experienced that the run-time 
+(as well as memory usage) can become an actual bottleneck. As an example, 
+consider the Penalty function I problem (Moré, Garbow, and Hillstrom (1981)) 
+included in the Illustrating code section.
+
+Response 2:08  In that we are in general agreement with this comment, we have added
+    some clarification to advise that it is worthwhile being aware of how long
+    calculations are taking.
+
+    The referee's commends ant Illustrating Code led us to try the Pen1 problem
+    (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
+    to which one of us recently added Hessian code). This showed that the R code
+    of nlsr::nlfb() was faster than the compiled Fortran of minpack.lm::nls.lm.
+    Moreover, the all-R code optimx::Rcgmin was 2 orders of magnitude faster --
+    it does not build a matrix!
+
 Minor comments
-• Avoid emphasizing words by writing in all capitals. Examples: NOT (pg. 14, 15), WEIGHTED (pg.
-13) and more.
-• The link https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211 (as well as the links to github)
-may be subject to change or become unavailable in the future.
+
+• Avoid emphasizing words by writing in all capitals. Examples: NOT (pg. 14, 15), 
+WEIGHTED (pg. 13) and more.
+
+Response 2:09  Fair enought. Done.
+
+• The link https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211 (as well 
+as the links to github) may be subject to change or become unavailable in the future.
+
+Response 2:10  We are not sure how to answer this, though we agree with the sentiment. Does
+     the editor have any advice?
+
 Illustrating code
+
 ## Hobbs weed example
@@
 tt = 1:12,
-weed = c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192,
-31.443, 38.558, 50.156, 62.948, 75.995, 91.972)
+weed = c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 38.558, 50.156, 62.948, 75.995,
+ 91.972)
 )
 ### nls-port all setups
-nls(
-formula = weed ~ b1 / (1 + b2 * exp(-b3 * tt)),
-data = weeddf,
-start = c(b1 = 1, b2 = 1, b3 = 1),
-algorithm = "port"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ b1/(1 + b2 * exp(-b3 * tt))
-#>
-data: weeddf
-#>
-b1
-b2
-b3
-#> 196.1863 49.0916
-0.3136
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm "port", convergence message: relative convergence (4)
-nls(
-formula = weed ~ 100 * c1 / (1 + 10 * c2 * exp(-0.1 * c3 * tt)),
-2data = weeddf,
-start = c(c1 = 1, c2 = 1, c3 = 1),
-algorithm = "port"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ 100 * c1/(1 + 10 * c2 * exp(-0.1 * c3 * tt))
-#>
-data: weeddf
-#>
-c1
-c2
-c3
-#> 1.962 4.909 3.136
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm "port", convergence message: relative convergence (4)
-nls(
-formula = weed ~ Asym /(1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-algorithm = "port"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm "port", convergence message: relative convergence (4)
+nls(formula = weed ~ b1 / (1 + b2 * exp(-b3 * tt)), data = weeddf,
+     start = c(b1 = 1, b2 = 1, b3 = 1), algorithm = "port")
+nls(formula = weed ~ 100 * c1 / (1 + 10 * c2 * exp(-0.1 * c3 * tt)),
+     data = weeddf, start = c(c1 = 1, c2 = 1, c3 = 1), algorithm = "port")
+nls(formula = weed ~ Asym /(1 + exp((xmid - tt) / scal)),data = weeddf,
+     start = c(Asym = 1, xmid = 1, scal = 1), algorithm = "port")
 ### gsl_nls Logistic3T setup
-gslnls::gsl_nls(
-fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-algorithm = "lmaccel"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm: multifit/levenberg-marquardt+accel, (scaling: more, solver: qr)
-#>
-#> Number of iterations to convergence: 22
-#> Achieved convergence tolerance: 1.18e-11
-gslnls::gsl_nls(
-fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-algorithm = "dogleg"
-)
-3#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm: multifit/dogleg, (scaling: more, solver: qr)
-#>
-#> Number of iterations to convergence: 16
-#> Achieved convergence tolerance: 1.599e-13
-gslnls::gsl_nls(
-fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-algorithm = "ddogleg"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm: multifit/double-dogleg, (scaling: more, solver: qr)
-#>
-#> Number of iterations to convergence: 16
-#> Achieved convergence tolerance: 1.14e-11
-gslnls::gsl_nls(
-fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-algorithm = "subspace2D"
-)
-#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm: multifit/2D-subspace, (scaling: more, solver: qr)
-#>
-#> Number of iterations to convergence: 19
-#> Achieved convergence tolerance: 9.783e-13
-gslnls::gsl_nls(
-fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),
-data = weeddf,
-start = c(Asym = 1, xmid = 1, scal = 1),
-control = list(scale = "levenberg")
-4)
-#> Nonlinear regression model
-#>
-model: weed ~ Asym/(1 + exp((xmid - tt)/scal))
-#>
-data: weeddf
-#>
-Asym
-xmid
-scal
-#> 196.186 12.417
-3.189
-#> residual sum-of-squares: 2.587
-#>
-#> Algorithm: multifit/levenberg-marquardt, (scaling: levenberg, solver: qr)
-#>
-#> Number of iterations to convergence: 24
-#> Achieved convergence tolerance: 7.065e-13
-## NOT RUN
+gslnls::gsl_nls(fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),data = weeddf,
+    start = c(Asym = 1, xmid = 1, scal = 1), algorithm = "lmaccel")
+gslnls::gsl_nls(fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),data = weeddf,
+    start = c(Asym = 1, xmid = 1, scal = 1),algorithm = "dogleg")
+gslnls::gsl_nls(fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),data = weeddf,
+    start = c(Asym = 1, xmid = 1, scal = 1), algorithm = "ddogleg")
+gslnls::gsl_nls(fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),data = weeddf,
+    start = c(Asym = 1, xmid = 1, scal = 1),algorithm = "subspace2D")
+gslnls::gsl_nls(fn = weed ~ Asym / (1 + exp((xmid - tt) / scal)),data = weeddf,
+    start = c(Asym = 1, xmid = 1, scal = 1),control = list(scale = "levenberg")
+
+Response 2:11  We had missed the "scale" setting i.e., had looked at the documentation
+    of `gsl_nls()` but not `gsl_nls_control()`, and in particular the original
+    Gnu Scientific Library documentation. It turns out the default scaling for the
+    `gsl_nls()` code is the value of the maximum diagonal element of J'J. (J. Mor'e).
+    Marquardt scaling is the diagonal of J'J, while Levenberg is a unit matrix.
+    nlsr uses a linear combination of Levenberg and Marquardt as described in 
+    Nash, John C., Minimizing a nonlinear sum of squares function on a small computer,
+    Journal of the Institute for Mathematics and its Applications, 1977,
+    volume 19, pages 231-237. Details are unfortunately important.
+
 ## Penalty function I [More, Garbow, Hillstrom 1981]
@@
 )
-# # A tibble: 1 × 13
-#
-expression
-min
-median `itr/sec` mem_alloc `gc/sec`
-#
-<bch:expr>
-<bch:tm> <bch:tm>
-<dbl> <bch:byt>
-<dbl>
-# 1 large problem
-3.24m
-3.24m
-0.00515
-9.23GB
-0.226
+
 References
-Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. “Testing Unconstrained Optimization
-Software.” ACM Transactions on Mathematical Software 7: 17–41.
-5
+
+Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. “Testing Unconstrained Optimization Software.” ACM Transactions on Mathematical Software 7: 17–41.

### 2023-07-03 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
Responses -- first draft

--- 2023-05-14 RefactoringNLS/ComparingNLStools.Rmd
@@
 ---
-title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
+title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
 author:
@@
     toc: no
-bibliography: ./ImproveNLS.bib
+bibliography: ./StateOfRToolsForNLS.bib
 link-citations: yes
@@
 
-Our Google Summer of Code project "Improvements to `nls()`" considered the features and limitations of the R function `nls()` with the aim 
+Our Google Summer of Code project "Improvements to `nls()`" considered the features 
+and limitations of the R function `nls()` with the aim 
 of improving and rationalizing R tools for nonlinear regression.
@@
 the code base that provides the functionality `nls()` claims to offer.
-Various packages, including our `nlsr`, provide alternative capabilities.
+Various packages, in particular our `nlsr`, provide alternative capabilities.
 We consider the differences in goals, approaches, and features of
@@
 primary software distribution
-from the Comprehensive R Archive Network (*https://cran.r-project.org*).
-The function dates to the 1980s and the work related to @bateswatts in S (see 
-*https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+from the Comprehensive R Archive Network (*https://cran.r-project.org*), hereafter
+abbreviated CRAN.
+Users do not have to request that `nls()` be loaded.
+The function dates to the 1980s and the work related to @bateswatts and others 
+in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
 
@@
 
-The tools we will consider are the base-R `nls()` function 
-and some packages that are available on the CRAN repository.
+Our Google Summer of Code project for 2021, "Improvements to `nls()`"
+had the goal of providing a version of the code that would address some
+of the deficiencies discovered over the long period that the `nls()`
+has been part of R. However, it was also critical to preserve legacy
+functionality, and we did not achieve a result that provided desired
+improvements and preserved legacy results. On the other hand, working
+with `nls()` suggested ways our own `nlsr` package could be improved
+in features, while also revealing how it has a very different internal
+structure and design.
+
+The primary goal of this paper is to present some of these issues,
+with a limited consideration of some other tools that exist for 
+nonlinear least squares calculations. A hope is that we will encourage
+others to engage, possibly with us, to find ways to bring improvements.
+We note that the referees' comments have been most helpful to focus
+the discussion here in ways that illustrate the best aspects of open
+source software development. We believe it is desirable to consolidate
+tools where reasonable so that users have an easier task in choosing
+which to use, but such consolidation demands careful and wide-ranging
+discussion of options. It is much simpler to create new packages than
+to bring together and simplify existing ones.
+
+### Tools considered
+
+Besides the base-R `nls()` function, we consider some packages that 
+are available on the CRAN repository.
 Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
 **minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
-general nonlinear least-squares solvers.
-
-While we will provide capsule comments for some other CRAN packages, 
-we will not pursue those in the Bioconductor (@Bioconductor) 
+general nonlinear least-squares solvers. We will 
+not pursue tools in the Bioconductor (@Bioconductor) 
 collection, nor those on repositories such as 
@@
 ## Some other CRAN packages for nonlinear modeling
+
+?? check wording
 
@@
 a nonlinear sum of 
-squares. NOT generally recommended if nonlinear 
+squares. not generally recommended if nonlinear 
 least squares programs can be easily used, but provides a check and alternative
@@
 $$ J' J \delta = - J' r$$
-though we can avoid some loss of accuracy by NOT forming the inner product matrix
+though we can avoid some loss of accuracy by not forming the inner product matrix
 $J' J$ and solving the linear least squares matrix problem 
@@
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are NOT equivalent and
+values for parameters. These starting points are not equivalent and
 are deliberately crude choices. Workers performing many calculations 
@@
 
+However, `nls()` also offers the option of using the "port" library algorithm
+choice (@FoxHallSchryer1978). There is a warning in the `nls()` documentation that
+this "appears unfinished", but it is successful in solving all three cases.
+
+```{r ex02nlsport}
+unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
+pnls(unls1p)
+snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
+pnls(snls1p)
+tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
+pnls(tnls1p)
+```
+
 ### Solution attempts with nlsr
@@
 values. Note that these are **displayed** by package `nlsr` in a single column in
-the output to provide a compact layout, but the values do **NOT** correspond to 
+the output to provide a compact layout, but the values do **not** correspond to 
 the individual parameters in 
@@
 pnls(tgslnls1) 
+## Other algorithms
+## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
+## pnls(tgslnls1a) 
+## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
+## pnls(tgslnls1d) 
+## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
+## pnls(tgslnls1dd) 
+## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
+## pnls(tgslnls1s2) 
 ```
@@
 `minpack.lm::nlsLM()` and `nls()`
-use `summary()`, which does NOT display the sum of squares, while `print()` 
+use `summary()`, which does not display the sum of squares, while `print()` 
 gives the sum of squares, but not the standard error of the residuals.
-
-The singular values allow us to gauge how "nearly singular" the 
+As pointed out by one of the referees, such behaviour is consistent
+with other modeling functions such as `lm()`.
+
+The importance of the  singular values is to allow us to gauge how "nearly singular" the 
 Jacobian is at the solution, and the ratio of the smallest to largest of the 
@@
 method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
-reported at this saddle point, `nlsr::nlxb()` cannot proceed.
+reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also note that
+changing the start to
+```
+stt2<-c(Asym=1, xmid=1, scal=2)
+```
+allows both `nlsLM` and `gsl_nls` to find the minimum.
 
@@
 
+We consider the code inside package `nlsr` to be competently coded, with
+the advantage over `nls()` of the Levenberg-Marquardt
+stabilization in common with `minpack.lm` and `glsnls`. However, our principal 
+message is that we obtain diagnostic information in the form of the singular values
+of the Jacobian that allow us to try alternative starts, algorithms or programs in 
+cases where there are indications of early termination.
+
 ## Functional specification of problems
+
+Let us write a slightly modified form of the nonlinear least squares problem:
+
+$$ S(p) = \sum_{i=1}^m { r_i(p)^2 }$$
+
+R users are likely to think almost always that the functions $r_i$ are all of
+the same form, differing in the details of the data they use. However, the
+famous Rosenbrock banana-shaped valley (@Rosenbrock1960, 
+also https://en.wikipedia.org/wiki/Rosenbrock_function) has 
+$$ r_1(p) = (1 - p_1)$$
+$$ r_2(p) = 10 (p_2 - {p_1}^2)$$
+
+Thus there can be problems where the idea of a "model" or a "prediction"
+makes no sense, though the nonlinear least squares problem is still
+well-posed. In our code, we still refer to the functions as "residuals",
+though this is purely a convenience.
 
@@
 By contrast, `nlsr::nlxb()` returns a much simpler structure of 11 items
-in one level. Moreover, `nlxb` explicitly returns the sum of squares, the residual vector, Jacobian, and counts of evaluations.
+in one level. Moreover, `nlxb` explicitly returns the sum of squares, the 
+residual vector, Jacobian, and counts of evaluations.
 
@@
 
-We are also surprised that the 
-analytic expressions for the Jacobian ("gradient") in the `SSLogis` function 
-and others save quantities in "hidden"
-variables, i.e., with names preceded by ".". 
-These are then not displayed by the `ls()`
-command, making them difficult to access by users who may wish to create
-their own selfStart model via copy and edit.
+We note that the analytic expressions for the Jacobian ("gradient") in the `SSLogis` 
+function and others save quantities by default in "hidden" variables, i.e., with 
+names preceded by ".".  A referee has indicated this behaviour is to avoid name 
+conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
+command, so users need to remember to use `ls(all.names=TRUE)` to see which quantities
+are defined, for example, in developing their own selfStart model via copy and edit.
+
 Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
@@
 
+Examples of the use of bounds constraints for nonlinear regression modeling are
+quite scarce. Traditionally, workers transform variables to ensure bounds are
+satisfied, such as squaring a parameter, or using a parameter which is the log()
+of the actual parameter, so that its exponential is positive. These transformations
+can create scaling issues, but seem to generally work.
+
+Bounds can also be helpful to avoid unanticipated excursions of the minimization
+trajectory into inadmissible parameter regions. Such bounds should not be too
+tight, as this may hinder progress to an optimum by causing the methods to 
+bump along the constraint in many cycles. Our aim is to avoid rubbish, not
+specify where the solution will be.
+
 ```{r ex10, echo=TRUE} 
@@
 pnls(anlshob1b) #  check the answer (short form)
-# nlsLM seems NOT to work with bounds in this example
+# nlsLM seems not to work with bounds in this example
 anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
@@
 
+One of the referees pointed out that nlsr::nlfb() is quite slow on a
+"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
+However, we would suggest that problems in large numbers of parameters
+will be problematic for Gauss-Newton methods because they create large
+computational matrices for the Jacobian or its inner product. The method
+"Rcgmin" from package `optimx` (@p-optimx) can be orders of magnitude
+faster, but is entirely written in R. Note that "Rcgmin" is a much more
+reliable conjugate gradient minimizer than method "CG" of `optim()`. One
+of us created both, but now deprecates the latter.
+
 ## Data sources for problems
@@
 search in the available environments (i.e., workspaces) for suitable data objects. 
-We do NOT like this approach, but it is "the R way". R allows users to leave many 
+We do not like this approach, but it is "the R way". R allows users to leave many 
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
@@
 
-The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
+The function `resid()` (an alias for `residuals()`) gives weighted residuals, 
 as does, for example, `result$m$resid()`. The function `nlsModel()`, which we 
@@
 Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # unweighted
+rnowt<-nowt$m$resid() # This has unweighted residual and Jacobian. Does not take coefficients.
 attr(rnowt, "gradient")<-NULL; rnowt
 usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+rusewt<-usewt$m$resid() # weighted. Does not take coefficients.
 attr(rusewt,"gradient")<-NULL; rusewt
@@
 Within `nls()`, consider the following FOUR different specifications for the same 
-problem, plus an intuitive choice, labeled `fm2a`, that does NOT work. 
+problem, plus an intuitive choice, labeled `fm2a`, that does not work. 
 In this failed attempt, putting the `Asym` parameter in the model causes the
@@
 like to see a consistent syntax. At the time of writing, we do
-not foresee a resolution for this issue. In the example, we have NOT evaluated
+not foresee a resolution for this issue. In the example, we have not evaluated
 the commands to save space.
@@
 
-## using conditional linearity AND Asym does NOT work
+## using conditional linearity AND Asym does not work
 fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
@@
 parameters are indexed but which uses the "plinear" option as an added complication.
-Running this example reveals that the answers for the parameters are NOT indexed
+Running this example reveals that the answers for the parameters are not indexed
 as in a vector. 
@@
 <!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
-<!--   ## where theta is constant. For now alpha and beta do NOT vary with Strip. -->
+<!--   ## where theta is constant. For now alpha and beta do not vary with Strip. -->
 <!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->

### 2023-07-03 RefactoringNLS/revision202307/BoxBOD.R

@@ -1,0 +1,54 @@
+# BoxBOD.R -- NIST/ITL StRD
+y<-c(109, 149, 149, 191, 213, 224)
+x<-c(1,2,3,5,7,10)
+boddta<-data.frame(x, y)
+## Reference:     Box, G. P., W. G. Hunter, and J. S. Hunter (1978).
+##               Statistics for Experimenters.  
+##               New York, NY: Wiley, pp. 483-487.
+# Formula               y = b1*(1-exp[-b2*x])  +  e
+boxbodfrm<-   y ~ b1*(1-exp(-b2*x))
+#      Starting values          Certified Values
+#
+#      Start 1  Start 2     Parameter     Standard Deviation
+#  b1 =   1       100   2.1380940889E+02  1.2354515176E+01
+#  b2 =   1       0.75  5.4723748542E-01  1.0455993237E-01
+
+#Residual Sum of Squares:          1.1680088766E+03
+#Residual Standard Deviation:      1.7088072423E+01
+#Degrees of Freedom:                                4
+#Number of Observations:                            6  
+
+st1<-c(b1=1, b2=1)
+st2<-c(b1=100, b2=0.75)
+library(nlsr)
+bbnlsr1<-nlsr::nlxb(formula=boxbodfrm, data=boddta, start=st1)
+pshort(bbnlsr1)
+bbnlsr2<-nlsr::nlxb(formula=boxbodfrm, data=boddta, start=st2)
+pshort(bbnlsr2)
+print(bbnlsr2)
+
+# nls
+
+bbnls1<-try(nls(formula=boxbodfrm, data=boddta, start=st1))
+pnls(bbnls1)
+bbnls2<-try(nls(formula=boxbodfrm, data=boddta, start=st2))
+pnls(bbnls2)
+bbnls1p<-try(nls(formula=boxbodfrm, data=boddta, start=st1, algorithm="port"))
+pnls(bbnls1p)
+bbnls2p<-try(nls(formula=boxbodfrm, data=boddta, start=st2, algorithm="port"))
+pnls(bbnls2p)
+
+library(minpack.lm)
+bbnlsLM1<-nlsLM(formula=boxbodfrm, data=boddta, start=st1)
+pnlslm(bbnlsLM1)
+bbnlsLM2<-nlsLM(formula=boxbodfrm, data=boddta, start=st2)
+pnlslm(bbnlsLM2)
+
+
+
+
+
+
+
+
+

### 2023-07-03 RefactoringNLS/revision202307/Pen1test.R

@@ -1,0 +1,79 @@
+## Penalty function I [More, Garbow, Hillstrom 1981]
+alpha <- 1e-5
+a <- 1e-5
+sqrta <- sqrt(alpha)
+### residuals
+fres <- function(x) {
+  c(sqrt(alpha) * (x - 1), sum(x^2) - 0.25)
+}
+### jacobian
+fjac <- function(x) {
+  jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
+  attr(jj, "gradient") <- jj
+  jj
+}
+
+ffn = function(par) {
+  n <- length(par)
+  if (n < 1) {
+    stop("Penalty Function I: n must be positive")
+  }
+  fsum <- 0
+  fn1 <- 0
+  for (i in 1:n) {
+    fi <- sqrta * (par[i] - 1)
+    fsum <- fsum + fi * fi
+    fn1 <- fn1 + par[i] * par[i]
+  }
+  
+  fn1 <- fn1 - 0.25
+  fsum <- fsum + fn1 * fn1
+  fsum
+}
+
+fgr = function(par) {
+  n <- length(par)
+  if (n < 1) {
+    stop("Penalty Function I: n must be positive")
+  }
+  grad <- rep(0, n)
+  fn1 <- 0
+  
+  for (i in 1:n) {
+    fi <- sqrta * (par[i] - 1)
+    grad[i] <- grad[i] + 2 * sqrta * fi
+    fn1 <- fn1 + par[i] * par[i]
+  }
+  fn1 <- fn1 - 0.25
+  grad <- grad + 4 * par * fn1
+  grad
+}
+
+n <- 10
+# ftest
+x0<-1:n
+cat("ffn=",ffn(x0),"\n")
+cat("fres:")
+fr<-fres(x0)
+print(fr)
+cat("sum(fr^2)=", sum(fr^2), "\n")
+cat("fgr:")
+print(fgr(x0))
+cat("Jacobian:\n")
+JJ<-fjac(x0)
+print(JJ)
+cat("grad from Jacobian:")
+grj <- 2*as.vector(t(JJ)%*%fr)
+print(grj)
+
+library(minpack.lm)
+library(optimx)
+n<-500
+x0<-as.numeric(1:n)
+
+tnlsr10<-bench::mark(nlsr10 = nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10)
+tnlsr10
[…]

### 2023-07-03 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
pen1

@@
 reliable conjugate gradient minimizer than method "CG" of `optim()`. One
-of us created both, but now deprecates the latter.
+of us created both, but now deprecates the latter. It should be noted
+that timings of CG methods are more sensitive to small differences in
+setup and arithmetic than Gauss-Newton methods.
 

### 2023-07-03 RefactoringNLS/revision202307/Pen1test.R

@@ -1,4 +1,5 @@
-## Penalty function I [More, Garbow, Hillstrom 1981]
+# Penalty function I [More, Garbow, Hillstrom 1981]
+sink("Pen1Times.txt", split=TRUE)
 alpha <- 1e-5
 a <- 1e-5
 sqrta <- sqrt(alpha)
@@ -48,8 +49,7 @@ fgr = function(par) {
   grad <- grad + 4 * par * fn1
   grad
 }
-
-n <- 10
+n<-10
 # ftest
 x0<-1:n
 cat("ffn=",ffn(x0),"\n")
@@ -66,14 +66,33 @@ cat("grad from Jacobian:")
 grj <- 2*as.vector(t(JJ)%*%fr)
 print(grj)
 
+
 library(minpack.lm)
 library(optimx)
-n<-500
-x0<-as.numeric(1:n)
 
-tnlsr10<-bench::mark(nlsr10 = nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10)
-tnlsr10
-tnlslm10<-bench::mark(nlslm10 = nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10)
-tnlslm10
-tcg10<-bench::mark(cg10 = optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10)
-tcg10
+nvals <- c(10, 25, 50, 100, 250, 500)
+nn<-length(nvals)
+TT <- matrix(NA, nrow=nn, ncol=4)
+MM <- TT
+for (i in 1:nn){
+n <- nvals[i]
+TT[i, 1] <- n
+MM[i, 1] <- n
+x0<-as.numeric(1:n)
+tnlsr10<-bench::mark(nlsr10 = nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10, time_unit='ms')
+print(tnlsr10)
+TT[i,2] <- tnlsr10$median
+MM[i,2] <- tnlsr10$min
+tnlslm10<-bench::mark(nlslm10 = nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
+print(tnlslm10)
+TT[i,3] <- tnlslm10$median
+MM[i,3] <- tnlslm10$min
[…]

### 2023-07-03 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
update to work

@@
 
-As mentioned, the output of `nls()` is an object of class "nls" which has a quite rich structure
-described in the manual file or revealed by applying the `str()` function to the
-result of `nls()`.
+As mentioned, the output of `nls()` is an object of class "nls" which has a quite 
+rich structure described in the manual file or revealed by applying the `str()` 
+function to the result of `nls()`.
 
@@
 `result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
-`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. To
-actually access the data from this element, we need to use the syntax:
+`gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. 
+If the user's environment exists, which it may not, we could see this data 
+via the syntax:
 ```
@@
 ```
-However, if the call is made with `model=TRUE`, then there is a returned element
+However, a better approach is to make the call to `nls()` with `model=TRUE`, 
+which returns an element 
 `model` which contains the data, and we can list its contents using:
@@
 in one level. Moreover, `nlxb` explicitly returns the sum of squares, the 
-residual vector, Jacobian, and counts of evaluations.
+residual vector, Jacobian, and counts of evaluations, but does not include
+the input data.
 

### 2023-07-03 PenFun1.R

@@ -1,0 +1,18 @@
+library(nlsr)
+## Penalty function I [More, Garbow, Hillstrom 1981]
+alpha <- 1e-5
+n <- 1000
+### residuals
+fres <- function(x) {
+  c(sqrt(alpha) * (x - 1), sum(x^2) - 0.25)
+}
+### jacobian
+fjac <- function(x) {
+  jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
+  attr(jj, "gradient") <- jj
+  jj
+}
+bench::mark(
+  "large problem" = nlsr::nlfb(start = 1:n, resfn = fres, jacfn = fjac, trace=TRUE, control=list(femax=1000)),
+  iterations = 1
+)

### 2023-07-03 RefactoringNLS/revision202307/ReviewResponses.txt
again

@@
 
-Response 2:05 ??
+Response 2:05 The section has been reworded to reflect the comments. I (JN) 
+   confess to having used `lm()` only a few times in a whole career, so
+   missed the parallel to that function. My work has focused on developing 
+   computational tools or diagnosing extreme cases where conventional 
+   statistical calculations seem to have misfired, and also do not work 
+   exclusively in R.
 
@@
     ensure positivity, this is likely a better approach than imposing bounds.
+    However, when there are more general bounds, transformations become more
+    tricky to apply, and may de-scale a problem. We have considered the
+    transfinite function used in CRAN package `adagio`, but consider that
+    this needs more extensive investigation before recommending in general.
 
@@
     to which one of us recently added Hessian code). This showed that the R code
-    of nlsr::nlfb() was faster than the compiled Fortran of minpack.lm::nls.lm.
-    Moreover, the all-R code optimx::Rcgmin was 2 orders of magnitude faster --
+    of nlsr::nlfb() was sometimes faster than the compiled Fortran of minpack.lm::nls.lm.
+    Moreover, the all-R code optimx::Rcgmin was much faster for larger problems --
     it does not build a matrix!

### 2023-07-03 PenFun1.R
remove PenFun1

@@ -1,18 +1,0 @@
-library(nlsr)
-## Penalty function I [More, Garbow, Hillstrom 1981]
-alpha <- 1e-5
-n <- 1000
-### residuals
-fres <- function(x) {
-  c(sqrt(alpha) * (x - 1), sum(x^2) - 0.25)
-}
-### jacobian
-fjac <- function(x) {
-  jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
-  attr(jj, "gradient") <- jj
-  jj
-}
-bench::mark(
-  "large problem" = nlsr::nlfb(start = 1:n, resfn = fres, jacfn = fjac, trace=TRUE, control=list(femax=1000)),
-  iterations = 1
-)

### 2023-07-04 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
mention scale=
minor updates

@@
 Below we see `nls()` does poorly on the example problem. To save page space,
-we use 1-line result display functions from package `nlsr`, namely `pnls()` 
+we use 1-line result display functions from package `nlsr`, namely `pnls0()` 
 and `pshort()`.
@@
 unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
-pnls(unls1p)
+pnls0(unls1p)
 snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
-pnls(snls1p)
+pnls0(snls1p)
 tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
-pnls(tnls1p)
+pnls0(tnls1p)
 ```
@@
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-pnls(unlm1)  # Short form of output
+pnls0(unlm1)  # Short form of output
 snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
-pnls(snlm1)
+pnls0(snlm1)
 tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
-pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
+pnls0(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
 ```
@@
 ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
-pnls(ugslnls1) # to get sum of squares
+pnls0(ugslnls1) # to get sum of squares
 sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
-pnls(sgslnls1) # Use summary() to get display
+pnls0(sgslnls1) # Use summary() to get display
 tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
-pnls(tgslnls1) 
+pnls0(tgslnls1) 
 ## Other algorithms
 ## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
-## pnls(tgslnls1a) 
+## pnls0(tgslnls1a) 
 ## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
-## pnls(tgslnls1d) 
+## pnls0(tgslnls1d) 
 ## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
-## pnls(tgslnls1dd) 
+## pnls0(tgslnls1dd) 
 ## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
-## pnls(tgslnls1s2) 
+## pnls0(tgslnls1s2) 
 ```
@@
 
-The importance of the  singular values is to allow us to gauge how "nearly singular" the 
-Jacobian is at the solution, and the ratio of the smallest to largest of the 
+The importance of the  singular values is to allow us to gauge how "nearly singular" 
+the Jacobian is at the solution, and the ratio of the smallest to largest of the 
 singular values is a simple but effective measure. The ratios are
@@
 
-The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a very 
-large sum of squares, which may suggest that these programs have failed. 
+The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a 
+very large sum of squares, which may suggest that these programs have failed. 
 Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
 offer singular values, we need to extract the Jacobian and compute its
-singular values. The following script shows how to do this, using as Jacobian what is called
-the `gradient` element in the returned solution for these solvers.
+singular values. The following script shows how to do this, using as Jacobian 
+what is called the `gradient` element in the returned solution for these solvers.
 
@@
 two out of three singular values are essentially 0. Our Jacobian is singular, and no 
-method of the Gauss-Newton type should be able to continue. Indeed, from the parameters
-reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also note that
-changing the start to
+method of the Gauss-Newton type should be able to continue. Indeed, from the 
+parameters reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also 
+note that changing the start to
 ```
@@
 ```
-allows both `nlsLM` and `gsl_nls` to find the minimum.
+allows both `nlsLM` and `gsl_nls` to find the minimum. `gsl_nls` also succeeds
+when the particular form of the Marquardt stabilization is changed from the
+default by the control setting `scale="levenberg"`.
 
@@
 a numerical approximation for the Jacobian if the argument `jac` 
-is missing. Note function `nlsr::pnlslm()` for a 1-line display of
+is missing. Note function `nlsr::pnls0()` for a 1-line display of
 the results of `minpack.lm::nls.lm()`.
@@
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnlslm(hobnlm)  
+pnls0(hobnlm)  
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
-pnls(hobgsln)
+pnls0(hobgsln)
 hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
-pnls(hobgsl) # using analytic Jacobian
+pnls0(hobgsl) # using analytic Jacobian
 ```
@@
              upper=c(2,6,3), algorithm='port')
-pnls(anlshob1b) #  check the answer (short form)
+pnls0(anlshob1b) #  check the answer (short form)
 # nlsLM seems not to work with bounds in this example
 anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls(anlsLM1b)
+pnls0(anlsLM1b)
 # also no warning if starting out of bounds, but gets a good answer!!
@@
 anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls(anlsLMob)
+pnls0(anlsLMob)
 # Try nlsr::nlxb()
@@
 			upper=c(200, 60, 3), trace=traceval))
-pnls(anlM1)
+pnls0(anlM1)
 anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
 		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
-pnls(anlsmsk1)
+pnls0(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
@@
              upper=c(2,6,30), algorithm='port')
-pnls(anlshmsk1)
+pnls0(anlshmsk1)
 anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
                  upper=c(2,6,30))
-pnls(anlsLMmsks1)
+pnls0(anlsLMmsks1)
 

### 2023-07-04 RefactoringNLS/revision202307/Pen1test.R
timings on HP21

@@ -66,11 +66,12 @@ cat("grad from Jacobian:")
 grj <- 2*as.vector(t(JJ)%*%fr)
 print(grj)
 
-
+library(nlsr)
 library(minpack.lm)
 library(optimx)
 
-nvals <- c(10, 25, 50, 100, 250, 500)
+nvals <- c(50, 100, 150, 200, 250, 300)
+#  nvals <- c(2, 10, 25, 50)
 nn<-length(nvals)
 TT <- matrix(NA, nrow=nn, ncol=4)
 MM <- TT
@@ -79,20 +80,21 @@ n <- nvals[i]
 TT[i, 1] <- n
 MM[i, 1] <- n
 x0<-as.numeric(1:n)
-tnlsr10<-bench::mark(nlsr10 = nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10, time_unit='ms')
-print(tnlsr10)
+tnlsr10<-bench::mark(nlsr10 <- nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10, time_unit='ms')
+pshort(nlsr10)
 TT[i,2] <- tnlsr10$median
 MM[i,2] <- tnlsr10$min
-tnlslm10<-bench::mark(nlslm10 = nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
-print(tnlslm10)
+tnlslm10<-bench::mark(nlslm10 <- nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
+pnlm0(nlslm10)
 TT[i,3] <- tnlslm10$median
 MM[i,3] <- tnlslm10$min
-tcg10<-bench::mark(cg10 = optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10, time_unit='ms')
-print(tcg10)
+tcg10<-bench::mark(cg10 <- optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10, time_unit='ms')
+proptimr(cg10)
 TT[i,4] <- tcg10$median
 MM[i,4] <- tcg10$min
 }
-
+colnames(TT)<-c("n","nlfb","nls.lm","Rcgmin")
+colnames(MM)<-c("n","nlfb","nls.lm","Rcgmin")
 print(TT)
 print(MM)
 sink()

### 2023-07-05 RefactoringNLS/revision202307/ReviewResponses.txt
minor
updates to rev1

@@
+Revew Responses to first review of R Journal paper 2023-15.
+
+We have created a text file with the content of both referees'
+reports and edited our responses to comments under headings
+
+   Response X:YY
+
+where X indicates the referee (1 or 2) and YY is an identifying
+number.
+
+Response 0:01  We noticed a factor of 2 error in the equations
+    in the discussion of the background to methods and have
+    (hopefully!) corrected them.
+
 Review 1
@@
 be a vector that just happens to have length 1.
+
+Response 1:07 Thanks for explaining this so clearly. We have added a
+    note in section "Estimation of models specified as formulas".
 
@@
       selfStart and getInitial, but do not see a mention in R 4.3.1.
-      ??Interim response --
+
+      Interim response: We have reworded the paragraph starting
+	"In considering `selfStart` models,"
+      to reflect the comments of the referee and to provide a reference
+      to the source of the starting approximations we used in SSlogisJN().
 
@@
 	in common with `minpack.lm` and `glsnls`. The success rate is high,
-        but is not perfect. However, our principal message is that we obtain 
-        diagnostic information in the form of the singular values of the Jacobian 
-	that allow us to try alternative starts, algorithms or programs in cases 
-	where there are indications of early termination.
+        but `nlsr` has its premature terminations failures too. Our principal 
+        message is that we obtain diagnostic information in the form of the 
+        singular values of the Jacobian that allow us to try alternative starts, 
+        algorithms or programs in cases where there are indications of early termination.
 
@@
    many ways, similar to the Hobbs example in its nasty characteristics. 
-   Some years ago, one of us created a package NISTopt to use the NIST examples 
-   on function minimization tools (as per package NISTnls). See, for example, 
-   https://rdrr.io/rforge/NISTopt/ 
    Experiences with these problems have led to an opinion (i.e., open to debate)
-   as to the usefulness of such extreme cases. They underline a point of view that
-   good diagnostics are more helpful than a "usually succeeds" level of reliability.
+   as to the usefulness of such extreme cases in measuring performance. We want 
+   to underline a point of view that good diagnostics are more helpful than a 
+   "usually succeeds" level of reliability.
 
@@
     is important and we thank the referee, as other readers would surely
-    have had similar concerns.
+    have had similar concerns. We have also added a comment to the 
+    "Problem setup" section to note that `nls()` is not suited to this
+    approach. Also a minor amendment to indicate `wrapnlsr()` is intended
+    for problems specified by formula.
 
@@
     to which one of us recently added Hessian code). This showed that the R code
-    of nlsr::nlfb() was sometimes faster than the compiled Fortran of minpack.lm::nls.lm.
-    Moreover, the all-R code optimx::Rcgmin was much faster for larger problems --
-    it does not build a matrix!
+    of nlsr::nlfb() was generally slower by a small factor than the compiled 
+    Fortran of minpack.lm::nls.lm. However, the all-R code optimx::Rcgmin was 
+    much faster for larger problems -- it does not build a matrix!
+
+	## On HP21 (Ryzen 7, 16GB)
+
+	> print(TT)  # median times
+	       n       nlfb    nls.lm     Rcgmin
+	[1,]  50   15.02918   7.95267   3.399167
+	[2,] 100   96.74012  40.05409  96.431357
+	[3,] 150  319.44073 127.58728   6.143749
+	[4,] 200  605.29889 290.00972   6.109964
+	[5,] 250 1373.51954 532.36149 182.678074
+	[6,] 300 2732.06920 874.97992  12.007694
+
+	> print(MM)  # minimum times
+	       n       nlfb     nls.lm     Rcgmin
+	[1,]  50   14.69422   7.895639   3.345110
+	[2,] 100   92.10252  39.718163  94.155118
+	[3,] 150  281.62738 125.972404   6.042982
+	[4,] 200  595.52840 284.833648   5.816762
+	[5,] 250 1300.00651 526.563457 182.678074
+	[6,] 300 2662.15242 869.606156  11.600069
 

### 2023-07-05 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
updates to rev1

--- 2023-07-03 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
@@
 
-?? check wording
-
 `onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
@@
 If we expand the Hessian for nonlinear least squares problems, we find
-$$  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
+$$  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}$$
 Let us use $D_{i,j}$ for the elements of the second term of this expression.
@@
 Applying this to a sum of squares problem gives
-$$ H \delta = (J'J + D) \delta = - J' r$$
+$$ H \delta / 2 = (J'J + D) \delta = - J' r$$
 In this expression, only the elements of $D$ have second partial derivatives. 
@@
 that these multiplied the second partial derivatives of $r$, so he approximated
-$$ H \approx J' J$$
+$$ H / 2 \approx J' J$$
 by assuming $D \approx 0$. This results in the Gauss-Newton method where we
@@
 It is also possible to provide R functions for the residual and Jacobian.
+`nls()` is not well-suited to using such a formulation, but `nlsr::nlfb()`
+and `minpack.lm::nls.lm()` work specifically with such a formulation.
 This is usually more work for the user if the formula setup is possible.
@@
 search. This works very efficiently as long as $J$ is not ill-conditioned.
-Below we see `nls()` does poorly on the example problem. To save page space,
-we use 1-line result display functions from package `nlsr`, namely `pnls()` 
-and `pshort()`.
+Below we see the default `nls()` algorithm does poorly on the example problem. 
+To save page space, we use 1-line result display functions from package `nlsr`, 
+namely `pnls()` and `pshort()`.
 
@@
 Here we see the infamous "singular gradient" termination message of `nls()`. 
-
-However, `nls()` also offers the option of using the "port" library algorithm
+Technically it is the Jacobian that is singular. A
+referee noted that R in this case cannot distinguish scalar and vector functions,
+which would yield partial derivatives that are a gradient or Jacobian respectively.
+
+`nls()` also offers the option of using the "port" library algorithm
 choice (@FoxHallSchryer1978). There is a warning in the `nls()` documentation that
@@
 
+While default `gsl_nls()` gives much the same results as `nlsLM()`, uncommenting the
+calls to the alternative algorithms shows it can get the global minimum with any of
+these choices.
+
 ### Comparison notes for formula-setup solutions
@@
 
-The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a very 
-large sum of squares, which may suggest that these programs have failed. 
+The results from `nlsLM` and `gsl_nls` for the transformed model Logistic3T have a 
+very large sum of squares, which may suggest that these programs have failed. 
 Since `nls()`, `nlsLM()`, and `gsl_nls()` do not 
 offer singular values, we need to extract the Jacobian and compute its
-singular values. The following script shows how to do this, using as Jacobian what is called
+singular values. The following script shows how to do this, using as Jacobian 
 the `gradient` element in the returned solution for these solvers.
@@
 
-We consider the code inside package `nlsr` to be competently coded, with
-the advantage over `nls()` of the Levenberg-Marquardt
-stabilization in common with `minpack.lm` and `glsnls`. However, our principal 
-message is that we obtain diagnostic information in the form of the singular values
-of the Jacobian that allow us to try alternative starts, algorithms or programs in 
-cases where there are indications of early termination.
+We consider the code inside package `nlsr` to be competently coded, 
+with the advantage over `nls()` of the Levenberg-Marquardt stabilization 
+in common with `minpack.lm` and `glsnls`. The success rate is high,
+but `nlsr` has its premature terminations too. Our principal message is that we obtain 
+diagnostic information in the form of the singular values of the Jacobian 
+that allow us to try alternative starts, algorithms or programs in cases 
+where there are indications of early termination.
 
@@
 a numerical approximation for the Jacobian if the argument `jac` 
-is missing. Note function `nlsr::pnlslm()` for a 1-line display of
-the results of `minpack.lm::nls.lm()`.
+is missing. Note function `nlsr::pnls()` for a 1-line display of
+the results.
 
@@
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnlslm(hobnlm)  
+pnls(hobnlm)  
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
@@
 a number of specialized modeling and diagnostic extensions. For compatibility, 
-the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias. 
+the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias,
+which is intended for problems specified as formulas.
 This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
@@
 since it was introduced. The default value 
-gives the legacy behaviour. This improvement has been in the R distributed code 
+gives `nls()` its legacy behaviour. This improvement has been in the R distributed code 
 since version 4.1.0.
@@
 Additional termination tests can be used. `nlsr` has a **small sum of 
-squares** test (**smallsstest**) that compares the latest evaluated sum of squared
+squares** test (`smallsstest`) that compares the latest evaluated sum of squared
 (weighted) residuals to `e4` times the initial sum of squares, where
@@
 ```
-However, a better approach is to make the call to `nls()` with `model=TRUE`, 
-which returns an element 
+However, since the evaluating environment might not exist (thanks to a referee
+for noting this possibility) a better approach is to make the call to `nls()` 
+with `model=TRUE`, which returns an element 
 `model` which contains the data, and we can list its contents using:
@@
 Thus the code in `nlsr` is likely of value for workers to copy and modify
-for customized tools.
+for customized tools. Coding in R is intended to support such uses.
 
@@
 
-Using the "gradient" attribute of the output of the Jacobian function to hold
+Using the `gradient` attribute of the output of the Jacobian function to hold
 the Jacobian matrix lets us embed this in the **residual** function as well,
@@
 There are also some in the now-archived package `NRAIA`.
-These provide the Jacobian in the "gradient" attribute of the "one-sided" formula
+These provide the Jacobian in the `gradient` attribute of the "one-sided" formula
 that defines each model, and these Jacobians are often the analytic forms.
@@
 computing the "right-hand side" or `rhs` of the residual, checks to see if the
-"gradient" attribute is defined, otherwise using `numericDeriv()` to compute a 
+`gradient` attribute is defined, otherwise using `numericDeriv()` to compute a 
 Jacobian into that attribute. This code is within the `nlsModel()` or
@@
 and finds a (full) solution. It then passes the solution coefficients to the default 
-algorithm unnecessarily. Moreover, the implicit double call is, in our view, prone to creating 
-errors in code maintenance. 
-To provide simpler starting parameters, the function `SSlogisJN`
-is now part of the package `nlsr`, but is most useful for `nls()`.
+algorithm which converges in one iteration and builds the standard output. 
+We found the implicit double call a challenge to unravel. 
+A more conventional approach to starting parameters, following the lead of @Ratkowsky1983, 
+is in the function `SSlogisJN`, now part of the package `nlsr`.
 
@@
 
-We note that the analytic expressions for the Jacobian ("gradient") in the `SSLogis` 
+We note that the analytic expressions for the Jacobian (`gradient`) in the `SSLogis` 
 function and others save quantities by default in "hidden" variables, i.e., with 
@@
 "large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
-However, we would suggest that problems in large numbers of parameters
+However, we note that problems in large numbers of parameters
 will be problematic for Gauss-Newton methods because they create large
 computational matrices for the Jacobian or its inner product. The method
-"Rcgmin" from package `optimx` (@p-optimx) can be orders of magnitude
-faster, but is entirely written in R. Note that "Rcgmin" is a much more
+"Rcgmin" from package `optimx` (@p-optimx) can be much 
+faster, despite being entirely written in R, and is a much more
 reliable conjugate gradient minimizer than method "CG" of `optim()`. One
-of us created both, but now deprecates the latter. It should be noted
-that timings of CG methods are more sensitive to small differences in
+of us created both, but now deprecates the latter. 
+Timings of CG methods are more sensitive to small differences in
 setup and arithmetic than Gauss-Newton methods.

### 2023-07-05 RefactoringNLS/revision202307/Pen1test.R

@@ -1,5 +1,5 @@
 # Penalty function I [More, Garbow, Hillstrom 1981]
-sink("Pen1Times.txt", split=TRUE)
+# sink("Pen1Times.txt", split=TRUE)
 alpha <- 1e-5
 a <- 1e-5
 sqrta <- sqrt(alpha)
@@ -85,7 +85,7 @@ pshort(nlsr10)
 TT[i,2] <- tnlsr10$median
 MM[i,2] <- tnlsr10$min
 tnlslm10<-bench::mark(nlslm10 <- nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
-pnlm0(nlslm10)
+pnlslm(nlslm10)
 TT[i,3] <- tnlslm10$median
 MM[i,3] <- tnlslm10$min
 tcg10<-bench::mark(cg10 <- optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10, time_unit='ms')

### 2023-07-05 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
finish draft1 review

@@
 Bounds change those assumptions. Hence, `nlsr::nlxb()` does not compute standard
-errors nor their derived statistics when bounds are active.
+errors nor their derived statistics when bounds are active. Transformations to
+implicitly enforce bounds avoid this issue, but at the cost of changing the 
+form of the model.
 
@@
 
-Let us try to fix (mask) the first parameter in the first two example problems.
+Let us try to fix (mask) the first parameter in the first two example problems. We might
+want to do this if it is considered that the asymptote is known from external sources,
+but we wish later to have it available to be estimated.
 
@@
 will be problematic for Gauss-Newton methods because they create large
-computational matrices for the Jacobian or its inner product. The method
+computational matrices for the Jacobian or its inner product. This was
+confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
+the Penalty 1 problem of various sizes, where the latter, which is 
+compiled Fortran internally, was moderately faster. However, the method
 "Rcgmin" from package `optimx` (@p-optimx) can be much 
@@
 `mf`. We feel that users could benefit from more extensive documentation and 
-examples of its use since it is used to implement features like `subset`.
+examples of its use since it is used to implement features like `subset`. Moreover,
+it provides access to resources that may be important in further calculations on
+models.
 
@@
 especially if a "singular gradient" (singular Jacobian) is found. However, 
-this may occur AFTER the function has made considerable progress in reducing 
+this may occur **after** considerable progress has been made in reducing 
 the sum of squared residuals. 

### 2023-07-11 RefactoringNLS/revision202307/ReviewResponses.txt
Revision with AB ideas

@@
+Arkajyoti responses -- delete before submission:
+
+I have read through your comments on the suggestions of the two referees and I agree with them. As usual, the document is well-organized, methodical, and nice way to check all the boxes.
+
+For “Response 2:10“, I don’t think any link would be a good choice based on the referees’ concern.
+An alternative article (maybe the reference manual or vignette of the corresponding package  in question, like for unit testing) may serve as a viable replacement.
+    Agree that no link is likely to be satisfactory. Modified to suggest search to find
+    articles.
+
+Some comments:
+
+    Maybe cite your nlsr package, as “our” nlsr seems confusing? -- DONE
+
+    There is some overlap between the Abstract and scope of our comparison, and I suggest making the 
+        latter more succinct (for example the paragraph before TOOLS CONSIDERED)?
+        DONE -- simplified.
+
+    Maybe talking about referees’ suggestions can go in the acknowledgements? DONE
+
+    I suggest common website links, like github.com, gitlab.com and cran be removed?
+        Agreed.
+
+    “Some other CRAN  packages…” can go later, after the main material of the paper?
+        DONE -- before unit tests paragraph.
+
+    Documentation and resources seem tangential to the main material and maybe removed?
+        DONE -- this was when we were talking more about the GSoC project.
+
+    Maybe add 5, 6 above in a separate r blog post or the main article’s supplementary article?
+       ??? What supplementary article? If we want to reference blog posts, they should
+       be available now. Perhaps set aside and focus on getting this published. Thoughts?
+
+=========================================================================
 Revew Responses to first review of R Journal paper 2023-15.
@@
 we have addressed the comments, essentially by clarifying the goals and how
-they arose.
+they arose. In reconsidering the focus of the article, we have removed the 
+"Documentation and Resources" section that was primarily reporting outputs 
+of our Google Summer of Code project. We have also moved the section 
+"Some other CRAN packages for nonlinear modeling" towards the end of the
+paper.
 
@@
 
-    The referee's commends ant Illustrating Code led us to try the Pen1 problem
+    The referee's commends and Illustrating Code led us to try the Pen1 problem
     (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
@@
 
+    Pen1test.R is appended at the bottom of this file.
+
 Minor comments
@@
 
-Response 2:10  We are not sure how to answer this, though we agree with the sentiment. Does
-     the editor have any advice?
+Response 2:10  We agree with this opinion. 
+    We have replaced the link with 'a number of blog posts found by searching "unit 
+    testing in R"' as there are several possibilities, of which we felt that of 
+    André Müller was helpful.
 
@@
 Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. “Testing Unconstrained Optimization Software.” ACM Transactions on Mathematical Software 7: 17–41.
+
+-----------------------------------------------------------------
+Pen1test.R code
+
+# Penalty function I [More, Garbow, Hillstrom 1981]
+# sink("Pen1Times.txt", split=TRUE)
+alpha <- 1e-5
+a <- 1e-5
+sqrta <- sqrt(alpha)
+### residuals
+fres <- function(x) {
+  c(sqrt(alpha) * (x - 1), sum(x^2) - 0.25)
+}
+### jacobian
+fjac <- function(x) {
+  jj <- rbind(diag(sqrt(alpha), nrow = length(x)), 2 * t(x))
+  attr(jj, "gradient") <- jj
+  jj
+}
+
+ffn = function(par) {
+  n <- length(par)
+  if (n < 1) {
+    stop("Penalty Function I: n must be positive")
+  }
+  fsum <- 0
+  fn1 <- 0
+  for (i in 1:n) {
+    fi <- sqrta * (par[i] - 1)
+    fsum <- fsum + fi * fi
+    fn1 <- fn1 + par[i] * par[i]
+  }
+
+  fn1 <- fn1 - 0.25
+  fsum <- fsum + fn1 * fn1
+  fsum
+}
+
+fgr = function(par) {
+  n <- length(par)
+  if (n < 1) {
+    stop("Penalty Function I: n must be positive")
+  }
+  grad <- rep(0, n)
+  fn1 <- 0
+
+  for (i in 1:n) {
+    fi <- sqrta * (par[i] - 1)
+    grad[i] <- grad[i] + 2 * sqrta * fi
+    fn1 <- fn1 + par[i] * par[i]
+  }
+  fn1 <- fn1 - 0.25
+  grad <- grad + 4 * par * fn1
+  grad
+}
+n<-10
+# ftest
+x0<-1:n
+cat("ffn=",ffn(x0),"\n")
+cat("fres:")
+fr<-fres(x0)
+print(fr)
+cat("sum(fr^2)=", sum(fr^2), "\n")
+cat("fgr:")
+print(fgr(x0))
+cat("Jacobian:\n")
+JJ<-fjac(x0)
+print(JJ)
+cat("grad from Jacobian:")
+grj <- 2*as.vector(t(JJ)%*%fr)
+print(grj)
+
+library(nlsr)
+library(minpack.lm)
+library(optimx)
+
+nvals <- c(50, 100, 150, 200, 250, 300)
+#  nvals <- c(2, 10, 25, 50)
+nn<-length(nvals)
+TT <- matrix(NA, nrow=nn, ncol=4)
+MM <- TT
+for (i in 1:nn){
+n <- nvals[i]
+TT[i, 1] <- n
+MM[i, 1] <- n
+x0<-as.numeric(1:n)
+tnlsr10<-bench::mark(nlsr10 <- nlsr::nlfb(start = x0, resfn = fres, jacfn = fjac),  iterations = 10, time_unit='ms')
+pshort(nlsr10)
+TT[i,2] <- tnlsr10$median
+MM[i,2] <- tnlsr10$min
+tnlslm10<-bench::mark(nlslm10 <- nls.lm(par = x0, fn = fres, jac = fjac),  iterations = 10, time_unit='ms')
+pnlslm(nlslm10)
+TT[i,3] <- tnlslm10$median
+MM[i,3] <- tnlslm10$min
+tcg10<-bench::mark(cg10 <- optimr(par = x0, fn = ffn, gr = fgr, method="Rcgmin"),  iterations = 10, time_unit='ms')
+proptimr(cg10)
+TT[i,4] <- tcg10$median
+MM[i,4] <- tcg10$min
+}
+colnames(TT)<-c("n","nlfb","nls.lm","Rcgmin")
+colnames(MM)<-c("n","nlfb","nls.lm","Rcgmin")
+print(TT)
+print(MM)
+sink()

### 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
minor updates

@@
 the code base that provides the functionality `nls()` claims to offer.
-Various packages, in particular our `nlsr`, provide alternative capabilities.
+Various packages, in particular `nlsr` (@nlsr2023manual), provide alternative capabilities.
 We consider the differences in goals, approaches, and features of
@@
 primary software distribution
-from the Comprehensive R Archive Network (*https://cran.r-project.org*), hereafter
-abbreviated CRAN.
+from the Comprehensive R Archive Network (CRAN).
 Users do not have to request that `nls()` be loaded.
 The function dates to the 1980s and the work related to @bateswatts and others 
-in S (see *https://en.wikipedia.org/wiki/S_%28programming_language%29*).
+in the S programming language.
 
@@
 - handles formulas that include R functions, even ones that call calculations in
-  other programming languages
-- allows data to be weighted or subset
-- can estimate bound-constrained parameters
-- provides a mechanism for handling partially linear models
-- permits parameters to be indexed over a set of related data
+  other programming languages;
+- allows data to be weighted or subset;
+- can estimate bound-constrained parameters;
+- provides a mechanism for handling partially linear models;
+- permits parameters to be indexed over a set of related data;
 - produces measures of variability (i.e., standard error estimates) for the 
-estimated parameters
+estimated parameters;
 - has related profiling capabilities for exploring the likelihood surface as 
-parameters are changed
+parameters are changed;
 - links to many pre-coded (`selfStart`) models that do not require
@@
 had the goal of providing a version of the code that would address some
-of the deficiencies discovered over the long period that the `nls()`
+of the deficiencies discovered over the long period that `nls()`
 has been part of R. However, it was also critical to preserve legacy
 functionality, and we did not achieve a result that provided desired
-improvements and preserved legacy results. On the other hand, working
-with `nls()` suggested ways our own `nlsr` package could be improved
+improvements and preserved legacy results simultaneously. On the other hand, working
+with `nls()` suggested ways package `nlsr` could be improved
 in features, while also revealing how it has a very different internal
@@
 The primary goal of this paper is to present some of these issues,
-with a limited consideration of some other tools that exist for 
-nonlinear least squares calculations. A hope is that we will encourage
-others to engage, possibly with us, to find ways to bring improvements.
-We note that the referees' comments have been most helpful to focus
-the discussion here in ways that illustrate the best aspects of open
-source software development. We believe it is desirable to consolidate
-tools where reasonable so that users have an easier task in choosing
-which to use, but such consolidation demands careful and wide-ranging
-discussion of options. It is much simpler to create new packages than
-to bring together and simplify existing ones.
+with a limited consideration of some other R tools that exist for 
+nonlinear least squares calculations. We hope to encourage
+others to engage, possibly with us, to find ways to bring improvements and
+to consolidate tools where reasonable. We want to ease the task of users
+in choosing which tools to use. Unfortunately, it is simpler to create 
+new packages than to bring together and simplify existing ones.
 
@@
 Besides the base-R `nls()` function, we consider some packages that 
-are available on the CRAN repository.
-Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
+are available on the CRAN repository. Of these, we will pay particular 
+attention to **nlsr** (@nlsr2023manual),
 **minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
@@
 collection, nor those on repositories such as 
-GitHub (https://github.com) and Gitlab (https://about.gitlab.com).
-
-## Some other CRAN packages for nonlinear modeling
-
-`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
-is the vertical difference between the dependent variable and the functional
-model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
-therefore different and involves nontrivial extra calculation. A vignette with
-the package and the blog article 
-https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
-give some description with illustrative graphs. `onls` appears to be limited to 
-problems with one independent and one dependent variable. The Wikipedia article
-https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
-ideas, with references to the literature. The approach needs a wider 
-discussion and tutorial examples to allow its merits to be judged than can be
-included here.
-
-`crsnls` (@R-crsnls-manual) -- This package allows nonlinear estimation by 
-controlled random search via two methods. There is unfortunately no vignette.
-A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
-a small fraction of the time required by either of the methods in `crsnls`. The
-method discussed in @tvrdik2007 claims better reliability in finding solutions
-than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
-on the extreme NIST examples mentioned next.
-
-`NISTnls` (@NISTnls-manual) -- This package provides R code and data for a set of
-(numerically ill-conditioned) nonlinear least squares problems from the U.S. National 
-Institute for Standards and Technology. These may not represent real-world situations.
-
-`nlshelper` (@nlshelper-manual) -- This package, which unfortunately lacks a
-vignette, provides a few utilities for summarizing, testing, and plotting 
-non-linear regression models estimated with nls(), nlsList() or nlme() that
-are linked or grouped in some way.
-
-`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
-equality and/or inequality constraints. It is clearly **not** about modeling, and
-the input and output are quite different from class `nls` methods. However, there do
-not appear to be other R packages with these capabilities.
-
-`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models 
-dedicated  to predictive microbiology, including a vignette, by authors of the
-`nlstools` package.
-
-`nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation 
-of nonlinear models, particularly using `nls()`. The vignette is actually a journal
-article,  and the authors have considerable experience in the subject.
-
-`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
-regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem 
-here is to minimize an objective that is modified from the traditional sum of squared
-residuals.
-
-`nls.multstart` (@nlsmultstart) -- Non-linear least squares regression using
-AIC scores with the Levenberg-Marquardt algorithm using multiple starting values 
-for increasing the chance that the minimum found is the global minimum.
-
-`nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
-motivations to `nls.multstart`, but uses `nls()` within multiple trials. The author
-has extensive expertise in R.
-
-`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
-Fernandez Torvisco et al. (2018) for fitting separable nonlinear regression curves. The 
-special class of problem for which this package is intended is an important and 
-difficult one. No vignette is provided, unfortunately.
-
-`easynls`  -- Fit and plot some nonlinear models. Thirteen models are treated, but 
-there is minimal documentation and no vignette. Package `nlraa` is to be preferred.
-
-`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models, primarily from
-agriculture. Most include analytic Jacobian code.
-
-`optimx` (@p-optimx) -- This provides 
-optimizers that can be applied to minimize a nonlinear function which could be
-a nonlinear sum of 
-squares. not generally recommended if nonlinear 
-least squares programs can be easily used, but provides a check and alternative
-solvers.
+GitHub and Gitlab.
 
@@
 The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
-task that seems straightforward but is quite nasty. Its very succinct statement 
+task that seems straightforward, but is quite nasty. Its very succinct statement 
 provides the "short reproducible example" much requested on R mailing lists. 
-This is a real problem from a field researcher. The data and graph follow.
+This is a real problem from a field researcher, though the units of measurement
+and background information are not available. The data and graph follow.
 
@@
 $$ J \delta \approx -r $$
-by one of several matrix decomposition methods.
-
-In reality, there are many problems were $D$ should not be ignored, but the work
+by one of several matrix decomposition methods. On the other hand, when $m >> n$,
+the matrix $J' J$ uses much less storage than $J$.
+
+In reality, there are many problems where $D$ should not be ignored, but the work
 to compute it precisely is considerable. Many work-arounds have been 
@@
 # model formulas
-frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
-frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
-frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100* c1 / (1 + 10 * c2 * exp(-0.1 *c3 * tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
 #
 # Starting parameter sets
-stu1<-c(b1=1, b2=1, b3=1)
-sts1<-c(c1=1, c2=1, c3=1)
-stt1<-c(Asym=1, xmid=1, scal=1)
+stu1<-c(b1 =1, b2 = 1, b3 = 1)
+sts1<-c(c1 = 1, c2 = 1, c3 = 1)
+stt1<-c(Asym = 1, xmid = 1, scal = 1)
 ```
@@
   tt  <-  1:12
-  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+  res  <-  x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
 }
@@
   tt  <-  1:12
-  yy  <-  exp(-x[3]*tt)
-  zz  <-  1.0/(1+x[2]*yy)
+  yy  <-  exp(-x[3] * tt)
+  zz  <-  1.0 / (1 + x[2] * yy)
   jj[tt,1]   <-   zz
-  jj[tt,2]   <-   -x[1]*zz*zz*yy
-  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
+  jj[tt,2]   <-   -x[1] * zz * zz * yy
+  jj[tt,3]   <-   x[1] * zz * zz * yy * x[2] * tt
   attr(jj, "gradient") <- jj
@@
 Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
-search. This works very efficiently as long as $J$ is not ill-conditioned.
+search (@Hartley1961). This works very efficiently as long as $J$ is not ill-conditioned.
 Below we see the default `nls()` algorithm does poorly on the example problem. 
@@
 ```{r ex02nls}
-unls1<-try(nls(formula=frmu, start=stu1, data=weeddf))
-snls1<-try(nls(formula=frms, start=sts1, data=weeddf))
-tnls1<-try(nls(formula=frmt, start=stt1, data=weeddf))
+unls1<-try(nls(formula = frmu, start = stu1, data = weeddf))
+snls1<-try(nls(formula = frms, start = sts1, data = weeddf))
+tnls1<-try(nls(formula = frmt, start = stt1, data = weeddf))
 ```
@@
 Here we see the infamous "singular gradient" termination message of `nls()`. 
-Technically it is the Jacobian that is singular. A
-referee noted that R in this case cannot distinguish scalar and vector functions,
-which would yield partial derivatives that are a gradient or Jacobian respectively.
+Technically, it is the Jacobian that is singular, but R does not distinguish scalar 
+and vector functions, which yield partial derivatives that are a gradient or Jacobian 
+respectively.
 
@@
 ```{r ex02nlsport}
-unls1p<-try(nls(formula=frmu, start=stu1, data=weeddf, algorithm="port"))
+unls1p<-try(nls(formula = frmu, start = stu1, data = weeddf, algorithm = "port"))
 pnls(unls1p)
-snls1p<-try(nls(formula=frms, start=sts1, data=weeddf, algorithm="port"))
+snls1p<-try(nls(formula = frms, start = sts1, data = weeddf, algorithm = "port"))
 pnls(snls1p)
-tnls1p<-try(nls(formula=frmt, start=stt1, data=weeddf, algorithm="port"))
+tnls1p<-try(nls(formula = frmt, start = stt1, data = weeddf, algorithm = "port"))
 pnls(tnls1p)
@@
 ```{r ex02nlsr}
-unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
+unlx1 <- try(nlxb(formula = frmu, start = stu1, data = weeddf))
 print(unlx1) 
-snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
+snlx1 <- try(nlxb(formula = frms, start = sts1, data = weeddf))
 pshort(snlx1) # a short-form output
-tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
+tnlx1 <- try(nlxb(formula = frmt, start = stt1, data = weeddf))
 pshort(tnlx1) # alternatively print(tnlx1)
@@
 ```{r ex02minpack}
-unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
+unlm1 <- try(nlsLM(formula = frmu, start = stu1, data = weeddf))
 pnls(unlm1)  # Short form of output
-snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
+snlm1 <- try(nlsLM(formula = frms, start = sts1, data = weeddf))
 pnls(snlm1)
-tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
+tnlm1 <- try(nlsLM(formula = frmt, start = stt1, data = weeddf))
 pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
@@
 library(gslnls)
-ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
+ugslnls1 <- try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
 pnls(ugslnls1) # to get sum of squares
-sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
+sgslnls1 <- try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
 pnls(sgslnls1) # Use summary() to get display
-tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
+tgslnls1 <- try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
 pnls(tgslnls1) 
@@
 
-While default `gsl_nls()` gives much the same results as `nlsLM()`, uncommenting the
+While default `gsl_nls()` gives much of the same results as `nlsLM()`, uncommenting the
 calls to the alternative algorithms shows it can get the global minimum with any of
@@
 gives the sum of squares, but not the standard error of the residuals.
-As pointed out by one of the referees, such behaviour is consistent
-with other modeling functions such as `lm()`.
+Such behaviour is consistent with other modeling functions such as `lm()`.
 
@@
 4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
-Logistic3T, so Logistic3S is the "least singular".
+Logistic3T; so Logistic3S is the "least singular".
 
@@
 method of the Gauss-Newton type should be able to continue. Indeed, from the 
-parameters reported at this saddle point, `nlsr::nlxb()` cannot proceed. We also 
+parameters reported at this saddle point, `nlsr::nlxb()` cannot proceed.
+
+```{r ex05, echo=TRUE}  
+stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
+print(badstart)
+```
+
+We also 
 note that changing the start to
 ```
-stt2<-c(Asym=1, xmid=1, scal=2)
+stt2 <- c(Asym=1, xmid=1, scal=2)
 ```
@@
 when the particular form of the Marquardt stabilization is changed from the
-default by the control setting `scale="levenberg"`.
-
-```{r ex05, echo=TRUE}  
-stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
-badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
-print(badstart)
-```
+default by the control setting `scale = "levenberg"`.
 
@@
 in common with `minpack.lm` and `glsnls`. The success rate is high,
-but `nlsr` has its premature terminations too. Our principal message is that we obtain 
+but `nlsr` has its premature terminations too. However, it provides 
 diagnostic information in the form of the singular values of the Jacobian 
-that allow us to try alternative starts, algorithms or programs in cases 
+that allows us to try alternative starts, algorithms or programs in cases 
 where there are indications of early termination.
@@
 the same form, differing in the details of the data they use. However, the
-famous Rosenbrock banana-shaped valley (@Rosenbrock1960, 
-also https://en.wikipedia.org/wiki/Rosenbrock_function) has 
+famous Rosenbrock banana-shaped valley (@Rosenbrock1960) has 
 $$ r_1(p) = (1 - p_1)$$
@@
 
-Thus there can be problems where the idea of a "model" or a "prediction"
+Thus, there can be problems where the idea of a "model" or a "prediction"
 makes no sense, though the nonlinear least squares problem is still
@@
 a vector `y` that is the expected value of what we have called the
-residual, but in this case, is actually the model. `gsl_nls` uses 
+residual, but in this case is actually the model. `gsl_nls` uses 
 a numerical approximation for the Jacobian if the argument `jac` 
-is missing. Note function `nlsr::pnls()` for a 1-line display of
-the results.
+is missing. Note functions `nlsr::pnls()` and `nslr::pnlslm()` for a 
+1-line display of the results.
 
 ```{r exhobbsfn, eval=TRUE}
-hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
+hobnlfb <- nlfb(start = stu1, resfn = hobbs.res, jacfn = hobbs.jac)
 pshort(hobnlfb) # use print(hobnlfb) for more detail
-hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnls(hobnlm)  
-hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
+hobnlm <- nls.lm(par = stu1, fn = hobbs.res, jac = hobbs.jac)
+pnlslm(hobnlm)  
+hobgsln <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12))
 pnls(hobgsln)
-hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
+hobgsl <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12), jac = hobbs.jac)
 pnls(hobgsl) # using analytic Jacobian
@@
 the `nlsr` package has the function `wrapnlsr()`, for which `nlsr()` is an alias,
-which is intended for problems specified as formulas.
+that is intended for problems specified as formulas.
 This uses `nlxb()` to find good parameters, then calls `nls()` to return the class 
@@
 (weighted) residuals to `e4` times the initial sum of squares, where
-`e4 <- (100*.Machine$double.eps)^4` is approximately 2.43e-55.
+`e4 <- (100 * .Machine$double.eps) ^ 4` is approximately 2.43e-55.
 
@@
 `nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
-examination of the code. We terminate execution when the number of residual or 
-Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the 
-default `maxiter = 50` of `nls()` to avoid stopping early.
+examination of the code. `nlsr` terminates when the number of residual or 
+Jacobian evaluations (`feval` and `jeval`) exceed set limits, and we prefer larger 
+limits (`jemax = 5000` is the default) than the `maxiter = 50` of `nls()` to avoid 
+stopping early. This may cause overly long run-times for `nlsr`.
 
@@
 
-The complexity of this object is a challenge to users. Let us use the
-`result` as the returned object from `nls()`, `minpack.lm::nlsLM()` or,
+The complexity of this object is a challenge to users. Let us use 
+`result` as the name of the returned object from `nls()`, `minpack.lm::nlsLM()` or,
 `gslnls::gsl_nls()`. For example, the `data` return element is an R symbol. 
@@
 ```
-eval(parse(text=result$data))
-```
-However, since the evaluating environment might not exist (thanks to a referee
-for noting this possibility) a better approach is to make the call to `nls()` 
-with `model=TRUE`, which returns an element 
-`model` which contains the data, and we can list its contents using:
+eval(parse(text = result$data))
+```
+However, since the evaluating environment might not exist, a better approach 
+is to make the call to `nls()` 
+with `model = TRUE`, which returns an element 
+`model` which contains the data, and we can list its contents using
 ```
@@
 Thus the code in `nlsr` is likely of value for workers to copy and modify
-for customized tools. Coding in R is intended to support such uses.
+for customized tools, with code in R to support such uses.
 
@@
 options for the Jacobian, with a further control `ndstep` for the size of the
-step used in the approximation.
+step used in the approximation. These features are mainly to allow for investigation
+of approximate derivatives. Regular users will rarely need them.
 
@@
   the derivative is undefined, but the model can be computed. In such cases, 
-  we observed that `nls()` and 
-  `minpack.lm::nlsLM` found a solution, though this seems to be a lucky accident.
+  we observed that `nls()` and `minpack.lm::nlsLM` found a solution through the 
+  lucky accident that the approximate derivative can be evaluated.
 
@@
 There are also some in the now-archived package `NRAIA`.
-These provide the Jacobian in the `gradient` attribute of the "one-sided" formula
+The Jacobian is provided via the `gradient` attribute of the "one-sided" formula
 that defines each model, and these Jacobians are often the analytic forms.
@@
 algorithm which converges in one iteration and builds the standard output. 
-We found the implicit double call a challenge to unravel. 
+We found the implicit double-call a challenge to unravel. 
 A more conventional approach to starting parameters, following the lead of @Ratkowsky1983, 
@@
 This does not appear to be possible with `nls()`. `nlsr::nlxb()` always requires 
-starting parameters, and can either use `getInitial()` to find them from the `selfStart`
-model or provide explicit values, but the `selfStart` model is used in the formula 
-for the `nlxb()` call.
+starting parameters. We could get these by separately calling `getInitial()` to 
+find them from the `selfStart` model, but could use other values. The `selfStart` 
+model is also used in the formula for the `nlxb()` call.
 
@@
 function and others save quantities by default in "hidden" variables, i.e., with 
-names preceded by ".".  A referee has indicated this behaviour is to avoid name 
+names preceded by ".".  This behaviour is to avoid name 
 conflicts with user quantities. Hidden variables are not displayed by the `ls()` 
-command, so users need to remember to use `ls(all.names=TRUE)` to see which quantities
+command, so users need to remember to use `ls(all.names = TRUE)` to see which quantities
 are defined, for example, in developing their own selfStart model via copy and edit.
@@
 
-- `nls()` can only impose bounds if the `algorithm="port"` argument is 
+- `nls()` can only impose bounds if the `algorithm = "port"` argument is 
 used in the call. Unfortunately, the documentation warns us:
@@
 quite scarce. Traditionally, workers transform variables to ensure bounds are
-satisfied, such as squaring a parameter, or using a parameter which is the log()
+satisfied, such as squaring a parameter, or using a parameter which is the $\log()$
 of the actual parameter, so that its exponential is positive. These transformations
@@
 bump along the constraint in many cycles. Our aim is to avoid rubbish, not
-specify where the solution will be.
+specify where the solution will be, and we hope the solution is complicated
+by having a parameter on a bound.
 
@@
 # Start MUST be feasible i.e. on or within bounds
-anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
-             upper=c(2,6,3), algorithm='port')
+anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
+             upper = c(2,6,3), algorithm = 'port')
 pnls(anlshob1b) #  check the answer (short form)
 # nlsLM seems not to work with bounds in this example
-anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pnls(anlsLM1b)
 # also no warning if starting out of bounds, but gets a good answer!!
-st4<-c(c1=4, c2=4, c3=4)
-anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+st4<-c(c1 = 4, c2 = 4, c3 = 4)
+anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pnls(anlsLMob)
 # Try nlsr::nlxb()
-anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pshort(anlx1b)
@@
 # Hobbsmaskx.R -- masks with formula specification of the problem
-require(nlsr); require(minpack.lm); traceval<-FALSE
-stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
-sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
+require(nlsr); require(minpack.lm); traceval <- FALSE
+stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
+sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
 # fix first parameter
-anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, 3), trace=traceval))
+anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+			upper = c(200, 60, 3), trace=traceval))
 print(anxbmsk1)
-anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, 3), trace=traceval))
+anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+			upper = c(200, 60, 3), trace=traceval))
 pnls(anlM1)
-anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
+anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+		upper = c(200, 60, 3),  algorithm="port", trace = traceval))
 pnls(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
-anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                  upper=c(2,6,30))
+anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                  upper = c(2,6,30))
 print(anlxmsks1)
-anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
-             upper=c(2,6,30), algorithm='port')
+anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, lower = c(2,0,0),
+             upper = c(2,6,30), algorithm = 'port')
 pnls(anlshmsk1)
-anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                 upper=c(2,6,30))
+anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                 upper = c(2,6,30))
 pnls(anlsLMmsks1)
@@
 # Test with all parameters masked
-anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+anlxmskall<- try(nlxb(frms, start = sts, data = weeddf, lower = sts, upper = sts))
 print(anlxmskall)
@@
 solutions in methods of the Gauss-Newton family, as `nls()` terminates 
-too frequently and unnecessarily with `singular gradient` errors.
-
-### Programming language
+too frequently and unnecessarily with `singular gradient` errors. Successful
+incorporation of Marquardt stabilization into `nls()` would be a major improvement.
+
+## Programming language
 
@@
 programming language. `nls()` uses a mix of R, C, and Fortran, as does `minpack.lm`. 
-`gslnls` is an R wrapper to various C-language routines in the Gnu Scientific 
+`gslnls` is an R wrapper to various C-language routines in the GNU Scientific 
 Library (@GSL-manual).
@@
 smaller performance cost than the aggressive approach to a solution,
-which can cause more iterations to be used.
-
-One of the referees pointed out that nlsr::nlfb() is quite slow on a
-"large" case of the Penalty 1 test problem of @More1981TUO. We agree. 
-However, we note that problems in large numbers of parameters
-will be problematic for Gauss-Newton methods because they create large
-computational matrices for the Jacobian or its inner product. This was
-confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
+such as large computational limits for residuals and Jacobians.
+
+There are problems where use of R will be slower. For example, a
+"large" case of the Penalty 1 test problem of @More1981TUO will take
+noticeable time to solve. Problems in large numbers of parameters
+create large computational matrices for the Jacobian or its inner product,
+which Gauss-Newton methods create and process. 
+This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
 the Penalty 1 problem of various sizes, where the latter, which is 
 compiled Fortran internally, was moderately faster. However, the method
-"Rcgmin" from package `optimx` (@p-optimx) can be much 
-faster, despite being entirely written in R, and is a much more
+"Rcgmin" from package `optimx` (@p-optimx) proved to be much 
+faster than either, despite being entirely written in R, and is a much more
 reliable conjugate gradient minimizer than method "CG" of `optim()`. One
@@
 Timings of CG methods are more sensitive to small differences in
-setup and arithmetic than Gauss-Newton methods.
+setup and arithmetic than Gauss-Newton methods, and the pattern of
+such timings varies much less regularly with problem size for CG methods.
 
@@
 `nls()` and other class `nls` tools accept an argument `subset`. This acts through 
-the mediation of `model.frame`, which not obvious in the 
+the mediation of `model.frame`, which is not obvious in the 
 source code files `/src/library/stats/R/nls.R` and `/src/library/stats/src/nls.C`. 
@@
 As with `subset`, our concern with `na.action` is that users may be unaware of the
-effects of an option they may not even be aware has been set. Should `na.fail` be the default?
-
-## Feature: model frame
+effects of an option they may not even know has been set. Should `na.fail` be the default?
+
+## Model frame
 
@@
 
-> **model** logical. If true, the model frame is returned as part of the object. Default is FALSE.
+  ***model** logical. If true, the model frame is returned as part of the object. Default is FALSE.*
 
@@
 Where available, the values returned by the
-`residuals()` function are weighted, and the `fitted()` or `predict()` function are 
+`residuals()` function are weighted, and the `fitted()` or `predict()` functions are 
 used to compute raw residuals.
@@
 we want weights that are determined at least partially from the model 
-parameters, for
+parameters; for
 example, a measure of the standard deviation of observations.
@@
 ```{r nlswtx, echo=TRUE}
-wts <- 0.5^tt # simple weights
-frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
-Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # unweighted
-rnowt<-nowt$m$resid() # This has unweighted residual and Jacobian. Does not take coefficients.
-attr(rnowt, "gradient")<-NULL; rnowt
+wts <- 0.5 ^ tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt) / scal))
+Asym <- 1; xmid <- 1; scal <- 1
+nowt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # unweighted
+rnowt <- nowt$m$resid() # This has unweighted residual and Jacobian. Does not take coefficients.
+attr(rnowt, "gradient") <- NULL; rnowt
 usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-rusewt<-usewt$m$resid() # weighted. Does not take coefficients.
-attr(rusewt,"gradient")<-NULL; rusewt
+rusewt <- usewt$m$resid() # weighted. Does not take coefficients.
+attr(rusewt,"gradient") <- NULL; rusewt
 source("nlsModel.R")
-nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
+nmod0 <- nlsModel(frmlogis, data = weeddf, start = c(Asym = 1, xmid = 1, scal = 1), wts = wts)
 rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0
-nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
-rn<-nmod$resid()
-attr(rn,"gradient")<-NULL; rn
+attr(rn0,"gradient") <- NULL; rn0
+nmod <- nlsModel(frmlogis, data = weeddf, start = coef(usewt), wts = wts)
+rn <- nmod$resid()
+attr(rn,"gradient") <- NULL; rn
 ```
@@
 where the relative offset convergence criterion is met, it may simply exit,
-especially if a "singular gradient" (singular Jacobian) is found. However, 
+especially if a "singular gradient" is found. However, 
 this may occur **after** considerable progress has been made in reducing 
@@
 ```{r tetrarun}
-time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
-conc <- c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
+time <- c(1,  2,  3,  4,  6 , 8, 10, 12, 16)
+conc <- c(0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
 NLSdata <- data.frame(time,conc)
@@
 are linear and communicating this information to estimating functions 
-is not a trivial task. `nls()` has an option `algorithm="plinear"` that allows some partially linear
+is not a trivial task. `nls()` has an option `algorithm = "plinear"` that allows some partially linear
 models to be solved. The other tools, as far as we are aware, do not offer any
@@
 different specifications with different `algorithm` options.
-For example, the explicit model `y ~ a * x + b` does not work with the linear
-modeling function `lm()`, which requires this model to be specified as `y ~ x`.
-Within `nls()`, consider the following FOUR different specifications for the same 
+For example, the explicit linear model `y ~ a * x + b` does not work. As with 
+the linear modeling function `lm()`, this must be specified as `y ~ x`.
+Within `nls()`, consider the following four different specifications for the same 
 problem, plus an intuitive choice, labeled `fm2a`, that does not work. 
@@
 to try to add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing, we do
-not foresee a resolution for this issue. In the example, we have not evaluated
+like to see a consistent syntax. We do not foresee a resolution for this issue, which
+arises from different approaches to model specification. In the example, we have not evaluated
 the commands to save space.
@@
 
-<!-- Ben Bolker mentioned a possibility of using a nonlinear model in  -->
-<!-- ??I think?? nlme with some sort of linear model form of the parameters. -->
-<!-- May need to mention. -->
-
-<!-- ===```{r nlsindx1} -->
-<!-- ## The muscle dataset in MASS is from an experiment on muscle -->
-<!-- ## contraction on 21 animals.  The observed variables are Strip -->
-<!-- ## (identifier of muscle), Conc (Cacl concentration) and Length -->
-<!-- ## (resulting length of muscle section). -->
-<!-- if(! requireNamespace("MASS", quietly = TRUE)) stop("Need MASS pkg") -->
-<!-- mm<- MASS::muscle[1:12,] # take only 1st few values of Strip (TRUNCATION OF EXAMPLE) -->
-<!-- mm<-droplevels(mm) # remove unused levels after truncation -->
-<!-- nlev <- nlevels(mm) -->
-<!-- withAutoprint({ -->
-<!--   ## The non linear model considered is -->
-<!--   ##       Length = alpha + beta*exp(-Conc/theta) + error -->
-<!--   ## where theta is constant. For now alpha and beta do not vary with Strip. -->
-<!--   with(mm, table(Strip)) # 2, 3 or 4 obs per strip -->
-<!--   nl <- nlevels(mm$Strip) -->
-<!--   ## We first use the plinear algorithm to fit an overall model, -->
-<!--   ## ignoring that alpha and beta might vary with Strip. -->
-<!--   musc.1 <- nls(Length ~ cbind(1, exp(-Conc/th)), mm, -->
-<!--                 start = list(th = 1), algorithm = "plinear") -->
-<!--   summary(musc.1) -->
-
-<!--   ## Then we use nls' indexing feature for parameters in non-linear -->
-<!--   ## models to use the conventional algorithm to fit a model in which -->
-<!--   ## alpha and beta vary with Strip.  The starting values are provided -->
-<!--   ## by the previously fitted model. -->
-<!--   ## Note that with indexed parameters, the starting values must be -->
-<!--   ## given in a list (with names): -->
-<!--   b <- coef(musc.1) -->
-<!--   musc.2 <- nls(Length ~ a[Strip] + b[Strip]*exp(-Conc/th), data=mm, -->
-<!--                 start = list(a = rep(b[2], nl), b = rep(b[3], nl), th = b[1])) -->
-<!--   summary(musc.2) -->
-<!-- }) -->
-<!-- === ``` -->
+## Some other CRAN packages for nonlinear modeling
+
+`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
+is the vertical difference between the dependent variable and the functional
+model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
+therefore different and involves nontrivial extra calculation. A vignette with
+the package and the blog article 
+https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
+give some description with illustrative graphs. `onls` appears to be limited to 
+problems with one independent and one dependent variable. The Wikipedia article
+https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
+ideas, with references to the literature. The approach needs a wider 
+discussion and tutorial examples to allow its merits to be judged than can be
+included here.
+
+`crsnls` (@R-crsnls-manual) -- This package allows nonlinear estimation by 
+controlled random search via two methods. There is unfortunately no vignette.
+A modest trial we carried out showed `nlsr::nlxb()` gave the same results in 
+a small fraction of the time required by either of the methods in `crsnls`. The
+method discussed in @tvrdik2007 claims better reliability in finding solutions
+than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
+on the extreme NIST examples mentioned next.
+
+`NISTnls` (@NISTnls-manual) -- This package provides R code and data for a set of
+(numerically ill-conditioned) nonlinear least squares problems from the U.S. National 
+Institute for Standards and Technology. These may not represent real-world situations.
+
+`nlshelper` (@nlshelper-manual) -- This package, which unfortunately lacks a
+vignette, provides a few utilities for summarizing, testing, and plotting 
+non-linear regression models estimated with nls(), nlsList() or nlme() that
+are linked or grouped in some way.
+
+`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
+equality and/or inequality constraints. It is clearly **not** about modeling, and
+the input and output are quite different from class `nls` methods. However, there do
+not appear to be other R packages with these capabilities.
+
+`nlsMicrobio` (@nlsMicrobio-manual) -- Data sets and nonlinear regression models 
+dedicated  to predictive microbiology, including a vignette, by authors of the
+`nlstools` package.
+
+`nlstools` (@p-nlstools) -- This package provides several tools for aiding the estimation 
+of nonlinear models, particularly using `nls()`. The vignette is actually a journal
+article,  and the authors have considerable experience in the subject.
+
+`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
+regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem 
+here is to minimize an objective that is modified from the traditional sum of squared
+residuals.
+
+`nls.multstart` (@nlsmultstart) -- Non-linear least squares regression using
+AIC scores with the Levenberg-Marquardt algorithm using multiple starting values 
+for increasing the chance that the minimum found is the global minimum.
+
+`nls2` (@nls2-22) -- Nonlinear least squares by brute force has similar
+motivations to `nls.multstart`, but uses `nls()` within multiple trials. The author
+has extensive expertise in R.
+
+`nlstac` (@m-nlstac) -- A set of functions implementing the algorithm described in
+Fernandez Torvisco et al. (2018) for fitting separable nonlinear regression curves. The 
+special class of problem for which this package is intended is an important and 
+difficult one. No vignette is provided, unfortunately.
+
+`easynls`  -- Fit and plot some nonlinear models. Thirteen models are treated, but 
+there is minimal documentation and no vignette. Package `nlraa` is to be preferred.
+
+`nlraa`  (@MiguezNLRAA2021) -- a set of nonlinear *selfStart* models, primarily from
+agriculture. Most include analytic Jacobian code.
+
+`optimx` (@p-optimx) -- This provides 
+optimizers that can be applied to minimize a nonlinear function which could be
+a nonlinear sum of squares. This is not generally recommended if nonlinear 
+least squares programs can be easily used, but provides a check and alternative
+solvers.
 
@@
 Such goals align with the aims of **unit testing** 
-(e.g., *https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211*, @HWtestthat11, @HWdevtools21
-and the conventional R package testing tools).
+(@HWtestthat11, @HWdevtools21, the conventional R package testing tools, and 
+a number of blog posts found by searching "unit testing in R").
 In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
@@
 a compact output. The prototype has a vignette to illustrate its use.
-
-## Documentation and resources
-
-In our investigation, we built several resources, which are now part of 
-the repository *https://github.com/nashjc/RNonlinearLS/*. Particular items
-are:
-
-- A BibTex bibliography for use with all documents in this project, but which 
-  has wider application to nonlinear least squares projects in general
-  (*https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib*).
-
-- `MachID.R` offers a suggested concise summary function
-  to identify a particular computational system used for tests. A discussion 
-  of how it was built and the resources needed across platforms is given in
-  at *https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary*.
-
-- @PkgFromRbase22 is an explanation of the construction of the `nlspkg` from the
-  `nls()` code in R-base.
-
-- As the 2021 Summer of Code period was ending, one of us (JN) was invited to prepare
-  a review of optimization in R. Ideas from the present work have been instrumental
-  in the creation of @NashWires22.
-
-<!-- ## Should nls() be refactored? -->
-
-<!-- When we started *Improvements to nls()* for the 2021 Google Summer of Code we wanted: -->
-
-<!-- - to provide a packaged version of `nls()` (call it `nlsalt`) coded entirely in R  -->
-<!--   that matches the features in base R or what is packaged in `nlspkg` as described  -->
-<!--   in @PkgFromRbase22. Our intent in providing a purely R version of the code was -->
-<!--   either to replace `nls()` or to serve as a reference version aiding maintenance  -->
-<!--   of C, Fortran, or other components of the software. -->
-
-<!-- - to streamline the overall `nls()` infrastructure. By this we mean a  -->
-<!--   refactoring of the routines so they are better suited to maintenance of both -->
-<!--   the existing `nls()` methods and features as well as new features we or others -->
-<!--   would like to add. -->
-
-<!-- - to explain what we do, either in comments, examples, or separate maintainer  -->
-<!--   documentation. While R and its packages are generally very well documented for -->
-<!--   usage, that documentation -->
-<!--   rarely extends as far as it might for code maintenance. The paucity of such -->
-<!--   documentation is exacerbated by the mixed R/C/Fortran code base, -->
-<!--   where seemingly trivial differences in things like representation of numbers or -->
-<!--   base index for arrays lead to traps for unwary programmers. -->
-
-<!-- These goals echo themes in the recent discussion by @Vidoni21 and accompanying  -->
-<!-- commentary. -->
-
-<!-- We found the legacy code and examples prevented us achieving a revised `nls()`  -->
-<!-- with all its current capabilities. Unfortunately, `nls()` is part of the base R -->
-<!-- distribution,  -->
-<!-- so users are drawn to using it even when it is less than ideal. A possible way forward -->
-<!-- is to  -->
-
-<!-- - use a packaged version of `nls()` to provide legacy features, with notification -->
-<!--   to users that there are other tools available -->
-
-<!-- - if maintainers are willing, to merge and rationalize the capabilities of  -->
-<!--   `minpack.lm`, `nlsr` and `gslnls` -->
-
-<!-- - to encourage the (separate) development of capabilities such as indexed -->
-<!--   parameter models if these are in demand. -->
-
-<!-- This direction does not, of course, resolve the inconsistencies in model  -->
-<!-- specification in formulas for nonlinear models, which undoubtedly is an issue -->
-<!-- for other modeling tools.   -->
-
-<!-- A possible workaround to allow consistency for the `plinear` algorithm  -->
-<!-- would be to specify the linear parameters when the `plinear` option is specified.  -->
-<!-- For example, we could use `algorithm="plinear(Asym)"` while keeping the general  -->
-<!-- model formula from the default specification. This would allow  -->
-<!-- for the output of different `algorithm` options to be consistent. However, we  -->
-<!-- have not yet tried to code such a change, and welcome discussion from those working -->
-<!-- with other packages using model formulas. -->
-
-<!-- A further complication of model specification in R is that some formulas require -->
-<!-- the user to surround a term with `I()` to inhibit interpretation of arithmetic  -->
-<!-- operators. We would prefer that formulas be usable without this complication. -->
 
@@
 vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
-attempts to succinctly summarize computing environments.
+attempts to succinctly summarize computing environments. Particularly thoughtful
+reviews by the referees, including details of inner workings of R, 
+have led to important revisions.
 
 # References
-
-<!-- ------------------------------ -->
-
-<!-- ## Comparison dimensions -->
-
-<!-- The tools for nonlinear estimation we consider have been built with different -->
-<!-- goals and contexts. We will make our comparison considering the overlapping -->
-<!-- headings summarized below. -->
-
-<!-- ### Modeling versus minimizing a sum of squares -->
-
-<!-- The goal of many statisticians and other workers is to estimate a model. The -->
-<!-- minimization of a sum of squared residuals -- the nonlinear least squares problem -- -->
-<!-- is the mechanism to estimate the model. Some tools focus on modeling,  -->
-<!-- so attempt to provide measures of uncertainty and other information.  -->
-<!-- The interface to modeling tools will generally try to streamline that task. By -->
-<!-- contrast, nonlinear least squares tools are trying to solve the more limited -->
-<!-- numerical task. -->
-
-<!-- ### Programming language(s) -->
-
-<!-- R tools are our interest, but they often call programs in other programming -->
-<!-- languages. There are advantages in using a well-established program that has -->
-<!-- been used and tested. Early versions of R were relatively sluggish in  -->
-<!-- computational speed. The downside is that we must interface between two or more -->
-<!-- programming languages, adding to the risk of errors. Changes to one or the  -->
-<!-- other language or its library support may -->
-<!-- force maintenance issues. Moreover, the pool of expertise with languages like C -->
-<!-- and Fortran is diminishing. -->
-
-<!-- ### Internal solver stabilization -->
-
-<!-- Early nonlinear least squares programs generally used the Gauss-Newton method  -->
-<!-- (@Hartley1961). This is often very effective and efficient, but for a quite large  -->
-<!-- proportion of problems can fail to find the solution. Stabilization of the  -->
-<!-- computations is relatively straightforward in principle, and overcomes most -->
-<!-- of the failures, though sometimes at the cost of more calculations. -->
-
-<!-- ### Jacobian calculation -->
-
-<!-- The necessarily iterative solution of a nonlinear least squares problem  -->
-<!-- to carry out nonlinear modeling requires that we can find smaller values -->
-<!-- of a sum of squared residuals. This implies there is a "downhill" direction. -->
-<!-- The gradient of the residual sum of squares is computed as its product -->
-<!-- with the Jacobian matrix: the partial derivatives of the residual vector  -->
-<!-- with respect to the modeling parameters. -->
-
-<!-- Most researchers are challenged to program the differential calculus  -->
-<!-- required to specify the Jacobian to a nonlinear estimation tool.  -->
-<!-- The workaround is generally to use a numerical approximation, -->
-<!-- but there is the possibility of using symbolic or automatic differentiation. -->
-<!-- There are also particular models where workers have provided code for the -->
-<!-- Jacobian. -->
-
-<!-- ### Algorithm and termination choices -->
-
-<!-- Iterative algorithms need to be stopped at some point, and how termination -->
-<!-- is forced may alter results. Furthermore, even supposedly similar algorithms -->
-<!-- may differ greatly in detail.  -->
-
-<!-- ### Constraints on parameters -->
-
-<!-- Reality imposes conditions on parameters. While general constraints can -->
-<!-- be difficult to include, bounds on parameters can be accommodated by many  -->
-<!-- solvers. For some reason, fixed parameters (masks) are not often considered -->
-<!-- but are convenient when a parameter has a commonly accepted value that we -->
-<!-- later wish to allow to be estimated from the data. -->
-
-<!-- ### Special capabilities -->
-
-<!-- Users often have particular problems that introduce other requirements. It is -->
-<!-- common to weight observations where their reliability is thought to differ. -->
-<!-- Some situations present a number of related situations where a combined model -->
-<!-- with some common and some local parameters are present, and the local parameters -->
-<!-- could be indexed. The common exponential decay and related **partially linear** -->
-<!-- models (sometimes called **separable** models) is a difficult one that can  -->
-<!-- benefit from special treatment.  -->
-
-<!-- ### Output  -->
-
-<!-- While nonlinear estimation aims to return a set of parameters for a -->
-<!-- particular model, there is a variety of additional information which may be -->
-<!-- provided. From the perspective of the nonlinear least squares problem, there -->
-<!-- are measures of work and information on the quality of the solution, while -->
-<!-- for statistical purposes measures of uncertainty in the parameters are helpful. -->
-
-<!-- Some tools allow constraints to be imposed on the solution, and it is helpful -->
-<!-- to know if these constraints are active in the proposed solution. -->
-
-<!-- ### Documentation -->
-
-<!-- We believe that the availability of good documentation with examples that users -->
-<!-- can modify to deal with their problems is essential. The presence of vignettes,  -->
-<!-- demos, and tests indicates a package is well-supported. -->

### 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
notes from AB

@@
   pdf_document:
-    keep_tex: yes
-    toc: no
+    pandoc_args: --listings
+    includes:
+      in_header: preamble.tex
   word_document:
@@
 $$ H \delta  = -g $$
-Given a set of parameters $p$, we solve this equation for $\delta$, adjust $p$
+Given a set of parameters $p$, which provide $H$ and $g$, we solve this equation for $\delta$, adjust $p$
 to $p + \delta$ and iterate, hopefully to converge on a solution. 
@@
 This is confirmed by comparing `nlsr::nlfb()` and `minpack.lm::nls.lm()` on
-the Penalty 1 problem of various sizes, where the latter, which is 
+the Penalty 1 problem of various sizes, where the latter, which uses 
 compiled Fortran internally, was moderately faster. However, the method

### 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
fixes as per AB

@@
   pdf_document:
-    pandoc_args: --listings
-    includes:
-      in_header: preamble.tex
+    keep_tex: yes
+    toc: no
   word_document:
@@
 tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
-print(tryit)
+# print(tryit) # gives no information on best sumsquares or model parameters
 ```

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/ReviewResponses.txt
Prep for resubmission

--- 2023-07-11 RefactoringNLS/revision202307/ReviewResponses.txt
@@
-Arkajyoti responses -- delete before submission:
+=========================================================================
+Revew Responses for R Journal paper 2023-15.
 
-I have read through your comments on the suggestions of the two referees and I agree with them. As usual, the document is well-organized, methodical, and nice way to check all the boxes.
-
-For “Response 2:10“, I don’t think any link would be a good choice based on the referees’ concern.
-An alternative article (maybe the reference manual or vignette of the corresponding package  in question, like for unit testing) may serve as a viable replacement.
-    Agree that no link is likely to be satisfactory. Modified to suggest search to find
-    articles.
-
-Some comments:
-
-    Maybe cite your nlsr package, as “our” nlsr seems confusing? -- DONE
-
-    There is some overlap between the Abstract and scope of our comparison, and I suggest making the 
-        latter more succinct (for example the paragraph before TOOLS CONSIDERED)?
-        DONE -- simplified.
-
-    Maybe talking about referees’ suggestions can go in the acknowledgements? DONE
-
-    I suggest common website links, like github.com, gitlab.com and cran be removed?
-        Agreed.
-
-    “Some other CRAN  packages…” can go later, after the main material of the paper?
-        DONE -- before unit tests paragraph.
-
-    Documentation and resources seem tangential to the main material and maybe removed?
-        DONE -- this was when we were talking more about the GSoC project.
-
-    Maybe add 5, 6 above in a separate r blog post or the main article’s supplementary article?
-       ??? What supplementary article? If we want to reference blog posts, they should
-       be available now. Perhaps set aside and focus on getting this published. Thoughts?
-
-=========================================================================
-Revew Responses to first review of R Journal paper 2023-15.
-
-We have created a text file with the content of both referees'
+We have created this text file with the content of both referees'
 reports and edited our responses to comments under headings
@@
 where X indicates the referee (1 or 2) and YY is an identifying
-number.
+number. If X=0, the response is something we felt should be corrected.
 
@@
 which topics are relevant to that purpose.
+
+Please see Response 2:02 which responds to a similar concern.
 
@@
 Response 1:02
-     This overlaps with comments by the second referee. See Response 2:01.
+     This overlaps with comments by the second referee. See Response 2:02.
 
@@
 
-      Interim response: We have reworded the paragraph starting
+      Interim action: We have reworded the paragraph starting
 	"In considering `selfStart` models,"
@@
 
-Response 1:05 The historical context could be very useful to the R community.
+I hope that this unconventional referee’s report is of use to the authors 
+and the editor.
+
+Response 1:05 The historical context given in the referee's report could 
+    be very useful to the R community.
     Would the editors consider asking that Referee 1 consider recasting
@@
 
-    See also 2:01 for clarification of purpose.
-
-I hope that this unconventional referee’s report is of use to the authors 
-and the editor.
+    See also 2:01 and 2:02 for clarification of purpose.
 
@@
         message is that we obtain diagnostic information in the form of the 
-        singular values of the Jacobian that allow us to try alternative starts, 
+        singular values of the Jacobian that allows us to try alternative starts, 
         algorithms or programs in cases where there are indications of early termination.
@@
 
-    The referee's commends and Illustrating Code led us to try the Pen1 problem
+    The referee's comments and Illustrating Code led us to try the Pen1 problem
     (it is part of the funconstrain package https://github.com/jlmelville/funconstrain
@@
 
-Response 2:09  Fair enought. Done.
+Response 2:09  Fair enough. Done.
 

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.Rmd
Prep for resubmission

--- 2023-07-15 RefactoringNLS/revision202307/StateOfRToolsForNLS.Rmd
@@
   Institute of Technology, Kanpur}
-date: "2023-3-2"
+date: "2023-7-16"
 output: 
@@
 Further considerations are the usability and  maintainability of
-the code base that provides the functionality `nls()` claims to offer.
+the code base that provides the functionality `nls()` offers.
 Various packages, in particular `nlsr` (@nlsr2023manual), provide alternative capabilities.
@@
 Besides the base-R `nls()` function, we consider some packages that 
-are available on the CRAN repository. Of these, we will pay particular 
+are available in the CRAN repository. Of these, we will pay particular 
 attention to **nlsr** (@nlsr2023manual),
@@
 limits (`jemax = 5000` is the default) than the `maxiter = 50` of `nls()` to avoid 
-stopping early. This may cause overly long run-times for `nlsr`.
+stopping early. This may cause overly long run-times for `nlsr`, but our work often
+involves problems that are sufficiently difficult that we want to be sure we do not
+terminate early.
 
@@
 bump along the constraint in many cycles. Our aim is to avoid rubbish, not
-specify where the solution will be, and we hope the solution is complicated
+specify where the solution will be, and we hope the solution is not complicated
 by having a parameter on a bound.
@@
 offers an accelerated Marquardt method and three alternative methods which we
-have not investigated.
+have not investigated beyond one example.
 Control settings for `nlxb()` or `nlfb()` allow exploration of Hartley and Marquardt 
-algorithm variants. In general, the Levenberg-Marquardt stabilization 
+algorithm variants, principally for the benefit of program developers. 
+In general, the Levenberg-Marquardt stabilization 
 is important in obtaining 
 solutions in methods of the Gauss-Newton family, as `nls()` terminates 
-too frequently and unnecessarily with `singular gradient` errors. Successful
+too frequently and unnecessarily with `singular gradient` errors. A tidy
 incorporation of Marquardt stabilization into `nls()` would be a major improvement.
@@
 
-There are problems where use of R will be slower. For example, a
+There are, of course, problems where use of R code will be slower. For example, a
 "large" case of the Penalty 1 test problem of @More1981TUO will take
@@
 a number of blog posts found by searching "unit testing in R").
-In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
+In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare*.
 A primary design objective of this is to allow the summarisation of multiple tests in

### 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.tex
Prep for resubmission

--- 2023-05-14 RefactoringNLS/ComparingNLStools.tex
@@
 \usepackage{amsmath,amssymb}
-\usepackage{lmodern}
 \usepackage{iftex}
@@
 \else % if luatex or xetex
-  \usepackage{unicode-math}
+  \usepackage{unicode-math} % this also loads fontspec
   \defaultfontfeatures{Scale=MatchLowercase}
   \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
+\fi
+\usepackage{lmodern}
+\ifPDFTeX\else
+  % xetex/luatex font selection
 \fi
@@
 \newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
+\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
 \newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
@@
 \newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
 \newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
@@
 \newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
 \newcommand{\ImportTok}[1]{#1}
@@
 \newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
+\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
 \newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
@@
 \IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same} % disable monospaced font for URLs
+\urlstyle{same}
 \hypersetup{
-  pdftitle={A comparison of R tools for nonlinear least squares modeling},
+  pdftitle={The Current State of R Tools for Nonlinear Least Squares Modeling},
   pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
@@
 
-\title{A comparison of R tools for nonlinear least squares modeling}
+\title{The Current State of R Tools for Nonlinear Least Squares
+Modeling}
 \author{John C. Nash
@@
 \thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
-\date{2023-3-2}
+\date{2023-7-16}
 
@@
 maintainability of the code base that provides the functionality
-\texttt{nls()} claims to offer. Various packages, including our
-\texttt{nlsr}, provide alternative capabilities. We consider the
-differences in goals, approaches, and features of different tools for
-nonlinear least squares modeling in R. Discussion of these matters is
-relevant to improving R generally as well as its nonlinear estimation
-tools.
+\texttt{nls()} offers. Various packages, in particular \texttt{nlsr}
+(John C Nash and Duncan Murdoch
+(\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative
+capabilities. We consider the differences in goals, approaches, and
+features of different tools for nonlinear least squares modeling in R.
+Discussion of these matters is relevant to improving R generally as well
+as its nonlinear estimation tools.
 
@@
 in base R, the primary software distribution from the Comprehensive R
-Archive Network (\emph{\url{https://cran.r-project.org}}). The function
-dates to the 1980s and the work related to Bates and Watts
-(\protect\hyperlink{ref-bateswatts}{1988}) in S (see
-\emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
+Archive Network (CRAN). Users do not have to request that \texttt{nls()}
+be loaded. The function dates to the 1980s and the work related to D. M.
+Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) and others in
+the S programming language.
 
@@
   handles formulas that include R functions, even ones that call
-  calculations in other programming languages
+  calculations in other programming languages;
 \item
-  allows data to be weighted or subset
+  allows data to be weighted or subset;
 \item
-  can estimate bound-constrained parameters
+  can estimate bound-constrained parameters;
 \item
-  provides a mechanism for handling partially linear models
+  provides a mechanism for handling partially linear models;
 \item
-  permits parameters to be indexed over a set of related data
+  permits parameters to be indexed over a set of related data;
 \item
   produces measures of variability (i.e., standard error estimates) for
-  the estimated parameters
+  the estimated parameters;
 \item
   has related profiling capabilities for exploring the likelihood
-  surface as parameters are changed
+  surface as parameters are changed;
 \item
@@
 
-The tools we will consider are the base-R \texttt{nls()} function and
-some packages that are available on the CRAN repository. Of these, we
-will pay particular attention to \textbf{nlsr} (John C Nash and Duncan
-Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
-\textbf{minpack.lm} (Elzhov et al.
-(\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau
-(\protect\hyperlink{ref-gslnls23}{2023})) which are general nonlinear
-least-squares solvers.
-
-While we will provide capsule comments for some other CRAN packages, we
-will not pursue those in the Bioconductor (Gentleman et al.
+Our Google Summer of Code project for 2021, ``Improvements to
+\texttt{nls()}'' had the goal of providing a version of the code that
+would address some of the deficiencies discovered over the long period
+that \texttt{nls()} has been part of R. However, it was also critical to
+preserve legacy functionality, and we did not achieve a result that
+provided desired improvements and preserved legacy results
+simultaneously. On the other hand, working with \texttt{nls()} suggested
+ways package \texttt{nlsr} could be improved in features, while also
+revealing how it has a very different internal structure and design.
+
+The primary goal of this paper is to present some of these issues, with
+a limited consideration of some other R tools that exist for nonlinear
+least squares calculations. We hope to encourage others to engage,
+possibly with us, to find ways to bring improvements and to consolidate
+tools where reasonable. We want to ease the task of users in choosing
+which tools to use. Unfortunately, it is simpler to create new packages
+than to bring together and simplify existing ones.
+
+\hypertarget{tools-considered}{%
+\subsubsection{Tools considered}\label{tools-considered}}
+
+Besides the base-R \texttt{nls()} function, we consider some packages
+that are available in the CRAN repository. Of these, we will pay
+particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch
+(\protect\hyperlink{ref-nlsr2023manual}{2023})), \textbf{minpack.lm}
+(Elzhov et al. (\protect\hyperlink{ref-minpacklm12}{2012})), and
+\textbf{gslnls} (Chau (\protect\hyperlink{ref-gslnls23}{2023})) which
+are general nonlinear least-squares solvers. We will not pursue tools in
+the Bioconductor (Gentleman et al.
 (\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
-repositories such as GitHub (\url{https://github.com}) and Gitlab
-(\url{https://about.gitlab.com}).
-
-\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
-\subsection{Some other CRAN packages for nonlinear
-modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
-
-\texttt{onls} (Andrej-Nikolai Spiess
-(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
-in estimating nonlinear models is the vertical difference between the
-dependent variable and the functional model. \texttt{onls} minimizes the
-sum of \textbf{orthogonal} residuals. The objective is therefore
-different and involves nontrivial extra calculation. A vignette with the
-package and the blog article
-\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
-give some description with illustrative graphs. \texttt{onls} appears to
-be limited to problems with one independent and one dependent variable.
-The Wikipedia article
-\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
-overview of some ideas, with references to the literature. The approach
-needs a wider discussion and tutorial examples to allow its merits to be
-judged than can be included here.
-
-\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
--- This package allows nonlinear estimation by controlled random search
-via two methods. There is unfortunately no vignette. A modest trial we
-carried out showed \texttt{nlsr::nlxb()} gave the same results in a
-small fraction of the time required by either of the methods in
-\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
-Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
-reliability in finding solutions than a Levenberg-Marquardt code
-(actually from Matlab), but the tests were conducted on the extreme NIST
-examples mentioned next.
-
-\texttt{NISTnls} (National Institutes for Standards and R port by
-Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
-package provides R code and data for a set of (numerically
-ill-conditioned) nonlinear least squares problems from the U.S. National
-Institute for Standards and Technology. These may not represent
-real-world situations.
-
-\texttt{nlshelper} (Duursma
-(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
-unfortunately lacks a vignette, provides a few utilities for
-summarizing, testing, and plotting non-linear regression models
-estimated with nls(), nlsList() or nlme() that are linked or grouped in
-some way.
-
-\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
-This solves nonlinear least squares problems with optional equality
-and/or inequality constraints. It is clearly \textbf{not} about
-modeling, and the input and output are quite different from class
-\texttt{nls} methods. However, there do not appear to be other R
-packages with these capabilities.
-
-\texttt{nlsMicrobio} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
-nonlinear regression models dedicated to predictive microbiology,
-including a vignette, by authors of the \texttt{nlstools} package.
-
-\texttt{nlstools} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
-several tools for aiding the estimation of nonlinear models,
-particularly using \texttt{nls()}. The vignette is actually a journal
-article, and the authors have considerable experience in the subject.
-
-\texttt{nlsmsn} (Prates, Lachos, and Garay
-(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
-scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
-Lachos, and Abanto-Valle (2011). The problem here is to minimize an
-objective that is modified from the traditional sum of squared
-residuals.
-
-\texttt{nls.multstart} (Padfield and Matheson
-(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
-squares regression using AIC scores with the Levenberg-Marquardt
-algorithm using multiple starting values for increasing the chance that
-the minimum found is the global minimum.
-
-\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
-Nonlinear least squares by brute force has similar motivations to
-\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
-The author has extensive expertise in R.
-
-\texttt{nlstac} (Rodriguez-Arias et al.
-(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
-implementing the algorithm described in Fernandez Torvisco et al.~(2018)
-for fitting separable nonlinear regression curves. The special class of
-problem for which this package is intended is an important and difficult
-one. No vignette is provided, unfortunately.
-
-\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models
-are treated, but there is minimal documentation and no vignette. Package
-\texttt{nlraa} is to be preferred.
-
-\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
--- a set of nonlinear \emph{selfStart} models, primarily from
-agriculture. Most include analytic Jacobian code.
-
-\texttt{optimx} (John C. Nash and Varadhan
-(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
-that can be applied to minimize a nonlinear function which could be a
-nonlinear sum of squares. NOT generally recommended if nonlinear least
-squares programs can be easily used, but provides a check and
-alternative solvers.
+repositories such as GitHub and Gitlab.
 
@@
 (\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
-task that seems straightforward but is quite nasty. Its very succinct
+task that seems straightforward, but is quite nasty. Its very succinct
 statement provides the ``short reproducible example'' much requested on
-R mailing lists. This is a real problem from a field researcher. The
-data and graph follow.
+R mailing lists. This is a real problem from a field researcher, though
+the units of measurement and background information are not available.
+The data and graph follow.
 
@@
 
-\includegraphics{ComparingNLStools_files/figure-latex/ex01-1.pdf}
+\includegraphics{StateOfRToolsForNLS_files/figure-latex/ex01-1.pdf}
 
@@
 function is \[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\] The gradient of
-\(S(p)\) is \[ 2 * J' r\] where the Jacobian \(J\) is given by elements
-\[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is defined
-by elements
+\(S(p)\) is \[ g = 2 * J' r\] where the Jacobian \(J\) is given by
+elements \[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is
+defined by elements
 \[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \] If we
 expand the Hessian for nonlinear least squares problems, we find
-\[  H_{i,j} = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
+\[  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
 Let us use \(D_{i,j}\) for the elements of the second term of this
@@
 stationary point of the function \(S(p)\)). This leads to
-\textbf{Newton's equation} \[ H \delta = -g \] Given a set of parameters
-\(p\), we solve this equation for \(\delta\), adjust \(p\) to
-\(p + \delta\) and iterate, hopefully to converge on a solution.
-Applying this to a sum of squares problem gives
-\[ H \delta = (J'J + D) \delta = - J' r\] In this expression, only the
-elements of \(D\) have second partial derivatives. Gauss, attempting to
-model planetary orbits, had small residuals, and noted that these
+\textbf{Newton's equation} \[ H \delta  = -g \] Given a set of
+parameters \(p\), which provide \(H\) and \(g\), we solve this equation
+for \(\delta\), adjust \(p\) to \(p + \delta\) and iterate, hopefully to
+converge on a solution. Applying this to a sum of squares problem gives
+\[ H \delta / 2 = (J'J + D) \delta = - J' r\] In this expression, only
+the elements of \(D\) have second partial derivatives. Gauss, attempting
+to model planetary orbits, had small residuals, and noted that these
 multiplied the second partial derivatives of \(r\), so he approximated
-\[ H \approx J' J\] by assuming \(D \approx 0\). This results in the
+\[ H / 2 \approx J' J\] by assuming \(D \approx 0\). This results in the
 Gauss-Newton method where we solve \[ J' J \delta = - J' r\] though we
-can avoid some loss of accuracy by NOT forming the inner product matrix
+can avoid some loss of accuracy by not forming the inner product matrix
 \(J' J\) and solving the linear least squares matrix problem
 \[ J \delta \approx -r \] by one of several matrix decomposition
-methods.
-
-In reality, there are many problems were \(D\) should not be ignored,
+methods. On the other hand, when \(m >> n\), the matrix \(J' J\) uses
+much less storage than \(J\).
+
+In reality, there are many problems where \(D\) should not be ignored,
 but the work to compute it precisely is considerable. Many work-arounds
@@
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are NOT equivalent and are
+values for parameters. These starting points are not equivalent and are
 deliberately crude choices. Workers performing many calculations of a
@@
 \CommentTok{\# model formulas}
-\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{b2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3}\SpecialCharTok{*}\NormalTok{tt))}
-\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{c1}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\DecValTok{10}\SpecialCharTok{*}\NormalTok{c2}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1}\SpecialCharTok{*}\NormalTok{c3}\SpecialCharTok{*}\NormalTok{tt))}
-\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid}\SpecialCharTok{{-}}\NormalTok{tt)}\SpecialCharTok{/}\NormalTok{scal))}
+\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3 }\SpecialCharTok{*}\NormalTok{ tt))}
+\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{ c1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ c2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{c3 }\SpecialCharTok{*}\NormalTok{ tt))}
+\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
 \CommentTok{\#}
 \CommentTok{\# Starting parameter sets}
-\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{1}\NormalTok{, }\AttributeTok{b2=}\DecValTok{1}\NormalTok{, }\AttributeTok{b3=}\DecValTok{1}\NormalTok{)}
-\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{1}\NormalTok{, }\AttributeTok{c2=}\DecValTok{1}\NormalTok{, }\AttributeTok{c3=}\DecValTok{1}\NormalTok{)}
-\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{)}
+\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =}\DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{)}
+\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{1}\NormalTok{)}
+\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{)}
 \end{Highlighting}
@@
 It is also possible to provide R functions for the residual and
-Jacobian. This is usually more work for the user if the formula setup is
-possible. To illustrate, we show the functions for the unscaled 3
-parameter logistic.
+Jacobian. \texttt{nls()} is not well-suited to using such a formulation,
+but \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} work
+specifically with such a formulation. This is usually more work for the
+user if the formula setup is possible. To illustrate, we show the
+functions for the unscaled 3 parameter logistic.
 
@@
 \NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
+\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
 \NormalTok{\}}
@@
 \NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt)}
-\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1}\SpecialCharTok{+}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{yy)}
+\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)}
+\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ yy)}
 \NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
-\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}
-\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{]}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{zz}\SpecialCharTok{*}\NormalTok{yy}\SpecialCharTok{*}\NormalTok{x[}\DecValTok{2}\NormalTok{]}\SpecialCharTok{*}\NormalTok{tt}
+\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy}
+\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy }\SpecialCharTok{*}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt}
   \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
@@
 uses a Gauss-Newton (i.e., unstabilized) iteration with a step reduction
-line search. This works very efficiently as long as \(J\) is not
-ill-conditioned. Below we see \texttt{nls()} does poorly on the example
+line search (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})). This
+works very efficiently as long as \(J\) is not ill-conditioned. Below we
+see the default \texttt{nls()} algorithm does poorly on the example
 problem. To save page space, we use 1-line result display functions from
@@
 \begin{Highlighting}[]
-\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \end{Highlighting}
@@
 Here we see the infamous ``singular gradient'' termination message of
-\texttt{nls()}.
+\texttt{nls()}. Technically, it is the Jacobian that is singular, but R
+does not distinguish scalar and vector functions, which yield partial
+derivatives that are a gradient or Jacobian respectively.
+
+\texttt{nls()} also offers the option of using the ``port'' library
+algorithm choice (Fox, Hall, and Schryer
+(\protect\hyperlink{ref-FoxHallSchryer1978}{1978})). There is a warning
+in the \texttt{nls()} documentation that this ``appears unfinished'',
+but it is successful in solving all three cases.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{unls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
+\FunctionTok{pnls}\NormalTok{(unls1p)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{snls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
+\FunctionTok{pnls}\NormalTok{(snls1p)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{tnls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
+\FunctionTok{pnls}\NormalTok{(tnls1p)}
+\end{Highlighting}
+\end{Shaded}
+
+\begin{verbatim}
+## tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
+\end{verbatim}
 
@@
 \begin{Highlighting}[]
-\NormalTok{unlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{unlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{print}\NormalTok{(unlx1) }
@@
 \begin{Highlighting}[]
-\NormalTok{snlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{snlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
@@
 \begin{Highlighting}[]
-\NormalTok{tnlx1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{tnlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
@@
 package \texttt{nlsr} in a single column in the output to provide a
-compact layout, but the values do \textbf{NOT} correspond to the
+compact layout, but the values do \textbf{not} correspond to the
 individual parameters in whose row they appear; they are a property of
@@
 \begin{Highlighting}[]
-\NormalTok{unlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmu, }\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{unlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
@@
 \begin{Highlighting}[]
-\NormalTok{snlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{snlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{pnls}\NormalTok{(snlm1)}
@@
 \begin{Highlighting}[]
-\NormalTok{tnlm1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stt1, }\AttributeTok{data=}\NormalTok{weeddf))}
+\NormalTok{tnlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
 \FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
@@
 \FunctionTok{library}\NormalTok{(gslnls)}
-\NormalTok{ugslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
+\NormalTok{ugslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
 \FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
@@
 \begin{Highlighting}[]
-\NormalTok{sgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
+\NormalTok{sgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
 \FunctionTok{pnls}\NormalTok{(sgslnls1) }\CommentTok{\# Use summary() to get display}
@@
 \begin{Highlighting}[]
-\NormalTok{tgslnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
+\NormalTok{tgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
 \FunctionTok{pnls}\NormalTok{(tgslnls1) }
@@
 \begin{verbatim}
-## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20847  scal = -1744.7; 47  itns
-\end{verbatim}
+## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
+\end{verbatim}
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\DocumentationTok{\#\# Other algorithms}
+\DocumentationTok{\#\# tgslnls1a\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))}
+\DocumentationTok{\#\# pnls(tgslnls1a) }
+\DocumentationTok{\#\# tgslnls1d\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))}
+\DocumentationTok{\#\# pnls(tgslnls1d) }
+\DocumentationTok{\#\# tgslnls1dd\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))}
+\DocumentationTok{\#\# pnls(tgslnls1dd) }
+\DocumentationTok{\#\# tgslnls1s2\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))}
+\DocumentationTok{\#\# pnls(tgslnls1s2) }
+\end{Highlighting}
+\end{Shaded}
+
+While default \texttt{gsl\_nls()} gives much of the same results as
+\texttt{nlsLM()}, uncommenting the calls to the alternative algorithms
+shows it can get the global minimum with any of these choices.
 
@@
 contrast, \texttt{minpack.lm::nlsLM()} and \texttt{nls()} use
-\texttt{summary()}, which does NOT display the sum of squares, while
+\texttt{summary()}, which does not display the sum of squares, while
 \texttt{print()} gives the sum of squares, but not the standard error of
-the residuals.
-
-The singular values allow us to gauge how ``nearly singular'' the
-Jacobian is at the solution, and the ratio of the smallest to largest of
-the singular values is a simple but effective measure. The ratios are
-4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
-Logistic3T, so Logistic3S is the ``least singular''.
+the residuals. Such behaviour is consistent with other modeling
+functions such as \texttt{lm()}.
+
+The importance of the singular values is to allow us to gauge how
+``nearly singular'' the Jacobian is at the solution, and the ratio of
+the smallest to largest of the singular values is a simple but effective
+measure. The ratios are 4.6641e-05 for Logistic3U, 0.021022 for
+Logistic3S, and 0.001055 for Logistic3T; so Logistic3S is the ``least
+singular''.
 
@@
 we need to extract the Jacobian and compute its singular values. The
-following script shows how to do this, using as Jacobian what is called
-the \texttt{gradient} element in the returned solution for these
-solvers.
+following script shows how to do this, using as Jacobian the
+\texttt{gradient} element in the returned solution for these solvers.
 
@@
 \begin{verbatim}
-## [1] 3.4641e+00 1.0111e-08 4.7443e-11
+## [1] 3.4641e+00 9.4495e-09 4.1324e-11
 \end{verbatim}
@@
 \begin{Highlighting}[]
-\NormalTok{stspecial}\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
-\NormalTok{badstart}\OtherTok{\textless{}{-}}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula=}\NormalTok{frmt, }\AttributeTok{start=}\NormalTok{stspecial, }\AttributeTok{data=}\NormalTok{weeddf)}
+\NormalTok{stspecial }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
+\NormalTok{badstart }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stspecial, }\AttributeTok{data =}\NormalTok{ weeddf)}
 \FunctionTok{print}\NormalTok{(badstart)}
@@
 
+We also note that changing the start to
+
+\begin{verbatim}
+stt2 <- c(Asym=1, xmid=1, scal=2)
+\end{verbatim}
+
+allows both \texttt{nlsLM} and \texttt{gsl\_nls} to find the minimum.
+\texttt{gsl\_nls} also succeeds when the particular form of the
+Marquardt stabilization is changed from the default by the control
+setting \texttt{scale\ =\ "levenberg"}.
+
+We consider the code inside package \texttt{nlsr} to be competently
+coded, with the advantage over \texttt{nls()} of the Levenberg-Marquardt
+stabilization in common with \texttt{minpack.lm} and \texttt{glsnls}.
+The success rate is high, but \texttt{nlsr} has its premature
+terminations too. However, it provides diagnostic information in the
+form of the singular values of the Jacobian that allows us to try
+alternative starts, algorithms or programs in cases where there are
+indications of early termination.
+
 \hypertarget{functional-specification-of-problems}{%
@@
 
+Let us write a slightly modified form of the nonlinear least squares
+problem:
+
+\[ S(p) = \sum_{i=1}^m { r_i(p)^2 }\]
+
+R users are likely to think almost always that the functions \(r_i\) are
+all of the same form, differing in the details of the data they use.
+However, the famous Rosenbrock banana-shaped valley (Rosenbrock
+(\protect\hyperlink{ref-Rosenbrock1960}{1960})) has
+\[ r_1(p) = (1 - p_1)\] \[ r_2(p) = 10 (p_2 - {p_1}^2)\]
+
+Thus, there can be problems where the idea of a ``model'' or a
+``prediction'' makes no sense, though the nonlinear least squares
+problem is still well-posed. In our code, we still refer to the
+functions as ``residuals'', though this is purely a convenience.
+
 We illustrate how to solve nonlinear least squares problems using a
@@
 a vector \texttt{y} that is the expected value of what we have called
-the residual, but in this case, is actually the model. \texttt{gsl\_nls}
+the residual, but in this case is actually the model. \texttt{gsl\_nls}
 uses a numerical approximation for the Jacobian if the argument
-\texttt{jac} is missing. Note function \texttt{nlsr::pnlslm()} for a
-1-line display of the results of \texttt{minpack.lm::nls.lm()}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobnlfb}\OtherTok{\textless{}{-}}\FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{resfn=}\NormalTok{hobbs.res, }\AttributeTok{jacfn=}\NormalTok{hobbs.jac)}
+\texttt{jac} is missing. Note functions \texttt{nlsr::pnls()} and
+\texttt{nslr::pnlslm()} for a 1-line display of the results.
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{hobnlfb }\OtherTok{\textless{}{-}} \FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{resfn =}\NormalTok{ hobbs.res, }\AttributeTok{jacfn =}\NormalTok{ hobbs.jac)}
 \FunctionTok{pshort}\NormalTok{(hobnlfb) }\CommentTok{\# use print(hobnlfb) for more detail}
@@
 \begin{Highlighting}[]
-\NormalTok{hobnlm}\OtherTok{\textless{}{-}}\FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
+\NormalTok{hobnlm }\OtherTok{\textless{}{-}} \FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{jac =}\NormalTok{ hobbs.jac)}
 \FunctionTok{pnlslm}\NormalTok{(hobnlm)  }
@@
 \begin{Highlighting}[]
-\NormalTok{hobgsln}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
+\NormalTok{hobgsln }\OtherTok{\textless{}{-}} \FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{y =} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
 \FunctionTok{pnls}\NormalTok{(hobgsln)}
@@
 \begin{Highlighting}[]
-\NormalTok{hobgsl}\OtherTok{\textless{}{-}}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start=}\NormalTok{stu1, }\AttributeTok{fn=}\NormalTok{hobbs.res, }\AttributeTok{y=}\FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{), }\AttributeTok{jac=}\NormalTok{hobbs.jac)}
+\NormalTok{hobgsl }\OtherTok{\textless{}{-}} \FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{y =} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{), }\AttributeTok{jac =}\NormalTok{ hobbs.jac)}
 \FunctionTok{pnls}\NormalTok{(hobgsl) }\CommentTok{\# using analytic Jacobian}
@@
 extensions. For compatibility, the \texttt{nlsr} package has the
-function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias.
-This uses \texttt{nlxb()} to find good parameters, then calls
-\texttt{nls()} to return the class \texttt{nls} object. Unless
-particular modeling features are needed, the use of \texttt{wrapnlsr()}
-is unnecessary and wasteful of resources.
+function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias,
+that is intended for problems specified as formulas. This uses
+\texttt{nlxb()} to find good parameters, then calls \texttt{nls()} to
+return the class \texttt{nls} object. Unless particular modeling
+features are needed, the use of \texttt{wrapnlsr()} is unnecessary and
+wasteful of resources.
 
@@
 been in \texttt{nlsr} since it was introduced. The default value gives
-the legacy behaviour. This improvement has been in the R distributed
-code since version 4.1.0.
+\texttt{nls()} its legacy behaviour. This improvement has been in the R
+distributed code since version 4.1.0.
 
 Additional termination tests can be used. \texttt{nlsr} has a
-\textbf{small sum of squares} test (\textbf{smallsstest}) that compares
+\textbf{small sum of squares} test (\texttt{smallsstest}) that compares
 the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
 times the initial sum of squares, where
-\texttt{e4\ \textless{}-\ (100*.Machine\$double.eps)\^{}4} is
+\texttt{e4\ \textless{}-\ (100\ *\ .Machine\$double.eps)\ \^{}\ 4} is
 approximately 2.43e-55.
@@
 important. \texttt{nls()} stops after \texttt{maxiter} ``iterations''.
-The meaning of ``iteration'' may require an examination of the code. We
-terminate execution when the number of residual or Jacobian evaluations
-exceed set limits. Generally, we prefer larger limits than the default
-\texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
+The meaning of ``iteration'' may require an examination of the code.
+\texttt{nlsr} terminates when the number of residual or Jacobian
+evaluations (\texttt{feval} and \texttt{jeval}) exceed set limits, and
+we prefer larger limits (\texttt{jemax\ =\ 5000} is the default) than
+the \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
+This may cause overly long run-times for \texttt{nlsr}, but our work
+often involves problems that are sufficiently difficult that we want to
+be sure we do not terminate early.
 
@@
 
-The complexity of this object is a challenge to users. Let us use the
-\texttt{result} as the returned object from \texttt{nls()},
+The complexity of this object is a challenge to users. Let us use
+\texttt{result} as the name of the returned object from \texttt{nls()},
 \texttt{minpack.lm::nlsLM()} or, \texttt{gslnls::gsl\_nls()}. For
-example, the \texttt{data} return element is an R symbol. To actually
-access the data from this element, we need to use the syntax:
-
-\begin{verbatim}
-eval(parse(text=result$data))
-\end{verbatim}
-
-However, if the call is made with \texttt{model=TRUE}, then there is a
-returned element \texttt{model} which contains the data, and we can list
-its contents using:
+example, the \texttt{data} return element is an R symbol. If the user's
+environment exists, which it may not, we could see this data via the
+syntax:
+
+\begin{verbatim}
+eval(parse(text = result$data))
+\end{verbatim}
+
+However, since the evaluating environment might not exist, a better
+approach is to make the call to \texttt{nls()} with
+\texttt{model\ =\ TRUE}, which returns an element \texttt{model} which
+contains the data, and we can list its contents using
 
@@
 sum of squares, the residual vector, Jacobian, and counts of
-evaluations.
+evaluations, but does not include the input data.
 
@@
 Thus the code in \texttt{nlsr} is likely of value for workers to copy
-and modify for customized tools.
+and modify for customized tools, with code in R to support such uses.
 
@@
 \texttt{ndstep} for the size of the step used in the approximation.
-
-Using the ``gradient'' attribute of the output of the Jacobian function
-to hold the Jacobian matrix lets us embed this in the \textbf{residual}
-function as well, so that the call to \texttt{nlsr::nlfb()} can be made
-with the same name used for both residual and Jacobian function
-arguments. This programming trick saves a lot of trouble for the package
-developer, but it can be a nuisance for users trying to understand the
-code.
+These features are mainly to allow for investigation of approximate
+derivatives. Regular users will rarely need them.
+
+Using the \texttt{gradient} attribute of the output of the Jacobian
+function to hold the Jacobian matrix lets us embed this in the
+\textbf{residual} function as well, so that the call to
+\texttt{nlsr::nlfb()} can be made with the same name used for both
+residual and Jacobian function arguments. This programming trick saves a
+lot of trouble for the package developer, but it can be a nuisance for
+users trying to understand the code.
 
@@
   be computed. In such cases, we observed that \texttt{nls()} and
-  \texttt{minpack.lm::nlsLM} found a solution, though this seems to be a
-  lucky accident.
+  \texttt{minpack.lm::nlsLM} found a solution through the lucky accident
+  that the approximate derivative can be evaluated.
 \end{itemize}
@@
 richer set. There are also some in the now-archived package
-\texttt{NRAIA}. These provide the Jacobian in the ``gradient'' attribute
-of the ``one-sided'' formula that defines each model, and these
-Jacobians are often the analytic forms.
+\texttt{NRAIA}. The Jacobian is provided via the \texttt{gradient}
+attribute of the ``one-sided'' formula that defines each model, and
+these Jacobians are often the analytic forms.
 
 The \texttt{nls()} function, after computing the ``right-hand side'' or
-\texttt{rhs} of the residual, checks to see if the ``gradient''
+\texttt{rhs} of the residual, checks to see if the \texttt{gradient}
 attribute is defined, otherwise using \texttt{numericDeriv()} to compute
@@
 \texttt{nls()} with the `plinear' algorithm and finds a (full) solution.
-It then passes the solution coefficients to the default algorithm
-unnecessarily. Moreover, the implicit double call is, in our view, prone
-to creating errors in code maintenance. To provide simpler starting
-parameters, the function \texttt{SSlogisJN} is now part of the package
-\texttt{nlsr}, but is most useful for \texttt{nls()}.
+It then passes the solution coefficients to the default algorithm which
+converges in one iteration and builds the standard output. We found the
+implicit double-call a challenge to unravel. A more conventional
+approach to starting parameters, following the lead of Ratkowsky
+(\protect\hyperlink{ref-Ratkowsky1983}{1983}), is in the function
+\texttt{SSlogisJN}, now part of the package \texttt{nlsr}.
 
@@
 possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
-starting parameters, and can either use \texttt{getInitial()} to find
-them from the \texttt{selfStart} model or provide explicit values, but
-the \texttt{selfStart} model is used in the formula for the
-\texttt{nlxb()} call.
-
-We are also surprised that the analytic expressions for the Jacobian
-(``gradient'') in the \texttt{SSLogis} function and others save
-quantities in ``hidden'' variables, i.e., with names preceded by ``.''.
-These are then not displayed by the \texttt{ls()} command, making them
-difficult to access by users who may wish to create their own selfStart
-model via copy and edit. Interactive tools, such as ``visual fitting''
-(John C. Nash and Velleman
+starting parameters. We could get these by separately calling
+\texttt{getInitial()} to find them from the \texttt{selfStart} model,
+but could use other values. The \texttt{selfStart} model is also used in
+the formula for the \texttt{nlxb()} call.
+
+We note that the analytic expressions for the Jacobian
+(\texttt{gradient}) in the \texttt{SSLogis} function and others save
+quantities by default in ``hidden'' variables, i.e., with names preceded
+by ``.''. This behaviour is to avoid name conflicts with user
+quantities. Hidden variables are not displayed by the \texttt{ls()}
+command, so users need to remember to use
+\texttt{ls(all.names\ =\ TRUE)} to see which quantities are defined, for
+example, in developing their own selfStart model via copy and edit.
+
+Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman
 (\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
@@
 \item
-  \texttt{nls()} can only impose bounds if the \texttt{algorithm="port"}
-  argument is used in the call. Unfortunately, the documentation warns
-  us:
+  \texttt{nls()} can only impose bounds if the
+  \texttt{algorithm\ =\ "port"} argument is used in the call.
+  Unfortunately, the documentation warns us:
 
@@
 
+Examples of the use of bounds constraints for nonlinear regression
+modeling are quite scarce. Traditionally, workers transform variables to
+ensure bounds are satisfied, such as squaring a parameter, or using a
+parameter which is the \(\log()\) of the actual parameter, so that its
+exponential is positive. These transformations can create scaling
+issues, but seem to generally work.
+
+Bounds can also be helpful to avoid unanticipated excursions of the
+minimization trajectory into inadmissible parameter regions. Such bounds
+should not be too tight, as this may hinder progress to an optimum by
+causing the methods to bump along the constraint in many cycles. Our aim
+is to avoid rubbish, not specify where the solution will be, and we hope
+the solution is not complicated by having a parameter on a bound.
+
 \begin{Shaded}
@@
 \CommentTok{\# Start MUST be feasible i.e. on or within bounds}
-\NormalTok{anlshob1b }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
+\NormalTok{anlshob1b }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
+             \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
 \FunctionTok{pnls}\NormalTok{(anlshob1b) }\CommentTok{\#  check the answer (short form)}
@@
 \begin{Highlighting}[]
-\CommentTok{\# nlsLM seems NOT to work with bounds in this example}
-\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
+\CommentTok{\# nlsLM seems not to work with bounds in this example}
+\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
 \FunctionTok{pnls}\NormalTok{(anlsLM1b)}
@@
 \CommentTok{\# also no warning if starting out of bounds, but gets a good answer!!}
-\NormalTok{st4}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{4}\NormalTok{, }\AttributeTok{c2=}\DecValTok{4}\NormalTok{, }\AttributeTok{c3=}\DecValTok{4}\NormalTok{)}
-\NormalTok{anlsLMob }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{st4, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
+\NormalTok{st4}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{4}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{4}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{4}\NormalTok{)}
+\NormalTok{anlsLMob }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ st4, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
 \FunctionTok{pnls}\NormalTok{(anlsLMob)}
@@
 \CommentTok{\# Try nlsr::nlxb()}
-\NormalTok{anlx1b }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts1, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
+\NormalTok{anlx1b }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
 \FunctionTok{pshort}\NormalTok{(anlx1b)}
@@
 compute standard errors nor their derived statistics when bounds are
-active.
+active. Transformations to implicitly enforce bounds avoid this issue,
+but at the cost of changing the form of the model.
 
@@
 Let us try to fix (mask) the first parameter in the first two example
-problems.
+problems. We might want to do this if it is considered that the
+asymptote is known from external sources, but we wish later to have it
+available to be estimated.
 
@@
 \CommentTok{\# Hobbsmaskx.R {-}{-} masks with formula specification of the problem}
-\FunctionTok{require}\NormalTok{(nlsr); }\FunctionTok{require}\NormalTok{(minpack.lm); traceval}\OtherTok{\textless{}{-}}\ConstantTok{FALSE}
-\NormalTok{stu }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1=}\DecValTok{200}\NormalTok{, }\AttributeTok{b2=}\DecValTok{50}\NormalTok{, }\AttributeTok{b3=}\FloatTok{0.3}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
-\NormalTok{sts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{c1=}\DecValTok{2}\NormalTok{, }\AttributeTok{c2=}\DecValTok{5}\NormalTok{, }\AttributeTok{c3=}\DecValTok{3}\NormalTok{) }\CommentTok{\# a default scaled starting vector (named!)}
+\FunctionTok{require}\NormalTok{(nlsr); }\FunctionTok{require}\NormalTok{(minpack.lm); traceval }\OtherTok{\textless{}{-}} \ConstantTok{FALSE}
+\NormalTok{stu }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{200}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{50}\NormalTok{, }\AttributeTok{b3 =} \FloatTok{0.3}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
+\NormalTok{sts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{2}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{5}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{3}\NormalTok{) }\CommentTok{\# a default scaled starting vector (named!)}
 \CommentTok{\# fix first parameter}
-\NormalTok{anxbmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
+\NormalTok{anxbmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
+            \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
 \FunctionTok{print}\NormalTok{(anxbmsk1)}
@@
 \begin{Highlighting}[]
-\NormalTok{anlM1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
+\NormalTok{anlM1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
+            \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
 \FunctionTok{pnls}\NormalTok{(anlM1)}
@@
 \begin{Highlighting}[]
-\NormalTok{anlsmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(frmu, }\AttributeTok{start=}\NormalTok{stu, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-        \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{),  }\AttributeTok{algorithm=}\StringTok{"port"}\NormalTok{, }\AttributeTok{trace=}\NormalTok{traceval))}
+\NormalTok{anlsmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
+        \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{),  }\AttributeTok{algorithm=}\StringTok{"port"}\NormalTok{, }\AttributeTok{trace =}\NormalTok{ traceval))}
 \FunctionTok{pnls}\NormalTok{(anlsmsk1)}
@@
 \CommentTok{\# Hobbs scaled problem with bounds, formula specification}
-\NormalTok{anlxmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                  \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
+\NormalTok{anlxmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
+                  \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
 \FunctionTok{print}\NormalTok{(anlxmsks1)}
@@
 \begin{Highlighting}[]
-\NormalTok{anlshmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{trace=}\NormalTok{traceval, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{), }\AttributeTok{algorithm=}\StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
+\NormalTok{anlshmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{trace =}\NormalTok{ traceval, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
+             \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
 \FunctionTok{pnls}\NormalTok{(anlshmsk1)}
@@
 \begin{Highlighting}[]
-\NormalTok{anlsLMmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                 \AttributeTok{upper=}\FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
+\NormalTok{anlsLMmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
+                 \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
 \FunctionTok{pnls}\NormalTok{(anlsLMmsks1)}
@@
 \CommentTok{\# Test with all parameters masked}
-\NormalTok{anlxmskall}\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start=}\NormalTok{sts, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{lower=}\NormalTok{sts, }\AttributeTok{upper=}\NormalTok{sts))}
+\NormalTok{anlxmskall}\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =}\NormalTok{ sts, }\AttributeTok{upper =}\NormalTok{ sts))}
 \end{Highlighting}
@@
 offers an accelerated Marquardt method and three alternative methods
-which we have not investigated. Control settings for \texttt{nlxb()} or
-\texttt{nlfb()} allow exploration of Hartley and Marquardt algorithm
-variants. In general, the Levenberg-Marquardt stabilization is important
-in obtaining solutions in methods of the Gauss-Newton family, as
-\texttt{nls()} terminates too frequently and unnecessarily with
-\texttt{singular\ gradient} errors.
+which we have not investigated beyond one example. Control settings for
+\texttt{nlxb()} or \texttt{nlfb()} allow exploration of Hartley and
+Marquardt algorithm variants, principally for the benefit of program
+developers. In general, the Levenberg-Marquardt stabilization is
+important in obtaining solutions in methods of the Gauss-Newton family,
+as \texttt{nls()} terminates too frequently and unnecessarily with
+\texttt{singular\ gradient} errors. A tidy incorporation of Marquardt
+stabilization into \texttt{nls()} would be a major improvement.
 
 \hypertarget{programming-language}{%
-\subsubsection{Programming language}\label{programming-language}}
+\subsection{Programming language}\label{programming-language}}
 
@@
 R, C, and Fortran, as does \texttt{minpack.lm}. \texttt{gslnls} is an R
-wrapper to various C-language routines in the Gnu Scientific Library
+wrapper to various C-language routines in the GNU Scientific Library
 (Galassi et al. (\protect\hyperlink{ref-GSL-manual}{2009})). Generally,
@@
 use of R may be a smaller performance cost than the aggressive approach
-to a solution, which can cause more iterations to be used.
+to a solution, such as large computational limits for residuals and
+Jacobians.
+
+There are, of course, problems where use of R code will be slower. For
+example, a ``large'' case of the Penalty 1 test problem of Moré, Garbow,
+and Hillstrom (\protect\hyperlink{ref-More1981TUO}{1981}) will take
+noticeable time to solve. Problems in large numbers of parameters create
+large computational matrices for the Jacobian or its inner product,
+which Gauss-Newton methods create and process. This is confirmed by
+comparing \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} on the
+Penalty 1 problem of various sizes, where the latter, which uses
+compiled Fortran internally, was moderately faster. However, the method
+``Rcgmin'' from package \texttt{optimx} (John C. Nash and Varadhan
+(\protect\hyperlink{ref-p-optimx}{2011})) proved to be much faster than
+either, despite being entirely written in R, and is a much more reliable
+conjugate gradient minimizer than method ``CG'' of \texttt{optim()}. One
+of us created both, but now deprecates the latter. Timings of CG methods
+are more sensitive to small differences in setup and arithmetic than
+Gauss-Newton methods, and the pattern of such timings varies much less
+regularly with problem size for CG methods.
 
@@
 argument. In this case, it will search in the available environments
-(i.e., workspaces) for suitable data objects. We do NOT like this
+(i.e., workspaces) for suitable data objects. We do not like this
 approach, but it is ``the R way''. R allows users to leave many objects
@@
 \texttt{subset}. This acts through the mediation of
-\texttt{model.frame}, which not obvious in the source code files
+\texttt{model.frame}, which is not obvious in the source code files
 \texttt{/src/library/stats/R/nls.R} and
@@
 As with \texttt{subset}, our concern with \texttt{na.action} is that
-users may be unaware of the effects of an option they may not even be
-aware has been set. Should \texttt{na.fail} be the default?
-
-\hypertarget{feature-model-frame}{%
-\subsection{Feature: model frame}\label{feature-model-frame}}
+users may be unaware of the effects of an option they may not even know
+has been set. Should \texttt{na.fail} be the default?
+
+\hypertarget{model-frame}{%
+\subsection{Model frame}\label{model-frame}}
 
@@
 
-\begin{quote}
-\textbf{model} logical. If true, the model frame is returned as part of
-the object. Default is FALSE.
-\end{quote}
+\emph{\textbf{model} logical. If true, the model frame is returned as
+part of the object. Default is FALSE.}
 
@@
 benefit from more extensive documentation and examples of its use since
-it is used to implement features like \texttt{subset}.
+it is used to implement features like \texttt{subset}. Moreover, it
+provides access to resources that may be important in further
+calculations on models.
 
@@
 \texttt{residuals()} function are weighted, and the \texttt{fitted()} or
-\texttt{predict()} function are used to compute raw residuals.
+\texttt{predict()} functions are used to compute raw residuals.
 
@@
 want weights that are determined at least partially from the model
-parameters, for example, a measure of the standard deviation of
+parameters; for example, a measure of the standard deviation of
 observations. Such dynamic weighting situations are discussed in the
@@
 The function \texttt{resid()} (an alias for \texttt{residuals()}) gives
-WEIGHTED residuals, as does, for example, \texttt{result\$m\$resid()}.
+weighted residuals, as does, for example, \texttt{result\$m\$resid()}.
 The function \texttt{nlsModel()}, which we have had to extract from the
@@
 \begin{Highlighting}[]
-\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5}\SpecialCharTok{\^{}}\NormalTok{tt }\CommentTok{\# simple weights}
-\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt)}\SpecialCharTok{/}\NormalTok{scal))}
-\NormalTok{Asym}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; xmid}\OtherTok{\textless{}{-}}\DecValTok{1}\NormalTok{; scal}\OtherTok{\textless{}{-}}\DecValTok{1}
-\NormalTok{nowt}\OtherTok{\textless{}{-}}\FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# UNWEIGHTED}
-\NormalTok{rnowt}\OtherTok{\textless{}{-}}\NormalTok{nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has UNWEIGHTED residual and Jacobian.
-Does NOT take coefficients.}
-\FunctionTok{attr}\NormalTok{(rnowt, }\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rnowt}
+\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5} \SpecialCharTok{\^{}}\NormalTok{ tt }\CommentTok{\# simple weights}
+\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
+\NormalTok{Asym }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{; xmid }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{; scal }\OtherTok{\textless{}{-}} \DecValTok{1}
+\NormalTok{nowt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# unweighted}
+\NormalTok{rnowt }\OtherTok{\textless{}{-}}\NormalTok{ nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has unweighted residual and Jacobian.
+Does not take coefficients.}
+\FunctionTok{attr}\NormalTok{(rnowt, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rnowt}
 \end{Highlighting}
@@
 \NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
-\NormalTok{rusewt}\OtherTok{\textless{}{-}}\NormalTok{usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# WEIGHTED. Does NOT take coefficients.}
-\FunctionTok{attr}\NormalTok{(rusewt,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rusewt}
+\NormalTok{rusewt }\OtherTok{\textless{}{-}}\NormalTok{ usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# weighted. Does not take coefficients.}
+\FunctionTok{attr}\NormalTok{(rusewt,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rusewt}
 \end{Highlighting}
@@
 \FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
-\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{1}\NormalTok{, }\AttributeTok{xmid=}\DecValTok{1}\NormalTok{, }\AttributeTok{scal=}\DecValTok{1}\NormalTok{), }\AttributeTok{wts=}\NormalTok{wts)}
+\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{start =} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{), }\AttributeTok{wts =}\NormalTok{ wts)}
 \NormalTok{rn0}\OtherTok{\textless{}{-}}\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
-\FunctionTok{attr}\NormalTok{(rn0,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn0}
+\FunctionTok{attr}\NormalTok{(rn0,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rn0}
 \end{Highlighting}
@@
 \begin{Highlighting}[]
-\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data=}\NormalTok{weeddf, }\AttributeTok{start=}\FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts=}\NormalTok{wts)}
-\NormalTok{rn}\OtherTok{\textless{}{-}}\NormalTok{nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
-\FunctionTok{attr}\NormalTok{(rn,}\StringTok{"gradient"}\NormalTok{)}\OtherTok{\textless{}{-}}\ConstantTok{NULL}\NormalTok{; rn}
+\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{start =} \FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts =}\NormalTok{ wts)}
+\NormalTok{rn }\OtherTok{\textless{}{-}}\NormalTok{ nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
+\FunctionTok{attr}\NormalTok{(rn,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rn}
 \end{Highlighting}
@@
 found a point where the relative offset convergence criterion is met, it
-may simply exit, especially if a ``singular gradient'' (singular
-Jacobian) is found. However, this may occur AFTER the function has made
-considerable progress in reducing the sum of squared residuals. Here is
-an abbreviated example:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
-\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{( }\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
+may simply exit, especially if a ``singular gradient'' is found.
+However, this may occur \textbf{after} considerable progress has been
+made in reducing the sum of squared residuals. Here is an abbreviated
+example:
+
+\begin{Shaded}
+\begin{Highlighting}[]
+\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
+\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
 \NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
@@
 \begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(tryit)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
-## attr(,"class")
-## [1] "try-error"
-## attr(,"condition")
-## <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
-\end{verbatim}
+\CommentTok{\# print(tryit) \# gives no information on best sumsquares or model parameters}
+\end{Highlighting}
+\end{Shaded}
 
@@
 (\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
-(\protect\hyperlink{ref-OlearyRust13}{2013})) is generally much more
+(\protect\hyperlink{ref-OlearyRust13}{2013})) is usually much more
 effective than general approaches in finding good solutions to nonlinear
@@
 information to estimating functions is not a trivial task.
-\texttt{nls()} has an option \texttt{algorithm="plinear"} that allows
-some partially linear models to be solved. The other tools, as far as we
-are aware, do not offer any such capability. The \texttt{nlstac} package
-uses a different algorithm for similar goals.
+\texttt{nls()} has an option \texttt{algorithm\ =\ "plinear"} that
+allows some partially linear models to be solved. The other tools, as
+far as we are aware, do not offer any such capability. The
+\texttt{nlstac} package uses a different algorithm for similar goals.
 
@@
 specifications with different \texttt{algorithm} options. For example,
-the explicit model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not
-work with the linear modeling function \texttt{lm()}, which requires
-this model to be specified as \texttt{y\ \textasciitilde{}\ x}. Within
-\texttt{nls()}, consider the following FOUR different specifications for
+the explicit linear model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b}
+does not work. As with the linear modeling function \texttt{lm()}, this
+must be specified as \texttt{y\ \textasciitilde{}\ x}. Within
+\texttt{nls()}, consider the following four different specifications for
 the same problem, plus an intuitive choice, labeled \texttt{fm2a}, that
-does NOT work. In this failed attempt, putting the \texttt{Asym}
+does not work. In this failed attempt, putting the \texttt{Asym}
 parameter in the model causes the \texttt{plinear} algorithm to try to
 add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. At the time of writing, we do not
-foresee a resolution for this issue. In the example, we have NOT
-evaluated the commands to save space.
+like to see a consistent syntax. We do not foresee a resolution for this
+issue, which arises from different approaches to model specification. In
+the example, we have not evaluated the commands to save space.
 
@@
 
-\DocumentationTok{\#\# using conditional linearity AND Asym does NOT work}
+\DocumentationTok{\#\# using conditional linearity AND Asym does not work}
 \NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
@@
 the ``plinear'' option as an added complication. Running this example
-reveals that the answers for the parameters are NOT indexed as in a
+reveals that the answers for the parameters are not indexed as in a
 vector. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
@@
 et al. (\protect\hyperlink{ref-p-nlme}{2013})), \texttt{bbmle} (Bolker
-and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4}
-(p-lme4) for estimating models of such type.
+and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4} (D.
+Bates et al. (\protect\hyperlink{ref-p-lme4}{2015})) for estimating
+models of such type.
+
+\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
+\subsection{Some other CRAN packages for nonlinear
+modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
+
+\texttt{onls} (Andrej-Nikolai Spiess
+(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
+in estimating nonlinear models is the vertical difference between the
+dependent variable and the functional model. \texttt{onls} minimizes the
+sum of \textbf{orthogonal} residuals. The objective is therefore
+different and involves nontrivial extra calculation. A vignette with the
+package and the blog article
+\url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
+give some description with illustrative graphs. \texttt{onls} appears to
+be limited to problems with one independent and one dependent variable.
+The Wikipedia article
+\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
+overview of some ideas, with references to the literature. The approach
+needs a wider discussion and tutorial examples to allow its merits to be
+judged than can be included here.
+
+\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
+-- This package allows nonlinear estimation by controlled random search
+via two methods. There is unfortunately no vignette. A modest trial we
+carried out showed \texttt{nlsr::nlxb()} gave the same results in a
+small fraction of the time required by either of the methods in
+\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
+Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
+reliability in finding solutions than a Levenberg-Marquardt code
+(actually from Matlab), but the tests were conducted on the extreme NIST
+examples mentioned next.
+
+\texttt{NISTnls} (National Institutes for Standards and R port by
+Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
+package provides R code and data for a set of (numerically
+ill-conditioned) nonlinear least squares problems from the U.S. National
+Institute for Standards and Technology. These may not represent
+real-world situations.
+
+\texttt{nlshelper} (Duursma
+(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
+unfortunately lacks a vignette, provides a few utilities for
+summarizing, testing, and plotting non-linear regression models
+estimated with nls(), nlsList() or nlme() that are linked or grouped in
+some way.
+
+\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
+This solves nonlinear least squares problems with optional equality
+and/or inequality constraints. It is clearly \textbf{not} about
+modeling, and the input and output are quite different from class
+\texttt{nls} methods. However, there do not appear to be other R
+packages with these capabilities.
+
+\texttt{nlsMicrobio} (Baty and Delignette-Muller
+(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
+nonlinear regression models dedicated to predictive microbiology,
+including a vignette, by authors of the \texttt{nlstools} package.
+
+\texttt{nlstools} (Baty and Delignette-Muller
+(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
+several tools for aiding the estimation of nonlinear models,
+particularly using \texttt{nls()}. The vignette is actually a journal
+article, and the authors have considerable experience in the subject.
+
+\texttt{nlsmsn} (Prates, Lachos, and Garay
+(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
+scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
+Lachos, and Abanto-Valle (2011). The problem here is to minimize an
+objective that is modified from the traditional sum of squared
+residuals.
+
+\texttt{nls.multstart} (Padfield and Matheson
+(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
+squares regression using AIC scores with the Levenberg-Marquardt
+algorithm using multiple starting values for increasing the chance that
+the minimum found is the global minimum.
+
+\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
+Nonlinear least squares by brute force has similar motivations to
+\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
+The author has extensive expertise in R.
+
+\texttt{nlstac} (Rodriguez-Arias et al.
+(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
+implementing the algorithm described in Fernandez Torvisco et al.~(2018)
+for fitting separable nonlinear regression curves. The special class of
+problem for which this package is intended is an important and difficult
+one. No vignette is provided, unfortunately.
+
+\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models
+are treated, but there is minimal documentation and no vignette. Package
+\texttt{nlraa} is to be preferred.
+
+\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
+-- a set of nonlinear \emph{selfStart} models, primarily from
+agriculture. Most include analytic Jacobian code.
+
+\texttt{optimx} (John C. Nash and Varadhan
+(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
+that can be applied to minimize a nonlinear function which could be a
+nonlinear sum of squares. This is not generally recommended if nonlinear
+least squares programs can be easily used, but provides a check and
+alternative solvers.
 
@@
 
-Such goals align with the aims of \textbf{unit testing} (e.g.,
-\emph{\url{https://towardsdatascience.com/unit-testing-in-r-68ab9cc8d211}},
-Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
-(\protect\hyperlink{ref-HWdevtools21}{2021}) and the conventional R
-package testing tools). In our work, one of us has developed a working
-prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}}
-. A primary design objective of this is to allow the summarisation of
-multiple tests in a compact output. The prototype has a vignette to
-illustrate its use.
-
-\hypertarget{documentation-and-resources}{%
-\subsection{Documentation and
-resources}\label{documentation-and-resources}}
-
-In our investigation, we built several resources, which are now part of
-the repository \emph{\url{https://github.com/nashjc/RNonlinearLS/}}.
-Particular items are:
-
-\begin{itemize}
-\item
-  A BibTex bibliography for use with all documents in this project, but
-  which has wider application to nonlinear least squares projects in
-  general
-  (\emph{\url{https://github.com/nashjc/RNonlinearLS/blob/main/BibSupport/ImproveNLS.bib}}).
-\item
-  \texttt{MachID.R} offers a suggested concise summary function to
-  identify a particular computational system used for tests. A
-  discussion of how it was built and the resources needed across
-  platforms is given in at
-  \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/MachineSummary}}.
-\item
-  John C. Nash and Bhattacharjee
-  (\protect\hyperlink{ref-PkgFromRbase22}{2022}) is an explanation of
-  the construction of the \texttt{nlspkg} from the \texttt{nls()} code
-  in R-base.
-\item
-  As the 2021 Summer of Code period was ending, one of us (JN) was
-  invited to prepare a review of optimization in R. Ideas from the
-  present work have been instrumental in the creation of John C. Nash
-  (\protect\hyperlink{ref-NashWires22}{2022}).
-\end{itemize}
+Such goals align with the aims of \textbf{unit testing} (Wickham
+(\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
+(\protect\hyperlink{ref-HWdevtools21}{2021}), the conventional R package
+testing tools, and a number of blog posts found by searching ``unit
+testing in R''). In our work, one of us has developed a working
+prototype package at
+\emph{\url{https://github.com/ArkaB-DS/nlsCompare}}. A primary design
+objective of this is to allow the summarisation of multiple tests in a
+compact output. The prototype has a vignette to illustrate its use.
 
@@
 helpful in guiding our attempts to succinctly summarize computing
-environments.
+environments. Particularly thoughtful reviews by the referees, including
+details of inner workings of R, have led to important revisions.
 
 \hypertarget{references}{%
-\section{References}\label{references}}
+\section*{References}\label{references}}
+\addcontentsline{toc}{section}{References}
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-p-lme4}{}}%
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015.
+{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of
+Statistical Software} 67 (1): 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-FoxHallSchryer1978}{}}%
+Fox, P. A., A. P. Hall, and N. L. Schryer. 1978. {``The PORT
+Mathematical Subroutine Library.''} \emph{ACM Trans. Math. Softw.} 4
+(2): 104--26. \url{https://doi.org/10.1145/355780.355783}.
+
 \leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-More1981TUO}{}}%
+Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981.
+{``Testing Unconstrained Optimization Software.''} \emph{J-Toms} 7 (1):
+17--41.
+
 \leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
@@
 Algebra and Function Minimisation}. Bristol: Adam Hilger.
-
-\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
----------. 2022. {``{Function minimization and nonlinear least squares
-in R}.''} \emph{WIREs Computational Statistics}, no. e1580.
-https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
-
-\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
-Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package
-from Base {R} Files.''} \emph{R-Bloggers}, January.
-\url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
 
@@
 
+\leavevmode\vadjust pre{\hypertarget{ref-Rosenbrock1960}{}}%
+Rosenbrock, H. H. 1960. {``An Automatic Method for Finding the Greatest
+or Least Value of a Function.''} \emph{The Computer Journal} 3: 175--84.
+
 \leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%

### 2023-07-20 RefactoringNLS/revision202307/RJournal2023-15Rev1ResubmissionRJstyle/StateOfRToolsForNLS.Rmd
fixup yaml

--- 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.Rmd
@@
 title: "The Current State of R Tools for Nonlinear Least Squares Modeling"
-author:
-- John C. Nash \thanks{ retired professor, Telfer School of Management, University
-  of Ottawa}
-- Arkajyoti Bhattacharjee \thanks{Department of Mathematics and Statistics, Indian
-  Institute of Technology, Kanpur}
 date: "2023-7-16"
-output: 
-  pdf_document:
-    keep_tex: yes
-    toc: no
-  word_document:
-    toc: no
-bibliography: ./StateOfRToolsForNLS.bib
-link-citations: yes
-linkcolor: red
-urlcolor: blue
----
-
-```{r setup, include=FALSE}
-knitr::opts_chunk$set(echo = TRUE)
-## require(bookdown) # language engine to display text - does not seem necessary
-```
-
-## Abstract
-
+abstract: 
   Our Google Summer of Code project "Improvements to `nls()`" considered the features 
@@
   nonlinear estimation tools.
+
+draft: true
+author:
+    - name: John C. Nash
+      affiliation: retired professor, University of Ottawa
+      address:
+      - Telfer School of Management
+      - Ottawa ON Canada K1N 6N5
+      orcid: 0000-0002-2762-8039
+      email: profjcnash@gmail.com
+    - name: Arkajyoti Bhattacharjee
+      affiliation: Indian Institute of Technology
+      address:
+      - Department of Mathematics and Statistics
+      - Kanpur
+      email: arkastat98@gmail.com
+type: package
+output: 
+  rjtools::rjournal_web_article:
+      self_contained: yes
+      toc: no
+bibliography: ./StateOfRToolsForNLS.bib
+link-citations: yes
+linkcolor: red
+urlcolor: blue
+---
+
+```{r setup, include=FALSE}
+knitr::opts_chunk$set(echo = TRUE)
+## require(bookdown) # language engine to display text - does not seem necessary
+```
 

### 2023-07-20 RefactoringNLS/revision202307/RJournal2023-15Rev1ResubmissionRJstyle/StateOfRToolsForNLS.tex
fixup yaml

--- 2023-07-16 RefactoringNLS/revision202307/RJournal2023-15Rev1Resubmission/StateOfRToolsForNLS.tex
@@
-% Options for packages loaded elsewhere
-\PassOptionsToPackage{unicode}{hyperref}
-\PassOptionsToPackage{hyphens}{url}
-\PassOptionsToPackage{dvipsnames,svgnames,x11names}{xcolor}
-%
-\documentclass[
-]{article}
-\usepackage{amsmath,amssymb}
-\usepackage{iftex}
-\ifPDFTeX
-  \usepackage[T1]{fontenc}
-  \usepackage[utf8]{inputenc}
-  \usepackage{textcomp} % provide euro and other symbols
-\else % if luatex or xetex
-  \usepackage{unicode-math} % this also loads fontspec
-  \defaultfontfeatures{Scale=MatchLowercase}
-  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
-\fi
-\usepackage{lmodern}
-\ifPDFTeX\else
-  % xetex/luatex font selection
-\fi
-% Use upquote if available, for straight quotes in verbatim environments
-\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
-\IfFileExists{microtype.sty}{% use microtype if available
-  \usepackage[]{microtype}
-  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
-}{}
-\makeatletter
-\@ifundefined{KOMAClassName}{% if non-KOMA class
-  \IfFileExists{parskip.sty}{%
-    \usepackage{parskip}
-  }{% else
-    \setlength{\parindent}{0pt}
-    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
-}{% if KOMA class
-  \KOMAoptions{parskip=half}}
-\makeatother
-\usepackage{xcolor}
-\usepackage[margin=1in]{geometry}
-\usepackage{color}
-\usepackage{fancyvrb}
-\newcommand{\VerbBar}{|}
-\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
-\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
-% Add ',fontsize=\small' for more characters per line
-\usepackage{framed}
-\definecolor{shadecolor}{RGB}{248,248,248}
-\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
-\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
-\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\BuiltInTok}[1]{#1}
-\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
-\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
-\newcommand{\ExtensionTok}[1]{#1}
-\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
-\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\ImportTok}[1]{#1}
-\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
-\newcommand{\NormalTok}[1]{#1}
-\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
-\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
-\newcommand{\RegionMarkerTok}[1]{#1}
-\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
-\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
-\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
-\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
-\usepackage{graphicx}
-\makeatletter
-\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
-\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
-\makeatother
-% Scale images if necessary, so that they will not overflow the page
-% margins by default, and it is still possible to overwrite the defaults
-% using explicit options in \includegraphics[width, height, ...]{}
-\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
-% Set default figure placement to htbp
-\makeatletter
-\def\fps@figure{htbp}
-\makeatother
-\setlength{\emergencystretch}{3em} % prevent overfull lines
-\providecommand{\tightlist}{%
-  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
-\setcounter{secnumdepth}{-\maxdimen} % remove section numbering
-\newlength{\cslhangindent}
-\setlength{\cslhangindent}{1.5em}
-\newlength{\csllabelwidth}
-\setlength{\csllabelwidth}{3em}
-\newlength{\cslentryspacingunit} % times entry-spacing
-\setlength{\cslentryspacingunit}{\parskip}
-\newenvironment{CSLReferences}[2] % #1 hanging-ident, #2 entry spacing
- {% don't indent paragraphs
-  \setlength{\parindent}{0pt}
-  % turn on hanging indent if param 1 is 1
-  \ifodd #1
-  \let\oldpar\par
-  \def\par{\hangindent=\cslhangindent\oldpar}
-  \fi
-  % set entry spacing
-  \setlength{\parskip}{#2\cslentryspacingunit}
- }%
- {}
-\usepackage{calc}
-\newcommand{\CSLBlock}[1]{#1\hfill\break}
-\newcommand{\CSLLeftMargin}[1]{\parbox[t]{\csllabelwidth}{#1}}
-\newcommand{\CSLRightInline}[1]{\parbox[t]{\linewidth - \csllabelwidth}{#1}\break}
-\newcommand{\CSLIndent}[1]{\hspace{\cslhangindent}#1}
-\ifLuaTeX
-  \usepackage{selnolig}  % disable illegal ligatures
-\fi
-\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
-\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
-\urlstyle{same}
-\hypersetup{
-  pdftitle={The Current State of R Tools for Nonlinear Least Squares Modeling},
-  pdfauthor={John C. Nash ; Arkajyoti Bhattacharjee },
-  colorlinks=true,
-  linkcolor={red},
-  filecolor={Maroon},
-  citecolor={Blue},
-  urlcolor={blue},
-  pdfcreator={LaTeX via pandoc}}
-
-\title{The Current State of R Tools for Nonlinear Least Squares
-Modeling}
-\author{John C. Nash
-\thanks{ retired professor, Telfer School of Management, University of Ottawa} \and Arkajyoti
-Bhattacharjee
-\thanks{Department of Mathematics and Statistics, Indian Institute of Technology, Kanpur}}
-\date{2023-7-16}
-
-\begin{document}
+% !TeX root = RJwrapper.tex
+\title{The Current State of R Tools for Nonlinear Least Squares Modeling}
+\author{by John C. Nash and Arkajyoti Bhattacharjee}
+
 \maketitle
 
-\hypertarget{abstract}{%
-\subsection{Abstract}\label{abstract}}
-
-Our Google Summer of Code project ``Improvements to \texttt{nls()}''
-considered the features and limitations of the R function \texttt{nls()}
-with the aim of improving and rationalizing R tools for nonlinear
-regression. The rich features of \texttt{nls()} are weakened by several
-deficiencies and inconsistencies such as a lack of stabilization of the
-Gauss-Newton solver. Further considerations are the usability and
-maintainability of the code base that provides the functionality
-\texttt{nls()} offers. Various packages, in particular \texttt{nlsr}
-(John C Nash and Duncan Murdoch
-(\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative
-capabilities. We consider the differences in goals, approaches, and
-features of different tools for nonlinear least squares modeling in R.
-Discussion of these matters is relevant to improving R generally as well
-as its nonlinear estimation tools.
+\abstract{%
+Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression.
+The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
+Further considerations are the usability and maintainability of the code base that provides the functionality \texttt{nls()} offers.
+Various packages, in particular \texttt{nlsr} (John C Nash and Duncan Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})), provide alternative capabilities.
+We consider the differences in goals, approaches, and features of different tools for nonlinear least squares modeling in R.
+Discussion of these matters is relevant to improving R generally as well as its nonlinear estimation tools.
+}
 
 \hypertarget{the-nls-function}{%
-\subsection{\texorpdfstring{The \texttt{nls()}
-function}{The nls() function}}\label{the-nls-function}}
-
-\texttt{nls()} is the tool for estimating nonlinear statistical models
-in base R, the primary software distribution from the Comprehensive R
-Archive Network (CRAN). Users do not have to request that \texttt{nls()}
-be loaded. The function dates to the 1980s and the work related to D. M.
-Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) and others in
-the S programming language.
-
-The \texttt{nls()} function has a remarkable and comprehensive set of
-capabilities for estimating nonlinear models that are expressed as
-\textbf{formulas}. In particular, we note that it
+\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}
+
+\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
+primary software distribution
+from the Comprehensive R Archive Network (CRAN).
+Users do not have to request that \texttt{nls()} be loaded.
+The function dates to the 1980s and the work related to D. M. Bates and Watts (\protect\hyperlink{ref-bateswatts}{1988}) and others
+in the S programming language.
+
+The \texttt{nls()} function has a remarkable and comprehensive set of capabilities for
+estimating nonlinear models that are expressed as \textbf{formulas}. In particular, we
+note that it
 
@@
 \item
-  handles formulas that include R functions, even ones that call
-  calculations in other programming languages;
+  handles formulas that include R functions, even ones that call calculations in
+  other programming languages;
 \item
@@
 \item
-  produces measures of variability (i.e., standard error estimates) for
-  the estimated parameters;
-\item
-  has related profiling capabilities for exploring the likelihood
-  surface as parameters are changed;
-\item
-  links to many pre-coded (\texttt{selfStart}) models that do not
-  require initial parameter values.
+  produces measures of variability (i.e., standard error estimates) for the
+  estimated parameters;
+\item
+  has related profiling capabilities for exploring the likelihood surface as
+  parameters are changed;
+\item
+  links to many pre-coded (\texttt{selfStart}) models that do not require
+  initial parameter values.
 \end{itemize}
 
-With such a range of features and long history, the code has become
-untidy and overly patched, difficult to maintain, and its underlying
-methods could be improved. Various workers have developed packages to
-overcome these concerns, and we will address some of these here.
+With such a range of features and long history, the code has become untidy
+and overly patched, difficult to maintain, and its underlying methods could
+be improved. Various workers have developed packages to overcome these
+concerns, and we will address some of these here.
 
@@
 
-Our Google Summer of Code project for 2021, ``Improvements to
-\texttt{nls()}'' had the goal of providing a version of the code that
-would address some of the deficiencies discovered over the long period
-that \texttt{nls()} has been part of R. However, it was also critical to
-preserve legacy functionality, and we did not achieve a result that
-provided desired improvements and preserved legacy results
-simultaneously. On the other hand, working with \texttt{nls()} suggested
-ways package \texttt{nlsr} could be improved in features, while also
-revealing how it has a very different internal structure and design.
-
-The primary goal of this paper is to present some of these issues, with
-a limited consideration of some other R tools that exist for nonlinear
-least squares calculations. We hope to encourage others to engage,
-possibly with us, to find ways to bring improvements and to consolidate
-tools where reasonable. We want to ease the task of users in choosing
-which tools to use. Unfortunately, it is simpler to create new packages
-than to bring together and simplify existing ones.
+Our Google Summer of Code project for 2021, ``Improvements to \texttt{nls()}''
+had the goal of providing a version of the code that would address some
+of the deficiencies discovered over the long period that \texttt{nls()}
+has been part of R. However, it was also critical to preserve legacy
+functionality, and we did not achieve a result that provided desired
+improvements and preserved legacy results simultaneously. On the other hand, working
+with \texttt{nls()} suggested ways package \texttt{nlsr} could be improved
+in features, while also revealing how it has a very different internal
+structure and design.
+
+The primary goal of this paper is to present some of these issues,
+with a limited consideration of some other R tools that exist for
+nonlinear least squares calculations. We hope to encourage
+others to engage, possibly with us, to find ways to bring improvements and
+to consolidate tools where reasonable. We want to ease the task of users
+in choosing which tools to use. Unfortunately, it is simpler to create
+new packages than to bring together and simplify existing ones.
 
@@
 
-Besides the base-R \texttt{nls()} function, we consider some packages
-that are available in the CRAN repository. Of these, we will pay
-particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch
-(\protect\hyperlink{ref-nlsr2023manual}{2023})), \textbf{minpack.lm}
-(Elzhov et al. (\protect\hyperlink{ref-minpacklm12}{2012})), and
-\textbf{gslnls} (Chau (\protect\hyperlink{ref-gslnls23}{2023})) which
-are general nonlinear least-squares solvers. We will not pursue tools in
-the Bioconductor (Gentleman et al.
-(\protect\hyperlink{ref-Bioconductor}{2004})) collection, nor those on
-repositories such as GitHub and Gitlab.
+Besides the base-R \texttt{nls()} function, we consider some packages that
+are available in the CRAN repository. Of these, we will pay particular
+attention to \textbf{nlsr} (John C Nash and Duncan Murdoch (\protect\hyperlink{ref-nlsr2023manual}{2023})),
+\textbf{minpack.lm} (Elzhov et al. (\protect\hyperlink{ref-minpacklm12}{2012})), and \textbf{gslnls} (Chau (\protect\hyperlink{ref-gslnls23}{2023})) which are
+general nonlinear least-squares solvers. We will
+not pursue tools in the Bioconductor (Gentleman et al. (\protect\hyperlink{ref-Bioconductor}{2004}))
+collection, nor those on repositories such as
+GitHub and Gitlab.
 
@@
 
-The Hobbs weed infestation problem (John C. Nash
-(\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
-task that seems straightforward, but is quite nasty. Its very succinct
-statement provides the ``short reproducible example'' much requested on
-R mailing lists. This is a real problem from a field researcher, though
-the units of measurement and background information are not available.
-The data and graph follow.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{weed }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{,}
-          \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{tt }\OtherTok{\textless{}{-}} \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{weeddf }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(tt, weed)}
-\FunctionTok{plot}\NormalTok{(weeddf, }\AttributeTok{main=}\StringTok{"Hobbs weed infestation data"}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
+The Hobbs weed infestation problem (John C. Nash (\protect\hyperlink{ref-cnm79}{1979, 120})) is a growth curve modeling
+task that seems straightforward, but is quite nasty. Its very succinct statement
+provides the ``short reproducible example'' much requested on R mailing lists.
+This is a real problem from a field researcher, though the units of measurement
+and background information are not available. The data and graph follow.
+
+\begin{verbatim}
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+\end{verbatim}
 
@@
 
-\textbf{Logistic3T}: \[ y \approx Asym / (1 + exp((xmid - t)/scal)) \]
-
-where we will use \texttt{weed} for \(y\) and \texttt{tt} for \(t\). The
-functions above are equivalent, but the first is generally more awkward
-to solve numerically due to its poor scaling. The parameters of the
-three forms are related as follows:
-
-\[   Asym =  b_1 = 100 * c_1 \] \[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
-\[ 1/scal  =  b_3 \] To allow for simpler discussion, let us say that
-the parameters form a (named) vector \(p\) and the model function is
-called \(model(p)\). The residuals can be written either as
-\(r_1 = y - model(p)\) or \(r = model(p) - y\) since their sum of
-squares has an identical value. The second form avoids a potential sign
-error if we need to evaluate derivatives.
-
-We wish to minimize the sum of squared residuals, which is our loss (or
-objective) function. Starting with some guess for the parameters, we aim
-to alter these parameters to obtain a smaller loss function. We then
-iterate until we can make no further progress.
-
-Let us consider there are \(n\) parameters and \(m\) residuals. The loss
-function is \[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\] The gradient of
-\(S(p)\) is \[ g = 2 * J' r\] where the Jacobian \(J\) is given by
-elements \[ J_{i,j} = \partial r_i / \partial p_j \] and the Hessian is
-defined by elements
-\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \] If we
-expand the Hessian for nonlinear least squares problems, we find
+\textbf{Logistic3T}:
+\[ y \approx Asym / (1 + exp((xmid - t)/scal)) \]
+
+where we will use \texttt{weed} for \(y\) and \texttt{tt} for \(t\).
+The functions above are equivalent, but the first is generally more awkward
+to solve numerically due to its poor scaling. The parameters of the three forms
+are related as follows:
+
+\[   Asym =  b_1 = 100 * c_1 \]
+\[ exp(xmid/scal)  =  b_2 = 10 * c_2 \]
+\[ 1/scal  =  b_3 \]
+To allow for simpler discussion, let us say that the parameters form a (named) vector
+\(p\) and the model function is called \(model(p)\). The residuals can be written
+either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
+since their sum of squares has an identical value. The second form
+avoids a potential sign error if we need to evaluate derivatives.
+
+We wish to minimize the sum of squared residuals, which is our loss (or objective) function.
+Starting with some guess for the parameters, we aim to alter these parameters to obtain a smaller
+loss function. We then iterate until we can make no further progress.
+
+Let us consider there are \(n\) parameters and \(m\) residuals.
+The loss function is
+\[ S(p) = r' r = \sum_{i=1}^m { r_i^2 }\]
+The gradient of \(S(p)\) is
+\[ g = 2 * J' r\]
+where the Jacobian \(J\) is given by elements
+\[ J_{i,j} = \partial r_i / \partial p_j \]
+and the Hessian is defined by elements
+\[ H_{i,j} = \partial ^2 S(p) / {\partial p_i \partial p_j} \]
+If we expand the Hessian for nonlinear least squares problems, we find
 \[  H_{i,j}/2 = \sum_{k = 1}^m{ J_{k,i} J_{k,j}}  + \sum_{k = 1}^m {r_k * \partial{r_k}/{\partial p_i \partial p_j}}\]
-Let us use \(D_{i,j}\) for the elements of the second term of this
-expression. What is generally called \textbf{Newton's method} for
-function minimization tries to set the gradient to zero (to find a
-stationary point of the function \(S(p)\)). This leads to
-\textbf{Newton's equation} \[ H \delta  = -g \] Given a set of
-parameters \(p\), which provide \(H\) and \(g\), we solve this equation
-for \(\delta\), adjust \(p\) to \(p + \delta\) and iterate, hopefully to
-converge on a solution. Applying this to a sum of squares problem gives
-\[ H \delta / 2 = (J'J + D) \delta = - J' r\] In this expression, only
-the elements of \(D\) have second partial derivatives. Gauss, attempting
-to model planetary orbits, had small residuals, and noted that these
-multiplied the second partial derivatives of \(r\), so he approximated
-\[ H / 2 \approx J' J\] by assuming \(D \approx 0\). This results in the
-Gauss-Newton method where we solve \[ J' J \delta = - J' r\] though we
-can avoid some loss of accuracy by not forming the inner product matrix
+Let us use \(D_{i,j}\) for the elements of the second term of this expression.
+What is generally called \textbf{Newton's method} for function minimization tries
+to set the gradient to zero (to find a stationary point of the function \(S(p)\)).
+This leads to \textbf{Newton's equation}
+\[ H \delta  = -g \]
+Given a set of parameters \(p\), which provide \(H\) and \(g\), we solve this equation for \(\delta\), adjust \(p\)
+to \(p + \delta\) and iterate, hopefully to converge on a solution.
+Applying this to a sum of squares problem gives
+\[ H \delta / 2 = (J'J + D) \delta = - J' r\]
+In this expression, only the elements of \(D\) have second partial derivatives.
+Gauss, attempting to model planetary orbits, had small residuals, and noted
+that these multiplied the second partial derivatives of \(r\), so he approximated
+\[ H / 2 \approx J' J\]
+by assuming \(D \approx 0\). This results in the Gauss-Newton method where we
+solve
+\[ J' J \delta = - J' r\]
+though we can avoid some loss of accuracy by not forming the inner product matrix
 \(J' J\) and solving the linear least squares matrix problem
-\[ J \delta \approx -r \] by one of several matrix decomposition
-methods. On the other hand, when \(m >> n\), the matrix \(J' J\) uses
-much less storage than \(J\).
-
-In reality, there are many problems where \(D\) should not be ignored,
-but the work to compute it precisely is considerable. Many work-arounds
-have been proposed, of which the Levenberg-Marquardt stabilization
-(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt
-(\protect\hyperlink{ref-Marquardt1963}{1963})) is the most commonly
-used. For convenience, we will use ``Marquardt'', as we believe he first
-incorporated the ideas into a practical computer program.
-
-The usual suggestion is that \(D\) be replaced by a multiple of the unit
-matrix or else a multiple of the diagonal part of \(J' J\). In low
-precision, some elements of \(J' J\) could underflow to zero (John C.
-Nash (\protect\hyperlink{ref-jn77ima}{1977})), and a linear combination
-of both choices is an effective compromise. Various choices for \(D\),
-as well as a possible line search along the direction \(\delta\) rather
-than a unit step (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})),
-give rise to several variant algorithms. ``Marquardt's method'' is a
-family of methods. Fortunately, most choices work well.
+\[ J \delta \approx -r \]
+by one of several matrix decomposition methods. On the other hand, when \(m >> n\),
+the matrix \(J' J\) uses much less storage than \(J\).
+
+In reality, there are many problems where \(D\) should not be ignored, but the work
+to compute it precisely is considerable. Many work-arounds have been
+proposed, of which the Levenberg-Marquardt stabilization
+(Levenberg (\protect\hyperlink{ref-Levenberg1944}{1944}), Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963})) is
+the most commonly used. For convenience, we will use ``Marquardt'', as we believe he
+first incorporated the ideas into a practical computer program.
+
+The usual suggestion is that \(D\) be replaced by a multiple of the unit matrix or else
+a multiple of the diagonal part of \(J' J\). In low precision, some elements of
+\(J' J\) could underflow to zero (John C. Nash (\protect\hyperlink{ref-jn77ima}{1977})), and a linear combination of both
+choices is an effective compromise. Various choices for
+\(D\), as well as a possible line search along the
+direction \(\delta\) rather than a unit step (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})),
+give rise to several variant
+algorithms. ``Marquardt's method'' is a family of methods. Fortunately, most choices
+work well.
 
@@
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are not equivalent and are
-deliberately crude choices. Workers performing many calculations of a
-similar nature should try to provide good starting points to reduce
+values for parameters. These starting points are not equivalent and
+are deliberately crude choices. Workers performing many calculations
+of a similar nature should try to provide good starting points to reduce
 computation time and avoid finding a false solution.
 
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# model formulas}
-\NormalTok{frmu }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ b1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ b2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{b3 }\SpecialCharTok{*}\NormalTok{ tt))}
-\NormalTok{frms }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}} \DecValTok{100}\SpecialCharTok{*}\NormalTok{ c1 }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \DecValTok{10} \SpecialCharTok{*}\NormalTok{ c2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FloatTok{0.1} \SpecialCharTok{*}\NormalTok{c3 }\SpecialCharTok{*}\NormalTok{ tt))}
-\NormalTok{frmt }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
-\CommentTok{\#}
-\CommentTok{\# Starting parameter sets}
-\NormalTok{stu1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =}\DecValTok{1}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{b3 =} \DecValTok{1}\NormalTok{)}
-\NormalTok{sts1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{1}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{1}\NormalTok{)}
-\NormalTok{stt1}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{)}
-\end{Highlighting}
-\end{Shaded}
-
-One of the useful features of \texttt{nls()} is the possibility of a
-\texttt{selfStart} model, where starting parameter values are not
-required. However, if a \texttt{selfStart} model is not available,
-\texttt{nls()} sets all the starting parameters to 1. This is tolerable,
-but could be improved by using a set of values that are all slightly
-different, which, in the case of the example model
-\(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\) would
-avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
-Program modifications to give a sequence like 1.0, 1.1, 1.2, 1.3 for the
-four parameters are fairly obvious.
-
-It is also possible to provide R functions for the residual and
-Jacobian. \texttt{nls()} is not well-suited to using such a formulation,
-but \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} work
-specifically with such a formulation. This is usually more work for the
-user if the formula setup is possible. To illustrate, we show the
-functions for the unscaled 3 parameter logistic.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Logistic3U}
-\NormalTok{hobbs.res  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x)\{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} residual}
-  \CommentTok{\# This variant uses looping}
-  \ControlFlowTok{if}\NormalTok{(}\FunctionTok{length}\NormalTok{(x) }\SpecialCharTok{!=} \DecValTok{3}\NormalTok{) }\FunctionTok{stop}\NormalTok{(}\StringTok{"hobbs.res {-}{-} parameter vector n!=3"}\NormalTok{)}
-\NormalTok{  y  }\OtherTok{\textless{}{-}}  \FunctionTok{c}\NormalTok{(}\FloatTok{5.308}\NormalTok{, }\FloatTok{7.24}\NormalTok{, }\FloatTok{9.638}\NormalTok{, }\FloatTok{12.866}\NormalTok{, }\FloatTok{17.069}\NormalTok{, }\FloatTok{23.192}\NormalTok{, }\FloatTok{31.443}\NormalTok{, }
-           \FloatTok{38.558}\NormalTok{, }\FloatTok{50.156}\NormalTok{, }\FloatTok{62.948}\NormalTok{, }\FloatTok{75.995}\NormalTok{, }\FloatTok{91.972}\NormalTok{)}
-\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  res  }\OtherTok{\textless{}{-}}\NormalTok{  x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)) }\SpecialCharTok{{-}}\NormalTok{ y}
-\NormalTok{\}}
-
-\NormalTok{hobbs.jac  }\OtherTok{\textless{}{-}}  \ControlFlowTok{function}\NormalTok{(x) \{ }\CommentTok{\# scaled Hobbs weeds problem {-}{-} Jacobian}
-\NormalTok{  jj  }\OtherTok{\textless{}{-}}  \FunctionTok{matrix}\NormalTok{(}\FloatTok{0.0}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{3}\NormalTok{)}
-\NormalTok{  tt  }\OtherTok{\textless{}{-}}  \DecValTok{1}\SpecialCharTok{:}\DecValTok{12}
-\NormalTok{  yy  }\OtherTok{\textless{}{-}}  \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{3}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt)}
-\NormalTok{  zz  }\OtherTok{\textless{}{-}}  \FloatTok{1.0} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ yy)}
-\NormalTok{  jj[tt,}\DecValTok{1}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   zz}
-\NormalTok{  jj[tt,}\DecValTok{2}\NormalTok{]   }\OtherTok{\textless{}{-}}   \SpecialCharTok{{-}}\NormalTok{x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy}
-\NormalTok{  jj[tt,}\DecValTok{3}\NormalTok{]   }\OtherTok{\textless{}{-}}\NormalTok{   x[}\DecValTok{1}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ zz }\SpecialCharTok{*}\NormalTok{ yy }\SpecialCharTok{*}\NormalTok{ x[}\DecValTok{2}\NormalTok{] }\SpecialCharTok{*}\NormalTok{ tt}
-  \FunctionTok{attr}\NormalTok{(jj, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}}\NormalTok{ jj}
-\NormalTok{  jj}
-\NormalTok{\}}
-\end{Highlighting}
-\end{Shaded}
+\begin{verbatim}
+# model formulas
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100* c1 / (1 + 10 * c2 * exp(-0.1 *c3 * tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
+#
+# Starting parameter sets
+stu1<-c(b1 =1, b2 = 1, b3 = 1)
+sts1<-c(c1 = 1, c2 = 1, c3 = 1)
+stt1<-c(Asym = 1, xmid = 1, scal = 1)
+\end{verbatim}
+
+One of the useful features of \texttt{nls()} is the possibility of a \texttt{selfStart} model,
+where starting parameter values are not required. However, if a \texttt{selfStart} model
+is not available, \texttt{nls()} sets all the starting parameters
+to 1. This is tolerable, but could be improved by using a set of values
+that are all slightly different, which, in the case of the example
+model \(y \,\sim\, a \,*\, exp(-b \,*\, x) + c\,*\,exp(-d \,*\, x)\)
+would avoid a singular Jacobian because \(b\) and \(d\) were equal in value.
+Program modifications to give a sequence like 1.0, 1.1, 1.2, 1.3 for the four
+parameters are fairly obvious.
+
+It is also possible to provide R functions for the residual and Jacobian.
+\texttt{nls()} is not well-suited to using such a formulation, but \texttt{nlsr::nlfb()}
+and \texttt{minpack.lm::nls.lm()} work specifically with such a formulation.
+This is usually more work for the user if the formula setup is possible.
+To illustrate, we show the functions for the unscaled 3 parameter logistic.
+
+\begin{verbatim}
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3] * tt)
+  zz  <-  1.0 / (1 + x[2] * yy)
+  jj[tt,1]   <-   zz
+  jj[tt,2]   <-   -x[1] * zz * zz * yy
+  jj[tt,3]   <-   x[1] * zz * zz * yy * x[2] * tt
+  attr(jj, "gradient") <- jj
+  jj
+}
+\end{verbatim}
 
 \hypertarget{estimation-of-models-specified-as-formulas}{%
-\subsection{Estimation of models specified as
-formulas}\label{estimation-of-models-specified-as-formulas}}
-
-Using a formula specification was a principal advantage made with
-\texttt{nls()} when it became available in S sometime in the 1980s. It
-uses a Gauss-Newton (i.e., unstabilized) iteration with a step reduction
-line search (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})). This
-works very efficiently as long as \(J\) is not ill-conditioned. Below we
-see the default \texttt{nls()} algorithm does poorly on the example
-problem. To save page space, we use 1-line result display functions from
-package \texttt{nlsr}, namely \texttt{pnls()} and \texttt{pshort()}.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frmu, start = stu1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frms, start = sts1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnls1}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Error in nls(formula = frmt, start = stt1, data = weeddf) : 
-##   singular gradient
-\end{verbatim}
-
-Here we see the infamous ``singular gradient'' termination message of
-\texttt{nls()}. Technically, it is the Jacobian that is singular, but R
-does not distinguish scalar and vector functions, which yield partial
-derivatives that are a gradient or Jacobian respectively.
-
-\texttt{nls()} also offers the option of using the ``port'' library
-algorithm choice (Fox, Hall, and Schryer
-(\protect\hyperlink{ref-FoxHallSchryer1978}{1978})). There is a warning
-in the \texttt{nls()} documentation that this ``appears unfinished'',
-but it is successful in solving all three cases.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(unls1p)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(snls1p)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnls1p}\OtherTok{\textless{}{-}}\FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(tnls1p)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
+\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}
+
+Using a formula specification was a principal advantage made
+with \texttt{nls()} when it became available in S sometime in the 1980s. It uses a
+Gauss-Newton (i.e., unstabilized) iteration with a step reduction line
+search (Hartley (\protect\hyperlink{ref-Hartley1961}{1961})). This works very efficiently as long as \(J\) is not ill-conditioned.
+Below we see the default \texttt{nls()} algorithm does poorly on the example problem.
+To save page space, we use 1-line result display functions from package \texttt{nlsr},
+namely \texttt{pnls()} and \texttt{pshort()}.
+
+\begin{verbatim}
+unls1<-try(nls(formula = frmu, start = stu1, data = weeddf))
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nls(formula = frmu, start = stu1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+snls1<-try(nls(formula = frms, start = sts1, data = weeddf))
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nls(formula = frms, start = sts1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+tnls1<-try(nls(formula = frmt, start = stt1, data = weeddf))
+\end{verbatim}
+
+\begin{verbatim}
+#> Error in nls(formula = frmt, start = stt1, data = weeddf) : 
+#>   singular gradient
+\end{verbatim}
+
+Here we see the infamous ``singular gradient'' termination message of \texttt{nls()}.
+Technically, it is the Jacobian that is singular, but R does not distinguish scalar
+and vector functions, which yield partial derivatives that are a gradient or Jacobian
+respectively.
+
+\texttt{nls()} also offers the option of using the ``port'' library algorithm
+choice (Fox, Hall, and Schryer (\protect\hyperlink{ref-FoxHallSchryer1978}{1978})). There is a warning in the \texttt{nls()} documentation that
+this ``appears unfinished'', but it is successful in solving all three cases.
+
+\begin{verbatim}
+unls1p<-try(nls(formula = frmu, start = stu1, data = weeddf, algorithm = "port"))
+pnls(unls1p)
+\end{verbatim}
+
+\begin{verbatim}
+#> unls1p  -- ss= 2.587277 :  b1 = 196.1863  b2 = 49.09164  b3 = 0.3135697; 17  itns
+\end{verbatim}
+
+\begin{verbatim}
+snls1p<-try(nls(formula = frms, start = sts1, data = weeddf, algorithm = "port"))
+pnls(snls1p)
+\end{verbatim}
+
+\begin{verbatim}
+#> snls1p  -- ss= 2.587277 :  c1 = 1.961863  c2 = 4.909164  c3 = 3.135697; 15  itns
+\end{verbatim}
+
+\begin{verbatim}
+tnls1p<-try(nls(formula = frmt, start = stt1, data = weeddf, algorithm = "port"))
+pnls(tnls1p)
+\end{verbatim}
+
+\begin{verbatim}
+#> tnls1p  -- ss= 2.587277 :  Asym = 196.1863  xmid = 12.4173  scal = 3.189083; 12  itns
 \end{verbatim}
@@
 \hypertarget{solution-attempts-with-nlsr}{%
-\subsubsection{Solution attempts with
-nlsr}\label{solution-attempts-with-nlsr}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{print}\NormalTok{(unlx1) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.5873  on  12 observations
-##     after  19    Jacobian and  25 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
-## b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
-## b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{pshort}\NormalTok{(snlx1) }\CommentTok{\# a short{-}form output}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnlx1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{pshort}\NormalTok{(tnlx1) }\CommentTok{\# alternatively print(tnlx1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
+\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+
+\begin{verbatim}
+unlx1 <- try(nlxb(formula = frmu, start = stu1, data = weeddf))
+print(unlx1) 
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  2.5873  on  12 observations
+#>     after  19    Jacobian and  25 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> b1               196.186         11.31      17.35  3.167e-08  -4.859e-09        1011  
+#> b2               49.0916         1.688      29.08  3.284e-10  -3.099e-08      0.4605  
+#> b3               0.31357      0.006863      45.69  5.768e-12   2.305e-06     0.04714
+\end{verbatim}
+
+\begin{verbatim}
+snlx1 <- try(nlxb(formula = frms, start = sts1, data = weeddf))
+pshort(snlx1) # a short-form output
+\end{verbatim}
+
+\begin{verbatim}
+#> snlx1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 34 res/ 23 jac
+\end{verbatim}
+
+\begin{verbatim}
+tnlx1 <- try(nlxb(formula = frmt, start = stt1, data = weeddf))
+pshort(tnlx1) # alternatively print(tnlx1)
+\end{verbatim}
+
+\begin{verbatim}
+#> tnlx1  -- ss= 2.5873 :  Asym = 196.19  xmid = 12.417  scal = 3.1891; 36 res/ 27 jac
 \end{verbatim}
@@
 Though we have found solutions, the Jacobian is essentially singular as
-shown by its singular values. Note that these are \textbf{displayed} by
-package \texttt{nlsr} in a single column in the output to provide a
-compact layout, but the values do \textbf{not} correspond to the
-individual parameters in whose row they appear; they are a property of
-the whole problem.
+shown by its singular
+values. Note that these are \textbf{displayed} by package \texttt{nlsr} in a single column in
+the output to provide a compact layout, but the values do \textbf{not} correspond to
+the individual parameters in
+whose row they appear; they are a property of the whole problem.
 
 \hypertarget{solution-attempts-with-minpack.lm}{%
-\subsubsection{Solution attempts with
-minpack.lm}\label{solution-attempts-with-minpack.lm}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{unlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmu, }\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{pnls}\NormalTok{(unlm1)  }\CommentTok{\# Short form of output}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{snlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{pnls}\NormalTok{(snlm1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tnlm1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stt1, }\AttributeTok{data =}\NormalTok{ weeddf))}
-\FunctionTok{pnls}\NormalTok{(tnlm1) }\CommentTok{\# short form to give sum of squares, else use summary(tnlm1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
+\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}
+
+\begin{verbatim}
+unlm1 <- try(nlsLM(formula = frmu, start = stu1, data = weeddf))
+pnls(unlm1)  # Short form of output
+\end{verbatim}
+
+\begin{verbatim}
+#> unlm1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
+\end{verbatim}
+
+\begin{verbatim}
+snlm1 <- try(nlsLM(formula = frms, start = sts1, data = weeddf))
+pnls(snlm1)
+\end{verbatim}
+
+\begin{verbatim}
+#> snlm1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 7  itns
+\end{verbatim}
+
+\begin{verbatim}
+tnlm1 <- try(nlsLM(formula = frmt, start = stt1, data = weeddf))
+pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
+\end{verbatim}
+
+\begin{verbatim}
+#> tnlm1  -- ss= 9205.4 :  Asym = 35.532  xmid = 43376  scal = -2935.4; 39  itns
 \end{verbatim}
@@
 \hypertarget{solution-attempts-with-gslnls}{%
-\subsubsection{Solution attempts with
-gslnls}\label{solution-attempts-with-gslnls}}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{library}\NormalTok{(gslnls)}
-\NormalTok{ugslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmu, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stu1))}
-\FunctionTok{pnls}\NormalTok{(ugslnls1) }\CommentTok{\# to get sum of squares}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{sgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frms, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ sts1))}
-\FunctionTok{pnls}\NormalTok{(sgslnls1) }\CommentTok{\# Use summary() to get display}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{tgslnls1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{fn =}\NormalTok{ frmt, }\AttributeTok{data =}\NormalTok{ weeddf,  }\AttributeTok{start =}\NormalTok{ stt1))}
-\FunctionTok{pnls}\NormalTok{(tgslnls1) }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\DocumentationTok{\#\# Other algorithms}
-\DocumentationTok{\#\# tgslnls1a\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))}
-\DocumentationTok{\#\# pnls(tgslnls1a) }
-\DocumentationTok{\#\# tgslnls1d\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))}
-\DocumentationTok{\#\# pnls(tgslnls1d) }
-\DocumentationTok{\#\# tgslnls1dd\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))}
-\DocumentationTok{\#\# pnls(tgslnls1dd) }
-\DocumentationTok{\#\# tgslnls1s2\textless{}{-}try(gsl\_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))}
-\DocumentationTok{\#\# pnls(tgslnls1s2) }
-\end{Highlighting}
-\end{Shaded}
-
-While default \texttt{gsl\_nls()} gives much of the same results as
-\texttt{nlsLM()}, uncommenting the calls to the alternative algorithms
-shows it can get the global minimum with any of these choices.
+\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}
+
+\begin{verbatim}
+library(gslnls)
+ugslnls1 <- try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
+pnls(ugslnls1) # to get sum of squares
+\end{verbatim}
+
+\begin{verbatim}
+#> ugslnls1  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+\end{verbatim}
+
+\begin{verbatim}
+sgslnls1 <- try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
+pnls(sgslnls1) # Use summary() to get display
+\end{verbatim}
+
+\begin{verbatim}
+#> sgslnls1  -- ss= 2.5873 :  c1 = 1.9619  c2 = 4.9092  c3 = 3.1357; 9  itns
+\end{verbatim}
+
+\begin{verbatim}
+tgslnls1 <- try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
+pnls(tgslnls1) 
+\end{verbatim}
+
+\begin{verbatim}
+#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
+\end{verbatim}
+
+\begin{verbatim}
+## Other algorithms
+## tgslnls1a<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="lmaccel"))
+## pnls(tgslnls1a) 
+## tgslnls1d<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="dogleg"))
+## pnls(tgslnls1d) 
+## tgslnls1dd<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="ddogleg"))
+## pnls(tgslnls1dd) 
+## tgslnls1s2<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1, algorithm="subspace2D"))
+## pnls(tgslnls1s2) 
+\end{verbatim}
+
+While default \texttt{gsl\_nls()} gives much of the same results as \texttt{nlsLM()}, uncommenting the
+calls to the alternative algorithms shows it can get the global minimum with any of
+these choices.
 
 \hypertarget{comparison-notes-for-formula-setup-solutions}{%
-\subsubsection{Comparison notes for formula-setup
-solutions}\label{comparison-notes-for-formula-setup-solutions}}
-
-\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors
-and singular values of the Jacobian (for diagnostic purposes). By
-contrast, \texttt{minpack.lm::nlsLM()} and \texttt{nls()} use
-\texttt{summary()}, which does not display the sum of squares, while
-\texttt{print()} gives the sum of squares, but not the standard error of
-the residuals. Such behaviour is consistent with other modeling
-functions such as \texttt{lm()}.
-
-The importance of the singular values is to allow us to gauge how
-``nearly singular'' the Jacobian is at the solution, and the ratio of
-the smallest to largest of the singular values is a simple but effective
-measure. The ratios are 4.6641e-05 for Logistic3U, 0.021022 for
-Logistic3S, and 0.001055 for Logistic3T; so Logistic3S is the ``least
-singular''.
-
-The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the
-transformed model Logistic3T have a very large sum of squares, which may
-suggest that these programs have failed. Since \texttt{nls()},
-\texttt{nlsLM()}, and \texttt{gsl\_nls()} do not offer singular values,
-we need to extract the Jacobian and compute its singular values. The
-following script shows how to do this, using as Jacobian the
-\texttt{gradient} element in the returned solution for these solvers.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# for nlsLM}
-\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tnlm1, }\StringTok{"try{-}error"}\NormalTok{))  \{}
-   \FunctionTok{print}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
-\NormalTok{ \} }\ControlFlowTok{else}\NormalTok{ \{  }
-\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tnlm1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{() }\CommentTok{\# actually the Jacobian}
-   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
-\NormalTok{\}   }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 3.4641e+00 3.2530e-10 8.9643e-12
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# for gsl\_nls}
-\ControlFlowTok{if}\NormalTok{ (}\FunctionTok{inherits}\NormalTok{(tgslnls1, }\StringTok{"try{-}error"}\NormalTok{)) \{}
-   \FunctionTok{cat}\NormalTok{(}\StringTok{"Cannot compute solution {-}{-} likely singular Jacobian"}\NormalTok{)}
-\NormalTok{\} }\ControlFlowTok{else}\NormalTok{ \{  }
-\NormalTok{   JtnlsLM }\OtherTok{\textless{}{-}}\NormalTok{ tgslnls1}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{gradient}\NormalTok{()}
-   \FunctionTok{svd}\NormalTok{(JtnlsLM)}\SpecialCharTok{$}\NormalTok{d }\CommentTok{\# Singular values}
-\NormalTok{\}   }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## [1] 3.4641e+00 9.4495e-09 4.1324e-11
-\end{verbatim}
-
-We see that there are differences in detail, but the more important
-result is that two out of three singular values are essentially 0. Our
-Jacobian is singular, and no method of the Gauss-Newton type should be
-able to continue. Indeed, from the parameters reported at this saddle
-point, \texttt{nlsr::nlxb()} cannot proceed.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{stspecial }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \FloatTok{35.532}\NormalTok{,  }\AttributeTok{xmid =} \DecValTok{43376}\NormalTok{,  }\AttributeTok{scal =} \SpecialCharTok{{-}}\FloatTok{2935.4}\NormalTok{)}
-\NormalTok{badstart }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(}\AttributeTok{formula =}\NormalTok{ frmt, }\AttributeTok{start =}\NormalTok{ stspecial, }\AttributeTok{data =}\NormalTok{ weeddf)}
-\FunctionTok{print}\NormalTok{(badstart)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  9205.4  on  12 observations
-##     after  2    Jacobian and  2 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## Asym             35.5321            NA         NA         NA  -9.694e-09       3.464  
-## xmid               43376            NA         NA         NA  -1.742e-09    2.61e-10  
-## scal             -2935.4            NA         NA         NA    -2.4e-08    7.12e-16
-\end{verbatim}
-
-We also note that changing the start to
+\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}
+
+\texttt{nlsr::nlxb()} uses \texttt{print()} to output standard errors and
+singular values of the Jacobian (for diagnostic purposes). By contrast,
+\texttt{minpack.lm::nlsLM()} and \texttt{nls()}
+use \texttt{summary()}, which does not display the sum of squares, while \texttt{print()}
+gives the sum of squares, but not the standard error of the residuals.
+Such behaviour is consistent with other modeling functions such as \texttt{lm()}.
+
+The importance of the singular values is to allow us to gauge how ``nearly singular''
+the Jacobian is at the solution, and the ratio of the smallest to largest of the
+singular values is a simple but effective measure. The ratios are
+4.6641e-05 for Logistic3U, 0.021022 for Logistic3S, and 0.001055 for
+Logistic3T; so Logistic3S is the ``least singular''.
+
+The results from \texttt{nlsLM} and \texttt{gsl\_nls} for the transformed model Logistic3T have a
+very large sum of squares, which may suggest that these programs have failed.
+Since \texttt{nls()}, \texttt{nlsLM()}, and \texttt{gsl\_nls()} do not
+offer singular values, we need to extract the Jacobian and compute its
+singular values. The following script shows how to do this, using as Jacobian
+the \texttt{gradient} element in the returned solution for these solvers.
+
+\begin{verbatim}
+# for nlsLM
+if (inherits(tnlm1, "try-error"))  {
+   print("Cannot compute solution -- likely singular Jacobian")
+ } else {  
+   JtnlsLM <- tnlm1$m$gradient() # actually the Jacobian
+   svd(JtnlsLM)$d # Singular values
+}   
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] 3.4641e+00 3.2530e-10 8.9643e-12
+\end{verbatim}
+
+\begin{verbatim}
+# for gsl_nls
+if (inherits(tgslnls1, "try-error")) {
+   cat("Cannot compute solution -- likely singular Jacobian")
+} else {  
+   JtnlsLM <- tgslnls1$m$gradient()
+   svd(JtnlsLM)$d # Singular values
+}   
+\end{verbatim}
+
+\begin{verbatim}
+#> [1] 3.4641e+00 9.4495e-09 4.1324e-11
+\end{verbatim}
+
+We see that there are differences in detail, but the more important result is that
+two out of three singular values are essentially 0. Our Jacobian is singular, and no
+method of the Gauss-Newton type should be able to continue. Indeed, from the
+parameters reported at this saddle point, \texttt{nlsr::nlxb()} cannot proceed.
+
+\begin{verbatim}
+stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
+print(badstart)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  9205.4  on  12 observations
+#>     after  2    Jacobian and  2 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> Asym             35.5321            NA         NA         NA  -9.694e-09       3.464  
+#> xmid               43376            NA         NA         NA  -1.742e-09    2.61e-10  
+#> scal             -2935.4            NA         NA         NA    -2.4e-08    7.12e-16
+\end{verbatim}
+
+We also
+note that changing the start to
 
@@
 
-allows both \texttt{nlsLM} and \texttt{gsl\_nls} to find the minimum.
-\texttt{gsl\_nls} also succeeds when the particular form of the
-Marquardt stabilization is changed from the default by the control
-setting \texttt{scale\ =\ "levenberg"}.
-
-We consider the code inside package \texttt{nlsr} to be competently
-coded, with the advantage over \texttt{nls()} of the Levenberg-Marquardt
-stabilization in common with \texttt{minpack.lm} and \texttt{glsnls}.
-The success rate is high, but \texttt{nlsr} has its premature
-terminations too. However, it provides diagnostic information in the
-form of the singular values of the Jacobian that allows us to try
-alternative starts, algorithms or programs in cases where there are
-indications of early termination.
+allows both \texttt{nlsLM} and \texttt{gsl\_nls} to find the minimum. \texttt{gsl\_nls} also succeeds
+when the particular form of the Marquardt stabilization is changed from the
+default by the control setting \texttt{scale\ =\ "levenberg"}.
+
+We consider the code inside package \texttt{nlsr} to be competently coded,
+with the advantage over \texttt{nls()} of the Levenberg-Marquardt stabilization
+in common with \texttt{minpack.lm} and \texttt{glsnls}. The success rate is high,
+but \texttt{nlsr} has its premature terminations too. However, it provides
+diagnostic information in the form of the singular values of the Jacobian
+that allows us to try alternative starts, algorithms or programs in cases
+where there are indications of early termination.
 
 \hypertarget{functional-specification-of-problems}{%
-\subsection{Functional specification of
-problems}\label{functional-specification-of-problems}}
-
-Let us write a slightly modified form of the nonlinear least squares
-problem:
+\subsection{Functional specification of problems}\label{functional-specification-of-problems}}
+
+Let us write a slightly modified form of the nonlinear least squares problem:
 
@@
 
-R users are likely to think almost always that the functions \(r_i\) are
-all of the same form, differing in the details of the data they use.
-However, the famous Rosenbrock banana-shaped valley (Rosenbrock
-(\protect\hyperlink{ref-Rosenbrock1960}{1960})) has
-\[ r_1(p) = (1 - p_1)\] \[ r_2(p) = 10 (p_2 - {p_1}^2)\]
-
-Thus, there can be problems where the idea of a ``model'' or a
-``prediction'' makes no sense, though the nonlinear least squares
-problem is still well-posed. In our code, we still refer to the
-functions as ``residuals'', though this is purely a convenience.
-
-We illustrate how to solve nonlinear least squares problems using a
-function to define the residual. Note that \texttt{gsl\_nls()} requires
-a vector \texttt{y} that is the expected value of what we have called
-the residual, but in this case is actually the model. \texttt{gsl\_nls}
-uses a numerical approximation for the Jacobian if the argument
-\texttt{jac} is missing. Note functions \texttt{nlsr::pnls()} and
-\texttt{nslr::pnlslm()} for a 1-line display of the results.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobnlfb }\OtherTok{\textless{}{-}} \FunctionTok{nlfb}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{resfn =}\NormalTok{ hobbs.res, }\AttributeTok{jacfn =}\NormalTok{ hobbs.jac)}
-\FunctionTok{pshort}\NormalTok{(hobnlfb) }\CommentTok{\# use print(hobnlfb) for more detail}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobnlm }\OtherTok{\textless{}{-}} \FunctionTok{nls.lm}\NormalTok{(}\AttributeTok{par =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{jac =}\NormalTok{ hobbs.jac)}
-\FunctionTok{pnlslm}\NormalTok{(hobnlm)  }
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobgsln }\OtherTok{\textless{}{-}} \FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{y =} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(hobgsln)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{hobgsl }\OtherTok{\textless{}{-}} \FunctionTok{gsl\_nls}\NormalTok{(}\AttributeTok{start =}\NormalTok{ stu1, }\AttributeTok{fn =}\NormalTok{ hobbs.res, }\AttributeTok{y =} \FunctionTok{rep}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{12}\NormalTok{), }\AttributeTok{jac =}\NormalTok{ hobbs.jac)}
-\FunctionTok{pnls}\NormalTok{(hobgsl) }\CommentTok{\# using analytic Jacobian}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+R users are likely to think almost always that the functions \(r_i\) are all of
+the same form, differing in the details of the data they use. However, the
+famous Rosenbrock banana-shaped valley (Rosenbrock (\protect\hyperlink{ref-Rosenbrock1960}{1960})) has
+\[ r_1(p) = (1 - p_1)\]
+\[ r_2(p) = 10 (p_2 - {p_1}^2)\]
+
+Thus, there can be problems where the idea of a ``model'' or a ``prediction''
+makes no sense, though the nonlinear least squares problem is still
+well-posed. In our code, we still refer to the functions as ``residuals'',
+though this is purely a convenience.
+
+We illustrate how to solve nonlinear least squares problems using
+a function to define the residual. Note that \texttt{gsl\_nls()} requires
+a vector \texttt{y} that is the expected value of what we have called the
+residual, but in this case is actually the model. \texttt{gsl\_nls} uses
+a numerical approximation for the Jacobian if the argument \texttt{jac}
+is missing. Note functions \texttt{nlsr::pnls()} and \texttt{nslr::pnlslm()} for a
+1-line display of the results.
+
+\begin{verbatim}
+hobnlfb <- nlfb(start = stu1, resfn = hobbs.res, jacfn = hobbs.jac)
+pshort(hobnlfb) # use print(hobnlfb) for more detail
+\end{verbatim}
+
+\begin{verbatim}
+#> hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
+\end{verbatim}
+
+\begin{verbatim}
+hobnlm <- nls.lm(par = stu1, fn = hobbs.res, jac = hobbs.jac)
+pnlslm(hobnlm)  
+\end{verbatim}
+
+\begin{verbatim}
+#> hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
+\end{verbatim}
+
+\begin{verbatim}
+hobgsln <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12))
+pnls(hobgsln)
+\end{verbatim}
+
+\begin{verbatim}
+#> hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
+\end{verbatim}
+
+\begin{verbatim}
+hobgsl <- gsl_nls(start = stu1, fn = hobbs.res, y = rep(0,12), jac = hobbs.jac)
+pnls(hobgsl) # using analytic Jacobian
+\end{verbatim}
+
+\begin{verbatim}
+#> hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
 \end{verbatim}
@@
 \hypertarget{design-goals-termination-tests-and-output-objects}{%
-\subsection{Design goals, termination tests, and output
-objects}\label{design-goals-termination-tests-and-output-objects}}
-
-The output object of \texttt{nlxb()} is smaller than the class
-\texttt{nls} object returned by \texttt{nls()}, \texttt{nlsLM()}, and
-\texttt{gsl\_nls()}. Package \texttt{nlsr} emphasizes the solution of
-the nonlinear least squares problem rather than the estimation of a
-nonlinear model that fits or explains the data. The object of class
-\texttt{nls} allows for a number of specialized modeling and diagnostic
-extensions. For compatibility, the \texttt{nlsr} package has the
-function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias,
-that is intended for problems specified as formulas. This uses
-\texttt{nlxb()} to find good parameters, then calls \texttt{nls()} to
-return the class \texttt{nls} object. Unless particular modeling
-features are needed, the use of \texttt{wrapnlsr()} is unnecessary and
-wasteful of resources.
-
-The design goals of the different tools may also be revealed in the
-so-called ``convergence tests'' for the iterative solvers. In the manual
-page for \texttt{nls()} in R 4.0.0 there was the warning:
+\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}
+
+The output object of \texttt{nlxb()} is smaller than the class \texttt{nls} object returned
+by \texttt{nls()}, \texttt{nlsLM()}, and \texttt{gsl\_nls()}. Package \texttt{nlsr} emphasizes the solution
+of the nonlinear least squares problem rather than the estimation of a nonlinear
+model that fits or explains the data. The object of class \texttt{nls} allows for
+a number of specialized modeling and diagnostic extensions. For compatibility,
+the \texttt{nlsr} package has the function \texttt{wrapnlsr()}, for which \texttt{nlsr()} is an alias,
+that is intended for problems specified as formulas.
+This uses \texttt{nlxb()} to find good parameters, then calls \texttt{nls()} to return the class
+\texttt{nls} object. Unless particular modeling features are needed, the use of
+\texttt{wrapnlsr()} is unnecessary and wasteful of resources.
+
+The design goals of the different tools may also be revealed in the so-called
+``convergence tests'' for the iterative solvers. In the manual page for \texttt{nls()}
+in R 4.0.0 there was the warning:
 
@@
 with suggested addition of small perturbations to the data. This admits
-\texttt{nls()} could not solve well-posed problems unless data is
-polluted with errors. Zero-residual problems are not always artificial,
+\texttt{nls()} could not solve well-posed problems unless data is polluted
+with errors. Zero-residual problems are not always artificial,
 since problems in function approximation and nonlinear equations can be
 approached with nonlinear least squares. Fortunately, a small adjustment
-to the ``termination test'' for the \textbf{program}, rather than for
-the ``convergence'' of the underlying \textbf{algorithm}, fixes the
-defect. The test is the Relative Offset Convergence Criterion (see
-Bates, Douglas M. and Watts, Donald G.
-(\protect\hyperlink{ref-BatesWatts81}{1981})). This scales an estimated
-reduction in the loss function by its current value. If the loss
-function is very small, we are close to a zero-divide. Adding a small
-quantity to the divisor avoids trouble. In 2021, one of us (J. Nash)
-proposed that \texttt{nls.control()} have an additional parameter
-\texttt{scaleOffset} with a default value of zero. Setting it to a small
-number -- 1.0 is a reasonable choice -- allows small-residual problems
+to the ``termination test'' for the \textbf{program}, rather than for the
+``convergence'' of the underlying \textbf{algorithm}, fixes the defect. The test
+is the Relative Offset Convergence
+Criterion (see Bates, Douglas M. and Watts, Donald G. (\protect\hyperlink{ref-BatesWatts81}{1981})). This scales an estimated reduction in the loss
+function by its current value. If the loss function is very small, we are close
+to a zero-divide. Adding a small quantity to the divisor avoids trouble.
+In 2021, one of us (J. Nash) proposed that \texttt{nls.control()} have an
+additional parameter \texttt{scaleOffset} with a default value of zero. Setting it to a
+small number -- 1.0 is a reasonable choice -- allows small-residual problems
 (i.e., near-exact fits) to be dealt with easily. We call this the
-\textbf{safeguarded relative offset convergence criterion}, and it has
-been in \texttt{nlsr} since it was introduced. The default value gives
-\texttt{nls()} its legacy behaviour. This improvement has been in the R
-distributed code since version 4.1.0.
-
-Additional termination tests can be used. \texttt{nlsr} has a
-\textbf{small sum of squares} test (\texttt{smallsstest}) that compares
-the latest evaluated sum of squared (weighted) residuals to \texttt{e4}
-times the initial sum of squares, where
-\texttt{e4\ \textless{}-\ (100\ *\ .Machine\$double.eps)\ \^{}\ 4} is
-approximately 2.43e-55.
-
-Termination after what may be considered excessive computation is also
-important. \texttt{nls()} stops after \texttt{maxiter} ``iterations''.
-The meaning of ``iteration'' may require an examination of the code.
-\texttt{nlsr} terminates when the number of residual or Jacobian
-evaluations (\texttt{feval} and \texttt{jeval}) exceed set limits, and
-we prefer larger limits (\texttt{jemax\ =\ 5000} is the default) than
-the \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
-This may cause overly long run-times for \texttt{nlsr}, but our work
-often involves problems that are sufficiently difficult that we want to
-be sure we do not terminate early.
+\textbf{safeguarded relative offset convergence criterion}, and it has been in \texttt{nlsr}
+since it was introduced. The default value
+gives \texttt{nls()} its legacy behaviour. This improvement has been in the R distributed code
+since version 4.1.0.
+
+Additional termination tests can be used. \texttt{nlsr} has a \textbf{small sum of
+squares} test (\texttt{smallsstest}) that compares the latest evaluated sum of squared
+(weighted) residuals to \texttt{e4} times the initial sum of squares, where
+\texttt{e4\ \textless{}-\ (100\ *\ .Machine\$double.eps)\ \^{}\ 4} is approximately 2.43e-55.
+
+Termination after what may be considered excessive computation is also important.
+\texttt{nls()} stops after \texttt{maxiter} ``iterations''. The meaning of ``iteration'' may require an
+examination of the code. \texttt{nlsr} terminates when the number of residual or
+Jacobian evaluations (\texttt{feval} and \texttt{jeval}) exceed set limits, and we prefer larger
+limits (\texttt{jemax\ =\ 5000} is the default) than the \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid
+stopping early. This may cause overly long run-times for \texttt{nlsr}, but our work often
+involves problems that are sufficiently difficult that we want to be sure we do not
+terminate early.
 
 \hypertarget{returned-results-of-nls-and-other-tools}{%
-\subsubsection{Returned results of nls() and other
-tools}\label{returned-results-of-nls-and-other-tools}}
-
-As mentioned, the output of \texttt{nls()} is an object of class ``nls''
-which has a quite rich structure described in the manual file or
-revealed by applying the \texttt{str()} function to the result of
-\texttt{nls()}.
+\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}
+
+As mentioned, the output of \texttt{nls()} is an object of class ``nls'' which has a quite
+rich structure described in the manual file or revealed by applying the \texttt{str()}
+function to the result of \texttt{nls()}.
 
 The complexity of this object is a challenge to users. Let us use
-\texttt{result} as the name of the returned object from \texttt{nls()},
-\texttt{minpack.lm::nlsLM()} or, \texttt{gslnls::gsl\_nls()}. For
-example, the \texttt{data} return element is an R symbol. If the user's
-environment exists, which it may not, we could see this data via the
-syntax:
+\texttt{result} as the name of the returned object from \texttt{nls()}, \texttt{minpack.lm::nlsLM()} or,
+\texttt{gslnls::gsl\_nls()}. For example, the \texttt{data} return element is an R symbol.
+If the user's environment exists, which it may not, we could see this data
+via the syntax:
 
@@
 
-However, since the evaluating environment might not exist, a better
-approach is to make the call to \texttt{nls()} with
-\texttt{model\ =\ TRUE}, which returns an element \texttt{model} which
-contains the data, and we can list its contents using
+However, since the evaluating environment might not exist, a better approach
+is to make the call to \texttt{nls()}
+with \texttt{model\ =\ TRUE}, which returns an element
+\texttt{model} which contains the data, and we can list its contents using
 
@@
 
-and if there is an element called \texttt{xdata}, then it can be
-accessed as \texttt{result\$model\$xdata}.
-
-By contrast, \texttt{nlsr::nlxb()} returns a much simpler structure of
-11 items in one level. Moreover, \texttt{nlxb} explicitly returns the
-sum of squares, the residual vector, Jacobian, and counts of
-evaluations, but does not include the input data.
+and if there is an element called \texttt{xdata}, then it can be accessed as
+\texttt{result\$model\$xdata}.
+
+By contrast, \texttt{nlsr::nlxb()} returns a much simpler structure of 11 items
+in one level. Moreover, \texttt{nlxb} explicitly returns the sum of squares, the
+residual vector, Jacobian, and counts of evaluations, but does not include
+the input data.
 
 \hypertarget{when-to-compute-ancillary-information}{%
-\subsubsection{When to compute ancillary
-information}\label{when-to-compute-ancillary-information}}
-
-Tools that produce a class \texttt{nls} output object create a rich set
-of functions and structures that are then used in a variety of modeling
-tasks, including the least squares solution. By contrast, \texttt{nlsr}
-computes quantities as they are requested or needed, with additional
-features in separate functions. For example, the singular values of the
-Jacobian are actually computed in the \texttt{print} and
-\texttt{summary} methods for the result. These two approaches lead to
-different consequences for performance and how features are provided.
-\texttt{nlsr} has antecedents in the methods of John C. Nash
-(\protect\hyperlink{ref-cnm79}{1979}), where storage for data and
-programs was at a ridiculous premium in the small computers of the era.
-Thus the code in \texttt{nlsr} is likely of value for workers to copy
-and modify for customized tools, with code in R to support such uses.
+\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}
+
+Tools that produce a class \texttt{nls} output object create a rich set of functions
+and structures that are then used in a variety of modeling tasks, including the
+least squares solution. By contrast,
+\texttt{nlsr} computes quantities as they are requested or needed, with additional
+features in separate functions. For example, the singular values of the Jacobian
+are actually computed in the \texttt{print} and \texttt{summary} methods for the result.
+These two approaches
+lead to different consequences for performance and how features are
+provided. \texttt{nlsr} has antecedents in the methods of John C. Nash (\protect\hyperlink{ref-cnm79}{1979}), where storage for
+data and programs was at a ridiculous premium in the small computers of the era.
+Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
+for customized tools, with code in R to support such uses.
 
@@
 
-Gauss-Newton/Marquardt methods all need a Jacobian matrix at each
-iteration. By default, \texttt{nlsr::nlxb()} will try to evaluate this
-using analytic expressions using symbolic and automatic differentiation
-tools. When using a formula specification of the model, \texttt{nls()},
-\texttt{minpack.lm::nlsLM()} and \texttt{gslnls::gsl\_nls()} use a
-finite difference approximation to compute the Jacobian, though
-\texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
-Package \texttt{nlsr} provides, via appropriate calling syntax, four
-numeric approximation options for the Jacobian, with a further control
-\texttt{ndstep} for the size of the step used in the approximation.
-These features are mainly to allow for investigation of approximate
-derivatives. Regular users will rarely need them.
-
-Using the \texttt{gradient} attribute of the output of the Jacobian
-function to hold the Jacobian matrix lets us embed this in the
-\textbf{residual} function as well, so that the call to
-\texttt{nlsr::nlfb()} can be made with the same name used for both
-residual and Jacobian function arguments. This programming trick saves a
-lot of trouble for the package developer, but it can be a nuisance for
-users trying to understand the code.
-
-As far as we can understand the logic in \texttt{nls()}, the Jacobian
-computation during parameter estimation is carried out within the called
-C-language program and its wrapper R code function
-\texttt{numericDeriv()}, part of \texttt{./src/library/stats/R/nls.R} in
-the R distribution source code. This is used to provide Jacobian
-information in the \texttt{nlsModel()} and \texttt{nlsModel.plinear()}
-functions, which are \textbf{not} exported for general use.
-\texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.
-
-\texttt{numericDeriv()} uses a simple forward difference approximation
-of derivatives, though a central difference approximation can be
-specified in control parameters. We are unclear why
-\texttt{numericDeriv()} in base R calls \texttt{C\_numeric\_deriv}, as
-we were easily able to create a more compact version entirely in R. See
-\emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
-
-\texttt{minpack.lm::nlsLM()} invokes \texttt{numericDeriv()} in its
-local version of \texttt{nlsModel()}, but it appears to use an internal
-approximate Jacobian code from the original Fortran \texttt{minpack}
-code, namely, \texttt{lmdif.f}. Such differences in approach can lead to
-different behaviour, usually minor, but sometimes annoying with
+Gauss-Newton/Marquardt methods all need a Jacobian matrix at each iteration.
+By default, \texttt{nlsr::nlxb()} will try to evaluate this using analytic
+expressions using symbolic and automatic differentiation tools.
+When using a formula specification of the model, \texttt{nls()}, \texttt{minpack.lm::nlsLM()}
+and \texttt{gslnls::gsl\_nls()} use a finite difference approximation to compute the
+Jacobian, though \texttt{gsl\_nls()} does have an option to attempt symbolic expressions.
+Package \texttt{nlsr} provides, via appropriate calling syntax, four numeric approximation
+options for the Jacobian, with a further control \texttt{ndstep} for the size of the
+step used in the approximation. These features are mainly to allow for investigation
+of approximate derivatives. Regular users will rarely need them.
+
+Using the \texttt{gradient} attribute of the output of the Jacobian function to hold
+the Jacobian matrix lets us embed this in the \textbf{residual} function as well,
+so that the call to \texttt{nlsr::nlfb()} can be made with the same name used for
+both residual and Jacobian function arguments. This programming trick saves a
+lot of trouble for the package developer, but it can be a nuisance for users
+trying to understand the code.
+
+As far as we can understand the logic in \texttt{nls()}, the Jacobian computation during
+parameter estimation is carried out within the called C-language program
+and its wrapper R code function \texttt{numericDeriv()}, part of
+\texttt{./src/library/stats/R/nls.R}
+in the R distribution source code. This is used to provide Jacobian information in
+the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are \textbf{not} exported for
+general use. \texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.
+
+\texttt{numericDeriv()} uses a
+simple forward difference approximation of derivatives, though a central
+difference approximation can be specified in control parameters.
+We are unclear why \texttt{numericDeriv()} in base R calls \texttt{C\_numeric\_deriv},
+as we were easily able to create a more compact version entirely in R.
+See \emph{\url{https://github.com/nashjc/RNonlinearLS/tree/main/DerivsNLS}}.
+
+\texttt{minpack.lm::nlsLM()} invokes \texttt{numericDeriv()} in its local
+version of \texttt{nlsModel()}, but it appears to use an internal approximate Jacobian
+code from
+the original Fortran \texttt{minpack} code, namely, \texttt{lmdif.f}. Such differences in approach
+can lead to different behaviour, usually minor, but sometimes annoying with
 ill-conditioned problems.
@@
 \item
-  A pasture regrowth problem (Huet et al.
-  (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky
-  (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a poorly
-  conditioned Jacobian and \texttt{nls()} fails with ``singular
-  gradient''. Worse, numerical approximation to the Jacobian gives the
-  error ``singular gradient matrix at initial parameter estimates'' for
-  \texttt{minpack.lm::nlsLM} so that the Marquardt stabilization is
-  unable to take effect, while the analytic derivatives of
-  \texttt{nlsr::nlxb} give a solution.
-\item
-  Karl Schilling (private communication) provided an example where a
-  model specified with the formula
-  \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes
-  \texttt{nlsr::nlxb} to fail because the partial derivative w.r.t.
-  \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for
-  which \texttt{x\ =\ 0}, the derivative is undefined, but the model can
-  be computed. In such cases, we observed that \texttt{nls()} and
-  \texttt{minpack.lm::nlsLM} found a solution through the lucky accident
-  that the approximate derivative can be evaluated.
+  A pasture regrowth problem (Huet et al. (\protect\hyperlink{ref-Huet2004}{2004}), page 1, based on Ratkowsky (\protect\hyperlink{ref-Ratkowsky1983}{1983})) has a
+  poorly conditioned Jacobian and \texttt{nls()} fails with ``singular gradient''.
+  Worse, numerical approximation to the Jacobian gives the error
+  ``singular gradient matrix at initial parameter estimates'' for \texttt{minpack.lm::nlsLM}
+  so that the Marquardt stabilization is unable to take effect, while the analytic
+  derivatives of \texttt{nlsr::nlxb} give a solution.
+\item
+  Karl Schilling (private communication) provided an example where a model specified
+  with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
+  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
+  the derivative is undefined, but the model can be computed. In such cases,
+  we observed that \texttt{nls()} and \texttt{minpack.lm::nlsLM} found a solution through the
+  lucky accident that the approximate derivative can be evaluated.
 \end{itemize}
@@
 \hypertarget{jacobian-code-in-selfstart-models}{%
-\subsubsection{Jacobian code in selfStart
-models}\label{jacobian-code-in-selfstart-models}}
+\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}
 
 Analytic Jacobian code can be provided to all the solvers discussed.
-Most \texttt{selfStart} models that automatically provide starting
-parameters also include such code. There is documentation in R of
-\texttt{selfStart} models, but their construction is non-trivial. A
-number of such models are included with base R in
-\texttt{./src/library/stats/R/zzModels.R}, with package \texttt{nlraa}
-(Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})) providing a
-richer set. There are also some in the now-archived package
-\texttt{NRAIA}. The Jacobian is provided via the \texttt{gradient}
-attribute of the ``one-sided'' formula that defines each model, and
-these Jacobians are often the analytic forms.
-
-The \texttt{nls()} function, after computing the ``right-hand side'' or
-\texttt{rhs} of the residual, checks to see if the \texttt{gradient}
-attribute is defined, otherwise using \texttt{numericDeriv()} to compute
-a Jacobian into that attribute. This code is within the
-\texttt{nlsModel()} or \texttt{nlsModel.plinear()} functions. The use of
-analytic Jacobians almost certainly contributes to the good performance
-of \texttt{nls()} on \texttt{selfStart} models.
-
-The use of \texttt{selfStart} models with \texttt{nlsr} is described in
-the ``Introduction to nlsr'' vignette. However, since \texttt{nlsr}
-generally can use very crude starting values, we have rarely needed
-them, though it should be pointed out that our work is primarily
-diagnostic. If we were carrying out a large number of similar
-estimations, such initial parameters are critical to efficiency.
-
-In considering \texttt{selfStart} models, we noted that the base-R
-function \texttt{SSlogis} is intended to solve problem
-\textbf{Logistic3T} above. When this function is used via
-\texttt{getInitial()} to find starting values, it actually calls
-\texttt{nls()} with the `plinear' algorithm and finds a (full) solution.
-It then passes the solution coefficients to the default algorithm which
-converges in one iteration and builds the standard output. We found the
-implicit double-call a challenge to unravel. A more conventional
-approach to starting parameters, following the lead of Ratkowsky
-(\protect\hyperlink{ref-Ratkowsky1983}{1983}), is in the function
-\texttt{SSlogisJN}, now part of the package \texttt{nlsr}.
-
-Users may also want to profit from the Jacobian code of
-\texttt{selfStart} models but supply explicit starting values other than
-those suggested by \texttt{getInitial()}. This does not appear to be
-possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
-starting parameters. We could get these by separately calling
-\texttt{getInitial()} to find them from the \texttt{selfStart} model,
-but could use other values. The \texttt{selfStart} model is also used in
-the formula for the \texttt{nlxb()} call.
-
-We note that the analytic expressions for the Jacobian
-(\texttt{gradient}) in the \texttt{SSLogis} function and others save
-quantities by default in ``hidden'' variables, i.e., with names preceded
-by ``.''. This behaviour is to avoid name conflicts with user
-quantities. Hidden variables are not displayed by the \texttt{ls()}
-command, so users need to remember to use
-\texttt{ls(all.names\ =\ TRUE)} to see which quantities are defined, for
-example, in developing their own selfStart model via copy and edit.
-
-Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman
-(\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be worth
-considering as another way to find starting parameters, but we know of
-no R capability of this type.
-
-As a side note, the introduction of \texttt{scaleOffset} in R 4.1.1 to
-deal with the convergence test for small residual problems now requires
-that the \texttt{getInitial()} function have dot-arguments
-(\texttt{...}) in its argument list. This illustrates the entanglement
-of many features in \texttt{nls()} that complicate its maintenance and
+Most \texttt{selfStart} models that automatically provide starting parameters
+also include such code. There is documentation in R
+of \texttt{selfStart} models, but their construction is non-trivial. A number
+of such models are included with base R in \texttt{./src/library/stats/R/zzModels.R},
+with package \texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})) providing a richer set.
+There are also some in the now-archived package \texttt{NRAIA}.
+The Jacobian is provided via the \texttt{gradient} attribute of the ``one-sided'' formula
+that defines each model, and these Jacobians are often the analytic forms.
+
+The \texttt{nls()} function, after
+computing the ``right-hand side'' or \texttt{rhs} of the residual, checks to see if the
+\texttt{gradient} attribute is defined, otherwise using \texttt{numericDeriv()} to compute a
+Jacobian into that attribute. This code is within the \texttt{nlsModel()} or
+\texttt{nlsModel.plinear()} functions. The use of analytic Jacobians
+almost certainly contributes to the good performance of \texttt{nls()} on \texttt{selfStart}
+models.
+
+The use of \texttt{selfStart} models with
+\texttt{nlsr} is described in the ``Introduction to nlsr'' vignette. However, since \texttt{nlsr}
+generally can use very crude starting values, we have rarely needed them, though
+it should be pointed out that our work is primarily diagnostic. If we were carrying
+out a large number of similar estimations, such initial parameters are
+critical to efficiency.
+
+In considering \texttt{selfStart} models, we noted that the base-R function \texttt{SSlogis}
+is intended to solve problem \textbf{Logistic3T} above. When this function is used
+via \texttt{getInitial()} to find
+starting values, it actually calls \texttt{nls()} with the `plinear' algorithm
+and finds a (full) solution. It then passes the solution coefficients to the default
+algorithm which converges in one iteration and builds the standard output.
+We found the implicit double-call a challenge to unravel.
+A more conventional approach to starting parameters, following the lead of Ratkowsky (\protect\hyperlink{ref-Ratkowsky1983}{1983}),
+is in the function \texttt{SSlogisJN}, now part of the package \texttt{nlsr}.
+
+Users may also want to profit from the Jacobian code of \texttt{selfStart} models
+but supply explicit starting values other than those suggested by \texttt{getInitial()}.
+This does not appear to be possible with \texttt{nls()}. \texttt{nlsr::nlxb()} always requires
+starting parameters. We could get these by separately calling \texttt{getInitial()} to
+find them from the \texttt{selfStart} model, but could use other values. The \texttt{selfStart}
+model is also used in the formula for the \texttt{nlxb()} call.
+
+We note that the analytic expressions for the Jacobian (\texttt{gradient}) in the \texttt{SSLogis}
+function and others save quantities by default in ``hidden'' variables, i.e., with
+names preceded by ``.''. This behaviour is to avoid name
+conflicts with user quantities. Hidden variables are not displayed by the \texttt{ls()}
+command, so users need to remember to use \texttt{ls(all.names\ =\ TRUE)} to see which quantities
+are defined, for example, in developing their own selfStart model via copy and edit.
+
+Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (\protect\hyperlink{ref-nash1996nonlinear}{1996})) might be
+worth considering as another way to find starting parameters, but we know of no
+R capability of this type.
+
+As a side note, the introduction of \texttt{scaleOffset} in R 4.1.1 to deal with the
+convergence test for small residual problems now requires that the \texttt{getInitial()}
+function have dot-arguments (\texttt{...}) in its argument list. This illustrates the
+entanglement of many features in \texttt{nls()} that complicate its maintenance and
 improvement.
@@
 \item
-  \texttt{nls()} can only impose bounds if the
-  \texttt{algorithm\ =\ "port"} argument is used in the call.
-  Unfortunately, the documentation warns us:
-
-  \emph{The algorithm = ``port'' code appears unfinished, and does not
-  even check that the starting value is within the bounds. Use with
+  \texttt{nls()} can only impose bounds if the \texttt{algorithm\ =\ "port"} argument is
+  used in the call. Unfortunately, the documentation warns us:
+
+  \emph{The algorithm = ``port'' code appears unfinished, and does not even
+  check that the starting value is within the bounds. Use with
   caution, especially where bounds are supplied.}
@@
 \item
-  \texttt{nlsLM()} includes bounds in the standard call, but we have
-  observed cases where it fails to get the correct answer. From an
-  examination of the code, we believe the authors have not taken into
-  account all possibilities, though all programs have some weakness
-  regarding constrained optimization in that programmers have to work
-  with assumptions on the scale of numbers, and some problems will be
-  outside the scope envisaged.
+  \texttt{nlsLM()} includes bounds in the standard call, but we have observed cases
+  where it fails to get the correct answer. From an examination of the code,
+  we believe the authors have not taken into account all possibilities, though
+  all programs have some weakness regarding constrained optimization in that
+  programmers have to work with assumptions on the scale of numbers,
+  and some problems will be outside the scope envisaged.
 \end{itemize}
 
-Examples of the use of bounds constraints for nonlinear regression
-modeling are quite scarce. Traditionally, workers transform variables to
-ensure bounds are satisfied, such as squaring a parameter, or using a
-parameter which is the \(\log()\) of the actual parameter, so that its
-exponential is positive. These transformations can create scaling
-issues, but seem to generally work.
-
-Bounds can also be helpful to avoid unanticipated excursions of the
-minimization trajectory into inadmissible parameter regions. Such bounds
-should not be too tight, as this may hinder progress to an optimum by
-causing the methods to bump along the constraint in many cycles. Our aim
-is to avoid rubbish, not specify where the solution will be, and we hope
-the solution is not complicated by having a parameter on a bound.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Start MUST be feasible i.e. on or within bounds}
-\NormalTok{anlshob1b }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
-\FunctionTok{pnls}\NormalTok{(anlshob1b) }\CommentTok{\#  check the answer (short form)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlshob1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 10  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# nlsLM seems not to work with bounds in this example}
-\NormalTok{anlsLM1b }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLM1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLM1b  -- ss= 881.02 :  c1 = 2  c2 = 6  c3 = 3; 2  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# also no warning if starting out of bounds, but gets a good answer!!}
-\NormalTok{st4}\OtherTok{\textless{}{-}}\FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{4}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{4}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{4}\NormalTok{)}
-\NormalTok{anlsLMob }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ st4, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLMob)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLMob  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Try nlsr::nlxb()}
-\NormalTok{anlx1b }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts1, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }\AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{3}\NormalTok{))}
-\FunctionTok{pshort}\NormalTok{(anlx1b)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlx1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 12 res/ 12 jac
+Examples of the use of bounds constraints for nonlinear regression modeling are
+quite scarce. Traditionally, workers transform variables to ensure bounds are
+satisfied, such as squaring a parameter, or using a parameter which is the \(\log()\)
+of the actual parameter, so that its exponential is positive. These transformations
+can create scaling issues, but seem to generally work.
+
+Bounds can also be helpful to avoid unanticipated excursions of the minimization
+trajectory into inadmissible parameter regions. Such bounds should not be too
+tight, as this may hinder progress to an optimum by causing the methods to
+bump along the constraint in many cycles. Our aim is to avoid rubbish, not
+specify where the solution will be, and we hope the solution is not complicated
+by having a parameter on a bound.
+
+\begin{verbatim}
+# Start MUST be feasible i.e. on or within bounds
+anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
+             upper = c(2,6,3), algorithm = 'port')
+pnls(anlshob1b) #  check the answer (short form)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlshob1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 10  itns
+\end{verbatim}
+
+\begin{verbatim}
+# nlsLM seems not to work with bounds in this example
+anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
+pnls(anlsLM1b)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLM1b  -- ss= 881.02 :  c1 = 2  c2 = 6  c3 = 3; 2  itns
+\end{verbatim}
+
+\begin{verbatim}
+# also no warning if starting out of bounds, but gets a good answer!!
+st4<-c(c1 = 4, c2 = 4, c3 = 4)
+anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
+pnls(anlsLMob)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLMob  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Try nlsr::nlxb()
+anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
+pshort(anlx1b)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlx1b  -- ss= 9.4726 :  c1 = 2  c2 = 4.4332  c3 = 3; 12 res/ 12 jac
 \end{verbatim}
@@
 \hypertarget{philosophical-considerations}{%
-\subsubsection{Philosophical
-considerations}\label{philosophical-considerations}}
+\subsubsection{Philosophical considerations}\label{philosophical-considerations}}
 
 Bounds on parameters raise some interesting questions about how
-uncertainty in parameter estimates should be computed or reported. That
-is, the traditional ``standard errors'' are generally taken to imply
-symmetric intervals about the point estimate in which the parameter may
-be expected to be found with some probability under certain assumptions.
-Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not
-compute standard errors nor their derived statistics when bounds are
-active. Transformations to implicitly enforce bounds avoid this issue,
-but at the cost of changing the form of the model.
+uncertainty in parameter estimates should be computed or reported. That is, the
+traditional ``standard errors'' are generally taken to imply symmetric intervals about
+the point estimate in which the parameter may be expected to be found with some
+probability under certain assumptions.
+Bounds change those assumptions. Hence, \texttt{nlsr::nlxb()} does not compute standard
+errors nor their derived statistics when bounds are active. Transformations to
+implicitly enforce bounds avoid this issue, but at the cost of changing the
+form of the model.
 
@@
 
-Let us try to fix (mask) the first parameter in the first two example
-problems. We might want to do this if it is considered that the
-asymptote is known from external sources, but we wish later to have it
-available to be estimated.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbsmaskx.R {-}{-} masks with formula specification of the problem}
-\FunctionTok{require}\NormalTok{(nlsr); }\FunctionTok{require}\NormalTok{(minpack.lm); traceval }\OtherTok{\textless{}{-}} \ConstantTok{FALSE}
-\NormalTok{stu }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{b1 =} \DecValTok{200}\NormalTok{, }\AttributeTok{b2 =} \DecValTok{50}\NormalTok{, }\AttributeTok{b3 =} \FloatTok{0.3}\NormalTok{) }\CommentTok{\# a default starting vector (named!)}
-\NormalTok{sts }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{c1 =} \DecValTok{2}\NormalTok{, }\AttributeTok{c2 =} \DecValTok{5}\NormalTok{, }\AttributeTok{c3 =} \DecValTok{3}\NormalTok{) }\CommentTok{\# a default scaled starting vector (named!)}
-\CommentTok{\# fix first parameter}
-\NormalTok{anxbmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
-\FunctionTok{print}\NormalTok{(anxbmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.6182  on  12 observations
-##     after  4    Jacobian and  4 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## b1                   200U M         NA         NA         NA           0          NA  
-## b2               49.5108          1.12      44.21  8.421e-13  -2.887e-07        1022  
-## b3              0.311461      0.002278      136.8  1.073e-17   0.0001635      0.4569
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlM1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlsLM}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-            \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{), }\AttributeTok{trace=}\NormalTok{traceval))}
-\FunctionTok{pnls}\NormalTok{(anlM1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlM1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(frmu, }\AttributeTok{start =}\NormalTok{ stu, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{), }
-        \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{200}\NormalTok{, }\DecValTok{60}\NormalTok{, }\DecValTok{3}\NormalTok{),  }\AttributeTok{algorithm=}\StringTok{"port"}\NormalTok{, }\AttributeTok{trace =}\NormalTok{ traceval))}
-\FunctionTok{pnls}\NormalTok{(anlsmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsmsk1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 5  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Hobbs scaled problem with bounds, formula specification}
-\NormalTok{anlxmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                  \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
-\FunctionTok{print}\NormalTok{(anlxmsks1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## residual sumsquares =  2.6182  on  12 observations
-##     after  4    Jacobian and  4 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## c1                     2U M         NA         NA         NA           0          NA  
-## c2               4.95108         0.112      44.21  8.421e-13  -2.981e-06       104.2  
-## c3               3.11461       0.02278      136.8  1.073e-17   1.583e-05       4.482
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlshmsk1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{trace =}\NormalTok{ traceval, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-             \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{), }\AttributeTok{algorithm =} \StringTok{\textquotesingle{}port\textquotesingle{}}\NormalTok{)}
-\FunctionTok{pnls}\NormalTok{(anlshmsk1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlshmsk1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 5  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{anlsLMmsks1 }\OtherTok{\textless{}{-}} \FunctionTok{nlsLM}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{0}\NormalTok{,}\DecValTok{0}\NormalTok{),}
-                 \AttributeTok{upper =} \FunctionTok{c}\NormalTok{(}\DecValTok{2}\NormalTok{,}\DecValTok{6}\NormalTok{,}\DecValTok{30}\NormalTok{))}
-\FunctionTok{pnls}\NormalTok{(anlsLMmsks1)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## anlsLMmsks1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 4  itns
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# Test with all parameters masked}
-\NormalTok{anlxmskall}\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nlxb}\NormalTok{(frms, }\AttributeTok{start =}\NormalTok{ sts, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{lower =}\NormalTok{ sts, }\AttributeTok{upper =}\NormalTok{ sts))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in nlfb(start = pnum, resfn = trjfn, jacfn = trjfn, trace = trace, :
-## All parameters are masked
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{print}\NormalTok{(anlxmskall)}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## Warning in summary.nlsr(x): No unmasked parameters
-\end{verbatim}
-
-\begin{verbatim}
-## residual sumsquares =  158.23  on  12 observations
-##     after  0    Jacobian and  1 function evaluations
-##   name            coeff          SE       tstat      pval      gradient    JSingval   
-## c1                     2U M         NA         NA         NA          NA          NA  
-## c2                     5U M         NA         NA         NA          NA          NA  
-## c3                     3U M         NA         NA         NA          NA          NA
-\end{verbatim}
-
-\texttt{nlsr} has an output format that indicates the constraint status
-of the parameter estimates. For \texttt{nlsr}, we have \textbf{chosen}
-to suppress the calculation of approximate standard errors in the
-parameters when constraints are active because their meaning under
-constraints is unclear, though we believe this policy worthy of
-discussion and further investigation.
+Let us try to fix (mask) the first parameter in the first two example problems. We might
+want to do this if it is considered that the asymptote is known from external sources,
+but we wish later to have it available to be estimated.
+
+\begin{verbatim}
+# Hobbsmaskx.R -- masks with formula specification of the problem
+require(nlsr); require(minpack.lm); traceval <- FALSE
+stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
+sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
+# fix first parameter
+anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+            upper = c(200, 60, 3), trace=traceval))
+print(anxbmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  2.6182  on  12 observations
+#>     after  4    Jacobian and  4 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> b1                   200U M         NA         NA         NA           0          NA  
+#> b2               49.5108          1.12      44.21  8.421e-13  -2.887e-07        1022  
+#> b3              0.311461      0.002278      136.8  1.073e-17   0.0001635      0.4569
+\end{verbatim}
+
+\begin{verbatim}
+anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+            upper = c(200, 60, 3), trace=traceval))
+pnls(anlM1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlM1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200,0,0), 
+        upper = c(200, 60, 3),  algorithm="port", trace = traceval))
+pnls(anlsmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsmsk1  -- ss= 2.6182 :  b1 = 200  b2 = 49.511  b3 = 0.31146; 5  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Hobbs scaled problem with bounds, formula specification
+anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                  upper = c(2,6,30))
+print(anlxmsks1)
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  2.6182  on  12 observations
+#>     after  4    Jacobian and  4 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> c1                     2U M         NA         NA         NA           0          NA  
+#> c2               4.95108         0.112      44.21  8.421e-13  -2.981e-06       104.2  
+#> c3               3.11461       0.02278      136.8  1.073e-17   1.583e-05       4.482
+\end{verbatim}
+
+\begin{verbatim}
+anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, lower = c(2,0,0),
+             upper = c(2,6,30), algorithm = 'port')
+pnls(anlshmsk1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlshmsk1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 5  itns
+\end{verbatim}
+
+\begin{verbatim}
+anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                 upper = c(2,6,30))
+pnls(anlsLMmsks1)
+\end{verbatim}
+
+\begin{verbatim}
+#> anlsLMmsks1  -- ss= 2.6182 :  c1 = 2  c2 = 4.9511  c3 = 3.1146; 4  itns
+\end{verbatim}
+
+\begin{verbatim}
+# Test with all parameters masked
+anlxmskall<- try(nlxb(frms, start = sts, data = weeddf, lower = sts, upper = sts))
+\end{verbatim}
+
+\begin{verbatim}
+#> Warning in nlfb(start = pnum, resfn = trjfn, jacfn = trjfn, trace = trace, :
+#> All parameters are masked
+\end{verbatim}
+
+\begin{verbatim}
+print(anlxmskall)
+\end{verbatim}
+
+\begin{verbatim}
+#> Warning in summary.nlsr(x): No unmasked parameters
+\end{verbatim}
+
+\begin{verbatim}
+#> residual sumsquares =  158.23  on  12 observations
+#>     after  0    Jacobian and  1 function evaluations
+#>   name            coeff          SE       tstat      pval      gradient    JSingval   
+#> c1                     2U M         NA         NA         NA          NA          NA  
+#> c2                     5U M         NA         NA         NA          NA          NA  
+#> c3                     3U M         NA         NA         NA          NA          NA
+\end{verbatim}
+
+\texttt{nlsr} has an output format that indicates the constraint status of the parameter
+estimates. For \texttt{nlsr}, we have \textbf{chosen} to suppress the calculation of
+approximate standard errors in the parameters when constraints are active because
+their meaning under constraints is unclear, though we believe this
+policy worthy of discussion and further investigation.
 
 \hypertarget{stabilization-of-gauss-newton-computations}{%
-\subsection{Stabilization of Gauss-Newton
-computations}\label{stabilization-of-gauss-newton-computations}}
+\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}}
 
 All four major tools illustrated solve some variant of the Gauss-Newton
-equations. \texttt{nls()} uses a modification of an approach suggested
-by Hartley (\protect\hyperlink{ref-Hartley1961}{1961}), while
-\texttt{nlsr}, \texttt{gslnls}, and \texttt{minpack.lm} use flavours of
-Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}). \texttt{gslnls}
-offers an accelerated Marquardt method and three alternative methods
-which we have not investigated beyond one example. Control settings for
-\texttt{nlxb()} or \texttt{nlfb()} allow exploration of Hartley and
-Marquardt algorithm variants, principally for the benefit of program
-developers. In general, the Levenberg-Marquardt stabilization is
-important in obtaining solutions in methods of the Gauss-Newton family,
-as \texttt{nls()} terminates too frequently and unnecessarily with
-\texttt{singular\ gradient} errors. A tidy incorporation of Marquardt
-stabilization into \texttt{nls()} would be a major improvement.
+equations. \texttt{nls()} uses a modification of an approach suggested by Hartley (\protect\hyperlink{ref-Hartley1961}{1961}),
+while \texttt{nlsr}, \texttt{gslnls}, and \texttt{minpack.lm} use flavours of Marquardt (\protect\hyperlink{ref-Marquardt1963}{1963}). \texttt{gslnls}
+offers an accelerated Marquardt method and three alternative methods which we
+have not investigated beyond one example.
+Control settings for \texttt{nlxb()} or \texttt{nlfb()} allow exploration of Hartley and Marquardt
+algorithm variants, principally for the benefit of program developers.
+In general, the Levenberg-Marquardt stabilization
+is important in obtaining
+solutions in methods of the Gauss-Newton family, as \texttt{nls()} terminates
+too frequently and unnecessarily with \texttt{singular\ gradient} errors. A tidy
+incorporation of Marquardt stabilization into \texttt{nls()} would be a major improvement.
 
@@
 
-An important choice made in developing \texttt{nlsr} was to code
-entirely within the R programming language. \texttt{nls()} uses a mix of
-R, C, and Fortran, as does \texttt{minpack.lm}. \texttt{gslnls} is an R
-wrapper to various C-language routines in the GNU Scientific Library
-(Galassi et al. (\protect\hyperlink{ref-GSL-manual}{2009})). Generally,
-keeping to a single programming language can allow for easier
-maintenance and upgrades. It also avoids some work when there are
-changes or upgrades to libraries for the non-R languages. R is usually
-considered slower than most computing environments because it keeps
-track of objects and because it is usually interpreted. In recent years,
-the performance penalty for using code entirely in R has been much
-reduced with the just-in-time compiler and other improvements. All-R
-computation may now offer acceptable performance. In \texttt{nlsr}, the
-use of R may be a smaller performance cost than the aggressive approach
-to a solution, such as large computational limits for residuals and
-Jacobians.
-
-There are, of course, problems where use of R code will be slower. For
-example, a ``large'' case of the Penalty 1 test problem of Moré, Garbow,
-and Hillstrom (\protect\hyperlink{ref-More1981TUO}{1981}) will take
-noticeable time to solve. Problems in large numbers of parameters create
-large computational matrices for the Jacobian or its inner product,
-which Gauss-Newton methods create and process. This is confirmed by
-comparing \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} on the
-Penalty 1 problem of various sizes, where the latter, which uses
+An important choice made in developing \texttt{nlsr} was to code entirely within the R
+programming language. \texttt{nls()} uses a mix of R, C, and Fortran, as does \texttt{minpack.lm}.
+\texttt{gslnls} is an R wrapper to various C-language routines in the GNU Scientific
+Library (Galassi et al. (\protect\hyperlink{ref-GSL-manual}{2009})).
+Generally, keeping to a single programming language can allow for
+easier maintenance and upgrades. It also avoids some work when there are changes
+or upgrades to libraries for the non-R languages.
+R is usually considered slower
+than most computing environments because it keeps track of objects and
+because it is usually interpreted. In recent years, the
+performance penalty for using code entirely in R has been much reduced
+with the just-in-time compiler and other improvements. All-R computation
+may now offer acceptable performance. In \texttt{nlsr}, the use of R may be a
+smaller performance cost than the aggressive approach to a solution,
+such as large computational limits for residuals and Jacobians.
+
+There are, of course, problems where use of R code will be slower. For example, a
+``large'' case of the Penalty 1 test problem of Moré, Garbow, and Hillstrom (\protect\hyperlink{ref-More1981TUO}{1981}) will take
+noticeable time to solve. Problems in large numbers of parameters
+create large computational matrices for the Jacobian or its inner product,
+which Gauss-Newton methods create and process.
+This is confirmed by comparing \texttt{nlsr::nlfb()} and \texttt{minpack.lm::nls.lm()} on
+the Penalty 1 problem of various sizes, where the latter, which uses
 compiled Fortran internally, was moderately faster. However, the method
-``Rcgmin'' from package \texttt{optimx} (John C. Nash and Varadhan
-(\protect\hyperlink{ref-p-optimx}{2011})) proved to be much faster than
-either, despite being entirely written in R, and is a much more reliable
-conjugate gradient minimizer than method ``CG'' of \texttt{optim()}. One
-of us created both, but now deprecates the latter. Timings of CG methods
-are more sensitive to small differences in setup and arithmetic than
-Gauss-Newton methods, and the pattern of such timings varies much less
-regularly with problem size for CG methods.
+``Rcgmin'' from package \texttt{optimx} (John C. Nash and Varadhan (\protect\hyperlink{ref-p-optimx}{2011})) proved to be much
+faster than either, despite being entirely written in R, and is a much more
+reliable conjugate gradient minimizer than method ``CG'' of \texttt{optim()}. One
+of us created both, but now deprecates the latter.
+Timings of CG methods are more sensitive to small differences in
+setup and arithmetic than Gauss-Newton methods, and the pattern of
+such timings varies much less regularly with problem size for CG methods.
 
@@
 
-\texttt{nls()} can be called without specifying the \texttt{data}
-argument. In this case, it will search in the available environments
-(i.e., workspaces) for suitable data objects. We do not like this
-approach, but it is ``the R way''. R allows users to leave many objects
-in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to
-actively suppress saving this workspace (\texttt{.RData}) on exit;
-otherwise, any such file in the path will be loaded on startup. R users
-in our acquaintance avoid saving the workspace because of lurking data
-and functions that may cause unwanted results.
+\texttt{nls()} can be called without specifying the \texttt{data} argument. In this case, it will
+search in the available environments (i.e., workspaces) for suitable data objects.
+We do not like this approach, but it is ``the R way''. R allows users to leave many
+objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
+suppress saving this workspace (\texttt{.RData}) on exit; otherwise, any such file in
+the path will be loaded on startup. R users in our
+acquaintance avoid saving the workspace because of lurking data and
+functions that may cause unwanted results.
 
@@
 
-\texttt{nls()} and other class \texttt{nls} tools accept an argument
-\texttt{subset}. This acts through the mediation of
-\texttt{model.frame}, which is not obvious in the source code files
-\texttt{/src/library/stats/R/nls.R} and
-\texttt{/src/library/stats/src/nls.C}. Having \texttt{subset} at the
-level of the call to a function like \texttt{nls()} saves effort, but it
-does mean that the programmer of the solver needs to be aware of the
+\texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
+the mediation of \texttt{model.frame}, which is not obvious in the
+source code files \texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
+Having \texttt{subset} at the level of the call to a function like
+\texttt{nls()} saves effort, but it does mean that the programmer of the
+solver needs to be aware of the
 origin (and value) of objects such as the data, residuals and Jacobian.
-By preference, we would implement subsetting by zero-value weights, with
-observation counts (and degrees of freedom) computed via the numbers of
-non-zero weights. Alternatively, we would extract a working dataframe
-from the relevant elements in the original.
+By preference,
+we would implement subsetting by zero-value weights, with observation counts
+(and degrees of freedom) computed via the numbers of non-zero weights. Alternatively,
+we would extract a working dataframe from the relevant elements in the original.
 
 \hypertarget{feature-na.action-missing-value-treatment}{%
-\subsubsection{Feature: na.action (missing value
-treatment)}\label{feature-na.action-missing-value-treatment}}
-
-\texttt{na.action} is an argument to the \texttt{nls()} function, but it
-does not appear obviously in the source code, often being handled behind
-the scenes after referencing the option \texttt{na.action}. This feature
-also changes the data supplied to our nonlinear least squares solver. A
-useful, but possibly dated, description is given in:
+\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}}
+
+\texttt{na.action} is an argument to the \texttt{nls()} function, but it does not appear
+obviously in the source code, often being handled behind the scenes after
+referencing the option \texttt{na.action}. This feature also changes the data
+supplied to our nonlinear least squares solver.
+A useful, but possibly dated, description is given in:
 \emph{\url{https://stats.idre.ucla.edu/r/faq/how-does-r-handle-missing-values/}}.
 The typical default action, which can be seen by using the command
-\texttt{getOption("na.action")} is \texttt{na.omit}. This option removes
-from computations any observations containing missing values (i.e.~any
-row of a data frame containing an NA). \texttt{na.exclude} does much of
-the same for solver computations, but keeps the rows with NA elements so
-that predictions are in the correct row position. We recommend that
-workers actually test output to verify the behaviour is as wanted. See
-\emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}.
-As with \texttt{subset}, our concern with \texttt{na.action} is that
-users may be unaware of the effects of an option they may not even know
-has been set. Should \texttt{na.fail} be the default?
+\texttt{getOption("na.action")}
+is \texttt{na.omit}. This option removes from computations any observations
+containing missing values (i.e.~any row of a data frame containing an NA).
+\texttt{na.exclude} does much of the same for solver computations, but keeps the rows with
+NA elements so that predictions are in the correct row position. We recommend that
+workers actually test output to verify the behaviour is as wanted.
+See \emph{\url{https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r}}.
+As with \texttt{subset}, our concern with \texttt{na.action} is that users may be unaware of the
+effects of an option they may not even know has been set. Should \texttt{na.fail} be the default?
 
@@
 
-\texttt{model} is an argument to the \texttt{nls()} and related
-functions, which is documented as:
-
-\emph{\textbf{model} logical. If true, the model frame is returned as
-part of the object. Default is FALSE.}
-
-Indeed, the argument only gets used when \texttt{nls()} is about to
-return its result object, and the element \texttt{model} is NULL unless
-the calling argument \texttt{model} is TRUE. (Using the same name could
-be confusing.) Despite this, the model frame is used within the function
-code in the form of the object \texttt{mf}. We feel that users could
-benefit from more extensive documentation and examples of its use since
-it is used to implement features like \texttt{subset}. Moreover, it
-provides access to resources that may be important in further
-calculations on models.
+\texttt{model} is an argument to the \texttt{nls()} and related functions, which is documented as:
+
+\emph{\textbf{model} logical. If true, the model frame is returned as part of the object. Default is FALSE.}
+
+Indeed, the argument only gets used when \texttt{nls()} is about to return its result
+object, and the
+element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. (Using the same
+name could
+be confusing.) Despite this, the model frame is used within the function code in
+the form of the object
+\texttt{mf}. We feel that users could benefit from more extensive documentation and
+examples of its use since it is used to implement features like \texttt{subset}. Moreover,
+it provides access to resources that may be important in further calculations on
+models.
 
@@
 
-All four main tools we consider here allow a \texttt{weights} argument
-that specifies a vector of fixed weights the same length as the number
-of residuals. Each residual is multiplied by the square root of the
-corresponding weight. Where available, the values returned by the
-\texttt{residuals()} function are weighted, and the \texttt{fitted()} or
-\texttt{predict()} functions are used to compute raw residuals.
-
-While fixed weights may be useful, there are many problems for which we
-want weights that are determined at least partially from the model
-parameters; for example, a measure of the standard deviation of
-observations. Such dynamic weighting situations are discussed in the
-vignette ``Introduction to nlsr'' of package \texttt{nlsr} in section
+All four main tools we consider here allow a \texttt{weights} argument that
+specifies a vector of fixed weights the same length as the number of residuals.
+Each residual is multiplied by the square root of the corresponding weight.
+Where available, the values returned by the
+\texttt{residuals()} function are weighted, and the \texttt{fitted()} or \texttt{predict()} functions are
+used to compute raw residuals.
+
+While fixed weights may be useful, there are many problems for which
+we want weights that are determined at least partially from the model
+parameters; for
+example, a measure of the standard deviation of observations.
+Such dynamic weighting situations are discussed in the vignette
+``Introduction to nlsr'' of package \texttt{nlsr} in section
 \emph{Weights that are functions of the model parameters}.
-\texttt{minpack.lm} offers a function \texttt{wfct()} to facilitate such
-weighting. Care is advised in applying such ideas.
+\texttt{minpack.lm} offers a function \texttt{wfct()} to
+facilitate such weighting. Care is advised
+in applying such ideas.
 
 \hypertarget{weights-in-returned-functions-from-nls}{%
-\subsubsection{Weights in returned functions from
-nls()}\label{weights-in-returned-functions-from-nls}}
-
-The function \texttt{resid()} (an alias for \texttt{residuals()}) gives
-weighted residuals, as does, for example, \texttt{result\$m\$resid()}.
-The function \texttt{nlsModel()}, which we have had to extract from the
-base R code and explicitly \texttt{source()} because it is not exported
-to the working namespace, allows us to compute residuals for particular
-coefficient sets.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{wts }\OtherTok{\textless{}{-}} \FloatTok{0.5} \SpecialCharTok{\^{}}\NormalTok{ tt }\CommentTok{\# simple weights}
-\NormalTok{frmlogis }\OtherTok{\textless{}{-}}\NormalTok{ weed }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym}\SpecialCharTok{/}\NormalTok{(}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}}\NormalTok{ tt) }\SpecialCharTok{/}\NormalTok{ scal))}
-\NormalTok{Asym }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{; xmid }\OtherTok{\textless{}{-}} \DecValTok{1}\NormalTok{; scal }\OtherTok{\textless{}{-}} \DecValTok{1}
-\NormalTok{nowt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal)) }\CommentTok{\# unweighted}
-\NormalTok{rnowt }\OtherTok{\textless{}{-}}\NormalTok{ nowt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# This has unweighted residual and Jacobian.
-Does not take coefficients.}
-\FunctionTok{attr}\NormalTok{(rnowt, }\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rnowt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
-##  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{usewt }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(weed }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(tt, Asym, xmid, scal), }\AttributeTok{weights=}\NormalTok{wts)}
-\NormalTok{rusewt }\OtherTok{\textless{}{-}}\NormalTok{ usewt}\SpecialCharTok{$}\NormalTok{m}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# weighted. Does not take coefficients.}
-\FunctionTok{attr}\NormalTok{(rusewt,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rusewt}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\FunctionTok{source}\NormalTok{(}\StringTok{"nlsModel.R"}\NormalTok{)}
-\NormalTok{nmod0 }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{start =} \FunctionTok{c}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{1}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{1}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{), }\AttributeTok{wts =}\NormalTok{ wts)}
-\NormalTok{rn0}\OtherTok{\textless{}{-}}\NormalTok{nmod0}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{() }\CommentTok{\# Parameters are supplied in nlsModel() \textasciigrave{}start\textasciigrave{} above.}
-\FunctionTok{attr}\NormalTok{(rn0,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rn0}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
-## [11] 1.6572 1.4214
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{nmod }\OtherTok{\textless{}{-}} \FunctionTok{nlsModel}\NormalTok{(frmlogis, }\AttributeTok{data =}\NormalTok{ weeddf, }\AttributeTok{start =} \FunctionTok{coef}\NormalTok{(usewt), }\AttributeTok{wts =}\NormalTok{ wts)}
-\NormalTok{rn }\OtherTok{\textless{}{-}}\NormalTok{ nmod}\SpecialCharTok{$}\FunctionTok{resid}\NormalTok{()}
-\FunctionTok{attr}\NormalTok{(rn,}\StringTok{"gradient"}\NormalTok{) }\OtherTok{\textless{}{-}} \ConstantTok{NULL}\NormalTok{; rn}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-##  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
-##  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}
+
+The function \texttt{resid()} (an alias for \texttt{residuals()}) gives weighted residuals,
+as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}, which we
+have had to extract from the base R code and explicitly \texttt{source()} because it
+is not exported to the working namespace,
+allows us to compute residuals for particular coefficient sets.
+
+\begin{verbatim}
+wts <- 0.5 ^ tt # simple weights
+frmlogis <- weed ~ Asym/(1 + exp((xmid - tt) / scal))
+Asym <- 1; xmid <- 1; scal <- 1
+nowt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # unweighted
+rnowt <- nowt$m$resid() # This has unweighted residual and Jacobian. Does not take coefficients.
+attr(rnowt, "gradient") <- NULL; rnowt
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] -0.011900  0.032755 -0.092030 -0.208782 -0.392634  0.057594  1.105728
+#>  [8] -0.715786  0.107647  0.348396 -0.652592  0.287569
+\end{verbatim}
+
+\begin{verbatim}
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
+rusewt <- usewt$m$resid() # weighted. Does not take coefficients.
+attr(rusewt,"gradient") <- NULL; rusewt
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
+\end{verbatim}
+
+\begin{verbatim}
+source("nlsModel.R")
+nmod0 <- nlsModel(frmlogis, data = weeddf, start = c(Asym = 1, xmid = 1, scal = 1), wts = wts)
+rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+attr(rn0,"gradient") <- NULL; rn0
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1] 3.3998 3.2545 3.0961 2.9784 2.8438 2.7748 2.6910 2.3474 2.1724 1.9359
+#> [11] 1.6572 1.4214
+\end{verbatim}
+
+\begin{verbatim}
+nmod <- nlsModel(frmlogis, data = weeddf, start = coef(usewt), wts = wts)
+rn <- nmod$resid()
+attr(rn,"gradient") <- NULL; rn
+\end{verbatim}
+
+\begin{verbatim}
+#>  [1]  0.0085640  0.0324442 -0.0176652 -0.0388479 -0.0579575  0.0163623
+#>  [7]  0.1042380 -0.0411766  0.0052509  0.0084324 -0.0194246 -0.0024053
 \end{verbatim}
@@
 \hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
-\subsection{Minor issues with nonlinear least-squares
-tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
+\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
 
 \hypertarget{interim-output-from-the-port-algorithm}{%
-\subsubsection{Interim output from the ``port''
-algorithm}\label{interim-output-from-the-port-algorithm}}
-
-As the \texttt{nls()} \textbf{man} page states, when the ``port''
-algorithm is used with the \texttt{trace} argument TRUE, the iterations
-display the objective function value which is 1/2 the sum of squares (or
-deviance). The trace display is likely embedded in the Fortran of the
-\texttt{nlminb} routine that is called to execute the ``port''
-algorithm, but the factor of 2 discrepancy is nonetheless unfortunate
-for users.
+\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}
+
+As the \texttt{nls()} \textbf{man} page states, when the ``port'' algorithm is used with the
+\texttt{trace} argument TRUE, the iterations display the objective function value which
+is 1/2 the sum of squares (or deviance). The trace display is likely
+embedded in the Fortran of the \texttt{nlminb} routine that is called to execute
+the ``port'' algorithm, but the factor of 2 discrepancy is nonetheless
+unfortunate for users.
 
 \hypertarget{failure-to-return-the-best-result-achieved}{%
-\subsubsection{Failure to return the best result
-achieved}\label{failure-to-return-the-best-result-achieved}}
-
-If \texttt{nls()} reaches a point where it cannot continue but has not
-found a point where the relative offset convergence criterion is met, it
-may simply exit, especially if a ``singular gradient'' is found.
-However, this may occur \textbf{after} considerable progress has been
-made in reducing the sum of squared residuals. Here is an abbreviated
-example:
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{time }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\DecValTok{1}\NormalTok{,  }\DecValTok{2}\NormalTok{,  }\DecValTok{3}\NormalTok{,  }\DecValTok{4}\NormalTok{,  }\DecValTok{6}\NormalTok{ , }\DecValTok{8}\NormalTok{, }\DecValTok{10}\NormalTok{, }\DecValTok{12}\NormalTok{, }\DecValTok{16}\NormalTok{)}
-\NormalTok{conc }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\FloatTok{0.7}\NormalTok{, }\FloatTok{1.2}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.4}\NormalTok{, }\FloatTok{1.1}\NormalTok{, }\FloatTok{0.8}\NormalTok{, }\FloatTok{0.6}\NormalTok{, }\FloatTok{0.5}\NormalTok{, }\FloatTok{0.3}\NormalTok{)}
-\NormalTok{NLSdata }\OtherTok{\textless{}{-}} \FunctionTok{data.frame}\NormalTok{(time,conc)}
-\NormalTok{NLSstart }\OtherTok{\textless{}{-}} \FunctionTok{c}\NormalTok{(}\AttributeTok{lrc1 =} \SpecialCharTok{{-}}\DecValTok{2}\NormalTok{, }\AttributeTok{lrc2 =} \FloatTok{0.25}\NormalTok{, }\AttributeTok{A1 =} \DecValTok{150}\NormalTok{, }\AttributeTok{A2 =} \DecValTok{50}\NormalTok{) }\CommentTok{\# a starting vector (named!)}
-\NormalTok{NLSformula }\OtherTok{\textless{}{-}}\NormalTok{ conc }\SpecialCharTok{\textasciitilde{}}\NormalTok{ A1 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc1) }\SpecialCharTok{*}\NormalTok{ time) }\SpecialCharTok{+}\NormalTok{ A2 }\SpecialCharTok{*} \FunctionTok{exp}\NormalTok{(}\SpecialCharTok{{-}}\FunctionTok{exp}\NormalTok{(lrc2) }\SpecialCharTok{*}\NormalTok{ time)}
-\NormalTok{tryit }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(NLSformula, }\AttributeTok{data =}\NormalTok{ NLSdata, }\AttributeTok{start =}\NormalTok{ NLSstart, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\end{Highlighting}
-\end{Shaded}
-
-\begin{verbatim}
-## 61216.    (3.56e+03): par = (-2 0.25 150 50)
-## 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
-## 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
-## Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
-##   singular gradient
-\end{verbatim}
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\CommentTok{\# print(tryit) \# gives no information on best sumsquares or model parameters}
-\end{Highlighting}
-\end{Shaded}
+\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}}
+
+If \texttt{nls()} reaches a point where it cannot continue but has not found a point
+where the relative offset convergence criterion is met, it may simply exit,
+especially if a ``singular gradient'' is found. However,
+this may occur \textbf{after} considerable progress has been made in reducing
+the sum of squared residuals.
+Here is an abbreviated example:
+
+\begin{verbatim}
+time <- c(1,  2,  3,  4,  6 , 8, 10, 12, 16)
+conc <- c(0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
+NLSdata <- data.frame(time,conc)
+NLSstart <- c(lrc1 = -2, lrc2 = 0.25, A1 = 150, A2 = 50) # a starting vector (named!)
+NLSformula <- conc ~ A1 * exp(-exp(lrc1) * time) + A2 * exp(-exp(lrc2) * time)
+tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
+\end{verbatim}
+
+\begin{verbatim}
+#> 61216.    (3.56e+03): par = (-2 0.25 150 50)
+#> 2.1757    (2.23e+01): par = (-1.9991 0.31711 2.6182 -1.3668)
+#> 1.6211    (7.14e+00): par = (-1.9605 -2.6203 2.5753 -0.55599)
+#> Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : 
+#>   singular gradient
+\end{verbatim}
+
+\begin{verbatim}
+# print(tryit) # gives no information on best sumsquares or model parameters
+\end{verbatim}
 
 Note that the sum of squares has been reduced from 61216 to 1.6211, but
-unless \texttt{trace} is invoked, the user will not get any information
-about this. This almost trivial change to the \texttt{nls()} function
-could be useful to R users.
+unless \texttt{trace} is invoked, the user will not get any information about this.
+This almost trivial change to the \texttt{nls()} function could be
+useful to R users.
 
 \hypertarget{estimating-models-that-are-partially-linear}{%
-\subsection{Estimating models that are partially
-linear}\label{estimating-models-that-are-partially-linear}}
-
-The variable projection method (Golub and Pereyra
-(\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust
-(\protect\hyperlink{ref-OlearyRust13}{2013})) is usually much more
-effective than general approaches in finding good solutions to nonlinear
-least squares problems when some of the parameters appear linearly. In
-our logistic examples, the asymptote parameters are an illustration.
-However, identifying which parameters are linear and communicating this
-information to estimating functions is not a trivial task.
-\texttt{nls()} has an option \texttt{algorithm\ =\ "plinear"} that
-allows some partially linear models to be solved. The other tools, as
-far as we are aware, do not offer any such capability. The
-\texttt{nlstac} package uses a different algorithm for similar goals.
-
-Within \texttt{nls()} itself we must, unfortunately, use different
-specifications with different \texttt{algorithm} options. For example,
-the explicit linear model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b}
-does not work. As with the linear modeling function \texttt{lm()}, this
-must be specified as \texttt{y\ \textasciitilde{}\ x}. Within
-\texttt{nls()}, consider the following four different specifications for
-the same problem, plus an intuitive choice, labeled \texttt{fm2a}, that
-does not work. In this failed attempt, putting the \texttt{Asym}
-parameter in the model causes the \texttt{plinear} algorithm to try to
-add another term to the model. We believe this is unfortunate, and would
-like to see a consistent syntax. We do not foresee a resolution for this
-issue, which arises from different approaches to model specification. In
-the example, we have not evaluated the commands to save space.
-
-\begin{Shaded}
-\begin{Highlighting}[]
-\NormalTok{DNase1 }\OtherTok{\textless{}{-}} \FunctionTok{subset}\NormalTok{(DNase, Run }\SpecialCharTok{==} \DecValTok{1}\NormalTok{) }\CommentTok{\# select the data}
-\DocumentationTok{\#\# using a selfStart model {-} do not specify the starting parameters}
-\NormalTok{fm1 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \FunctionTok{SSlogis}\NormalTok{(}\FunctionTok{log}\NormalTok{(conc), Asym, xmid, scal), DNase1)}
-\FunctionTok{summary}\NormalTok{(fm1)}
-
-\DocumentationTok{\#\# using conditional linearity {-} leave out the Asym parameter}
-\NormalTok{fm2 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}} \DecValTok{1} \SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{)}
-\FunctionTok{summary}\NormalTok{(fm2)}
-
-\DocumentationTok{\#\# without conditional linearity}
-\NormalTok{fm3 }\OtherTok{\textless{}{-}} \FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1,}
-                 \AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm3)}
-
-\DocumentationTok{\#\# using Port\textquotesingle{}s nl2sol algorithm}
-\NormalTok{fm4 }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)),}
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym =} \DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"port"}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm4)}
-
-\DocumentationTok{\#\# using conditional linearity AND Asym does not work}
-\NormalTok{fm2a }\OtherTok{\textless{}{-}} \FunctionTok{try}\NormalTok{(}\FunctionTok{nls}\NormalTok{(density }\SpecialCharTok{\textasciitilde{}}\NormalTok{ Asym }\SpecialCharTok{/}\NormalTok{ (}\DecValTok{1} \SpecialCharTok{+} \FunctionTok{exp}\NormalTok{((xmid }\SpecialCharTok{{-}} \FunctionTok{log}\NormalTok{(conc)) }\SpecialCharTok{/}\NormalTok{ scal)), }
-                 \AttributeTok{data =}\NormalTok{ DNase1, }\AttributeTok{start =} \FunctionTok{list}\NormalTok{(}\AttributeTok{Asym=}\DecValTok{3}\NormalTok{, }\AttributeTok{xmid =} \DecValTok{0}\NormalTok{, }\AttributeTok{scal =} \DecValTok{1}\NormalTok{),}
-                 \AttributeTok{algorithm =} \StringTok{"plinear"}\NormalTok{, }\AttributeTok{trace =} \ConstantTok{TRUE}\NormalTok{))}
-\FunctionTok{summary}\NormalTok{(fm2a)}
-\end{Highlighting}
-\end{Shaded}
+\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}}
+
+The variable projection method (Golub and Pereyra (\protect\hyperlink{ref-Golub1973}{1973}), O'Leary and Rust (\protect\hyperlink{ref-OlearyRust13}{2013})) is
+usually much more effective than general approaches in finding good solutions
+to nonlinear least squares
+problems when some of the parameters appear linearly. In our logistic examples,
+the asymptote parameters are an illustration. However, identifying which parameters
+are linear and communicating this information to estimating functions
+is not a trivial task. \texttt{nls()} has an option \texttt{algorithm\ =\ "plinear"} that allows some partially linear
+models to be solved. The other tools, as far as we are aware, do not offer any
+such capability. The \texttt{nlstac} package uses a different algorithm for similar goals.
+
+Within \texttt{nls()} itself we must, unfortunately, use
+different specifications with different \texttt{algorithm} options.
+For example, the explicit linear model \texttt{y\ \textasciitilde{}\ a\ *\ x\ +\ b} does not work. As with
+the linear modeling function \texttt{lm()}, this must be specified as \texttt{y\ \textasciitilde{}\ x}.
+Within \texttt{nls()}, consider the following four different specifications for the same
+problem, plus an intuitive choice, labeled \texttt{fm2a}, that does not work.
+In this failed attempt, putting the \texttt{Asym} parameter in the model causes the
+\texttt{plinear} algorithm
+to try to add another term to the model. We believe this is unfortunate, and would
+like to see a consistent syntax. We do not foresee a resolution for this issue, which
+arises from different approaches to model specification. In the example, we have not evaluated
+the commands to save space.
+
+\begin{verbatim}
+DNase1 <- subset(DNase, Run == 1) # select the data
+## using a selfStart model - do not specify the starting parameters
+fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
+summary(fm1)
+
+## using conditional linearity - leave out the Asym parameter
+fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(xmid = 0, scal = 1),
+                 algorithm = "plinear")
+summary(fm2)
+
+## without conditional linearity
+fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1,
+                 start = list(Asym = 3, xmid = 0, scal = 1))
+summary(fm3)
+
+## using Port's nl2sol algorithm
+fm4 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+                 data = DNase1, start = list(Asym = 3, xmid = 0, scal = 1),
+                 algorithm = "port"))
+summary(fm4)
+
+## using conditional linearity AND Asym does not work
+fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
+                 data = DNase1, start = list(Asym=3, xmid = 0, scal = 1),
+                 algorithm = "plinear", trace = TRUE))
+summary(fm2a)
+\end{verbatim}
 
 \hypertarget{models-with-indexed-parameters}{%
-\subsection{Models with indexed
-parameters}\label{models-with-indexed-parameters}}
-
-Some models have several common parameters and others that are tied to
-particular cases. The \textbf{man} file for \texttt{nls()} includes an
-example of a situation in which parameters are indexed but which uses
-the ``plinear'' option as an added complication. Running this example
-reveals that the answers for the parameters are not indexed as in a
-vector. That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}}
-but \texttt{a1,\ a2,\ a3}. This is no doubt because programming for
-indexed parameters is challenging. We note that there are capabilities
-in packages for mixed-effects modeling such as \texttt{nlme} (Pinheiro
-et al. (\protect\hyperlink{ref-p-nlme}{2013})), \texttt{bbmle} (Bolker
-and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4} (D.
-Bates et al. (\protect\hyperlink{ref-p-lme4}{2015})) for estimating
-models of such type.
+\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}}
+
+Some models have several common parameters and
+others that are tied to particular cases.
+The \textbf{man} file for \texttt{nls()} includes an example of a situation in which
+parameters are indexed but which uses the ``plinear'' option as an added complication.
+Running this example reveals that the answers for the parameters are not indexed
+as in a vector.
+That is, we do not see \texttt{a{[}1{]},\ a{[}2{]},\ a{[}3{]}} but \texttt{a1,\ a2,\ a3}.
+This is no doubt because programming for indexed parameters is challenging.
+We note that there are capabilities in packages for mixed-effects modeling
+such as \texttt{nlme} (Pinheiro et al.
+(\protect\hyperlink{ref-p-nlme}{2013})), \texttt{bbmle} (Bolker and Team (\protect\hyperlink{ref-p-bbmle}{2013})), and \texttt{lme4} (D.
+Bates et al.
+(\protect\hyperlink{ref-p-lme4}{2015})) for
+estimating models of such type.
 
 \hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
-\subsection{Some other CRAN packages for nonlinear
-modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
-
-\texttt{onls} (Andrej-Nikolai Spiess
-(\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization
-in estimating nonlinear models is the vertical difference between the
-dependent variable and the functional model. \texttt{onls} minimizes the
-sum of \textbf{orthogonal} residuals. The objective is therefore
-different and involves nontrivial extra calculation. A vignette with the
-package and the blog article
+\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
+
+\texttt{onls} (Andrej-Nikolai Spiess (\protect\hyperlink{ref-R-onls-manual}{2022})) -- The usual optimization in estimating nonlinear models
+is the vertical difference between the dependent variable and the functional
+model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
+therefore different and involves nontrivial extra calculation. A vignette with
+the package and the blog article
 \url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
-give some description with illustrative graphs. \texttt{onls} appears to
-be limited to problems with one independent and one dependent variable.
-The Wikipedia article
-\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an
-overview of some ideas, with references to the literature. The approach
-needs a wider discussion and tutorial examples to allow its merits to be
-judged than can be included here.
-
-\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016}))
--- This package allows nonlinear estimation by controlled random search
-via two methods. There is unfortunately no vignette. A modest trial we
-carried out showed \texttt{nlsr::nlxb()} gave the same results in a
-small fraction of the time required by either of the methods in
-\texttt{crsnls}. The method discussed in Josef Tvrdík and Ivan Křivý and
-Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better
-reliability in finding solutions than a Levenberg-Marquardt code
-(actually from Matlab), but the tests were conducted on the extreme NIST
-examples mentioned next.
-
-\texttt{NISTnls} (National Institutes for Standards and R port by
-Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This
-package provides R code and data for a set of (numerically
-ill-conditioned) nonlinear least squares problems from the U.S. National
-Institute for Standards and Technology. These may not represent
-real-world situations.
-
-\texttt{nlshelper} (Duursma
-(\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which
-unfortunately lacks a vignette, provides a few utilities for
-summarizing, testing, and plotting non-linear regression models
-estimated with nls(), nlsList() or nlme() that are linked or grouped in
-some way.
-
-\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) --
-This solves nonlinear least squares problems with optional equality
-and/or inequality constraints. It is clearly \textbf{not} about
-modeling, and the input and output are quite different from class
-\texttt{nls} methods. However, there do not appear to be other R
-packages with these capabilities.
-
-\texttt{nlsMicrobio} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and
-nonlinear regression models dedicated to predictive microbiology,
-including a vignette, by authors of the \texttt{nlstools} package.
-
-\texttt{nlstools} (Baty and Delignette-Muller
-(\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides
-several tools for aiding the estimation of nonlinear models,
-particularly using \texttt{nls()}. The vignette is actually a journal
+give some description with illustrative graphs. \texttt{onls} appears to be limited to
+problems with one independent and one dependent variable. The Wikipedia article
+\url{https://en.wikipedia.org/wiki/Total_least_squares} presents an overview of some
+ideas, with references to the literature. The approach needs a wider
+discussion and tutorial examples to allow its merits to be judged than can be
+included here.
+
+\texttt{crsnls} (Tvrdík (\protect\hyperlink{ref-R-crsnls-manual}{2016})) -- This package allows nonlinear estimation by
+controlled random search via two methods. There is unfortunately no vignette.
+A modest trial we carried out showed \texttt{nlsr::nlxb()} gave the same results in
+a small fraction of the time required by either of the methods in \texttt{crsnls}. The
+method discussed in Josef Tvrdík and Ivan Křivý and Ladislav Mišík (\protect\hyperlink{ref-tvrdik2007}{2007}) claims better reliability in finding solutions
+than a Levenberg-Marquardt code (actually from Matlab), but the tests were conducted
+on the extreme NIST examples mentioned next.
+
+\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (\protect\hyperlink{ref-NISTnls-manual}{2012})) -- This package provides R code and data for a set of
+(numerically ill-conditioned) nonlinear least squares problems from the U.S. National
+Institute for Standards and Technology. These may not represent real-world situations.
+
+\texttt{nlshelper} (Duursma (\protect\hyperlink{ref-nlshelper-manual}{2017})) -- This package, which unfortunately lacks a
+vignette, provides a few utilities for summarizing, testing, and plotting
+non-linear regression models estimated with nls(), nlsList() or nlme() that
+are linked or grouped in some way.
+
+\texttt{nlsic} (Sokol (\protect\hyperlink{ref-nlsic-manual}{2022})) -- This solves nonlinear least squares problems with optional
+equality and/or inequality constraints. It is clearly \textbf{not} about modeling, and
+the input and output are quite different from class \texttt{nls} methods. However, there do
+not appear to be other R packages with these capabilities.
+
+\texttt{nlsMicrobio} (Baty and Delignette-Muller (\protect\hyperlink{ref-nlsMicrobio-manual}{2014})) -- Data sets and nonlinear regression models
+dedicated to predictive microbiology, including a vignette, by authors of the
+\texttt{nlstools} package.
+
+\texttt{nlstools} (Baty and Delignette-Muller (\protect\hyperlink{ref-p-nlstools}{2013})) -- This package provides several tools for aiding the estimation
+of nonlinear models, particularly using \texttt{nls()}. The vignette is actually a journal
 article, and the authors have considerable experience in the subject.
 
-\texttt{nlsmsn} (Prates, Lachos, and Garay
-(\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear
-scale mixture of skew-normal(NL-SMSN) regression, with details in Garay,
-Lachos, and Abanto-Valle (2011). The problem here is to minimize an
-objective that is modified from the traditional sum of squared
+\texttt{nlsmsn} (Prates, Lachos, and Garay (\protect\hyperlink{ref-m-nlsmsn}{2021})) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN)
+regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
+here is to minimize an objective that is modified from the traditional sum of squared
 residuals.
 
-\texttt{nls.multstart} (Padfield and Matheson
-(\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least
-squares regression using AIC scores with the Levenberg-Marquardt
-algorithm using multiple starting values for increasing the chance that
-the minimum found is the global minimum.
-
-\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) --
-Nonlinear least squares by brute force has similar motivations to
-\texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials.
-The author has extensive expertise in R.
-
-\texttt{nlstac} (Rodriguez-Arias et al.
-(\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions
-implementing the algorithm described in Fernandez Torvisco et al.~(2018)
-for fitting separable nonlinear regression curves. The special class of
-problem for which this package is intended is an important and difficult
-one. No vignette is provided, unfortunately.
-
-\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models
-are treated, but there is minimal documentation and no vignette. Package
-\texttt{nlraa} is to be preferred.
-
-\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021}))
--- a set of nonlinear \emph{selfStart} models, primarily from
+\texttt{nls.multstart} (Padfield and Matheson (\protect\hyperlink{ref-nlsmultstart}{2020})) -- Non-linear least squares regression using
+AIC scores with the Levenberg-Marquardt algorithm using multiple starting values
+for increasing the chance that the minimum found is the global minimum.
+
+\texttt{nls2} (Grothendieck (\protect\hyperlink{ref-nls2-22}{2022})) -- Nonlinear least squares by brute force has similar
+motivations to \texttt{nls.multstart}, but uses \texttt{nls()} within multiple trials. The author
+has extensive expertise in R.
+
+\texttt{nlstac} (Rodriguez-Arias et al. (\protect\hyperlink{ref-m-nlstac}{2020})) -- A set of functions implementing the algorithm described in
+Fernandez Torvisco et al.~(2018) for fitting separable nonlinear regression curves. The
+special class of problem for which this package is intended is an important and
+difficult one. No vignette is provided, unfortunately.
+
+\texttt{easynls} -- Fit and plot some nonlinear models. Thirteen models are treated, but
+there is minimal documentation and no vignette. Package \texttt{nlraa} is to be preferred.
+
+\texttt{nlraa} (Miguez (\protect\hyperlink{ref-MiguezNLRAA2021}{2021})) -- a set of nonlinear \emph{selfStart} models, primarily from
 agriculture. Most include analytic Jacobian code.
 
-\texttt{optimx} (John C. Nash and Varadhan
-(\protect\hyperlink{ref-p-optimx}{2011})) -- This provides optimizers
-that can be applied to minimize a nonlinear function which could be a
-nonlinear sum of squares. This is not generally recommended if nonlinear
-least squares programs can be easily used, but provides a check and
-alternative solvers.
+\texttt{optimx} (John C. Nash and Varadhan (\protect\hyperlink{ref-p-optimx}{2011})) -- This provides
+optimizers that can be applied to minimize a nonlinear function which could be
+a nonlinear sum of squares. This is not generally recommended if nonlinear
+least squares programs can be easily used, but provides a check and alternative
+solvers.
 
 \hypertarget{tests-and-use-case-examples}{%
-\subsection{Tests and use-case
-examples}\label{tests-and-use-case-examples}}
-
-Maintainers of packages need suitable tests and use-case examples in
-order
+\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}}
+
+Maintainers of packages need suitable tests and use-case examples in order
 
@@
 \item
-  to ensure packages work properly, in particular, giving results
-  comparable to or better than the functions they are to replace.
-\item
-  to test individual solver functions to ensure they work across the
-  range of calling mechanisms, that is, different ways of supplying
-  inputs to the solver(s);
-\item
-  to pose ``silly'' inputs to see if these bad inputs are caught by the
-  programs.
+  to ensure packages work properly, in particular, giving
+  results comparable to or better than the functions they are to replace.
+\item
+  to test individual solver functions to ensure they work across the range of
+  calling mechanisms, that is, different ways of supplying inputs to the solver(s);
+\item
+  to pose ``silly'' inputs to see if these bad inputs are caught by the programs.
 \end{itemize}
 
-Such goals align with the aims of \textbf{unit testing} (Wickham
-(\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al.
-(\protect\hyperlink{ref-HWdevtools21}{2021}), the conventional R package
-testing tools, and a number of blog posts found by searching ``unit
-testing in R''). In our work, one of us has developed a working
-prototype package at
-\emph{\url{https://github.com/ArkaB-DS/nlsCompare}}. A primary design
-objective of this is to allow the summarisation of multiple tests in a
-compact output. The prototype has a vignette to illustrate its use.
+Such goals align with the aims of \textbf{unit testing}
+(Wickham (\protect\hyperlink{ref-HWtestthat11}{2011}), Wickham et al. (\protect\hyperlink{ref-HWdevtools21}{2021}), the conventional R package testing tools, and
+a number of blog posts found by searching ``unit testing in R'').
+In our work, one of us has developed a working prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}}.
+A primary design objective of this is to allow the summarisation of multiple tests in
+a compact output. The prototype has a vignette to illustrate its use.
 
 \hypertarget{future-of-nonlinear-model-estimation-in-r}{%
-\subsection{Future of nonlinear model estimation in
-R}\label{future-of-nonlinear-model-estimation-in-r}}
-
-Given its importance to R, it is possible that \texttt{nls()} will
-remain more or less as it has been for the past several decades. If so,
-the focus of discussion should be the measures needed to secure its
-continued operation for legacy purposes and how that may be
-accomplished. We welcome an opportunity to participate in such
-conversations.
-
-To advance the stability and maintainability of R, we believe the
-program objects (R functions) that are created by tools such as
-\texttt{nls()} should have minimal cross-linkages and side-effects. The
-aspects of \texttt{nls()} that concern us follow.
+\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}}
+
+Given its importance to R, it is possible that \texttt{nls()} will remain more or less
+as it has been for the past several decades. If so, the focus of discussion
+should be the measures needed to secure its continued operation for legacy
+purposes and how that may be accomplished.
+We welcome an opportunity to participate in such conversations.
+
+To advance the stability and maintainability of R, we believe the program
+objects (R functions)
+that are created by tools such as \texttt{nls()} should have minimal
+cross-linkages and side-effects.
+The aspects of \texttt{nls()} that concern us follow.
 
@@
 
-Hans Werner Borchers was helpful in developing the GSoC project
-motivating this article and in comments on this and related work.
-Heather Turner co-mentored the project and helped guide the progress of
-the work. Exchanges with Fernando Miguez helped to clarify aspects of
-\texttt{selfStart} models and instigated the ``Introduction to nlsr''
-vignette. Colin Gillespie (package \texttt{benchmarkme}) has been
-helpful in guiding our attempts to succinctly summarize computing
-environments. Particularly thoughtful reviews by the referees, including
-details of inner workings of R, have led to important revisions.
+Hans Werner Borchers was helpful in developing the GSoC project motivating this article
+and in comments on this and related work. Heather Turner co-mentored the project
+and helped guide the progress of the work. Exchanges with Fernando Miguez helped to
+clarify aspects of \texttt{selfStart} models and instigated the ``Introduction to nlsr''
+vignette. Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our
+attempts to succinctly summarize computing environments. Particularly thoughtful
+reviews by the referees, including details of inner workings of R,
+have led to important revisions.
 
 \hypertarget{references}{%
-\section*{References}\label{references}}
+\section{References}\label{references}}
+
+\hypertarget{references-1}{%
+\section*{References}\label{references-1}}
 \addcontentsline{toc}{section}{References}
@@
 \leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
-Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear
-Least-Squares Regression}.
-\url{https://CRAN.R-project.org/package=onls}.
+Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
-Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis
-and Its Applications}. Wiley.
+Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
 
 \leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
-Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset
-Orthogonality Convergence Criterion for Nonlinear Least Squares.''}
-\emph{Technometrics} 23 (2): 179--83.
+Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-lme4}{}}%
-Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker. 2015.
-{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of
-Statistical Software} 67 (1): 1--48.
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker.
+2015.
+{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of Statistical Software} 67 (1): 1--48.
 \url{https://doi.org/10.18637/jss.v067.i01}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
-Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools:
-Tools for Nonlinear Regression Diagnostics}.
+Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
----------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression
-Models Dedicated to Predictive Microbiology}.
+---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
-Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for
-General Maximum Likelihood Estimation}.
-\url{http://CRAN.R-project.org/package=bbmle}.
+Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
-Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}.
-\url{https://CRAN.R-project.org/package=gslnls}.
+Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
-Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for
-Non-Linear Regression}.
-\url{https://CRAN.R-project.org/package=nlshelper}.
+Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
-Elzhov, Timur V., Katharine M. Mullen, Andrej-Nikolai Spiess, and Ben
-Bolker. 2012. \emph{Minpack.lm: R Interface to the Levenberg-Marquardt
-Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for
-Bounds}. R Project for Statistical Computing.
+Elzhov, Timur V., Katharine M.
+Mullen, Andrej-Nikolai Spiess, and Ben Bolker.
+2012.
+\emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}.
+R Project for Statistical Computing.
 \url{http://CRAN.R-project.org/package=minpack.lm}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-FoxHallSchryer1978}{}}%
-Fox, P. A., A. P. Hall, and N. L. Schryer. 1978. {``The PORT
-Mathematical Subroutine Library.''} \emph{ACM Trans. Math. Softw.} 4
-(2): 104--26. \url{https://doi.org/10.1145/355780.355783}.
+Fox, P. A., A. P. Hall, and N. L. Schryer. 1978. {``The PORT Mathematical Subroutine Library.''} \emph{ACM Trans. Math. Softw.} 4 (2): 104--26. \url{https://doi.org/10.1145/355780.355783}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
-Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard
-Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third
-Edition, for GSL Version 1.12 (3. Ed.).}
+Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
 
 \leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
-Gentleman, Robert C., Vincent J. Carey, Douglas M. Bates, Ben Bolstad,
-Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al. 2004.
-{``Bioconductor: Open Software Development for Computational Biology and
-Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
+Gentleman, Robert C., Vincent J.
+Carey, Douglas M.
+Bates, Ben Bolstad, Marcel Dettling, Sandrine Dudoit, Byron Ellis, et al.
+2004.
+{``Bioconductor: Open Software Development for Computational Biology and Bioinformatics.''} \emph{{Genome Biology}} 5 (R80).
 \url{https://doi.org/10.1186/gb-2004-5-10-r80}.
@@
 \leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
-Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain
-Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied
-Mathematics} 2: 164-\/-168.
+Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
-Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation
-of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics}
-11 (2): 431--41.
+Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
 
 \leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
-Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for
-Agricultural Applications}}.
-\url{https://CRAN.R-project.org/package=nlraa}.
+Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-More1981TUO}{}}%
-Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981.
-{``Testing Unconstrained Optimization Software.''} \emph{J-Toms} 7 (1):
-17--41.
+Moré, Jorge J., Burton S. Garbow, and Kenneth E. Hillstrom. 1981. {``Testing Unconstrained Optimization Software.''} \emph{J-Toms} 7 (1): 17--41.
 
 \leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
-Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on
-a Small Computer.''} \emph{Journal of the Institute for Mathematics and
-Its Applications} 19: 231--37.
+Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
 
 \leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
----------. 1979. \emph{Compact Numerical Methods for Computers: Linear
-Algebra and Function Minimisation}. Bristol: Adam Hilger.
+---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
-Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A
-Replacement and Extension of the {optim()} Function}. Nash Information
-Services Inc.; Johns Hopkins University.
+Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
-Nash, John C., and Paul Velleman. 1996. {``Nonlinear Estimation
-Combining Visual Fitting with Optimization Methods.''} In
-\emph{Proceedings of the Section on Physical and Engineering Sciences of
-the American Statistical Association}, 256--61. American Statistical
-Association.
+Nash, John C., and Paul Velleman.
+1996.
+{``Nonlinear Estimation Combining Visual Fitting with Optimization Methods.''} In \emph{Proceedings of the Section on Physical and Engineering Sciences of the American Statistical Association}, 256--61.
+American Statistical Association.
 
 \leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
-National Institutes for Standards, original from, and Technology
-http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by
-Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples
-from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.
+National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.
+2012.
+\emph{NISTnls: Nonlinear Least Squares Examples from NIST}.
+\url{https://CRAN.R-project.org/package=NISTnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
-O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for
-Nonlinear Least Squares Problems.''} \emph{Computational Optimization
-and Applications} 54 (3): 579--93.
+O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
-Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart:
-Robust Non-Linear Regression Using AIC Scores}.
-\url{https://CRAN.R-project.org/package=nls.multstart}.
+Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
-Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R
-Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
+Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
-Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn:
-Fitting Nonlinear Models with Scale Mixture of Skew-Normal
-Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
+Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
-Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified
-Practical Approach}. New York; Basel: Marcel Dekker Inc.
+Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
 \leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
-Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and
-Rafael Benitez. 2020. \emph{Nlstac: An r Package for Fitting Separable
-Nonlinear Models}. \url{https://CRAN.R-project.org/package=nlstac}.
+Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez.
+2020.
+\emph{Nlstac: An r Package for Fitting Separable Nonlinear Models}.
+\url{https://CRAN.R-project.org/package=nlstac}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-Rosenbrock1960}{}}%
-Rosenbrock, H. H. 1960. {``An Automatic Method for Finding the Greatest
-or Least Value of a Function.''} \emph{The Computer Journal} 3: 175--84.
+Rosenbrock, H. H. 1960. {``An Automatic Method for Finding the Greatest or Least Value of a Function.''} \emph{The Computer Journal} 3: 175--84.
 
 \leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%
-Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with
-Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
+Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
-Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters
-Estimation by 'CRS4HC' and 'CRS4HCe'}.
-\url{https://CRAN.R-project.org/package=crsnls}.
+Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
-Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''}
-\emph{The R Journal} 3: 5--10.
-\url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
+Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
 
 \leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
-Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021.
-\emph{{devtools: Tools to Make Developing R Packages Easier}}.
-\url{https://CRAN.R-project.org/package=devtools}.
+Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.
 
@@
 
-\end{document}
+\bibliography{./StateOfRToolsForNLS.bib}
+
+\address{%
+John C. Nash\\
+retired professor, University of Ottawa\\%
+Telfer School of Management\\ Ottawa ON Canada K1N 6N5\\
+%
+%
+\textit{ORCiD: \href{https://orcid.org/0000-0002-2762-8039}{0000-0002-2762-8039}}\\%
+\href{mailto:profjcnash@gmail.com}{\nolinkurl{profjcnash@gmail.com}}%
+}
+
+\address{%
+Arkajyoti Bhattacharjee\\
+Indian Institute of Technology\\%
+Department of Mathematics and Statistics\\ Kanpur\\
+%
+%
+%
+\href{mailto:arkastat98@gmail.com}{\nolinkurl{arkastat98@gmail.com}}%
+}

### 2023-07-20 RefactoringNLS/revision202307/RJournal2023-15Rev1ResubmissionRJstyle/StateOfRToolsForNLS.R

@@ -1,0 +1,250 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit StateOfRToolsForNLS.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = TRUE)
+## require(bookdown) # language engine to display text - does not seem necessary
+
+
+## ----ex01, echo=TRUE----------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE-------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100* c1 / (1 + 10 * c2 * exp(-0.1 *c3 * tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
+#
+# Starting parameter sets
+stu1<-c(b1 =1, b2 = 1, b3 = 1)
+sts1<-c(c1 = 1, c2 = 1, c3 = 1)
+stt1<-c(Asym = 1, xmid = 1, scal = 1)
+
+
+## ----ex02fn, echo=TRUE--------------------------------------------------------
+# Logistic3U
+hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
+  # This variant uses looping
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
+}
+
+hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
+  yy  <-  exp(-x[3] * tt)
+  zz  <-  1.0 / (1 + x[2] * yy)
+  jj[tt,1]   <-   zz
[…]

### 2023-10-18 JNABRev20231018/JNAB2review-response.txt
2nd revision

Response by Nash and Bhattacharjee to second set of referees' reports.

Our responses are prefaced by ">> " on each line.

2-review-2:

I believe the authors have adequately addressed my main comments/questions in the revised version of the paper. I have no other (major) concerns that still need clarification.

>> No response necessary.

2-review-1

Referee report of “The Current State of R Tools for Nonlinear Least Squares Modeling”.

I appreciate the effort that the authors have made in this revision. I believe the paper has
been considerably improved.

>> Thanks. Appreciated.

I think it would useful to expand a bit on some of the comments in the section “An illustrative
example”. In particular, the choice of starting estimates, at least for the third formulation of
the model, doesn’t make sense to me.

>> Added explanatory comment.
>> <!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
>> We thank one of the referees for expressing the viewpoint that a nonlinear regression
>> approach would use background knowledge and graphs to obtain an understanding of
>> the general magnitude of the parameters. Indeed, the "visual fitting" approach
>> (@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
>> of the **Logistic3T** variant of the model. 
>> 
>> Our emphasis on the software resilience to starting parameters that are, in the
>> sense of statistical modeling, "silly" comes from over half a century of dealing
>> with users whose interests and understanding are very far from those of statistical
>> modelers. Thus we seek methods and codes that obtain reasonable answers under
>> highly unfavourable conditions. Nevertheless, a proper approach to 
>> nonlinear modeling is to apply all available knowledge to the task.
>> <!-- end 231018 work -->

Purpose of nonlinear least squares fits

In contrast to linear least squares or techniques like generalized additive models (GAMs), which
provide empirical or data-driven models, I feel that the goal of a nonlinear regression analysis,
at least in part, centers on the parameter estimates and the interpretation of the parameters.
That is, the model is usually derived from a mechanistic model of how the data are generated,
this model depends on certain parameters, and the estimates of those parameters are of interest
in and of themselves.

If the parameters are of interest, the analyst should be able to “ball park” the values of the
parameters from the data, probably aided by plotting the data.
I feel there are three stages in a nonlinear least squares fit:

1. Determine starting estimates of the parameters from the data and knowledge of the
interpretation of the parameters.

2. Starting at these estimates use an iterative algorithm that seeks to reduce the sum of
squared residuals at each iteration.

3. Determine if convergence has been achieved at the current iteration and, if not, go to
step 2.

Admittedly, the vast majority of research and software development is focussed on step 2 but
steps 1 and 3 are also important and an analyst performing a nonlinear regression analysis
should be able to formulate starting estimates for a model.

Figure 1: Hobbs weed infestation data with fitted 3-parameter logistic curve (blue) and curve
from the starting estimates in third parameterization of the model (green).
Interpretation of parameters in the three-parameter logistic model
The logistic growth model is based on vertically scaling and possibly shifting the cumulative
distribution function of the logistic distribution. If the lower limit of the predicted response is
set at zero, there is only a vertical scaling constant which determines the asymptotic predicted
response as x → ∞. The three-parameter logistic model incorporates this vertical scaling
parameter with location and scale parameters on the horizontal axis. (The four-parameter
logistic model incorporates both shift and scale parameters on the vertical axis.)
The common way of writing location/scale parameters on the horizontal axis results in the
2location being the value on the x-axis of the inflection point (when the fitted value is midway
between the lower and upper vertical limits) and the scale being the x-axis distance between
the midpoint and the point where the fitted value is roughly one-quarter of the distance from
the lower limit and the upper vertical limits.
It is straightforward to look at a data plot, such as Figure 1, and choose appropriate starting
values for the parameters in this representation of the three-parameter logistic.
However, one does not look at this plot and decide that (1, 1, 1), which would give the green
line in the plot, is a reasonable set of parameter values for this model applied to these data.
As we can see the fitted curve is a very good fit to these data but the curve from the unit
starting estimates is not even close to the observed data. It doesn’t even overlap on the y-axis
with the observed data.

>> We agree the starting points are bad, for the reason given above.

So these are egregiously bad starting estimates for the parameters. The authors state that
fitting this model is very tricky but in my previous report I showed that is was trivial to fit
this model using the self-starting model function SSlogis, because it goes through steps of
using the data to obtain good starting values.

>> Added comment to point out selfStart should be used. 
>> 
>> Users should, of course, be using the `SSlogis` *selfStart* tool, but ignorance
>> of this possibility or a slight variant in the model can easily lead to outcomes
>> similar to those seen here.

A good portion of this paper is studying whether different nonlinear least squares software
implementations can converge to appropriate parameters estimates from these ridiculously
bad starting values and, quite frankly, I don’t see the point.

>> To editor: We are not sure that we can resolve this difference of opinion with the
>>  referee. From a modeling point of view, he/she is definitely correct. From a software
>>  reliability / resilience perspective, we feel it is important that codes keep going
>>  and get decent answers if possible, and we have shown that even for very nasty cases,
>>  software can "work" successfully. Moreover, even if software cannot proceed, it should
>>  provide more guidance if possible than "Singular gradient", which is not meaningful to
>>  most current R users.

It is the author’s prerogative to use whatever criterion for comparison of nonlinear least squares
implementations that they feel is appropriate but I think they should clarify to the reader that
they are using intentionally bad and inappropriate starting estimates as a way of stressing
algorithms.

>> Agreed, and we believe we have now done so. To be honest, the original c(1,1,1) parameters
>> of the 1970s were likely pure laziness and quick to type.

On page 4 the authors state that if a selfStart model is not available nls() sets the starting
values to 1. That may be the case now but the original intention was certainly that a user
calling nls() would provide starting estimates suitable to the model and data.

>> To editor: Referee may be revealing their identity!

Also on page 4 I don’t understand the comment in the hobbs.res function that This variant
uses looping. I don’t see any explicit loops. There will naturally be looping in the C code
that implements the arithmetic operations on vectors but that is inevitable.

>> The comment was a left-over from an earlier variant of the functions and should have been
>> removed. Done. Thanks for spotting this.
>> We also noted that the model is UNscaled, and edited the comment.
>> An explanatory comment has been added
>> 
>> The particular form of these explicit residual and Jacobian functions comes from
>> their translation from BASIC codes of the 1970s, as adapted in @jnmws87. The use
>> of the Laplace form of the residual and the inclusion of data within the functions
>> reflects choices of that era that are at odds with current practice. Some users
>> do still want or need to provide problems as explicit functions, particularly
>> for problems that are NOT regressions. For example, the Rosenbrock banana-valley
>> test problem can be provided this way, where the two "residuals" are different
>> functional forms.

Is it really intended that anyone using the function-based specification of the model will encode
their data into the model function like hobbs.res and hobbs.jac? Generally we would expect
to separate the specification of the data from the model - otherwise there is a great risk of,
say, the residual and the Jacobian function accidentally getting out of sync.

>> In October 2023, one of us (JN) had a query on usage of this formulation where 
>> performance was important. Moreover, the query asked how to extract standard errors
>> and has led to a minor but important addition to the "Intro to nlsr" vignette.
>> Generally we agree that users today are much less likely to use explicit code.

Also, this is the first time that the authors have invoked their non-standard definition of
“residual” as the negative of what most others mean by “residual”, and they haven’t mentioned
this. I would discourage the authors from using an idiosyncratic definition of “residual” simply
to save writing a negative sign in a formula for the Jacobian.

>> We believe our comment above addresses this. We have explained our reasoning for the use
>> of the Laplace residual in other works, but present usage or commentary is a red herring
>> that is irrelevant to today's software and practice. Indeed, the main advance that package
>> 'nlsr' represents is to automatically provide analytic derivatives that caused so much work
>> and introduced so many errors in the 1970s and 1980s.

### 2023-10-18 JNABRev20231018/ComparingNLStools.Rmd
2nd revision

--- 2023-05-22 JNABsub20230518/ComparingNLStools.Rmd
@@
 ```
+
+<!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
+We thank one of the referees for expressing the viewpoint that a nonlinear regression
+approach would use background knowledge and graphs to obtain an understanding of
+the general magnitude of the parameters. Indeed, the "visual fitting" approach
+(@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
+of the **Logistic3T** variant of the model.
+
+Our emphasis on the software resilience to starting parameters that are, in the
+sense of statistical modeling, "silly" comes from over half a century of dealing
+with users whose interests and understanding are very far from those of statistical
+modelers. Thus we seek methods and codes that obtain reasonable answers under
+highly unfavourable conditions. Nevertheless, a proper approach to 
+nonlinear modeling is to apply all available knowledge to the task.
+<!-- end 231018 work -->
 
@@
 It is also possible to provide R functions for the residual and Jacobian. 
-This is usually more work for the user if the formula setup is possible.
+This is usually much more work for the user if the formula setup is possible.
 To illustrate, we show the functions for the unscaled 3 parameter logistic.
+<!-- 231018: added a note to explain deficiencies -->
+The particular form of these explicit residual and Jacobian functions comes from
+their translation from BASIC codes of the 1970s, as adapted in @jnmws87. The use
+of the Laplace form of the residual and the inclusion of data within the functions
+reflects choices of that era that are at odds with current practice. Some users
+do still want or need to provide problems as explicit functions, particularly
+for problems that are NOT regressions. For example, the Rosenbrock banana-valley
+test problem can be provided this way, where the two "residuals" are different
+functional forms.
+
+<!-- 231018: removed "this variant uses looping" -->
+<!-- Added note on form of the residual. See Stigler (1986) -->
+<!--   History of Statistics pg.34 -->
+<!-- Note: UNscaled, not scaled -->
 
@@
 # Logistic3U
-hobbs.res  <-  function(x){ # scaled Hobbs weeds problem -- residual
-  # This variant uses looping
+hobbs.res  <-  function(x){ # unscaled Hobbs weeds problem -- residual
   if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
@@
   res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+  # Note: this form of the residual, from Laplace (1788) in the form
+  # of "fitted - observed" has been used in our software for half
+  # a century, but sometimes offends users of the more traditional
+  # "observed - fitted" form
 }
 
-hobbs.jac  <-  function(x) { # scaled Hobbs weeds problem -- Jacobian
+hobbs.jac  <-  function(x) { # unscaled Hobbs weeds problem -- Jacobian
   jj  <-  matrix(0.0, 12, 3)
@@
 Here we see the infamous "singular gradient" termination message of `nls()`. 
+<!-- Added comment to point out selfStart should be used. -->
+Users should, of course, be using the `SSlogis` *selfStart* tool, but ignorance
+of this possibility or a slight variant in the model can easily lead to outcomes
+similar to those seen here.
+<!-- End added comment -->
 

### 2023-10-18 JNABRev20231018/ComparingNLStools.tex
2nd revision

--- 2023-05-18 JNABsub20230518/ComparingNLStools.tex
@@
 
-\hypertarget{the-nls-function}{%
-\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}
+\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}
 
@@
 from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
-The function dates to the 1980s and the work related to Bates and Watts (1988) in S (see
+The function dates to the 1980s and the work related to D. M. Bates and Watts (1988) in S (see
 \emph{\url{https://en.wikipedia.org/wiki/S_\%28programming_language\%29}}).
@@
 
-\hypertarget{scope-of-our-comparison}{%
-\subsection{Scope of our comparison}\label{scope-of-our-comparison}}
+\subsection{Scope of our comparison}\label{scope-of-our-comparison}
 
@@
 
-\hypertarget{principal-messages}{%
-\subsubsection{Principal messages}\label{principal-messages}}
+\subsubsection{Principal messages}\label{principal-messages}
 
@@
 
-\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
-\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
+\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}
 
@@
 
-\hypertarget{an-illustrative-example}{%
-\subsection{An illustrative example}\label{an-illustrative-example}}
+\subsection{An illustrative example}\label{an-illustrative-example}
 
@@
 
+We thank one of the referees for expressing the viewpoint that a nonlinear regression
+approach would use background knowledge and graphs to obtain an understanding of
+the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
+(John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
+of the \textbf{Logistic3T} variant of the model.
+
+Our emphasis on the software resilience to starting parameters that are, in the
+sense of statistical modeling, ``silly'' comes from over half a century of dealing
+with users whose interests and understanding are very far from those of statistical
+modelers. Thus we seek methods and codes that obtain reasonable answers under
+highly unfavourable conditions. Nevertheless, a proper approach to
+nonlinear modeling is to apply all available knowledge to the task.
+
 Three suggested models for this data are (with names to allow for easy reference)
@@
 
-\hypertarget{problem-setup}{%
-\subsubsection{Problem setup}\label{problem-setup}}
+\subsubsection{Problem setup}\label{problem-setup}
 
@@
 It is also possible to provide R functions for the residual and Jacobian.
-This is usually more work for the user if the formula setup is possible.
+This is usually much more work for the user if the formula setup is possible.
 To illustrate, we show the functions for the unscaled 3 parameter logistic.
+The particular form of these explicit residual and Jacobian functions comes from
+their translation from BASIC codes of the 1970s, as adapted in John C. Nash and Walker-Smith (1987). The use
+of the Laplace form of the residual and the inclusion of data within the functions
+reflects choices of that era that are at odds with current practice. Some users
+do still want or need to provide problems as explicit functions, particularly
+for problems that are NOT regressions. For example, the Rosenbrock banana-valley
+test problem can be provided this way, where the two ``residuals'' are different
+functional forms.
 
@@
 
-\hypertarget{estimation-of-models-specified-as-formulas}{%
-\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}
+\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}
 
@@
 Here we see the infamous ``singular gradient'' termination message of \texttt{nls()}.
-
-\hypertarget{solution-attempts-with-nlsr}{%
-\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+Users should, of course, be using the \texttt{SSlogis} \emph{selfStart} tool, but ignorance
+of this possibility or a slight variant in the model can easily lead to outcomes
+similar to those seen here.
+
+\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}
+
+\begin{verbatim}
+#> unlx1:
+\end{verbatim}
 
@@
 
-\hypertarget{solution-attempts-with-minpack.lm}{%
-\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}
+\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}
 
@@
 
-\hypertarget{solution-attempts-with-gslnls}{%
-\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}
+\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}
 
@@
 
-\hypertarget{comparison-notes-for-formula-setup-solutions}{%
-\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}
+\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}
 
@@
 
-\hypertarget{functional-specification-of-problems}{%
-\subsection{Functional specification of problems}\label{functional-specification-of-problems}}
+\subsection{Functional specification of problems}\label{functional-specification-of-problems}
 
@@
 
-\hypertarget{design-goals-termination-tests-and-output-objects}{%
-\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}
+\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}
 
@@
 
-\hypertarget{returned-results-of-nls-and-other-tools}{%
-\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}
+\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}
 
@@
 
-\hypertarget{when-to-compute-ancillary-information}{%
-\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}
+\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}
 
@@
 
-\hypertarget{jacobian-calculation}{%
-\subsection{Jacobian calculation}\label{jacobian-calculation}}
+\subsection{Jacobian calculation}\label{jacobian-calculation}
 
@@
 
-\hypertarget{jacobian-code-in-selfstart-models}{%
-\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}
+\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}
 
@@
 
-\hypertarget{bounds-constraints-on-parameters}{%
-\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}}
+\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}
 
@@
 
-\hypertarget{philosophical-considerations}{%
-\subsubsection{Philosophical considerations}\label{philosophical-considerations}}
+\subsubsection{Philosophical considerations}\label{philosophical-considerations}
 
@@
 
-\hypertarget{fixed-parameters-masks}{%
-\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}}
+\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}
 
@@
 
-\hypertarget{stabilization-of-gauss-newton-computations}{%
-\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}}
+\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}
 
@@
 
-\hypertarget{programming-language}{%
-\subsubsection{Programming language}\label{programming-language}}
+\subsubsection{Programming language}\label{programming-language}
 
@@
 
-\hypertarget{data-sources-for-problems}{%
-\subsection{Data sources for problems}\label{data-sources-for-problems}}
+\subsection{Data sources for problems}\label{data-sources-for-problems}
 
@@
 
-\hypertarget{feature-subsetting}{%
-\subsubsection{Feature: Subsetting}\label{feature-subsetting}}
+\subsubsection{Feature: Subsetting}\label{feature-subsetting}
 
 \texttt{nls()} and other class \texttt{nls} tools accept an argument \texttt{subset}. This acts through
-the mediation of \texttt{model.frame}, which not obvious in the
-source code files \texttt{/src/library/stats/R/nls.R} and \texttt{/src/library/stats/src/nls.C}.
+the mediation of \texttt{model.frame}, which is not obvious in the
+source code files
+\texttt{/src/library/stats/R/nls.R} and
+\texttt{/src/library/stats/src/nls.C}.
 Having \texttt{subset} at the level of the call to a function like
@@
 
-\hypertarget{feature-na.action-missing-value-treatment}{%
-\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}}
+\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}
 
@@
 
-\hypertarget{feature-model-frame}{%
-\subsection{Feature: model frame}\label{feature-model-frame}}
+\subsection{Feature: model frame}\label{feature-model-frame}
 
@@
 
-\hypertarget{weights-on-observations}{%
-\subsection{Weights on observations}\label{weights-on-observations}}
+\subsection{Weights on observations}\label{weights-on-observations}
 
@@
 
-\hypertarget{weights-in-returned-functions-from-nls}{%
-\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}
+\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}
 
@@
 
-\hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
-\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
-
-\hypertarget{interim-output-from-the-port-algorithm}{%
-\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}
+\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}
+
+\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}
 
@@
 
-\hypertarget{failure-to-return-the-best-result-achieved}{%
-\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}}
+\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}
 
@@
 unless \texttt{trace} is invoked, the user will not get any information about this.
-This almost trivial change to the \texttt{nls()} function could be
-useful to R users.
-
-\hypertarget{estimating-models-that-are-partially-linear}{%
-\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}}
+Changing this in the \texttt{nls()} function could be useful to R users, and would
+be almost trivial.
+
+\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}
 
 The variable projection method (Golub and Pereyra (1973), O'Leary and Rust (2013)) is
-generally much more effective than general approaches in finding good solutions
+usually much more effective than general approaches in finding good solutions
 to nonlinear least squares
@@
 
-\hypertarget{models-with-indexed-parameters}{%
-\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}}
+\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}
 
@@
 We note that there are capabilities in packages for mixed-effects modeling
-such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (p-lme4) for
+such as \texttt{nlme} (Pinheiro et al. (2013)), \texttt{bbmle} (Bolker and Team (2013)), and \texttt{lme4} (D. Bates et al. (2015)) for
 estimating models of such type.
 
-\hypertarget{tests-and-use-case-examples}{%
-\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}}
+\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}
 
@@
 
-\hypertarget{documentation-and-resources}{%
-\subsection{Documentation and resources}\label{documentation-and-resources}}
+\subsection{Documentation and resources}\label{documentation-and-resources}
 
@@
 
-\hypertarget{future-of-nonlinear-model-estimation-in-r}{%
-\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}}
+\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}
 
@@
 
-\hypertarget{acknowledgments}{%
-\section{Acknowledgments}\label{acknowledgments}}
+\section{Acknowledgments}\label{acknowledgments}
 
@@
 
-\hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{references-1}{%
-\section*{References}\label{references-1}}
+\section*{References}\label{references}
 \addcontentsline{toc}{section}{References}
 
-\hypertarget{refs}{}
-\begin{CSLReferences}{1}{0}
-\leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
+\phantomsection\label{refs}
+\setlength{\cslentryspacing}{0em}
+\begin{CSLReferences}
+\bibitem[\citeproctext]{ref-R-onls-manual}
 Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
+\bibitem[\citeproctext]{ref-bateswatts}
 Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
 
-\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
+\bibitem[\citeproctext]{ref-BatesWatts81}
 Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
 
-\leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
+\bibitem[\citeproctext]{ref-p-lme4}
+Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker.
+2015.
+{``Fitting Linear Mixed-Effects Models Using {lme4}.''} \emph{Journal of Statistical Software} 67 (1): 1--48.
+\url{https://doi.org/10.18637/jss.v067.i01}.
+
+\bibitem[\citeproctext]{ref-p-nlstools}
 Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
+\bibitem[\citeproctext]{ref-nlsMicrobio-manual}
 ---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
+\bibitem[\citeproctext]{ref-p-bbmle}
 Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
+\bibitem[\citeproctext]{ref-gslnls23}
 Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
+\bibitem[\citeproctext]{ref-nlshelper-manual}
 Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
+\bibitem[\citeproctext]{ref-minpacklm12}
 Elzhov, Timur V., Katharine M.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
+\bibitem[\citeproctext]{ref-GSL-manual}
 Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
 
-\leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
+\bibitem[\citeproctext]{ref-Bioconductor}
 Gentleman, Robert C., Vincent J.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
+\bibitem[\citeproctext]{ref-Golub1973}
 Golub, G.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
+\bibitem[\citeproctext]{ref-nls2-22}
 Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
+\bibitem[\citeproctext]{ref-Hartley1961}
 Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
+\bibitem[\citeproctext]{ref-Huet2004}
 Huet, S., A.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
+\bibitem[\citeproctext]{ref-nlsr2023manual}
 John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
+\bibitem[\citeproctext]{ref-tvrdik2007}
 Josef Tvrdík and Ivan Křivý and Ladislav Mišík.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
+\bibitem[\citeproctext]{ref-Levenberg1944}
 Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
+\bibitem[\citeproctext]{ref-Marquardt1963}
 Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
 
-\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
+\bibitem[\citeproctext]{ref-MiguezNLRAA2021}
 Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
+\bibitem[\citeproctext]{ref-jn77ima}
 Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
 
-\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
+\bibitem[\citeproctext]{ref-cnm79}
 ---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
-\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
+\bibitem[\citeproctext]{ref-NashWires22}
 ---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
+\bibitem[\citeproctext]{ref-PkgFromRbase22}
 Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
+\bibitem[\citeproctext]{ref-p-optimx}
 Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
+\bibitem[\citeproctext]{ref-nash1996nonlinear}
 Nash, John C., and Paul Velleman.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
+\bibitem[\citeproctext]{ref-jnmws87}
+Nash, John C., and Mary Walker-Smith. 1987. \emph{Nonlinear Parameter Estimation: An Integrated System in BASIC}. Book. New York: Marcel Dekker.
+
+\bibitem[\citeproctext]{ref-NISTnls-manual}
 National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
+\bibitem[\citeproctext]{ref-OlearyRust13}
 O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
+\bibitem[\citeproctext]{ref-nlsmultstart}
 Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
+\bibitem[\citeproctext]{ref-p-nlme}
 Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
+\bibitem[\citeproctext]{ref-m-nlsmsn}
 Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
+\bibitem[\citeproctext]{ref-Ratkowsky1983}
 Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
-\leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
+\bibitem[\citeproctext]{ref-m-nlstac}
 Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-nlsic-manual}{}}%
+\bibitem[\citeproctext]{ref-nlsic-manual}
 Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-Torvisco2018}{}}%
+\bibitem[\citeproctext]{ref-Torvisco2018}
 Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez.
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
+\bibitem[\citeproctext]{ref-R-crsnls-manual}
 Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
+\bibitem[\citeproctext]{ref-HWtestthat11}
 Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
 
-\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
+\bibitem[\citeproctext]{ref-HWdevtools21}
 Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.

### 2023-10-18 JNABRev20231018/ComparingNLStools.R

@@ -1,0 +1,408 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit ComparingNLStools.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----ex01, echo=TRUE----------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE-------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
+frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
+frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+#
+# Starting parameter sets
+stu1<-c(b1=1, b2=1, b3=1)
+sts1<-c(c1=1, c2=1, c3=1)
+stt1<-c(Asym=1, xmid=1, scal=1)
+
+
+## ----ex02fn, echo=TRUE--------------------------------------------------------
+# Logistic3U
+hobbs.res  <-  function(x){ # unscaled Hobbs weeds problem -- residual
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y  <-  c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt  <-  1:12
+  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+  # Note: this form of the residual, from Laplace (1788) in the form
+  # of "fitted - observed" has been used in our software for half
+  # a century, but sometimes offends users of the more traditional
+  # "observed - fitted" form
+}
+
+hobbs.jac  <-  function(x) { # unscaled Hobbs weeds problem -- Jacobian
+  jj  <-  matrix(0.0, 12, 3)
+  tt  <-  1:12
[…]

### 2023-10-22 JNABRev20231018/JNAB2review-response.txt
various updates for JNAB revision

@@
 >> (@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
->> of the **Logistic3T** variant of the model. 
+>> of the **Logistic3T** variant of the model, as well as discussions in @Ross90, 
+>> @Seber89Wild, @bateswatts, and @Gallant87.
 >> 
@@
 
->> To editor: Referee may be revealing their identity!
-
 Also on page 4 I don’t understand the comment in the hobbs.res function that This variant
@@
 >> We also noted that the model is UNscaled, and edited the comment.
->> An explanatory comment has been added
+>> An explanatory comment has been added.
 >> 

### 2023-10-22 JNABRev20231018/ComparingNLStools.Rmd
various updates for JNAB revision

@@
       toc: no
+  # pdf_document
 bibliography: ./ComparingNLStools.bib
@@
 
-`nlsic` (@nlsic-manual) -- This solves nonlinear least squares problems with optional
+`nlsic` (@nlsic22) -- This solves nonlinear least squares problems with optional
 equality and/or inequality constraints. It is clearly **not** about modeling, and
@@
 (@nash1996nonlinear) mentioned later accords with this viewpoint, as does use
-of the **Logistic3T** variant of the model.
+of the **Logistic3T** variant of the model, as well as discussions in @Ross90, 
+@Seber89Wild, @bateswatts, and @Gallant87.
 

### 2023-10-22 JNABRev20231018/ComparingNLStools.tex
various updates for JNAB revision

@@
 
-\texttt{NISTnls} (National Institutes for Standards and R port by Douglas Bates (2012)) -- This package provides R code and data for a set of
+\texttt{NISTnls} ((ed.) (2012)) -- This package provides R code and data for a set of
 (numerically ill-conditioned) nonlinear least squares problems from the U.S. National
@@
 \texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
-Torvisco, Fernandez, and Sanchez (2018) for fitting separable nonlinear regression curves. The
+Torvisco, Rodrıguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
 special class of problem for which this package is intended is an important and
@@
 (John C. Nash and Velleman (1996)) mentioned later accords with this viewpoint, as does use
-of the \textbf{Logistic3T} variant of the model.
+of the \textbf{Logistic3T} variant of the model, as well as discussions in Ross (1990),
+Seber and Wild (1989), D. M. Bates and Watts (1988), and Gallant (1987).
 
@@
 
+\bibitem[\citeproctext]{ref-NISTnls-manual}
+(ed.), Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.
+
 \bibitem[\citeproctext]{ref-minpacklm12}
@@
 Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
+
+\bibitem[\citeproctext]{ref-Gallant87}
+Gallant, A. Ronald. 1987. \emph{Nonlinear Statistical Models}. Wiley.
 
@@
 
-\bibitem[\citeproctext]{ref-NISTnls-manual}
-National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.
-2012.
-\emph{NISTnls: Nonlinear Least Squares Examples from NIST}.
-\url{https://CRAN.R-project.org/package=NISTnls}.
-
 \bibitem[\citeproctext]{ref-OlearyRust13}
@@
 
-\bibitem[\citeproctext]{ref-nlsic-manual}
+\bibitem[\citeproctext]{ref-Ross90}
+Ross, Gavin J. S. 1990. \emph{Nonlinear Estimation.} New York etc.: Springer-Verlag.
+
+\bibitem[\citeproctext]{ref-Seber89Wild}
+Seber, G. A. F., and C. J. Wild. 1989. \emph{{Nonlinear regression}}. New York: Wiley.
+
+\bibitem[\citeproctext]{ref-nlsic22}
 Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
@@
 \bibitem[\citeproctext]{ref-Torvisco2018}
-Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias Fernandez, and Javier Cabello Sanchez.
+Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias, and Javier Cabello Sanchez.
 2018.
@@
 \bibitem[\citeproctext]{ref-R-crsnls-manual}
-Tvrdík, Josef. 2016. \emph{Crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}. \url{https://CRAN.R-project.org/package=crsnls}.
+Tvrdík, Josef. 2016. \emph{{crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}}. \url{https://CRAN.R-project.org/package=crsnls}.
 

### 2023-10-22 JNABsub20230518/ComparingNLStools.tex
various updates for JNAB revision

--- 2023-05-18 JNABsub20230518/ComparingNLStools.tex
@@
 
-\hypertarget{solution-attempts-with-nlsr}{%
-\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}
+
+\begin{verbatim}
+#> unlx1:
+\end{verbatim}
 
@@
 
-\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
+\bibitem[\citeproctext]{ref-NISTnls-manual}
 National Institutes for Standards, original from, and Technology http://www.itl.nist.gov/div898/strd/nls/nls\_main.shtml R port by Douglas Bates.

### 2023-10-23 JNABRev20231018/ComparingNLStools.Rmd
fixup bibliography

@@
 
-`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN) 
+`nlsmsn` (@m-nlsmsn) -- Fit univariate non-linear scale mixture of skew-normal (NL-SMSN) 
 regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem 
@@
 attempts to succinctly summarize computing environments.
+
+# References

### 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
minor updates

@@
 
-For R **developers** we invite and encourage discussion of the design choices, since these
-have downstream implications for ease of use, adaptation to new features and 
+For R **developers**, we invite and encourage discussion of the design choices, since these
+have downstream implications for ease of use, adaptation to new features, and 
 efficiency of ongoing maintenance.
@@
 
-The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth curve modeling 
+The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth-curve modeling 
 task that seems straightforward but is quite nasty. Its very succinct statement 
@@
 with users whose interests and understanding are very far from those of statistical
-modelers. Thus we seek methods and codes that obtain reasonable answers under
+modelers. Thus, we seek methods and codes that obtain reasonable answers under
 highly unfavourable conditions. Nevertheless, a proper approach to 
@@
 # model formulas
-frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
-frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
-frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
 #
 # Starting parameter sets
-stu1<-c(b1=1, b2=1, b3=1)
-sts1<-c(c1=1, c2=1, c3=1)
-stt1<-c(Asym=1, xmid=1, scal=1)
+stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
+sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
+stt1 <- c(Asym = 1, xmid = 1, scal = 1)
 ```
@@
   tt <- 1:12
-  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
   # Note: this form of the residual, from Laplace (1788) in the form
@@
   tt <- 1:12
-  yy  <-  exp(-x[3]*tt)
-  zz  <-  1.0/(1+x[2]*yy)
-  jj[tt,1]   <-   zz
-  jj[tt,2]   <-   -x[1]*zz*zz*yy
-  jj[tt,3]   <-   x[1]*zz*zz*yy*x[2]*tt
+  yy <- exp(-x[3] * tt)
+  zz <- 1.0 / (1 + x[2] * yy)
+  jj[tt, 1] <- zz
+  jj[tt, 2] <- -x[1] * zz * zz * yy
+  jj[tt, 3] <- x[1] * zz * zz * yy * x[2] * tt
   attr(jj, "gradient") <- jj
@@
 ```{r ex05, echo=TRUE}  
-stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
-badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
+stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
 print(badstart)
@@
 "convergence tests" for the iterative solvers. In the manual page for `nls()` 
-in R 4.0.0 there was the warning:
+in R 4.0.0, there was the warning:
 
@@
 use for example `result <- snlm1` as the returned object from `nlsLM()` for the 
-Logistic3S problem. . The `data` return element 
+Logistic3S problem. The `data` return element 
 is an R symbol. To actually access the data from this element, we need to use 
@@
 ```
-eval(parse(text=result$data))
-```
-However, if the call is made with `model=TRUE`, then there is a returned element
+eval(parse(text = result$data))
+```
+However, if the call is made with `model = TRUE`, then there is a returned element
 `model` which contains the data, and we can list its contents using:
@@
 data and programs was at a ridiculous premium in the small computers of the era.
-Thus the code in `nlsr` is likely of value for workers to copy and modify
+Thus, the code in `nlsr` is likely of value for workers to copy and modify
 for customized tools.
@@
   with the formula `y ~ a * (x ˆ b)` causes `nlsr::nlxb` to fail because the partial
-  derivative w.r.t. `b` is `a * (x^b * log(x))`. If there is data for which `x = 0`,
+  derivative w.r.t. `b` is `a * (x ^ b * log(x))`. If there is data for which `x = 0`,
   the derivative is undefined, but the model can be computed. In such cases, 
@@
 # Start MUST be feasible i.e. on or within bounds
-anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
-             upper=c(2,6,3), algorithm='port')
+anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
+             upper = c(2,6,3), algorithm = 'port')
 pnls(anlshob1b) #  check the answer (short form)
 # nlsLM seems NOT to work with bounds in this example
-anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pnls(anlsLM1b)
 # also no warning if starting out of bounds, but gets a good answer!!
-st4<-c(c1=4, c2=4, c3=4)
-anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+st4 <- c(c1 = 4, c2 = 4, c3 = 4)
+anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pnls(anlsLMob)
 # Try nlsr::nlxb()
-anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
+anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
 pshort(anlx1b)
@@
 # Hobbsmaskx.R -- masks with formula specification of the problem
-require(nlsr); require(minpack.lm); traceval<-FALSE
-stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
-sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
+require(nlsr); require(minpack.lm); traceval <- FALSE
+stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
+sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
 # fix first parameter
-anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, 3), trace=traceval))
+anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+			upper = c(200, 60, 3), trace=traceval))
 print(anxbmsk1)
-anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-			upper=c(200, 60, 3), trace=traceval))
+anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+			upper=c(200, 60, 3), trace = traceval))
 pnls(anlM1)
-anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-		upper=c(200, 60, 3),  algorithm="port", trace=traceval))
+anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+		upper = c(200, 60, 3),  algorithm = "port", trace = traceval))
 pnls(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
-anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                  upper=c(2,6,30))
+anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2, 0, 0),
+                  upper = c(2, 6, 30))
 print(anlxmsks1)
-anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
-             upper=c(2,6,30), algorithm='port')
+anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, 
+                 lower = c(2, 0, 0), upper = c(2, 6, 30), algorithm = 'port')
 pnls(anlshmsk1)
-anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                 upper=c(2,6,30))
+anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                 upper = c(2,6,30))
 pnls(anlsLMmsks1)
@@
 # Test with all parameters masked
-anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+anlxmskall <- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
 print(anlxmskall)
@@
 
-nlsModel <- function(form, data, start, wts, upper=NULL, scaleOffset = 0, nDcentral = FALSE)
+nlsModel <- function(form, data, start, wts, upper = NULL, scaleOffset = 0,
+                     nDcentral = FALSE)
 {
@@
     parLength <- 0L
-    for(i in names(ind) ) {
+    for(i in names(ind)) {
         temp <- start[[i]]
@@
 ```{r nlswtx, echo=TRUE}
-wts <- 0.5^tt # simple weights
-frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
-Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
-attr(rnowt, "gradient")<-NULL; rnowt
-usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
-attr(rusewt,"gradient")<-NULL; rusewt
+wts <- 0.5 ^ tt # simple weights
+frmlogis <- weed ~ Asym / (1 + exp((xmid - tt)/scal))
+Asym <- 1; xmid <- 1; scal <- 1
+nowt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+rnowt <- nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+attr(rnowt, "gradient") <- NULL; rnowt
+usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights = wts)
+rusewt <- usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+attr(rusewt, "gradient") <- NULL; rusewt
 ## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
-nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
-rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
-nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
-rn<-nmod$resid()
+nmod0 <- nlsModel(frmlogis, data = weeddf, start = c(Asym = 1, xmid = 1, scal = 1), wts = wts)
+rn0 <- nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+attr(rn0, "gradient") <- NULL; rn0 # weighted residuals at starting coefficients
+nmod <- nlsModel(frmlogis, data = weeddf, start = coef(usewt), wts = wts)
+rn <- nmod$resid()
 attr(rn,"gradient")<-NULL; rn # same as rusewt
@@
 and the conventional R package testing tools).
-In our work, one of us has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
+In our work, one of us (AB) has developed a working prototype package at *https://github.com/ArkaB-DS/nlsCompare* .
 A primary design objective of this is to allow the summarization of multiple tests in

### 2023-10-25 JNABRev20231018/JNAB2review-response.txt
Adjustments 20231025

@@
 >> of the Laplace residual in other works, but present usage or commentary is a red herring
->> that is irrelevant to today's software and practice. Indeed, the main advance that package
->> 'nlsr' represents is to automatically provide analytic derivatives that caused so much work
->> and introduced so many errors in the 1970s and 1980s.
+>> that is irrelevant to today's software and practice except for a note in the hobbs.res
+>> function definition. Indeed, the main advance that package 'nlsr' represents is to 
+>> automatically provide analytic derivatives that caused so much work and introduced 
+>> so many errors in the 1970s and 1980s.
+
+>> We have also removed the sentences (just after the definition of the Logistic models):
+>> "The residuals can be written either as r1=y−model(p) or r=model(p)−y since their sum of 
+>> squares has an identical value. The second form avoids a potential sign error if we need 
+>> to evaluate derivatives."
+
+>> Other notes:
+>>
+>> In re-reading our paper, made some minor wording changes to avoid what we believe may be
+>> confusing. In particular, we reworded the explanation of the `y` vector for gsl_nls() 
+>> under "Functional specification of problems" to accord better with the man page.
+
+>> We noticed the citation to the "illustrative example" lacked a "p." to indicate the
+>> page number. We added "csl: ./apa.csl" to the YAML header and downloaded and have
+>> included this file. 
+>> We have also changed all "NOT" occurrences to simply "not", and removed some boldface
+>> where it is unnecessary.

### 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
Adjustments 20231025

--- 2023-10-23 JNABRev20231018/ComparingNLStools.Rmd
@@
 title: "A Comparison of R Tools for Nonlinear Least Squares Modeling"
-date: "2023-5-18"
+date:  \`r format(Sys.Date(), "%B %d, %Y")`\
 abstract: >
@@
 bibliography: ./ComparingNLStools.bib
+csl: ./apa.csl
 ---
@@
 
-## The `nls()` function
+# The `nls()` function
 
@@
 
-## Scope of our comparison
+# Scope of our comparison
 
@@
 
-### Principal messages
+## Principal messages
 
@@
 
-## Some other CRAN packages for nonlinear modeling
+# Some other CRAN packages for nonlinear modeling
 
@@
 `nlsic` (@nlsic22) -- This solves nonlinear least squares problems with optional
-equality and/or inequality constraints. It is clearly **not** about modeling, and
+equality and/or inequality constraints. It is clearly not about modeling, and
 the input and output are quite different from class `nls` methods. However, there do
@@
 
-## An illustrative example
+# An illustrative example
 
@@
 <!-- 231018: attempt to address comments in 2report1 on modeling vs least squares -->
-We thank one of the referees for expressing the viewpoint that a nonlinear regression
-approach would use background knowledge and graphs to obtain an understanding of
+While model estimation is increasingly automated, nonlinear regression
+should use background knowledge and graphs to obtain an understanding of
 the general magnitude of the parameters. Indeed, the "visual fitting" approach
@@
 @Seber89Wild, @bateswatts, and @Gallant87.
-
 Our emphasis on the software resilience to starting parameters that are, in the
@@
 
-### Problem setup
+## Problem setup
 
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are NOT equivalent and
+values for parameters. These starting points are not equivalent and
 are deliberately crude choices. Workers performing many calculations 
@@
 do still want or need to provide problems as explicit functions, particularly
-for problems that are NOT regressions. For example, the Rosenbrock banana-valley
+for problems that are not regressions. For example, the Rosenbrock banana-valley
 test problem can be provided this way, where the two "residuals" are different
@@
 
-## Estimation of models specified as formulas
+# Estimation of models specified as formulas
 
@@
 
-### Solution attempts with nlsr
+## Solution attempts with nlsr
 
@@
 shown by its singular
-values. Note that these are **displayed** by package `nlsr` in a single column in
-the output to provide a compact layout, but the values do **NOT** correspond to 
+values. Note that these are displayed by package `nlsr` in a single column in
+the output to provide a compact layout, but the values do not correspond to 
 the individual parameters in 
@@
 
-### Solution attempts with minpack.lm
+## Solution attempts with minpack.lm
 
@@
 
-### Solution attempts with gslnls
+## Solution attempts with gslnls
 
@@
 
-### Comparison notes for formula-setup solutions
+## Comparison notes for formula-setup solutions
 
@@
 `minpack.lm::nlsLM()` and `nls()`
-use `summary()`, which does NOT display the sum of squares, while `print()` 
+use `summary()`, which does not display the sum of squares, while `print()` 
 gives the sum of squares, but not the standard error of the residuals.
@@
 
-## Functional specification of problems
+# Functional specification of problems
 
@@
 a function to define the residual. Note that `gsl_nls()` requires
-a vector `y` that is the expected value of what we have called the
-residual, but in this case, is actually the model. `gsl_nls` uses 
+a vector `y` that is the length of the vector returned by the
+function supplied e.g., `hobbs.res()`. `gsl_nls` uses 
 a numerical approximation for the Jacobian if the argument `jac` 
@@
 
-## Design goals, termination tests, and output objects
+# Design goals, termination tests, and output objects
 
@@
 
->  *Do not use `nls` on artificial "zero-residual" data.*
+>  *Do not use `nls` on artificial "zero-residual" data*
 
@@
 approached with nonlinear least squares. Fortunately, a small adjustment
-to the "termination test" for the **program**, rather than for the
-"convergence" of the underlying **algorithm**, fixes the defect. The test
+to the "termination test" for the program, rather than for the
+"convergence" of the underlying algorithm, fixes the defect. The test
 is the Relative Offset Convergence
@@
 
-### Returned results of nls() and other tools
+## Returned results of nls() and other tools
 
@@
 
-### When to compute ancillary information
+## When to compute ancillary information
 
@@
 
-## Jacobian calculation
+# Jacobian calculation
 
@@
 Using the "gradient" attribute of the output of the Jacobian function to hold
-the Jacobian matrix lets us embed this in the **residual** function as well,
+the Jacobian matrix lets us embed this in the residual function as well,
 so that the call to `nlsr::nlfb()` can be made with the same name used for 
@@
 in the R distribution source code. This is used to provide Jacobian information in 
-the `nlsModel()` and `nlsModel.plinear()` functions, which are **not** exported for 
+the `nlsModel()` and `nlsModel.plinear()` functions, which are not exported for 
 general use. `gsl_nls()` also appears to use `numericDeriv()`.
@@
 
-### Jacobian code in selfStart models
+## Jacobian code in selfStart models
 
@@
 command, making them difficult to access by users who may wish to create
-their own selfStart model via copy and edit.
+their own `selfStart` model via copy and edit.
 Interactive tools, such as "visual fitting" (@nash1996nonlinear) might be
@@
 
-## Bounds constraints on parameters
+# Bounds constraints on parameters
 
@@
 
-### Philosophical considerations
+## Philosophical considerations
 
@@
 
-### Fixed parameters (masks)
+## Fixed parameters (masks)
 
@@
 `nlsr` has an output format that indicates the constraint status of the parameter
-estimates. For `nlsr`, we have **chosen** to suppress the calculation of 
+estimates. For `nlsr`, we have chosen to suppress the calculation of 
 approximate standard errors in the parameters when constraints are active because
@@
 
-## Stabilization of Gauss-Newton computations
+# Stabilization of Gauss-Newton computations
 
@@
 
-### Programming language
+## Programming language
 
@@
 
-## Data sources for problems
+# Data sources for problems
 
@@
 
-### Feature: Subsetting
+## Feature: Subsetting
 
@@
 
-### Feature: na.action (missing value treatment)
+## Feature: na.action (missing value treatment)
 
@@
 
-### Weights in returned functions from nls()
+## Weights in returned functions from nls()
 
@@
 nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does not take coefficients.
 attr(rnowt, "gradient")<-NULL; rnowt
 usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+rusewt<-usewt$m$resid() # WEIGHTED. Does not take coefficients.
 attr(rusewt,"gradient")<-NULL; rusewt
@@
 
-## Minor issues with nonlinear least-squares tools
-
-### Interim output from the "port" algorithm
+# Minor issues with nonlinear least-squares tools
+
+## Interim output from the "port" algorithm
 
@@
 
-### Failure to return the best result achieved
+## Failure to return the best result achieved
 
@@
 
-## Estimating models that are partially linear
+# Estimating models that are partially linear
 
@@
 
-## Models with indexed parameters
+# Models with indexed parameters
 
@@
 
-## Tests and use-case examples
+# Tests and use-case examples
 
@@
 
-## Documentation and resources
+# Documentation and resources
 
@@
 
-## Future of nonlinear model estimation in R
+# Future of nonlinear model estimation in R
 
@@
 vignette. Colin Gillespie (package `benchmarkme`) has been helpful in guiding our 
-attempts to succinctly summarize computing environments.
+attempts to succinctly summarize computing environments. We thank one of the referees 
+for pointing out that it is important to use background knowledge and graphs to
+guide modeling.
 

### 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
offends --> concerns

--- 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
@@
   # of "fitted - observed" has been used in our software for half
-  # a century, but sometimes offends users of the more traditional
+  # a century, but sometimes concerns users of the more traditional
   # "observed - fitted" form

### 2023-11-13 JNABresubmission20231101/JNAB2review-response.txt
resub 231101

--- 2023-10-25 JNABRev20231018/JNAB2review-response.txt
@@
 
+>> In the example output, we have added a printout of the calling sequence to document 
+>> the results better.
+
 Is it really intended that anyone using the function-based specification of the model will encode
@@
 >> to evaluate derivatives."
+>>    but added the definition of the residual as r = y - model(p).
 
@@
 >> We noticed the citation to the "illustrative example" lacked a "p." to indicate the
->> page number. We added "csl: ./apa.csl" to the YAML header and downloaded and have
->> included this file. 
+>> page number. We added "csl: ./apa.csl" to the YAML header and downloaded it and have
+>> included this file. The output of rendering the Rmd file asks that custom CSL not be
+>> used. We are not sure how to resolve this?
+
 >> We have also changed all "NOT" occurrences to simply "not", and removed some boldface
 >> where it is unnecessary.
+
+>> A minor clarification of the iteration limit termination tests has been added, namely,
+>>  "examination of the code for the different algorithms" since the different `nls()`
+>> algorithms may use different tests.

### 2023-11-13 JNABresubmission20231101/ComparingNLStools.Rmd
resub 231101

--- 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
@@
 output: 
-  rjtools::rjournal_web_article:
+  rjtools::rjournal_article:
     self_contained: yes
     toc: no
-  # pdf_document
 bibliography: ./ComparingNLStools.bib
-csl: ./apa.csl
 ---
@@
 optimizers that can be applied to minimize a nonlinear function which could be
-a nonlinear sum of 
-squares. not generally recommended if nonlinear 
+a nonlinear sum of squares. Not generally recommended if nonlinear 
 least squares programs can be easily used, but provides a check and alternative
@@
 To allow for simpler discussion, let us say that the parameters form a (named) vector
-$p$ and the model function is called $model(p)$. The residuals can be written
-either as $r_1 = y - model(p)$ or $r = model(p) - y$
-since their sum of squares has an identical value. The second form
-avoids a potential sign error if we need to evaluate derivatives.
+$p$ and the model function is called $model(p)$ with $r = y - model(p)$.
 
@@
 unlx1<-try(nlxb(formula=frmu, start=stu1, data=weeddf))
-cat("unlx1:")
-print(unlx1) 
+if (! inherits(unlx1, "try-error")) {cat("unlx1:"); print(unlx1) }
 snlx1<-try(nlxb(formula=frms, start=sts1, data=weeddf))
-pshort(snlx1) # a short-form output
+if (! inherits(snlx1, "try-error"))pshort(snlx1) # a short-form output
 tnlx1<-try(nlxb(formula=frmt, start=stt1, data=weeddf))
-pshort(tnlx1) # alternatively print(tnlx1)
+if (! inherits(tnlx1, "try-error"))pshort(tnlx1) # alternatively print(tnlx1)
 ```
@@
 unlm1<-try(nlsLM(formula=frmu, start=stu1, data=weeddf))
-pnls(unlm1)  # Short form of output
+if (! inherits(unlm1, "try-error"))pnls(unlm1)  # Short form of output
 snlm1<-try(nlsLM(formula=frms, start=sts1, data=weeddf))
-pnls(snlm1)
+if (! inherits(snlm1, "try-error"))pnls(snlm1)
 tnlm1<-try(nlsLM(formula=frmt, start=stt1, data=weeddf))
-pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
+if (! inherits(tnlm1, "try-error"))pnls(tnlm1) # short form to give sum of squares, else use summary(tnlm1)
 ```
@@
 ugslnls1<-try(gsl_nls(fn = frmu, data = weeddf,  start = stu1))
-pnls(ugslnls1) # to get sum of squares
+if (! inherits(ugslnls1, "try-error")) pnls(ugslnls1) # to get sum of squares
 sgslnls1<-try(gsl_nls(fn = frms, data = weeddf,  start = sts1))
-pnls(sgslnls1) # Use summary() to get display
+if (! inherits(sgslnls1, "try-error")) pnls(sgslnls1) # Use summary() to get display
 tgslnls1<-try(gsl_nls(fn = frmt, data = weeddf,  start = stt1))
-pnls(tgslnls1) 
+if (! inherits(tgslnls1, "try-error")) pnls(tgslnls1) 
 ```
@@
 stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
-badstart <- nlxb(formula = frmt, start = stspecial, data = weeddf)
-print(badstart)
+badstart <- try(nlxb(formula = frmt, start = stspecial, data = weeddf))
+if (! inherits(badstart, "try-error")) print(badstart)
 ```
@@
 a vector `y` that is the length of the vector returned by the
-function supplied e.g., `hobbs.res()`. `gsl_nls` uses 
+function supplied, e.g., `hobbs.res()`. `gsl_nls` uses 
 a numerical approximation for the Jacobian if the argument `jac` 
@@
 ```{r exhobbsfn, eval=TRUE}
-hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
-pshort(hobnlfb) # use print(hobnlfb) for more detail
-hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
-pnlslm(hobnlm)  
-hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
-pnls(hobgsln)
-hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
-pnls(hobgsl) # using analytic Jacobian
+cat("hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)\n")
+hobnlfb<-try(nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac))
+if (! inherits(hobnlfb, "try-error")) pshort(hobnlfb) # use print(hobnlfb) for more detail
+cat("hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)\n")
+hobnlm<-try(nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac))
+if (! inherits(hobnlm, "try-error")) pnlslm(hobnlm)
+cat("hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))\n")
+hobgsln<-try(gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12)))
+if (! inherits(hobgsln, "try-error")) pnls(hobgsln)
+cat("hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)\n")
+hobgsl<-try(gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac))
+if (! inherits(hobgsl, "try-error")) pnls(hobgsl) # using analytic Jacobian
 ```
@@
 `nls()` stops after `maxiter` "iterations". The meaning of "iteration" may require an 
-examination of the code. We terminate execution when the number of residual or 
-Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the 
-default `maxiter = 50` of `nls()` to avoid stopping early.
+examination of the code for the different algorithms. `nlsr` terminates execution 
+when the number of residual or Jacobian evaluations exceed set limits. Generally, 
+we prefer larger limits than the default `maxiter = 50` of `nls()` to avoid stopping 
+early, though this may result in some unnecessary computations.
 
@@
 # Start MUST be feasible i.e. on or within bounds
-anlshob1b <- nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
-             upper = c(2,6,3), algorithm = 'port')
-pnls(anlshob1b) #  check the answer (short form)
+anlshob1b <- try(nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
+             upper = c(2,6,3), algorithm = 'port'))
+if (! inherits(anlshob1b, "try-error")) pnls(anlshob1b) #  check the answer (short form)
 # nlsLM seems NOT to work with bounds in this example
-anlsLM1b <- nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
-pnls(anlsLM1b)
+anlsLM1b <- try(nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlsLM1b, "try-error")) pnls(anlsLM1b)
 # also no warning if starting out of bounds, but gets a good answer!!
 st4 <- c(c1 = 4, c2 = 4, c3 = 4)
-anlsLMob <- nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
-pnls(anlsLMob)
+anlsLMob <- try(nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlsLMob, "try-error")) pnls(anlsLMob)
 # Try nlsr::nlxb()
-anlx1b <- nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3))
-pshort(anlx1b)
+anlx1b <- try(nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlx1b, "try-error")) pshort(anlx1b)
 ```
@@
 			upper = c(200, 60, 3), trace=traceval))
-print(anxbmsk1)
+if (! inherits(anxbmsk1, "try-error")) print(anxbmsk1)
 anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
 			upper=c(200, 60, 3), trace = traceval))
-pnls(anlM1)
+if (! inherits(anlM1, "try-error")) pnls(anlM1)
 anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
 		upper = c(200, 60, 3),  algorithm = "port", trace = traceval))
-pnls(anlsmsk1)
+if (! inherits(anlsmsk1, "try-error")) pnls(anlsmsk1)
 # Hobbs scaled problem with bounds, formula specification
-anlxmsks1 <- nlxb(frms, start = sts, data = weeddf, lower = c(2, 0, 0),
-                  upper = c(2, 6, 30))
-print(anlxmsks1)
-anlshmsk1 <- nls(frms, start = sts, trace = traceval, data = weeddf, 
-                 lower = c(2, 0, 0), upper = c(2, 6, 30), algorithm = 'port')
-pnls(anlshmsk1)
-anlsLMmsks1 <- nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
-                 upper = c(2,6,30))
-pnls(anlsLMmsks1)
+anlxmsks1 <- try(nlxb(frms, start = sts, data = weeddf, lower = c(2, 0, 0),
+                  upper = c(2, 6, 30)))
+if (! inherits(anlxmsks1, "try-error")) print(anlxmsks1)
+anlshmsk1 <- try(nls(frms, start = sts, trace = traceval, data = weeddf, 
+                 lower = c(2, 0, 0), upper = c(2, 6, 30), algorithm = 'port'))
+if (! inherits(anlshmsk1, "try-error")) pnls(anlshmsk1)
+anlsLMmsks1 <- try(nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                 upper = c(2,6,30)))
+if (! inherits(anlsLMmsks1, "try-error")) pnls(anlsLMmsks1)
 
@@
 anlxmskall <- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
-print(anlxmskall)
+if (! inherits(anlxmskall, "try-error")) print(anlxmskall)
 ```
@@
 objects in the default (`.GlobalEnv`) workspace. Moreover, users have to actively 
-suppress saving this workspace (`.RData`) on exit; otherwise any such file in 
+suppress saving this workspace (`.RData`) on exit, otherwise any such file in 
 the path will be loaded on startup. R users in our
@@
 Asym <- 1; xmid <- 1; scal <- 1
-nowt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
+nowt <- try(nls(weed ~ SSlogis(tt, Asym, xmid, scal))) # UNWEIGHTED
+if (! inherits(nowt, "try-error")) {
   rnowt <- nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
-attr(rnowt, "gradient") <- NULL; rnowt
-usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights = wts)
+  attr(rnowt, "gradient") <- NULL 
+} else rnowt <- NULL
+rnowt
+usewt <- try(nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights = wts))
+if (! inherits(usewt, "try-error")) {
   rusewt <- usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
-attr(rusewt, "gradient") <- NULL; rusewt
+  attr(rusewt, "gradient") <- NULL
+} else rusewt <- NULL
+rusewt
 ## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
@@
 
-```{r tetrarun}
+```{r tetrarun, echo=TRUE}
 time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
@@
 tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
-print(tryit)
+if (! inherits(tryit, "try-error")) print(tryit)
 ```
@@
 ## using a selfStart model - do not specify the starting parameters
-fm1 <- nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1)
-summary(fm1)
+fm1 <- try(nls(density ~ SSlogis(log(conc), Asym, xmid, scal), DNase1))
+if (! inherits(fm1, "try-error")) summary(fm1)
 
 ## using conditional linearity - leave out the Asym parameter
-fm2 <- nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
+fm2 <- try(nls(density ~ 1 / (1 + exp((xmid - log(conc)) / scal)),
                  data = DNase1, start = list(xmid = 0, scal = 1),
-                 algorithm = "plinear")
-summary(fm2)
+                 algorithm = "plinear"))
+if (! inherits(fm2, "try-error")) summary(fm2)
 
 ## without conditional linearity
-fm3 <- nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
+fm3 <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)),
                  data = DNase1,
-                 start = list(Asym = 3, xmid = 0, scal = 1))
-summary(fm3)
+                 start = list(Asym = 3, xmid = 0, scal = 1)))
+if (! inherits(fm3, "try-error")) summary(fm3)
 
@@
                  algorithm = "port"))
-summary(fm4)
+if (! inherits(fm4, "try-error")) summary(fm4)
 
@@
                  algorithm = "plinear", trace = TRUE))
-summary(fm2a)
+if (! inherits(fm2a, "try-error")) summary(fm2a)
 ```

### 2023-11-13 JNABresubmission20231101/ComparingNLStools.tex
resub 231101

--- 2023-10-22 JNABRev20231018/ComparingNLStools.tex
@@
 \title{A Comparison of R Tools for Nonlinear Least Squares Modeling}
+
 \author{by John C. Nash and Arkajyoti Bhattacharjee}
@@
 
-\subsection{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}
+\hypertarget{the-nls-function}{%
+\section{\texorpdfstring{The \texttt{nls()} function}{The nls() function}}\label{the-nls-function}}
 
@@
 
-\subsection{Scope of our comparison}\label{scope-of-our-comparison}
+\hypertarget{scope-of-our-comparison}{%
+\section{Scope of our comparison}\label{scope-of-our-comparison}}
 
@@
 and some packages that are available on the CRAN repository.
-Of these, we will pay particular attention to \textbf{nlsr} (John C Nash and Duncan Murdoch (2023)),
+Of these, we will pay particular attention to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
 \textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
@@
 
-\subsubsection{Principal messages}\label{principal-messages}
+\hypertarget{principal-messages}{%
+\subsection{Principal messages}\label{principal-messages}}
 
@@
 
-For R \textbf{developers} we invite and encourage discussion of the design choices, since these
-have downstream implications for ease of use, adaptation to new features and
+For R \textbf{developers}, we invite and encourage discussion of the design choices, since these
+have downstream implications for ease of use, adaptation to new features, and
 efficiency of ongoing maintenance.
 
-\subsection{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}
-
-\texttt{onls} (Andrej-Nikolai Spiess (2022)) -- The usual optimization in estimating nonlinear models
+\hypertarget{some-other-cran-packages-for-nonlinear-modeling}{%
+\section{Some other CRAN packages for nonlinear modeling}\label{some-other-cran-packages-for-nonlinear-modeling}}
+
+\texttt{onls} (Spiess, Andrej-Nikolai (2022)) -- The usual optimization in estimating nonlinear models
 is the vertical difference between the dependent variable and the functional
@@
 
-\texttt{NISTnls} ((ed.) (2012)) -- This package provides R code and data for a set of
+\texttt{NISTnls} (D. M. Bates (2012)) -- This package provides R code and data for a set of
 (numerically ill-conditioned) nonlinear least squares problems from the U.S. National
@@
 \texttt{nlsic} (Sokol (2022)) -- This solves nonlinear least squares problems with optional
-equality and/or inequality constraints. It is clearly \textbf{not} about modeling, and
+equality and/or inequality constraints. It is clearly not about modeling, and
 the input and output are quite different from class \texttt{nls} methods. However, there do
@@
 
-\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal(NL-SMSN)
+\texttt{nlsmsn} (Prates, Lachos, and Garay (2021)) -- Fit univariate non-linear scale mixture of skew-normal (NL-SMSN)
 regression, with details in Garay, Lachos, and Abanto-Valle (2011). The problem
@@
 \texttt{nlstac} (Rodriguez-Arias et al. (2020)) -- A set of functions implementing the algorithm described in
-Torvisco, Rodrıguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
+Torvisco, Rodriguez-Arias, and Sanchez (2018) for fitting separable nonlinear regression curves. The
 special class of problem for which this package is intended is an important and
@@
 optimizers that can be applied to minimize a nonlinear function which could be
-a nonlinear sum of
-squares. NOT generally recommended if nonlinear
+a nonlinear sum of squares. Not generally recommended if nonlinear
 least squares programs can be easily used, but provides a check and alternative
@@
 
-\subsection{An illustrative example}\label{an-illustrative-example}
-
-The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth curve modeling
+\hypertarget{an-illustrative-example}{%
+\section{An illustrative example}\label{an-illustrative-example}}
+
+The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth-curve modeling
 task that seems straightforward but is quite nasty. Its very succinct statement
@@
 
-We thank one of the referees for expressing the viewpoint that a nonlinear regression
-approach would use background knowledge and graphs to obtain an understanding of
+While model estimation is increasingly automated, nonlinear regression
+should use background knowledge and graphs to obtain an understanding of
 the general magnitude of the parameters. Indeed, the ``visual fitting'' approach
@@
 Seber and Wild (1989), D. M. Bates and Watts (1988), and Gallant (1987).
-
 Our emphasis on the software resilience to starting parameters that are, in the
@@
 with users whose interests and understanding are very far from those of statistical
-modelers. Thus we seek methods and codes that obtain reasonable answers under
+modelers. Thus, we seek methods and codes that obtain reasonable answers under
 highly unfavourable conditions. Nevertheless, a proper approach to
@@
 To allow for simpler discussion, let us say that the parameters form a (named) vector
-\(p\) and the model function is called \(model(p)\). The residuals can be written
-either as \(r_1 = y - model(p)\) or \(r = model(p) - y\)
-since their sum of squares has an identical value. The second form
-avoids a potential sign error if we need to evaluate derivatives.
+\(p\) and the model function is called \(model(p)\) with \(r = y - model(p)\).
 
@@
 \[ J' J \delta = - J' r\]
-though we can avoid some loss of accuracy by NOT forming the inner product matrix
+though we can avoid some loss of accuracy by not forming the inner product matrix
 \(J' J\) and solving the linear least squares matrix problem
@@
 
-\subsubsection{Problem setup}\label{problem-setup}
+\hypertarget{problem-setup}{%
+\subsection{Problem setup}\label{problem-setup}}
 
 Let us specify in R the three model formulas and set some starting
-values for parameters. These starting points are NOT equivalent and
+values for parameters. These starting points are not equivalent and
 are deliberately crude choices. Workers performing many calculations
@@
 # model formulas
-frmu <- weed ~ b1/(1+b2*exp(-b3*tt))
-frms <- weed ~ 100*c1/(1+10*c2*exp(-0.1*c3*tt))
-frmt <- weed ~ Asym /(1 + exp((xmid-tt)/scal))
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
 #
 # Starting parameter sets
-stu1<-c(b1=1, b2=1, b3=1)
-sts1<-c(c1=1, c2=1, c3=1)
-stt1<-c(Asym=1, xmid=1, scal=1)
+stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
+sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
+stt1 <- c(Asym = 1, xmid = 1, scal = 1)
 \end{verbatim}
@@
 do still want or need to provide problems as explicit functions, particularly
-for problems that are NOT regressions. For example, the Rosenbrock banana-valley
+for problems that are not regressions. For example, the Rosenbrock banana-valley
 test problem can be provided this way, where the two ``residuals'' are different
@@
   tt <- 1:12
-  res  <-  x[1]/(1+x[2]*exp(-x[3]*tt)) - y
+  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
   # Note: this form of the residual, from Laplace (1788) in the form
   # of "fitted - observed" has been used in our software for half
-  # a century, but sometimes offends users of the more traditional
+  # a century, but sometimes concerns users of the more traditional
   # "observed - fitted" form
@@
 
-\subsection{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}
+\hypertarget{estimation-of-models-specified-as-formulas}{%
+\section{Estimation of models specified as formulas}\label{estimation-of-models-specified-as-formulas}}
 
@@
 
-\subsubsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}
-
-\begin{verbatim}
-#> unlx1:
-\end{verbatim}
-
-\begin{verbatim}
-#> residual sumsquares =  2.5873  on  12 observations
+\hypertarget{solution-attempts-with-nlsr}{%
+\subsection{Solution attempts with nlsr}\label{solution-attempts-with-nlsr}}
+
+\begin{verbatim}
+#> unlx1:residual sumsquares =  2.5873  on  12 observations
 #>     after  19    Jacobian and  25 function evaluations
@@
 shown by its singular
-values. Note that these are \textbf{displayed} by package \texttt{nlsr} in a single column in
-the output to provide a compact layout, but the values do \textbf{NOT} correspond to
+values. Note that these are displayed by package \texttt{nlsr} in a single column in
+the output to provide a compact layout, but the values do not correspond to
 the individual parameters in
@@
 
-\subsubsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}
+\hypertarget{solution-attempts-with-minpack.lm}{%
+\subsection{Solution attempts with minpack.lm}\label{solution-attempts-with-minpack.lm}}
 
@@
 
-\subsubsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}
+\hypertarget{solution-attempts-with-gslnls}{%
+\subsection{Solution attempts with gslnls}\label{solution-attempts-with-gslnls}}
 
@@
 
-\subsubsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}
+\hypertarget{comparison-notes-for-formula-setup-solutions}{%
+\subsection{Comparison notes for formula-setup solutions}\label{comparison-notes-for-formula-setup-solutions}}
 
@@
 \texttt{minpack.lm::nlsLM()} and \texttt{nls()}
-use \texttt{summary()}, which does NOT display the sum of squares, while \texttt{print()}
+use \texttt{summary()}, which does not display the sum of squares, while \texttt{print()}
 gives the sum of squares, but not the standard error of the residuals.
@@
 \begin{verbatim}
-stspecial<- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
-badstart<-nlxb(formula=frmt, start=stspecial, data=weeddf)
-print(badstart)
+stspecial <- c(Asym = 35.532,  xmid = 43376,  scal = -2935.4)
+badstart <- try(nlxb(formula = frmt, start = stspecial, data = weeddf))
+if (! inherits(badstart, "try-error")) print(badstart)
 \end{verbatim}
@@
 
-\subsection{Functional specification of problems}\label{functional-specification-of-problems}
+\hypertarget{functional-specification-of-problems}{%
+\section{Functional specification of problems}\label{functional-specification-of-problems}}
 
@@
 a function to define the residual. Note that \texttt{gsl\_nls()} requires
-a vector \texttt{y} that is the expected value of what we have called the
-residual, but in this case, is actually the model. \texttt{gsl\_nls} uses
+a vector \texttt{y} that is the length of the vector returned by the
+function supplied, e.g., \texttt{hobbs.res()}. \texttt{gsl\_nls} uses
 a numerical approximation for the Jacobian if the argument \texttt{jac}
@@
 \begin{verbatim}
+#> hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
+\end{verbatim}
+
+\begin{verbatim}
 #> hobnlfb  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25 res/ 19 jac
@@
 \begin{verbatim}
+#> hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
+\end{verbatim}
+
+\begin{verbatim}
 #> hobnlm  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 17  itns
@@
 \begin{verbatim}
+#> hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
+\end{verbatim}
+
+\begin{verbatim}
 #> hobgsln  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
@@
 \begin{verbatim}
+#> hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)
+\end{verbatim}
+
+\begin{verbatim}
 #> hobgsl  -- ss= 2.5873 :  b1 = 196.19  b2 = 49.092  b3 = 0.31357; 25  itns
@@
 
-\subsection{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}
+\hypertarget{design-goals-termination-tests-and-output-objects}{%
+\section{Design goals, termination tests, and output objects}\label{design-goals-termination-tests-and-output-objects}}
 
@@
 ``convergence tests'' for the iterative solvers. In the manual page for \texttt{nls()}
-in R 4.0.0 there was the warning:
+in R 4.0.0, there was the warning:
 
 \begin{quote}
-\emph{Do not use \texttt{nls} on artificial ``zero-residual'' data.}
+\emph{Do not use \texttt{nls} on artificial ``zero-residual'' data}
 \end{quote}
@@
 approached with nonlinear least squares. Fortunately, a small adjustment
-to the ``termination test'' for the \textbf{program}, rather than for the
-``convergence'' of the underlying \textbf{algorithm}, fixes the defect. The test
+to the ``termination test'' for the program, rather than for the
+``convergence'' of the underlying algorithm, fixes the defect. The test
 is the Relative Offset Convergence
-Criterion (see Bates, Douglas M. and Watts, Donald G. (1981)). This scales an estimated reduction in the loss
+Criterion (see D. M. Bates and Watts (1981)). This scales an estimated reduction in the loss
 function by its current value. If the loss function is very small, we are close
@@
 \texttt{nls()} stops after \texttt{maxiter} ``iterations''. The meaning of ``iteration'' may require an
-examination of the code. We terminate execution when the number of residual or
-Jacobian evaluations exceed set limits. Generally, we prefer larger limits than the
-default \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping early.
-
-\subsubsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}
+examination of the code for the different algorithms. \texttt{nlsr} terminates execution
+when the number of residual or Jacobian evaluations exceed set limits. Generally,
+we prefer larger limits than the default \texttt{maxiter\ =\ 50} of \texttt{nls()} to avoid stopping
+early, though this may result in some unnecessary computations.
+
+\hypertarget{returned-results-of-nls-and-other-tools}{%
+\subsection{Returned results of nls() and other tools}\label{returned-results-of-nls-and-other-tools}}
 
@@
 use for example \texttt{result\ \textless{}-\ snlm1} as the returned object from \texttt{nlsLM()} for the
-Logistic3S problem. . The \texttt{data} return element
+Logistic3S problem. The \texttt{data} return element
 is an R symbol. To actually access the data from this element, we need to use
@@
 \begin{verbatim}
-eval(parse(text=result$data))
-\end{verbatim}
-
-However, if the call is made with \texttt{model=TRUE}, then there is a returned element
+eval(parse(text = result$data))
+\end{verbatim}
+
+However, if the call is made with \texttt{model\ =\ TRUE}, then there is a returned element
 \texttt{model} which contains the data, and we can list its contents using:
@@
 
-\subsubsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}
+\hypertarget{when-to-compute-ancillary-information}{%
+\subsection{When to compute ancillary information}\label{when-to-compute-ancillary-information}}
 
@@
 data and programs was at a ridiculous premium in the small computers of the era.
-Thus the code in \texttt{nlsr} is likely of value for workers to copy and modify
+Thus, the code in \texttt{nlsr} is likely of value for workers to copy and modify
 for customized tools.
 
-\subsection{Jacobian calculation}\label{jacobian-calculation}
+\hypertarget{jacobian-calculation}{%
+\section{Jacobian calculation}\label{jacobian-calculation}}
 
@@
 Using the ``gradient'' attribute of the output of the Jacobian function to hold
-the Jacobian matrix lets us embed this in the \textbf{residual} function as well,
+the Jacobian matrix lets us embed this in the residual function as well,
 so that the call to \texttt{nlsr::nlfb()} can be made with the same name used for
@@
 in the R distribution source code. This is used to provide Jacobian information in
-the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are \textbf{not} exported for
+the \texttt{nlsModel()} and \texttt{nlsModel.plinear()} functions, which are not exported for
 general use. \texttt{gsl\_nls()} also appears to use \texttt{numericDeriv()}.
@@
   with the formula \texttt{y\ \textasciitilde{}\ a\ *\ (x\ ˆ\ b)} causes \texttt{nlsr::nlxb} to fail because the partial
-  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\^{}b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
+  derivative w.r.t. \texttt{b} is \texttt{a\ *\ (x\ \^{}\ b\ *\ log(x))}. If there is data for which \texttt{x\ =\ 0},
   the derivative is undefined, but the model can be computed. In such cases,
@@
 
-\subsubsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}
+\hypertarget{jacobian-code-in-selfstart-models}{%
+\subsection{Jacobian code in selfStart models}\label{jacobian-code-in-selfstart-models}}
 
@@
 command, making them difficult to access by users who may wish to create
-their own selfStart model via copy and edit.
+their own \texttt{selfStart} model via copy and edit.
 Interactive tools, such as ``visual fitting'' (John C. Nash and Velleman (1996)) might be
@@
 
-\subsection{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}
+\hypertarget{bounds-constraints-on-parameters}{%
+\section{Bounds constraints on parameters}\label{bounds-constraints-on-parameters}}
 
@@
 \item
-  \texttt{nls()} can only impose bounds if the \texttt{algorithm="port"} argument is
+  \texttt{nls()} can only impose bounds if the \texttt{algorithm\ =\ "port"} argument is
   used in the call. Unfortunately, the documentation warns us:
@@
 # Start MUST be feasible i.e. on or within bounds
-anlshob1b <- nls(frms, start=sts1, data=weeddf, lower=c(0,0,0),
-             upper=c(2,6,3), algorithm='port')
-pnls(anlshob1b) #  check the answer (short form)
+anlshob1b <- try(nls(frms, start = sts1, data = weeddf, lower = c(0,0,0),
+             upper = c(2,6,3), algorithm = 'port'))
+if (! inherits(anlshob1b, "try-error")) pnls(anlshob1b) #  check the answer (short form)
 \end{verbatim}
@@
 # nlsLM seems NOT to work with bounds in this example
-anlsLM1b <- nlsLM(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls(anlsLM1b)
+anlsLM1b <- try(nlsLM(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlsLM1b, "try-error")) pnls(anlsLM1b)
 \end{verbatim}
@@
 # also no warning if starting out of bounds, but gets a good answer!!
-st4<-c(c1=4, c2=4, c3=4)
-anlsLMob <- nlsLM(frms, start=st4, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pnls(anlsLMob)
+st4 <- c(c1 = 4, c2 = 4, c3 = 4)
+anlsLMob <- try(nlsLM(frms, start = st4, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlsLMob, "try-error")) pnls(anlsLMob)
 \end{verbatim}
@@
 # Try nlsr::nlxb()
-anlx1b <- nlxb(frms, start=sts1, data=weeddf, lower=c(0,0,0), upper=c(2,6,3))
-pshort(anlx1b)
+anlx1b <- try(nlxb(frms, start = sts1, data = weeddf, lower = c(0,0,0), upper = c(2,6,3)))
+if (! inherits(anlx1b, "try-error")) pshort(anlx1b)
 \end{verbatim}
@@
 
-\subsubsection{Philosophical considerations}\label{philosophical-considerations}
+\hypertarget{philosophical-considerations}{%
+\subsection{Philosophical considerations}\label{philosophical-considerations}}
 
@@
 
-\subsubsection{Fixed parameters (masks)}\label{fixed-parameters-masks}
+\hypertarget{fixed-parameters-masks}{%
+\subsection{Fixed parameters (masks)}\label{fixed-parameters-masks}}
 
@@
 # Hobbsmaskx.R -- masks with formula specification of the problem
-require(nlsr); require(minpack.lm); traceval<-FALSE
-stu <- c(b1=200, b2=50, b3=0.3) # a default starting vector (named!)
-sts <- c(c1=2, c2=5, c3=3) # a default scaled starting vector (named!)
+require(nlsr); require(minpack.lm); traceval <- FALSE
+stu <- c(b1 = 200, b2 = 50, b3 = 0.3) # a default starting vector (named!)
+sts <- c(c1 = 2, c2 = 5, c3 = 3) # a default scaled starting vector (named!)
 # fix first parameter
-anxbmsk1 <- try(nlxb(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-            upper=c(200, 60, 3), trace=traceval))
-print(anxbmsk1)
+anxbmsk1 <- try(nlxb(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+            upper = c(200, 60, 3), trace=traceval))
+if (! inherits(anxbmsk1, "try-error")) print(anxbmsk1)
 \end{verbatim}
@@
 \begin{verbatim}
-anlM1 <- try(nlsLM(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-            upper=c(200, 60, 3), trace=traceval))
-pnls(anlM1)
+anlM1 <- try(nlsLM(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+            upper=c(200, 60, 3), trace = traceval))
+if (! inherits(anlM1, "try-error")) pnls(anlM1)
 \end{verbatim}
@@
 \begin{verbatim}
-anlsmsk1 <- try(nls(frmu, start=stu, data=weeddf, lower=c(200,0,0), 
-        upper=c(200, 60, 3),  algorithm="port", trace=traceval))
-pnls(anlsmsk1)
+anlsmsk1 <- try(nls(frmu, start = stu, data = weeddf, lower = c(200, 0, 0), 
+        upper = c(200, 60, 3),  algorithm = "port", trace = traceval))
+if (! inherits(anlsmsk1, "try-error")) pnls(anlsmsk1)
 \end{verbatim}
@@
 # Hobbs scaled problem with bounds, formula specification
-anlxmsks1 <- nlxb(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                  upper=c(2,6,30))
-print(anlxmsks1)
+anlxmsks1 <- try(nlxb(frms, start = sts, data = weeddf, lower = c(2, 0, 0),
+                  upper = c(2, 6, 30)))
+if (! inherits(anlxmsks1, "try-error")) print(anlxmsks1)
 \end{verbatim}
@@
 \begin{verbatim}
-anlshmsk1 <- nls(frms, start=sts, trace=traceval, data=weeddf, lower=c(2,0,0),
-             upper=c(2,6,30), algorithm='port')
-pnls(anlshmsk1)
+anlshmsk1 <- try(nls(frms, start = sts, trace = traceval, data = weeddf, 
+                 lower = c(2, 0, 0), upper = c(2, 6, 30), algorithm = 'port'))
+if (! inherits(anlshmsk1, "try-error")) pnls(anlshmsk1)
 \end{verbatim}
@@
 \begin{verbatim}
-anlsLMmsks1 <- nlsLM(frms, start=sts, data=weeddf, lower=c(2,0,0),
-                 upper=c(2,6,30))
-pnls(anlsLMmsks1)
+anlsLMmsks1 <- try(nlsLM(frms, start = sts, data = weeddf, lower = c(2,0,0),
+                 upper = c(2,6,30)))
+if (! inherits(anlsLMmsks1, "try-error")) pnls(anlsLMmsks1)
 \end{verbatim}
@@
 # Test with all parameters masked
-anlxmskall<- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
-print(anlxmskall)
+anlxmskall <- try(nlxb(frms, start=sts, data=weeddf, lower=sts, upper=sts))
+if (! inherits(anlxmskall, "try-error")) print(anlxmskall)
 \end{verbatim}
@@
 \texttt{nlsr} has an output format that indicates the constraint status of the parameter
-estimates. For \texttt{nlsr}, we have \textbf{chosen} to suppress the calculation of
+estimates. For \texttt{nlsr}, we have chosen to suppress the calculation of
 approximate standard errors in the parameters when constraints are active because
@@
 
-\subsection{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}
+\hypertarget{stabilization-of-gauss-newton-computations}{%
+\section{Stabilization of Gauss-Newton computations}\label{stabilization-of-gauss-newton-computations}}
 
@@
 
-\subsubsection{Programming language}\label{programming-language}
+\hypertarget{programming-language}{%
+\subsection{Programming language}\label{programming-language}}
 
@@
 programming language. \texttt{nls()} uses a mix of R, C, and Fortran, as does \texttt{minpack.lm}.
-\texttt{gslnls} is an R wrapper to various C-language routines in the Gnu Scientific
+\texttt{gslnls} is an R wrapper to various C-language routines in the GNU Scientific
 Library (Galassi et al. (2009)).
@@
 
-\subsection{Data sources for problems}\label{data-sources-for-problems}
+\hypertarget{data-sources-for-problems}{%
+\section{Data sources for problems}\label{data-sources-for-problems}}
 
@@
 search in the available environments (i.e., workspaces) for suitable data objects.
-We do NOT like this approach, but it is ``the R way''. R allows users to leave many
+We do not like this approach, but it is ``the R way''. R allows users to leave many
 objects in the default (\texttt{.GlobalEnv}) workspace. Moreover, users have to actively
-suppress saving this workspace (\texttt{.RData}) on exit; otherwise any such file in
+suppress saving this workspace (\texttt{.RData}) on exit, otherwise any such file in
 the path will be loaded on startup. R users in our
@@
 
-\subsubsection{Feature: Subsetting}\label{feature-subsetting}
+\hypertarget{feature-subsetting}{%
+\subsection{Feature: Subsetting}\label{feature-subsetting}}
 
@@
 
-\subsubsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}
+\hypertarget{feature-na.action-missing-value-treatment}{%
+\subsection{Feature: na.action (missing value treatment)}\label{feature-na.action-missing-value-treatment}}
 
@@
 
-\subsection{Feature: model frame}\label{feature-model-frame}
+\hypertarget{feature-model-frame}{%
+\subsection{Feature: model frame}\label{feature-model-frame}}
 
@@
 
-\subsection{Weights on observations}\label{weights-on-observations}
+\hypertarget{weights-on-observations}{%
+\subsection{Weights on observations}\label{weights-on-observations}}
 
@@
 
-\subsubsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}
+\hypertarget{weights-in-returned-functions-from-nls}{%
+\subsection{Weights in returned functions from nls()}\label{weights-in-returned-functions-from-nls}}
 
@@
 \begin{verbatim}
-wts <- 0.5^tt # simple weights
-frmlogis <- weed ~ Asym/(1 + exp((xmid - tt)/scal))
-Asym<-1; xmid<-1; scal<-1
-nowt<-nls(weed ~ SSlogis(tt, Asym, xmid, scal)) # UNWEIGHTED
-rnowt<-nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
-attr(rnowt, "gradient")<-NULL; rnowt
+wts <- 0.5 ^ tt # simple weights
+frmlogis <- weed ~ Asym / (1 + exp((xmid - tt)/scal))
+Asym <- 1; xmid <- 1; scal <- 1
+nowt <- try(nls(weed ~ SSlogis(tt, Asym, xmid, scal))) # UNWEIGHTED
+if (! inherits(nowt, "try-error")) {
+  rnowt <- nowt$m$resid() # This has UNWEIGHTED residual and Jacobian. Does NOT take coefficients.
+  attr(rnowt, "gradient") <- NULL 
+} else rnowt <- NULL
+rnowt
 \end{verbatim}
@@
 \begin{verbatim}
-usewt <- nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights=wts)
-rusewt<-usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
-attr(rusewt,"gradient")<-NULL; rusewt
+usewt <- try(nls(weed ~ SSlogis(tt, Asym, xmid, scal), weights = wts))
+if (! inherits(usewt, "try-error")) {
+  rusewt <- usewt$m$resid() # WEIGHTED. Does NOT take coefficients.
+  attr(rusewt, "gradient") <- NULL
+} else rusewt <- NULL
+rusewt
 \end{verbatim}
@@
 ## source("nlsModel.R") # or use {r nlsmodelsource, echo=FALSE} code chunk
-nmod0 <- nlsModel(frmlogis, data=weeddf, start=c(Asym=1, xmid=1, scal=1), wts=wts)
-rn0<-nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
-attr(rn0,"gradient")<-NULL; rn0 # weighted residuals at starting coefficients
+nmod0 <- nlsModel(frmlogis, data = weeddf, start = c(Asym = 1, xmid = 1, scal = 1), wts = wts)
+rn0 <- nmod0$resid() # Parameters are supplied in nlsModel() `start` above.
+attr(rn0, "gradient") <- NULL; rn0 # weighted residuals at starting coefficients
 \end{verbatim}
@@
 \begin{verbatim}
-nmod <- nlsModel(frmlogis, data=weeddf, start=coef(usewt), wts=wts)
-rn<-nmod$resid()
+nmod <- nlsModel(frmlogis, data = weeddf, start = coef(usewt), wts = wts)
+rn <- nmod$resid()
 attr(rn,"gradient")<-NULL; rn # same as rusewt
@@
 
-\subsection{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}
-
-\subsubsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}
+\hypertarget{minor-issues-with-nonlinear-least-squares-tools}{%
+\section{Minor issues with nonlinear least-squares tools}\label{minor-issues-with-nonlinear-least-squares-tools}}
+
+\hypertarget{interim-output-from-the-port-algorithm}{%
+\subsection{Interim output from the ``port'' algorithm}\label{interim-output-from-the-port-algorithm}}
 
@@
 
-\subsubsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}
+\hypertarget{failure-to-return-the-best-result-achieved}{%
+\subsection{Failure to return the best result achieved}\label{failure-to-return-the-best-result-achieved}}
 
@@
 \begin{verbatim}
+time <- c( 1,  2,  3,  4,  6 , 8, 10, 12, 16)
+conc <- c( 0.7, 1.2, 1.4, 1.4, 1.1, 0.8, 0.6, 0.5, 0.3)
+NLSdata <- data.frame(time,conc)
+NLSstart <- c(lrc1 = -2, lrc2 = 0.25, A1 = 150, A2 = 50) # a starting vector (named!)
+NLSformula <- conc ~ A1 * exp(-exp(lrc1) * time) + A2 * exp(-exp(lrc2) * time)
+tryit <- try(nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE))
+\end{verbatim}
+
+\begin{verbatim}
 #> 61216.    (3.56e+03): par = (-2 0.25 150 50)
@@
 \begin{verbatim}
-#> [1] "Error in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE) : \n  singular gradient\n"
-#> attr(,"class")
-#> [1] "try-error"
-#> attr(,"condition")
-#> <simpleError in nls(NLSformula, data = NLSdata, start = NLSstart, trace = TRUE): singular gradient>
+if (! inherits(tryit, "try-error")) print(tryit)
 \end{verbatim}
@@
 
-\subsection{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}
+\hypertarget{estimating-models-that-are-partially-linear}{%
+\section{Estimating models that are partially linear}\label{estimating-models-that-are-partially-linear}}
 
@@
 are linear and communicating this information to estimating functions
-is not a trivial task. \texttt{nls()} has an option \texttt{algorithm="plinear"} that allows some partially linear
+is not a trivial task. \texttt{nls()} has an option \texttt{algorithm\ =\ "plinear"} that allows some partially linear
 models to be solved. The other tools, as far as we are aware, do not offer any
@@
 Within \texttt{nls()}, consider the following FOUR different specifications for the same
-problem, plus an intuitive choice, labeled \texttt{fm2a}, that does NOT work.
+problem, plus an intuitive choice, labeled \texttt{fm2a}, that does not work.
 In this failed attempt, putting the \texttt{Asym} parameter in the model causes the
@@
                  algorithm = "port"))
-summary(fm4)
-
-## using conditional linearity AND Asym does NOT work
+if (! inherits(fm4, "try-error")) summary(fm4)
+
+## using conditional linearity AND Asym does not work
 fm2a <- try(nls(density ~ Asym / (1 + exp((xmid - log(conc)) / scal)), 
@@
                  algorithm = "plinear", trace = TRUE))
-summary(fm2a)
-\end{verbatim}
-
-\subsection{Models with indexed parameters}\label{models-with-indexed-parameters}
+if (! inherits(fm2a, "try-error")) summary(fm2a)
+\end{verbatim}
+
+\hypertarget{models-with-indexed-parameters}{%
+\section{Models with indexed parameters}\label{models-with-indexed-parameters}}
 
@@
 parameters are indexed but which uses the ``plinear'' option as an added complication.
-Running this example reveals that the answers for the parameters are NOT indexed
+Running this example reveals that the answers for the parameters are not indexed
 as in a vector.
@@
 
-\subsection{Tests and use-case examples}\label{tests-and-use-case-examples}
+\hypertarget{tests-and-use-case-examples}{%
+\section{Tests and use-case examples}\label{tests-and-use-case-examples}}
 
@@
 and the conventional R package testing tools).
-In our work, one of us has developed a working prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}} .
+In our work, one of us (AB) has developed a working prototype package at \emph{\url{https://github.com/ArkaB-DS/nlsCompare}} .
 A primary design objective of this is to allow the summarization of multiple tests in
@@
 
-\subsection{Documentation and resources}\label{documentation-and-resources}
+\hypertarget{documentation-and-resources}{%
+\section{Documentation and resources}\label{documentation-and-resources}}
 
@@
 
-\subsection{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}
+\hypertarget{future-of-nonlinear-model-estimation-in-r}{%
+\section{Future of nonlinear model estimation in R}\label{future-of-nonlinear-model-estimation-in-r}}
 
@@
 
-\section{Acknowledgments}\label{acknowledgments}
+\hypertarget{acknowledgments}{%
+\section{Acknowledgments}\label{acknowledgments}}
 
@@
 vignette. Colin Gillespie (package \texttt{benchmarkme}) has been helpful in guiding our
-attempts to succinctly summarize computing environments.
-
-\section*{References}\label{references}
+attempts to succinctly summarize computing environments. We thank one of the referees
+for pointing out that it is important to use background knowledge and graphs to
+guide modeling.
+
+\hypertarget{references}{%
+\section{References}\label{references}}
+
+\hypertarget{references-1}{%
+\section*{References}\label{references-1}}
 \addcontentsline{toc}{section}{References}
 
-\phantomsection\label{refs}
-\setlength{\cslentryspacing}{0em}
-\begin{CSLReferences}
-\bibitem[\citeproctext]{ref-R-onls-manual}
-Andrej-Nikolai Spiess. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
-
-\bibitem[\citeproctext]{ref-bateswatts}
-Bates, D. M., and D. G. Watts. 1988. \emph{Nonlinear Regression Analysis and Its Applications}. Wiley.
-
-\bibitem[\citeproctext]{ref-BatesWatts81}
-Bates, Douglas M., and Watts, Donald G. 1981. {``A Relative Offset Orthogonality Convergence Criterion for Nonlinear Least Squares.''} \emph{Technometrics} 23 (2): 179--83.
-
-\bibitem[\citeproctext]{ref-p-lme4}
+\hypertarget{refs}{}
+\begin{CSLReferences}{1}{0}
+\leavevmode\vadjust pre{\hypertarget{ref-NISTnls-manual}{}}%
+Bates, Douglas M. 2012. \emph{{NISTnls: Nonlinear least squares examples from NIST}}. \url{https://CRAN.R-project.org/package=NISTnls}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-BatesWatts81}{}}%
+Bates, Douglas M., and Donald G. Watts. 1981. {``{A Relative Offset Orthogonality Convergence Criterion for Nonlinear least Squares}.''} \emph{Technometrics} 23 (2): 179--83.
+
+\leavevmode\vadjust pre{\hypertarget{ref-bateswatts}{}}%
+---------. 1988. \emph{{Nonlinear Regression Analysis and Its Applications}}. Wiley.
+
+\leavevmode\vadjust pre{\hypertarget{ref-p-lme4}{}}%
 Bates, Douglas, Martin Mächler, Ben Bolker, and Steve Walker.
@@
 
-\bibitem[\citeproctext]{ref-p-nlstools}
+\leavevmode\vadjust pre{\hypertarget{ref-p-nlstools}{}}%
 Baty, Florent, and Marie-Laure Delignette-Muller. 2013. \emph{Nlstools: Tools for Nonlinear Regression Diagnostics}.
 
-\bibitem[\citeproctext]{ref-nlsMicrobio-manual}
+\leavevmode\vadjust pre{\hypertarget{ref-nlsMicrobio-manual}{}}%
 ---------. 2014. \emph{nlsMicrobio: Data Sets and Nonlinear Regression Models Dedicated to Predictive Microbiology}.
 
-\bibitem[\citeproctext]{ref-p-bbmle}
+\leavevmode\vadjust pre{\hypertarget{ref-p-bbmle}{}}%
 Bolker, Ben, and R Development Core Team. 2013. \emph{Bbmle: Tools for General Maximum Likelihood Estimation}. \url{http://CRAN.R-project.org/package=bbmle}.
 
-\bibitem[\citeproctext]{ref-gslnls23}
-Chau, Joris. 2023. \emph{Gslnls: GSL Nonlinear Least-Squares Fitting}. \url{https://CRAN.R-project.org/package=gslnls}.
-
-\bibitem[\citeproctext]{ref-nlshelper-manual}
+\leavevmode\vadjust pre{\hypertarget{ref-gslnls23}{}}%
+Chau, Joris. 2023. \emph{{gslnls: GSL Nonlinear Least-Squares Fitting}}. \url{https://CRAN.R-project.org/package=gslnls}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-nlshelper-manual}{}}%
 Duursma, Remko. 2017. \emph{Nlshelper: Convenient Functions for Non-Linear Regression}. \url{https://CRAN.R-project.org/package=nlshelper}.
 
-\bibitem[\citeproctext]{ref-NISTnls-manual}
-(ed.), Douglas Bates. 2012. \emph{NISTnls: Nonlinear Least Squares Examples from NIST}. \url{https://CRAN.R-project.org/package=NISTnls}.
-
-\bibitem[\citeproctext]{ref-minpacklm12}
+\leavevmode\vadjust pre{\hypertarget{ref-minpacklm12}{}}%
 Elzhov, Timur V., Katharine M.
@@
 2012.
-\emph{Minpack.lm: R Interface to the Levenberg-Marquardt Nonlinear Least-Squares Algorithm Found in MINPACK, Plus Support for Bounds}.
+\emph{{minpack.lm: R interface to the Levenberg-Marquardt nonlinear least-squares algorithm found in MINPACK, plus support for bounds}}.
 R Project for Statistical Computing.
@@
 
-\bibitem[\citeproctext]{ref-GSL-manual}
-Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman. 2009. \emph{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3. Ed.).}
-
-\bibitem[\citeproctext]{ref-Gallant87}
+\leavevmode\vadjust pre{\hypertarget{ref-GSL-manual}{}}%
+Galassi, Mark, Jim Davies, James Theiler, Brian Gough, and Gerard Jungman.
+2009.
+\emph{{GNU Scientific Library - Reference Manual, Third Edition, for GSL Version 1.12 (3.
+ed.)}}.
+Free Software Foundation.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Gallant87}{}}%
 Gallant, A. Ronald. 1987. \emph{Nonlinear Statistical Models}. Wiley.
 
-\bibitem[\citeproctext]{ref-Bioconductor}
+\leavevmode\vadjust pre{\hypertarget{ref-Bioconductor}{}}%
 Gentleman, Robert C., Vincent J.
@@
 
-\bibitem[\citeproctext]{ref-Golub1973}
+\leavevmode\vadjust pre{\hypertarget{ref-Golub1973}{}}%
 Golub, G.
@@
 
-\bibitem[\citeproctext]{ref-nls2-22}
+\leavevmode\vadjust pre{\hypertarget{ref-nls2-22}{}}%
 Grothendieck, G. 2022. \emph{Nls2: Non-Linear Regression with Brute Force}. \url{https://CRAN.R-project.org/package=nls2}.
 
-\bibitem[\citeproctext]{ref-Hartley1961}
-Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269-\/-280.
-
-\bibitem[\citeproctext]{ref-Huet2004}
+\leavevmode\vadjust pre{\hypertarget{ref-Hartley1961}{}}%
+Hartley, H. O. 1961. {``The Modified Gauss-Newton Method for the Fitting of Non-Linear Regression Functions by Least Squares.''} \emph{Technometrics} 3: 269--80.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Huet2004}{}}%
 Huet, S., A.
@@
 
-\bibitem[\citeproctext]{ref-nlsr2023manual}
-John C Nash, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
-
-\bibitem[\citeproctext]{ref-tvrdik2007}
+\leavevmode\vadjust pre{\hypertarget{ref-tvrdik2007}{}}%
 Josef Tvrdík and Ivan Křivý and Ladislav Mišík.
@@
 
-\bibitem[\citeproctext]{ref-Levenberg1944}
-Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164-\/-168.
-
-\bibitem[\citeproctext]{ref-Marquardt1963}
+\leavevmode\vadjust pre{\hypertarget{ref-Levenberg1944}{}}%
+Levenberg, Kenneth. 1944. {``A Method for the Solution of Certain Non-Linear Problems in Least Squares.''} \emph{Quarterly of Applied Mathematics} 2: 164--68.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Marquardt1963}{}}%
 Marquardt, Donald W. 1963. {``{An Algorithm for Least-Squares Estimation of Nonlinear Parameters}.''} \emph{SIAM Journal on Applied Mathematics} 11 (2): 431--41.
 
-\bibitem[\citeproctext]{ref-MiguezNLRAA2021}
+\leavevmode\vadjust pre{\hypertarget{ref-MiguezNLRAA2021}{}}%
 Miguez, Fernando. 2021. \emph{{nlraa: Nonlinear Regression for Agricultural Applications}}. \url{https://CRAN.R-project.org/package=nlraa}.
 
-\bibitem[\citeproctext]{ref-jn77ima}
+\leavevmode\vadjust pre{\hypertarget{ref-jn77ima}{}}%
 Nash, John C. 1977. {``Minimizing a Nonlinear Sum of Squares Function on a Small Computer.''} \emph{Journal of the Institute for Mathematics and Its Applications} 19: 231--37.
 
-\bibitem[\citeproctext]{ref-cnm79}
+\leavevmode\vadjust pre{\hypertarget{ref-cnm79}{}}%
 ---------. 1979. \emph{Compact Numerical Methods for Computers: Linear Algebra and Function Minimisation}. Bristol: Adam Hilger.
 
-\bibitem[\citeproctext]{ref-NashWires22}
----------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics}, no. e1580. https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
-
-\bibitem[\citeproctext]{ref-PkgFromRbase22}
+\leavevmode\vadjust pre{\hypertarget{ref-NashWires22}{}}%
+---------. 2022. {``{Function minimization and nonlinear least squares in R}.''} \emph{WIREs Computational Statistics} 14 (e1580). https://doi.org/\url{https://doi.org/10.1002/wics.1580}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-PkgFromRbase22}{}}%
 Nash, John C., and Arkajyoti Bhattacharjee. 2022. {``Making a Package from Base {R} Files.''} \emph{R-Bloggers}, January. \url{https://www.r-bloggers.com/2022/01/making-a-package-from-base-r-files/}.
 
-\bibitem[\citeproctext]{ref-p-optimx}
+\leavevmode\vadjust pre{\hypertarget{ref-nlsr2023manual}{}}%
+Nash, John C, and Duncan Murdoch. 2023. \emph{{nlsr: Functions for Nonlinear Least Squares Solutions}}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-p-optimx}{}}%
 Nash, John C, and Ravi Varadhan. 2011. \emph{\emph{Optimx}: A Replacement and Extension of the {optim()} Function}. Nash Information Services Inc.; Johns Hopkins University.
 
-\bibitem[\citeproctext]{ref-nash1996nonlinear}
+\leavevmode\vadjust pre{\hypertarget{ref-nash1996nonlinear}{}}%
 Nash, John C., and Paul Velleman.
@@
 
-\bibitem[\citeproctext]{ref-jnmws87}
-Nash, John C., and Mary Walker-Smith. 1987. \emph{Nonlinear Parameter Estimation: An Integrated System in BASIC}. Book. New York: Marcel Dekker.
-
-\bibitem[\citeproctext]{ref-OlearyRust13}
+\leavevmode\vadjust pre{\hypertarget{ref-jnmws87}{}}%
+Nash, John C., and Mary Walker-Smith. 1987. \emph{{Nonlinear Parameter Estimation: An Integrated System in BASIC}}. New York: Marcel Dekker.
+
+\leavevmode\vadjust pre{\hypertarget{ref-OlearyRust13}{}}%
 O'Leary, Dianne P., and Bert W. Rust. 2013. {``Variable Projection for Nonlinear Least Squares Problems.''} \emph{Computational Optimization and Applications} 54 (3): 579--93.
 
-\bibitem[\citeproctext]{ref-nlsmultstart}
-Padfield, Daniel, and Granville Matheson. 2020. \emph{Nls.multstart: Robust Non-Linear Regression Using AIC Scores}. \url{https://CRAN.R-project.org/package=nls.multstart}.
-
-\bibitem[\citeproctext]{ref-p-nlme}
+\leavevmode\vadjust pre{\hypertarget{ref-nlsmultstart}{}}%
+Padfield, Daniel, and Granville Matheson. 2020. \emph{{nls.multstart: Robust Non-Linear Regression using AIC Scores}}. \url{https://CRAN.R-project.org/package=nls.multstart}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-p-nlme}{}}%
 Pinheiro, Jose, Douglas Bates, Saikat DebRoy, Deepayan Sarkar, and R Core Team. 2013. \emph{Nlme: Linear and Nonlinear Mixed Effects Models}.
 
-\bibitem[\citeproctext]{ref-m-nlsmsn}
-Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{Nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}. \url{https://CRAN.R-project.org/package=nlsmsn}.
-
-\bibitem[\citeproctext]{ref-Ratkowsky1983}
+\leavevmode\vadjust pre{\hypertarget{ref-m-nlsmsn}{}}%
+Prates, Marcos, Victor Lachos, and Aldo Garay. 2021. \emph{{nlsmsn: Fitting Nonlinear Models with Scale Mixture of Skew-Normal Distributions}}. \url{https://CRAN.R-project.org/package=nlsmsn}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Ratkowsky1983}{}}%
 Ratkowsky, David A. 1983. \emph{Nonlinear Regression Modeling: A Unified Practical Approach}. New York; Basel: Marcel Dekker Inc.
 
-\bibitem[\citeproctext]{ref-m-nlstac}
+\leavevmode\vadjust pre{\hypertarget{ref-m-nlstac}{}}%
 Rodriguez-Arias, Mariano, Juan Antonio Fernandez, Javier Cabello, and Rafael Benitez.
@@
 
-\bibitem[\citeproctext]{ref-Ross90}
+\leavevmode\vadjust pre{\hypertarget{ref-Ross90}{}}%
 Ross, Gavin J. S. 1990. \emph{Nonlinear Estimation.} New York etc.: Springer-Verlag.
 
-\bibitem[\citeproctext]{ref-Seber89Wild}
+\leavevmode\vadjust pre{\hypertarget{ref-Seber89Wild}{}}%
 Seber, G. A. F., and C. J. Wild. 1989. \emph{{Nonlinear regression}}. New York: Wiley.
 
-\bibitem[\citeproctext]{ref-nlsic22}
+\leavevmode\vadjust pre{\hypertarget{ref-nlsic22}{}}%
 Sokol, Serguei. 2022. \emph{Nlsic: Non Linear Least Squares with Inequality Constraints}. \url{https://CRAN.R-project.org/package=nlsic}.
 
-\bibitem[\citeproctext]{ref-Torvisco2018}
-Torvisco, Juan Antonio Fernandez, Mariano Rodrıguez-Arias, and Javier Cabello Sanchez.
+\leavevmode\vadjust pre{\hypertarget{ref-R-onls-manual}{}}%
+Spiess, Andrej-Nikolai. 2022. \emph{Onls: Orthogonal Nonlinear Least-Squares Regression}. \url{https://CRAN.R-project.org/package=onls}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-Torvisco2018}{}}%
+Torvisco, Juan Antonio Fernandez, Mariano Rodriguez-Arias, and Javier Cabello Sanchez.
 2018.
-{``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Niˇs, Serbia}.
-\url{https://doi.org/doi.org/10.2298/FIL1812233T}.
-
-\bibitem[\citeproctext]{ref-R-crsnls-manual}
+{``A New Algorithm to Fit Exponential Decays Without Initial Guess.''} {Faculty of Sciences and Mathematics, University of Nis, Serbia}.
+\url{https://doi.org/10.2298/fil1812233t}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-R-crsnls-manual}{}}%
 Tvrdík, Josef. 2016. \emph{{crsnls: Nonlinear Regression Parameters Estimation by 'CRS4HC' and 'CRS4HCe'}}. \url{https://CRAN.R-project.org/package=crsnls}.
 
-\bibitem[\citeproctext]{ref-HWtestthat11}
-Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal_2011-1_Wickham.pdf}.
-
-\bibitem[\citeproctext]{ref-HWdevtools21}
+\leavevmode\vadjust pre{\hypertarget{ref-HWtestthat11}{}}%
+Wickham, Hadley. 2011. {``{testthat: Get Started with Testing}.''} \emph{The R Journal} 3: 5--10. \url{https://journal.r-project.org/archive/2011-1/RJournal/_2011-1/_Wickham.pdf}.
+
+\leavevmode\vadjust pre{\hypertarget{ref-HWdevtools21}{}}%
 Wickham, Hadley, Jim Hester, Winston Chang, and Jennifer Bryan. 2021. \emph{{devtools: Tools to Make Developing R Packages Easier}}. \url{https://CRAN.R-project.org/package=devtools}.

### 2023-11-13 JNABresubmission20231101/ComparingNLStools.R

@@ -1,0 +1,410 @@
+## ----setup, include=FALSE--------------------------------------------------------------------------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----ex01, echo=TRUE-------------------------------------------------------------------------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE----------------------------------------------------------------------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
+#
+# Starting parameter sets
+stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
+sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
+stt1 <- c(Asym = 1, xmid = 1, scal = 1)
+
+
+## ----ex02fn, echo=TRUE-----------------------------------------------------------------------------------------------------------------------
+# Logistic3U
+hobbs.res <- function(x){ # unscaled Hobbs weeds problem -- residual
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt <- 1:12
+  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
+  # Note: this form of the residual, from Laplace (1788) in the form
+  # of "fitted - observed" has been used in our software for half
+  # a century, but sometimes concerns users of the more traditional
+  # "observed - fitted" form
+}
+
+hobbs.jac <- function(x) { # unscaled Hobbs weeds problem -- Jacobian
+  jj <- matrix(0.0, 12, 3)
+  tt <- 1:12
+  yy <- exp(-x[3] * tt)
+  zz <- 1.0 / (1 + x[2] * yy)
+  jj[tt, 1] <- zz
[…]

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.Rmd
copyedit temp

--- 2023-11-13 JNABresubmission20231101/ComparingNLStools.Rmd
@@
 abstract: >
-  Our Google Summer of Code project "Improvements to `nls()`" considered the 
-  features and limitations of the R function `nls()` with the aim 
-  of improving and rationalizing R tools for nonlinear regression.
+  Our Google Summer of Code project "Improvements to `nls()`" investigated
+  rationalizing R tools for nonlinear regression and nonlinear estimation tools by
+  considering usability, maintainability and functionality, especially for a
+  Gauss-Newton solver.
+  <!-- JN: copyedit essentially accepted -->
   The rich features of `nls()` are weakened by several deficiencies and
@@
 
-`nls()` is the tool for estimating nonlinear statistical models in base R, the 
-primary software distribution
-from the Comprehensive R Archive Network (*https://cran.r-project.org*).
+`nls()` is a Comprehensive R Archive Network (CRAN: *https://cran.r-project.org*) 
+tool that has remarkable and wide-ranging features for estimating nonlinear
+statistical models that are expressed as **formulas**. 
+<!-- JN: copyedit comment integrated -->
 The function dates to the 1980s and the work related to @bateswatts in S (see 
@@
 
-The `nls()` function has a remarkable and comprehensive set of capabilities for 
-estimating nonlinear models that are expressed as **formulas**. In particular, we
+In particular, we
 note that it
 
-- handles formulas that include R functions, even ones that call calculations in
-  other programming languages
-- allows data to be weighted or subset
+- can call calculations in other programming languages
+<!-- JN: copyedit comment integrated -->
+- allows weighted or subset data
+<!-- JN: copyedit comment integrated -->
 - can estimate bound-constrained parameters
-- provides a mechanism for handling partially linear models
+- provides partially linear model handling mechanism
+<!-- JN: copyedit comment integrated -->
 - permits parameters to be indexed over a set of related data
@@
 
-With such a range of features and long history, the code has become untidy 
-and overly patched, difficult to maintain, and its underlying methods could
-be improved. Various workers have developed packages to overcome these 
-concerns, and we will address some of these here.
+Due to its extensive range of features and prolonged history, the code has become untidy and overly patched, making it challenging to maintain and ripe for improvement in its underlying methods. 
+<!-- JN: copyedit comment integrated -->
 
@@
 
-The tools we will consider are the base-R `nls()` function 
-and some packages that are available on the CRAN repository.
-Of these, we will pay particular attention to **nlsr** (@nlsr2023manual),
+Besides the base-R `nls()` function, we will pay particular attention 
+to **nlsr** (@nlsr2023manual),
 **minpack.lm** (@minpacklm12), and **gslnls** (@gslnls23) which are
-general nonlinear least-squares solvers.
-
+general nonlinear least-squares solvers in the CRAN repository.
 While we will provide capsule comments for some other CRAN packages, 
-we will not pursue those in the Bioconductor (@Bioconductor) 
-collection, nor those on repositories such as 
-GitHub (https://github.com) and Gitlab (https://about.gitlab.com).
+those in the Bioconductor (@Bioconductor) collection are more
+specialized, and those on repositories such as 
+GitHub (https://github.com) and Gitlab (https://about.gitlab.com),
+while interesting, do not have the checking applied to CRAN packages.
+<!-- JN: revised text combining 2 copyedit comments -->
 
@@
 
-Our work was aimed at unifying nonlinear modeling functionality in R,
+Our work aimed at unifying nonlinear modeling functionality in R,
+<!-- JN: removed "was" as suggesgted -->
 ideally in a refactored `nls()` function.
@@
 with documentation and using recent tools. If there is a suspicion that there
-may be ill-conditioning, package `nlsr` or the example we give below of how to
+may be ill-conditioning, package `nlsr` or the example we give in the 
+section "Comparison notes for formula-setup solutions" below of how to
 find singular values of the Jacobian allow these diagnostics to be calculated.
-
-For R **developers**, we invite and encourage discussion of the design choices, since these
+<!-- JN: reference to section added ??have I done this correctly? -->
+
+For R **developers**, we invite and encourage discussion of the design choices, 
+since these
 have downstream implications for ease of use, adaptation to new features, and 
 efficiency of ongoing maintenance.
+<!-- JN: Could not understand how copyedit "such as" would be applied -->
 

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.tex
copyedit temp

--- 2023-11-13 JNABresubmission20231101/ComparingNLStools.tex
@@
 \abstract{%
-Our Google Summer of Code project ``Improvements to \texttt{nls()}'' considered the features and limitations of the R function \texttt{nls()} with the aim of improving and rationalizing R tools for nonlinear regression.
+Our Google Summer of Code project ``Improvements to \texttt{nls()}'' investigated rationalizing R tools for nonlinear regression and nonlinear estimation tools by considering usability, maintainability and functionality, especially for a Gauss-Newton solver.
 The rich features of \texttt{nls()} are weakened by several deficiencies and inconsistencies such as a lack of stabilization of the Gauss-Newton solver.
@@
 
-\texttt{nls()} is the tool for estimating nonlinear statistical models in base R, the
-primary software distribution
-from the Comprehensive R Archive Network (\emph{\url{https://cran.r-project.org}}).
+\texttt{nls()} is a Comprehensive R Archive Network (CRAN: \emph{\url{https://cran.r-project.org}})
+tool that has remarkable and wide-ranging features for estimating nonlinear
+statistical models that are expressed as \textbf{formulas}.
 The function dates to the 1980s and the work related to D. M. Bates and Watts (1988) in S (see
@@
 
-The \texttt{nls()} function has a remarkable and comprehensive set of capabilities for
-estimating nonlinear models that are expressed as \textbf{formulas}. In particular, we
+In particular, we
 note that it
@@
 \item
-  handles formulas that include R functions, even ones that call calculations in
-  other programming languages
-\item
-  allows data to be weighted or subset
+  can call calculations in other programming languages
+\item
+  allows weighted or subset data
 \item
@@
 \item
-  provides a mechanism for handling partially linear models
+  provides partially linear model handling mechanism
 \item
@@
 
-With such a range of features and long history, the code has become untidy
-and overly patched, difficult to maintain, and its underlying methods could
-be improved. Various workers have developed packages to overcome these
-concerns, and we will address some of these here.
+Due to its extensive range of features and prolonged history, the code has become untidy and overly patched, making it challenging to maintain and ripe for improvement in its underlying methods.
 
@@
 
-The tools we will consider are the base-R \texttt{nls()} function
-and some packages that are available on the CRAN repository.
-Of these, we will pay particular attention to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
+Besides the base-R \texttt{nls()} function, we will pay particular attention
+to \textbf{nlsr} (John C. Nash and Murdoch (2023)),
 \textbf{minpack.lm} (Elzhov et al. (2012)), and \textbf{gslnls} (Chau (2023)) which are
-general nonlinear least-squares solvers.
-
+general nonlinear least-squares solvers in the CRAN repository.
 While we will provide capsule comments for some other CRAN packages,
-we will not pursue those in the Bioconductor (Gentleman et al. (2004))
-collection, nor those on repositories such as
-GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}).
+those in the Bioconductor (Gentleman et al. (2004)) collection are more
+specialized, and those on repositories such as
+GitHub (\url{https://github.com}) and Gitlab (\url{https://about.gitlab.com}),
+while interesting, do not have the checking applied to CRAN packages.
 
@@
 
-Our work was aimed at unifying nonlinear modeling functionality in R,
+Our work aimed at unifying nonlinear modeling functionality in R,
 ideally in a refactored \texttt{nls()} function.
@@
 with documentation and using recent tools. If there is a suspicion that there
-may be ill-conditioning, package \texttt{nlsr} or the example we give below of how to
+may be ill-conditioning, package \texttt{nlsr} or the example we give in the
+section ``Comparison notes for formula-setup solutions'' below of how to
 find singular values of the Jacobian allow these diagnostics to be calculated.
 
-For R \textbf{developers}, we invite and encourage discussion of the design choices, since these
+For R \textbf{developers}, we invite and encourage discussion of the design choices,
+since these
 have downstream implications for ease of use, adaptation to new features, and
@@
 
-\includegraphics{ComparingNLStools_files/figure-latex/ex01-1.pdf}
+\includegraphics{20230-15-copyeditresponse_files/figure-latex/ex01-1.pdf}
 
@@
 \begin{verbatim}
-#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1741.1; 47  itns
+#> tgslnls1  -- ss= 9205.4 :  Asym = 35.532  xmid = 20846  scal = -1740.4; 47  itns
 \end{verbatim}
@@
 \begin{verbatim}
-#> [1] 3.4641e+00 9.4495e-09 4.1324e-11
+#> [1] 3.4641e+00 1.0020e-08 2.9378e-11
 \end{verbatim}

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.R

@@ -1,0 +1,418 @@
+# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
+# Please edit 20230-15-copyeditresponse.Rmd to modify this file
+
+## ----setup, include=FALSE-----------------------------------------------------
+knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
+
+
+## ----ex01, echo=TRUE----------------------------------------------------------
+weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
+          38.558, 50.156, 62.948, 75.995, 91.972)
+tt <- 1:12
+weeddf <- data.frame(tt, weed)
+plot(weeddf, main="Hobbs weed infestation data")
+
+
+## ----ex02set, echo=TRUE-------------------------------------------------------
+# model formulas
+frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
+frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
+frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
+#
+# Starting parameter sets
+stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
+sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
+stt1 <- c(Asym = 1, xmid = 1, scal = 1)
+
+
+## ----ex02fn, echo=TRUE--------------------------------------------------------
+# Logistic3U
+hobbs.res <- function(x){ # unscaled Hobbs weeds problem -- residual
+  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
+  y <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
+           38.558, 50.156, 62.948, 75.995, 91.972)
+  tt <- 1:12
+  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
+  # Note: this form of the residual, from Laplace (1788) in the form
+  # of "fitted - observed" has been used in our software for half
+  # a century, but sometimes concerns users of the more traditional
+  # "observed - fitted" form
+}
+
+hobbs.jac <- function(x) { # unscaled Hobbs weeds problem -- Jacobian
+  jj <- matrix(0.0, 12, 3)
+  tt <- 1:12
+  yy <- exp(-x[3] * tt)
+  zz <- 1.0 / (1 + x[2] * yy)
[…]

### 2024-02-08 JNABfinalcheck20231117/Run231117.sh
update finalcheck

@@ -1,0 +1,4 @@
+# run html processing
+Rscript -e 'rmarkdown::render("ComparingNLStools.Rmd", "rjtools::rjournal_web_article", output_options=list(), clean=FALSE)'
+Rscript -e 'rmarkdown::render("ComparingNLStools.Rmd", "rjtools::rjournal_pdf_article", output_options=list(), clean=FALSE)'
+pdflatex RJwrapper.tex

### 2024-02-08 JNABresubmission20231027/ComparingNLStools.Rmd
misc

--- 2023-10-25 JNABRev20231018/ComparingNLStools.Rmd
@@
 ```{r exhobbsfn, eval=TRUE}
+cat("hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)\n")
 hobnlfb<-nlfb(start=stu1, resfn=hobbs.res, jacfn=hobbs.jac)
 pshort(hobnlfb) # use print(hobnlfb) for more detail
+cat("hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)\n")
 hobnlm<-nls.lm(par=stu1, fn=hobbs.res, jac=hobbs.jac)
 pnlslm(hobnlm)  
+cat("hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))\n")
 hobgsln<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12))
 pnls(hobgsln)
+cat("hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)\n")
 hobgsl<-gsl_nls(start=stu1, fn=hobbs.res, y=rep(0,12), jac=hobbs.jac)

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.Rmd
first pass complete

@@
 
-## Principal messages
+<!-- ## Principal messages -->
+<!-- Ed: How does this relate to the comparisons? -->
+<!-- JN: It is a precis of the content.  -->
 
@@
 <!-- JN: removed "was" as suggesgted -->
-ideally in a refactored `nls()` function.
-
-For R **users**, we would advise adapting working scripts, preferably those
+ideally in a refactored `nls()` function. The primary messages from this work
+are:
+
+- For R **users**, we would advise that it is most efficient to carry out
+nonlinear modeling or least squares by adapting working scripts, preferably those
+<!-- JN: added wording to clarify what we are advising users to do -->
 with documentation and using recent tools. If there is a suspicion that there
@@
 
-For R **developers**, we invite and encourage discussion of the design choices, 
+- For R **developers**, we invite and encourage discussion of the design choices, 
 since these
@@
 
-`onls` (@R-onls-manual) -- The usual optimization in estimating nonlinear models
-is the vertical difference between the dependent variable and the functional
-model. `onls` minimizes the sum of **orthogonal** residuals. The objective is
-therefore different and involves nontrivial extra calculation. A vignette with
-the package and the blog article 
+<!-- JN: copyedit suggestion incorporated -->
+`onls` (@R-onls-manual) is used for optimising and estimating nonlinear models 
+by minimizing the sum of squares of **orthogonal** residuals rather than vertical
+residuals. The objective is therefore different and involves nontrivial extra
+calculation. A vignette with the package and the blog article 
 https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/ 
@@
 https://en.wikipedia.org/wiki/Total_least_squares presents an overview of some
+<!-- Ed. Anything specific from the Wikipedia article that is worth citing? -->
+<!-- JN: Commentary would add too much material of tangential nature. -->
 ideas, with references to the literature. The approach needs a wider 
@@
 The Hobbs weed infestation problem (@cnm79 [page 120]) is a growth-curve modeling 
-task that seems straightforward but is quite nasty. Its very succinct statement 
+task. Its very succinct statement 
 provides the "short reproducible example" much requested on R mailing lists. 
-This is a real problem from a field researcher. The data and graph follow.
+The problem is small and seemingly straightforward, yet presents such 
+difficulties that optimization researchers have asked if it is contrived
+rather than a real problem from a field researcher. The data and graph follow.
+<!-- JN: merged copyedit comments to explain "nasty". -->
 
@@
 object, and the
-element `model` is NULL unless the calling argument `model` is TRUE. (Using the same 
-name could
-be confusing.) Despite this, the model frame is used within the function code in 
+element `model` is NULL unless the calling argument `model` is TRUE. Using the 
+same name for both function argument and object element could be confusing.
+<!-- JN: reworded to make clear our concern. -->
+Despite this, the model frame is used within the function code in 
 the form of the object
@@
 
-The function `resid()` (an alias for `residuals()`) gives WEIGHTED residuals, 
-as does, for example, `result$m$resid()`. The function `nlsModel()` 
+The function `resid()` (an alias for `residuals()`) gives weighted residuals
+like `result$m$resid()`. The function `nlsModel()` 
+<!-- JN: copyedit suggestion accepted -->
 allows us to compute residuals for particular coefficient sets.
@@
 for pointing out that it is important to use background knowledge and graphs to
-guide modeling.
-
-# References
+guide modeling. Simon Urbanek gave some important practical help with preparing
+files for publication.
+<!-- JN: added Simon. Removed # References which duplicated subhead -->

### 2024-02-08 copyedit240208/20230-15-copyeditresponse.tex
first pass complete

@@
 
-\hypertarget{principal-messages}{%
-\subsection{Principal messages}\label{principal-messages}}
-
 Our work aimed at unifying nonlinear modeling functionality in R,
-ideally in a refactored \texttt{nls()} function.
-
-For R \textbf{users}, we would advise adapting working scripts, preferably those
+ideally in a refactored \texttt{nls()} function. The primary messages from this work
+are:
+
+\begin{itemize}
+\item
+  For R \textbf{users}, we would advise that it is most efficient to carry out
+  nonlinear modeling or least squares by adapting working scripts, preferably those
   with documentation and using recent tools. If there is a suspicion that there
@@
   find singular values of the Jacobian allow these diagnostics to be calculated.
-
+\item
   For R \textbf{developers}, we invite and encourage discussion of the design choices,
@@
   efficiency of ongoing maintenance.
+\end{itemize}
 
@@
 
-\texttt{onls} (Spiess, Andrej-Nikolai (2022)) -- The usual optimization in estimating nonlinear models
-is the vertical difference between the dependent variable and the functional
-model. \texttt{onls} minimizes the sum of \textbf{orthogonal} residuals. The objective is
-therefore different and involves nontrivial extra calculation. A vignette with
-the package and the blog article
+\texttt{onls} (Spiess, Andrej-Nikolai (2022)) is used for optimising and estimating nonlinear models
+by minimizing the sum of squares of \textbf{orthogonal} residuals rather than vertical
+residuals. The objective is therefore different and involves nontrivial extra
+calculation. A vignette with the package and the blog article
 \url{https://www.r-bloggers.com/2015/01/introducing-orthogonal-nonlinear-least-squares-regression-in-r/}
@@
 The Hobbs weed infestation problem (John C. Nash (1979, 120)) is a growth-curve modeling
-task that seems straightforward but is quite nasty. Its very succinct statement
+task. Its very succinct statement
 provides the ``short reproducible example'' much requested on R mailing lists.
-This is a real problem from a field researcher. The data and graph follow.
+The problem is small and seemingly straightforward, yet presents such
+difficulties that optimization researchers have asked if it is contrived
+rather than a real problem from a field researcher. The data and graph follow.
 
@@
 object, and the
-element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. (Using the same
-name could
-be confusing.) Despite this, the model frame is used within the function code in
+element \texttt{model} is NULL unless the calling argument \texttt{model} is TRUE. Using the
+same name for both function argument and object element could be confusing.
+Despite this, the model frame is used within the function code in
 the form of the object
@@
 
-The function \texttt{resid()} (an alias for \texttt{residuals()}) gives WEIGHTED residuals,
-as does, for example, \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
+The function \texttt{resid()} (an alias for \texttt{residuals()}) gives weighted residuals
+like \texttt{result\$m\$resid()}. The function \texttt{nlsModel()}
 allows us to compute residuals for particular coefficient sets.
@@
 for pointing out that it is important to use background knowledge and graphs to
-guide modeling.
+guide modeling. Simon Urbanek gave some important practical help with preparing
+files for publication.
 
 \hypertarget{references}{%
-\section{References}\label{references}}
-
-\hypertarget{references-1}{%
-\section*{References}\label{references-1}}
+\section*{References}\label{references}}
 \addcontentsline{toc}{section}{References}

### 2024-02-12 copyedit240208/20230-15-copyeditresponse.R
fixups

@@ -1,418 +1,0 @@
-# Generated by `rjournal_pdf_article()` using `knitr::purl()`: do not edit by hand
-# Please edit 20230-15-copyeditresponse.Rmd to modify this file
-
-## ----setup, include=FALSE-----------------------------------------------------
-knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
-
-
-## ----ex01, echo=TRUE----------------------------------------------------------
-weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,
-          38.558, 50.156, 62.948, 75.995, 91.972)
-tt <- 1:12
-weeddf <- data.frame(tt, weed)
-plot(weeddf, main="Hobbs weed infestation data")
-
-
-## ----ex02set, echo=TRUE-------------------------------------------------------
-# model formulas
-frmu <- weed ~ b1 / (1 + b2 * exp(-b3 * tt))
-frms <- weed ~ 100 * c1/(1 + 10* c2* exp(-0.1 * c3* tt))
-frmt <- weed ~ Asym / (1 + exp((xmid - tt) / scal))
-#
-# Starting parameter sets
-stu1 <- c(b1 = 1, b2 = 1, b3 = 1)
-sts1 <- c(c1 = 1, c2 = 1, c3 = 1)
-stt1 <- c(Asym = 1, xmid = 1, scal = 1)
-
-
-## ----ex02fn, echo=TRUE--------------------------------------------------------
-# Logistic3U
-hobbs.res <- function(x){ # unscaled Hobbs weeds problem -- residual
-  if(length(x) != 3) stop("hobbs.res -- parameter vector n!=3")
-  y <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443, 
-           38.558, 50.156, 62.948, 75.995, 91.972)
-  tt <- 1:12
-  res <- x[1] / (1 + x[2] * exp(-x[3] * tt)) - y
-  # Note: this form of the residual, from Laplace (1788) in the form
-  # of "fitted - observed" has been used in our software for half
-  # a century, but sometimes concerns users of the more traditional
-  # "observed - fitted" form
-}
-
-hobbs.jac <- function(x) { # unscaled Hobbs weeds problem -- Jacobian
-  jj <- matrix(0.0, 12, 3)
-  tt <- 1:12
-  yy <- exp(-x[3] * tt)
-  zz <- 1.0 / (1 + x[2] * yy)
[…]

### 2024-02-12 copyedit240208/2023-15-copyeditresponse.Rmd
add AB resp

--- 2024-02-08 copyedit240208/20230-15-copyeditresponse.Rmd
@@
   rationalizing R tools for nonlinear regression and nonlinear estimation tools by
-  considering usability, maintainability and functionality, especially for a
+  considering usability, maintainability, and functionality, especially for a
   Gauss-Newton solver.
@@
 find singular values of the Jacobian allow these diagnostics to be calculated.
-<!-- JN: reference to section added ??have I done this correctly? -->
+<!-- JN: reference to section added ?have I done this correctly? -->
 
@@
 $$ 1/scal  =  b_3 = 0.1*c3 $$
-To allow for simpler discussion, let us say that the parameters form a (named) vector
+To allow for a simpler discussion, let us say that the parameters form a 
+<!-- add "a" -->
+(named) vector
 $p$ and the model function is called $model(p)$ with $r = y - model(p)$.
@@
 
-In reality, there are many problems were $D$ should not be ignored, but the work
+In reality, there are many problems where $D$ should not be ignored, but the work
+<!-- were to where -->
 to compute it precisely is considerable. Many work-arounds have been 
@@
 is not available, `nls()` sets all the starting parameters 
-to 1. This is tolerable, but could be improved by using a set of values
+to 1. This is tolerable but could be improved by using a set of values
 that are all slightly different, which, in the case of the example 
@@
 reflects choices of that era that are at odds with current practice. Some users
-do still want or need to provide problems as explicit functions, particularly
+still want or need to provide problems as explicit functions, particularly
 for problems that are not regressions. For example, the Rosenbrock banana-valley
@@
 in R 4.0.0, there was the warning:  
-
 >  *Do not use `nls` on artificial "zero-residual" data*  
-
-with suggested addition of small perturbations to the data. This admits 
+<!-- Used double space to force linefeed -->
+with the suggested addition of small perturbations to the data. This admits 
 `nls()` could not solve well-posed problems unless data is polluted
@@
 NA elements so that predictions are in the correct row position. We recommend that 
-workers actually test output to verify the behavior is as wanted.
+workers test output to verify the behavior is as wanted.
 See *https://stats.stackexchange.com/questions/492955/should-i-use-na-omit-or-na-exclude-in-a-linear-model-in-r*.
@@
 <!-- An example is to be found in the `Tetra_1.R` example from the `nlsCompare` package. -->
-<!-- ?? AB: Are we far enough along to include reference? -->
+<!-- AB: Are we far enough along to include reference? -->
 Here is an abbreviated example:

### 2024-02-12 copyedit240208/mkarticle.sh

@@ -1,0 +1,5 @@
+#!/bin/bash
+## make article
+Rscript -e 'rmarkdown::render("2023-15-copyeditresponse.Rmd", "rjtools::rjournal_web_article", output_options=list(), clean=FALSE)'
+pdflatex RJwrapper.tex
+pdflatex RJwrapper.tex

### 2024-05-10 copyedit240208/2023-15-copyeditresponse.Rmd
uncommitted

@@
 
-```{r ex01, echo=TRUE}
+```{r ex01, echo=TRUE, out.width="90%"}
 weed <- c(5.308, 7.24, 9.638, 12.866, 17.069, 23.192, 31.443,

### 2024-05-10 copyedit240208/mkarticle.sh

@@ -1,5 +1,8 @@
 #!/bin/bash
 ## make article
 Rscript -e 'rmarkdown::render("2023-15-copyeditresponse.Rmd", "rjtools::rjournal_web_article", output_options=list(), clean=FALSE)'
+echo "Files:"
+ls
+read tmp
 pdflatex RJwrapper.tex
 pdflatex RJwrapper.tex
